### 2021-08-30 chapters/abstract.tex
add : my old files

\thispagestyle{empty} % this line to make the hide the page number of the first page of the abstract. it is a UCCS requirement

this is the abstract of my thesis

### 2021-08-30 chapters/conclusion.tex
add : my old files

\newpage
\chapter{Discussion and Conclusion}
\label{chapter:Discussion_Future_Work}

\section{Conclusion}
\label{section:Conclusion}

\section{Summary of Contributions}
\label{section:SummaryofContributions}

This section will describe the contributions of this thesis.  These can be summarized as follows:

\begin{enumerate}
    \item \textbf{First Idea:}
          We proposed ...
    \item \textbf{Second Idea:}
          We investigated ...
    \item \textbf{Third Idea:}
          We addressed ...

\end{enumerate}

\section{Limitations and Challenges}
\label{section:limitations}

\section{Future Work}
.... Some potential areas for future efforts could include the following:

\begin{enumerate}
    \item ...
    \item ...
    \item ...

\end{enumerate}

\vfill \strut  % to fill the rest of the page with blank lines
\cleardoublepage

### 2021-08-30 chapters/introduction.tex
add : my old files

\newpage
\chapter{Introduction}
\label{chapter:introduction}

Basically we want to save the planet, and save your pockets within it.
We know that humans don't care about anything else except their pockets. So let's try to push them into saving our mother earth by making them reduce their costs and increase their benifits.
How to do so ?
well we will work on reducing the elecricity consumption of their services. without making them to change their machines, their providers or impacting the quality of service that they provide.

Our approach is is to reduce the energy consumption of the software services by changing certains parameters, such as palteform, programming langauge, and/or the design pattern.

well to do so we have multiple approachs ,
1- by using formal apporachs such as static analysers, complexity , nad programming theories -- lets check maybe there is a state of the art -
2- second by a more interactive appreach where based on tests and feedback

This theis is about the second one, So basically all we do is testing programms, measuring their energy and then provide a feedback based on our observations.
To do so we first need to setup a testing enviroment. something that allows us to have accurate, precise and reproductible tests.
Therefore the first chapter our the thesis will be dedicated to provide green coders with a set of elements in orders to test their hypothesis.

....

\section{Motivation}
....

\section{Research Contributions}
..... This work makes the following contributions:

\begin{enumerate}

    \item Introduces a new ...
    \item Shows how ....
    \item Proposes ...

\end{enumerate}

\section{Publications}

\begin{enumerate}
    \item ...

    \item ...

    \item ...

\end{enumerate}
\cleardoublepage

### 2021-08-30 chapters/testing/accuracy.tex
add : my old files

This is the chapter where i will put all the works related to the a successful testing

% \begin{abstract}
%     The past decade witnessed the inclusion of power measurements to evaluate the energy efficiency of software systems, thus making energy a prime indicator along with performance.
%     Nevertheless, measuring the energy consumption of a software system remains a tedious task for practitioners.
%     In particular, the energy measurement process may be subject to a lot of variations that hinder the relevance of potential comparisons.
%     While the state of the art mostly acknowledged the impact of hardware factors (chip printing process, CPU temperature), this paper investigates the impact of controllable factors on these variations.
%     More specifically, we conduct an empirical study of multiple controllable parameters that one can easily tune to tame the energy consumption variations when benchmarking software systems.

%     To better understand the causes of such variations, we ran more than a $1,000$ experiments on more than $100$ nodes with different workloads and configurations.
%     The main factors we studied encompass: experimental protocol, CPU features (C-states, Turbo~Boost, core pinning) and generations, as well as the operating system.
%     Our experiments showed that, for some workloads, it is possible to tighten the energy variation by up to $30\times$.
%     Finally, we summarize our results as guidelines to tame energy consumption variations.
%     We argue that the guidelines we deliver are the minimal requirements to be considered prior to any energy efficiency evaluation.
% \end{abstract}

\section{Introduction}
%\lettrine[nindent=0em,lines=3]{C} 
To conduct robust evaluations, practitioners often try to ensure reproducible environmental conditions in order to properly benchmark their software systems.
In this area, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.

In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should favour reproducible experiments.
However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only on homogeneous nodes, but even within a single node.
This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.

Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
One can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.

\begin{figure}%[!htb]
    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
\end{figure}

Most of the state of the art has been investigating this power consumption issue from a hardware perspective~\cite{borkar_designing_2005,tschanz_adaptive_2002} and reported that the causes of such energy variations are CMOS manufacturing process of transistors in a chip, differences in node assembly and data center hot spots.
Additionally, \cite{heinrich_predicting} described it as a combination of parameters, mentioning a list of candidate factors, such as the thermal effect or the CPU frequency, but failing to deliver a deeper analysis of these factors.

Unfortunately, hardware factors can be hardly tuned to tame the energy variation that can be observed in systems benchmarking.
For example, managing the CPU temperature or a server position in a cluster, are not actions that one can easily do, especially on the modern data centers and cloud platforms.
Therefore, the goal of this paper is to investigate the spectrum of factors that can cause or increase the variability of energy consumption in systems benchmarking, and to propose effective guidelines to control such factors in order to mitigate this variability.
While this paper does not question the benefits of established CPU features, like C-states or Turbo~Boost, it delivers a deeper analysis of the effects they may introduce on a wide set of experiments and nodes.
By quantifying potential energy variations induced by controllable factors, we intend to identify the proper configurations that minimize energy variations, depending on the workload characteristics.
These guidelines aim at supporting practitioners in conducting robust system benchmarks and reporting reproducible energy consumption, rather than recommending production-scale practices or solutions to reduce the power consumption of a given system.
% 
% The key contributions of this paper can therefore be summarized as:
% \begin{enumerate}
%     \item Providing a better understanding of the energy variation, by using different generations of CPU deployed in $4$ clusters with more than $100$ physical nodes, and by considering existing systems benchmarks with diverse workloads;
%     \item Identifying controllable factors that contribute to the variation in CPU energy consumption, comparing them against the state of the art, and completing them with other uncovered assumptions;
%     \item Reporting on some guidelines on how to conduct reproducible experiments with less energy variations;
%     \item Discussing the differences between inter-nodes and intra-nodes variations.
% \end{enumerate}

%\paragraph{}
% The remainder of this paper is organized as follows.
% Section~\ref{sec:related_work} discusses the related works in the area of power measurements.
% Section~\ref{sec:research_qs} formalizes our research questions.
% Section~\ref{sec:setup} reports on the experimental setup (hardware, benchmarks, tools and methodology) we used in this work.
% Section~\ref{sec:parameters} analyzes the causes of the variations we observed along experiments.
% Finally, we discuss the results of the different experimented factors, and their impact on the energy variation in Section~\ref{sec:guidelines}, while Sections~\ref{sec:threats} \&~\ref{sec:conclusion} cover the validity threats and our conclusions, respectively.

%------------------------------------------------
\section{Related Work}\label{sec:related_work}
In this section, we review the state of the art in the domain of energy consumption analysis.
In particular, one can observe that several studies considered the evaluation of the processor variation while executing the same workload, in particular to test distributed applications running in homogeneous clusters.

\paragraph{Studying Hardware Factors.}

This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
On the one hand, the correlation between the processor temperature and the energy consumption was one of the most explored paths.
Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the correlation between the CPU temperature and the energy consumption variation is very tight.

\note{add the corelation value}

This makes the processor temperature a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.

The ambient temperature was also discussed in many papers as a candidate factor for the energy variation of a processor.
In~\cite{ranka_energy_2009}, the authors claimed that energy consumption may vary due to fluctuations caused by the external environment.
These fluctuations may alter the processor temperature and its energy consumption.
However, the temperature inside a data center does not show major variations from one node to another.
In~\cite{el_mehdi_diouri_your_2013}, El~Mehdi~Dirouri~\emph{et~al.} showed that switching the spot of two servers does not affect their energy consumption.
Moreover, changing hardware components, such as the hard drive, the memory or even the power supply, does not affect the energy variation of a node, making it mainly related to the processor.
This result was recently assessed by~\cite{wang_potential_2018}, where the rack placement and power supply introduced a maximum of $2.8\,\%$ variation in the observed energy consumption.

Beyond hardware components, the accuracy of power meters has also been questioned.
Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.

\paragraph{Mitigating Energy Variations.}
Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
They also proposed some guidelines to reduce this variation by replacing the old slow chips, by load balancing the workload on the CPU cores and leaving one core idle.
They claimed that the variation between the processor cores is insignificant.
In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.

In~\cite{marathe_empirical_2017_m}, the authors highlight the increase of energy variation across the latest Intel micro-architectures by a factor of $4$ from Sandy~Bridge to Broadwell, a $15\,\%$ of run-to-run variation within the same processor and the increase of the inter-cores variation from $2.5\,\%$ to $5\,\%$ due to hardware-enforced constraints, concluding with some recommendations for Broadwell usage, such as running one hyper-thread per core.
\note{this might be interesting for greeFaaS}

\paragraph{Objective.}
The aim of our paper is not to reproduce all the exposed results and explored paths.
Instead, we intend to check some of these results on latest hardware components, and to consider new potential controllable factors that could also contribute to the energy variation.
We also aim to provide practitioners with guidelines to conduct robust experiments in a controlled environment, thus reducing the variation overhead to share trustable results and build credible comparisons.

\section{Research Questions}\label{sec:research_qs}
Worth saying, we are aware that part of the energy consumption variation is due to chip manufacturing differences or some of the previously discussed enforced factors, such as the thermal effect or the servers placement.
Those parameters are often tricky to manage, as we cannot have a perfect chips manufacturing process, or assume that two identical processors have the same thermal behavior.
We will therefore focus on providing the practitioners with an empirical study of some controllable parameters that can be tuned to conduct experimental evaluations of energy consumption with less variation, especially if the practitioners do not have a physical access to the data center or BIOS configuration, which is the case on most of the modern cloud platforms and data centers.
These parameters span the choice of the benchmarking protocol, processor frequencies management, operating system tuning or some other parameters.
The authors of~\cite{heinrich_predicting} mentioned some of these potential parameters.

In this paper, we will therefore investigate the following controllable factors, which we formalize as 4 research questions:
\begin{description}
    \item[\textsc{RQ}~1:] Does the benchmarking protocol affect the energy variation?
    \item[\textsc{RQ}~2:] How important is the impact of the processor features on the energy variation?
    \item[\textsc{RQ}~3:] What is the impact of the operating system on the energy variation? and finally
    \item[\textsc{RQ}~4:] Does the choice of the processor matter to mitigate the energy variation?
\end{description}

\section{Experimental Setup}\label{sec:setup}
This section describes our detailed experimental environment, covering the clusters and nodes configuration, the benchmarks we used and justifying our experimental methodology.

\subsection{Hardware Platform}
We considered 4 distinct clusters of variable sizes and different generations of CPU, as summarized in Table~\ref{table:g5k}.
In particular, we used the Grid5000 (G5K) platform in our experiments~\cite{grid5000,margery:hal-00965708}.
G5K is a bare metal cloud platform that can be used to provision clusters of identical nodes.
In our study, we mainly used the cluster \textsf{Dahu} located in Grenoble to run most of our tests, as it has one of the newest Xeon CPUs.
We also used the clusters \textsf{Chetemi}, \textsf{Ecotype} and \textsf{Paranoia} in some of our experiments.
Table~\ref{table:g5k} describes the configurations of the clusters we considered.
% We note that each cluster is geographically independent.

\begin{table}
    \centering
    \caption{Description of clusters included in the study}
    \label{table:g5k}
    \small
    \begin{tabular}{|l|l|r|r|}
        \hline
        \textbf{Cluster}  & \textbf{Processor}               & \textbf{Nodes} & \textbf{RAM} \\
        \hline
        \hline
        \textsf{Dahu}     & $2\times$ Intel Xeon\,Gold\,6130 & 32             & 192\,GiB     \\
        \hline
        \textsf{Chetemi}  & $2\times$ Intel Xeon\,E5-2630v4  & 15             & 768\,GiB     \\
        \hline
        \textsf{Ecotype}  & $2\times$ Intel Xeon\,E5-2630Lv4 & 48             & 128\,GiB     \\
        \hline
        \textsf{Paranoia} & $2\times$ Intel Xeon\,E5-2660v2  & 8              & 128\,GiB     \\
        \hline
    \end{tabular}
\end{table}

As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or socket to designate one of the two sockets and \textsf{PU} for the operating system \emph{processing unit}.
The number of \textsf{PU} often doubles the number of available cores because of the hyper-threading support, as the OS considers 2 hyper-threads sharing the same core as 2 different \textsf{PU}.
Figure~\ref{fig:topo} illustrates a detailed topology of a node belonging to the cluster \textsf{Dahu}.

\begin{figure}%[!htb]
    \center{\includegraphics[width=.9\linewidth]{imgs/lstopo}}
    \caption{Topology of the nodes of the cluster \textsf{Dahu}}\label{fig:topo}
\end{figure}
% The advantage of using G5K is having an access to a wide set of physical nodes with different capabilities. 
% The usage procedure is to book a machine for period of time with the chosen OS in less than 5 minutes, ship and run the test script, then release the machine. When a machine is released everything is discarded even the OS.

\subsection{Systems Benchmarks}
Our first criterion to choose the systems benchmarks was the scalability, as we need to run tests at different workloads by choosing the right number of used PU.
The other criteria are the documentation, the accuracy and the references to the benchmark.
\emph{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925} is one of the most used suite of benchmarks in the HPC literature and it fulfills our benchmarking requirements.
We mainly used the pseudo application \emph{Lower-Upper symmetric Gauss-Seidel} (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
These are the main benchmarks used in many similar works, such as~\cite{heinrich_predicting}.
Nonetheless, in order to validate our results on a wider set of benchmarks and applications, we also used \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}\footnote{\url{https://linux.die.net/man/1/sha256sum}} as representative systems benchmarks to conduct our experiments with a broad diversity of workloads.

\subsection{Measurement Tools \& Methodology}
To study the energy consumption of nodes, we considered Intel \emph{Running Average Power Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754}, which is one of the most accurate tools to report the CPU/DRAM global energy consumption.
We also used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, which is a power monitoring toolkit that builds a model over RAPL to compute the energy consumption at process-level when we needed to isolate energy consumption of a single process.
Our clusters are provisioned with a minimal version of Debian\,9 (4.9.0 kernel version) where we install Docker (version 18.09.5), which will be used to run the RAPL sensor and the benchmark itself.
The energy sensor collects RAPL reports and stores them in a remote \textsc{MongoDB} instance, allowing us to perform \emph{post-mortem} analysis in a dedicated environment.
Using Docker makes the deployment process easier on the one hand, and provides us with a built-in control group encapsulation of the conducted tests on the other hand.
This allows \textsc{PowerAPI} to measures all the running containers, even the RAPL sensor consumption, as it is isolated in a container.
One potential threat covers the impact of Docker on the energy variation.
We therefore conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.

\begin{figure}
    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
\end{figure}

Every experiment is conducted on $100$ iterations, on multiple nodes and using the 3 NPB benchmarks we mentioned, with a warmup phase of 10 iterations for each experiment.
In most cases, we were seeking to evaluate the \emph{STandard Deviation} (STD), which is the most representative factor of the energy variation.
We tried to be very careful, while running our experiments, not to fall in the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
As we study the STD difference of measurements we observed from empirical experiments, we use the bootstrap method~\cite{efron2000bootstrap} to randomly build multiple subsets of data from the original dataset, and we draw the STD density of those sets, as illustrated in Figure~\ref{fig:docker}.
Given the space constraints, this paper reports on aggregated results for nodes, benchmarks and workloads, but the raw data we collected remains available through the public repository we published.\footnote{\url{https://github.com/anonymous-data/Energy-Variation}}
We believe this can help to achieve better and more reliable comparisons.

We mainly consider 3 different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium and high CPU usage when analyzing the studied parameters effect, respectively.
These workloads reflect the ratio of used PU count to the total available PU.

\section{Energy Variation Analysis}\label{sec:parameters}
In this section, we aim to establish experimental guidelines to reduce the CPU energy variation.
We therefore explore many potential factors and parameters that could have a considerable effect on the energy variation.

\subsection{\textsc{RQ}~1: Benchmarking Protocol}
To achieve a robust and reproducible experiment, practitioners often tend to repeat their tests multiple times, in order to analyze the related performance indicators, such as execution time, memory consumption or energy consumption.
We therefore aim to study the benchmarking protocol to identify how to efficiently iterate the tests to capture a trustable energy consumption evaluation.
% This process of repeating executions should not be random in order to avoid different results because of the CPU behavior, the cache data, the memory size and plenty of other aspects.

In this first experiment, we investigate if changing the testing protocol affects the energy variation.
To achieve this, we considered 3 execution modes:
In the \textsf{"normal"} mode, we iteratively run the benchmark 100 times without any extra command, while the \textsf{"sleep"} mode suspends the execution script for 60\,seconds between iterations.
Finally, the \textsf{"reboot"} mode automatically reboots the machine after each iteration.
The difference between the \textsf{normal} and \textsf{sleep} modes intends to highlight that the CPU needs some rest before starting another iteration, especially for an intense workload.
Putting the CPU into sleep for several seconds could give it some time to reach a lower frequency state or/and reduce its temperature, which could have an impact on the energy variation.
The \textsf{reboot} mode, on the other hand, is the most straightforward way to reset the machine state after every iteration.
It could also be beneficial to reset the CPU frequency and temperature, the stored data, the cache or the CPU registries.
However, the reboot task takes a considerable amount of time, so rebooting the node after every single operation is not the fastest nor the most eco-friendly solution, but it deserves to be checked to investigate if it effectively enhances the overall energy variation or not.

Figure~\ref{fig:running-process} reports on $300$ aggregated executions of the  benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}, on 4 machines of the cluster \textsf{Dahu} (cf. Table~\ref{table:g5k}) for different workloads.
We note that the results have been executed with different datasets sizes (\textsf{B}, \textsf{C} and \textsf{D} for single process, 50\,\% and 100\,\% respectively) to remedy to the brief execution times at high workloads for small datasets.
This justifies the scale differences of reported energy consumptions between the 3 modes in Figure~\ref{fig:running-process}.
As one can observe, picking one of these strategies does not have a strong impact on the energy variation for most workloads.
In fact, all the strategies seem to exhibit the same variation with all the workloads we considered---\emph{i.e.}, the STD is tightly close between the three modes.
The only exception is the \textsf{reboot} mode at 100\,\% load, where the STD is 150\,\% times worst, due to an important amount of outliers.
This goes against our expectation, even when setting a warm-up time after reboot to stabilize the OS.

\begin{figure*}
    \center{\includegraphics[width=\linewidth]{imgs/running-process}}
    \caption{Energy variation with the \textsf{normal}, \textsf{sleep} and \textsf{reboot} modes}\label{fig:running-process}
\end{figure*}

In Figure~\ref{fig:reboot}, we study the standard deviation of the three modes by constituting $5,000$ random 30-iterations sets from the previous executions set and we compute the STD in each case, considering mainly the 100\,\% workload as the STD was 150\,\% higher for the \textsf{reboot} mode with that load.
We can observe that the considerable amount of outliers in the \textsf{reboot} mode is not negligible, as the STD density is clearly higher than the two other modes.
This makes the \textsf{reboot} mode as the less appropriate for the energy variation at high workloads.

\begin{figure}
    \center{\includegraphics[width=.9\linewidth]{imgs/reboot}}
    \caption{STD analysis of the \textsf{normal}, \textsf{sleep} and \textsf{reboot} modes}\label{fig:reboot}
\end{figure}

\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]

    To answer \textsc{RQ~1}, we conclude that the benchmarking protocol \textbf{partially affects} the energy variation, as highlighted by the \textsf{reboot} mode results for high workloads.

\end{mdframed}

\subsection{\textsc{RQ}~2: Processor Features}
The C-states provide the ability to switch the CPU between more or less consuming states upon activities.
Turning the C-states on or off have been subject of many discussions~\cite{5463056}, because of its dynamic frequency mechanism but, to the best of our knowledge, there have been no fully conducted C-states behavior analysis on CPU energy variation.

We intend to investigate how much the energy consumption varies when disabling the C-states (thus, keeping the CPU in the \texttt{C0} state) and at which workload.
Figure~\ref{fig:c-states} depicts the results of the experiments we executed on three nodes of the cluster \textsf{Dahu}.
On each node, we ran the same set of benchmarks with two modes: \textsf{C-states on}, which is the default mode, and \textsf{C-states off}.
Each iteration includes 100 executions of the same benchmark at a given workload, with three workload levels.
We note that our results have been confirmed with the benchmarks \texttt{LU}, \texttt{CG} and \texttt{EP}.

We can clearly see the effect that has the \textsf{C-states off} mode when running a single-process application/benchmark.
The energy consumption varies 5~times less than the default mode.
In this case, only one CPU core is used among $2\times16$ physical cores.
The other cores are switched to a low consumption state when C-states are on, the switching operation causes an important energy consumption difference between the cores, and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
On the other hand, switching off the C-states would keep all the cores---even the unused ones---at a high frequency usage.
This highly reduces the variation, but causes up to 50\,\% of extra energy consumption in this test ($Mean_{C-states-off}=11,665 mJ$,$Mean_{C-states-on}=7,641 mJ$).

At a 100\,\% workload, disabling the C-states seems to have no effect on the total energy consumption nor its variation.
In fact, all the cores are used at 100\,\% and the C-states module would have no effect, as the cores are not idle.
The same reason would apply for the 50\,\% load, as the hyper-threading is active on all cores, thus causing the usage of most of them.
For single process workloads, disabling the C-states causes the process to consume 50\,\% more energy as reported in Figure~\ref{fig:c-states}, but reduces the variation by 5~times compared to the \textsf{C-states on} mode.
This leads to mainly two questions: Can a process pinning method reduce/increase the energy variation? And, how does the energy consumption variation evolve at different PU usage level?

\begin{figure*}
    \center{\includegraphics[width=\linewidth]{imgs/c-states}}
    \caption{Energy variation when disabling the C-states}\label{fig:c-states}
\end{figure*}

\subsubsection{Cores Pinning}
To answer the first question, we repeated the previous test at 50\,\% workload.
In this experiment, we considered three cores usage strategies, the first one (\textsf{S1}) would pin the processes on all the PU of one of the two sockets (including hyper-threads), so it will be used at 100\,\%, and leave the other CPU idle.
The second strategy (\textsf{S2}) splits the workload on the two sockets so each CPU will handle 50\,\% of the load.
In this strategy, we only use the core PU and not the hyper-threads PU, so every process would not share his core usage (all the cores are being used).
The third strategy (\textsf{S3}) consists also on splitting the workload between the two sockets, but considering the usage of the hyper-threads on each core---\emph{i.e.}, half of the cores are being used over the two CPU.
Figure~\ref{fig:cores-pinning} reports on the energy consumption of the three strategies when running the benchmark \texttt{CG} on the cluster \textsf{Dahu}.
We can notice the big difference between these three execution modes that we obtained only by changing the PU pinning method (that we acknowledged with more than 100 additional runs over more than 30 machines and with the benchmarks \texttt{LU} and \texttt{EP}).
For example, \textsf{S2} is the least power consuming strategy.
We argue that the reason is related to the isolation of every process on a single physical core, reducing the context switch operations.
In the first and third strategy, $32$ processes are being scheduled on $16$ physical cores using the hyper-threads PU, which will introduce more context switching, and thus more energy consumption.

\begin{figure}
    \center{\includegraphics[width=.8\linewidth]{imgs/cores_pinning}}
    \caption{Energy variation considering the three cores pinning strategies at 50\,\% workload}\label{fig:cores-pinning}
\end{figure}

We note that even if the first and third strategies are very similar (both use hyper-threads, but only on one CPU for the first and on two CPU for the third), the gap between them is considerable variation-wise, as the variation is 30 times lower in the first strategy ($STD_{S1}=116 mJ$,$STD_{S3}=3,452 mJ$).
This shows that the usage of the hyper-threads technology is not the main reason behind the variation, the first strategy has even less variation than the second one and still uses the hyper-threading.

The reason for the \textsf{S1} low energy consumption is that one of the two sockets is idle and will likely be in a lower power P-state, even with the disabled C-states.
The \textsf{S2} case is also low energy consuming because by distributing the threads across all the cores, it completes the task faster than in the other cases.
Hence, it consumes less energy.
The \textsf{S3} is a high consuming strategy because both sockets are being used, but only half the cores are active.
This means that we pay the energy cost for both sockets being operational and for the experiments taking longer to run because of the recurrent context switching.

Our hypothesis regarding the worst results that we observed when using the third strategy is the recurrent context switching, added to the OS scheduling that could reschedule processes from a socket to another, which invalids the cache usage as a process can not take profit of the socket local L3 cache when it moves from a CPU to another (cf. Figure~\ref{fig:topo}).

Moreover, the fact that the variation is 4--5 times higher when using the strategy \textsf{S2} compared to \textsf{S1} ($STD_{S1}=116 mJ$, $STD_{S3}=575 mJ$), gives another reason to believe that swapping a process from a CPU to another increases the variation due to CPU micro~differences, cache misses and cache coherency.
While the mean execution time for the strategy \textsf{S3} is very high ($MeanTime_{S3}= 46 s$) compared to the two other strategies ($MeanTime_{S1}= 11 s$, $MeanTime_{S2}= 7 s$), we see no correlation between the execution time and the energy variation, as the \textsf{S1} still give less variations than \textsf{S2} even if it takes 36\,\% more time to run.

Table~\ref{table:corespinning} reports on additional aggregated results for the STD comparison on four other nodes of the cluster \textsf{Dahu} at 50\,\%, with the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
In fact, the CPU usage strategy \textsf{S1} is by far the experimentation mode that gave the least variation.
The STD is almost 5 times better than the strategy \textsf{S2}, but is up to 10\,\% more energy consuming ($Mean_{S1}=4469 mJ$, $Mean_{S2}=4016 mJ$).
On the other hand, the strategy \textsf{S3} is the worst, where the energy consumption can be up to 5 times higher than the strategy \textsf{S2} ($Mean_{S2}=4016 mJ$, $Mean_{S3}=21645 mJ$) and the variation is much worst (30 times compared to the first strategy).
These results allow us to have a better understanding of the different processes-to-PU pinning strategies, where isolating the workload on a single CPU is the best strategy.
Using the hyper-threads PU on multiple sockets seems to be a bad recommendation, while keeping the hyper-threading enabled on the machine is not problematic, as long as the processes are correctly pinned on the PU.
Our experiments show that running one hyper-thread per core is not always the best to do, at the opposite of the claims of~\cite{marathe_empirical_2017_m}.

\begin{table}
    \centering
    \caption{STD (mJ) comparison for 3 pinning strategies}
    \label{table:corespinning}
    \small
    \begin{tabular}{|l|l|l|l|}
        \hline
        \textbf{Strategy} & \textbf{S1} & \textbf{S2} & \textbf{S3} \\
        \hline
        \hline
        \textbf{Node~1}   & 88          & 270         & 1,654       \\
        \hline
        \textbf{Node~2}   & 79          & 283         & 2,096       \\
        \hline
        \textbf{Node~3}   & 58          & 287         & 1,725       \\
        \hline
        \textbf{Node~4}   & 51          & 229         & 1,334       \\
        \hline
    \end{tabular}
\end{table}

\subsubsection{Processes Threshold}
To answer the second question regarding the evolution of the energy variation at different levels of CPU usage, we varied the used PU's count to track the EV evolution.
Figure~\ref{fig:process-tresh} compares the aggregated energy variation when the C-states are on and off using 2, 4 and 8 processes for the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
This figure confirms that disabling the CPU C-states does not decrease the variation for all the workloads, as we can clearly observe, the variation is increasing along with the number of processes.
When running only 2 processes, turning off the C-states reduces the STD up to 6~times, but consumes 20\,\% more energy ($Mean_{C-states-on}=10,334 mJ$, $Mean_{C-states-off}=12,594 mJ$).
This variation is 4~times lower when running 4 processes and almost equal to the \textsf{C-states on} mode when running 8 processes.
In fact, running more processes implies to use more CPU cores, which reduces the idle cores count, so the cores will more likely stay at a higher consumption state even if the C-states mechanism is on.

In our case, using 4 PU reduces the variation by 4~times and consumes almost the same energy as keeping the C-states mechanism on ($Mean_{C-states-on}=7,048 mJ$, $Mean_{C-states-off}=7,119 mJ$).
This case would be the closest to reality as we do not want to increase the energy consumption while reducing the variation, but using a lower number of PU still results in less variation, even if it increases the overall energy consumption.

\begin{figure*}
    \center{\includegraphics[width=\linewidth]{imgs/tresh}}
    \caption{C-states effect on the energy variation, regarding the application processes count}\label{fig:process-tresh}
\end{figure*}

We note that disabling the C-states is not recommended in production environments, as it introduces extra energy consumption for low workloads (around 50\,\% in our case for a single process job).
However, our goal is not to optimize the energy consumption, but to minimize the energy variation.
Thus, disabling the C-states is very important to stabilize the measurements in some cases when the variation matters the most.
Comparing the energy consumptions of two algorithms or two versions of a software systems is an example of use case benefiting from this recommendation.

\subsubsection{Turbo Boost}
The Turbo Boost---also known as \emph{Dynamic Overclocking}---is a feature that has been incorporated in Intel CPU since the Sandy~Bridge micro-architecture, and is now widely available on all of the Core\,i5, Core\,i7, Core\,i9 and Xeon series.
It automatically raises some of the CPU cores operating frequency for short periods of time, and thus boost performances under specific constraints.
When demanding tasks are running, the operating system decides on using the highest performance state of the processor.

Disabling or enabling the Turbo~Boost has a direct impact on the CPU frequency behavior, as enabling it allows the CPU to reach higher frequencies in order to execute some tasks for a short period of time.
However, its usage does not have a trivial impact on the energy variation.
Acun~\emph{et~al.}~\cite{acun_variation_2016} tried to track the Turbo~Boost impact on the Ivy~Bridge and the Sandy~Bridge architectures.
They concluded that it is one of the main responsible for the energy variation, as it increases the variation from 1\,\% to 16\,\%.
In our study, we included a Turbo~Boost experiment in our testbed, to check this property on the recent Xeon Gold processors, covering various workloads.

The experiment we conducted showed that disabling the Turbo Boost does not exhibit any considerable positive or negative effect on the energy variation.
Table~\ref{table:turboboost} compares the STD when enabling/disabling the Turbo~Boost, where the columns are a combination of workload and benchmark.
In fact, we only got some minor measurements differences when switching on and off the Turbo~Boost, and where in favor or against the usage of the Turbo~Boost while repeating tests, considering multiple nodes and benchmarks.
This behavior is mainly related to the \emph{thermal design power} (TDP), especially at high workloads executions.
When a CPU is used at its maximum capacity, the cores would be heating up very fast and would hit the maximum TDP limit.
In this case, the Turbo~Boost cannot offer more power to the CPU because of the CPU thermal restrictions.
At lower workloads, the tests we conducted proved that the Turbo~Boost is not one of the main reasons of the energy variation.
In fact, the variation difference is barely noticeable when disabling the Turbo~Boost, which cannot be considered as a result regarding the OS activity and the measurement error margin.
We cannot affirm that the Turbo~Boost does not have an impact on all the CPU, as we only tested on two recent Xeon CPU (clusters \textsf{Chetemi} and \textsf{Dahu}).
We confirmed our experiments on these machines 100 times at 5\,\%, 25\,\%, 50\,\% and 100\,\% workloads.

\begin{table}%[h!]
    \centering
    \caption{STD (mJ) comparison when enabling/disabling the Turbo~Boost}
    \label{table:turboboost}
    \small
    \begin{tabular}{|c|r|r|}
        \hline
        \textbf{Turbo Boost}  & \textbf{Enabled} & \textbf{Disabled} \\
        \hline
        \hline
        \textsf{EP} / 5\,\%   & 310              & 308               \\
        \hline
        \textsf{CG} / 25\,\%  & 95               & 140               \\
        \hline
        \textsf{LU} / 25\,\%  & 204              & 240               \\
        \hline
        \textsf{EP} / 50\,\%  & 84               & 79                \\
        \hline
        \textsf{EP} / 100\,\% & 125              & 110               \\
        \hline
    \end{tabular}
\end{table}

\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
    We conclude that CPU features \textbf{highly impact} the energy variation as an answer for \textsc{RQ~2}.
\end{mdframed}

\subsection{\textsc{RQ}~3: Operating System}
The \emph{operating system} (OS) is the layer that exploits the hardware capabilities efficiently.
It has been designed to ease the execution of most tasks with multitasking and resource sharing.
In some delicate tests and measurements, the OS activity and processes can cause a significant overhead and therefore a potential threat to the validity. % experiments validity
The purpose behind this experiment is to determine if the sampled consumption can be reliably related to the tested application, especially for low-workload applications where CPU resources are not heavily used by the application.

The first way to do is to evaluate the OS idle activity consumption, and to compare it to a low workload running job.
Therefore, we ran 100 iterations of a single process benchmark \textsf{EP}, \textsf{LU} and \textsf{CG} on multiple nodes from the cluster \textsf{Dahu}, and compared the energy behavior of the node with its idle state on the same duration.
The aggregated results, illustrated in Figure~\ref{fig:os-idle}, depict that the idle energy variation is up to 140\,\% worst than when running a job, even if it consumes 120\,\% less energy ($Mean_{Job}=8,746 mJ$, $Mean_{Idle}=3,927 mJ$).
In fact, for the three nodes, randomly picked from the cluster \textsf{Dahu}, the idle variation is way more important than when a test was running, even if it is a single process test on a $32$-cores node.
This result shows that OS idle consumption varies widely, due to the lack of activity and the different CPU frequencies states, but it does not mean that this variation is the main responsible for the overall energy variation.
The OS behaves differently when a job is running, even if the amount of available cores is more than enough for the OS to keep his idle behavior when running a single process.

\begin{figure}
    \center{\includegraphics[width=.9\linewidth]{imgs/idle}}
    \caption{OS consumption between idle and when running a single process job}\label{fig:os-idle}
\end{figure}

Inspecting the OS idle energy variation is not sufficient to relate the energy variation to the active job.
In fact, the OS can behave differently regarding the resource usage when running a task.
To evaluate the OS and the job energy consumption separately, we used the \textsc{PowerAPI} toolkit.
This fine-grained power~meter allows the distribution of the RAPL global energy across all the Cgroups of the OS using a power model.
Thus, it is possible to isolate the job energy consumption instead of the global energy consumption delivered by RAPL.
To do so, we ran tests with a single process workload on the cluster \textsf{Dahu}, and used the \textsc{PowerAPI} toolkit to measure the energy consumption.
Then, we compared the job energy consumption to the global RAPL data.
We calculated the Pearson correlation~\cite{ref1} of the energy consumption and variation between global RAPL and \textsc{PowerAPI}, as illustrated in Figure~\ref{fig:correlation}.
The job energy consumption and variation are strongly correlated with the global energy consumption and variation with the coefficients 93.6\,\% and 85.3\,\%, respectively.
However, this does not completely exclude the OS activity, especially if the jobs have tight interaction with the OS through the signals and system calls.
This brings a new question on whether applying extra-tuning on a minimal OS would reduce the variation? As well as what is the effect of the Meltdown security patch---that is known to be causing some performance degradation~\cite{Kocher2018spectre,Lipp2018meltdown}---on the energy variation?

\begin{figure}
    \center{\includegraphics[width=.8\linewidth]{imgs/correlation}}
    \caption{The correlation between the RAPL and the job consumption and variation}\label{fig:correlation}
\end{figure}

\subsubsection{OS Tuning}
An OS is a pack of running processes and services that might or not be required its execution.
In fact, even using a minimal version of a Debian Linux, we could list many OS running services and process that could be disabled/stopped without impacting the test execution.
This extra-tuning may not be the same depending on the nature of the test or the OS.
Thus, we conducted a test with a deeply-tuned OS version.
We disabled all the services/processes that are not essential to the OS/test running, including the OS networking interfaces and logging modules, and we only kept the strict minimum required to the experiment's execution.
Table~\ref{table:osmin} reports on the aggregated results for running single process measurements with the benchmarks \textsf{CG}, \textsf{LU} and \textsf{EP}, on three servers of the cluster \textsf{Dahu}, before and after tuning the OS.
Every cell contains the \emph{STD} value before the tuning, plus/minus a ratio of the energy variation after the tuning.
We notice that the energy variation varies less than 10\,\% after the extra-tuning.
We argue that this variation is not substantial, as it is not stable from a node to another.
Moreover, 10\,\% of variation is not a representative difference, due to many factors that can affect it as the CPU temperature or the measurement errors.

\begin{table}
    \centering
    \caption{STD (mJ) comparison before/after tuning the OS}
    \label{table:osmin}
    \small
    \begin{tabular}{|c|c|c|c|}
        \hline
        \textbf{Node} & \textbf{\sf EP} & \textbf{\sf CG} & \textbf{\sf LU} \\
        \hline
        \hline
        \textbf{N1}   & 1370~-9\,\%     & 78~+7\,\%       & 128~+2\,\%      \\
        \hline
        \textbf{N2}   & 1278~-7\,\%     & 64~-1\,\%       & 120~+9\,\%      \\
        \hline
        \textbf{N3}   & 1118~+1\,\%     & 83~+2\,\%       & 93~+7\,\%       \\
        \hline
    \end{tabular}
\end{table}
% \subsubsection{Meltdown Security Patch}

\subsubsection{Speculative Executions}
Meltdown and Spectre are two of the most famous hardware vulnerabilities discovered in 2018, and exploiting them allows a malicious process to access others processes data that is supposed to be private~\cite{Kocher2018spectre,Lipp2018meltdown}.
They both exploit the speculative execution technique where a process anticipates some upcoming tasks, which are not guaranteed to be executed, when extra resources are available, and revert those changes if not.
Some OS-level patches had been applied to prevent/reduce the criticality of these vulnerabilities.
On the Linux kernel, the patch has been automatically applied since the version 4.14.12.
It mitigates the risk by isolating the kernel and the user space and preventing the mapping of most of the kernel memory in the user space.
% These patches have been reported as decreasing the performance of a system. 
Nikolay~\emph{et~al.} have studied in~\cite{DBLP:journals/corr/abs-1801-04329} the impact of patching the OS on the performance.
The results showed that the overall performance decrease is around 2--3\,\% for most of the benchmarks and real-world applications, only some specific functions can meet a high performance decrease.
In our study, we are interested in the applied patch's impact on the energy variation, as the performance decrease could mean an energy consumption increase.
Thus, we ran the same benchmarks \textsf{LU}, \textsf{CG} ad \textsf{EP} on the cluster \textsf{Dahu} with different workloads, using the same OS, with and without the security patch.
Table~\ref{table:meltdown} reports on the STD values before disabling the security patch.
A minus means that the energy varies less without the patch being applied, while a plus means that it varies more.
These results help us to conclude that the security patch's effect on the energy variation is not substantial and can be absorbed through the error margin for the tested benchmarks.
In fact, the best case to consider is the benchmark \textsf{LU} where the energy variation is less than 10\,\% when we disable the security patch, but this difference is still moderate.
The little performance difference discussed in~\cite{Kocher2018spectre,Lipp2018meltdown} may only be responsible of a small variation, which will be absorbed through the measurement tools and external noise error margin in most cases.

\begin{table}[h!]
    \centering
    \caption{STD (mJ) comparison with/without the security patch}
    \label{table:meltdown}
    \small
    \begin{tabular}{|l|c|c|c|}
        \hline
        \textbf{Node} & \textbf{\sf EP} & \textbf{\sf CG} & \textbf{\sf LU} \\
        \hline
        \hline
        \textbf{N1}   & 269~+2\,\%      & 83~+1\,\%       & 108~-6\,\%      \\
        \hline
        \textbf{N2}   & 195~+1\,\%      & 84~-5\,\%       & 121~-9\,\%      \\
        \hline
        \textbf{N3}   & 223~+/-1\,\%    & 72~-4\,\%       & 117~+8\,\%      \\
        \hline
        \textbf{N4}   & 276~+3\,\%      & 60~+0\,\%       & 113~-3\,\%      \\
        \hline
    \end{tabular}
\end{table}

\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
    To answer \textsc{RQ~3}, we conclude that the OS \textbf{should not be the main focus} of the energy variation taming efforts.
\end{mdframed}

\subsection{\textsc{RQ}~4: Processor Generation}
Intel microprocessors have noticeably evolved during these last 20 years.
Most of the new CPU come with new enhancements to the chip density, the maximum Frequency or some optimization features like the C-states or the Turbo~Boost.
This active evolution caused that different generations of CPU can handle a task differently.
The aim of this expriment is not to justify the evolution of the variation across CPU versions/generations, but to observe if the user can choose the best node to execute her experiments.
Previous papers have discussed the evolution of the energy consumption variation across CPU generations and concluded that the variation is getting higher with the latest CPU generations~\cite{wang_experimental_nodate,marathe_empirical_2017_m}, which makes measurements stability even worse.
In this experiment, we therefore compare four different generations of CPU with the aim to evaluate the energy variation for each CPU and its correlation with the generation.
Table~\ref{table:cpus} indicates the characteristics of each of the tested CPU.

\begin{table}
    \centering
    \caption{STD (mJ) comparison of experiments from 4 clusters}
    \small
    \begin{tabular}{|l|c|c|c|c|}
        \hline
        \textbf{Cluster} & \textbf{Dahu} & \textbf{Chetemi} & \textbf{Ecotype} & \textbf{Paranoia} \\
        \hline
        \hline
        Arch             & Skylake       & Broadwell        & Broadwell        & Ivy Bridge        \\
        \hline
        Freq             & 3.7\,GHz      & 3.1\,GHz         & 2.9\,GHz         & 3.0\,GHz          \\
        \hline
        TDP              & 125 W         & 85 W             & 55 W             & 95 W              \\
        \hline
        \hline
        5\%              & 364           & 210              & \textbf{75}      & \textbf{76}       \\
        \hline
        50\%             & 98            & 86               & \textbf{49}      & 244               \\
        \hline
        100\%            & 119           & 116              & \textbf{106}     & 240               \\
        \hline
    \end{tabular}
    \label{table:cpus}
\end{table}

Table~\ref{table:cpus} also shows the aggregated energy variation of the different generations of nodes for the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
The results attest that the latest versions of CPU do not necessarily cause more variation.
In the experiments we ran, the nodes from the cluster \textsf{Paranoia} tend to cause more variation at high workloads, even if they are from the latest generation.
While the Skylake CPU of the cluster \textsf{Dahu} cause often more energy variation than \textsf{Chetemi} and the Ecotype Broadwell CPU.
We argue that the hypothesis "\emph{the energy consumption on newer CPU varies more}" could be true or not depending on the compared generations, but most importantly, the chips energy behaviors.
On the other hand, our experiments showed the lowest energy variation when using the \textsf{Ecotype CPU}, these CPU are not the oldest nor the latest, but are tagged with "\texttt{L}" for their low power/TDP.
This result rises another hypothesis when considering CPU choice, which implies selecting the CPU with a low TDP.
This hypothesis has been confirmed on all the \textsf{Ecotype} cluster nodes, especially at low and medium workloads.

Figure~\ref{fig:cpugen} is an illustration of the aggregated STD density of more than $5,000$-random values sets taken from all the conducted experiments.
This shows that the cluster \textsf{Paranoia} reports the worst variation in most cases, and that \textsf{Ecotype} is the best cluster to consider to get the least variations, as it has a higher density for small variation values.

\begin{figure}
    \center{\includegraphics[width=.9\linewidth]{imgs/cpugen}}
    \caption{Energy consumption STD density of the 4 clusters}\label{fig:cpugen}
\end{figure}

\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
    We conclude on \textbf{affirming \textsc{RQ}~4}, as selecting the right CPU can help to get less variations.
\end{mdframed}

\section{Experimental Guidelines}\label{sec:guidelines}
To summarize our experiments, we provide some experimental guidelines in Table~\ref{table:guidelines}, based on the multiple experiments and analysis we did.
These guidelines constitute a set of minimal requirements or best practices, depending on the workload and the criticality of the energy measurement precision.
It therefore intends to help practitioners in taming the energy variation on the selected CPU, and conduct the experiments with the least variations.

\begin{table}[h!]
    \centering
    \caption{Experimental Guidelines for Energy Variations}
    \small
    \begin{tabular}{|p{4.7cm}|c|c|}
        \hline
        \textbf{Guideline}                                                                                                                                                & \textbf{Load} & \textbf{Gain}     \\
        \hline
        \hline
        Use a low TDP CPU                                                                                                                                                 & Low \& medium & Up to $3\times$   \\
        \hline
        Disable the CPU C-states                                                                                                                                          & Low           & Up to $6\times$   \\
        \hline
        Use the least of sockets in a case of multiple CPU                                                                                                                & Medium        & Up to $30\times$  \\
        \hline
        Avoid the usage of hyper-threading whenever possible                                                                                                              & Medium        & Up to $5\times$   \\
        \hline
        Avoid rebooting the machine between tests                                                                                                                         & High          & Up to $1.5\times$ \\
        \hline
        Do not relate to the machine idle variation to isolate a test EC, the CPU/OS changes its behavior when a test is running and can exhibit less variation than idle & Any           & ---               \\
        \hline
        Rather focus the optimization efforts on the system under test than the OS                                                                                        & Any           & ---               \\
        \hline
        Execute all the similar and comparable experiments on a same machine. Identical machines can exhibit many differences regarding their energy behavior             & Any           & Up to $1.3\times$ \\
        \hline
    \end{tabular}
    \label{table:guidelines}
\end{table}

Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and its variation reduction at low workloads.
However, it also lists some new factors that we identified along the analysis we conducted in Section~\ref{sec:parameters}, such as the results related to the OS or the reboot mode.
Some of the guidelines are more useful/efficient for specific workloads, as showed in our experiments.
Thus, qualifying the workload before conducting the experiments can help in choosing the proper guidelines to apply.
Other studied factors are not been mentioned in the guidelines, like the Turbo~Boost or the Speculative execution, due to the small effect that has been observed in our study.

In order to validate the accuracy of our guidelines among a varied set of benchmarks on one hand, and their effect on the variation between identical machines on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
Half of these experiments has been performed at a 50\,\% workload and the other half on single process jobs.
The choice of these two workloads is related to the optimization guidelines that are mainly effective at low and medium workloads.
We note that we used the cluster \textsf{Dahu} over \textsf{Ecotype} to highlight the guidelines effect on the nodes where the variation is susceptible to be higher.

\begin{figure*}%[!htb]
    \center{\includegraphics[width=\linewidth]{imgs/all}}
    \caption{Energy variation comparison with/without applying our guidelines}\label{fig:optimized}
\end{figure*}

\begin{figure}%[!htb]
    \center{\includegraphics[width=.9\linewidth]{imgs/stressng}}
    \caption{Energy variation comparison with/without applying our guidelines for \textsc{Stress-NG}}\label{fig:stress}
\end{figure}

Figure~\ref{fig:optimized} and~\ref{fig:stress} highlight the improvement brought by the adoption of our guidelines.
They demonstrate the intra-node STD reduction at low and medium workloads for all the benchmarks used at different levels.
Concretely, for low workloads, the energy variation is 2--6 times lower after applying the optimization guidelines for the benchmarks \textsf{LU} and \textsf{EP}, as well as \textsc{Linpack}, while it is 1.2--1.8 times better for \texttt{Sha256}.
For this workload, the overall energy consumption after optimization can be up to 80\,\% higher due to disabling the C-states to keep all the unused cores at a high power consumption state ($Mean_{LU-normal-Dahu2}=11,500 mJ$, $Mean_{LU-optimized-Dahu2}=20,508 mJ$).
For medium workloads, the STD, and thus variation, is up to 100\,\% better for the benchmark \textsf{CG}, 20--150\,\% better for the \texttt{pbzip2} application and up to 100\% for \textsc{Stress-NG}.
We note that the optimized version consumes less energy thanks to an appropriate core pinning method.

Figures~\ref{fig:optimized} and~\ref{fig:stress} also highlight that applying the guidelines does not reduce the inter-nodes variation in all the cases.
This variation can be up to 30\,\% in modern CPU~\cite{wang_experimental_nodate}.
However, taming the intra-node variation is a good strategy to identify more relevant mediums and medians, and then perform accurate comparisons between the nodes variation.
Even though, using the same node is always better, to avoid the extra inter-nodes variation and thus improve the stability of measurements.

\section{Threats to Validity}\label{sec:threats}
A number of issues affect the validity of our work.
For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPU, but still adds an important overhead if we adopt a sampling at high frequency.
The other fine-grained tool we used for measurements is \textsc{PowerAPI}.
It allows to measure the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
The usage of \textsc{PowerAPI} adds an error margin because of the power model built over RAPL.
The RAPL tool mainly measures the CPU and DRAM energy consumption.
However, even running CPU/RAM intensive benchmarks would keep a degree on uncertainty concerning the hard disk and networking energy consumption.
In addition, the operating system adds a layer of confusion and uncertainty.

The Intel CPU chip manufacturing process and the materials micro-heterogeneity is one of the biggest issues, as we cannot track or justify some of the energy variation between identical CPU or cores.
These CPU/cores might handle frequencies and temperature differently and behave consequently.
This hardware heterogeneity also makes reproduction complex and requires the usage of the same nodes on the cluster with the same OS.

% A more subtle issue may arise due to the values of the measurements that we achieved.
% In fact, the energy measures are quite small, and iterations may be taking a few milliseconds more or less to run.
% A thing we cannot measure using our measurement tools.
% How generalizable are our results? As a set of energy variation optimization guidelines, we argue that our results applied on most of the modern Intel CPU.
% However, using and comparing identical CPU is still tricky and is very dependent to chips.

\section{Conclusion}\label{sec:conclusion}
In this paper, we conducted an empirical study of controllable factors that can increase the energy variations on platforms with some of the latest CPU, and for several workloads.
We provide a set of guidelines that can be implemented and tuned (through the OS GRUB for example), especially with the new data centers isolation trend and the cloud usage, even for scientific and R\&D purposes.
Our guidelines aim at helping the user in reducing the CPU energy variation during systems benchmarking, and conduct more stable experiments when the variation is critical.
For example, when comparing the energy consumption of two versions of an algorithm or a software system, where the difference can be tight and need to be measured accurately.

Overall, our results are not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
The aim of our work is to be able to tame and mitigate this variability along controlled experiments.
We studied some previously discussed aspects on some recent CPU, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to achieve the variability mitigating purpose.
Some of these factors, like the C-states usage, can reduce the energy variation up to 500\,\% at low workloads, while choosing the wrong cores/PU strategy can cause up to $30\times$ more variability.

We believe that our approach can also be used to study/discover other potential variability factors, and extend our results to alternative CPU generations/brands.
Most importantly, this should motivate future works on creating a better knowledge on the variability due to CPU manufacturing process and other factors.

### 2021-08-30 chapters/testing/reproducibility.tex
add : my old files

\section{Reproducibility}

%%%%%%%%% preliminary taughts and section structure 
% The need of reproducibility in our field - software optimization based on empirical studies -

% The importance of Virtualisation for reproducibility \cite{howe_virtual_2012}
% some of the most important parts are
% - fewer constraints on research methods
% - on-demand backups
% - virtual Machines as Publications
% - more Variables captured

One of the most challenges which face the reaserchers is the reproducibility of their tests.Actually many results are not reproducible \footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.}, which let to a \emph{replication crisis}.When this crisis hit most of the empirical studies,most of the reviews now includes reproducibility as one of the minimal standard for judging scientific \cite{peng2011reproducible}. one of the creterions for a reproducidibility is the publication of the dataset and the algorithm run on the raw data to condlude the results. There is even some disagreement about what the terms "reproducibility" or "replicability" by themselves mean \cite{goodman2016does} .According to \cite{echtler2018open} \emph{replicability} extends \emph{reproducibility} to the ability to collect a new raw dataset comparable to the original one by reexecuting the experiment under similar condition, Instead of just the ability to get the same results by ruruning the statistical analyses on the orginal data set.

One major problem that hinders the reproducibility of the empirical tests is the interaction with the external environment, either as concurency or dependencies. Therefore testers won't have the same results unless they duplicate the same environment.
% introduction of vms 
\subsection{Virtualisation}
To resolve this problem, practitioners tend to use Virtualisation. Using virtual machines aka VM gives reaserchers the freedom to choose their own tools, software and operating system that they are the most confortable with without paying the price to change the actual working environment, which will give them eventually more controle over the dependencies and the execution environement. Moreover, using a vm will solve the \emph{replication crisis} thanks to the virtual images, even the most complex architecture can be reproduced easily by just instantiating a copy of the image. Since the virtual machines are agnostic to the host architecture, reaserchers won't have to worry about where and how their experiments are replicated because they have already setup the execution environement. Another advantage to the virtual machines is the snapshot mechanism, it allows reaserchers to create backups and revert some changes with simple clicks. Last but not least,thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables -controled and uncontroled-  and do other analysis without dealing with any dependencies. In his paper \cite{howe_virtual_2012} bill howe lists the advantages of using virtual machines in reaserchers experements including the economical impact and cultural limitation to a such approach.

which allow them to have control over the ressouces, the dependencies and the execution environment. Moreover, thanks to the snapshots, deploying a software is easily done by instantiating a copy of that image.
However, this choice comes with a certain cost. Because the intervention of the hypervisor,the software will use two kernels , the virtual machine one and the host machine one, which will provide a noticiable overhead, and will impact the performances of the tests.Therefore, we can't use virtual machines for exepements that are related to performance. Another limiation with the virtual machine is the the isolation. It is true that this feature will prevent the experience environement with any undesirable interference from the outside world. but sometimes this contact is needed, especially when the experiment is dependent to an external part, such as sensors. In energicial tests we tend to use hardware powermeters which will make it difficult to use the virtual machines in this case .

% introduction of docker

\subsection{Dockerisation}
another solution would be using something that allows us to have the isolation from the host os and the ease of replication that virtual machines offer, and the direct interaction with the hardware that the classical method give.

One interessting solution would be using containers.

\subsection{Testing Protocol}
the results of this protocil can be found in the project of the continious integrtation that was made by the intern MAMADou to give a feedback about the evolution of the energyconsumption within time

### 2021-08-30 chapters/testing/testing.tex
add : my old files

\newpage
\chapter{Testing}
\label{chapter:testing}

In order to optimize the energy consumption of software we first want to setup a testing enviroment. This chapter will be dedicated to provide a set of guidelines to provide a \em{reproducible, accurate} and \em{representative} tests.
Each aspect will be detailed in the sections below

\import{chapters/testing/}{reproducibility}
\import{chapters/testing/}{accuracy}
\import{chapters/testing/}{representativeness}

### 2021-09-03 chapters/testing/reproducibility.tex
add : docker vs vm machines

@@
-\section{Reproducibility}
+\section{Empirical Evaluation Protocol}
+
+This thesis is based on the empirical analyses to measure the energy consumption.
+As we mentioned earlier, we have chosen the \emph{agile} methode.
+so basically we will reduce the energy consumption with trial and error.
 
@@
 % - more Variables captured
+\subsection{Threats and Challenges}
 
@@
 
-\subsection{Dockerisation}
+\begin{figure}
+    \center{\includegraphics[width=1\linewidth]{imgs/virtualization_techniques}}
+    \caption{Different Methods of Virtualzation}\label{reproducibility:virtualization_technique}
+\end{figure}
+
+\subsection{Containers}
+
 another solution would be using something that allows us to have the isolation from the host os and the ease of replication that virtual machines offer, and the direct interaction with the hardware that the classical method give.
+contarization offers a such advantages while keeping the isolation and the ease of replication for application.
 
-One interessting solution would be using containers.
+Figure \ref{reproducibility:virtualization_technique} explains the differents in architecture between the clasic types2 of Virtualisation and Containers.
+\begin{itemize}
+    \item Type 1: runs directly on the hardware, it is mainly used by the cloud providers where there is no main OS, but just virtaul machines, we can site for this the open-srouce XEN and VMware ESX
+    \item Type 2: runs over the hostmachine Operating System, mostly used for personal comptuers, VMware server and virtualBox are famous examples of this type, most of the reaserchers expermentation are run witht this type, howerver due to the 2 Operatings syestms the applications tend to be more slower
+    \item containers : Instead of its own kernel, containers used the hots kernel to run their Os, which makes them ligher, quickers and use the full pontial use of the hardware.
+For this we can cite \emph{Docker}, \emph{Linux LXC}\emph{LXD} \cite{abuabdo_virtualization_2019}
+\end{itemize}
+\note{Rephrase so i include it before VMs+ add links to the technologies }
+
+\subsection{Docker vs Virtual Machine}
+Depsite that Type 1 is more performant than type 2, the second one is the most used in reaserch, since most researchers conducts their experements in their own machine.
+In the other hand, docker is the most famous thechonology for for containers,
+
+%%%%% THINGS TO ADD : 
+% the impact of docker on the reprodudicibility 
+% Limits of docker 
+% The docker files and the clarification of the methodology
+
+\subsection{Docker and energy}
+%  basically just rephrase the paper
+
+\subsection{docker and accuracy}
 
 \subsection{Testing Protocol}
-the results of this protocil can be found in the project of the continious integrtation that was made by the intern MAMADou to give a feedback about the evolution of the energyconsumption within time
+the results of this protocil can be found in the project of the continious integrtation that was made by the intern Mamadou to give a feedback about the evolution of the energyconsumption within time

### 2021-09-03 chapters/testing/testing.tex
add : docker vs vm machines

@@
 \newpage
+
 \chapter{Testing}
@@
 
-In order to optimize the energy consumption of software we first want to setup a testing enviroment.
+In order to optimize the energy consumption of software we first want to setup a experemental enviroment.
 This chapter will be dedicated to provide a set of guidelines to provide a \em{reproducible, accurate} and \em{representative} tests.
-Each aspect will be detailed in the sections below
+Each aspect will be detailed in the sections below.
 

### 2021-10-21 chapters/testing/challenges.tex
draft. first chapter

\section{Threats and Challenges}

A seccusefull test has three creterions tp fullfill. First, it has to be \emph{reproducible} in order for others to replecate. Second, the results should be \emph{accurate}, which means each time we rerun the test we are expecting the same results. Finally, it should \emph{represent} the real world. In other words, the conclusions brought from the experemnt should be valide outside the expermentation as well. In our case the real world is the production enviroment, Therefore our experements should reflect what is happening in the production environements.
In this section we will dig deeper in each aspect, present what has been done by others and ourselfs as well, and finally we will propose some perspectives, to make the experements better .

%%%%%%%%% preliminary taughts and section structure 
% The need of reproducibility in our field - software optimization based on empirical studies -

% The importance of Virtualisation for reproducibility \cite{howe_virtual_2012}
% some of the most important parts are
% - fewer constraints on research methods
% - on-demand backups
% - virtual Machines as Publications
% - more Variables captured

\subsection{Reproducibiliy}

One of the most challenges which face the reaserchers is the reproducibility of their tests.Actually many results are not reproducible \footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.}, which let to a \emph{replication crisis}.When this crisis hit most of the empirical studies,most of the reviews now includes reproducibility as one of the minimal standard for judging scientific \cite{peng2011reproducible}. one of the creterions for a reproducidibility is the publication of the dataset and the algorithm run on the raw data to condlude the results. There is even some disagreement about what the terms "reproducibility" or "replicability" by themselves mean \cite{goodman2016does} .According to \cite{echtler2018open} \emph{replicability} extends \emph{reproducibility} to the ability to collect a new raw dataset comparable to the original one by reexecuting the experiment under similar condition, Instead of just the ability to get the same results by ruruning the statistical analyses on the orginal data set.

In this area, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only on homogeneous nodes, but even within a single node.

One major problem that hinders the reproducibility of the empirical tests is the interaction with the external environment, either as concurency or dependencies. Therefore testers won't have the same results unless they duplicate the same environment.

\subsection{accuracy}
according to oxford, \emph{accuracy} means \href{https://www.lexico.com/definition/accuracy}{"technical The degree to which the result of a measurement, calculation, or specification conforms to the correct value or a standard"}.
Compare with precision. In our case this means the ability to run the benchmark multiple times with a low variation.

Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.

In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of the measures. Unfortunately, this is not possible  when it comes to measuring the energy consumption of a system. applying the benchmarking guidelines and repeating the same experiment with in the same configuration are not enought to reproduce the the same energy measures,not only between identical machines but even within the same machine.
This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
\begin{figure}%[!htb]
    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
\end{figure}

%TODO:  rephrase this figure, maybe change it a little bit 
Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
we can observe a large variation of the energy consumption, not only among homogeneous machiness, but also at the scale of a single machines, reaching up to $25\,\%$ in this example.

Some reaserchers started investigating the hardware impact of the energy variation of power consumption. As an example we cite ~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the main cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
\cite{heinrich_predicting}. decribed this variation as a set of parameters such as CPU Frequency and the termal effect.

\subsection{representativeness}
As obvious as it seems, the reason of the doing tests is to validate ideas so we can use them in the real life. Howerver this is means that those tests have to repesent reality somehow.
Basically when we want to test something we create a mock up version of the situation that we want it to wrok. But how can we assure that the test is representative ?. honestly i don't know. and we can't generalize this. but there are some works that have been done for this.
First we can talk about the benchmarking and their selection, then we will talk about the tress test for some applications. and finally we will bring this representativeness in our case and how can we get closer to the energy consumption behaviour of our software between the test environement and the production one.

As Stephen M. Blackburn et al cited in their paper "evaluate collaboratory" \cite{stephen_evaluate_2012} one of major pitfalls of the measurement contexts is the inconsistancy which is translated here with the fact that the production context is not the same as the testing one.

another difficult part for practicioners to generlize their claimes that they have got in the lab outside is the the the workloads. are they appripriate ? are they consistent and are they reproducible ?. To answer those questions the community agreed on some well known benchmarks that they represents an aspect of the the production world. we can site as an exemple the Dacapo set and Renaissance for JAVA applciations. the CLBG benchmark set for programming langauges.
Although they  do not cover all the cases. the community agrees on them. for the moment

In addition to those benchmarks a new category of testings has been born to simulate the worst case of the production enviromenet, Stress tests, are tests meant to evaluate the behaviour of a software under extreme situations.
we can cite as an exemple Gatling for web application and stress-ng for hardwares.
%% YEP I KNOW, THIS IS JUST FOR ME TO READ IT LATER AND GIVE A BETTER CONTEXT

### 2021-10-21 chapters/testing/conclusion.tex
draft. first chapter

--- 2021-08-30 chapters/testing/accuracy.tex
@@
-This is the chapter where i will put all the works related to the a successful testing
-
-% \begin{abstract}
-%     The past decade witnessed the inclusion of power measurements to evaluate the energy efficiency of software systems, thus making energy a prime indicator along with performance.
-%     Nevertheless, measuring the energy consumption of a software system remains a tedious task for practitioners.
-%     In particular, the energy measurement process may be subject to a lot of variations that hinder the relevance of potential comparisons.
-%     While the state of the art mostly acknowledged the impact of hardware factors (chip printing process, CPU temperature), this paper investigates the impact of controllable factors on these variations.
-%     More specifically, we conduct an empirical study of multiple controllable parameters that one can easily tune to tame the energy consumption variations when benchmarking software systems.
-
-%     To better understand the causes of such variations, we ran more than a $1,000$ experiments on more than $100$ nodes with different workloads and configurations.
-%     The main factors we studied encompass: experimental protocol, CPU features (C-states, Turbo~Boost, core pinning) and generations, as well as the operating system.
-%     Our experiments showed that, for some workloads, it is possible to tighten the energy variation by up to $30\times$.
-%     Finally, we summarize our results as guidelines to tame energy consumption variations.
-%     We argue that the guidelines we deliver are the minimal requirements to be considered prior to any energy efficiency evaluation.
-% \end{abstract}
-
-\section{Introduction}
-%\lettrine[nindent=0em,lines=3]{C} 
-To conduct robust evaluations, practitioners often try to ensure reproducible environmental conditions in order to properly benchmark their software systems.
-In this area, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
-Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.
-
-In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should favour reproducible experiments.
-However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only on homogeneous nodes, but even within a single node.
-This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
-
-Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
-One can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
-    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
-\end{figure}
-
-Most of the state of the art has been investigating this power consumption issue from a hardware perspective~\cite{borkar_designing_2005,tschanz_adaptive_2002} and reported that the causes of such energy variations are CMOS manufacturing process of transistors in a chip, differences in node assembly and data center hot spots.
-Additionally, \cite{heinrich_predicting} described it as a combination of parameters, mentioning a list of candidate factors, such as the thermal effect or the CPU frequency, but failing to deliver a deeper analysis of these factors.
-
-Unfortunately, hardware factors can be hardly tuned to tame the energy variation that can be observed in systems benchmarking.
-For example, managing the CPU temperature or a server position in a cluster, are not actions that one can easily do, especially on the modern data centers and cloud platforms.
-Therefore, the goal of this paper is to investigate the spectrum of factors that can cause or increase the variability of energy consumption in systems benchmarking, and to propose effective guidelines to control such factors in order to mitigate this variability.
-While this paper does not question the benefits of established CPU features, like C-states or Turbo~Boost, it delivers a deeper analysis of the effects they may introduce on a wide set of experiments and nodes.
-By quantifying potential energy variations induced by controllable factors, we intend to identify the proper configurations that minimize energy variations, depending on the workload characteristics.
-These guidelines aim at supporting practitioners in conducting robust system benchmarks and reporting reproducible energy consumption, rather than recommending production-scale practices or solutions to reduce the power consumption of a given system.
-% 
-% The key contributions of this paper can therefore be summarized as:
-% \begin{enumerate}
-%     \item Providing a better understanding of the energy variation, by using different generations of CPU deployed in $4$ clusters with more than $100$ physical nodes, and by considering existing systems benchmarks with diverse workloads;
-%     \item Identifying controllable factors that contribute to the variation in CPU energy consumption, comparing them against the state of the art, and completing them with other uncovered assumptions;
-%     \item Reporting on some guidelines on how to conduct reproducible experiments with less energy variations;
-%     \item Discussing the differences between inter-nodes and intra-nodes variations.
-% \end{enumerate}
-
-%\paragraph{}
-% The remainder of this paper is organized as follows.
-% Section~\ref{sec:related_work} discusses the related works in the area of power measurements.
-% Section~\ref{sec:research_qs} formalizes our research questions.
-% Section~\ref{sec:setup} reports on the experimental setup (hardware, benchmarks, tools and methodology) we used in this work.
-% Section~\ref{sec:parameters} analyzes the causes of the variations we observed along experiments.
-% Finally, we discuss the results of the different experimented factors, and their impact on the energy variation in Section~\ref{sec:guidelines}, while Sections~\ref{sec:threats} \&~\ref{sec:conclusion} cover the validity threats and our conclusions, respectively.
-
-%------------------------------------------------
-\section{Related Work}\label{sec:related_work}
-In this section, we review the state of the art in the domain of energy consumption analysis.
-In particular, one can observe that several studies considered the evaluation of the processor variation while executing the same workload, in particular to test distributed applications running in homogeneous clusters.
-
-\paragraph{Studying Hardware Factors.}
-
-This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
-On the one hand, the correlation between the processor temperature and the energy consumption was one of the most explored paths.
-Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
-On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the correlation between the CPU temperature and the energy consumption variation is very tight.
-
-\note{add the corelation value}
-
-This makes the processor temperature a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.
-
-The ambient temperature was also discussed in many papers as a candidate factor for the energy variation of a processor.
-In~\cite{ranka_energy_2009}, the authors claimed that energy consumption may vary due to fluctuations caused by the external environment.
-These fluctuations may alter the processor temperature and its energy consumption.
-However, the temperature inside a data center does not show major variations from one node to another.
-In~\cite{el_mehdi_diouri_your_2013}, El~Mehdi~Dirouri~\emph{et~al.} showed that switching the spot of two servers does not affect their energy consumption.
-Moreover, changing hardware components, such as the hard drive, the memory or even the power supply, does not affect the energy variation of a node, making it mainly related to the processor.
-This result was recently assessed by~\cite{wang_potential_2018}, where the rack placement and power supply introduced a maximum of $2.8\,\%$ variation in the observed energy consumption.
-
-Beyond hardware components, the accuracy of power meters has also been questioned.
-Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
-All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
-
-\paragraph{Mitigating Energy Variations.}
-Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
-In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
-The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
-Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
-They also proposed some guidelines to reduce this variation by replacing the old slow chips, by load balancing the workload on the CPU cores and leaving one core idle.
-They claimed that the variation between the processor cores is insignificant.
-In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.
-
-In~\cite{marathe_empirical_2017_m}, the authors highlight the increase of energy variation across the latest Intel micro-architectures by a factor of $4$ from Sandy~Bridge to Broadwell, a $15\,\%$ of run-to-run variation within the same processor and the increase of the inter-cores variation from $2.5\,\%$ to $5\,\%$ due to hardware-enforced constraints, concluding with some recommendations for Broadwell usage, such as running one hyper-thread per core.
-\note{this might be interesting for greeFaaS}
-
-\paragraph{Objective.}
-The aim of our paper is not to reproduce all the exposed results and explored paths.
-Instead, we intend to check some of these results on latest hardware components, and to consider new potential controllable factors that could also contribute to the energy variation.
-We also aim to provide practitioners with guidelines to conduct robust experiments in a controlled environment, thus reducing the variation overhead to share trustable results and build credible comparisons.
-
-\section{Research Questions}\label{sec:research_qs}
-Worth saying, we are aware that part of the energy consumption variation is due to chip manufacturing differences or some of the previously discussed enforced factors, such as the thermal effect or the servers placement.
-Those parameters are often tricky to manage, as we cannot have a perfect chips manufacturing process, or assume that two identical processors have the same thermal behavior.
-We will therefore focus on providing the practitioners with an empirical study of some controllable parameters that can be tuned to conduct experimental evaluations of energy consumption with less variation, especially if the practitioners do not have a physical access to the data center or BIOS configuration, which is the case on most of the modern cloud platforms and data centers.
-These parameters span the choice of the benchmarking protocol, processor frequencies management, operating system tuning or some other parameters.
-The authors of~\cite{heinrich_predicting} mentioned some of these potential parameters.
-
-In this paper, we will therefore investigate the following controllable factors, which we formalize as 4 research questions:
-\begin{description}
-    \item[\textsc{RQ}~1:] Does the benchmarking protocol affect the energy variation?
-    \item[\textsc{RQ}~2:] How important is the impact of the processor features on the energy variation?
-    \item[\textsc{RQ}~3:] What is the impact of the operating system on the energy variation? and finally
-    \item[\textsc{RQ}~4:] Does the choice of the processor matter to mitigate the energy variation?
-\end{description}
-
-\section{Experimental Setup}\label{sec:setup}
-This section describes our detailed experimental environment, covering the clusters and nodes configuration, the benchmarks we used and justifying our experimental methodology.
-
-\subsection{Hardware Platform}
-We considered 4 distinct clusters of variable sizes and different generations of CPU, as summarized in Table~\ref{table:g5k}.
-In particular, we used the Grid5000 (G5K) platform in our experiments~\cite{grid5000,margery:hal-00965708}.
-G5K is a bare metal cloud platform that can be used to provision clusters of identical nodes.
-In our study, we mainly used the cluster \textsf{Dahu} located in Grenoble to run most of our tests, as it has one of the newest Xeon CPUs.
-We also used the clusters \textsf{Chetemi}, \textsf{Ecotype} and \textsf{Paranoia} in some of our experiments.
-Table~\ref{table:g5k} describes the configurations of the clusters we considered.
-% We note that each cluster is geographically independent.
-
-\begin{table}
-    \centering
-    \caption{Description of clusters included in the study}
-    \label{table:g5k}
-    \small
-    \begin{tabular}{|l|l|r|r|}
-        \hline
-        \textbf{Cluster}  & \textbf{Processor}               & \textbf{Nodes} & \textbf{RAM} \\
-        \hline
-        \hline
-        \textsf{Dahu}     & $2\times$ Intel Xeon\,Gold\,6130 & 32             & 192\,GiB     \\
-        \hline
-        \textsf{Chetemi}  & $2\times$ Intel Xeon\,E5-2630v4  & 15             & 768\,GiB     \\
-        \hline
-        \textsf{Ecotype}  & $2\times$ Intel Xeon\,E5-2630Lv4 & 48             & 128\,GiB     \\
-        \hline
-        \textsf{Paranoia} & $2\times$ Intel Xeon\,E5-2660v2  & 8              & 128\,GiB     \\
-        \hline
-    \end{tabular}
-\end{table}
-
-As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or socket to designate one of the two sockets and \textsf{PU} for the operating system \emph{processing unit}.
-The number of \textsf{PU} often doubles the number of available cores because of the hyper-threading support, as the OS considers 2 hyper-threads sharing the same core as 2 different \textsf{PU}.
-Figure~\ref{fig:topo} illustrates a detailed topology of a node belonging to the cluster \textsf{Dahu}.
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/lstopo}}
-    \caption{Topology of the nodes of the cluster \textsf{Dahu}}\label{fig:topo}
-\end{figure}
-% The advantage of using G5K is having an access to a wide set of physical nodes with different capabilities. 
-% The usage procedure is to book a machine for period of time with the chosen OS in less than 5 minutes, ship and run the test script, then release the machine.
-When a machine is released everything is discarded even the OS.
-
-\subsection{Systems Benchmarks}
-Our first criterion to choose the systems benchmarks was the scalability, as we need to run tests at different workloads by choosing the right number of used PU.
-The other criteria are the documentation, the accuracy and the references to the benchmark.
-\emph{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925} is one of the most used suite of benchmarks in the HPC literature and it fulfills our benchmarking requirements.
-We mainly used the pseudo application \emph{Lower-Upper symmetric Gauss-Seidel} (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
-These are the main benchmarks used in many similar works, such as~\cite{heinrich_predicting}.
-Nonetheless, in order to validate our results on a wider set of benchmarks and applications, we also used \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}\footnote{\url{https://linux.die.net/man/1/sha256sum}} as representative systems benchmarks to conduct our experiments with a broad diversity of workloads.
-
-\subsection{Measurement Tools \& Methodology}
-To study the energy consumption of nodes, we considered Intel \emph{Running Average Power Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754}, which is one of the most accurate tools to report the CPU/DRAM global energy consumption.
-We also used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, which is a power monitoring toolkit that builds a model over RAPL to compute the energy consumption at process-level when we needed to isolate energy consumption of a single process.
-Our clusters are provisioned with a minimal version of Debian\,9 (4.9.0 kernel version) where we install Docker (version 18.09.5), which will be used to run the RAPL sensor and the benchmark itself.
-The energy sensor collects RAPL reports and stores them in a remote \textsc{MongoDB} instance, allowing us to perform \emph{post-mortem} analysis in a dedicated environment.
-Using Docker makes the deployment process easier on the one hand, and provides us with a built-in control group encapsulation of the conducted tests on the other hand.
-This allows \textsc{PowerAPI} to measures all the running containers, even the RAPL sensor consumption, as it is isolated in a container.
-One potential threat covers the impact of Docker on the energy variation.
-We therefore conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
-Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
-In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
-The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
-    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
-\end{figure}
-
-Every experiment is conducted on $100$ iterations, on multiple nodes and using the 3 NPB benchmarks we mentioned, with a warmup phase of 10 iterations for each experiment.
-In most cases, we were seeking to evaluate the \emph{STandard Deviation} (STD), which is the most representative factor of the energy variation.
-We tried to be very careful, while running our experiments, not to fall in the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
-As we study the STD difference of measurements we observed from empirical experiments, we use the bootstrap method~\cite{efron2000bootstrap} to randomly build multiple subsets of data from the original dataset, and we draw the STD density of those sets, as illustrated in Figure~\ref{fig:docker}.
-Given the space constraints, this paper reports on aggregated results for nodes, benchmarks and workloads, but the raw data we collected remains available through the public repository we published.\footnote{\url{https://github.com/anonymous-data/Energy-Variation}}
-We believe this can help to achieve better and more reliable comparisons.
-
-We mainly consider 3 different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium and high CPU usage when analyzing the studied parameters effect, respectively.
-These workloads reflect the ratio of used PU count to the total available PU.
-
-\section{Energy Variation Analysis}\label{sec:parameters}
-In this section, we aim to establish experimental guidelines to reduce the CPU energy variation.
-We therefore explore many potential factors and parameters that could have a considerable effect on the energy variation.
-
-\subsection{\textsc{RQ}~1: Benchmarking Protocol}
-To achieve a robust and reproducible experiment, practitioners often tend to repeat their tests multiple times, in order to analyze the related performance indicators, such as execution time, memory consumption or energy consumption.
-We therefore aim to study the benchmarking protocol to identify how to efficiently iterate the tests to capture a trustable energy consumption evaluation.
-% This process of repeating executions should not be random in order to avoid different results because of the CPU behavior, the cache data, the memory size and plenty of other aspects.
-
-In this first experiment, we investigate if changing the testing protocol affects the energy variation.
-To achieve this, we considered 3 execution modes:
-In the \textsf{"normal"} mode, we iteratively run the benchmark 100 times without any extra command, while the \textsf{"sleep"} mode suspends the execution script for 60\,seconds between iterations.
-Finally, the \textsf{"reboot"} mode automatically reboots the machine after each iteration.
-The difference between the \textsf{normal} and \textsf{sleep} modes intends to highlight that the CPU needs some rest before starting another iteration, especially for an intense workload.
-Putting the CPU into sleep for several seconds could give it some time to reach a lower frequency state or/and reduce its temperature, which could have an impact on the energy variation.
-The \textsf{reboot} mode, on the other hand, is the most straightforward way to reset the machine state after every iteration.
-It could also be beneficial to reset the CPU frequency and temperature, the stored data, the cache or the CPU registries.
-However, the reboot task takes a considerable amount of time, so rebooting the node after every single operation is not the fastest nor the most eco-friendly solution, but it deserves to be checked to investigate if it effectively enhances the overall energy variation or not.
-
-Figure~\ref{fig:running-process} reports on $300$ aggregated executions of the  benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}, on 4 machines of the cluster \textsf{Dahu} (cf.
-Table~\ref{table:g5k}) for different workloads.
-We note that the results have been executed with different datasets sizes (\textsf{B}, \textsf{C} and \textsf{D} for single process, 50\,\% and 100\,\% respectively) to remedy to the brief execution times at high workloads for small datasets.
-This justifies the scale differences of reported energy consumptions between the 3 modes in Figure~\ref{fig:running-process}.
-As one can observe, picking one of these strategies does not have a strong impact on the energy variation for most workloads.
-In fact, all the strategies seem to exhibit the same variation with all the workloads we considered---\emph{i.e.}, the STD is tightly close between the three modes.
-The only exception is the \textsf{reboot} mode at 100\,\% load, where the STD is 150\,\% times worst, due to an important amount of outliers.
-This goes against our expectation, even when setting a warm-up time after reboot to stabilize the OS.
-
-\begin{figure*}
-    \center{\includegraphics[width=\linewidth]{imgs/running-process}}
-    \caption{Energy variation with the \textsf{normal}, \textsf{sleep} and \textsf{reboot} modes}\label{fig:running-process}
-\end{figure*}
-
-In Figure~\ref{fig:reboot}, we study the standard deviation of the three modes by constituting $5,000$ random 30-iterations sets from the previous executions set and we compute the STD in each case, considering mainly the 100\,\% workload as the STD was 150\,\% higher for the \textsf{reboot} mode with that load.
-We can observe that the considerable amount of outliers in the \textsf{reboot} mode is not negligible, as the STD density is clearly higher than the two other modes.
-This makes the \textsf{reboot} mode as the less appropriate for the energy variation at high workloads.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/reboot}}
-    \caption{STD analysis of the \textsf{normal}, \textsf{sleep} and \textsf{reboot} modes}\label{fig:reboot}
-\end{figure}
-
-\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
-
-    To answer \textsc{RQ~1}, we conclude that the benchmarking protocol \textbf{partially affects} the energy variation, as highlighted by the \textsf{reboot} mode results for high workloads.
-
-\end{mdframed}
-
-\subsection{\textsc{RQ}~2: Processor Features}
-The C-states provide the ability to switch the CPU between more or less consuming states upon activities.
-Turning the C-states on or off have been subject of many discussions~\cite{5463056}, because of its dynamic frequency mechanism but, to the best of our knowledge, there have been no fully conducted C-states behavior analysis on CPU energy variation.
-
-We intend to investigate how much the energy consumption varies when disabling the C-states (thus, keeping the CPU in the \texttt{C0} state) and at which workload.
-Figure~\ref{fig:c-states} depicts the results of the experiments we executed on three nodes of the cluster \textsf{Dahu}.
-On each node, we ran the same set of benchmarks with two modes: \textsf{C-states on}, which is the default mode, and \textsf{C-states off}.
-Each iteration includes 100 executions of the same benchmark at a given workload, with three workload levels.
-We note that our results have been confirmed with the benchmarks \texttt{LU}, \texttt{CG} and \texttt{EP}.
-
-We can clearly see the effect that has the \textsf{C-states off} mode when running a single-process application/benchmark.
-The energy consumption varies 5~times less than the default mode.
-In this case, only one CPU core is used among $2\times16$ physical cores.
-The other cores are switched to a low consumption state when C-states are on, the switching operation causes an important energy consumption difference between the cores, and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
-On the other hand, switching off the C-states would keep all the cores---even the unused ones---at a high frequency usage.
-This highly reduces the variation, but causes up to 50\,\% of extra energy consumption in this test ($Mean_{C-states-off}=11,665 mJ$,$Mean_{C-states-on}=7,641 mJ$).
-
-At a 100\,\% workload, disabling the C-states seems to have no effect on the total energy consumption nor its variation.
-In fact, all the cores are used at 100\,\% and the C-states module would have no effect, as the cores are not idle.
-The same reason would apply for the 50\,\% load, as the hyper-threading is active on all cores, thus causing the usage of most of them.
-For single process workloads, disabling the C-states causes the process to consume 50\,\% more energy as reported in Figure~\ref{fig:c-states}, but reduces the variation by 5~times compared to the \textsf{C-states on} mode.
-This leads to mainly two questions: Can a process pinning method reduce/increase the energy variation? And, how does the energy consumption variation evolve at different PU usage level?
-
-\begin{figure*}
-    \center{\includegraphics[width=\linewidth]{imgs/c-states}}
-    \caption{Energy variation when disabling the C-states}\label{fig:c-states}
-\end{figure*}
-
-\subsubsection{Cores Pinning}
-To answer the first question, we repeated the previous test at 50\,\% workload.
-In this experiment, we considered three cores usage strategies, the first one (\textsf{S1}) would pin the processes on all the PU of one of the two sockets (including hyper-threads), so it will be used at 100\,\%, and leave the other CPU idle.
-The second strategy (\textsf{S2}) splits the workload on the two sockets so each CPU will handle 50\,\% of the load.
-In this strategy, we only use the core PU and not the hyper-threads PU, so every process would not share his core usage (all the cores are being used).
-The third strategy (\textsf{S3}) consists also on splitting the workload between the two sockets, but considering the usage of the hyper-threads on each core---\emph{i.e.}, half of the cores are being used over the two CPU.
-Figure~\ref{fig:cores-pinning} reports on the energy consumption of the three strategies when running the benchmark \texttt{CG} on the cluster \textsf{Dahu}.
-We can notice the big difference between these three execution modes that we obtained only by changing the PU pinning method (that we acknowledged with more than 100 additional runs over more than 30 machines and with the benchmarks \texttt{LU} and \texttt{EP}).
-For example, \textsf{S2} is the least power consuming strategy.
-We argue that the reason is related to the isolation of every process on a single physical core, reducing the context switch operations.
-In the first and third strategy, $32$ processes are being scheduled on $16$ physical cores using the hyper-threads PU, which will introduce more context switching, and thus more energy consumption.
-
-\begin{figure}
-    \center{\includegraphics[width=.8\linewidth]{imgs/cores_pinning}}
-    \caption{Energy variation considering the three cores pinning strategies at 50\,\% workload}\label{fig:cores-pinning}
-\end{figure}
-
-We note that even if the first and third strategies are very similar (both use hyper-threads, but only on one CPU for the first and on two CPU for the third), the gap between them is considerable variation-wise, as the variation is 30 times lower in the first strategy ($STD_{S1}=116 mJ$,$STD_{S3}=3,452 mJ$).
-This shows that the usage of the hyper-threads technology is not the main reason behind the variation, the first strategy has even less variation than the second one and still uses the hyper-threading.
-
-The reason for the \textsf{S1} low energy consumption is that one of the two sockets is idle and will likely be in a lower power P-state, even with the disabled C-states.
-The \textsf{S2} case is also low energy consuming because by distributing the threads across all the cores, it completes the task faster than in the other cases.
-Hence, it consumes less energy.
-The \textsf{S3} is a high consuming strategy because both sockets are being used, but only half the cores are active.
-This means that we pay the energy cost for both sockets being operational and for the experiments taking longer to run because of the recurrent context switching.
-
-Our hypothesis regarding the worst results that we observed when using the third strategy is the recurrent context switching, added to the OS scheduling that could reschedule processes from a socket to another, which invalids the cache usage as a process can not take profit of the socket local L3 cache when it moves from a CPU to another (cf.
-Figure~\ref{fig:topo}).
-
-Moreover, the fact that the variation is 4--5 times higher when using the strategy \textsf{S2} compared to \textsf{S1} ($STD_{S1}=116 mJ$, $STD_{S3}=575 mJ$), gives another reason to believe that swapping a process from a CPU to another increases the variation due to CPU micro~differences, cache misses and cache coherency.
-While the mean execution time for the strategy \textsf{S3} is very high ($MeanTime_{S3}= 46 s$) compared to the two other strategies ($MeanTime_{S1}= 11 s$, $MeanTime_{S2}= 7 s$), we see no correlation between the execution time and the energy variation, as the \textsf{S1} still give less variations than \textsf{S2} even if it takes 36\,\% more time to run.
-
-Table~\ref{table:corespinning} reports on additional aggregated results for the STD comparison on four other nodes of the cluster \textsf{Dahu} at 50\,\%, with the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
-In fact, the CPU usage strategy \textsf{S1} is by far the experimentation mode that gave the least variation.
-The STD is almost 5 times better than the strategy \textsf{S2}, but is up to 10\,\% more energy consuming ($Mean_{S1}=4469 mJ$, $Mean_{S2}=4016 mJ$).
-On the other hand, the strategy \textsf{S3} is the worst, where the energy consumption can be up to 5 times higher than the strategy \textsf{S2} ($Mean_{S2}=4016 mJ$, $Mean_{S3}=21645 mJ$) and the variation is much worst (30 times compared to the first strategy).
-These results allow us to have a better understanding of the different processes-to-PU pinning strategies, where isolating the workload on a single CPU is the best strategy.
-Using the hyper-threads PU on multiple sockets seems to be a bad recommendation, while keeping the hyper-threading enabled on the machine is not problematic, as long as the processes are correctly pinned on the PU.
-Our experiments show that running one hyper-thread per core is not always the best to do, at the opposite of the claims of~\cite{marathe_empirical_2017_m}.
-
-\begin{table}
-    \centering
-    \caption{STD (mJ) comparison for 3 pinning strategies}
-    \label{table:corespinning}
-    \small
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \textbf{Strategy} & \textbf{S1} & \textbf{S2} & \textbf{S3} \\
-        \hline
-        \hline
-        \textbf{Node~1}   & 88          & 270         & 1,654       \\
-        \hline
-        \textbf{Node~2}   & 79          & 283         & 2,096       \\
-        \hline
-        \textbf{Node~3}   & 58          & 287         & 1,725       \\
-        \hline
-        \textbf{Node~4}   & 51          & 229         & 1,334       \\
-        \hline
-    \end{tabular}
-\end{table}
-
-\subsubsection{Processes Threshold}
-To answer the second question regarding the evolution of the energy variation at different levels of CPU usage, we varied the used PU's count to track the EV evolution.
-Figure~\ref{fig:process-tresh} compares the aggregated energy variation when the C-states are on and off using 2, 4 and 8 processes for the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
-This figure confirms that disabling the CPU C-states does not decrease the variation for all the workloads, as we can clearly observe, the variation is increasing along with the number of processes.
-When running only 2 processes, turning off the C-states reduces the STD up to 6~times, but consumes 20\,\% more energy ($Mean_{C-states-on}=10,334 mJ$, $Mean_{C-states-off}=12,594 mJ$).
-This variation is 4~times lower when running 4 processes and almost equal to the \textsf{C-states on} mode when running 8 processes.
-In fact, running more processes implies to use more CPU cores, which reduces the idle cores count, so the cores will more likely stay at a higher consumption state even if the C-states mechanism is on.
-
-In our case, using 4 PU reduces the variation by 4~times and consumes almost the same energy as keeping the C-states mechanism on ($Mean_{C-states-on}=7,048 mJ$, $Mean_{C-states-off}=7,119 mJ$).
-This case would be the closest to reality as we do not want to increase the energy consumption while reducing the variation, but using a lower number of PU still results in less variation, even if it increases the overall energy consumption.
-
-\begin{figure*}
-    \center{\includegraphics[width=\linewidth]{imgs/tresh}}
-    \caption{C-states effect on the energy variation, regarding the application processes count}\label{fig:process-tresh}
-\end{figure*}
-
-We note that disabling the C-states is not recommended in production environments, as it introduces extra energy consumption for low workloads (around 50\,\% in our case for a single process job).
-However, our goal is not to optimize the energy consumption, but to minimize the energy variation.
-Thus, disabling the C-states is very important to stabilize the measurements in some cases when the variation matters the most.
-Comparing the energy consumptions of two algorithms or two versions of a software systems is an example of use case benefiting from this recommendation.
-
-\subsubsection{Turbo Boost}
-The Turbo Boost---also known as \emph{Dynamic Overclocking}---is a feature that has been incorporated in Intel CPU since the Sandy~Bridge micro-architecture, and is now widely available on all of the Core\,i5, Core\,i7, Core\,i9 and Xeon series.
-It automatically raises some of the CPU cores operating frequency for short periods of time, and thus boost performances under specific constraints.
-When demanding tasks are running, the operating system decides on using the highest performance state of the processor.
-
-Disabling or enabling the Turbo~Boost has a direct impact on the CPU frequency behavior, as enabling it allows the CPU to reach higher frequencies in order to execute some tasks for a short period of time.
-However, its usage does not have a trivial impact on the energy variation.
-Acun~\emph{et~al.}~\cite{acun_variation_2016} tried to track the Turbo~Boost impact on the Ivy~Bridge and the Sandy~Bridge architectures.
-They concluded that it is one of the main responsible for the energy variation, as it increases the variation from 1\,\% to 16\,\%.
-In our study, we included a Turbo~Boost experiment in our testbed, to check this property on the recent Xeon Gold processors, covering various workloads.
-
-The experiment we conducted showed that disabling the Turbo Boost does not exhibit any considerable positive or negative effect on the energy variation.
-Table~\ref{table:turboboost} compares the STD when enabling/disabling the Turbo~Boost, where the columns are a combination of workload and benchmark.
-In fact, we only got some minor measurements differences when switching on and off the Turbo~Boost, and where in favor or against the usage of the Turbo~Boost while repeating tests, considering multiple nodes and benchmarks.
-This behavior is mainly related to the \emph{thermal design power} (TDP), especially at high workloads executions.
-When a CPU is used at its maximum capacity, the cores would be heating up very fast and would hit the maximum TDP limit.
-In this case, the Turbo~Boost cannot offer more power to the CPU because of the CPU thermal restrictions.
-At lower workloads, the tests we conducted proved that the Turbo~Boost is not one of the main reasons of the energy variation.
-In fact, the variation difference is barely noticeable when disabling the Turbo~Boost, which cannot be considered as a result regarding the OS activity and the measurement error margin.
-We cannot affirm that the Turbo~Boost does not have an impact on all the CPU, as we only tested on two recent Xeon CPU (clusters \textsf{Chetemi} and \textsf{Dahu}).
-We confirmed our experiments on these machines 100 times at 5\,\%, 25\,\%, 50\,\% and 100\,\% workloads.
-
-\begin{table}%[h!]
-    \centering
-    \caption{STD (mJ) comparison when enabling/disabling the Turbo~Boost}
-    \label{table:turboboost}
-    \small
-    \begin{tabular}{|c|r|r|}
-        \hline
-        \textbf{Turbo Boost}  & \textbf{Enabled} & \textbf{Disabled} \\
-        \hline
-        \hline
-        \textsf{EP} / 5\,\%   & 310              & 308               \\
-        \hline
-        \textsf{CG} / 25\,\%  & 95               & 140               \\
-        \hline
-        \textsf{LU} / 25\,\%  & 204              & 240               \\
-        \hline
-        \textsf{EP} / 50\,\%  & 84               & 79                \\
-        \hline
-        \textsf{EP} / 100\,\% & 125              & 110               \\
-        \hline
-    \end{tabular}
-\end{table}
-
-\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
-    We conclude that CPU features \textbf{highly impact} the energy variation as an answer for \textsc{RQ~2}.
-\end{mdframed}
-
-\subsection{\textsc{RQ}~3: Operating System}
-The \emph{operating system} (OS) is the layer that exploits the hardware capabilities efficiently.
-It has been designed to ease the execution of most tasks with multitasking and resource sharing.
-In some delicate tests and measurements, the OS activity and processes can cause a significant overhead and therefore a potential threat to the validity. % experiments validity
-The purpose behind this experiment is to determine if the sampled consumption can be reliably related to the tested application, especially for low-workload applications where CPU resources are not heavily used by the application.
-
-The first way to do is to evaluate the OS idle activity consumption, and to compare it to a low workload running job.
-Therefore, we ran 100 iterations of a single process benchmark \textsf{EP}, \textsf{LU} and \textsf{CG} on multiple nodes from the cluster \textsf{Dahu}, and compared the energy behavior of the node with its idle state on the same duration.
-The aggregated results, illustrated in Figure~\ref{fig:os-idle}, depict that the idle energy variation is up to 140\,\% worst than when running a job, even if it consumes 120\,\% less energy ($Mean_{Job}=8,746 mJ$, $Mean_{Idle}=3,927 mJ$).
-In fact, for the three nodes, randomly picked from the cluster \textsf{Dahu}, the idle variation is way more important than when a test was running, even if it is a single process test on a $32$-cores node.
-This result shows that OS idle consumption varies widely, due to the lack of activity and the different CPU frequencies states, but it does not mean that this variation is the main responsible for the overall energy variation.
-The OS behaves differently when a job is running, even if the amount of available cores is more than enough for the OS to keep his idle behavior when running a single process.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/idle}}
-    \caption{OS consumption between idle and when running a single process job}\label{fig:os-idle}
-\end{figure}
-
-Inspecting the OS idle energy variation is not sufficient to relate the energy variation to the active job.
-In fact, the OS can behave differently regarding the resource usage when running a task.
-To evaluate the OS and the job energy consumption separately, we used the \textsc{PowerAPI} toolkit.
-This fine-grained power~meter allows the distribution of the RAPL global energy across all the Cgroups of the OS using a power model.
-Thus, it is possible to isolate the job energy consumption instead of the global energy consumption delivered by RAPL.
-To do so, we ran tests with a single process workload on the cluster \textsf{Dahu}, and used the \textsc{PowerAPI} toolkit to measure the energy consumption.
-Then, we compared the job energy consumption to the global RAPL data.
-We calculated the Pearson correlation~\cite{ref1} of the energy consumption and variation between global RAPL and \textsc{PowerAPI}, as illustrated in Figure~\ref{fig:correlation}.
-The job energy consumption and variation are strongly correlated with the global energy consumption and variation with the coefficients 93.6\,\% and 85.3\,\%, respectively.
-However, this does not completely exclude the OS activity, especially if the jobs have tight interaction with the OS through the signals and system calls.
-This brings a new question on whether applying extra-tuning on a minimal OS would reduce the variation?
-As well as what is the effect of the Meltdown security patch---that is known to be causing some performance degradation~\cite{Kocher2018spectre,Lipp2018meltdown}---on the energy variation?
-
-\begin{figure}
-    \center{\includegraphics[width=.8\linewidth]{imgs/correlation}}
-    \caption{The correlation between the RAPL and the job consumption and variation}\label{fig:correlation}
-\end{figure}
-
-\subsubsection{OS Tuning}
-An OS is a pack of running processes and services that might or not be required its execution.
-In fact, even using a minimal version of a Debian Linux, we could list many OS running services and process that could be disabled/stopped without impacting the test execution.
-This extra-tuning may not be the same depending on the nature of the test or the OS.
-Thus, we conducted a test with a deeply-tuned OS version.
-We disabled all the services/processes that are not essential to the OS/test running, including the OS networking interfaces and logging modules, and we only kept the strict minimum required to the experiment's execution.
-Table~\ref{table:osmin} reports on the aggregated results for running single process measurements with the benchmarks \textsf{CG}, \textsf{LU} and \textsf{EP}, on three servers of the cluster \textsf{Dahu}, before and after tuning the OS.
-Every cell contains the \emph{STD} value before the tuning, plus/minus a ratio of the energy variation after the tuning.
-We notice that the energy variation varies less than 10\,\% after the extra-tuning.
-We argue that this variation is not substantial, as it is not stable from a node to another.
-Moreover, 10\,\% of variation is not a representative difference, due to many factors that can affect it as the CPU temperature or the measurement errors.
-
-\begin{table}
-    \centering
-    \caption{STD (mJ) comparison before/after tuning the OS}
-    \label{table:osmin}
-    \small
-    \begin{tabular}{|c|c|c|c|}
-        \hline
-        \textbf{Node} & \textbf{\sf EP} & \textbf{\sf CG} & \textbf{\sf LU} \\
-        \hline
-        \hline
-        \textbf{N1}   & 1370~-9\,\%     & 78~+7\,\%       & 128~+2\,\%      \\
-        \hline
-        \textbf{N2}   & 1278~-7\,\%     & 64~-1\,\%       & 120~+9\,\%      \\
-        \hline
-        \textbf{N3}   & 1118~+1\,\%     & 83~+2\,\%       & 93~+7\,\%       \\
-        \hline
-    \end{tabular}
-\end{table}
-% \subsubsection{Meltdown Security Patch}
-
-\subsubsection{Speculative Executions}
-Meltdown and Spectre are two of the most famous hardware vulnerabilities discovered in 2018, and exploiting them allows a malicious process to access others processes data that is supposed to be private~\cite{Kocher2018spectre,Lipp2018meltdown}.
-They both exploit the speculative execution technique where a process anticipates some upcoming tasks, which are not guaranteed to be executed, when extra resources are available, and revert those changes if not.
-Some OS-level patches had been applied to prevent/reduce the criticality of these vulnerabilities.
-On the Linux kernel, the patch has been automatically applied since the version 4.14.12.
-It mitigates the risk by isolating the kernel and the user space and preventing the mapping of most of the kernel memory in the user space.
-% These patches have been reported as decreasing the performance of a system. 
-Nikolay~\emph{et~al.} have studied in~\cite{DBLP:journals/corr/abs-1801-04329} the impact of patching the OS on the performance.
-The results showed that the overall performance decrease is around 2--3\,\% for most of the benchmarks and real-world applications, only some specific functions can meet a high performance decrease.
-In our study, we are interested in the applied patch's impact on the energy variation, as the performance decrease could mean an energy consumption increase.
-Thus, we ran the same benchmarks \textsf{LU}, \textsf{CG} ad \textsf{EP} on the cluster \textsf{Dahu} with different workloads, using the same OS, with and without the security patch.
-Table~\ref{table:meltdown} reports on the STD values before disabling the security patch.
-A minus means that the energy varies less without the patch being applied, while a plus means that it varies more.
-These results help us to conclude that the security patch's effect on the energy variation is not substantial and can be absorbed through the error margin for the tested benchmarks.
-In fact, the best case to consider is the benchmark \textsf{LU} where the energy variation is less than 10\,\% when we disable the security patch, but this difference is still moderate.
-The little performance difference discussed in~\cite{Kocher2018spectre,Lipp2018meltdown} may only be responsible of a small variation, which will be absorbed through the measurement tools and external noise error margin in most cases.
-
-\begin{table}[h!]
-    \centering
-    \caption{STD (mJ) comparison with/without the security patch}
-    \label{table:meltdown}
-    \small
-    \begin{tabular}{|l|c|c|c|}
-        \hline
-        \textbf{Node} & \textbf{\sf EP} & \textbf{\sf CG} & \textbf{\sf LU} \\
-        \hline
-        \hline
-        \textbf{N1}   & 269~+2\,\%      & 83~+1\,\%       & 108~-6\,\%      \\
-        \hline
-        \textbf{N2}   & 195~+1\,\%      & 84~-5\,\%       & 121~-9\,\%      \\
-        \hline
-        \textbf{N3}   & 223~+/-1\,\%    & 72~-4\,\%       & 117~+8\,\%      \\
-        \hline
-        \textbf{N4}   & 276~+3\,\%      & 60~+0\,\%       & 113~-3\,\%      \\
-        \hline
-    \end{tabular}
-\end{table}
-
-\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
-    To answer \textsc{RQ~3}, we conclude that the OS \textbf{should not be the main focus} of the energy variation taming efforts.
-\end{mdframed}
-
-\subsection{\textsc{RQ}~4: Processor Generation}
-Intel microprocessors have noticeably evolved during these last 20 years.
-Most of the new CPU come with new enhancements to the chip density, the maximum Frequency or some optimization features like the C-states or the Turbo~Boost.
-This active evolution caused that different generations of CPU can handle a task differently.
-The aim of this expriment is not to justify the evolution of the variation across CPU versions/generations, but to observe if the user can choose the best node to execute her experiments.
-Previous papers have discussed the evolution of the energy consumption variation across CPU generations and concluded that the variation is getting higher with the latest CPU generations~\cite{wang_experimental_nodate,marathe_empirical_2017_m}, which makes measurements stability even worse.
-In this experiment, we therefore compare four different generations of CPU with the aim to evaluate the energy variation for each CPU and its correlation with the generation.
-Table~\ref{table:cpus} indicates the characteristics of each of the tested CPU.
-
-\begin{table}
-    \centering
-    \caption{STD (mJ) comparison of experiments from 4 clusters}
-    \small
-    \begin{tabular}{|l|c|c|c|c|}
-        \hline
-        \textbf{Cluster} & \textbf{Dahu} & \textbf{Chetemi} & \textbf{Ecotype} & \textbf{Paranoia} \\
-        \hline
-        \hline
-        Arch             & Skylake       & Broadwell        & Broadwell        & Ivy Bridge        \\
-        \hline
-        Freq             & 3.7\,GHz      & 3.1\,GHz         & 2.9\,GHz         & 3.0\,GHz          \\
-        \hline
-        TDP              & 125 W         & 85 W             & 55 W             & 95 W              \\
-        \hline
-        \hline
-        5\%              & 364           & 210              & \textbf{75}      & \textbf{76}       \\
-        \hline
-        50\%             & 98            & 86               & \textbf{49}      & 244               \\
-        \hline
-        100\%            & 119           & 116              & \textbf{106}     & 240               \\
-        \hline
-    \end{tabular}
-    \label{table:cpus}
-\end{table}
-
-Table~\ref{table:cpus} also shows the aggregated energy variation of the different generations of nodes for the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
-The results attest that the latest versions of CPU do not necessarily cause more variation.
-In the experiments we ran, the nodes from the cluster \textsf{Paranoia} tend to cause more variation at high workloads, even if they are from the latest generation.
-While the Skylake CPU of the cluster \textsf{Dahu} cause often more energy variation than \textsf{Chetemi} and the Ecotype Broadwell CPU.
-We argue that the hypothesis "\emph{the energy consumption on newer CPU varies more}" could be true or not depending on the compared generations, but most importantly, the chips energy behaviors.
-On the other hand, our experiments showed the lowest energy variation when using the \textsf{Ecotype CPU}, these CPU are not the oldest nor the latest, but are tagged with "\texttt{L}" for their low power/TDP.
-This result rises another hypothesis when considering CPU choice, which implies selecting the CPU with a low TDP.
-This hypothesis has been confirmed on all the \textsf{Ecotype} cluster nodes, especially at low and medium workloads.
-
-Figure~\ref{fig:cpugen} is an illustration of the aggregated STD density of more than $5,000$-random values sets taken from all the conducted experiments.
-This shows that the cluster \textsf{Paranoia} reports the worst variation in most cases, and that \textsf{Ecotype} is the best cluster to consider to get the least variations, as it has a higher density for small variation values.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/cpugen}}
-    \caption{Energy consumption STD density of the 4 clusters}\label{fig:cpugen}
-\end{figure}
-
-\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
-    We conclude on \textbf{affirming \textsc{RQ}~4}, as selecting the right CPU can help to get less variations.
-\end{mdframed}
+%  TODO: add the environement part,
+%  TODO: add perspective
 
@@
 Most importantly, this should motivate future works on creating a better knowledge on the variability due to CPU manufacturing process and other factors.
+
+\section{Perspectives}
+By the end of this study we have gathered enought guidelines to make the tests more reproducible, accurate.
+We created a set of new tests named \textbf{energy tests} which are more similar to performance tests.
+Thanks to the work of two interns [ mamadou and adrien] we created a CI/CD plateform to measure the energy consumption of Java projects and we could track the evolution of the this energy accross different stages of the project.
+In the figure below we see an example of this plugin.
+For more details please visit the gitlab repository ... add link.
+
+\begin{figure}%[!htb]
+    \center{\includegraphics[width=.9\linewidth]{imgs/JunitSonarplugin}}
+    \caption{Example of the Junit Sonar Plugin}\label{fig:sonar_plugin}
+\end{figure}
+
+% However for them to be more representatives we wanted to simulate the production enviromenet.
+%% should i add  the thing about green faas ? 
+%% same about what we have done with python workers and lazy ? basically just a graph

### 2021-10-21 chapters/testing/environement.tex
draft. first chapter

--- 2021-09-03 chapters/testing/reproducibility.tex
@@
-\section{Empirical Evaluation Protocol}
-
-This thesis is based on the empirical analyses to measure the energy consumption.
-As we mentioned earlier, we have chosen the \emph{agile} methode.
-so basically we will reduce the energy consumption with trial and error.
-
-%%%%%%%%% preliminary taughts and section structure 
-% The need of reproducibility in our field - software optimization based on empirical studies -
-
-% The importance of Virtualisation for reproducibility \cite{howe_virtual_2012}
-% some of the most important parts are
-% - fewer constraints on research methods
-% - on-demand backups
-% - virtual Machines as Publications
-% - more Variables captured
-\subsection{Threats and Challenges}
-
-One of the most challenges which face the reaserchers is the reproducibility of their tests.Actually many results are not reproducible \footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.}, which let to a \emph{replication crisis}.When this crisis hit most of the empirical studies,most of the reviews now includes reproducibility as one of the minimal standard for judging scientific \cite{peng2011reproducible}.
-one of the creterions for a reproducidibility is the publication of the dataset and the algorithm run on the raw data to condlude the results.
-There is even some disagreement about what the terms "reproducibility" or "replicability" by themselves mean \cite{goodman2016does} .According to \cite{echtler2018open} \emph{replicability} extends \emph{reproducibility} to the ability to collect a new raw dataset comparable to the original one by reexecuting the experiment under similar condition, Instead of just the ability to get the same results by ruruning the statistical analyses on the orginal data set.
-
-One major problem that hinders the reproducibility of the empirical tests is the interaction with the external environment, either as concurency or dependencies.
-Therefore testers won't have the same results unless they duplicate the same environment.
 % introduction of vms 
@@
     \center{\includegraphics[width=1\linewidth]{imgs/virtualization_techniques}}
-    \caption{Different Methods of Virtualzation}\label{reproducibility:virtualization_technique}
+    \caption{Different Methods of Virtualzation}\label{environement:virtualization_technique}
 \end{figure}
@@
 
-Figure \ref{reproducibility:virtualization_technique} explains the differents in architecture between the clasic types2 of Virtualisation and Containers.
+Figure \ref{environement:virtualization_technique} explains the differents in architecture between the clasic types2 of Virtualisation and Containers.
 \begin{itemize}
@@
 \end{itemize}
-\note{Rephrase so i include it before VMs+ add links to the technologies }
+%  TODO Rephrase so i include it before VMs+ add links to the technologie
 
@@
 Depsite that Type 1 is more performant than type 2, the second one is the most used in reaserch, since most researchers conducts their experements in their own machine.
-In the other hand, docker is the most famous thechonology for for containers,
-
+In the other hand, docker is the most famous thechonology for for containers.
+In ower case we are more prone to docker for two reasons.
+\begin{enumerate}
+    \item we need a litetweight orchestrator to not affect the energy consumption of ower tests. As prior work mentioned [cite Morabito (2015) and van Kessel et al. (2016)]
+          % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
+    \item since we are using the hardware itself to measure the energy consumption, we are required to interract with the host OS itself.
+\end{enumerate}
+Special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that allows us to retireve the energy consumption of a virtual machine.
 %%%%% THINGS TO ADD : 
+%% -------------- Dont think so 
 % the impact of docker on the reprodudicibility 
@@
 % The docker files and the clarification of the methodology 
+\subsection{Docker and energy}
+Now that we have chosen to go with the containers technorolgy to encapsulate our tests. What would be the impact of this solution on the energy consumption of our tests.
 
-\subsection{Docker and energy}
+Based on the studies of \cite{eddie_antonio_santos_how_nodate}, who analysed the impact of adding the docker layer on the energy consumption.
+In their experement. Eddie Antonio et al run multiple benchmarks with and without docker. and compared their energy consumption and execution time.
+The first step was to see the impact of docker deamon while there is no work. to see the impact of the orchestrator alone. Later they had the experiment with the following benchmarks
+\begin{itemize}
+    \item wordpress
+    \item reddis
+    \item postfresSQL
+\end{itemize}
+The following figures represents the energy consumption of the system while it is idle. As we can see  in figure \ref{fig:docker_idle} Docker braught arround 1000joules overhead.
+\begin{figure}
+    \center{\includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/idle_energy}}
+    \caption{energy consumption of Idle system with and without docker \cite{eddie_antonio_santos_how_nodate}}\label{fig:docker_idle}
+\end{figure}
+In the other hand as we can see in figure \ref{fig:docker_reddis}.
+docker increased the execution time of the benchmark by 50 seconds which caused an increase in energy since they are highly correlated.
+The authors also highlated the fact that this increase of energy consumption is due to the docker deamon and not the fact that the application is in a container.
+Moreover they estimted the price of this extra energy and it was less than 0.15\$ in the worst case.
+Which is non significant compared to the advantages that docker bring for isolation and reproducibility.\\
+if we  recap this study in one sentence,it would be the following one.
+The dockerised softwares tend to consume more energy, because mainly they take more time to be executed.
+The average power consumption is higher with only \textbf{2Watts} and it is due to the docker deamon.This overhead can be up to 5\% for IO intensive application, but it is mearly noticiable when it comes to CPU or DRMA intesive works
+
+\begin{figure}
+    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_time}
+    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_energy}
+    \caption{execution time and energy consumption of Redis  with and without docker \cite{eddie_antonio_santos_how_nodate}}\label{fig:docker_reddis}
+\end{figure}
+
 %  basically just rephrase the paper
@@
 \subsection{docker and accuracy}
+%% TODO : add an introduction, maybe some more details 
+And now Since the stat of the art has aggreed on the impact of docker on the energy consumption,Let's discuss it's impact on the accuracy. In other words\\
+\textbf{RQ :} does Docker affect the energy variation of the exepements ?
 
-\subsection{Testing Protocol}
-the results of this protocil can be found in the project of the continious integrtation that was made by the intern Mamadou to give a feedback about the evolution of the energyconsumption within time
+To Answer this question we have conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
+Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
+In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
+The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
+
+\begin{figure}
+    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
+    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
+\end{figure}

### 2021-10-21 chapters/testing/experement.tex
draft. first chapter

--- 2021-08-30 chapters/testing/accuracy.tex
@@
-This is the chapter where i will put all the works related to the a successful testing
+% To satisfy those three requirements. we elaborated a benchmarking protocol that aims to help reaserchers and practicioners in their experementations.
 
-% \begin{abstract}
-%     The past decade witnessed the inclusion of power measurements to evaluate the energy efficiency of software systems, thus making energy a prime indicator along with performance.
-%     Nevertheless, measuring the energy consumption of a software system remains a tedious task for practitioners.
-%     In particular, the energy measurement process may be subject to a lot of variations that hinder the relevance of potential comparisons.
-%     While the state of the art mostly acknowledged the impact of hardware factors (chip printing process, CPU temperature), this paper investigates the impact of controllable factors on these variations.
-%     More specifically, we conduct an empirical study of multiple controllable parameters that one can easily tune to tame the energy consumption variations when benchmarking software systems.
+% for each aspect we present a set of guidelines and tools.
+%% This is a conclusion maaan  !! 
+\subsection{Goal}
+In this part we will try to answer the following Research questions.
 
-%     To better understand the causes of such variations, we ran more than a $1,000$ experiments on more than $100$ nodes with different workloads and configurations.
-%     The main factors we studied encompass: experimental protocol, CPU features (C-states, Turbo~Boost, core pinning) and generations, as well as the operating system.
-%     Our experiments showed that, for some workloads, it is possible to tighten the energy variation by up to $30\times$.
-%     Finally, we summarize our results as guidelines to tame energy consumption variations.
-%     We argue that the guidelines we deliver are the minimal requirements to be considered prior to any energy efficiency evaluation.
-% \end{abstract}
-
-\section{Introduction}
-%\lettrine[nindent=0em,lines=3]{C} 
-To conduct robust evaluations, practitioners often try to ensure reproducible environmental conditions in order to properly benchmark their software systems.
-In this area, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
-Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.
-
-In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should favour reproducible experiments.
-However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only on homogeneous nodes, but even within a single node.
-This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
-
-Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
-One can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
-    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
-\end{figure}
-
-Most of the state of the art has been investigating this power consumption issue from a hardware perspective~\cite{borkar_designing_2005,tschanz_adaptive_2002} and reported that the causes of such energy variations are CMOS manufacturing process of transistors in a chip, differences in node assembly and data center hot spots.
-Additionally, \cite{heinrich_predicting} described it as a combination of parameters, mentioning a list of candidate factors, such as the thermal effect or the CPU frequency, but failing to deliver a deeper analysis of these factors.
-
-Unfortunately, hardware factors can be hardly tuned to tame the energy variation that can be observed in systems benchmarking.
-For example, managing the CPU temperature or a server position in a cluster, are not actions that one can easily do, especially on the modern data centers and cloud platforms.
-Therefore, the goal of this paper is to investigate the spectrum of factors that can cause or increase the variability of energy consumption in systems benchmarking, and to propose effective guidelines to control such factors in order to mitigate this variability.
-While this paper does not question the benefits of established CPU features, like C-states or Turbo~Boost, it delivers a deeper analysis of the effects they may introduce on a wide set of experiments and nodes.
-By quantifying potential energy variations induced by controllable factors, we intend to identify the proper configurations that minimize energy variations, depending on the workload characteristics.
-These guidelines aim at supporting practitioners in conducting robust system benchmarks and reporting reproducible energy consumption, rather than recommending production-scale practices or solutions to reduce the power consumption of a given system.
-% 
-% The key contributions of this paper can therefore be summarized as:
-% \begin{enumerate}
-%     \item Providing a better understanding of the energy variation, by using different generations of CPU deployed in $4$ clusters with more than $100$ physical nodes, and by considering existing systems benchmarks with diverse workloads;
-%     \item Identifying controllable factors that contribute to the variation in CPU energy consumption, comparing them against the state of the art, and completing them with other uncovered assumptions;
-%     \item Reporting on some guidelines on how to conduct reproducible experiments with less energy variations;
-%     \item Discussing the differences between inter-nodes and intra-nodes variations.
-% \end{enumerate}
-
-%\paragraph{}
-% The remainder of this paper is organized as follows.
-% Section~\ref{sec:related_work} discusses the related works in the area of power measurements.
-% Section~\ref{sec:research_qs} formalizes our research questions.
-% Section~\ref{sec:setup} reports on the experimental setup (hardware, benchmarks, tools and methodology) we used in this work.
-% Section~\ref{sec:parameters} analyzes the causes of the variations we observed along experiments.
-% Finally, we discuss the results of the different experimented factors, and their impact on the energy variation in Section~\ref{sec:guidelines}, while Sections~\ref{sec:threats} \&~\ref{sec:conclusion} cover the validity threats and our conclusions, respectively.
-
-%------------------------------------------------
-\section{Related Work}\label{sec:related_work}
-In this section, we review the state of the art in the domain of energy consumption analysis.
-In particular, one can observe that several studies considered the evaluation of the processor variation while executing the same workload, in particular to test distributed applications running in homogeneous clusters.
-
-\paragraph{Studying Hardware Factors.}
-
-This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
-On the one hand, the correlation between the processor temperature and the energy consumption was one of the most explored paths.
-Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
-On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the correlation between the CPU temperature and the energy consumption variation is very tight.
-
-\note{add the corelation value}
-
-This makes the processor temperature a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.
-
-The ambient temperature was also discussed in many papers as a candidate factor for the energy variation of a processor.
-In~\cite{ranka_energy_2009}, the authors claimed that energy consumption may vary due to fluctuations caused by the external environment.
-These fluctuations may alter the processor temperature and its energy consumption.
-However, the temperature inside a data center does not show major variations from one node to another.
-In~\cite{el_mehdi_diouri_your_2013}, El~Mehdi~Dirouri~\emph{et~al.} showed that switching the spot of two servers does not affect their energy consumption.
-Moreover, changing hardware components, such as the hard drive, the memory or even the power supply, does not affect the energy variation of a node, making it mainly related to the processor.
-This result was recently assessed by~\cite{wang_potential_2018}, where the rack placement and power supply introduced a maximum of $2.8\,\%$ variation in the observed energy consumption.
-
-Beyond hardware components, the accuracy of power meters has also been questioned.
-Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
-All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
-
-\paragraph{Mitigating Energy Variations.}
-Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
-In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
-The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
-Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
-They also proposed some guidelines to reduce this variation by replacing the old slow chips, by load balancing the workload on the CPU cores and leaving one core idle.
-They claimed that the variation between the processor cores is insignificant.
-In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.
-
-In~\cite{marathe_empirical_2017_m}, the authors highlight the increase of energy variation across the latest Intel micro-architectures by a factor of $4$ from Sandy~Bridge to Broadwell, a $15\,\%$ of run-to-run variation within the same processor and the increase of the inter-cores variation from $2.5\,\%$ to $5\,\%$ due to hardware-enforced constraints, concluding with some recommendations for Broadwell usage, such as running one hyper-thread per core.
-\note{this might be interesting for greeFaaS}
-
-\paragraph{Objective.}
-The aim of our paper is not to reproduce all the exposed results and explored paths.
-Instead, we intend to check some of these results on latest hardware components, and to consider new potential controllable factors that could also contribute to the energy variation.
-We also aim to provide practitioners with guidelines to conduct robust experiments in a controlled environment, thus reducing the variation overhead to share trustable results and build credible comparisons.
-
-\section{Research Questions}\label{sec:research_qs}
-Worth saying, we are aware that part of the energy consumption variation is due to chip manufacturing differences or some of the previously discussed enforced factors, such as the thermal effect or the servers placement.
-Those parameters are often tricky to manage, as we cannot have a perfect chips manufacturing process, or assume that two identical processors have the same thermal behavior.
-We will therefore focus on providing the practitioners with an empirical study of some controllable parameters that can be tuned to conduct experimental evaluations of energy consumption with less variation, especially if the practitioners do not have a physical access to the data center or BIOS configuration, which is the case on most of the modern cloud platforms and data centers.
-These parameters span the choice of the benchmarking protocol, processor frequencies management, operating system tuning or some other parameters.
-The authors of~\cite{heinrich_predicting} mentioned some of these potential parameters.
-
-In this paper, we will therefore investigate the following controllable factors, which we formalize as 4 research questions:
 \begin{description}
@@
 
-\section{Experimental Setup}\label{sec:setup}
-This section describes our detailed experimental environment, covering the clusters and nodes configuration, the benchmarks we used and justifying our experimental methodology.
+\subsection{Experimental Setup}\label{subsec:setup}
+This section describes our detailed experimental environment, covering the clusters configurationand the benchmarks we used  justifying our experimental methodology.
 
-\subsection{Hardware Platform}
-We considered 4 distinct clusters of variable sizes and different generations of CPU, as summarized in Table~\ref{table:g5k}.
-In particular, we used the Grid5000 (G5K) platform in our experiments~\cite{grid5000,margery:hal-00965708}.
-G5K is a bare metal cloud platform that can be used to provision clusters of identical nodes.
-In our study, we mainly used the cluster \textsf{Dahu} located in Grenoble to run most of our tests, as it has one of the newest Xeon CPUs.
-We also used the clusters \textsf{Chetemi}, \textsf{Ecotype} and \textsf{Paranoia} in some of our experiments.
-Table~\ref{table:g5k} describes the configurations of the clusters we considered.
-% We note that each cluster is geographically independent.
+\subsubsection{Measurement Context}
+There are three main contexts.
+\begin{itemize}
+    \item Different machines with different configuration
+    \item Different machines with the same configuration
+    \item Same machine
+\end{itemize}
 
-\begin{table}
+To satisfy those requirements we have used the plateform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708},a large-scale and flexible testbed for experiment-driven research distributed accross all france.
+Grid5000 offers  miultiple clusters composed with 4 upto 124 identical machine with different configurations for each cluster.
+For our experement we have considered 4 clusters. Our main creterions was the CPU Configuration.
+the table below~\ref{table:g5k}, presents a description of the 4 clusters
+
+\begin{table}[!hbt]
     \centering
@@
 
-As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or socket to designate one of the two sockets and \textsf{PU} for the operating system \emph{processing unit}.
-The number of \textsf{PU} often doubles the number of available cores because of the hyper-threading support, as the OS considers 2 hyper-threads sharing the same core as 2 different \textsf{PU}.
-Figure~\ref{fig:topo} illustrates a detailed topology of a node belonging to the cluster \textsf{Dahu}.
+As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or \textsf{socket} to designate one of the two sockets and \textsf{PU} for the \emph{processing unit}.
+For our study we coonsider hyper-threads as distincts \textbf{PU}
+
+As an example, Figure~\ref{fig:topo} illustrates a detailed topology of a node belonging to the cluster \textsf{Dahu}.
 
@@
 \end{figure}
-% The advantage of using G5K is having an access to a wide set of physical nodes with different capabilities. 
-% The usage procedure is to book a machine for period of time with the chosen OS in less than 5 minutes, ship and run the test script, then release the machine.
-When a machine is released everything is discarded even the OS.
 
-\subsection{Systems Benchmarks}
-Our first criterion to choose the systems benchmarks was the scalability, as we need to run tests at different workloads by choosing the right number of used PU.
-The other criteria are the documentation, the accuracy and the references to the benchmark.
-\emph{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925} is one of the most used suite of benchmarks in the HPC literature and it fulfills our benchmarking requirements.
-We mainly used the pseudo application \emph{Lower-Upper symmetric Gauss-Seidel} (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
-These are the main benchmarks used in many similar works, such as~\cite{heinrich_predicting}.
-Nonetheless, in order to validate our results on a wider set of benchmarks and applications, we also used \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}\footnote{\url{https://linux.die.net/man/1/sha256sum}} as representative systems benchmarks to conduct our experiments with a broad diversity of workloads.
+\subsection{Workload}
+Our choice for the benchmarks was based on two creterions.
 
-\subsection{Measurement Tools \& Methodology}
-To study the energy consumption of nodes, we considered Intel \emph{Running Average Power Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754}, which is one of the most accurate tools to report the CPU/DRAM global energy consumption.
-We also used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, which is a power monitoring toolkit that builds a model over RAPL to compute the energy consumption at process-level when we needed to isolate energy consumption of a single process.
-Our clusters are provisioned with a minimal version of Debian\,9 (4.9.0 kernel version) where we install Docker (version 18.09.5), which will be used to run the RAPL sensor and the benchmark itself.
+First, \textbf{scalability} : We wanted to gather the highest possible amount insights (regarding time spent for the experiment, therefore we wanted some benchmarks who can scale according to the number of PUs and can fulfill different scenarios.
+Second creterion is \textbf{representativeness} : as menstionned in the challenges, a workload has to be representative otherwise the experiment would be inconsistent \cite{evaluate_Collaboratory}.
+To satisfy those creterions we went through state of the art and looked for the most used benchmarks for testing the performance of the hardware, and then we selected the scallable ones.
+Our candidate is  emph{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925}: one of the most used benchmarks for \emph{HPC}.
+We used the applications (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
+Further more we have used other applications to validate our results using more applications such as \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}\footnote{\url{https://linux.die.net/man/1/sha256sum}}.
+
+\subsection{Metrics \& Measurement Tools}
+Our metric for the accuracy of the test is the Standard deviation aka \textbf{STD} of the energy consumption. Therefore wether the tests consumes more or less energy is out of our scope.
+To study this variation we need first a tool to measure the energy consumption.
+For this we used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, which is a power monitoring toolkit that is based on \emph{Intel Running Average Power Limit} (RAPL) \cite{Khan:2018:RAE:3199681.3177754}.
+The advantage of PowerAPI is that it reports the Energy consumption of CPU and DRAM at a socket level.
+
+Our testbeds are run with a minimal version of Debian \,9 (4.9.0 kernel version)\footnote{\url{https://github.com/grid5000/environments-recipes/blob/master/debian9-x64-min.yaml}} where we install Docker (version 18.09.5), which will be used to run the RAPL sensor and the benchmark itself.
 The energy sensor collects RAPL reports and stores them in a remote \textsc{MongoDB} instance, allowing us to perform \emph{post-mortem} analysis in a dedicated environment.
@@
 This allows \textsc{PowerAPI} to measures all the running containers, even the RAPL sensor consumption, as it is isolated in a container.
-One potential threat covers the impact of Docker on the energy variation.
-We therefore conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
-Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
-In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
-The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
-    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
-\end{figure}
 
@@
 We believe this can help to achieve better and more reliable comparisons.
-
 We mainly consider 3 different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium and high CPU usage when analyzing the studied parameters effect, respectively.
@@
 
-\section{Energy Variation Analysis}\label{sec:parameters}
-In this section, we aim to establish experimental guidelines to reduce the CPU energy variation.
+\subsection{Analysis}\label{subsec:parameters}
+In this part, we aim to establish experimental guidelines to reduce the CPU energy variation.
 We therefore explore many potential factors and parameters that could have a considerable effect on the energy variation.
+
+%% TODO: Proof read this paragraph
 
@@
 \end{mdframed}
-
-\section{Experimental Guidelines}\label{sec:guidelines}
-To summarize our experiments, we provide some experimental guidelines in Table~\ref{table:guidelines}, based on the multiple experiments and analysis we did.
-These guidelines constitute a set of minimal requirements or best practices, depending on the workload and the criticality of the energy measurement precision.
-It therefore intends to help practitioners in taming the energy variation on the selected CPU, and conduct the experiments with the least variations.
-
-\begin{table}[h!]
-    \centering
-    \caption{Experimental Guidelines for Energy Variations}
-    \small
-    \begin{tabular}{|p{4.7cm}|c|c|}
-        \hline
-        \textbf{Guideline}                                                                                                                                                & \textbf{Load} & \textbf{Gain}     \\
-        \hline
-        \hline
-        Use a low TDP CPU                                                                                                                                                 & Low \& medium & Up to $3\times$   \\
-        \hline
-        Disable the CPU C-states                                                                                                                                          & Low           & Up to $6\times$   \\
-        \hline
-        Use the least of sockets in a case of multiple CPU                                                                                                                & Medium        & Up to $30\times$  \\
-        \hline
-        Avoid the usage of hyper-threading whenever possible                                                                                                              & Medium        & Up to $5\times$   \\
-        \hline
-        Avoid rebooting the machine between tests                                                                                                                         & High          & Up to $1.5\times$ \\
-        \hline
-        Do not relate to the machine idle variation to isolate a test EC, the CPU/OS changes its behavior when a test is running and can exhibit less variation than idle & Any           & ---               \\
-        \hline
-        Rather focus the optimization efforts on the system under test than the OS                                                                                        & Any           & ---               \\
-        \hline
-        Execute all the similar and comparable experiments on a same machine.
-Identical machines can exhibit many differences regarding their energy behavior             & Any           & Up to $1.3\times$ \\
-        \hline
-    \end{tabular}
-    \label{table:guidelines}
-\end{table}
-
-Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and its variation reduction at low workloads.
-However, it also lists some new factors that we identified along the analysis we conducted in Section~\ref{sec:parameters}, such as the results related to the OS or the reboot mode.
-Some of the guidelines are more useful/efficient for specific workloads, as showed in our experiments.
-Thus, qualifying the workload before conducting the experiments can help in choosing the proper guidelines to apply.
-Other studied factors are not been mentioned in the guidelines, like the Turbo~Boost or the Speculative execution, due to the small effect that has been observed in our study.
-
-In order to validate the accuracy of our guidelines among a varied set of benchmarks on one hand, and their effect on the variation between identical machines on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
-Half of these experiments has been performed at a 50\,\% workload and the other half on single process jobs.
-The choice of these two workloads is related to the optimization guidelines that are mainly effective at low and medium workloads.
-We note that we used the cluster \textsf{Dahu} over \textsf{Ecotype} to highlight the guidelines effect on the nodes where the variation is susceptible to be higher.
-
-\begin{figure*}%[!htb]
-    \center{\includegraphics[width=\linewidth]{imgs/all}}
-    \caption{Energy variation comparison with/without applying our guidelines}\label{fig:optimized}
-\end{figure*}
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/stressng}}
-    \caption{Energy variation comparison with/without applying our guidelines for \textsc{Stress-NG}}\label{fig:stress}
-\end{figure}
-
-Figure~\ref{fig:optimized} and~\ref{fig:stress} highlight the improvement brought by the adoption of our guidelines.
-They demonstrate the intra-node STD reduction at low and medium workloads for all the benchmarks used at different levels.
-Concretely, for low workloads, the energy variation is 2--6 times lower after applying the optimization guidelines for the benchmarks \textsf{LU} and \textsf{EP}, as well as \textsc{Linpack}, while it is 1.2--1.8 times better for \texttt{Sha256}.
-For this workload, the overall energy consumption after optimization can be up to 80\,\% higher due to disabling the C-states to keep all the unused cores at a high power consumption state ($Mean_{LU-normal-Dahu2}=11,500 mJ$, $Mean_{LU-optimized-Dahu2}=20,508 mJ$).
-For medium workloads, the STD, and thus variation, is up to 100\,\% better for the benchmark \textsf{CG}, 20--150\,\% better for the \texttt{pbzip2} application and up to 100\% for \textsc{Stress-NG}.
-We note that the optimized version consumes less energy thanks to an appropriate core pinning method.
-
-Figures~\ref{fig:optimized} and~\ref{fig:stress} also highlight that applying the guidelines does not reduce the inter-nodes variation in all the cases.
-This variation can be up to 30\,\% in modern CPU~\cite{wang_experimental_nodate}.
-However, taming the intra-node variation is a good strategy to identify more relevant mediums and medians, and then perform accurate comparisons between the nodes variation.
-Even though, using the same node is always better, to avoid the extra inter-nodes variation and thus improve the stability of measurements.
-
-\section{Threats to Validity}\label{sec:threats}
-A number of issues affect the validity of our work.
-For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPU, but still adds an important overhead if we adopt a sampling at high frequency.
-The other fine-grained tool we used for measurements is \textsc{PowerAPI}.
-It allows to measure the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
-The usage of \textsc{PowerAPI} adds an error margin because of the power model built over RAPL.
-The RAPL tool mainly measures the CPU and DRAM energy consumption.
-However, even running CPU/RAM intensive benchmarks would keep a degree on uncertainty concerning the hard disk and networking energy consumption.
-In addition, the operating system adds a layer of confusion and uncertainty.
-
-The Intel CPU chip manufacturing process and the materials micro-heterogeneity is one of the biggest issues, as we cannot track or justify some of the energy variation between identical CPU or cores.
-These CPU/cores might handle frequencies and temperature differently and behave consequently.
-This hardware heterogeneity also makes reproduction complex and requires the usage of the same nodes on the cluster with the same OS.
-
-% A more subtle issue may arise due to the values of the measurements that we achieved.
-% In fact, the energy measures are quite small, and iterations may be taking a few milliseconds more or less to run.
-% A thing we cannot measure using our measurement tools.
-% How generalizable are our results? As a set of energy variation optimization guidelines, we argue that our results applied on most of the modern Intel CPU.
-% However, using and comparing identical CPU is still tricky and is very dependent to chips.
-
-\section{Conclusion}\label{sec:conclusion}
-In this paper, we conducted an empirical study of controllable factors that can increase the energy variations on platforms with some of the latest CPU, and for several workloads.
-We provide a set of guidelines that can be implemented and tuned (through the OS GRUB for example), especially with the new data centers isolation trend and the cloud usage, even for scientific and R\&D purposes.
-Our guidelines aim at helping the user in reducing the CPU energy variation during systems benchmarking, and conduct more stable experiments when the variation is critical.
-For example, when comparing the energy consumption of two versions of an algorithm or a software system, where the difference can be tight and need to be measured accurately.
-
-Overall, our results are not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
-The aim of our work is to be able to tame and mitigate this variability along controlled experiments.
-We studied some previously discussed aspects on some recent CPU, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to achieve the variability mitigating purpose.
-Some of these factors, like the C-states usage, can reduce the energy variation up to 500\,\% at low workloads, while choosing the wrong cores/PU strategy can cause up to $30\times$ more variability.
-
-We believe that our approach can also be used to study/discover other potential variability factors, and extend our results to alternative CPU generations/brands.
-Most importantly, this should motivate future works on creating a better knowledge on the variability due to CPU manufacturing process and other factors.

### 2021-10-21 chapters/testing/stateoftheart.tex
draft. first chapter

--- 2021-08-30 chapters/testing/accuracy.tex
@@
-This is the chapter where i will put all the works related to the a successful testing
+%  State of the art. 
+%  TODO :
 
-% \begin{abstract}
-%     The past decade witnessed the inclusion of power measurements to evaluate the energy efficiency of software systems, thus making energy a prime indicator along with performance.
-%     Nevertheless, measuring the energy consumption of a software system remains a tedious task for practitioners.
-%     In particular, the energy measurement process may be subject to a lot of variations that hinder the relevance of potential comparisons.
-%     While the state of the art mostly acknowledged the impact of hardware factors (chip printing process, CPU temperature), this paper investigates the impact of controllable factors on these variations.
-%     More specifically, we conduct an empirical study of multiple controllable parameters that one can easily tune to tame the energy consumption variations when benchmarking software systems.
-
-%     To better understand the causes of such variations, we ran more than a $1,000$ experiments on more than $100$ nodes with different workloads and configurations.
-%     The main factors we studied encompass: experimental protocol, CPU features (C-states, Turbo~Boost, core pinning) and generations, as well as the operating system.
-%     Our experiments showed that, for some workloads, it is possible to tighten the energy variation by up to $30\times$.
-%     Finally, we summarize our results as guidelines to tame energy consumption variations.
-%     We argue that the guidelines we deliver are the minimal requirements to be considered prior to any energy efficiency evaluation.
-% \end{abstract}
-
-\section{Introduction}
-%\lettrine[nindent=0em,lines=3]{C} 
-To conduct robust evaluations, practitioners often try to ensure reproducible environmental conditions in order to properly benchmark their software systems.
-In this area, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
-Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.
-
-In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should favour reproducible experiments.
-However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only on homogeneous nodes, but even within a single node.
-This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
-
-Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
-One can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
-    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
-\end{figure}
-
-Most of the state of the art has been investigating this power consumption issue from a hardware perspective~\cite{borkar_designing_2005,tschanz_adaptive_2002} and reported that the causes of such energy variations are CMOS manufacturing process of transistors in a chip, differences in node assembly and data center hot spots.
-Additionally, \cite{heinrich_predicting} described it as a combination of parameters, mentioning a list of candidate factors, such as the thermal effect or the CPU frequency, but failing to deliver a deeper analysis of these factors.
-
-Unfortunately, hardware factors can be hardly tuned to tame the energy variation that can be observed in systems benchmarking.
-For example, managing the CPU temperature or a server position in a cluster, are not actions that one can easily do, especially on the modern data centers and cloud platforms.
-Therefore, the goal of this paper is to investigate the spectrum of factors that can cause or increase the variability of energy consumption in systems benchmarking, and to propose effective guidelines to control such factors in order to mitigate this variability.
-While this paper does not question the benefits of established CPU features, like C-states or Turbo~Boost, it delivers a deeper analysis of the effects they may introduce on a wide set of experiments and nodes.
-By quantifying potential energy variations induced by controllable factors, we intend to identify the proper configurations that minimize energy variations, depending on the workload characteristics.
-These guidelines aim at supporting practitioners in conducting robust system benchmarks and reporting reproducible energy consumption, rather than recommending production-scale practices or solutions to reduce the power consumption of a given system.
-% 
-% The key contributions of this paper can therefore be summarized as:
-% \begin{enumerate}
-%     \item Providing a better understanding of the energy variation, by using different generations of CPU deployed in $4$ clusters with more than $100$ physical nodes, and by considering existing systems benchmarks with diverse workloads;
-%     \item Identifying controllable factors that contribute to the variation in CPU energy consumption, comparing them against the state of the art, and completing them with other uncovered assumptions;
-%     \item Reporting on some guidelines on how to conduct reproducible experiments with less energy variations;
-%     \item Discussing the differences between inter-nodes and intra-nodes variations.
-% \end{enumerate}
-
-%\paragraph{}
-% The remainder of this paper is organized as follows.
-% Section~\ref{sec:related_work} discusses the related works in the area of power measurements.
-% Section~\ref{sec:research_qs} formalizes our research questions.
-% Section~\ref{sec:setup} reports on the experimental setup (hardware, benchmarks, tools and methodology) we used in this work.
-% Section~\ref{sec:parameters} analyzes the causes of the variations we observed along experiments.
-% Finally, we discuss the results of the different experimented factors, and their impact on the energy variation in Section~\ref{sec:guidelines}, while Sections~\ref{sec:threats} \&~\ref{sec:conclusion} cover the validity threats and our conclusions, respectively.
-
-%------------------------------------------------
-\section{Related Work}\label{sec:related_work}
-In this section, we review the state of the art in the domain of energy consumption analysis.
-In particular, one can observe that several studies considered the evaluation of the processor variation while executing the same workload, in particular to test distributed applications running in homogeneous clusters.
+% - add : Evaluate Collaboratory article 
+% - add : the  hardware papers
 
@@
 All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
-
+%% this is to prove that the problem is REAL !!! 
 \paragraph{Mitigating Energy Variations.}
@@
 In~\cite{marathe_empirical_2017_m}, the authors highlight the increase of energy variation across the latest Intel micro-architectures by a factor of $4$ from Sandy~Bridge to Broadwell, a $15\,\%$ of run-to-run variation within the same processor and the increase of the inter-cores variation from $2.5\,\%$ to $5\,\%$ due to hardware-enforced constraints, concluding with some recommendations for Broadwell usage, such as running one hyper-thread per core.
-\note{this might be interesting for greeFaaS}
+%  NOTE :/this might be interesting for greeFaaS
 
-\paragraph{Objective.}
-The aim of our paper is not to reproduce all the exposed results and explored paths.
-Instead, we intend to check some of these results on latest hardware components, and to consider new potential controllable factors that could also contribute to the energy variation.
-We also aim to provide practitioners with guidelines to conduct robust experiments in a controlled environment, thus reducing the variation overhead to share trustable results and build credible comparisons.
+\subsection*{Our position}
+Meanwhile the previoius work was to create a large umbrella tht englobes all emprical tests related to computer science, we want to narrow our study to energy testing only.
+In other words we will study pitfalls that hinders the energy claimes of software .
 
-\section{Research Questions}\label{sec:research_qs}
-Worth saying, we are aware that part of the energy consumption variation is due to chip manufacturing differences or some of the previously discussed enforced factors, such as the thermal effect or the servers placement.
-Those parameters are often tricky to manage, as we cannot have a perfect chips manufacturing process, or assume that two identical processors have the same thermal behavior.
-We will therefore focus on providing the practitioners with an empirical study of some controllable parameters that can be tuned to conduct experimental evaluations of energy consumption with less variation, especially if the practitioners do not have a physical access to the data center or BIOS configuration, which is the case on most of the modern cloud platforms and data centers.
-These parameters span the choice of the benchmarking protocol, processor frequencies management, operating system tuning or some other parameters.
-The authors of~\cite{heinrich_predicting} mentioned some of these potential parameters.
-
-In this paper, we will therefore investigate the following controllable factors, which we formalize as 4 research questions:
-\begin{description}
-    \item[\textsc{RQ}~1:] Does the benchmarking protocol affect the energy variation?
-    \item[\textsc{RQ}~2:] How important is the impact of the processor features on the energy variation?
-    \item[\textsc{RQ}~3:] What is the impact of the operating system on the energy variation? and finally
-    \item[\textsc{RQ}~4:] Does the choice of the processor matter to mitigate the energy variation?
-\end{description}
-
-\section{Experimental Setup}\label{sec:setup}
-This section describes our detailed experimental environment, covering the clusters and nodes configuration, the benchmarks we used and justifying our experimental methodology.
-
-\subsection{Hardware Platform}
-We considered 4 distinct clusters of variable sizes and different generations of CPU, as summarized in Table~\ref{table:g5k}.
-In particular, we used the Grid5000 (G5K) platform in our experiments~\cite{grid5000,margery:hal-00965708}.
-G5K is a bare metal cloud platform that can be used to provision clusters of identical nodes.
-In our study, we mainly used the cluster \textsf{Dahu} located in Grenoble to run most of our tests, as it has one of the newest Xeon CPUs.
-We also used the clusters \textsf{Chetemi}, \textsf{Ecotype} and \textsf{Paranoia} in some of our experiments.
-Table~\ref{table:g5k} describes the configurations of the clusters we considered.
-% We note that each cluster is geographically independent.
-
-\begin{table}
-    \centering
-    \caption{Description of clusters included in the study}
-    \label{table:g5k}
-    \small
-    \begin{tabular}{|l|l|r|r|}
-        \hline
-        \textbf{Cluster}  & \textbf{Processor}               & \textbf{Nodes} & \textbf{RAM} \\
-        \hline
-        \hline
-        \textsf{Dahu}     & $2\times$ Intel Xeon\,Gold\,6130 & 32             & 192\,GiB     \\
-        \hline
-        \textsf{Chetemi}  & $2\times$ Intel Xeon\,E5-2630v4  & 15             & 768\,GiB     \\
-        \hline
-        \textsf{Ecotype}  & $2\times$ Intel Xeon\,E5-2630Lv4 & 48             & 128\,GiB     \\
-        \hline
-        \textsf{Paranoia} & $2\times$ Intel Xeon\,E5-2660v2  & 8              & 128\,GiB     \\
-        \hline
-    \end{tabular}
-\end{table}
-
-As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or socket to designate one of the two sockets and \textsf{PU} for the operating system \emph{processing unit}.
-The number of \textsf{PU} often doubles the number of available cores because of the hyper-threading support, as the OS considers 2 hyper-threads sharing the same core as 2 different \textsf{PU}.
-Figure~\ref{fig:topo} illustrates a detailed topology of a node belonging to the cluster \textsf{Dahu}.
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/lstopo}}
-    \caption{Topology of the nodes of the cluster \textsf{Dahu}}\label{fig:topo}
-\end{figure}
-% The advantage of using G5K is having an access to a wide set of physical nodes with different capabilities. 
-% The usage procedure is to book a machine for period of time with the chosen OS in less than 5 minutes, ship and run the test script, then release the machine.
-When a machine is released everything is discarded even the OS.
-
-\subsection{Systems Benchmarks}
-Our first criterion to choose the systems benchmarks was the scalability, as we need to run tests at different workloads by choosing the right number of used PU.
-The other criteria are the documentation, the accuracy and the references to the benchmark.
-\emph{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925} is one of the most used suite of benchmarks in the HPC literature and it fulfills our benchmarking requirements.
-We mainly used the pseudo application \emph{Lower-Upper symmetric Gauss-Seidel} (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
-These are the main benchmarks used in many similar works, such as~\cite{heinrich_predicting}.
-Nonetheless, in order to validate our results on a wider set of benchmarks and applications, we also used \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}\footnote{\url{https://linux.die.net/man/1/sha256sum}} as representative systems benchmarks to conduct our experiments with a broad diversity of workloads.
-
-\subsection{Measurement Tools \& Methodology}
-To study the energy consumption of nodes, we considered Intel \emph{Running Average Power Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754}, which is one of the most accurate tools to report the CPU/DRAM global energy consumption.
-We also used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, which is a power monitoring toolkit that builds a model over RAPL to compute the energy consumption at process-level when we needed to isolate energy consumption of a single process.
-Our clusters are provisioned with a minimal version of Debian\,9 (4.9.0 kernel version) where we install Docker (version 18.09.5), which will be used to run the RAPL sensor and the benchmark itself.
-The energy sensor collects RAPL reports and stores them in a remote \textsc{MongoDB} instance, allowing us to perform \emph{post-mortem} analysis in a dedicated environment.
-Using Docker makes the deployment process easier on the one hand, and provides us with a built-in control group encapsulation of the conducted tests on the other hand.
-This allows \textsc{PowerAPI} to measures all the running containers, even the RAPL sensor consumption, as it is isolated in a container.
-One potential threat covers the impact of Docker on the energy variation.
-We therefore conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
-Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
-In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
-The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
-    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
-\end{figure}
-
-Every experiment is conducted on $100$ iterations, on multiple nodes and using the 3 NPB benchmarks we mentioned, with a warmup phase of 10 iterations for each experiment.
-In most cases, we were seeking to evaluate the \emph{STandard Deviation} (STD), which is the most representative factor of the energy variation.
-We tried to be very careful, while running our experiments, not to fall in the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
-As we study the STD difference of measurements we observed from empirical experiments, we use the bootstrap method~\cite{efron2000bootstrap} to randomly build multiple subsets of data from the original dataset, and we draw the STD density of those sets, as illustrated in Figure~\ref{fig:docker}.
-Given the space constraints, this paper reports on aggregated results for nodes, benchmarks and workloads, but the raw data we collected remains available through the public repository we published.\footnote{\url{https://github.com/anonymous-data/Energy-Variation}}
-We believe this can help to achieve better and more reliable comparisons.
-
-We mainly consider 3 different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium and high CPU usage when analyzing the studied parameters effect, respectively.
-These workloads reflect the ratio of used PU count to the total available PU.
-
-\section{Energy Variation Analysis}\label{sec:parameters}
-In this section, we aim to establish experimental guidelines to reduce the CPU energy variation.
-We therefore explore many potential factors and parameters that could have a considerable effect on the energy variation.
-
-\subsection{\textsc{RQ}~1: Benchmarking Protocol}
-To achieve a robust and reproducible experiment, practitioners often tend to repeat their tests multiple times, in order to analyze the related performance indicators, such as execution time, memory consumption or energy consumption.
-We therefore aim to study the benchmarking protocol to identify how to efficiently iterate the tests to capture a trustable energy consumption evaluation.
-% This process of repeating executions should not be random in order to avoid different results because of the CPU behavior, the cache data, the memory size and plenty of other aspects.
-
-In this first experiment, we investigate if changing the testing protocol affects the energy variation.
-To achieve this, we considered 3 execution modes:
-In the \textsf{"normal"} mode, we iteratively run the benchmark 100 times without any extra command, while the \textsf{"sleep"} mode suspends the execution script for 60\,seconds between iterations.
-Finally, the \textsf{"reboot"} mode automatically reboots the machine after each iteration.
-The difference between the \textsf{normal} and \textsf{sleep} modes intends to highlight that the CPU needs some rest before starting another iteration, especially for an intense workload.
-Putting the CPU into sleep for several seconds could give it some time to reach a lower frequency state or/and reduce its temperature, which could have an impact on the energy variation.
-The \textsf{reboot} mode, on the other hand, is the most straightforward way to reset the machine state after every iteration.
-It could also be beneficial to reset the CPU frequency and temperature, the stored data, the cache or the CPU registries.
-However, the reboot task takes a considerable amount of time, so rebooting the node after every single operation is not the fastest nor the most eco-friendly solution, but it deserves to be checked to investigate if it effectively enhances the overall energy variation or not.
-
-Figure~\ref{fig:running-process} reports on $300$ aggregated executions of the  benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}, on 4 machines of the cluster \textsf{Dahu} (cf.
-Table~\ref{table:g5k}) for different workloads.
-We note that the results have been executed with different datasets sizes (\textsf{B}, \textsf{C} and \textsf{D} for single process, 50\,\% and 100\,\% respectively) to remedy to the brief execution times at high workloads for small datasets.
-This justifies the scale differences of reported energy consumptions between the 3 modes in Figure~\ref{fig:running-process}.
-As one can observe, picking one of these strategies does not have a strong impact on the energy variation for most workloads.
-In fact, all the strategies seem to exhibit the same variation with all the workloads we considered---\emph{i.e.}, the STD is tightly close between the three modes.
-The only exception is the \textsf{reboot} mode at 100\,\% load, where the STD is 150\,\% times worst, due to an important amount of outliers.
-This goes against our expectation, even when setting a warm-up time after reboot to stabilize the OS.
-
-\begin{figure*}
-    \center{\includegraphics[width=\linewidth]{imgs/running-process}}
-    \caption{Energy variation with the \textsf{normal}, \textsf{sleep} and \textsf{reboot} modes}\label{fig:running-process}
-\end{figure*}
-
-In Figure~\ref{fig:reboot}, we study the standard deviation of the three modes by constituting $5,000$ random 30-iterations sets from the previous executions set and we compute the STD in each case, considering mainly the 100\,\% workload as the STD was 150\,\% higher for the \textsf{reboot} mode with that load.
-We can observe that the considerable amount of outliers in the \textsf{reboot} mode is not negligible, as the STD density is clearly higher than the two other modes.
-This makes the \textsf{reboot} mode as the less appropriate for the energy variation at high workloads.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/reboot}}
-    \caption{STD analysis of the \textsf{normal}, \textsf{sleep} and \textsf{reboot} modes}\label{fig:reboot}
-\end{figure}
-
-\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
-
-    To answer \textsc{RQ~1}, we conclude that the benchmarking protocol \textbf{partially affects} the energy variation, as highlighted by the \textsf{reboot} mode results for high workloads.
-
-\end{mdframed}
-
-\subsection{\textsc{RQ}~2: Processor Features}
-The C-states provide the ability to switch the CPU between more or less consuming states upon activities.
-Turning the C-states on or off have been subject of many discussions~\cite{5463056}, because of its dynamic frequency mechanism but, to the best of our knowledge, there have been no fully conducted C-states behavior analysis on CPU energy variation.
-
-We intend to investigate how much the energy consumption varies when disabling the C-states (thus, keeping the CPU in the \texttt{C0} state) and at which workload.
-Figure~\ref{fig:c-states} depicts the results of the experiments we executed on three nodes of the cluster \textsf{Dahu}.
-On each node, we ran the same set of benchmarks with two modes: \textsf{C-states on}, which is the default mode, and \textsf{C-states off}.
-Each iteration includes 100 executions of the same benchmark at a given workload, with three workload levels.
-We note that our results have been confirmed with the benchmarks \texttt{LU}, \texttt{CG} and \texttt{EP}.
-
-We can clearly see the effect that has the \textsf{C-states off} mode when running a single-process application/benchmark.
-The energy consumption varies 5~times less than the default mode.
-In this case, only one CPU core is used among $2\times16$ physical cores.
-The other cores are switched to a low consumption state when C-states are on, the switching operation causes an important energy consumption difference between the cores, and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
-On the other hand, switching off the C-states would keep all the cores---even the unused ones---at a high frequency usage.
-This highly reduces the variation, but causes up to 50\,\% of extra energy consumption in this test ($Mean_{C-states-off}=11,665 mJ$,$Mean_{C-states-on}=7,641 mJ$).
-
-At a 100\,\% workload, disabling the C-states seems to have no effect on the total energy consumption nor its variation.
-In fact, all the cores are used at 100\,\% and the C-states module would have no effect, as the cores are not idle.
-The same reason would apply for the 50\,\% load, as the hyper-threading is active on all cores, thus causing the usage of most of them.
-For single process workloads, disabling the C-states causes the process to consume 50\,\% more energy as reported in Figure~\ref{fig:c-states}, but reduces the variation by 5~times compared to the \textsf{C-states on} mode.
-This leads to mainly two questions: Can a process pinning method reduce/increase the energy variation? And, how does the energy consumption variation evolve at different PU usage level?
-
-\begin{figure*}
-    \center{\includegraphics[width=\linewidth]{imgs/c-states}}
-    \caption{Energy variation when disabling the C-states}\label{fig:c-states}
-\end{figure*}
-
-\subsubsection{Cores Pinning}
-To answer the first question, we repeated the previous test at 50\,\% workload.
-In this experiment, we considered three cores usage strategies, the first one (\textsf{S1}) would pin the processes on all the PU of one of the two sockets (including hyper-threads), so it will be used at 100\,\%, and leave the other CPU idle.
-The second strategy (\textsf{S2}) splits the workload on the two sockets so each CPU will handle 50\,\% of the load.
-In this strategy, we only use the core PU and not the hyper-threads PU, so every process would not share his core usage (all the cores are being used).
-The third strategy (\textsf{S3}) consists also on splitting the workload between the two sockets, but considering the usage of the hyper-threads on each core---\emph{i.e.}, half of the cores are being used over the two CPU.
-Figure~\ref{fig:cores-pinning} reports on the energy consumption of the three strategies when running the benchmark \texttt{CG} on the cluster \textsf{Dahu}.
-We can notice the big difference between these three execution modes that we obtained only by changing the PU pinning method (that we acknowledged with more than 100 additional runs over more than 30 machines and with the benchmarks \texttt{LU} and \texttt{EP}).
-For example, \textsf{S2} is the least power consuming strategy.
-We argue that the reason is related to the isolation of every process on a single physical core, reducing the context switch operations.
-In the first and third strategy, $32$ processes are being scheduled on $16$ physical cores using the hyper-threads PU, which will introduce more context switching, and thus more energy consumption.
-
-\begin{figure}
-    \center{\includegraphics[width=.8\linewidth]{imgs/cores_pinning}}
-    \caption{Energy variation considering the three cores pinning strategies at 50\,\% workload}\label{fig:cores-pinning}
-\end{figure}
-
-We note that even if the first and third strategies are very similar (both use hyper-threads, but only on one CPU for the first and on two CPU for the third), the gap between them is considerable variation-wise, as the variation is 30 times lower in the first strategy ($STD_{S1}=116 mJ$,$STD_{S3}=3,452 mJ$).
-This shows that the usage of the hyper-threads technology is not the main reason behind the variation, the first strategy has even less variation than the second one and still uses the hyper-threading.
-
-The reason for the \textsf{S1} low energy consumption is that one of the two sockets is idle and will likely be in a lower power P-state, even with the disabled C-states.
-The \textsf{S2} case is also low energy consuming because by distributing the threads across all the cores, it completes the task faster than in the other cases.
-Hence, it consumes less energy.
-The \textsf{S3} is a high consuming strategy because both sockets are being used, but only half the cores are active.
-This means that we pay the energy cost for both sockets being operational and for the experiments taking longer to run because of the recurrent context switching.
-
-Our hypothesis regarding the worst results that we observed when using the third strategy is the recurrent context switching, added to the OS scheduling that could reschedule processes from a socket to another, which invalids the cache usage as a process can not take profit of the socket local L3 cache when it moves from a CPU to another (cf.
-Figure~\ref{fig:topo}).
-
-Moreover, the fact that the variation is 4--5 times higher when using the strategy \textsf{S2} compared to \textsf{S1} ($STD_{S1}=116 mJ$, $STD_{S3}=575 mJ$), gives another reason to believe that swapping a process from a CPU to another increases the variation due to CPU micro~differences, cache misses and cache coherency.
-While the mean execution time for the strategy \textsf{S3} is very high ($MeanTime_{S3}= 46 s$) compared to the two other strategies ($MeanTime_{S1}= 11 s$, $MeanTime_{S2}= 7 s$), we see no correlation between the execution time and the energy variation, as the \textsf{S1} still give less variations than \textsf{S2} even if it takes 36\,\% more time to run.
-
-Table~\ref{table:corespinning} reports on additional aggregated results for the STD comparison on four other nodes of the cluster \textsf{Dahu} at 50\,\%, with the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
-In fact, the CPU usage strategy \textsf{S1} is by far the experimentation mode that gave the least variation.
-The STD is almost 5 times better than the strategy \textsf{S2}, but is up to 10\,\% more energy consuming ($Mean_{S1}=4469 mJ$, $Mean_{S2}=4016 mJ$).
-On the other hand, the strategy \textsf{S3} is the worst, where the energy consumption can be up to 5 times higher than the strategy \textsf{S2} ($Mean_{S2}=4016 mJ$, $Mean_{S3}=21645 mJ$) and the variation is much worst (30 times compared to the first strategy).
-These results allow us to have a better understanding of the different processes-to-PU pinning strategies, where isolating the workload on a single CPU is the best strategy.
-Using the hyper-threads PU on multiple sockets seems to be a bad recommendation, while keeping the hyper-threading enabled on the machine is not problematic, as long as the processes are correctly pinned on the PU.
-Our experiments show that running one hyper-thread per core is not always the best to do, at the opposite of the claims of~\cite{marathe_empirical_2017_m}.
-
-\begin{table}
-    \centering
-    \caption{STD (mJ) comparison for 3 pinning strategies}
-    \label{table:corespinning}
-    \small
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \textbf{Strategy} & \textbf{S1} & \textbf{S2} & \textbf{S3} \\
-        \hline
-        \hline
-        \textbf{Node~1}   & 88          & 270         & 1,654       \\
-        \hline
-        \textbf{Node~2}   & 79          & 283         & 2,096       \\
-        \hline
-        \textbf{Node~3}   & 58          & 287         & 1,725       \\
-        \hline
-        \textbf{Node~4}   & 51          & 229         & 1,334       \\
-        \hline
-    \end{tabular}
-\end{table}
-
-\subsubsection{Processes Threshold}
-To answer the second question regarding the evolution of the energy variation at different levels of CPU usage, we varied the used PU's count to track the EV evolution.
-Figure~\ref{fig:process-tresh} compares the aggregated energy variation when the C-states are on and off using 2, 4 and 8 processes for the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
-This figure confirms that disabling the CPU C-states does not decrease the variation for all the workloads, as we can clearly observe, the variation is increasing along with the number of processes.
-When running only 2 processes, turning off the C-states reduces the STD up to 6~times, but consumes 20\,\% more energy ($Mean_{C-states-on}=10,334 mJ$, $Mean_{C-states-off}=12,594 mJ$).
-This variation is 4~times lower when running 4 processes and almost equal to the \textsf{C-states on} mode when running 8 processes.
-In fact, running more processes implies to use more CPU cores, which reduces the idle cores count, so the cores will more likely stay at a higher consumption state even if the C-states mechanism is on.
-
-In our case, using 4 PU reduces the variation by 4~times and consumes almost the same energy as keeping the C-states mechanism on ($Mean_{C-states-on}=7,048 mJ$, $Mean_{C-states-off}=7,119 mJ$).
-This case would be the closest to reality as we do not want to increase the energy consumption while reducing the variation, but using a lower number of PU still results in less variation, even if it increases the overall energy consumption.
-
-\begin{figure*}
-    \center{\includegraphics[width=\linewidth]{imgs/tresh}}
-    \caption{C-states effect on the energy variation, regarding the application processes count}\label{fig:process-tresh}
-\end{figure*}
-
-We note that disabling the C-states is not recommended in production environments, as it introduces extra energy consumption for low workloads (around 50\,\% in our case for a single process job).
-However, our goal is not to optimize the energy consumption, but to minimize the energy variation.
-Thus, disabling the C-states is very important to stabilize the measurements in some cases when the variation matters the most.
-Comparing the energy consumptions of two algorithms or two versions of a software systems is an example of use case benefiting from this recommendation.
-
-\subsubsection{Turbo Boost}
-The Turbo Boost---also known as \emph{Dynamic Overclocking}---is a feature that has been incorporated in Intel CPU since the Sandy~Bridge micro-architecture, and is now widely available on all of the Core\,i5, Core\,i7, Core\,i9 and Xeon series.
-It automatically raises some of the CPU cores operating frequency for short periods of time, and thus boost performances under specific constraints.
-When demanding tasks are running, the operating system decides on using the highest performance state of the processor.
-
-Disabling or enabling the Turbo~Boost has a direct impact on the CPU frequency behavior, as enabling it allows the CPU to reach higher frequencies in order to execute some tasks for a short period of time.
-However, its usage does not have a trivial impact on the energy variation.
-Acun~\emph{et~al.}~\cite{acun_variation_2016} tried to track the Turbo~Boost impact on the Ivy~Bridge and the Sandy~Bridge architectures.
-They concluded that it is one of the main responsible for the energy variation, as it increases the variation from 1\,\% to 16\,\%.
-In our study, we included a Turbo~Boost experiment in our testbed, to check this property on the recent Xeon Gold processors, covering various workloads.
-
-The experiment we conducted showed that disabling the Turbo Boost does not exhibit any considerable positive or negative effect on the energy variation.
-Table~\ref{table:turboboost} compares the STD when enabling/disabling the Turbo~Boost, where the columns are a combination of workload and benchmark.
-In fact, we only got some minor measurements differences when switching on and off the Turbo~Boost, and where in favor or against the usage of the Turbo~Boost while repeating tests, considering multiple nodes and benchmarks.
-This behavior is mainly related to the \emph{thermal design power} (TDP), especially at high workloads executions.
-When a CPU is used at its maximum capacity, the cores would be heating up very fast and would hit the maximum TDP limit.
-In this case, the Turbo~Boost cannot offer more power to the CPU because of the CPU thermal restrictions.
-At lower workloads, the tests we conducted proved that the Turbo~Boost is not one of the main reasons of the energy variation.
-In fact, the variation difference is barely noticeable when disabling the Turbo~Boost, which cannot be considered as a result regarding the OS activity and the measurement error margin.
-We cannot affirm that the Turbo~Boost does not have an impact on all the CPU, as we only tested on two recent Xeon CPU (clusters \textsf{Chetemi} and \textsf{Dahu}).
-We confirmed our experiments on these machines 100 times at 5\,\%, 25\,\%, 50\,\% and 100\,\% workloads.
-
-\begin{table}%[h!]
-    \centering
-    \caption{STD (mJ) comparison when enabling/disabling the Turbo~Boost}
-    \label{table:turboboost}
-    \small
-    \begin{tabular}{|c|r|r|}
-        \hline
-        \textbf{Turbo Boost}  & \textbf{Enabled} & \textbf{Disabled} \\
-        \hline
-        \hline
-        \textsf{EP} / 5\,\%   & 310              & 308               \\
-        \hline
-        \textsf{CG} / 25\,\%  & 95               & 140               \\
-        \hline
-        \textsf{LU} / 25\,\%  & 204              & 240               \\
-        \hline
-        \textsf{EP} / 50\,\%  & 84               & 79                \\
-        \hline
-        \textsf{EP} / 100\,\% & 125              & 110               \\
-        \hline
-    \end{tabular}
-\end{table}
-
-\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
-    We conclude that CPU features \textbf{highly impact} the energy variation as an answer for \textsc{RQ~2}.
-\end{mdframed}
-
-\subsection{\textsc{RQ}~3: Operating System}
-The \emph{operating system} (OS) is the layer that exploits the hardware capabilities efficiently.
-It has been designed to ease the execution of most tasks with multitasking and resource sharing.
-In some delicate tests and measurements, the OS activity and processes can cause a significant overhead and therefore a potential threat to the validity. % experiments validity
-The purpose behind this experiment is to determine if the sampled consumption can be reliably related to the tested application, especially for low-workload applications where CPU resources are not heavily used by the application.
-
-The first way to do is to evaluate the OS idle activity consumption, and to compare it to a low workload running job.
-Therefore, we ran 100 iterations of a single process benchmark \textsf{EP}, \textsf{LU} and \textsf{CG} on multiple nodes from the cluster \textsf{Dahu}, and compared the energy behavior of the node with its idle state on the same duration.
-The aggregated results, illustrated in Figure~\ref{fig:os-idle}, depict that the idle energy variation is up to 140\,\% worst than when running a job, even if it consumes 120\,\% less energy ($Mean_{Job}=8,746 mJ$, $Mean_{Idle}=3,927 mJ$).
-In fact, for the three nodes, randomly picked from the cluster \textsf{Dahu}, the idle variation is way more important than when a test was running, even if it is a single process test on a $32$-cores node.
-This result shows that OS idle consumption varies widely, due to the lack of activity and the different CPU frequencies states, but it does not mean that this variation is the main responsible for the overall energy variation.
-The OS behaves differently when a job is running, even if the amount of available cores is more than enough for the OS to keep his idle behavior when running a single process.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/idle}}
-    \caption{OS consumption between idle and when running a single process job}\label{fig:os-idle}
-\end{figure}
-
-Inspecting the OS idle energy variation is not sufficient to relate the energy variation to the active job.
-In fact, the OS can behave differently regarding the resource usage when running a task.
-To evaluate the OS and the job energy consumption separately, we used the \textsc{PowerAPI} toolkit.
-This fine-grained power~meter allows the distribution of the RAPL global energy across all the Cgroups of the OS using a power model.
-Thus, it is possible to isolate the job energy consumption instead of the global energy consumption delivered by RAPL.
-To do so, we ran tests with a single process workload on the cluster \textsf{Dahu}, and used the \textsc{PowerAPI} toolkit to measure the energy consumption.
-Then, we compared the job energy consumption to the global RAPL data.
-We calculated the Pearson correlation~\cite{ref1} of the energy consumption and variation between global RAPL and \textsc{PowerAPI}, as illustrated in Figure~\ref{fig:correlation}.
-The job energy consumption and variation are strongly correlated with the global energy consumption and variation with the coefficients 93.6\,\% and 85.3\,\%, respectively.
-However, this does not completely exclude the OS activity, especially if the jobs have tight interaction with the OS through the signals and system calls.
-This brings a new question on whether applying extra-tuning on a minimal OS would reduce the variation?
-As well as what is the effect of the Meltdown security patch---that is known to be causing some performance degradation~\cite{Kocher2018spectre,Lipp2018meltdown}---on the energy variation?
-
-\begin{figure}
-    \center{\includegraphics[width=.8\linewidth]{imgs/correlation}}
-    \caption{The correlation between the RAPL and the job consumption and variation}\label{fig:correlation}
-\end{figure}
-
-\subsubsection{OS Tuning}
-An OS is a pack of running processes and services that might or not be required its execution.
-In fact, even using a minimal version of a Debian Linux, we could list many OS running services and process that could be disabled/stopped without impacting the test execution.
-This extra-tuning may not be the same depending on the nature of the test or the OS.
-Thus, we conducted a test with a deeply-tuned OS version.
-We disabled all the services/processes that are not essential to the OS/test running, including the OS networking interfaces and logging modules, and we only kept the strict minimum required to the experiment's execution.
-Table~\ref{table:osmin} reports on the aggregated results for running single process measurements with the benchmarks \textsf{CG}, \textsf{LU} and \textsf{EP}, on three servers of the cluster \textsf{Dahu}, before and after tuning the OS.
-Every cell contains the \emph{STD} value before the tuning, plus/minus a ratio of the energy variation after the tuning.
-We notice that the energy variation varies less than 10\,\% after the extra-tuning.
-We argue that this variation is not substantial, as it is not stable from a node to another.
-Moreover, 10\,\% of variation is not a representative difference, due to many factors that can affect it as the CPU temperature or the measurement errors.
-
-\begin{table}
-    \centering
-    \caption{STD (mJ) comparison before/after tuning the OS}
-    \label{table:osmin}
-    \small
-    \begin{tabular}{|c|c|c|c|}
-        \hline
-        \textbf{Node} & \textbf{\sf EP} & \textbf{\sf CG} & \textbf{\sf LU} \\
-        \hline
-        \hline
-        \textbf{N1}   & 1370~-9\,\%     & 78~+7\,\%       & 128~+2\,\%      \\
-        \hline
-        \textbf{N2}   & 1278~-7\,\%     & 64~-1\,\%       & 120~+9\,\%      \\
-        \hline
-        \textbf{N3}   & 1118~+1\,\%     & 83~+2\,\%       & 93~+7\,\%       \\
-        \hline
-    \end{tabular}
-\end{table}
-% \subsubsection{Meltdown Security Patch}
-
-\subsubsection{Speculative Executions}
-Meltdown and Spectre are two of the most famous hardware vulnerabilities discovered in 2018, and exploiting them allows a malicious process to access others processes data that is supposed to be private~\cite{Kocher2018spectre,Lipp2018meltdown}.
-They both exploit the speculative execution technique where a process anticipates some upcoming tasks, which are not guaranteed to be executed, when extra resources are available, and revert those changes if not.
-Some OS-level patches had been applied to prevent/reduce the criticality of these vulnerabilities.
-On the Linux kernel, the patch has been automatically applied since the version 4.14.12.
-It mitigates the risk by isolating the kernel and the user space and preventing the mapping of most of the kernel memory in the user space.
-% These patches have been reported as decreasing the performance of a system. 
-Nikolay~\emph{et~al.} have studied in~\cite{DBLP:journals/corr/abs-1801-04329} the impact of patching the OS on the performance.
-The results showed that the overall performance decrease is around 2--3\,\% for most of the benchmarks and real-world applications, only some specific functions can meet a high performance decrease.
-In our study, we are interested in the applied patch's impact on the energy variation, as the performance decrease could mean an energy consumption increase.
-Thus, we ran the same benchmarks \textsf{LU}, \textsf{CG} ad \textsf{EP} on the cluster \textsf{Dahu} with different workloads, using the same OS, with and without the security patch.
-Table~\ref{table:meltdown} reports on the STD values before disabling the security patch.
-A minus means that the energy varies less without the patch being applied, while a plus means that it varies more.
-These results help us to conclude that the security patch's effect on the energy variation is not substantial and can be absorbed through the error margin for the tested benchmarks.
-In fact, the best case to consider is the benchmark \textsf{LU} where the energy variation is less than 10\,\% when we disable the security patch, but this difference is still moderate.
-The little performance difference discussed in~\cite{Kocher2018spectre,Lipp2018meltdown} may only be responsible of a small variation, which will be absorbed through the measurement tools and external noise error margin in most cases.
-
-\begin{table}[h!]
-    \centering
-    \caption{STD (mJ) comparison with/without the security patch}
-    \label{table:meltdown}
-    \small
-    \begin{tabular}{|l|c|c|c|}
-        \hline
-        \textbf{Node} & \textbf{\sf EP} & \textbf{\sf CG} & \textbf{\sf LU} \\
-        \hline
-        \hline
-        \textbf{N1}   & 269~+2\,\%      & 83~+1\,\%       & 108~-6\,\%      \\
-        \hline
-        \textbf{N2}   & 195~+1\,\%      & 84~-5\,\%       & 121~-9\,\%      \\
-        \hline
-        \textbf{N3}   & 223~+/-1\,\%    & 72~-4\,\%       & 117~+8\,\%      \\
-        \hline
-        \textbf{N4}   & 276~+3\,\%      & 60~+0\,\%       & 113~-3\,\%      \\
-        \hline
-    \end{tabular}
-\end{table}
-
-\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
-    To answer \textsc{RQ~3}, we conclude that the OS \textbf{should not be the main focus} of the energy variation taming efforts.
-\end{mdframed}
-
-\subsection{\textsc{RQ}~4: Processor Generation}
-Intel microprocessors have noticeably evolved during these last 20 years.
-Most of the new CPU come with new enhancements to the chip density, the maximum Frequency or some optimization features like the C-states or the Turbo~Boost.
-This active evolution caused that different generations of CPU can handle a task differently.
-The aim of this expriment is not to justify the evolution of the variation across CPU versions/generations, but to observe if the user can choose the best node to execute her experiments.
-Previous papers have discussed the evolution of the energy consumption variation across CPU generations and concluded that the variation is getting higher with the latest CPU generations~\cite{wang_experimental_nodate,marathe_empirical_2017_m}, which makes measurements stability even worse.
-In this experiment, we therefore compare four different generations of CPU with the aim to evaluate the energy variation for each CPU and its correlation with the generation.
-Table~\ref{table:cpus} indicates the characteristics of each of the tested CPU.
-
-\begin{table}
-    \centering
-    \caption{STD (mJ) comparison of experiments from 4 clusters}
-    \small
-    \begin{tabular}{|l|c|c|c|c|}
-        \hline
-        \textbf{Cluster} & \textbf{Dahu} & \textbf{Chetemi} & \textbf{Ecotype} & \textbf{Paranoia} \\
-        \hline
-        \hline
-        Arch             & Skylake       & Broadwell        & Broadwell        & Ivy Bridge        \\
-        \hline
-        Freq             & 3.7\,GHz      & 3.1\,GHz         & 2.9\,GHz         & 3.0\,GHz          \\
-        \hline
-        TDP              & 125 W         & 85 W             & 55 W             & 95 W              \\
-        \hline
-        \hline
-        5\%              & 364           & 210              & \textbf{75}      & \textbf{76}       \\
-        \hline
-        50\%             & 98            & 86               & \textbf{49}      & 244               \\
-        \hline
-        100\%            & 119           & 116              & \textbf{106}     & 240               \\
-        \hline
-    \end{tabular}
-    \label{table:cpus}
-\end{table}
-
-Table~\ref{table:cpus} also shows the aggregated energy variation of the different generations of nodes for the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
-The results attest that the latest versions of CPU do not necessarily cause more variation.
-In the experiments we ran, the nodes from the cluster \textsf{Paranoia} tend to cause more variation at high workloads, even if they are from the latest generation.
-While the Skylake CPU of the cluster \textsf{Dahu} cause often more energy variation than \textsf{Chetemi} and the Ecotype Broadwell CPU.
-We argue that the hypothesis "\emph{the energy consumption on newer CPU varies more}" could be true or not depending on the compared generations, but most importantly, the chips energy behaviors.
-On the other hand, our experiments showed the lowest energy variation when using the \textsf{Ecotype CPU}, these CPU are not the oldest nor the latest, but are tagged with "\texttt{L}" for their low power/TDP.
-This result rises another hypothesis when considering CPU choice, which implies selecting the CPU with a low TDP.
-This hypothesis has been confirmed on all the \textsf{Ecotype} cluster nodes, especially at low and medium workloads.
-
-Figure~\ref{fig:cpugen} is an illustration of the aggregated STD density of more than $5,000$-random values sets taken from all the conducted experiments.
-This shows that the cluster \textsf{Paranoia} reports the worst variation in most cases, and that \textsf{Ecotype} is the best cluster to consider to get the least variations, as it has a higher density for small variation values.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/cpugen}}
-    \caption{Energy consumption STD density of the 4 clusters}\label{fig:cpugen}
-\end{figure}
-
-\begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
-    We conclude on \textbf{affirming \textsc{RQ}~4}, as selecting the right CPU can help to get less variations.
-\end{mdframed}
-
-\section{Experimental Guidelines}\label{sec:guidelines}
-To summarize our experiments, we provide some experimental guidelines in Table~\ref{table:guidelines}, based on the multiple experiments and analysis we did.
-These guidelines constitute a set of minimal requirements or best practices, depending on the workload and the criticality of the energy measurement precision.
-It therefore intends to help practitioners in taming the energy variation on the selected CPU, and conduct the experiments with the least variations.
-
-\begin{table}[h!]
-    \centering
-    \caption{Experimental Guidelines for Energy Variations}
-    \small
-    \begin{tabular}{|p{4.7cm}|c|c|}
-        \hline
-        \textbf{Guideline}                                                                                                                                                & \textbf{Load} & \textbf{Gain}     \\
-        \hline
-        \hline
-        Use a low TDP CPU                                                                                                                                                 & Low \& medium & Up to $3\times$   \\
-        \hline
-        Disable the CPU C-states                                                                                                                                          & Low           & Up to $6\times$   \\
-        \hline
-        Use the least of sockets in a case of multiple CPU                                                                                                                & Medium        & Up to $30\times$  \\
-        \hline
-        Avoid the usage of hyper-threading whenever possible                                                                                                              & Medium        & Up to $5\times$   \\
-        \hline
-        Avoid rebooting the machine between tests                                                                                                                         & High          & Up to $1.5\times$ \\
-        \hline
-        Do not relate to the machine idle variation to isolate a test EC, the CPU/OS changes its behavior when a test is running and can exhibit less variation than idle & Any           & ---               \\
-        \hline
-        Rather focus the optimization efforts on the system under test than the OS                                                                                        & Any           & ---               \\
-        \hline
-        Execute all the similar and comparable experiments on a same machine.
-Identical machines can exhibit many differences regarding their energy behavior             & Any           & Up to $1.3\times$ \\
-        \hline
-    \end{tabular}
-    \label{table:guidelines}
-\end{table}
-
-Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and its variation reduction at low workloads.
-However, it also lists some new factors that we identified along the analysis we conducted in Section~\ref{sec:parameters}, such as the results related to the OS or the reboot mode.
-Some of the guidelines are more useful/efficient for specific workloads, as showed in our experiments.
-Thus, qualifying the workload before conducting the experiments can help in choosing the proper guidelines to apply.
-Other studied factors are not been mentioned in the guidelines, like the Turbo~Boost or the Speculative execution, due to the small effect that has been observed in our study.
-
-In order to validate the accuracy of our guidelines among a varied set of benchmarks on one hand, and their effect on the variation between identical machines on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
-Half of these experiments has been performed at a 50\,\% workload and the other half on single process jobs.
-The choice of these two workloads is related to the optimization guidelines that are mainly effective at low and medium workloads.
-We note that we used the cluster \textsf{Dahu} over \textsf{Ecotype} to highlight the guidelines effect on the nodes where the variation is susceptible to be higher.
-
-\begin{figure*}%[!htb]
-    \center{\includegraphics[width=\linewidth]{imgs/all}}
-    \caption{Energy variation comparison with/without applying our guidelines}\label{fig:optimized}
-\end{figure*}
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/stressng}}
-    \caption{Energy variation comparison with/without applying our guidelines for \textsc{Stress-NG}}\label{fig:stress}
-\end{figure}
-
-Figure~\ref{fig:optimized} and~\ref{fig:stress} highlight the improvement brought by the adoption of our guidelines.
-They demonstrate the intra-node STD reduction at low and medium workloads for all the benchmarks used at different levels.
-Concretely, for low workloads, the energy variation is 2--6 times lower after applying the optimization guidelines for the benchmarks \textsf{LU} and \textsf{EP}, as well as \textsc{Linpack}, while it is 1.2--1.8 times better for \texttt{Sha256}.
-For this workload, the overall energy consumption after optimization can be up to 80\,\% higher due to disabling the C-states to keep all the unused cores at a high power consumption state ($Mean_{LU-normal-Dahu2}=11,500 mJ$, $Mean_{LU-optimized-Dahu2}=20,508 mJ$).
-For medium workloads, the STD, and thus variation, is up to 100\,\% better for the benchmark \textsf{CG}, 20--150\,\% better for the \texttt{pbzip2} application and up to 100\% for \textsc{Stress-NG}.
-We note that the optimized version consumes less energy thanks to an appropriate core pinning method.
-
-Figures~\ref{fig:optimized} and~\ref{fig:stress} also highlight that applying the guidelines does not reduce the inter-nodes variation in all the cases.
-This variation can be up to 30\,\% in modern CPU~\cite{wang_experimental_nodate}.
-However, taming the intra-node variation is a good strategy to identify more relevant mediums and medians, and then perform accurate comparisons between the nodes variation.
-Even though, using the same node is always better, to avoid the extra inter-nodes variation and thus improve the stability of measurements.
-
-\section{Threats to Validity}\label{sec:threats}
-A number of issues affect the validity of our work.
-For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPU, but still adds an important overhead if we adopt a sampling at high frequency.
-The other fine-grained tool we used for measurements is \textsc{PowerAPI}.
-It allows to measure the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
-The usage of \textsc{PowerAPI} adds an error margin because of the power model built over RAPL.
-The RAPL tool mainly measures the CPU and DRAM energy consumption.
-However, even running CPU/RAM intensive benchmarks would keep a degree on uncertainty concerning the hard disk and networking energy consumption.
-In addition, the operating system adds a layer of confusion and uncertainty.
-
-The Intel CPU chip manufacturing process and the materials micro-heterogeneity is one of the biggest issues, as we cannot track or justify some of the energy variation between identical CPU or cores.
-These CPU/cores might handle frequencies and temperature differently and behave consequently.
-This hardware heterogeneity also makes reproduction complex and requires the usage of the same nodes on the cluster with the same OS.
-
-% A more subtle issue may arise due to the values of the measurements that we achieved.
-% In fact, the energy measures are quite small, and iterations may be taking a few milliseconds more or less to run.
-% A thing we cannot measure using our measurement tools.
-% How generalizable are our results? As a set of energy variation optimization guidelines, we argue that our results applied on most of the modern Intel CPU.
-% However, using and comparing identical CPU is still tricky and is very dependent to chips.
-
-\section{Conclusion}\label{sec:conclusion}
-In this paper, we conducted an empirical study of controllable factors that can increase the energy variations on platforms with some of the latest CPU, and for several workloads.
-We provide a set of guidelines that can be implemented and tuned (through the OS GRUB for example), especially with the new data centers isolation trend and the cloud usage, even for scientific and R\&D purposes.
-Our guidelines aim at helping the user in reducing the CPU energy variation during systems benchmarking, and conduct more stable experiments when the variation is critical.
-For example, when comparing the energy consumption of two versions of an algorithm or a software system, where the difference can be tight and need to be measured accurately.
-
-Overall, our results are not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
-The aim of our work is to be able to tame and mitigate this variability along controlled experiments.
-We studied some previously discussed aspects on some recent CPU, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to achieve the variability mitigating purpose.
-Some of these factors, like the C-states usage, can reduce the energy variation up to 500\,\% at low workloads, while choosing the wrong cores/PU strategy can cause up to $30\times$ more variability.
-
-We believe that our approach can also be used to study/discover other potential variability factors, and extend our results to alternative CPU generations/brands.
-Most importantly, this should motivate future works on creating a better knowledge on the variability due to CPU manufacturing process and other factors.
+We are fully aware of the impact that hardware has on energy variation.
+However, we believe that there is still a room to reduce this energy variation for practicioners using only paramenters that they are in cotrol.
+To do so, we have inducted a set of empirical experiments using the guidelines provided by state of the art, to determine which controllable factors can reduce the variation of the energy consumption of tests.
+We will first start with evaluating the paramenters mentioned by \cite{heinrich_predicting}.

### 2021-10-22 chapters/programming_languages/programming_languages.tex
add : langauges chapter structure

\newpage

\chapter{The Energy footprints of programming languages}
\label{chapter:porgramming_langauges}
In this chapter we will dicuss the impact of the choice of the programming language on the energy consumption of the software.

\import{chapters/programming_languages/}{challanges}
\import{chapters/programming_languages/}{microbenchmarks}
\import{chapters/programming_languages/}{conclusion}
\import{chapters/programming_languages/}{rpc}
\import{chapters/programming_languages/}{web_frameworks}

### 2021-10-29 chapters/programming_languages/challanges.tex
add : rpc figures

The ultimate goal of this chapter is to provide a guideline on which programming language the developers should chose based on the charecteristics of the project in order to minimize the energy foot print of their product.
No answer is evidedent for a such question. However, we can extract some feartures of each programming langues such as
\begin{itemize}
    \item performance
    \item community support
    \item scalability
    \item energy consumptiom
    \item memory usage
    \item etc
\end{itemize}

first we will start but analysing the behaviour of the general purpose programming langauges with some micro benchkaks, principally for the CLBG game and others from rosetta code base

.

### 2021-10-29 chapters/programming_languages/programming_languages.tex
add : rpc figures

@@
 In this chapter we will dicuss the impact of the choice of the programming language on the energy consumption of the software.
+To do so we suggest to start with the general micro benchmarking and see how each programming languages interact with the CPU/ Memory
 
@@
 \import{chapters/programming_languages/}{microbenchmarks}
-\import{chapters/programming_languages/}{conclusion}
+As we have seen in the previous chapter, one of the most important feature of a test is to be \textbf{representative}.
+Therefore, we extend this study to some reallife use case.
+The sections belows will provide two study cases.
 \import{chapters/programming_languages/}{rpc}
 \import{chapters/programming_languages/}{web_frameworks}
+\import{chapters/programming_languages/}{conclusion}

### 2021-10-29 chapters/programming_languages/rpc.tex
add : rpc figures

\section{Remote Procedure Call}

\subsection{Definition}

\subsection{Motivation}
%  this is just to warmup 
With the emerging technology to the cloud many protocols wanted to take the lead.
Nowdays most of the architectures are based on multi-servicces and micro services.
And to have higher versatility of developpers mulptile companies choose to be open to different programming langauges. basically it would be more efficient if we take advantage of each programming language to satisfy a specific need.
However the callenge nowdays is to make the bridge between those plateforms.
We have many initiatives such as openapi that try to create a taxonomy for rest apis.
other approachs is to impelement all the different interfaces of the protocol by them selfs such as the RPC
\subsection{State of the art}

\subsection{Research Questions}
In this section we will first explore the ease of implemenation of this protocol
then we will try to answer the following Research questions

\begin{description}
    \item[\textsc{RQ}~1:] How do RPC implementations react to the size of the request ?
    \item[\textsc{RQ}~2:] How do RPC implementations react to the number of clients ?

\end{description}
\subsection{Experimental protocol}
\subsubsection{Hardware settings}
All the experiments are run on the cluster paravance of the G5K platform. This cluster is composed of 72 identical machines, each one is equipped with 2 Intel Xeon E5-2630 V3, with 128 GIB of RAM. For more accuracy, our SUT (System Under Test) is equipped with a minimal version of Debian 9 (4.9.0 kernel version), which enforces the core processes required for the purpose of our experiment. Furthermore, we used Docker containers technology for reproducibility of the experiments and the isolation of the servers.
\subsubsection{Energy measurements}
To report on the energy consumption, we used HWPC sensor [], which is based on Intel RAPL technology, one of the most accurate tools to measure the energy consumption of the CPU and DRAM []. For better accuracy, we ran the HPWC sensor with a frequency of 10 Hz, and we used the same machine for all the experiments in order to reduce the variability [].

\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=.8\linewidth]{imgs/rpcprotocol}
    \end{center}
    \caption{Experemenal software architecture}\label{fig:rpcprotocol}
\end{figure}
\subsubsection{Client and server environments}

To limit the impact of the network on the experiments, we run both the client and the server on the same machine. However, we isolate each one on a different socket, in order to reduce the effect that the client might have on the server ones and vise-versa. To do so, for each iteration, we always run the same client on Socket 0 and the server that we want to test on socket 1. Both the server and the client take the whole socket for their experiment. In addition all the extra services, such as OS, hwpc, etc. are run on Socket 0. Therefore, the only process being executed in Socket 1 is the server that we benchmark.

\paragraph{Client :}
% TODO : add the github repository links 
For better accuracy and more details, We use an updated version of the open source RPC benchmarking tool, named GHZ (https://ghz.sh/).
The modified version allows us to get the average power for each request form both the server and the client sides.
The new version is available in the repo.
The client will take the protocol description to generate an implementation for the message helloworld, and then fork 50 instances that will send the same request to the server simultaneously.
\paragraph{Server :}
The server implementations are based on the official implementation by Google for most of the languages . Each server uses 16 cores  and is limited to 512 MB of RAM

\subsection{Results and finding}
[\textsc{RQ}~1:] How do RPC implementations react to the size of the data ?
The purpose of this question is to study the behaviour of the server for transferring large objects. To do so, we send to the server 80,000 requests with a size scaling from 10 bytes up to 10 Megabytes, which gives 10,000 requests per size per server. To eliminate extra factors, we let the server handle the rate at which it can answer each request. However, we put a 20 sec timeout limit for each request. Therefore, our boundary condition is only the number of requests received by the server.  For this experiment, we investigate 4 observable variables :
\begin{enumerate}
    \item The average power consumption during the the process : this will indicate the overall behaviour of the server in working mode for long durations ;
    \item The tail latency for the 99th percentile : which indicates how performant is the server ;
    \item The average number of requests per second : which indicates the  average number of clients that  the server can handle ;
    \item The average energy cost of a single request : unlike the first indicator, this one shows how green is the implementation taking performance into consideration.
\end{enumerate}

% TODO : ADD the figures 
The above figure depicts the overall behaviour of each framework based on the size of request (Payload). For each framework, we can distinguish three modes, and they all depend on the payload :
\begin{enumerate}
    \item Stress free mode: when the server has enough resources to satisfy the requests because they require a memory less than a certain threshold (depends on the language and the platform),
    \item Escalation mode: where the requests tend to be bigger, however the server can still manage to handle them, and here where we can see a change in the energetic and performance behaviour.
    \item Broken state mode: when the requests are much heavier and the server break—like 10 MB.
\end{enumerate}

\subsubsection{Stress free mode}
In this mode, the compiled languages tend to consume less resources (av power). JVM-based languages tend to consume more energy, especially Scala. However, we do not observe the same behaviour when it comes to efficiency. Unlike the other interpreted programming languages, PHP performances could be compared to the compiled ones, such as CPP or GO, and even better to some others, such as Swift. JVM-based languages tend to have better performances than the interpreted ones. Furthermore, OpenJDK has shown more efficiency than GraalVM []. Overall, we can have 3 groups when it comes the cost of each request:
\begin{itemize}
    \item Green languages: CPP, GO, RUST, ELIXIR, and PHP
    \item Middle class: Most of the interpreted languages and vm-based ones
    \item Energivore class: Crystal and Scala
\end{itemize}

\subsubsection{Escalation mode}
In this mode, the behaviour of the server depends on the payload. We observe three behaviours
\begin{enumerate}
    \item Drop in performances without an increased power, such as .Net core, Java micronaut, Crystal, and Dart. In this case the server keeps using the same ressources, and sometimes less, because it takes more time to handle the less requests. This class of languages tend to be the most energivore when it comes to the cost per request ;
    \item Increase in power without affecting the performances: such as Go, .Net. The energy consumption of a single request, is affected slightly but still increase ;
    \item Increase in power and drop in performances: Despite the increase of the power consumption, the server becomes slightly slower, which increases the cost of the energy cost per request. This cost is still better than the first case, which concludes  that the servers in the first category are on the verge of breaking.
\end{enumerate}

Special mention to Elixir that kept scaling despite the lack of performances compared  to other compiled languages (Go, CPP)

\subsubsection{Broken state mode }
Only four of the 25 configurations could parse the 10 MB files. And only 1 from those could achieve a 76\% acceptance rate which is Elixir, the other 3 had less than 3\% success rate (Rust, Swift and Dart).
The rest could be divided into two categories:
\begin{itemize}
    \item Timeout : where requests took too much time that the client canceled them, in this category we find most of dynamic codes such as: openJDK, Kotlin ;
    \item The size of request exceeded the maximum size by implementation : this is where the implementation could not handle requests with large size, such as .Net, Go, .Net core, CPP, PHP, Scala, Nodejs, Ruby, Python.
\end{itemize}

[\textsc{RQ}~2:] How do RPC implementations react to the number of clients ?

\subsubsection{Power behaviour}
based on the heatmap, we can distinguish two main modes.
\begin{itemize}
    \item Low number of clients : where the total number of simultaneous clients is less than 100
    \item Moderate to high number of clients : the number of clients exceeds 100
\end{itemize}

\paragraph{Lite mode}
The implementations can be grouped into two main categories :
\begin{enumerate}
    \item Green Frameworks: most of the framework's power consumption is around 33 Watts.
    \item Energivore frameworks : where the average power consumption is higher than 37 Watts.
\end{enumerate}

In each programming category we observe both behaviours green and energivore. Therefore, we conclude that it depends more on the implementation of the library itself rather than the category of the programming language. Scala and Kotlin are an excellent example to support this hypothesis since both of them run on the same virtual machine as Java (openjdk 16.1).Yet, their average power is 130\% higher than the Java implementation .

\paragraph{Stressed mode}
Although the same classes remained the same , Not all the languages had the same evolution. and here we can clearly see that it is correlated with the category of the programming language rather than the implementation itself.\\
We can clearly highlight that after VM based languages have a significant increase in the average power consumption after they receive more than 100 simultaneous clients. This increase reaches almost double.
Except PHP, all the interpreted languages preserved their energetic behaviour. Same for the compiled ones.
Our Hypothesis points to the JIT, since it will compile the code and make it run faster so it will stress the CPU more.\\
A Remarquable behaviour has been noticed for the grallVM, is the decrease of the energy consumption when we increase the number of the clients. This is related to the drop of the performances which was probably due to the bottleneck situation where the GraalVM couldn't handle more than 100 clients simultaneously.

\subsubsection{Performance Behaviour}
In this section we study only the number of requests per seconds processed by the server without looking at its energy.
We have three observable variables
\begin{itemize}
    \item Satisfaction ratio : how many requerts have been satisfied among the total requests
    \item Request Per Seconds : The number of the requests that have been answered from the server
    \item TailLatency at 99\% : one of the best metrics to evaluate the performances of a server
\end{itemize}

\paragraph{Satisfaction ratio}
Most of the frameworks tried to satisfy all the requests, by either reducing the number of requests per second or by increasing the time of the treatment. However, there are some frameworks that have chosen a different approach such as dart or scalla where the choice was to keep a certain limit of latency even if not all the requests are answered.\\
Furthermore, we tend to see this behaviour among other frameworks such as Python or Asynchronous NodeJs when the number of the client exceeds 800.
\paragraph{RPS}
Most of the servers hit their RPS limit after 5 clients and 100 clients for vm based servers, and after this the number keeps constant. which will decrease the average RPS per client.\\
.net Server is the most performant one, and after him we see JAVA and Go in second place, on the other end Python and ruby are the least performant.
\paragraph{Tail LAtency}
However, the increase in the number of requests per second, doesn't necessarily mean a lower Latency.
As we see in the table ... until the 1000 clients, Go provides the least Latency beside .net.\\
GraalVM provides the highest latency, on average. However, dart tends to become slower when we increase the number of clients, until we pass the 600 simultaneous clients, and there it changes its behaviour, instead of satisfying most the requests it notify the clients directly that the server is saturated, hence a drop in satisfaction ratio, and an amelioration for the Average Latency.

\subsubsection{Energy Per Request}
Now after we made a separation between The energy and The performances, we have seen that most of the performance servers tend to be energy hungry, so we propose to investigate this trade off between the energy and the performance. Todo so we will give an average cost of a single Request in Joules.
Except GraalVM when the cost of the a single request increases with when we add more clients, All the frameworks keep a constant cost,
Java, .net and go are the greenest. and Python, ruby are the most costly with 10x higher.
Therefore we conclude that the number of clients won't impact the energy that much.
Next study will be the payload and how the size of the requests will affect the energy consumption of the framework

\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/energy_cost_clients}
    \end{center}
    \caption{Experemenal software architecture}\label{fig:rpcprotocol}
\end{figure}

\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/energy_cost_payload}
    \end{center}
    \caption{Experemenal software architecture}\label{fig:energy_cost_payload}
\end{figure}

\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/power_consumption_clients}
    \end{center}
    \caption{Experemenal software architecture}\label{fig:power_consumption_clients}
\end{figure}

\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/power_consumption_payload}
    \end{center}
    \caption{Experemenal software architecture}\label{fig:power_consumption_payload}
\end{figure}

\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/rps_clients}
    \end{center}
    \caption{Experemenal software architecture}\label{fig:rps_clients}
\end{figure}

\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/rps_payload}
    \end{center}
    \caption{Experemenal software architecture}\label{fig:rps_payload}
\end{figure}

\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/tail99_clients}
    \end{center}
    \caption{Experemenal software architecture}\label{fig:tail99_clients}
\end{figure}

\begin{figure}[!hbt]
    \begin{center}
        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/tail99_payload}
    \end{center}
    \caption{Experemenal software architecture}\label{fig:tail99_payload}
\end{figure}

\subsection{Threads to validity}
\subsection{Conclusion}

### 2021-11-02 chapters/execution_environement/execution_environement.tex
adding : grpc

\newpage

\chapter{Execution enviroment}
\label{chapter:enviroment}
After studying the impact of the programming languages choice on the energy consumption of softwares. we wanted to dig in more deeper. Therefore we have chosen two of the most popular programming lanaguages \textbf{Python} and \textbf{Java}
In this chapter we will reduce the energy consumption of those following lanaguages by doing some tuning
\import{chapters/execution_environement/java}{java}

### 2022-05-03 chapters/execution_environement/java/java_article.tex
add : python

\begin{abstract}
    \vspace{0pt}\noindent\textbf{Background.}
    The \emph{Java Virtual Machine} (JVM) platforms have known multiple evolutions along the last decades to enhance both the performance they exhibit and the features they offer.
    With regards to energy consumption, few studies have investigated the energy consumption of code and data structures.
    Yet, we keep missing an evaluation of the energy efficiency of existing JVM platforms and an identification of the configurations that minimize the energy consumption of software hosted on the JVM.

    \vspace{1pt}\noindent\textbf{Aims.}
    The purpose of this paper is to investigate the variations in energy consumption between different JVM distributions and parameters to help developers configuring the least consuming environment for their Java application.

    \vspace{1pt}\noindent\textbf{Method.}
    We thus assess the energy consumption of some of the most popular and supported JVM platforms using 12 Java benchmarks that explore different performance objectives.
    Moreover, we investigate the impact of the different JVM parameters and configurations on the energy consumption of software.

    \vspace{1pt}\noindent\textbf{Results.}
    Our results show that some JVM platforms can exhibit up to 100\% more energy consumption.
    JVM configurations can also play a substantial role to reduce the energy consumption during the software execution.
    Interestingly, the default configuration of the garbage collector was energy efficient in only 50\% of our experiments.

    \vspace{1pt}\noindent\textbf{Conclusion.}
    Finally, we provide an OSS tool, named \textsf{J-Referral} that recommends an energy-efficient JVM distribution and configuration for any Java application.
\end{abstract}

% \maketitle

\section{Introduction}\label{sec:intro}
Software services are widely deployed to support our daily activities, being mobile or hosted in the cloud.
Yet, beyond this undeniable success, the environmental impact of ICT is raising concerns and calls for solutions to reduce the energy footprint of software services~\cite{the_shift_project_lean_2019}.

Software developers often report that such solutions should come from more energy-efficient hardware components or optimized algorithms~\cite{wang_potential_2018,opaper} but, given the complexity of modern software environments, the composition of software layers makes this sustainability objective particularly challenging.

Given this context, this paper more specifically investigates the impact of one of these layers, the runtime environment and its settings, on the energy consumption of an hosted software service.
More precisely, we aim at revealing the importance of carefully selecting and configuring the \emph{Java Virtual Machine} (JVM) to reduce the energy consumption of any software service built from any language compatible with the Java ecosystem (Java, Kotlin, Scala, Groovy, Clojure, Jython, etc.).

The empirical study we conduct in this paper reports on the energy footprint of several versions of popular JVM distributions that are freely available for download.
Beyond the choice of an appropriate runtime and its most energy-efficient version, we also consider the impact of exposed JVM settings to maximize the energy savings for a given software service.
The observations of this study aim to quantify the role played by internal JVM mechanisms, like the \emph{Just in Time} (JIT) compiler and the \emph{Garbage Collector} (GC), in the reduction of the energy consumption of hosted applications.
More formally, we formulate the following research questions:
\begin{compactenum}[\indent\bf RQ\,1:]
    \item \emph{What is the impact of existing JVM distributions on the energy consumption of Java-based software services?}
    \item \emph{What are the relevant JVM settings that can reduce the energy consumption of a given software service?}
\end{compactenum}

By answering these questions, we envision supporting application developers and administrators in the configuration of their production environment by substantially reducing the energy consumption of hosted services at large.
We also hope that our results will encourage the JVM developers to keep investing in the integration of further optimizations that can benefit a large population of software services.
This paper comes with a set of contributions that can be summarized as:
\begin{compactenum}[\indent\em (a)]
    \item Reporting on the energy-efficiency of a large panel of JVM when running acknowledged benchmarks,
    \item Identifying and assessing the key JVM settings that can influence the energy consumption of a software service,
    \item Sharing guidelines and prerequisites that will help in configuring the most energy-efficiency environment before deployment,
    \item Providing a JVM benchmarking environment to evaluate the energy-efficiency of upcoming JVM distributions and their settings,
    \item Delivering an open-source tool, named \textsf{J-Referral}, to recommend the most energy efficient JVM distribution among 85 JVMs/versions and hundreds of configurations for Java applications.
\end{compactenum}

The remainder of this paper is organized as follows.
\Cref{sec:meth} introduces the experimental protocol and methodology (hardware, projects, tools, and methodology) we adopted in this study.
\Cref{sec:exp} analyzes the results of our experiments on the energy consumption of the different JVM configurations.
\Cref{sec:rw} discusses the related works to reduce the energy consumption of Java-based Software Services.
%Finally, \Cref{sec:threats,sec:ccl} cover the validity threats and our conclusions, respectively.
Finally, \Cref{sec:ccl} covers our conclusions.

% Energy efficiency is a major concern in software engineering.
% Indeed, the huge leap in technology with all the mobile devices, IoT, cloud computing, etc. raised serious questions about the way energy is consumed. 
% In most domains, we often aim to reduce the consumption of the electronic devices/parts while improving.
% In computer science, many researches were interested in the EC optimization.
% Some of the works focused on reducing the EC at the hardware or the OS level, others by optimizing software's source codes consumption.
% In this paper, we focus on the JVM based languages, especially Java.
% Many works wave been conducted to reduce the Java and JVM based software energy consumption \cite{}.
% These works focused on many aspects of the source code such as the data collections, control structures,  code refactoring ,etc.
% Other works focused on aspects that are closer to the JVM execution such as bytecode energetic cost analysis \cite{brownlee_search_based_2017,grass_energy_2006} and the consumption that can exhibit the same source code written in many languages that can use the JVM \cite{pereira_energy_2017}.
% However, even with the increasing concern about software energetic efficiency, we still lack some basic knowloedge on how to parameter and configure the execution environment do the software can run in ideal conditions and thus consume less energy.
% Hence, our study aims at constituting a set of knowledge on the basic choices and configurations that one can apply on a JVM to reduce the energy consumption.
% Concretely, we benchmark different JVM plateforms from different providers, along with different configurations and parameters of threads management, Just In Time compiler (JIT) and Garbage Collector (GC).
% The puropose of this study is to check if some configurations can play a substantial role in reducing software energy consumption.
% In particular, if the use the default JVM with a default configuration can be a reason for energy consumption increase and how often can that be. 
% This study, therefore, aims to answer the following research questions:
% \begin{description}
% 	\item[\textbf{RQ\,1}:] What is the impact that can a JVM exhibit on the energy consumption of the running software?
% 	\item[\textbf{RQ\,2}:] How can one properly configure his JMV to to reduce software energy consumption?
% \end{description}

\section{The Java Virtual Machine}\label{sec:background}
Java was originally developed by James Gosling at Sun~Microsystems and released in 1995, before being acquired by Oracle in 2010.
One key design goal of Java is portability, which means that Java applications must run similarly on any combination of hardware and operating system with adequate runtime support.
This is achieved by compiling the Java language code to an intermediate representation, called Java \emph{bytecode}, instead of machine code.
Java bytecode instructions are analogous to machine code, but executed by a \emph{Java Virtual Machine} (JVM), which is specific to the host machine.
For example, Oracle keeps offering the \textsc{HotSpot}\,JVM, while the official reference implementation is now the \textsc{OpenJDK}\,JVM---a free and open-source software used by most developers.

Programs written in Java have a reputation for being slower and requiring more memory than those written in C++, but \emph{Just-in-Time} (JIT) compilation, embedded in the JVM, delivers a boost of performance by opportunistically compiling bytecode to machine code at runtime.
The JIT combines two compilers, C1 and C2 (also known as Client \& Server\,VM), which are triggered based on the activity of the hosted application.
Additionally, Java uses an automatic \emph{garbage collector} (GC) to manage memory in the object lifecycle and recovering the memory once objects are no longer in use.
Each JVM usually includes multiple GC, each designed to satisfy different requirements.

By default, \textsc{HotSpot} uses both C1 and C2 as tiered compilers,\footnote{\url{http://www.ittc.ku.edu/~kulkarni/teaching/EECS768/19-Spring/Idhaya_Elango_JIT.pdf}} and the \emph{Garbage-First} (G1) GC with a maximum number of GC threads limited by available CPU resources and heap size, whose initial size is set $1/64^{th}$ of physical memory and maximum size may reach up to $1/4^{th}$ of physical memory.\footnote{https://docs.oracle.com/en/java/javase/15/gctuning/ergonomics.html}

%\begin{figure}%[ht]
%	\centering
%	\includegraphics[width=\linewidth]{imgs/hotspot-jvm.pdf}
%	\caption{Overview of the HotSpot JVM key components.}
%	\label{fig:hotspot}
%\end{figure}

At the time of writing this paper, the latest JVM version is Java\,15, released in September~2020, while Java\,11 is the current \emph{Long-Term Support} (LTS) version.
As Java\,9, 10, 12, and 13 are no longer supported, Oracle advises developers to immediately transition to the latest version (currently Java\,15), or an LTS release.

Beyond \textsc{HotSpot}, one can observe that the initial JVM design leads to numerous initiatives to improve the performances of Java applications, including new hotswapping strategies---with the \emph{Dynamic Code Evolution Virtual Machine} (DCE\,VM)~\cite{DBLP:journals/scp/WurthingerWS13}---or alternative JIT---with \textsc{GraalVM}.\footnote{\url{https://www.graalvm.org}}
This also includes alternative implementations, like IBM\,J9 JVM, which is currently distributed as part of the Eclipse foundation, and known as \textsc{J9}.\footnote{https://www.eclipse.org/openj9}
Given the wide diversity of distributions and related settings, this paper aims to study the impact of the features implemented by available JVM distributions on the energy consumption of the hosted Java software services.

\section{Experimental Protocol}\label{sec:meth}
To investigate the effect that could have the JVM distribution choice and/or parameters on software energy consumption, we conducted a wide set of experiments on a cluster of machines and using several established Java benchmarks and JVM configurations.

\vspace{6pt}
\noindent\textbf{Hardware Settings.}
To report on reproducible measurements, we used the cluster \textsf{Dahu} of the G5K platform~\cite{grid5000} for most of our experiments.
This cluster is composed of $32$ identical compute nodes, which are equipped with $2$ Intel Xeon\,Gold\,6130 and 192\,GB of RAM.
Our experimental protocol enforces that the software under test is the only process executed on the node configured with a very minimal Linux Debian\,9 (4.9.0 kernel version).
The minimal OS configuration ensures that only mandatory services and daemons are kept active to conduct robust experiments and reduce the factors that can affect the energy consumption measurements during our experiments~\cite{opaper}.

\vspace{6pt}
\noindent\textbf{Energy Measurements.}
We used Intel\,RAPL as a physical power~meter to analyze the energy consumption of the CPU package and the DRAM.
RAPL is one of the most accurate tools to report on the global energy consumption of a processor~\cite{Khan:2018:RAE:3199681.3177754,10.1145/2989081.2989088}.
We note that, due to CPU energy consumption variations issues~\cite{opaper}, we used the same node for all our experiments.
% In other words, we used the same node to ran all the JVM from the different providers and with numerous configurations for each single benchmark application.
Moreover, we tried to be very careful, while running our experiments, not to fall in most common benchmarking "crimes"~\cite{crimes}.
Every single experiment, therefore, reports on energy metrics obtained from at least $20$ executions of $50$ iterations per benchmark.
All of our experiments are available for use/reproducibility from our anonymous repository.\footnote{\url{https://anonymous.4open.science/r/jvm-comparaison-213E/Readme.md}}

\vspace{6pt}
\noindent\textbf{Java Virtual Machines.}
We considered a set of $52$ JVM distributions taken from $8$ different providers/packagers mostly obtained from \textsf{SDKMAN!},\footnote{\url{https://sdkman.io/}} as listed in \Cref{tab:JVM}.
Depending on providers, either all the versions, majors, or LTS are made available by \textsf{SDKMAN!}.

\begin{table}
    \centering
    \caption{List of selected JVM distributions.}
    \label{tab:JVM}
    \resizebox{\linewidth}{!}{
        \begin{tabular}{|c|c|c|l|}
            \hline
            \textbf{Distribution} & \textbf{Provider}       & \textbf{Support} & \textbf{Selected versions}                                               \\
            \hline
            \hline
            \textsc{HotSpot}      & \textbf{Adopt\,OpenJDK} & {\sc All}        & 8.0.275, 11.0.9, 12.0.2, 13.0.2, 14.0.2, 15.0.1                          \\
            \hline
            \textsc{HotSpot}      & \textbf{Oracle}         & {\sc All}        & 8.0.265, 9.0.4, 10.0.2, 11.0.2, 12.0.2, 13.0.2, 14.0.2, 15.0.1, 16.ea.24 \\
            \hline
            \textsc{Zulu}         & \textbf{Azul\,Systems}  & {\sc All}        & 8.0.272, 9.0.7, 10.0.2, 11.0.9, 12.0.2, 13.0.5, 14.0.2, 15.0.1           \\
            \hline
            \textsc{SapMachine}   & \textbf{SAP}            & {\sc All}        & 11.0.9, 12.0.2, 13.0.2, 14.0.2, 15.0.1                                   \\
            \hline
            \textsc{Librca}       & \textbf{BellSoft}       & {\sc All}        & 8.0.275, 11.0.9, 12.0.2, 13.0.2, 14.0.2, 15.0.1                          \\
            \hline
            \textsc{Corretto}     & \textbf{Amazon}         & {\sc Mjr}        & 8.0.275, 11.0.9, 15.0.1                                                  \\
            \hline
            \textsc{HotSpot}      & \textbf{Trava\,OpenJDK} & LTS              & 8.0.232, 11.0.9                                                          \\
            \hline
            \textsc{Dragonwell}   & \textbf{Alibaba}        & LTS              & 8.0.272, 11.0.8                                                          \\
            \hline
            \textsc{OpenJ9}       & \textbf{Eclipse}        & {\sc All}        & 8.0.275, 11.0.9, 12.0.2, 13.0.2, 14.0.2, 15.0.1                          \\
            \hline
            \textsc{GraalVM}      & \textbf{Oracle}         & LTS              & 19.3.4.r8, 19.3.4.r11, 20.2.0.r8, 20.2.0.r11                             \\
            \hline
            \textsc{Mandrel}      & \textbf{Redhat}         & LTS              & 20.2.0.0                                                                 \\
            \hline
        \end{tabular}
    }
\end{table}

\begin{table*}[t]
    \centering
    \caption{List of selected open-source Java benchmarks taken from \textsc{Dacapo} and \textsc{Renaissance}.}
    \label{tab:benchmarks}
    \resizebox{\linewidth}{!}{
        \begin{tabular}{|p{0.1\linewidth}|p{0.6\linewidth}|p{0.3\linewidth}|}
            \hline
            \textbf{Benchmark}    & {\bf Description}                                                                                                    & \textbf{Focus}                             \\
            \hline
            \hline
            \textsf{ALS}          & Factorize a matrix using the alternating least square algorithm on spark                                             & Data-parallel, compute-bound               \\
            \hline
            \textsf{Avrora}       & Simulates and analyses for AVR microcontrollers                                                                      & Fine-grained multi-threading, events queue \\
            \hline
            \textsf{Dotty}        & Uses the dotty Scala compiler to compile a Scala codebase                                                            & Data structure, synchronization            \\
            \hline
            \textsf{Fj-Kmeans}    & Runs K-means algorithm using a fork-join framework                                                                   & Concurrent data structure, task parallel   \\
            \hline
            \textsf{H2}           & Simulates an SQL database by executing a TPC-C like benchmark written by Apache                                      & Query processing, transactions             \\
            \hline
            \textsf{Lusearch}     & Searches keywords over a corpus of data comprising the works of Shakespeare and the King James bible                 & Externally multi-threaded                  \\
            \hline
            \textsf{Neo4j}        & Runs analytical queries and transactions on the Neo4j database                                                       & Query Processing, Transactions             \\
            \hline
            \textsf{Philosophers} & Solves dining philosophers problem                                                                                   & Atomic, guarded blocks                     \\
            \hline
            \textsf{PMD}          & Analyzes a list of Java classes for a range of source code problems                                                  & Internally multi-threaded                  \\
            \hline
            \textsf{Reactors}     & Runs a set of message-passing workloads based on the reactors framework                                              & Message-passing, critical-sections         \\
            \hline
            \textsf{Scrabble}     & Solves a scrabble puzzle using Java streams                                                                          & Data-parallel, memory-bound                \\
            \hline
            \textsf{Sunflow}      & Renders a classic Cornell box; a simple scene comprising two teapots and two glass spheres within an illuminated box & Compute-bound                              \\
            \hline
        \end{tabular}
    }
\end{table*}

\vspace{6pt}
\noindent\textbf{Java Benchmarks.}
We ran our experiments across $12$ Java benchmarks we picked from \textsf{OpenBenchmarking.org}.\footnote{\url{https://openbenchmarking.org}}
This includes $5$ acknowledged benchmarks from the \textsc{Dacapo} benchmark suite v.\,9.12~\cite{DaCapo:paper}, namely \textsf{Avrora}, \textsf{H2}, \textsf{Lusearch}, \textsf{Sunflow} and \textsf{PMD}, that have been widely used in previous studies and proven to be accurate for memory management and computer architecture communities~\cite{DBLP:conf/wosp/LengauerBMW17,DBLP:conf/oopsla/KaliberaMJV12}.
It consists of open-source and real-world applications with non-trivial memory loads.
% 
Then, we also considered $7$ additional benchmarks from the \textsc{Renaissance} benchmark suite~\cite{renaissance,DBLP:conf/pldi/ProkopecRLD0SBZ19}, namely \textsf{ALS}, \textsf{Dotty}, \textsf{Fj-kmeans}, \textsf{Neo4j}, \textsf{Philosophers}, \textsf{Reaction} and \textsf{Scrabble}, which offers a diversified set of benchmarks aimed at testing JIT, GC, profilers, analyzers, and other tools.
The benchmarks we picked from both suites exercise a broad range of programming paradigms, including concurrent, parallel, functional, and object-oriented programming.
\Cref{tab:benchmarks} summarizes the selected benchmarks with a short description.

\section{Experiments \& Results}\label{sec:exp}
\subsection{Energy Impact of JVM Distributions}
\noindent\textbf{Job-oriented applications.}
To answer our first research question, we executed $62,400$ experiments by combining the $52$ JVM distributions with the $12$ Java benchmarks, thus reasoning on $100$ energy samples acquired for each of these combinations.
\Cref{fig:JVMs} first depicts the accumulated energy consumption of the $12$ Java benchmarks per JVM distribution and major versions (or LTS when unavailable).
Concretely, We measure the energy consumption of each of the benchmarks and compute the ratio of energy consumption compared to \textsc{HotSpot-8}, which we consider as the baseline in this experiment.
Then, we sum the ratios of the $12$ benchmarks and depict them as percentages in \Cref{fig:JVMs}.

\begin{figure*}%[ht]
    \centering
    \includegraphics[width=\linewidth]{imgs/alljvms_chetemi8_baseon8.pdf}
    \caption{Energy consumption evolution of selected JVM distributions along versions.}
    \label{fig:JVMs}
\end{figure*}

One can observe that, along with time and versions, the energy efficiency of JVM distributions tends to improve (10\% savings), thus demonstrating the benefits of optimizations delivered by the communities.
Yet, one can also observe that energy consumption may differ from one distribution to another, thus showing that the choice of a JVM distribution may have a substantial impact on the energy consumption of the deployed software services.
For example, one can note that \textsc{J9} can exhibit up to 15\% of energy consumption overhead, while other distributions seem to converge towards a lower energy footprint for the latest version of Java.
As \textsc{GraalVM} adopts a different strategy focused on LTS support, one can observe that its recent releases provide the best energy efficiency for Java\,11, but recent releases of other distributions seem to reach similar efficiency for Java\,13 and above, which are recent versions not supported by \textsc{GraalVM} yet.

Interestingly, this convergence of distributions has been observed since Java\,11 and coincides with the adoption of DCE\,VM by \textsc{HotSpot}.
Ultimately, 3 clusters of JVMs that encompass JVMs with similar energy consumption can be seen through \Cref{fig:JVMs}: \textsc{J9}, the \textsc{HotSpot} and its variants, and \textsc{GraalVM}.
Additional detailed figures to illustrate the evolution of energy consumption per benchmark/JVM are made available from the anonymous repository.\footnote{\url{https://anonymous.4open.science/r/jvm-comparaison-213E/Readme.md}}

Then, Figure~\ref{hotspot8} depicts the evolution of the energy consumption of the 12 benchmarks, when executed on the \textsc{HotSpot} JVM.
\Cref{fig:hotspot8} reports on the energy consumption variation of individual benchmarks, using to \textsc{HotSpot-8}\ as the baseline.
Our results show that the JVM version can severely impact the energy consumption of the application.
However, unlike \Cref{fig:JVMs}, one can observe that, depending on applications, latest JVM versions can consume less energy (60\% less energy for \textsf{Scrabble}) or more energy (25\% more energy for the \textsf{Neo4J}).
It is worth noticing that the energy consumption of some benchmarks, such as \textsf{Reactors}, exhibit large variations across JVM versions due to experimental features and changes that are not always kept when releasing LTS versions (version 11 here).
For example, the introduction of \texttt{VarHandle} to allow low-level access to the memory order modes available in JDK 9 and work along \texttt{Unsafe Classe} that was removed from from JVM 11.\footnote{\url{https://blogs.oracle.com/javamagazine/the-unsafe-class-unsafe-at-any-speed}}

\begin{figure*}%[ht]
    \centering
    \includegraphics[width=\linewidth]{imgs/lineplothotspot_chetemi8_baseon8}
    \caption{Energy consumption of the HotSpot JVM along versions.}
    \label{fig:hotspot8}
\end{figure*}

% The purpose of this experiment is to check whether the JVM choice can affect the energy consumption of software.
% Yet, the purpose is also to retrieve some mainline information on the JVMs in order to achieve a better energy efficiency.
% Figures \ref{avrora} and \ref{scrabble} depict examples of the energy consumption end execution time of the Avrora and ALS benchmarks respectively.
% We noticed that many of the tested JVMs exhibit similar behaviors and consumption due to the high similarities between these versions.
% In fact, many versions share the same JVM core, and the differences between them are minor (sometimes just packaging changes from a constructor to another) to have any substantial impact in performance or energy consumption.

% \begin{figure*}%[ht]
% 	\begin{subfigure}{\textwidth}
% 		\includegraphics[width=\linewidth]{imgs/energy_pkg_all_avrorachetemi-8.pdf}
% 		\centering
% 		\captionsetup{justification=centering}
% 		\caption{Energy consumption of the Avrora benchmark using the 52 JVMs}\label{fig:avrora_energy}
% 	\end{subfigure}
% 	\begin{subfigure}{\textwidth}
% 		\includegraphics[width=\linewidth]{imgs/execution_time_all_avrorachetemi-8}
% 		\centering
% 		\captionsetup{justification=centering}
% 		\caption{Execution time of the Avrora benchmark using the 52 JVMs}\label{fig:avrora_time}
% 	\end{subfigure}
% 	\caption{Energy consumption and execution time of the Avrora benchmark using the 52 JVMs}
% 	\label{fig:avrora}
% \end{figure*}

% 	\begin{figure*}%[ht]
% 	\begin{subfigure}{\textwidth}
% 		\includegraphics[width=\linewidth]{imgs/energy_pkg_all_scrabblechetemi-8.pdf}
% 		\centering
% 		\captionsetup{justification=centering}
% 		\caption{Energy consumption of the Scrabble benchmark using the 52 JVMs}\label{fig:scrabble_energy}
% 	\end{subfigure}
% 	\begin{subfigure}{\textwidth}
% 		\includegraphics[width=\linewidth]{imgs/execution_time_all_scrabblechetemi-8}
% 		\centering
% 		\captionsetup{justification=centering}
% 		\caption{Execution time of the Scrabble benchmark using the 52 JVMs}\label{fig:scrabble_time}
% 	\end{subfigure}
% 	\caption{Energy consumption and execution time of the Scrabble benchmark using the 52 JVMs}
% 	\label{fig:scrabble}
% \end{figure*}

% We applied those comparison experiments on the whole benchmark set.
% The full results can be found through the anonymous link[footnote].
% Hence, Only 2 JVMs  gave energy consumption results that are different enough to be distinguished from the OpenJDK (GraalVM and J9).

Given that the wide set of distributions and versions seems to highlight 3 classes of energy behaviors, the remainder of this paper considers the following distributions as relevant samples of JVM to be further evaluated: \textsf{20.2.0.r11-grl} (\textsc{GraalVM}), \textsf{15.0.1-open} (\textsc{HotSpot-15}), \textsf{15.0.21.j9} (\textsc{J9}).
We also decided to keep the \textsf{8.0.275-open} (\textsc{HotSpot-8}) as a baseline JVM for some figures to highlight the evolution of energy consumption over time/versions.

\Cref{fig:all_benchs} further explores the comparison of energy efficiency of the JVM distributions per benchmark.
One can observe that, depending on the benchmark's focus, the energy efficiency of JVM distributions may strongly vary.
When considering individual benchmarks, \textsc{J9} performs the worst for at least 6 out of 12 benchmarks---\emph{i.e.}, worst ratio among the 4 tested distributions.
Even though, \textsc{J9} can still exhibit a significant energy saving for some benchmarks, such as \textsf{Avrora}, where it consumes 38\% less energy than \textsc{HotSpot} and others.

\begin{figure*}%[ht]
    \includegraphics[width=.87\linewidth]{imgs/bar_plot_all_basedon8}
    \centering
    \captionsetup{justification=centering}
    \caption{Energy consumption comparison across Java benchmarks for \textsc{HotSpot}, \textsc{GraalVM} \& \textsc{J9}.}
    \label{fig:all_benchs}
\end{figure*}

Interestingly, \textsc{GraalVM} delivers good results overall, being among the distributions with a low energy consumption for all benchmarks, except for \textsf{Reactors} and \textsf{Avrora}.
Yet, some differences still can be observed with \textsc{HotSpot} depending on applications.
The newer version of \textsc{HotSpot-15} was averagely good and, compared to \textsc{HotSpot-8}, it significantly enhances energy consumption for most scenarios.
Finally, \textsf{Neo4J} is the only selected benchmark where \textsc{HotSpot-8} is more energy efficient than \textsc{HotSpot-15}.

% We can clearly see through the previous figures \ref{avrora} and \ref{scrabble} that choosing OpenJDK, J9 or GraalVM can have a substantial impact the energy consumption and the execution time.
% % also it is still wildy used 
% In figure \ref{avrora}, the consumed energy to run the Avrora benchmark is 44\% less when using the J9 JVM.
% Moreover, we can even see that Avrora executions are running faster on J9 compared to the other platforms .
% In figure\ref{scrabble}, It is noticeable that the GraalVm with version 11 of Java is running 10\% faster and consuming 15\% less energy compared to other JVMs.

% These results can also be confirmed through many of the other benchmarks as each JVM can exhibit a better energy efficiency for some benchmarks and/or experiments.[ref to the footnote again].
% The results showed that we still can distinguish the 3 main platforms (OpenJDK, J9 and Graal) where noticeable energy consumption differences can be detected.
% Figure \ref{all_benchs} illustrates the ratio of consumed energy for the J9 and GraalVM platforms compared to OpenJDK 15.0.1.
% It clearly shows that the three JVM platforms rarely consume the same energy when running the same task,
% A platform can exhibit the best or worst performance/energy consumption depending the benchmark we tested.
% Overall, and among the 12 benchmarks we tested, GraalVM seemed to be the most interesting JVM energy-consumption wise.
% In fact, GraalVm is the most energy efficient in 5 of the 12 benchmarks, by up to 40\% for the Lusearch benchmark compared to OpenJDK.
% J9 on the other hand recorded the worst energy consumption for 7 of the 12 benchmarks. but gave some interesting results with ~45\% and ~25\% improvement for the Avrora and Pmd benchmarks respectively.
% This endorses the further investigation that has to be done in order to determine the different parameters that can contribute to this difference.
% While openJDK~15 and OpenJDK~8 often show close energy consumption (9 of the 12 benchmarks).
% However the oldest version of OpenJDK can consume half the energy consumption for some cases such as Neo4J, and up to ~80\% and ~130\% more for some other cases such as Reactors an Scrabble respectively.
% This proves that the later versions of OpenJDK doesn't essentially outperforms the older ones.

% \subsubsection{Energy Impact of JVM Distributions on services}
\vspace{6pt}
\noindent\textbf{Service-oriented applications.}
% The previous experiments have been conducted on deterministic jobs where we measured the energy consumption of many runs.
In this section, instead of considering bounded execution of benchmarks, we run the same benchmarks as services for 20 minutes, and we compare the average power and total requests processed by each of the 3 JVM distributions.
Globally, the results showed that the average power when using \textsc{GraalVM}, \textsc{HotSpot}, and \textsc{OpenJ9} is often equivalent and stable over time.
This means that the energy efficiency observed for some JVM distributions with Job-oriented applications is mainly related to shorter execution times, which incidentally results in energy savings.
Nonetheless, we can highlight two interesting observations for two benchmarks whose behaviors differ from others.
First, the analysis of the \textsf{Scrabble} benchmark experiments showed that, in some scenarios, some JVMs can exhibit different power consumptions.
\Cref{fig:servicescrabble} depicts the power consumed by the 3 JVM distributions for the \textsf{Scrabble} benchmark.
One can clearly see that \textsc{GraalVM} requires an average power of 109\,W, which is 9\,W higher than \textsc{HotSpot-15} and 15\,W higher than \textsc{J9}.
% This average power consumption is stable over time.
When it comes to the number of requests processed by \textsf{Scrabbles} during that same amount of time, \textsc{GraalVM} completes $5,336$ requests, against $3,595$ for \textsc{HotSpot} and $2,603$ for \textsc{J9}, as shown in \Cref{table:power/request}.
The higher power usage for \textsc{GraalVM} helped in achieving a high amount of requests, but also the fastest execution of every, request which was 40\% faster on \textsc{GraalVM}.
Thus, \textsc{GraalVM} was more energy efficient, even if it uses more power, which confirms the results observed in \Cref{fig:all_benchs} for this benchmark.

\begin{table}
    \centering
    \caption{Power per request for \textsc{HotSpot}, \textsc{GraalVM} \& \textsc{J9}.}
    \label{table:power/request}
    \small
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \textbf{Benchmark}            & \textbf{JVM}     & \textbf{Power\,(P)} & \textbf{Requests\,(R)} & $P/R\,\times\,10^{-3}$ \\
        \hline
        \hline
        \multirow{3}{*}{\sf Scrabble} & \textsc{GraalVM} & 109\,W              & \best 5,336\,req       & \best 20\,mW           \\
        \cline{2-5}
                                      & \textsc{HotSpot} & 98\,W               & 3,595\,req             & 27\,mW                 \\
        \cline{2-5}
                                      & \textsc{J9}      & \best 92\,W         & 2,603\,req             & 35\,mW                 \\
        \hline
        \multirow{3}{*}{\sf Dotty}    & \textsc{GraalVM} & \best 45\,W         & 510\,req               & 88\,mW                 \\
        \cline{2-5}
                                      & \textsc{HotSpot} & \best 45\,W         & \best 597\,req         & \best 75\,mW           \\
        \cline{2-5}
                                      & \textsc{J9}      & 46\,W               & 381\,req               & 120\,mW                \\
        \hline
    \end{tabular}
\end{table}

\begin{figure*}%[ht]
    \includegraphics[width=.9\linewidth]{imgs/powers_chetemi-2-scrabble.pdf}
    \centering
    \captionsetup{justification=centering}
    \caption{Power consumption of \textsf{Scrabble} as a service for \textsc{HotSpot}, \textsc{GraalVM} \& \textsc{J9}.}
    \label{fig:servicescrabble}
\end{figure*}

The second interesting situation was observed on the \textsf{Dotty} benchmark.
More specifically, during the first 100 seconds of the execution of the \textsf{Dotty} benchmark on all evaluated JVMs.
At the beginning of the execution, \textsc{GraalVM} has a slightly lower power consumption, is faster, and consumes 10\% less energy.
After about 150 seconds, the power differences between the 3 JVMs is barely noticeable.
One can, however, notice the effect of the JIT, as \textsc{HotSpot} takes the advantage over \textsc{GraalVM} and becomes more energy efficient.
In total, \textsc{HotSpot} completes $597$ requests against $510$ for \textsc{GraalVM} and $381$ for \textsc{J9}, as shown in \Cref{table:power/request}.
\textsc{HotSpot} was thus the best choice on the long term, which explains why it is always necessary to consider a warm-up phase and wait for the JIT to be triggered before evaluating the effect of the JVM or the performance of an application.
This is exactly what we did in our experiments, and why \textsc{HotSpot} was more energy efficient than \textsc{GraalVM} in \Cref{fig:all_benchs}, thus ignoring the warm-up phase would have misleading.

\begin{figure*}%[ht]
    \includegraphics[width=.9\linewidth]{imgs/powers_chetemi-2-dotty.pdf}
    \centering
    \captionsetup{justification=centering}
    \caption{Power consumption of \textsf{Dotty} as a service for \textsc{HotSpot}, \textsc{GraalVM} \& \textsc{J9}.}
    \label{fig:servicedotty}
\end{figure*}
\bigbreak
\begin{mdframed}[]
    To answer \textbf{RQ\,1}, we conclude that---while most of the JVM platforms perform similarly---we can cluster JVMs in 3 classes: \textsc{HotSpot}, \textsc{J9}, and \textsc{GraalVM}.
    The choice of one JVM of these classes can have a major impact on software energy consumption, that strongly depends on the application context.
    When it comes to the JVM version, latest releases tend to offer the lowest power consumption, but experimental features should be carefully configured, thus further questioning the impact of JVM parameters.
\end{mdframed}

\subsection{Energy Impact of JVM Settings}
The purpose of our study is not only to investigate the impact of the JVM platform on the energy consumption, but also the different JVM parameters and configurations that might have a positive or negative effect, with a focus on 3 available settings: multi-threading, JIT, and GC.

\subsubsection{Multithreading}
The purpose behind this phase is to investigate the impact JVM thread management strategies on the energy consumption.
This encompasses exploring if the management strategies of application-level parallelism (so called \emph{threads}) results in different energy efficiencies, depending on JVM distributions.

Investigating such an hypothesis requires a selection of highly parallel and CPU-intensive benchmarks, which is one of the main criteria for our benchmark selection.
As no tool can accurately monitor the energy consumption at a thread level, we monitor the global power consumption and CPU utilization during the execution using RAPL for the energy, and several Linux tools for the CPU-utilization (\texttt{htop}, \texttt{cpufreq}).
%The experiments showed differences in energy consumption between the 3 JVMs and even between 2 version on the HotSpot VM.
Knowing that most of the benchmarks are multi-threaded jobs that use multiple cores, further analysis of thread management is required to understand the results of our previous experiments.
We thus selected the benchmarks that highlighted the highest differences along JVM distributions from \Cref{fig:all_benchs}, namely \textsf{Avrora} and \textsf{Reactors}.
We studied their multi-threaded behavior to optimize their energy efficiency.

\begin{figure*}%[ht]
    \begin{subfigure}{\textwidth}
        \includegraphics[width=.9\linewidth]{imgs/active_threads_avrora_chetemi_3}
        \centering
        \captionsetup{justification=centering}
        \caption{Active threads of \textsf{Avrora} when using \textsc{HotSpot}, \textsc{GraalVM}, or \textsc{J9}.}
        \label{fig:avrora_threads}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
        \includegraphics[width=.9\linewidth]{imgs/active_threads_reactors_chetemi_3}
        \centering
        \captionsetup{justification=centering}
        \caption{Active threads of \textsf{Reactors} when using \textsc{HotSpot}, \textsc{GraalVM}, or \textsc{J9}.}
        \label{fig:reactors_threads}
    \end{subfigure}
    %	\begin{subfigure}{\textwidth}
    %		\includegraphics[width=.9\linewidth]{imgs/active_threads_scrabble_chetemi_3.pdf}
    %		\centering
    %		\captionsetup{justification=centering}
    %		\caption{Active threads of \textsf{Scrabble} when using \textsc{HotSpot}, \textsc{GraalVM}, or \textsc{J9}}
    %		\label{fig:scrabble_threads}
    %	\end{subfigure}
    \caption{Active threads evolution when using \textsc{HotSpot}, \textsc{GraalVM}, or \textsc{J9}.}
    \label{fig:threads}
\end{figure*}

\Cref{fig:threads} delivers a closer look to the thread allocation strategies adopted by JVM.
First, \Cref{fig:avrora_threads} illustrates the active threads count evolution over time (excluding the JVM-related threads, usually 1 or 2 extra threads depending on the execution phase) for \textsf{Avrora}.
One can notice through the figure that \textsc{J9} exploits the CPU more intensively by running much more parallel threads compared to other JVMs (an average of 5.1 threads per second for \textsc{J9} while the other JVMs do not exceed 1.5 thread per second).
Furthermore, the number of context switches is twice bigger for \textsc{J9}, while the number of soft page faults is twice smaller.
The efficient \textsc{J9} thread management explains why running the \textsf{Avrora} benchmark took much less time and consumed less energy, given that no other difference for the JIT or GC configuration was spotted between the JVMs.
Another key reasons of the \textsc{J9}'s efficiency for the \textsf{Avrora} benchmark is memory allocation, as \textsf{OpenJ9} adopts a different policy for the heap allocation.
It creates a non-collectable \emph{thread local heap} (TLH) within the main heap for each active thread.
The benefit of cloning a dedicated TLH is the fast memory access for independent threads: each thread has its own heap and no deadlock can occur.

The second example in \Cref{fig:reactors_threads} depicts the active threads evolution over time of the \textsf{Reactors} benchmark.
In this case, all the JVMs have a close average of threads per second.
Nevertheless, one can still observe that \textsc{HotSpot-15} and \textsc{J9} keep running faster, which confirms the results of \Cref{fig:all_benchs}, where both JVMs consume much less energy compared to \textsc{GraalVM} and \textsc{HotSpot-8}.
This difference in energy consumption between benchmarks can be less likely caused by thread management for the \textsf{Reactors} benchmark, as \textsc{HotSpot-8} reports on a higher average of active threads.
However, the TLH mechanism was not as efficient as for the \textsf{Avrora} benchmark, as dedicating a heap for each thread can also cause some extra memory usage for data duplication and synchronization, especially if a lot of data is shared between threads.

%In the case of the \textsf{Scrabble} benchmark illustrated in \Cref{fig:scrabble_threads}, one can see that \textsc{GraalVM} executed the benchmark much faster, and with even less threads.
%With only 5.2 threads/sec, \textsc{GraalVM} was able to be 50\% faster than \textsc{HotSpot-8} and \textsc{J9} and consuming the least energy, while the other JVMs ran more threads per second.
% This advocates that threads management was not the key factor here, thus requiring to further investigate other JVM settings. 
% \textsc{GraalVM} was faster and more energy efficient due to other JVM optimizations.

In conclusion, JVMs thread management can sometimes constitute a key factor that impacts software energy consumption.
However, we suggest to check and compare JVMs before deploying a software, especially if the target application is parallel and multi-threaded.

\subsubsection{Just-in-Time Compilation}
The purpose of experiments on JIT is to highlight the different strategies that can impact software energy consumption within a JVM and between JVMs.
We identified a set of JIT compiler parameters for every JVM platform.

For \textsc{J9}, we considered fixing the intensity of the JIT compiler at multiple levels (\textsf{cold}, \textsf{warm}, \textsf{hot}, \textsf{veryhot}, and \textsf{scorching}).\footnote{\url{[https://www.eclipse.org/openj9/docs/jit/]}}
The hotter the JIT, the more code optimization to be triggered.
We also varied the minimum count method calls before a JIT compilation occurs (\textsf{10}, \textsf{50}, \textsf{100}), and the number of JIT instances threads (from \textsf{1} to \textsf{7}).
For \textsc{Hotspot-15}, we conducted experiments while disabling the tiered complication (that generates compiled versions of methods that collect profiling information about themselves), and we also varied the JIT maximum compilation level from \textsf{0} to \textsf{4}, we also tried out \textsc{HotSpot} with a basic \textsc{GraalVM} JIT.
We note that the level 0 of JIT compilation only uses the interpreter, with no real JIT compilation.
Levels 1, 2, and 3 use the C1 compiler (called client-side) with different amount of extra tuning.
% Furthermore, the level 3 performs a full profiling.
The JIT C2 (also called server-side JIT) compiler only kicks-in at level 4.

For \textsc{GraalVM}, we conducted experiments with and without the JVMCI (a Java-based JVM compiler interface enabling a compiler written in Java to be used by the JVM as a dynamic compiler).
We also considered both the community and economy configurations (no enterprise).
A JIT+AOT (\emph{Ahead Of Time}) disabling experiment has also been considered for all of the 3 JVM platforms.
\Cref{table:JIT} reports on the energy consumption of the experiments we conducted for most of the benchmarks and JIT configurations under study.

\input{table-jit.tex}

The $p$-values are computed with the mann-whitney test, with a null hypothesis of the energy consumption being equal to the default configuration.
The $p$-values in bold show the values that are significantly different from the default configuration with a 95\% confidence, where the values in green highlight the strategies that consumed significantly less energy than default (less energy and significant $p$-value).

%The results show that none of the 3 JVMs has the best or worst performance for all benchmarks.

For \textsc{J9}, we noticed that adopting the default JIT configuration is always better than specifying a custom JIT intensity.
The \textsf{warm} configuration delivers the closest results to the best results observed with the default configuration.
Moreover, choosing a low minimum count of method calls seems to have a negative effect on the execution time and the energy consumption.
The only parameter that can give better performance than the default configuration in some cases is the number of parallel JIT threads---using 3 and 7 parallel threads---but is not statistically significant.

For \textsc{GraalVM}, the default community configuration is often the one that consumes the least energy.
Disabling the JVMCI can---in some cases---have a benefit (16\% of energy consumption reduction for the \textsf{H2} benchmark), but still gave overall worst results (80\% more energy consumption for the \textsf{Neo4J} benchmark).
In addition, switching the economy version of the \textsc{GraalVM} JIT often results in consuming more energy and delaying the execution.

For \textsc{HotSpot}, keeping the default configuration of the JIT is also mostly good.
In fact, the usage of the C2 JIT is often beneficial (JIT level 4) in most cases, while using the \textsc{GraalVM} JIT reported similar energy efficiency.
Yet, some benchmarks showed that using only the C1 JIT (JIT level 1) is more efficient and even outperforms the usage of the C2 compiler.
10\% on \textsf{Avrora} and 30\% on \textsf{Pmd} are examples of energy savings observed by using the C1 compiler.
However, being limited to the C1 compiler can also cause a huge degradation in energy consumption, such as 32\% and 34\% of additional energy consumed for the \textsf{Dotty} and \textsf{FJ-kmeans} benchmarks, respectively.
Hence, if it is a matter of not using the C2 JIT, the experiments have shown that the level 1 JIT is always the best, compared to levels 2 or 3 that also use the C1 JIT, but with more options, such as code profiling that impacts negatively the performance and the energy efficiency.
Level 0 JIT compilation should never be an option to consider.
No $p$-value has been computed for Level 0, due to the limited amount of iterations executed with this mode (very high execution time, clearly much more consumed energy).

Globally, we conclude through these experiments that keeping the default JIT configuration was more energy efficient in ~80\% of our experiments and for the 3 classes of JVMs.
This advocates that using the default JIT configuration that can often deliver near-optimal energy efficiency.
Although, some other configurations, such as using only the C1 JIT or disabling the JVMCI could be advantageous in some cases.

\subsubsection{Garbage Collection}
Changing or tuning the GC strategy has been acknowledged to impact the JVM performances~\cite{10.1145/2568088.2568097}.
To investigate if this impact also benefits to energy consumption, we conducted a set of experiments on the selected JVMs.
We considered different garbage collector strategies with a limited memory quantity of 2\,GB, and recorded the execution time and the energy consumption.
The tested GC strategies options mainly vary between \textsc{J9} and the other 2 JVMs, as detailed in \Cref{table:GCJ9}.

\begin{table}
    \centering
    \caption{The different \textsc{J9} GC policies}
    \label{table:GCJ9}
    \small
    \begin{tabular}{|p{0.2\columnwidth}|p{0.7\columnwidth}|}
        \hline
        \textbf{Policy}              & \textbf{Description}                                                                                                         \\
        \hline
        \hline
        \textsf{Balanced}            & Evens out pause times \& reduces the overhead of the costlier operations associated with GC                                  \\
        \hline
        \textsf{Metronome}           & GC occurs in small interruptible steps to avoid stop-the-world pauses                                                        \\
        \hline
        \textsf{Nogc}                & Handles only memory allocation \& heap expansion, with no memory reclaim                                                     \\
        \hline
        \textsf{Gencon}\,(default)   & Minimizes GC pause times without compromising throughput, best for short-lived objects                                       \\
        \hline
        \textsf{Concurrent Scavenge} & Minimizes the time spent in stop-the-world pauses by collecting nursery garbage in parallel with running application threads \\
        \hline
        \textsf{optthruput}          & Optimized for throughput, stopping applications for long pauses while GC takes place                                         \\
        \hline
        \textsf{Optavgpause}         & Sacrifices performance throughput to reduce pause times compared to \textsf{optthruput}                                      \\
        \hline
    \end{tabular}
\end{table}

For \textsc{HotSpot} and \textsc{GraalVM}, we also considered many GC policies, as described in \Cref{table:GCHS}.
Furthermore, other GC settings have also been tested for all JVM platforms, such as the \emph{pause time}, the \emph{number of parallel threads} and \emph{concurrent threads} and \emph{tenure age}.

\begin{table}
    \centering
    \caption{The different \textsc{HotSpot}/\textsc{GraalVM} GC policies}
    \label{table:GCHS}
    \small
    \begin{tabular}{|p{0.2\columnwidth}|p{0.7\columnwidth}|}
        \hline
        \textbf{Policy}          & \textbf{Description}                                                                                                                 \\
        \hline
        \hline
        \textsf{G1GC}\,(default) & Uses concurrent \& parallel phases to achieve low-pauses GC and maintain good throughput                                             \\
        \hline
        \textsf{SerialGC}        & Uses a single thread to perform all garbage collection work (no threads communication overhead)                                      \\
        \hline
        \textsf{ParallelGC}      & Known as throughput collector: similar to \textsf{SerialGC}, but uses multiple threads to speed up garbage collections for scavenges \\
        \hline
        \textsf{parallelOldGC}   & Use parallel garbage collection for the full collections, enabling it automatically enables the \textsf{ParallelGC}                  \\
        \hline
    \end{tabular}
\end{table}

\input{table-gc.tex}

\Cref{table:GC} summarizes the results of all the tested GC strategies with our selected benchmarks and the $p$-values of the mann-whitney test, with a null hypothesis of the energy consumption being equal to the default configuration with a 95\% confidence.
The $p$-values in bold show the values that are significantly different from the default configuration, where the values in green highlight the strategies that consumed significantly less energy than default.
For \textsc{GraalVM}, one can see that the GC default configuration is efficient in most experiments, compared to other strategies.
The main noticeable impact is related to the \textsf{ParallelGC} and \textsf{ParallelOldGC}.
In fact, the \textsf{ParallelGC} can be 13\% more energy efficient in some applications with a significant $p$-value, such as \textsf{Reactors}, compared to default.
However, the same GC strategy can cause the software to consume twice times more, as for the \textsf{Neo4j} benchmark, due to the high communications between the GC threads, and the fragmentation of the memory.

For \textsc{J9}, the default \textsf{Gencon} GC causes the software to report an overall good energy efficiency among the tested benchmarks.
However, other GC can cause better or worse energy consumption than \textsf{Gencon} depending on workloads.
Using the \textsf{Metronome} GC consumes ~35\% less energy for the \textsf{ALS} benchmark and ~17\% less energy for the \textsf{Sunflow} benchmark, but it also consumes twice energy for the \textsf{Neo4j} benchmark and 28\% more energy for \textsf{Reactors}.
The reason is that \textsf{Metronome} occurs in small preemptible steps to reduce the GC cycles composed of many GC quanta.
This suits well for real-time applications and can be very beneficial when long GC pauses are not desired, as observed for \textsf{ALS}.
However, if the heap space is insufficient after a GC cycle, another cycle will be triggered with the same ID.
As \textsf{Metronome} supports class unloading in the standard way, there might be pause time outliers during GC activities, inducing a negative impact on the \textsf{Neo4j} execution time and energy consumption.

The same goes for the \textsf{Balanced} GC that tries to reduce the maximum pause time on the heap by dividing it into individually managed regions.
The \textsf{Balanced} strategy is preferred to reduce the pause times that are caused by global GC, but can also be disadvantageous due to the separate management of the heap regions, such as for \textsf{ALS} where it consumed about three times the energy consumption, compared to the default \textsf{Gencon} GC.
On the other hand, the \textsf{Optthruput} GC, which stops the application longer and less frequently, gave very good overall results and sometimes even outperformed the \textsf{Gencon} GC by a small margin.
Other JVM parameters, such as the \textsf{ConcurrentScavenge} or \textsf{noAdaptiveTenure} did not have a substantial impact during our experiments.

Finally, the results of \textsc{HotSpot} shared similarities with \textsc{GraalVM}.
The \textsf{ParallelGC} happened to give better (6\% for \textsf{Dotty}) or worst (10\% for \textsf{Neo4j}) energy efficiency compared to the default GC.
On the other hand, \textsf{ParallelOldGC} and \textsf{Serial} GC gave better results than the default G1 GC.
More specifically, the second one consumed 30\% and 6\% less energy than default GC for the \textsf{Neo4j} and \textsf{Dotty} benchmarks, respectively.
%Moreover, the Epsilon GC gave worst results compared to the default GC, and also used much more memory( 63 Gb for Epsilon GC against 4Gb for default GC).
%It even stopped working for many benchmarks due to the heap memory being saturated.
The most interesting result for \textsc{HotSpot} is the 30\% energy reduction obtained with the \textsf{Serial} GC.
This last was also more efficient on ALS (6\% less energy), compared to the default G1 GC, due to its single-threaded GC that only uses one CPU core.
%Moreover, it's single threaded GC that uses one CPU core gave it the edge by quite a decent margin of 30\% on the h2 benchmark.
%Disabling the explicit GC also trends to give better results than the default GC configurations. Same as for GraalVM, using the concurrent configuration of the default GC is often better than the parallel one or default one (up to 50\% for neo4j benchmark).

Unfortunately, we cannot convey predictive patterns on how to configure the GC to optimize energy efficiency.
However, some considerations should be taken into account when choosing the GC, such as the garbage collection time, the throughput, etc.
Other settings are less trivial to determine, such as tenure age, memory size, and GC threads count.
Experiments should thus be conducted on the software to tune the most convenient GC configuration to achieve a better energy efficiency in production.

Therefore, we noticed during our experiments that, even if using the default GC configuration ensures an overall steady and correct energy consumption, we still found other settings that reduce that energy consumption in ~50\% of our experiments.
Tuning the GC according to the hosted app/benchmark is thus critical to reduce the energy consumption.

\bigbreak
\begin{mdframed}[]
    To answer \textbf{RQ\,2}, we conclude that users should be careful while choosing and configuring the garbage collector as substantial energy enhancements can be recorded from a configuration to another.
    The default GC consumed more energy than other strategies in most of the situations.
    % choosing the most convenient GC strategy in thus very important to achieve a better software energy efficiency. 
    However, keeping the default JIT parameters often delivers near-optimal energy efficiency.
    In addition, the JVM platforms can handle differently multi-threaded applications and thus consume a different amount of time/energy.
    Dedicated performance tuning evaluations should therefore be conducted on such software to identify the most energy-efficient platform and settings.
\end{mdframed}

\subsubsection{J-Referral}
To help developers and practitioners choosing an energy efficient JVM distribution and configuration, we propose a tool named \textsf{J-Referral}.
This tool takes a Java application as an input and a launch script to tune the JVM configuration.
The recommendation is is based on the 85 JVM distributions and versions made available by \textsf{SDKMAN!}.\footnote{\url{https://sdkman.io/}}
Beyond the JVM distribution and version recommendation, \textsf{J-Referral} automatically explores many GC and JIT options for each distribution.
Hundreds of combinations are thus compared to recommend an energy efficient JVM distribution and the associated configuration parameters.

\Cref{table:j-referral} illustrates an example of the final report of \textsf{J-Referral}.
The tool was tested for 2 real Java projects: \textsf{Zip4J} and\footnote{\url{https://github.com/srikanth-lingala/zip4j}} and \textsf{K-nucleotide}.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/performance/knucleotide.html}}.
\textsf{Zip4J} runs a large file compression, while \textsf{K-nucleotide}  extracts a DNA sequence, and updates a hashtable of k-nucleotide keys to count specific values.
The short report presented in \Cref{table:j-referral} shows the ratio of potential energy saving between the most and least energy consuming tested JVM (40\% and 70\% energy savings for \textsf{Zip4J} and \textsf{K-nucleotide} respectively).
Options are available for \textsf{J-Referral} to obtain much more detailed reports including execution time, DRAM usage, split DRAM vs. CPU consumption, etc.
The tool is available as \emph{open-source software} (OSS) from our anonymous repository.\footnote{\url{https://anonymous.4open.science/r/jreferral/Readme.md}}

\begin{table}
    \centering
    \caption{\textsf{J-Referral} recommendations.}
    \label{table:j-referral}
    \small
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \textbf{Project}                 & \textbf{Metric} & \textbf{Energy} & \textbf{JVM}       & \textbf{Execution flags} \\
        \hline
        \multirow{2}{*}{\textsf{Zip4J}}  & Least energy    & \best 2210\,J   & \best 16-sapmchn   & \best default            \\
        \cline{2-5}
                                         & Most energy     & 3680\,J         & 8.0.292-J9         & default                  \\
        \hline
        \hline
        \multirow{2}{*}{\textsf{K-nucl}} & Least energy    & \best 1296\,J   & \best 21.1.r16-grl & \best default            \\
        \cline{2-5}
                                         & Most energy     & 4433\,J         & 15.0.1-J9          & -Xjit:optlevel=cold      \\
        \hline
    \end{tabular}
\end{table}

\section{Related Work}\label{sec:rw}
Practitioners keep looking for development tools and methods to deliver software that meets both performance and quality requirements.
In this context, energy consumption is becoming of growing importance when deploying software services in the cloud, which generally promotes a \emph{pay-as-you-go} pricing model, thus challenging the resource-efficiency of deployed services.
So far, software engineering tackles this optimization challenge along three axes:
\begin{inparaenum}
    \item programming languages,
    \item source code,
    \item execution platforms
\end{inparaenum}

In this section, we review the state of the art of studies related to JVM energy consumption.

\vspace{4pt}\noindent\textbf{At code level.}
Many works investigated software energy consumption efficiency through source code changes and optimizations.
For example, \cite{pinto_comprehensive_2016,fernandes_assisting_2017} studied the effect of Java collections on energy consumption, with regards to the collection size and/or the most executed tasks on the collection (insertion, removal, search).
In~\cite{samir_hasan_energy_2016}, Hasan~\textit{et~al.} compared the energy consumption of several Java data structures, analyzing the bytecode using the Wala framework\footnote{\url{http://wala.sourceforge.net/wiki/index.php}} and assessing the evolution of the energy consumption in different scenarios (insertion at the beginning, iteration, etc.).
They also used some automated replacement of \texttt{LinkedList} and \texttt{ArrayList} to simulate best- and worst-case energy consumption scenarios on real production applications.
Their study showed that using inappropriate collection can cause an energy consumption inefficiency of 300\% for the worst-case scenario.

SEEDS and SEEDS-API is a fully automated framework, proposed in~\cite{manotas_seeds_2014}, to analyze source code (at bytecode level for Java) and auto-tunes apps to have reduce their energy consumption, with a focus on Java collection tuning.
The authors reported on an improvement up to 17\%.
In~\cite{pereira_helping_2017}, the same authors presented SPELL, the energy leaks detector tool.
The tool uses \textsc{JRapl}~\cite{alves_experiments_2020,pinto_comprehensive_2016} to detect energy-inefficient code fragments by using a statistical spectrum-based energy red spots localization.
% It has the  advantage of being language and context independent.
Their evaluation reported up to 18\% energy savings on Java applications.

Some researchers tried to highlight the impact that has some atomic code changes on the energy consumption.
Many papers studied the impact of code refactoring on energy consumption~\cite{DBLP:conf/seke/ParkHL14}.
Other papers investigated the energy consumption of Java primitive types, operations on strings, usage of exceptions, loops, and arrays~\cite{kumar_energy_2017}.

\vspace{4pt}\noindent\textbf{At JVM level.}
In~\cite{pereira_energy_2017}, Pereira~\textit{et~al.} investigated several programming languages, including JVM-based languages like Java and Scala, and compared them on different dimensions (energy, memory, and time) using \textsf{CLBG} benchmarks.
Their observation reported that compiled languages are more energy-efficient than interpreted ones, overall.
More advanced results expose how the selected languages may satisfy one, two, or three dimensions of the equation.
%The paper also tells finding no correlation between time and energy, so couldn't affirm or deny if "faster" is "greener". 
The authors of~\cite{6122743} conducted a performance analysis and comparison between \textsc{HotSpot} and \textsc{J9}.
They claimed in their results that the relative performance of \textsc{HotSpot} ranges from 44\% to 289\% of J9, while the dynamic power consumption varies from 2.7W to 7.2W using the \textsf{SPECjvm2008} benchmarks.

The difference between \textsf{HotSpot} and \textsc{J9} was also reported in other studies.
Chiba~\textit{et~al.}~\cite{8457806} evaluated the effect that could have those 2 JVM platforms on the performance of a combination of big data query engines (\textsc{Spark} and \textsc{Tez}) using \textsf{TPC-DS} benchmark.
They reported on a 3-fold drawback that can exhibit one JVM compared to the other.

On another note, the authors of~\cite{grass_energy_2006} attempted to assess and assign an energy cost to atomic bytecode instructions together with a constant overhead of using the JVM.
Their claims are rather surprising, as the energy cost should be impacted by the execution environment, workload, core frequencies, and can hardly be fixed.
A similar idea was used in~\cite{brownlee_search_based_2017} to design a model for JVM-based software energy consumption, using a bytecode-level model.
The authors described their tool, named \textsf{OPACITOR}, as being deterministic, accurate, and robust to the surrounding noise.
However, they disable the JIT in their experiments to maintain the deterministic behavior of their tool, which does not reflect a real software execution given all the optimizations that the JVM triggers to optimize the performances.

%\section{Threats to Validity}\label{sec:threats}
%A number of issues may affect the validity of our work.
%First, we have the use of the Intel RAPL, one of the most accurate available tools to measure the energy consumption of software~\cite{Khan:2018:RAE:3199681.3177754,10.1145/2989081.2989088}.
%However, RAPL only gives the global energy consumption and no fine-grained measures at process or thread levels.
%We used baremetal hardware with a minimal OS and tuned off all the non-essential services and daemons to limit the overhead that the OS may add to the execution, even if it is not substantial~\cite{opaper}.
%
%Another measurement issue is the CPU energy variation within machines~\cite{opaper}, thus we executed all the comparable tests on the same node and with the recommended settings to mitigate this threat. 
%
%Benchmarks execution time could also constitute on more subtle threat to the validity of our work, especially for some benchmarks that run fast, such as the \textsf{Pmd} benchmark.
%We thus gave a lot of attention on how long the benchmark is running for the hardware we used, and we tuned the input data workloads to execute benchmarks for at least many (from 10 to hundreds) seconds.
%Experiments ran at least 30 times to compute the average consumption and the associated standard deviation, therefore reasoning over reasonable dispersion around the average. 
%
%How generalizable are our results? We believe that our study conclusions and guideline remain empirical, as we do not intend to generalize any result we obtained for some JVM or benchmark.
%We provide practioners with some prerequisites to check before software deployment to reduce the software energy footprint by considering the JVM and its settings.

\section{Conclusion}\label{sec:ccl}
This paper reports on an empirical investigation of the key differences in energy consumption that some of the most famous and supported JVM platforms can exhibit, in addition to the key settings that can impact this energy consumption positively or negatively.
During our experiments, we considered a total of 12 well-known and diversified-purposes Java benchmarks together with a total of 52 JVMs, including many versions of 11 different distributions.
The results of our investigations showed that many JVMs share energy efficiencies and can grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
The 3 selected JVM classes can however report a different energy efficiency for different software and/or workloads, sometimes by a large margin.
While we did not observed a unique champion when it comes to energy consumption, \textsc{GraalVM} reported the best energy efficiency for a majority of benchmarks.
Nonetheless, each JVM can achieve the best or the worst depending on the hosted application.
One cause can be thread management strategies, as observed with J9 when advantageously running \textsf{Avrora}.
Moreover, some JVM settings can cause energy consumption variations.
Our experiments showed that the default JIT compiler of the JVM is often near-optimal, in at least ~80\% of our experiments.
The default GC, however, was outperforming alternative strategies in half of our experiments, with some large gains observed when using some alternative GC depending on the application characteristics.

Our main conclusions and guidelines can be thus summarized as:
\begin{inparaenum}[\em i)]
    \item testing software on the 3 classes of JVM and identifying the one that consumes the least is a good practice, especially for multi-threading purposes,
    \item while the JVM default JIT give often good energy consumption results, some settings may improve the energy consumption and could be tested,
    \item the choice of the GC may lead to a large impact on the energy consumption in many situations, thus encouraging a careful tuning of this parameter prior to deployment.
\end{inparaenum}
To ease the integration of the above guidelines, we propose a tool, named \textsf{J-Referral}, to recommend the most energy-efficient JVM distribution and configuration among more than a hundred considered possibilities.
It establishes a full report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to help the user to choose the one with the least consumption for a Java Software.

### 2022-05-03 chapters/execution_environement/python/python.tex
add : python

\subsection{Introduction}
% reasons why we have chosen python

\subsection{python popularity}

%%%% python is a popular programming language  based TIOBE PYPI   add graphs

\subsection{environements}
in this section we will describe the chosen python interpreter

\begin{figure}[thb]
    \centering
    \includegraphics[width=\linewidth]{imgs/python-implementations-tree}
    \caption{Python interpeters}
    \label{fig:interpreters}
\end{figure}

\subsection{benchmarks}

\subsection{first results}

\subsubsection{micro-benchmarks}

\subsubsection{benchmarks}

\subsection{results and foundings}

\subsection{tools and package}

\subsection{Threads to validity}

\subsection{conclusion}

### 2022-05-06 chapters/literature/draft_citations.tex
add : citations

% this sections containes all the references for each section :

% 1. group articles by field 
hello my world!!

\section{testing}

this is \cite{arcuri_practical_2011}

difference tests  \cite{sonnenwald2003evaluating}
%further references 
\cite{lilja2005measuring}
\cite{bukh1992art}
%%%%%%%%%%%%%%%%%

list of benchmarks \cite{van_der_kouwe_benchmarking_2018}:
\begin{itemize}
    \item  Not evaluating potential performance degradation
    \item  Benchmark subsetting without proper justification
    \item  Selective data set hiding deficiencies
    \item  Microbenchmarks representing overall performance
    \item  Throughput degraded by x\% => overhead is \%
    \item  Creative overhead accounting
    \item  No indication of significance of data
    \item  Incorrect averaging across benchmark scores
    \item  Benchmarking of simplified simulated system
    \item  Inappropriate and misleading benchmarks
    \item  Same dataset for calibration and validation
    \item  No proper baseline
    \item  Only evaluate against yourself
    \item  Unfair benchmarking of competitors
    \item  Not all contributions evaluated
    \item  Only measure runtime overhead
    \item  False positives/negatives not tested
    \item  Elements of solution not tested incrementally
    \item  Missing platform specification
    \item  Missing software versions
    \item  Subbenchmarks not listed
    \item  Relative numbers only
\end{itemize}
% further references 
\cite{mytkowicz2009producing}
%%%%%%%%%%

startup performance vs steady performance \cite{buytaert_statistically_nodate}

%%%%% ** reasons behind the energy variation ****

simgrid simulation of power consumption of parallel aplication using simgrid \cite{heinrich2017predicting}

impact of manifacturing process on the energy variation \cite{coles2014comparing}

different energy consumption between identical processers [ idle and hight load ]\cite{von2016variations}

the temperature is the main raison a high energy variation \cite{wang2018potential}

external factors are impact on the energy variation \cite{mukherjee2009spatio}

the place of the server in the room doesn't affect the energy variation \cite{diouri2013your}

comparaison of three measuring tools and they did exhebit 10\% variation each time \cite{inadomi2015analyzing}

%%****** mitigating the energy variation ******
low-cost scalable variation-aware algorithm \cite{inadomi2015analyzing}

reducing the energy variation by disabling turbo boost \cite{acun2016variation}

reducing the energy consumption in parallel systems \cite{chasapis2016runtime}

\cite{marathe2017empirical}

%\\\ https://www.cpubenchmark.net/compare/Intel-i7-12700KF-vs-Intel-i7-12700F/4608vs4692 | hardware benchmarking 
\section{energy consumption}

\subsection{hardware tools}

WattsUp Pro \cite{hirst2013watts}

PowerMon \cite{bedard2010powermon}

PowerInsight \cite{laros2013powerinsight}

GreenMiner \cite{hindle2014greenminer}

powerpack \cite{ge2009powerpack}

hdeem \cite{hackenberg2014hdeem}

PDU ,RAPL ,AMD \cite{hackenberg2013power}

Intel RAPL Haswell \cite{hackenberg2015energy}

GPU TESLA \cite{burtscher2014measuring}

NVML\cite{fahad2019comparative}

\subsection{software tools}

powerapi \cite{colmant2018next}

smartwatts formula \cite{fieni2020smartwatts}

Wattwatcher \cite{lebeane2015watt}

joulemeter \cite{kothari2009joulemeter}\cite{jagroep2017energy}

jrapl \cite{guimaraes2016some} \cite{liu2015data}

joulinar \cite{islam2016measuring} \cite{noureddine2016jolinar}

jalen \cite{noureddine2015monitoring}

selfwatts \cite{fieni2021selfwatts}

powerjoular \cite{noureddine2022powerjoular}
\subsection{hybrid tools}
% split hardware tools into extern and embeded tools

\section{programming languages and performances}
\subsection{energy}

energy comparaison of programming languages in the game benchmark by debian \cite{pereira2017energy}

toward green ranking \cite{couto2017towards}

java vs kotlin in web performances \cite{bujnowski2020java}

rosetta code \cite{nanz2015comparative} \cite{mirowski2020rosetta}

network energy comparaison \cite{balasubramanian2009energy}

Aequitas \cite{ribic2016aequitas}

vm placement \cite{mishra2018energy}

\citeauthor{mishra2018energy}

static cost of Vms \cite{kurpicz2016much}

carbon footprint of training neural network \cite{strubell2019energy}

SPELL, the energy leaks detector tool \cite{pereira2017helping}

impact of energy profiling on software \cite{jagroep2017energy}

impact of maintability on software energy evolution \cite{calero2021does}

\subsection{jvm}

java vs python \cite{destefanis_statistical_2016}

hotspot vs J9 \cite{oi2011power} - same but for big querries \cite{chiba2018towards}

constant overheat of the jvm \cite{lafond2006energy}

infer the energy cost based on the byte code instructions \cite{ma2017biogeography}

java collection energy footprint  \cite{pinto2016comprehensive} \cite{fernandes2017assisting}

java classes energy footprint \cite{hasan2016energy}

framework to reduce java collections \cite{manotas2014seeds}

energy consumption of java dev tools \cite{baskar2013experimental}

Microbenchmarks on jvm \cite{longo2019reducing} \cite{baskar2013experimental}

android automatic refactoring to reduce energy consumption \cite{banerjee2016automated} \cite{rodriguez2017reducing}

java vs native c in android \cite{corral2014method}

\subsection{benchmarking}

NAS \cite{bailey_nas_nodate}

statistics \cite{he_statistics-based_2019}

benchmark crimes \cite{van_der_kouwe_benchmarking_2018}

microservices \cite{grambow_benchmarking_2020}

impact of webservers on web applications energy \cite{manotas_investigating_2013}

\subsection{frameworks}

impact of website implemenation on energy consumption \cite{philippot_characterization_2014} \cite{manotas2013investigating}

PHP \cite{benmoussa_new_2019} \cite{das_comparison_2016}

simulation of web cache \cite{cardenas_performance_2005}

java spring analysis \cite{gajewski_analysis_2019}

performance \cite{mishra2021web}

cros compiler bencharking \cite{yet2016cross}

\subsection{python}

webframeworks on python \cite{pankiv_concurrent_nodate}

hope library \cite{akeret_hope_2015}

bytecode effect \cite{ben_asher_effect_2009}

numba \cite{crist_dask_2016}

intel python \cite{li_boosting_2016}

python 2 vs 3 \cite{modzelewski_pyston_2020}

python runtime performances \cite{redondo_comprehensive_2015} \cite{murri_performance_2013}

static vs dynamic languages \cite{pang_what_nodate}

concurent benchmark framework \cite{pankiv2019concurrent}

\subsection{grpc}

defintion \cite{wang1993grpc}

comparaison \cite{chamas2017comparing}

emprical study \cite{de2021empirical}

### 2022-05-06 chapters/testing/conclusion.tex
add : citations

@@
 Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and its variation reduction at low workloads.
-However, it also lists some new factors that we identified along the analysis we conducted in Section~\ref{sec:parameters}, such as the results related to the OS or the reboot mode.
+However, it also lists some new factors that we identified along the analysis we conducted in Section, such as the results related to the OS or the reboot mode.
 Some of the guidelines are more useful/efficient for specific workloads, as showed in our experiments.

### 2022-05-11 chapters/programming_languages/programming_languages.tex
Proofreading RPC

--- 2021-10-22 chapters/programming_languages/programming_languages.tex
@@
 
-\chapter{The Energy footprints of programming languages}
+\chapter{Energy footprint of programming languages}
 \label{chapter:porgramming_langauges}
-In this chapter we will dicuss the impact of the choice of the programming language on the energy consumption of the software.
+In this chapter, we discuss the impact of the choice of the programming language on the energy consumption of the resulting software.
+To do so, we suggest to start with the general micro benchmarking and see how each programming languages interact with the CPU/Memory.
 
@@
 \import{chapters/programming_languages/}{microbenchmarks}
-\import{chapters/programming_languages/}{conclusion}
+As we have seen in the previous chapter, one of the most important feature of a test is to be \textbf{representative}.
+Therefore, we extend this study to some real-life use case, and the following sections below provide two case studies.
+
 \import{chapters/programming_languages/}{rpc}
 \import{chapters/programming_languages/}{web_frameworks}
+\import{chapters/programming_languages/}{conclusion}

### 2022-05-11 chapters/literature/draft_citations.tex
update : state of the art 60%

@@
 % 1. group articles by field
-hello my world!!
 
@@
 
+to be able to reduce the energy cost of running programs we should first be able to estame this energy consumption.
+many studies have been conducted to estimate a such energy consumption.
+that varries from static analysis of the source code to enfere it's energy consumption like \citeauthor{pereira_helping_2017} where they provided a tool to highlight the most energy consuming parts of the code \cite{pereira_helping_2017}.the main advantage of this apporach is to be able to estimate the energy consumption of a program without running it.
+however.
+unlike the conplexity of programs, the energy consumption is highly related the execusion environment.
+Therefor, Static analysis might not represent the real behaviour of the same program when it is run in the production environment.
+To treat the problem of representativeness, many researchers tend to measure the energy consumption of the programs while they are running them. Hence we will get more accurate results.
+there is a variety of tools to measure such energy, and they cover a large spectrium of usage depends on how \textbf{accurate} and \textbf{pricise} those results are needed to be in tone hand, and the price that practinoners are ready to pay for a such accuray/pricision.
+Below we will present some tools that are well known in the literature and discuss their main advantages / drawbacks for our case.
 \subsection{hardware tools}
 
-WattsUp Pro \cite{hirst2013watts}
+% TODO Add : diagram of the possible point where we can measure the energy
 
-PowerMon \cite{bedard2010powermon}
+% depends on the scientific usage there are two apporaches :
+% 1. extern one or independent : when the most of researchs are using
+% the benifits of this is the simplicity of instalation however we pay the price either with a lack of temporal resolution ( low frequency ) or the spatial granularity
+% ( most of them measure the energy consumption of the whole machine )
+% 2.
+dedicated one , they are based on the instrumentation of some parts of the computer , however they have a high cost and they are defficult to scale since they require an instrumentation of the execusion environment.
+according to \citeauthor{hackenberg2014hdeem} there are four main creterions to evaluate an energy measurement tool \cite{hackenberg2014hdeem}
+\begin{itemize}
+    \item \em{spatial granularity}: the more specific the target of monitoring we are able to measure, the more efficient we can do optimization since we will know what causes the pitfalls of the energy consumption.
+    \item \em{temporal granularity}: same as spatial granularity, temporal granularity helps us to identif the sequence of code that need to be optimized.
+    \item \em{scalability}: this is mainly related to the cost of the tools and the ease of their integration for our system.
+    \item  \em{accuracy}:to elimanate extra hazards and get more representative measurement.
+\end{itemize}
 
-PowerInsight \cite{laros2013powerinsight}
+We believe that accuracy can be extracted from those creterions since it is a result of the combaination of the two first ones. therefore, we will focus more on the first 3 creterions from later on.
+In the follwong section we will discuss some of the well known tools that are used in literature.
 
-GreenMiner \cite{hindle2014greenminer}
+% TODO : add a triangle to classify all those tools
 
-powerpack \cite{ge2009powerpack}
+Nowdays, most of the high performance computing systems (HPC) implements a tool to report the energy consumption of the nodes for the sake of monitoring and administration.
+Those tools are mainly integrated withing the power suply units (PSU) or the power distributuin units (PDU).
+then , they provide an interface and a log to follow the history of the energy consumption.
+Despite their scalability and ease of integration.
+such tools lack both in spatial and temporal granularity since they monitor the whole energy of the nodes, and most of the times they have a very low sampling frequency.
+most of those tools are provided directly by the manufacturers.
+such as \em{IBM EnergyScale technology} \cite{mccreary2007energyscale} \cite{caldeira2014ibm} \cite{caldeiraibm} or \em{Dell poweredge} \cite{lovicott2009thermal}, MEGware Cluststafe \cite{breitbart2015case}
+As we said earlier the true purpose of those tools is more monitoring than analysing the energy consumption.
+%% add something to link with wattsup pro 
+WattsUp Pro, is a device that can be installed  between the  powerer source of the machine and the system under test.
+it allows a sampling frequency up to 1Hz and has a internal memory to store a wide variety of data, such as the maximum voltage, current..
+etc that later can be exported via USB port for personal usage or lined to some graph programs like Logger Pro or LabQuest.The main advantage of this tool is the ability to monitor the energy consumption from a different device which will reduce the risk of interference with the energy consumption of the test % a revoir
+\cite{hirst2013watts}
 
-hdeem \cite{hackenberg2014hdeem}
+despite his high temporal granularity  , wattsup pro lacks in term of spatial granularity since it monitors the energy consumption of the whole system.
+To have a finer granularity we need to isolate the energy consumption of each component.
+.
+For this we will need more intrusive tools such.
 
-PDU ,RAPL ,AMD \cite{hackenberg2013power}
+%% THIS IS FOR GPU %%% 
+% GPU TESLA \cite{burtscher2014measuring}
 
-Intel RAPL Haswell \cite{hackenberg2015energy}
+% NVML\cite{fahad2019comparative}
+%%%%
 
-GPU TESLA \cite{burtscher2014measuring}
+PowerMon and its upgradded version powerMon2 \cite{bedard2010powermon} are based on an microcontroler chip that can monitor up to 6 channel ( 8 for powermon2) simultanetly.
+therefore we are able to monitor the powerconsumption of 4 devices at the same time.
+the frequency sample of this tool is up to 50Hz with an accuracy of 1.2\%.
+Powermore2 comes with smaller size that can fit within 3.5 inches rack drive.
 
-NVML\cite{fahad2019comparative}
+PowerInsight \cite{laros2013powerinsight} is another finegrained measurement tool that is based on an ARM bagelbone processor \cite{coley2012beaglebone} which is able to measure up to 30 channel simultanetly with a frequency of 1KHz per channel.
+
+on the other hand GreenMiner \cite{hindle2014greenminer}, %%i dont think that  mobile measurement will be relevent for my case
+
+powerpack \cite{ge2009powerpack} in the other hand is an api that synchronizes the data gathered from other monitoring tools such as Watt’s Up Pro, NI and RadioShack pro and the lines of code %% maybe remove this one from the list because it is not well known.
+
+other monitor tools have been relized by the manufactures such as IBM Power executive \cite{koomey2011growth} which allows their customer to monitor the power consumption and thermal behaviour of the of BladeCenter systems in the datacenter.
+
+%% TODO : add diagram where the distribution of the energy consumption is more related to the CPU rather than other tools - u can cite the tools dedicated to the GPU since it is is included in the software aggregators ?
+power api, pythoules ...etc
+
+as we see in % two diagrams one for research the other one for power usage 
+The CPU is the part responsible of the most energy consumption of softwares. hence the finner we go to measure this energy consumption the better it is for our work.
+Fortunately, Intel and later AMD proposed a tool that estimate the power consumption of different part of the CPI based on counter performances.
+RAPL (RUnning Average Power Limit) \cite{hackenberg2013power} \cite{hackenberg2015energy} is a set of registers that was introduced by intel in their CPU since Sandy bridge generation, and later it was fllowed by AMD since Family 17h Zen.
+It is capable of monitoring different components of CPU , such as DRAM, Cores , and the integrated GPU for desktop processors.
+
+%% TODO :ADD figure of RAPL + references
+
+The advanage of a such apporach is that the absence of any intrusive measurement tools.
+further more they have a high temporal granurality with a sampling frequency up to 1KHz \cite{ilsche_power_2015}.
+
+with similar approach we can find NVIDIA reporting tools such as GPU TESLA \cite{burtscher2014measuring} and the NVML library \cite{fahad2019comparative}
 
 \subsection{software tools}
+Now that we got our hands on a precise tools that allows us to monitor
 
@@
 
-cros compiler bencharking \cite{yet2016cross}
+cross compiler bencharking \cite{yet2016cross}
 

### 2022-05-11 compile-thesis.sh

@@ -1,12 +1,46 @@
 #!/bin/bash
-# A script to compile the PhD Thesis - Krishna Kumar 
+# A script to compile the PhD Thesis - Krishna Kumar
 # Distributed under GPLv2.0 License
 
-compile="compile";
-clean="clean";
+compile="compile"
+clean="clean"
 
-if test -z "$2"
-then
+if test -z "$2"; then
+	if [ $1 = $clean ]; then
+		echo "Cleaning please wait ..."
+		rm -f *~
+		rm -rf *.aux
+		rm -rf *.bbl
+		rm -rf *.blg
+		rm -rf *.d
+		rm -rf *.fls
+		rm -rf *.ilg
+		rm -rf *.ind
+		rm -rf *.toc*
+		rm -rf *.lot*
+		rm -rf *.lof*
+		rm -rf *.log
+		rm -rf *.idx
+		rm -rf *.out*
+		rm -rf *.nlo
+		rm -rf *.nls
+		rm -rf $filename.pdf
+		rm -rf $filename.ps
+		rm -rf $filename.dvi
+		rm -rf *#*
+		echo "Cleaning complete!"
+		exit
+	else
+		echo "Shell script for compiling the PhD Thesis"
+		echo "Usage: sh ./compile-thesis.sh [OPTIONS] [filename]"
+		echo "[option]  compile: Compiles the PhD Thesis"
+		echo "[option]  clean: removes temporary files no filename required"
+		exit
+	fi
+fi
+
+filename=$2
+
 if [ $1 = $clean ]; then
 	echo "Cleaning please wait ..."
 	rm -f *~
@@ -28,48 +62,13 @@ if [ $1 = $clean ]; then
 	rm -rf $filename.pdf
 	rm -rf $filename.ps
 	rm -rf $filename.dvi
-	rm -rf *#* 
-	echo "Cleaning complete!"
-	exit
-else
-	echo "Shell script for compiling the PhD Thesis"
-	echo "Usage: sh ./compile-thesis.sh [OPTIONS] [filename]"
-	echo "[option]  compile: Compiles the PhD Thesis"
-	echo "[option]  clean: removes temporary files no filename required"
-	exit
-fi
-fi
-
-filename=$2;
-
-if [ $1 = $clean ]; then
-	echo "Cleaning please wait ..."
-	rm -f *~
[…]

### 2022-05-11 chapters/programming_languages/rpc.tex

@@
-\section{Remote Procedure Call}
+\section{Remote Procedure Call Frameworks}
 
@@
 %  this is just to warmup 
-With the emerging technology to the cloud many protocols wanted to take the lead.
-Nowdays most of the architectures are based on multi-servicces and micro services.
-And to have higher versatility of developpers mulptile companies choose to be open to different programming langauges.
-basically it would be more efficient if we take advantage of each programming language to satisfy a specific need.
-However the callenge nowdays is to make the bridge between those plateforms.
-We have many initiatives such as openapi that try to create a taxonomy for rest apis.
-other approachs is to impelement all the different interfaces of the protocol by them selfs such as the RPC
+With the emergence of cloud technologies, many protocols wanted to take the lead.
+Nowdays, most of the architectures are based on multi-services and microservices.
+And, to have higher versatility of developers multiple companies choose to be open to different programming languages.
+Basically, it would be more efficient if we take advantage of each programming language to satisfy a specific need.
+However, the challenge nowadays is to make the bridge between those platforms.
+We have many initiatives, such as OpenAPI, that try to create a taxonomy for RESTful APIs.
+Other approaches implement all the different interfaces of the protocol by themselves, such as RPC.
+
 \subsection{State of the art}
+TODO
 
 \subsection{Research Questions}
-In this section we will first explore the ease of implemenation of this protocol
-then we will try to answer the following Research questions
+In this section, we first explore the ease of implementation of this protocol then we will try to answer the following research questions:
+\begin{description}
+    \item[\textsc{RQ}~1:] How do RPC implementations consume with regards to the size of the incoming request?
+    \item[\textsc{RQ}~2:] How do RPC implementations consume with regards to the number of concurrent clients?
+\end{description}
 
-\begin{description}
-    \item[\textsc{RQ}~1:] How do RPC implementations react to the size of the request ?
-    \item[\textsc{RQ}~2:] How do RPC implementations react to the number of clients ?
-
-\end{description}
 \subsection{Experimental protocol}
 \subsubsection{Hardware settings}
-All the experiments are run on the cluster paravance of the G5K platform.
-This cluster is composed of 72 identical machines, each one is equipped with 2 Intel Xeon E5-2630 V3, with 128 GIB of RAM.
-For more accuracy, our SUT (System Under Test) is equipped with a minimal version of Debian 9 (4.9.0 kernel version), which enforces the core processes required for the purpose of our experiment.
+All the experiments are run on the cluster \textsf{paravance} of the G5K platform.
+This cluster is composed of 72 identical machines, each one is equipped with 2 Intel Xeon E5-2630 V3, with 128\,GB of RAM.
+For more accuracy, our SUT (\emph{System Under Test}) is equipped with a minimal version of Debian\,9 (4.9.0 kernel version), which enforces the core processes required for the purpose of our experiment.
 Furthermore, we used Docker containers technology for reproducibility of the experiments and the isolation of the servers.
+
 \subsubsection{Energy measurements}
-To report on the energy consumption, we used HWPC sensor [], which is based on Intel RAPL technology, one of the most accurate tools to measure the energy consumption of the CPU and DRAM [].
-For better accuracy, we ran the HPWC sensor with a frequency of 10 Hz, and we used the same machine for all the experiments in order to reduce the variability [].
+To report on the energy consumption, we used HwPC sensor [TOCITE], which is based on Intel RAPL technology, one of the most accurate tools to measure the energy consumption of the CPU and DRAM [TOCITE].
+For better accuracy, we ran the HwPC sensor with a frequency of 10~Hz, and we used the same machine for all the experiments in order to reduce the variability [TOCITE].
 
@@
     \end{center}
-    \caption{Experemenal software architecture}\label{fig:rpcprotocol}
+    \caption{Experimental software architecture}\label{fig:rpcprotocol}
 \end{figure}
+
 \subsubsection{Client and server environments}
+To limit the impact of the network on the experiments, we run both the client and the server on the same machine.
+However, we isolate each part on a separate socket, in order to reduce the effect that the client might have on the server and \emph{vise versa}.
+To do so, for each iteration, we always run the same client on \textsf{socket\,0} and the server that we want to test on \textsf{socket\,1}.
+Both the server and the client use the whole socket for their experiment.
+In addition, all the additional services, such as the kernel and HwPC sensor, are run on \textsf{socket\,0}.
+Therefore, the only process being executed in \textsf{socket\,1} is the server that we benchmark and monitor.
 
-To limit the impact of the network on the experiments, we run both the client and the server on the same machine.
-However, we isolate each one on a different socket, in order to reduce the effect that the client might have on the server ones and vise-versa.
-To do so, for each iteration, we always run the same client on Socket 0 and the server that we want to test on socket 1.
-Both the server and the client take the whole socket for their experiment.
-In addition all the extra services, such as OS, hwpc, etc.
-are run on Socket 0.
-Therefore, the only process being executed in Socket 1 is the server that we benchmark.
+\paragraph{Client:}
+% TODO : add the github repository links 
+For better accuracy and more details, We use an updated version of the open source RPC benchmarking tool, named GHZ (\url{https://ghz.sh/}).
+The modified version allows us to monitor the average power for each request from both the server and the client sides.
+The new version is available in the repository.%ROMAIN: Which repo?
+The client consumes the protocol description to generate an implementation for the message \texttt{helloworld}, and then forks 50 instances that send the same request to the server concurrently.
 
-\paragraph{Client :}
-% TODO : add the github repository links 
-For better accuracy and more details, We use an updated version of the open source RPC benchmarking tool, named GHZ (https://ghz.sh/).
-The modified version allows us to get the average power for each request form both the server and the client sides.
-The new version is available in the repo.
-The client will take the protocol description to generate an implementation for the message helloworld, and then fork 50 instances that will send the same request to the server simultaneously.
-\paragraph{Server :}
-The server implementations are based on the official implementation by Google for most of the languages . Each server uses 16 cores  and is limited to 512 MB of RAM
+\paragraph{Server:}
+The server implementations are based on the official implementation by Google for most of the languages.
+Each server uses 16 cores and is limited to 512\,MB of RAM.
 
-\subsection{Results and finding}
+\subsection{Results and findings}
 [\textsc{RQ}~1:] How do RPC implementations react to the size of the data ?
-The purpose of this question is to study the behaviour of the server for transferring large objects.
-To do so, we send to the server 80,000 requests with a size scaling from 10 bytes up to 10 Megabytes, which gives 10,000 requests per size per server.
+The purpose of this question is to study the behaviour of the server when transferring large objects.
+To do so, we send 80,000 requests to the server whose size scales from 10 bytes up to 10 Megabytes, which results in 10,000 requests per size per server.
 To eliminate extra factors, we let the server handle the rate at which it can answer each request.
-However, we put a 20 sec timeout limit for each request.
+However, we put a 20 seconds timeout limit for each request.
 Therefore, our boundary condition is only the number of requests received by the server.
-For this experiment, we investigate 4 observable variables :
+For this experiment, we investigate 4 observable variables~:
 \begin{enumerate}
-    \item The average power consumption during the the process : this will indicate the overall behaviour of the server in working mode for long durations ;
-    \item The tail latency for the 99th percentile : which indicates how performant is the server ;
-    \item The average number of requests per second : which indicates the  average number of clients that  the server can handle ;
-    \item The average energy cost of a single request : unlike the first indicator, this one shows how green is the implementation taking performance into consideration.
+    \item the \textsf{average power consumption} during the the process: this will indicate the overall behaviour of the server in working mode for long durations,
+    \item the \textsf{tail latency} for the 99th percentile, which indicates how performant is the server,
+    \item the \textsf{average number of requests per second}, which indicates the average number of clients that the server can handle,
+    \item the \textsf{average energy cost} of a single request: unlike the first indicator, this one shows how green is the implementation taking performance into consideration.
 \end{enumerate}
@@
 % TODO : ADD the figures 
-The above figure depicts the overall behaviour of each framework based on the size of request (Payload). For each framework, we can distinguish three modes, and they all depend on the payload :
+The above figure depicts the overall behaviour of each framework based on the size of request (payload).
+For each framework, we can distinguish three modes, and they all depend on the payload size:
 \begin{enumerate}
-    \item Stress free mode: when the server has enough resources to satisfy the requests because they require a memory less than a certain threshold (depends on the language and the platform),
-    \item Escalation mode: where the requests tend to be bigger, however the server can still manage to handle them, and here where we can see a change in the energetic and performance behaviour.
-    \item Broken state mode: when the requests are much heavier and the server break—like 10 MB.
+    \item \textsf{Stress free} mode when the server has enough resources to satisfy the requests because they require a memory less than a certain threshold (depends on the language and the platform),
+    \item \textsf{Escalation} mode when the requests tend to be bigger, however the server can still manage to handle them, and here where we can see a change in the energetic and performance behaviour,
+    \item \textsf{Broken state} mode when the requests are much heavier and the server break—like 10 MB.%ROMAIN unclear to me what this means
 \end{enumerate}
@@
 \subsubsection{Stress free mode}
-In this mode, the compiled languages tend to consume less resources (av power).
+In this mode, the compiled languages tend to consume less resources (average power).
 JVM-based languages tend to consume more energy, especially Scala.
@@
 \begin{itemize}
-    \item Green languages: CPP, GO, RUST, ELIXIR, and PHP
-    \item Middle class: Most of the interpreted languages and vm-based ones
-    \item Energivore class: Crystal and Scala
+    \item Energy-efficient class: C++, GO, RUST, ELIXIR, and PHP,
+    \item Middle class: Most of the interpreted languages and VM-based ones,
+    \item Energy-greedy class: Crystal and Scala.
 \end{itemize}
@@
 \subsubsection{Escalation mode}
-In this mode, the behaviour of the server depends on the payload. We observe three behaviours
+In this mode, the behaviour of the server depends on the payload. We observe three behaviours:
 \begin{enumerate}
     \item Drop in performances without an increased power, such as .Net core, Java micronaut, Crystal, and Dart.
-In this case the server keeps using the same ressources, and sometimes less, because it takes more time to handle the less requests.
-This class of languages tend to be the most energivore when it comes to the cost per request ;
-    \item Increase in power without affecting the performances: such as Go, .Net. The energy consumption of a single request, is affected slightly but still increase ;
+    In this case, the server keeps using the same resources, and sometimes less, because it takes more time to handle the less requests.
+    This class of languages tend to be the most energy consuming when it comes to the cost per request,
+    \item Increase in power without affecting the performances: such as Go, .Net.
+    The energy consumption of a single request, is affected slightly but still increase,
     \item Increase in power and drop in performances: Despite the increase of the power consumption, the server becomes slightly slower, which increases the cost of the energy cost per request.
@@
 
-Special mention to Elixir that kept scaling despite the lack of performances compared  to other compiled languages (Go, CPP)
+We can mention the case of Elixir that kept scaling despite the lack of performances compared to other compiled languages (Go, CPP).
 
 \subsubsection{Broken state mode }
-Only four of the 25 configurations could parse the 10 MB files.
-And only 1 from those could achieve a 76\% acceptance rate which is Elixir, the other 3 had less than 3\% success rate (Rust, Swift and Dart).
+Only four of the 25 configurations could parse the 10 MB files, and only 1 from those could achieve a 76\% acceptance rate which is Elixir, the other 3 had less than 3\% success rate (Rust, Swift and Dart).
 The rest could be divided into two categories:
 \begin{itemize}
-    \item Timeout : where requests took too much time that the client canceled them, in this category we find most of dynamic codes such as: openJDK, Kotlin ;
-    \item The size of request exceeded the maximum size by implementation : this is where the implementation could not handle requests with large size, such as .Net, Go, .Net core, CPP, PHP, Scala, Nodejs, Ruby, Python.
+    \item \textsf{Timeout} where requests took too much time that the client canceled them, in this category we find most of dynamic codes, such as OpenJDK and Kotlin,
+    \item \textsf{Size of request} exceeded the maximum size when the implementation could not handle requests with large size, as observed with .Net, Go, .Net core, CPP, PHP, Scala, Nodejs, Ruby, Python.
 \end{itemize}
@@
 \subsubsection{Power behaviour}
-based on the heatmap, we can distinguish two main modes.
+Based on the heatmap, we can distinguish two modes:
 \begin{itemize}
-    \item Low number of clients : where the total number of simultaneous clients is less than 100
-    \item Moderate to high number of clients : the number of clients exceeds 100
+    \item \textsf{Low number of clients} when the number of concurrent clients is below 100,
+    \item \textsf{Moderate to high number of clients} when the number of clients exceeds 100.
 \end{itemize}
@@
 \paragraph{Lite mode}
-The implementations can be grouped into two main categories :
+The benchmarked implementations can be grouped into two classes:
 \begin{enumerate}
-    \item Green Frameworks: most of the framework's power consumption is around 33 Watts.
-    \item Energivore frameworks : where the average power consumption is higher than 37 Watts.
+    \item \textsf{Energy-efficient frameworks} where most of the framework's power consumption is around 33 Watts.
+    \item \textsf{Energy-greedy frameworks} where the average power consumption is higher than 37 Watts.
 \end{enumerate}
 
-In each programming category we observe both behaviours green and energivore.
-Therefore, we conclude that it depends more on the implementation of the library itself rather than the category of the programming language.
-Scala and Kotlin are an excellent example to support this hypothesis since both of them run on the same virtual machine as Java (openjdk 16.1).Yet, their average power is 130\% higher than the Java implementation .
+In each programming category, we observe both energy-efficient and energy-greedy behaviours.
+Therefore, we conclude that it depends more on the implementation of the library itself, rather than the category of the programming language.
+Scala and Kotlin are an excellent example to support this hypothesis, as both of them run on the same virtual machine as Java (OpenJDK 16.1).
+Yet, their average power is 130\% higher than the Java implementation.
 
 \paragraph{Stressed mode}
-Although the same classes remained the same , Not all the languages had the same evolution.
-and here we can clearly see that it is correlated with the category of the programming language rather than the implementation itself.\\
-We can clearly highlight that after VM based languages have a significant increase in the average power consumption after they receive more than 100 simultaneous clients.
-This increase reaches almost double.
-Except PHP, all the interpreted languages preserved their energetic behaviour. Same for the compiled ones.
-Our Hypothesis points to the JIT, since it will compile the code and make it run faster so it will stress the CPU more.\\
-A Remarquable behaviour has been noticed for the grallVM, is the decrease of the energy consumption when we increase the number of the clients.
-This is related to the drop of the performances which was probably due to the bottleneck situation where the GraalVM couldn't handle more than 100 clients simultaneously.
+Although the same classes remained the same, not all the languages had the same evolution and here we can clearly observe a correlation with the category of the programming language rather than the implementation itself.
+We can clearly highlight that VM-based languages have a significant increase (double) in the average power consumption after they receive more than 100 concurrent clients.
+Except PHP, all the interpreted and compiled languages preserved their energetic behaviour.
+Our hypothesis points to the JIT, since it compiles the code and makes it run faster, hence stressing the CPU.
+A interesting behaviour has been noticed for the GraalVM: the decrease of energy consumption when increasing the number of the clients.
+This is related to the drop of the performances, which was probably due to the bottleneck situation where the GraalVM could not handle more than 100 clients simultaneously.
 
 \subsubsection{Performance Behaviour}
-In this section we study only the number of requests per seconds processed by the server without looking at its energy.
-We have three observable variables
+In this section, we study only the number of requests per seconds processed by the server without looking at its energy.
+We consider three observable variables:
 \begin{itemize}
-    \item Satisfaction ratio : how many requerts have been satisfied among the total requests
-    \item Request Per Seconds : The number of the requests that have been answered from the server
-    \item TailLatency at 99\% : one of the best metrics to evaluate the performances of a server
+    \item \textsf{Satisfaction ratio}: how many requests have been satisfied among the total requests,
+    \item \textsf{Request Per Seconds}: The number of the requests that have been answered from the server,
+    \item \textsf{Tail Latency at 99\%}: one of the best metrics to evaluate the performances of a server.
 \end{itemize}
@@
 \paragraph{Satisfaction ratio}
-Most of the frameworks tried to satisfy all the requests, by either reducing the number of requests per second or by increasing the time of the treatment.
-However, there are some frameworks that have chosen a different approach such as dart or scalla where the choice was to keep a certain limit of latency even if not all the requests are answered.\\
-Furthermore, we tend to see this behaviour among other frameworks such as Python or Asynchronous NodeJs when the number of the client exceeds 800.
+Most of the considered frameworks satisfy all the requests, by either reducing the number of requests per second or by increasing the processing time.
+However, there are some frameworks that have chosen a different approach, such as Dart or Scala, where the choice was to keep a certain limit of latency even if not all the requests are answered.
+Furthermore, we tend to observe this behaviour among other frameworks, such as Python or Asynchronous NodeJs, when the number of the client exceeds 800.
+
 \paragraph{RPS}
-Most of the servers hit their RPS limit after 5 clients and 100 clients for vm based servers, and after this the number keeps constant. which will decrease the average RPS per client.\\
-.net Server is the most performant one, and after him we see JAVA and Go in second place, on the other end Python and ruby are the least performant.
-\paragraph{Tail LAtency}
-However, the increase in the number of requests per second, doesn't necessarily mean a lower Latency.
-As we see in the table ... until the 1000 clients, Go provides the least Latency beside .net.\\
+Most of the servers hit their RPS limit after 5 clients and 100 clients for vm based servers, and after this the number keeps constant, which will decrease the average RPS per client.
+.Net server is the most performant, followed by Java and Go, while Python and Ruby are the least performant.
+
+\paragraph{Tail Latency}
+The increase in the number of requests per second, does not necessarily mean a lower latency.
+As one can notice in Table~\ref{...}, until the 1000 clients, Go provides the least latency beside .Net.
 GraalVM provides the highest latency, on average.
-However, dart tends to become slower when we increase the number of clients, until we pass the 600 simultaneous clients, and there it changes its behaviour, instead of satisfying most the requests it notify the clients directly that the server is saturated, hence a drop in satisfaction ratio, and an amelioration for the Average Latency.
+However, Dart tends to become slower when we increase the number of clients, until we pass the 600 simultaneous clients, and there it changes its behaviour, instead of satisfying most the requests it notify the clients directly that the server is saturated, hence a drop in satisfaction ratio, and an amelioration for the average latency.
 
 \subsubsection{Energy Per Request}
-Now after we made a separation between The energy and The performances, we have seen that most of the performance servers tend to be energy hungry, so we propose to investigate this trade off between the energy and the performance.
-Todo so we will give an average cost of a single Request in Joules.
-Except GraalVM when the cost of the a single request increases with when we add more clients, All the frameworks keep a constant cost,
-Java, .net and go are the greenest. and Python, ruby are the most costly with 10x higher.
-Therefore we conclude that the number of clients won't impact the energy that much.
-Next study will be the payload and how the size of the requests will affect the energy consumption of the framework
+Now, after we made a separation between the energy and the performances, we have seen that most of the performance servers tend to be energy-greedy, so we propose to investigate this trade-off between energy and performances.
+To do so, we report on an average cost of a single request, in Joules.
+Except GraalVM when the cost of the a single request increases with when we add more clients, all the frameworks report on a constant cost, Java, .net and go are the most energy efficient, while Python and Ruby may cost up to 10x more.
+Therefore, we conclude that the number of clients does not impact the energy significantly.
+Then, we study how the payload size of the requests impact the energy consumption of the framework.
 
@@
     \end{center}
-    \caption{Experemenal software architecture}\label{fig:rpcprotocol}
+    \caption{Experimental software architecture}\label{fig:rpcprotocol}
 \end{figure}
@@
     \end{center}
-    \caption{Experemenal software architecture}\label{fig:energy_cost_payload}
+    \caption{Experimental software architecture}\label{fig:energy_cost_payload}
 \end{figure}
@@
     \end{center}
-    \caption{Experemenal software architecture}\label{fig:power_consumption_clients}
+    \caption{Experimental software architecture}\label{fig:power_consumption_clients}
 \end{figure}
@@
     \end{center}
-    \caption{Experemenal software architecture}\label{fig:power_consumption_payload}
+    \caption{Experimental software architecture}\label{fig:power_consumption_payload}
 \end{figure}
@@
     \end{center}
-    \caption{Experemenal software architecture}\label{fig:rps_clients}
+    \caption{Experimental software architecture}\label{fig:rps_clients}
 \end{figure}
@@
     \end{center}
-    \caption{Experemenal software architecture}\label{fig:rps_payload}
+    \caption{Experimental software architecture}\label{fig:rps_payload}
 \end{figure}
@@
     \end{center}
-    \caption{Experemenal software architecture}\label{fig:tail99_clients}
+    \caption{Experimental software architecture}\label{fig:tail99_clients}
 \end{figure}
@@
     \end{center}
-    \caption{Experemenal software architecture}\label{fig:tail99_payload}
+    \caption{Experimental software architecture}\label{fig:tail99_payload}
 \end{figure}
@@
 \subsection{Threads to validity}
+
 \subsection{Conclusion}

### 2022-05-18 chapters/execution_environement/execution_environement.tex
add : python
Revising Java

@@
 
-\chapter{Execution enviroment}
+\chapter{Execution environment}
 \label{chapter:enviroment}
-After studying the impact of the programming languages choice on the energy consumption of softwares.
-we wanted to dig in more deeper.
-Therefore we have chosen two of the most popular programming lanaguages \textbf{Python} and \textbf{Java}
-In this chapter we will reduce the energy consumption of those following lanaguages by doing some tuning
+After studying the impact of the programming languages choice on the energy consumption of software, we wanted to dig in more deeper.
+Therefore, we have chosen two of the most popular programming languages \textbf{Python} and \textbf{Java}.
+In this chapter, we explore the reduction of energy consumption of those languages by tuning their configuration.
+
 \import{chapters/execution_environement/java}{java}
+\import{chapters/execution_environement/python}{python}

### 2022-05-18 chapters/execution_environement/java/java.tex
Revising Java

--- 2022-05-03 chapters/execution_environement/java/java_article.tex
@@
-\begin{abstract}
-    \vspace{0pt}\noindent\textbf{Background.}
-    The \emph{Java Virtual Machine} (JVM) platforms have known multiple evolutions along the last decades to enhance both the performance they exhibit and the features they offer.
-    With regards to energy consumption, few studies have investigated the energy consumption of code and data structures.
-    Yet, we keep missing an evaluation of the energy efficiency of existing JVM platforms and an identification of the configurations that minimize the energy consumption of software hosted on the JVM.
+\section{Introduction}
+As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.%ROMAIN: Add source
+Furthermore, if we take into consideration legacy applications, Java becomes the most used programming langues.%ROMAIN: Add source
+In addition to its popularity, Java exhibits an interesting behaviour when it comes to energy consumption and performances, basically Java applications can be at the same time one of the most energy efficient or hungry solution.
+As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
+Therefore, we wanted to dig deeper into this aspect of Java and study its runtime.
+This chapter thus focuses on the impact of the runtime of Java applications on the energy consumption.
 
-    \vspace{1pt}\noindent\textbf{Aims.}
-    The purpose of this paper is to investigate the variations in energy consumption between different JVM distributions and parameters to help developers configuring the least consuming environment for their Java application.
-
-    \vspace{1pt}\noindent\textbf{Method.}
-    We thus assess the energy consumption of some of the most popular and supported JVM platforms using 12 Java benchmarks that explore different performance objectives.
-    Moreover, we investigate the impact of the different JVM parameters and configurations on the energy consumption of software.
-
-    \vspace{1pt}\noindent\textbf{Results.}
-    Our results show that some JVM platforms can exhibit up to 100\% more energy consumption.
-    JVM configurations can also play a substantial role to reduce the energy consumption during the software execution.
-    Interestingly, the default configuration of the garbage collector was energy efficient in only 50\% of our experiments.
-
-    \vspace{1pt}\noindent\textbf{Conclusion.}
-    Finally, we provide an OSS tool, named \textsf{J-Referral} that recommends an energy-efficient JVM distribution and configuration for any Java application.
-\end{abstract}
-
-% \maketitle
-
-\section{Introduction}\label{sec:intro}
-Software services are widely deployed to support our daily activities, being mobile or hosted in the cloud.
-Yet, beyond this undeniable success, the environmental impact of ICT is raising concerns and calls for solutions to reduce the energy footprint of software services~\cite{the_shift_project_lean_2019}.
-
-Software developers often report that such solutions should come from more energy-efficient hardware components or optimized algorithms~\cite{wang_potential_2018,opaper} but, given the complexity of modern software environments, the composition of software layers makes this sustainability objective particularly challenging.
-
-Given this context, this paper more specifically investigates the impact of one of these layers, the runtime environment and its settings, on the energy consumption of an hosted software service.
-More precisely, we aim at revealing the importance of carefully selecting and configuring the \emph{Java Virtual Machine} (JVM) to reduce the energy consumption of any software service built from any language compatible with the Java ecosystem (Java, Kotlin, Scala, Groovy, Clojure, Jython, etc.).
-
-The empirical study we conduct in this paper reports on the energy footprint of several versions of popular JVM distributions that are freely available for download.
-Beyond the choice of an appropriate runtime and its most energy-efficient version, we also consider the impact of exposed JVM settings to maximize the energy savings for a given software service.
-The observations of this study aim to quantify the role played by internal JVM mechanisms, like the \emph{Just in Time} (JIT) compiler and the \emph{Garbage Collector} (GC), in the reduction of the energy consumption of hosted applications.
-More formally, we formulate the following research questions:
+\subsection{Goal}
+In this section, we will investigate the following research questions:
 \begin{compactenum}[\indent\bf RQ\,1:]
+    %  TODO : generalize this format of Rq  to all the thesis 
     \item \emph{What is the impact of existing JVM distributions on the energy consumption of Java-based software services?}
@@
 
-By answering these questions, we envision supporting application developers and administrators in the configuration of their production environment by substantially reducing the energy consumption of hosted services at large.
-We also hope that our results will encourage the JVM developers to keep investing in the integration of further optimizations that can benefit a large population of software services.
-This paper comes with a set of contributions that can be summarized as:
-\begin{compactenum}[\indent\em (a)]
-    \item Reporting on the energy-efficiency of a large panel of JVM when running acknowledged benchmarks,
-    \item Identifying and assessing the key JVM settings that can influence the energy consumption of a software service,
-    \item Sharing guidelines and prerequisites that will help in configuring the most energy-efficiency environment before deployment,
-    \item Providing a JVM benchmarking environment to evaluate the energy-efficiency of upcoming JVM distributions and their settings,
-    \item Delivering an open-source tool, named \textsf{J-Referral}, to recommend the most energy efficient JVM distribution among 85 JVMs/versions and hundreds of configurations for Java applications.
-\end{compactenum}
+To answer those research question we will run an empirical study to highlight the impact of this runtime.
 
-The remainder of this paper is organized as follows.
-\Cref{sec:meth} introduces the experimental protocol and methodology (hardware, projects, tools, and methodology) we adopted in this study.
-\Cref{sec:exp} analyzes the results of our experiments on the energy consumption of the different JVM configurations.
-\Cref{sec:rw} discusses the related works to reduce the energy consumption of Java-based Software Services.
-%Finally, \Cref{sec:threats,sec:ccl} cover the validity threats and our conclusions, respectively.
-Finally, \Cref{sec:ccl} covers our conclusions.
-
-% Energy efficiency is a major concern in software engineering.
-% Indeed, the huge leap in technology with all the mobile devices, IoT, cloud computing, etc. raised serious questions about the way energy is consumed. 
-% In most domains, we often aim to reduce the consumption of the electronic devices/parts while improving.
-% In computer science, many researches were interested in the EC optimization.
-% Some of the works focused on reducing the EC at the hardware or the OS level, others by optimizing software's source codes consumption.
-% In this paper, we focus on the JVM based languages, especially Java.
-% Many works wave been conducted to reduce the Java and JVM based software energy consumption \cite{}.
-% These works focused on many aspects of the source code such as the data collections, control structures,  code refactoring ,etc.
-% Other works focused on aspects that are closer to the JVM execution such as bytecode energetic cost analysis \cite{brownlee_search_based_2017,grass_energy_2006} and the consumption that can exhibit the same source code written in many languages that can use the JVM \cite{pereira_energy_2017}.
-% However, even with the increasing concern about software energetic efficiency, we still lack some basic knowloedge on how to parameter and configure the execution environment do the software can run in ideal conditions and thus consume less energy.
-% Hence, our study aims at constituting a set of knowledge on the basic choices and configurations that one can apply on a JVM to reduce the energy consumption.
-% Concretely, we benchmark different JVM plateforms from different providers, along with different configurations and parameters of threads management, Just In Time compiler (JIT) and Garbage Collector (GC).
-% The puropose of this study is to check if some configurations can play a substantial role in reducing software energy consumption.
-% In particular, if the use the default JVM with a default configuration can be a reason for energy consumption increase and how often can that be. 
-% This study, therefore, aims to answer the following research questions:
-% \begin{description}
-% 	\item[\textbf{RQ\,1}:] What is the impact that can a JVM exhibit on the energy consumption of the running software?
-% 	\item[\textbf{RQ\,2}:] How can one properly configure his JMV to to reduce software energy consumption?
-% \end{description}
-
-\section{The Java Virtual Machine}\label{sec:background}
-Java was originally developed by James Gosling at Sun~Microsystems and released in 1995, before being acquired by Oracle in 2010.
-One key design goal of Java is portability, which means that Java applications must run similarly on any combination of hardware and operating system with adequate runtime support.
-This is achieved by compiling the Java language code to an intermediate representation, called Java \emph{bytecode}, instead of machine code.
-Java bytecode instructions are analogous to machine code, but executed by a \emph{Java Virtual Machine} (JVM), which is specific to the host machine.
-For example, Oracle keeps offering the \textsc{HotSpot}\,JVM, while the official reference implementation is now the \textsc{OpenJDK}\,JVM---a free and open-source software used by most developers.
-
-Programs written in Java have a reputation for being slower and requiring more memory than those written in C++, but \emph{Just-in-Time} (JIT) compilation, embedded in the JVM, delivers a boost of performance by opportunistically compiling bytecode to machine code at runtime.
-The JIT combines two compilers, C1 and C2 (also known as Client \& Server\,VM), which are triggered based on the activity of the hosted application.
-Additionally, Java uses an automatic \emph{garbage collector} (GC) to manage memory in the object lifecycle and recovering the memory once objects are no longer in use.
-Each JVM usually includes multiple GC, each designed to satisfy different requirements.
-
-By default, \textsc{HotSpot} uses both C1 and C2 as tiered compilers,\footnote{\url{http://www.ittc.ku.edu/~kulkarni/teaching/EECS768/19-Spring/Idhaya_Elango_JIT.pdf}} and the \emph{Garbage-First} (G1) GC with a maximum number of GC threads limited by available CPU resources and heap size, whose initial size is set $1/64^{th}$ of physical memory and maximum size may reach up to $1/4^{th}$ of physical memory.\footnote{https://docs.oracle.com/en/java/javase/15/gctuning/ergonomics.html}
-
-%\begin{figure}%[ht]
-%	\centering
-%	\includegraphics[width=\linewidth]{imgs/hotspot-jvm.pdf}
-%	\caption{Overview of the HotSpot JVM key components.}
-%	\label{fig:hotspot}
-%\end{figure}
-
-At the time of writing this paper, the latest JVM version is Java\,15, released in September~2020, while Java\,11 is the current \emph{Long-Term Support} (LTS) version.
-As Java\,9, 10, 12, and 13 are no longer supported, Oracle advises developers to immediately transition to the latest version (currently Java\,15), or an LTS release.
-
-Beyond \textsc{HotSpot}, one can observe that the initial JVM design leads to numerous initiatives to improve the performances of Java applications, including new hotswapping strategies---with the \emph{Dynamic Code Evolution Virtual Machine} (DCE\,VM)~\cite{DBLP:journals/scp/WurthingerWS13}---or alternative JIT---with \textsc{GraalVM}.\footnote{\url{https://www.graalvm.org}}
-This also includes alternative implementations, like IBM\,J9 JVM, which is currently distributed as part of the Eclipse foundation, and known as \textsc{J9}.\footnote{https://www.eclipse.org/openj9}
-Given the wide diversity of distributions and related settings, this paper aims to study the impact of the features implemented by available JVM distributions on the energy consumption of the hosted Java software services.
+\subsection{definition of the JVM}
+\begin{figure}%[ht]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/Jvm_architecture}
+    \caption{JVM architecture}
+    \label{fig:hotspot}
+\end{figure}
 
@@
 
+\subsection{Measurement Contexts}
+
+\noindent\textbf{Software Settings.}
+For the sake of reproducibility, each experiment runs within a Docker container based on SDKMAN image and Alpline docker.
+
 \vspace{6pt}
 \noindent\textbf{Hardware Settings.}
-To report on reproducible measurements, we used the cluster \textsf{Dahu} of the G5K platform~\cite{grid5000} for most of our experiments.
+To report on reproducible measurements, we used the cluster \textsf{Dahu} from the G5K platform~\cite{grid5000} for most of our experiments.
 This cluster is composed of $32$ identical compute nodes, which are equipped with $2$ Intel Xeon\,Gold\,6130 and 192\,GB of RAM.
@@
 \vspace{6pt}
-\noindent\textbf{Energy Measurements.}
-We used Intel\,RAPL as a physical power~meter to analyze the energy consumption of the CPU package and the DRAM.
-RAPL is one of the most accurate tools to report on the global energy consumption of a processor~\cite{Khan:2018:RAE:3199681.3177754,10.1145/2989081.2989088}.
-We note that, due to CPU energy consumption variations issues~\cite{opaper}, we used the same node for all our experiments.
-% In other words, we used the same node to ran all the JVM from the different providers and with numerous configurations for each single benchmark application.
-Moreover, we tried to be very careful, while running our experiments, not to fall in most common benchmarking "crimes"~\cite{crimes}.
-Every single experiment, therefore, reports on energy metrics obtained from at least $20$ executions of $50$ iterations per benchmark.
-All of our experiments are available for use/reproducibility from our anonymous repository.\footnote{\url{https://anonymous.4open.science/r/jvm-comparaison-213E/Readme.md}}
-
-\vspace{6pt}
-\noindent\textbf{Java Virtual Machines.}
+\noindent\textbf{Java Virtual Machines Candidates.}
 We considered a set of $52$ JVM distributions taken from $8$ different providers/packagers mostly obtained from \textsf{SDKMAN!},\footnote{\url{https://sdkman.io/}} as listed in \Cref{tab:JVM}.
@@
 
-\begin{table*}[t]
+\subsection{Workload}
+We ran our experiments across $12$ Java benchmarks we picked from \textsf{OpenBenchmarking.org}.\footnote{\url{https://openbenchmarking.org}}
+This includes $5$ acknowledged benchmarks from the \textsc{Dacapo} benchmark suite v.\,9.12~\cite{DaCapo:paper}, namely \textsf{Avrora}, \textsf{H2}, \textsf{Lusearch}, \textsf{Sunflow} and \textsf{PMD}, that have been widely used in previous studies and proven to be accurate for memory management and computer architecture communities~\cite{DBLP:conf/wosp/LengauerBMW17,DBLP:conf/oopsla/KaliberaMJV12}.
+It consists of open-source and real-world applications with non-trivial memory loads.
+% 
+Then, we also considered $7$ additional benchmarks from the \textsc{Renaissance} benchmark suite~\cite{renaissance,DBLP:conf/pldi/ProkopecRLD0SBZ19}, namely \textsf{ALS}, \textsf{Dotty}, \textsf{Fj-kmeans}, \textsf{Neo4j}, \textsf{Philosophers}, \textsf{Reaction} and \textsf{Scrabble}, which offers a diversified set of benchmarks aimed at testing JIT, GC, profilers, analyzers, and other tools.
+The benchmarks we picked from both suites exercise a broad range of programming paradigms, including concurrent, parallel, functional, and object-oriented programming.
+\Cref{tab:benchmarks} summarizes the selected benchmarks with a short description.
+
+\begin{table*}
     \centering
@@
 
-\vspace{6pt}
-\noindent\textbf{Java Benchmarks.}
-We ran our experiments across $12$ Java benchmarks we picked from \textsf{OpenBenchmarking.org}.\footnote{\url{https://openbenchmarking.org}}
-This includes $5$ acknowledged benchmarks from the \textsc{Dacapo} benchmark suite v.\,9.12~\cite{DaCapo:paper}, namely \textsf{Avrora}, \textsf{H2}, \textsf{Lusearch}, \textsf{Sunflow} and \textsf{PMD}, that have been widely used in previous studies and proven to be accurate for memory management and computer architecture communities~\cite{DBLP:conf/wosp/LengauerBMW17,DBLP:conf/oopsla/KaliberaMJV12}.
-It consists of open-source and real-world applications with non-trivial memory loads.
-% 
-Then, we also considered $7$ additional benchmarks from the \textsc{Renaissance} benchmark suite~\cite{renaissance,DBLP:conf/pldi/ProkopecRLD0SBZ19}, namely \textsf{ALS}, \textsf{Dotty}, \textsf{Fj-kmeans}, \textsf{Neo4j}, \textsf{Philosophers}, \textsf{Reaction} and \textsf{Scrabble}, which offers a diversified set of benchmarks aimed at testing JIT, GC, profilers, analyzers, and other tools.
-The benchmarks we picked from both suites exercise a broad range of programming paradigms, including concurrent, parallel, functional, and object-oriented programming.
-\Cref{tab:benchmarks} summarizes the selected benchmarks with a short description.
+\subsection{metrics and measurement}
+Since the goal of this study is the green aspect Java virtual machines, our key metric will be the energy consumption of the job completed by each configuration.
+In addition to the energy consumption, we collected additional metrics to explain the reasons behind the behaviour of each experiment.
+Those extra metrics are:
+\begin{itemize}
+    \item execution time,
+    \item number of threads.
+\end{itemize}
+
+% Our protocol allow the user to record extra metrics, such as the energy consumption of the memory, the CPU usage, etc.
+% However, we focus only the the previous metrics.
+
+\noindent\textbf{Energy Measurements.}
+We used Intel\,RAPL as a physical power~meter to analyze the energy consumption of the CPU package and the DRAM.
+RAPL is one of the most accurate tools to report on the global energy consumption of a processor~\cite{Khan:2018:RAE:3199681.3177754,10.1145/2989081.2989088}.
+We note that, due to CPU energy consumption variations issues~\cite{opaper}, we used the same node for all our experiments.
+% In other words, we used the same node to ran all the JVM from the different providers and with numerous configurations for each single benchmark application.
+Moreover, we tried to be very careful, while running our experiments, not to fall in most common benchmarking "crimes"~\cite{crimes}.
+Every single experiment, therefore, reports on energy metrics obtained from at least $20$ executions of $50$ iterations per benchmark.
+All of our experiments are available for use/reproducibility from our anonymous repository.\footnote{\url{https://anonymous.4open.science/r/jvm-comparaison-213E/Readme.md}}
+
+\noindent\textbf{Number of threads.}
+To collect the number of active threads used by the an experiment, we use the command \texttt{top} and record at fixed intervals.
 
@@
 Ultimately, 3 clusters of JVMs that encompass JVMs with similar energy consumption can be seen through \Cref{fig:JVMs}: \textsc{J9}, the \textsc{HotSpot} and its variants, and \textsc{GraalVM}.
-Additional detailed figures to illustrate the evolution of energy consumption per benchmark/JVM are made available from the anonymous repository.\footnote{\url{https://anonymous.4open.science/r/jvm-comparaison-213E/Readme.md}}
+Additional detailed figures to illustrate the evolution of energy consumption per benchmark/JVM are made available from the online repository.\footnote{\url{https://anonymous.4open.science/r/jvm-comparaison-213E/Readme.md}}
 
-Then, Figure~\ref{hotspot8} depicts the evolution of the energy consumption of the 12 benchmarks, when executed on the \textsc{HotSpot} JVM.
+Then, Figure~\ref{hotspot8} depicts the evolution of the energy consumption of the $12$ benchmarks, when executed on the \textsc{HotSpot} JVM.
 \Cref{fig:hotspot8} reports on the energy consumption variation of individual benchmarks, using to \textsc{HotSpot-8}\ as the baseline.
@@
 
-Given that the wide set of distributions and versions seems to highlight 3 classes of energy behaviors, the remainder of this paper considers the following distributions as relevant samples of JVM to be further evaluated: \textsf{20.2.0.r11-grl} (\textsc{GraalVM}), \textsf{15.0.1-open} (\textsc{HotSpot-15}), \textsf{15.0.21.j9} (\textsc{J9}).
+Given that the wide set of distributions and versions seems to highlight 3 classes of energy behaviors, the remainder of this chapter considers the following distributions as relevant samples of JVM to be further evaluated: \textsf{20.2.0.r11-grl} (\textsc{GraalVM}), \textsf{15.0.1-open} (\textsc{HotSpot-15}), \textsf{15.0.21.j9} (\textsc{J9}).
 We also decided to keep the \textsf{8.0.275-open} (\textsc{HotSpot-8}) as a baseline JVM for some figures to highlight the evolution of energy consumption over time/versions.
@@
 
-\subsubsection{J-Referral}
-To help developers and practitioners choosing an energy efficient JVM distribution and configuration, we propose a tool named \textsf{J-Referral}.
-This tool takes a Java application as an input and a launch script to tune the JVM configuration.
-The recommendation is is based on the 85 JVM distributions and versions made available by \textsf{SDKMAN!}.\footnote{\url{https://sdkman.io/}}
-Beyond the JVM distribution and version recommendation, \textsf{J-Referral} automatically explores many GC and JIT options for each distribution.
-Hundreds of combinations are thus compared to recommend an energy efficient JVM distribution and the associated configuration parameters.
+\section{Threats to Validity}\label{sec:threats}
+A number of issues may affect the validity of our work.
+First, we have the use of the Intel RAPL, one of the most accurate available tools to measure the energy consumption of software~\cite{Khan:2018:RAE:3199681.3177754,10.1145/2989081.2989088}.
+However, RAPL only gives the global energy consumption and no fine-grained measures at process or thread levels.
+We used baremetal hardware with a minimal OS and tuned off all the non-essential services and daemons to limit the overhead that the OS may add to the execution, even if it is not substantial~\cite{opaper}.
 
+Another measurement issue is the CPU energy variation within machines~\cite{opaper}, thus we executed all the comparable tests on the same node and with the recommended settings to mitigate this threat.
+
+Benchmarks execution time could also constitute on more subtle threat to the validity of our work, especially for some benchmarks that run fast, such as the \textsf{Pmd} benchmark.
+We thus gave a lot of attention on how long the benchmark is running for the hardware we used, and we tuned the input data workloads to execute benchmarks for at least many (from 10 to hundreds) seconds.
+Experiments ran at least 30 times to compute the average consumption and the associated standard deviation, therefore reasoning over reasonable dispersion around the average.
+
+How generalizable are our results? We believe that our study conclusions and guideline remain empirical, as we do not intend to generalize any result we obtained for some JVM or benchmark.
+We provide practitioners with some prerequisites to check before software deployment to reduce the software energy footprint by considering the JVM and its settings.
+
+\section{Conclusion}
+The results of our investigations showed that many JVMs share energy efficiencies and can grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
+The 3 selected JVM classes can however report a different energy efficiency for different software and/or workloads, sometimes by a large margin.
+While we did not observed a unique champion when it comes to energy consumption, \textsc{GraalVM} reported the best energy efficiency for a majority of benchmarks.
+Nonetheless, each JVM can achieve the best or the worst depending on the hosted application.
+One cause can be thread management strategies, as observed with J9 when advantageously running \textsf{Avrora}.
+Moreover, some JVM settings can cause energy consumption variations.
+Our experiments showed that the default JIT compiler of the JVM is often near-optimal, in at least ~80\% of our experiments.
+The default GC, however, was outperforming alternative strategies in half of our experiments, with some large gains observed when using some alternative GC depending on the application characteristics.
+
+Our main conclusions and guidelines can be thus summarized as:
+\begin{inparaenum}[\em i)]
+    \item testing software on the 3 classes of JVM and identifying the one that consumes the least is a good practice, especially for multi-threading purposes,
+    \item while the JVM default JIT give often good energy consumption results, some settings may improve the energy consumption and could be tested,
+    \item the choice of the GC may lead to a large impact on the energy consumption in many situations, thus encouraging a careful tuning of this parameter prior to deployment.
+\end{inparaenum}
+To ease the integration of the above guidelines, we propose a tool, named \textsf{J-Referral}, to recommend the most energy-efficient JVM distribution and configuration among more than a hundred considered possibilities.
+It establishes a full report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to help the user to choose the one with the least consumption for a Java Software.
+
+\section{ Tools and contributions }
 \Cref{table:j-referral} illustrates an example of the final report of \textsf{J-Referral}.
@@
 
-\section{Related Work}\label{sec:rw}
-Practitioners keep looking for development tools and methods to deliver software that meets both performance and quality requirements.
-In this context, energy consumption is becoming of growing importance when deploying software services in the cloud, which generally promotes a \emph{pay-as-you-go} pricing model, thus challenging the resource-efficiency of deployed services.
-So far, software engineering tackles this optimization challenge along three axes:
-\begin{inparaenum}
-    \item programming languages,
-    \item source code,
-    \item execution platforms
-\end{inparaenum}
-
-In this section, we review the state of the art of studies related to JVM energy consumption.
-
-\vspace{4pt}\noindent\textbf{At code level.}
-Many works investigated software energy consumption efficiency through source code changes and optimizations.
-For example, \cite{pinto_comprehensive_2016,fernandes_assisting_2017} studied the effect of Java collections on energy consumption, with regards to the collection size and/or the most executed tasks on the collection (insertion, removal, search).
-In~\cite{samir_hasan_energy_2016}, Hasan~\textit{et~al.} compared the energy consumption of several Java data structures, analyzing the bytecode using the Wala framework\footnote{\url{http://wala.sourceforge.net/wiki/index.php}} and assessing the evolution of the energy consumption in different scenarios (insertion at the beginning, iteration, etc.).
-They also used some automated replacement of \texttt{LinkedList} and \texttt{ArrayList} to simulate best- and worst-case energy consumption scenarios on real production applications.
-Their study showed that using inappropriate collection can cause an energy consumption inefficiency of 300\% for the worst-case scenario.
-
-SEEDS and SEEDS-API is a fully automated framework, proposed in~\cite{manotas_seeds_2014}, to analyze source code (at bytecode level for Java) and auto-tunes apps to have reduce their energy consumption, with a focus on Java collection tuning.
-The authors reported on an improvement up to 17\%.
-In~\cite{pereira_helping_2017}, the same authors presented SPELL, the energy leaks detector tool.
-The tool uses \textsc{JRapl}~\cite{alves_experiments_2020,pinto_comprehensive_2016} to detect energy-inefficient code fragments by using a statistical spectrum-based energy red spots localization.
-% It has the  advantage of being language and context independent.
-Their evaluation reported up to 18\% energy savings on Java applications.
-
-Some researchers tried to highlight the impact that has some atomic code changes on the energy consumption.
-Many papers studied the impact of code refactoring on energy consumption~\cite{DBLP:conf/seke/ParkHL14}.
-Other papers investigated the energy consumption of Java primitive types, operations on strings, usage of exceptions, loops, and arrays~\cite{kumar_energy_2017}.
-
-\vspace{4pt}\noindent\textbf{At JVM level.}
-In~\cite{pereira_energy_2017}, Pereira~\textit{et~al.} investigated several programming languages, including JVM-based languages like Java and Scala, and compared them on different dimensions (energy, memory, and time) using \textsf{CLBG} benchmarks.
-Their observation reported that compiled languages are more energy-efficient than interpreted ones, overall.
-More advanced results expose how the selected languages may satisfy one, two, or three dimensions of the equation.
-%The paper also tells finding no correlation between time and energy, so couldn't affirm or deny if "faster" is "greener". 
-The authors of~\cite{6122743} conducted a performance analysis and comparison between \textsc{HotSpot} and \textsc{J9}.
-They claimed in their results that the relative performance of \textsc{HotSpot} ranges from 44\% to 289\% of J9, while the dynamic power consumption varies from 2.7W to 7.2W using the \textsf{SPECjvm2008} benchmarks.
-
-The difference between \textsf{HotSpot} and \textsc{J9} was also reported in other studies.
-Chiba~\textit{et~al.}~\cite{8457806} evaluated the effect that could have those 2 JVM platforms on the performance of a combination of big data query engines (\textsc{Spark} and \textsc{Tez}) using \textsf{TPC-DS} benchmark.
-They reported on a 3-fold drawback that can exhibit one JVM compared to the other.
-
-On another note, the authors of~\cite{grass_energy_2006} attempted to assess and assign an energy cost to atomic bytecode instructions together with a constant overhead of using the JVM.
-Their claims are rather surprising, as the energy cost should be impacted by the execution environment, workload, core frequencies, and can hardly be fixed.
-A similar idea was used in~\cite{brownlee_search_based_2017} to design a model for JVM-based software energy consumption, using a bytecode-level model.
-The authors described their tool, named \textsf{OPACITOR}, as being deterministic, accurate, and robust to the surrounding noise.
-However, they disable the JIT in their experiments to maintain the deterministic behavior of their tool, which does not reflect a real software execution given all the optimizations that the JVM triggers to optimize the performances.
-
-%\section{Threats to Validity}\label{sec:threats}
-%A number of issues may affect the validity of our work.
-%First, we have the use of the Intel RAPL, one of the most accurate available tools to measure the energy consumption of software~\cite{Khan:2018:RAE:3199681.3177754,10.1145/2989081.2989088}.
-%However, RAPL only gives the global energy consumption and no fine-grained measures at process or thread levels.
-%We used baremetal hardware with a minimal OS and tuned off all the non-essential services and daemons to limit the overhead that the OS may add to the execution, even if it is not substantial~\cite{opaper}.
-%
-%Another measurement issue is the CPU energy variation within machines~\cite{opaper}, thus we executed all the comparable tests on the same node and with the recommended settings to mitigate this threat.
-%
-%Benchmarks execution time could also constitute on more subtle threat to the validity of our work, especially for some benchmarks that run fast, such as the \textsf{Pmd} benchmark.
-%We thus gave a lot of attention on how long the benchmark is running for the hardware we used, and we tuned the input data workloads to execute benchmarks for at least many (from 10 to hundreds) seconds.
-%Experiments ran at least 30 times to compute the average consumption and the associated standard deviation, therefore reasoning over reasonable dispersion around the average. 
-%
-%How generalizable are our results? We believe that our study conclusions and guideline remain empirical, as we do not intend to generalize any result we obtained for some JVM or benchmark.
-%We provide practioners with some prerequisites to check before software deployment to reduce the software energy footprint by considering the JVM and its settings.
-
-\section{Conclusion}\label{sec:ccl}
-This paper reports on an empirical investigation of the key differences in energy consumption that some of the most famous and supported JVM platforms can exhibit, in addition to the key settings that can impact this energy consumption positively or negatively.
-During our experiments, we considered a total of 12 well-known and diversified-purposes Java benchmarks together with a total of 52 JVMs, including many versions of 11 different distributions.
-The results of our investigations showed that many JVMs share energy efficiencies and can grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
-The 3 selected JVM classes can however report a different energy efficiency for different software and/or workloads, sometimes by a large margin.
-While we did not observed a unique champion when it comes to energy consumption, \textsc{GraalVM} reported the best energy efficiency for a majority of benchmarks.
-Nonetheless, each JVM can achieve the best or the worst depending on the hosted application.
-One cause can be thread management strategies, as observed with J9 when advantageously running \textsf{Avrora}.
-Moreover, some JVM settings can cause energy consumption variations.
-Our experiments showed that the default JIT compiler of the JVM is often near-optimal, in at least ~80\% of our experiments.
-The default GC, however, was outperforming alternative strategies in half of our experiments, with some large gains observed when using some alternative GC depending on the application characteristics.
-
-Our main conclusions and guidelines can be thus summarized as:
-\begin{inparaenum}[\em i)]
-    \item testing software on the 3 classes of JVM and identifying the one that consumes the least is a good practice, especially for multi-threading purposes,
-    \item while the JVM default JIT give often good energy consumption results, some settings may improve the energy consumption and could be tested,
-    \item the choice of the GC may lead to a large impact on the energy consumption in many situations, thus encouraging a careful tuning of this parameter prior to deployment.
-\end{inparaenum}
-To ease the integration of the above guidelines, we propose a tool, named \textsf{J-Referral}, to recommend the most energy-efficient JVM distribution and configuration among more than a hundred considered possibilities.
-It establishes a full report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to help the user to choose the one with the least consumption for a Java Software.
+% TODO : add or change the graphes of the use case 
+% TODO : add the description and a diagrame for the experimental protocol

### 2022-05-18 chapters/testing/testing.tex
draft. first chapter
Proofreading benchmarks

--- 2021-08-30 chapters/testing/testing.tex
@@
 \newpage
-\chapter{Testing}
+
+\chapter{Empirical Evaluation Protocol}
 \label{chapter:testing}
 
-In order to optimize the energy consumption of software we first want to setup a testing enviroment.
-This chapter will be dedicated to provide a set of guidelines to provide a \em{reproducible, accurate} and \em{representative} tests.
-Each aspect will be detailed in the sections below
+To optimize the software energy consumption, a reproducible setup environment requires to be designed.
+I chose to adopt an empirical approach to explore energy consumption reductions through a trial and error methodology.
 
-\import{chapters/testing/}{reproducibility}
-\import{chapters/testing/}{accuracy}
-\import{chapters/testing/}{representativeness}
+%  This chapter will be dedicated to provide a set of guidelines to provide a \em{reproducible, accurate} and \em{representative} tests. Each aspect will be detailed in the sections below.
+
+\import{chapters/testing/}{challenges}
+\import{chapters/testing}{stateoftheart}
+\import{chapters/testing/}{environement}
+\import{chapters/testing/}{experement}
+\import{chapters/testing/}{conclusion}

### 2022-05-23 chapters/literature/draft_citations.tex
refactoring : structure

@@
 
-\section{testing}
+\section{benchmarking}
 
@@
 In the follwong section we will discuss some of the well known tools that are used in literature.
-
-% TODO : add a triangle to classify all those tools
 
@@
 emprical study \cite{de2021empirical}
+
+\section{Conclusion}
+As we have seen in the state of the art, many methodes have been proposed to reduce the energy footprint of the the ICT,which they can be applied in deffirent parts of the lifecycle of the program, from consumption to the execution.
+Furthermore, the execution phase took attention of many researchers because it's the part where the most energy is consumed.
+This thesis will focus on that aspect as well.
+However, unlike the most work that have been done on the hardware aspect, we will target the software impact on this energy, starting from the choice of the programming language up to how to tune some features of a frameworks in order to make the software consumes less energy.To do so we use the emprical apporach due to it's concinstency for the moment.Unlike the perfromance which essentialy related to the complexity of the algorithm, the energy consumption is more impacted by the hardware.
+Therefore, to optimize the enrgy consumption we choose a spiral methode. based on 3 phases:
+\begin{enumerate}
+    \item execute the code
+    \item measure the program
+    \item infere the guidelines
+\end{enumerate}
+
+\begin{figure}[!hbt]
+    \center{\includegraphics[scale=0.3]{images/greenbenchmarkcycle}}
+    \caption{the spiral methode of energy optimization }\label{fig:spirals}
+\end{figure}
+the aim of this work is to present a set of guidelines to create a benchmarking system to to measure the energy consumption of different programs.
+After that, we will use this system to compare the energy consumption of different programming languages.
+We will extend the work of \citeauthor{pereira2017energy} to a closer distance to production environment by comparing a set of usecases.
+starting by GRPC framework, and a set of Web Framewroks.
+Finally we will discuss the impact of the execution environment on the energy consumption of two of the most famous programming languages JAVA and Python.
+and present how can tunning the Virtual Machine can reduce the energy consumption.

### 2022-05-25 chapters/execution_environement/python/python.tex
refactoring : structure
add : uncompleted papers

@@
+\section{python}
+
 \subsection{Introduction}
-% reasons why we have chosen python
 
+When it comes to developing software systems, one can notice that---except Perl---dynamic programming languages have taken over compiled languages along the last decade in terms of developers popularity (cf.
+Figure~\ref{fig:pypl}).
+However, it remains unclear if this class of dynamic programming languages can reasonably compete with compiled ones when it comes to power efficiency.
+
+\begin{figure}[htbp]
+    \includegraphics[width=\linewidth]{imgs/programminglanguangespopularity.png}
+    \caption{PYPL Popularity of Programming Languages~\cite{noauthor_pypl_nodate}}
+    \label{fig:pypl}
+\end{figure}
+
+In particular, Noureddine~\emph{et~al.}~\cite{noureddine_preliminary_2012} in 2012, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
+According to their experiments, an interpreted programming language like Python can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. Figure~\ref{fig:hannoi}).
+% However, the operational conditions under which these measurements may threaten the validity of their results.
+
+\begin{figure}[htbp]
+    \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
+    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
+    \label{fig:hannoi}
+\end{figure}
 \subsection{python popularity}
+
+However, Python seems to attract a large community of developers interested in data analysis, web development, system administration, machine learning---according to a survey conducted in 2018 by JetBrains---\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}}one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.
+
+As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
+Rather, we believe that a better option consists in leveraging the strength of this rich ecosystem to promote energy-efficient solutions in order to improve the power consumption of legacy software systems.
+
+In this chapter, we therefore explore the oblivious optimizations that can be applied to Python legacy applications in order to reduce their energy footprint.
+As Python is widely adopted by software services deployed in public and private Cloud infrastructures, we believe that our contributions will benefit to wide diversity of legacy systems and not only favorably contribute to reduce the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
+More specifically, this chapter focuses on runtime optimizations that can adopted by developers to leverage the power consumption of Python applications.
+These optimizations include not only alternative interpreters, but also \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
+
+% The contributions of this chapter can therefore be summarized as:
+% \begin{compactenum}
+%     \item a classification of state-of-the-art Python optimizations,
+%     \item an evaluation of the power efficiency of these optimizations on well-known micro-benchmarks,
+%     \item a comparison of the effective energy efficiency of Python compared to compiled languages,
+%     \item an assessment of the energy speedup of these optimizations on representative applications and workloads.
+% \end{compactenum}
 
@@
 
-\subsection{environements}
-in this section we will describe the chosen python interpreter
+he simplicity, flexibility and the clarity of Python code led this programming language to a huge success.
+Therefore, a larger community and a wide range of support and libraries that allowed it to touch several fields like data analyses, machine learning, natural language processing, web development, astrophysics etc.
+% python is not efficient 
+% 1 Energy Efficiency Across Programming Languages: How Do Energy, Time, and Memory Relate?
+%  \begin{figure}[hbt]
+%     \includegraphics[width=\linewidth]{figures/hannoiimplementation.png}
+%     \caption{Energy consumption of the recursive implementation of Tower of Hanoi program in different languages (using a base 10 logarithmic scale) \cite{noureddine_preliminary_2012}}      
+%     \label{fig:hannoi}
+% \end{figure}
+
+However, and according to \cite{pinto_energy_2017,noureddine_preliminary_2012} and [1].
+Python tends to be one more energy hungry .
+In figure \ref{fig:hannoi}, python consumes 30 times more than C or C++.
+The test was done on an implementation of \fnurl{hannoi test}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi}  30 disks.
+
+Python consume a lot of energy mainly because it is slow in execution.
+Its Flexibility and simplicity caused it to drop off in performances, because Python gains its flexibility from being a dynamic language, therefore it needs an interpreter to execute its programs, which makes them much slower compared to the other ones that are written in compiled programming languages such as C and C++ or semi-compiled languages like Java.
+
+As shown in Table \ref{fig:clbg}, we can see that in the most of the test cases Python takes more time to execute -the only case that he wasn't the last one was in the benchmark regx-redux where he beaten Go-, and in some cases the gab was huge such as in the n-body test where python took around 100 times more than C++.
+
+\begin{table}[hbt]
+    \begin{tabular}{l|*{5}c}
+                           & C    & C++   & Java & python & Go    \\
+        \hline
+        pidigits           & 1.75 & 1.89  & 3.13 & 3.51   & 2.04  \\
+        reverse-complement & 1.75 & 2.95  & 3.31 & 16.76  & 4.00  \\
+        regx-redux         & 1.45 & 1.66  & 10.5 & 15.56  & 28.69 \\
+        k-nucleotide       & 5.07 & 3.66  & 8.66 & 79.79  & 15.36 \\
+        binary-trees       & 2.55 & 2.63  & 8.28 & 92.72  & 28.90 \\
+        fasta              & 1.32 & 1.33  & 2.32 & 62.88  & 2.07  \\
+        Fannkuch-redux     & 8.72 & 10.62 & 17.9 & 547.23 & 17.82 \\
+        n-body             & 9.17 & 8.24  & 22.0 & 882.00 & 21.00 \\
+        spectral-norm      & 1.99 & 1.98  & 4.27 & 193.86 & 3.95  \\
+        Mandelbort         & 1.64 & 1.51  & 6.96 & 279.68 & 5.47
+    \end{tabular}
+    \caption{Comparison of the execution time between different programming languages using the computer language benchmarks game \cite{noauthor_which_nodate}}
+    \label{fig:clbg}
+\end{table}
+
+\subsection{Python Runtimes}
+
+we used the following implementations.
+
+We should explain the methodology we used to select these
+
+\begin{enumerate}
+    \item \textbf{CPython}:\footnote{\fnurl{https://www.python.org/}} a Python interpreter written in C programming language.
+It is the reference interpreter of Python.
+CPython compiles the source code into a bytecode and then interprets it.
+The CPython project still supports both versions of Python 2 and 3;
+    \item \textbf{PyPy}:\footnote{\url{http://pypy.org}} An alternative implementation of the Python interpreter.
+It is written using \emph{RPython} in order to use the JIT.
+It compiles the most used portions of the Python code into a binary code for better performance.
+To benefit from these optimizations, the program has to be executed for at least for few seconds so the JIT has enough time to warm up, the JIT optimization will be applied only the code written by the programmer and not external libraries;
+    \item \textbf{Cython}:\footnote{\url{https://github.com/cython/cython}} a static compiler for Python.
+It translates the Python code into a C code, and then compiles it using a C compiler.
+It also support(s) an extended version of the Python language that allows programmers to call \emph{C functions}, declare \emph{C types} and use static types which will help the translation of Python objects into native types, such as integers, float, which often means a better performances, since native C libraries are almost all the time faster than the python written once~\cite{pereira_energy_2017}.
+    \item \textbf{Intel\,Python}:\footnote{\url{https://software.intel.com/en-us/distribution-for-python}} A customized interpreter developed by Intel in order to enhance performances of Python programs.
+It is dedicated to data sciences, high-Performance computing.
+It uses some Intel kernel libraries, such as Math Kernel Library (Intel MKL\footnote{\url{https://software.intel.com/en-us/mkl}}) and data analytics acceleration library (Intel DAAL\footnote{\url{https://software.intel.com/en-us/intel-daal}}).
+It supports both versions of Python;
+    \item \textbf{Active\,Python}:\footnote{\url{https://www.activestate.com/products/activepython/}} developed by the Activestates company, it provides a standardized Python distribution to ensure license compliance, security, compatibility and performance.
+Therefore, ActivePython has its own built-in packages (more than 300 packages) and supports both versions of Python;
+    \item \textbf{IronPython}:\footnote{\url{https://ironpython.net}} A .Net-based Python interpretation platform written in C\# in order to be used over the .Net vm or mono, it benefits from all the optimizations of .Net virtual machine, such as the JIT and garbage collector mechanisms.
+    \item \textbf{GraalPython}:\footnote{\url{https://github.com/graalvm/graalpython/}} A Python interpreter that is based on GraalVM \footnote{\url{https://www.graalvm.org/docs/why-graal/}} (a universal virtual machine developed by oracle for running applications written in different programming languages).
+For the time being, it only supports Python~3 and it is still in the experimental stage;
+    \item \textbf{Jython}:\footnote{\url{https://jython.github.io}} Implementation of Python programming language written in Java for the \emph{Java virtual machine} (JVM).
+Similar to IronPython and GraalPython, it leverages the optimization mechanisms provided by the JVM to enhance the Python programs;
+    \item \textbf{MicroPython}:\footnote{\url{http://micropython.org}} a lightweight Python version dedicated to embedded systems and micro-controllers;
+    \item \textbf{Nuitka}:\footnote{\url{http://nuitka.net/pages/overview.html}} a Python compiler written in Python that generates a binary executable from Python code.
+It translates the Python code into a C program that is then compiled into a binary executable;
+    \item \textbf{Numba}:\footnote{\url{https://numba.pydata.org}} a library that includes JIT compiler in order to enhance the performances of Python functions using the industry-standard LLVM compiler library;
+    \item \textbf{Shedskin}:\footnote{\url{https://github.com/shedskin/shedskin}} a static transpiler that translates implicitly statically typed python into C++ code.
+    \item \textbf{Hope}~\cite{akeret_hope_2015}: a Python library that aims to introduce JIT compiler into the Python code;
+    \item \textbf{Parakeet}~\cite{DBLP:conf/hotpar/RubinsteynHWS12}: a runtime accelerator for an array-oriented subset of Python;
+    \item \textbf{Stackless\,Python}:\footnote{\url{https://github.com/stackless-dev/stackless/wiki}} an interpreter that focus on enhancing multi-threading programming.
+    \item \textbf{Pyjion}:\footnote{\url{https://github.com/microsoft/pyjion}} JIT API for CPython, same purpose as Parakeet and Hope.
+    \item \textbf{Pyston}:\footnote{\url{https://blog.pyston.org}} performance-oriented Python implementation built using LLVM and modern JIT techniques. the project is funded by dropbox;
+    \item \textbf{Grumpy}:\footnote{\url{https://github.com/google/grumpy}} a source-to-source transpiler that translates the Python code into a Go code that will be compiled to a binary executable.
+It also offers an interpreter called \emph{grumprun} which can directly execute the Python code.
+Unfortunately, we cannot use it because the project is already outdated (last commit is in 2017) and it has a lot of limitation in term of supporting the Python language, such as some built-in functions and standard libraries;
+    \item \textbf{Psyco}:\footnote{\url{http://psyco.sourceforge.net}} a JIT compiler for Python;
+    \item \textbf{Unladen\,Swallow}:\footnote{\url{https://unladen-swallow.readthedocs.io/en/latest/}} an attempt to (use) LLVM  as  JIT compiler for CPython.
+\end{enumerate}
+
+\subsection{Runtime Classification}
+Before futher proceeding with the list of candidate runtimes for Python applications, we propose a classification according to several criterions:
+\begin{description}
+    \item[\bf Type] refers to the category of runtime infrastructure that supports the execution of a Python application.
+In particular, we consider 3 types of environments: \emph{Interpreter}, \emph{Compiler} and \emph{Library}.
+\emph{Interpreter} refers to the class of environment that does not require any preprocessing of Python source code.
+\emph{Compiler} introduces a compilation phase prior to the execution of the application.
+Finally, \emph{Library} induces some modification of the source code.
+    \item[\bf Runtime] refers to the technology supporting the execution of a Python application.
+This technology can refer to the programming language used to program the interpreter, the target language for a compiler or a library.
+    \item[\bf JIT optimisation] refers to the support of \emph{just-in-time} compilation in the runtime infrastructure supporting the execution of the application.
+    \item[\bf GC optimisation] refers to the support of \emph{garbage collection} in the runtime infrastructure supporting the execution of the application.
+    \item[\bf Python version(s)] refers to the list of Python source code versions supported by the runtime environment.
+\end{description}
+
+We should explain the classification of these runtimes
+
+\begin{table}
+    \caption{Classification of Python implementations}
+    \label{fig:python-classes}
+    \center
+    \begin{tabular}{|l|c|c|c|c|c|c|}
+        \hline
+        \multirow{2}{*}{\bf Name} & \multirow{2}{*}{\bf Type} & \multirow{2}{*}{\bf Runtime} & \multicolumn{2}{c|}{\bf Optimisations} & \multicolumn{2}{c|}{\bf Python}                     \\
+        % \cline{1-5} 
+                                  &                           &                              & {\bf JIT}                              & {\bf GC}                        & {\bf 2} & {\bf 3} \\
+        \hline
+        \hline
+        CPython                   & Interpreter               & C                            & \no                                    & \no                             & \yes    & \yes    \\
+        \hline
+        Intel\,Python             & Interpreter               & C                            & \no                                    & \no                             & \yes    & \yes    \\
+        \hline
+        ActivePython              & Interpreter               & C                            & \no                                    & \yes                            & \yes    & \yes    \\
+        \hline
+        PyPy                      & Interpreter               & Python                       & \yes                                   & \yes                            & \yes    & \yes    \\
+        \hline
+        IronPython                & Interpreter               & .Net                         & \yes                                   & \yes                            & \yes    & \yes    \\
+        \hline
+        GraalPython               & Interpreter               & GraalVM                      & \yes                                   & \yes                            & \no     & \yes    \\
+        \hline
+        Jython                    & Interpreter               & Java                         & \yes                                   & \yes                            & \yes    & \no     \\
+        \hline
+        Stackless\,Python         & Interpreter               & python                       & \no                                    & \no                             & \yes    & \no     \\
+        \hline
+        MicroPython               & interpreter               & c                            & \no                                    & \no                             & \no     & \yes    \\
+        \hline
+        Pyston                    & interpreter               & LLVM                         & \yes                                   & \no                             & \yes    & \no     \\
+        \hline
+        Unladen\,Swallow          & Interpreter               & LLVM                         & \yes                                   & \no                             & \yes    & \no     \\
+        \hline
+        Cython                    & Compiler                  & C                            & \no                                    & \no                             & \yes    & \yes    \\
+        \hline
+        Nuitka                    & Compiler                  & C                            & \no                                    & \no                             & \yes    & \yes    \\
+        \hline
+        Shedskin                  & Compiler                  & C++                          & \no                                    & \no                             & \yes    & \yes    \\
+        \hline
+        Grumpy                    & Compiler                  & Go                           & \no                                    & \no                             & \yes    & \yes    \\
+        \hline
+        Numba                     & Library                   & C                            & \yes                                   & \no                             & \yes    & \yes    \\
+        \hline
+        Hope                      & Library                   & python                       & \yes                                   & \no                             & \yes    & \yes    \\
+        \hline
+        Psyco                     & Library                   & python                       & \yes                                   & \no                             & \yes    & \yes    \\
+        \hline
+        Pyjion                    & Library                   & .NET Core                    & \yes                                   & \no                             & \yes    & \yes    \\
+        \hline
+        Parakeet                  & library                   & C                            & -                                      & \no                             & \yes    & \no     \\
+        \hline
+    \end{tabular}
+\end{table}
+
+\begin{table}[hbt]
+    \caption{Classification of Python implementations}
+    \label{fig:pythonimplementations}
+    \center
+    \begin{tabular}{|l|c|c|c|}
+        \hline
+        Version  & Interpreter     & Transpiler/Compiler & Jit library \\
+        \hline
+        \hline
+        Python~2 & Cpython2        & Cython2             & Numba 2     \\
+        % \cline{2-4} 
+                 & Pypy2           & Shesdskin           & Hope        \\
+        % \cline{2-4} 
+                 & Pytson          & Grumpy              & Parakeet    \\
+                 & Ironpython      &                     & Psyco       \\
+                 & Jython          &                     & Pyjion      \\
+        % &  &  &  \\
+                 & Micropython     &                     &             \\
+                 & Pysec           &                     &             \\
+                 & StacklessPython &                     &             \\
+        % \cline{2-4} 
+        \hline
+        Python~3 & Cpython3        & Nuitka              & Numba3      \\
+                 & Pypy3           &                     &             \\
+                 & GraalPython     &                     &             \\
+        \hline
+    \end{tabular}
+\end{table}
+
+There are other implementations that we didn't considerate because either the project aborted many years ago or it has a very limited support for python features.
+After the collection of those implementation we filtered them. in order to keep only the versions that are still maintained and support most of python features.
+and we classified them into 3 categories depending of their integration with the python code.
+In the \ref{fig:pythonimplementations} we describe the implementations that we kept and the version of each implementation and its category.
 
@@
 \end{figure}
+
+\subsection{tests preparation}
+
+To study the behaviour of the python implementation regarding the energy consumption, we have to focus on the effect of the implementation and mitigate the maximum any side effects such as the organisation of the code or any extra consumption due to the operating system or tier libraries.
+Therefore, for each test we took the implementation written in python2 as a reference and tried to use it in other implementations as it is.
+If it is not supported by python3, we transformed the code using the officiel library  \emph{2to3}\footnote{\url{https://docs.python.org/3.7/library/2to3.html}}.
+In the case of the libraries that use \emph{JIT} adding a decorator to the function that we want to optimize was enough, if there are other changes we assume that they alter the original code which is against our purpose.
+
+Each test is implemented in a Docker container for the several reasons
+\begin{itemize}
+    \item Isolation: each container has only the test program implemented with a single python runtime to remove any interference between different implementations.
+    \item Deplpoyment: to use the testing machine without extra configurations that may alter the behaviour of the os toward the enrgy consumtion
+    \item Reproductibility: One of the most frequent benchmark crimes~\cite{DBLP:journals/corr/abs-1801-02381} in research is the lack of Reproductibility, by using Docker we ensure that each test has an Image that will be accessible in public.
+\end{itemize}
+
+Despite the presence of the official docker images for the most of the runtimes, we prefered to build our own using the same reference image in order to remove any bias due to the Os used in the offical image.We used ArchLinux with the kernel version 4.9.184 as a base image.
+
+\begin{table}[hbt]
+    \begin{tabular}{ll}
+        \hline
+        CPU     & Intel Xeon Gold 6130 (Skylake, 2.10GHz, 2 CPUs/node, 16 cores/CPU)                                                     \\
+        Memory  & 192 GiB                                                                                                                \\
+        Storage & 240 GB SSD SATA Samsung MZ7KM240HMHQ0D3                                                                                \\
+                & 480 GB SSD SATA Samsung MZ7KM480HMHQ0D3                                                                                \\
+                & 4.0 TB HDD SATA Seagate                                                                                                \\
+        Network & eth0/enp24s0f0, Ethernet, configured rate: 10 Gbps, model: Intel Ethernet Controller X710 for 10GbE SFP+, driver: i40e \\
+                & ib0, Omni-Path, configured rate: 100 Gbps, model: Intel Omni-Path HFI Silicon 100 Series [discrete], driver: hfi1      \\
+        \hline
+    \end{tabular}
+    \caption{Testing Machine Configuration }
+    \label{fig:dahuconfig}
+\end{table}
+
+\subsection{experimental setup}
+
+all our tests have been executed in a Dell PowerEdge C6420 server machine.
+A summary of its hardware is listed in table \ref{fig:dahuconfig}.
+The machine is equiped with a minimal version of Debian\,9 (4.9.0 kernel version) where we install Docker (version 18.09.5).
+
+\begin{figure}[hbt]
+    \includegraphics[width=\linewidth]{imgs/SmartWatts.png}
+    \caption{powerapi architecture}
+    \label{fig:powerapi}
+
+\end{figure}
+\subsection{Power Measurements}
+
+As we know, the energy of a program is the integrale of its power overtime.
+For us ower case, we used Intel \emph{Running Power Average Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754} to collect the power samples of the running tests,
+
+We used used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, to report Data collected by intel Rapl and send it into another machine that we call computing machine.
+then we calculate the Energy using  the the trapezoidal rule.
+
+\ref{fig:trapezrule}
+\begin{figure}[hbt]
+    \centering
+    \begin{equation}
+        E = \int^a_b P(t)dt \simeq \sum^n_{k=1} \frac{P(t_k-1)+P(t_k)}{2}
+    \end{equation}
+    \caption{}
+    \label{fig:trapezrule}
+\end{figure}
+
+figure \ref{fig:powerapi} shows the architecture of our testing model.
+
+The reason of separation between data collection and energy calculation is to minimise any interference with the test so our sensor is a light c program running inside a docker container.
 

### 2022-05-25 chapters/programming_languages/web_frameworks.tex
add : uncompleted papers

\section{Introduction}
Nowadays, web applications are dominating online systems.
From Google, to Facebook and others, web applications are widely deployed across organizations and continuously accessed by end-users, both for their personal and professional daily tasks.
In practice, the development of these web applications heavily rely on a wide ecosystem of \emph{web frameworks}, which are intended to ease and foster the development process.
However, once deployed, the applications developed with such web frameworks do not exhibit the same performances, as reported by the \emph{Web Framework Benchmarks} periodically published by the \textsc{TechEmpower} company.\footnote{\url{https://www.techempower.com/benchmarks}}
Thanks to such benchmarks, developers can take informed decisions on the most performing technology to adopt to implement their web applications.
Unfortunately, one can regret that developers and benchmark providers focus mostly on popularity and performance criteria when picking a web framework, with less considerations for the resource consumption implications of their choice.
This is all the more regrettable that cloud providers are more and more adopted by developers to host these web applications.
While cloud providers offer a convenient elastic provision of resources to scale according to application requirements, this convenience may induce critical cost for their business.
% , their pricing model focuses on this resource consumption by charging the application owner.

Beyond the economical cost of web applications, one can also question the more global impact of web applications on worldwide carbon emissions.
Given the tremendous success of web applications, their deployment has severely increased over last years, thus causing a rebound effect on the power consumption of server infrastructure---being hosted or supported by cloud providers.
While one can challenge the relevance of features that are continuously deployed by developers to keep engaging end-users, reconciling economical and environmental concerns remains an open challenge to address.

Given this context, this paper intends to contribute to this challenge by investigating the energy footprint of web frameworks.
In particular, we aim to support the developers of web applications with relevant guidelines that can help them to choose the web framework that is not only the most popular or provide the best performances, but also exhibits a low energy footprint.
By minimizing the energy consumed to process user requests, with no service quality penalty, developers can reduce the operational cost of their web applications and contribute to reduce worldwide carbon emissions of ICT.
% While we assume this choice does not conflict with the appropriate selection of relevant user features, this topic remains out of the scope of this paper.

To achieve this objective, we leverage the \textsc{TechEmpower} \emph{Web Framework Benchmarks} to incorporate server-side energy measurements obtained from a software-defined power meter, named \textsc{PowerAPI}~\cite{}.
These measurements are then analyzed in depth to understand the key criteria that can amper the power consumption of web frameworks and derive guidelines to can help developers to pick the most energy efficient web frameworks according to their requirements.

The remainder of this paper is organized as follows.

\subsection{Comparaison between web frameworks}

Many studies have been conducted to compare the performance of web frameworks.

We can cite \cite{gajewski_analysis_2019} where they made a comparison between two of the most famous java frameworks, Play and Spring, or the work of \cite{benmoussa_new_2019} when they compare different PHP frameworks using 6 creterions intrinsic durability, industrialized solution, technical adaptability, strategy, technical architecture, and Speed, In the previous paper we find that all those 6 creterions have their own wait when it comes to choose which framework to take for a project. In our study we want to push a 7th criterion that impacts the economic outcome of the project.

\subsection{energy efficiency in software engineering}

In their paper \cite{pereira_energy_2017} the authors studied the impact of programming languages on energy, time, and Memory by using the CLBG benchmark,
where they executed 10 different benchmarks [ add citation] accross 27 wel-know programming languages [ add reference]
the work of the authors was an extension of the a research marled by [ add sitation[6 in the authers paper ]]. we continued  their work  with measuring the impact of programming languages choice in real life application instead of microbenchmark

Irene et al.\cite{manotas_investigating_2013} investigated the impact of web servers on energy when handleling web applications . they analysed 7 applications executed within 4 serveces with 38 different scenarios . the authors showed that the energy greatly depends on the web server , howerver the impact of the application may influence this energitical behaviour of the server.

In their approach they used measured the energy consumption during the integration tests and for us we measure with a simulation of a website when there is a client in another machine that requests the website, therefore we isolate the energy consumption of the server from the the client's one.

%notes 
% instead of focusing on energy or execution time we change the paradigm into efficiency and average power 
Other works have been done on the client side, as an example \cite{philippot_characterization_2014} as they concluded that theres is a variation among the different websites and the impact of the browser on this energy.

\begin{longtable}{lll}
    \toprule
    framework        & language    & alternative implementations                        \\
    \midrule
    \endhead
    \midrule
    \multicolumn{3}{r}{{Continued on next page}}                                        \\
    \midrule
    \endfoot

    \bottomrule
    \endlastfoot
    pico.v           & V           &                                                    \\
    dylan            & Dylan       &                                                    \\
    sinatra-sequel   & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
    agoo             & Ruby        &                                                    \\
    rack             & Ruby        & unicorn falcon                                     \\
    h2o\_mruby       & Ruby        &                                                    \\
    rails            & Ruby        & postgresql unicorn                                 \\
    grape            & Ruby        & unicorn                                            \\
    roda-sequel      & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
    rack-sequel      & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
    sinatra          & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
    padrino          & Ruby        & unicorn                                            \\
    ninglex          & Common Lisp &                                                    \\
    ningle           & Common Lisp &                                                    \\
    woo              & Common Lisp &                                                    \\
    grails           & Groovy      &                                                    \\
    hot              & Groovy      & mongodb postgres mysql                             \\
    plack            & Perl        & async                                              \\
    kelp             & Perl        & mongodb                                            \\
    dancer           & Perl        &                                                    \\
    mojolicious      & Perl        &                                                    \\
    web-simple       & Perl        &                                                    \\
    cowboy           & Erlang      &                                                    \\
    elli             & Erlang      &                                                    \\
    chicagoboss      & Erlang      &                                                    \\
    mochiweb         & Erlang      &                                                    \\
    opium            & OCaml       & haproxy                                            \\
    morph            & OCaml       & static                                             \\
    webmachine       & OCaml       & haproxy                                            \\
    vertx-web-scala  & Scala       &                                                    \\
    http4s           & Scala       &                                                    \\
    akka-http        & Scala       & slick-postgres                                     \\
    fintrospect      & Scala       &                                                    \\
    finagle          & Scala       &                                                    \\
    colossus         & Scala       &                                                    \\
    finatra          & Scala       &                                                    \\
    blaze            & Scala       &                                                    \\
    play2-scala      & Scala       & netty anorm anorm-netty reactivemongo reactiv...   \\
    finch            & Scala       &                                                    \\
    scalene          & Scala       &                                                    \\
    cask             & Scala       &                                                    \\
    youi             & Scala       &                                                    \\
    httpbeast        & Nim         &                                                    \\
    prologue         & Nim         &                                                    \\
    basolato         & Nim         &                                                    \\
    jester           & Nim         &                                                    \\
    falco            & F\#         &                                                    \\
    frank            & F\#         &                                                    \\
    suave            & F\#         &                                                    \\
    giraffe          & F\#         & utf8json utf8direct stripped                       \\
    zebra            & F\#         & simple                                             \\
    vapor            & Swift       & fluent sql-kit postgres                            \\
    swift-nio        & Swift       &                                                    \\
    perfect          & Swift       & mysql postgresql mongodb                           \\
    kitura           & Swift       & postgres postgres-orm postgres-orm-codable mo...   \\
    racket           & Racket      &                                                    \\
    vsgi             & Vala        &                                                    \\
    valum            & Vala        &                                                    \\
    wsgi             & Python      &                                                    \\
    hug              & Python      &                                                    \\
    crax             & Python      &                                                    \\
    falcon           & Python      & py3 pypy2                                          \\
    uvicorn          & Python      &                                                    \\
    flask            & Python      & raw pypy pypy2 pypy2-raw nginx-uwsgi               \\
    responder        & Python      &                                                    \\
    klein            & Python      &                                                    \\
    blacksheep       & Python      &                                                    \\
    apidaora         & Python      & core                                               \\
    bottle           & Python      & pypy2 raw nginx-uwsgi                              \\
    japronto         & Python      &                                                    \\
    api\_hour        & Python      & mysql json dbs plaintext                           \\
    weppy            & Python      & py3 pypy2 nginx-uwsgi                              \\
    starlette        & Python      &                                                    \\
    emmett           & Python      &                                                    \\
    sanic            & Python      &                                                    \\
    aiohttp          & Python      & pg-raw                                             \\
    tornado          & Python      & pypy2 postgresql-raw py3 py3-uvloop                \\
    uwsgi            & Python      & nginx-uwsgi                                        \\
    pyramid          & Python      & py2                                                \\
    morepath         & Python      &                                                    \\
    vibora           & Python      &                                                    \\
    web2py           & Python      & optimized                                          \\
    django           & Python      & postgresql                                         \\
    cherrypy         & Python      & py3                                                \\
    eve              & Python      &                                                    \\
    webware          & Python      &                                                    \\
    spyne            & Python      & raw nginx-uwsgi                                    \\
    fastapi          & Python      & orjson                                             \\
    quart            & Python      & uvicorn                                            \\
    turbogears       & Python      &                                                    \\
    m-web-server     & mumps       &                                                    \\
    urweb            & Ur          & mysql cache mysql-cache                            \\
    rouille          & Rust        &                                                    \\
    warp-rust        & Rust        &                                                    \\
    thruster         & Rust        &                                                    \\
    ntex             & Rust        & db raw                                             \\
    iron             & Rust        &                                                    \\
    actix            & Rust        & core raw diesel pg                                 \\
    roa              & Rust        & core tokio diesel pg sqlx                          \\
    saphir           & Rust        &                                                    \\
    may-minihttp     & Rust        &                                                    \\
    rocket           & Rust        &                                                    \\
    tokio-minihttp   & Rust        &                                                    \\
    nickel           & Rust        &                                                    \\
    hyper            & Rust        & db                                                 \\
    gotham           & Rust        &                                                    \\
    kooby            & Kotlin      &                                                    \\
    pronghorn        & Kotlin      &                                                    \\
    http4k           & Kotlin      & apache apache4 jetty ktorcio ktornetty netty ...   \\
    ktor             & Kotlin      & jetty cio jasync reactivepg                        \\
    hexagon          & Kotlin      & resin-mongodb jetty-postgresql resin-postgresql    \\
    elixir-plug-ecto & Elixir      &                                                    \\
    phoenix          & Elixir      &                                                    \\
    fiber            & Go          & prefork                                            \\
    webgo            & Go          &                                                    \\
    gnet             & Go          &                                                    \\
    goframe          & Go          &                                                    \\
    revel            & Go          & fast raw jet qbs                                   \\
    echo             & Go          &                                                    \\
    gramework        & Go          &                                                    \\
    gin              & Go          & scratch                                            \\
    atreugo          & Go          & prefork                                            \\
    beego            & Go          & orm-mysql                                          \\
    kami             & Go          &                                                    \\
    falcore          & Go          &                                                    \\
    goji             & Go          &                                                    \\
    clevergo         & Go          &                                                    \\
    fasthttp         & Go          & prefork                                            \\
    aah              & Go          & mysql postgresql                                   \\
    gearbox          & Go          & prefork                                            \\
    martini          & Go          &                                                    \\
    go               & Go          & pgx pgx-easyjson pgx-quicktemplate pgx-prefor...   \\
    evio             & Go          & stdlib                                             \\
    chi              & Go          & scratch prefork gojay gojay-prefork sjson sjs...   \\
    warp             & Haskell     & hasql mysql-haskell                                \\
    yesod            & Haskell     & mongodb-raw                                        \\
    wizzardo-inline  & Haskell     &                                                    \\
    snap             & Haskell     &                                                    \\
    spock            & Haskell     &                                                    \\
    ihp              & Haskell     &                                                    \\
    servant          & Haskell     & beam psql-simple                                   \\
    stream           & Dart        &                                                    \\
    start            & Dart        &                                                    \\
    angel            & Dart        & postgres                                           \\
    redstone         & Dart        & mongodb                                            \\
    dart             & Dart        &                                                    \\
    aqueduct         & Dart        &                                                    \\
    aspcore-vb-mw    & VB          & ado-pg ado-my                                      \\
    spider-gazelle   & Crystal     &                                                    \\
    toro             & Crystal     &                                                    \\
    onyx             & Crystal     &                                                    \\
    h2o.cr           & Crystal     &                                                    \\
    kemal            & Crystal     & concurrent-queries                                 \\
    lucky            & Crystal     &                                                    \\
    crystal          & Crystal     & radix                                              \\
    grip             & Crystal     &                                                    \\
    orion            & Crystal     &                                                    \\
    amber            & Crystal     &                                                    \\
    raze             & Crystal     &                                                    \\
    collie           & D           & ldc                                                \\
    vibed            & D           & dmd-pgsql ldc ldc-pgsql                            \\
    hunt             & D           &                                                    \\
    macchiato        & Clojure     &                                                    \\
    aleph            & Clojure     &                                                    \\
    luminus          & Clojure     &                                                    \\
    pedestal         & Clojure     &                                                    \\
    compojure        & Clojure     & raw                                                \\
    http-kit         & Clojure     & raw                                                \\
    duct             & Clojure     & mongodb httpkit aleph immutant                     \\
    reitit           & Clojure     & jdbc async                                         \\
    nestjs           & TypeScript  & mysql mongo fastify fastify-mysql fastify-mongo    \\
    typescript-rest  & TypeScript  &                                                    \\
    deno             & TypeScript  &                                                    \\
    dropwizard       & Java        & mongodb postgres jdbi-postgres                     \\
    httpserver       & Java        & postgres                                           \\
    quarkus          & Java        & pgclient                                           \\
    jooby            & Java        & netty jaxrs jetty pgclient                         \\
    undertow         & Java        & postgresql                                         \\
    tapestry         & Java        &                                                    \\
    vertx-web        & Java        & mongodb postgres susom-postgres                    \\
    jlhttp           & Java        & postgres                                           \\
    voovan           & Java        &                                                    \\
    play2-java       & Java        & netty ebean-hikaricp ebean-hikaricp-netty jpa...   \\
    officefloor      & Java        & raw tpr micro thread\_affinity netty spring\_data  \\
    jersey           & Java        &                                                    \\
    proteus          & Java        & mysql                                              \\
    light-4j         & Java        &                                                    \\
    activeweb        & Java        & jackson                                            \\
    minijax          & Java        &                                                    \\
    jetty            & Java        & servlet                                            \\
    spark            & Java        &                                                    \\
    grizzly          & Java        & jersey                                             \\
    servlet          & Java        & lib-native mysql postgresql                        \\
    tio-mvc          & Java        &                                                    \\
    baratine         & Java        &                                                    \\
    servicetalk      & Java        &                                                    \\
    play1            & Java        &                                                    \\
    ninja-standalone & Java        &                                                    \\
    nanohttpd        & Java        &                                                    \\
    simple-server    & Java        &                                                    \\
    restexpress      & Java        & mysql-raw                                          \\
    javalin          & Java        &                                                    \\
    jawn             & Java        &                                                    \\
    redkale          & Java        & postgres                                           \\
    revenj-jvm       & Java        &                                                    \\
    spring-webflux   & Java        & mongo pgclient rxjdbc jdbc                         \\
    firenio          & Java        & http-lite                                          \\
    curacao          & Java        &                                                    \\
    bayou            & Java        &                                                    \\
    wizzardo-http    & Java        &                                                    \\
    netty            & Java        &                                                    \\
    armeria          & Java        &                                                    \\
    vertx            & Java        & postgres                                           \\
    comsat           & Java        & servlet-undertow webactors-netty webactors-un...   \\
    ratpack          & Java        & jdbc pgclient                                      \\
    wicket           & Java        &                                                    \\
    wildfly-ee       & Java        &                                                    \\
    undertow-jersey  & Java        & hikaricp                                           \\
    micronaut        & Java        &                                                    \\
    servlet3         & Java        & sync                                               \\
    act              & Java        & hibernate-pgsql hibernate-pgsql-rythm hiberna...   \\
    blade            & Java        &                                                    \\
    rapidoid         & Java        & mysql postgres http-fast                           \\
    smart-socket     & Java        &                                                    \\
    gemini           & Java        & mysql postgres                                     \\
    greenlightning   & Java        &                                                    \\
    helidon          & Java        &                                                    \\
    spring           & Java        & jpa mongo                                          \\
    codeigniter      & PHP         &                                                    \\
    fat-free         & PHP         & raw                                                \\
    php              & PHP         & activerecord raw7-tcp pgsql-raw h2o unit pool...   \\
    cakephp          & PHP         &                                                    \\
    ubiquity         & PHP         & workerman swoole swoole-mysql roadrunner road...   \\
    fuel             & PHP         &                                                    \\
    sw-fw-less       & PHP         &                                                    \\
    hyperf           & PHP         & raw micro                                          \\
    kumbiaphp        & PHP         & raw workerman workerman-mysql                      \\
    workerman        & PHP         & pgsql async                                        \\
    lumen            & PHP         & swoole                                             \\
    peachpie         & PHP         &                                                    \\
    yii2             & PHP         & raw                                                \\
    php-ngx          & PHP         & mysql pgsql async                                  \\
    laravel          & PHP         & swoole                                             \\
    symfony          & PHP         & raw swoole                                         \\
    imi              & PHP         & query-builder raw                                  \\
    webman           & PHP         &                                                    \\
    phalcon          & PHP         & mongodb micro                                      \\
    slim             & PHP         &                                                    \\
    amp              & PHP         &                                                    \\
    hhvm             & PHP         &                                                    \\
    spiral           & PHP         &                                                    \\
    swoole           & PHP         & postgres no-async                                  \\
    hamlet           & PHP         & swoole workerman                                   \\
    simps            & PHP         & micro                                              \\
    one              & PHP         &                                                    \\
    zend             & PHP         &                                                    \\
    phpixie          & PHP         &                                                    \\
    comet            & PHP         & mysql                                              \\
    nginx            & C           &                                                    \\
    onion            & C           &                                                    \\
    duda             & C           &                                                    \\
    libreactor       & C           & server                                             \\
    lwan             & C           & lua                                                \\
    h2o              & C           &                                                    \\
    facil\_io        & C           &                                                    \\
    octane           & C           &                                                    \\
    polkadot         & JavaScript  & mysql postgres mongodb                             \\
    restify          & JavaScript  &                                                    \\
    hapi             & JavaScript  & nginx mysql postgres                               \\
    nodejs           & JavaScript  & chakra mongodb mongodb-raw mysql postgres          \\
    restana          & JavaScript  &                                                    \\
    es4x             & JavaScript  &                                                    \\
    sailsjs          & JavaScript  & postgres                                           \\
    just             & JavaScript  &                                                    \\
    fastify          & JavaScript  & mysql postgres                                     \\
    ringojs          & JavaScript  & convenient                                         \\
    express          & JavaScript  & chakra mongodb mysql postgres graphql-mysql g...   \\
    koa              & JavaScript  & mysql postgres                                     \\
    0http            & JavaScript  &                                                    \\
    openresty        & Lua         &                                                    \\
    octopus          & Lua         &                                                    \\
    lapis            & Lua         &                                                    \\
    easyrpc          & C\#         &                                                    \\
    aspcore-mono     & C\#         & pg mw mw-pg mw-my mvc mvc-pg mvc-my                \\
    servicestack     & C\#         &                                                    \\
    tetsuweb         & C\#         &                                                    \\
    beetlex          & C\#         & core core-updb debug                               \\
    aspcore          & C\#         & pgo ado-pg ado-pg-up pgo-pg pgo-pg-up mw mw-j...   \\
    revenj           & C\#         &                                                    \\
    carter           & C\#         &                                                    \\
    aspnet-mono-ngx  & C\#         & jn ss my my-ef                                     \\
    genhttp          & C\#         &                                                    \\
    ben              & C\#         & pg                                                 \\
    evhttp-sharp     & C\#         &                                                    \\
    zysocket-v       & C\#         &                                                    \\
    nancy            & C\#         & netcore                                            \\
    aspcore-corert   & C\#         &                                                    \\
    cuehttp          & C++         &                                                    \\
    libhttpserver    & C++         &                                                    \\
    ffead-cpp        & C++         & mongo-raw lithium cinatra drogon libreactor h...   \\
    lithium          & C++         & postgres postgres-batch                            \\
    wt               & C++         & postgres                                           \\
    treefrog         & C++         & postgres mongodb                                   \\
    ulib             & C++         & plaintext\_fit json json\_fit mysql postgres po... \\
    cppcms           & C++         & postgres                                           \\
    oatpp            & C++         & no-mp async async-no-mp                            \\
    poco             & C++         &                                                    \\
    silicon          & C++         &                                                    \\
    cutelyst         & C++         & pf-apg pf-pg pf-my thread thread-apg thread-p...   \\
    drogon           & C++         & core                                               \\
    cinatra          & C++         &                                                    \\
    libsniper        & C++         &                                                    \\
    cpoll\_cppsp     & C++         & raw postgres-raw postgres-raw-threadpool           \\
\end{longtable}

\newcommand\duration{20}
\newcommand\parallelclient{512}
\section{Experimental Protocol}

In this section, we describe the environement used during the experiments covering hardware material, experiments of the framework and the methodology.

\subsection{experiment workflow}
To mesure compare the energy consumption and performance effeciency between mutiple frameworks, each framework used to implement the samewebsite answering the same urls and requesting the same database. Then we run the same algorthim for all impelentations.
\begin{enumerate}
    \item lunch the framework
    \item wait for \duration s for the warmup
    \item measure the average power when the framework is in idle state
    \item using multiple clients. we send the same request in simultaneously during \duration s
    \item increase the number of parallel request
    \item measure the enrgy during this execution
    \item change the request type
    \item repeat from 3 rd step
\end{enumerate}
the section will detail each type of experiments  and the purpose behind it giving some examples of the expecteed response

\subsection{Test categories }
We have 7 categories of tests :
\\ \todo{ADD expected response for each one of them }
\subsubsection{Idle}
In this test we measure the idle energy consumption of the framework: this reflects the average energy consumption of an application during the periods when we have no clients.
For example, a company website beyond working hours or a small website at night. % maybe i should remove it

\subsubsection{Singe query}
During this test, each request is processed by fetching a single row from a simple database table. That row is then serialized as a JSON response.

\subsubsection{Multiple queries}
this test aims to test the behaviour of a framework when it process multiple entries from the database.
Therefore each request is processed by fetching multiple rows from a simple database table and serializing these rows as a JSON response.
%The test is run multiple times: testing 1, 5, 10, 15, and 20 queries per request. All tests are run at 512 concurrency.
%//TODO  add the cuncurency levels 
\subsubsection{Cached queries}
To evaluate the performance of a framework when the data is already cashed  each request is processed by fetching multiple cached objects from an in-memory database and serializing these objects as a JSON response.
%  The test is run multiple times: testing 1, 5, 10, 15, and 20 cached object fetches per request. All tests are run at 512 concurrency. Conceptually, this is similar to the multiple-queries test except the fact that it uses a caching layer.
\subsubsection{Fortunes}
In this test, the framework's ORM is used to fetch all rows from a database table containing an unknown number of Unix fortune cookie messages (the table has 12 rows, but the code cannot have foreknowledge of the table's size). An additional fortune cookie message is inserted into the list at runtime and then the list is sorted by the message text. Finally, the list is delivered to the client using a server-side HTML template. The message text must be considered untrusted and properly escaped and the UTF-8 fortune messages must be rendered properly.

\subsubsection{Update queries}

This test exercises database writes. Each request is processed by fetching multiple rows from a simple database table, converting the rows to in-memory objects, modifying one attribute of each object in memory, updating each associated row in the database individually, and then serializing the list of objects as a JSON response.
%  The test is run multiple times: testing 1, 5, 10, 15, and 20 updates per request. It is important that the number of statements per request is twice the number of updates since each update is paired with one query to fetch the object. All tests are run at 512 concurrency.

The response is analogous to the multiple-query test.

\subsubsection{Plain text}
In this test, the framework responds with the simplest of responses: a "Hello, World" message rendered as plain text. The size of the response is kept small so that gigabit Ethernet is not the limiting factor for all implementations. HTTP pipelining is enabled and higher client-side concurrency levels are used for this test.

\subsubsection{JSON Serialization}
In this test, each response is a JSON serialization of a freshly-instantiated object that maps the key message to the value "Hello, World!"

\subsection{Architecture}
The aim is to compare the energy consumption of different frameworks. For this we consider the framework as a blackbox and we take into account the response to the 6 previous requests using the same database.
In order to isolate the energy consumption of the framework the test is run in a seperate machine with the minimum services and the powermeter.
Figure \ref{fig:architecture} illustrates the architecture of our system

\begin{figure}\centering
    \includegraphics[width=.8\columnwidth]{imgs/architecture.png}
    \caption[Architecture]{Architecture of the experiments}
    \label{fig:architecture}
\end{figure}

\begin{itemize}
    \item orchestrator : the part that is responisble for creating docker images, selecting the benchmarks and launching the tests;
    \item web server : or the system under test (SUT) is the machine  responsible for launching the framework  by means of the pre-installed powermeter;
    \item database server : a offers the same data base that will be used by all the frameworks during the tests
          %(is seperated from the application server to neglect its contribution in energy calculation?) 
    \item client machines : To avoid the bottleneck on the client's side, client requests are sent from another machine (one or many) that simulates hundreds of concurrent connections to the framework.
          %Does even one machine simulate hundreds? 
    \item recorder : the party responsible for collecting the power data from the SUT and the performance metrics collected by the clients \todo{add link to the measurement process}
\end{itemize}

The tests have been executed in machines from the cluster chetemi of the \reference{Grid500} plateform .
\todo{add hardware description}

\paragraph{Note}
It has been proven in {cite } that docker does not influence on the energy cosumption. Thus, using its containers and isolation allow to avoid any alteration of the operation system after testing one benchmark and to take advantage of its reproducibility.

\subsection{Measurement Workflow}
We focus on comparing the energitical behaviour of different frameworks in multiple scenarios

To measure the energy consumption of those frameworks, we launch each one for a fixed duration, then all the clients send multiple requests simultaneously.
We calculate %calculate - with paper and pen or calculator, compute - with computer (programs)
the number of satisfied responses which reflects the performance of the framework, the average latency and the global energy consumed during the whole period to deduce the energy cost of each request.

\subsubsection{Runtime Measurements}
\begin{itemize}
    \item energy measurement :
          we use \reference{powerapi}, a software powermeter  to gather the runninng power of the SUT, after we project the timestamps of each experiment phase to calculate the energy consumption of the framework during this phase.
          Energy is an integral of power over time, so we use a numerical apporoach to estimate this energy. \todo{add formula}
          after this we devide the calculated energy per number of responses
          during this phase
          we have 4 mertrics :

    \item Total cost of the energy during each period
    \item Total number of requests
    \item Average latency
    \item Average energy cost per request
\end{itemize}

\subsubsection{Bias analysis}
we are aware of bias analysis regarding the estimation of the total energy cost , the interference of other system processess during the execution and some external events. Thus, we run experiments multiple times and compute the average.

\subsection{tools }

### 2022-05-26 chapters/execution_environement/python/python.tex
add : tommti benchmark just a single draft have to readit later

@@
 
-\subsection{benchmarks}
+\begin{figure}
+    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
+    \label{table:bitops}
+    \begin{tabular}{lrrrrr}
+        \toprule
+        benchmark    & A          & GG         & LL         & Or         & XOR        \\
+        \midrule
+        activepython & 676.980672 & 763.208040 & 651.783048 & 743.016425 & 728.828481 \\
+        cpython2     & 441.082298 & 435.886950 & 430.846171 & 415.247383 & 419.081447 \\
+        cpython3     & 595.209019 & 685.085315 & 563.839300 & 657.972734 & 655.560574 \\
+        cython       & 35.077408  & 182.688395 & 274.177830 & 34.868014  & 34.504778  \\
+        nuitka       & 33.260991  & 32.980380  & 33.256450  & 33.472770  & 33.030889  \\
+        numba2       & 9.102939   & 8.411176   & 9.460620   & 9.375920   & 9.755952   \\
+        numba3       & 9.566646   & 10.144212  & 9.219703   & 9.344613   & 9.665108   \\
+        pypy2        & 8.456867   & 7.844918   & 8.286278   & 8.138692   & 7.952999   \\
+        pypy3        & 7.552731   & 8.093884   & 8.108508   & 8.669448   & 8.623737   \\
+        shedskin     & 8.024198   & 8.070870   & 8.399917   & 8.126236   & 8.277546   \\
+        \bottomrule
+    \end{tabular}
 
-\subsection{first results}
+\end{figure}
+
+\begin{figure}[thb]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/bitopts_mean}
+    \caption{Mean consumption of different implementations of bit operations (Joule) }
+    \label{fig:bitops}
+\end{figure}
+
+\begin{figure}[thb]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/alltomti_performance}
+    \caption{different interpererts optimisation }
+    \label{fig:tommi_all}
+\end{figure}
+\begin{figure}[thb]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/tommti_compare__pypy2_pypy3}
+    \caption{green factor of pypy }
+    \label{fig:pypy2vspypy3}
+\end{figure}
+
+\begin{figure}[thb]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/tommti_compare__cpython3_numba3_pypy3}
+    \caption{comparaison of pypy vs python vs numba }
+    \label{fig:p3}
+\end{figure}
+
+\begin{figure}[thb]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/tommti_compare__cpython2_pypy2_numba2}
+    \caption{comparaison of pypy vs python vs numba }
+    \label{fig:p2}
+\end{figure}
+
+\begin{figure}[thb]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/bar_plot_tommtivector}
+    \caption{energy consumption of different impetions treating vectors }
+    \label{fig:bar_tommti_vector}
+\end{figure}
 
@@
 \subsection{results and foundings}
+As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
+Intelpython and active python both folows the same behaviour. one can conclude that the work that have been on those interpreters is mainly
+to improve a specific purpose, Active python claims that their version is focused on secuirty which explains the lack of some performances due to the introduction of more reonforcment
+.Morver Intel published their version of python as a dedicated for machine learning. Unfortunately the tommti benchmark is a set that focuses on the general purpose programming
+which does not reflect the performance of the machine learning.
+Another aspect of a such behaviour might be due to the fact of the processors that were used in the testbed.
+and it might change in the future if we include the GPU part for the machine learning benchmarks
+% reference the work of the intern 
+As for nuitka. there were no optimization in the energy consumpotion despite the fact that is it a complier.
+However if we dig through the nuitka mechanisms. they basically embed the python code with an interpreter.
+Unllike nuitka shedskin exhebit a the best energy consumption pattern when it comes to the arithmetic operartions.
+One can conclude it is due to the fact of the native type of the variables.
+unlike the interpreters where they are treated as object in the begining.
+
+for the other interpeters pypy is very promising especially when it comes to data maniupulation as one can see in the figure \ref{fig:bar_tommti_vector}
+pypy is by far the best interperer when it comes to treating vectors.
+numba2 introduced the JIT but wasn't as promising as numba3.
+
+for the other vm based interpereters.
+jython and ipy lacked in term of energy optimisation which was kinda expected since they were in their the begining of the stage and the main puprose of such implementation is to link the bytecode generated by jython and ironpython with their respectives virtual machin.
+
+Unlike the the previous interpeters. graal exhebits a certin promises when it comes to complex algorithms - nested loops -
+micro python is dedicated to embeded systems so lunching it is powerful cpu machines will be misleading.
+
+Most of the interperters had the same behaviour when it comes to the input outputs. except for jython which was kinda of a anomaly probably due to the lack of the optimizations.
+% pypy handles exceptions very well
+
+%% todo 
+% add performance 
+% add dram \ a glimps 
+% add bitops
 

### 2022-05-26 chapters/programming_languages/web_frameworks.tex
add : tommti benchmark just a single draft have to readit later

@@
 
-\begin{longtable}{lll}
-    \toprule
-    framework        & language    & alternative implementations                        \\
-    \midrule
-    \endhead
-    \midrule
-    \multicolumn{3}{r}{{Continued on next page}}                                        \\
-    \midrule
-    \endfoot
+% \begin{longtable}{lll}
+%     \toprule
+%     framework        & language    & alternative implementations                        \\
+%     \midrule
+%     \endhead
+%     \midrule
+%     \multicolumn{3}{r}{{Continued on next page}}                                        \\
+%     \midrule
+%     \endfoot
 
-    \bottomrule
-    \endlastfoot
-    pico.v           & V           &                                                    \\
-    dylan            & Dylan       &                                                    \\
-    sinatra-sequel   & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
-    agoo             & Ruby        &                                                    \\
-    rack             & Ruby        & unicorn falcon                                     \\
-    h2o\_mruby       & Ruby        &                                                    \\
-    rails            & Ruby        & postgresql unicorn                                 \\
-    grape            & Ruby        & unicorn                                            \\
-    roda-sequel      & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
-    rack-sequel      & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
-    sinatra          & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
-    padrino          & Ruby        & unicorn                                            \\
-    ninglex          & Common Lisp &                                                    \\
-    ningle           & Common Lisp &                                                    \\
-    woo              & Common Lisp &                                                    \\
-    grails           & Groovy      &                                                    \\
-    hot              & Groovy      & mongodb postgres mysql                             \\
-    plack            & Perl        & async                                              \\
-    kelp             & Perl        & mongodb                                            \\
-    dancer           & Perl        &                                                    \\
-    mojolicious      & Perl        &                                                    \\
-    web-simple       & Perl        &                                                    \\
-    cowboy           & Erlang      &                                                    \\
-    elli             & Erlang      &                                                    \\
-    chicagoboss      & Erlang      &                                                    \\
-    mochiweb         & Erlang      &                                                    \\
-    opium            & OCaml       & haproxy                                            \\
-    morph            & OCaml       & static                                             \\
-    webmachine       & OCaml       & haproxy                                            \\
-    vertx-web-scala  & Scala       &                                                    \\
-    http4s           & Scala       &                                                    \\
-    akka-http        & Scala       & slick-postgres                                     \\
-    fintrospect      & Scala       &                                                    \\
-    finagle          & Scala       &                                                    \\
-    colossus         & Scala       &                                                    \\
-    finatra          & Scala       &                                                    \\
-    blaze            & Scala       &                                                    \\
-    play2-scala      & Scala       & netty anorm anorm-netty reactivemongo reactiv...   \\
-    finch            & Scala       &                                                    \\
-    scalene          & Scala       &                                                    \\
-    cask             & Scala       &                                                    \\
-    youi             & Scala       &                                                    \\
-    httpbeast        & Nim         &                                                    \\
-    prologue         & Nim         &                                                    \\
-    basolato         & Nim         &                                                    \\
-    jester           & Nim         &                                                    \\
-    falco            & F\#         &                                                    \\
-    frank            & F\#         &                                                    \\
-    suave            & F\#         &                                                    \\
-    giraffe          & F\#         & utf8json utf8direct stripped                       \\
-    zebra            & F\#         & simple                                             \\
-    vapor            & Swift       & fluent sql-kit postgres                            \\
-    swift-nio        & Swift       &                                                    \\
-    perfect          & Swift       & mysql postgresql mongodb                           \\
-    kitura           & Swift       & postgres postgres-orm postgres-orm-codable mo...   \\
-    racket           & Racket      &                                                    \\
-    vsgi             & Vala        &                                                    \\
-    valum            & Vala        &                                                    \\
-    wsgi             & Python      &                                                    \\
-    hug              & Python      &                                                    \\
-    crax             & Python      &                                                    \\
-    falcon           & Python      & py3 pypy2                                          \\
-    uvicorn          & Python      &                                                    \\
-    flask            & Python      & raw pypy pypy2 pypy2-raw nginx-uwsgi               \\
-    responder        & Python      &                                                    \\
-    klein            & Python      &                                                    \\
-    blacksheep       & Python      &                                                    \\
-    apidaora         & Python      & core                                               \\
-    bottle           & Python      & pypy2 raw nginx-uwsgi                              \\
-    japronto         & Python      &                                                    \\
-    api\_hour        & Python      & mysql json dbs plaintext                           \\
-    weppy            & Python      & py3 pypy2 nginx-uwsgi                              \\
-    starlette        & Python      &                                                    \\
-    emmett           & Python      &                                                    \\
-    sanic            & Python      &                                                    \\
-    aiohttp          & Python      & pg-raw                                             \\
-    tornado          & Python      & pypy2 postgresql-raw py3 py3-uvloop                \\
-    uwsgi            & Python      & nginx-uwsgi                                        \\
-    pyramid          & Python      & py2                                                \\
-    morepath         & Python      &                                                    \\
-    vibora           & Python      &                                                    \\
-    web2py           & Python      & optimized                                          \\
-    django           & Python      & postgresql                                         \\
-    cherrypy         & Python      & py3                                                \\
-    eve              & Python      &                                                    \\
-    webware          & Python      &                                                    \\
-    spyne            & Python      & raw nginx-uwsgi                                    \\
-    fastapi          & Python      & orjson                                             \\
-    quart            & Python      & uvicorn                                            \\
-    turbogears       & Python      &                                                    \\
-    m-web-server     & mumps       &                                                    \\
-    urweb            & Ur          & mysql cache mysql-cache                            \\
-    rouille          & Rust        &                                                    \\
-    warp-rust        & Rust        &                                                    \\
-    thruster         & Rust        &                                                    \\
-    ntex             & Rust        & db raw                                             \\
-    iron             & Rust        &                                                    \\
-    actix            & Rust        & core raw diesel pg                                 \\
-    roa              & Rust        & core tokio diesel pg sqlx                          \\
-    saphir           & Rust        &                                                    \\
-    may-minihttp     & Rust        &                                                    \\
-    rocket           & Rust        &                                                    \\
-    tokio-minihttp   & Rust        &                                                    \\
-    nickel           & Rust        &                                                    \\
-    hyper            & Rust        & db                                                 \\
-    gotham           & Rust        &                                                    \\
-    kooby            & Kotlin      &                                                    \\
-    pronghorn        & Kotlin      &                                                    \\
-    http4k           & Kotlin      & apache apache4 jetty ktorcio ktornetty netty ...   \\
-    ktor             & Kotlin      & jetty cio jasync reactivepg                        \\
-    hexagon          & Kotlin      & resin-mongodb jetty-postgresql resin-postgresql    \\
-    elixir-plug-ecto & Elixir      &                                                    \\
-    phoenix          & Elixir      &                                                    \\
-    fiber            & Go          & prefork                                            \\
-    webgo            & Go          &                                                    \\
-    gnet             & Go          &                                                    \\
-    goframe          & Go          &                                                    \\
-    revel            & Go          & fast raw jet qbs                                   \\
-    echo             & Go          &                                                    \\
-    gramework        & Go          &                                                    \\
-    gin              & Go          & scratch                                            \\
-    atreugo          & Go          & prefork                                            \\
-    beego            & Go          & orm-mysql                                          \\
-    kami             & Go          &                                                    \\
-    falcore          & Go          &                                                    \\
-    goji             & Go          &                                                    \\
-    clevergo         & Go          &                                                    \\
-    fasthttp         & Go          & prefork                                            \\
-    aah              & Go          & mysql postgresql                                   \\
-    gearbox          & Go          & prefork                                            \\
-    martini          & Go          &                                                    \\
-    go               & Go          & pgx pgx-easyjson pgx-quicktemplate pgx-prefor...   \\
-    evio             & Go          & stdlib                                             \\
-    chi              & Go          & scratch prefork gojay gojay-prefork sjson sjs...   \\
-    warp             & Haskell     & hasql mysql-haskell                                \\
-    yesod            & Haskell     & mongodb-raw                                        \\
-    wizzardo-inline  & Haskell     &                                                    \\
-    snap             & Haskell     &                                                    \\
-    spock            & Haskell     &                                                    \\
-    ihp              & Haskell     &                                                    \\
-    servant          & Haskell     & beam psql-simple                                   \\
-    stream           & Dart        &                                                    \\
-    start            & Dart        &                                                    \\
-    angel            & Dart        & postgres                                           \\
-    redstone         & Dart        & mongodb                                            \\
-    dart             & Dart        &                                                    \\
-    aqueduct         & Dart        &                                                    \\
-    aspcore-vb-mw    & VB          & ado-pg ado-my                                      \\
-    spider-gazelle   & Crystal     &                                                    \\
-    toro             & Crystal     &                                                    \\
-    onyx             & Crystal     &                                                    \\
-    h2o.cr           & Crystal     &                                                    \\
-    kemal            & Crystal     & concurrent-queries                                 \\
-    lucky            & Crystal     &                                                    \\
-    crystal          & Crystal     & radix                                              \\
-    grip             & Crystal     &                                                    \\
-    orion            & Crystal     &                                                    \\
-    amber            & Crystal     &                                                    \\
-    raze             & Crystal     &                                                    \\
-    collie           & D           & ldc                                                \\
-    vibed            & D           & dmd-pgsql ldc ldc-pgsql                            \\
-    hunt             & D           &                                                    \\
-    macchiato        & Clojure     &                                                    \\
-    aleph            & Clojure     &                                                    \\
-    luminus          & Clojure     &                                                    \\
-    pedestal         & Clojure     &                                                    \\
-    compojure        & Clojure     & raw                                                \\
-    http-kit         & Clojure     & raw                                                \\
-    duct             & Clojure     & mongodb httpkit aleph immutant                     \\
-    reitit           & Clojure     & jdbc async                                         \\
-    nestjs           & TypeScript  & mysql mongo fastify fastify-mysql fastify-mongo    \\
-    typescript-rest  & TypeScript  &                                                    \\
-    deno             & TypeScript  &                                                    \\
-    dropwizard       & Java        & mongodb postgres jdbi-postgres                     \\
-    httpserver       & Java        & postgres                                           \\
-    quarkus          & Java        & pgclient                                           \\
-    jooby            & Java        & netty jaxrs jetty pgclient                         \\
-    undertow         & Java        & postgresql                                         \\
-    tapestry         & Java        &                                                    \\
-    vertx-web        & Java        & mongodb postgres susom-postgres                    \\
-    jlhttp           & Java        & postgres                                           \\
-    voovan           & Java        &                                                    \\
-    play2-java       & Java        & netty ebean-hikaricp ebean-hikaricp-netty jpa...   \\
-    officefloor      & Java        & raw tpr micro thread\_affinity netty spring\_data  \\
-    jersey           & Java        &                                                    \\
-    proteus          & Java        & mysql                                              \\
-    light-4j         & Java        &                                                    \\
-    activeweb        & Java        & jackson                                            \\
-    minijax          & Java        &                                                    \\
-    jetty            & Java        & servlet                                            \\
-    spark            & Java        &                                                    \\
-    grizzly          & Java        & jersey                                             \\
-    servlet          & Java        & lib-native mysql postgresql                        \\
-    tio-mvc          & Java        &                                                    \\
-    baratine         & Java        &                                                    \\
-    servicetalk      & Java        &                                                    \\
-    play1            & Java        &                                                    \\
-    ninja-standalone & Java        &                                                    \\
-    nanohttpd        & Java        &                                                    \\
-    simple-server    & Java        &                                                    \\
-    restexpress      & Java        & mysql-raw                                          \\
-    javalin          & Java        &                                                    \\
-    jawn             & Java        &                                                    \\
-    redkale          & Java        & postgres                                           \\
-    revenj-jvm       & Java        &                                                    \\
-    spring-webflux   & Java        & mongo pgclient rxjdbc jdbc                         \\
-    firenio          & Java        & http-lite                                          \\
-    curacao          & Java        &                                                    \\
-    bayou            & Java        &                                                    \\
-    wizzardo-http    & Java        &                                                    \\
-    netty            & Java        &                                                    \\
-    armeria          & Java        &                                                    \\
-    vertx            & Java        & postgres                                           \\
-    comsat           & Java        & servlet-undertow webactors-netty webactors-un...   \\
-    ratpack          & Java        & jdbc pgclient                                      \\
-    wicket           & Java        &                                                    \\
-    wildfly-ee       & Java        &                                                    \\
-    undertow-jersey  & Java        & hikaricp                                           \\
-    micronaut        & Java        &                                                    \\
-    servlet3         & Java        & sync                                               \\
-    act              & Java        & hibernate-pgsql hibernate-pgsql-rythm hiberna...   \\
-    blade            & Java        &                                                    \\
-    rapidoid         & Java        & mysql postgres http-fast                           \\
-    smart-socket     & Java        &                                                    \\
-    gemini           & Java        & mysql postgres                                     \\
-    greenlightning   & Java        &                                                    \\
-    helidon          & Java        &                                                    \\
-    spring           & Java        & jpa mongo                                          \\
-    codeigniter      & PHP         &                                                    \\
-    fat-free         & PHP         & raw                                                \\
-    php              & PHP         & activerecord raw7-tcp pgsql-raw h2o unit pool...   \\
-    cakephp          & PHP         &                                                    \\
-    ubiquity         & PHP         & workerman swoole swoole-mysql roadrunner road...   \\
-    fuel             & PHP         &                                                    \\
-    sw-fw-less       & PHP         &                                                    \\
-    hyperf           & PHP         & raw micro                                          \\
-    kumbiaphp        & PHP         & raw workerman workerman-mysql                      \\
-    workerman        & PHP         & pgsql async                                        \\
-    lumen            & PHP         & swoole                                             \\
-    peachpie         & PHP         &                                                    \\
-    yii2             & PHP         & raw                                                \\
-    php-ngx          & PHP         & mysql pgsql async                                  \\
-    laravel          & PHP         & swoole                                             \\
-    symfony          & PHP         & raw swoole                                         \\
-    imi              & PHP         & query-builder raw                                  \\
-    webman           & PHP         &                                                    \\
-    phalcon          & PHP         & mongodb micro                                      \\
-    slim             & PHP         &                                                    \\
-    amp              & PHP         &                                                    \\
-    hhvm             & PHP         &                                                    \\
-    spiral           & PHP         &                                                    \\
-    swoole           & PHP         & postgres no-async                                  \\
-    hamlet           & PHP         & swoole workerman                                   \\
-    simps            & PHP         & micro                                              \\
-    one              & PHP         &                                                    \\
-    zend             & PHP         &                                                    \\
-    phpixie          & PHP         &                                                    \\
-    comet            & PHP         & mysql                                              \\
-    nginx            & C           &                                                    \\
-    onion            & C           &                                                    \\
-    duda             & C           &                                                    \\
-    libreactor       & C           & server                                             \\
-    lwan             & C           & lua                                                \\
-    h2o              & C           &                                                    \\
-    facil\_io        & C           &                                                    \\
-    octane           & C           &                                                    \\
-    polkadot         & JavaScript  & mysql postgres mongodb                             \\
-    restify          & JavaScript  &                                                    \\
-    hapi             & JavaScript  & nginx mysql postgres                               \\
-    nodejs           & JavaScript  & chakra mongodb mongodb-raw mysql postgres          \\
-    restana          & JavaScript  &                                                    \\
-    es4x             & JavaScript  &                                                    \\
-    sailsjs          & JavaScript  & postgres                                           \\
-    just             & JavaScript  &                                                    \\
-    fastify          & JavaScript  & mysql postgres                                     \\
-    ringojs          & JavaScript  & convenient                                         \\
-    express          & JavaScript  & chakra mongodb mysql postgres graphql-mysql g...   \\
-    koa              & JavaScript  & mysql postgres                                     \\
-    0http            & JavaScript  &                                                    \\
-    openresty        & Lua         &                                                    \\
-    octopus          & Lua         &                                                    \\
-    lapis            & Lua         &                                                    \\
-    easyrpc          & C\#         &                                                    \\
-    aspcore-mono     & C\#         & pg mw mw-pg mw-my mvc mvc-pg mvc-my                \\
-    servicestack     & C\#         &                                                    \\
-    tetsuweb         & C\#         &                                                    \\
-    beetlex          & C\#         & core core-updb debug                               \\
-    aspcore          & C\#         & pgo ado-pg ado-pg-up pgo-pg pgo-pg-up mw mw-j...   \\
-    revenj           & C\#         &                                                    \\
-    carter           & C\#         &                                                    \\
-    aspnet-mono-ngx  & C\#         & jn ss my my-ef                                     \\
-    genhttp          & C\#         &                                                    \\
-    ben              & C\#         & pg                                                 \\
-    evhttp-sharp     & C\#         &                                                    \\
-    zysocket-v       & C\#         &                                                    \\
-    nancy            & C\#         & netcore                                            \\
-    aspcore-corert   & C\#         &                                                    \\
-    cuehttp          & C++         &                                                    \\
-    libhttpserver    & C++         &                                                    \\
-    ffead-cpp        & C++         & mongo-raw lithium cinatra drogon libreactor h...   \\
-    lithium          & C++         & postgres postgres-batch                            \\
-    wt               & C++         & postgres                                           \\
-    treefrog         & C++         & postgres mongodb                                   \\
-    ulib             & C++         & plaintext\_fit json json\_fit mysql postgres po... \\
-    cppcms           & C++         & postgres                                           \\
-    oatpp            & C++         & no-mp async async-no-mp                            \\
-    poco             & C++         &                                                    \\
-    silicon          & C++         &                                                    \\
-    cutelyst         & C++         & pf-apg pf-pg pf-my thread thread-apg thread-p...   \\
-    drogon           & C++         & core                                               \\
-    cinatra          & C++         &                                                    \\
-    libsniper        & C++         &                                                    \\
-    cpoll\_cppsp     & C++         & raw postgres-raw postgres-raw-threadpool           \\
-\end{longtable}
+%     \bottomrule
+%     \endlastfoot
+%     pico.v           & V           &                                                    \\
+%     dylan            & Dylan       &                                                    \\
+%     sinatra-sequel   & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
+%     agoo             & Ruby        &                                                    \\
+%     rack             & Ruby        & unicorn falcon                                     \\
+%     h2o\_mruby       & Ruby        &                                                    \\
+%     rails            & Ruby        & postgresql unicorn                                 \\
+%     grape            & Ruby        & unicorn                                            \\
+%     roda-sequel      & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
+%     rack-sequel      & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
+%     sinatra          & Ruby        & postgres passenger-mri postgres-passenger-mri...   \\
+%     padrino          & Ruby        & unicorn                                            \\
+%     ninglex          & Common Lisp &                                                    \\
+%     ningle           & Common Lisp &                                                    \\
+%     woo              & Common Lisp &                                                    \\
+%     grails           & Groovy      &                                                    \\
+%     hot              & Groovy      & mongodb postgres mysql                             \\
+%     plack            & Perl        & async                                              \\
+%     kelp             & Perl        & mongodb                                            \\
+%     dancer           & Perl        &                                                    \\
+%     mojolicious      & Perl        &                                                    \\
+%     web-simple       & Perl        &                                                    \\
+%     cowboy           & Erlang      &                                                    \\
+%     elli             & Erlang      &                                                    \\
+%     chicagoboss      & Erlang      &                                                    \\
+%     mochiweb         & Erlang      &                                                    \\
+%     opium            & OCaml       & haproxy                                            \\
+%     morph            & OCaml       & static                                             \\
+%     webmachine       & OCaml       & haproxy                                            \\
+%     vertx-web-scala  & Scala       &                                                    \\
+%     http4s           & Scala       &                                                    \\
+%     akka-http        & Scala       & slick-postgres                                     \\
+%     fintrospect      & Scala       &                                                    \\
+%     finagle          & Scala       &                                                    \\
+%     colossus         & Scala       &                                                    \\
+%     finatra          & Scala       &                                                    \\
+%     blaze            & Scala       &                                                    \\
+%     play2-scala      & Scala       & netty anorm anorm-netty reactivemongo reactiv...   \\
+%     finch            & Scala       &                                                    \\
+%     scalene          & Scala       &                                                    \\
+%     cask             & Scala       &                                                    \\
+%     youi             & Scala       &                                                    \\
+%     httpbeast        & Nim         &                                                    \\
+%     prologue         & Nim         &                                                    \\
+%     basolato         & Nim         &                                                    \\
+%     jester           & Nim         &                                                    \\
+%     falco            & F\#         &                                                    \\
+%     frank            & F\#         &                                                    \\
+%     suave            & F\#         &                                                    \\
+%     giraffe          & F\#         & utf8json utf8direct stripped                       \\
+%     zebra            & F\#         & simple                                             \\
+%     vapor            & Swift       & fluent sql-kit postgres                            \\
+%     swift-nio        & Swift       &                                                    \\
+%     perfect          & Swift       & mysql postgresql mongodb                           \\
+%     kitura           & Swift       & postgres postgres-orm postgres-orm-codable mo...   \\
+%     racket           & Racket      &                                                    \\
+%     vsgi             & Vala        &                                                    \\
+%     valum            & Vala        &                                                    \\
+%     wsgi             & Python      &                                                    \\
+%     hug              & Python      &                                                    \\
+%     crax             & Python      &                                                    \\
+%     falcon           & Python      & py3 pypy2                                          \\
+%     uvicorn          & Python      &                                                    \\
+%     flask            & Python      & raw pypy pypy2 pypy2-raw nginx-uwsgi               \\
+%     responder        & Python      &                                                    \\
+%     klein            & Python      &                                                    \\
+%     blacksheep       & Python      &                                                    \\
+%     apidaora         & Python      & core                                               \\
+%     bottle           & Python      & pypy2 raw nginx-uwsgi                              \\
+%     japronto         & Python      &                                                    \\
+%     api\_hour        & Python      & mysql json dbs plaintext                           \\
+%     weppy            & Python      & py3 pypy2 nginx-uwsgi                              \\
+%     starlette        & Python      &                                                    \\
+%     emmett           & Python      &                                                    \\
+%     sanic            & Python      &                                                    \\
+%     aiohttp          & Python      & pg-raw                                             \\
+%     tornado          & Python      & pypy2 postgresql-raw py3 py3-uvloop                \\
+%     uwsgi            & Python      & nginx-uwsgi                                        \\
+%     pyramid          & Python      & py2                                                \\
+%     morepath         & Python      &                                                    \\
+%     vibora           & Python      &                                                    \\
+%     web2py           & Python      & optimized                                          \\
+%     django           & Python      & postgresql                                         \\
+%     cherrypy         & Python      & py3                                                \\
+%     eve              & Python      &                                                    \\
+%     webware          & Python      &                                                    \\
+%     spyne            & Python      & raw nginx-uwsgi                                    \\
+%     fastapi          & Python      & orjson                                             \\
+%     quart            & Python      & uvicorn                                            \\
+%     turbogears       & Python      &                                                    \\
+%     m-web-server     & mumps       &                                                    \\
+%     urweb            & Ur          & mysql cache mysql-cache                            \\
+%     rouille          & Rust        &                                                    \\
+%     warp-rust        & Rust        &                                                    \\
+%     thruster         & Rust        &                                                    \\
+%     ntex             & Rust        & db raw                                             \\
+%     iron             & Rust        &                                                    \\
+%     actix            & Rust        & core raw diesel pg                                 \\
+%     roa              & Rust        & core tokio diesel pg sqlx                          \\
+%     saphir           & Rust        &                                                    \\
+%     may-minihttp     & Rust        &                                                    \\
+%     rocket           & Rust        &                                                    \\
+%     tokio-minihttp   & Rust        &                                                    \\
+%     nickel           & Rust        &                                                    \\
+%     hyper            & Rust        & db                                                 \\
+%     gotham           & Rust        &                                                    \\
+%     kooby            & Kotlin      &                                                    \\
+%     pronghorn        & Kotlin      &                                                    \\
+%     http4k           & Kotlin      & apache apache4 jetty ktorcio ktornetty netty ...   \\
+%     ktor             & Kotlin      & jetty cio jasync reactivepg                        \\
+%     hexagon          & Kotlin      & resin-mongodb jetty-postgresql resin-postgresql    \\
+%     elixir-plug-ecto & Elixir      &                                                    \\
+%     phoenix          & Elixir      &                                                    \\
+%     fiber            & Go          & prefork                                            \\
+%     webgo            & Go          &                                                    \\
+%     gnet             & Go          &                                                    \\
+%     goframe          & Go          &                                                    \\
+%     revel            & Go          & fast raw jet qbs                                   \\
+%     echo             & Go          &                                                    \\
+%     gramework        & Go          &                                                    \\
+%     gin              & Go          & scratch                                            \\
+%     atreugo          & Go          & prefork                                            \\
+%     beego            & Go          & orm-mysql                                          \\
+%     kami             & Go          &                                                    \\
+%     falcore          & Go          &                                                    \\
+%     goji             & Go          &                                                    \\
+%     clevergo         & Go          &                                                    \\
+%     fasthttp         & Go          & prefork                                            \\
+%     aah              & Go          & mysql postgresql                                   \\
+%     gearbox          & Go          & prefork                                            \\
+%     martini          & Go          &                                                    \\
+%     go               & Go          & pgx pgx-easyjson pgx-quicktemplate pgx-prefor...   \\
+%     evio             & Go          & stdlib                                             \\
+%     chi              & Go          & scratch prefork gojay gojay-prefork sjson sjs...   \\
+%     warp             & Haskell     & hasql mysql-haskell                                \\
+%     yesod            & Haskell     & mongodb-raw                                        \\
+%     wizzardo-inline  & Haskell     &                                                    \\
+%     snap             & Haskell     &                                                    \\
+%     spock            & Haskell     &                                                    \\
+%     ihp              & Haskell     &                                                    \\
+%     servant          & Haskell     & beam psql-simple                                   \\
+%     stream           & Dart        &                                                    \\
+%     start            & Dart        &                                                    \\
+%     angel            & Dart        & postgres                                           \\
+%     redstone         & Dart        & mongodb                                            \\
+%     dart             & Dart        &                                                    \\
+%     aqueduct         & Dart        &                                                    \\
+%     aspcore-vb-mw    & VB          & ado-pg ado-my                                      \\
+%     spider-gazelle   & Crystal     &                                                    \\
+%     toro             & Crystal     &                                                    \\
+%     onyx             & Crystal     &                                                    \\
+%     h2o.cr           & Crystal     &                                                    \\
+%     kemal            & Crystal     & concurrent-queries                                 \\
+%     lucky            & Crystal     &                                                    \\
+%     crystal          & Crystal     & radix                                              \\
+%     grip             & Crystal     &                                                    \\
+%     orion            & Crystal     &                                                    \\
+%     amber            & Crystal     &                                                    \\
+%     raze             & Crystal     &                                                    \\
+%     collie           & D           & ldc                                                \\
+%     vibed            & D           & dmd-pgsql ldc ldc-pgsql                            \\
+%     hunt             & D           &                                                    \\
+%     macchiato        & Clojure     &                                                    \\
+%     aleph            & Clojure     &                                                    \\
+%     luminus          & Clojure     &                                                    \\
+%     pedestal         & Clojure     &                                                    \\
+%     compojure        & Clojure     & raw                                                \\
+%     http-kit         & Clojure     & raw                                                \\
+%     duct             & Clojure     & mongodb httpkit aleph immutant                     \\
+%     reitit           & Clojure     & jdbc async                                         \\
+%     nestjs           & TypeScript  & mysql mongo fastify fastify-mysql fastify-mongo    \\
+%     typescript-rest  & TypeScript  &                                                    \\
+%     deno             & TypeScript  &                                                    \\
+%     dropwizard       & Java        & mongodb postgres jdbi-postgres                     \\
+%     httpserver       & Java        & postgres                                           \\
+%     quarkus          & Java        & pgclient                                           \\
+%     jooby            & Java        & netty jaxrs jetty pgclient                         \\
+%     undertow         & Java        & postgresql                                         \\
+%     tapestry         & Java        &                                                    \\
+%     vertx-web        & Java        & mongodb postgres susom-postgres                    \\
+%     jlhttp           & Java        & postgres                                           \\
+%     voovan           & Java        &                                                    \\
+%     play2-java       & Java        & netty ebean-hikaricp ebean-hikaricp-netty jpa...   \\
+%     officefloor      & Java        & raw tpr micro thread\_affinity netty spring\_data  \\
+%     jersey           & Java        &                                                    \\
+%     proteus          & Java        & mysql                                              \\
+%     light-4j         & Java        &                                                    \\
+%     activeweb        & Java        & jackson                                            \\
+%     minijax          & Java        &                                                    \\
+%     jetty            & Java        & servlet                                            \\
+%     spark            & Java        &                                                    \\
+%     grizzly          & Java        & jersey                                             \\
+%     servlet          & Java        & lib-native mysql postgresql                        \\
+%     tio-mvc          & Java        &                                                    \\
+%     baratine         & Java        &                                                    \\
+%     servicetalk      & Java        &                                                    \\
+%     play1            & Java        &                                                    \\
+%     ninja-standalone & Java        &                                                    \\
+%     nanohttpd        & Java        &                                                    \\
+%     simple-server    & Java        &                                                    \\
+%     restexpress      & Java        & mysql-raw                                          \\
+%     javalin          & Java        &                                                    \\
+%     jawn             & Java        &                                                    \\
+%     redkale          & Java        & postgres                                           \\
+%     revenj-jvm       & Java        &                                                    \\
+%     spring-webflux   & Java        & mongo pgclient rxjdbc jdbc                         \\
+%     firenio          & Java        & http-lite                                          \\
+%     curacao          & Java        &                                                    \\
+%     bayou            & Java        &                                                    \\
+%     wizzardo-http    & Java        &                                                    \\
+%     netty            & Java        &                                                    \\
+%     armeria          & Java        &                                                    \\
+%     vertx            & Java        & postgres                                           \\
+%     comsat           & Java        & servlet-undertow webactors-netty webactors-un...   \\
+%     ratpack          & Java        & jdbc pgclient                                      \\
+%     wicket           & Java        &                                                    \\
+%     wildfly-ee       & Java        &                                                    \\
+%     undertow-jersey  & Java        & hikaricp                                           \\
+%     micronaut        & Java        &                                                    \\
+%     servlet3         & Java        & sync                                               \\
+%     act              & Java        & hibernate-pgsql hibernate-pgsql-rythm hiberna...   \\
+%     blade            & Java        &                                                    \\
+%     rapidoid         & Java        & mysql postgres http-fast                           \\
+%     smart-socket     & Java        &                                                    \\
+%     gemini           & Java        & mysql postgres                                     \\
+%     greenlightning   & Java        &                                                    \\
+%     helidon          & Java        &                                                    \\
+%     spring           & Java        & jpa mongo                                          \\
+%     codeigniter      & PHP         &                                                    \\
+%     fat-free         & PHP         & raw                                                \\
+%     php              & PHP         & activerecord raw7-tcp pgsql-raw h2o unit pool...   \\
+%     cakephp          & PHP         &                                                    \\
+%     ubiquity         & PHP         & workerman swoole swoole-mysql roadrunner road...   \\
+%     fuel             & PHP         &                                                    \\
+%     sw-fw-less       & PHP         &                                                    \\
+%     hyperf           & PHP         & raw micro                                          \\
+%     kumbiaphp        & PHP         & raw workerman workerman-mysql                      \\
+%     workerman        & PHP         & pgsql async                                        \\
+%     lumen            & PHP         & swoole                                             \\
+%     peachpie         & PHP         &                                                    \\
+%     yii2             & PHP         & raw                                                \\
+%     php-ngx          & PHP         & mysql pgsql async                                  \\
+%     laravel          & PHP         & swoole                                             \\
+%     symfony          & PHP         & raw swoole                                         \\
+%     imi              & PHP         & query-builder raw                                  \\
+%     webman           & PHP         &                                                    \\
+%     phalcon          & PHP         & mongodb micro                                      \\
+%     slim             & PHP         &                                                    \\
+%     amp              & PHP         &                                                    \\
+%     hhvm             & PHP         &                                                    \\
+%     spiral           & PHP         &                                                    \\
+%     swoole           & PHP         & postgres no-async                                  \\
+%     hamlet           & PHP         & swoole workerman                                   \\
+%     simps            & PHP         & micro                                              \\
+%     one              & PHP         &                                                    \\
+%     zend             & PHP         &                                                    \\
+%     phpixie          & PHP         &                                                    \\
+%     comet            & PHP         & mysql                                              \\
+%     nginx            & C           &                                                    \\
+%     onion            & C           &                                                    \\
+%     duda             & C           &                                                    \\
+%     libreactor       & C           & server                                             \\
+%     lwan             & C           & lua                                                \\
+%     h2o              & C           &                                                    \\
+%     facil\_io        & C           &                                                    \\
+%     octane           & C           &                                                    \\
+%     polkadot         & JavaScript  & mysql postgres mongodb                             \\
+%     restify          & JavaScript  &                                                    \\
+%     hapi             & JavaScript  & nginx mysql postgres                               \\
+%     nodejs           & JavaScript  & chakra mongodb mongodb-raw mysql postgres          \\
+%     restana          & JavaScript  &                                                    \\
+%     es4x             & JavaScript  &                                                    \\
+%     sailsjs          & JavaScript  & postgres                                           \\
+%     just             & JavaScript  &                                                    \\
+%     fastify          & JavaScript  & mysql postgres                                     \\
+%     ringojs          & JavaScript  & convenient                                         \\
+%     express          & JavaScript  & chakra mongodb mysql postgres graphql-mysql g...   \\
+%     koa              & JavaScript  & mysql postgres                                     \\
+%     0http            & JavaScript  &                                                    \\
+%     openresty        & Lua         &                                                    \\
+%     octopus          & Lua         &                                                    \\
+%     lapis            & Lua         &                                                    \\
+%     easyrpc          & C\#         &                                                    \\
+%     aspcore-mono     & C\#         & pg mw mw-pg mw-my mvc mvc-pg mvc-my                \\
+%     servicestack     & C\#         &                                                    \\
+%     tetsuweb         & C\#         &                                                    \\
+%     beetlex          & C\#         & core core-updb debug                               \\
+%     aspcore          & C\#         & pgo ado-pg ado-pg-up pgo-pg pgo-pg-up mw mw-j...   \\
+%     revenj           & C\#         &                                                    \\
+%     carter           & C\#         &                                                    \\
+%     aspnet-mono-ngx  & C\#         & jn ss my my-ef                                     \\
+%     genhttp          & C\#         &                                                    \\
+%     ben              & C\#         & pg                                                 \\
+%     evhttp-sharp     & C\#         &                                                    \\
+%     zysocket-v       & C\#         &                                                    \\
+%     nancy            & C\#         & netcore                                            \\
+%     aspcore-corert   & C\#         &                                                    \\
+%     cuehttp          & C++         &                                                    \\
+%     libhttpserver    & C++         &                                                    \\
+%     ffead-cpp        & C++         & mongo-raw lithium cinatra drogon libreactor h...   \\
+%     lithium          & C++         & postgres postgres-batch                            \\
+%     wt               & C++         & postgres                                           \\
+%     treefrog         & C++         & postgres mongodb                                   \\
+%     ulib             & C++         & plaintext\_fit json json\_fit mysql postgres po... \\
+%     cppcms           & C++         & postgres                                           \\
+%     oatpp            & C++         & no-mp async async-no-mp                            \\
+%     poco             & C++         &                                                    \\
+%     silicon          & C++         &                                                    \\
+%     cutelyst         & C++         & pf-apg pf-pg pf-my thread thread-apg thread-p...   \\
+%     drogon           & C++         & core                                               \\
+%     cinatra          & C++         &                                                    \\
+%     libsniper        & C++         &                                                    \\
+%     cpoll\_cppsp     & C++         & raw postgres-raw postgres-raw-threadpool           \\
+% \end{longtable}
 

### 2022-05-28 chapters/execution_environement/python/python.tex
add: python grpahs

@@
+\newpage
 \section{python}
@@
 \end{table}
+
+\subsection{python insights}
+The goal of this section is to find some insights where we can optimize the energy consumption without impacting the source code.
+so we started by extending the work of \citeauthor{hasan_energy_2016} and \citeauthor{oliveira_recommending_nodate} to the python environement.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/pythonloops}
+    \caption{energy behaviour of different python loops }
+    \label{fig:pythonloops}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/pythonarrays}
+    \caption{energy behaviour of different python data structures}
+    \label{fig:pythonarrays}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
+    \caption{energy behaviour based on multiprocessing}
+    \label{fig:python_multiprocessing}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django}
+    \caption{energy behaviour based on multiprocessing}
+    \label{fig:python_multiprocessing}
+\end{figure}
+
+Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
+As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
+
+asd we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
+where different sub process have to compete for the CPU Ressources.
+Another finding is when we hit the number of physical cores.
+there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
+basicaly he favorits the hyper threads than the physical core wich will lead to some context switches which cause the slow behaviour, howerver in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the eneryg consumption in this case.
+in the Chapter %% reference the work of intern from lyon  
+we discuss a deeper this behaviour of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
 
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/barplot_binarry_tree}
+    \caption{energy behaviour based on multiprocessing}
+    \label{fig:python_multiprocessing}
+\end{figure}
+
 \begin{figure}[thb]
@@
     \label{fig:p2}
-\end{figure}
-
-\begin{figure}[thb]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/bar_plot_tommtivector}
-    \caption{energy consumption of different impetions treating vectors }
-    \label{fig:bar_tommti_vector}
 \end{figure}

### 2022-06-01 chapters/execution_environement/python/python.tex
add : python loops

@@
 \newpage
-\section{python}
+\chapter{python}
 
-\subsection{Introduction}
+\section{Introduction}
 
@@
 More specifically, this chapter focuses on runtime optimizations that can adopted by developers to leverage the power consumption of Python applications.
-These optimizations include not only alternative interpreters, but also \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
+We start by studuing the impact of some of programmers choice such as the type of data structures or loops, on the global energy consumption of the execution code.
+then we discuss some other factors such as level of concurrency etc.
+later we we will talk about another non intrusive ways to optimize the energy consumption of this code.
+one kind of those optimizations includes alternative interpertersand some libraries that are dedicated to optimize the code without changing its structures such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
 
@@
 
-he simplicity, flexibility and the clarity of Python code led this programming language to a huge success.
+\subsection{python gluttony}
+
+the simplicity, flexibility and the clarity of Python code led this programming language to a huge success.
 Therefore, a larger community and a wide range of support and libraries that allowed it to touch several fields like data analyses, machine learning, natural language processing, web development, astrophysics etc.
@@
 
-Python consume a lot of energy mainly because it is slow in execution.
+Python consumes a lot of energy mainly because it is slow in execution.
 Its Flexibility and simplicity caused it to drop off in performances, because Python gains its flexibility from being a dynamic language, therefore it needs an interpreter to execute its programs, which makes them much slower compared to the other ones that are written in compiled programming languages such as C and C++ or semi-compiled languages like Java.
 
-As shown in Table \ref{fig:clbg}, we can see that in the most of the test cases Python takes more time to execute -the only case that he wasn't the last one was in the benchmark regx-redux where he beaten Go-, and in some cases the gab was huge such as in the n-body test where python took around 100 times more than C++.
+As shown in Table \ref{fig:clbg}, one can see that in the most of the test cases Python takes more time to execute -the only case that he wasn't the last one was in the benchmark regx-redux where he beaten Go-, and in some cases the gab was huge such as in the n-body test where python took around 100 times more than C++.
 
@@
     \caption{Comparison of the execution time between different programming languages using the computer language benchmarks game \cite{noauthor_which_nodate}}
+    %  TODO : ADD reference 
     \label{fig:clbg}
@@
 
-\subsection{python insights}
+In this chapter we will target the following research question :
+\begin{enumerate}
+    \item Do the data structres and controle structures impact the energy consumption ?
+    \item how does the parallelism affect the energy behaviour of python programs ?
+    \item are they any non intrusive methodes to reduce the imapct of the energy consumption of existing python softwares ?
+    \item how can one apply those findings in the real world ?
+\end{enumerate}
+
+\section{experimetal protocole}
+
+as we have discussed in the previous chapter \ref{chapter:benchmarking}.
+instead of running the tests the idea was to design a system that allows practiouners to reproduce and extends our tests.
+and then we use the same system to run answer some of researchs questions
+%%%% probably ill resue the same things from other chapters
+
+\section{python insights}
+
 The goal of this section is to find some insights where we can optimize the energy consumption without impacting the source code.
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/pythonloops}
-    \caption{energy behaviour of different python loops }
-    \label{fig:pythonloops}
-\end{figure}
+%%%% concurency
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/pythonarrays}
-    \caption{energy behaviour of different python data structures}
-    \label{fig:pythonarrays}
-\end{figure}
+Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
+As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
 
-\begin{figure}
+\begin{figure}[hbt]
     \centering
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/django}
-    \caption{energy behaviour based on multiprocessing}
-    \label{fig:python_multiprocessing}
-\end{figure}
-
-Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
-As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
-
-asd we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
+as we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
 where different sub process have to compete for the CPU Ressources.
@@
 
+another field of investigation is the type data and structure conroles. that might impact the energy consumption.
+To do so. We iterate over a list using three methodes.
+first the classical for ( for i in range (len(n))).
+howerver as we can see here unlike other porgramming langauges it requires extra operations such as determining the length of the collection , and then using the iterator range.
+so we tried the more adapted version
+( for element in collection).
+Morever in most programming languages the for loop is translated to a while loop ( transfirmation from D type to B type -- asm -- ), therefore we wanted to compare this with a ( while ) version.
+After determining the main ways to iterate over a loop we run the algorithms  collections of different data types to see wether it will impact of the energy consumption of the code, same thing for the size of the collection.
+
+As we can see in figure~\ref{fig:pythonloops} the type of the data hasn't an impact on the energy consumption.In the other hand the way we iterate over the collection has a huge factor.
+interestely, the for in range loop was by far the optimal one with following by the regural for in collection, and the while part was the least one with an overhead of 400\% compared to the first option.
+
+%%% putting the idea, later i ll reformulate it 
+the reason behind a such behaviour is mainly related on how python interpeter is made.
+% add refenece https://www.python.org/doc/essays/list2str/ 
+%% https://www.pythonpool.com/for-vs-while-loop-python/  
+% https://wiki.python.org/moin/TimeComplexity
+%% http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1
+In order to reduce the latency of the python code.
+most of the buitl-in functions an operations are written in C, same thing goes for the function range, further more the function len has a complexity of O(1) because it is based on the function Py\_SIZE  of C which stores the length in a field for the object.
+
+so basically for the for in range is basically creating a new iterator that has the same length of the first one.
+and for each iteration we will have to do a second acceess ( l[i]) instead of one one access hence the double time.
+
+for the wile is even slower due to the implicite incrementation of the variable which will cause an extra operation during the loop .
+
+to confirm the hyphothsis.
+we tried to construct a new list by editing the elements of the previous one.
+And as predicted , the builtin methodes are the best energy saving, while the customize while loop is the heaviest.
+an interesting finding is.
+the impact of annonymous functions ( lambda expressions) on the energy consumption.
+the reseason behind this is the fact that python treats those functions as local variables unlike the predifined ones which are global in our case.
+therefore they are faster and consumes less energy  % same reference https://www.python.org/doc/essays/list2str/
+
+\begin{figure}[htb]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/python_iterations}
+    \caption{energy behaviour of different python loops }
+    \label{fig:pythonloops}
+\end{figure}
+
+\begin{figure}[htb]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/python_treatemens}
+    \caption{energy behaviour of different methodes to change a list}
+    \label{fig:pythontreatement}
+\end{figure}
+
+%% should i include the part with the exceptions ???
+
+\newpage
 \subsection{Python Runtimes}
@@
     \label{fig:powerapi}
+\end{figure}
 
-\end{figure}
 \subsection{Power Measurements}
@@
 
+\subsection{use cases}
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django}
+    \caption{energy behaviour based on multiprocessing}
+    \label{fig:django}
+\end{figure}
+
 %% todo 

### 2022-06-09 chapters/execution_environement/python/python.tex
add : python usage

@@
 
+\subsection{use cases}
+according to the survey made by Jetbrains company \footnote
+{\fnurl{https://www.jetbrains.com/lp/python-developers-survey-2020}} data analysis and web develoepment are the top two purposes for python\ref{fig:usecase}
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/python_use_cases}
+    \caption{use cases of python }
+    \label{fig:usecase}
+\end{figure}
+
+\subsubsection*{web developement}
+first we wanted to test the energy consumption of python using on djagon,one of the most popular web frameworks.
+therefore we create a simple website where we try to analyse the cost of a single request using different mechanisms to fetch the data from the database.
+the idea behind this use case is to see wether the choice of the database and the orm ( object relational mapper) impacts the energy consumption and the perfromances of the website.
+
+we considered using two different databases postgresSQL and SQlite3 that containes the same data, and three different ways to fetch the data.
+\begin{enumerate}
+    \item the naive version: which relies only on the ORM to get the data
+    \item Prefetch : basically we just prefetch the data before we request it
+    \item optimized : and here we optimize our request on the SQL level without passing by the ORM
+\end{enumerate}
+
+\begin{figure}[!hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django}
+    \caption{energy behaviour based on multiprocessing}
+    \label{fig:django}
+\end{figure}
+
+%%%%% have to rephrase this
+
+As one can see in figure~\ref{fig:django}, different choices of requesting the data have a huge impact on the energy consumption.
+as the naive version can consume up to 10x more energy than the optimized one.
+In the other hand the choice of the data base didnt have huge impact on the total energy despite their different behaviour regarding the execution time and the average power.
+This can be useful to help develeppers make a choice regarding which data base they can use based on the number of the expected requests and the expectation of the performance.
+
+Another interesting observation is the impact of the interpreter, as figure~\ref{fig:django} highligh.
+the use of Pypy interpter instead of the default one helped reducing the amount of the energy consumption even when we are dealing with the naive version.
+
+\subsubsection*{machine learning}
+the other part is to study the impact of machine learning algorthims
+
+%% ADD the work of alex 
+\newpage
 \section{experimetal protocole}
@@
 %% should i include the part with the exceptions ???
+\subsection{conclusion}
+This study has shown us, that the optimial way to reduce the energy consumption of the python code is to follow the guidelines and use the builtin functions, which is kinda happy news for the developers since they don't have to make extra effort to make their code green
+%% except the parallellism part , which i dont know how to link it here yet
 
 \newpage
-\subsection{Python Runtimes}
+\section{Python Runtimes}
+One main inconvienent of the previous methode to reduce the energy consumption, is the hard work that should be done in order to alter the existing code base in order to redcue the energy consumption.
+To avoid such hustle we tried to find a non intrusive way to make the python code more ecofriendly without changing its structure.
+One aspect of python is the fact that it is an interptered language which led many works to create their own implementation of the interpeter to enhance one or many aspect of the python code.
+in the following section we will discuss the impact of those implementations on the energy consumption of python programs, and in which case, one should use a non conventional interpter to save the energy consumption of their program.
 
-we used the following implementations.
+To do so, we gathered a list of interpeters, transpilers and some optimization libraries.
 
-We should explain the methodology we used to select these
+the following list, presents the most famous ones .
+% We should explain the methodology we used to select these
 
@@
 
-\subsubsection{micro-benchmarks}
+% \subsubsection{micro-benchmarks}
 
@@
 
-\subsection{use cases}
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/django}
-    \caption{energy behaviour based on multiprocessing}
-    \label{fig:django}
-\end{figure}
+the energy consumption.
 
@@
 % add dram \ a glimps
-% add bitops
 
@@
 \subsection{conclusion}
+
+one can notice that the choice of the right interpeter can impact havely the energy consumption of the python programs.
+The fact that there is no general solution make this study more interesting.
+however the main drawback is the lack of the compatibility for some of those solutions which lead to some sacrifices when we want something generic

### 2022-06-10 chapters/benchmarking/challenges.tex
refactoring

\section{Threats and Challenges}
A successful benchmark has three criterions to fullfil.
First, it has to be \emph{reproducible} for others to replicate.
Second, the results should be \emph{accurate}, which means each time we rerun the benchmark we are expecting the same results.
Finally, it should \emph{represent} the real world.
In other words, the conclusions brought from the experiment should be valid outside the experimentation as well.
In my case, the real world is the production environment.
Therefore, our experiments should reflect what is happening in the production environments.
In this section, I will deeper dig in each aspect, present what has been done by others and my contributions, and finally I will propose some perspectives, to improve such experiments.

%%%%%%%%% preliminary taughts and section structure 
% The need of reproducibility in our field - software optimization based on empirical studies -

% The importance of Virtualisation for reproducibility \cite{howe_virtual_2012}
% some of the most important parts are
% - fewer constraints on research methods
% - on-demand backups
% - virtual Machines as Publications
% - more Variables captured

\subsection{Reproducibility}
One of the most difficult challenges faced by researchers is the reproducibility of their benchmarks.
Actually, many results fail to be reproduced,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
As this crisis hit most of the empirical studies, most of the reviews now includes reproducibility as one of the minimal standard for judging scientific merit.\cite{peng2011reproducible}
One of the criterions supporting reproducibility is the publication of the dataset and the algorithms run on the raw data to conclude the results.
There is even some disagreement about what the terms "reproducibility" or "replicability" by themselves mean \cite{goodman2016does}.
According to \cite{echtler2018open}, \emph{replicability} extends \emph{reproducibility} with the ability to collect a new raw dataset comparable to the original one by re-executing the experiment under similar condition, instead of just the ability to get the same results by running the statistical analyses on the original data set.

In the area covered by this PhD thesis, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only among homogeneous nodes, but even within a single node.

One major problem that hinders the reproducibility of the empirical benchmarks is the interaction with the external environment, either as concurrency or dependencies.
Therefore, researchers cannot observe the same results, unless they duplicate the same environment.

\subsection{Accuracy}
According to Oxford, \emph{accuracy} means \href{https://www.lexico.com/definition/accuracy}{"technical The degree to which the result of a measurement, calculation, or specification conforms to the correct value or a standard"}.
In ny case, this means the ability to run the benchmark multiple times with a low variation.

Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.

In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of physical measurements.
Unfortunately, this is not possible when it comes to measuring the energy consumption of a system.
Applying the benchmarking guidelines and repeating the same experiment with in the same configuration are not sufficient to reproduce the the same energy measurements, not only between identical machines, but even within the same machine.
This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.

%TODO:  rephrase this figure, maybe change it a little bit 
Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
We can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.

\begin{figure}%[!htb]
    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
\end{figure}

Some researchers started investigating the hardware impact of the energy variation of power consumption.
As an example, one can cite~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the main cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
\cite{heinrich_predicting} described this variation as a set of parameters, such as CPU Frequency and the thermal effect.

\subsection{Representativeness}
As obvious as it seems, the reason of executing benchmarks is to validate ideas so we can use them in the real life.
However, this is means that those benchmarks have to represent reality somehow.
Basically, when we aim to benchmark something, we create a mock up version of the situation that we want it to work.
% But, how can we assure that the benchmark is representative? honestly i don't know. and we can't generalize this. but there are some works that have been done for this.
First, we can talk about the benchmarking and their selection, then we will talk about the stress benchmark for some applications, and finally we will bring this representativeness in our case and how can we get closer to the energy consumption behavior of a software between the benchmark environnement and the production one.

As Stephen M. Blackburn~\emph{et~al.} cited in their paper "evaluate collaboratory"~\cite{stephen_evaluate_2012}, one of the major pitfalls of the measurement contexts is the inconsistency, which can be translated here by the fact that the production context is not the same as the benchmarking one.

Another difficult part for practitioners is to generalize the claims they reached beyond the lab conditions.
Are they appropriate?
Are they consistent and are they reproducible?
To answer those questions, the community agreed on some wellknown benchmarks to represent a specific concern of the production world.
One can site as an example the Dacapo and Renaissance benchmark suites for Java applications, or the CLBG benchmark suite for comparing programming languages.
Although they do not cover all the cases, the community agrees on their relevance and representativeness.

In addition to those benchmarks, a new category of benchmarking technique emerged to simulate the worst case of the production environments, inlcuding performance tests, which are benchmarks meant to evaluate the behavior of a software under stress situations.
We can cite as an example the Gatling for web application and stress-ng for hardwares.%ROMAIN: Add citations.
%% YEP I KNOW, THIS IS JUST FOR ME TO READ IT LATER AND GIVE A BETTER CONTEXT

### 2022-06-10 chapters/benchmarking/environement.tex
refactoring

--- 2021-10-21 chapters/testing/environement.tex
@@
 
-Based on the studies of \cite{eddie_antonio_santos_how_nodate}, who analysed the impact of adding the docker layer on the energy consumption.
+Based on the studies of \cite{santos2018does}, who analysed the impact of adding the docker layer on the energy consumption.
 In their experement. Eddie Antonio et al run multiple benchmarks with and without docker. and compared their energy consumption and execution time.
@@
     \center{\includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/idle_energy}}
-    \caption{energy consumption of Idle system with and without docker \cite{eddie_antonio_santos_how_nodate}}\label{fig:docker_idle}
+    \caption{energy consumption of Idle system with and without docker \cite{santos2018does}}\label{fig:docker_idle}
 \end{figure}
@@
     \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_energy}
-    \caption{execution time and energy consumption of Redis  with and without docker \cite{eddie_antonio_santos_how_nodate}}\label{fig:docker_reddis}
+    \caption{execution time and energy consumption of Redis  with and without docker \cite{santos2018does}}\label{fig:docker_reddis}
 \end{figure}

### 2022-06-10 chapters/benchmarking/experement.tex
refactoring

--- 2021-10-21 chapters/testing/experement.tex
@@
 First, \textbf{scalability} : We wanted to gather the highest possible amount insights (regarding time spent for the experiment, therefore we wanted some benchmarks who can scale according to the number of PUs and can fulfill different scenarios.
-Second creterion is \textbf{representativeness} : as menstionned in the challenges, a workload has to be representative otherwise the experiment would be inconsistent \cite{evaluate_Collaboratory}.
+Second creterion is \textbf{representativeness} : as menstionned in the challenges, a workload has to be representative otherwise the experiment would be inconsistent \cite{stephen_evaluate_2012}.
 To satisfy those creterions we went through state of the art and looked for the most used benchmarks for testing the performance of the hardware, and then we selected the scallable ones.

### 2022-06-10 chapters/benchmarking/stateoftheart.tex
refactoring

--- 2021-10-21 chapters/testing/stateoftheart.tex
@@
 \paragraph{Studying Hardware Factors.}
-
 This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
@@
 Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
-On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the correlation between the CPU temperature and the energy consumption variation is very tight.
+On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the CPU temperature and the energy consumption variation are tightly coupled.
 
@@
 
-This makes the processor temperature a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.
+This exposes the processor temperature as a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.
 
@@
 %% this is to prove that the problem is REAL !!!
+
 \paragraph{Mitigating Energy Variations.}
@@
 Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
-They also proposed some guidelines to reduce this variation by replacing the old slow chips, by load balancing the workload on the CPU cores and leaving one core idle.
+They also proposed some guidelines to reduce this variation by replacing the old---slower---chips, by load balancing the workload on the CPU cores and leaving one core idle.
 They claimed that the variation between the processor cores is insignificant.
@@
 
-\subsection*{Our position}
-Meanwhile the previoius work was to create a large umbrella tht englobes all emprical tests related to computer science, we want to narrow our study to energy testing only.
-In other words we will study pitfalls that hinders the energy claimes of software .
-
-We are fully aware of the impact that hardware has on energy variation.
-However, we believe that there is still a room to reduce this energy variation for practicioners using only paramenters that they are in cotrol.
-To do so, we have inducted a set of empirical experiments using the guidelines provided by state of the art, to determine which controllable factors can reduce the variation of the energy consumption of tests.
-We will first start with evaluating the paramenters mentioned by \cite{heinrich_predicting}.
+\subsection*{Objective}
+While the state of the art aims to create a large umbrella that covers all empirical benchmarks related to computer science, we rather aim to narrow our study to energy benchmarking only.
+In other words, we will focus on pitfalls that hinders the energy claims of software.
+We are clearly aware of the impact of hardware on energy variations.
+However, we believe that there is still a room to control this energy variation for practitioners by using only parameters that can be tuned.
+To do so, we have inducted a set of empirical experiments using the guidelines provided by state of the art, to determine which controllable factors can reduce the variation of the energy consumption of benchmarks.
+We will first start with evaluating the parameters mentioned by \cite{heinrich_predicting}.

### 2022-06-10 chapters/introduction.tex
refactoring

@@
 
-Basically we want to save the planet, and save your pockets within it.
-We know that humans don't care about anything else except their pockets. So let's try to push them into saving our mother earth by making them reduce their costs and increase their benifits.
-How to do so ?
-well we will work on reducing the elecricity consumption of their services. without making them to change their machines, their providers or impacting the quality of service that they provide.
+Nowdays, computers are invasing our daily life from work to leasure, from fancy smartphone to an embded peacemaker that regulatres the heartbeat of some people.
+As human beigns we are known to use tools to enhance our bodies.
+And thanks to computers we pushed that step even further where now we are using machines to extend our brains.
+Some even argue that one day they will replace us, and therefore we created our own ending.
+Well this is the problem for the future generations.
+For the moment, the main concern of humanity is to keep this planet livebale until we find another alternative.
 
+Unfortunately those machines doesn't help that much. as according to .... ....
+
+\note{add the impact of the ict ..etc}
+
+Researchers are trying to solve this issue through different angles.
+While some scientists are trying to find an alternative green source to generate energy.
+others focus on reducing it's consumption.
+In computer science we are concerned with the ladder solution.
+Therefore most of the works are done on tunning the software and the hardware in ordrer to have a more efficient way use this energy.
+
+In this thesis we are trying to find a way to make the energy consumption of the computer to be as low as possible.
 Our approach is is to reduce the energy consumption of the software services by changing certains parameters, such as palteform, programming langauge, and/or the design pattern.
 
-well to do so we have multiple approachs ,
-1- by using formal apporachs such as static analysers, complexity , nad programming theories -- lets check maybe there is a state of the art -
-2- second by a more interactive appreach where based on tests and feedback
+The best way to do so is to formulate a theory behind the energy consumption of algorthms such as the complexity and the o notation.
+Unfortunately this is not possible in the current state of the art.
+due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.
 
-This theis is about the second one, So basically all we do is testing programms, measuring their energy and then provide a feedback based on our observations.
-To do so we first need to setup a testing enviroment. something that allows us to have accurate, precise and reproductible tests.
-Therefore the first chapter our the thesis will be dedicated to provide green coders with a set of elements in orders to test their hypothesis.
+Unlike algrothim optimization in the field of perfornacne which is agnostic toward the plateform. the energy consumption of the algorithms is dependent on execution environement.
+Therefore, for the moment we will start by formulating some hypthosis and explore them using empirical analysis.
+
+Our work will be presented through the following chapters :
+
+\begin{enumerate}
+    \item \ref{chapter:literature_review}: Where we discuss the work done on the energy consumption and optimization in software engineering
+    \item \ref{chapter:benchmarking}: It will present a set of guidelines and tools to help practitioners measure the energy consumption of their algorithms.
+    \item : we will present the impact of programming languages on the energy consumption of the algorithms.
+    \item : it will discuss the behaviour of python and the possibile ways to tune it in ordrer to reduce the energy consumpion
+    \item : will present a study on java programming language and the impact of the JVM choice on the energy consumpion
+    \item : as a prespective we introduce the impact of parallelisme on the enrgy consumption on time agnostic cases
+\end{enumerate}
 
@@
 \section{Research Contributions}
-..... This work makes the following contributions:
 
@@
 
-    \item Introduces a new ...
+    \item Introduces
     \item Shows how ....

### 2022-06-10 chapters/java/java-dir/java.tex
refactoring

--- 2022-05-18 chapters/execution_environement/java/java.tex
@@
-\section{Introduction}
+\section*{JAVA}
+\subsubsection{Introduction}
 As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.%ROMAIN: Add source
@@
 
-\subsection{Goal}
+\paragraph{Goal}
 In this section, we will investigate the following research questions:
@@
 
-\subsection{definition of the JVM}
+\subsubsection{definition of the JVM}
 \begin{figure}%[ht]
@@
 
-\section{Experimental Protocol}\label{sec:meth}
+\subsection{Experimental Protocol}\label{sec:meth}
 To investigate the effect that could have the JVM distribution choice and/or parameters on software energy consumption, we conducted a wide set of experiments on a cluster of machines and using several established Java benchmarks and JVM configurations.
 
-\subsection{Measurement Contexts}
+\subsubsection{Measurement Contexts}
 
@@
 
-\subsection{Workload}
+\subsubsection{Workload}
 We ran our experiments across $12$ Java benchmarks we picked from \textsf{OpenBenchmarking.org}.\footnote{\url{https://openbenchmarking.org}}
@@
 
-\subsection{metrics and measurement}
+\subsubsection{metrics and measurement}
 Since the goal of this study is the green aspect Java virtual machines, our key metric will be the energy consumption of the job completed by each configuration.
@@
 
-\section{Experiments \& Results}\label{sec:exp}
-\subsection{Energy Impact of JVM Distributions}
+\subsection{Experiments \& Results}\label{sec:exp}
+\subsubsection{Energy Impact of JVM Distributions}
 \noindent\textbf{Job-oriented applications.}
@@
 
-% \subsubsection{Energy Impact of JVM Distributions on services}
+% \paragraph{Energy Impact of JVM Distributions on services}
 \vspace{6pt}
@@
 \end{figure*}
-\bigbreak
+% \bigbreak
 \begin{mdframed}[]
@@
 
-\subsection{Energy Impact of JVM Settings}
+\subsubsection{Energy Impact of JVM Settings}
 The purpose of our study is not only to investigate the impact of the JVM platform on the energy consumption, but also the different JVM parameters and configurations that might have a positive or negative effect, with a focus on 3 available settings: multi-threading, JIT, and GC.
 
-\subsubsection{Multithreading}
+\paragraph{Multithreading}
 The purpose behind this phase is to investigate the impact JVM thread management strategies on the energy consumption.
@@
 
-\subsubsection{Just-in-Time Compilation}
+\paragraph{Just-in-Time Compilation}
 The purpose of experiments on JIT is to highlight the different strategies that can impact software energy consumption within a JVM and between JVMs.
@@
 
-\subsubsection{Garbage Collection}
+\paragraph{Garbage Collection}
 Changing or tuning the GC strategy has been acknowledged to impact the JVM performances~\cite{10.1145/2568088.2568097}.
@@
 
-\section{Threats to Validity}\label{sec:threats}
+\subsection{Threats to Validity}\label{sec:threats}
 A number of issues may affect the validity of our work.
@@
 
-\section{Conclusion}
-The results of our investigations showed that many JVMs share energy efficiencies and can grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
-The 3 selected JVM classes can however report a different energy efficiency for different software and/or workloads, sometimes by a large margin.
-While we did not observed a unique champion when it comes to energy consumption, \textsc{GraalVM} reported the best energy efficiency for a majority of benchmarks.
-Nonetheless, each JVM can achieve the best or the worst depending on the hosted application.
-One cause can be thread management strategies, as observed with J9 when advantageously running \textsf{Avrora}.
-Moreover, some JVM settings can cause energy consumption variations.
-Our experiments showed that the default JIT compiler of the JVM is often near-optimal, in at least ~80\% of our experiments.
-The default GC, however, was outperforming alternative strategies in half of our experiments, with some large gains observed when using some alternative GC depending on the application characteristics.
-
-Our main conclusions and guidelines can be thus summarized as:
-\begin{inparaenum}[\em i)]
-    \item testing software on the 3 classes of JVM and identifying the one that consumes the least is a good practice, especially for multi-threading purposes,
-    \item while the JVM default JIT give often good energy consumption results, some settings may improve the energy consumption and could be tested,
-    \item the choice of the GC may lead to a large impact on the energy consumption in many situations, thus encouraging a careful tuning of this parameter prior to deployment.
-\end{inparaenum}
-To ease the integration of the above guidelines, we propose a tool, named \textsf{J-Referral}, to recommend the most energy-efficient JVM distribution and configuration among more than a hundred considered possibilities.
-It establishes a full report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to help the user to choose the one with the least consumption for a Java Software.
-
-\section{ Tools and contributions }
+\subsection{Tools and contributions}
 \Cref{table:j-referral} illustrates an example of the final report of \textsf{J-Referral}.
@@
 
+\subsection{Conclusion}
+The results of our investigations showed that many JVMs share energy efficiencies and can grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
+The 3 selected JVM classes can however report a different energy efficiency for different software and/or workloads, sometimes by a large margin.
+While we did not observed a unique champion when it comes to energy consumption, \textsc{GraalVM} reported the best energy efficiency for a majority of benchmarks.
+Nonetheless, each JVM can achieve the best or the worst depending on the hosted application.
+One cause can be thread management strategies, as observed with J9 when advantageously running \textsf{Avrora}.
+Moreover, some JVM settings can cause energy consumption variations.
+Our experiments showed that the default JIT compiler of the JVM is often near-optimal, in at least ~80\% of our experiments.
+The default GC, however, was outperforming alternative strategies in half of our experiments, with some large gains observed when using some alternative GC depending on the application characteristics.
+
+Our main conclusions and guidelines can be thus summarized as:
+\begin{inparaenum}[\em i)]
+    \item testing software on the 3 classes of JVM and identifying the one that consumes the least is a good practice, especially for multi-threading purposes,
+    \item while the JVM default JIT give often good energy consumption results, some settings may improve the energy consumption and could be tested,
+    \item the choice of the GC may lead to a large impact on the energy consumption in many situations, thus encouraging a careful tuning of this parameter prior to deployment.
+\end{inparaenum}
+To ease the integration of the above guidelines, we propose a tool, named \textsf{J-Referral}, to recommend the most energy-efficient JVM distribution and configuration among more than a hundred considered possibilities.
+It establishes a full report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to help the user to choose the one with the least consumption for a Java Software.
 % TODO : add or change the graphes of the use case 

### 2022-06-10 chapters/programming_languages/programming_languages_chapter.tex
refactoring

--- 2021-10-29 chapters/programming_languages/challanges.tex
@@
+\chapter{Energy footprint of programming languages}
+\label{chapter:porgramming_langauges}
+In this chapter, we discuss the impact of the choice of the programming language on the energy consumption of the resulting software.
+To do so, we suggest to start with the general micro benchmarking and see how each programming languages interact with the CPU/Memory.
+
 The ultimate goal of this chapter is to provide a guideline on which programming language the developers should chose based on the charecteristics of the project in order to minimize the energy foot print of their product.
@@
 
-.
+As we have seen in the previous chapter, one of the most important feature of a test is to be \textbf{representative}.
+Therefore, we extend this study to some real-life use case, and the following sections below provide two case studies.
+\section{Remote Procedure Call Frameworks}
+\import{\currfiledir/rpc-dir}{rpc}
+\section{web Frameworks}
+\import{\currfiledir/web_frameworks-dir}{web_frameworks}

### 2022-06-10 chapters/programming_languages/rpc-dir/rpc.tex
refactoring

--- 2022-05-11 chapters/programming_languages/rpc.tex
@@
-\section{Remote Procedure Call Frameworks}
-
-\subsection{Definition}
-
-\subsection{Motivation}
+\subsection*{introduction}
 %  this is just to warmup 
@@
 Other approaches implement all the different interfaces of the protocol by themselves, such as RPC.
-
-\subsection{State of the art}
-TODO
 
@@
     \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/energy_cost_clients}
+        \includegraphics[width=1.2\linewidth]{imgs/energy_cost_clients}
     \end{center}
@@
     \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/energy_cost_payload}
+        \includegraphics[width=1.2\linewidth]{imgs/energy_cost_payload}
     \end{center}
@@
     \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/power_consumption_clients}
+        \includegraphics[width=1.2\linewidth]{imgs/power_consumption_clients}
     \end{center}
@@
     \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/power_consumption_payload}
+        \includegraphics[width=1.2\linewidth]{imgs/power_consumption_payload}
     \end{center}
@@
     \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/rps_clients}
+        \includegraphics[width=1.2\linewidth]{imgs/rps_clients}
     \end{center}
@@
     \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/rps_payload}
+        \includegraphics[width=1.2\linewidth]{imgs/rps_payload}
     \end{center}
@@
     \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/tail99_clients}
+        \includegraphics[width=1.2\linewidth]{imgs/tail99_clients}
     \end{center}
@@
     \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rpc_images/tail99_payload}
+        \includegraphics[width=1.2\linewidth]{imgs/tail99_payload}
     \end{center}

### 2022-06-10 chapters/benchmarking/draft.tex
refactoring : continuation

--- 2022-05-06 chapters/testing/conclusion.tex
@@
+To optimize the software energy consumption, a reproducible setup environment requires to be designed.
+
+This chapter will be dedicated to provide a set of guidelines to provide a \em{reproducible, accurate} and \em{representative} tests. Each aspect will be detailed in the sections below.
+
+\section{Threats and Challenges}
+A successful benchmark has three criterions to fullfil.
+First, it has to be \emph{reproducible} for others to replicate.
+Second, the results should be \emph{accurate}, which means each time we rerun the benchmark we are expecting the same results.
+Finally, it should \emph{represent} the real world.
+In other words, the conclusions brought from the experiment should be valid outside the experimentation as well.
+In my case, the real world is the production environment.
+Therefore, our experiments should reflect what is happening in the production environments.
+In this section, I will deeper dig in each aspect, present what has been done by others and my contributions, and finally I will propose some perspectives, to improve such experiments.
+
+%%%%%%%%% preliminary taughts and section structure 
+% The need of reproducibility in our field - software optimization based on empirical studies -
+
+% The importance of Virtualisation for reproducibility \cite{howe_virtual_2012}
+% some of the most important parts are
+% - fewer constraints on research methods
+% - on-demand backups
+% - virtual Machines as Publications
+% - more Variables captured
+
+\subsection{Reproducibility}
+One of the most difficult challenges faced by researchers is the reproducibility of their benchmarks.
+Actually, many results fail to be reproduced,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
+As this crisis hit most of the empirical studies, most of the reviews now includes reproducibility as one of the minimal standard for judging scientific merit.\cite{peng2011reproducible}
+One of the criterions supporting reproducibility is the publication of the dataset and the algorithms run on the raw data to conclude the results.
+There is even some disagreement about what the terms "reproducibility" or "replicability" by themselves mean \cite{goodman2016does}.
+According to \cite{echtler2018open}, \emph{replicability} extends \emph{reproducibility} with the ability to collect a new raw dataset comparable to the original one by re-executing the experiment under similar condition, instead of just the ability to get the same results by running the statistical analyses on the original data set.
+
+In the area covered by this PhD thesis, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
+However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only among homogeneous nodes, but even within a single node.
+
+One major problem that hinders the reproducibility of the empirical benchmarks is the interaction with the external environment, either as concurrency or dependencies.
+Therefore, researchers cannot observe the same results, unless they duplicate the same environment.
+
+\subsection{Accuracy}
+According to Oxford, \emph{accuracy} means \href{https://www.lexico.com/definition/accuracy}{"technical The degree to which the result of a measurement, calculation, or specification conforms to the correct value or a standard"}.
+In ny case, this means the ability to run the benchmark multiple times with a low variation.
+
+Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.
+
+In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of physical measurements.
+Unfortunately, this is not possible when it comes to measuring the energy consumption of a system.
+Applying the benchmarking guidelines and repeating the same experiment with in the same configuration are not sufficient to reproduce the the same energy measurements, not only between identical machines, but even within the same machine.
+This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
+
+%TODO:  rephrase this figure, maybe change it a little bit 
+Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
+We can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
+
+\begin{figure}%[!htb]
+    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
+    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
+\end{figure}
+
+Some researchers started investigating the hardware impact of the energy variation of power consumption.
+As an example, one can cite~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the main cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
+\cite{heinrich_predicting} described this variation as a set of parameters, such as CPU Frequency and the thermal effect.
+
+\subsection{Representativeness}
+As obvious as it seems, the reason of executing benchmarks is to validate ideas so we can use them in the real life.
+However, this is means that those benchmarks have to represent reality somehow.
+Basically, when we aim to benchmark something, we create a mock up version of the situation that we want it to work.
+% But, how can we assure that the benchmark is representative? honestly i don't know. and we can't generalize this. but there are some works that have been done for this.
+First, we can talk about the benchmarking and their selection, then we will talk about the stress benchmark for some applications, and finally we will bring this representativeness in our case and how can we get closer to the energy consumption behavior of a software between the benchmark environnement and the production one.
+
+As Stephen M.
+Blackburn~\emph{et~al.} cited in their paper "evaluate collaboratory"~\cite{stephen_evaluate_2012}, one of the major pitfalls of the measurement contexts is the inconsistency, which can be translated here by the fact that the production context is not the same as the benchmarking one.
+
+Another difficult part for practitioners is to generalize the claims they reached beyond the lab conditions.
+Are they appropriate?
+Are they consistent and are they reproducible?
+To answer those questions, the community agreed on some wellknown benchmarks to represent a specific concern of the production world.
+One can site as an example the Dacapo and Renaissance benchmark suites for Java applications, or the CLBG benchmark suite for comparing programming languages.
+Although they do not cover all the cases, the community agrees on their relevance and representativeness.
+
+In addition to those benchmarks, a new category of benchmarking technique emerged to simulate the worst case of the production environments, inlcuding performance tests, which are benchmarks meant to evaluate the behavior of a software under stress situations.
+We can cite as an example the Gatling for web application and stress-ng for hardwares.%ROMAIN: Add citations.
+%% YEP I KNOW, THIS IS JUST FOR ME TO READ IT LATER AND GIVE A BETTER CONTEXT
+
 %  TODO: add the environement part,
@@
 \end{figure}
-
-% However for them to be more representatives we wanted to simulate the production enviromenet.
-%% should i add  the thing about green faas ? 
-%% same about what we have done with python workers and lazy ? basically just a graph

### 2022-06-10 chapters/benchmarking/stateoftheart.tex
refactoring : continuation

@@
 We will first start with evaluating the parameters mentioned by \cite{heinrich_predicting}.
+
+% introduction of vms 
+\subsection{Virtualisation}
+To resolve this problem, practitioners tend to use Virtualisation.
+Using virtual machines aka VM gives reaserchers the freedom to choose their own tools, software and operating system that they are the most confortable with without paying the price to change the actual working environment, which will give them eventually more controle over the dependencies and the execution environement.
+Moreover, using a vm will solve the \emph{replication crisis} thanks to the virtual images, even the most complex architecture can be reproduced easily by just instantiating a copy of the image.
+Since the virtual machines are agnostic to the host architecture, reaserchers won't have to worry about where and how their experiments are replicated because they have already setup the execution environement.
+Another advantage to the virtual machines is the snapshot mechanism, it allows reaserchers to create backups and revert some changes with simple clicks.
+Last but not least,thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables -controled and uncontroled-  and do other analysis without dealing with any dependencies.
+In his paper \cite{howe_virtual_2012} bill howe lists the advantages of using virtual machines in reaserchers experements including the economical impact and cultural limitation to a such approach.
+
+which allow them to have control over the ressouces, the dependencies and the execution environment.
+Moreover, thanks to the snapshots, deploying a software is easily done by instantiating a copy of that image.
+However, this choice comes with a certain cost.
+Because the intervention of the hypervisor,the software will use two kernels , the virtual machine one and the host machine one, which will provide a noticiable overhead, and will impact the performances of the tests.Therefore, we can't use virtual machines for exepements that are related to performance.
+Another limiation with the virtual machine is the the isolation.
+It is true that this feature will prevent the experience environement with any undesirable interference from the outside world.
+but sometimes this contact is needed, especially when the experiment is dependent to an external part, such as sensors.
+In energicial tests we tend to use hardware powermeters which will make it difficult to use the virtual machines in this case .
+
+% introduction of docker
+
+\begin{figure}
+    \center{\includegraphics[width=1\linewidth]{imgs/virtualization_techniques}}
+    \caption{Different Methods of Virtualzation}\label{environement:virtualization_technique}
+\end{figure}
+
+\subsection{Containers}
+
+another solution would be using something that allows us to have the isolation from the host os and the ease of replication that virtual machines offer, and the direct interaction with the hardware that the classical method give.
+contarization offers a such advantages while keeping the isolation and the ease of replication for application.
+
+Figure \ref{environement:virtualization_technique} explains the differents in architecture between the clasic types2 of Virtualisation and Containers.
+\begin{itemize}
+    \item Type 1: runs directly on the hardware, it is mainly used by the cloud providers where there is no main OS, but just virtaul machines, we can site for this the open-srouce XEN and VMware ESX
+    \item Type 2: runs over the hostmachine Operating System, mostly used for personal comptuers, VMware server and virtualBox are famous examples of this type, most of the reaserchers expermentation are run witht this type, howerver due to the 2 Operatings syestms the applications tend to be more slower
+    \item containers : Instead of its own kernel, containers used the hots kernel to run their Os, which makes them ligher, quickers and use the full pontial use of the hardware.
+For this we can cite \emph{Docker}, \emph{Linux LXC}\emph{LXD} \cite{abuabdo_virtualization_2019}
+\end{itemize}
+%  TODO Rephrase so i include it before VMs+ add links to the technologie
+
+\subsection{Docker vs Virtual Machine}
+Depsite that Type 1 is more performant than type 2, the second one is the most used in reaserch, since most researchers conducts their experements in their own machine.
+In the other hand, docker is the most famous thechonology for for containers.
+In ower case we are more prone to docker for two reasons.
+\begin{enumerate}
+    \item we need a litetweight orchestrator to not affect the energy consumption of ower tests. As prior work mentioned [cite Morabito (2015) and van Kessel et al. (2016)]
+          % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
+    \item since we are using the hardware itself to measure the energy consumption, we are required to interract with the host OS itself.
+\end{enumerate}
+Special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that allows us to retireve the energy consumption of a virtual machine.
+%%%%% THINGS TO ADD : 
+%% -------------- Dont think so 
+% the impact of docker on the reprodudicibility 
+% Limits of docker 
+% The docker files and the clarification of the methodology 
+\subsection{Docker and energy}
+Now that we have chosen to go with the containers technorolgy to encapsulate our tests. What would be the impact of this solution on the energy consumption of our tests.
+
+Based on the studies of \cite{santos2018does}, who analysed the impact of adding the docker layer on the energy consumption.
+In their experement. Eddie Antonio et al run multiple benchmarks with and without docker. and compared their energy consumption and execution time.
+The first step was to see the impact of docker deamon while there is no work. to see the impact of the orchestrator alone. Later they had the experiment with the following benchmarks
+\begin{itemize}
+    \item wordpress
+    \item reddis
+    \item postfresSQL
+\end{itemize}
+The following figures represents the energy consumption of the system while it is idle. As we can see  in figure \ref{fig:docker_idle} Docker braught arround 1000joules overhead.
+\begin{figure}
+    \center{\includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/idle_energy}}
+    \caption{energy consumption of Idle system with and without docker \cite{santos2018does}}\label{fig:docker_idle}
+\end{figure}
+In the other hand as we can see in figure \ref{fig:docker_reddis}.
+docker increased the execution time of the benchmark by 50 seconds which caused an increase in energy since they are highly correlated.
+The authors also highlated the fact that this increase of energy consumption is due to the docker deamon and not the fact that the application is in a container.
+Moreover they estimted the price of this extra energy and it was less than 0.15\$ in the worst case.
+Which is non significant compared to the advantages that docker bring for isolation and reproducibility.\\
+if we  recap this study in one sentence,it would be the following one.
+The dockerised softwares tend to consume more energy, because mainly they take more time to be executed.
+The average power consumption is higher with only \textbf{2Watts} and it is due to the docker deamon.This overhead can be up to 5\% for IO intensive application, but it is mearly noticiable when it comes to CPU or DRMA intesive works
+
+\begin{figure}
+    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_time}
+    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_energy}
+    \caption{execution time and energy consumption of Redis  with and without docker \cite{santos2018does}}\label{fig:docker_reddis}
+\end{figure}
+
+%  basically just rephrase the paper
+
+\subsection{docker and accuracy}
+%% TODO : add an introduction, maybe some more details 
+And now Since the stat of the art has aggreed on the impact of docker on the energy consumption,Let's discuss it's impact on the accuracy. In other words\\
+\textbf{RQ :} does Docker affect the energy variation of the exepements ?
+
+To Answer this question we have conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
+Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
+In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
+The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
+
+\begin{figure}
+    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
+    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
+\end{figure}

### 2022-06-10 chapters/programming_languages/programming_languages.tex
refactoring : continuation

@@
-\newpage
-
 \chapter{Energy footprint of programming languages}
@@
 
-\import{chapters/programming_languages/}{challanges}
-\import{chapters/programming_languages/}{microbenchmarks}
+The ultimate goal of this chapter is to provide a guideline on which programming language the developers should chose based on the charecteristics of the project in order to minimize the energy foot print of their product.
+No answer is evidedent for a such question. However, we can extract some feartures of each programming langues such as
+\begin{itemize}
+    \item performance
+    \item community support
+    \item scalability
+    \item energy consumptiom
+    \item memory usage
+    \item etc
+\end{itemize}
+
+first we will start but analysing the behaviour of the general purpose programming langauges with some micro benchkaks, principally for the CLBG game and others from rosetta code base
+
 As we have seen in the previous chapter, one of the most important feature of a test is to be \textbf{representative}.
 Therefore, we extend this study to some real-life use case, and the following sections below provide two case studies.
-
-\import{chapters/programming_languages/}{rpc}
-\import{chapters/programming_languages/}{web_frameworks}
-\import{chapters/programming_languages/}{conclusion}
+\section{Remote Procedure Call Frameworks}
+\import{\currfiledir/rpc}{rpc}
+\section{web Frameworks}
+\import{\currfiledir/web_frameworks}{web_frameworks}

### 2022-06-11 chapters/benchmarking/benchmarking.tex
refactoring
refactoring : continuation
refactoring : stateoftheart

--- 2021-08-30 chapters/testing/testing.tex
@@
 \newpage
-\chapter{Testing}
-\label{chapter:testing}
+\label{chapter:benchmarking}
+\chapter{Benchmarking Protocol for measuring Energy Consumption of stoftwares }
 
-In order to optimize the energy consumption of software we first want to setup a testing enviroment.
-This chapter will be dedicated to provide a set of guidelines to provide a \em{reproducible, accurate} and \em{representative} tests.
-Each aspect will be detailed in the sections below
+\import{\currfiledir}{draft}
 
-\import{chapters/testing/}{reproducibility}
-\import{chapters/testing/}{accuracy}
-\import{chapters/testing/}{representativeness}
+\import{\currfiledir}{reproducibility}
+\import{\currfiledir}{experiment}
+
+\import{\currfiledir}{conclusion}
+
+% However for them to be more representatives we wanted to simulate the production enviromenet.
+%% should i add  the thing about green faas ? 
+%% same about what we have done with python workers and lazy ? basically just a graph

### 2022-06-11 chapters/benchmarking/draft.tex
refactoring : stateoftheart

@@
+\section{Introduction}
 To optimize the software energy consumption, a reproducible setup environment requires to be designed.
@@
 \section{Threats and Challenges}
+
 A successful benchmark has three criterions to fullfil.
@@
 
-%TODO:  rephrase this figure, maybe change it a little bit 
 Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
@@
 %% YEP I KNOW, THIS IS JUST FOR ME TO READ IT LATER AND GIVE A BETTER CONTEXT
-
-%  TODO: add the environement part,
-%  TODO: add perspective
-
-\section{Experimental Guidelines}\label{sec:guidelines}
-To summarize our experiments, we provide some experimental guidelines in Table~\ref{table:guidelines}, based on the multiple experiments and analysis we did.
-These guidelines constitute a set of minimal requirements or best practices, depending on the workload and the criticality of the energy measurement precision.
-It therefore intends to help practitioners in taming the energy variation on the selected CPU, and conduct the experiments with the least variations.
-
-\begin{table}[h!]
-    \centering
-    \caption{Experimental Guidelines for Energy Variations}
-    \small
-    \begin{tabular}{|p{4.7cm}|c|c|}
-        \hline
-        \textbf{Guideline}                                                                                                                                                & \textbf{Load} & \textbf{Gain}     \\
-        \hline
-        \hline
-        Use a low TDP CPU                                                                                                                                                 & Low \& medium & Up to $3\times$   \\
-        \hline
-        Disable the CPU C-states                                                                                                                                          & Low           & Up to $6\times$   \\
-        \hline
-        Use the least of sockets in a case of multiple CPU                                                                                                                & Medium        & Up to $30\times$  \\
-        \hline
-        Avoid the usage of hyper-threading whenever possible                                                                                                              & Medium        & Up to $5\times$   \\
-        \hline
-        Avoid rebooting the machine between tests                                                                                                                         & High          & Up to $1.5\times$ \\
-        \hline
-        Do not relate to the machine idle variation to isolate a test EC, the CPU/OS changes its behavior when a test is running and can exhibit less variation than idle & Any           & ---               \\
-        \hline
-        Rather focus the optimization efforts on the system under test than the OS                                                                                        & Any           & ---               \\
-        \hline
-        Execute all the similar and comparable experiments on a same machine.
-Identical machines can exhibit many differences regarding their energy behavior             & Any           & Up to $1.3\times$ \\
-        \hline
-    \end{tabular}
-    \label{table:guidelines}
-\end{table}
-
-Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and its variation reduction at low workloads.
-However, it also lists some new factors that we identified along the analysis we conducted in Section, such as the results related to the OS or the reboot mode.
-Some of the guidelines are more useful/efficient for specific workloads, as showed in our experiments.
-Thus, qualifying the workload before conducting the experiments can help in choosing the proper guidelines to apply.
-Other studied factors are not been mentioned in the guidelines, like the Turbo~Boost or the Speculative execution, due to the small effect that has been observed in our study.
-
-In order to validate the accuracy of our guidelines among a varied set of benchmarks on one hand, and their effect on the variation between identical machines on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
-Half of these experiments has been performed at a 50\,\% workload and the other half on single process jobs.
-The choice of these two workloads is related to the optimization guidelines that are mainly effective at low and medium workloads.
-We note that we used the cluster \textsf{Dahu} over \textsf{Ecotype} to highlight the guidelines effect on the nodes where the variation is susceptible to be higher.
-
-\begin{figure*}%[!htb]
-    \center{\includegraphics[width=\linewidth]{imgs/all}}
-    \caption{Energy variation comparison with/without applying our guidelines}\label{fig:optimized}
-\end{figure*}
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/stressng}}
-    \caption{Energy variation comparison with/without applying our guidelines for \textsc{Stress-NG}}\label{fig:stress}
-\end{figure}
-
-Figure~\ref{fig:optimized} and~\ref{fig:stress} highlight the improvement brought by the adoption of our guidelines.
-They demonstrate the intra-node STD reduction at low and medium workloads for all the benchmarks used at different levels.
-Concretely, for low workloads, the energy variation is 2--6 times lower after applying the optimization guidelines for the benchmarks \textsf{LU} and \textsf{EP}, as well as \textsc{Linpack}, while it is 1.2--1.8 times better for \texttt{Sha256}.
-For this workload, the overall energy consumption after optimization can be up to 80\,\% higher due to disabling the C-states to keep all the unused cores at a high power consumption state ($Mean_{LU-normal-Dahu2}=11,500 mJ$, $Mean_{LU-optimized-Dahu2}=20,508 mJ$).
-For medium workloads, the STD, and thus variation, is up to 100\,\% better for the benchmark \textsf{CG}, 20--150\,\% better for the \texttt{pbzip2} application and up to 100\% for \textsc{Stress-NG}.
-We note that the optimized version consumes less energy thanks to an appropriate core pinning method.
-
-Figures~\ref{fig:optimized} and~\ref{fig:stress} also highlight that applying the guidelines does not reduce the inter-nodes variation in all the cases.
-This variation can be up to 30\,\% in modern CPU~\cite{wang_experimental_nodate}.
-However, taming the intra-node variation is a good strategy to identify more relevant mediums and medians, and then perform accurate comparisons between the nodes variation.
-Even though, using the same node is always better, to avoid the extra inter-nodes variation and thus improve the stability of measurements.
-
-\section{Threats to Validity}\label{sec:threats}
-A number of issues affect the validity of our work.
-For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPU, but still adds an important overhead if we adopt a sampling at high frequency.
-The other fine-grained tool we used for measurements is \textsc{PowerAPI}.
-It allows to measure the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
-The usage of \textsc{PowerAPI} adds an error margin because of the power model built over RAPL.
-The RAPL tool mainly measures the CPU and DRAM energy consumption.
-However, even running CPU/RAM intensive benchmarks would keep a degree on uncertainty concerning the hard disk and networking energy consumption.
-In addition, the operating system adds a layer of confusion and uncertainty.
-
-The Intel CPU chip manufacturing process and the materials micro-heterogeneity is one of the biggest issues, as we cannot track or justify some of the energy variation between identical CPU or cores.
-These CPU/cores might handle frequencies and temperature differently and behave consequently.
-This hardware heterogeneity also makes reproduction complex and requires the usage of the same nodes on the cluster with the same OS.
+%% sorry caps ^^"
 
@@
 % However, using and comparing identical CPU is still tricky and is very dependent to chips.
-
-\section{Conclusion}\label{sec:conclusion}
-In this paper, we conducted an empirical study of controllable factors that can increase the energy variations on platforms with some of the latest CPU, and for several workloads.
-We provide a set of guidelines that can be implemented and tuned (through the OS GRUB for example), especially with the new data centers isolation trend and the cloud usage, even for scientific and R\&D purposes.
-Our guidelines aim at helping the user in reducing the CPU energy variation during systems benchmarking, and conduct more stable experiments when the variation is critical.
-For example, when comparing the energy consumption of two versions of an algorithm or a software system, where the difference can be tight and need to be measured accurately.
-
-Overall, our results are not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
-The aim of our work is to be able to tame and mitigate this variability along controlled experiments.
-We studied some previously discussed aspects on some recent CPU, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to achieve the variability mitigating purpose.
-Some of these factors, like the C-states usage, can reduce the energy variation up to 500\,\% at low workloads, while choosing the wrong cores/PU strategy can cause up to $30\times$ more variability.
-
-We believe that our approach can also be used to study/discover other potential variability factors, and extend our results to alternative CPU generations/brands.
-Most importantly, this should motivate future works on creating a better knowledge on the variability due to CPU manufacturing process and other factors.
-
-\section{Perspectives}
-By the end of this study we have gathered enought guidelines to make the tests more reproducible, accurate.
-We created a set of new tests named \textbf{energy tests} which are more similar to performance tests.
-Thanks to the work of two interns [ mamadou and adrien] we created a CI/CD plateform to measure the energy consumption of Java projects and we could track the evolution of the this energy accross different stages of the project.
-In the figure below we see an example of this plugin.
-For more details please visit the gitlab repository ... add link.
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/JunitSonarplugin}}
-    \caption{Example of the Junit Sonar Plugin}\label{fig:sonar_plugin}
-\end{figure}

### 2022-06-11 chapters/benchmarking/experiment.tex
refactoring : continuation
refactoring : stateoftheart

--- 2022-06-10 chapters/benchmarking/experement.tex
@@
 %% This is a conclusion maaan  !! 
-\subsection{Goal}
+\section{Taming the energy variation for the benchmarking protocol}
+
+\subsection{Objective}
+While the previous part aims to create a large umbrella that covers all empirical benchmarks related to computer science, we rather aim to narrow our study to energy benchmarking only.
+In other words, we will focus on pitfalls that hinders the energy claims of software.
+We are clearly aware of the impact of hardware on energy variations.
+However, we believe that there is still a room to control this energy variation for practitioners by using only parameters that can be tuned.
+To do so, we have inducted a set of empirical experiments using the guidelines provided by state of the art, to determine which controllable factors can reduce the variation of the energy consumption of benchmarks.
+% We will first start with evaluating the parameters mentioned by \cite{heinrich_predicting}.
+
 In this part we will try to answer the following Research questions.
@@
 
-\subsection{Workload}
+\subsubsection{Workload}
 Our choice for the benchmarks was based on two creterions.
@@
 
-\subsection{Metrics \& Measurement Tools}
+\subsubsection{Metrics \& Measurement Tools}
 Our metric for the accuracy of the test is the Standard deviation aka \textbf{STD} of the energy consumption. Therefore wether the tests consumes more or less energy is out of our scope.
@@
 
-\subsection{\textsc{RQ}~1: Benchmarking Protocol}
+\subsubsection{\textsc{RQ}~1: Benchmarking Protocol}
 To achieve a robust and reproducible experiment, practitioners often tend to repeat their tests multiple times, in order to analyze the related performance indicators, such as execution time, memory consumption or energy consumption.
@@
 
-\subsection{\textsc{RQ}~2: Processor Features}
+\subsubsection{\textsc{RQ}~2: Processor Features}
 The C-states provide the ability to switch the CPU between more or less consuming states upon activities.
@@
 
-\subsubsection{Cores Pinning}
+\paragraph{Cores Pinning}
 To answer the first question, we repeated the previous test at 50\,\% workload.
@@
 
-\subsubsection{Processes Threshold}
+\paragraph{Processes Threshold}
 To answer the second question regarding the evolution of the energy variation at different levels of CPU usage, we varied the used PU's count to track the EV evolution.
@@
 
-\subsubsection{Turbo Boost}
+\paragraph{Turbo Boost}
 The Turbo Boost---also known as \emph{Dynamic Overclocking}---is a feature that has been incorporated in Intel CPU since the Sandy~Bridge micro-architecture, and is now widely available on all of the Core\,i5, Core\,i7, Core\,i9 and Xeon series.
@@
 
-\subsection{\textsc{RQ}~3: Operating System}
+\subsubsection{\textsc{RQ}~3: Operating System}
 The \emph{operating system} (OS) is the layer that exploits the hardware capabilities efficiently.
@@
 
-\subsubsection{OS Tuning}
+\paragraph{OS Tuning}
 An OS is a pack of running processes and services that might or not be required its execution.
@@
 
-\subsubsection{Speculative Executions}
+\paragraph{Speculative Executions}
 Meltdown and Spectre are two of the most famous hardware vulnerabilities discovered in 2018, and exploiting them allows a malicious process to access others processes data that is supposed to be private~\cite{Kocher2018spectre,Lipp2018meltdown}.
@@
 
-\subsection{\textsc{RQ}~4: Processor Generation}
+\subsubsection{\textsc{RQ}~4: Processor Generation}
 Intel microprocessors have noticeably evolved during these last 20 years.
@@
 \end{mdframed}
+
+\subsection{Experimental Guidelines}
+To summarize our experiments, we provide some experimental guidelines in Table~\ref{table:guidelines}, based on the multiple experiments and analysis we did.
+These guidelines constitute a set of minimal requirements or best practices, depending on the workload and the criticality of the energy measurement precision.
+It therefore intends to help practitioners in taming the energy variation on the selected CPU, and conduct the experiments with the least variations.
+
+\begin{table}[h!]
+    \centering
+    \caption{Experimental Guidelines for Energy Variations}
+    \small
+    \begin{tabular}{|p{4.7cm}|c|c|}
+        \hline
+        \textbf{Guideline}                                                                                                                                                & \textbf{Load} & \textbf{Gain}     \\
+        \hline
+        \hline
+        Use a low TDP CPU                                                                                                                                                 & Low \& medium & Up to $3\times$   \\
+        \hline
+        Disable the CPU C-states                                                                                                                                          & Low           & Up to $6\times$   \\
+        \hline
+        Use the least of sockets in a case of multiple CPU                                                                                                                & Medium        & Up to $30\times$  \\
+        \hline
+        Avoid the usage of hyper-threading whenever possible                                                                                                              & Medium        & Up to $5\times$   \\
+        \hline
+        Avoid rebooting the machine between tests                                                                                                                         & High          & Up to $1.5\times$ \\
+        \hline
+        Do not relate to the machine idle variation to isolate a test EC, the CPU/OS changes its behavior when a test is running and can exhibit less variation than idle & Any           & ---               \\
+        \hline
+        Rather focus the optimization efforts on the system under test than the OS                                                                                        & Any           & ---               \\
+        \hline
+        Execute all the similar and comparable experiments on a same machine.
+Identical machines can exhibit many differences regarding their energy behavior             & Any           & Up to $1.3\times$ \\
+        \hline
+    \end{tabular}
+    \label{table:guidelines}
+\end{table}
+
+Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and its variation reduction at low workloads.
+However, it also lists some new factors that we identified along the analysis we conducted in Section, such as the results related to the OS or the reboot mode.
+Some of the guidelines are more useful/efficient for specific workloads, as showed in our experiments.
+Thus, qualifying the workload before conducting the experiments can help in choosing the proper guidelines to apply.
+Other studied factors are not been mentioned in the guidelines, like the Turbo~Boost or the Speculative execution, due to the small effect that has been observed in our study.
+
+In order to validate the accuracy of our guidelines among a varied set of benchmarks on one hand, and their effect on the variation between identical machines on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
+Half of these experiments has been performed at a 50\,\% workload and the other half on single process jobs.
+The choice of these two workloads is related to the optimization guidelines that are mainly effective at low and medium workloads.
+We note that we used the cluster \textsf{Dahu} over \textsf{Ecotype} to highlight the guidelines effect on the nodes where the variation is susceptible to be higher.
+
+\begin{figure*}%[!htb]
+    \center{\includegraphics[width=\linewidth]{imgs/all}}
+    \caption{Energy variation comparison with/without applying our guidelines}\label{fig:optimized}
+\end{figure*}
+
+\begin{figure}%[!htb]
+    \center{\includegraphics[width=.9\linewidth]{imgs/stressng}}
+    \caption{Energy variation comparison with/without applying our guidelines for \textsc{Stress-NG}}\label{fig:stress}
+\end{figure}
+
+Figure~\ref{fig:optimized} and~\ref{fig:stress} highlight the improvement brought by the adoption of our guidelines.
+They demonstrate the intra-node STD reduction at low and medium workloads for all the benchmarks used at different levels.
+Concretely, for low workloads, the energy variation is 2--6 times lower after applying the optimization guidelines for the benchmarks \textsf{LU} and \textsf{EP}, as well as \textsc{Linpack}, while it is 1.2--1.8 times better for \texttt{Sha256}.
+For this workload, the overall energy consumption after optimization can be up to 80\,\% higher due to disabling the C-states to keep all the unused cores at a high power consumption state ($Mean_{LU-normal-Dahu2}=11,500 mJ$, $Mean_{LU-optimized-Dahu2}=20,508 mJ$).
+For medium workloads, the STD, and thus variation, is up to 100\,\% better for the benchmark \textsf{CG}, 20--150\,\% better for the \texttt{pbzip2} application and up to 100\% for \textsc{Stress-NG}.
+We note that the optimized version consumes less energy thanks to an appropriate core pinning method.
+
+Figures~\ref{fig:optimized} and~\ref{fig:stress} also highlight that applying the guidelines does not reduce the inter-nodes variation in all the cases.
+This variation can be up to 30\,\% in modern CPU~\cite{wang_experimental_nodate}.
+However, taming the intra-node variation is a good strategy to identify more relevant mediums and medians, and then perform accurate comparisons between the nodes variation.
+Even though, using the same node is always better, to avoid the extra inter-nodes variation and thus improve the stability of measurements.
+
+\subsection{Threats to Validity}
+A number of issues affect the validity of our work.
+For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPU, but still adds an important overhead if we adopt a sampling at high frequency.
+The other fine-grained tool we used for measurements is \textsc{PowerAPI}.
+It allows to measure the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
+The usage of \textsc{PowerAPI} adds an error margin because of the power model built over RAPL.
+The RAPL tool mainly measures the CPU and DRAM energy consumption.
+However, even running CPU/RAM intensive benchmarks would keep a degree on uncertainty concerning the hard disk and networking energy consumption.
+In addition, the operating system adds a layer of confusion and uncertainty.
+
+The Intel CPU chip manufacturing process and the materials micro-heterogeneity is one of the biggest issues, as we cannot track or justify some of the energy variation between identical CPU or cores.
+These CPU/cores might handle frequencies and temperature differently and behave consequently.
+This hardware heterogeneity also makes reproduction complex and requires the usage of the same nodes on the cluster with the same OS.
+
+\subsection{Conclusion}
+In this part, we conducted an empirical study of controllable factors that can increase the energy variations on platforms with some of the latest CPU, and for several workloads.
+We provide a set of guidelines that can be implemented and tuned (through the OS GRUB for example), especially with the new data centers isolation trend and the cloud usage, even for scientific and R\&D purposes.
+Our guidelines aim at helping the user in reducing the CPU energy variation during systems benchmarking, and conduct more stable experiments when the variation is critical.
+For example, when comparing the energy consumption of two versions of an algorithm or a software system, where the difference can be tight and need to be measured accurately.
+
+Overall, our results are not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
+The aim of our work is to be able to tame and mitigate this variability along controlled experiments.
+We studied some previously discussed aspects on some recent CPU, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to achieve the variability mitigating purpose.
+Some of these factors, like the C-states usage, can reduce the energy variation up to 500\,\% at low workloads, while choosing the wrong cores/PU strategy can cause up to $30\times$ more variability.
+
+We believe that our approach can also be used to study/discover other potential variability factors, and extend our results to alternative CPU generations/brands.
+Most importantly, this should motivate future works on creating a better knowledge on the variability due to CPU manufacturing process and other factors.

### 2022-06-11 chapters/benchmarking/reproducibility.tex
refactoring : stateoftheart

--- 2021-10-21 chapters/testing/stateoftheart.tex
@@
 
-\paragraph{Studying Hardware Factors.}
+%  basically just rephrase the paper
 
-This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
-On the one hand, the correlation between the processor temperature and the energy consumption was one of the most explored paths.
-Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
-On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the correlation between the CPU temperature and the energy consumption variation is very tight.
+%% add this to my contribution 
+\subsection{docker and accuracy}
+And now Since the stat of the art has aggreed on the impact of docker on the energy consumption,Let's discuss it's impact on the accuracy. In other words\\
+\textbf{RQ :} does Docker affect the energy variation of the exepements ?
 
-\note{add the corelation value}
+To Answer this question we have conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
+Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
+In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
+The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
 
-This makes the processor temperature a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.
-
-The ambient temperature was also discussed in many papers as a candidate factor for the energy variation of a processor.
-In~\cite{ranka_energy_2009}, the authors claimed that energy consumption may vary due to fluctuations caused by the external environment.
-These fluctuations may alter the processor temperature and its energy consumption.
-However, the temperature inside a data center does not show major variations from one node to another.
-In~\cite{el_mehdi_diouri_your_2013}, El~Mehdi~Dirouri~\emph{et~al.} showed that switching the spot of two servers does not affect their energy consumption.
-Moreover, changing hardware components, such as the hard drive, the memory or even the power supply, does not affect the energy variation of a node, making it mainly related to the processor.
-This result was recently assessed by~\cite{wang_potential_2018}, where the rack placement and power supply introduced a maximum of $2.8\,\%$ variation in the observed energy consumption.
-
-Beyond hardware components, the accuracy of power meters has also been questioned.
-Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
-All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
-%% this is to prove that the problem is REAL !!! 
-\paragraph{Mitigating Energy Variations.}
-Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
-In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
-The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
-Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
-They also proposed some guidelines to reduce this variation by replacing the old slow chips, by load balancing the workload on the CPU cores and leaving one core idle.
-They claimed that the variation between the processor cores is insignificant.
-In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.
-
-In~\cite{marathe_empirical_2017_m}, the authors highlight the increase of energy variation across the latest Intel micro-architectures by a factor of $4$ from Sandy~Bridge to Broadwell, a $15\,\%$ of run-to-run variation within the same processor and the increase of the inter-cores variation from $2.5\,\%$ to $5\,\%$ due to hardware-enforced constraints, concluding with some recommendations for Broadwell usage, such as running one hyper-thread per core.
-%  NOTE :/this might be interesting for greeFaaS
-
-\subsection*{Our position}
-Meanwhile the previoius work was to create a large umbrella tht englobes all emprical tests related to computer science, we want to narrow our study to energy testing only.
-In other words we will study pitfalls that hinders the energy claimes of software .
-
-We are fully aware of the impact that hardware has on energy variation.
-However, we believe that there is still a room to reduce this energy variation for practicioners using only paramenters that they are in cotrol.
-To do so, we have inducted a set of empirical experiments using the guidelines provided by state of the art, to determine which controllable factors can reduce the variation of the energy consumption of tests.
-We will first start with evaluating the paramenters mentioned by \cite{heinrich_predicting}.
+\begin{figure}
+    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
+    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
+\end{figure}

### 2022-06-11 chapters/introduction.tex
refactoring : stateoftheart

@@
 Researchers are trying to solve this issue through different angles.
-While some scientists are trying to find an alternative green source to generate energy.
-others focus on reducing it's consumption.
-In computer science we are concerned with the ladder solution.
-Therefore most of the works are done on tunning the software and the hardware in ordrer to have a more efficient way use this energy.
-
-In this thesis we are trying to find a way to make the energy consumption of the computer to be as low as possible.
-Our approach is is to reduce the energy consumption of the software services by changing certains parameters, such as palteform, programming langauge, and/or the design pattern.
+While some scientists are trying to find an alternative green source to generate energy. others focus on reducing it's%whose?
+consumption. In computer science we are concerned with the ladder solution %reference.
+Therefore most of the works are done on tunning the software and the hardware in ordrer to have a more efficient way use this energy. %more efficient way of producing programs from electricity? 
+%say about hardware (i.e. superconductive materials with zero resistance)
+In this thesis we are trying to find a way to make the energy consumption of the computer to be as low as possible. %on the software level
+%More precisely, the purpose of this thesis is to highlight (attract attention to?) the impact of programers' choices on energy comsumption of software while running the code and to advise them with the strategies to write the greenest possible code.
+Our approach is to reduce the energy consumption of the software services by changing certains parameters, such as palteform, programming langauge, and/or the design pattern.
 

### 2022-06-11 chapters/literature/draft_citations.tex
refactoring : stateoftheart

@@
 %%%%%%%%%%%%%%%%%
-
+% TODO : REPHRASE THIS other wise it will be plagiarism
 list of benchmarks \cite{van_der_kouwe_benchmarking_2018}:
@@
 
+defintion \cite{wang1993grpc}
+
+comparaison \cite{chamas2017comparing}
+
+emprical study \cite{de2021empirical}
+
+impact of website implemenation on energy consumption \cite{philippot_characterization_2014} \cite{manotas2013investigating}
+
+PHP \cite{benmoussa_new_2019} \cite{das_comparison_2016}
+
+simulation of web cache \cite{cardenas_performance_2005}
+
+java spring analysis \cite{gajewski_analysis_2019}
+
+performance \cite{mishra2021web}
+
+cross compiler benchmarking \cite{yet2016cross}
+
+\subsection{python}
+
+webframeworks on python \cite{pankiv_concurrent_nodate}
+
+hope library \cite{akeret_hope_2015}
+
+bytecode effect \cite{ben_asher_effect_2009}
+
+numba \cite{crist_dask_2016}
+
+intel python \cite{li_boosting_2016}
+
+python 2 vs 3 \cite{modzelewski_pyston_2020}
+
+python runtime performances \cite{redondo_comprehensive_2015} \cite{murri_performance_2013}
+
+static vs dynamic languages \cite{pang_what_nodate}
+
+concurent benchmark framework \cite{pankiv2019concurrent}
+
 \subsection{jvm}
@@
 \subsection{benchmarking}
+%%%%%%%%%%% further references 
+NAS \cite{bailey_nas_nodate}
+statistics \cite{he_statistics-based_2019}
+benchmark crimes \cite{van_der_kouwe_benchmarking_2018}
+microservices \cite{grambow_benchmarking_2020}
+impact of webservers on web applications energy \cite{manotas_investigating_2013}
+%%%%%%%%%%%%%%%%%  virtualisation
 
-NAS \cite{bailey_nas_nodate}
+\section{Reproducibility in benchmarking }
 
-statistics \cite{he_statistics-based_2019}
+%% just a placeholder title for the moment 
+% introduction of vms
 
-benchmark crimes \cite{van_der_kouwe_benchmarking_2018}
+\subsection{Virtual machines}
+To resolve this problem, practitioners tend to use Virtualisation.
+Using virtual machines aka VM gives reaserchers the freedom to choose their own tools, software and operating system that they are the most confortable with without paying the price to change the actual working environment, which will give them eventually more controle over the dependencies and the execution environement.
+Moreover, using a vm will solve the \emph{replication crisis} thanks to the virtual images, even the most complex architecture can be reproduced easily by just instantiating a copy of the image.
+Since the virtual machines are agnostic to the host architecture, reaserchers won't have to worry about where and how their experiments are replicated because they have already setup the execution environement.
+Another advantage to the virtual machines is the snapshot mechanism, it allows reaserchers to create backups and revert some changes with simple clicks.
+Last but not least,thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables -controled and uncontroled-  and do other analysis without dealing with any dependencies.
+In his paper \cite{howe_virtual_2012} bill howe lists the advantages of using virtual machines in reaserchers experements including the economical impact and cultural limitation to a such approach.
 
-microservices \cite{grambow_benchmarking_2020}
+which allow them to have control over the ressouces, the dependencies and the execution environment.
+Moreover, thanks to the snapshots, deploying a software is easily done by instantiating a copy of that image.
+However, this choice comes with a certain cost.
+Because the intervention of the hypervisor,the software will use two kernels , the virtual machine one and the host machine one, which will provide a noticiable overhead, and will impact the performances of the tests.Therefore, we can't use virtual machines for exepements that are related to performance.
+Another limiation with the virtual machine is the the isolation.
+It is true that this feature will prevent the experience environement with any undesirable interference from the outside world.
+but sometimes this contact is needed, especially when the experiment is dependent to an external part, such as sensors.
+In energicial tests we tend to use hardware powermeters which will make it difficult to use the virtual machines in this case .
 
-impact of webservers on web applications energy \cite{manotas_investigating_2013}
+% introduction of docker
 
-\subsection{frameworks}
+\begin{figure}
+    \center{\includegraphics[width=1\linewidth]{images/virtualization_techniques}}
+    \caption{Different Methods of Virtualzation}\label{environement:virtualization_technique}
+\end{figure}
 
-impact of website implemenation on energy consumption \cite{philippot_characterization_2014} \cite{manotas2013investigating}
+\subsection{Containers}
 
-PHP \cite{benmoussa_new_2019} \cite{das_comparison_2016}
+another solution would be using something that allows us to have the isolation from the host os and the ease of replication that virtual machines offer, and the direct interaction with the hardware that the classical method give.
+contarization offers a such advantages while keeping the isolation and the ease of replication for application.
 
-simulation of web cache \cite{cardenas_performance_2005}
+Figure \ref{environement:virtualization_technique} explains the differents in architecture between the clasic types2 of Virtualisation and Containers.
+\begin{itemize}
+    \item Type 1: runs directly on the hardware, it is mainly used by the cloud providers where there is no main OS, but just virtaul machines, we can site for this the open-srouce XEN and VMware ESX
+    \item Type 2: runs over the hostmachine Operating System, mostly used for personal comptuers, VMware server and virtualBox are famous examples of this type, most of the reaserchers expermentation are run witht this type, howerver due to the 2 Operatings syestms the applications tend to be more slower
+    \item containers : Instead of its own kernel, containers used the hots kernel to run their Os, which makes them ligher, quickers and use the full pontial use of the hardware.
+For this we can cite \emph{Docker}, \emph{Linux LXC}\emph{LXD} \cite{abuabdo_virtualization_2019}
+\end{itemize}
+%  TODO Rephrase so i include it before VMs+ add links to the technologie
 
-java spring analysis \cite{gajewski_analysis_2019}
+\subsection{Docker vs Virtual Machine}
+Depsite that Type 1 is more performant than type 2, the second one is the most used in reaserch, since most researchers conducts their experements in their own machine.
+In the other hand, docker is the most famous thechonology for for containers.
+In ower case we are more prone to docker for two reasons.
+\begin{enumerate}
+    \item we need a litetweight orchestrator to not affect the energy consumption of ower tests. As prior work mentioned [cite Morabito (2015) and van Kessel et al. (2016)]
+          % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
+    \item since we are using the hardware itself to measure the energy consumption, we are required to interract with the host OS itself.
+\end{enumerate}
+Special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that allows us to retireve the energy consumption of a virtual machine.
+%%%%% THINGS TO ADD : 
+%% -------------- Dont think so 
+% the impact of docker on the reprodudicibility 
+% Limits of docker 
+% The docker files and the clarification of the methodology 
+\subsection{Docker and energy}
+Now that we have chosen to go with the containers technorolgy to encapsulate our tests. What would be the impact of this solution on the energy consumption of our tests.
 
-performance \cite{mishra2021web}
+Based on the studies of \cite{santos2018does}, who analysed the impact of adding the docker layer on the energy consumption.
+In their experement. Eddie Antonio et al run multiple benchmarks with and without docker. and compared their energy consumption and execution time.
+The first step was to see the impact of docker deamon while there is no work. to see the impact of the orchestrator alone. Later they had the experiment with the following benchmarks
+\begin{itemize}
+    \item wordpress
+    \item reddis
+    \item postfresSQL
+\end{itemize}
+The following figures represents the energy consumption of the system while it is idle. As we can see  in figure \ref{fig:docker_idle} Docker braught arround 1000joules overhead.
+\begin{figure}
+    \center{\includegraphics[width=.5\linewidth]{images/docker_vs_vm_energy_paper/idle_energy}}
+    \caption{energy consumption of Idle system with and without docker \cite{santos2018does}}\label{fig:docker_idle}
+\end{figure}
+In the other hand as we can see in figure \ref{fig:docker_reddis}.
+docker increased the execution time of the benchmark by 50 seconds which caused an increase in energy since they are highly correlated.
+The authors also highlated the fact that this increase of energy consumption is due to the docker deamon and not the fact that the application is in a container.
+Moreover they estimted the price of this extra energy and it was less than 0.15\$ in the worst case.
+Which is non significant compared to the advantages that docker bring for isolation and reproducibility.\\
+if we  recap this study in one sentence,it would be the following one.
+The dockerised softwares tend to consume more energy, because mainly they take more time to be executed.
+The average power consumption is higher with only \textbf{2Watts} and it is due to the docker deamon.This overhead can be up to 5\% for IO intensive application, but it is mearly noticiable when it comes to CPU or DRMA intesive works
 
-cross compiler bencharking \cite{yet2016cross}
+\begin{figure}
+    \includegraphics[width=.5\linewidth]{images/docker_vs_vm_energy_paper/reddis_time}
+    \includegraphics[width=.5\linewidth]{images/docker_vs_vm_energy_paper/reddis_energy}
+    \caption{execution time and energy consumption of Redis  with and without docker \cite{santos2018does}}\label{fig:docker_reddis}
+\end{figure}
 
-\subsection{python}
+%%%%%%%%%%%%% energy metigation
 
-webframeworks on python \cite{pankiv_concurrent_nodate}
+\section{Energy Variation}
+\subsection{Studying Hardware Factors}
+This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
+On the one hand, the correlation between the processor temperature and the energy consumption was one of the most explored paths.
+Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
+On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the CPU temperature and the energy consumption variation are tightly coupled.
 
-hope library \cite{akeret_hope_2015}
+\note{add the corelation value}
 
-bytecode effect \cite{ben_asher_effect_2009}
+This exposes the processor temperature as a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.
 
-numba \cite{crist_dask_2016}
+The ambient temperature was also discussed in many papers as a candidate factor for the energy variation of a processor.
+In~\cite{ranka_energy_2009}, the authors claimed that energy consumption may vary due to fluctuations caused by the external environment.
+These fluctuations may alter the processor temperature and its energy consumption.
+However, the temperature inside a data center does not show major variations from one node to another.
+In~\cite{el_mehdi_diouri_your_2013}, El~Mehdi~Dirouri~\emph{et~al.} showed that switching the spot of two servers does not affect their energy consumption.
+Moreover, changing hardware components, such as the hard drive, the memory or even the power supply, does not affect the energy variation of a node, making it mainly related to the processor.
+This result was recently assessed by~\cite{wang_potential_2018}, where the rack placement and power supply introduced a maximum of $2.8\,\%$ variation in the observed energy consumption.
 
-intel python \cite{li_boosting_2016}
+Beyond hardware components, the accuracy of power meters has also been questioned.
+Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
+All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
+%% this is to prove that the problem is REAL !!!
 
-python 2 vs 3 \cite{modzelewski_pyston_2020}
+\subsection{Mitigating Energy Variations.}
+Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
+In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
+The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
+Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
+They also proposed some guidelines to reduce this variation by replacing the old---slower---chips, by load balancing the workload on the CPU cores and leaving one core idle.
+They claimed that the variation between the processor cores is insignificant.
+In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.
 
-python runtime performances \cite{redondo_comprehensive_2015} \cite{murri_performance_2013}
-
-static vs dynamic languages \cite{pang_what_nodate}
-
-concurent benchmark framework \cite{pankiv2019concurrent}
-
-\subsection{grpc}
-
-defintion \cite{wang1993grpc}
-
-comparaison \cite{chamas2017comparing}
-
-emprical study \cite{de2021empirical}
+In~\cite{marathe_empirical_2017_m}, the authors highlight the increase of energy variation across the latest Intel micro-architectures by a factor of $4$ from Sandy~Bridge to Broadwell, a $15\,\%$ of run-to-run variation within the same processor and the increase of the inter-cores variation from $2.5\,\%$ to $5\,\%$ due to hardware-enforced constraints, concluding with some recommendations for Broadwell usage, such as running one hyper-thread per core.
+%  NOTE :/this might be interesting for greeFaaS
 

### 2022-06-14 chapters/optimization/java-dir/java.tex
add : python

--- 2022-05-18 chapters/execution_environement/java/java.tex
@@
+\section*{JAVA}
 \section{Introduction}
@@
 \subsection{definition of the JVM}
+% TODO : ADD definition here 
 \begin{figure}%[ht]
@@
 
-\section{Experimental Protocol}\label{sec:meth}
+\section{Experimental Protocol}\label{sec:javameth}
 To investigate the effect that could have the JVM distribution choice and/or parameters on software energy consumption, we conducted a wide set of experiments on a cluster of machines and using several established Java benchmarks and JVM configurations.
@@
 
-\subsection{metrics and measurement}
+\subsection{Metrics and measurement}
 Since the goal of this study is the green aspect Java virtual machines, our key metric will be the energy consumption of the job completed by each configuration.
@@
 
-\section{Conclusion}
-The results of our investigations showed that many JVMs share energy efficiencies and can grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
-The 3 selected JVM classes can however report a different energy efficiency for different software and/or workloads, sometimes by a large margin.
-While we did not observed a unique champion when it comes to energy consumption, \textsc{GraalVM} reported the best energy efficiency for a majority of benchmarks.
-Nonetheless, each JVM can achieve the best or the worst depending on the hosted application.
-One cause can be thread management strategies, as observed with J9 when advantageously running \textsf{Avrora}.
-Moreover, some JVM settings can cause energy consumption variations.
-Our experiments showed that the default JIT compiler of the JVM is often near-optimal, in at least ~80\% of our experiments.
-The default GC, however, was outperforming alternative strategies in half of our experiments, with some large gains observed when using some alternative GC depending on the application characteristics.
-
-Our main conclusions and guidelines can be thus summarized as:
-\begin{inparaenum}[\em i)]
-    \item testing software on the 3 classes of JVM and identifying the one that consumes the least is a good practice, especially for multi-threading purposes,
-    \item while the JVM default JIT give often good energy consumption results, some settings may improve the energy consumption and could be tested,
-    \item the choice of the GC may lead to a large impact on the energy consumption in many situations, thus encouraging a careful tuning of this parameter prior to deployment.
-\end{inparaenum}
-To ease the integration of the above guidelines, we propose a tool, named \textsf{J-Referral}, to recommend the most energy-efficient JVM distribution and configuration among more than a hundred considered possibilities.
-It establishes a full report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to help the user to choose the one with the least consumption for a Java Software.
-
-\section{ Tools and contributions }
+\section{Tools and contributions}
 \Cref{table:j-referral} illustrates an example of the final report of \textsf{J-Referral}.
@@
 
+\section{Conclusion}
+The results of our investigations showed that many JVMs share energy efficiencies and can grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
+The 3 selected JVM classes can however report a different energy efficiency for different software and/or workloads, sometimes by a large margin.
+While we did not observed a unique champion when it comes to energy consumption, \textsc{GraalVM} reported the best energy efficiency for a majority of benchmarks.
+Nonetheless, each JVM can achieve the best or the worst depending on the hosted application.
+One cause can be thread management strategies, as observed with J9 when advantageously running \textsf{Avrora}.
+Moreover, some JVM settings can cause energy consumption variations.
+Our experiments showed that the default JIT compiler of the JVM is often near-optimal, in at least ~80\% of our experiments.
+The default GC, however, was outperforming alternative strategies in half of our experiments, with some large gains observed when using some alternative GC depending on the application characteristics.
+
+Our main conclusions and guidelines can be thus summarized as:
+\begin{inparaenum}[\em i)]
+    \item testing software on the 3 classes of JVM and identifying the one that consumes the least is a good practice, especially for multi-threading purposes,
+    \item while the JVM default JIT give often good energy consumption results, some settings may improve the energy consumption and could be tested,
+    \item the choice of the GC may lead to a large impact on the energy consumption in many situations, thus encouraging a careful tuning of this parameter prior to deployment.
+\end{inparaenum}
+To ease the integration of the above guidelines, we propose a tool, named \textsf{J-Referral}, to recommend the most energy-efficient JVM distribution and configuration among more than a hundred considered possibilities.
+It establishes a full report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to help the user to choose the one with the least consumption for a Java Software.
 % TODO : add or change the graphes of the use case 

### 2022-06-14 chapters/optimization/python-dir/python.tex
add : python

--- 2022-06-09 chapters/execution_environement/python/python.tex
@@
-\newpage
-\chapter{python}
-
 \section{Introduction}
@@
 According to their experiments, an interpreted programming language like Python can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. Figure~\ref{fig:hannoi}).
-% However, the operational conditions under which these measurements may threaten the validity of their results.
-
-\begin{figure}[htbp]
-    \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
-    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
-    \label{fig:hannoi}
-\end{figure}
-\subsection{python popularity}
-
-However, Python seems to attract a large community of developers interested in data analysis, web development, system administration, machine learning---according to a survey conducted in 2018 by JetBrains---\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}}one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.
-
-As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
-Rather, we believe that a better option consists in leveraging the strength of this rich ecosystem to promote energy-efficient solutions in order to improve the power consumption of legacy software systems.
-
 In this chapter, we therefore explore the oblivious optimizations that can be applied to Python legacy applications in order to reduce their energy footprint.
@@
 one kind of those optimizations includes alternative interpertersand some libraries that are dedicated to optimize the code without changing its structures such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
+% However, the operational conditions under which these measurements may threaten the validity of their results.
+
+\section{Motivation}
+\subsection{python popularity}
+
+Nowdays, Python seems to attract a large community of developers interested in data analysis, web development, system administration, machine learning---according to a survey conducted in 2018 by JetBrains---\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}}one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.
+
+As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
+Rather, we believe that a better option consists in leveraging the strength of this rich ecosystem to promote energy-efficient solutions in order to improve the power consumption of legacy software systems.
 
@@
 Therefore, a larger community and a wide range of support and libraries that allowed it to touch several fields like data analyses, machine learning, natural language processing, web development, astrophysics etc.
-% python is not efficient 
-% 1 Energy Efficiency Across Programming Languages: How Do Energy, Time, and Memory Relate?
-%  \begin{figure}[hbt]
-%     \includegraphics[width=\linewidth]{figures/hannoiimplementation.png}
-%     \caption{Energy consumption of the recursive implementation of Tower of Hanoi program in different languages (using a base 10 logarithmic scale) \cite{noureddine_preliminary_2012}}      
-%     \label{fig:hannoi}
-% \end{figure}
+\begin{figure}[htbp]
+    \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
+    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
+    \label{fig:hannoi}
+\end{figure}
 
-However, and according to \cite{pinto_energy_2017,noureddine_preliminary_2012} and [1].
+However, and according to \cite{pinto_energy_2017} and \cite{noureddine_preliminary_2012}.
 Python tends to be one more energy hungry .
@@
 
-In this chapter we will target the following research question :
-\begin{enumerate}
-    \item Do the data structres and controle structures impact the energy consumption ?
-    \item how does the parallelism affect the energy behaviour of python programs ?
-    \item are they any non intrusive methodes to reduce the imapct of the energy consumption of existing python softwares ?
-    \item how can one apply those findings in the real world ?
-\end{enumerate}
-
 \subsection{use cases}
-according to the survey made by Jetbrains company \footnote
-{\fnurl{https://www.jetbrains.com/lp/python-developers-survey-2020}} data analysis and web develoepment are the top two purposes for python\ref{fig:usecase}
-
+to reduce the energy consumption of python we started by targeting the main usage of this programming language, wich is revealed to be data sciences and web developpement.
+Figure~\ref{fig:usecase} illustrate a study made by the jetbrain company on python developers \footnote{\fnurl{https://www.jetbrains.com/lp/python-developers-survey-2020}}.
+In this survey, multiple multiple answers were accepted.
 \begin{figure}[hbt]
@@
 
-\subsubsection*{web developement}
-first we wanted to test the energy consumption of python using on djagon,one of the most popular web frameworks.
-therefore we create a simple website where we try to analyse the cost of a single request using different mechanisms to fetch the data from the database.
-the idea behind this use case is to see wether the choice of the database and the orm ( object relational mapper) impacts the energy consumption and the perfromances of the website.
+\subsection*{Goal}
+To reduce the energy consumption of python programs, we decided to start with studying its behaviour during the most common use cases.
+After this we will targtet the structures of the language itself in the hope to find some generic guidelines for more generic purpose as \citeauthor{hasan_energy_2016-1} did in their paper \cite{hasan_energy_2016-1}.
+After this we will measure the energy consumption of different python implementation for the goal of finding a non intrusive way to enhance this energy efficiency.
 
-we considered using two different databases postgresSQL and SQlite3 that containes the same data, and three different ways to fetch the data.
-\begin{enumerate}
-    \item the naive version: which relies only on the ORM to get the data
-    \item Prefetch : basically we just prefetch the data before we request it
-    \item optimized : and here we optimize our request on the SQL level without passing by the ORM
-\end{enumerate}
+%% should we add Research questions ?
 
-\begin{figure}[!hbt]
+\newpage
+
+\paragraph{Energy measurement}
+
+As we know, the energy of a program is the integrale of its power overtime.
+For us ower case, we used Intel \emph{Running Power Average Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754} to collect the power samples of the running tests,
+
+We used used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, to report Data collected by intel Rapl and send it into another machine that we call computing machine.
+then we calculate the Energy using  the the trapezoidal rule.
+
+\ref{fig:trapezrule}
+\begin{figure}[hbt]
     \centering
-    \includegraphics[width=\linewidth]{imgs/django}
-    \caption{energy behaviour based on multiprocessing}
-    \label{fig:django}
+    \begin{equation}
+        E = \int^a_b P(t)dt \simeq \sum^n_{k=1} \frac{P(t_k-1)+P(t_k)}{2}
+    \end{equation}
+    \caption{}
+    \label{fig:trapezrule}
 \end{figure}
 
-%%%%% have to rephrase this
+figure \ref{fig:powerapi} shows the architecture of our testing model.
 
-As one can see in figure~\ref{fig:django}, different choices of requesting the data have a huge impact on the energy consumption.
-as the naive version can consume up to 10x more energy than the optimized one.
-In the other hand the choice of the data base didnt have huge impact on the total energy despite their different behaviour regarding the execution time and the average power.
-This can be useful to help develeppers make a choice regarding which data base they can use based on the number of the expected requests and the expectation of the performance.
+The reason of separation between data collection and energy calculation is to minimise any interference with the test so our sensor is a light c program running inside a docker container.
 
-Another interesting observation is the impact of the interpreter, as figure~\ref{fig:django} highligh.
-the use of Pypy interpter instead of the default one helped reducing the amount of the energy consumption even when we are dealing with the naive version.
+%% should i include the part with the exceptions ??? naah no need so
 
-\subsubsection*{machine learning}
-the other part is to study the impact of machine learning algorthims
+%% except the parallellism part , which i dont know how to link it here yet 
+\section{Workload} : 
+for this chapter we are consediring 4 seperated, yet related studies . 
+First we are going to see the energy behavriour of python within two study cases.
+Web servers and Machine learning.
+later we will dig deeper into the energy consumption of python basic structures where finally we will conclude with the impact of python interpterer on the energy consumption.
+Due to the lack of of support for most of  non conventional pyton interpeters, our main focus was only on micro benchmarks. 
+Except pypy most of the python implementations don't support extra python libraries, Despite that most of the cases those extra implementations were made to optimize a specific library such as numba with numpy, or intelpython with machine learning algorthms.
 
-%% ADD the work of alex 
-\newpage
-\section{experimetal protocole}
-
-as we have discussed in the previous chapter \ref{chapter:benchmarking}.
-instead of running the tests the idea was to design a system that allows practiouners to reproduce and extends our tests.
-and then we use the same system to run answer some of researchs questions
-%%%% probably ill resue the same things from other chapters
-
-\section{python insights}
-
-The goal of this section is to find some insights where we can optimize the energy consumption without impacting the source code.
-so we started by extending the work of \citeauthor{hasan_energy_2016} and \citeauthor{oliveira_recommending_nodate} to the python environement.
-
-%%%% concurency
-
-Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
-As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
-
-\begin{figure}[hbt]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-    \caption{energy behaviour based on multiprocessing}
-    \label{fig:python_multiprocessing}
-\end{figure}
-
-as we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
-where different sub process have to compete for the CPU Ressources.
-Another finding is when we hit the number of physical cores.
-there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
-basicaly he favorits the hyper threads than the physical core wich will lead to some context switches which cause the slow behaviour, howerver in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the eneryg consumption in this case.
-in the Chapter %% reference the work of intern from lyon  
-we discuss a deeper this behaviour of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
-
-another field of investigation is the type data and structure conroles. that might impact the energy consumption.
-To do so. We iterate over a list using three methodes.
-first the classical for ( for i in range (len(n))).
-howerver as we can see here unlike other porgramming langauges it requires extra operations such as determining the length of the collection , and then using the iterator range.
-so we tried the more adapted version
-( for element in collection).
-Morever in most programming languages the for loop is translated to a while loop ( transfirmation from D type to B type -- asm -- ), therefore we wanted to compare this with a ( while ) version.
-After determining the main ways to iterate over a loop we run the algorithms  collections of different data types to see wether it will impact of the energy consumption of the code, same thing for the size of the collection.
-
-As we can see in figure~\ref{fig:pythonloops} the type of the data hasn't an impact on the energy consumption.In the other hand the way we iterate over the collection has a huge factor.
-interestely, the for in range loop was by far the optimal one with following by the regural for in collection, and the while part was the least one with an overhead of 400\% compared to the first option.
-
-%%% putting the idea, later i ll reformulate it 
-the reason behind a such behaviour is mainly related on how python interpeter is made.
-% add refenece https://www.python.org/doc/essays/list2str/ 
-%% https://www.pythonpool.com/for-vs-while-loop-python/  
-% https://wiki.python.org/moin/TimeComplexity
-%% http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1
-In order to reduce the latency of the python code.
-most of the buitl-in functions an operations are written in C, same thing goes for the function range, further more the function len has a complexity of O(1) because it is based on the function Py\_SIZE  of C which stores the length in a field for the object.
-
-so basically for the for in range is basically creating a new iterator that has the same length of the first one.
-and for each iteration we will have to do a second acceess ( l[i]) instead of one one access hence the double time.
-
-for the wile is even slower due to the implicite incrementation of the variable which will cause an extra operation during the loop .
-
-to confirm the hyphothsis.
-we tried to construct a new list by editing the elements of the previous one.
-And as predicted , the builtin methodes are the best energy saving, while the customize while loop is the heaviest.
-an interesting finding is.
-the impact of annonymous functions ( lambda expressions) on the energy consumption.
-the reseason behind this is the fact that python treats those functions as local variables unlike the predifined ones which are global in our case.
-therefore they are faster and consumes less energy  % same reference https://www.python.org/doc/essays/list2str/
-
-\begin{figure}[htb]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/python_iterations}
-    \caption{energy behaviour of different python loops }
-    \label{fig:pythonloops}
-\end{figure}
-
-\begin{figure}[htb]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/python_treatemens}
-    \caption{energy behaviour of different methodes to change a list}
-    \label{fig:pythontreatement}
-\end{figure}
-
-%% should i include the part with the exceptions ???
-\subsection{conclusion}
-This study has shown us, that the optimial way to reduce the energy consumption of the python code is to follow the guidelines and use the builtin functions, which is kinda happy news for the developers since they don't have to make extra effort to make their code green
-%% except the parallellism part , which i dont know how to link it here yet
-
-\newpage
-\section{Python Runtimes}
+\paragraph{preliminary studies}
+for the first studies we used only the official version of python. because the goal was mainly to highlight the impact of the structure of code on the energy consumption.  
 One main inconvienent of the previous methode to reduce the energy consumption, is the hard work that should be done in order to alter the existing code base in order to redcue the energy consumption.
@@
 
-\subsection{tests preparation}
+\section{experimetal protocole}
 
-To study the behaviour of the python implementation regarding the energy consumption, we have to focus on the effect of the implementation and mitigate the maximum any side effects such as the organisation of the code or any extra consumption due to the operating system or tier libraries.
-Therefore, for each test we took the implementation written in python2 as a reference and tried to use it in other implementations as it is.
-If it is not supported by python3, we transformed the code using the officiel library  \emph{2to3}\footnote{\url{https://docs.python.org/3.7/library/2to3.html}}.
-In the case of the libraries that use \emph{JIT} adding a decorator to the function that we want to optimize was enough, if there are other changes we assume that they alter the original code which is against our purpose.
+As we have discussed in the previous chapter \ref{chapter:benchmarking}.
+instead of running the tests, the idea was to design a system that allows practiouners to reproduce and extends our tests.
+and then we use the same system to run answer some of researchs question.
+%%%% probably ill resue the same things from other chapters 
+\subsection{measurement context}
+\paragraph{Hardware settings}
 
-Each test is implemented in a Docker container for the several reasons
-\begin{itemize}
-    \item Isolation: each container has only the test program implemented with a single python runtime to remove any interference between different implementations.
-    \item Deplpoyment: to use the testing machine without extra configurations that may alter the behaviour of the os toward the enrgy consumtion
-    \item Reproductibility: One of the most frequent benchmark crimes~\cite{DBLP:journals/corr/abs-1801-02381} in research is the lack of Reproductibility, by using Docker we ensure that each test has an Image that will be accessible in public.
-\end{itemize}
-
-Despite the presence of the official docker images for the most of the runtimes, we prefered to build our own using the same reference image in order to remove any bias due to the Os used in the offical image.We used ArchLinux with the kernel version 4.9.184 as a base image.
+all our tests have been executed in a Dell PowerEdge C6420 server machine.
+A summary of its hardware is listed in table \ref{fig:dahuconfig}.
+The machine is equiped with a minimal version of Debian\,9 (4.9.0 kernel version) where we install Docker (version 18.09.5).
 
@@
 
-\subsection{experimental setup}
-
-all our tests have been executed in a Dell PowerEdge C6420 server machine.
-A summary of its hardware is listed in table \ref{fig:dahuconfig}.
-The machine is equiped with a minimal version of Debian\,9 (4.9.0 kernel version) where we install Docker (version 18.09.5).
-
 \begin{figure}[hbt]
@@
 
-\subsection{Power Measurements}
+\paragraph{software settings}
+For the sake of reproducibility, each experiment runs within a Docker container.
+for each test we create a docker image. 
+\subsection{Metrics}
+Our focus will be mainly on CPU energy consumption because it is ten folds more than the DRAM one, since it is finit job benchmarking, time is highly correlated within the energy, and it will be only useful to explain certain energytical behaviour so we wont put a lot of focus on this metric.
 
-As we know, the energy of a program is the integrale of its power overtime.
-For us ower case, we used Intel \emph{Running Power Average Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754} to collect the power samples of the running tests,
+%% basically i what i want to say is that we we care mostly about energy, because there are studies that were made on the performance, plus the DRAM is not that interested even within the heavy memory  benchmakrs
+% TODO : ADD DRAM graphs for vectors and matrix multiply
 
-We used used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, to report Data collected by intel Rapl and send it into another machine that we call computing machine.
-then we calculate the Energy using  the the trapezoidal rule.
+\subsection{tests preparation}
 
-\ref{fig:trapezrule}
+To study the behaviour of the python implementation regarding the energy consumption, we have to focus on the effect of the implementation and mitigate the maximum any side effects such as the organisation of the code or any extra consumption due to the operating system or tier libraries.
+Therefore, for each test we took the implementation written in python2 as a reference and tried to use it in other implementations as it is.
+If it is not supported by python3, we transformed the code using the officiel library  \emph{2to3}\footnote{\url{https://docs.python.org/3.7/library/2to3.html}}.
+In the case of the libraries that use \emph{JIT} adding a decorator to the function that we want to optimize was enough, if there are other changes we assume that they alter the original code which is against our purpose.
+
+Each test is implemented in a Docker container for the several reasons
+\begin{itemize}
+    \item Isolation: each container has only the test program implemented with a single python runtime to remove any interference between different implementations.
+    \item Deplpoyment: to use the testing machine without extra configurations that may alter the behaviour of the os toward the enrgy consumtion
+    \item Reproductibility: One of the most frequent benchmark crimes~\cite{DBLP:journals/corr/abs-1801-02381} in research is the lack of Reproductibility, by using Docker we ensure that each test has an Image that will be accessible in public.
+\end{itemize}
+
+Despite the presence of the official docker images for the most of the runtimes, we prefered to build our own using the same reference image in order to remove any bias due to the Os used in the offical image.We used ArchLinux with the kernel version 4.9.184 as a base image.
+\section{results and finding}
+
+% \subsubsection{micro-benchmarks}
+
+% \subsubsection{benchmarks}
+
+\subsubsection{web developement}
+first we wanted to test the energy consumption of python using on djagon,one of the most popular web frameworks.
+therefore we create a simple website where we try to analyse the cost of a single request using different mechanisms to fetch the data from the database.
+the idea behind this use case is to see wether the choice of the database and the orm ( object relational mapper) impacts the energy consumption and the perfromances of the website.
+
+we considered using two different databases postgresSQL and SQlite3 that containes the same data, and three different ways to fetch the data.
+\begin{enumerate}
+    \item the naive version: which relies only on the ORM to get the data
+    \item Prefetch : basically we just prefetch the data before we request it
+    \item optimized : and here we optimize our request on the SQL level without passing by the ORM
+\end{enumerate}
+
+\begin{figure}[!hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django}
+    \caption{energy behaviour based on multiprocessing}
+    \label{fig:django}
+\end{figure}
+
+%%%%% have to rephrase this
+
+As one can see in figure~\ref{fig:django}, different choices of requesting the data have a huge impact on the energy consumption.
+as the naive version can consume up to 10x more energy than the optimized one.
+In the other hand the choice of the data base didnt have huge impact on the total energy despite their different behaviour regarding the execution time and the average power.
+This can be useful to help develeppers make a choice regarding which data base they can use based on the number of the expected requests and the expectation of the performance.
+
+Another interesting observation is the impact of the interpreter, as figure~\ref{fig:django} highligh.
+the use of Pypy interpter instead of the default one helped reducing the amount of the energy consumption even when we are dealing with the naive version.
+
+Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
+As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
+
 \begin{figure}[hbt]
     \centering
-    \begin{equation}
-        E = \int^a_b P(t)dt \simeq \sum^n_{k=1} \frac{P(t_k-1)+P(t_k)}{2}
-    \end{equation}
-    \caption{}
-    \label{fig:trapezrule}
+    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
+    \caption{energy behaviour based on multiprocessing}
+    \label{fig:python_multiprocessing}
 \end{figure}
 
-figure \ref{fig:powerapi} shows the architecture of our testing model.
+as we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
+where different sub process have to compete for the CPU Ressources.
+Another finding is when we hit the number of physical cores.
+there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
+basicaly he favorits the hyper threads than the physical core wich will lead to some context switches which cause the slow behaviour, howerver in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the eneryg consumption in this case.
+in the Chapter %% reference the work of intern from lyon  
+we discuss a deeper this behaviour of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
 
-The reason of separation between data collection and energy calculation is to minimise any interference with the test so our sensor is a light c program running inside a docker container.
+\subsubsection{python and multiprocessing}
+Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
+As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
+    \caption{energy behaviour based on multiprocessing}
+    \label{fig:python_multiprocessing}
+\end{figure}
+
+as we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
+where different sub process have to compete for the CPU Ressources.
+Another finding is when we hit the number of physical cores.
+there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
+basicaly he favorits the hyper threads than the physical core wich will lead to some context switches which cause the slow behaviour, howerver in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the eneryg consumption in this case.
+in the Chapter %% reference the work of intern from lyon  
+we discuss a deeper this behaviour of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
+
+%% todo 
+% add performance 
+% add dram \ a glimps
+
+\subsection{python insights}
+
+The goal of this section is to find some insights where we can optimize the energy consumption without impacting the source code.
+so we started by extending the work of \citeauthor{hasan_energy_2016} and \citeauthor{oliveira_recommending_nodate} to the python environement.
+
+%%%% concurency
+
+another field of investigation is the type data and structure conroles. that might impact the energy consumption.
+To do so. We iterate over a list using three methodes.
+first the classical for ( for i in range (len(n))).
+howerver as we can see here unlike other porgramming langauges it requires extra operations such as determining the length of the collection , and then using the iterator range.
+so we tried the more adapted version
+( for element in collection).
+Morever in most programming languages the for loop is translated to a while loop ( transfirmation from D type to B type -- asm -- ), therefore we wanted to compare this with a ( while ) version.
+After determining the main ways to iterate over a loop we run the algorithms  collections of different data types to see wether it will impact of the energy consumption of the code, same thing for the size of the collection.
+
+As we can see in figure~\ref{fig:pythonloops} the type of the data hasn't an impact on the energy consumption.In the other hand the way we iterate over the collection has a huge factor.
+interestely, the for in range loop was by far the optimal one with following by the regural for in collection, and the while part was the least one with an overhead of 400\% compared to the first option.
+
+%%% putting the idea, later i ll reformulate it 
+the reason behind a such behaviour is mainly related on how python interpeter is made.
+% add refenece https://www.python.org/doc/essays/list2str/ 
+%% https://www.pythonpool.com/for-vs-while-loop-python/  
+% https://wiki.python.org/moin/TimeComplexity
+%% http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1
+In order to reduce the latency of the python code.
+most of the buitl-in functions an operations are written in C, same thing goes for the function range, further more the function len has a complexity of O(1) because it is based on the function Py\_SIZE  of C which stores the length in a field for the object.
+
+so basically for the for in range is basically creating a new iterator that has the same length of the first one.
+and for each iteration we will have to do a second acceess ( l[i]) instead of one one access hence the double time.
+
+for the wile is even slower due to the implicite incrementation of the variable which will cause an extra operation during the loop .
+
+to confirm the hyphothsis.
+we tried to construct a new list by editing the elements of the previous one.
+And as predicted , the builtin methodes are the best energy saving, while the customize while loop is the heaviest.
+an interesting finding is.
+the impact of annonymous functions ( lambda expressions) on the energy consumption.
+the reseason behind this is the fact that python treats those functions as local variables unlike the predifined ones which are global in our case.
+therefore they are faster and consumes less energy  % same reference https://www.python.org/doc/essays/list2str/
+
+\begin{figure}[htb]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/python_iterations}
+    \caption{energy behaviour of different python loops }
+    \label{fig:pythonloops}
+\end{figure}
+
+\begin{figure}[htb]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/python_treatemens}
+    \caption{energy behaviour of different methodes to change a list}
+    \label{fig:pythontreatement}
+\end{figure}
+
+\paragraph{conclusion}
+This study has shown us, that the optimial way to reduce the energy consumption of the python code is to follow the guidelines and use the builtin functions, which is kinda happy news for the developers since they don't have to make extra effort to make their code green
+
+\subsection{python runtimes}
+
+As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
+Intelpython and active python both folows the same behaviour. one can conclude that the work that have been on those interpreters is mainly
+to improve a specific purpose, Active python claims that their version is focused on secuirty which explains the lack of some performances due to the introduction of more reonforcment
+.Morver Intel published their version of python as a dedicated for machine learning. Unfortunately the tommti benchmark is a set that focuses on the general purpose programming
+which does not reflect the performance of the machine learning.
+Another aspect of a such behaviour might be due to the fact of the processors that were used in the testbed.
+and it might change in the future if we include the GPU part for the machine learning benchmarks
+% reference the work of the intern 
+As for nuitka. there were no optimization in the energy consumpotion despite the fact that is it a complier.
+However if we dig through the nuitka mechanisms. they basically embed the python code with an interpreter.
+Unllike nuitka shedskin exhebit a the best energy consumption pattern when it comes to the arithmetic operartions.
+One can conclude it is due to the fact of the native type of the variables.
+unlike the interpreters where they are treated as object in the begining.
+
+for the other interpeters pypy is very promising especially when it comes to data maniupulation as one can see in the figure \ref{fig:bar_tommti_vector}
+pypy is by far the best interperer when it comes to treating vectors.
+numba2 introduced the JIT but wasn't as promising as numba3.
+
+for the other vm based interpereters.
+jython and ipy lacked in term of energy optimisation which was kinda expected since they were in their the begining of the stage and the main puprose of such implementation is to link the bytecode generated by jython and ironpython with their respectives virtual machin.
+
+Unlike the the previous interpeters. graal exhebits a certin promises when it comes to complex algorithms - nested loops -
+micro python is dedicated to embeded systems so lunching it is powerful cpu machines will be misleading.
+
+Most of the interperters had the same behaviour when it comes to the input outputs. except for jython which was kinda of a anomaly probably due to the lack of the optimizations.
+% pypy handles exceptions very well
 
@@
 
-% \subsubsection{micro-benchmarks}
+\section{Threads to validity}
+\section{tools and contributions}
 
-\subsubsection{benchmarks}
-
-\subsection{results and foundings}
-As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
-Intelpython and active python both folows the same behaviour. one can conclude that the work that have been on those interpreters is mainly
-to improve a specific purpose, Active python claims that their version is focused on secuirty which explains the lack of some performances due to the introduction of more reonforcment
-.Morver Intel published their version of python as a dedicated for machine learning. Unfortunately the tommti benchmark is a set that focuses on the general purpose programming
-which does not reflect the performance of the machine learning.
-Another aspect of a such behaviour might be due to the fact of the processors that were used in the testbed.
-and it might change in the future if we include the GPU part for the machine learning benchmarks
-% reference the work of the intern 
-As for nuitka. there were no optimization in the energy consumpotion despite the fact that is it a complier.
-However if we dig through the nuitka mechanisms. they basically embed the python code with an interpreter.
-Unllike nuitka shedskin exhebit a the best energy consumption pattern when it comes to the arithmetic operartions.
-One can conclude it is due to the fact of the native type of the variables.
-unlike the interpreters where they are treated as object in the begining.
-
-for the other interpeters pypy is very promising especially when it comes to data maniupulation as one can see in the figure \ref{fig:bar_tommti_vector}
-pypy is by far the best interperer when it comes to treating vectors.
-numba2 introduced the JIT but wasn't as promising as numba3.
-
-for the other vm based interpereters.
-jython and ipy lacked in term of energy optimisation which was kinda expected since they were in their the begining of the stage and the main puprose of such implementation is to link the bytecode generated by jython and ironpython with their respectives virtual machin.
-
-Unlike the the previous interpeters. graal exhebits a certin promises when it comes to complex algorithms - nested loops -
-micro python is dedicated to embeded systems so lunching it is powerful cpu machines will be misleading.
-
-Most of the interperters had the same behaviour when it comes to the input outputs. except for jython which was kinda of a anomaly probably due to the lack of the optimizations.
-% pypy handles exceptions very well
-
-the energy consumption.
-
-%% todo 
-% add performance 
-% add dram \ a glimps
-
-\subsection{tools and package}
-
-\subsection{Threads to validity}
-
-\subsection{conclusion}
+\section{conclusion}
 

### 2022-06-27 chapters/benchmarking/conclusion.tex
add : extension P1

--- 2022-05-06 chapters/testing/conclusion.tex
@@
-%  TODO: add the environement part,
-%  TODO: add perspective
-
-\section{Experimental Guidelines}\label{sec:guidelines}
-To summarize our experiments, we provide some experimental guidelines in Table~\ref{table:guidelines}, based on the multiple experiments and analysis we did.
-These guidelines constitute a set of minimal requirements or best practices, depending on the workload and the criticality of the energy measurement precision.
-It therefore intends to help practitioners in taming the energy variation on the selected CPU, and conduct the experiments with the least variations.
-
-\begin{table}[h!]
-    \centering
-    \caption{Experimental Guidelines for Energy Variations}
-    \small
-    \begin{tabular}{|p{4.7cm}|c|c|}
-        \hline
-        \textbf{Guideline}                                                                                                                                                & \textbf{Load} & \textbf{Gain}     \\
-        \hline
-        \hline
-        Use a low TDP CPU                                                                                                                                                 & Low \& medium & Up to $3\times$   \\
-        \hline
-        Disable the CPU C-states                                                                                                                                          & Low           & Up to $6\times$   \\
-        \hline
-        Use the least of sockets in a case of multiple CPU                                                                                                                & Medium        & Up to $30\times$  \\
-        \hline
-        Avoid the usage of hyper-threading whenever possible                                                                                                              & Medium        & Up to $5\times$   \\
-        \hline
-        Avoid rebooting the machine between tests                                                                                                                         & High          & Up to $1.5\times$ \\
-        \hline
-        Do not relate to the machine idle variation to isolate a test EC, the CPU/OS changes its behavior when a test is running and can exhibit less variation than idle & Any           & ---               \\
-        \hline
-        Rather focus the optimization efforts on the system under test than the OS                                                                                        & Any           & ---               \\
-        \hline
-        Execute all the similar and comparable experiments on a same machine.
-Identical machines can exhibit many differences regarding their energy behavior             & Any           & Up to $1.3\times$ \\
-        \hline
-    \end{tabular}
-    \label{table:guidelines}
-\end{table}
-
-Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and its variation reduction at low workloads.
-However, it also lists some new factors that we identified along the analysis we conducted in Section, such as the results related to the OS or the reboot mode.
-Some of the guidelines are more useful/efficient for specific workloads, as showed in our experiments.
-Thus, qualifying the workload before conducting the experiments can help in choosing the proper guidelines to apply.
-Other studied factors are not been mentioned in the guidelines, like the Turbo~Boost or the Speculative execution, due to the small effect that has been observed in our study.
-
-In order to validate the accuracy of our guidelines among a varied set of benchmarks on one hand, and their effect on the variation between identical machines on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
-Half of these experiments has been performed at a 50\,\% workload and the other half on single process jobs.
-The choice of these two workloads is related to the optimization guidelines that are mainly effective at low and medium workloads.
-We note that we used the cluster \textsf{Dahu} over \textsf{Ecotype} to highlight the guidelines effect on the nodes where the variation is susceptible to be higher.
-
-\begin{figure*}%[!htb]
-    \center{\includegraphics[width=\linewidth]{imgs/all}}
-    \caption{Energy variation comparison with/without applying our guidelines}\label{fig:optimized}
-\end{figure*}
+\section{Extension}
+In computer science, things are changing fast, which leads to a raise of obsolete results. Specially when it comes to Comparison studies,One can't cover all the candidates. 
+Moreover, between the initial experiment and the published results there might be new candidates and an evolution of
+others. Therefore we Propose new creterioin for a successful experiment ~Extension~.
+In our content Extension means the ability to provide the necessary tools to not only Repeat the experiment but to be able to add extra candidates, Workloads or metrics.
 
 \begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/stressng}}
-    \caption{Energy variation comparison with/without applying our guidelines for \textsc{Stress-NG}}\label{fig:stress}
+    \center{\includegraphics[width=.9\linewidth]{imgs/benchmarkingprotocol}}
+    \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
 \end{figure}
-
-Figure~\ref{fig:optimized} and~\ref{fig:stress} highlight the improvement brought by the adoption of our guidelines.
-They demonstrate the intra-node STD reduction at low and medium workloads for all the benchmarks used at different levels.
-Concretely, for low workloads, the energy variation is 2--6 times lower after applying the optimization guidelines for the benchmarks \textsf{LU} and \textsf{EP}, as well as \textsc{Linpack}, while it is 1.2--1.8 times better for \texttt{Sha256}.
-For this workload, the overall energy consumption after optimization can be up to 80\,\% higher due to disabling the C-states to keep all the unused cores at a high power consumption state ($Mean_{LU-normal-Dahu2}=11,500 mJ$, $Mean_{LU-optimized-Dahu2}=20,508 mJ$).
-For medium workloads, the STD, and thus variation, is up to 100\,\% better for the benchmark \textsf{CG}, 20--150\,\% better for the \texttt{pbzip2} application and up to 100\% for \textsc{Stress-NG}.
-We note that the optimized version consumes less energy thanks to an appropriate core pinning method.
-
-Figures~\ref{fig:optimized} and~\ref{fig:stress} also highlight that applying the guidelines does not reduce the inter-nodes variation in all the cases.
-This variation can be up to 30\,\% in modern CPU~\cite{wang_experimental_nodate}.
-However, taming the intra-node variation is a good strategy to identify more relevant mediums and medians, and then perform accurate comparisons between the nodes variation.
-Even though, using the same node is always better, to avoid the extra inter-nodes variation and thus improve the stability of measurements.
-
-\section{Threats to Validity}\label{sec:threats}
-A number of issues affect the validity of our work.
-For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPU, but still adds an important overhead if we adopt a sampling at high frequency.
-The other fine-grained tool we used for measurements is \textsc{PowerAPI}.
-It allows to measure the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
-The usage of \textsc{PowerAPI} adds an error margin because of the power model built over RAPL.
-The RAPL tool mainly measures the CPU and DRAM energy consumption.
-However, even running CPU/RAM intensive benchmarks would keep a degree on uncertainty concerning the hard disk and networking energy consumption.
-In addition, the operating system adds a layer of confusion and uncertainty.
-
-The Intel CPU chip manufacturing process and the materials micro-heterogeneity is one of the biggest issues, as we cannot track or justify some of the energy variation between identical CPU or cores.
-These CPU/cores might handle frequencies and temperature differently and behave consequently.
-This hardware heterogeneity also makes reproduction complex and requires the usage of the same nodes on the cluster with the same OS.
-
-% A more subtle issue may arise due to the values of the measurements that we achieved.
-% In fact, the energy measures are quite small, and iterations may be taking a few milliseconds more or less to run.
-% A thing we cannot measure using our measurement tools.
-% How generalizable are our results? As a set of energy variation optimization guidelines, we argue that our results applied on most of the modern Intel CPU.
-% However, using and comparing identical CPU is still tricky and is very dependent to chips.
-
-\section{Conclusion}\label{sec:conclusion}
-In this paper, we conducted an empirical study of controllable factors that can increase the energy variations on platforms with some of the latest CPU, and for several workloads.
-We provide a set of guidelines that can be implemented and tuned (through the OS GRUB for example), especially with the new data centers isolation trend and the cloud usage, even for scientific and R\&D purposes.
-Our guidelines aim at helping the user in reducing the CPU energy variation during systems benchmarking, and conduct more stable experiments when the variation is critical.
-For example, when comparing the energy consumption of two versions of an algorithm or a software system, where the difference can be tight and need to be measured accurately.
-
-Overall, our results are not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
-The aim of our work is to be able to tame and mitigate this variability along controlled experiments.
-We studied some previously discussed aspects on some recent CPU, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to achieve the variability mitigating purpose.
-Some of these factors, like the C-states usage, can reduce the energy variation up to 500\,\% at low workloads, while choosing the wrong cores/PU strategy can cause up to $30\times$ more variability.
-
-We believe that our approach can also be used to study/discover other potential variability factors, and extend our results to alternative CPU generations/brands.
-Most importantly, this should motivate future works on creating a better knowledge on the variability due to CPU manufacturing process and other factors.
 

### 2022-06-28 chapters/optimization/java-dir/java.tex
add : extention to JAVA

@@
 
-\section{Experimental Protocol}\label{sec:javameth}
+\section{Experimental Protocol}\label{sec:javaprotocol}
 To investigate the effect that could have the JVM distribution choice and/or parameters on software energy consumption, we conducted a wide set of experiments on a cluster of machines and using several established Java benchmarks and JVM configurations.
@@
 To collect the number of active threads used by the an experiment, we use the command \texttt{top} and record at fixed intervals.
+
+\subsection{Extension}
+We also added an extension to the protocol to allow the user to run the same experiment with different configurations.
+The package is available in the github repository \footnote{\url{https://github.com/chakib-belgaid/jvm-comparaison}}.
+To add extra \textbf{candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the jvm to be used.
+\note{The jvm must be provided by sdkman\footnote{\url{https://sdkman.io/}}}
+For the \textbf{workload},the benchmarks should be provided in the \texttt{benchmarks} directory.
+As for extra \textbf{metrics} One can create a new script file that monitor the experiment and record the metrics.
+We provide some examples such as \texttt{recordpower.sh} to measure the instantanious power and \texttt{recordthreads.s} to measure the number of active threads during the experiment.
+
+For faster experements we propose \textbf{Jreferral}\footnote{\url{https://github.com/chakib-belgaid/jreferral}} an open source that provides an opensource to compare the energy consumption of differents jvms for any java application.
+Later we will discuss this tool in the section \ref{section:SummaryofContributions}
 
@@
 Options are available for \textsf{J-Referral} to obtain much more detailed reports including execution time, DRAM usage, split DRAM vs. CPU consumption, etc.
-The tool is available as \emph{open-source software} (OSS) from our anonymous repository.\footnote{\url{https://anonymous.4open.science/r/jreferral/Readme.md}}
+The tool is available as \emph{open-source software} (OSS) from our anonymous repository.\footnote{\url{https://github.com/chakib-belgaid/jreferral}}
 

### 2022-06-28 chapters/programming_languages/rpc/rpc.tex
add : extension to RPC

--- 2022-06-10 chapters/programming_languages/rpc-dir/rpc.tex
@@
 \subsection*{introduction}
-%  this is just to warmup 
 With the emergence of cloud technologies, many protocols wanted to take the lead.
@@
 \subsection{Experimental protocol}
-\subsubsection{Hardware settings}
+\subsubsection{measurement context}
+\paragraph{Hardware settings}
 All the experiments are run on the cluster \textsf{paravance} of the G5K platform.
@@
 Furthermore, we used Docker containers technology for reproducibility of the experiments and the isolation of the servers.
+\paragraph{Client and server environments}
+To limit the impact of the network on the experiments, we run both the client and the server on the same machine.
+However, we isolate each part on a separate socket, in order to reduce the effect that the client might have on the server and \emph{vise versa}.
+To do so, for each iteration, we always run the same client on \textsf{socket\,0} and the server that we want to test on \textsf{socket\,1}.
+Both the server and the client use the whole socket for their experiment.
+In addition, all the additional services, such as the kernel and HwPC sensor, are run on \textsf{socket\,0}.
+Therefore, the only process being executed in \textsf{socket\,1} is the server that we benchmark and monitor.
 
-\subsubsection{Energy measurements}
+\subsubsection{metrics}
+\paragraph{Energy measurements}
 To report on the energy consumption, we used HwPC sensor [TOCITE], which is based on Intel RAPL technology, one of the most accurate tools to measure the energy consumption of the CPU and DRAM [TOCITE].
@@
 
-\subsubsection{Client and server environments}
-To limit the impact of the network on the experiments, we run both the client and the server on the same machine.
-However, we isolate each part on a separate socket, in order to reduce the effect that the client might have on the server and \emph{vise versa}.
-To do so, for each iteration, we always run the same client on \textsf{socket\,0} and the server that we want to test on \textsf{socket\,1}.
-Both the server and the client use the whole socket for their experiment.
-In addition, all the additional services, such as the kernel and HwPC sensor, are run on \textsf{socket\,0}.
-Therefore, the only process being executed in \textsf{socket\,1} is the server that we benchmark and monitor.
-
-\paragraph{Client:}
-% TODO : add the github repository links 
+\paragraph{performances}
 For better accuracy and more details, We use an updated version of the open source RPC benchmarking tool, named GHZ (\url{https://ghz.sh/}).
 The modified version allows us to monitor the average power for each request from both the server and the client sides.
-The new version is available in the repository.%ROMAIN: Which repo?
-The client consumes the protocol description to generate an implementation for the message \texttt{helloworld}, and then forks 50 instances that send the same request to the server concurrently.
+The new version is available in the repository\footnote{\url{https://github.com/chakib-belgaid/energy_ghz}}.
 
-\paragraph{Server:}
+\subsubsection{workload}
+The purpose of the experiment is to analyse the behaviour of different GRPC implementations. therefore we have two kind of worlkoads 
+\paragraph{number of clients}: this is the number of concurrent clients that we want to test. It is automatically handled with the ghz client 
+\paragraph{payload}: the size of each request . this varries from 50 Bytes up to 10 MB.
+
+The client consumes the protocol description which can be found if the file \texttt{helloworld.proto} to generate an implementation for the message  and then forks multiple instances that send the same request to the server
+
+\subsubsection*{candidates}
 The server implementations are based on the official implementation by Google for most of the languages.
 Each server uses 16 cores and is limited to 512\,MB of RAM.
+each implementation is created through a docker image, which will allow us to add new implementations easily.
+
+\subsubsection{Extension}: 
+for the sake of extedind the experement we provide a github repository \footnote{\url{https://github.com/chakib-belgaid/energy_benchmarking_grpc2}} that containes the implementation of the experement.
+Adding \textbf{candidates} can be achieved by creating a new docker image and putting it in a new folder named after the language.
+As for the \textbf{workload} it can be extended easly by adding new files in the folder \textbf{payload}.
 

### 2022-06-28 chapters/optimization/python-dir/python.tex
add ; frameworks table
update
add : extension for python

@@
+%% TODO : refactor the experemental protocol to have the 3 main components measurement context, workload and metrics 
 \section{Introduction}
@@
 Despite the presence of the official docker images for the most of the runtimes, we prefered to build our own using the same reference image in order to remove any bias due to the Os used in the offical image.We used ArchLinux with the kernel version 4.9.184 as a base image.
-\section{results and finding}
+
+\subsection{Extension}
+As we have done with the previous chapters. we provide a tool that allows to extend the tests with new workloads and new candidates. 
+In the repository \footnote{\url{https://github.com/chakib-belgaid/python-implementations}}.
+we have a tool that allows to generate new workloads and new candidates.
+The script \texttt{generator.py} allows to create new benchmarks by implementing a python code within different interpeters.
+then it generates \texttt{launcher-benchmark.sh} that can be used directly to run the expermient.
+Furthermore all the successful implementations are stored in seperate directory and the that couldn't work ( mostly because of compatibility issues) stored in a recap file called \texttt{benchmarkTest.md}.
+where benchmark is the name of the new workload;
+To add extra \textbf{candidates} one should add a base docker file 
+that containes the new implementation and if there should be extra manipulation that should be added to the workload files ,such as adding new decorator or changing some parameters, then they should be added as an extra function in the script \texttt{generator.py}.
+Finally they should be included in the python candidates.
+
+\section{Results and finding}
 

### 2022-06-30 chapters/literature/draft_citations.tex
rename : images folder

@@
 
+Efficiency in energy usage is a well-known topic.
+In the majority of fields, the purpose is to minimize the energy consumption of electrical devices/components.
+Modern times even see energy classification (A, B,...
+F) for Homes, cars and electronic products, to provide the consumer an indication of the energy consumption of his devices, which will reflect on his power bill.
+This creterions is extended even to the hardware components of computer.
+figure\ref{fig:comparaisoncpu} \footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}}
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{imgs/cpu_cost_comparaison}
+    \caption{electrical cost comparasion between two cpus }
+    \label{fig:comparaisoncpu}
+\end{figure}
+
+In computer science, the objective is essentially same.
+Numerous studies have been conducted on energy optimization.
+Some of these studies concentrate on minimizing energy consumption at the hardware level, while others optimize energy consumption via software.
+% displays the digital energy consumption distribution in 2017 \cite{andrae2015global}. ("p" for production and "u" for usage). 
+% It demonstrates that both the production and consumption stages have substantial effects.
+% In addition, the utilization expenses of data centers, networks, and terminals are also substantial (19 percent , 16 percent and 20 percent respectively).
+% TODO : as romain about the graphics for energy distribution
+
+As an example \citeauthor{avgerinou2017trends} evaluated the development of power use effectiveness (PUE) in data centers that belongs to various organizations participating in the European code of conduct for energy efficiency program \cite{avgerinou2017trends}.
+The research found a gradual decline in the PUE of data centers, which measures the ratio of the overall energy supplied to the energy used by IT equipment.
+A low PUE implies that the majority of energy is utilized to power the data center's IT equipment, while just a little amount is needed for cooling and lighting.
+
+In this thesis, we will place a greater emphasis on the software level in order to decrease the amount of energy that is used.
+In order to do this, we provide a benchmarking process and a set of tools that are intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
+
+Thus, Section \ref{sec:energymeasurement}  describes software energy measurements.
+It provides examples of hardware and software measuring instruments and describes their differences, benefits, and drawbacks.
+It also examines the sources of energy measurement variations, which represent a significant obstacle to achieving precise readings.
+
 \section{benchmarking}
 
+definition :benchmark: [noun] something that serves as a standard by which others may be measured or judged.
+that serves as a basis for evaluation or comparison (as of computer system performance)
 this is \cite{arcuri_practical_2011}
@@
 %%%%%%%%%%%%%%%%%
-% TODO : REPHRASE THIS other wise it will be plagiarism
+%% generate an introduction for this chapter 
+%%%%%%%%%%%%%%%%%
+\section{introduction}
+
 list of benchmarks \cite{van_der_kouwe_benchmarking_2018}:
@@
 %\\\ https://www.cpubenchmark.net/compare/Intel-i7-12700KF-vs-Intel-i7-12700F/4608vs4692 | hardware benchmarking 
-\section{energy consumption}
+\section{energy measurement}\label{sec:energymeasurement}
 
@@
 \begin{figure}
-    \center{\includegraphics[width=1\linewidth]{images/virtualization_techniques}}
+    \center{\includegraphics[width=1\linewidth]{imgs/virtualization_techniques}}
     \caption{Different Methods of Virtualzation}\label{environement:virtualization_technique}
@@
 \begin{figure}
-    \center{\includegraphics[width=.5\linewidth]{images/docker_vs_vm_energy_paper/idle_energy}}
+    \center{\includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/idle_energy}}
     \caption{energy consumption of Idle system with and without docker \cite{santos2018does}}\label{fig:docker_idle}
@@
 \begin{figure}
-    \includegraphics[width=.5\linewidth]{images/docker_vs_vm_energy_paper/reddis_time}
-    \includegraphics[width=.5\linewidth]{images/docker_vs_vm_energy_paper/reddis_energy}
+    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_time}
+    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_energy}
     \caption{execution time and energy consumption of Redis  with and without docker \cite{santos2018does}}\label{fig:docker_reddis}
@@
 \begin{figure}[!hbt]
-    \center{\includegraphics[scale=0.3]{images/greenbenchmarkcycle}}
+    \center{\includegraphics[scale=0.3]{imgs/greenbenchmarkcycle}}
     \caption{the spiral methode of energy optimization }\label{fig:spirals}

### 2022-06-30 chapters/literature/draft_citations.tex
add : introduction to soa chapter

@@
 % In addition, the utilization expenses of data centers, networks, and terminals are also substantial (19 percent , 16 percent and 20 percent respectively).
-% TODO : as romain about the graphics for energy distribution
+% TODO : ask romain about the graphics for energy distribution
 
-As an example \citeauthor{avgerinou2017trends} evaluated the development of power use effectiveness (PUE) in data centers that belongs to various organizations participating in the European code of conduct for energy efficiency program \cite{avgerinou2017trends}.
+As an example \citeauthor{avgerinou2017trends} evaluated the development of power use effectiveness (PUE) in data centers thkat belongs to various organizations participating in the European code of conduct for energy efficiency program \cite{avgerinou2017trends}.
 The research found a gradual decline in the PUE of data centers, which measures the ratio of the overall energy supplied to the energy used by IT equipment.
@@
 
-In this thesis, we will place a greater emphasis on the software level in order to decrease the amount of energy that is used.
-In order to do this, we provide a benchmarking process and a set of tools that are intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
+we will place a greater emphasis on the software level in order to decrease the amount of energy that is used.more particularly on the execution phase of the program cycle.
+We will be proceeding through an empirical analysis of the energy consumption of the software while changing some components of the source code without impacting its behavior.
+In order to do this, we will elaborate a benchmarking process and a set of tools that are intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
+Thus, we will begin by examining the state of the art of empirical analysis and retrieving the best empirical experimentation methodologies in the research field.Then, we will narrow these practices down to computer science so that we can finally adapt them to energy consumption.
 
-Thus, Section \ref{sec:energymeasurement}  describes software energy measurements.
+Section~\ref{sec:soa_benchmarking} will discuss the pitfalls and best practices associated with empirical research before applying them to our field of interest.
+After that Section~\ref{sec:energymeasurement} describes software energy measurements.
 It provides examples of hardware and software measuring instruments and describes their differences, benefits, and drawbacks.
-It also examines the sources of energy measurement variations, which represent a significant obstacle to achieving precise readings.
-
-\section{benchmarking}
+It also examines the sources of energy measurement variations, which represent a significant obstacle to achieving precise readings and a higher accuracy.
+Then, in section \ref{section:soa_optimization}, we will go through some of the previous work on improving the energy consumption of softwares.
+\section{benchmarking}\label{sec:soa_benchmarking}
 
@@
 
-\section{programming languages and performances}
-\subsection{energy}
+\section{programming languages and performances}\label{section:soa_optimization}
+\subsection{energy}\
 

### 2022-06-30 chapters/literature/draft_citations.tex
state of the art : benchmarking crimes

@@
 This creterions is extended even to the hardware components of computer.
-figure\ref{fig:comparaisoncpu} \footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}}
+figure\ref{fig:soa_comparaisoncpu} \footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}}
 
@@
     \caption{electrical cost comparasion between two cpus }
-    \label{fig:comparaisoncpu}
+    \label{fig:soa_comparaisoncpu}
 \end{figure}
@@
 Then, in section \ref{section:soa_optimization}, we will go through some of the previous work on improving the energy consumption of softwares.
+
+\newpage
 \section{benchmarking}\label{sec:soa_benchmarking}
+This section will go through the flaws and best practices of empirical research before applying them to our topic of study.
 
-definition :benchmark: [noun] something that serves as a standard by which others may be measured or judged.
-that serves as a basis for evaluation or comparison (as of computer system performance)
-this is \cite{arcuri_practical_2011}
+% We first start by defining the word \textbf{benchmark},  
+% definition :benchmark: [noun] something that serves as a standard by which others may be measured or judged.
+% that serves as a basis for evaluation or comparison (as of computer system performance)
+In their work~\cite{van_der_kouwe_benchmarking_2018}, \citeauthor{van_der_kouwe_benchmarking_2018} Investigated 50 papers published in top venues
+to findout that tier-1 papers commit an average of five benchmarking crimes.
+To analyze the magnitude of the phenomenon, they have identified a set of 22 "benchmarking crimes" that threaten the validity of the system.
+% TO [ROMAIN] should i include  this list ? 
+% thoes benchmark crimes that one can commit intentionally or by accident and that could jeopardize the integrity of their paper.t are listed below.
+% \begin{itemize}
+%     \item  Not evaluating potential performance degradation
+%     \item  Benchmark subsetting without proper justification
+%     \item  Selective data set hiding deficiencies
+%     \item  Microbenchmarks representing overall performance
+%     \item  Throughput degraded by x\% => overhead is \%
+%     \item  Creative overhead accounting
+%     \item  No indication of significance of data
+%     \item  Incorrect averaging across benchmark scores
+%     \item  Benchmarking of simplified simulated system
+%     \item  Inappropriate and misleading benchmarks
+%     \item  Same dataset for calibration and validation
+%     \item  No proper baseline
+%     \item  Only evaluate against yourself
+%     \item  Unfair benchmarking of competitors
+%     \item  Not all contributions evaluated
+%     \item  Only measure runtime overhead
+%     \item  False positives/negatives not tested
+%     \item  Elements of solution not tested incrementally
+%     \item  Missing platform specification
+%     \item  Missing software versions
+%     \item  Subbenchmarks not listed
+%     \item  Relative numbers only
+% \end{itemize}
 
-difference tests  \cite{sonnenwald2003evaluating}
+%% BENCHMARKING FRAMEWORK -- ADD IT later \cite{sonnenwald2003evaluating}
 %further references 
-\cite{lilja2005measuring}
-\cite{bukh1992art}
-%%%%%%%%%%%%%%%%%
-%% generate an introduction for this chapter 
-%%%%%%%%%%%%%%%%%
-\section{introduction}
+In the book "Measuring computer performance: a practitioner's guide"\cite{lilja2005measuring},\citeauthor{lilja2005measuring} examines performance indicators and gives in-depth treatment of benchmark program tactics.
+He provides clear explanations of the basic statistical methods required for interpreting measured performance data.
+He also outlines the overall 'design of experiments' method and demonstrates how to collect the most information with the least amount of work.
+This practical book will appeal to anybody seeking a comprehensive, yet intuitive, grasp of computer system performance analysis.
+\citeauthor{mytkowicz2009producing} made an evaluation  of 133 studies from ASPLOS, PACT, PLDI, and CGO, to find out that none of the experimental findings papers appropriately considered measurement bias.
+Which can lead derive incorrect results from an experiment if a seemingly insignificant feature of the experimental design is altered.
+They treated this problem by proposing two strategies for detecting measurement bias by using causal analysis and preventing it with setup randomization \cite{mytkowicz2009producing}.
+\citeauthor{bukh1992art} wrote a book about computer performance analysis, where he discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems .
 
-list of benchmarks \cite{van_der_kouwe_benchmarking_2018}:
-\begin{itemize}
-    \item  Not evaluating potential performance degradation
-    \item  Benchmark subsetting without proper justification
-    \item  Selective data set hiding deficiencies
-    \item  Microbenchmarks representing overall performance
-    \item  Throughput degraded by x\% => overhead is \%
-    \item  Creative overhead accounting
-    \item  No indication of significance of data
-    \item  Incorrect averaging across benchmark scores
-    \item  Benchmarking of simplified simulated system
-    \item  Inappropriate and misleading benchmarks
-    \item  Same dataset for calibration and validation
-    \item  No proper baseline
-    \item  Only evaluate against yourself
-    \item  Unfair benchmarking of competitors
-    \item  Not all contributions evaluated
-    \item  Only measure runtime overhead
-    \item  False positives/negatives not tested
-    \item  Elements of solution not tested incrementally
-    \item  Missing platform specification
-    \item  Missing software versions
-    \item  Subbenchmarks not listed
-    \item  Relative numbers only
-\end{itemize}
 % further references
-\cite{mytkowicz2009producing}
+
 %%%%%%%%%% 
-
-startup performance vs steady performance \cite{buytaert_statistically_nodate}
-
-%%%%% ** reasons behind the energy variation ****
-
-simgrid simulation of power consumption of parallel aplication using simgrid \cite{heinrich2017predicting}
-
-impact of manifacturing process on the energy variation \cite{coles2014comparing}
-
-different energy consumption between identical processers [ idle and hight load ]\cite{von2016variations}
-
-the temperature is the main raison a high energy variation \cite{wang2018potential}
-
-external factors are impact on the energy variation \cite{mukherjee2009spatio}
-
-the place of the server in the room doesn't affect the energy variation \cite{diouri2013your}
-
-comparaison of three measuring tools and they did exhebit 10\% variation each time \cite{inadomi2015analyzing}
-
-%%****** mitigating the energy variation ******
-low-cost scalable variation-aware algorithm \cite{inadomi2015analyzing}
-
-reducing the energy variation by disabling turbo boost \cite{acun2016variation}
-
-reducing the energy consumption in parallel systems \cite{chasapis2016runtime}
-
-\cite{marathe2017empirical}
-
+In their paper~\cite{buytaert_statistically_nodate},\citeauthor{buytaert_statistically_nodate} they separated the start-up measurement from the steady state when comparing different Java optimizations.
+\\
 %\\\ https://www.cpubenchmark.net/compare/Intel-i7-12700KF-vs-Intel-i7-12700F/4608vs4692 | hardware benchmarking 
@@
 % split hardware tools into extern and embeded tools
+
+simgrid simulation of power consumption of parallel aplication using simgrid \cite{heinrich2017predicting}
+
+impact of manifacturing process on the energy variation \cite{coles2014comparing}
+
+different energy consumption between identical processers [ idle and hight load ]\cite{von2016variations}
+
+the temperature is the main raison a high energy variation \cite{wang2018potential}
+
+external factors are impact on the energy variation \cite{mukherjee2009spatio}
+
+the place of the server in the room doesn't affect the energy variation \cite{diouri2013your}
+
+comparaison of three measuring tools and they did exhebit 10\% variation each time \cite{inadomi2015analyzing}
+
+%%****** mitigating the energy variation ******
+low-cost scalable variation-aware algorithm \cite{inadomi2015analyzing}
+
+reducing the energy variation by disabling turbo boost \cite{acun2016variation}
+
+reducing the energy consumption in parallel systems \cite{chasapis2016runtime}
+
+\cite{marathe2017empirical}
 

### 2022-07-01 chapters/literature/draft_citations.tex
add : soa software tools

@@
 \subsection{software tools}
-Now that we got our hands on a precise tools that allows us to monitor
 
-powerapi \cite{colmant2018next}
+Software-based measurement tools are based on other hardware tools to monitor energy use.
+Granularity is the core value of these technologies, unlike hardware tools, which only provide the total energy usage of the system/component (computer, server, motherboard, etc.) for most cases.
+Because they are frequently constructed on empirical estimations and data learning methods, they drop in accuracy.
 
-smartwatts formula \cite{fieni2020smartwatts}
+Many software measuring tools understand the behavior of a power model and provide estimates of energy usage.
+This model is then used to allocate the observed energy consumption among different execution entities, such as processes, control groups, threads, or even code lines.
 
-Wattwatcher \cite{lebeane2015watt}
+The first examples of software measurement tools are PowerAPI~\cite{colmant2018next}, SmartWatts formula~\cite{fieni2020smartwatts} and selfWatts~\cite{fieni2021selfwatts}.
+These tools collect global energy consumption measurements from RAPL and use other system events such as cache misses/hits and CPU frequency evolution (DVFS) via a sensor to construct a power model of the control groups (system control groups, docker containers, kubernetes pods, etc.) using a Ridge regression.
+The model continuously learns and improves its precision of real-time energy usage data with a maximum frequency of 100 Hz.
+The instrument has a decentralized, lightweight design.
+Only the lightest sensors required for data collection and transmission are put into the monitored devices.
+The SmartWatts formula is then executed on the primary server in order to construct the model that permits assigning the energy usage for each operating control group.
+PowerAPI is only compatible with Linux on a bare-metal physical computer.
+\\
+WattWatcher~\cite{lebeane2015watt} is a multi-core power measuring framework that provides process-level energy measurements.
+This program uses power models to predict process energy usage.
+It uses CPU events which are passed from the measured node to a model generator node in order to construct the power model.
+It works by combining a description of the CPU with a list of the hardware events through multiple calibration phases in order to build a robust model.
 
-joulemeter \cite{kothari2009joulemeter}\cite{jagroep2017energy}
+Joulemeter~\cite{kothari2009joulemeter,jagroep2017energy} is a Microsoft software that estimates the energy usage of Windows running applications down to the process level by using power models (for CPU, memory, and drives).
+It employs low-overhead power models to infer power consumption from resource utilization during runtime, and it provides power limiting features for virtual machines.
+Previous Joulemeter tests~\cite{jagroep2015profiling} demonstrated that the instrument provides a less accurate estimation of energy use that differs greatly from the real one.
+To adjust its models to the hardware on which it operates, Joulemeter must first go through a calibration step.
+It is able to only monitor one process at a time with a frequency of 1 HZ.
 
-jrapl \cite{guimaraes2016some} \cite{liu2015data}
+JRAPL is another example of an energy measurement tool estimating tool that has been utilized in a variety of publications\cite{guimaraes2016some,liu2015data}.
+This software enables for the energy usage of Java programs, functions, or even a block of code lines to be profiled and measured.
+The measurements are heavily reliant on the data supplied by RAPL.
+As a result, the global energy consumption collected by RAPL between two timestamps (the start and finish of the code to measure) is used to calculate the energy consumption of the Java code.
+Tests using JRAPL should be done on a well-configured machine to minimize the impact of the operating system and user operations on the overall energy consumption of JRAPL.
 
-joulinar \cite{islam2016measuring} \cite{noureddine2016jolinar}
+Another process-level energy usage measuring tool is Jolinar \cite{islam2016measuring,noureddine2016jolinar}.
+The tool does not need a calibration phase and relies on pre-established power models based on hardware metrics (TDP, disk I/O rate, and so on).
+These settings must be identified and supplied by the user for his machine.
+Jolinar is only capable of measuring the energy consumption of one application at a time.
+At the end of the execution, the tool provides the CPU, DRAM, and disk energy usage of the main process.
+joulinar
 
-jalen \cite{noureddine2015monitoring}
+Jalen\cite{noureddine2015monitoring} is another tool that profiles and monitors the energy usage of a Java program. Unlike JRAPL, Jalen is able to reduce the scope up to the function level.
+It gathers data using code instrumentation and statistical sampling at a predetermined pace.
+Because of the overhead that code instrumentation may incur, the authors recommend utilizing the second option.
+Every 10 ms, Jalen records the JVM's stack trace together with the CPU time of threads and computes statistics about method calls.
+These statistics are then utilized to calculate each method's energy usage.
 
 selfwatts \cite{fieni2021selfwatts}
+powerjoular \cite{noureddine2022powerjoular}
 
-powerjoular \cite{noureddine2022powerjoular}
 \subsection{hybrid tools}
@@
 
-simgrid simulation of power consumption of parallel aplication using simgrid \cite{heinrich2017predicting}
+Simgrid simulation of power consumption of parallel aplication using simgrid \cite{heinrich2017predicting}
 

### 2022-07-02 chapters/benchmarking/conclusion.tex
add : extention to JAVA
refactoring

@@
-\section{Extension}
-In computer science, things are changing fast, which leads to a raise of obsolete results. Specially when it comes to Comparison studies,One can't cover all the candidates. 
-Moreover, between the initial experiment and the published results there might be new candidates and an evolution of
-others. Therefore we Propose new creterioin for a successful experiment ~Extension~.
-In our content Extension means the ability to provide the necessary tools to not only Repeat the experiment but to be able to add extra candidates, Workloads or metrics.
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/benchmarkingprotocol}}
-    \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
-\end{figure}
-
 \section{Perspectives}

### 2022-07-02 chapters/benchmarking/draft.tex
refactoring

@@
 \section{Introduction}
-To optimize the software energy consumption, a reproducible setup environment requires to be designed.
+A reproducible setup environment needs to be established in order to maximize the software's potential to reduce energy consumption.
 
-This chapter will be dedicated to provide a set of guidelines to provide a \em{reproducible, accurate} and \em{representative} tests. Each aspect will be detailed in the sections below.
+This chapter will be devoted to giving a set of guidelines with the intention of delivering a test that is accurate, representative, and replicable.
+
+% This chapter will go over each component in further detail, describe it, and explain how it is used in the field of empirical analysis for software energy optimization.
 
@@
 
-A successful benchmark has three criterions to fullfil.
-First, it has to be \emph{reproducible} for others to replicate.
-Second, the results should be \emph{accurate}, which means each time we rerun the benchmark we are expecting the same results.
-Finally, it should \emph{represent} the real world.
-In other words, the conclusions brought from the experiment should be valid outside the experimentation as well.
-In my case, the real world is the production environment.
-Therefore, our experiments should reflect what is happening in the production environments.
-In this section, I will deeper dig in each aspect, present what has been done by others and my contributions, and finally I will propose some perspectives, to improve such experiments.
+A successful benchmark must meet three criteria.
+First, it must be \textbf{reproducible} in order for others to imitate it.
+Second, the findings should be \textbf{accurate},which implies that we should expect the same results each time we run the benchmark.
+Finally, it should \textbf{represent} reality.
+In other words, the experiment's findings should be applicable outside of the research lab as well.
+The aim of \textbf{representativeness} in this thesis is the manufacturing environment.
+As a result, the experiments should reflect what happens in production contexts.
 
@@
 \subsection{Reproducibility}
-One of the most difficult challenges faced by researchers is the reproducibility of their benchmarks.
-Actually, many results fail to be reproduced,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
+The reproducibility of experiments is often ranked as one of the most difficult difficulties faced by researchers.
+In point of fact, many of the outcomes cannot be duplicated.,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
 As this crisis hit most of the empirical studies, most of the reviews now includes reproducibility as one of the minimal standard for judging scientific merit.\cite{peng2011reproducible}
@@
 
-In addition to those benchmarks, a new category of benchmarking technique emerged to simulate the worst case of the production environments, inlcuding performance tests, which are benchmarks meant to evaluate the behavior of a software under stress situations.
-We can cite as an example the Gatling for web application and stress-ng for hardwares.%ROMAIN: Add citations.
-%% YEP I KNOW, THIS IS JUST FOR ME TO READ IT LATER AND GIVE A BETTER CONTEXT
-%% sorry caps ^^"
+In addition to these benchmarks, a new category of testing techniques has emerged to simulate the worst cases of the production environments.
+This new category of benchmarking called performance tests\cite{pradeep2019pragmatic}, which are benchmarks meant to evaluate the behavior of software under stress situations.
+We can use the Gatling\footnote{\url{https://gatling.io/}} as an illustration for web applications stresser and stress-ng for hardware heavy workload perormance measurements~\cite{king2017stress}.
+
+\subsection{Extension}
+Morover, The field of computer science is seeing rapid advancements, which has led to an increase in the number of obsolete results.
+Even more so, when it comes to studies that compare multiple solutions, it is impossible to cover them all.
+In addition, between the preliminary experiment and the results that were published, there may have been the appearance of new candidates as well as the development of others.
+As a result, we would like to provide a fresh idea for a productive experiment that we will call "Extension." The ability to give the required tools to not only repeat the experiment but also to add additional candidates, workloads, or key metrics.
+
+\subsection{conclusion}
+
+We propose making certain enhancements to the benchmarking framework that was suggested by the collaboratory on Experimental Evaluation of Software \footnote{\url{http://evaluate.inf.usi.ch/}}.
+and Systems in Computer Science in order to address those problems.
+Instead of only presenting the four primary aspects of their guidelines, which are \emph{measurement contexts} that indicate the software and hardware components that will alter or remain constant during the experiment.
+\emph{workloads} which identify the benchmarks to use in the experiment, as well as their inputs; \emph{metrics} that specify the attributes to be measured and how to assess them.
+\emph{Data analysis} show how to examine data and evaluate the outcomes of the analysis to offer insight into the assertions that arise from the study.
+
+\begin{itemize}
+    \item \textbf{candidates}: a set of candidates that will be used in the experiment. The candidates should be agnostic of the experiment context and they all
+\end{itemize}
+
+\begin{figure}%[!htb]
+    \center{\includegraphics[width=.9\linewidth]{imgs/benchmarkingprotocol}}
+    \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
+\end{figure}
 

### 2022-07-02 chapters/benchmarking/reproducibility.tex
add : python
refactoring

--- 2022-06-10 chapters/benchmarking/stateoftheart.tex
@@
 %  State of the art.
-%  TODO :
 
@@
 
-\paragraph{Studying Hardware Factors.}
-This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
-On the one hand, the correlation between the processor temperature and the energy consumption was one of the most explored paths.
-Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
-On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the CPU temperature and the energy consumption variation are tightly coupled.
+\section{Reproducibility in benchmarking }
 
-\note{add the corelation value}
+%% just a placeholder title for the moment 
+% introduction of vms
 
-This exposes the processor temperature as a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.
-
-The ambient temperature was also discussed in many papers as a candidate factor for the energy variation of a processor.
-In~\cite{ranka_energy_2009}, the authors claimed that energy consumption may vary due to fluctuations caused by the external environment.
-These fluctuations may alter the processor temperature and its energy consumption.
-However, the temperature inside a data center does not show major variations from one node to another.
-In~\cite{el_mehdi_diouri_your_2013}, El~Mehdi~Dirouri~\emph{et~al.} showed that switching the spot of two servers does not affect their energy consumption.
-Moreover, changing hardware components, such as the hard drive, the memory or even the power supply, does not affect the energy variation of a node, making it mainly related to the processor.
-This result was recently assessed by~\cite{wang_potential_2018}, where the rack placement and power supply introduced a maximum of $2.8\,\%$ variation in the observed energy consumption.
-
-Beyond hardware components, the accuracy of power meters has also been questioned.
-Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
-All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
-%% this is to prove that the problem is REAL !!!
-
-\paragraph{Mitigating Energy Variations.}
-Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
-In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
-The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
-Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
-They also proposed some guidelines to reduce this variation by replacing the old---slower---chips, by load balancing the workload on the CPU cores and leaving one core idle.
-They claimed that the variation between the processor cores is insignificant.
-In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.
-
-In~\cite{marathe_empirical_2017_m}, the authors highlight the increase of energy variation across the latest Intel micro-architectures by a factor of $4$ from Sandy~Bridge to Broadwell, a $15\,\%$ of run-to-run variation within the same processor and the increase of the inter-cores variation from $2.5\,\%$ to $5\,\%$ due to hardware-enforced constraints, concluding with some recommendations for Broadwell usage, such as running one hyper-thread per core.
-%  NOTE :/this might be interesting for greeFaaS
-
-\subsection*{Objective}
-While the state of the art aims to create a large umbrella that covers all empirical benchmarks related to computer science, we rather aim to narrow our study to energy benchmarking only.
-In other words, we will focus on pitfalls that hinders the energy claims of software.
-We are clearly aware of the impact of hardware on energy variations.
-However, we believe that there is still a room to control this energy variation for practitioners by using only parameters that can be tuned.
-To do so, we have inducted a set of empirical experiments using the guidelines provided by state of the art, to determine which controllable factors can reduce the variation of the energy consumption of benchmarks.
-We will first start with evaluating the parameters mentioned by \cite{heinrich_predicting}.
-
-% introduction of vms 
-\subsection{Virtualisation}
+\subsection{Virtual machines}
 To resolve this problem, practitioners tend to use Virtualisation.
@@
 
-%  basically just rephrase the paper
-
 \subsection{docker and accuracy}
-%% TODO : add an introduction, maybe some more details 
 And now Since the stat of the art has aggreed on the impact of docker on the energy consumption,Let's discuss it's impact on the accuracy. In other words\\

### 2022-07-02 chapters/literature/draft_citations.tex
refactoring

@@
 
-\section{Reproducibility in benchmarking }
-
-%% just a placeholder title for the moment 
-% introduction of vms
-
-\subsection{Virtual machines}
-To resolve this problem, practitioners tend to use Virtualisation.
-Using virtual machines aka VM gives reaserchers the freedom to choose their own tools, software and operating system that they are the most confortable with without paying the price to change the actual working environment, which will give them eventually more controle over the dependencies and the execution environement.
-Moreover, using a vm will solve the \emph{replication crisis} thanks to the virtual images, even the most complex architecture can be reproduced easily by just instantiating a copy of the image.
-Since the virtual machines are agnostic to the host architecture, reaserchers won't have to worry about where and how their experiments are replicated because they have already setup the execution environement.
-Another advantage to the virtual machines is the snapshot mechanism, it allows reaserchers to create backups and revert some changes with simple clicks.
-Last but not least,thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables -controled and uncontroled-  and do other analysis without dealing with any dependencies.
-In his paper \cite{howe_virtual_2012} bill howe lists the advantages of using virtual machines in reaserchers experements including the economical impact and cultural limitation to a such approach.
-
-which allow them to have control over the ressouces, the dependencies and the execution environment.
-Moreover, thanks to the snapshots, deploying a software is easily done by instantiating a copy of that image.
-However, this choice comes with a certain cost.
-Because the intervention of the hypervisor,the software will use two kernels , the virtual machine one and the host machine one, which will provide a noticiable overhead, and will impact the performances of the tests.Therefore, we can't use virtual machines for exepements that are related to performance.
-Another limiation with the virtual machine is the the isolation.
-It is true that this feature will prevent the experience environement with any undesirable interference from the outside world.
-but sometimes this contact is needed, especially when the experiment is dependent to an external part, such as sensors.
-In energicial tests we tend to use hardware powermeters which will make it difficult to use the virtual machines in this case .
-
-% introduction of docker
-
-\begin{figure}
-    \center{\includegraphics[width=1\linewidth]{imgs/virtualization_techniques}}
-    \caption{Different Methods of Virtualzation}\label{environement:virtualization_technique}
-\end{figure}
-
-\subsection{Containers}
-
-another solution would be using something that allows us to have the isolation from the host os and the ease of replication that virtual machines offer, and the direct interaction with the hardware that the classical method give.
-contarization offers a such advantages while keeping the isolation and the ease of replication for application.
-
-Figure \ref{environement:virtualization_technique} explains the differents in architecture between the clasic types2 of Virtualisation and Containers.
-\begin{itemize}
-    \item Type 1: runs directly on the hardware, it is mainly used by the cloud providers where there is no main OS, but just virtaul machines, we can site for this the open-srouce XEN and VMware ESX
-    \item Type 2: runs over the hostmachine Operating System, mostly used for personal comptuers, VMware server and virtualBox are famous examples of this type, most of the reaserchers expermentation are run witht this type, howerver due to the 2 Operatings syestms the applications tend to be more slower
-    \item containers : Instead of its own kernel, containers used the hots kernel to run their Os, which makes them ligher, quickers and use the full pontial use of the hardware.
-For this we can cite \emph{Docker}, \emph{Linux LXC}\emph{LXD} \cite{abuabdo_virtualization_2019}
-\end{itemize}
-%  TODO Rephrase so i include it before VMs+ add links to the technologie
-
-\subsection{Docker vs Virtual Machine}
-Depsite that Type 1 is more performant than type 2, the second one is the most used in reaserch, since most researchers conducts their experements in their own machine.
-In the other hand, docker is the most famous thechonology for for containers.
-In ower case we are more prone to docker for two reasons.
-\begin{enumerate}
-    \item we need a litetweight orchestrator to not affect the energy consumption of ower tests. As prior work mentioned [cite Morabito (2015) and van Kessel et al. (2016)]
-          % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
-    \item since we are using the hardware itself to measure the energy consumption, we are required to interract with the host OS itself.
-\end{enumerate}
-Special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that allows us to retireve the energy consumption of a virtual machine.
-%%%%% THINGS TO ADD : 
-%% -------------- Dont think so 
-% the impact of docker on the reprodudicibility 
-% Limits of docker 
-% The docker files and the clarification of the methodology 
-\subsection{Docker and energy}
-Now that we have chosen to go with the containers technorolgy to encapsulate our tests. What would be the impact of this solution on the energy consumption of our tests.
-
-Based on the studies of \cite{santos2018does}, who analysed the impact of adding the docker layer on the energy consumption.
-In their experement. Eddie Antonio et al run multiple benchmarks with and without docker. and compared their energy consumption and execution time.
-The first step was to see the impact of docker deamon while there is no work. to see the impact of the orchestrator alone. Later they had the experiment with the following benchmarks
-\begin{itemize}
-    \item wordpress
-    \item reddis
-    \item postfresSQL
-\end{itemize}
-The following figures represents the energy consumption of the system while it is idle. As we can see  in figure \ref{fig:docker_idle} Docker braught arround 1000joules overhead.
-\begin{figure}
-    \center{\includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/idle_energy}}
-    \caption{energy consumption of Idle system with and without docker \cite{santos2018does}}\label{fig:docker_idle}
-\end{figure}
-In the other hand as we can see in figure \ref{fig:docker_reddis}.
-docker increased the execution time of the benchmark by 50 seconds which caused an increase in energy since they are highly correlated.
-The authors also highlated the fact that this increase of energy consumption is due to the docker deamon and not the fact that the application is in a container.
-Moreover they estimted the price of this extra energy and it was less than 0.15\$ in the worst case.
-Which is non significant compared to the advantages that docker bring for isolation and reproducibility.\\
-if we  recap this study in one sentence,it would be the following one.
-The dockerised softwares tend to consume more energy, because mainly they take more time to be executed.
-The average power consumption is higher with only \textbf{2Watts} and it is due to the docker deamon.This overhead can be up to 5\% for IO intensive application, but it is mearly noticiable when it comes to CPU or DRMA intesive works
-
-\begin{figure}
-    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_time}
-    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_energy}
-    \caption{execution time and energy consumption of Redis  with and without docker \cite{santos2018does}}\label{fig:docker_reddis}
-\end{figure}
-
 %%%%%%%%%%%%% energy metigation

### 2022-07-02 chapters/benchmarking/draft.tex
update

@@
 \section{Introduction}
+Since the objective of this thesis is to study various alternatives for reducing the energy consumption of software programs, this chapter will focus on this topic.
+We shall conduct extensive empirical research.
+This chapter describes the instruments required to conduct a successful experiment to measure the energy consumption of software applications.
+In addition, the majority of the time we will conduct an analytical comparison; therefore, the goal is to develop a framework to assist practitioners in comparing various solutions in terms of energy consumption.
+
 A reproducible setup environment needs to be established in order to maximize the software's potential to reduce energy consumption.
@@
 % This chapter will go over each component in further detail, describe it, and explain how it is used in the field of empirical analysis for software energy optimization.
-
-\section{Threats and Challenges}
-
-A successful benchmark must meet three criteria.
-First, it must be \textbf{reproducible} in order for others to imitate it.
-Second, the findings should be \textbf{accurate},which implies that we should expect the same results each time we run the benchmark.
-Finally, it should \textbf{represent} reality.
-In other words, the experiment's findings should be applicable outside of the research lab as well.
-The aim of \textbf{representativeness} in this thesis is the manufacturing environment.
-As a result, the experiments should reflect what happens in production contexts.
 
@@
 
-\subsection{Reproducibility}
-The reproducibility of experiments is often ranked as one of the most difficult difficulties faced by researchers.
-In point of fact, many of the outcomes cannot be duplicated.,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
-As this crisis hit most of the empirical studies, most of the reviews now includes reproducibility as one of the minimal standard for judging scientific merit.\cite{peng2011reproducible}
-One of the criterions supporting reproducibility is the publication of the dataset and the algorithms run on the raw data to conclude the results.
-There is even some disagreement about what the terms "reproducibility" or "replicability" by themselves mean \cite{goodman2016does}.
-According to \cite{echtler2018open}, \emph{replicability} extends \emph{reproducibility} with the ability to collect a new raw dataset comparable to the original one by re-executing the experiment under similar condition, instead of just the ability to get the same results by running the statistical analyses on the original data set.
+%%%% part of the state of the art 
+% In the area covered by this PhD thesis, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
+%%% why it won't work for energy consumption 
+% However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only among homogeneous nodes, but even within a single node.
+%% reseasons why 
+% One major problem that hinders the reproducibility of the empirical benchmarks is the interaction with the external environment, either as concurrency or dependencies.
+% Therefore, researchers cannot observe the same results, unless they duplicate the same environment.
 
-In the area covered by this PhD thesis, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
-However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only among homogeneous nodes, but even within a single node.
-
-One major problem that hinders the reproducibility of the empirical benchmarks is the interaction with the external environment, either as concurrency or dependencies.
-Therefore, researchers cannot observe the same results, unless they duplicate the same environment.
-
-\subsection{Accuracy}
-According to Oxford, \emph{accuracy} means \href{https://www.lexico.com/definition/accuracy}{"technical The degree to which the result of a measurement, calculation, or specification conforms to the correct value or a standard"}.
-In ny case, this means the ability to run the benchmark multiple times with a low variation.
-
-Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.
-
+%%%% this is relative about my thesis 
 In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of physical measurements.
@@
 \end{figure}
-
+%% state of the art 
 Some researchers started investigating the hardware impact of the energy variation of power consumption.
@@
 \cite{heinrich_predicting} described this variation as a set of parameters, such as CPU Frequency and the thermal effect.
-
-\subsection{Representativeness}
-As obvious as it seems, the reason of executing benchmarks is to validate ideas so we can use them in the real life.
-However, this is means that those benchmarks have to represent reality somehow.
-Basically, when we aim to benchmark something, we create a mock up version of the situation that we want it to work.
-% But, how can we assure that the benchmark is representative? honestly i don't know. and we can't generalize this. but there are some works that have been done for this.
-First, we can talk about the benchmarking and their selection, then we will talk about the stress benchmark for some applications, and finally we will bring this representativeness in our case and how can we get closer to the energy consumption behavior of a software between the benchmark environnement and the production one.
 
@@
 To answer those questions, the community agreed on some wellknown benchmarks to represent a specific concern of the production world.
-One can site as an example the Dacapo and Renaissance benchmark suites for Java applications, or the CLBG benchmark suite for comparing programming languages.
+One can site as an example the Dacapo~\cite{DaCapo:paper} and Renaissance~\cite{renaissance} benchmark suites for Java applications, or the CLBG benchmark suite for comparing programming languages\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}.
 Although they do not cover all the cases, the community agrees on their relevance and representativeness.
@@
 We can use the Gatling\footnote{\url{https://gatling.io/}} as an illustration for web applications stresser and stress-ng for hardware heavy workload perormance measurements~\cite{king2017stress}.
+
+%%%% THIS Should the introduction of my chapter
 
@@
 
-\subsection{conclusion}
+%% how to extend
 

### 2022-07-02 chapters/literature/draft_citations.tex
update

@@
 This section will go through the flaws and best practices of empirical research before applying them to our topic of study.
+\subsection{Threats and Challenges}
+
+A successful benchmark must meet three criteria.
+First, it must be \textbf{reproducible} in order for others to imitate it.
+Second, the findings should be \textbf{accurate},which implies that we should expect the same results each time we run the benchmark.
+Finally, it should \textbf{represent} reality.
+In other words, the experiment's findings should be applicable outside of the research lab as well.
+The aim of \textbf{representativeness} in this thesis is the manufacturing environment.
+As a result, the experiments should reflect what happens in production contexts.
+
+\subsubsection{Reproducibility}
+The reproducibility of experiments is often ranked as one of the most difficult difficulties faced by researchers.
+In point of fact, many of the outcomes cannot be duplicated.,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
+As this crisis hit most of the empirical studies, most of the reviews now includes reproducibility as one of the minimal standard for judging scientific merit.\cite{peng2011reproducible}
+One of the criterions supporting reproducibility is the publication of the dataset and the algorithms run on the raw data to conclude the results.
+There is even some disagreement about what the terms "reproducibility" or "replicability" by themselves mean \cite{goodman2016does}.
+According to \cite{echtler2018open}, \emph{replicability} extends \emph{reproducibility} with the ability to collect a new raw dataset comparable to the original one by re-executing the experiment under similar condition, instead of just the ability to get the same results by running the statistical analyses on the original data set.
+\subsubsection{Accuracy}
+According to Oxford, \emph{accuracy} means \href{https://www.lexico.com/definition/accuracy}{"technical The degree to which the result of a measurement, calculation, or specification conforms to the correct value or a standard"}.
+In ny case, this means the ability to run the benchmark multiple times with a low variation.
+
+\subsubsection{Representativeness}
+As obvious as it seems, the reason of executing benchmarks is to validate ideas so we can use them in the real life.
+However, this is means that those benchmarks have to represent reality somehow.
+Basically, when we aim to benchmark something, we create a mock up version of the situation that we want it to work.
+% But, how can we assure that the benchmark is representative? honestly i don't know. and we can't generalize this. but there are some works that have been done for this.
+First, we can talk about the benchmarking and their selection, then we will talk about the stress benchmark for some applications, and finally we will bring this representativeness in our case and how can we get closer to the energy consumption behavior of a software between the benchmark environnement and the production one.
+Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.
 
@@
 % that serves as a basis for evaluation or comparison (as of computer system performance)
+\subsubsection{the impact of these challenges in the empirical research }
 In their work~\cite{van_der_kouwe_benchmarking_2018}, \citeauthor{van_der_kouwe_benchmarking_2018} Investigated 50 papers published in top venues
@@
 % \end{itemize}
+
+\subsection{proposed solutions}
 

### 2022-07-02 chapters/literature/draft_citations.tex
add : extra references for the state of the art

@@
 \subsubsection{Reproducibility}
-The reproducibility of experiments is often ranked as one of the most difficult difficulties faced by researchers.
+The reproducibility of experiments is often ranked as one of the most difficult difficulties faced by researchers .This issue has been one of the major tools science has used to help establish the validity and importance of scientific findings since the Philosophical Transactions of the Royal Society were established in 1665~\cite{hankins1986debate} .
 In point of fact, many of the outcomes cannot be duplicated.,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
@@
 To analyze the magnitude of the phenomenon, they have identified a set of 22 "benchmarking crimes" that threaten the validity of the system.
+
 % TO [ROMAIN] should i include  this list ? 
@@
 They treated this problem by proposing two strategies for detecting measurement bias by using causal analysis and preventing it with setup randomization \cite{mytkowicz2009producing}.
-\citeauthor{bukh1992art} wrote a book about computer performance analysis, where he discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems .
 
-% further references
+\citeauthor{bukh1992art} wrote a book about computer performance analysis, where he discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems.
 
-%%%%%%%%%% 
-In their paper~\cite{buytaert_statistically_nodate},\citeauthor{buytaert_statistically_nodate} they separated the start-up measurement from the steady state when comparing different Java optimizations.
-\\
-%\\\ https://www.cpubenchmark.net/compare/Intel-i7-12700KF-vs-Intel-i7-12700F/4608vs4692 | hardware benchmarking 
-\section{energy measurement}\label{sec:energymeasurement}
+the article \cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistake in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often seen errors.
 
+another article~\cite{patil2018training} expands on this line of thought by evaluating the uncertainty caused by replications in new research.
+They offer strategies for capturing uncertainty in inferential investigations, such as cross-study validation and ensemble models.
+
+In their paper\cite{stodden2018empirical}, the authors find that, even though it's a big move in the right direction, journal policies that require authors to give back digital scholarly objects after publication, like the data and code that back up the claims, don't get more than half of these objects back.
+Then, using these artifacts, about a fourth of the published computational claims in the study could be made.
+They suggest putting out the claims in the literature and the digital scholarly objects that back them up at the same time.
+
+% as for computer sience. In their paper~\cite{buytaert_statistically_nodate}, the authors  separated the start-up measurement from the steady state when comparing different Java optimizations.
+
+\begin{figure}[!hbt]
+    \center{\includegraphics[scale=0.3]{imgs/componentsofanexperiment}}
+    \caption{The proposed decomposition of an experement\footnote{\url{http://evaluate.inf.usi.ch/}}}\label{fig:soa_expermiment component}
+\end{figure}
+
+\section{energy measurement}
 to be able to reduce the energy cost of running programs we should first be able to estame this energy consumption.

### 2022-07-02 chapters/benchmarking/draft.tex
fix : bug where i cant add references

@@
 In addition, the majority of the time we will conduct an analytical comparison; therefore, the goal is to develop a framework to assist practitioners in comparing various solutions in terms of energy consumption.
+
+First we will project the challenges of benchmarking to the case of energy consumption and see when it deffers from the previous contexts.
+then we will propose a framework to deal with those challenges.
+later we will elaborate each part of the framework and how to solve certain of these challenges.
 
@@
 % Therefore, researchers cannot observe the same results, unless they duplicate the same environment.
+
+%%% FOR METRICS AND ACCURACY
 
@@
 % However, using and comparing identical CPU is still tricky and is very dependent to chips.
+
+\section{Reproducibility in benchmarking }
+
+%% just a placeholder title for the moment 
+% introduction of vms
+
+\subsection{Virtual machines}
+To resolve this problem, practitioners tend to use Virtualisation.
+Using virtual machines aka VM gives reaserchers the freedom to choose their own tools, software and operating system that they are the most confortable with without paying the price to change the actual working environment, which will give them eventually more controle over the dependencies and the execution environement.
+Moreover, using a vm will solve the \emph{replication crisis} thanks to the virtual images, even the most complex architecture can be reproduced easily by just instantiating a copy of the image.
+Since the virtual machines are agnostic to the host architecture, reaserchers won't have to worry about where and how their experiments are replicated because they have already setup the execution environement.
+Another advantage to the virtual machines is the snapshot mechanism, it allows reaserchers to create backups and revert some changes with simple clicks.
+Last but not least,thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables -controled and uncontroled-  and do other analysis without dealing with any dependencies.
+In his paper \cite{howe_virtual_2012} bill howe lists the advantages of using virtual machines in reaserchers experements including the economical impact and cultural limitation to a such approach.
+
+which allow them to have control over the ressouces, the dependencies and the execution environment.
+Moreover, thanks to the snapshots, deploying a software is easily done by instantiating a copy of that image.
+However, this choice comes with a certain cost.
+Because the intervention of the hypervisor,the software will use two kernels , the virtual machine one and the host machine one, which will provide a noticiable overhead, and will impact the performances of the tests.Therefore, we can't use virtual machines for exepements that are related to performance.
+Another limiation with the virtual machine is the the isolation.
+It is true that this feature will prevent the experience environement with any undesirable interference from the outside world.
+but sometimes this contact is needed, especially when the experiment is dependent to an external part, such as sensors.
+In energicial tests we tend to use hardware powermeters which will make it difficult to use the virtual machines in this case .
+
+% introduction of docker
+
+\begin{figure}
+    \center{\includegraphics[width=1\linewidth]{imgs/virtualization_techniques}}
+    \caption{Different Methods of Virtualzation}\label{environement:virtualization_technique}
+\end{figure}
+
+\subsection{Containers}
+
+another solution would be using something that allows us to have the isolation from the host os and the ease of replication that virtual machines offer, and the direct interaction with the hardware that the classical method give.
+contarization offers a such advantages while keeping the isolation and the ease of replication for application.
+
+Figure \ref{environement:virtualization_technique} explains the differents in architecture between the clasic types2 of Virtualisation and Containers.
+\begin{itemize}
+    \item Type 1: runs directly on the hardware, it is mainly used by the cloud providers where there is no main OS, but just virtaul machines, we can site for this the open-srouce XEN and VMware ESX
+    \item Type 2: runs over the hostmachine Operating System, mostly used for personal comptuers, VMware server and virtualBox are famous examples of this type, most of the reaserchers expermentation are run witht this type, howerver due to the 2 Operatings syestms the applications tend to be more slower
+    \item containers : Instead of its own kernel, containers used the hots kernel to run their Os, which makes them ligher, quickers and use the full pontial use of the hardware.
+For this we can cite \emph{Docker}, \emph{Linux LXC}\emph{LXD} \cite{abuabdo_virtualization_2019}
+\end{itemize}
+%  TODO Rephrase so i include it before VMs+ add links to the technologie
+
+\subsection{Docker vs Virtual Machine}
+Depsite that Type 1 is more performant than type 2, the second one is the most used in reaserch, since most researchers conducts their experements in their own machine.
+In the other hand, docker is the most famous thechonology for for containers.
+In ower case we are more prone to docker for two reasons.
+\begin{enumerate}
+    \item we need a litetweight orchestrator to not affect the energy consumption of ower tests. As prior work mentioned [cite Morabito (2015) and van Kessel et al. (2016)]
+          % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
+    \item since we are using the hardware itself to measure the energy consumption, we are required to interract with the host OS itself.
+\end{enumerate}
+Special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that allows us to retireve the energy consumption of a virtual machine.
+%%%%% THINGS TO ADD : 
+%% -------------- Dont think so 
+% the impact of docker on the reprodudicibility 
+% Limits of docker 
+% The docker files and the clarification of the methodology 
+\subsection{Docker and energy}
+Now that we have chosen to go with the containers technorolgy to encapsulate our tests. What would be the impact of this solution on the energy consumption of our tests.
+
+Based on the studies of \cite{santos2018does}, who analysed the impact of adding the docker layer on the energy consumption.
+In their experement. Eddie Antonio et al run multiple benchmarks with and without docker. and compared their energy consumption and execution time.
+The first step was to see the impact of docker deamon while there is no work. to see the impact of the orchestrator alone. Later they had the experiment with the following benchmarks
+\begin{itemize}
+    \item wordpress
+    \item reddis
+    \item postfresSQL
+\end{itemize}
+The following figures represents the energy consumption of the system while it is idle. As we can see  in figure \ref{fig:docker_idle} Docker braught arround 1000joules overhead.
+\begin{figure}
+    \center{\includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/idle_energy}}
+    \caption{energy consumption of Idle system with and without docker \cite{santos2018does}}\label{fig:docker_idle}
+\end{figure}
+In the other hand as we can see in figure \ref{fig:docker_reddis}.
+docker increased the execution time of the benchmark by 50 seconds which caused an increase in energy since they are highly correlated.
+The authors also highlated the fact that this increase of energy consumption is due to the docker deamon and not the fact that the application is in a container.
+Moreover they estimted the price of this extra energy and it was less than 0.15\$ in the worst case.
+Which is non significant compared to the advantages that docker bring for isolation and reproducibility.\\
+if we  recap this study in one sentence,it would be the following one.
+The dockerised softwares tend to consume more energy, because mainly they take more time to be executed.
+The average power consumption is higher with only \textbf{2Watts} and it is due to the docker deamon.This overhead can be up to 5\% for IO intensive application, but it is mearly noticiable when it comes to CPU or DRMA intesive works
+
+\begin{figure}
+    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_time}
+    \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_energy}
+    \caption{execution time and energy consumption of Redis  with and without docker \cite{santos2018does}}\label{fig:docker_reddis}
+\end{figure}
+
+\subsection{docker and accuracy}
+And now Since the stat of the art has aggreed on the impact of docker on the energy consumption,Let's discuss it's impact on the accuracy. In other words\\
+\textbf{RQ :} does Docker affect the energy variation of the exepements ?
+
+To Answer this question we have conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
+Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
+In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
+The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
+
+\begin{figure}
+    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
+    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
+\end{figure}

### 2022-07-02 chapters/benchmarking/reproducibility.tex
update
fix : bug where i cant add references

--- 2022-06-11 chapters/benchmarking/reproducibility.tex
@@
 
-%  basically just rephrase the paper
-
-%% add this to my contribution 
-\subsection{docker and accuracy}
-And now Since the stat of the art has aggreed on the impact of docker on the energy consumption,Let's discuss it's impact on the accuracy. In other words\\
-\textbf{RQ :} does Docker affect the energy variation of the exepements ?
-
-To Answer this question we have conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
-Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
-In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
-The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
-    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
-\end{figure}
+%% this is might be the part that treats reproducibility for my case

### 2022-07-03 chapters/benchmarking/draft.tex
add : soa

@@
 \section{Introduction}
-Since the objective of this thesis is to study various alternatives for reducing the energy consumption of software programs, this chapter will focus on this topic.
-We shall conduct extensive empirical research.
-This chapter describes the instruments required to conduct a successful experiment to measure the energy consumption of software applications.
+Because the goal of this thesis is to investigate various possibilities for improving the energy consumption of software programs, this chapter will be focusing on how to conduct effective experiments when comparing the energy consumption of softwares.
+We will carry out substantial empirical investigation. This chapter discusses the tools needed to do a successful experiment measuring the energy consumption of software programs.
 In addition, the majority of the time we will conduct an analytical comparison; therefore, the goal is to develop a framework to assist practitioners in comparing various solutions in terms of energy consumption.
@@
 and Systems in Computer Science in order to address those problems.
-Instead of only presenting the four primary aspects of their guidelines, which are \emph{measurement contexts} that indicate the software and hardware components that will alter or remain constant during the experiment.
-\emph{workloads} which identify the benchmarks to use in the experiment, as well as their inputs; \emph{metrics} that specify the attributes to be measured and how to assess them.
-\emph{Data analysis} show how to examine data and evaluate the outcomes of the analysis to offer insight into the assertions that arise from the study.
+Instead of only presenting the four primary aspects of their guidelines, which are
 

### 2022-07-03 chapters/literature/draft_citations.tex
fix : bug where i cant add references
add : soa

@@
 % as for computer sience. In their paper~\cite{buytaert_statistically_nodate}, the authors  separated the start-up measurement from the steady state when comparing different Java optimizations.
+Another work, to make unify the bemchmarking methodology across different research works in the field of computer science, we can cite the paper~\cite{stephen_evaluate_2012}.
+In their approach they  divided any emprical experiment into four  components:
+
+\begin{enumerate}
+    \item \emph{measurement contexts} that indicate the software and hardware components that will alter or remain constant during the experiment.
+    \item \emph{workloads} which identify the benchmarks to use in the experiment, as well as their inputs; 
+    \item \emph{metrics} that specify the attributes to be measured and how to assess them. 
+    \item \emph{Data analysis} show how to examine data and evaluate the outcomes of the analysis to offer insight into the assertions that arise from the study.
+\end{enumerate}
+
+Figure \ref{fig:soa_experiment_component} shows the four component presented in the article~~\cite{stephen_evaluate_2012}
 
@@
     \center{\includegraphics[scale=0.3]{imgs/componentsofanexperiment}}
-    \caption{The proposed decomposition of an experement\footnote{\url{http://evaluate.inf.usi.ch/}}}\label{fig:soa_expermiment component}
+    \caption{The proposed decomposition of an experement~\cite{stephen_evaluate_2012}}\label{fig:soa_expermiment_component}
 \end{figure}

### 2022-07-03 chapters/benchmarking/draft.tex
update : benchmarking protocl

@@
 \section{Introduction}
-Because the goal of this thesis is to investigate various possibilities for improving the energy consumption of software programs, this chapter will be focusing on how to conduct effective experiments when comparing the energy consumption of softwares.
-We will carry out substantial empirical investigation. This chapter discusses the tools needed to do a successful experiment measuring the energy consumption of software programs.
-In addition, the majority of the time we will conduct an analytical comparison; therefore, the goal is to develop a framework to assist practitioners in comparing various solutions in terms of energy consumption.
+We are conducting empirical tests to determine the effect of programmer choice on software's energy consumption.
+To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state of the art.
+Moreover, when it comes to measuring the energy consumption of software, other obstacles arise, such as the impact of hardware, the difference between the machine used in the research and the machines that may apply the findings, etc.
+Moreover, significant improvements in computer science have resulted in an increase in the quantity of obsolete results.
+It is even more difficult to address all of them in studies that compare multiple solutions.
+Furthermore, there may have been the appearance of new candidates as well as the development of others between the preliminary experiment and the published results.
+As a result, it is not only required to ensure the "reproducibility" of the results, but also to make way for new candidates.
+As a result, we'd like to propose a new concept for a fruitful experiment, which we'll call "Extension." The capacity to provide the tools required not only to repeat the experiment but also to add additional candidates, workloads, or crucial Key metrics.
 
-First we will project the challenges of benchmarking to the case of energy consumption and see when it deffers from the previous contexts.
-then we will propose a framework to deal with those challenges.
-later we will elaborate each part of the framework and how to solve certain of these challenges.
+This chapter discusses how to adapt the state-of-the-art of empirical analysis in the energy consumption field, highlighting the difficulties and the various techniques to overcome them.
 
-A reproducible setup environment needs to be established in order to maximize the software's potential to reduce energy consumption.
-
-This chapter will be devoted to giving a set of guidelines with the intention of delivering a test that is accurate, representative, and replicable.
-
-% This chapter will go over each component in further detail, describe it, and explain how it is used in the field of empirical analysis for software energy optimization.
+%% this is might be the part that treats reproducibility for my case
 
@@
 
-%%% FOR METRICS AND ACCURACY
+\subsection{Extension}
 
-%%%% this is relative about my thesis 
-In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of physical measurements.
-Unfortunately, this is not possible when it comes to measuring the energy consumption of a system.
-Applying the benchmarking guidelines and repeating the same experiment with in the same configuration are not sufficient to reproduce the the same energy measurements, not only between identical machines, but even within the same machine.
-This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
+We propose making certain enhancements to the benchmarking framework that was suggested by the collaboratory on Experimental Evaluation of Software and Systems in Computer Science in order to address those problems \footnote{\url{http://evaluate.inf.usi.ch/}}.
+Instead of only presenting the four primary aspects of their guidelines that were mentioned in state of the art\cref{sec:soa_reproducibility} we propose an abstract model to represent the experiment, 
+\emph{context} : the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor .
+An \emph{orchestrator} which its job to run the experiment. 
+The orchestrator provides three intrafaces :
 
-Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
-We can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
+Workload interface  : Wich provides a set of functions that should be implemented by the workload  
+Candidate interface : 
+Observer interface : provides a set of metrics that are collected by the orchestrator during the expermentation.
+note that some of those metrics are external, such the energy consumption of the system , execution time, execution stack ..etc .
 
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
-    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
-\end{figure}
-%% state of the art 
-Some researchers started investigating the hardware impact of the energy variation of power consumption.
-As an example, one can cite~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the main cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
-\cite{heinrich_predicting} described this variation as a set of parameters, such as CPU Frequency and the thermal effect.
+% \begin{itemize}
+%     \item textbf{expermimental protocol}: Which describes, the context and  an orchestrator that handles the experiment, this orchestrator should take as an entry 
+%     \item \textbf{candidates}: a set of candidates that will be used in the experiment. The candidates should be agnostic of the experiment context and they all
+% \end{itemize}
 
-As Stephen M.
-Blackburn~\emph{et~al.} cited in their paper "evaluate collaboratory"~\cite{stephen_evaluate_2012}, one of the major pitfalls of the measurement contexts is the inconsistency, which can be translated here by the fact that the production context is not the same as the benchmarking one.
+% \begin{figure}%[!htb]
+%     \center{\includegraphics[width=.9\linewidth]{imgs/benchmarkingprotocol}}
+%     \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
+% \end{figure}
 
-Another difficult part for practitioners is to generalize the claims they reached beyond the lab conditions.
-Are they appropriate?
-Are they consistent and are they reproducible?
-To answer those questions, the community agreed on some wellknown benchmarks to represent a specific concern of the production world.
-One can site as an example the Dacapo~\cite{DaCapo:paper} and Renaissance~\cite{renaissance} benchmark suites for Java applications, or the CLBG benchmark suite for comparing programming languages\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}.
-Although they do not cover all the cases, the community agrees on their relevance and representativeness.
-
-In addition to these benchmarks, a new category of testing techniques has emerged to simulate the worst cases of the production environments.
-This new category of benchmarking called performance tests\cite{pradeep2019pragmatic}, which are benchmarks meant to evaluate the behavior of software under stress situations.
-We can use the Gatling\footnote{\url{https://gatling.io/}} as an illustration for web applications stresser and stress-ng for hardware heavy workload perormance measurements~\cite{king2017stress}.
-
-%%%% THIS Should the introduction of my chapter
-
-\subsection{Extension}
-Morover, The field of computer science is seeing rapid advancements, which has led to an increase in the number of obsolete results.
-Even more so, when it comes to studies that compare multiple solutions, it is impossible to cover them all.
-In addition, between the preliminary experiment and the results that were published, there may have been the appearance of new candidates as well as the development of others.
-As a result, we would like to provide a fresh idea for a productive experiment that we will call "Extension." The ability to give the required tools to not only repeat the experiment but also to add additional candidates, workloads, or key metrics.
-
-%% how to extend
-
-We propose making certain enhancements to the benchmarking framework that was suggested by the collaboratory on Experimental Evaluation of Software \footnote{\url{http://evaluate.inf.usi.ch/}}.
-and Systems in Computer Science in order to address those problems.
-Instead of only presenting the four primary aspects of their guidelines, which are
-
-\begin{itemize}
-    \item \textbf{candidates}: a set of candidates that will be used in the experiment. The candidates should be agnostic of the experiment context and they all
-\end{itemize}
-
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/benchmarkingprotocol}}
-    \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
-\end{figure}
-
-% A more subtle issue may arise due to the values of the measurements that we achieved.
-% In fact, the energy measures are quite small, and iterations may be taking a few milliseconds more or less to run.
-% A thing we cannot measure using our measurement tools.
-% How generalizable are our results? As a set of energy variation optimization guidelines, we argue that our results applied on most of the modern Intel CPU.
-% However, using and comparing identical CPU is still tricky and is very dependent to chips.
+later in this chapter we will discuss each aspect of those interfaces
 
 \section{Reproducibility in benchmarking }
-
-%% just a placeholder title for the moment 
-% introduction of vms
+As we have discussed in the previuous section , the candidates shouldn't be related to the experiment, all they need to know is the interface of the workload, therefore.
+This section will discuss different ways to ensure a reproducibility of the experiment by comparing different ways to encapsulate the candidates
 
 \subsection{Virtual machines}
-To resolve this problem, practitioners tend to use Virtualisation.
-Using virtual machines aka VM gives reaserchers the freedom to choose their own tools, software and operating system that they are the most confortable with without paying the price to change the actual working environment, which will give them eventually more controle over the dependencies and the execution environement.
+First option should be using virtual machines,  Using virtual machines aka VM gives reaserchers the freedom to choose their own tools, software and operating system that they are the most confortable with without paying the price to change the actual working environment, which will give them eventually more controle over the dependencies and the execution environement.
 Moreover, using a vm will solve the \emph{replication crisis} thanks to the virtual images, even the most complex architecture can be reproduced easily by just instantiating a copy of the image.
@@
 In his paper \cite{howe_virtual_2012} bill howe lists the advantages of using virtual machines in reaserchers experements including the economical impact and cultural limitation to a such approach.
-
 which allow them to have control over the ressouces, the dependencies and the execution environment.

### 2022-07-03 chapters/literature/draft_citations.tex
update : benchmarking protocl

@@
 
-\subsection{proposed solutions}
+\subsection{proposed solutions}\label[]{sec:soa_reproducibility}
 
@@
 \end{figure}
+
+%%%% TODO : Elaborate this section and check what should be added 
+% As Stephen M.
+Blackburn~\emph{et~al.} cited in their paper "evaluate collaboratory"~\cite{stephen_evaluate_2012}, one of the major pitfalls of the measurement contexts is the inconsistency, which can be translated here by the fact that the production context is not the same as the benchmarking one.
+
+% Another difficult part for practitioners is to generalize the claims they reached beyond the lab conditions.
+% Are they appropriate?
+% Are they consistent and are they reproducible?
+% To answer those questions, the community agreed on some wellknown benchmarks to represent a specific concern of the production world.
+% One can site as an example the Dacapo~\cite{DaCapo:paper} and Renaissance~\cite{renaissance} benchmark suites for Java applications, or the CLBG benchmark suite for comparing programming languages\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}.
+% Although they do not cover all the cases, the community agrees on their relevance and representativeness.
+
+% In addition to these benchmarks, a new category of testing techniques has emerged to simulate the worst cases of the production environments.
+This new category of benchmarking called performance tests\cite{pradeep2019pragmatic}, which are benchmarks meant to evaluate the behavior of software under stress situations.
+We can use the Gatling\footnote{\url{https://gatling.io/}} as an illustration for web applications stresser and stress-ng for hardware heavy workload perormance measurements~\cite{king2017stress}.
 
@@
 \section{Energy Variation}
+In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of physical measurements.
+Unfortunately, this is not possible when it comes to measuring the energy consumption of a system.
+Applying the benchmarking guidelines and repeating the same experiment with in the same configuration are not sufficient to reproduce the the same energy measurements, not only between identical machines, but even within the same machine.
+This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
+
+Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
+We can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
+
+\begin{figure}%[!htb]
+    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
+    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
+\end{figure}
+%% state of the art 
+Some researchers started investigating the hardware impact of the energy variation of power consumption.
+As an example, one can cite~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the main cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
+\cite{heinrich_predicting} described this variation as a set of parameters, such as CPU Frequency and the thermal effect.
 \subsection{Studying Hardware Factors}

### 2022-07-04 chapters/benchmarking/draft.tex
proofreading

@@
-\section{Introduction}
-We are conducting empirical tests to determine the effect of programmer choice on software's energy consumption.
+\section{Motivation}
+Empirical measurements are critical tasks to capture the effect of developers' choice on software energy consumption.
 To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state of the art.
-Moreover, when it comes to measuring the energy consumption of software, other obstacles arise, such as the impact of hardware, the difference between the machine used in the research and the machines that may apply the findings, etc.
+Moreover, when it comes to measuring the energy consumption of software, other obstacles arise, such as the impact of hardware, the difference between the machines used for measurements purpose and the ones that may apply the findings, etc.
 Moreover, significant improvements in computer science have resulted in an increase in the quantity of obsolete results.
 It is even more difficult to address all of them in studies that compare multiple solutions.
-Furthermore, there may have been the appearance of new candidates as well as the development of others between the preliminary experiment and the published results.
+Furthermore, there may have been the appearance of new candidates, as well as the development of others between the preliminary experiment and the published results.
 As a result, it is not only required to ensure the "reproducibility" of the results, but also to make way for new candidates.
-As a result, we'd like to propose a new concept for a fruitful experiment, which we'll call "Extension." The capacity to provide the tools required not only to repeat the experiment but also to add additional candidates, workloads, or crucial Key metrics.
+Hence, we would like to propose a new concept to leverage experiments, which we call "Extension".
+The capacity to provide the tools required not only to repeat the experiments, but also to include additional candidates, workloads, or relevant key performance indicators.
 
@@
 \subsection{Extension}
+We propose to extend the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
+Instead of only presenting the four primary aspects of their guidelines that were mentioned in state of the art~\cref{sec:soa_reproducibility}, we propose an abstract model to describe an empirical experiment:
 
-We propose making certain enhancements to the benchmarking framework that was suggested by the collaboratory on Experimental Evaluation of Software and Systems in Computer Science in order to address those problems \footnote{\url{http://evaluate.inf.usi.ch/}}.
-Instead of only presenting the four primary aspects of their guidelines that were mentioned in state of the art\cref{sec:soa_reproducibility} we propose an abstract model to represent the experiment, 
-\emph{context} : the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor .
+\emph{context}: the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor .
 An \emph{orchestrator} which its job to run the experiment. 
-The orchestrator provides three intrafaces :
+The orchestrator provides three interfaces:
 
@@
 
-\section{Reproducibility in benchmarking }
-As we have discussed in the previuous section , the candidates shouldn't be related to the experiment, all they need to know is the interface of the workload, therefore.
-This section will discuss different ways to ensure a reproducibility of the experiment by comparing different ways to encapsulate the candidates
+\section{Reproducibility in Benchmarking}
+As discussed in the previous section, the candidates should not be related to the experiment, all they need to know is the interface of the workload.
+This section thus discusses different ways to enforce the reproducibility of experiment by comparing different ways to encapsulate the systems-under-test.
 
-\subsection{Virtual machines}
-First option should be using virtual machines,  Using virtual machines aka VM gives reaserchers the freedom to choose their own tools, software and operating system that they are the most confortable with without paying the price to change the actual working environment, which will give them eventually more controle over the dependencies and the execution environement.
-Moreover, using a vm will solve the \emph{replication crisis} thanks to the virtual images, even the most complex architecture can be reproduced easily by just instantiating a copy of the image.
-Since the virtual machines are agnostic to the host architecture, reaserchers won't have to worry about where and how their experiments are replicated because they have already setup the execution environement.
-Another advantage to the virtual machines is the snapshot mechanism, it allows reaserchers to create backups and revert some changes with simple clicks.
-Last but not least,thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables -controled and uncontroled-  and do other analysis without dealing with any dependencies.
-In his paper \cite{howe_virtual_2012} bill howe lists the advantages of using virtual machines in reaserchers experements including the economical impact and cultural limitation to a such approach.
-which allow them to have control over the ressouces, the dependencies and the execution environment.
-Moreover, thanks to the snapshots, deploying a software is easily done by instantiating a copy of that image.
+\subsection{Virtual Machines}
+First option should be using \emph{Virtual Machines} (VM), which gives researchers the freedom to choose the most appropriate tools, software and operating system that they are the most comfortable with, without paying the price to change the actual working environment, which will give them eventually more control over the dependencies and the execution environment.
+Moreover, using a VM addresses the \emph{replication crisis} thanks to the virtual images, as even the most complex architecture can be reproduced easily by instantiating a copy of the image.
+Since the virtual machines are agnostic to the host architecture, researchers will not have to worry about where and how their experiments are replicated, because they have already setup the execution environment.
+Another advantage of VM is the snapshot mechanism, as it allows researchers to create backups and revert some changes with simple clicks. 
+Last but not least, thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables---controlled and uncontrolled---and do other analysis without dealing with any dependencies.
+In his paper~\cite{howe_virtual_2012}, Bill Howe lists the benefits of using VM in researchers experiments, including the economical impact and cultural limitation to a such approach.
+In particular, VM allow them to have control over the resources, the dependencies, and the execution environment.
+Moreover, thanks to snapshots, deploying a software is easily achieved by instantiating a copy of that image.
 However, this choice comes with a certain cost.
-Because the intervention of the hypervisor,the software will use two kernels , the virtual machine one and the host machine one, which will provide a noticiable overhead, and will impact the performances of the tests.Therefore, we can't use virtual machines for exepements that are related to performance.
-Another limiation with the virtual machine is the the isolation.
-It is true that this feature will prevent the experience environement with any undesirable interference from the outside world.
-but sometimes this contact is needed, especially when the experiment is dependent to an external part, such as sensors.
-In energicial tests we tend to use hardware powermeters which will make it difficult to use the virtual machines in this case .
+Because of the hypervisor, software will build on two kernels---the virtual machine (guest) one and the host machine one---which will provide a noticeable overhead, and will impact the performances of the system-under-test.
+Therefore, we cannot use VM for experiments that are related to performance.
+Another limitation of VM is the the isolation.
+While this feature prevents the experimental environment from any undesirable interference from the outside world, this interaction may be required---especially when the experiment is dependent to an external part, such as sensors.
+For energy measurements, one tend to use hardware power~meters, which will make it difficult to support VM in such a case.
 
 % introduction of docker 
+\subsection{Containers}
+Another solution would be using something that allows us to benefit from the isolation of the host OS, while easing of replication proposed by VM, and the direct interaction with the hardware that the classical method give.
+Containers offer such an advantage, while ensuring the isolation and the ease of replication for application.
+
+Figure~\ref{environement:virtualization_technique} explains the differences in architecture between the types of virtualization and container technologies:
+\begin{itemize}
+    \item \textsf{Type\,1} runs directly on the hardware, it is mainly used by the cloud providers where there is no host OS, but just VM, use the open-source Xen or VMware ESX hypervisors to be executed;
+    \item \textsf{Type\,2} runs over the host OS, mostly used for personal computers.
+    VMware server and virtualBox are popular examples of this type, and most of the researchers experimentation are run with this type. However, due to the 2 OS, the applications tend to be slower.
+    \item \textsf{Containers}. Instead of its own kernel, containers use the host kernel to run their OS, which makes them lighter, faster and use the full potential of hardware.
+    For this category, one can cite \emph{Docker}, \emph{Linux LXC}, or \emph{LXD}~\cite{abuabdo_virtualization_2019}.
+\end{itemize}
+%  TODO Rephrase so i include it before VMs+ add links to the technologie
 
@@
     \center{\includegraphics[width=1\linewidth]{imgs/virtualization_techniques}}
-    \caption{Different Methods of Virtualzation}\label{environement:virtualization_technique}
+    \caption{Different Methods of Virtualization}\label{environement:virtualization_technique}
 \end{figure}
 
-\subsection{Containers}
+\subsection{Docker Vs. Virtual Machine}
+Despite the fact that \textsf{Type\,1} is more performant than \textsf{Type\,2}, the second one is the most used in research, as most researchers tend to conduct their experiments in their own machine.
+In the other hand, Docker is the most famous container technology.
+In our case, we are more prone to promote Docker for two reasons:
+\begin{enumerate}
+    \item we need a lightweight orchestrator to limit the overhead on energy consumption of our tests, as prior work mentioned [cite Morabito (2015) and van Kessel et al. (2016)],
+          % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
+    \item as we are using hardware sensors to measure the energy consumption, we need to interact with the host OS.
+\end{enumerate}
 
-another solution would be using something that allows us to have the isolation from the host os and the ease of replication that virtual machines offer, and the direct interaction with the hardware that the classical method give.
-contarization offers a such advantages while keeping the isolation and the ease of replication for application.
-
-Figure \ref{environement:virtualization_technique} explains the differents in architecture between the clasic types2 of Virtualisation and Containers.
-\begin{itemize}
-    \item Type 1: runs directly on the hardware, it is mainly used by the cloud providers where there is no main OS, but just virtaul machines, we can site for this the open-srouce XEN and VMware ESX
-    \item Type 2: runs over the hostmachine Operating System, mostly used for personal comptuers, VMware server and virtualBox are famous examples of this type, most of the reaserchers expermentation are run witht this type, howerver due to the 2 Operatings syestms the applications tend to be more slower
-    \item containers : Instead of its own kernel, containers used the hots kernel to run their Os, which makes them ligher, quickers and use the full pontial use of the hardware.
-For this we can cite \emph{Docker}, \emph{Linux LXC}\emph{LXD} \cite{abuabdo_virtualization_2019}
-\end{itemize}
-%  TODO Rephrase so i include it before VMs+ add links to the technologie
-
-\subsection{Docker vs Virtual Machine}
-Depsite that Type 1 is more performant than type 2, the second one is the most used in reaserch, since most researchers conducts their experements in their own machine.
-In the other hand, docker is the most famous thechonology for for containers.
-In ower case we are more prone to docker for two reasons.
-\begin{enumerate}
-    \item we need a litetweight orchestrator to not affect the energy consumption of ower tests. As prior work mentioned [cite Morabito (2015) and van Kessel et al. (2016)]
-          % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
-    \item since we are using the hardware itself to measure the energy consumption, we are required to interract with the host OS itself.
-\end{enumerate}
 Special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that allows us to retireve the energy consumption of a virtual machine.
@@
 % The docker files and the clarification of the methodology
-\subsection{Docker and energy}
-Now that we have chosen to go with the containers technorolgy to encapsulate our tests. What would be the impact of this solution on the energy consumption of our tests.
 
-Based on the studies of \cite{santos2018does}, who analysed the impact of adding the docker layer on the energy consumption.
-In their experement. Eddie Antonio et al run multiple benchmarks with and without docker. and compared their energy consumption and execution time.
-The first step was to see the impact of docker deamon while there is no work. to see the impact of the orchestrator alone. Later they had the experiment with the following benchmarks
+\subsection{Docker \& Energy}
+Now that we have chosen to go with the containers technology to encapsulate our tests.
+What would be the impact of this solution on the energy consumption of our tests.
+
+Based on the studies of \cite{santos2018does}, who analysed the impact of adding the Docker layer on the energy consumption.
+In their experiment, Eddie Antonio~\emph{et~ al.} run multiple benchmarks with and without Ddocker. 
+They compared the resulting energy consumption and execution time.
+The first step was to see the impact of Docker deamon, while there is no work, to observe the impact of the orchestrator.
+Then, they conduct their experiments with the following benchmarks:
 \begin{itemize}
-    \item wordpress
-    \item reddis
-    \item postfresSQL
+    \item Wordpress,
+    \item Reddis,
+    \item PostgreSQL.
 \end{itemize}
-The following figures represents the energy consumption of the system while it is idle. As we can see  in figure \ref{fig:docker_idle} Docker braught arround 1000joules overhead.
+
+The following figures represents the energy consumption of the system-under-test while it is idle.
+As one can observe in Figure~\ref{fig:docker_idle}, Docker brings an overhead of around $1,000$ Joules.
+
 \begin{figure}
@@
 \end{figure}
-In the other hand as we can see in figure \ref{fig:docker_reddis}.
-docker increased the execution time of the benchmark by 50 seconds which caused an increase in energy since they are highly correlated.
-The authors also highlated the fact that this increase of energy consumption is due to the docker deamon and not the fact that the application is in a container.
-Moreover they estimted the price of this extra energy and it was less than 0.15\$ in the worst case.
-Which is non significant compared to the advantages that docker bring for isolation and reproducibility.\\
-if we  recap this study in one sentence,it would be the following one.
-The dockerised softwares tend to consume more energy, because mainly they take more time to be executed.
-The average power consumption is higher with only \textbf{2Watts} and it is due to the docker deamon.This overhead can be up to 5\% for IO intensive application, but it is mearly noticiable when it comes to CPU or DRMA intesive works
+
+In the other hand, as one can see in Figure~\ref{fig:docker_reddis}.
+Docker increased the execution time of the benchmark by 50 seconds, hence increasing the energy consumption.
+The authors reported that this overhead is mostly due to the Docker deamon and not the fact that the application is in a container.
+Moreover, they estimated the cost of this extra energy and it was less than 0.15\$ in the worst case, which is non-significant compared to the benefits that Docker brings for isolation and reproducibility.
+
+To summarize, Docker-based software tends to consume more energy, because mainly they take more time to be executed.
+The average power consumption is higher with only \textbf{2\,Watts}, due to the execution of the Docker deamon.
+This overhead can reach up to 5\% for IO-intensive application, but it is merely noticeable when it comes to CPU- or DRMA-intesive workloads.
 
@@
     \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_energy}
-    \caption{execution time and energy consumption of Redis  with and without docker \cite{santos2018does}}\label{fig:docker_reddis}
+    \caption{Execution time \& energy consumption of Redis with and without Docker~\cite{santos2018does}}\label{fig:docker_reddis}
 \end{figure}
 
-\subsection{docker and accuracy}
-And now Since the stat of the art has aggreed on the impact of docker on the energy consumption,Let's discuss it's impact on the accuracy. In other words\\
-\textbf{RQ :} does Docker affect the energy variation of the exepements ?
+\subsection{Docker \& Accuracy}
+As the state of the art assesses the impact of Docker on the energy consumption, one can also consider its impact on accuracy.
+In other words:\\
+\textbf{RQ:} does Docker affect the energy variation of the experiments? %La RQ n'est pas lié au reste du paragraphe, non ?
 
-To Answer this question we have conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
+To answer this question we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
 Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.

### 2022-07-04 chapters/introduction.tex
proofreading

@@
 \newpage
-\chapter{Introduction}
+\chapter{Motivation}
 \label{chapter:introduction}
 
-Nowdays, computers are invasing our daily life from work to leasure, from fancy smartphone to an embded peacemaker that regulatres the heartbeat of some people.
-As human beigns we are known to use tools to enhance our bodies.
+Nowadays, computers are invading our daily life from work to leisure, from fancy smartphone to an embedded peacemaker that regulates the heartbeat of people.
+As human beings, we are known to use tools to enhance our bodies.
 And thanks to computers we pushed that step even further where now we are using machines to extend our brains.
@@
 Well this is the problem for the future generations.
-For the moment, the main concern of humanity is to keep this planet livebale until we find another alternative.
+For the moment, the main concern of humanity is to keep this planet liveable until we find another alternative.
 
@@
 consumption. In computer science we are concerned with the ladder solution %reference.
-Therefore most of the works are done on tunning the software and the hardware in ordrer to have a more efficient way use this energy. %more efficient way of producing programs from electricity? 
+Therefore most of the works are done on tunning the software and the hardware in order to have a more efficient way use this energy. %more efficient way of producing programs from electricity? 
 %say about hardware (i.e. superconductive materials with zero resistance)
@@
 %More precisely, the purpose of this thesis is to highlight (attract attention to?) the impact of programers' choices on energy comsumption of software while running the code and to advise them with the strategies to write the greenest possible code.
-Our approach is to reduce the energy consumption of the software services by changing certains parameters, such as palteform, programming langauge, and/or the design pattern.
+Our approach is to reduce the energy consumption of the software services by changing certain parameters, such as platform, programming language, and/or the design pattern.
 
-The best way to do so is to formulate a theory behind the energy consumption of algorthms such as the complexity and the o notation.
+The best way to do so is to formulate a theory behind the energy consumption of algorithm, such as the complexity and the o notation.
 Unfortunately this is not possible in the current state of the art.
@@
 
-Unlike algrothim optimization in the field of perfornacne which is agnostic toward the plateform. the energy consumption of the algorithms is dependent on execution environement.
-Therefore, for the moment we will start by formulating some hypthosis and explore them using empirical analysis.
+Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on execution environment.
+Therefore, for the moment we will start by formulating some hypothesis and explore them using empirical analysis.
 
-Our work will be presented through the following chapters :
-
+Our work will be presented through the following chapters:
 \begin{enumerate}

### 2022-07-04 chapters/literature/draft_citations.tex
proofreading

@@
 \newpage
-\section{benchmarking}\label{sec:soa_benchmarking}
+\section{Benchmarking}\label{sec:soa_benchmarking}
 This section will go through the flaws and best practices of empirical research before applying them to our topic of study.
-\subsection{Threats and Challenges}
 
+\subsection{Threats \& Challenges}
 A successful benchmark must meet three criteria.
@@
 % that serves as a basis for evaluation or comparison (as of computer system performance)
-\subsubsection{the impact of these challenges in the empirical research }
-In their work~\cite{van_der_kouwe_benchmarking_2018}, \citeauthor{van_der_kouwe_benchmarking_2018} Investigated 50 papers published in top venues
-to findout that tier-1 papers commit an average of five benchmarking crimes.
+\subsubsection{Impact of these challenges in the empirical research}
+In their work~\cite{van_der_kouwe_benchmarking_2018}, Van-der-Kouwe~\emph{et~al.} investigated 50 papers published in top venues
+to find out that Tier-1 papers commit an average of five benchmarking crimes.
 To analyze the magnitude of the phenomenon, they have identified a set of 22 "benchmarking crimes" that threaten the validity of the system.
@@
 
-\subsection{proposed solutions}\label[]{sec:soa_reproducibility}
-
+\subsection{Proposed Solutions}\label[]{sec:soa_reproducibility}
 %% BENCHMARKING FRAMEWORK -- ADD IT later \cite{sonnenwald2003evaluating}
@@
 % as for computer sience. In their paper~\cite{buytaert_statistically_nodate}, the authors  separated the start-up measurement from the steady state when comparing different Java optimizations.
-Another work, to make unify the bemchmarking methodology across different research works in the field of computer science, we can cite the paper~\cite{stephen_evaluate_2012}.
-In their approach they  divided any emprical experiment into four  components:
+Another work, to make unify the benchmarking methodology across different research works in the field of computer science, we can cite the paper~\cite{stephen_evaluate_2012}.
+In their approach they  divided any empirical experiment into four  components:
 
@@
     \center{\includegraphics[scale=0.3]{imgs/componentsofanexperiment}}
-    \caption{The proposed decomposition of an experement~\cite{stephen_evaluate_2012}}\label{fig:soa_expermiment_component}
+    \caption{The proposed decomposition of an experiment~\cite{stephen_evaluate_2012}}\label{fig:soa_expermiment_component}
 \end{figure}
@@
 
-\section{energy measurement}
-to be able to reduce the energy cost of running programs we should first be able to estame this energy consumption.
-many studies have been conducted to estimate a such energy consumption.
-that varries from static analysis of the source code to enfere it's energy consumption like \citeauthor{pereira_helping_2017} where they provided a tool to highlight the most energy consuming parts of the code \cite{pereira_helping_2017}.the main advantage of this apporach is to be able to estimate the energy consumption of a program without running it.
-however.
-unlike the conplexity of programs, the energy consumption is highly related the execusion environment.
-Therefor, Static analysis might not represent the real behaviour of the same program when it is run in the production environment.
+\section{Energy Measurement}
+To be able to reduce the energy cost of running programs we should first be able to estimate this energy consumption.
+Many studies have been conducted to estimate a such energy consumption.
+that varies from static analysis of the source code to infer its energy consumption like \citeauthor{pereira_helping_2017} where they provided a tool to highlight the most energy consuming parts of the code \cite{pereira_helping_2017}.
+The main advantage of this approach is to be able to estimate the energy consumption of a program without running it.
+Unlike the complexity of programs, the energy consumption is highly related the execution environment.
+Therefore, static analysis might not represent the real behavior of the same program when it is run in the production environment.
 To treat the problem of representativeness, many researchers tend to measure the energy consumption of the programs while they are running them. Hence we will get more accurate results.
-there is a variety of tools to measure such energy, and they cover a large spectrium of usage depends on how \textbf{accurate} and \textbf{pricise} those results are needed to be in tone hand, and the price that practinoners are ready to pay for a such accuray/pricision.
+there is a variety of tools to measure such energy, and they cover a large spectrum of usage depends on how \textbf{accurate} and \textbf{precise} those results are needed to be in tone hand, and the price that practitioners are ready to pay for a such accuracy/precision.
 Below we will present some tools that are well known in the literature and discuss their main advantages / drawbacks for our case.
-\subsection{hardware tools}
 
+\subsection{Hardware Tools}
 % TODO Add : diagram of the possible point where we can measure the energy 
-
 % depends on the scientific usage there are two apporaches :
@@
 dedicated one , they are based on the instrumentation of some parts of the computer , however they have a high cost and they are defficult to scale since they require an instrumentation of the execusion environment.
-according to \citeauthor{hackenberg2014hdeem} there are four main creterions to evaluate an energy measurement tool \cite{hackenberg2014hdeem}
+according to \citeauthor{hackenberg2014hdeem} there are four main criterions to evaluate an energy measurement tool \cite{hackenberg2014hdeem}
 \begin{itemize}
     \item \em{spatial granularity}: the more specific the target of monitoring we are able to measure, the more efficient we can do optimization since we will know what causes the pitfalls of the energy consumption.
-    \item \em{temporal granularity}: same as spatial granularity, temporal granularity helps us to identif the sequence of code that need to be optimized.
+    \item \em{temporal granularity}: same as spatial granularity, temporal granularity helps us to identify the sequence of code that need to be optimized.
     \item \em{scalability}: this is mainly related to the cost of the tools and the ease of their integration for our system.
-    \item  \em{accuracy}:to elimanate extra hazards and get more representative measurement.
+    \item  \em{accuracy}: to eliminate extra hazards and get more representative measurement.
 \end{itemize}
 
-We believe that accuracy can be extracted from those creterions since it is a result of the combaination of the two first ones. therefore, we will focus more on the first 3 creterions from later on.
-In the follwong section we will discuss some of the well known tools that are used in literature.
+We believe that accuracy can be extracted from those criterions since it is a result of the combination of the two first ones. therefore, we will focus more on the first 3 criterions from later on.
+In the following section, we will discuss some of the well known tools that are used in literature.
 
-Nowdays, most of the high performance computing systems (HPC) implements a tool to report the energy consumption of the nodes for the sake of monitoring and administration.
-Those tools are mainly integrated withing the power suply units (PSU) or the power distributuin units (PDU).
-then , they provide an interface and a log to follow the history of the energy consumption.
+Nowadays, most of the high performance computing systems (HPC) implements a tool to report the energy consumption of the nodes for the sake of monitoring and administration.
+Those tools are mainly integrated withing the power supply units (PSU) or the power distribution units (PDU).
+Then, they provide an interface and a log to follow the history of the energy consumption.
 Despite their scalability and ease of integration.
-such tools lack both in spatial and temporal granularity since they monitor the whole energy of the nodes, and most of the times they have a very low sampling frequency.
-most of those tools are provided directly by the manufacturers.
-such as \em{IBM EnergyScale technology} \cite{mccreary2007energyscale} \cite{caldeira2014ibm} \cite{caldeiraibm} or \em{Dell poweredge} \cite{lovicott2009thermal}, MEGware Cluststafe \cite{breitbart2015case}
-As we said earlier the true purpose of those tools is more monitoring than analysing the energy consumption.
+Such tools lack both in spatial and temporal granularity since they monitor the whole energy of the nodes, and most of the times they have a very low sampling frequency. 
+Most of those tools are provided directly by the manufacturers. 
+Such as \em{IBM EnergyScale technology}~\cite{mccreary2007energyscale,caldeira2014ibm,caldeiraibm} or \em{Dell poweredge}~\cite{lovicott2009thermal}, MEGware Cluststafe~\cite{breitbart2015case}.
+As said earlier, the true purpose of those tools is more monitoring than analyzing the energy consumption.
 %% add something to link with wattsup pro 
-WattsUp Pro, is a device that can be installed  between the  powerer source of the machine and the system under test.
-it allows a sampling frequency up to 1Hz and has a internal memory to store a wide variety of data, such as the maximum voltage, current..
-etc that later can be exported via USB port for personal usage or lined to some graph programs like Logger Pro or LabQuest.The main advantage of this tool is the ability to monitor the energy consumption from a different device which will reduce the risk of interference with the energy consumption of the test % a revoir
+WattsUp Pro, is a device that can be installed between the power source of the machine and the system under test.
+It allows a sampling frequency up to 1\,Hz and has a internal memory to store a wide variety of data, such as the maximum voltage, current, that later can be exported via USB port for personal usage or lined to some graph programs like Logger Pro or LabQuest.
+The main advantage of this tool is the ability to monitor the energy consumption from a different device which will reduce the risk of interference with the energy consumption of the test % a revoir 
 \cite{hirst2013watts}
 
-despite his high temporal granularity  , wattsup pro lacks in term of spatial granularity since it monitors the energy consumption of the whole system.
+despite his high temporal granularity, wattsup pro lacks in term of spatial granularity since it monitors the energy consumption of the whole system.
 To have a finer granularity we need to isolate the energy consumption of each component.
-.
-For this we will need more intrusive tools such.
+For this, we will need more intrusive tools such.
 
@@
 
-PowerMon and its upgradded version powerMon2 \cite{bedard2010powermon} are based on an microcontroler chip that can monitor up to 6 channel ( 8 for powermon2) simultanetly.
-therefore we are able to monitor the powerconsumption of 4 devices at the same time.
-the frequency sample of this tool is up to 50Hz with an accuracy of 1.2\%.
+PowerMon and its upgraded version powerMon2 \cite{bedard2010powermon} are based on an micro-controller chip that can monitor up to 6 channel (8 for powermon2), simultanetly. 
+Therefore, we are able to monitor the power consumption of 4 devices at the same time. 
+The frequency sample of this tool is up to 50\,Hz, with an accuracy of 1.2\%.
 Powermore2 comes with smaller size that can fit within 3.5 inches rack drive.
 
-PowerInsight \cite{laros2013powerinsight} is another finegrained measurement tool that is based on an ARM bagelbone processor \cite{coley2012beaglebone} which is able to measure up to 30 channel simultanetly with a frequency of 1KHz per channel.
+PowerInsight~\cite{laros2013powerinsight} is another fine-grained measurement tool that is based on an ARM bagelbone processor~\cite{coley2012beaglebone}, which is able to measure up to 30 channel simultaneously with a frequency of 1KHz per channel.
 
-on the other hand GreenMiner \cite{hindle2014greenminer}, %%i dont think that  mobile measurement will be relevent for my case
+On the other hand, GreenMiner \cite{hindle2014greenminer}, %%i dont think that  mobile measurement will be relevent for my case 
+powerpack~\cite{ge2009powerpack} in the other hand is an api that synchronizes the data gathered from other monitoring tools such as Watt’s Up Pro, NI and RadioShack pro and the lines of code.
+%% maybe remove this one from the list because it is not well known.
 
-powerpack \cite{ge2009powerpack} in the other hand is an api that synchronizes the data gathered from other monitoring tools such as Watt’s Up Pro, NI and RadioShack pro and the lines of code %% maybe remove this one from the list because it is not well known.
-
-other monitor tools have been relized by the manufactures such as IBM Power executive \cite{koomey2011growth} which allows their customer to monitor the power consumption and thermal behaviour of the of BladeCenter systems in the datacenter.
+Other monitor tools have been realized by the manufactures, such as IBM Power executive~\cite{koomey2011growth}, which allows their customer to monitor the power consumption and thermal behavior of the of BladeCenter systems in the data center.
 
@@
 as we see in % two diagrams one for research the other one for power usage 
-The CPU is the part responsible of the most energy consumption of softwares. hence the finner we go to measure this energy consumption the better it is for our work.
+The CPU is the part responsible of the most energy consumption of softwares. hence the finer we go to measure this energy consumption the better it is for our work.
 Fortunately, Intel and later AMD proposed a tool that estimate the power consumption of different part of the CPI based on counter performances.
-RAPL (RUnning Average Power Limit) \cite{hackenberg2013power} \cite{hackenberg2015energy} is a set of registers that was introduced by intel in their CPU since Sandy bridge generation, and later it was fllowed by AMD since Family 17h Zen.
-It is capable of monitoring different components of CPU , such as DRAM, Cores , and the integrated GPU for desktop processors.
+RAPL (\emph{RUnning Average Power Limit}) \cite{hackenberg2013power,hackenberg2015energy} is a set of registers that was introduced by Intel in their CPU since Sandy bridge generation, and later it was followed by AMD, since Family 17h Zen.
+It is capable of monitoring different components of CPU, such as DRAM, cores, and the integrated GPU for desktop processors.
 
@@
 
-The advanage of a such apporach is that the absence of any intrusive measurement tools.
-further more they have a high temporal granurality with a sampling frequency up to 1KHz \cite{ilsche_power_2015}.
+The advantage of such an approach is that the absence of any intrusive measurement tools. 
+Furthermore, they have a high temporal granularity with a sampling frequency up to 1\,KHz~\cite{ilsche_power_2015}.
 
-with similar approach we can find NVIDIA reporting tools such as GPU TESLA \cite{burtscher2014measuring} and the NVML library \cite{fahad2019comparative}
+With similar approach, we can find NVIDIA reporting tools, such as GPU TESLA \cite{burtscher2014measuring} and the NVML library~\cite{fahad2019comparative}.
 
-\subsection{software tools}
-
+\subsection{Software Tools}
 Software-based measurement tools are based on other hardware tools to monitor energy use.
@@
 
-The first examples of software measurement tools are PowerAPI~\cite{colmant2018next}, SmartWatts formula~\cite{fieni2020smartwatts} and selfWatts~\cite{fieni2021selfwatts}.
+The first examples of software measurement tools are PowerAPI~\cite{colmant2018next}, SmartWatts formula~\cite{fieni2020smartwatts} and SelfWatts~\cite{fieni2021selfwatts}.
 These tools collect global energy consumption measurements from RAPL and use other system events such as cache misses/hits and CPU frequency evolution (DVFS) via a sensor to construct a power model of the control groups (system control groups, docker containers, kubernetes pods, etc.) using a Ridge regression.
@@
 To adjust its models to the hardware on which it operates, Joulemeter must first go through a calibration step.
-It is able to only monitor one process at a time with a frequency of 1 HZ.
+It is able to only monitor one process at a time with a frequency of 1\,Hz.
 
-JRAPL is another example of an energy measurement tool estimating tool that has been utilized in a variety of publications\cite{guimaraes2016some,liu2015data}.
+jRAPL is another example of an energy measurement tool estimating tool that has been utilized in a variety of publications\cite{guimaraes2016some,liu2015data}.
 This software enables for the energy usage of Java programs, functions, or even a block of code lines to be profiled and measured.
@@
 As a result, the global energy consumption collected by RAPL between two timestamps (the start and finish of the code to measure) is used to calculate the energy consumption of the Java code.
-Tests using JRAPL should be done on a well-configured machine to minimize the impact of the operating system and user operations on the overall energy consumption of JRAPL.
+Tests using jRAPL should be done on a well-configured machine to minimize the impact of the operating system and user operations on the overall energy consumption of jRAPL.
 
-Another process-level energy usage measuring tool is Jolinar \cite{islam2016measuring,noureddine2016jolinar}.
+Another process-level energy usage measuring tool is Jolinar~\cite{islam2016measuring,noureddine2016jolinar}.
 The tool does not need a calibration phase and relies on pre-established power models based on hardware metrics (TDP, disk I/O rate, and so on).
@@
 At the end of the execution, the tool provides the CPU, DRAM, and disk energy usage of the main process.
-joulinar
 
-Jalen\cite{noureddine2015monitoring} is another tool that profiles and monitors the energy usage of a Java program. Unlike JRAPL, Jalen is able to reduce the scope up to the function level.
+Jalen~\cite{noureddine2015monitoring} is another tool that profiles and monitors the energy usage of a Java program.
+Unlike jRAPL, Jalen is able to cover the scope down to the function level.
 It gathers data using code instrumentation and statistical sampling at a predetermined pace.
@@
 
-\subsection{hybrid tools}
+\subsection{Hybrid Tools}
 % split hardware tools into extern and embeded tools
@@
 
-\section{programming languages and performances}\label{section:soa_optimization}
-\subsection{energy}\
+\section{Programming Languages \& Performances}\label{section:soa_optimization}
 
+\subsection{Energy}
 energy comparaison of programming languages in the game benchmark by debian \cite{pereira2017energy}
@@
 
-impact of maintability on software energy evolution \cite{calero2021does}
+impact of maintainability on software energy evolution \cite{calero2021does}
 
-defintion \cite{wang1993grpc}
+definition \cite{wang1993grpc}
 
-comparaison \cite{chamas2017comparing}
+comparison \cite{chamas2017comparing}
 
-emprical study \cite{de2021empirical}
+empirical study \cite{de2021empirical}
 
-impact of website implemenation on energy consumption \cite{philippot_characterization_2014} \cite{manotas2013investigating}
+impact of website implementation on energy consumption \cite{philippot_characterization_2014} \cite{manotas2013investigating}
 
@@
 
-\subsection{python}
+\subsection{Python}
 
-webframeworks on python \cite{pankiv_concurrent_nodate}
+web frameworks on python \cite{pankiv_concurrent_nodate}
 
@@
 
-\subsection{jvm}
+\subsection{JVM}
 
@@
 
-%%%%%%%%%%%%% energy metigation
-
+%%%%%%%%%%%%% energy mitigation
 \section{Energy Variation}
@@
 \section{Conclusion}
-As we have seen in the state of the art, many methodes have been proposed to reduce the energy footprint of the the ICT,which they can be applied in deffirent parts of the lifecycle of the program, from consumption to the execution.
-Furthermore, the execution phase took attention of many researchers because it's the part where the most energy is consumed.
+As we have seen in the state of the art, many methods have been proposed to reduce the energy footprint of the the ICT, which they can be applied in different parts of the lifecycle of the program, from consumption to the execution.
+Furthermore, the execution phase took attention of many researchers because it is the part where the most energy is consumed. 
 This thesis will focus on that aspect as well. 
-However, unlike the most work that have been done on the hardware aspect, we will target the software impact on this energy, starting from the choice of the programming language up to how to tune some features of a frameworks in order to make the software consumes less energy.To do so we use the emprical apporach due to it's concinstency for the moment.Unlike the perfromance which essentialy related to the complexity of the algorithm, the energy consumption is more impacted by the hardware.
-Therefore, to optimize the enrgy consumption we choose a spiral methode. based on 3 phases:
+However, unlike the most work that have been done on the hardware aspect, we will target the software impact on this energy, starting from the choice of the programming language up to how to tune some features of a frameworks in order to make the software consumes less energy.
+To do so we use the empirical approach due to its consistency for the moment.
+Unlike the performance, which is essentially related to the complexity of the algorithm, the energy consumption is more impacted by the hardware.
+Therefore, to optimize the energy consumption we choose a spiral method based on 3 phases:
 \begin{enumerate}
-    \item execute the code
-    \item measure the program
-    \item infere the guidelines
+    \item execute the code,
+    \item measure the program,
+    \item infer the guidelines.
 \end{enumerate}
@@
     \center{\includegraphics[scale=0.3]{imgs/greenbenchmarkcycle}}
-    \caption{the spiral methode of energy optimization }\label{fig:spirals}
+    \caption{the spiral method of energy optimization }\label{fig:spirals}
 \end{figure}
-the aim of this work is to present a set of guidelines to create a benchmarking system to to measure the energy consumption of different programs.
+
+The aim of this work is to present a set of guidelines to create a benchmarking system to to measure the energy consumption of different programs.
 After that, we will use this system to compare the energy consumption of different programming languages.
-We will extend the work of \citeauthor{pereira2017energy} to a closer distance to production environment by comparing a set of usecases.
-starting by GRPC framework, and a set of Web Framewroks.
-Finally we will discuss the impact of the execution environment on the energy consumption of two of the most famous programming languages JAVA and Python.
-and present how can tunning the Virtual Machine can reduce the energy consumption.
+We will extend the work of \citeauthor{pereira2017energy} to a closer distance to production environment by comparing a set of use cases.
+Starting by GRPC framework, and a set of Web Frameworks.
+Finally, we will discuss the impact of the execution environment on the energy consumption of two of the most famous programming languages Java and Python and present how can tunning the Virtual Machine can reduce the energy consumption.

### 2022-07-04 chapters/benchmarking/conclusion.tex

@@
 \section{Perspectives}
-By the end of this study we have gathered enought guidelines to make the tests more reproducible, accurate.
+By the end of this study, we have gathered enough guidelines to make the tests more reproducible, accurate.
 We created a set of new tests named \textbf{energy tests} which are more similar to performance tests.
-Thanks to the work of two interns [ mamadou and adrien] we created a CI/CD plateform to measure the energy consumption of Java projects and we could track the evolution of the this energy accross different stages of the project.
+Thanks to the work of two interns [ mamadou and adrien] we created a CI/CD platform to measure the energy consumption of Java projects and we could track the evolution of the this energy across different stages of the project.
 In the figure below we see an example of this plugin.

### 2022-07-04 chapters/benchmarking/experiment.tex

@@
 \subsection{Experimental Setup}\label{subsec:setup}
-This section describes our detailed experimental environment, covering the clusters configurationand the benchmarks we used  justifying our experimental methodology.
+This section describes our detailed experimental environment, covering the clusters configuration and the benchmarks we used  justifying our experimental methodology.
 
@@
 
-To satisfy those requirements we have used the plateform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708},a large-scale and flexible testbed for experiment-driven research distributed accross all france.
-Grid5000 offers  miultiple clusters composed with 4 upto 124 identical machine with different configurations for each cluster.
-For our experement we have considered 4 clusters. Our main creterions was the CPU Configuration.
-the table below~\ref{table:g5k}, presents a description of the 4 clusters
+To satisfy those requirements we have used the platform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708},a large-scale and flexible testbed for experiment-driven research distributed across all france.
+Grid5000 offers  multiple clusters composed with 4 up to 124 identical machine with different configurations for each cluster.
+For our experiment, we have considered 4 clusters.
+Our main criterions was the CPU Configuration.
+the table below~\ref{table:g5k}, presents a description of the 4 clusters.
 
@@
 As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or \textsf{socket} to designate one of the two sockets and \textsf{PU} for the \emph{processing unit}.
-For our study we coonsider hyper-threads as distincts \textbf{PU}
+For our study we consider hyper-threads as distinct \textbf{PU}
 
@@
 \subsubsection{Workload}
-Our choice for the benchmarks was based on two creterions.
+Our choice for the benchmarks was based on two criterions.
 
 First, \textbf{scalability} : We wanted to gather the highest possible amount insights (regarding time spent for the experiment, therefore we wanted some benchmarks who can scale according to the number of PUs and can fulfill different scenarios.
-Second creterion is \textbf{representativeness} : as menstionned in the challenges, a workload has to be representative otherwise the experiment would be inconsistent \cite{stephen_evaluate_2012}.
-To satisfy those creterions we went through state of the art and looked for the most used benchmarks for testing the performance of the hardware, and then we selected the scallable ones.
+Second criterion is \textbf{representativeness} : as mentioned in the challenges, a workload has to be representative otherwise the experiment would be inconsistent \cite{stephen_evaluate_2012}.
+To satisfy those criterions we went through state of the art and looked for the most used benchmarks for testing the performance of the hardware, and then we selected the scalable ones.
 Our candidate is  emph{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925}: one of the most used benchmarks for \emph{HPC}.
@@
         \hline
-        Use the least of sockets in a case of multiple CPU                                                                                                                & Medium        & Up to $30\times$  \\
+        Use the least of sockets in a case of multiple CPU   & Medium        & Up to $30\times$\\
         \hline
-        Avoid the usage of hyper-threading whenever possible                                                                                                              & Medium        & Up to $5\times$   \\
+        Avoid the usage of hyper-threading whenever possible & Medium        & Up to $5\times$\\
         \hline
@@
         \hline
-        Do not relate to the machine idle variation to isolate a test EC, the CPU/OS changes its behavior when a test is running and can exhibit less variation than idle & Any           & ---               \\
+        Do not relate to the machine idle variation to isolate a test EC, the CPU/OS changes its behavior when a test is running and can exhibit less variation than idle & Any & ---\\
         \hline

### 2022-07-05 chapters/benchmarking/draft.tex
refactoring : undo the changes for the benchmarking protocol

@@
 
-This chapter discusses how to adapt the state-of-the-art of empirical analysis in the energy consumption field, highlighting the difficulties and the various techniques to overcome them.
+This chapter discusses how to solve the challanges of empirical analysis when it comes to the energy consumption experiments,
+
+first we will dicuss how to ensure the three aspects of successful benchmark within the energy consumption. 
+So section~\ref{sec:benchmarking_reproducibility} will focus on the "reproducibility" aspect and how to provide reproducible experiments without impacting the energy measurement. 
+Section~\ref{sec:taming-the-energy-variation} will be about the accuracy when it comes to software energy consumption. and finally 
+Section~\ref{sec:extension} will be about the tools to include new candidates, workloads, or relevant key performance indicators.
+
+As for "representativeness" of the benchamrks, we will discuss it further in the Perspectives section.
 
@@
 
-\subsection{Extension}
-We propose to extend the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
-Instead of only presenting the four primary aspects of their guidelines that were mentioned in state of the art~\cref{sec:soa_reproducibility}, we propose an abstract model to describe an empirical experiment:
-
-\emph{context}: the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor .
-An \emph{orchestrator} which its job to run the experiment. 
-The orchestrator provides three interfaces:
-
-Workload interface  : Wich provides a set of functions that should be implemented by the workload  
-Candidate interface : 
-Observer interface : provides a set of metrics that are collected by the orchestrator during the expermentation.
-note that some of those metrics are external, such the energy consumption of the system , execution time, execution stack ..etc .
-
 % \begin{itemize}
@@
 % \end{itemize}
-
-% \begin{figure}%[!htb]
-%     \center{\includegraphics[width=.9\linewidth]{imgs/benchmarkingprotocol}}
-%     \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
-% \end{figure}
-
-later in this chapter we will discuss each aspect of those interfaces
-
-\section{Reproducibility in Benchmarking}
-As discussed in the previous section, the candidates should not be related to the experiment, all they need to know is the interface of the workload.
+\section{Reproducibility within the context of energy}\label{sec:benchmarking_reproducibility}
+% Now that we have covered the general strucutre of the protocol, this section details the specific aspects of the protocol that are relevant to the energy consumption analysis.
+% We will first discuss how to insure the extension of the candidates.
+later we will talk about the energy metrics and how to ensure the accuracy when it comes to the measurement of the energy consumption.
 This section thus discusses different ways to enforce the reproducibility of experiment by comparing different ways to encapsulate the systems-under-test.
 
-\subsection{Virtual Machines}
-First option should be using \emph{Virtual Machines} (VM), which gives researchers the freedom to choose the most appropriate tools, software and operating system that they are the most comfortable with, without paying the price to change the actual working environment, which will give them eventually more control over the dependencies and the execution environment.
+\paragraph{Virtual Machines}
+First option should be using \emph{Virtual Machines} (VM), which gives researchers the freedom to choose the most appropriate tools, software and operating system that they are the most comfortable with, without paying the price to change the actual working environment, which will give them eventually more control over the dependencies and the execution environment.Therefore, Adding a new candidate is as simple as adding a new VM to the experiment.
 Moreover, using a VM addresses the \emph{replication crisis} thanks to the virtual images, as even the most complex architecture can be reproduced easily by instantiating a copy of the image. 
-Since the virtual machines are agnostic to the host architecture, researchers will not have to worry about where and how their experiments are replicated, because they have already setup the execution environment.
-Another advantage of VM is the snapshot mechanism, as it allows researchers to create backups and revert some changes with simple clicks. 
-Last but not least, thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables---controlled and uncontrolled---and do other analysis without dealing with any dependencies.
-In his paper~\cite{howe_virtual_2012}, Bill Howe lists the benefits of using VM in researchers experiments, including the economical impact and cultural limitation to a such approach.
-In particular, VM allow them to have control over the resources, the dependencies, and the execution environment.
-Moreover, thanks to snapshots, deploying a software is easily achieved by instantiating a copy of that image.
 However, this choice comes with a certain cost.
@@
 Another limitation of VM is the the isolation.
-While this feature prevents the experimental environment from any undesirable interference from the outside world, this interaction may be required---especially when the experiment is dependent to an external part, such as sensors.
-For energy measurements, one tend to use hardware power~meters, which will make it difficult to support VM in such a case.
+While this feature prevents the experimental environment from any undesirable interference from the outside world, this interaction may be required---especially when the experiment is dependent to an external
 
-% introduction of docker 
-\subsection{Containers}
+\paragraph{Containers}
 Another solution would be using something that allows us to benefit from the isolation of the host OS, while easing of replication proposed by VM, and the direct interaction with the hardware that the classical method give.
@@
 \end{itemize}
-%  TODO Rephrase so i include it before VMs+ add links to the technologie
 
@@
 \begin{enumerate}
-    \item we need a lightweight orchestrator to limit the overhead on energy consumption of our tests, as prior work mentioned [cite Morabito (2015) and van Kessel et al. (2016)],
+    \item we need a lightweight orchestrator to limit the overhead on energy consumption of our tests, as prior work mentioned
+    \cite{van2016power}
+    % TODO : ask romain about the title of the paper and the reference to the paper
+    [cite Morabito (2015) and van Kessel et al. (2016)],
           % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
@@
 Special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that allows us to retireve the energy consumption of a virtual machine.
-%%%%% THINGS TO ADD : 
-%% -------------- Dont think so 
-% the impact of docker on the reprodudicibility 
-% Limits of docker 
-% The docker files and the clarification of the methodology
 
@@
 
-\subsection{Docker \& Accuracy}
-As the state of the art assesses the impact of Docker on the energy consumption, one can also consider its impact on accuracy.
-In other words:\\
-\textbf{RQ:} does Docker affect the energy variation of the experiments? %La RQ n'est pas lié au reste du paragraphe, non ?
-
-To answer this question we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
-Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
-In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
-The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
-
-\begin{figure}
-    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
-    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
-\end{figure}
+\subsection{Conclusion}
+As one can see, Docker impact on energy is more constat due to the introduction of the supervisor.
+which will be applied to all the experimenets evenly.
+Therefore when it comes to comparison analysis, it will mitigate its impact autimatically.
+Moreover since we have access to the host Hardware, we dont have to worry about catching the energy consumption for the SUT. 
+Therefore we will be conducting all our experiments using docker as an encapsuation technology. wich will ensure the reproducibility of our experiments with a simple and easy way.

### 2022-07-05 chapters/benchmarking/conclusion.tex
refactoring : revert the changes for the benchmarking protocol chapter

@@
+\subsection{Extension}
+We propose to extend the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
+Instead of only presenting the four primary aspects of their guidelines that were mentioned in state of the art~\ref{sec:soa_reproducibility}, we propose an abstract model to describe an empirical experiment:
+
+\emph{context}: the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor .
+An \emph{orchestrator} which its job to run the experiment. 
+The orchestrator provides three interfaces:
+
+Workload interface  : Wich provides a set of functions that should be implemented by the workload  and called by the each candidate in the experiment. 
+Observer interface : provides a set of metrics that are collected by the orchestrator during the expermentation.
+note that some of those metrics are external, such the energy consumption of the system , execution time, execution stack ..etc .
+
+Figure\ref{fig:benchmarkingprotocol} explains in detail the relation ship within each component of the Protocol. 
+First of all, we should start with the orchestrator which is the kernel of the experiment. Most of the time this kernel should represent the constants parts of the experiment. 
+The orchestrator is responsible for the following  for the coordination between the workload and the candidate, therefore it provides an interface that is implemented by each workload. 
+Later this interface will be provided to each candidate by the orchestrator.
+Like this we will remove any bias of dependecy between the workload and the candidates, which will alow us to extend both parts without any change in the code.
+Instead of retrieving the data directly from the experiment, we suggest to use an observator to allow the extension of the metrics, this observer in general should be outside the experiment to be able to gather as many informations without impacting the the code.
+\begin{figure}[!htb]
+    \center{\includegraphics[width=\linewidth]{imgs/benchmarkingprotocol}}
+    \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
+\end{figure}
+
 \section{Perspectives}

### 2022-07-05 chapters/benchmarking/experiment.tex
refactoring : revert the changes for the benchmarking protocol chapter

@@
 %% This is a conclusion maaan  !! 
-\section{Taming the energy variation for the benchmarking protocol}
+\section{Taming the energy variation for the benchmarking protocol}\label{sec:taming-the-energy-variation}
 
@@
 
-%% TODO: Proof read this paragraph
+\subsection{Docker \& Accuracy}
+As the state of the art assesses the impact of Docker on the energy consumption, one can also consider its impact on accuracy.
+In other words:\\
+\textbf{RQ:} does Docker affect the energy variation of the experiments?
+
+To answer this question we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
+Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
+In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
+The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
+
+\begin{figure}
+    \center{\includegraphics[width=.9\linewidth]{imgs/docvsbin}}
+    \caption{Comparing the variation of binary and Docker versions of aggregated \textsf{LU}, \textsf{CG} and \textsf{EP} benchmarks}\label{fig:docker}
+\end{figure}
 

### 2022-07-07 chapters/benchmarking/conclusion.tex
fix: bug when it incliudes a graphic
proof read :  chapter benchmarking

@@
 
-\section{Perspectives}
-By the end of this study, we have gathered enough guidelines to make the tests more reproducible, accurate.
+\section{Perspectives / representativeness}
+By the end of this study, we have gathered enough guidelines to make the tests more reproducible, accurate and extensible.
 We created a set of new tests named \textbf{energy tests} which are more similar to performance tests.
-Thanks to the work of two interns [ mamadou and adrien] we created a CI/CD platform to measure the energy consumption of Java projects and we could track the evolution of the this energy across different stages of the project.
+Thanks to the work of two interns [Mamadou and Adrien] we created a CI/CD platform to measure the energy consumption of Java projects and we could track the evolution of the this energy across different stages of the project.
 In the figure below we see an example of this plugin.
-For more details please visit the gitlab repository ... add link.
+For more details please visit the gitlab repository\footnote{\url{https://gitlab.inria.fr/mamdiall/J-Joules}}~\footnote{\url{https://gitlab.inria.fr/mamdiall/jjoules-plugin}} .
 

### 2022-07-07 chapters/benchmarking/draft.tex
proof read :  chapter benchmarking

@@
 \section{Motivation}
-Empirical measurements are critical tasks to capture the effect of developers' choice on software energy consumption.
+Empirical measurements are critical to capture the effect of developers' choices on software energy consumption.
 To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state of the art.
-Moreover, when it comes to measuring the energy consumption of software, other obstacles arise, such as the impact of hardware, the difference between the machines used for measurements purpose and the ones that may apply the findings, etc.
-Moreover, significant improvements in computer science have resulted in an increase in the quantity of obsolete results.
-It is even more difficult to address all of them in studies that compare multiple solutions.
-Furthermore, there may have been the appearance of new candidates, as well as the development of others between the preliminary experiment and the published results.
-As a result, it is not only required to ensure the "reproducibility" of the results, but also to make way for new candidates.
-Hence, we would like to propose a new concept to leverage experiments, which we call "Extension".
-The capacity to provide the tools required not only to repeat the experiments, but also to include additional candidates, workloads, or relevant key performance indicators.
+Furthermore, when it comes to evaluating the energy consumption of software, other challenges occur, such as the impact of hardware, the difference between the machines used for measurements and those that may apply the results, and so on.
+Moreover, considerable advances in computer science have resulted in an increase in the number of obsolete results.
+Addressing all of them becomes significantly more difficult in studies that compare multiple treatments.
 
-This chapter discusses how to solve the challanges of empirical analysis when it comes to the energy consumption experiments,
+Furthermore, between the exploratory experiment and the published results, there may have been the appearance of new candidates as well as the evolution of others.
+As a result, it is not only necessary to assure the "reproducibility" of the results, but it is also essential to provide room for new candidates.
+As a result, we'd like to present "Extension," a new approach for leveraging experimentation.
+The ability to give not only the tools needed to replicate the tests but also to integrate more candidates, workloads, or important key performance indicators.
 
-first we will dicuss how to ensure the three aspects of successful benchmark within the energy consumption. 
-So section~\ref{sec:benchmarking_reproducibility} will focus on the "reproducibility" aspect and how to provide reproducible experiments without impacting the energy measurement. 
-Section~\ref{sec:taming-the-energy-variation} will be about the accuracy when it comes to software energy consumption. and finally 
-Section~\ref{sec:extension} will be about the tools to include new candidates, workloads, or relevant key performance indicators.
+This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
+First, we'll go through the three components of a successful benchmark within the energy consumption experiment.
+Section 1 will therefore concentrate on the "reproducibility" issue and how to deliver reproducible experiments without interfering with the energy measurement.
+Section 2 will discuss accuracy in software energy consumption.
+Finally, Section 3 will discuss tools for introducing new candidates, workloads, or other key performance indicators.
 
-As for "representativeness" of the benchamrks, we will discuss it further in the Perspectives section.
+Regarding the "representativeness" of the benchmarks, we will elaborate in the section entitled Perspectives.
 
+This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
+First, we'll go through the three components of a successful benchmark within the energy consumption experiment.
+Section~\ref{sec:benchmarking_reproducibility} will therefore concentrate on the "reproducibility" issue and how to deliver reproducible experiments without interfering with the energy measurement.
+Section~\ref{sec:taming-the-energy-variation} will discuss accuracy in software energy consumption.
+Finally, Section~\ref{sec:extension} will discuss tools for introducing new candidates, workloads, or other key performance indicators.
+
+Regarding the "representativeness" of the benchmarks, we will elaborate in the section entitled Perspectives.
 %% this is might be the part that treats reproducibility for my case
@@
 later we will talk about the energy metrics and how to ensure the accuracy when it comes to the measurement of the energy consumption.
-This section thus discusses different ways to enforce the reproducibility of experiment by comparing different ways to encapsulate the systems-under-test.
+This section discusses various techniques to encapsulate the systems-under-test in order to assure the reproducibility of an experiment.
 
 \paragraph{Virtual Machines}
-First option should be using \emph{Virtual Machines} (VM), which gives researchers the freedom to choose the most appropriate tools, software and operating system that they are the most comfortable with, without paying the price to change the actual working environment, which will give them eventually more control over the dependencies and the execution environment.Therefore, Adding a new candidate is as simple as adding a new VM to the experiment.
-Moreover, using a VM addresses the \emph{replication crisis} thanks to the virtual images, as even the most complex architecture can be reproduced easily by instantiating a copy of the image. 
-However, this choice comes with a certain cost.
-Because of the hypervisor, software will build on two kernels---the virtual machine (guest) one and the host machine one---which will provide a noticeable overhead, and will impact the performances of the system-under-test.
-Therefore, we cannot use VM for experiments that are related to performance.
-Another limitation of VM is the the isolation.
-While this feature prevents the experimental environment from any undesirable interference from the outside world, this interaction may be required---especially when the experiment is dependent to an external
+FThe very first choice should be to use \emph{Virtual Machines} (VM), which allow researchers to select the most appropriate tools, software, and operating system that they are most comfortable with without incurring the cost of changing the actual working environment, giving them more control over dependencies and the execution environment.
+Furthermore, adopting a VM addresses the \emph{replication crisis} since virtual images allow even the most sophisticated architecture to be simply replicated by instantiating a copy of the image.
+
+This option, however, comes at a cost.
+
+Because of the hypervisor, software will be built on two kernels: one for the virtual machine (guest) and one for the host machine, resulting in a visible overhead and a negative influence on the performance of the system-under-test.
+
+As a result, we cannot use VM for performance-related tests.
+
+Isolation is another drawback of VM.
+
+While this feature protects the experimental setting from unwanted interference from the outside world, it is possible that this interaction is required—-especially if the experiment is dependent on an external source such as energy montitors .
 
 \paragraph{Containers}
-Another solution would be using something that allows us to benefit from the isolation of the host OS, while easing of replication proposed by VM, and the direct interaction with the hardware that the classical method give.
-Containers offer such an advantage, while ensuring the isolation and the ease of replication for application.
+Another option would be to use something that allows us to benefit from the host OS's isolation while simultaneously simplifying replication as proposed by VM and the direct interface with the hardware provided by the traditional techniques.
 
-Figure~\ref{environement:virtualization_technique} explains the differences in architecture between the types of virtualization and container technologies:
+Containers provide such an advantage by ensuring application separation and ease of replication.
+Figure~\ref{environement:virtualization_technique}depicts the architectural differences between virtualization and container technologies. There are three main types of virtualisation.
 \begin{itemize}
-    \item \textsf{Type\,1} runs directly on the hardware, it is mainly used by the cloud providers where there is no host OS, but just VM, use the open-source Xen or VMware ESX hypervisors to be executed;
-    \item \textsf{Type\,2} runs over the host OS, mostly used for personal computers.
-    VMware server and virtualBox are popular examples of this type, and most of the researchers experimentation are run with this type. However, due to the 2 OS, the applications tend to be slower.
-    \item \textsf{Containers}. Instead of its own kernel, containers use the host kernel to run their OS, which makes them lighter, faster and use the full potential of hardware.
-    For this category, one can cite \emph{Docker}, \emph{Linux LXC}, or \emph{LXD}~\cite{abuabdo_virtualization_2019}.
+    \item \textsf{Type\,1}: runs on the hardware directly.
+          It is primarily utilized by cloud providers where there is no host OS and only VMs that run on the open-source Xen or VMware ESX hypervisors.
+
+    \item \textsf{Type\,2}: runs on top of the host operating system and is most commonly found on personal computers.
+
+          VMware server and virtualBox are notable examples of this category, and most researchers' experiments use them.
+          However, because of the two operating systems, the applications are typically slower.
+
+    \item \textsf{Containers}: run their operating systems on the host kernel rather than their own , which makes them smaller, faster, and more efficient in terms of hardware utilization.
+          one can cite \emph{Docker}, \emph{Linux LXC}, or \emph{LXD}~\cite{abuabdo_virtualization_2019}.
 \end{itemize}
@@
 \subsection{Docker Vs. Virtual Machine}
-Despite the fact that \textsf{Type\,1} is more performant than \textsf{Type\,2}, the second one is the most used in research, as most researchers tend to conduct their experiments in their own machine.
-In the other hand, Docker is the most famous container technology.
-In our case, we are more prone to promote Docker for two reasons:
+Despite the fact that Type 1 is more performant than Type 2, the latter is the most used in research, as most researchers tend to conduct their experiments on their own machines.
+Docker, on the other hand, is the most well-known container technology.
+In our case, we are more likely to promote Docke for two reasons:
 \begin{enumerate}
-    \item we need a lightweight orchestrator to limit the overhead on energy consumption of our tests, as prior work mentioned
+    \item As previously stated [143] [reference Morabito (2015) and van Kessel et al. (2016)], we require a lightweight orchestrator to limit the overhead on energy usage of our experiments
           \cite{van2016power}
@@
           % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
-    \item as we are using hardware sensors to measure the energy consumption, we need to interact with the host OS.
+    \item We need to communicate with the host OS because we are using hardware sensors to measure the energy consumption..
 \end{enumerate}
 
-Special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that allows us to retireve the energy consumption of a virtual machine.
+A special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that enables us to retrieve the energy consumption of a virtual machine.
 
 \subsection{Docker \& Energy}
-Now that we have chosen to go with the containers technology to encapsulate our tests.
-What would be the impact of this solution on the energy consumption of our tests.
+Now that we've decided to use container technology to enclose our tests, what effect will this have on the amount of energy our tests use?
+%%%%%%%%%
 
-Based on the studies of \cite{santos2018does}, who analysed the impact of adding the Docker layer on the energy consumption.
-In their experiment, Eddie Antonio~\emph{et~ al.} run multiple benchmarks with and without Ddocker. 
-They compared the resulting energy consumption and execution time.
-The first step was to see the impact of Docker deamon, while there is no work, to observe the impact of the orchestrator.
-Then, they conduct their experiments with the following benchmarks:
+Using research from \cite{santos2018does} who examined how adding the Docker layer affected energy consumption.
+\citeauthor{santos2018does}~\emph{et~ al.}conducted their experiment by running numerous benchmarks both with and without Ddocker.They contrasted the energy usage and execution time that resulted.
+The first step was to observe the effects of the orchestrator and the Docker deamon while there was no work to be done.
+Then, they use the following benchmarks in their experiments:
+%%
+
 \begin{itemize}
@@
 
-The following figures represents the energy consumption of the system-under-test while it is idle.
-As one can observe in Figure~\ref{fig:docker_idle}, Docker brings an overhead of around $1,000$ Joules.
+The values below show the system under test's energy consumption while it is idle.
+Docker has an overhead of roughly $1,000$  joules, as seen in Figure~\ref{fig:docker_idle}
+
+Yet, as seen in Figure 3.3, Docker increased the execution duration of the benchmark by 50 seconds, increasing energy usage.
+According to the authors, this overhead is primarily caused by the Docker deamon and not by the fact that the
 
@@
 
-In the other hand, as one can see in Figure~\ref{fig:docker_reddis}.
-Docker increased the execution time of the benchmark by 50 seconds, hence increasing the energy consumption.
-The authors reported that this overhead is mostly due to the Docker deamon and not the fact that the application is in a container.
-Moreover, they estimated the cost of this extra energy and it was less than 0.15\$ in the worst case, which is non-significant compared to the benefits that Docker brings for isolation and reproducibility.
+Yet, as seen in Figure~\ref{fig:docker_reddis}, Docker increased the execution duration of the benchmark by 50 seconds,which led to a significant rise in the energy usage.
+According to the authors, this overhead is primarily caused by the Docker deamon and not by the fact that  the application is in a container.
 
-To summarize, Docker-based software tends to consume more energy, because mainly they take more time to be executed.
-The average power consumption is higher with only \textbf{2\,Watts}, due to the execution of the Docker deamon.
-This overhead can reach up to 5\% for IO-intensive application, but it is merely noticeable when it comes to CPU- or DRMA-intesive workloads.
+Furthermore, they calculated the cost of this extra energy, which was less than 0.15\$ in the worst-case scenario, which is insignificant in comparison to the benefits of Docker for isolation and reproducibility.
 
-\begin{figure}
+To summarize, Docker-based software tendst to consume more energy since it takes longer to execute.
+The execution of the Docker deamon causes an increase in average power consumption of only \textbf{2\,Watts}.
+This overhead can reach up to 5\% in IO-intensive applications, while it is barely visible in CPU- or DRAM-intensive workloads.
+
+\begin{figure}[!bht]
     \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_time}
@@
 \subsection{Conclusion}
-As one can see, Docker impact on energy is more constat due to the introduction of the supervisor.
-which will be applied to all the experimenets evenly.
+As can be seen, the introduction of the supervisor has increased the impact of Docker on energy.
+which will be applied equitably to all experiments.
 Therefore when it comes to comparison analysis, it will mitigate its impact autimatically.
-Moreover since we have access to the host Hardware, we dont have to worry about catching the energy consumption for the SUT. 
-Therefore we will be conducting all our experiments using docker as an encapsuation technology. wich will ensure the reproducibility of our experiments with a simple and easy way.
+Furthermore, because we have access to the host Hardware, we do not need to worry about capturing the SUT's energy use.
+As a result, we will use Docker as an encapsuation solution for all of our experiments.
+It would ensure the reproducibility of our experiments in a simple and straightforward manner.
+\clearpage

### 2022-07-07 chapters/benchmarking/experiment.tex
proof read :  chapter benchmarking

@@
+\newpage
 % To satisfy those three requirements. we elaborated a benchmarking protocol that aims to help reaserchers and practicioners in their experementations.
-
 % for each aspect we present a set of guidelines and tools.
 %% This is a conclusion maaan  !! 
-\section{Taming the energy variation for the benchmarking protocol}\label{sec:taming-the-energy-variation}
+\section{Improving accuracy by Taming the energy variation}\label{sec:taming-the-energy-variation}
 
-\subsection{Objective}
-While the previous part aims to create a large umbrella that covers all empirical benchmarks related to computer science, we rather aim to narrow our study to energy benchmarking only.
-In other words, we will focus on pitfalls that hinders the energy claims of software.
-We are clearly aware of the impact of hardware on energy variations.
-However, we believe that there is still a room to control this energy variation for practitioners by using only parameters that can be tuned.
-To do so, we have inducted a set of empirical experiments using the guidelines provided by state of the art, to determine which controllable factors can reduce the variation of the energy consumption of benchmarks.
+While the previous section aimed to ensure the reproducibility of our software energy consumption experiments, this section will provide a collection of tips and tools to help you increase the accuracy of these experiments.
+We are well aware of the effect that hardware has on energy fluctuations.
+However, we feel that there is still an opportunity for practitioners to minimize this energy variance by employing solely tuneable factors.
+To that end, we conducted a series of empirical studies utilizing state-of-the-art recommendations to discover which controllable elements can limit the variations of benchmark energy usage.
 % We will first start with evaluating the parameters mentioned by \cite{heinrich_predicting}.
 
-In this part we will try to answer the following Research questions.
-
+In this section we will try to answer the following Research questions.
 \begin{description}
@@
     \item[\textsc{RQ}~2:] How important is the impact of the processor features on the energy variation?
-    \item[\textsc{RQ}~3:] What is the impact of the operating system on the energy variation? and finally
-    \item[\textsc{RQ}~4:] Does the choice of the processor matter to mitigate the energy variation?
+    \item[\textsc{RQ}~3:] What effect does the operating system have on energy variation?
+    \item[\textsc{RQ}~4:] Does the choice of processor make a difference in reducing the energy variation?
 \end{description}
@@
 \subsection{Experimental Setup}\label{subsec:setup}
-This section describes our detailed experimental environment, covering the clusters configuration and the benchmarks we used  justifying our experimental methodology.
+This section describes our detailed experimental environment, covering the cluster configuration and the benchmarks we used to justify our experimental methodology.
 
@@
 \begin{itemize}
-    \item Different machines with different configuration
-    \item Different machines with the same configuration
-    \item Same machine
+    \item different machines with different settings;
+    \item different machines with the same settings ;
+    \item the same machine. 
 \end{itemize}
 
-To satisfy those requirements we have used the platform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708},a large-scale and flexible testbed for experiment-driven research distributed across all france.
-Grid5000 offers  multiple clusters composed with 4 up to 124 identical machine with different configurations for each cluster.
-For our experiment, we have considered 4 clusters.
-Our main criterions was the CPU Configuration.
-the table below~\ref{table:g5k}, presents a description of the 4 clusters.
+We used the platform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708}, which is a large, flexible testbed for experiment-driven research that is spread out across all of France, to meet these needs.Grid5000 has several clusters that are made up of 4 to 124 identical machines with different configurations for each cluster.For our experiment, we looked at four groups.
+Our main criterion was the CPU configuration.
+The table below~\ref{table:g5k}, presents a description of the four cluster that have been chosen for our experiments.
 
@@
 As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or \textsf{socket} to designate one of the two sockets and \textsf{PU} for the \emph{processing unit}.
-For our study we consider hyper-threads as distinct \textbf{PU}
+For our study we consider hyper-threads as distinct \textbf{PUs}.
 
-As an example, Figure~\ref{fig:topo} illustrates a detailed topology of a node belonging to the cluster \textsf{Dahu}.
+Figure~\ref{fig:topo}  shows the detailed topology of a node in the cluster \textsf{Dahu}. as an example.
 
@@
 To satisfy those criterions we went through state of the art and looked for the most used benchmarks for testing the performance of the hardware, and then we selected the scalable ones.
+%%%%%% 
+We picked the benchmarks based on two criteria.
+
+First, \textbf{scalability}: We wanted to learn as much as possible about the experiment in the time we had, so we needed some benchmarks that could grow or shrink with the number of PUs and work in different situations.
+
+The second criterion is whether or not the workload is \textbf{representative}.
+As stated in the challenges, a workload must be representative or the experiment would be inconsistent~\cite{stephen_evaluate_2012}.
+
+To meet these criteria, we looked at the "state of the art" and found the most common benchmarks used to test the performance of hardware. We then chose the ones that could be scaled up. 
 Our candidate is  emph{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925}: one of the most used benchmarks for \emph{HPC}.
@@
 \subsubsection{Metrics \& Measurement Tools}
-Our metric for the accuracy of the test is the Standard deviation aka \textbf{STD} of the energy consumption. Therefore wether the tests consumes more or less energy is out of our scope.
+Our metric for the accuracy of the test is the Standard deviation aka \textbf{STD} of the energy consumption. Therefore whether the tests consumes more or less energy is out of our scope.
 To study this variation we need first a tool to measure the energy consumption.
@@
         \hline
-        Use the least of sockets in a case of multiple CPU   & Medium        & Up to $30\times$\\
+        Use the least of sockets in a case of multiple CPU                                                                                                                & Medium        & Up to $30\times$  \\
         \hline
-        Avoid the usage of hyper-threading whenever possible & Medium        & Up to $5\times$\\
+        Avoid the usage of hyper-threading whenever possible                                                                                                              & Medium        & Up to $5\times$   \\
         \hline
@@
         \hline
-        Do not relate to the machine idle variation to isolate a test EC, the CPU/OS changes its behavior when a test is running and can exhibit less variation than idle & Any & ---\\
+        Do not relate to the machine idle variation to isolate a test EC, the CPU/OS changes its behavior when a test is running and can exhibit less variation than idle & Any           & ---               \\
         \hline

### 2022-07-07 chapters/benchmarking/draft.tex
fix : repeated paragarpahs

@@
 As a result, we'd like to present "Extension," a new approach for leveraging experimentation.
-The ability to give not only the tools needed to replicate the tests but also to integrate more candidates, workloads, or important key performance indicators.
-
-This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
-First, we'll go through the three components of a successful benchmark within the energy consumption experiment.
-Section 1 will therefore concentrate on the "reproducibility" issue and how to deliver reproducible experiments without interfering with the energy measurement.
-Section 2 will discuss accuracy in software energy consumption.
-Finally, Section 3 will discuss tools for introducing new candidates, workloads, or other key performance indicators.
-
-Regarding the "representativeness" of the benchmarks, we will elaborate in the section entitled Perspectives.
+The ability to give not only the tools needed to replicate the tests but also to integrate more candidates, workloads, or important key performance indicators
 
@@
 \paragraph{Virtual Machines}
-FThe very first choice should be to use \emph{Virtual Machines} (VM), which allow researchers to select the most appropriate tools, software, and operating system that they are most comfortable with without incurring the cost of changing the actual working environment, giving them more control over dependencies and the execution environment.
+The very first choice should be to use \emph{Virtual Machines} (VM), which allow researchers to select the most appropriate tools, software, and operating system that they are most comfortable with without incurring the cost of changing the actual working environment, giving them more control over dependencies and the execution environment.
 Furthermore, adopting a VM addresses the \emph{replication crisis} since virtual images allow even the most sophisticated architecture to be simply replicated by instantiating a copy of the image.
@@
 
-Yet, as seen in Figure 3.3, Docker increased the execution duration of the benchmark by 50 seconds, increasing energy usage.
-According to the authors, this overhead is primarily caused by the Docker deamon and not by the fact that the
-
 \begin{figure}

### 2022-07-07 chapters/benchmarking/experiment.tex
fix : repeated paragarpahs

@@
 \subsubsection{Workload}
-Our choice for the benchmarks was based on two criterions.
 
-First, \textbf{scalability} : We wanted to gather the highest possible amount insights (regarding time spent for the experiment, therefore we wanted some benchmarks who can scale according to the number of PUs and can fulfill different scenarios.
-Second criterion is \textbf{representativeness} : as mentioned in the challenges, a workload has to be representative otherwise the experiment would be inconsistent \cite{stephen_evaluate_2012}.
-To satisfy those criterions we went through state of the art and looked for the most used benchmarks for testing the performance of the hardware, and then we selected the scalable ones.
-%%%%%% 
 We picked the benchmarks based on two criteria.
@@
 First, \textbf{scalability}: We wanted to learn as much as possible about the experiment in the time we had, so we needed some benchmarks that could grow or shrink with the number of PUs and work in different situations.
-
 The second criterion is whether or not the workload is \textbf{representative}.
-As stated in the challenges, a workload must be representative or the experiment would be inconsistent~\cite{stephen_evaluate_2012}.
+As stated in the challenges, a workload must represent the production enviromenet or the experiment would be inconsistent~\cite{stephen_evaluate_2012}.
 

### 2022-07-07 chapters/programming_languages/programming_languages.tex
fix : repeated paragarpahs

@@
 \label{chapter:porgramming_langauges}
-In this chapter, we discuss the impact of the choice of the programming language on the energy consumption of the resulting software.
-To do so, we suggest to start with the general micro benchmarking and see how each programming languages interact with the CPU/Memory.
-
-The ultimate goal of this chapter is to provide a guideline on which programming language the developers should chose based on the charecteristics of the project in order to minimize the energy foot print of their product.
-No answer is evidedent for a such question. However, we can extract some feartures of each programming langues such as
+In this chapter, we talk about how the programming language affects how much energy the software consumes.We suggest starting with general microbenchmarking and watching how each programming language performs with the CPU and memory.The main goal of this chapter is to advise developers on how to choose a programming language based on their project's needs in order to make their product use the least amount of energy possible.
+For such a question, no answer is evidedent.But there are some features we can take from each programming language, such as:
 \begin{itemize}
@@
     \item memory usage
-    \item etc
 \end{itemize}
 
-first we will start but analysing the behaviour of the general purpose programming langauges with some micro benchkaks, principally for the CLBG game and others from rosetta code base
-
-As we have seen in the previous chapter, one of the most important feature of a test is to be \textbf{representative}.
-Therefore, we extend this study to some real-life use case, and the following sections below provide two case studies.
+As we saw in the last chapter, one of the most important things about a test is how well it \textsc{represents} the production environment.
+Therefore, we extend this study to include real-world use cases, with two case studies provided in the parts that follow.
 \section{Remote Procedure Call Frameworks}

### 2022-07-08 chapters/optimization/python-dir/python.tex
add : ml figures

@@
 
-When it comes to developing software systems, one can notice that---except Perl---dynamic programming languages have taken over compiled languages along the last decade in terms of developers popularity (cf.
+Dynamic programming languages, with the exception of Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf.
 Figure~\ref{fig:pypl}).
-However, it remains unclear if this class of dynamic programming languages can reasonably compete with compiled ones when it comes to power efficiency.
-
+However, it remains unclear whether this category of dynamic programming languages can truly compete with compiled ones in terms of power consumption. 
 \begin{figure}[htbp]
@@
 one kind of those optimizations includes alternative interpertersand some libraries that are dedicated to optimize the code without changing its structures such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
+
+%%%%%%%%%%% READ AGAIN 
+articular, Noureddine in 2012 and Pereira in 2017 conducted empirical power measurements on this topic, and both concluded that compiled programming languages consume less power than dynamic ones.They found that a programming language that is interpreted, like Python, can use up to 7,584 times more energy than C.
+As a result, in this chapter, we will look at the obvious optimizations that may be done to Python legacy programs to lower their energy footprint.Since Python is used by many software services in both public and private cloud infrastructures, we think that our contributions will help a wide range of legacy systems by reducing ICT carbon emissions and lowering cloud bills for the resources used by these services.
+This chapter focuses on runtime improvements that developers can employ to reduce the power consumption of Python applications.We start by looking at how the choices made by programmers, like the type of data structure or loop, affect how much energy the code as a whole uses.
+Then we discuss additional topics, such as the level of concurrency, and so on.We will discuss another non-intrusive technique to optimize this code's energy consumption.Alternative interpreters and libraries dedicated to optimizing code without affecting its structures, such as ahead-of-time (AOT) compilation and just-in-time (JIT) libraries maintained by the community, are examples of these optimizations.
 % However, the operational conditions under which these measurements may threaten the validity of their results.
@@
 
-\begin{figure}[thb]
+\begin{figure}
     \centering
@@
 most of the buitl-in functions an operations are written in C, same thing goes for the function range, further more the function len has a complexity of O(1) because it is based on the function Py\_SIZE  of C which stores the length in a field for the object.
-
 so basically for the for in range is basically creating a new iterator that has the same length of the first one.
 and for each iteration we will have to do a second acceess ( l[i]) instead of one one access hence the double time.
-
 for the wile is even slower due to the implicite incrementation of the variable which will cause an extra operation during the loop .
-
 to confirm the hyphothsis.
@@
 
-\begin{figure}[htb]
+\begin{figure}
     \centering
@@
 
+\subsection{Python and machine learning }
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/accuracy_basedonepoch}
+    \caption{accuracy based on epoch  }
+    \label{fig:p2}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/power_gpu_baedonepoche}
+    \caption{evolution of average gpu power based on epoch  }
+    \label{fig:p2}
+\end{figure}
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/cumulative_energy_fast10}
+    \caption{cumulative energy of fast10 benchmark within different configurations }
+    \label{fig:p2}
+\end{figure}
+
+\paragraph{conclusion}
+We discovered that when a model's accuracy improves, so does the energy required for a subsequent accuracy gain.
+This begs the question of when we should discontinue training.
+Is a 10\% increase in accuracy worth it if we have to spend three times the energy?
+Even while utilizing a GPU consumes more total power, the training time is significantly reduced.
+As a result, the utilization of a GPU is required to lower the energy consumption of the training.
+Some experiments revealed unusual behavior that was not studied during this internship.
+These could be examined in future studies and perhaps used to achieve more energy savings.
+The execution time did not depend on the number of cores employed in several studies.
+This parameter had a significant impact on training time in others.
+As a result, the best core configuration will be determined by the script.
+The next steps in this area could include exploring the reasons for these discrepancies and possibly creating a script that can automatically find the ideal core configuration for a specific script.
+
 \subsection{python runtimes}
-
-As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
-Intelpython and active python both folows the same behaviour. one can conclude that the work that have been on those interpreters is mainly
-to improve a specific purpose, Active python claims that their version is focused on secuirty which explains the lack of some performances due to the introduction of more reonforcment
-.Morver Intel published their version of python as a dedicated for machine learning. Unfortunately the tommti benchmark is a set that focuses on the general purpose programming
-which does not reflect the performance of the machine learning.
-Another aspect of a such behaviour might be due to the fact of the processors that were used in the testbed.
-and it might change in the future if we include the GPU part for the machine learning benchmarks
-% reference the work of the intern 
-As for nuitka. there were no optimization in the energy consumpotion despite the fact that is it a complier.
-However if we dig through the nuitka mechanisms. they basically embed the python code with an interpreter.
-Unllike nuitka shedskin exhebit a the best energy consumption pattern when it comes to the arithmetic operartions.
-One can conclude it is due to the fact of the native type of the variables.
-unlike the interpreters where they are treated as object in the begining.
-
-for the other interpeters pypy is very promising especially when it comes to data maniupulation as one can see in the figure \ref{fig:bar_tommti_vector}
-pypy is by far the best interperer when it comes to treating vectors.
-numba2 introduced the JIT but wasn't as promising as numba3.
-
-for the other vm based interpereters.
-jython and ipy lacked in term of energy optimisation which was kinda expected since they were in their the begining of the stage and the main puprose of such implementation is to link the bytecode generated by jython and ironpython with their respectives virtual machin.
-
-Unlike the the previous interpeters. graal exhebits a certin promises when it comes to complex algorithms - nested loops -
-micro python is dedicated to embeded systems so lunching it is powerful cpu machines will be misleading.
-
-Most of the interperters had the same behaviour when it comes to the input outputs. except for jython which was kinda of a anomaly probably due to the lack of the optimizations.
-% pypy handles exceptions very well
 
@@
 \end{figure}
-\begin{figure}[thb]
+\begin{figure}
     \centering
@@
 
+As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
+Intelpython and active python both folows the same behaviour. one can conclude that the work that have been on those interpreters is mainly
+to improve a specific purpose, Active python claims that their version is focused on secuirty which explains the lack of some performances due to the introduction of more reonforcment
+.Morver Intel published their version of python as a dedicated for machine learning. Unfortunately the tommti benchmark is a set that focuses on the general purpose programming
+which does not reflect the performance of the machine learning.
+Another aspect of a such behaviour might be due to the fact of the processors that were used in the testbed.
+and it might change in the future if we include the GPU part for the machine learning benchmarks
+% reference the work of the intern 
+As for nuitka. there were no optimization in the energy consumpotion despite the fact that is it a complier.
+However if we dig through the nuitka mechanisms. they basically embed the python code with an interpreter.
+Unllike nuitka shedskin exhebit a the best energy consumption pattern when it comes to the arithmetic operartions.
+One can conclude it is due to the fact of the native type of the variables.
+unlike the interpreters where they are treated as object in the begining.
+
+for the other interpeters pypy is very promising especially when it comes to data maniupulation as one can see in the figure \ref{fig:bar_tommti_vector}
+pypy is by far the best interperer when it comes to treating vectors.
+numba2 introduced the JIT but wasn't as promising as numba3.
+
+for the other vm based interpereters.
+jython and ipy lacked in term of energy optimisation which was kinda expected since they were in their the begining of the stage and the main puprose of such implementation is to link the bytecode generated by jython and ironpython with their respectives virtual machin.
+
+Unlike the the previous interpeters. graal exhebits a certin promises when it comes to complex algorithms - nested loops -
+micro python is dedicated to embeded systems so lunching it is powerful cpu machines will be misleading.
+
+Most of the interperters had the same behaviour when it comes to the input outputs. except for jython which was kinda of a anomaly probably due to the lack of the optimizations.
+% pypy handles exceptions very well
+
 \section{Threads to validity}
@@
 \section{conclusion}
-
-one can notice that the choice of the right interpeter can impact havely the energy consumption of the python programs.
-The fact that there is no general solution make this study more interesting.
-however the main drawback is the lack of the compatibility for some of those solutions which lead to some sacrifices when we want something generic
+One may observe that the choice of Python interpreter has a significant impact on the programs' energy consumption.
+This investigation is made more intriguing by the absence of a universal solution.
+The primary downside is the incompatibility of some of these solutions, which causes us to make concessions when we need a generic answer.

### 2022-07-10 chapters/benchmarking/conclusion.tex
correections : chapter 3 part 1/2

@@
-\subsection{Extension}
+\subsection{Extension}\label{sec:bench_extension}
 We propose to extend the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
-Instead of only presenting the four primary aspects of their guidelines that were mentioned in state of the art~\ref{sec:soa_reproducibility}, we propose an abstract model to describe an empirical experiment:
+Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in section ~\ref{sec:soa_reproducibility}, we propose an abstract model to describe an empirical experiment:
 
@@
 
-Workload interface  : Wich provides a set of functions that should be implemented by the workload  and called by the each candidate in the experiment. 
-Observer interface : provides a set of metrics that are collected by the orchestrator during the expermentation.
+\emph{Workload interface}: provides a set of functions that should be implemented by the workload  and called by the each candidate in the experiment. 
+Observer interface: provides a set of metrics that are collected by the orchestrator during the expermentation.
 note that some of those metrics are external, such the energy consumption of the system , execution time, execution stack ..etc .
 
-Figure\ref{fig:benchmarkingprotocol} explains in detail the relation ship within each component of the Protocol. 
+Figure~\ref{fig:benchmarkingprotocol} explains in detail the relationship within each component of the protocol. 
 First of all, we should start with the orchestrator which is the kernel of the experiment. Most of the time this kernel should represent the constants parts of the experiment. 
@@
 Later this interface will be provided to each candidate by the orchestrator.
-Like this we will remove any bias of dependecy between the workload and the candidates, which will alow us to extend both parts without any change in the code.
+Like this we will remove any bias of dependency between the workload and the candidates, which will alow us to extend both parts without any change in the code.
 Instead of retrieving the data directly from the experiment, we suggest to use an observator to allow the extension of the metrics, this observer in general should be outside the experiment to be able to gather as many informations without impacting the the code.
@@
 \end{figure}
+
+%% TODO : make a generic conclusion , 
+%% maybe add extension as a part of the conclusion for this chapter
 
@@
 
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/JunitSonarplugin}}
-    \caption{Example of the Junit Sonar Plugin}\label{fig:sonar_plugin}
-\end{figure}
+% \begin{figure}%[!htb]
+%     \center{\includegraphics[width=.9\linewidth]{imgs/JunitSonarplugin}}
+%     \caption{Example of the Junit Sonar Plugin}\label{fig:sonar_plugin}
+% \end{figure}

### 2022-07-10 chapters/benchmarking/draft.tex
correction : chapter 3 part 1/2

@@
 Empirical measurements are critical to capture the effect of developers' choices on software energy consumption.
-To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state of the art.
+To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state of the art~\cite{van_der_kouwe_benchmarking_2018}.
 Furthermore, when it comes to evaluating the energy consumption of software, other challenges occur, such as the impact of hardware, the difference between the machines used for measurements and those that may apply the results, and so on.
@@
 Furthermore, between the exploratory experiment and the published results, there may have been the appearance of new candidates as well as the evolution of others.
-As a result, it is not only necessary to assure the "reproducibility" of the results, but it is also essential to provide room for new candidates.
+As a result, it is not only necessary to ensure the "reproducibility" of the results, but it is also essential to provide room for new candidates.
 As a result, we'd like to present "Extension," a new approach for leveraging experimentation.
@@
 This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
-First, we'll go through the three components of a successful benchmark within the energy consumption experiment.
-Section~\ref{sec:benchmarking_reproducibility} will therefore concentrate on the "reproducibility" issue and how to deliver reproducible experiments without interfering with the energy measurement.
-Section~\ref{sec:taming-the-energy-variation} will discuss accuracy in software energy consumption.
-Finally, Section~\ref{sec:extension} will discuss tools for introducing new candidates, workloads, or other key performance indicators.
+First, we  go through the three components of a successful benchmark within the energy consumption experiment.
+Section~\ref{sec:benchmarking_reproducibility}  therefore concentrates on the "reproducibility" issue and how to deliver reproducible experiments without interfering with the energy measurement.
+Section~\ref{sec:taming-the-energy-variation}  discusses accuracy in software energy consumption.
+Finally, Section~\ref{sec:bench_extension} presents tools for introducing new candidates, workloads, or other key performance indicators.
 
@@
 This option, however, comes at a cost.
-
 Because of the hypervisor, software will be built on two kernels: one for the virtual machine (guest) and one for the host machine, resulting in a visible overhead and a negative influence on the performance of the system-under-test.
-
 As a result, we cannot use VM for performance-related tests.
-
 Isolation is another drawback of VM.
-
 While this feature protects the experimental setting from unwanted interference from the outside world, it is possible that this interaction is required—-especially if the experiment is dependent on an external source such as energy montitors .
@@
 Containers provide such an advantage by ensuring application separation and ease of replication.
-Figure~\ref{environement:virtualization_technique}depicts the architectural differences between virtualization and container technologies. There are three main types of virtualisation.
+Figure~\ref{environement:virtualization_technique} depicts the architectural differences between virtualization and container technologies. There are three main types of virtualisation.
 \begin{itemize}
@@
     \item \textsf{Type\,2}: runs on top of the host operating system and is most commonly found on personal computers.
-
 VMware server and virtualBox are notable examples of this category, and most researchers' experiments use them.
@@
     \item \textsf{Containers}: run their operating systems on the host kernel rather than their own , which makes them smaller, faster, and more efficient in terms of hardware utilization.
-          one can cite \emph{Docker}, \emph{Linux LXC}, or \emph{LXD}~\cite{abuabdo_virtualization_2019}.
+One can cite \emph{Docker}, \emph{Linux LXC}, or \emph{LXD}~\cite{abuabdo_virtualization_2019}.
 \end{itemize}
@@
 Docker, on the other hand, is the most well-known container technology.
-In our case, we are more likely to promote Docke for two reasons:
+In our case, we are more likely to promote Docker for two reasons:
 \begin{enumerate}
@@
           % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
-    \item We need to communicate with the host OS because we are using hardware sensors to measure the energy consumption..
+    \item We need to communicate with the host OS because we are using hardware sensors to measure the energy consumption.
 \end{enumerate}
 
-A special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that enables us to retrieve the energy consumption of a virtual machine.
+%  TODO : Rephrase this 
+% A special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that enables us to retrieve the energy consumption of a virtual machine.
 
@@
 Using research from \cite{santos2018does} who examined how adding the Docker layer affected energy consumption.
-\citeauthor{santos2018does}~\emph{et~ al.}conducted their experiment by running numerous benchmarks both with and without Ddocker.They contrasted the energy usage and execution time that resulted.
+\citeauthor{santos2018does}~\emph{et~ al} conducted their experiment by running numerous benchmarks both with and without Ddocker.
+They contrasted the energy usage and execution time that resulted.
 The first step was to observe the effects of the orchestrator and the Docker deamon while there was no work to be done.
-Then, they use the following benchmarks in their experiments:
-%%
-
-\begin{itemize}
-    \item Wordpress,
-    \item Reddis,
-    \item PostgreSQL.
-\end{itemize}
-
+Then, they use three benchmarks in their experiments: Wordpress, Reddis and PostgreSQL.
 The values below show the system under test's energy consumption while it is idle.
-Docker has an overhead of roughly $1,000$  joules, as seen in Figure~\ref{fig:docker_idle}
+Docker has an overhead of roughly $1,000$  joules, as seen in Figure~\ref{fig:docker_idle} .
 
@@
 
-To summarize, Docker-based software tendst to consume more energy since it takes longer to execute.
+To summarize, Docker-based software tends to consume more energy since it takes longer to execute.
 The execution of the Docker deamon causes an increase in average power consumption of only \textbf{2\,Watts}.
@@
 \subsection{Conclusion}
-As can be seen, the introduction of the supervisor has increased the impact of Docker on energy.
-which will be applied equitably to all experiments.
-Therefore when it comes to comparison analysis, it will mitigate its impact autimatically.
+As can be seen, the introduction of the supervisor has increased the impact of Docker on energy that will be applied equitably to all experiments.
+Therefore when it comes to comparison analysis, it will mitigate its impact automatically.
 Furthermore, because we have access to the host Hardware, we do not need to worry about capturing the SUT's energy use.
 As a result, we will use Docker as an encapsuation solution for all of our experiments.
-It would ensure the reproducibility of our experiments in a simple and straightforward manner.
+It will ensure the reproducibility of our experiments in a simple and straightforward manner.
 \clearpage

### 2022-07-10 chapters/benchmarking/experiment.tex
correction : chapter 3 part 1/2

@@
 %% This is a conclusion maaan  !! 
-\section{Improving accuracy by Taming the energy variation}\label{sec:taming-the-energy-variation}
+\section{Improving accuracy by taming the energy variation}\label{sec:taming-the-energy-variation}
 
-While the previous section aimed to ensure the reproducibility of our software energy consumption experiments, this section will provide a collection of tips and tools to help you increase the accuracy of these experiments.
+While the previous section aimed to ensure the reproducibility of our software energy consumption experiments, this section provides a collection of tips and tools to help increase the accuracy of these experiments.
 We are well aware of the effect that hardware has on energy fluctuations.
@@
 
-In this section we will try to answer the following Research questions.
+In this section, we investigate the following research questions:
 \begin{description}
@@
 
-We used the platform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708}, which is a large, flexible testbed for experiment-driven research that is spread out across all of France, to meet these needs.Grid5000 has several clusters that are made up of 4 to 124 identical machines with different configurations for each cluster.For our experiment, we looked at four groups.
+We used the platform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708}, which is a large, flexible testbed for experiment-driven research that is spread out across all of France, to meet these needs.
+Grid5000 has several clusters that are made up of 4 to 124 identical machines with different configurations for each cluster.
+For our experiment, we looked at four groups.
 Our main criterion was the CPU configuration.
-The table below~\ref{table:g5k}, presents a description of the four cluster that have been chosen for our experiments.
+The table~\ref{table:g5k}  below, presents a description of the four clusters that have been chosen for our experiments.
 
@@
 
-Figure~\ref{fig:topo}  shows the detailed topology of a node in the cluster \textsf{Dahu}. as an example.
+Figure~\ref{fig:topo}  shows the detailed topology of a node in the cluster \textsf{Dahu} as an example.
 
@@
 The second criterion is whether or not the workload is \textbf{representative}.
-As stated in the challenges, a workload must represent the production enviromenet or the experiment would be inconsistent~\cite{stephen_evaluate_2012}.
+As stated in the challenges, a workload must represent the production environment or the experiment would be inconsistent~\cite{stephen_evaluate_2012}.
 
@@
 We used the applications (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
-Further more we have used other applications to validate our results using more applications such as \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}\footnote{\url{https://linux.die.net/man/1/sha256sum}}.
+Furthermore we have used other applications to validate our results using more applications such as \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}\footnote{\url{https://linux.die.net/man/1/sha256sum}}.
 
 \subsubsection{Metrics \& Measurement Tools}
-Our metric for the accuracy of the test is the Standard deviation aka \textbf{STD} of the energy consumption. Therefore whether the tests consumes more or less energy is out of our scope.
+Our metric for the accuracy of the test is the Standard deviation aka \textbf{STD} of the energy consumption. Therefore whether the tests consume more or less energy is out of our scope.
 To study this variation we need first a tool to measure the energy consumption.
@@
 Using Docker makes the deployment process easier on the one hand, and provides us with a built-in control group encapsulation of the conducted tests on the other hand.
-This allows \textsc{PowerAPI} to measures all the running containers, even the RAPL sensor consumption, as it is isolated in a container.
+This allows \textsc{PowerAPI} to measure all the running containers, even the RAPL sensor consumption, as it is isolated in a container.
 
@@
 As we study the STD difference of measurements we observed from empirical experiments, we use the bootstrap method~\cite{efron2000bootstrap} to randomly build multiple subsets of data from the original dataset, and we draw the STD density of those sets, as illustrated in Figure~\ref{fig:docker}.
-Given the space constraints, this paper reports on aggregated results for nodes, benchmarks and workloads, but the raw data we collected remains available through the public repository we published.\footnote{\url{https://github.com/anonymous-data/Energy-Variation}}
+Given the space constraints, we report on aggregated results for nodes, benchmarks and workloads.
+However, the raw data we collected is available through the public repository we published.\footnote{\url{https://github.com/anonymous-data/Energy-Variation}}
 We believe this can help to achieve better and more reliable comparisons.

### 2022-07-12 chapters/optimization/python-dir/python.tex
add : some notes

@@
 % \end{compactenum}
-
-%%%% python is a popular programming language  based TIOBE PYPI   add graphs
 
@@
 
+\section{placeholder}
+we performed a shapiro normality test for the first on the results, and almost all the p-values smaller than alpha=0.01\% 
+therefore the distrubution is not normal - which is kinda true because if it was normal then we dont have any comparaison --
+
+for arrays and vectors  graalpython took so much time that we decided to remove it
+
+the startup cost for ironpython was too high 
+for graalpython and ipy they couldnt handle big numbers hence the overflow  when it comes to execution time 
+but since we measure the energy outside the values of the energy aren't impacted
+
+as usual there is  correlaction between DRAM and CPU so no need to classify based on DRAM + the DRAM consumes less 10\% of the  energy compared to the CPU
+
+so the plan is to prove that there is a statistical difference 
+then we use the multiparameter optimization , howerver we stop in the phase where we calculate the score 
+because we dont know the weight of each parameter ( of tommti microbenchmark) and instead of doing the sum with the coefissients we use the radar plot to let the reader decide which one is the most suitable for his usage.
+howerver since the differences are gigantics we change the scale to logarithmic to be easier for the eye to read
+
 \begin{figure}

### 2022-07-14 chapters/optimization/python-dir/python.tex
update:  introduction of python

@@
 %%%%%%%%%%% READ AGAIN 
-articular, Noureddine in 2012 and Pereira in 2017 conducted empirical power measurements on this topic, and both concluded that compiled programming languages consume less power than dynamic ones.They found that a programming language that is interpreted, like Python, can use up to 7,584 times more energy than C.
+% articular, Noureddine in 2012 and Pereira in 2017 conducted empirical power measurements on this topic, and both concluded that compiled programming languages consume less power than dynamic ones.They found that a programming language that is interpreted, like Python, can use up to 7,584 times more energy than C.
 As a result, in this chapter, we will look at the obvious optimizations that may be done to Python legacy programs to lower their energy footprint.Since Python is used by many software services in both public and private cloud infrastructures, we think that our contributions will help a wide range of legacy systems by reducing ICT carbon emissions and lowering cloud bills for the resources used by these services.
@@
 
-Nowdays, Python seems to attract a large community of developers interested in data analysis, web development, system administration, machine learning---according to a survey conducted in 2018 by JetBrains---\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}}one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.
-
+Nowadays, Python seems to attract a large community of developers who are interested in data analysis, web development, system administration, and machine learning—according to a survey conducted in 2018 by JetBrains ---\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}}one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.
 As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
-Rather, we believe that a better option consists in leveraging the strength of this rich ecosystem to promote energy-efficient solutions in order to improve the power consumption of legacy software systems.
+Rather, we believe that a better option consists of leveraging the strength of this rich ecosystem to promote energy-efficient solutions in order to improve the power consumption of legacy software systems.
 
@@
 
-the simplicity, flexibility and the clarity of Python code led this programming language to a huge success.
-Therefore, a larger community and a wide range of support and libraries that allowed it to touch several fields like data analyses, machine learning, natural language processing, web development, astrophysics etc.
+In the other hand,because of the interpreted nature of Python code, this programming language has paid a high price in terms of performance.  \ref{fig:clbg}, memory and energy consumption.
+
 \begin{figure}[htbp]
@@
 
-However, and according to \cite{pinto_energy_2017} and \cite{noureddine_preliminary_2012}.
-Python tends to be one more energy hungry .
-In figure \ref{fig:hannoi}, python consumes 30 times more than C or C++.
+According to \cite{pinto_energy_2017} and \cite{noureddine_preliminary_2012}, Python tends to be one more energy hungry .
+As onne can notice in figure \ref{fig:hannoi}, python consumes 30 times more than C or C++.
 The test was done on an implementation of \fnurl{hannoi test}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi}  30 disks.
 
-Python consumes a lot of energy mainly because it is slow in execution.
-Its Flexibility and simplicity caused it to drop off in performances, because Python gains its flexibility from being a dynamic language, therefore it needs an interpreter to execute its programs, which makes them much slower compared to the other ones that are written in compiled programming languages such as C and C++ or semi-compiled languages like Java.
+Python consumes a lot of energy, mainly because it is slow in execution.
+Its flexibility and simplicity caused it to drop off in performance, because Python gains its flexibility from being a dynamic language.
+Therefore, it needs an interpreter to execute its programs, which makes them much slower compared to the others that are written in compiled programming languages such as C and C++ or semi-compiled languages like Java.
 
-As shown in Table \ref{fig:clbg}, one can see that in the most of the test cases Python takes more time to execute -the only case that he wasn't the last one was in the benchmark regx-redux where he beaten Go-, and in some cases the gab was huge such as in the n-body test where python took around 100 times more than C++.
-
+As shown in Table\ref{fig:clbg}, one can see that in most of the test cases, Python takes more time to execute-the only case that he wasn't the last one was in the benchmark regx-redux where he beat Go-, and in some cases the gap was huge, such as in the n-body test where Python took around 100 times more than C++.
 \begin{table}[hbt]
@@
     \end{tabular}
-    \caption{Comparison of the execution time between different programming languages using the computer language benchmarks game \cite{noauthor_which_nodate}}
-    %  TODO : ADD reference 
+    \caption{Comparison of the execution time between different programming languages using the computer language benchmarks game\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}}
     \label{fig:clbg}
@@
 
-\subsection{use cases}
-to reduce the energy consumption of python we started by targeting the main usage of this programming language, wich is revealed to be data sciences and web developpement.
-Figure~\ref{fig:usecase} illustrate a study made by the jetbrain company on python developers \footnote{\fnurl{https://www.jetbrains.com/lp/python-developers-survey-2020}}.
-In this survey, multiple multiple answers were accepted.
+\subsection{Goal}
+
+To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
+Figure~\ref{fig:usecase} illustrates a study made by the JetBrain company on Python developers.
+\footnote{\fnurl{https://www.jetbrains.com/lp/python-developers-survey-2020}}.In this survey, multiple answers were accepted.
+As one can see, 57\% of the respondents said that they use Python for data science, 51\% said they are using it for web development, and around 40\% said they are using it for system administration.
 \begin{figure}[hbt]
@@
     \includegraphics[width=\linewidth]{imgs/python_use_cases}
-    \caption{use cases of python }
+    \caption{Use cases of python }
     \label{fig:usecase}
@@
 
-\subsection*{Goal}
-To reduce the energy consumption of python programs, we decided to start with studying its behaviour during the most common use cases.
-After this we will targtet the structures of the language itself in the hope to find some generic guidelines for more generic purpose as \citeauthor{hasan_energy_2016-1} did in their paper \cite{hasan_energy_2016-1}.
-After this we will measure the energy consumption of different python implementation for the goal of finding a non intrusive way to enhance this energy efficiency.
+In this chapter, we will first investigate its behavior in the most common usage cases.Following that, we will focus on the language's structures in the hope of discovering some fundamental guidelines for more general purposes, as \citeauthor{hasan_energy_2016-1} did in their article~\cite{hasan_energy_2016-1}.
+Following that, we will measure the energy consumption of several Python implementations in order to determine a non-intrusive technique to improve energy efficiency.
+
+Therefore this chapter will target the following research questions :
+\begin{compactenum}[\indent\bf RQ\,1:]
+    \item \emph{What is the behavior of Python when it is used as a data science tool?}
+    \item \emph{Are the default guidelines of Python energy efficient?}
+    \item \emph{Can we reduce the energy consumption of Python programs without impacting the source code?}
+\end{compactenum}
 

### 2022-07-14 chapters/optimization/python-dir/python.tex
refactoring :python

@@
-%% TODO : refactor the experemental protocol to have the 3 main components measurement context, workload and metrics 
 \section{Introduction}
@@
 As onne can notice in figure \ref{fig:hannoi}, python consumes 30 times more than C or C++.
-The test was done on an implementation of \fnurl{hannoi test}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi}  30 disks.
+The test was done on an implementation of \url{hannoi test}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi}  30 disks.
 
@@
 
-As shown in Table\ref{fig:clbg}, one can see that in most of the test cases, Python takes more time to execute-the only case that he wasn't the last one was in the benchmark regx-redux where he beat Go-, and in some cases the gap was huge, such as in the n-body test where Python took around 100 times more than C++.
+As shown in Table\ref{fig:clbg}, one can see that in most of the test cases, Python takes more time to execute-the only case that he wasn't the last one was in the benchmark regx-redux where he beat Go-, and in some cases the gap was huge, such as in the n-body test where Python took around 100 times more than C++.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}
 \begin{table}[hbt]
@@
         spectral-norm      & 1.99 & 1.98  & 4.27 & 193.86 & 3.95  \\
-        Mandelbort         & 1.64 & 1.51  & 6.96 & 279.68 & 5.47
+        Mandelbort         & 1.64 & 1.51  & 6.96 & 279.68 & 5.47  \\
     \end{tabular}
-    \caption{Comparison of the execution time between different programming languages using the computer language benchmarks game\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}}
+    \caption{Comparison of the execution time between different programming languages using the computer language benchmarks game}
     \label{fig:clbg}
@@
 Figure~\ref{fig:usecase} illustrates a study made by the JetBrain company on Python developers.
-\footnote{\fnurl{https://www.jetbrains.com/lp/python-developers-survey-2020}}.In this survey, multiple answers were accepted.
+\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}.In this survey, multiple answers were accepted.
 As one can see, 57\% of the respondents said that they use Python for data science, 51\% said they are using it for web development, and around 40\% said they are using it for system administration.
@@
 
-%% should we add Research questions ?
+% \section{Experimental Protocol}\label{sec:pythonprotocol}
+% In this section, we discuss the experimental setting, including hardware components, software components, and the methodology that we adopted to answer the previous research questions.
+% For each of the questions that follow, we will first talk about how the experiment was done, and then we will talk about the results.
 
-\newpage
+% %% should i include the part with the exceptions ??? naah no need so
 
-\paragraph{Energy measurement}
-
-As we know, the energy of a program is the integrale of its power overtime.
-For us ower case, we used Intel \emph{Running Power Average Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754} to collect the power samples of the running tests,
-
-We used used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, to report Data collected by intel Rapl and send it into another machine that we call computing machine.
-then we calculate the Energy using  the the trapezoidal rule.
-
-\ref{fig:trapezrule}
-\begin{figure}[hbt]
-    \centering
-    \begin{equation}
-        E = \int^a_b P(t)dt \simeq \sum^n_{k=1} \frac{P(t_k-1)+P(t_k)}{2}
-    \end{equation}
-    \caption{}
-    \label{fig:trapezrule}
-\end{figure}
-
-figure \ref{fig:powerapi} shows the architecture of our testing model.
-
-The reason of separation between data collection and energy calculation is to minimise any interference with the test so our sensor is a light c program running inside a docker container.
-
-%% should i include the part with the exceptions ??? naah no need so
-
-%% except the parallellism part , which i dont know how to link it here yet 
-\section{Workload} :
-for this chapter we are consediring 4 seperated, yet related studies .
-First we are going to see the energy behavriour of python within two study cases.
+% %% except the parallellism part , which i dont know how to link it here yet 
+% \section{Workload} :
+% for this chapter we are consediring 4 seperated, yet related studies .
+% First we are going to see the energy behavriour of python within two study cases.
 Web servers and Machine learning.
 later we will dig deeper into the energy consumption of python basic structures where finally we will conclude with the impact of python interpterer on the energy consumption.
-Due to the lack of of support for most of  non conventional pyton interpeters, our main focus was only on micro benchmarks.
-Except pypy most of the python implementations don't support extra python libraries, Despite that most of the cases those extra implementations were made to optimize a specific library such as numba with numpy, or intelpython with machine learning algorthms.
-
-\paragraph{preliminary studies}
-for the first studies we used only the official version of python. because the goal was mainly to highlight the impact of the structure of code on the energy consumption.
-One main inconvienent of the previous methode to reduce the energy consumption, is the hard work that should be done in order to alter the existing code base in order to redcue the energy consumption.
-To avoid such hustle we tried to find a non intrusive way to make the python code more ecofriendly without changing its structure.
-One aspect of python is the fact that it is an interptered language which led many works to create their own implementation of the interpeter to enhance one or many aspect of the python code.
-in the following section we will discuss the impact of those implementations on the energy consumption of python programs, and in which case, one should use a non conventional interpter to save the energy consumption of their program.
-
-To do so, we gathered a list of interpeters, transpilers and some optimization libraries.
-
-the following list, presents the most famous ones .
-% We should explain the methodology we used to select these
-
-\begin{enumerate}
-    \item \textbf{CPython}:\footnote{\fnurl{https://www.python.org/}} a Python interpreter written in C programming language.
-It is the reference interpreter of Python.
-CPython compiles the source code into a bytecode and then interprets it.
-The CPython project still supports both versions of Python 2 and 3;
-    \item \textbf{PyPy}:\footnote{\url{http://pypy.org}} An alternative implementation of the Python interpreter.
-It is written using \emph{RPython} in order to use the JIT.
-It compiles the most used portions of the Python code into a binary code for better performance.
-To benefit from these optimizations, the program has to be executed for at least for few seconds so the JIT has enough time to warm up, the JIT optimization will be applied only the code written by the programmer and not external libraries;
-    \item \textbf{Cython}:\footnote{\url{https://github.com/cython/cython}} a static compiler for Python.
-It translates the Python code into a C code, and then compiles it using a C compiler.
-It also support(s) an extended version of the Python language that allows programmers to call \emph{C functions}, declare \emph{C types} and use static types which will help the translation of Python objects into native types, such as integers, float, which often means a better performances, since native C libraries are almost all the time faster than the python written once~\cite{pereira_energy_2017}.
-    \item \textbf{Intel\,Python}:\footnote{\url{https://software.intel.com/en-us/distribution-for-python}} A customized interpreter developed by Intel in order to enhance performances of Python programs.
-It is dedicated to data sciences, high-Performance computing.
-It uses some Intel kernel libraries, such as Math Kernel Library (Intel MKL\footnote{\url{https://software.intel.com/en-us/mkl}}) and data analytics acceleration library (Intel DAAL\footnote{\url{https://software.intel.com/en-us/intel-daal}}).
-It supports both versions of Python;
-    \item \textbf{Active\,Python}:\footnote{\url{https://www.activestate.com/products/activepython/}} developed by the Activestates company, it provides a standardized Python distribution to ensure license compliance, security, compatibility and performance.
-Therefore, ActivePython has its own built-in packages (more than 300 packages) and supports both versions of Python;
-    \item \textbf{IronPython}:\footnote{\url{https://ironpython.net}} A .Net-based Python interpretation platform written in C\# in order to be used over the .Net vm or mono, it benefits from all the optimizations of .Net virtual machine, such as the JIT and garbage collector mechanisms.
-    \item \textbf{GraalPython}:\footnote{\url{https://github.com/graalvm/graalpython/}} A Python interpreter that is based on GraalVM \footnote{\url{https://www.graalvm.org/docs/why-graal/}} (a universal virtual machine developed by oracle for running applications written in different programming languages).
-For the time being, it only supports Python~3 and it is still in the experimental stage;
-    \item \textbf{Jython}:\footnote{\url{https://jython.github.io}} Implementation of Python programming language written in Java for the \emph{Java virtual machine} (JVM).
-Similar to IronPython and GraalPython, it leverages the optimization mechanisms provided by the JVM to enhance the Python programs;
-    \item \textbf{MicroPython}:\footnote{\url{http://micropython.org}} a lightweight Python version dedicated to embedded systems and micro-controllers;
-    \item \textbf{Nuitka}:\footnote{\url{http://nuitka.net/pages/overview.html}} a Python compiler written in Python that generates a binary executable from Python code.
-It translates the Python code into a C program that is then compiled into a binary executable;
-    \item \textbf{Numba}:\footnote{\url{https://numba.pydata.org}} a library that includes JIT compiler in order to enhance the performances of Python functions using the industry-standard LLVM compiler library;
-    \item \textbf{Shedskin}:\footnote{\url{https://github.com/shedskin/shedskin}} a static transpiler that translates implicitly statically typed python into C++ code.
-    \item \textbf{Hope}~\cite{akeret_hope_2015}: a Python library that aims to introduce JIT compiler into the Python code;
-    \item \textbf{Parakeet}~\cite{DBLP:conf/hotpar/RubinsteynHWS12}: a runtime accelerator for an array-oriented subset of Python;
-    \item \textbf{Stackless\,Python}:\footnote{\url{https://github.com/stackless-dev/stackless/wiki}} an interpreter that focus on enhancing multi-threading programming.
-    \item \textbf{Pyjion}:\footnote{\url{https://github.com/microsoft/pyjion}} JIT API for CPython, same purpose as Parakeet and Hope.
-    \item \textbf{Pyston}:\footnote{\url{https://blog.pyston.org}} performance-oriented Python implementation built using LLVM and modern JIT techniques. the project is funded by dropbox;
-    \item \textbf{Grumpy}:\footnote{\url{https://github.com/google/grumpy}} a source-to-source transpiler that translates the Python code into a Go code that will be compiled to a binary executable.
-It also offers an interpreter called \emph{grumprun} which can directly execute the Python code.
-Unfortunately, we cannot use it because the project is already outdated (last commit is in 2017) and it has a lot of limitation in term of supporting the Python language, such as some built-in functions and standard libraries;
-    \item \textbf{Psyco}:\footnote{\url{http://psyco.sourceforge.net}} a JIT compiler for Python;
-    \item \textbf{Unladen\,Swallow}:\footnote{\url{https://unladen-swallow.readthedocs.io/en/latest/}} an attempt to (use) LLVM  as  JIT compiler for CPython.
-\end{enumerate}
-
-\subsection{Runtime Classification}
-Before futher proceeding with the list of candidate runtimes for Python applications, we propose a classification according to several criterions:
-\begin{description}
-    \item[\bf Type] refers to the category of runtime infrastructure that supports the execution of a Python application.
-In particular, we consider 3 types of environments: \emph{Interpreter}, \emph{Compiler} and \emph{Library}.
-\emph{Interpreter} refers to the class of environment that does not require any preprocessing of Python source code.
-\emph{Compiler} introduces a compilation phase prior to the execution of the application.
-Finally, \emph{Library} induces some modification of the source code.
-    \item[\bf Runtime] refers to the technology supporting the execution of a Python application.
-This technology can refer to the programming language used to program the interpreter, the target language for a compiler or a library.
-    \item[\bf JIT optimisation] refers to the support of \emph{just-in-time} compilation in the runtime infrastructure supporting the execution of the application.
-    \item[\bf GC optimisation] refers to the support of \emph{garbage collection} in the runtime infrastructure supporting the execution of the application.
-    \item[\bf Python version(s)] refers to the list of Python source code versions supported by the runtime environment.
-\end{description}
-
-We should explain the classification of these runtimes
-
-\begin{table}
-    \caption{Classification of Python implementations}
-    \label{fig:python-classes}
-    \center
-    \begin{tabular}{|l|c|c|c|c|c|c|}
-        \hline
-        \multirow{2}{*}{\bf Name} & \multirow{2}{*}{\bf Type} & \multirow{2}{*}{\bf Runtime} & \multicolumn{2}{c|}{\bf Optimisations} & \multicolumn{2}{c|}{\bf Python}                     \\
-        % \cline{1-5} 
-                                  &                           &                              & {\bf JIT}                              & {\bf GC}                        & {\bf 2} & {\bf 3} \\
-        \hline
-        \hline
-        CPython                   & Interpreter               & C                            & \no                                    & \no                             & \yes    & \yes    \\
-        \hline
-        Intel\,Python             & Interpreter               & C                            & \no                                    & \no                             & \yes    & \yes    \\
-        \hline
-        ActivePython              & Interpreter               & C                            & \no                                    & \yes                            & \yes    & \yes    \\
-        \hline
-        PyPy                      & Interpreter               & Python                       & \yes                                   & \yes                            & \yes    & \yes    \\
-        \hline
-        IronPython                & Interpreter               & .Net                         & \yes                                   & \yes                            & \yes    & \yes    \\
-        \hline
-        GraalPython               & Interpreter               & GraalVM                      & \yes                                   & \yes                            & \no     & \yes    \\
-        \hline
-        Jython                    & Interpreter               & Java                         & \yes                                   & \yes                            & \yes    & \no     \\
-        \hline
-        Stackless\,Python         & Interpreter               & python                       & \no                                    & \no                             & \yes    & \no     \\
-        \hline
-        MicroPython               & interpreter               & c                            & \no                                    & \no                             & \no     & \yes    \\
-        \hline
-        Pyston                    & interpreter               & LLVM                         & \yes                                   & \no                             & \yes    & \no     \\
-        \hline
-        Unladen\,Swallow          & Interpreter               & LLVM                         & \yes                                   & \no                             & \yes    & \no     \\
-        \hline
-        Cython                    & Compiler                  & C                            & \no                                    & \no                             & \yes    & \yes    \\
-        \hline
-        Nuitka                    & Compiler                  & C                            & \no                                    & \no                             & \yes    & \yes    \\
-        \hline
-        Shedskin                  & Compiler                  & C++                          & \no                                    & \no                             & \yes    & \yes    \\
-        \hline
-        Grumpy                    & Compiler                  & Go                           & \no                                    & \no                             & \yes    & \yes    \\
-        \hline
-        Numba                     & Library                   & C                            & \yes                                   & \no                             & \yes    & \yes    \\
-        \hline
-        Hope                      & Library                   & python                       & \yes                                   & \no                             & \yes    & \yes    \\
-        \hline
-        Psyco                     & Library                   & python                       & \yes                                   & \no                             & \yes    & \yes    \\
-        \hline
-        Pyjion                    & Library                   & .NET Core                    & \yes                                   & \no                             & \yes    & \yes    \\
-        \hline
-        Parakeet                  & library                   & C                            & -                                      & \no                             & \yes    & \no     \\
-        \hline
-    \end{tabular}
-\end{table}
-
-\begin{table}[hbt]
-    \caption{Classification of Python implementations}
-    \label{fig:pythonimplementations}
-    \center
-    \begin{tabular}{|l|c|c|c|}
-        \hline
-        Version  & Interpreter     & Transpiler/Compiler & Jit library \\
-        \hline
-        \hline
-        Python~2 & Cpython2        & Cython2             & Numba 2     \\
-        % \cline{2-4} 
-                 & Pypy2           & Shesdskin           & Hope        \\
-        % \cline{2-4} 
-                 & Pytson          & Grumpy              & Parakeet    \\
-                 & Ironpython      &                     & Psyco       \\
-                 & Jython          &                     & Pyjion      \\
-        % &  &  &  \\
-                 & Micropython     &                     &             \\
-                 & Pysec           &                     &             \\
-                 & StacklessPython &                     &             \\
-        % \cline{2-4} 
-        \hline
-        Python~3 & Cpython3        & Nuitka              & Numba3      \\
-                 & Pypy3           &                     &             \\
-                 & GraalPython     &                     &             \\
-        \hline
-    \end{tabular}
-\end{table}
-
-There are other implementations that we didn't considerate because either the project aborted many years ago or it has a very limited support for python features.
-After the collection of those implementation we filtered them. in order to keep only the versions that are still maintained and support most of python features.
-and we classified them into 3 categories depending of their integration with the python code.
-In the \ref{fig:pythonimplementations} we describe the implementations that we kept and the version of each implementation and its category.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/python-implementations-tree}
-    \caption{Python interpeters}
-    \label{fig:interpreters}
-\end{figure}
-
-\section{experimetal protocole}
-
-As we have discussed in the previous chapter \ref{chapter:benchmarking}.
-instead of running the tests, the idea was to design a system that allows practiouners to reproduce and extends our tests.
-and then we use the same system to run answer some of researchs question.
-%%%% probably ill resue the same things from other chapters 
-\subsection{measurement context}
-\paragraph{Hardware settings}
-
-all our tests have been executed in a Dell PowerEdge C6420 server machine.
-A summary of its hardware is listed in table \ref{fig:dahuconfig}.
-The machine is equiped with a minimal version of Debian\,9 (4.9.0 kernel version) where we install Docker (version 18.09.5).
-
-\begin{table}[hbt]
-    \begin{tabular}{ll}
-        \hline
-        CPU     & Intel Xeon Gold 6130 (Skylake, 2.10GHz, 2 CPUs/node, 16 cores/CPU)                                                     \\
-        Memory  & 192 GiB                                                                                                                \\
-        Storage & 240 GB SSD SATA Samsung MZ7KM240HMHQ0D3                                                                                \\
-                & 480 GB SSD SATA Samsung MZ7KM480HMHQ0D3                                                                                \\
-                & 4.0 TB HDD SATA Seagate                                                                                                \\
-        Network & eth0/enp24s0f0, Ethernet, configured rate: 10 Gbps, model: Intel Ethernet Controller X710 for 10GbE SFP+, driver: i40e \\
-                & ib0, Omni-Path, configured rate: 100 Gbps, model: Intel Omni-Path HFI Silicon 100 Series [discrete], driver: hfi1      \\
-        \hline
-    \end{tabular}
-    \caption{Testing Machine Configuration }
-    \label{fig:dahuconfig}
-\end{table}
-
-\begin{figure}[hbt]
-    \includegraphics[width=\linewidth]{imgs/SmartWatts.png}
-    \caption{powerapi architecture}
-    \label{fig:powerapi}
-\end{figure}
-
-\paragraph{software settings}
-For the sake of reproducibility, each experiment runs within a Docker container.
-for each test we create a docker image.
-\subsection{Metrics}
-Our focus will be mainly on CPU energy consumption because it is ten folds more than the DRAM one, since it is finit job benchmarking, time is highly correlated within the energy, and it will be only useful to explain certain energytical behaviour so we wont put a lot of focus on this metric.
-
-%% basically i what i want to say is that we we care mostly about energy, because there are studies that were made on the performance, plus the DRAM is not that interested even within the heavy memory  benchmakrs
-% TODO : ADD DRAM graphs for vectors and matrix multiply
-
-\subsection{tests preparation}
-
-To study the behaviour of the python implementation regarding the energy consumption, we have to focus on the effect of the implementation and mitigate the maximum any side effects such as the organisation of the code or any extra consumption due to the operating system or tier libraries.
-Therefore, for each test we took the implementation written in python2 as a reference and tried to use it in other implementations as it is.
-If it is not supported by python3, we transformed the code using the officiel library  \emph{2to3}\footnote{\url{https://docs.python.org/3.7/library/2to3.html}}.
-In the case of the libraries that use \emph{JIT} adding a decorator to the function that we want to optimize was enough, if there are other changes we assume that they alter the original code which is against our purpose.
-
-Each test is implemented in a Docker container for the several reasons
-\begin{itemize}
-    \item Isolation: each container has only the test program implemented with a single python runtime to remove any interference between different implementations.
-    \item Deplpoyment: to use the testing machine without extra configurations that may alter the behaviour of the os toward the enrgy consumtion
-    \item Reproductibility: One of the most frequent benchmark crimes~\cite{DBLP:journals/corr/abs-1801-02381} in research is the lack of Reproductibility, by using Docker we ensure that each test has an Image that will be accessible in public.
-\end{itemize}
-
-Despite the presence of the official docker images for the most of the runtimes, we prefered to build our own using the same reference image in order to remove any bias due to the Os used in the offical image.We used ArchLinux with the kernel version 4.9.184 as a base image.
-
-\subsection{Extension}
-As we have done with the previous chapters. we provide a tool that allows to extend the tests with new workloads and new candidates.
-In the repository \footnote{\url{https://github.com/chakib-belgaid/python-implementations}}.
-we have a tool that allows to generate new workloads and new candidates.
-The script \texttt{generator.py} allows to create new benchmarks by implementing a python code within different interpeters.
-then it generates \texttt{launcher-benchmark.sh} that can be used directly to run the expermient.
-Furthermore all the successful implementations are stored in seperate directory and the that couldn't work ( mostly because of compatibility issues) stored in a recap file called \texttt{benchmarkTest.md}.
-where benchmark is the name of the new workload;
-To add extra \textbf{candidates} one should add a base docker file
-that containes the new implementation and if there should be extra manipulation that should be added to the workload files ,such as adding new decorator or changing some parameters, then they should be added as an extra function in the script \texttt{generator.py}.
-Finally they should be included in the python candidates.
-
-\section{Results and finding}
 
@@
 
-\subsubsection{web developement}
+\section{web developement}
 first we wanted to test the energy consumption of python using on djagon,one of the most popular web frameworks.
@@
 we discuss a deeper this behaviour of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
-
-\subsubsection{python and multiprocessing}
-Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
-As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
-
-\begin{figure}[hbt]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-    \caption{energy behaviour based on multiprocessing}
-    \label{fig:python_multiprocessing}
-\end{figure}
-
-as we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
-where different sub process have to compete for the CPU Ressources.
-Another finding is when we hit the number of physical cores.
-there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
-basicaly he favorits the hyper threads than the physical core wich will lead to some context switches which cause the slow behaviour, howerver in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the eneryg consumption in this case.
-in the Chapter %% reference the work of intern from lyon  
-we discuss a deeper this behaviour of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
-
-%% todo 
-% add performance 
-% add dram \ a glimps
 
@@
 
+\subsection{python and multiprocessing}
+Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
+As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
+    \caption{energy behaviour based on multiprocessing}
+    \label{fig:python_multiprocessing}
+\end{figure}
+
+as we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
+where different sub process have to compete for the CPU Ressources.
+Another finding is when we hit the number of physical cores.
+there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
+basicaly he favorits the hyper threads than the physical core wich will lead to some context switches which cause the slow behaviour, howerver in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the eneryg consumption in this case.
+in the Chapter %% reference the work of intern from lyon  
+we discuss a deeper this behaviour of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
+
+%% todo 
+% add performance 
+% add dram \ a glimps
+
 \subsection{Python and machine learning }
@@
 
-\subsection{python runtimes}
-
-\section{placeholder}
-we performed a shapiro normality test for the first on the results, and almost all the p-values smaller than alpha=0.01\%
-therefore the distrubution is not normal - which is kinda true because if it was normal then we dont have any comparaison --
-
-for arrays and vectors  graalpython took so much time that we decided to remove it
-
-the startup cost for ironpython was too high
-for graalpython and ipy they couldnt handle big numbers hence the overflow  when it comes to execution time
-but since we measure the energy outside the values of the energy aren't impacted
-
-as usual there is  correlaction between DRAM and CPU so no need to classify based on DRAM + the DRAM consumes less 10\% of the  energy compared to the CPU
-
-so the plan is to prove that there is a statistical difference
-then we use the multiparameter optimization , howerver we stop in the phase where we calculate the score
-because we dont know the weight of each parameter ( of tommti microbenchmark) and instead of doing the sum with the coefissients we use the radar plot to let the reader decide which one is the most suitable for his usage.
-howerver since the differences are gigantics we change the scale to logarithmic to be easier for the eye to read
-
-\begin{figure}
-    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
-    \label{table:bitops}
-    \begin{tabular}{lrrrrr}
-        \toprule
-        benchmark    & A          & GG         & LL         & Or         & XOR        \\
-        \midrule
-        activepython & 676.980672 & 763.208040 & 651.783048 & 743.016425 & 728.828481 \\
-        cpython2     & 441.082298 & 435.886950 & 430.846171 & 415.247383 & 419.081447 \\
-        cpython3     & 595.209019 & 685.085315 & 563.839300 & 657.972734 & 655.560574 \\
-        cython       & 35.077408  & 182.688395 & 274.177830 & 34.868014  & 34.504778  \\
-        nuitka       & 33.260991  & 32.980380  & 33.256450  & 33.472770  & 33.030889  \\
-        numba2       & 9.102939   & 8.411176   & 9.460620   & 9.375920   & 9.755952   \\
-        numba3       & 9.566646   & 10.144212  & 9.219703   & 9.344613   & 9.665108   \\
-        pypy2        & 8.456867   & 7.844918   & 8.286278   & 8.138692   & 7.952999   \\
-        pypy3        & 7.552731   & 8.093884   & 8.108508   & 8.669448   & 8.623737   \\
-        shedskin     & 8.024198   & 8.070870   & 8.399917   & 8.126236   & 8.277546   \\
-        \bottomrule
-    \end{tabular}
-
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/barplot_binarry_tree}
-    \caption{energy behaviour based on multiprocessing}
-    \label{fig:python_multiprocessing}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/bitopts_mean}
-    \caption{Mean consumption of different implementations of bit operations (Joule) }
-    \label{fig:bitops}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/alltomti_performance}
-    \caption{different interpererts optimisation }
-    \label{fig:tommi_all}
-\end{figure}
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/tommti_compare__pypy2_pypy3}
-    \caption{green factor of pypy }
-    \label{fig:pypy2vspypy3}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/tommti_compare__cpython3_numba3_pypy3}
-    \caption{comparaison of pypy vs python vs numba }
-    \label{fig:p3}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/tommti_compare__cpython2_pypy2_numba2}
-    \caption{comparaison of pypy vs python vs numba }
-    \label{fig:p2}
-\end{figure}
-
-As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
-Intelpython and active python both folows the same behaviour. one can conclude that the work that have been on those interpreters is mainly
-to improve a specific purpose, Active python claims that their version is focused on secuirty which explains the lack of some performances due to the introduction of more reonforcment
-.Morver Intel published their version of python as a dedicated for machine learning. Unfortunately the tommti benchmark is a set that focuses on the general purpose programming
-which does not reflect the performance of the machine learning.
-Another aspect of a such behaviour might be due to the fact of the processors that were used in the testbed.
-and it might change in the future if we include the GPU part for the machine learning benchmarks
-% reference the work of the intern 
-As for nuitka. there were no optimization in the energy consumpotion despite the fact that is it a complier.
-However if we dig through the nuitka mechanisms. they basically embed the python code with an interpreter.
-Unllike nuitka shedskin exhebit a the best energy consumption pattern when it comes to the arithmetic operartions.
-One can conclude it is due to the fact of the native type of the variables.
-unlike the interpreters where they are treated as object in the begining.
-
-for the other interpeters pypy is very promising especially when it comes to data maniupulation as one can see in the figure \ref{fig:bar_tommti_vector}
-pypy is by far the best interperer when it comes to treating vectors.
-numba2 introduced the JIT but wasn't as promising as numba3.
-
-for the other vm based interpereters.
-jython and ipy lacked in term of energy optimisation which was kinda expected since they were in their the begining of the stage and the main puprose of such implementation is to link the bytecode generated by jython and ironpython with their respectives virtual machin.
-
-Unlike the the previous interpeters. graal exhebits a certin promises when it comes to complex algorithms - nested loops -
-micro python is dedicated to embeded systems so lunching it is powerful cpu machines will be misleading.
-
-Most of the interperters had the same behaviour when it comes to the input outputs. except for jython which was kinda of a anomaly probably due to the lack of the optimizations.
-% pypy handles exceptions very well
-
-\section{Threads to validity}
-\section{tools and contributions}
-
-\section{conclusion}
-One may observe that the choice of Python interpreter has a significant impact on the programs' energy consumption.
-This investigation is made more intriguing by the absence of a universal solution.
-The primary downside is the incompatibility of some of these solutions, which causes us to make concessions when we need a generic answer.
+\input{python_interperters.tex}

### 2022-07-14 chapters/optimization/python-dir/python_interperters.tex
refactoring :python

--- 2022-05-28 chapters/execution_environement/python/python.tex
@@
-\newpage
-\section{python}
+\section{Python interpreters}\label{sec:pythoninterpreters}
 
-\subsection{Introduction}
+%%% note : add this to the section of python interpreters 
+Due to the lack of of support for most of  non conventional pyton interpeters, our main focus was only on micro benchmarks.
+Except pypy most of the python implementations don't support extra python libraries, Despite that most of the cases those extra implementations were made to optimize a specific library such as numba with numpy, or intelpython with machine learning algorthms.
 
-When it comes to developing software systems, one can notice that---except Perl---dynamic programming languages have taken over compiled languages along the last decade in terms of developers popularity (cf.
-Figure~\ref{fig:pypl}).
-However, it remains unclear if this class of dynamic programming languages can reasonably compete with compiled ones when it comes to power efficiency.
+\paragraph{preliminary studies}
+For the first studies we used only the official version of python. because the goal was mainly to highlight the impact of the structure of code on the energy consumption.
+One main inconvienent of the previous methode to reduce the energy consumption, is the hard work that should be done in order to alter the existing code base in order to redcue the energy consumption.
+To avoid such hustle we tried to find a non intrusive way to make the python code more ecofriendly without changing its structure.
+One aspect of python is the fact that it is an interptered language which led many works to create their own implementation of the interpeter to enhance one or many aspect of the python code.
+in the following section we will discuss the impact of those implementations on the energy consumption of python programs, and in which case, one should use a non conventional interpter to save the energy consumption of their program.
 
-\begin{figure}[htbp]
-    \includegraphics[width=\linewidth]{imgs/programminglanguangespopularity.png}
-    \caption{PYPL Popularity of Programming Languages~\cite{noauthor_pypl_nodate}}
-    \label{fig:pypl}
-\end{figure}
+To do so, we gathered a list of interpeters, transpilers and some optimization libraries.
 
-In particular, Noureddine~\emph{et~al.}~\cite{noureddine_preliminary_2012} in 2012, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
-According to their experiments, an interpreted programming language like Python can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. Figure~\ref{fig:hannoi}).
-% However, the operational conditions under which these measurements may threaten the validity of their results.
-
-\begin{figure}[htbp]
-    \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
-    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
-    \label{fig:hannoi}
-\end{figure}
-\subsection{python popularity}
-
-However, Python seems to attract a large community of developers interested in data analysis, web development, system administration, machine learning---according to a survey conducted in 2018 by JetBrains---\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}}one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.
-
-As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
-Rather, we believe that a better option consists in leveraging the strength of this rich ecosystem to promote energy-efficient solutions in order to improve the power consumption of legacy software systems.
-
-In this chapter, we therefore explore the oblivious optimizations that can be applied to Python legacy applications in order to reduce their energy footprint.
-As Python is widely adopted by software services deployed in public and private Cloud infrastructures, we believe that our contributions will benefit to wide diversity of legacy systems and not only favorably contribute to reduce the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
-More specifically, this chapter focuses on runtime optimizations that can adopted by developers to leverage the power consumption of Python applications.
-These optimizations include not only alternative interpreters, but also \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
-
-% The contributions of this chapter can therefore be summarized as:
-% \begin{compactenum}
-%     \item a classification of state-of-the-art Python optimizations,
-%     \item an evaluation of the power efficiency of these optimizations on well-known micro-benchmarks,
-%     \item a comparison of the effective energy efficiency of Python compared to compiled languages,
-%     \item an assessment of the energy speedup of these optimizations on representative applications and workloads.
-% \end{compactenum}
-
-%%%% python is a popular programming language  based TIOBE PYPI   add graphs
-
-he simplicity, flexibility and the clarity of Python code led this programming language to a huge success.
-Therefore, a larger community and a wide range of support and libraries that allowed it to touch several fields like data analyses, machine learning, natural language processing, web development, astrophysics etc.
-% python is not efficient 
-% 1 Energy Efficiency Across Programming Languages: How Do Energy, Time, and Memory Relate?
-%  \begin{figure}[hbt]
-%     \includegraphics[width=\linewidth]{figures/hannoiimplementation.png}
-%     \caption{Energy consumption of the recursive implementation of Tower of Hanoi program in different languages (using a base 10 logarithmic scale) \cite{noureddine_preliminary_2012}}      
-%     \label{fig:hannoi}
-% \end{figure}
-
-However, and according to \cite{pinto_energy_2017,noureddine_preliminary_2012} and [1].
-Python tends to be one more energy hungry .
-In figure \ref{fig:hannoi}, python consumes 30 times more than C or C++.
-The test was done on an implementation of \fnurl{hannoi test}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi}  30 disks.
-
-Python consume a lot of energy mainly because it is slow in execution.
-Its Flexibility and simplicity caused it to drop off in performances, because Python gains its flexibility from being a dynamic language, therefore it needs an interpreter to execute its programs, which makes them much slower compared to the other ones that are written in compiled programming languages such as C and C++ or semi-compiled languages like Java.
-
-As shown in Table \ref{fig:clbg}, we can see that in the most of the test cases Python takes more time to execute -the only case that he wasn't the last one was in the benchmark regx-redux where he beaten Go-, and in some cases the gab was huge such as in the n-body test where python took around 100 times more than C++.
-
-\begin{table}[hbt]
-    \begin{tabular}{l|*{5}c}
-                           & C    & C++   & Java & python & Go    \\
-        \hline
-        pidigits           & 1.75 & 1.89  & 3.13 & 3.51   & 2.04  \\
-        reverse-complement & 1.75 & 2.95  & 3.31 & 16.76  & 4.00  \\
-        regx-redux         & 1.45 & 1.66  & 10.5 & 15.56  & 28.69 \\
-        k-nucleotide       & 5.07 & 3.66  & 8.66 & 79.79  & 15.36 \\
-        binary-trees       & 2.55 & 2.63  & 8.28 & 92.72  & 28.90 \\
-        fasta              & 1.32 & 1.33  & 2.32 & 62.88  & 2.07  \\
-        Fannkuch-redux     & 8.72 & 10.62 & 17.9 & 547.23 & 17.82 \\
-        n-body             & 9.17 & 8.24  & 22.0 & 882.00 & 21.00 \\
-        spectral-norm      & 1.99 & 1.98  & 4.27 & 193.86 & 3.95  \\
-        Mandelbort         & 1.64 & 1.51  & 6.96 & 279.68 & 5.47
-    \end{tabular}
-    \caption{Comparison of the execution time between different programming languages using the computer language benchmarks game \cite{noauthor_which_nodate}}
-    \label{fig:clbg}
-\end{table}
-
-\subsection{python insights}
-The goal of this section is to find some insights where we can optimize the energy consumption without impacting the source code.
-so we started by extending the work of \citeauthor{hasan_energy_2016} and \citeauthor{oliveira_recommending_nodate} to the python environement.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/pythonloops}
-    \caption{energy behaviour of different python loops }
-    \label{fig:pythonloops}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/pythonarrays}
-    \caption{energy behaviour of different python data structures}
-    \label{fig:pythonarrays}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-    \caption{energy behaviour based on multiprocessing}
-    \label{fig:python_multiprocessing}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/django}
-    \caption{energy behaviour based on multiprocessing}
-    \label{fig:python_multiprocessing}
-\end{figure}
-
-Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
-As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
-
-asd we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
-where different sub process have to compete for the CPU Ressources.
-Another finding is when we hit the number of physical cores.
-there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
-basicaly he favorits the hyper threads than the physical core wich will lead to some context switches which cause the slow behaviour, howerver in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the eneryg consumption in this case.
-in the Chapter %% reference the work of intern from lyon  
-we discuss a deeper this behaviour of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
-
-\subsection{Python Runtimes}
-
-we used the following implementations.
-
-We should explain the methodology we used to select these
+the following list, presents the most famous ones .
+% We should explain the methodology we used to select these
 
 \begin{enumerate}
-    \item \textbf{CPython}:\footnote{\fnurl{https://www.python.org/}} a Python interpreter written in C programming language.
+    \item \textbf{CPython}:\footnote{\url{https://www.python.org/}} a Python interpreter written in C programming language.
 It is the reference interpreter of Python.
@@
     \label{fig:powerapi}
+\end{figure}
 
-\end{figure}
-\subsection{Power Measurements}
+\paragraph{software settings}
+For the sake of reproducibility, each experiment runs within a Docker container.
+for each test we create a docker image.
+\subsection{Metrics}
+Our focus will be mainly on CPU energy consumption because it is ten folds more than the DRAM one, since it is finit job benchmarking, time is highly correlated within the energy, and it will be only useful to explain certain energytical behaviour so we wont put a lot of focus on this metric.
+
+%% basically i what i want to say is that we we care mostly about energy, because there are studies that were made on the performance, plus the DRAM is not that interested even within the heavy memory  benchmakrs
+% TODO : ADD DRAM graphs for vectors and matrix multiply
+
+\paragraph{Energy measurement}
 
@@
 The reason of separation between data collection and energy calculation is to minimise any interference with the test so our sensor is a light c program running inside a docker container.
+\subsection{tests preparation}
+
+To study the behaviour of the python implementation regarding the energy consumption, we have to focus on the effect of the implementation and mitigate the maximum any side effects such as the organisation of the code or any extra consumption due to the operating system or tier libraries.
+Therefore, for each test we took the implementation written in python2 as a reference and tried to use it in other implementations as it is.
+If it is not supported by python3, we transformed the code using the officiel library  \emph{2to3}\footnote{\url{https://docs.python.org/3.7/library/2to3.html}}.
+In the case of the libraries that use \emph{JIT} adding a decorator to the function that we want to optimize was enough, if there are other changes we assume that they alter the original code which is against our purpose.
+
+Each test is implemented in a Docker container for the several reasons
+\begin{itemize}
+    \item Isolation: each container has only the test program implemented with a single python runtime to remove any interference between different implementations.
+    \item Deplpoyment: to use the testing machine without extra configurations that may alter the behaviour of the os toward the enrgy consumtion
+    \item Reproductibility: One of the most frequent benchmark crimes~\cite{DBLP:journals/corr/abs-1801-02381} in research is the lack of Reproductibility, by using Docker we ensure that each test has an Image that will be accessible in public.
+\end{itemize}
+
+Despite the presence of the official docker images for the most of the runtimes, we prefered to build our own using the same reference image in order to remove any bias due to the Os used in the offical image.We used ArchLinux with the kernel version 4.9.184 as a base image.
+
+\subsection{Extension}
+As we have done with the previous chapters. we provide a tool that allows to extend the tests with new workloads and new candidates.
+In the repository \footnote{\url{https://github.com/chakib-belgaid/python-implementations}}.
+we have a tool that allows to generate new workloads and new candidates.
+The script \texttt{generator.py} allows to create new benchmarks by implementing a python code within different interpeters.
+then it generates \texttt{launcher-benchmark.sh} that can be used directly to run the expermient.
+Furthermore all the successful implementations are stored in seperate directory and the that couldn't work ( mostly because of compatibility issues) stored in a recap file called \texttt{benchmarkTest.md}.
+where benchmark is the name of the new workload;
+To add extra \textbf{candidates} one should add a base docker file
+that containes the new implementation and if there should be extra manipulation that should be added to the workload files ,such as adding new decorator or changing some parameters, then they should be added as an extra function in the script \texttt{generator.py}.
+Finally they should be included in the python candidates.
+
+\section{Results and finding}
+
+\section{placeholder}
+we performed a shapiro normality test for the first on the results, and almost all the p-values smaller than alpha=0.01\%
+therefore the distrubution is not normal - which is kinda true because if it was normal then we dont have any comparaison --
+
+for arrays and vectors  graalpython took so much time that we decided to remove it
+
+the startup cost for ironpython was too high
+for graalpython and ipy they couldnt handle big numbers hence the overflow  when it comes to execution time
+but since we measure the energy outside the values of the energy aren't impacted
+
+as usual there is  correlaction between DRAM and CPU so no need to classify based on DRAM + the DRAM consumes less 10\% of the  energy compared to the CPU
+
+so the plan is to prove that there is a statistical difference
+then we use the multiparameter optimization , howerver we stop in the phase where we calculate the score
+because we dont know the weight of each parameter ( of tommti microbenchmark) and instead of doing the sum with the coefissients we use the radar plot to let the reader decide which one is the most suitable for his usage.
+howerver since the differences are gigantics we change the scale to logarithmic to be easier for the eye to read
 
@@
 
-\subsubsection{micro-benchmarks}
-
-\subsubsection{benchmarks}
-
-\subsection{results and foundings}
 As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
@@
 
-%% todo 
-% add performance 
-% add dram \ a glimps 
-% add bitops
-
-\subsection{tools and package}
-
-\subsection{Threads to validity}
-
-\subsection{conclusion}
+\section{conclusion}
+One may observe that the choice of Python interpreter has a significant impact on the programs' energy consumption.
+This investigation is made more intriguing by the absence of a universal solution.
+The primary downside is the incompatibility of some of these solutions, which causes us to make concessions when we need a generic answer.

### 2022-07-17 chapters/optimization/python-dir/python.tex
refactoring : fix a link
add : experemental setup for the ml experiment

@@
 As onne can notice in figure \ref{fig:hannoi}, python consumes 30 times more than C or C++.
-The test was done on an implementation of \url{hannoi test}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi}  30 disks.
+The test was done on an implementation of \fnurl{hannoi test}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi}  30 disks.
 
@@
 Following that, we will measure the energy consumption of several Python implementations in order to determine a non-intrusive technique to improve energy efficiency.
-
 Therefore this chapter will target the following research questions :
@@
 \end{compactenum}
+
+For this chapter we are considering 4 separate yet related studies .
+First, we are going to see the energy behavior of the python within two study cases.
+Web servers and machine learning.
+We will then delve deeper into the energy consumption of Python basic structures before concluding with the impact of Python interpterer on energy consumption.
 
@@
 \section{web developement}
-first we wanted to test the energy consumption of python using on djagon,one of the most popular web frameworks.
-therefore we create a simple website where we try to analyse the cost of a single request using different mechanisms to fetch the data from the database.
-the idea behind this use case is to see wether the choice of the database and the orm ( object relational mapper) impacts the energy consumption and the perfromances of the website.
+The purpose of this study is to understand the impact of the \emph{ORMs} (Object Relational Mapping)~\cite{o2008object}. We have chosen django as our web framework
 
-we considered using two different databases postgresSQL and SQlite3 that containes the same data, and three different ways to fetch the data.
+As a result, we create a simple website in which we attempt to analyze the cost of a single request using various mechanisms to retrieve data from the database.
+
+The idea behind this use case is to see whether the choice of the database and the orm (object relational mapper) impacts the energy consumption and the performance of the website.
+
+We considered using two different databases, PostgreSQL and SQlite3, that contained the same data, and three different ways to fetch the data.
 \begin{enumerate}
@@
 \subsection{Python and machine learning }
+Machine learning is becoming an integral part of our daily lives, growing more potent and energy-hungry each year.
+
+As machine learning can have a significant impact on climate change, it is vital to investigate mitigation techniques.
+
+\paragraph{hardware }
+Chifflot 8 from Grid 5000's Lille site was used for all of the trials.
+The machine is outfitted with two Intel Xeon Gold 6126 CPUs, each having 12 physical cores, 192 GB of RAM, and two 32 GB Tesla V100 GPUs.
+
+\paragraph{software}
+
+For the sake of reproducibility, each experiment is done within a Docker container using Jupyter lab .
+These tests are run on top of a minimal version of Debian-10 in order to increase the accuracy of the tests by elimanting any unnecessary processes.
+
+\subsubsection{Workload}
+\paragraph{Models}
+
+Several models were developed, however only two were used in the final trials because they achieved 94 percent accuracy in an acceptable length of time.
+David Page's cifar10-fast and Woonhyuk Baek's torchskeleton are shown here.
+\paragraph{Datasets}
+
+The CIFAR-10 dataset was the major source of data for the studies.
+It is made up of 60000 32x32 color images grouped into ten categories.
+
+Some experiments were done using the MNIST dataset of handwritten digits to validate the results acquired from the first dataset.
+The model did not need to be updated because the 60000 28x28 grayscale photos were padded.
+
+\subsubsection{candidates}
+The experiments were run with several different CPU and GPU configurations:
+
+\begin{itemize}
+    \item With and without GPU
+    \item With and without hyperthreading
+    \item Different number of physical cores
+\end{itemize}
+
+\subsubsection{metrics}
+
+We used Pytorch 1.10.0  to train those models and pyjoules to measure the energy consumption of the GPU and CPU.
+\begin{itemize}
+    \item accuracy : in \%
+    \item execution time : in seconds for both the duration of each epoch and the total duration to a chieve a certain accuracy
+    \item total energy consumption : for the CPU and the  GPU
+    \item
+\end{itemize}
+
+\section{finding and results}
 \begin{figure}

### 2022-07-19 chapters/benchmarking/draft.tex
Proof reading

@@
 As a result, it is not only necessary to ensure the "reproducibility" of the results, but it is also essential to provide room for new candidates.
-As a result, we'd like to present "Extension," a new approach for leveraging experimentation.
+As a result, we'd like to present "Extension", a new approach for leveraging experimentation.
 The ability to give not only the tools needed to replicate the tests but also to integrate more candidates, workloads, or important key performance indicators
@@
 First, we go through the three components of a successful benchmark within the energy consumption experiment.
-Section~\ref{sec:benchmarking_reproducibility}  therefore concentrates on the "reproducibility" issue and how to deliver reproducible experiments without interfering with the energy measurement.
+Section~\ref{sec:benchmarking_reproducibility} concentrates on the "reproducibility" issue and how to deliver reproducible experiments without interfering with the energy measurement.
 Section~\ref{sec:taming-the-energy-variation} discusses accuracy in software energy consumption.
@@
 \subsection{Docker \& Energy}
-Now that we've decided to use container technology to enclose our tests, what effect will this have on the amount of energy our tests use?
+Now that we have decided to use container technology to enclose our tests, what effect will this have on the amount of energy our tests use?
 %%%%%%%%%
@@
 Using research from \cite{santos2018does} who examined how adding the Docker layer affected energy consumption.
-\citeauthor{santos2018does}~\emph{et~ al} conducted their experiment by running numerous benchmarks both with and without Ddocker.
+\citeauthor{santos2018does} conducted their experiment by running numerous benchmarks both with and without Ddocker.
 They contrasted the energy usage and execution time that resulted.

### 2022-07-21 chapters/optimization/python-dir/python.tex
add comments

@@
 \section{finding and results}
+As the model's accuracy increases, so does the energy required for the next accuracy increment.
+
+Figure \ref{fig:cum_energy_vs_accuracy} depicts how the curve steepens as training progresses.
+For example, training to 90\% accuracy requires three times the energy required for training to 80% accuracy. 
 \begin{figure}
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/cumulative_energy_vs_accuracy}
+    \caption{cumulative energy consumption vs accuracy
+    }
+    \label{fig:cum_energy_vs_accuracy}
+\end{figure}
 \begin{figure}

### 2022-07-26 chapters/optimization/python-dir/python.tex
add : p-values for python

@@
 In this chapter, we therefore explore the oblivious optimizations that can be applied to Python legacy applications in order to reduce their energy footprint.
-As Python is widely adopted by software services deployed in public and private Cloud infrastructures, we believe that our contributions will benefit to wide diversity of legacy systems and not only favorably contribute to reduce the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
+As Python is widely adopted by software services deployed in public and private Cloud infrastructures, we believe that our contributions will benefit to wide diversity of legacy systems and not only favorably contribute to reduce the carbon emissions of ICT, b kmoiut also reduce their cloud invoice for the resources consumed by these services.
 More specifically, this chapter focuses on runtime optimizations that can adopted by developers to leverage the power consumption of Python applications.
@@
 Figure \ref{fig:cum_energy_vs_accuracy} depicts how the curve steepens as training progresses.
-For example, training to 90\% accuracy requires three times the energy required for training to 80% accuracy. 
+For example, training to 90\% accuracy requires three times the energy required for training to 80\% accuracy.
 \begin{figure}

### 2022-08-02 chapters/benchmarking/draft.tex
fix: Proofreading

--- 2022-07-10 chapters/benchmarking/draft.tex
@@
 As a result, it is not only necessary to ensure the "reproducibility" of the results, but it is also essential to provide room for new candidates.
-As a result, we'd like to present "Extension," a new approach for leveraging experimentation.
+As a result, we would like to present "Extension," a new approach for leveraging experimentation.
 The ability to give not only the tools needed to replicate the tests but also to integrate more candidates, workloads, or important key performance indicators
@@
 As a result, we cannot use VM for performance-related tests.
-Isolation is another drawback of VM.
-While this feature protects the experimental setting from unwanted interference from the outside world, it is possible that this interaction is required—-especially if the experiment is dependent on an external source such as energy montitors .
+Isolation is another drawback of VM: while this feature protects the experimental setting from unwanted interference from the outside world, it is possible that this interaction is required—-especially if the experiment is dependent on an external source such as energy monitors.
 
@@
 Containers provide such an advantage by ensuring application separation and ease of replication.
-Figure~\ref{environement:virtualization_technique} depicts the architectural differences between virtualization and container technologies. There are three main types of virtualisation.
+Figure~\ref{environement:virtualization_technique} depicts the architectural differences between virtualization and container technologies. There are three main types of virtualization.
 \begin{itemize}
@@
 Using research from \cite{santos2018does} who examined how adding the Docker layer affected energy consumption.
-\citeauthor{santos2018does}~\emph{et~ al} conducted their experiment by running numerous benchmarks both with and without Ddocker.
+\citeauthor{santos2018does}~\emph{et~ al} conducted their experiment by running numerous benchmarks both with and without Docker.
 They contrasted the energy usage and execution time that resulted.
-The first step was to observe the effects of the orchestrator and the Docker deamon while there was no work to be done.
-Then, they use three benchmarks in their experiments: Wordpress, Reddis and PostgreSQL.
+The first step was to observe the effects of the orchestrator and the Docker daemon while there was no work to be done.
+Then, they use three benchmarks in their experiments: Wordpress, Reddis, and PostgreSQL.
 The values below show the system under test's energy consumption while it is idle.
@@
 Yet, as seen in Figure~\ref{fig:docker_reddis}, Docker increased the execution duration of the benchmark by 50 seconds,which led to a significant rise in the energy usage.
-According to the authors, this overhead is primarily caused by the Docker deamon and not by the fact that  the application is in a container.
+According to the authors, this overhead is primarily caused by the Docker daemon and not by the fact that  the application is in a container.
 
@@
 To summarize, Docker-based software tends to consume more energy since it takes longer to execute.
-The execution of the Docker deamon causes an increase in average power consumption of only \textbf{2\,Watts}.
+The execution of the Docker daemon causes an increase in average power consumption of only \textbf{2\,Watts}.
 This overhead can reach up to 5\% in IO-intensive applications, while it is barely visible in CPU- or DRAM-intensive workloads.
@@
 Furthermore, because we have access to the host Hardware, we do not need to worry about capturing the SUT's energy use.
-As a result, we will use Docker as an encapsuation solution for all of our experiments.
+As a result, we will use Docker as an encapsulation solution for all of our experiments.
 It will ensure the reproducibility of our experiments in a simple and straightforward manner.

### 2022-08-02 chapters/optimization/java-dir/java.tex
add : extension to RPC
fix: Proofreading Java chapter

--- 2022-06-14 chapters/optimization/java-dir/java.tex
@@
-\section*{JAVA}
+% \section*{JAVA}
 \section{Introduction}
@@
 Furthermore, if we take into consideration legacy applications, Java becomes the most used programming langues.%ROMAIN: Add source
-In addition to its popularity, Java exhibits an interesting behaviour when it comes to energy consumption and performances, basically Java applications can be at the same time one of the most energy efficient or hungry solution.
+In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performances, basically Java applications can be at the same time one of the most energy efficient or hungry solution.
 As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
@@
 \noindent\textbf{Software Settings.}
-For the sake of reproducibility, each experiment runs within a Docker container based on SDKMAN image and Alpline docker.
+For the sake of reproducibility, each experiment runs within a Docker container based on \textsf{SDKMAN!}\footnote{\url{https://sdkman.io}} image and Alpine docker.\footnote{\url{https://github.com/alpinelinux/docker-alpine}}
 
@@
 \noindent\textbf{Java Virtual Machines Candidates.}
-We considered a set of $52$ JVM distributions taken from $8$ different providers/packagers mostly obtained from \textsf{SDKMAN!},\footnote{\url{https://sdkman.io/}} as listed in \Cref{tab:JVM}.
+We considered a set of $52$ JVM distributions taken from $8$ different providers/packagers mostly obtained from \textsf{SDKMAN!}, as listed in \Cref{tab:JVM}.
 Depending on providers, either all the versions, majors, or LTS are made available by \textsf{SDKMAN!}.
@@
 
-\subsection{Metrics and measurement}
-Since the goal of this study is the green aspect Java virtual machines, our key metric will be the energy consumption of the job completed by each configuration.
-In addition to the energy consumption, we collected additional metrics to explain the reasons behind the behaviour of each experiment.
-Those extra metrics are:
+\Cref{fig:jvm-benchmarks-classification} highlights the scope of each benchmark from the test suite.
+
+\begin{figure*}%[ht]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/jvm_benchmarks}
+    \caption{Target scope of \textsc{Dacapo} and \textsc{Renaissance} benchmarks.}
+    \label{fig:jvm-benchmarks-classification}
+\end{figure*}
+
+\subsection{Metrics and Measurement}
+Since the goal of this study is the green aspect of JVM, our key metric will be the energy consumed by job completed for each JVM configuration.
+In addition to the energy consumption, we collected additional metrics to explain the reasons behind the behavior of each experiment.
+Those additional metrics are:
 \begin{itemize}
@@
 To collect the number of active threads used by the an experiment, we use the command \texttt{top} and record at fixed intervals.
+
+\subsection{Extension}
+We also added an extension to the protocol to allow the user to run the same experiment with different configurations.
+The package is available in the GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jvm-comparaison}}
+To add extra \textbf{jvm candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the jvm to be used.\note{The jvm must be provided by \textsf{SDKMAN!}}
+For the \textbf{input workload},the benchmarks should be provided in the \texttt{benchmarks} directory.
+As for extra \textbf{metrics}, one can create a new script file that monitor the experiment and record the metrics.
+We provide some examples, such as \texttt{recordpower.sh} to measure the instantaneous power and \texttt{recordthreads.sh} to measure the number of active threads during the experiment.
+
+For faster experiments, we propose \textsc{Jreferral}\footnote{\url{https://github.com/chakib-belgaid/jreferral}}, an open source tool that automatically compares the JVM energy consumption and recommends the most energy-efficient configuration for a given Java application.
+We further discuss this tool in Section~\ref{section:SummaryofContributions}.
 
@@
 
-Then, Figure~\ref{hotspot8} depicts the evolution of the energy consumption of the $12$ benchmarks, when executed on the \textsc{HotSpot} JVM.
+Then, \Cref{fig:hotspot8} depicts the evolution of the energy consumption of the $12$ benchmarks, when executed on the \textsc{HotSpot} JVM.
 \Cref{fig:hotspot8} reports on the energy consumption variation of individual benchmarks, using to \textsc{HotSpot-8}\ as the baseline.
@@
 It is worth noticing that the energy consumption of some benchmarks, such as \textsf{Reactors}, exhibit large variations across JVM versions due to experimental features and changes that are not always kept when releasing LTS versions (version 11 here).
-For example, the introduction of \texttt{VarHandle} to allow low-level access to the memory order modes available in JDK 9 and work along \texttt{Unsafe Classe} that was removed from from JVM 11.\footnote{\url{https://blogs.oracle.com/javamagazine/the-unsafe-class-unsafe-at-any-speed}}
+For example, the introduction of \texttt{VarHandle} to allow low-level access to the memory order modes available in JDK 9 and work along \texttt{Unsafe Class} that was removed from from JVM 11.\footnote{\url{https://blogs.oracle.com/javamagazine/the-unsafe-class-unsafe-at-any-speed}}
 
@@
 The second interesting situation was observed on the \textsf{Dotty} benchmark.
-More specifically, during the first 100 seconds of the execution of the \textsf{Dotty} benchmark on all evaluated JVMs.
+More specifically, during the first $100$ seconds of the execution of the \textsf{Dotty} benchmark on all evaluated JVMs.
 At the beginning of the execution, \textsc{GraalVM} has a slightly lower power consumption, is faster, and consumes 10\% less energy.
-After about 150 seconds, the power differences between the 3 JVMs is barely noticeable.
+After about $150$ seconds, the power differences between the 3 JVMs is barely noticeable.
 One can, however, notice the effect of the JIT, as \textsc{HotSpot} takes the advantage over \textsc{GraalVM} and becomes more energy efficient.
@@
 
-Another measurement issue is the CPU energy variation within machines~\cite{opaper}, thus we executed all the comparable tests on the same node and with the recommended settings to mitigate this threat.
+Another measurement issue is the CPU energy variation within machines (cf.
+\Cref{chapter:benchmarking}), thus we executed all the comparable tests on the same node and with the recommended settings to mitigate this threat.
 
@@
 
-How generalizable are our results? We believe that our study conclusions and guideline remain empirical, as we do not intend to generalize any result we obtained for some JVM or benchmark.
+How generalizable are our results?
+We believe that our study conclusions and guideline remain empirical, as we do not intend to generalize any result we obtained for some JVM or benchmark.
 We provide practitioners with some prerequisites to check before software deployment to reduce the software energy footprint by considering the JVM and its settings.
@@
 \Cref{table:j-referral} illustrates an example of the final report of \textsf{J-Referral}.
-The tool was tested for 2 real Java projects: \textsf{Zip4J} and\footnote{\url{https://github.com/srikanth-lingala/zip4j}} and \textsf{K-nucleotide}.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/performance/knucleotide.html}}.
+The tool was tested for 2 real Java projects: \textsf{Zip4J}\footnote{\url{https://github.com/srikanth-lingala/zip4j}} and \textsf{K-nucleotide}.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/performance/knucleotide.html}}.
 \textsf{Zip4J} runs a large file compression, while \textsf{K-nucleotide} extracts a DNA sequence, and updates a hashtable of k-nucleotide keys to count specific values.
-The short report presented in \Cref{table:j-referral} shows the ratio of potential energy saving between the most and least energy consuming tested JVM (40\% and 70\% energy savings for \textsf{Zip4J} and \textsf{K-nucleotide} respectively).
+The short report presented in \Cref{table:j-referral} shows the ratio of potential energy saving between the most and least energy consuming tested JVM (40\% and 70\% energy savings for \textsf{Zip4J} and \textsf{K-nucleotide}, respectively).
 Options are available for \textsf{J-Referral} to obtain much more detailed reports including execution time, DRAM usage, split DRAM vs. CPU consumption, etc.
-The tool is available as \emph{open-source software} (OSS) from our anonymous repository.\footnote{\url{https://anonymous.4open.science/r/jreferral/Readme.md}}
+The tool is available as \emph{open-source software} (OSS) from our GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jreferral}}
 

### 2022-08-02 chapters/optimization/java-dir/java.tex
Proofreading

@@
 As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.%ROMAIN: Add source
-Furthermore, if we take into consideration legacy applications, Java becomes the most used programming langues.%ROMAIN: Add source
-In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performances, basically Java applications can be at the same time one of the most energy efficient or hungry solution.
+Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language.%ROMAIN: Add source
+In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performance, Java applications can be at the same time one of the most energy-efficient or hungry solutions.
 As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
 Therefore, we wanted to dig deeper into this aspect of Java and study its runtime.
-This chapter thus focuses on the impact of the runtime of Java applications on the energy consumption.
+This chapter thus focuses on the impact of the runtime of Java applications on energy consumption.
 
@@
 
-To answer those research question we will run an empirical study to highlight the impact of this runtime.
+To answer those research questions, we conduct an empirical study to highlight the impact of this runtime.
 
@@
 \section{Experimental Protocol}\label{sec:javaprotocol}
-To investigate the effect that could have the JVM distribution choice and/or parameters on software energy consumption, we conducted a wide set of experiments on a cluster of machines and using several established Java benchmarks and JVM configurations.
+To investigate the effect that could have the JVM distribution choice and/or parameters on software energy consumption, we conducted a wide set of experiments on a cluster of machines and used several established Java benchmarks and JVM configurations.
 
@@
 % In other words, we used the same node to ran all the JVM from the different providers and with numerous configurations for each single benchmark application.
-Moreover, we tried to be very careful, while running our experiments, not to fall in most common benchmarking "crimes"~\cite{crimes}.
+Moreover, we tried to be very careful, while running our experiments, not to fall in the most common benchmarking "crimes"~\cite{crimes}.
 Every single experiment, therefore, reports on energy metrics obtained from at least $20$ executions of $50$ iterations per benchmark.
@@
 \noindent\textbf{Number of threads.}
-To collect the number of active threads used by the an experiment, we use the command \texttt{top} and record at fixed intervals.
+To collect the number of active threads used by the experiment, we use the command \texttt{top} and record at fixed intervals.
 
@@
 The package is available in the GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jvm-comparaison}}
-To add extra \textbf{jvm candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the jvm to be used.\note{The jvm must be provided by \textsf{SDKMAN!}}
+To add extra \textbf{jvm candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the jvm to be used.\note{The JVM must be supported by \textsf{SDKMAN!}}
 For the \textbf{input workload},the benchmarks should be provided in the \texttt{benchmarks} directory.
-As for extra \textbf{metrics}, one can create a new script file that monitor the experiment and record the metrics.
+As for extra \textbf{metrics}, one can create a new script file that monitors the experiment and record the metrics.
 We provide some examples, such as \texttt{recordpower.sh} to measure the instantaneous power and \texttt{recordthreads.sh} to measure the number of active threads during the experiment.
@@
 Our results show that the JVM version can severely impact the energy consumption of the application.
-However, unlike \Cref{fig:JVMs}, one can observe that, depending on applications, latest JVM versions can consume less energy (60\% less energy for \textsf{Scrabble}) or more energy (25\% more energy for the \textsf{Neo4J}).
+However, unlike \Cref{fig:JVMs}, one can observe that, depending on applications, the latest JVM versions can consume less energy (60\% less energy for \textsf{Scrabble}) or more energy (25\% more energy for the \textsf{Neo4J}).
 It is worth noticing that the energy consumption of some benchmarks, such as \textsf{Reactors}, exhibit large variations across JVM versions due to experimental features and changes that are not always kept when releasing LTS versions (version 11 here).
-For example, the introduction of \texttt{VarHandle} to allow low-level access to the memory order modes available in JDK 9 and work along \texttt{Unsafe Class} that was removed from from JVM 11.\footnote{\url{https://blogs.oracle.com/javamagazine/the-unsafe-class-unsafe-at-any-speed}}
+For example, the introduction of \texttt{VarHandle} to allow low-level access to the memory order modes available in JDK 9 and work along \texttt{Unsafe Class} was removed from JVM 11.\footnote{\url{https://blogs.oracle.com/javamagazine/the-unsafe-class-unsafe-at-any-speed}}
 
@@
 One can observe that, depending on the benchmark's focus, the energy efficiency of JVM distributions may strongly vary.
-When considering individual benchmarks, \textsc{J9} performs the worst for at least 6 out of 12 benchmarks---\emph{i.e.}, worst ratio among the 4 tested distributions.
+When considering individual benchmarks, \textsc{J9} performs the worst for at least 6 out of 12 benchmarks---\emph{i.e.}, the worst ratio among the 4 tested distributions.
 Even though, \textsc{J9} can still exhibit a significant energy saving for some benchmarks, such as \textsf{Avrora}, where it consumes 38\% less energy than \textsc{HotSpot} and others.
@@
 
-Interestingly, \textsc{GraalVM} delivers good results overall, being among the distributions with a low energy consumption for all benchmarks, except for \textsf{Reactors} and \textsf{Avrora}.
+Interestingly, \textsc{GraalVM} delivers good results overall, being among the distributions with low energy consumption for all benchmarks, except for \textsf{Reactors} and \textsf{Avrora}.
 Yet, some differences still can be observed with \textsc{HotSpot} depending on applications.
@@
 When it comes to the number of requests processed by \textsf{Scrabbles} during that same amount of time, \textsc{GraalVM} completes $5,336$ requests, against $3,595$ for \textsc{HotSpot} and $2,603$ for \textsc{J9}, as shown in \Cref{table:power/request}.
-The higher power usage for \textsc{GraalVM} helped in achieving a high amount of requests, but also the fastest execution of every, request which was 40\% faster on \textsc{GraalVM}.
+The higher power usage for \textsc{GraalVM} helped in achieving a high amount of requests, but also the fastest execution of every request, which was 40\% faster on \textsc{GraalVM}.
 Thus, \textsc{GraalVM} was more energy efficient, even if it uses more power, which confirms the results observed in \Cref{fig:all_benchs} for this benchmark.
@@
 In total, \textsc{HotSpot} completes $597$ requests against $510$ for \textsc{GraalVM} and $381$ for \textsc{J9}, as shown in \Cref{table:power/request}.
-\textsc{HotSpot} was thus the best choice on the long term, which explains why it is always necessary to consider a warm-up phase and wait for the JIT to be triggered before evaluating the effect of the JVM or the performance of an application.
-This is exactly what we did in our experiments, and why \textsc{HotSpot} was more energy efficient than \textsc{GraalVM} in \Cref{fig:all_benchs}, thus ignoring the warm-up phase would have misleading.
+\textsc{HotSpot} was thus the best choice in the long term, which explains why it is always necessary to consider a warm-up phase and wait for the JIT to be triggered before evaluating the effect of the JVM or the performance of an application.
+This is exactly what we did in our experiments, and why \textsc{HotSpot} was more energy efficient than \textsc{GraalVM} in \Cref{fig:all_benchs}, thus ignoring the warm-up phase would have been misleading.
 
@@
     To answer \textbf{RQ\,1}, we conclude that---while most of the JVM platforms perform similarly---we can cluster JVMs in 3 classes: \textsc{HotSpot}, \textsc{J9}, and \textsc{GraalVM}.
-    The choice of one JVM of these classes can have a major impact on software energy consumption, that strongly depends on the application context.
-    When it comes to the JVM version, latest releases tend to offer the lowest power consumption, but experimental features should be carefully configured, thus further questioning the impact of JVM parameters.
+    The choice of one JVM of these classes can have a major impact on software energy consumption, which strongly depends on the application context.
+    When it comes to the JVM version, the latest releases tend to offer the lowest power consumption, but experimental features should be carefully configured, thus further questioning the impact of JVM parameters.
 \end{mdframed}
@@
 \subsection{Energy Impact of JVM Settings}
-The purpose of our study is not only to investigate the impact of the JVM platform on the energy consumption, but also the different JVM parameters and configurations that might have a positive or negative effect, with a focus on 3 available settings: multi-threading, JIT, and GC.
+The purpose of our study is not only to investigate the impact of the JVM platform on energy consumption, but also the different JVM parameters and configurations that might have a positive or negative effect, with a focus on 3 available settings: multi-threading, JIT, and GC.
 
 \subsubsection{Multithreading}
-The purpose behind this phase is to investigate the impact JVM thread management strategies on the energy consumption.
-This encompasses exploring if the management strategies of application-level parallelism (so called \emph{threads}) results in different energy efficiencies, depending on JVM distributions.
+The purpose of this phase is to investigate the impact JVM thread management strategies on energy consumption.
+This encompasses exploring if the management strategies of application-level parallelism (so-called \emph{threads}) result in different energy efficiencies, depending on JVM distributions.
 
-Investigating such an hypothesis requires a selection of highly parallel and CPU-intensive benchmarks, which is one of the main criteria for our benchmark selection.
+Investigating such a hypothesis requires a selection of highly parallel and CPU-intensive benchmarks, which is one of the main criteria for our benchmark selection.
 As no tool can accurately monitor the energy consumption at a thread level, we monitor the global power consumption and CPU utilization during the execution using RAPL for the energy, and several Linux tools for the CPU-utilization (\texttt{htop}, \texttt{cpufreq}).
@@
 The efficient \textsc{J9} thread management explains why running the \textsf{Avrora} benchmark took much less time and consumed less energy, given that no other difference for the JIT or GC configuration was spotted between the JVMs.
-Another key reasons of the \textsc{J9}'s efficiency for the \textsf{Avrora} benchmark is memory allocation, as \textsf{OpenJ9} adopts a different policy for the heap allocation.
-It creates a non-collectable \emph{thread local heap} (TLH) within the main heap for each active thread.
-The benefit of cloning a dedicated TLH is the fast memory access for independent threads: each thread has its own heap and no deadlock can occur.
+Another key reason for the \textsc{J9}'s efficiency for the \textsf{Avrora} benchmark is memory allocation, as \textsf{OpenJ9} adopts a different policy for the heap allocation.
+It creates a non-collectible \emph{thread local heap} (TLH) within the main heap for each active thread.
+The benefit of cloning a dedicated TLH is the fast memory access for independent threads: each thread has its heap and no deadlock can occur.
 
@@
 In conclusion, JVMs thread management can sometimes constitute a key factor that impacts software energy consumption.
-However, we suggest to check and compare JVMs before deploying a software, especially if the target application is parallel and multi-threaded.
+However, we suggest checking and comparing JVMs before deploying a software, especially if the target application is parallel and multi-threaded.
 
@@
 For \textsc{Hotspot-15}, we conducted experiments while disabling the tiered complication (that generates compiled versions of methods that collect profiling information about themselves), and we also varied the JIT maximum compilation level from \textsf{0} to \textsf{4}, we also tried out \textsc{HotSpot} with a basic \textsc{GraalVM} JIT.
-We note that the level 0 of JIT compilation only uses the interpreter, with no real JIT compilation.
-Levels 1, 2, and 3 use the C1 compiler (called client-side) with different amount of extra tuning.
+We note that level 0 of JIT compilation only uses the interpreter, with no real JIT compilation.
+Levels 1, 2, and 3 use the C1 compiler (called client-side) with different amounts of extra tuning.
 % Furthermore, the level 3 performs a full profiling.
-The JIT C2 (also called server-side JIT) compiler only kicks-in at level 4.
+The JIT C2 (also called server-side JIT) compiler only kicks in at level 4.
 
@@
 
-The $p$-values are computed with the mann-whitney test, with a null hypothesis of the energy consumption being equal to the default configuration.
-The $p$-values in bold show the values that are significantly different from the default configuration with a 95\% confidence, where the values in green highlight the strategies that consumed significantly less energy than default (less energy and significant $p$-value).
+The $p$-values are computed with the Mann-Whitney test, with a null hypothesis of the energy consumption being equal to the default configuration.
+The $p$-values in bold show the values that are significantly different from the default configuration with a 95\% confidence, where the values in green highlight the strategies that consumed significantly less energy than the default (less energy and significant $p$-value).
 
@@
 The \textsf{warm} configuration delivers the closest results to the best results observed with the default configuration.
-Moreover, choosing a low minimum count of method calls seems to have a negative effect on the execution time and the energy consumption.
+Moreover, choosing a low minimum count of method calls seems to have a negative effect on the execution time and energy consumption.
 The only parameter that can give better performance than the default configuration in some cases is the number of parallel JIT threads---using 3 and 7 parallel threads---but is not statistically significant.
@@
 For \textsc{HotSpot}, keeping the default configuration of the JIT is also mostly good.
-In fact, the usage of the C2 JIT is often beneficial (JIT level 4) in most cases, while using the \textsc{GraalVM} JIT reported similar energy efficiency.
+The usage of the C2 JIT is often beneficial (JIT level 4) in most cases while using the \textsc{GraalVM} JIT reported similar energy efficiency.
 Yet, some benchmarks showed that using only the C1 JIT (JIT level 1) is more efficient and even outperforms the usage of the C2 compiler.
@@
 Globally, we conclude through these experiments that keeping the default JIT configuration was more energy efficient in ~80\% of our experiments and for the 3 classes of JVMs.
-This advocates that using the default JIT configuration that can often deliver near-optimal energy efficiency.
+This advocates using the default JIT configuration that can often deliver near-optimal energy efficiency.
 Although, some other configurations, such as using only the C1 JIT or disabling the JVMCI could be advantageous in some cases.
@@
 Changing or tuning the GC strategy has been acknowledged to impact the JVM performances~\cite{10.1145/2568088.2568097}.
-To investigate if this impact also benefits to energy consumption, we conducted a set of experiments on the selected JVMs.
+To investigate if this impact also benefits energy consumption, we conducted a set of experiments on the selected JVMs.
 We considered different garbage collector strategies with a limited memory quantity of 2\,GB, and recorded the execution time and the energy consumption.
@@
 
-\Cref{table:GC} summarizes the results of all the tested GC strategies with our selected benchmarks and the $p$-values of the mann-whitney test, with a null hypothesis of the energy consumption being equal to the default configuration with a 95\% confidence.
-The $p$-values in bold show the values that are significantly different from the default configuration, where the values in green highlight the strategies that consumed significantly less energy than default.
+\Cref{table:GC} summarizes the results of all the tested GC strategies with our selected benchmarks and the $p$-values of the Mann-Whitney test, with a null hypothesis of the energy consumption being equal to the default configuration with a 95\% confidence.
+The $p$-values in bold show the values that are significantly different from the default configuration, whereas the values in green highlight the strategies that consumed significantly less energy than the default.
 For \textsc{GraalVM}, one can see that the GC default configuration is efficient in most experiments, compared to other strategies.
@@
 On the other hand, \textsf{ParallelOldGC} and \textsf{Serial} GC gave better results than the default G1 GC.
-More specifically, the second one consumed 30\% and 6\% less energy than default GC for the \textsf{Neo4j} and \textsf{Dotty} benchmarks, respectively.
+More specifically, the second one consumed 30\% and 6\% less energy than the default GC for the \textsf{Neo4j} and \textsf{Dotty} benchmarks, respectively.
 %Moreover, the Epsilon GC gave worst results compared to the default GC, and also used much more memory( 63 Gb for Epsilon GC against 4Gb for default GC).
@@
 Other settings are less trivial to determine, such as tenure age, memory size, and GC threads count.
-Experiments should thus be conducted on the software to tune the most convenient GC configuration to achieve a better energy efficiency in production.
+Experiments should thus be conducted on the software to tune the most convenient GC configuration to achieve better energy efficiency in production.
 
 Therefore, we noticed during our experiments that, even if using the default GC configuration ensures an overall steady and correct energy consumption, we still found other settings that reduce that energy consumption in ~50\% of our experiments.
-Tuning the GC according to the hosted app/benchmark is thus critical to reduce the energy consumption.
+Tuning the GC according to the hosted app/benchmark is thus critical to reducing energy consumption.
 
@@
 \begin{mdframed}[]
-    To answer \textbf{RQ\,2}, we conclude that users should be careful while choosing and configuring the garbage collector as substantial energy enhancements can be recorded from a configuration to another.
-    The default GC consumed more energy than other strategies in most of the situations.
+    To answer \textbf{RQ\,2}, we conclude that users should be careful while choosing and configuring the garbage collector as substantial energy enhancements can be recorded from one configuration to another.
+    The default GC consumed more energy than other strategies in most situations.
     % choosing the most convenient GC strategy in thus very important to achieve a better software energy efficiency. 
@@
 \section{Threats to Validity}\label{sec:threats}
-A number of issues may affect the validity of our work.
-First, we have the use of the Intel RAPL, one of the most accurate available tools to measure the energy consumption of software~\cite{Khan:2018:RAE:3199681.3177754,10.1145/2989081.2989088}.
+Several issues may affect the validity of our work.
+First, we have the use of the Intel\,RAPL, one of the most accurate available tools to measure the energy consumption of software~\cite{Khan:2018:RAE:3199681.3177754,10.1145/2989081.2989088}.
 However, RAPL only gives the global energy consumption and no fine-grained measures at process or thread levels.
-We used baremetal hardware with a minimal OS and tuned off all the non-essential services and daemons to limit the overhead that the OS may add to the execution, even if it is not substantial~\cite{opaper}.
+We used bare-metal hardware with a minimal OS and turned off all the non-essential services and daemons to limit the overhead that the OS may add to the execution, even if it is not substantial~\cite{opaper}.
 
@@
 
-Benchmarks execution time could also constitute on more subtle threat to the validity of our work, especially for some benchmarks that run fast, such as the \textsf{Pmd} benchmark.
-We thus gave a lot of attention on how long the benchmark is running for the hardware we used, and we tuned the input data workloads to execute benchmarks for at least many (from 10 to hundreds) seconds.
+Benchmarks' execution time could also constitute a more subtle threat to the validity of our work, especially for some benchmarks that run fast, such as the \textsf{Pmd} benchmark.
+We thus gave a lot of attention to how long the benchmark is running for the hardware we used, and we tuned the input data workloads to execute benchmarks for at least many (from 10 to hundreds) seconds.
 Experiments ran at least 30 times to compute the average consumption and the associated standard deviation, therefore reasoning over reasonable dispersion around the average.
@@
 \Cref{table:j-referral} illustrates an example of the final report of \textsf{J-Referral}.
-The tool was tested for 2 real Java projects: \textsf{Zip4J}\footnote{\url{https://github.com/srikanth-lingala/zip4j}} and \textsf{K-nucleotide}.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/performance/knucleotide.html}}.
+The tool was tested for 2 Java projects: \textsf{Zip4J}\footnote{\url{https://github.com/srikanth-lingala/zip4j}} and \textsf{K-nucleotide}.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/performance/knucleotide.html}}
 \textsf{Zip4J} runs a large file compression, while \textsf{K-nucleotide} extracts a DNA sequence, and updates a hashtable of k-nucleotide keys to count specific values.
@@
 \section{Conclusion}
-The results of our investigations showed that many JVMs share energy efficiencies and can grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
+The results of our investigations showed that many JVMs share energy efficiencies and can be grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
 The 3 selected JVM classes can however report a different energy efficiency for different software and/or workloads, sometimes by a large margin.
-While we did not observed a unique champion when it comes to energy consumption, \textsc{GraalVM} reported the best energy efficiency for a majority of benchmarks.
+While we did not observe a unique champion when it comes to energy consumption, \textsc{GraalVM} reported the best energy efficiency for a majority of benchmarks.
 Nonetheless, each JVM can achieve the best or the worst depending on the hosted application.
@@
     \item while the JVM default JIT give often good energy consumption results, some settings may improve the energy consumption and could be tested,
-    \item the choice of the GC may lead to a large impact on the energy consumption in many situations, thus encouraging a careful tuning of this parameter prior to deployment.
+    \item the choice of the GC may lead to a large impact on the energy consumption in many situations, thus encouraging a careful tuning of this parameter before deployment.
 \end{inparaenum}
 To ease the integration of the above guidelines, we propose a tool, named \textsf{J-Referral}, to recommend the most energy-efficient JVM distribution and configuration among more than a hundred considered possibilities.
-It establishes a full report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to help the user to choose the one with the least consumption for a Java Software.
+It establishes a full report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to help the user to choose the one with the least consumption for Java software.
 % TODO : add or change the graphes of the use case 

### 2022-08-02 chapters/optimization/python-dir/python.tex
Proofreading

@@
 \section{Introduction}
+Dynamic programming languages, except Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf. \Cref{fig:pypl}).
+However, it remains unclear whether this category of dynamic programming languages can truly compete with compiled ones in terms of power consumption.
 
-Dynamic programming languages, with the exception of Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf.
-Figure~\ref{fig:pypl}).
-However, it remains unclear whether this category of dynamic programming languages can truly compete with compiled ones in terms of power consumption.
 \begin{figure}[htbp]
     \includegraphics[width=\linewidth]{imgs/programminglanguangespopularity.png}
-    \caption{PYPL Popularity of Programming Languages~\cite{noauthor_pypl_nodate}}
+    \caption{PYPL Popularity of Programming Languages~\cite{noauthor_pypl_nodate}.}
     \label{fig:pypl}
@@
 In particular, Noureddine~\emph{et~al.}~\cite{noureddine_preliminary_2012} in 2012, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
-According to their experiments, an interpreted programming language like Python can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. Figure~\ref{fig:hannoi}).
-In this chapter, we therefore explore the oblivious optimizations that can be applied to Python legacy applications in order to reduce their energy footprint.
-As Python is widely adopted by software services deployed in public and private Cloud infrastructures, we believe that our contributions will benefit to wide diversity of legacy systems and not only favorably contribute to reduce the carbon emissions of ICT, b kmoiut also reduce their cloud invoice for the resources consumed by these services.
-More specifically, this chapter focuses on runtime optimizations that can adopted by developers to leverage the power consumption of Python applications.
-We start by studuing the impact of some of programmers choice such as the type of data structures or loops, on the global energy consumption of the execution code.
-then we discuss some other factors such as level of concurrency etc.
-later we we will talk about another non intrusive ways to optimize the energy consumption of this code.
-one kind of those optimizations includes alternative interpertersand some libraries that are dedicated to optimize the code without changing its structures such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
-
-%%%%%%%%%%% READ AGAIN 
-% articular, Noureddine in 2012 and Pereira in 2017 conducted empirical power measurements on this topic, and both concluded that compiled programming languages consume less power than dynamic ones.They found that a programming language that is interpreted, like Python, can use up to 7,584 times more energy than C.
-As a result, in this chapter, we will look at the obvious optimizations that may be done to Python legacy programs to lower their energy footprint.Since Python is used by many software services in both public and private cloud infrastructures, we think that our contributions will help a wide range of legacy systems by reducing ICT carbon emissions and lowering cloud bills for the resources used by these services.
-This chapter focuses on runtime improvements that developers can employ to reduce the power consumption of Python applications.We start by looking at how the choices made by programmers, like the type of data structure or loop, affect how much energy the code as a whole uses.
-Then we discuss additional topics, such as the level of concurrency, and so on.We will discuss another non-intrusive technique to optimize this code's energy consumption.Alternative interpreters and libraries dedicated to optimizing code without affecting its structures, such as ahead-of-time (AOT) compilation and just-in-time (JIT) libraries maintained by the community, are examples of these optimizations.
-% However, the operational conditions under which these measurements may threaten the validity of their results.
-
-\section{Motivation}
-\subsection{python popularity}
-
-Nowadays, Python seems to attract a large community of developers who are interested in data analysis, web development, system administration, and machine learning—according to a survey conducted in 2018 by JetBrains ---\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}}one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.
-As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
-Rather, we believe that a better option consists of leveraging the strength of this rich ecosystem to promote energy-efficient solutions in order to improve the power consumption of legacy software systems.
-
-% The contributions of this chapter can therefore be summarized as:
-% \begin{compactenum}
-%     \item a classification of state-of-the-art Python optimizations,
-%     \item an evaluation of the power efficiency of these optimizations on well-known micro-benchmarks,
-%     \item a comparison of the effective energy efficiency of Python compared to compiled languages,
-%     \item an assessment of the energy speedup of these optimizations on representative applications and workloads.
-% \end{compactenum}
-
-\subsection{python gluttony}
-
-In the other hand,because of the interpreted nature of Python code, this programming language has paid a high price in terms of performance.  \ref{fig:clbg}, memory and energy consumption.
+According to their experiments, an interpreted programming language, like Python, can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. \Cref{fig:hannoi}).
+In this chapter, we explore the oblivious optimizations that can be applied to Python legacy applications to reduce their energy footprint.
+As Python is widely adopted by software services deployed in public and private cloud infrastructures, we believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
+More specifically, this chapter focuses on runtime optimizations that can be adopted by developers to leverage the power consumption of Python applications.
+We start by studying the impact of programmer choices, such as the type of data or control structures, on the global energy consumption of the execution code.
+Then, we discuss other factors, such as the levels of concurrency, before we investigate other non-intrusive approaches to optimize the energy consumption of applications.
+One type of optimization includes alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
 
@@
 
-According to \cite{pinto_energy_2017} and \cite{noureddine_preliminary_2012}, Python tends to be one more energy hungry .
-As onne can notice in figure \ref{fig:hannoi}, python consumes 30 times more than C or C++.
-The test was done on an implementation of \fnurl{hannoi test}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi}  30 disks.
+\section{Motivation}
+
+\subsection{Python Popularity}
+Nowadays, Python seems to attract a large community of developers who are interested in data analysis, web development, system administration, and machine learning.
+According to a survey conducted in 2018 by JetBrains,\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}} one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.
+As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
+Rather, we believe that a better option consists of leveraging the strength of this rich ecosystem to promote energy-efficient solutions to improve the power consumption of legacy software systems.
+
+\subsection{Python Gluttony}
+% In the other hand,because of the interpreted nature of Python code, this programming language has paid a high price in terms of performance.  \ref{fig:clbg}, memory and energy consumption.
+According to~\cite{pinto_energy_2017} and~\cite{noureddine_preliminary_2012}, Python tends to be one more energy hungry programming language.
+As one can notice in \Cref{fig:hannoi}, Python consumes $30$ times more than C or C++.
+The benchmark was done with an implementation of the \fnurl{Tower of Hanoi}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi} of 30 disks.
 
 Python consumes a lot of energy, mainly because it is slow in execution.
-Its flexibility and simplicity caused it to drop off in performance, because Python gains its flexibility from being a dynamic language.
-Therefore, it needs an interpreter to execute its programs, which makes them much slower compared to the others that are written in compiled programming languages such as C and C++ or semi-compiled languages like Java.
+Its flexibility and simplicity caused it to drop off in performance because Python gains its flexibility from being a dynamic language.
+Therefore, it needs an interpreter to execute its programs, which makes them much slower compared to the others that are written in compiled programming languages, such as C and C++ or semi-compiled languages like Java.
 
-As shown in Table\ref{fig:clbg}, one can see that in most of the test cases, Python takes more time to execute-the only case that he wasn't the last one was in the benchmark regx-redux where he beat Go-, and in some cases the gap was huge, such as in the n-body test where Python took around 100 times more than C++.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}
+As shown in \Cref{fig:clbg}, one can observe that, for most of the applications taken for the \emph{Computer Language Benchmark Game} (CLBG), Python takes more time to execute---the only case that he was not the worst one was in the benchmark \textsf{regx-redux} where he beat Go---and in some cases the gap was huge, such as in \textsf{n-body} where Python took around $100$ times more than C++.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}
+
 \begin{table}[hbt]
-    \begin{tabular}{l|*{5}c}
-                           & C    & C++   & Java & python & Go    \\
+    \centering
+    \caption{Comparison of CLBG execution times (in seconds) depending on programming languages.}
+    \label{fig:clbg}
+    \begin{tabular}{l|*{5}r}
         \hline
-        pidigits           & 1.75 & 1.89  & 3.13 & 3.51   & 2.04  \\
-        reverse-complement & 1.75 & 2.95  & 3.31 & 16.76  & 4.00  \\
-        regx-redux         & 1.45 & 1.66  & 10.5 & 15.56  & 28.69 \\
-        k-nucleotide       & 5.07 & 3.66  & 8.66 & 79.79  & 15.36 \\
-        binary-trees       & 2.55 & 2.63  & 8.28 & 92.72  & 28.90 \\
-        fasta              & 1.32 & 1.33  & 2.32 & 62.88  & 2.07  \\
-        Fannkuch-redux     & 8.72 & 10.62 & 17.9 & 547.23 & 17.82 \\
-        n-body             & 9.17 & 8.24  & 22.0 & 882.00 & 21.00 \\
-        spectral-norm      & 1.99 & 1.98  & 4.27 & 193.86 & 3.95  \\
-        Mandelbort         & 1.64 & 1.51  & 6.96 & 279.68 & 5.47  \\
+                                    & \bf C    & \bf C++  & \bf Java & \bf Python & \bf Go \\
+        \hline
+        \hline
+        \textsf{pidigits}           & \best{1.75} & 1.89  & 3.13 & \worst{3.51}   & 2.04  \\
+        \textsf{reverse-complement} & \best{1.75} & 2.95  & 3.31 & \worst{16.76}  & 4.00  \\
+        \textsf{regx-redux}         & \best{1.45} & 1.66  & 10.5 & 15.56  & \worst{28.69} \\
+        \textsf{k-nucleotide}       & 5.07 & \best{3.66}  & 8.66 & \worst{79.79}  & 15.36 \\
+        \textsf{binary-trees}       & \best{2.55} & 2.63  & 8.28 & \worst{92.72}  & 28.90 \\
+        \textsf{fasta}              & \best{1.32} & 1.33  & 2.32 & \worst{62.88}  & 2.07  \\
+        \textsf{Fannkuch-redux}     & \best{8.72} & 10.62 & 17.9 & \worst{547.23} & 17.82 \\
+        \textsf{n-body}             & 9.17 & \best{8.24}  & 22.0 & \worst{882.00} & 21.00 \\
+        \textsf{spectral-norm}      & 1.99 & \best{1.98}  & 4.27 & \worst{193.86} & 3.95  \\
+        \textsf{Mandelbort}         & 1.64 & \best{1.51}  & 6.96 & \worst{279.68} & 5.47  \\
+        \hline
     \end{tabular}
-    \caption{Comparison of the execution time between different programming languages using the computer language benchmarks game}
-    \label{fig:clbg}
 \end{table}
 
-\subsection{Goal}
+\subsection{Python Limits}
+To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
+\Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers.\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}.
+In this survey, multiple answers were accepted.
+57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration.
 
-To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
-Figure~\ref{fig:usecase} illustrates a study made by the JetBrain company on Python developers.
-\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}.In this survey, multiple answers were accepted.
-As one can see, 57\% of the respondents said that they use Python for data science, 51\% said they are using it for web development, and around 40\% said they are using it for system administration.
 \begin{figure}[hbt]
@@
     \includegraphics[width=\linewidth]{imgs/python_use_cases}
-    \caption{Use cases of python }
+    \caption{Use cases of Python (source: JetBrain).}
     \label{fig:usecase}
@@
 
-In this chapter, we will first investigate its behavior in the most common usage cases.Following that, we will focus on the language's structures in the hope of discovering some fundamental guidelines for more general purposes, as \citeauthor{hasan_energy_2016-1} did in their article~\cite{hasan_energy_2016-1}.
-Following that, we will measure the energy consumption of several Python implementations in order to determine a non-intrusive technique to improve energy efficiency.
-Therefore this chapter will target the following research questions :
+In this chapter, we investigate the energy footprint of Python in its most popular domains of adoption.
+We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016-1} did in~\cite{hasan_energy_2016-1}.
+Then, we measure the energy consumption of several Python implementations to propose a non-intrusive technique to improve energy efficiency.
+Therefore, this chapter will address the following research questions:
 \begin{compactenum}[\indent\bf RQ\,1:]
-    \item \emph{What is the behavior of Python when it is used as a data science tool?}
-    \item \emph{Are the default guidelines of Python energy efficient?}
-    \item \emph{Can we reduce the energy consumption of Python programs without impacting the source code?}
+    \item \emph{What is the energy footprint of Python when used in data science?}
+    \item \emph{Are the Python guidelines energy efficient by construction?}
+    \item \emph{Can we reduce the energy consumption of Python programs without altering the source code?}
 \end{compactenum}
 
-For this chapter we are considering 4 separate yet related studies .
-First, we are going to see the energy behavior of the python within two study cases.
-Web servers and machine learning.
-We will then delve deeper into the energy consumption of Python basic structures before concluding with the impact of Python interpterer on energy consumption.
+In this chapter, we report on 4 case studies that intend to answer these research questions.
+First, we study the energy behavior of Python in two application contexts: web applications (cf. \Cref{sec:webdev}) and machine learning.
+Then, we dive deeper into the energy consumption of Python core structures, before concluding with the impact of the Python interpreter on energy consumption.
 
-% \section{Experimental Protocol}\label{sec:pythonprotocol}
-% In this section, we discuss the experimental setting, including hardware components, software components, and the methodology that we adopted to answer the previous research questions.
-% For each of the questions that follow, we will first talk about how the experiment was done, and then we will talk about the results.
+\section{Green Web Development}\label{sec:webdev}
+We start by studying the impact of the \emph{Object Relational Mapping} (ORM)~\cite{o2008object} on energy consumption.
+% We have chosen Django as our web framework.
+The idea behind this use case is to observe if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
+We, therefore, create a Django website to analyze the cost of a single request using various mechanisms to retrieve data from the database.
 
-% %% should i include the part with the exceptions ??? naah no need so
+We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records.
+\begin{enumerate}
+    \item \textsf{Vanilla} relies only on the ORM to retrieve the data,
+    \item \textsf{Prefetch} queries the data before being requested,
+    \item \textsf{Optimized} leverages SQL without passing by the ORM.
+\end{enumerate}
 
-% %% except the parallellism part , which i dont know how to link it here yet 
-% \section{Workload} :
-% for this chapter we are consediring 4 seperated, yet related studies .
-% First we are going to see the energy behavriour of python within two study cases.
-Web servers and Machine learning.
-later we will dig deeper into the energy consumption of python basic structures where finally we will conclude with the impact of python interpterer on the energy consumption.
-
-% \subsubsection{micro-benchmarks}
-
-% \subsubsection{benchmarks}
-
-\section{web developement}
-The purpose of this study is to understand the impact of the \emph{ORMs} (Object Relational Mapping)~\cite{o2008object}. We have chosen django as our web framework
-
-As a result, we create a simple website in which we attempt to analyze the cost of a single request using various mechanisms to retrieve data from the database.
-
-The idea behind this use case is to see whether the choice of the database and the orm (object relational mapper) impacts the energy consumption and the performance of the website.
-
-We considered using two different databases, PostgreSQL and SQlite3, that contained the same data, and three different ways to fetch the data.
-\begin{enumerate}
-    \item the naive version: which relies only on the ORM to get the data
-    \item Prefetch : basically we just prefetch the data before we request it
-    \item optimized : and here we optimize our request on the SQL level without passing by the ORM
-\end{enumerate}
+%%%%% have to rephrase this 
+As one can observe in \Cref{fig:django}, the strategy to query stored data has a huge impact on the energy consumption.
+As the \textsf{Vanilla} strategy can consume up to $10\times$ more energy than the \textsf{Optimized} one.
+Conversely, the choice of the database does not exhibit a key impact on the total energy despite their different behavior regarding the execution time and the average power.
+This can be useful to support developers in choosing which database engine they can adopt, based on the number of expected requests and the targeted performance.
 
@@
     \includegraphics[width=\linewidth]{imgs/django}
-    \caption{energy behaviour based on multiprocessing}
+    \caption{Energy behavior resulting from data access strategies.}
     \label{fig:django}
@@
 
-%%%%% have to rephrase this
+Another interesting observation is the impact of the interpreter, as \Cref{fig:django} highlights.
+For example, using the \textsf{Pypy} interpreter reduces the energy consumption, even when we adopting the \textsf{Vanilla} strategy.
 
-As one can see in figure~\ref{fig:django}, different choices of requesting the data have a huge impact on the energy consumption.
-as the naive version can consume up to 10x more energy than the optimized one.
-In the other hand the choice of the data base didnt have huge impact on the total energy despite their different behaviour regarding the execution time and the average power.
-This can be useful to help develeppers make a choice regarding which data base they can use based on the number of the expected requests and the expectation of the performance.
-
-Another interesting observation is the impact of the interpreter, as figure~\ref{fig:django} highligh.
-the use of Pypy interpter instead of the default one helped reducing the amount of the energy consumption even when we are dealing with the naive version.
-
-Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
-As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
+\Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
+Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).
+However, due to the increase of cores/threads per CPU, many Python libraries started to take advantage of this hardware feature by allowing concurrent execution of Python processes.
+For example, the multiprocessing library\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}} spawns sub-processes to increase the degree of concurrency of Python applications.
+As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of threads until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.
 
@@
     \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-    \caption{energy behaviour based on multiprocessing}
+    \caption{Energy consumption of Python multiprocessing depending on the number of exploited threads.}
     \label{fig:python_multiprocessing}
@@
 
-as we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
-where different sub process have to compete for the CPU Ressources.
-Another finding is when we hit the number of physical cores.
+Before reaching the limit of physical cores, we also observe that the scheduler of the operating system tends to favor the execution of processes on the same physical core, by taking advantage of the hyper-thread feature.
+While this strategy aims to save energy by leveraging the ACPI P-states and C-states of unallocated cores, this leads to increase execution times.
+
+% Another key finding is when we hit the number of physical cores.
 there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
-basicaly he favorits the hyper threads than the physical core wich will lead to some context switches which cause the slow behaviour, howerver in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the eneryg consumption in this case.
-in the Chapter %% reference the work of intern from lyon  
-we discuss a deeper this behaviour of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
+Basically it favors the hyper threads than the physical core which will lead to some context switches which cause the slow behavior, however in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the energy consumption in this case.
+% in the Chapter %% reference the work of intern from lyon  
+% we discuss a deeper this behavior of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
 
-\subsection{python insights}
-
-The goal of this section is to find some insights where we can optimize the energy consumption without impacting the source code.
-so we started by extending the work of \citeauthor{hasan_energy_2016} and \citeauthor{oliveira_recommending_nodate} to the python environement.
+\subsection{Python Insights}
+This section aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
+Therefore, we start by extending the work of \citeauthor{hasan_energy_2016} and \citeauthor{oliveira_recommending_nodate} to the context of Python environments.
 
 %%%% concurency 
+Another field of investigation is the type of data and control structures that might impact the energy consumption.
+We iterate over a list using 3 methods.
+First, the classical \texttt{for(i in range(len(n)))}.
+However, as we can see here, unlike other programming languages, it requires extra operations, such as determining the length of the collection and then using the iterator range.
+So, we tried the more adapted version \texttt{for(element in collection)}.
+Moreover, in most programming languages.
+the \texttt{for} loop is translated to a \texttt{while} loop ( transformation from D type to B type -- asm -- ), therefore we wanted to compare this with a \texttt{while} version.
+After determining the main ways to iterate over a \texttt{loop}, we run the collection algorithms of different data types to observe if they impact the energy consumption of the code, and we repeated it for the size of the collection.
 
-another field of investigation is the type data and structure conroles. that might impact the energy consumption.
-To do so. We iterate over a list using three methodes.
-first the classical for ( for i in range (len(n))).
-howerver as we can see here unlike other porgramming langauges it requires extra operations such as determining the length of the collection , and then using the iterator range.
-so we tried the more adapted version
-( for element in collection).
-Morever in most programming languages the for loop is translated to a while loop ( transfirmation from D type to B type -- asm -- ), therefore we wanted to compare this with a ( while ) version.
-After determining the main ways to iterate over a loop we run the algorithms  collections of different data types to see wether it will impact of the energy consumption of the code, same thing for the size of the collection.
+As one can see in \Cref{fig:pythonloops}, the type of the data has no impact on the energy consumption.
+However, the way one iterates over the collection has a huge factor.
+Interestingly, the \texttt{for in range} loop was by far the optimal one, followed by the regular \texttt{for in collection}, and the \texttt{while} part was the last one with an overhead of 400\% compared to the first option.
 
-As we can see in figure~\ref{fig:pythonloops} the type of the data hasn't an impact on the energy consumption.In the other hand the way we iterate over the collection has a huge factor.
-interestely, the for in range loop was by far the optimal one with following by the regural for in collection, and the while part was the least one with an overhead of 400\% compared to the first option.
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/python_iterations}
+    \caption{Comparison of the energy consumption of different Python loops.}
+    \label{fig:pythonloops}
+\end{figure}
 
 %%% putting the idea, later i ll reformulate it 
-the reason behind a such behaviour is mainly related on how python interpeter is made.
+The reason behind such a behavior is mainly related to how the Python interpreter is implemented.
 % add refenece https://www.python.org/doc/essays/list2str/ 
@@
 %% http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1
-In order to reduce the latency of the python code.
-most of the buitl-in functions an operations are written in C, same thing goes for the function range, further more the function len has a complexity of O(1) because it is based on the function Py\_SIZE  of C which stores the length in a field for the object.
-so basically for the for in range is basically creating a new iterator that has the same length of the first one.
-and for each iteration we will have to do a second acceess ( l[i]) instead of one one access hence the double time.
-for the wile is even slower due to the implicite incrementation of the variable which will cause an extra operation during the loop .
-to confirm the hyphothsis.
-we tried to construct a new list by editing the elements of the previous one.
-And as predicted , the builtin methodes are the best energy saving, while the customize while loop is the heaviest.
-an interesting finding is.
-the impact of annonymous functions ( lambda expressions) on the energy consumption.
-the reseason behind this is the fact that python treats those functions as local variables unlike the predifined ones which are global in our case.
-therefore they are faster and consumes less energy  % same reference https://www.python.org/doc/essays/list2str/
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/python_iterations}
-    \caption{energy behaviour of different python loops }
-    \label{fig:pythonloops}
-\end{figure}
+To reduce the latency of Python applications, most of the built-in functions and operations are written in C, and the same goes for the function \texttt{range}.
+Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$ as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object.
+Therefore, the \texttt{for in range} is creating a new iterator that has the same length as the first one and, for each iteration, requires a second access (\texttt{l[i]}) instead of one---explaining the doubled time.
+The \texttt{while} is even slower due to the implicit increment of the variable, which causes an extra operation during the loop.
+To confirm this hypothesis, we tried to construct a new list by editing the elements of the previous one (cf. \Cref{fig:pythontreatement}).
+And, as predicted, the built-in methods are the most energy-saving ones, while the customize \texttt{while} loop is the heaviest.
 
@@
     \includegraphics[width=\linewidth]{imgs/python_treatemens}
-    \caption{energy behaviour of different methodes to change a list}
+    \caption{Comparison of the energy consumption of different methods to convert a list.}
     \label{fig:pythontreatement}
@@
 
-\paragraph{conclusion}
-This study has shown us, that the optimial way to reduce the energy consumption of the python code is to follow the guidelines and use the builtin functions, which is kinda happy news for the developers since they don't have to make extra effort to make their code green
+Another interesting finding is the impact of anonymous functions (also known as lambda expressions) on energy consumption.
+The reason is that Python treats these functions as local variables, unlike the predefined ones which are global in our case.
+Therefore, they are faster and consume less energy.  % same reference https://www.python.org/doc/essays/list2str/
 
-\subsection{python and multiprocessing}
-Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
+\paragraph{Conclusion}
+This study demonstrated that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
+
+\subsection{Python \& Multiprocessing}
+\Cref{fig:python_multiprocessing} compares the behaviour of python programs when we try to introduce the parallelism.
 As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY

### 2022-08-02 chapters/optimization/python-dir/python_interperters.tex
add : experemental setup for the ml experiment
add : p-values for python
fix: Python chapter

@@
-\section{Python interpreters}\label{sec:pythoninterpreters}
+\section{Python Interpreters}\label{sec:pythoninterpreters}
+%%% note : add this to the section of python interpreters 
+Due to the lack of support for most non-conventional Python interpreters, we mainly focus on micro-benchmarks.
+Except for \textsc{PyPy}, most of the Python implementations do not support extra Python libraries, despite those extra implementations being developed to optimize a specific library, such as \textsf{Numba} with \textsf{Numpy}, or \textsf{intelpython} with machine learning algorithms.
 
-%%% note : add this to the section of python interpreters 
-Due to the lack of of support for most of  non conventional pyton interpeters, our main focus was only on micro benchmarks.
-Except pypy most of the python implementations don't support extra python libraries, Despite that most of the cases those extra implementations were made to optimize a specific library such as numba with numpy, or intelpython with machine learning algorthms.
+\paragraph{Preliminary studies}
+For the first studies, we used the official version of Python, because the goal was mainly to highlight the impact of the structure of code on energy consumption.
+One main drawback of the previous method is the work to be done to update the existing code base to reduce  energy consumption.
+To avoid such hustle, we tried to find a non-intrusive approach to make the Python code more eco-friendly without altering its structure.
+Python is an interpreted language, which led many initiatives to implement their own interpreter to improve one or many aspects of the Python code.
+In the following section, we discuss the impact of those implementations on the energy consumption of python programs, and in which case, one should use a non-conventional interpreter to save the energy consumption of their application.
 
-\paragraph{preliminary studies}
-For the first studies we used only the official version of python. because the goal was mainly to highlight the impact of the structure of code on the energy consumption.
-One main inconvienent of the previous methode to reduce the energy consumption, is the hard work that should be done in order to alter the existing code base in order to redcue the energy consumption.
-To avoid such hustle we tried to find a non intrusive way to make the python code more ecofriendly without changing its structure.
-One aspect of python is the fact that it is an interptered language which led many works to create their own implementation of the interpeter to enhance one or many aspect of the python code.
-in the following section we will discuss the impact of those implementations on the energy consumption of python programs, and in which case, one should use a non conventional interpter to save the energy consumption of their program.
-
-To do so, we gathered a list of interpeters, transpilers and some optimization libraries.
-
-the following list, presents the most famous ones .
-% We should explain the methodology we used to select these
-
+To do so, we gathered a list of interpreters, transpilers and other optimization libraries that can contribute to reduce the energy consumption of legacy Python applications:
 \begin{enumerate}
-    \item \textbf{CPython}:\footnote{\url{https://www.python.org/}} a Python interpreter written in C programming language.
-It is the reference interpreter of Python.
-CPython compiles the source code into a bytecode and then interprets it.
-The CPython project still supports both versions of Python 2 and 3;
-    \item \textbf{PyPy}:\footnote{\url{http://pypy.org}} An alternative implementation of the Python interpreter.
-It is written using \emph{RPython} in order to use the JIT.
+    \item \textbf{CPython}:\footnote{\url{https://www.python.org/}}
+    This Python interpreter, written in C, is the reference interpreter of Python.
+    CPython compiles the source code into byte-code and then interprets it.
+    The CPython project supports both versions of Python 2 and 3;
+    \item \textbf{PyPy}:\footnote{\url{http://pypy.org}}
+    An alternative implementation of the Python interpreter.
+    It is written using \emph{RPython} to use the JIT.
     It compiles the most used portions of the Python code into a binary code for better performance.
-To benefit from these optimizations, the program has to be executed for at least for few seconds so the JIT has enough time to warm up, the JIT optimization will be applied only the code written by the programmer and not external libraries;
-    \item \textbf{Cython}:\footnote{\url{https://github.com/cython/cython}} a static compiler for Python.
-It translates the Python code into a C code, and then compiles it using a C compiler.
-It also support(s) an extended version of the Python language that allows programmers to call \emph{C functions}, declare \emph{C types} and use static types which will help the translation of Python objects into native types, such as integers, float, which often means a better performances, since native C libraries are almost all the time faster than the python written once~\cite{pereira_energy_2017}.
-    \item \textbf{Intel\,Python}:\footnote{\url{https://software.intel.com/en-us/distribution-for-python}} A customized interpreter developed by Intel in order to enhance performances of Python programs.
-It is dedicated to data sciences, high-Performance computing.
+    To benefit from these optimizations, the program has to be executed for at least for few seconds so the JIT has enough time to warm up, the JIT optimization are only applied to the code written by the developer and not to external libraries;
+    \item \textbf{Cython}:\footnote{\url{https://github.com/cython/cython}}
+    A static compiler for Python.
+    It translates the Python code into C, and then compiles it using a C compiler.
+    It also supports an extended version of the Python language that allows programmers to call \emph{C functions}, declare \emph{C types} and use static types, which will help the translation of Python objects into native types, such as integers, float.
+    This often means better performances, since native C libraries are almost all the time faster than the Python written once~\cite{pereira_energy_2017};
+    \item \textbf{Intel\,Python}:\footnote{\url{https://software.intel.com/en-us/distribution-for-python}}
+    A customized interpreter developed by Intel to enhance performances of Python programs.
+    It is dedicated to data sciences and high-performance computing.
     It uses some Intel kernel libraries, such as Math Kernel Library (Intel MKL\footnote{\url{https://software.intel.com/en-us/mkl}}) and data analytics acceleration library (Intel DAAL\footnote{\url{https://software.intel.com/en-us/intel-daal}}).
     It supports both versions of Python;
-    \item \textbf{Active\,Python}:\footnote{\url{https://www.activestate.com/products/activepython/}} developed by the Activestates company, it provides a standardized Python distribution to ensure license compliance, security, compatibility and performance.
-Therefore, ActivePython has its own built-in packages (more than 300 packages) and supports both versions of Python;
-    \item \textbf{IronPython}:\footnote{\url{https://ironpython.net}} A .Net-based Python interpretation platform written in C\# in order to be used over the .Net vm or mono, it benefits from all the optimizations of .Net virtual machine, such as the JIT and garbage collector mechanisms.
-    \item \textbf{GraalPython}:\footnote{\url{https://github.com/graalvm/graalpython/}} A Python interpreter that is based on GraalVM \footnote{\url{https://www.graalvm.org/docs/why-graal/}} (a universal virtual machine developed by oracle for running applications written in different programming languages).
+    \item \textbf{Active\,Python}:\footnote{\url{https://www.activestate.com/products/activepython/}}
+    It is developed by the Activestates company and provides a standardized Python distribution to ensure license compliance, security, compatibility and performance.
+    Therefore, ActivePython implements its built-in packages (more than 300 packages) and supports both versions of Python;
+    \item \textbf{IronPython}:\footnote{\url{https://ironpython.net}}
+    A .Net-based Python interpretation platform written in C\# that is used with the .Net virtual machine or Mono.
+    It benefits from all the optimizations of .Net virtual machines, such as the JIT and garbage collector mechanisms;
+    \item \textbf{GraalPython}:\footnote{\url{https://github.com/graalvm/graalpython/}}
+    A Python interpreter that is based on GraalVM\footnote{\url{https://www.graalvm.org/docs/why-graal/}} (a universal virtual machine developed by oracle for running applications written in different programming languages).
     For the time being, it only supports Python~3 and it is still in the experimental stage;
-    \item \textbf{Jython}:\footnote{\url{https://jython.github.io}} Implementation of Python programming language written in Java for the \emph{Java virtual machine} (JVM).
-Similar to IronPython and GraalPython, it leverages the optimization mechanisms provided by the JVM to enhance the Python programs;
-    \item \textbf{MicroPython}:\footnote{\url{http://micropython.org}} a lightweight Python version dedicated to embedded systems and micro-controllers;
-    \item \textbf{Nuitka}:\footnote{\url{http://nuitka.net/pages/overview.html}} a Python compiler written in Python that generates a binary executable from Python code.
+    \item \textbf{Jython}:\footnote{\url{https://jython.github.io}}
+    An implementation of Python programming language written in Java for the \emph{Java Virtual Machine} (JVM).
+    Similar to IronPython and GraalPython, it leverages the optimization mechanisms provided by the JVM to enhance the Python performances;
+    \item \textbf{MicroPython}:\footnote{\url{http://micropython.org}}
+    A lightweight Python version dedicated to embedded systems and micro-controllers;
+    \item \textbf{Nuitka}:\footnote{\url{http://nuitka.net/pages/overview.html}}
+    A Python compiler written in Python that generates a binary executable from Python code.
     It translates the Python code into a C program that is then compiled into a binary executable;
-    \item \textbf{Numba}:\footnote{\url{https://numba.pydata.org}} a library that includes JIT compiler in order to enhance the performances of Python functions using the industry-standard LLVM compiler library;
-    \item \textbf{Shedskin}:\footnote{\url{https://github.com/shedskin/shedskin}} a static transpiler that translates implicitly statically typed python into C++ code.
-    \item \textbf{Hope}~\cite{akeret_hope_2015}: a Python library that aims to introduce JIT compiler into the Python code;
-    \item \textbf{Parakeet}~\cite{DBLP:conf/hotpar/RubinsteynHWS12}: a runtime accelerator for an array-oriented subset of Python;
-    \item \textbf{Stackless\,Python}:\footnote{\url{https://github.com/stackless-dev/stackless/wiki}} an interpreter that focus on enhancing multi-threading programming.
-    \item \textbf{Pyjion}:\footnote{\url{https://github.com/microsoft/pyjion}} JIT API for CPython, same purpose as Parakeet and Hope.
-    \item \textbf{Pyston}:\footnote{\url{https://blog.pyston.org}} performance-oriented Python implementation built using LLVM and modern JIT techniques. the project is funded by dropbox;
-    \item \textbf{Grumpy}:\footnote{\url{https://github.com/google/grumpy}} a source-to-source transpiler that translates the Python code into a Go code that will be compiled to a binary executable.
-It also offers an interpreter called \emph{grumprun} which can directly execute the Python code.
-Unfortunately, we cannot use it because the project is already outdated (last commit is in 2017) and it has a lot of limitation in term of supporting the Python language, such as some built-in functions and standard libraries;
-    \item \textbf{Psyco}:\footnote{\url{http://psyco.sourceforge.net}} a JIT compiler for Python;
-    \item \textbf{Unladen\,Swallow}:\footnote{\url{https://unladen-swallow.readthedocs.io/en/latest/}} an attempt to (use) LLVM  as  JIT compiler for CPython.
+    \item \textbf{Numba}:\footnote{\url{https://numba.pydata.org}}
+    A library that includes JIT compiler to enhance the performances of Python functions using the industry-standard LLVM compiler library;
+    \item \textbf{Shedskin}:\footnote{\url{https://github.com/shedskin/shedskin}}
+    A static transpiler that translates implicitly statically typed python into C++ code;
+    \item \textbf{Hope}~\cite{akeret_hope_2015}:
+    A Python library that aims to introduce JIT compiler into the Python code;
+    \item \textbf{Parakeet}~\cite{DBLP:conf/hotpar/RubinsteynHWS12}:
+    A runtime accelerator for an array-oriented subset of Python;
+    \item \textbf{Stackless\,Python}:\footnote{\url{https://github.com/stackless-dev/stackless/wiki}}
+    An interpreter that focuses on enhancing multi-threading programming;
+    \item \textbf{Pyjion}:\footnote{\url{https://github.com/microsoft/pyjion}}
+    A JIT API for CPython, same purpose as Parakeet and Hope;
+    \item \textbf{Pyston}:\footnote{\url{https://blog.pyston.org}} 
+    A performance-oriented Python implementation built using LLVM and modern JIT techniques.
+    The project is funded by Dropbox;
+    \item \textbf{Grumpy}:\footnote{\url{https://github.com/google/grumpy}} 
+    A source-to-source transpiler that translates the Python code into Go before being compiled to a binary executable.
+    It also offers an interpreter, called \emph{grumprun}, which can directly execute the Python code. 
+    Unfortunately, we cannot use it because the project is already outdated (last commit is in 2017) and it has a lot of limitations in terms of supporting the Python language, such as some built-in functions and standard libraries;
+    \item \textbf{Psyco}:\footnote{\url{http://psyco.sourceforge.net}}
+    A JIT compiler for Python;
+    \item \textbf{Unladen\,Swallow}:\footnote{\url{https://unladen-swallow.readthedocs.io/en/latest/}}
+    An attempt to (use) LLVM as a JIT compiler for CPython.
 \end{enumerate}
@@
 \subsection{Runtime Classification}
-Before futher proceeding with the list of candidate runtimes for Python applications, we propose a classification according to several criterions:
+Before further proceeding with the list of candidate runtime for Python applications, we propose a classification according to several criteria:
 \begin{description}
     \item[\bf Type] refers to the category of runtime infrastructure that supports the execution of a Python application.
-In particular, we consider 3 types of environments: \emph{Interpreter}, \emph{Compiler} and \emph{Library}.
-\emph{Interpreter} refers to the class of environment that does not require any preprocessing of Python source code.
-\emph{Compiler} introduces a compilation phase prior to the execution of the application.
-Finally, \emph{Library} induces some modification of the source code.
+    In particular, we consider 3 types of environments: \emph{Interpreter}, \emph{Compiler} and \emph{Library};
+    \emph{Interpreter} refers to the class of environment that does not require any preprocessing of Python source code;
+    \emph{Compiler} introduces a compilation phase before the execution of the application.
+    Finally, \emph{Library} requires some modification of the source code;
     \item[\bf Runtime] refers to the technology supporting the execution of a Python application.
-This technology can refer to the programming language used to program the interpreter, the target language for a compiler or a library.
-    \item[\bf JIT optimisation] refers to the support of \emph{just-in-time} compilation in the runtime infrastructure supporting the execution of the application.
-    \item[\bf GC optimisation] refers to the support of \emph{garbage collection} in the runtime infrastructure supporting the execution of the application.
+    This technology can refer to the programming language used to program the interpreter, the target language for a compiler or a library;
+    \item[\bf JIT optimization] refers to the support of \emph{just-in-time} compilation in the runtime infrastructure supporting the execution of the application;
+    \item[\bf GC optimization] refers to the support of \emph{garbage collection} in the runtime infrastructure supporting the execution of the application;
     \item[\bf Python version(s)] refers to the list of Python source code versions supported by the runtime environment.
@@
 We should explain the classification of these runtimes
-
 \begin{table}
@@
         \hline
-        Stackless\,Python         & Interpreter               & python                       & \no                                    & \no                             & \yes    & \no     \\
+        Stackless\,Python         & Interpreter               & Python                       & \no                                    & \no                             & \yes    & \no     \\
         \hline
-        MicroPython               & interpreter               & c                            & \no                                    & \no                             & \no     & \yes    \\
+        MicroPython               & Interpreter               & c                            & \no                                    & \no                             & \no     & \yes    \\
         \hline
-        Pyston                    & interpreter               & LLVM                         & \yes                                   & \no                             & \yes    & \no     \\
+        Pyston                    & Interpreter               & LLVM                         & \yes                                   & \no                             & \yes    & \no     \\
         \hline
         Unladen\,Swallow          & Interpreter               & LLVM                         & \yes                                   & \no                             & \yes    & \no     \\
+        \hline
         \hline
@@
         \hline
+        \hline
         Numba                     & Library                   & C                            & \yes                                   & \no                             & \yes    & \yes    \\
         \hline
-        Hope                      & Library                   & python                       & \yes                                   & \no                             & \yes    & \yes    \\
+        Hope                      & Library                   & Python                       & \yes                                   & \no                             & \yes    & \yes    \\
         \hline
-        Psyco                     & Library                   & python                       & \yes                                   & \no                             & \yes    & \yes    \\
+        Psyco                     & Library                   & Python                       & \yes                                   & \no                             & \yes    & \yes    \\
         \hline
@@
         \hline
-        Parakeet                  & library                   & C                            & -                                      & \no                             & \yes    & \no     \\
+        Parakeet                  & Library                   & C                            & -                                      & \no                             & \yes    & \no     \\
         \hline
@@
 
-There are other implementations that we didn't considerate because either the project aborted many years ago or it has a very limited support for python features.
-After the collection of those implementation we filtered them. in order to keep only the versions that are still maintained and support most of python features.
-and we classified them into 3 categories depending of their integration with the python code.
-In the \ref{fig:pythonimplementations} we describe the implementations that we kept and the version of each implementation and its category.
+There are other implementations that we did not consider because either the project aborted many years ago or it has very limited support for Python features.
+After the collection of those implementations, we filtered them. To keep only the versions that are still maintained and support most Python features.
+and we classified them into 3 categories depending on their integration with the python code.
+In \Cref{fig:pythonimplementations}, we describe the implementations that we kept and the version of each implementation and its category.
 
@@
 For the sake of reproducibility, each experiment runs within a Docker container.
-for each test we create a docker image.
+% for each test we create a docker image.
 \subsection{Metrics}
@@
 The primary downside is the incompatibility of some of these solutions, which causes us to make concessions when we need a generic answer.
+
+\input{table-envs.tex}

### 2022-08-07 chapters/optimization/python-dir/python_interperters.tex
add : dendogram

@@
 \section{placeholder}
-we performed a shapiro normality test for the first on the results, and almost all the p-values smaller than alpha=0.01\%
-therefore the distrubution is not normal - which is kinda true because if it was normal then we dont have any comparaison --
+we performed a shapiro normality test for the first on the results, and almost all the p-values smaller than alpha=0.01\% therefore the distrubution is not normal
 
@@
 \begin{figure}
-    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
+
     \label{table:bitops}
-    \begin{tabular}{lrrrrr}
+    \begin{tabular}{|lrrrrr|}
         \toprule
@@
     \end{tabular}
+    \vfill
+    \centering
+    \includegraphics[width=\linewidth]{imgs/bitopts_mean}
+    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
+\end{figure}
+
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\linewidth]{imgs/barplot_binarry_tree}
+%     \caption{energy behaviour of different python implementations for the binarry Tree benchmark}
+%     \label{fig:binarrytree}
+% \end{figure}
+
+\begin{figure}
 
 \end{figure}
-
 \begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{imgs/barplot_binarry_tree}
-    \caption{energy behaviour based on multiprocessing}
-    \label{fig:python_multiprocessing}
+    \includegraphics[width=\linewidth]{imgs/dendogram_interpreters}
+    \caption{Dendogram of the different implementations}
+    \label{fig:dendogram}
 \end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/bitopts_mean}
-    \caption{Mean consumption of different implementations of bit operations (Joule) }
-    \label{fig:bitops}
-\end{figure}
-
 \begin{figure}
@@
 
-for the other interpeters pypy is very promising especially when it comes to data maniupulation as one can see in the figure \ref{fig:bar_tommti_vector}
+for the other interpeters pypy is very promising especially when it comes to data maniupulation as one can see in the figure \ref{fig:bar_tommtei_vector}
 pypy is by far the best interperer when it comes to treating vectors.

### 2022-08-07 chapters/optimization/python-dir/python_interperters.tex
trim energy values

@@
 
+\subsection{Preliminary stduies}
+This part will be dedicated to analyse the initial behaviour of different python interpereters.
+First we started with a typical benchmark \em binarrytree. where we compare the energy consumption of the different implementations.
+
 \section{placeholder}
+
 we performed a shapiro normality test for the first on the results, and almost all the p-values smaller than alpha=0.01\% therefore the distrubution is not normal
@@
         \midrule
-        activepython & 676.980672 & 763.208040 & 651.783048 & 743.016425 & 728.828481 \\
-        cpython2     & 441.082298 & 435.886950 & 430.846171 & 415.247383 & 419.081447 \\
-        cpython3     & 595.209019 & 685.085315 & 563.839300 & 657.972734 & 655.560574 \\
-        cython       & 35.077408  & 182.688395 & 274.177830 & 34.868014  & 34.504778  \\
-        nuitka       & 33.260991  & 32.980380  & 33.256450  & 33.472770  & 33.030889  \\
-        numba2       & 9.102939   & 8.411176   & 9.460620   & 9.375920   & 9.755952   \\
-        numba3       & 9.566646   & 10.144212  & 9.219703   & 9.344613   & 9.665108   \\
-        pypy2        & 8.456867   & 7.844918   & 8.286278   & 8.138692   & 7.952999   \\
-        pypy3        & 7.552731   & 8.093884   & 8.108508   & 8.669448   & 8.623737   \\
-        shedskin     & 8.024198   & 8.070870   & 8.399917   & 8.126236   & 8.277546   \\
+        activepython & 676.980 & 763.208 & 651.783 & 743.016 & 728.828 \\
+        cpython2     & 441.082 & 435.886 & 430.846 & 415.247 & 419.081 \\
+        cpython3     & 595.209 & 685.085 & 563.839 & 657.972 & 655.560 \\
+        cython       & 35.077  & 182.688 & 274.177 & 34.868  & 34.504  \\
+        nuitka       & 33.260  & 32.980  & 33.256  & 33.472  & 33.030  \\
+        numba2       & 9.102   & 8.411   & 9.460   & 9.375   & 9.755   \\
+        numba3       & 9.566   & 10.144  & 9.219   & 9.344   & 9.665   \\
+        pypy2        & 8.456   & 7.844   & 8.286   & 8.138   & 7.952   \\
+        pypy3        & 7.552   & 8.093   & 8.108   & 8.669   & 8.623   \\
+        shedskin     & 8.024   & 8.070   & 8.399   & 8.126   & 8.277   \\
         \bottomrule

### 2022-08-08 chapters/optimization/python-dir/python.tex

@@
 
-\subsection{Python and machine learning }
+\subsection{Python \& Machine Learning}
 Machine learning is becoming an integral part of our daily lives, growing more potent and energy-hungry each year.
@@
 
-\paragraph{hardware }
+\paragraph{Hardware settings}
 Chifflot 8 from Grid 5000's Lille site was used for all of the trials.
@@
 
-\paragraph{software}
+\paragraph{Software settings}
+For the sake of reproducibility, each experiment is done within a Docker container using Jupyter lab.
+These tests are run on top of a minimal version of Debian-10 to increase the accuracy of the tests by eliminating any unnecessary processes.
 
-For the sake of reproducibility, each experiment is done within a Docker container using Jupyter lab .
-These tests are run on top of a minimal version of Debian-10 in order to increase the accuracy of the tests by elimanting any unnecessary processes.
+\subsubsection{Input Workload}
+\paragraph{Models}
+Several models were developed, however, only two were used in the final trials because they achieved 94 percent accuracy in an acceptable length of time.
+David Page's cifar10-fast and Woonhyuk Baek's torch skeleton are shown here.
 
-\subsubsection{Workload}
-\paragraph{Models}
-
-Several models were developed, however only two were used in the final trials because they achieved 94 percent accuracy in an acceptable length of time.
-David Page's cifar10-fast and Woonhyuk Baek's torchskeleton are shown here.
 \paragraph{Datasets}
-
 The CIFAR-10 dataset was the major source of data for the studies.
@@
 The experiments were run with several different CPU and GPU configurations:
-
 \begin{itemize}
-    \item With and without GPU
-    \item With and without hyperthreading
-    \item Different number of physical cores
+    \item with and without GPU,
+    \item with and without CPU hyper-threading,
+    \item different number of CPU physical cores.
 \end{itemize}
 
-\subsubsection{metrics}
-
+\subsubsection{Key Performance Metrics}
 We used Pytorch 1.10.0 to train those models and pyjoules to measure the energy consumption of the GPU and CPU.
@@
 
-\section{finding and results}
+\section{Results \& Findings}
 As the model's accuracy increases, so does the energy required for the next accuracy increment.
@@
 
-\paragraph{conclusion}
+\paragraph{Conclusion}
 We discovered that when a model's accuracy improves, so does the energy required for a subsequent accuracy gain.

### 2022-08-08 chapters/programming_languages/rpc/rpc.tex

@@
-\subsection*{introduction}
+\subsection{Introduction}
 With the emergence of cloud technologies, many protocols wanted to take the lead.
@@
 
-\subsection{Experimental protocol}
-\subsubsection{measurement context}
+\subsection{Experimental Protocol}
+\subsubsection{Measurement Context}
 \paragraph{Hardware settings}
@@
 
-\subsubsection{metrics}
+\subsubsection{Key Performance Metrics}
 \paragraph{Energy measurements}
@@
 
-\subsubsection{workload}
+\subsubsection{Workload}
 The purpose of the experiment is to analyse the behaviour of different GRPC implementations. therefore we have two kind of worlkoads 
@@
 
-\subsection{Results and findings}
+\subsection{Results \& Findings}
 [\textsc{RQ}~1:] How do RPC implementations react to the size of the data ?
@@
 
-\subsubsection{Stress free mode}
+\subsubsection{Stress Free Mode}
 In this mode, the compiled languages tend to consume less resources (average power).
@@
 
-\subsubsection{Escalation mode}
+\subsubsection{Escalation Mode}
 In this mode, the behaviour of the server depends on the payload. We observe three behaviours:
@@
 
-\subsection{Threads to validity}
+\subsection{Threads to Validity}
 

### 2022-08-08 chapters/benchmarking/draft.tex

--- 2022-07-19 chapters/benchmarking/draft.tex
@@
 The ability to give not only the tools needed to replicate the tests but also to integrate more candidates, workloads, or important key performance indicators
-
 This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
@@
 Finally, Section~\ref{sec:bench_extension} presents tools for introducing new candidates, workloads, or other key performance indicators.
-
 Regarding the "representativeness" of the benchmarks, we will elaborate in the section entitled Perspectives.

### 2022-08-25 chapters/optimization/java-dir/java.tex
add : bitops analysis

@@
 \noindent\textbf{Software Settings.}
-For the sake of reproducibility, each experiment runs within a Docker container based on \textsf{SDKMAN!}\footnote{\url{https://sdkman.io}} image and Alpine docker.\footnote{\url{https://github.com/alpinelinux/docker-alpine}}
+For the sake of reproducibility, each experiment runs within a Docker container based on \textsf{SDKMAN}\footnote{\url{https://sdkman.io}} image and Alpine docker.\footnote{\url{https://github.com/alpinelinux/docker-alpine}}
 
@@
 \noindent\textbf{Java Virtual Machines Candidates.}
-We considered a set of $52$ JVM distributions taken from $8$ different providers/packagers mostly obtained from \textsf{SDKMAN!}, as listed in \Cref{tab:JVM}.
-Depending on providers, either all the versions, majors, or LTS are made available by \textsf{SDKMAN!}.
+We considered a set of $52$ JVM distributions taken from $8$ different providers/packagers mostly obtained from \textsf{SDKMAN}, as listed in \Cref{tab:JVM}.
+Depending on providers, either all the versions, majors, or LTS are made available by \textsf{SDKMAN}.
 
@@
 The package is available in the GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jvm-comparaison}}
-To add extra \textbf{jvm candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the jvm to be used.\note{The JVM must be supported by \textsf{SDKMAN!}}
+To add extra \textbf{jvm candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the jvm to be used.
 For the \textbf{input workload},the benchmarks should be provided in the \texttt{benchmarks} directory.

### 2022-08-25 chapters/optimization/python-dir/python_interperters.tex
refactoring
check
add : bitops analysis

@@
 This part will be dedicated to analyse the initial behaviour of different python interpereters.
-First we started with a typical benchmark \em binarrytree. where we compare the energy consumption of the different implementations.
+First we started with a typical benchmark binarrytree. where we compare the energy consumption of the different implementations.
+
+\subsection{python and binarry operations }
+The goal of this study is to look at how different Python implementations work while treating different binary operations .
+
+\begin{figure}[!hbt]
+    \begin{tabular}{|lrrrrr|c|}
+        \toprule
+        benchmark    & Addition & Right Rotation & Left Rotation & Or      & XOR     & Average \\
+        \midrule
+        activepython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76  \\
+        cpython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43  \\
+        cpython3     & 595.209  & 685.085        & 563.839       & 657.972 & 655.560 & 531.53  \\
+        \hline
+        cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26  \\
+        nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2    \\
+        \hline
+        numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & 9.22    \\
+        numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & 9.59    \\
+        pypy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & 8.13    \\
+        pypy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & 8.21    \\
+        shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & 8.18    \\
+        \bottomrule
+    \end{tabular}
+    % should i add the interpretation 
+    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
+    \label{table:bitops}
+\end{figure}
+
+Table~ provides a detailed report on the energy consumption of different implementations of the code.
+
+Except for \emph{Cython}, each implementation treats the binary operation the same. As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
+
+However, when it comes to Cython, the rotation operations tend to consume 5 times more energy than the other operations, and left rotations consume even more than the right ones.
+
+The reason behind such behavior is the fact that Cython does not have a native implementation for these two operations, so it replaces them with more complex operations, which leads to an extra time of 6s for the right rotation and 10s for the left, compared to 1.5 for the other operations.
+-
+
+Therefore, it is safe to summarize all the operations into one that we call binary.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/bitopts_mean}
+    \caption{Summary of the binary operations for different python implementations }
+    \label{figure:bitops}
+\end{figure}
+Figure \ref{figure:bitops}  shows the average energy consumption of different implementations for the binary operations.
+
+One can notice 3 clusters.
+
+The first, which is the interpreted one, uses approximately 500 joules per operation.
+The second is a non-optimized compilation that includes Cython and Nuitka and uses 10 times less energy than the first.
+And the final one, which has been fully optimized, uses 50 percent less energy.
+This category includes jit libraries, also known as numba, Python implementations that include JIT, also known as pypy, and shedskin, which transpile the source code into C and then use a C/C++ compiler to generate the binary, as opposed to cython and nuitka, which directly compile the python code.
+
+To conclude this. JIT has a huge impact when it comes to binary functions, and the wrong choice of the implementation can lead to up to a 100x increase in energy consumption.
 
 \section{placeholder}
-
+introdution to the stufy
 we performed a shapiro normality test for the first on the results, and almost all the p-values smaller than alpha=0.01\% therefore the distrubution is not normal
@@
 
-\begin{figure}
-
-    \label{table:bitops}
-    \begin{tabular}{|lrrrrr|}
-        \toprule
-        benchmark    & A       & GG      & LL      & Or      & XOR     \\
-        \midrule
-        activepython & 676.980 & 763.208 & 651.783 & 743.016 & 728.828 \\
-        cpython2     & 441.082 & 435.886 & 430.846 & 415.247 & 419.081 \\
-        cpython3     & 595.209 & 685.085 & 563.839 & 657.972 & 655.560 \\
-        cython       & 35.077  & 182.688 & 274.177 & 34.868  & 34.504  \\
-        nuitka       & 33.260  & 32.980  & 33.256  & 33.472  & 33.030  \\
-        numba2       & 9.102   & 8.411   & 9.460   & 9.375   & 9.755   \\
-        numba3       & 9.566   & 10.144  & 9.219   & 9.344   & 9.665   \\
-        pypy2        & 8.456   & 7.844   & 8.286   & 8.138   & 7.952   \\
-        pypy3        & 7.552   & 8.093   & 8.108   & 8.669   & 8.623   \\
-        shedskin     & 8.024   & 8.070   & 8.399   & 8.126   & 8.277   \\
-        \bottomrule
-    \end{tabular}
-    \vfill
-    \centering
-    \includegraphics[width=\linewidth]{imgs/bitopts_mean}
-    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
-\end{figure}
-
 % \begin{figure}
@@
 
-\begin{figure}
-
-\end{figure}
+\input{table-envs.tex}
 \begin{figure}
@@
 
+%   this is after , when we come to the conclusion
 As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
@@
 The primary downside is the incompatibility of some of these solutions, which causes us to make concessions when we need a generic answer.
-
-\input{table-envs.tex}

### 2022-08-29 chapters/optimization/python-dir/python_interperters.tex
add : python bitops

@@
 The reason of separation between data collection and energy calculation is to minimise any interference with the test so our sensor is a light c program running inside a docker container.
+
 \subsection{tests preparation}
+\paragraph{workload}
+To test our implementations, we employed the Tommti microbenchmarks suite~\footnote{\url{http://www.tommti-systems.de/main-Dateien/reviews/languages/benchmarks.html}}.
+A set of 13 microbenchmarks that examine common language features such as arithmetic operations, data structures and input/output manipulations, among others.
+In addition to these tests ,we implemented some binary operations to examine the behavior of the previous implementations when it comes to low-level operations that works only with registries.
 
@@
 \section{Results and finding}
-
-\subsection{Preliminary stduies}
-This part will be dedicated to analyse the initial behaviour of different python interpereters.
-First we started with a typical benchmark binarrytree. where we compare the energy consumption of the different implementations.
-
-\subsection{python and binarry operations }
+This section will be dedicated to the results of the experiments and the finding of the best implementation.
+\paragraph{python and binarry operations }
 The goal of this study is to look at how different Python implementations work while treating different binary operations .
@@
 
-Table~ provides a detailed report on the energy consumption of different implementations of the code.
-
-Except for \emph{Cython}, each implementation treats the binary operation the same. As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
+Table~~\ref{table:bitops} provides a detailed report on the energy consumption of different implementations of the code.
+Besides \emph{Cython}, each implementation treats the binary operation the same. As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
 
@@
 This category includes jit libraries, also known as numba, Python implementations that include JIT, also known as pypy, and shedskin, which transpile the source code into C and then use a C/C++ compiler to generate the binary, as opposed to cython and nuitka, which directly compile the python code.
+To conclude this.
+JIT has a huge impact when it comes to binary operations, and the wrong choice of the implementation can lead to up to a 100x increase in energy consumption.
+In the other hand,Different binary operations exhibit the same behaviour when it comes to the energy consumption for all the implementations except Cython.
 
-To conclude this. JIT has a huge impact when it comes to binary functions, and the wrong choice of the implementation can lead to up to a 100x increase in energy consumption.
+\section{introduction to the study}
+Since we are in a comparative study, we first performed a shapiro normality test on the results, and almost all the p-values are  smaller than alpha=0.01\% therefore the distrubution is not normal.
+To make a comparaison we used Manwithyu's tests.
+Table \ref{table:manwithyu} .
 
-\section{placeholder}
-introdution to the stufy
-we performed a shapiro normality test for the first on the results, and almost all the p-values smaller than alpha=0.01\% therefore the distrubution is not normal
-
+\input{table-envs-based-cpython.tex}
 for arrays and vectors  graalpython took so much time that we decided to remove it
@@
 
-so the plan is to prove that there is a statistical difference
 then we use the multiparameter optimization , howerver we stop in the phase where we calculate the score
@@
 \end{figure}
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/tommti_compare__pypy2_pypy3}
-    \caption{green factor of pypy }
-    \label{fig:pypy2vspypy3}
-\end{figure}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/tommti_compare__cpython3_numba3_pypy3}
-    \caption{comparaison of pypy vs python vs numba }
-    \label{fig:p3}
-\end{figure}
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\linewidth]{imgs/tommti_compare__pypy2_pypy3}
+%     \caption{green factor of pypy }
+%     \label{fig:pypy2vspypy3}
+% \end{figure}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/tommti_compare__cpython2_pypy2_numba2}
-    \caption{comparaison of pypy vs python vs numba }
-    \label{fig:p2}
-\end{figure}
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\linewidth]{imgs/tommti_compare__cpython3_numba3_pypy3}
+%     \caption{comparaison of pypy vs python vs numba }
+%     \label{fig:p3}
+% \end{figure}
+
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\linewidth]{imgs/tommti_compare__cpython2_pypy2_numba2}
+%     \caption{comparaison of pypy vs python vs numba }
+%     \label{fig:p2}
+% \end{figure}
 

### 2022-09-01 chapters/optimization/java-dir/java.tex
fix : notes that were hilighlted by Lionel

@@
 \section{Introduction}
-As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.%ROMAIN: Add source
-Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language.%ROMAIN: Add source
+As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners. %ROMAIN: Add source
+Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language. %ROMAIN:  Add source
 In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performance, Java applications can be at the same time one of the most energy-efficient or hungry solutions.
 As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
-Therefore, we wanted to dig deeper into this aspect of Java and study its runtime.
+in this chapter, we want to dig deeper into this aspect of Java and study its runtime.
 This chapter thus focuses on the impact of the runtime of Java applications on energy consumption.
@@
 
-\subsection{definition of the JVM}
+% \subsection{definition of the JVM}
 % TODO : ADD definition here 
@@
 \section{Experimental Protocol}\label{sec:javaprotocol}
-To investigate the effect that could have the JVM distribution choice and/or parameters on software energy consumption, we conducted a wide set of experiments on a cluster of machines and used several established Java benchmarks and JVM configurations.
+To investigate the effect that can have the JVM distribution choice and/or parameters on software energy consumption, we conducted a wide set of experiments on a cluster of machines and used several established Java benchmarks and JVM configurations.
 
@@
 We ran our experiments across $12$ Java benchmarks we picked from \textsf{OpenBenchmarking.org}.\footnote{\url{https://openbenchmarking.org}}
-This includes $5$ acknowledged benchmarks from the \textsc{Dacapo} benchmark suite v.\,9.12~\cite{DaCapo:paper}, namely \textsf{Avrora}, \textsf{H2}, \textsf{Lusearch}, \textsf{Sunflow} and \textsf{PMD}, that have been widely used in previous studies and proven to be accurate for memory management and computer architecture communities~\cite{DBLP:conf/wosp/LengauerBMW17,DBLP:conf/oopsla/KaliberaMJV12}.
+This includes $5$ acknowledged benchmarks from the \textsc{Dacapo} benchmark suite v.\,9.12~\cite{DaCapo:paper}, namely \textsf{Avrora}, \textsf{H2}, \textsf{Lusearch}, \textsf{Sunflow} and \textsf{PMD}, that have been widely used in previous studies and proven to be accurate for memory management and computer architecture~\cite{DBLP:conf/wosp/LengauerBMW17,DBLP:conf/oopsla/KaliberaMJV12}.
 It consists of open-source and real-world applications with non-trivial memory loads.
 % 
-Then, we also considered $7$ additional benchmarks from the \textsc{Renaissance} benchmark suite~\cite{renaissance,DBLP:conf/pldi/ProkopecRLD0SBZ19}, namely \textsf{ALS}, \textsf{Dotty}, \textsf{Fj-kmeans}, \textsf{Neo4j}, \textsf{Philosophers}, \textsf{Reaction} and \textsf{Scrabble}, which offers a diversified set of benchmarks aimed at testing JIT, GC, profilers, analyzers, and other tools.
+Then, we also considered $7$ additional benchmarks from the \textsc{Renaissance} benchmark suite~\cite{renaissance,DBLP:conf/pldi/ProkopecRLD0SBZ19}, namely \textsf{ALS}, \textsf{Dotty}, \textsf{Fj-kmeans}, \textsf{Neo4j}, \textsf{Philosophers}, \textsf{Reaction} and \textsf{Scrabble}, which offer a diversified set of benchmarks aimed at testing JIT, GC, profilers, analyzers, and other tools.
 The benchmarks we picked from both suites exercise a broad range of programming paradigms, including concurrent, parallel, functional, and object-oriented programming.
-\Cref{tab:benchmarks} summarizes the selected benchmarks with a short description.
+\Cref{tab:benchmarks} summarizes the selected benchmarks with a short description. Meanwhile, Figure~\ref{fig:jvm-benchmarks-classification} highlights the scope of each benchmark from the test suite.
 
@@
 \end{table*}
-
-\Cref{fig:jvm-benchmarks-classification} highlights the scope of each benchmark from the test suite.
 
@@
 Ultimately, 3 clusters of JVMs that encompass JVMs with similar energy consumption can be seen through \Cref{fig:JVMs}: \textsc{J9}, the \textsc{HotSpot} and its variants, and \textsc{GraalVM}.
-Additional detailed figures to illustrate the evolution of energy consumption per benchmark/JVM are made available from the online repository.\footnote{\url{https://anonymous.4open.science/r/jvm-comparaison-213E/Readme.md}}
+Additional detailed figures to illustrate the evolution of energy consumption per benchmark/JVM are made available from the online repository.\footnote{\url{https://github.com/chakib-belgaid/jvm-comparaison}}
 
 Then, \Cref{fig:hotspot8} depicts the evolution of the energy consumption of the $12$ benchmarks, when executed on the \textsc{HotSpot} JVM.
-\Cref{fig:hotspot8} reports on the energy consumption variation of individual benchmarks, using to \textsc{HotSpot-8}\ as the baseline.
+\Cref{fig:hotspot8} reports on the energy consumption variation of individual benchmarks, using \textsc{HotSpot-8}\ as the baseline.
 Our results show that the JVM version can severely impact the energy consumption of the application.
@@
 It is worth noticing that the energy consumption of some benchmarks, such as \textsf{Reactors}, exhibit large variations across JVM versions due to experimental features and changes that are not always kept when releasing LTS versions (version 11 here).
-For example, the introduction of \texttt{VarHandle} to allow low-level access to the memory order modes available in JDK 9 and work along \texttt{Unsafe Class} was removed from JVM 11.\footnote{\url{https://blogs.oracle.com/javamagazine/the-unsafe-class-unsafe-at-any-speed}}
+For example, the introduction of \texttt{VarHandle} to allow low-level access to the memory order modes available in JDK 9~\footnote{\url{https://gee.cs.oswego.edu/dl/html/j9mm.html}} and work along \texttt{Unsafe Class} was removed from JVM 11.\footnote{\url{https://blogs.oracle.com/javamagazine/the-unsafe-class-unsafe-at-any-speed}}
 
@@
 Globally, the results showed that the average power when using \textsc{GraalVM}, \textsc{HotSpot}, and \textsc{OpenJ9} is often equivalent and stable over time.
-This means that the energy efficiency observed for some JVM distributions with Job-oriented applications is mainly related to shorter execution times, which incidentally results in energy savings.
+This means that the energy efficiency observed for some JVM distributions with job-oriented applications is mainly related to shorter execution times, which incidentally results in energy savings.
 Nonetheless, we can highlight two interesting observations for two benchmarks whose behaviors differ from others.
@@
 \textsc{HotSpot} was thus the best choice in the long term, which explains why it is always necessary to consider a warm-up phase and wait for the JIT to be triggered before evaluating the effect of the JVM or the performance of an application.
-This is exactly what we did in our experiments, and why \textsc{HotSpot} was more energy efficient than \textsc{GraalVM} in \Cref{fig:all_benchs}, thus ignoring the warm-up phase would have been misleading.
+This is precisely what we did in our experiments, and it is why \textsc{HotSpo}t was more energy efficient than \textsc{GraalVM} in \Cref{fig:all_benchs}; therefore, ignoring the warm-up phase would have been misleading.
 
@@
 We also varied the minimum count method calls before a JIT compilation occurs (\textsf{10}, \textsf{50}, \textsf{100}), and the number of JIT instances threads (from \textsf{1} to \textsf{7}).
-For \textsc{Hotspot-15}, we conducted experiments while disabling the tiered complication (that generates compiled versions of methods that collect profiling information about themselves), and we also varied the JIT maximum compilation level from \textsf{0} to \textsf{4}, we also tried out \textsc{HotSpot} with a basic \textsc{GraalVM} JIT.
+For \textsc{Hotspot-15}, we conducted experiments while disabling the tiered compilation (that generates compiled versions of methods that collect profiling information about themselves), and we also varied the JIT maximum compilation level from \textsf{0} to \textsf{4}, we also tried out \textsc{HotSpot} with a basic \textsc{GraalVM} JIT.
 We note that level 0 of JIT compilation only uses the interpreter, with no real JIT compilation.
@@
 However, other GC can cause better or worse energy consumption than \textsf{Gencon} depending on workloads.
-Using the \textsf{Metronome} GC consumes ~35\% less energy for the \textsf{ALS} benchmark and ~17\% less energy for the \textsf{Sunflow} benchmark, but it also consumes twice energy for the \textsf{Neo4j} benchmark and 28\% more energy for \textsf{Reactors}.
+Using the \textsf{Metronome} GC consumes ~35\% less energy for the \textsf{ALS} benchmark and ~17\% less energy for the \textsf{Sunflow} benchmark, but it also consumes 100\% more energy for the \textsf{Neo4j} benchmark and 28\% more energy for \textsf{Reactors}.
 The reason is that \textsf{Metronome} occurs in small preemptible steps to reduce the GC cycles composed of many GC quanta.
@@
     To answer \textbf{RQ\,2}, we conclude that users should be careful while choosing and configuring the garbage collector as substantial energy enhancements can be recorded from one configuration to another.
-    The default GC consumed more energy than other strategies in most situations.
+    The default GC consumes more energy than other strategies in most situations.
     % choosing the most convenient GC strategy in thus very important to achieve a better software energy efficiency. 
@@
 \section{Tools and contributions}
-\Cref{table:j-referral} illustrates an example of the final report of \textsf{J-Referral}.
+\textsf{J-Referral} is an open-source tool designed to assist developers and practitioners in selecting the most energy-efficient JVM configuration for their software..
+\Cref{table:j-referral} illustrates an example of the final report returned by \textsf{J-Referral}.
 The tool was tested for 2 Java projects: \textsf{Zip4J}\footnote{\url{https://github.com/srikanth-lingala/zip4j}} and \textsf{K-nucleotide}.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/performance/knucleotide.html}}
@@
 \section{Conclusion}
-The results of our investigations showed that many JVMs share energy efficiencies and can be grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
+The results of our investigations show that many JVMs share energy efficiencies and can be grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
 The 3 selected JVM classes can however report a different energy efficiency for different software and/or workloads, sometimes by a large margin.

### 2022-09-01 chapters/optimization/python-dir/python.tex
fix : notes that were hilighlted by Lionel

@@
 
-As shown in \Cref{fig:clbg}, one can observe that, for most of the applications taken for the \emph{Computer Language Benchmark Game} (CLBG), Python takes more time to execute---the only case that he was not the worst one was in the benchmark \textsf{regx-redux} where he beat Go---and in some cases the gap was huge, such as in \textsf{n-body} where Python took around $100$ times more than C++.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}
+As shown in \Cref{fig:clbg}, one can observe that, for most of the applications taken for the \emph{Computer Language Benchmark Game} (CLBG), Python takes more time to execute---the only case that he was not the worst one was in the benchmark \textsf{regx-redux} where he beats Go---and in some cases the gap was huge, such as in \textsf{n-body} where Python took around $100$ times more than C++.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}
 
@@
 To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
-\Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers.\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}.
-In this survey, multiple answers were accepted.
+\Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}.
 57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration.
+\note{The options in this survey were not mutually exclusive. As a result, the total of the percentages is greater than $100\%$.}
 
@@
 In this chapter, we investigate the energy footprint of Python in its most popular domains of adoption.
-We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016-1} did in~\cite{hasan_energy_2016-1}.
+We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{s} did in~\cite{hasan_energy_2016-1}.
 Then, we measure the energy consumption of several Python implementations to propose a non-intrusive technique to improve energy efficiency.
@@
 
-In this chapter, we report on 4 case studies that intend to answer these research questions.
+To answer these questions, we report on 4 case studies that intend to answer these research questions.
 First, we study the energy behavior of Python in two application contexts: web applications (cf. \Cref{sec:webdev}) and machine learning.
@@
 \section{Green Web Development}\label{sec:webdev}
+% TODO : ADD introduction to the orms 
 We start by studying the impact of the \emph{Object Relational Mapping} (ORM)~\cite{o2008object} on energy consumption.
@@
 \paragraph{Conclusion}
-This study demonstrated that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
+This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
 
@@
 \Cref{fig:python_multiprocessing} compares the behaviour of python programs when we try to introduce the parallelism.
-As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY
+As we know python is a single threaded program thanks to ghe GLS (Global Lock System) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances of the processor % TODO: TO VERIFY
 
@@
 
-as we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
+As we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
 where different sub process have to compete for the CPU Ressources.
@@
 Several models were developed, however, only two were used in the final trials because they achieved 94 percent accuracy in an acceptable length of time.
-David Page's cifar10-fast and Woonhyuk Baek's torch skeleton are shown here.
+David Page's cifar10-fast~\footnote{\url{https://github.com/davidcpage/cifar10-fast}} and Woonhyuk Baek's torch skeleton~\footnote{\url{https://github.com/wbaek/torchskeleton}} are shown here.
 

### 2022-09-03 chapters/optimization/python-dir/python_interperters.tex
fix : notes that were hilighlted by Lionel
update:  python - part 1
python interpreters

@@
 %% basically i what i want to say is that we we care mostly about energy, because there are studies that were made on the performance, plus the DRAM is not that interested even within the heavy memory  benchmakrs
-% TODO : ADD DRAM graphs for vectors and matrix multiply
 
@@
 \section{Results and finding}
-This section will be dedicated to the results of the experiments and the finding of the best implementation.
-\paragraph{python and binarry operations }
-The goal of this study is to look at how different Python implementations work while treating different binary operations .
 
-\begin{figure}[!hbt]
-    \begin{tabular}{|lrrrrr|c|}
-        \toprule
-        benchmark    & Addition & Right Rotation & Left Rotation & Or      & XOR     & Average \\
-        \midrule
-        activepython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76  \\
-        cpython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43  \\
-        cpython3     & 595.209  & 685.085        & 563.839       & 657.972 & 655.560 & 531.53  \\
-        \hline
-        cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26  \\
-        nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2    \\
-        \hline
-        numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & 9.22    \\
-        numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & 9.59    \\
-        pypy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & 8.13    \\
-        pypy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & 8.21    \\
-        shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & 8.18    \\
-        \bottomrule
-    \end{tabular}
-    % should i add the interpretation 
-    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
-    \label{table:bitops}
+This section will be dedicated to the results of the experiments and a statistical analysis of these results.
+
+Because we are conducting a comparison study, we first did a Shapiro-Wilk~\cite{shapiro1968comparative} test to see if the data follows a normal distribution.
+Despite, the fact some implementations had a p-value bigger than a=0.05 (such as cpython and active python),other implementaions like pypy and numba presented a p-values smaller than 0.01 which leads to reject the hypothesis H0 that all the distributions follow a normal distribution.
+Therefore,to compare the different implementations, we performed the non-parametric Mann-Whitney U test~\cite{zimmerman1987comparative}, with CPython2 and CPython3 as base references.
+
+The Mann-Whitney U test results for each implementation are shown in Table \ref{table:manwithyu}.
+The first column displays the implementation's average energy usage, while the second provides the p-value of the test when compared to the reference Python implementation.
+If the p-value is less than 0.05, the test result is significant.
+This implies that the given implementation's energy consumption differs significantly from the reference implementation's.As one can notice, most of the implemnetations differe significantly from the reference implementation.
+\input{table-envs-based-cpython.tex}
+We want to study the position of these implementations toward the references. To do so, we proceed with an agglomerative hierarchical cluster analysis (HCA)~\cite{johnson1967hierarchical}.
+The HCA is a method that allows grouping the different implementations based on their energy consumption.
+This technique is a bottom-up approach that starts with each implementation as a single cluster and then merges the two most similar clusters until all the clusters are merged into a single cluster.
+The similarity between two clusters is measured by the maximum distance between their furthest points.
+The distance between two points is calculated as the Euclidean distance between the two points.
+Figure \ref{fig:dendogram} displays the dendogram produced by the HCA
+\begin{figure}
+      \centering
+      \includegraphics[width=\linewidth]{imgs/dendogram_interpreters}
+      \caption{Dendogram of the different implementations}
+      \label{fig:dendogram}
 \end{figure}
 
-Table~~\ref{table:bitops} provides a detailed report on the energy consumption of different implementations of the code.
-Besides \emph{Cython}, each implementation treats the binary operation the same. As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
+Except GraalPython, one can notice 3 main clusters.
+The first cluster contains the reference implementations (CPython2 and CPython3) and the implementations that are based on interpreters (Ironpython,Intelpython, ActivePython, Jython, and Nuitka).
+Moreover, each implementation is the closest to its reference Python version.
+The interpreters that are based on other virtual machines, such as Jython with JVM and IronPython with.NET, behave slightly differently from the reference implementation.
+As for the micropython, It is the furthest from the references due to his behavior toward exceptions.
+The second cluster contains the interpreter implementations that are based on a JIT compiler (Pypy2 and Pypy3).
+Closer to them we find shedskin, which is a transaltor that converts the Python code into C++ code to be later compiled into binary code.
+The results table~\ref{table:manwithyu} shows that this cluster is by far the greenest one.
+While numba2 and numba3 are also JIT libraries, they differ from PyPy due to their manual optimisation.
+Unlike pypy, which will decide by itself which part of Python to optimize.
+In Numba, the developer must target the portion of the code that will be passed to JIT, while the rest of the code will be executed by the regural interpreter (aka CPython).
 
-However, when it comes to Cython, the rotation operations tend to consume 5 times more energy than the other operations, and left rotations consume even more than the right ones.
+\paragraph{Remark :}
+Because graalPython was still in its early stages, some techniques required an abnormally lengthy time, affecting the clustering Algorithm.
 
-The reason behind such behavior is the fact that Cython does not have a native implementation for these two operations, so it replaces them with more complex operations, which leads to an extra time of 6s for the right rotation and 10s for the left, compared to 1.5 for the other operations.
--
+To help practiouners to choose the best implementation for their use case, we have created a chart that describes the behavior of each implementation toward a specific aspect of programming. 
+Figure~\ref{fig:tommi_all} present a radar plot for all the implementations that have been used in this experemation.
+\begin{figure}[!htb]
+      \centering
+      \includegraphics[width=\linewidth]{imgs/alltomti_performance}
+      \caption{Different interpererts optimisation score }
+      \label{fig:tommi_all}
+\end{figure}
 
-Therefore, it is safe to summarize all the operations into one that we call binary.
+This graph depicts each implementation's energy score in relation to each benchmark.
+The lower the energy consumption, the higher the score.
+We adopted a logarithmic scale to allow the reader to compare numerous implementations due to the large differences between them.
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/bitopts_mean}
-    \caption{Summary of the binary operations for different python implementations }
-    \label{figure:bitops}
-\end{figure}
-Figure \ref{figure:bitops}  shows the average energy consumption of different implementations for the binary operations.
+As one can notice in figure~\ref{fig:tommi_all}, there is no evolution between CPython2 and CPython3.
+Moreover, Intelpython, and Active Python all behave similarly to the reference implementation .
+Therefore, one can conclude that the work done on those interpreters is primarily to improve a specific purpose and not the core interpreter.
+Active Python states that their version is focused on security and prepackaged libraries, which explains why it is slower than other versions due to the addition of more security reinforcements\footnote{\url{https://www.activestate.com/solutions/why-activestate/}} .
+As for IntelPython,it is designed for machine learning. Unfortunately, the Tommti benchmark is geared for general-purpose programming.
+Despite the fact that Nuitka is a compiler, there was no discernible difference in energy consumption. It was even more similar to the reference Python implementation than other interpreters.
+However, if we look at the nuitka techniques, we can see that they simply encapsulate the Python code with an interpreter into a single executable.
+However, shedskin exhebit has the best energy consumption pattern when it comes to arithmetic operations.
+One can conclude it is due to the fact of the native type of the variables.
+Unlike in the JIT, where they are treated as objects in the beginning.
 
-One can notice 3 clusters.
+for the other VM-based interpereters.
+Jython and Ipy lacked in terms of energy optimisation, which was kind of expected since they were at the beginning of the development stage and the main purpose of such implementation is to link the bytecode generated by Jython and IronPython with their respective virtual machines.
 
-The first, which is the interpreted one, uses approximately 500 joules per operation.
-The second is a non-optimized compilation that includes Cython and Nuitka and uses 10 times less energy than the first.
-And the final one, which has been fully optimized, uses 50 percent less energy.
-This category includes jit libraries, also known as numba, Python implementations that include JIT, also known as pypy, and shedskin, which transpile the source code into C and then use a C/C++ compiler to generate the binary, as opposed to cython and nuitka, which directly compile the python code.
-To conclude this.
-JIT has a huge impact when it comes to binary operations, and the wrong choice of the implementation can lead to up to a 100x increase in energy consumption.
-In the other hand,Different binary operations exhibit the same behaviour when it comes to the energy consumption for all the implementations except Cython.
+Unlike the previous interpeters, GraalPython exhibits a certain promise when it comes to complex algorithms (nested loops).
+Finally,micro python is dedicated to embedded systems, so claiming it is a powerful CPU machine will be misleading.
 
-\section{introduction to the study}
-Since we are in a comparative study, we first performed a shapiro normality test on the results, and almost all the p-values are  smaller than alpha=0.01\% therefore the distrubution is not normal.
-To make a comparaison we used Manwithyu's tests.
-Table \ref{table:manwithyu} .
+When it comes to  the input outputs operations , 
+when it came to input/output manupulation. Except for Jython,most of the interpreters behaved similarly . Because these operations are more dependent on the hardware than the interpreter.
 
-\input{table-envs-based-cpython.tex}
-for arrays and vectors  graalpython took so much time that we decided to remove it
+% However, Jython is the only implementation that has a significant difference in energy consumption.
+This is due to the fact that Jython is based on the JVM, which is a virtual machine that is designed to run Java bytecode.
+Therefore, the energy consumption of the input/output operations is not the same as the other implementations.
 
-the startup cost for ironpython was too high
-for graalpython and ipy they couldnt handle big numbers hence the overflow  when it comes to execution time
-but since we measure the energy outside the values of the energy aren't impacted
+% \begin{figure}
+%       \centering
+%       \includegraphics[width=\linewidth]{imgs/bar_plot_tommtivector}
+%       \caption{Average energy consumption of different python implementaions for vector treatement}
+%       \label{fig:bar_tommti_vector}
+%   \end{figure}
+% as usual there is  correlaction between DRAM and CPU so no need to classify based on DRAM + the DRAM consumes less 10\% of the  energy compared to the CPU
 
-as usual there is  correlaction between DRAM and CPU so no need to classify based on DRAM + the DRAM consumes less 10\% of the  energy compared to the CPU
-
-then we use the multiparameter optimization , howerver we stop in the phase where we calculate the score
-because we dont know the weight of each parameter ( of tommti microbenchmark) and instead of doing the sum with the coefissients we use the radar plot to let the reader decide which one is the most suitable for his usage.
+% then we use the multiparameter optimization , howerver we stop in the phase where we calculate the score
+% because we dont know the weight of each parameter ( of tommti microbenchmark) and instead of doing the sum with the coefissients we use the radar plot to let the reader decide which one is the most suitable for his usage.
 howerver since the differences are gigantics we change the scale to logarithmic to be easier for the eye to read
@@
 
-\input{table-envs.tex}
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/dendogram_interpreters}
-    \caption{Dendogram of the different implementations}
-    \label{fig:dendogram}
-\end{figure}
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/alltomti_performance}
-    \caption{different interpererts optimisation }
-    \label{fig:tommi_all}
-\end{figure}
+% \input{table-envs.tex}
+
+%% distance euclidean 
+%% linkage complete Voor Hees Algorithm.
+
+%%%%%%%%%%%%%%%%%
 
@@
 %   this is after , when we come to the conclusion
-As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
-Intelpython and active python both folows the same behaviour. one can conclude that the work that have been on those interpreters is mainly
-to improve a specific purpose, Active python claims that their version is focused on secuirty which explains the lack of some performances due to the introduction of more reonforcment
-.Morver Intel published their version of python as a dedicated for machine learning. Unfortunately the tommti benchmark is a set that focuses on the general purpose programming
-which does not reflect the performance of the machine learning.
-Another aspect of a such behaviour might be due to the fact of the processors that were used in the testbed.
-and it might change in the future if we include the GPU part for the machine learning benchmarks
-% reference the work of the intern
-As for nuitka. there were no optimization in the energy consumpotion despite the fact that is it a complier.
-However if we dig through the nuitka mechanisms. they basically embed the python code with an interpreter.
-Unllike nuitka shedskin exhebit a the best energy consumption pattern when it comes to the arithmetic operartions.
-One can conclude it is due to the fact of the native type of the variables.
-unlike the interpreters where they are treated as object in the begining.
-
-for the other interpeters pypy is very promising especially when it comes to data maniupulation as one can see in the figure \ref{fig:bar_tommtei_vector}
-pypy is by far the best interperer when it comes to treating vectors.
-numba2 introduced the JIT but wasn't as promising as numba3.
-
-for the other vm based interpereters.
-jython and ipy lacked in term of energy optimisation which was kinda expected since they were in their the begining of the stage and the main puprose of such implementation is to link the bytecode generated by jython and ironpython with their respectives virtual machin.
-
-Unlike the the previous interpeters. graal exhebits a certin promises when it comes to complex algorithms - nested loops -
-micro python is dedicated to embeded systems so lunching it is powerful cpu machines will be misleading.
-
-Most of the interperters had the same behaviour when it comes to the input outputs. except for jython which was kinda of a anomaly probably due to the lack of the optimizations.
 % pypy handles exceptions very well
 
+\subsection{python and binarry operations }
+The second part of this study, is to look at how different Python implementations work while treating different binary operations .
+
+\begin{figure}[!hbt]
+    \begin{tabular}{|lrrrrr|c|}
+        \toprule
+        benchmark    & Addition & Right Rotation & Left Rotation & Or      & XOR     & Average \\
+        \midrule
+        activepython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76  \\
+        cpython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43  \\
+        cpython3     & 595.209  & 685.085        & 563.839       & 657.972 & 655.560 & 531.53  \\
+        \hline
+        cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26  \\
+        nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2    \\
+        \hline
+        numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & 9.22    \\
+        numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & 9.59    \\
+        pypy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & 8.13    \\
+        pypy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & 8.21    \\
+        shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & 8.18    \\
+        \bottomrule
+    \end{tabular}
+    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
+    \label{table:bitops}
+\end{figure}
+
+Table~~\ref{table:bitops} provides a detailed report on the energy consumption of different implementations of the code.
+Besides \emph{Cython}, each implementation treats the binary operation the same. As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
+
+However, when it comes to Cython, the rotation operations tend to consume 5 times more energy than the other operations, and left rotations consume even more than the right ones.
+
+The reason behind such behavior is the fact that Cython does not have a native implementation for these two operations, so it replaces them with more complex operations, which leads to an extra time of 6s for the right rotation and 10s for the left, compared to 1.5 for the other operations.
+-
+
+Therefore, it is safe to summarize all the operations into one that we call \em{binary}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/bitopts_mean}
+    \caption{Summary of the binary operations for different python implementations }
+    \label{fig:bitops}
+\end{figure}
+Figure \ref{fig:bitops}  shows the average energy consumption of different implementations for the binary operations.One can notice 3 clusters.
+
+The first, which is the interpreted one, uses approximately 500 joules per operation.
+The second is a non-optimized compilation that includes Cython and Nuitka and uses 10 times less energy than the first.
+And the final one, which has been fully optimized, uses 50 percent less energy.
+This category includes jit libraries, also known as numba, Python implementations that include JIT, also known as pypy, and shedskin, which transpile the source code into C and then use a C/C++ compiler to generate the binary, as opposed to cython and nuitka, which directly compile the python code.
+To conclude this.
+JIT has a huge impact when it comes to binary operations, and the wrong choice of the implementation can lead to up to a 100x increase in energy consumption.
+In the other hand,Different binary operations exhibit the same behaviour when it comes to the energy consumption for all the implementations except Cython.
+
 \section{conclusion}
-One may observe that the choice of Python interpreter has a significant impact on the programs' energy consumption.
+One can conclude that the choice of Python interpreter has a significant impact on the programs' energy consumption.
 This investigation is made more intriguing by the absence of a universal solution.

### 2022-09-03 chapters/appendices/benchmarking_paper.tex

--- 2021-08-30 chapters/testing/accuracy.tex
@@
 
-\section{Introduction}
+% \section{Introduction}
 %\lettrine[nindent=0em,lines=3]{C} 
-To conduct robust evaluations, practitioners often try to ensure reproducible environmental conditions in order to properly benchmark their software systems.
-In this area, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
-Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.
+% To conduct robust evaluations, practitioners often try to ensure reproducible environmental conditions in order to properly benchmark their software systems.
+% 
+% Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.
 
-In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should favour reproducible experiments.
-However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only on homogeneous nodes, but even within a single node.
-This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
+% In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should favour reproducible experiments.
+% However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only on homogeneous nodes, but even within a single node.
+% This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
 
-Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
-One can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
+% Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
+% One can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
 
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
-    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
-\end{figure}
+% \begin{figure}%[!htb]
+%     \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
+%     \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
+% \end{figure}
+
+% TODO Should I add the hardware mitigation  for energy variation ?
 
@@
 Therefore, the goal of this paper is to investigate the spectrum of factors that can cause or increase the variability of energy consumption in systems benchmarking, and to propose effective guidelines to control such factors in order to mitigate this variability.
-While this paper does not question the benefits of established CPU features, like C-states or Turbo~Boost, it delivers a deeper analysis of the effects they may introduce on a wide set of experiments and nodes.
-By quantifying potential energy variations induced by controllable factors, we intend to identify the proper configurations that minimize energy variations, depending on the workload characteristics.
-These guidelines aim at supporting practitioners in conducting robust system benchmarks and reporting reproducible energy consumption, rather than recommending production-scale practices or solutions to reduce the power consumption of a given system.
+% While this paper does not question the benefits of established CPU features, like C-states or Turbo~Boost, it delivers a deeper analysis of the effects they may introduce on a wide set of experiments and nodes.
+% By quantifying potential energy variations induced by controllable factors, we intend to identify the proper configurations that minimize energy variations, depending on the workload characteristics.
+% These guidelines aim at supporting practitioners in conducting robust system benchmarks and reporting reproducible energy consumption, rather than recommending production-scale practices or solutions to reduce the power consumption of a given system.
 % 
@@
 
-\paragraph{Objective.}
-The aim of our paper is not to reproduce all the exposed results and explored paths.
-Instead, we intend to check some of these results on latest hardware components, and to consider new potential controllable factors that could also contribute to the energy variation.
-We also aim to provide practitioners with guidelines to conduct robust experiments in a controlled environment, thus reducing the variation overhead to share trustable results and build credible comparisons.
+% \paragraph{Objective.}
+% The aim of our paper is not to reproduce all the exposed results and explored paths.
+% Instead, we intend to check some of these results on latest hardware components, and to consider new potential controllable factors that could also contribute to the energy variation.
+% We also aim to provide practitioners with guidelines to conduct robust experiments in a controlled environment, thus reducing the variation overhead to share trustable results and build credible comparisons.
 
@@
 Most importantly, this should motivate future works on creating a better knowledge on the variability due to CPU manufacturing process and other factors.
+
+\subsection{Testing Protocol}
+the results of this protocil can be found in the project of the continious integrtation that was made by the intern Mamadou to give a feedback about the evolution of the energyconsumption within time

### 2022-09-03 chapters/benchmarking/experiment.tex

@@
 %% This is a conclusion maaan  !! 
-\section{Improving accuracy by taming the energy variation}\label{sec:taming-the-energy-variation}
+\section{Improving Accuracy by Taming Energy Variations}\label{sec:taming-the-energy-variation}
 
@@
 We used the applications (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
-Furthermore we have used other applications to validate our results using more applications such as \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}\footnote{\url{https://linux.die.net/man/1/sha256sum}}.
+Furthermore we have used other applications to validate our results using more applications such as \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}.\footnote{\url{https://linux.die.net/man/1/sha256sum}}
 

### 2022-09-03 chapters/literature/draft_citations.tex
refactoring : revert the changes for the benchmarking protocol chapter

@@
 F) for Homes, cars and electronic products, to provide the consumer an indication of the energy consumption of his devices, which will reflect on his power bill.
-This creterions is extended even to the hardware components of computer.
-figure\ref{fig:soa_comparaisoncpu} \footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}}
+This criterion is extended even to the hardware components of computer.
+Figure~\ref{fig:soa_comparaisoncpu}\footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}}
 
@@
 % that serves as a basis for evaluation or comparison (as of computer system performance)
-\subsubsection{Impact of these challenges in the empirical research}
+\subsubsection{Impact of these Challenges in the Empirical Research}
 In their work~\cite{van_der_kouwe_benchmarking_2018}, Van-der-Kouwe~\emph{et~al.} investigated 50 papers published in top venues
@@
 
+% %%%%% move it to SOA maybe 
+% \subsection{Virtual Machines}
+% First option should be using \emph{Virtual Machines} (VM), which gives researchers the freedom to choose the most appropriate tools, software and operating system that they are the most comfortable with, without paying the price to change the actual working environment, which will give them eventually more control over the dependencies and the execution environment.
+% Moreover, using a VM addresses the \emph{replication crisis} thanks to the virtual images, as even the most complex architecture can be reproduced easily by instantiating a copy of the image.
+% Since the virtual machines are agnostic to the host architecture, researchers will not have to worry about where and how their experiments are replicated, because they have already setup the execution environment.
+% Another advantage of VM is the snapshot mechanism, as it allows researchers to create backups and revert some changes with simple clicks. 
+% Last but not least, thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables---controlled and uncontrolled---and do other analysis without dealing with any dependencies.
+% In his paper~\cite{howe_virtual_2012}, Bill Howe lists the benefits of using VM in researchers experiments, including the economical impact and cultural limitation to a such approach.
+% In particular, VM allow them to have control over the resources, the dependencies, and the execution environment.
+% Moreover, thanks to snapshots, deploying a software is easily achieved by instantiating a copy of that image.
+% However, this choice comes with a certain cost.
+% Because of the hypervisor, software will build on two kernels---the virtual machine (guest) one and the host machine one---which will provide a noticeable overhead, and will impact the performances of the system-under-test.
+% Therefore, we cannot use VM for experiments that are related to performance.
+% Another limitation of VM is the the isolation.
+% While this feature prevents the experimental environment from any undesirable interference from the outside world, this interaction may be required---especially when the experiment is dependent to an external
 %%%% TODO : Elaborate this section and check what should be added 
@@
 
-PowerMon and its upgraded version powerMon2 \cite{bedard2010powermon} are based on an micro-controller chip that can monitor up to 6 channel (8 for powermon2), simultanetly. 
+PowerMon and its upgraded version powerMon2 \cite{bedard2010powermon} are based on an micro-controller chip that can monitor up to 6 channel (8 for powermon2), simultaneously. 
 Therefore, we are able to monitor the power consumption of 4 devices at the same time. 

### 2022-09-03 chapters/optimization/java-dir/java.tex

@@
 \section{Introduction}
-As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners. %ROMAIN: Add source
-Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language. %ROMAIN:  Add source
+As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.In fact, in 2022, Java will be second only to Python, according to PYPY~\footnote{\url{https://pypl.github.io/PYPL.html}}.
+Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language.According to the TIOBE index, Java was the most frequently used language from 2002 until 2017, and it remained in the top 5 after that~\footnote{\url{https://www.tiobe.com/tiobe-index/}}.
 In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performance, Java applications can be at the same time one of the most energy-efficient or hungry solutions.
@@
 To answer those research questions, we conduct an empirical study to highlight the impact of this runtime.
+Figure
 
 % \subsection{definition of the JVM}
-% TODO : ADD definition here 
-\begin{figure}%[ht]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/Jvm_architecture}
-    \caption{JVM architecture}
-    \label{fig:hotspot}
-\end{figure}
+% TODO : ADD it to the introduction to 
+% \begin{figure}%[ht]
+%     \centering
+%     \includegraphics[width=\linewidth]{imgs/Jvm_architecture}
+%     \caption{JVM architecture}
+%     \label{fig:hotspot}
+% \end{figure}
 

### 2022-09-03 chapters/optimization/python-dir/python.tex

@@
 \section{Green Web Development}\label{sec:webdev}
-% TODO : ADD introduction to the orms 
-We start by studying the impact of the \emph{Object Relational Mapping} (ORM)~\cite{o2008object} on energy consumption.
+Django~\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}}  are the most popular frameworks for web development.
+According to a Jetbrains poll\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} , Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.
+
+In contrast to Flask, which is still a microframework, Django is a high-level web framework that provides a standard method for creating and maintaining complex and scalable database-driven websites quickly and effectively.
+Therefore, this study will focus on The latter one.
+
+\subsection{life cycle of a request in Django}
+
+Django is a MVT (Model-View-Template) framework, which means that it follows the MVC (Model-View-Controller) pattern.
+Figure~\ref{fig:django-life-cycle} describes the life cycle of a request in this framework.
+Whenever a request arrives in Django, it is processed by several middlewares, one at a time.
+These middlewares are in charge of authentication, security, and so on.
+Once the request is processed by these middlewares, it is passed to the URL Router, which will simply extract the URL from the request and will try to match it to the defined URLs.
+After getting the matching URL, the corresponding view function is called.
+This function is responsible for treating the request, gathering the data, and then generating the response that will be put inside a template to be returned as a response.
+However, as mentioned before, Django is an MVT model. which offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django_request_lifecycle}
+    \caption{Request-Response life cycle in Django}
+    \label{fig:django-life-cycle}
+\end{figure}
+
+First, we start by investigating the energy consumption of the request-response life cycle in Django, in ordrer to determine which part consumes the most energy.
+To do so, we created a simple Django application that simply returns the response of the request.
+and then tracked its energy consumption using joulehunter~\footnote{\url{https://github.com/powerapi-ng/joulehunter}}.
+
+Joulehunter is an open source library that we developed in order to help practitioners find the energy hotspots in their applications using statistical profilers.
+In the case of Django, it can be added as middleware without any extra setup or changes to the source code.
+The energy consumption of the request-response life cycle in Django is shown in the figure~\ref{fig:django_life_cycle_naive}.As we can notice, 91.4\% of the total energy consumption is spent to resolve the request by retrieving the data while only 5.27\% is spent to render the response.
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django_life_cycle_naive}
+    \caption{Tree representation of the energy consumption of a single request in django (naive version)}
+    \label{fig:django_life_cycle_naive}
+\end{figure}
+
+%% REMARQUE the experiment with joulehunter isn't conducted within the same machine that is used for generating the boxplot graphs 
 % We have chosen Django as our web framework.
-The idea behind this use case is to observe if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
-We, therefore, create a Django website to analyze the cost of a single request using various mechanisms to retrieve data from the database.
-
+Therefore,we chose to observe if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
+To accomplish so, we examined the cost of a single request of the prior website using various ways to extract data from the database.
 We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records.
@@
 For example, using the \textsf{Pypy} interpreter reduces the energy consumption, even when we adopting the \textsf{Vanilla} strategy.
+
+Finally, we run the same experiment with the \textsf{Optimized} strategy 
+using joulehunter
+.Figure~\ref{fig:django_profiled_optimized} shows the new graph of the energy consumption.
+As one can notice.
+while the rendering method consumed the same amount of energy as in  the \textsf{Vanilla} strategy (arround 1.3KJ), the resolve part droped by $20\times$.
+
+\paragraph{Conclusion}
+We conclude this study by stating that the database and ORM selection have a major impact on the website's performance and energy consumption.
+And unlike the rendering portion, there are many options available for the database, the ORM, and the data handling strategy.
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django_profiled_optimized}
+    \caption{Tree representation of the energy consumption of a single request in django (naive version)}
+    \label{fig:django_profiled_optimized}
+\end{figure}
 
@@
 %%% putting the idea, later i ll reformulate it 
-The reason behind such a behavior is mainly related to how the Python interpreter is implemented.
-% add refenece https://www.python.org/doc/essays/list2str/ 
-%% https://www.pythonpool.com/for-vs-while-loop-python/  
-% https://wiki.python.org/moin/TimeComplexity
-%% http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1
-To reduce the latency of Python applications, most of the built-in functions and operations are written in C, and the same goes for the function \texttt{range}.
-Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$ as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object.
+The reason behind such a behavior is mainly related to how the Python interpreter is implemented\footnote{\url{https://www.python.org/doc/essays/list2str}},\footnote{\url{https://www.pythonpool.com/for-vs-while-loop-python/}}.
+To reduce the latency of Python applications, most of the built-in functions and operations are written in C, and the same goes for the function \texttt{range}.\footnote{\url{ http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1}}
+Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$ as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object~\footnote{\url{https://wiki.python.org/moin/TimeComplexity}}.
 Therefore, the \texttt{for in range} is creating a new iterator that has the same length as the first one and, for each iteration, requires a second access (\texttt{l[i]}) instead of one---explaining the doubled time.

### 2022-09-03 chapters/programming_languages/programming_languages.tex

@@
-\chapter{Energy footprint of programming languages}
+\chapter{Energy Footprint of Programming Languages}
 \label{chapter:porgramming_langauges}
-In this chapter, we talk about how the programming language affects how much energy the software consumes.We suggest starting with general microbenchmarking and watching how each programming language performs with the CPU and memory.The main goal of this chapter is to advise developers on how to choose a programming language based on their project's needs in order to make their product use the least amount of energy possible.
-For such a question, no answer is evidedent.But there are some features we can take from each programming language, such as:
+In this chapter, we talk about how the programming language affects how much energy the software consumes.We suggest starting with general micro-benchmarking and watching how each programming language performs with the CPU and memory.The main goal of this chapter is to advise developers on how to choose a programming language based on their project's needs in order to make their product use the least amount of energy possible.
+For such a question, no answer is obvious.
+Nonetheless, there are some features we can take from each programming language, such as: 
 \begin{itemize}
-    \item performance
-    \item community support
-    \item scalability
-    \item energy consumptiom
-    \item memory usage
+    \item performance,
+    \item community support,
+    \item scalability,
+    \item energy consumption,
+    \item memory usage.
 \end{itemize}
@@
 Therefore, we extend this study to include real-world use cases, with two case studies provided in the parts that follow.
-\section{Remote Procedure Call Frameworks}
+
+\section{Investigating Remote Procedure Call Frameworks}
 \import{\currfiledir/rpc}{rpc}
-\section{web Frameworks}
+
+\section{Investigating Web Application Frameworks}
 \import{\currfiledir/web_frameworks}{web_frameworks}

### 2022-09-08 chapters/optimization/python-dir/python_interperters.tex
update : techempower

@@
 \subsection{python and binarry operations }
-The second part of this study, is to look at how different Python implementations work while treating different binary operations .
+The second part of this study is to look at how different Python implementations work while treating different binary operations .
 

### 2022-09-12 chapters/literature/draft_citations.tex
update : benhcmarking soa

@@
 
+\section{Introduction}
 Efficiency in energy usage is a well-known topic.
-In the majority of fields, the purpose is to minimize the energy consumption of electrical devices/components.
+In the majority of fields, the purpose is to minimize the energy consumption of electrical devices and components.
 Modern times even see energy classification (A, B,...
-F) for Homes, cars and electronic products, to provide the consumer an indication of the energy consumption of his devices, which will reflect on his power bill.
-This criterion is extended even to the hardware components of computer.
-Figure~\ref{fig:soa_comparaisoncpu}\footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}}
+F) for homes, cars, and electronic products, to provide the consumer an indication of the energy consumption of their devices, which will reflect on their power bill.
+This criterion is extended even to the hardware components of a computer.
+Figure~\ref{fig:soa_comparaisoncpu}\footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}} Presents a comparaison between to Intel CPU, the i9-12900KS and the i9-12900KF.The difference between these two CPUs is that the KS has an unlocked multiplier, allowing it to be overclocked.
+As a result, the basic consumption is less than the KF.
+This statistic also estimates the average power consumption of these two CPUs each day, as well as the monetary equivalent, to make people more aware of the values of energy consumption rather than the raw data.
 
@@
 
-In computer science, the objective is essentially same.
-Numerous studies have been conducted on energy optimization.
+In computer science, the objective is essentially the same.Numerous studies have been conducted on energy optimization.
 Some of these studies concentrate on minimizing energy consumption at the hardware level, while others optimize energy consumption via software.
+
 % displays the digital energy consumption distribution in 2017 \cite{andrae2015global}. ("p" for production and "u" for usage). 
@@
 The research found a gradual decline in the PUE of data centers, which measures the ratio of the overall energy supplied to the energy used by IT equipment.
-A low PUE implies that the majority of energy is utilized to power the data center's IT equipment, while just a little amount is needed for cooling and lighting.
+A low PUE implies that the majority of energy is utilized to power the data center's IT equipment, while just a small amount is needed for cooling and lighting.
 
@@
 \subsubsection{Reproducibility}
-The reproducibility of experiments is often ranked as one of the most difficult difficulties faced by researchers .This issue has been one of the major tools science has used to help establish the validity and importance of scientific findings since the Philosophical Transactions of the Royal Society were established in 1665~\cite{hankins1986debate} .
-In point of fact, many of the outcomes cannot be duplicated.,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
-As this crisis hit most of the empirical studies, most of the reviews now includes reproducibility as one of the minimal standard for judging scientific merit.\cite{peng2011reproducible}
-One of the criterions supporting reproducibility is the publication of the dataset and the algorithms run on the raw data to conclude the results.
+Experiment reproducibility is frequently listed as one of the most difficult challenges faced by researchers.Reproducing an experiment has been one of the major tools science has used to help establish the validity and importance of scientific findings since the Philosophical Transactions of the Royal Society were established in 1665~\cite{hankins1986debate}.
+In point of fact, many of the outcomes are not duplicable,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
+As a result of the crisis affecting the majority of empirical studies, most reviews now include reproducibility as a minimum standard for judging scientific merit~\cite{peng2011reproducible}.
+One of the criteria for  supporting reproducibility is the publication of the dataset and the algorithms run on the raw data to derive the results.
 There is even some disagreement about what the terms "reproducibility" or "replicability" by themselves mean \cite{goodman2016does}.
-According to \cite{echtler2018open}, \emph{replicability} extends \emph{reproducibility} with the ability to collect a new raw dataset comparable to the original one by re-executing the experiment under similar condition, instead of just the ability to get the same results by running the statistical analyses on the original data set.
+According to \cite{echtler2018open}, \emph{replicability} extends \emph{reproducibility} with the ability to collect a new raw dataset comparable to the original one by re-executing the experiment under similar conditions, instead of just the ability to get the same results by running the statistical analyses on the original data set.
 \subsubsection{Accuracy}
 According to Oxford, \emph{accuracy} means \href{https://www.lexico.com/definition/accuracy}{"technical The degree to which the result of a measurement, calculation, or specification conforms to the correct value or a standard"}.
-In ny case, this means the ability to run the benchmark multiple times with a low variation.
+In our case, this means the ability to run the benchmark multiple times with low variation. This can be achieved by controlling the experiment environment, allowing less room for random factors.
+In biology, chimestry, and electronics, they use clean rooms, which are environments where pollutants like dust, airborne microbes, and aerosol particles are filtered out and factors like humidity, air flow, and temperature can be regulated.
+As for empirical analysis, the accuracy can be measured by numeric metrics such as the variance,the standard deviation (STD), and the interval inter quartile (IRQ).
+Section~\ref{sec:energy_variation} will go over the accuracy in the subject of energy optimisation.
 
 \subsubsection{Representativeness}
-As obvious as it seems, the reason of executing benchmarks is to validate ideas so we can use them in the real life.
-However, this is means that those benchmarks have to represent reality somehow.
-Basically, when we aim to benchmark something, we create a mock up version of the situation that we want it to work.
+As obvious as it seems, the reason for executing benchmarks is to validate ideas so we can use them in real life.
+However, this means that those benchmarks have to represent reality somehow.
+In Social sciences, this can be achieved by selecting a representative sample size.
+\citeauthor{omair2014sample} presents a guideline on to achieve a such representativeness.
+As for computer science.
+The field of benchmarking is still in its infancy, and there is no consensus on how to achieve representativeness.
+howerver many attemps have been made to provide a set of benchmarks for specific purpose.
+For example, the \href{https://www.spec.org/}{Standard Performance Evaluation Corporation} (SPEC) provides a set of benchmarks for CPU performance evaluation.
+this sets covers a wide range of use cases such as  CPU17 for testing the CPU \footnote{\url{https://www.spec.org/cpu2017/}}, SPECviewperf\footnote{\url{https://gwpg.spec.org/benchmarks/benchmark/specviewperf-2020-v3-1/}} for Graphic usage and
+one can cite stressNg when it comes to benchmark the hardware compoment, the SPEC benchmark when it comes to benchmark the software performance,and SPECPower\footnote{\url{https://www.spec.org/power_ssj2008} }.
+NASA Parallel Benchmarks (NPB~\footnote{\url{https://www.nas.nasa.gov/publications/npb.html}} )
+and HPCchallenge \footnote{\url{https://www.hpcchallenge.org/}} are two other examples of benchmarking sets that are created to represent the high performance computing.
+As for programming languages we can cite Computer Language Benchmarks Game~\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}, which is a collection of benchmarks for various programming languages.
+The benchmarks are designed to be small, self-contained, and easy to implement in any language.
+The benchmarks are also ,made in a way to represent of most of the typical real-world workloads in an isolated manner.
+Dacapo~\cite{blackburn2006dacapo} and renaissance~\cite{prokopec2019renaissance} are another example of a benchmarking set that are created to represent the Java Virtual Machine (JVM) performance.
+On the other hand, a new sort of test has arisen within the software development life cycle.
+This type is known as stress testing, and it is used to assess the software's robustness and reliability before releasing it to the public.
+gatling~\footnote{\url{https://gatling.io}} and TCPCopy\footnote{\url{https://github.com/session-replay-tools/tcpcopy}} are great  examples of a stress testing tools that are used to test the performance server applications.
+
 % But, how can we assure that the benchmark is representative? honestly i don't know. and we can't generalize this. but there are some works that have been done for this.
-First, we can talk about the benchmarking and their selection, then we will talk about the stress benchmark for some applications, and finally we will bring this representativeness in our case and how can we get closer to the energy consumption behavior of a software between the benchmark environnement and the production one.
-Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.
+% Basically, when we aim to benchmark something, we create a mockup version of the situation in which we want it to work.
+% First, we can talk about the benchmarking and their selection, then we will talk about the stress benchmark for some applications, and finally we will bring this representativeness in our case and how can we get closer to the energy consumption behavior of a software between the benchmark environnement and the production one.
+% Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.
 
@@
 
-Figure \ref{fig:soa_experiment_component} shows the four component presented in the article~~\cite{stephen_evaluate_2012}
+Figure \ref{fig:soa_expermiment_component} shows the four component presented in the article~~\cite{stephen_evaluate_2012}
 
@@
 RAPL (\emph{RUnning Average Power Limit}) \cite{hackenberg2013power,hackenberg2015energy} is a set of registers that was introduced by Intel in their CPU since Sandy bridge generation, and later it was followed by AMD, since Family 17h Zen.
-It is capable of monitoring different components of CPU, such as DRAM, cores, and the integrated GPU for desktop processors.
 
-%% TODO :ADD figure of RAPL + references
+\begin{figure}[!hbt]
+    \centering{\includegraphics{imgs/power-planes-rapl}}
+    \caption{INTEL RAPL scopes
+    }\label{fig:rapl-domaine}
+\end{figure}
 
+Figre~\ref{fig:rapl-domaine} shows the different scopes  thhat can be monitored using RAPL.
+The CPU package can be monitored for both server and desktop processors. However, DRAM is only available on server CPUs, whilst the integrated GPU is only available on desktop processors.
 The advantage of such an approach is that the absence of any intrusive measurement tools.
@@
 Because of the overhead that code instrumentation may incur, the authors recommend utilizing the second option.
-Every 10 ms, Jalen records the JVM's stack trace together with the CPU time of threads and computes statistics about method calls.
-These statistics are then utilized to calculate each method's energy usage.
+Every 10 ms, Jalen records the JVM's stack trace together with the CPU time of threads and computes statistics about method calls.These statistics are then utilized to calculate each method's energy usage.
 
-selfwatts \cite{fieni2021selfwatts}
-powerjoular \cite{noureddine2022powerjoular}
+% selfwatts \cite{fieni2021selfwatts}
+% powerjoular \cite{noureddine2022powerjoular}
 
-\subsection{Hybrid Tools}
+% \subsection{Hybrid Tools}
 % split hardware tools into extern and embeded tools
 
-Simgrid simulation of power consumption of parallel aplication using simgrid \cite{heinrich2017predicting}
+% Simgrid simulation of power consumption of parallel aplication using simgrid \cite{heinrich2017predicting}
 
-impact of manifacturing process on the energy variation \cite{coles2014comparing}
+% impact of manifacturing process on the energy variation \cite{coles2014comparing}
 
-different energy consumption between identical processers [ idle and hight load ]\cite{von2016variations}
+% different energy consumption between identical processers [ idle and hight load ]\cite{von2016variations}
 
-the temperature is the main raison a high energy variation \cite{wang2018potential}
+% the temperature is the main raison a high energy variation \cite{wang2018potential}
 
-external factors are impact on the energy variation \cite{mukherjee2009spatio}
+% external factors are impact on the energy variation \cite{mukherjee2009spatio}
 
-the place of the server in the room doesn't affect the energy variation \cite{diouri2013your}
+% the place of the server in the room doesn't affect the energy variation \cite{diouri2013your}
 
-comparaison of three measuring tools and they did exhebit 10\% variation each time \cite{inadomi2015analyzing}
+% comparaison of three measuring tools and they did exhebit 10\% variation each time \cite{inadomi2015analyzing}
 
@@
 \cite{marathe2017empirical}
-
-\section{Programming Languages \& Performances}\label{section:soa_optimization}
 
@@
 %%%%%%%%%%%%% energy mitigation
-\section{Energy Variation}
+\section{Energy Variation}\label{sec:energy_variation}
 In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of physical measurements.
@@
 
-\note{add the corelation value}
+%  TODO : add the corelation value
 
@@
 All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
-%% this is to prove that the problem is REAL !!!
 

### 2022-09-12 chapters/literature/draft_citations.tex
update : soa / energy measurement

@@
 \subsubsection{Reproducibility}
+
 Experiment reproducibility is frequently listed as one of the most difficult challenges faced by researchers.Reproducing an experiment has been one of the major tools science has used to help establish the validity and importance of scientific findings since the Philosophical Transactions of the Royal Society were established in 1665~\cite{hankins1986debate}.
@@
 According to \cite{echtler2018open}, \emph{replicability} extends \emph{reproducibility} with the ability to collect a new raw dataset comparable to the original one by re-executing the experiment under similar conditions, instead of just the ability to get the same results by running the statistical analyses on the original data set.
+
 \subsubsection{Accuracy}
@@
 The benchmarks are also ,made in a way to represent of most of the typical real-world workloads in an isolated manner.
-Dacapo~\cite{blackburn2006dacapo} and renaissance~\cite{prokopec2019renaissance} are another example of a benchmarking set that are created to represent the Java Virtual Machine (JVM) performance.
+Dacapo~\cite{blackburn2006dacapo} and renaissance~\cite{renaissance} are another example of a benchmarking set that are created to represent the Java Virtual Machine (JVM) performance.
 On the other hand, a new sort of test has arisen within the software development life cycle.
@@
 
-\subsection{Proposed Solutions}\label[]{sec:soa_reproducibility}
+\subsection{Proposed Solutions}\label{sec:soa_reproducibility}
+Researchers have proposed several solutions to overcome these challenges in the computer science field. We will discuss some of them below.
 %% BENCHMARKING FRAMEWORK -- ADD IT later \cite{sonnenwald2003evaluating}
 %further references 
-In the book "Measuring computer performance: a practitioner's guide"\cite{lilja2005measuring},\citeauthor{lilja2005measuring} examines performance indicators and gives in-depth treatment of benchmark program tactics.
-He provides clear explanations of the basic statistical methods required for interpreting measured performance data.
-He also outlines the overall 'design of experiments' method and demonstrates how to collect the most information with the least amount of work.
-This practical book will appeal to anybody seeking a comprehensive, yet intuitive, grasp of computer system performance analysis.
-\citeauthor{mytkowicz2009producing} made an evaluation  of 133 studies from ASPLOS, PACT, PLDI, and CGO, to find out that none of the experimental findings papers appropriately considered measurement bias.
+First, we start with the work of \citeauthor{mytkowicz2009producing} ,where they made an evaluation  of 133 studies from ASPLOS, PACT, PLDI, and CGO, to find out that none of the experimental findings papers appropriately considered measurement bias.
 Which can lead derive incorrect results from an experiment if a seemingly insignificant feature of the experimental design is altered.
 They treated this problem by proposing two strategies for detecting measurement bias by using causal analysis and preventing it with setup randomization \cite{mytkowicz2009producing}.
+another study that was published in the book "Measuring computer performance: a practitioner's guide"\cite{lilja2005measuring},\citeauthor{lilja2005measuring} examinated performance indicators and gave in-depth treatment of benchmark program tactics.
+they provided clear explanations of the basic statistical methods required for interpreting measured performance data.
+they also outlined the overall 'design of experiments' method and demonstrated how to collect the most information with the least amount of work.
+This practical book will appeal to anybody seeking a comprehensive, yet intuitive, grasp of computer system performance analysis.
 
-\citeauthor{bukh1992art} wrote a book about computer performance analysis, where he discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems.
+\citeauthor{bukh1992art} wrote a book about computer performance analysis, where they discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems.
+The article \cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistake in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often seen errors.
 
-the article \cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistake in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often seen errors.
+Another article~\cite{patil2018training} expands on this line of thought by evaluating the uncertainty caused by replications in new research.
+They offered some strategies to capture uncertainty in inferential investigations, such as cross-study validation and ensemble models.
 
-another article~\cite{patil2018training} expands on this line of thought by evaluating the uncertainty caused by replications in new research.
-They offer strategies for capturing uncertainty in inferential investigations, such as cross-study validation and ensemble models.
-
-In their paper\cite{stodden2018empirical}, the authors find that, even though it's a big move in the right direction, journal policies that require authors to give back digital scholarly objects after publication, like the data and code that back up the claims, don't get more than half of these objects back.
+In their paper\cite{stodden2018empirical}, the authors found that, even though it's a big step in the right direction, journal policies that require authors to give back digital scholarly objects after publication, like the data and code that back up the claims, don't get more than half of these objects back.
 Then, using these artifacts, about a fourth of the published computational claims in the study could be made.
-They suggest putting out the claims in the literature and the digital scholarly objects that back them up at the same time.
+They suggested putting out the claims in the literature and the digital scholarly objects that back them up at the same time.
 
 % as for computer sience. In their paper~\cite{buytaert_statistically_nodate}, the authors  separated the start-up measurement from the steady state when comparing different Java optimizations.
-Another work, to make unify the benchmarking methodology across different research works in the field of computer science, we can cite the paper~\cite{stephen_evaluate_2012}.
-In their approach they  divided any empirical experiment into four  components:
-
+\begin{figure}[!hbt]
+    \center{\includegraphics[scale=0.3]{imgs/componentsofanexperiment}}
+    \caption{The proposed decomposition of an experiment~\cite{stephen_evaluate_2012}}\label{fig:soa_expermiment_component}
+\end{figure}
+Finally, to make unify the benchmarking methodology across different research works in the field of computer science, we can cite the paper~\cite{stephen_evaluate_2012}.
+In their approach they  divided any empirical experiment into four  components.
+Figure \ref{fig:soa_expermiment_component} presents these components, which are:
 \begin{enumerate}
@@
 
-Figure \ref{fig:soa_expermiment_component} shows the four component presented in the article~~\cite{stephen_evaluate_2012}
-
-\begin{figure}[!hbt]
-    \center{\includegraphics[scale=0.3]{imgs/componentsofanexperiment}}
-    \caption{The proposed decomposition of an experiment~\cite{stephen_evaluate_2012}}\label{fig:soa_expermiment_component}
-\end{figure}
+The work of this thesis will be based on this approach, since it provides a unified methodology for benchmarking and evaluating the performance of different systems.
 
@@
 \section{Energy Measurement}
-To be able to reduce the energy cost of running programs we should first be able to estimate this energy consumption.
+Now that we have discussed the importance of benchmarking and the different approaches that can be used to evaluate the performance of a system, we will focus on energy measurement, which will be the main metric in this thesis.Therefore,in this section, we will discuss the different approaches that can be used to measure the energy consumption of a system.
 Many studies have been conducted to estimate a such energy consumption.
 that varies from static analysis of the source code to infer its energy consumption like \citeauthor{pereira_helping_2017} where they provided a tool to highlight the most energy consuming parts of the code \cite{pereira_helping_2017}.
-The main advantage of this approach is to be able to estimate the energy consumption of a program without running it.
-Unlike the complexity of programs, the energy consumption is highly related the execution environment.
-Therefore, static analysis might not represent the real behavior of the same program when it is run in the production environment.
-To treat the problem of representativeness, many researchers tend to measure the energy consumption of the programs while they are running them. Hence we will get more accurate results.
-there is a variety of tools to measure such energy, and they cover a large spectrum of usage depends on how \textbf{accurate} and \textbf{precise} those results are needed to be in tone hand, and the price that practitioners are ready to pay for a such accuracy/precision.
-Below we will present some tools that are well known in the literature and discuss their main advantages / drawbacks for our case.
+The key advantage of this approach is that it allows you to estimate a program's energy usage without executing it.
+Unlike program complexity, energy consumption is strongly dependent to the execution environment.
+As a result, static analysis may not accurately represent the behavior of the same program when run in a production context.
+To address the issue of representativeness, many researchers measure the energy consumption of programs as they run. As a result, we will get more accurate results.
+There are various tools for measuring energy, and they cover a wide range of applications depending on how accurate and precise the results must be on the one hand, and the price that practitioners are prepared to pay for such accuracy and precision on the other.
 
-\subsection{Hardware Tools}
 % TODO Add : diagram of the possible point where we can measure the energy 
@@
 \begin{itemize}
-    \item \em{spatial granularity}: the more specific the target of monitoring we are able to measure, the more efficient we can do optimization since we will know what causes the pitfalls of the energy consumption.
-    \item \em{temporal granularity}: same as spatial granularity, temporal granularity helps us to identify the sequence of code that need to be optimized.
-    \item \em{scalability}: this is mainly related to the cost of the tools and the ease of their integration for our system.
-    \item  \em{accuracy}: to eliminate extra hazards and get more representative measurement.
+    \item \em{Spatial granularity}: the more specific the target of monitoring we are able to measure, the more efficient we can do optimization since we will know what causes the pitfalls of the energy consumption.
+    \item \em{Temporal granularity}: same as spatial granularity, temporal granularity helps us to identify the sequence of code that need to be optimized.
+    \item \em{Scalability}: this is mainly related to the cost of the tools and the ease of their integration for our system.
+    \item  \em{Accuracy}: to eliminate extra hazards and get more representative measurement.
 \end{itemize}
@@
 We believe that accuracy can be extracted from those criterions since it is a result of the combination of the two first ones. therefore, we will focus more on the first 3 criterions from later on.
-In the following section, we will discuss some of the well known tools that are used in literature.
-
+below, we will discuss some of the well known tools that are used in literature.
+\subsection{Hardware Tools}
 Nowadays, most of the high performance computing systems (HPC) implements a tool to report the energy consumption of the nodes for the sake of monitoring and administration.

### 2022-09-12 chapters/optimization/python-dir/python.tex
update
Proofreading Python runtimes

@@
 Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$ as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object~\footnote{\url{https://wiki.python.org/moin/TimeComplexity}}.
-Therefore, the \texttt{for in range} is creating a new iterator that has the same length as the first one and, for each iteration, requires a second access (\texttt{l[i]}) instead of one---explaining the doubled time.
+Therefore, the \texttt{for in range} is creating a new iterator that has the same length as the first one and, for each iteration, requires second access (\texttt{l[i]}) instead of one---explaining the doubled time.
 The \texttt{while} is even slower due to the implicit increment of the variable, which causes an extra operation during the loop.
@@
 \subsection{Python \& Multiprocessing}
-\Cref{fig:python_multiprocessing} compares the behaviour of python programs when we try to introduce the parallelism.
-As we know python is a single threaded program thanks to ghe GLS (Global Lock System) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances of the processor % TODO: TO VERIFY
+\Cref{fig:python_multiprocessing} compares the behavior of python programs when we try to introduce parallelism.
+As we know python is a single-threaded program thanks to the GLS (\emph{Global Lock System}) however, due to the increase in the number of cores/threads per CPU, many libraries started to take advantage of this hardware feature by simulating multithreading as multiple instances of the processor.
+% TODO: TO VERIFY
 
@@
     \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-    \caption{energy behaviour based on multiprocessing}
+    \caption{Energy behavior when using multiprocessing.}
     \label{fig:python_multiprocessing}
@@
 
-As we can see in the graph the energy consumption is correlate with the number of threads until we arrive to the limit of the cores and then we lose the adventage of the multiprocessing , well in that case we pass from parallelism to concurency.
-where different sub process have to compete for the CPU Ressources.
-Another finding is when we hit the number of physical cores.
-there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
-basicaly he favorits the hyper threads than the physical core wich will lead to some context switches which cause the slow behaviour, howerver in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the eneryg consumption in this case.
-in the Chapter %% reference the work of intern from lyon  
-we discuss a deeper this behaviour of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
+As we can see in Figure~\ref{fig:python_multiprocessing}, the energy consumption correlates with the number of threads until reaching the number of physical cores, then losing the benefit from multiprocessing by facing concurency issues imposed by different sub-processes competing for the CPU resources.
+Another finding is observed when hitting the number of physical cores: there is an increase of execution time, but with a reduced energy.
+The reason behind this is the scheduler of the operating system.
+More specifically, it favors hyper-threads instead of physical cores, hence increasing context switches which cause the performance bottleneck.
+Howerver,the other two physical cores are not consuming energy, neither their hyper-threads, which explains the energy savings.
+In the Chapter , %% reference the work of intern from lyon  
+we discuss this behavior of the scheduler more deeply and we assess that it is not due to Python, but is a more generic behavior. %% maybe even this graph should go there
 
@@
 Machine learning is becoming an integral part of our daily lives, growing more potent and energy-hungry each year.
-
 As machine learning can have a significant impact on climate change, it is vital to investigate mitigation techniques.
@@
 
-Figure \ref{fig:cum_energy_vs_accuracy} depicts how the curve steepens as training progresses.
+Figure~\ref{fig:cum_energy_vs_accuracy} depicts how the curve steepens as training progresses.
 For example, training to 90\% accuracy requires three times the energy required for training to 80\% accuracy.
+
 \begin{figure}
@@
     \includegraphics[width=\linewidth]{imgs/accuracy_basedonepoch}
-    \caption{accuracy based on epoch  }
+    \caption{Accuracy along epochs.}
     \label{fig:p2}
@@
     \includegraphics[width=\linewidth]{imgs/cumulative_energy_vs_accuracy}
-    \caption{cumulative energy consumption vs accuracy
-    }
+    \caption{Cumulative energy consumption vs accuracy.}
     \label{fig:cum_energy_vs_accuracy}
@@
     \includegraphics[width=\linewidth]{imgs/power_gpu_baedonepoche}
-    \caption{evolution of average gpu power based on epoch  }
+    \caption{Evolution of average GPU power along epochs.}
     \label{fig:p2}
@@
 \paragraph{Conclusion}
-We discovered that when a model's accuracy improves, so does the energy required for a subsequent accuracy gain.
+We discovered that when the model's accuracy improves, so does the energy required for a subsequent accuracy gain.
 This begs the question of when we should discontinue training.
@@
 \input{python_interperters.tex}
+%% TODO : ADD General conclusion for the chapter

### 2022-09-12 chapters/optimization/python-dir/python_interperters.tex
Proofreading Python runtimes

@@
     \label{fig:python-classes}
+    \small
     \center
@@
     \label{fig:pythonimplementations}
+    \small
     \center
@@
         % \cline{2-4}
-                 & Pytson          & Grumpy              & Parakeet    \\
+                 & Pyston          & Grumpy              & Parakeet    \\
                  & Ironpython      &                     & Psyco       \\
@@
 There are other implementations that we did not consider because either the project aborted many years ago or it has very limited support for Python features.
-After the collection of those implementations, we filtered them. To keep only the versions that are still maintained and support most Python features.
-and we classified them into 3 categories depending on their integration with the python code.
-In \Cref{fig:pythonimplementations}, we describe the implementations that we kept and the version of each implementation and its category.
+After the inventory of those implementations, we filtered them.
+To keep only the versions that are still maintained and support most Python features.
+We then classified them into 3 categories depending on their integration with the Python code.
+In \Cref{fig:pythonimplementations}, we describe the implementations that we kept, the version of each implementation and its category.
 
@@
     \centering
-    \includegraphics[width=\linewidth]{imgs/python-implementations-tree}
+    \includegraphics[width=.7\linewidth]{imgs/python-implementations-tree}
     \caption{Python interpeters}
@@
 
-\section{experimetal protocole}
+\section{Experimental Protocol}
+As discussed in the previous chapter, our idea is to design a benchmarking solution that allows practitioners to reproduce and extends our benchmarks.
+This benchmarking solution is also the one we use to answer the research questions addressed in this manuscript.
+%%%% probably ill resue the same things from other chapters
 
-As we have discussed in the previous chapter \ref{chapter:benchmarking}.
-instead of running the tests, the idea was to design a system that allows practiouners to reproduce and extends our tests.
-and then we use the same system to run answer some of researchs question.
-%%%% probably ill resue the same things from other chapters
-\subsection{measurement context}
-\paragraph{Hardware settings}
-
-all our tests have been executed in a Dell PowerEdge C6420 server machine.
-A summary of its hardware is listed in table \ref{fig:dahuconfig}.
-The machine is equiped with a minimal version of Debian\,9 (4.9.0 kernel version) where we install Docker (version 18.09.5).
+\subsection{Measurement Context}
+\paragraph{Hardware settings.}
+All our benchmarks have been executed on a Dell PowerEdge C6420 server, whose hardware features are summarized in Table~\ref{fig:dahuconfig}.
+The server uses a minimal version of Debian\,9 (4.9.0 kernel version) where we install Docker (version 18.09.5).
 
 \begin{table}[hbt]
+    \resizebox*{\linewidth}{!}{
     \begin{tabular}{ll}
@@
     \end{tabular}
-    \caption{Testing Machine Configuration }
+    }
+    \caption{Benchmarking server configuration.}
     \label{fig:dahuconfig}
@@
 
+\paragraph{Software settings.}
+For the sake of reproducibility, each experiment runs within a Docker container.
+% for each test we create a docker image.
+\subsection{Key Performance Metrics}
+Our focus will be mainly on CPU energy consumption as it is ten folds more than the DRAM one, since it is a task-based benchmarking, time is highly correlated within the energy, and it will be only useful to explain some specific energy behaviors, thus we do not put a lot of focus on this metric.
+%% basically i what i want to say is that we we care mostly about energy, because there are studies that were made on the performance, plus the DRAM is not that interested even within the heavy memory  benchmakrs
+
+\paragraph{Energy measurement.}
+As we know, the energy of a program is integral of its power over time.
+For our case, we used Intel \emph{Running Power Average Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754} to collect the power samples of the running tests.
+We run \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, to report on measurements collected by Intel RAPL and upload them to a so-called \emph{computing machine}, then we calculate the Energy using the trapezoidal rule:
+% \begin{figure}[hbt]
+    % \centering
+    \begin{equation}
+        E = \int^a_b P(t)dt \simeq \sum^n_{k=1} \frac{P(t_k-1)+P(t_k)}{2}
+    \end{equation}
+    % \caption{}
+    % \label{fig:trapezrule}
+% \end{figure}
+Figure~\ref{fig:powerapi} overviews the architecture of our benchmarking infrastructure.
+
 \begin{figure}[hbt]
-    \includegraphics[width=\linewidth]{imgs/SmartWatts.png}
-    \caption{powerapi architecture}
+    \includegraphics[width=.9\linewidth]{imgs/SmartWatts.png}
+    \caption{Benchmarking architecture deployed with \textsc{PowerAPI}.}
     \label{fig:powerapi}
@@
 
-\paragraph{software settings}
-For the sake of reproducibility, each experiment runs within a Docker container.
-% for each test we create a docker image.
-\subsection{Metrics}
-Our focus will be mainly on CPU energy consumption because it is ten folds more than the DRAM one, since it is finit job benchmarking, time is highly correlated within the energy, and it will be only useful to explain certain energytical behaviour so we wont put a lot of focus on this metric.
+The motivation for separating measurement collection from energy computations is to reduce any interference with the benchmark, our sensor being a lightweight C program running as a Docker container.
 
-%% basically i what i want to say is that we we care mostly about energy, because there are studies that were made on the performance, plus the DRAM is not that interested even within the heavy memory  benchmakrs
+\subsection{Benchmark Preparation}
+\paragraph{Input workload.}
+To benchmark our implementations, we employed the \textsc{Tommti} microbenchmarks suite.\footnote{\url{http://www.tommti-systems.de/main-Dateien/reviews/languages/benchmarks.html}}
+\textsc{Tommti} is a set of 13 microbenchmarks that examine common language features, such as arithmetic operations, data structures and input/output manipulations, among others.
+In addition to these microbenchmarks, we implemented some binary operations to investigate the behavior of the previous implementations when it comes to low-level operations that only works with registries.
 
-\paragraph{Energy measurement}
-
-As we know, the energy of a program is the integrale of its power overtime.
-For us ower case, we used Intel \emph{Running Power Average Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754} to collect the power samples of the running tests,
-
-We used used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, to report Data collected by intel Rapl and send it into another machine that we call computing machine.
-then we calculate the Energy using  the the trapezoidal rule.
-
-\ref{fig:trapezrule}
-\begin{figure}[hbt]
-    \centering
-    \begin{equation}
-        E = \int^a_b P(t)dt \simeq \sum^n_{k=1} \frac{P(t_k-1)+P(t_k)}{2}
-    \end{equation}
-    \caption{}
-    \label{fig:trapezrule}
-\end{figure}
-
-figure \ref{fig:powerapi} shows the architecture of our testing model.
-
-The reason of separation between data collection and energy calculation is to minimise any interference with the test so our sensor is a light c program running inside a docker container.
-
-\subsection{tests preparation}
-\paragraph{workload}
-To test our implementations, we employed the Tommti microbenchmarks suite~\footnote{\url{http://www.tommti-systems.de/main-Dateien/reviews/languages/benchmarks.html}}.
-A set of 13 microbenchmarks that examine common language features such as arithmetic operations, data structures and input/output manipulations, among others.
-In addition to these tests ,we implemented some binary operations to examine the behavior of the previous implementations when it comes to low-level operations that works only with registries.
-
-To study the behaviour of the python implementation regarding the energy consumption, we have to focus on the effect of the implementation and mitigate the maximum any side effects such as the organisation of the code or any extra consumption due to the operating system or tier libraries.
-Therefore, for each test we took the implementation written in python2 as a reference and tried to use it in other implementations as it is.
-If it is not supported by python3, we transformed the code using the officiel library  \emph{2to3}\footnote{\url{https://docs.python.org/3.7/library/2to3.html}}.
-In the case of the libraries that use \emph{JIT} adding a decorator to the function that we want to optimize was enough, if there are other changes we assume that they alter the original code which is against our purpose.
-
-Each test is implemented in a Docker container for the several reasons
+To study the energy behavior of the Python implementations, we have to focus on the effect of the implementations and mitigate any side effect, such as the organization of the code or any extra consumption due to the operating system or third-party libraries.
+Therefore, for each benchmark, we took the implementation written in Python2 as a reference and tried to use it in other implementations as it is.
+If it is not supported by Python3, we converted the code using the official library \emph{2to3}.\footnote{\url{https://docs.python.org/3.7/library/2to3.html}}
+In the case of the libraries using \emph{JIT} adding a decorator to the function that we want to optimize was enough, if there are other changes, we assume that they alter the original code which is against our purpose.
+Each benchmark is isolated in a Docker container for several reasons:
 \begin{itemize}
-    \item Isolation: each container has only the test program implemented with a single python runtime to remove any interference between different implementations.
-    \item Deplpoyment: to use the testing machine without extra configurations that may alter the behaviour of the os toward the enrgy consumtion
-    \item Reproductibility: One of the most frequent benchmark crimes~\cite{DBLP:journals/corr/abs-1801-02381} in research is the lack of Reproductibility, by using Docker we ensure that each test has an Image that will be accessible in public.
+    \item Isolation: each container has only the benchmark program implemented with a single python runtime to remove any interference between different implementations,
+    \item Deployment: to use the benchmarking machine without extra configurations that may alter the behavior of the operating system toward energy consumption,
+    \item Reproducibility: one of the most frequent benchmark crimes~\cite{DBLP:journals/corr/abs-1801-02381} in research is the lack of reproducibility---by using Docker we ensure that each benchmark has an image that will be publicly accessible.
 \end{itemize}
 
-Despite the presence of the official docker images for the most of the runtimes, we prefered to build our own using the same reference image in order to remove any bias due to the Os used in the offical image.We used ArchLinux with the kernel version 4.9.184 as a base image.
+Despite the presence of the official docker images for most of the runtimes, we preferred to build our own using the same reference image to remove any bias due to the OS used in the official images.
+We used ArchLinux with kernel version 4.9.184 as a base image for all our benchmarks.
 
-\subsection{Extension}
-As we have done with the previous chapters. we provide a tool that allows to extend the tests with new workloads and new candidates.
-In the repository \footnote{\url{https://github.com/chakib-belgaid/python-implementations}}.
-we have a tool that allows to generate new workloads and new candidates.
-The script \texttt{generator.py} allows to create new benchmarks by implementing a python code within different interpeters.
-then it generates \texttt{launcher-benchmark.sh} that can be used directly to run the expermient.
-Furthermore all the successful implementations are stored in seperate directory and the that couldn't work ( mostly because of compatibility issues) stored in a recap file called \texttt{benchmarkTest.md}.
-where benchmark is the name of the new workload;
-To add extra \textbf{candidates} one should add a base docker file
-that containes the new implementation and if there should be extra manipulation that should be added to the workload files ,such as adding new decorator or changing some parameters, then they should be added as an extra function in the script \texttt{generator.py}.
-Finally they should be included in the python candidates.
+\paragraph{Benchmark extension.}
+As we have done with the previous chapters, we provide a tool that allows extending the benchmarks with new workloads and new candidates.
+In the repository listing the Python implementations under study,\footnote{\url{https://github.com/chakib-belgaid/python-implementations}} we propose a dedicated tool to generate new workloads and new candidates.
+The script \texttt{generator.py} allows practitioners to create new benchmarks by implementing a Python code within different interpreters.
+Then, it generates \texttt{launcher-benchmark.sh} that can be executed to run the associated benchmark.
+Furthermore, all the successful implementations are stored in a separate directory and the ones that failed (\emph{e.g.}, mostly because of compatibility issues) are stored in a recap file called \texttt{benchmarkTest.md}, where \texttt{benchmark} is the name of the new workload.
+To add new candidate runtimes, one should add a base Docker file that contains the new implementation and, if there should be extra manipulation that should be added to the workload files such as adding new decorator or changing some parameters, then they should be added as an extra function in the script \texttt{generator.py}.
+Finally, they should be included in the Python candidates.
 
-\section{Results and finding}
-
-This section will be dedicated to the results of the experiments and a statistical analysis of these results.
-
+\section{Results \& Findings}
+This section will be dedicated to the results of the experiments and statistical analysis of these results. 
 Because we are conducting a comparison study, we first did a Shapiro-Wilk~\cite{shapiro1968comparative} test to see if the data follows a normal distribution.
-Despite, the fact some implementations had a p-value bigger than a=0.05 (such as cpython and active python),other implementaions like pypy and numba presented a p-values smaller than 0.01 which leads to reject the hypothesis H0 that all the distributions follow a normal distribution.
-Therefore,to compare the different implementations, we performed the non-parametric Mann-Whitney U test~\cite{zimmerman1987comparative}, with CPython2 and CPython3 as base references.
+Despite the fact that some implementations had a $p$-value higher than $a=0.05$ (such as CPython and ActivePython), other implementations like PyPy and Numba presented a $p$-values smaller than 0.01, which leads to reject the hypothesis $H0$ that all the distributions follow a normal distribution.
+Therefore, to compare the different implementations, we performed the non-parametric Mann-Whitney U test~\cite{zimmerman1987comparative}, with CPython2 and CPython3 as base references.
 
@@
 If the p-value is less than 0.05, the test result is significant.
-This implies that the given implementation's energy consumption differs significantly from the reference implementation's.As one can notice, most of the implemnetations differe significantly from the reference implementation.
+This implies that the given implementation's energy consumption differs significantly from the reference implementation's.As one can notice, most of the implementations differ significantly from the reference implementation.
 \input{table-envs-based-cpython.tex}
-We want to study the position of these implementations toward the references. To do so, we proceed with an agglomerative hierarchical cluster analysis (HCA)~\cite{johnson1967hierarchical}.
-The HCA is a method that allows grouping the different implementations based on their energy consumption.
-This technique is a bottom-up approach that starts with each implementation as a single cluster and then merges the two most similar clusters until all the clusters are merged into a single cluster.
+
+As we want to study the position of these implementations with regards to the references, we proceed with agglomerative \emph{hierarchical cluster analysis} (HCA)~\cite{johnson1967hierarchical}.
+HCA is a method that groups the different implementations based on their energy consumption.
+This technique is a bottom-up approach as it starts with each implementation being a single cluster, and then merges the two most similar clusters until all the clusters are merged into a single cluster.
 The similarity between two clusters is measured by the maximum distance between their furthest points.
 The distance between two points is calculated as the Euclidean distance between the two points.
-Figure \ref{fig:dendogram} displays the dendogram produced by the HCA
+Figure~\ref{fig:dendogram} displays the Dendogram produced by the HCA.
+
 \begin{figure}
@@
 Except GraalPython, one can notice 3 main clusters.
-The first cluster contains the reference implementations (CPython2 and CPython3) and the implementations that are based on interpreters (Ironpython,Intelpython, ActivePython, Jython, and Nuitka).
+The first cluster contains the reference implementations (CPython2 and CPython3) and the implementations that are based on interpreters (IronPython, IntelPython, ActivePython, Jython, and Nuitka).
 Moreover, each implementation is the closest to its reference Python version.
-The interpreters that are based on other virtual machines, such as Jython with JVM and IronPython with.NET, behave slightly differently from the reference implementation.
-As for the micropython, It is the furthest from the references due to his behavior toward exceptions.
-The second cluster contains the interpreter implementations that are based on a JIT compiler (Pypy2 and Pypy3).
-Closer to them we find shedskin, which is a transaltor that converts the Python code into C++ code to be later compiled into binary code.
-The results table~\ref{table:manwithyu} shows that this cluster is by far the greenest one.
-While numba2 and numba3 are also JIT libraries, they differ from PyPy due to their manual optimisation.
-Unlike pypy, which will decide by itself which part of Python to optimize.
-In Numba, the developer must target the portion of the code that will be passed to JIT, while the rest of the code will be executed by the regural interpreter (aka CPython).
+The interpreters that are based on other virtual machines, such as Jython with JVM and IronPython with .NET, behave slightly differently from the reference implementation.
+As for the MicroPython, It is the furthest from the references due to his behavior toward exceptions.
+The second cluster groups the interpreter implementations that are based on a JIT compiler (PyPy2 and PyPy3).
+Closer to them we find Shedskin, which is a translator that converts the Python code into C++ code to be later compiled into binary code.
+Table~\ref{table:manwithyu} shows that this cluster is by far the most energy-efficient one.
+While Numba2 and Numba3 are also JIT libraries, they differ from PyPy due to their manual optimization.
+Unlike PyPy, which will decide by itself which part of Python to optimize.
+In Numba, the developer must target the portion of the code that will be passed to JIT, while the rest of the code will be executed by the regular interpreter (aka CPython).
 
-\paragraph{Remark :}
-Because graalPython was still in its early stages, some techniques required an abnormally lengthy time, affecting the clustering Algorithm.
+\paragraph{Remark:}
+Because GraalPython was still in its early stages, some techniques required an abnormally lengthy time, affecting the clustering algorithm.
 
-To help practiouners to choose the best implementation for their use case, we have created a chart that describes the behavior of each implementation toward a specific aspect of programming. 
-Figure~\ref{fig:tommi_all} present a radar plot for all the implementations that have been used in this experemation.
+To help practitioners to choose the best implementation for their use case, we have created a chart that describes the behavior of each implementation toward a specific aspect of programming. 
+Figure~\ref{fig:tommi_all} introduces a radar plot for all the implementations that have been used in this experimentation.
+
 \begin{figure}[!htb]
@@
       \includegraphics[width=\linewidth]{imgs/alltomti_performance}
-      \caption{Different interpererts optimisation score }
+      \caption{Different interpreter's energy scores.}
       \label{fig:tommi_all}
@@
 
-This graph depicts each implementation's energy score in relation to each benchmark.
+This graph summarizes each implementation energy score when executing our benchmark.
 The lower the energy consumption, the higher the score.
-We adopted a logarithmic scale to allow the reader to compare numerous implementations due to the large differences between them.
+We adopted a logarithmic scale to help the practitioners to compare numerous implementations due to the large differences between them.
 
-As one can notice in figure~\ref{fig:tommi_all}, there is no evolution between CPython2 and CPython3.
-Moreover, Intelpython, and Active Python all behave similarly to the reference implementation .
+As one can notice in Figure~\ref{fig:tommi_all}, there is no evolution between CPython2 and CPython3.
+Moreover, IntelPython, and ActivePython all behave similarly to the reference implementation.
 Therefore, one can conclude that the work done on those interpreters is primarily to improve a specific purpose and not the core interpreter.
-Active Python states that their version is focused on security and prepackaged libraries, which explains why it is slower than other versions due to the addition of more security reinforcements\footnote{\url{https://www.activestate.com/solutions/why-activestate/}} .
-As for IntelPython,it is designed for machine learning. Unfortunately, the Tommti benchmark is geared for general-purpose programming.
-Despite the fact that Nuitka is a compiler, there was no discernible difference in energy consumption. It was even more similar to the reference Python implementation than other interpreters.
-However, if we look at the nuitka techniques, we can see that they simply encapsulate the Python code with an interpreter into a single executable.
-However, shedskin exhebit has the best energy consumption pattern when it comes to arithmetic operations.
-One can conclude it is due to the fact of the native type of the variables.
-Unlike in the JIT, where they are treated as objects in the beginning.
+ActivePython states that their version is focused on security and prepackaged libraries, which explains why it is slower than other versions due to the addition of this security support.\footnote{\url{https://www.activestate.com/solutions/why-activestate/}}
+As for IntelPython, it is designed for machine learning.
+Unfortunately, the \textsc{Tommti} benchmark is geared for general-purpose programming.
+Despite the fact that Nuitka is a compiler, there was no discernible difference in energy consumption.
+It was even more similar to the reference Python implementation than other interpreters.
+However, if we look at the Nuitka techniques, we can see that they simply encapsulate the Python code with an interpreter into a single executable.
+Finally, Shedskin reports on the best energy consumption pattern when it comes to arithmetic operations.
+One can conclude it is due to the fact of the native type of the variables, unlike in the JIT, where they are treated as objects in the beginning.
 
-for the other VM-based interpereters.
-Jython and Ipy lacked in terms of energy optimisation, which was kind of expected since they were at the beginning of the development stage and the main purpose of such implementation is to link the bytecode generated by Jython and IronPython with their respective virtual machines.
+Regarding the other VM-based interpereters, Jython and Ipy lacked in terms of energy optimization, which expected as they were at the beginning of the development stage and the main purpose of such implementation is to link the bytecode generated by Jython and IronPython with their respective virtual machines.
 
-Unlike the previous interpeters, GraalPython exhibits a certain promise when it comes to complex algorithms (nested loops).
-Finally,micro python is dedicated to embedded systems, so claiming it is a powerful CPU machine will be misleading.
+Unlike the previous interpeters, GraalPython exhibits a certain promise when it comes to complex algorithms (nested loops in particular).
+% Finally, MicroPython is dedicated to embedded systems, so claiming it is a powerful CPU machine will be misleading.
 
-When it comes to  the input outputs operations , 
-when it came to input/output manupulation. Except for Jython,most of the interpreters behaved similarly . Because these operations are more dependent on the hardware than the interpreter.
+% When it comes to  the input outputs operations , 
+% when it came to input/output manupulation. Except for Jython,most of the interpreters behaved similarly . Because these operations are more dependent on the hardware than the interpreter.
 
@@
 
-\subsection{python and binarry operations }
-The second part of this study is to look at how different Python implementations work while treating different binary operations .
+\subsection{Python \& Binary Operations}
+The second part of this study is to look at how different Python implementations work when executing different binary operations.
 
 \begin{figure}[!hbt]
-    \begin{tabular}{|lrrrrr|c|}
+    \small
+    \center
+    \begin{tabular}{|lrrrrr||r|}
         \toprule
@@
         \midrule
-        activepython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76  \\
-        cpython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43  \\
+        ActivePython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76  \\
+        CPython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43  \\
         cpython3     & 595.209  & 685.085        & 563.839       & 657.972 & 655.560 & 531.53  \\
         \hline
-        cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26  \\
-        nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2    \\
+        Cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26  \\
+        Nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2    \\
         \hline
-        numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & 9.22    \\
-        numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & 9.59    \\
-        pypy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & 8.13    \\
-        pypy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & 8.21    \\
-        shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & 8.18    \\
+        Numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & \best{9.22}    \\
+        Numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & \best{9.59}    \\
+        PyPy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & \best{8.13}    \\
+        PyPy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & \best{8.21}    \\
+        Shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & \best{8.18}    \\
         \bottomrule
     \end{tabular}
-    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
+    \caption{Energy consumption of different implementations using bit operations benchmark (in Joules).}
     \label{table:bitops}
@@
 
-Table~~\ref{table:bitops} provides a detailed report on the energy consumption of different implementations of the code.
-Besides \emph{Cython}, each implementation treats the binary operation the same. As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
-
+Table~\ref{table:bitops} provides a detailed report on the energy consumption of different implementations of the code.
+Besides \emph{Cython}, each implementation treats the binary operation the same.
+As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
 However, when it comes to Cython, the rotation operations tend to consume 5 times more energy than the other operations, and left rotations consume even more than the right ones.
-
-The reason behind such behavior is the fact that Cython does not have a native implementation for these two operations, so it replaces them with more complex operations, which leads to an extra time of 6s for the right rotation and 10s for the left, compared to 1.5 for the other operations.
--
-
-Therefore, it is safe to summarize all the operations into one that we call \em{binary}.
+The reason behind such behavior is the fact that Cython does not have a native implementation for these two operations, so it replaces them with more complex operations, which leads to an overhead of $6s$ for the right rotation and $10s$ for the left, compared to $1.5s$ for the other operations.
+Therefore, it is safe to summarize all the operations into one that we call \emph{binary}.
 
@@
 \end{figure}
-Figure \ref{fig:bitops}  shows the average energy consumption of different implementations for the binary operations.One can notice 3 clusters.
 
-The first, which is the interpreted one, uses approximately 500 joules per operation.
+Figure~\ref{fig:bitops} depicts the average energy consumption of different implementations for the binary operations.
+One can notice 3 clusters.
+The first, which is the interpreted one, uses approximately 500 Joules per operation.
 The second is a non-optimized compilation that includes Cython and Nuitka and uses 10 times less energy than the first.
 And the final one, which has been fully optimized, uses 50 percent less energy.
-This category includes jit libraries, also known as numba, Python implementations that include JIT, also known as pypy, and shedskin, which transpile the source code into C and then use a C/C++ compiler to generate the binary, as opposed to cython and nuitka, which directly compile the python code.
-To conclude this.
-JIT has a huge impact when it comes to binary operations, and the wrong choice of the implementation can lead to up to a 100x increase in energy consumption.
-In the other hand,Different binary operations exhibit the same behaviour when it comes to the energy consumption for all the implementations except Cython.
+This category includes JIT libraries, also known as Numba, Python implementations that include JIT, also known as PyPy, and Shedskin, which transpile the source code into C and then use a C/C++ compiler to generate the binary, as opposed to Cython and Nuitka, which directly compile the Python code.
+To conclude this experiment, JIT has a huge impact when it comes to binary operations, and an inadequate runtime selection can lead to a 100x increase in energy consumption.
+Furthermore, we observed that different binary operations exhibit the same energy footprint for all the implementations, except Cython.
 
-\section{conclusion}
+\section{Conclusion}
 One can conclude that the choice of Python interpreter has a significant impact on the programs' energy consumption.

### 2022-09-12 chapters/programming_languages/programming_languages.tex
Proofreading RPC

@@
 \label{chapter:porgramming_langauges}
-In this chapter, we talk about how the programming language affects how much energy the software consumes.We suggest starting with general micro-benchmarking and watching how each programming language performs with the CPU and memory.The main goal of this chapter is to advise developers on how to choose a programming language based on their project's needs in order to make their product use the least amount of energy possible.
+In this chapter, we study how the programming language affects the energy the software consumes.
+We suggest starting with general micro-benchmarking and watching how each programming language performs with the CPU and memory.
+The main goal of this chapter is to advise developers on how to choose a programming language based on their project's needs in order to make their product use the least amount of energy possible.
 For such a question, no answer is obvious.

### 2022-09-12 chapters/programming_languages/rpc/rpc.tex
update
Proofreading RPC

--- 2022-06-10 chapters/programming_languages/rpc-dir/rpc.tex
@@
-\subsection*{introduction}
-%  this is just to warmup 
-With the emergence of cloud technologies, many protocols wanted to take the lead.
-Nowdays, most of the architectures are based on multi-services and microservices.
-And, to have higher versatility of developers multiple companies choose to be open to different programming languages.
-Basically, it would be more efficient if we take advantage of each programming language to satisfy a specific need.
+\subsection{Introduction}
+With the success of Internet and the emergence of cloud technologies, many communication protocols compete to take the lead.
+In particular, most of the software architectures are now based on multi-services and micro-service technologies.
+And, to cope with the higher versatility of developers, multiple companies choose to open their services to different programming languages.
+Interestingly, this approach intends to take advantage of each programming language to satisfy specific requirements.
 However, the challenge nowadays is to make the bridge between those platforms.
-We have many initiatives, such as OpenAPI, that try to create a taxonomy for RESTful APIs.
-Other approaches implement all the different interfaces of the protocol by themselves, such as RPC.
+We have many initiatives, such as OpenAPI that try to create a taxonomy for RESTful APIs, while other approaches implement all the different interfaces of the protocol by themselves, such as RPC.
 
 \subsection{Research Questions}
-In this section, we first explore the ease of implementation of this protocol then we will try to answer the following research questions:
+In this section, we first explore the ease of implementation of this protocol, and then we will try to answer the following research questions:
 \begin{description}
-    \item[\textsc{RQ}~1:] How do RPC implementations consume with regards to the size of the incoming request?
-    \item[\textsc{RQ}~2:] How do RPC implementations consume with regards to the number of concurrent clients?
+    \item[\textsc{RQ}~1:] How do RPC implementations consume with regard to the size of the incoming request?
+    \item[\textsc{RQ}~2:] How do RPC implementations consume with regard to the number of concurrent clients?
 \end{description}
 
-\subsection{Experimental protocol}
-\subsubsection{Hardware settings}
+\subsection{Experimental Protocol}
+\subsubsection{Measurement Context}
+\paragraph{Hardware settings.}
 All the experiments are run on the cluster \textsf{paravance} of the G5K platform.
 This cluster is composed of 72 identical machines, each one is equipped with 2 Intel Xeon E5-2630 V3, with 128\,GB of RAM.
-For more accuracy, our SUT (\emph{System Under Test}) is equipped with a minimal version of Debian\,9 (4.9.0 kernel version), which enforces the core processes required for the purpose of our experiment.
+For more accuracy, our SUT (\emph{System Under Test}) runs a minimal version of Debian\,9 (4.9.0 kernel version), which enforces the core processes required for our experiment.
 Furthermore, we used Docker containers technology for reproducibility of the experiments and the isolation of the servers.
 
-\subsubsection{Energy measurements}
+\paragraph{Client \& server environments.}
+To limit the impact of the network on the experiments, we run both the client and the server on the same machine.
+However, we isolate each part on a separate CPU socket to reduce the noise that the client might have on the server and \emph{vice-versa}.
+To do so, for each iteration, we always run the same client on \textsf{socket\,0} and the server that we want to test on \textsf{socket\,1}.
+Both the server and the client use the whole socket for their experiment.
+In addition, all the additional services, such as the kernel and HwPC sensor, are run on \textsf{socket\,0}.
+Therefore, the only process being executed in \textsf{socket\,1} is the server that we benchmark and monitor.
+
+\subsubsection{Key Performance Metrics}
+\paragraph{Energy measurements.}
 To report on the energy consumption, we used HwPC sensor [TOCITE], which is based on Intel RAPL technology, one of the most accurate tools to measure the energy consumption of the CPU and DRAM [TOCITE].
-For better accuracy, we ran the HwPC sensor with a frequency of 10~Hz, and we used the same machine for all the experiments in order to reduce the variability [TOCITE].
+For better accuracy, we ran the HwPC sensor with a frequency of 1~Hz, and we used the same machine for all the experiments to reduce the variability [TOCITE].
 
@@
     \end{center}
-    \caption{Experimental software architecture}\label{fig:rpcprotocol}
+    \caption{Experimental software architecture.}\label{fig:rpcprotocol}
 \end{figure}
 
-\subsubsection{Client and server environments}
-To limit the impact of the network on the experiments, we run both the client and the server on the same machine.
-However, we isolate each part on a separate socket, in order to reduce the effect that the client might have on the server and \emph{vise versa}.
-To do so, for each iteration, we always run the same client on \textsf{socket\,0} and the server that we want to test on \textsf{socket\,1}.
-Both the server and the client use the whole socket for their experiment.
-In addition, all the additional services, such as the kernel and HwPC sensor, are run on \textsf{socket\,0}.
-Therefore, the only process being executed in \textsf{socket\,1} is the server that we benchmark and monitor.
-
-\paragraph{Client:}
-% TODO : add the github repository links 
+\paragraph{Performances.}
 For better accuracy and more details, We use an updated version of the open source RPC benchmarking tool, named GHZ (\url{https://ghz.sh/}).
 The modified version allows us to monitor the average power for each request from both the server and the client sides.
-The new version is available in the repository.%ROMAIN: Which repo?
-The client consumes the protocol description to generate an implementation for the message \texttt{helloworld}, and then forks 50 instances that send the same request to the server concurrently.
+The new version is available in the repository.\footnote{\url{https://github.com/chakib-belgaid/energy_ghz}}
 
-\paragraph{Server:}
+\subsubsection{Input Workload}
+The purpose of the experiment is to analyze the behavior of different GRPC implementations.
+Therefore, we have two kind of worlkoads: \emph{number of clients} is the number of concurrent clients that we want to benchmark (handled by the GHZ client) and the \emph{payload} reflects the size of each request (varies from 50 Bytes up to 10 MB).
+The client consumes the protocol description found in the file \texttt{helloworld.proto} to generate an implementation for the message and then forks multiple instances that send the same request to the server.
+
+\subsubsection*{Candidates}
 The server implementations are based on the official implementation by Google for most of the languages.
 Each server uses 16 cores and is limited to 512\,MB of RAM.
+Each implementation is packaged as a Docker image, which will allow us to add new implementations easily.
 
-\subsection{Results and findings}
-[\textsc{RQ}~1:] How do RPC implementations react to the size of the data ?
-The purpose of this question is to study the behaviour of the server when transferring large objects.
-To do so, we send 80,000 requests to the server whose size scales from 10 bytes up to 10 Megabytes, which results in 10,000 requests per size per server.
+\subsubsection{Extension}
+For the sake of experimental extensibility, we provide a GitHub repository that contains the implementation of all our experiments.\footnote{\url{https://github.com/chakib-belgaid/energy_benchmarking_grpc2}}
+Adding a new RPC \textbf{candidate} can be achieved by creating a new Docker image and putting it in a new folder named after the language.
+As for the \textbf{workload}, it can be extended easily by adding new files in the folder \textbf{payload}.
+
+\subsection{Results \& Findings}
+\paragraph{\textsc{RQ}~1: How do RPC implementations consume with regard to the size of the incoming request?}
+The purpose of this question is to study the energy consumption of the RPC server when transferring large objects.
+To do so, we send $80,000$ requests to the server whose size scales from 10 bytes up to 10 Megabytes, resulting in $10,000$ requests per size per server.
 To eliminate extra factors, we let the server handle the rate at which it can answer each request.
-However, we put a 20 seconds timeout limit for each request.
+However, we put a 20 seconds timeout deadline for each request.
 Therefore, our boundary condition is only the number of requests received by the server.
-For this experiment, we investigate 4 observable variables~:
+For this experiment, we investigate 4 observable variables:
 \begin{enumerate}
-    \item the \textsf{average power consumption} during the the process: this will indicate the overall behaviour of the server in working mode for long durations,
-    \item the \textsf{tail latency} for the 99th percentile, which indicates how performant is the server,
+    \item the \textsf{average power consumption} during the scenario indicates the overall behavior of the server when working for long durations,
+    \item the \textsf{tail latency} for the 99th percentile, which indicates how efficient is the server,
     \item the \textsf{average number of requests per second}, which indicates the average number of clients that the server can handle,
-    \item the \textsf{average energy cost} of a single request: unlike the first indicator, this one shows how green is the implementation taking performance into consideration.
+    \item the \textsf{average energy cost} of a single request: unlike the first indicator, this one shows how green is the implementation considering performance.
 \end{enumerate}
 
-% TODO : ADD the figures 
-The above figure depicts the overall behaviour of each framework based on the size of request (payload).
+% The above figure depicts the overall behavior of each framework based on the size of request (payload).
 For each framework, we can distinguish three modes, and they all depend on the payload size:
 \begin{enumerate}
-    \item \textsf{Stress free} mode when the server has enough resources to satisfy the requests because they require a memory less than a certain threshold (depends on the language and the platform),
-    \item \textsf{Escalation} mode when the requests tend to be bigger, however the server can still manage to handle them, and here where we can see a change in the energetic and performance behaviour,
+    \item \textsf{Stress free} mode when the server has enough resources to satisfy the requests, as they require a memory less than a certain threshold (depends on the language and the platform),
+    \item \textsf{Escalation} mode when the requests tend to be bigger, but the server can still manage to handle them, at the price of a change in the energy and performance behaviors,
     \item \textsf{Broken state} mode when the requests are much heavier and the server break—like 10 MB.%ROMAIN unclear to me what this means
@@
 
-\subsubsection{Stress free mode}
-In this mode, the compiled languages tend to consume less resources (average power).
+\subsubsection{Stress Free Mode}
+In this mode, the compiled languages tend to consume fewer resources (average power).
 JVM-based languages tend to consume more energy, especially Scala.
-However, we do not observe the same behaviour when it comes to efficiency.
+However, we do not observe the same behavior when it comes to efficiency.
 Unlike the other interpreted programming languages, PHP performances could be compared to the compiled ones, such as CPP or GO, and even better to some others, such as Swift.
@@
 
-\subsubsection{Escalation mode}
-In this mode, the behaviour of the server depends on the payload. We observe three behaviours:
+\subsubsection{Escalation Mode}
+In this mode, the behavior of the server depends on the payload. We observe three cases:
 \begin{enumerate}
     \item Drop in performances without an increased power, such as .Net core, Java micronaut, Crystal, and Dart.
-          In this case, the server keeps using the same resources, and sometimes less, because it takes more time to handle the less requests.
-          This class of languages tend to be the most energy consuming when it comes to the cost per request,
-    \item Increase in power without affecting the performances: such as Go, .Net.
-          The energy consumption of a single request, is affected slightly but still increase,
-    \item Increase in power and drop in performances: Despite the increase of the power consumption, the server becomes slightly slower, which increases the cost of the energy cost per request.
+          In this case, the server keeps using the same resources, and sometimes less, because it takes more time to handle the fewer requests.
+          This class of languages tends to be the most energy-consuming when it comes to the cost per request;
+    \item Increase in power without affecting the performances, such as Go or .Net.
+          The energy consumption of a single request, is affected slightly but still increases;
+    \item Increase in power and drop in performances.
+        Despite the increase of power consumption, the server becomes slightly slower, which increases the energy cost per request.
         This cost is still better than the first case, which concludes that the servers in the first category are on the verge of breaking.
@@
 
-We can mention the case of Elixir that kept scaling despite the lack of performances compared to other compiled languages (Go, CPP).
+We can mention the case of Elixir that keeps scaling despite the lack of performances compared to other compiled languages (Go, CPP).
 

### 2022-09-13 chapters/literature/draft_citations.tex
update :soa

@@
 
-as we see in % two diagrams one for research the other one for power usage 
-The CPU is the part responsible of the most energy consumption of softwares. hence the finer we go to measure this energy consumption the better it is for our work.
-Fortunately, Intel and later AMD proposed a tool that estimate the power consumption of different part of the CPI based on counter performances.
+As we see in % two diagrams one for research the other one for power usage 
+The CPU is the part responsible for the most energy consumption in software. Hence, the finer we go to measure this energy consumption, the better it is for our work.
+Fortunately, Intel and later AMD proposed a tool that estimates the power consumption of different parts of the CPI based on counter performances.
 RAPL (\emph{RUnning Average Power Limit}) \cite{hackenberg2013power,hackenberg2015energy} is a set of registers that was introduced by Intel in their CPU since Sandy bridge generation, and later it was followed by AMD, since Family 17h Zen.
@@
 
-jRAPL is another example of an energy measurement tool estimating tool that has been utilized in a variety of publications\cite{guimaraes2016some,liu2015data}.
+JRAPL is another example of an energy measurement tool estimating tool that has been utilized in a variety of publications\cite{guimaraes2016some,liu2015data}.
 This software enables for the energy usage of Java programs, functions, or even a block of code lines to be profiled and measured.
@@
 At the end of the execution, the tool provides the CPU, DRAM, and disk energy usage of the main process.
-
 Jalen~\cite{noureddine2015monitoring} is another tool that profiles and monitors the energy usage of a Java program.
-Unlike jRAPL, Jalen is able to cover the scope down to the function level.
-It gathers data using code instrumentation and statistical sampling at a predetermined pace.
+Unlike jRAPL, Jalen is able to cover the scope down to the function level.It gathers data using code instrumentation and statistical sampling at a predetermined pace.
 Because of the overhead that code instrumentation may incur, the authors recommend utilizing the second option.
@@
 
-%%****** mitigating the energy variation ******
-low-cost scalable variation-aware algorithm \cite{inadomi2015analyzing}
+\section{Energy Variation}\label{sec:energy_variation}
+In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of physical measurements.
+Unfortunately, this is not possible when it comes to measuring the energy consumption of a system.
+Applying the benchmarking guidelines and repeating the same experiment with in the same configuration are not sufficient to reproduce the the same energy measurements, not only between identical machines, but even within the same machine.
+This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
 
-reducing the energy variation by disabling turbo boost \cite{acun2016variation}
+Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
+We can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
 
-reducing the energy consumption in parallel systems \cite{chasapis2016runtime}
+\begin{figure}%[!htb]
+    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
+    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
+\end{figure}
+%% state of the art 
+Some researchers started investigating the hardware impact of the energy variation of power consumption.
+As an example, one can cite~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the main cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
+\cite{heinrich_predicting} described this variation as a set of parameters, such as CPU Frequency and the thermal effect.
+\subsection{Studying Hardware Factors}
+This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
+On the one hand, the correlation between the processor temperature and the energy consumption was one of the most explored paths.
+Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
+On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the CPU temperature and the energy consumption variation are tightly coupled.
 
-\cite{marathe2017empirical}
+%  TODO : add the corelation value
 
-\subsection{Energy}
-energy comparaison of programming languages in the game benchmark by debian \cite{pereira2017energy}
+This exposes the processor temperature as a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.
 
-toward green ranking \cite{couto2017towards}
+The ambient temperature was also discussed in many papers as a candidate factor for the energy variation of a processor.
+In~\cite{ranka_energy_2009}, the authors claimed that energy consumption may vary due to fluctuations caused by the external environment.
+These fluctuations may alter the processor temperature and its energy consumption.
+However, the temperature inside a data center does not show major variations from one node to another.
+In~\cite{el_mehdi_diouri_your_2013}, El~Mehdi~Dirouri~\emph{et~al.} showed that switching the spot of two servers does not affect their energy consumption.
+Moreover, changing hardware components, such as the hard drive, the memory or even the power supply, does not affect the energy variation of a node, making it mainly related to the processor.
+This result was recently assessed by~\cite{wang_potential_2018}, where the rack placement and power supply introduced a maximum of $2.8\,\%$ variation in the observed energy consumption.
 
-java vs kotlin in web performances \cite{bujnowski2020java}
+Beyond hardware components, the accuracy of power meters has also been questioned.
+Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
+All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
 
-rosetta code \cite{nanz2015comparative} \cite{mirowski2020rosetta}
+\subsection{Mitigating Energy Variations.}
+Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
+In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
+The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
+Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
+They also proposed some guidelines to reduce this variation by replacing the old---slower---chips, by load balancing the workload on the CPU cores and leaving one core idle.
+They claimed that the variation between the processor cores is insignificant.
+In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.
 
-network energy comparaison \cite{balasubramanian2009energy}
+In~\cite{marathe_empirical_2017_m}, the authors highlight the increase of energy variation across the latest Intel micro-architectures by a factor of $4$ from Sandy~Bridge to Broadwell, a $15\,\%$ of run-to-run variation within the same processor and the increase of the inter-cores variation from $2.5\,\%$ to $5\,\%$ due to hardware-enforced constraints, concluding with some recommendations for Broadwell usage, such as running one hyper-thread per core.
+%  NOTE :/this might be interesting for greeFaaS
 
-Aequitas \cite{ribic2016aequitas}
+\section{Energy optimisation}
 
-vm placement \cite{mishra2018energy}
+We will now turn our focus to the energy optimisation challenge after considering the various ways for measuring energy consumption in computers and understanding the energy variation problem.
+Over the previous decade, there has been a considerable increase of interest in this field, with many papers proposing different approaches to reduce the energy consumption of software applications.
+This section will passthrough the main contributions in this field, and will focus on the two following parts.
 
-\citeauthor{mishra2018energy}
+\subsection{Energy optimization in the conception phase}
+The first part of this section will focus on energy optimisation in the conception phase, where the goal is to make the final product use less energy by choosing the best set of programming languages, tools, libraries, etc.
+It also includes all the work and optimizations that developers can do to the source code to make the software use less energy when it is running.
 
-static cost of Vms \cite{kurpicz2016much}
+We start with the work of \citeauthor{pereira2017energy} where they  did an energy consumption comparison analysis of the most popular programming languages.
+Using the pareto optimum, the paper  provides recommendations on how to combine some of these languages to enhance code quality while taking into account execution time, memory utilization, and energy consumption using the pareto optimum.
+Some of the study's findings demonstrate that interpreted languages, such as Python, have lower energy efficiency than compiled languages, such as C or Rust.
+The research also offers language combinations that developers might use together to improve energy economy, execution speed, and memory utilization.
+This paper's findings were based on the game benchmark, which is the most famous set of benchmarks that compares several programming languages.
+
+In this paper, we define a ranking of energy efficiency in programming languages.
+We considered a set of computing problems implemented in ten well-known programming languages and monitored the energy consumed when executing each language.
+Our preliminary results show that although the fastest languages tend to be the lowest energy-consuming ones, there are other interesting cases where slower languages are more energy-efficient than faster ones.
+
+\citeauthor{couto2017towards} investigated the influence of programming language choice on the energy consumption of software during execution.In their paper~\cite{couto2017towards}, they examined a set of computing problems written in ten well-known programming languages while observing the energy required when running each language.
+They also found that there are interesting situations in which slower languages use less energy than faster ones, even though fast languages are usually the ones that use the least energy.
+Finally, they produced an energy efficiency rating of programming languages.
+The paper~\cite{bujnowski2020java} compared the energy, performance and database response time of web applications written in Java versus those written in Kotlin.
+They discovered that there is no statistically significant difference in CPU load between individual measurements( less than 2\%) 2.
+Howerver, Kotlin implementation has never earned the best results in any collection of measurements.
+
+Other server works that have studied the impact of the website technology on the energy conusmptuon,\cite{philippot_characterization_2014},\cite{manotas2013investigating}.
+In their work~\cite{philippot_characterization_2014},\citeauthor{philippot_characterization_2014} measured the computer resources used during the loading of a website in a browser, such as memory utilization and energy consumption, for over 500 websites and proposed some best practices for developers.
+
+% rosetta code \cite{nanz2015comparative} \cite{mirowski2020rosetta}
+
+% network energy comparaison \cite{balasubramanian2009energy}
+
+% Aequitas \cite{ribic2016aequitas}
+
+% vm placement \cite{mishra2018energy}
+
+% \citeauthor{mishra2018energy}
+
+% static cost of Vms \cite{kurpicz2016much}
 
@@
 %%%%%%%%%%%%% energy mitigation
-\section{Energy Variation}\label{sec:energy_variation}
-In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of physical measurements.
-Unfortunately, this is not possible when it comes to measuring the energy consumption of a system.
-Applying the benchmarking guidelines and repeating the same experiment with in the same configuration are not sufficient to reproduce the the same energy measurements, not only between identical machines, but even within the same machine.
-This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
+%%****** mitigating the energy variation ******
+% low-cost scalable variation-aware algorithm \cite{inadomi2015analyzing}
 
-Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
-We can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
+% reducing the energy variation by disabling turbo boost \cite{acun2016variation}
 
-\begin{figure}%[!htb]
-    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
-    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
-\end{figure}
-%% state of the art 
-Some researchers started investigating the hardware impact of the energy variation of power consumption.
-As an example, one can cite~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the main cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
-\cite{heinrich_predicting} described this variation as a set of parameters, such as CPU Frequency and the thermal effect.
-\subsection{Studying Hardware Factors}
-This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
-On the one hand, the correlation between the processor temperature and the energy consumption was one of the most explored paths.
-Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
-On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the CPU temperature and the energy consumption variation are tightly coupled.
+% reducing the energy consumption in parallel systems \cite{chasapis2016runtime}
 
-%  TODO : add the corelation value
-
-This exposes the processor temperature as a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.
-
-The ambient temperature was also discussed in many papers as a candidate factor for the energy variation of a processor.
-In~\cite{ranka_energy_2009}, the authors claimed that energy consumption may vary due to fluctuations caused by the external environment.
-These fluctuations may alter the processor temperature and its energy consumption.
-However, the temperature inside a data center does not show major variations from one node to another.
-In~\cite{el_mehdi_diouri_your_2013}, El~Mehdi~Dirouri~\emph{et~al.} showed that switching the spot of two servers does not affect their energy consumption.
-Moreover, changing hardware components, such as the hard drive, the memory or even the power supply, does not affect the energy variation of a node, making it mainly related to the processor.
-This result was recently assessed by~\cite{wang_potential_2018}, where the rack placement and power supply introduced a maximum of $2.8\,\%$ variation in the observed energy consumption.
-
-Beyond hardware components, the accuracy of power meters has also been questioned.
-Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
-All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
-
-\subsection{Mitigating Energy Variations.}
-Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
-In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
-The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
-Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
-They also proposed some guidelines to reduce this variation by replacing the old---slower---chips, by load balancing the workload on the CPU cores and leaving one core idle.
-They claimed that the variation between the processor cores is insignificant.
-In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.
-
-In~\cite{marathe_empirical_2017_m}, the authors highlight the increase of energy variation across the latest Intel micro-architectures by a factor of $4$ from Sandy~Bridge to Broadwell, a $15\,\%$ of run-to-run variation within the same processor and the increase of the inter-cores variation from $2.5\,\%$ to $5\,\%$ due to hardware-enforced constraints, concluding with some recommendations for Broadwell usage, such as running one hyper-thread per core.
-%  NOTE :/this might be interesting for greeFaaS
+\cite{marathe2017empirical}
 

### 2022-09-13 chapters/optimization/python-dir/python.tex
reformat text
update :soa

@@
 57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration.
-\note{The options in this survey were not mutually exclusive. As a result, the total of the percentages is greater than $100\%$.}
+\paragraph{Remark}
+The options in this survey were not mutually exclusive. As a result, the total of the percentages is greater than $100\%$.
 
@@
 In this chapter, we investigate the energy footprint of Python in its most popular domains of adoption.
-We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{s} did in~\cite{hasan_energy_2016-1}.
+We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016-1} did in~\cite{hasan_energy_2016-1}.
 Then, we measure the energy consumption of several Python implementations to propose a non-intrusive technique to improve energy efficiency.
@@
     \item total energy consumption : for the CPU and the  GPU
-    \item
 \end{itemize}

### 2022-09-13 chapters/literature/draft_citations.tex
add : software optimization

@@
 
-JRAPL is another example of an energy measurement tool estimating tool that has been utilized in a variety of publications\cite{guimaraes2016some,liu2015data}.
+JRAPL is another example of an energy measurement tool estimating tool that has been utilized in a variety of publications~\cite{guimaraes2016some,liu2015data}.
 This software enables for the energy usage of Java programs, functions, or even a block of code lines to be profiled and measured.
@@
 Beyond hardware components, the accuracy of power meters has also been questioned.
-Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
+\citeauthor{inadomi_analyzing_2015}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
 All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
@@
 
-We start with the work of \citeauthor{pereira2017energy} where they  did an energy consumption comparison analysis of the most popular programming languages.
+We start with the work of \citeauthor{pereira_energy_2017} where they  did an energy consumption comparison analysis of the most popular programming languages.
 Using the pareto optimum, the paper  provides recommendations on how to combine some of these languages to enhance code quality while taking into account execution time, memory utilization, and energy consumption using the pareto optimum.
@@
 
-Other server works that have studied the impact of the website technology on the energy conusmptuon,\cite{philippot_characterization_2014},\cite{manotas2013investigating}.
+Other works that have studied the impact of the website technology on the energy conusmptuon,\cite{philippot_characterization_2014,manotas_investigating_2013}.
 In their work~\cite{philippot_characterization_2014},\citeauthor{philippot_characterization_2014} measured the computer resources used during the loading of a website in a browser, such as memory utilization and energy consumption, for over 500 websites and proposed some best practices for developers.
 
-% rosetta code \cite{nanz2015comparative} \cite{mirowski2020rosetta}
+Other works examined software energy consumption efficiency through source code changes and optimizations.
+For example, some papers [42, 122] studied the effect of Java collections on energy consumption, depending on the collection size and/or the most executed tasks on the collection (insertion, removal, search).
+They provided some insights into the energy efficiency of some collections for multiple scenarios.
+For example, Hasan et al.
+[52] compared the energy consumption of several Java data structures, analyzing the bytecode using the Wala framework8 and assessing the evolution of the energy consumption in different scenarios (insertion at the beginning, iteration, etc.).
+They also used some automated replacement of LinkedList and ArrayList to simulate best-and worst-case energy consumption scenarios in real production applications.
+Their study showed that using inappropriate collection can cause up to 300\% of energy consumption inefficiency.
 
-% network energy comparaison \cite{balasubramanian2009energy}
+As for the impact of the source code on the energy consumption, we can cite~\cite{pinto_comprehensive_2016,fernandes_assisting_2017} where the authors investigated the impact of Java collections on energy usage based on collection size and its usage (insertion, removal, search) and They provide some insights into the energy efficiency of specific collections under various scenarios.
 
-% Aequitas \cite{ribic2016aequitas}
+\citeauthor{hasan_energy_2016} examinind the energy consumption of multiple Java data structures, analyzing the bytecode and evaluating the change in energy consumption in various circumstances(research,insertion, deletion, etc.).
+They also simulated best- and worst-case energy usage scenarios in real-world production systems by replacing the LinkedList and ArrayList, and discovered that incorrect collection can result in a 300\% increase in energy consumption\cite{hasan_energy_2016}.
 
-% vm placement \cite{mishra2018energy}
+Other studies\cite{longo_reducing_2019,kumar_energy_2017} looked into the energy use of Java primitive types, string operations, and the use of exceptions, loops, and arrays.
+\citeauthor{kumar_energy_2017}, for example, examined the energy usage of code snippets and micro-benchmarks and provided several conclusions, such as string concatenation would use less energy than StringBuilder and StringBuffer and static variables tends to consume 60\% more energy than instance variables.
 
-% \citeauthor{mishra2018energy}
+\citeauthor{pereira_helping_2017} \cite{pereira_helping_2017} presented SPELL, a tool that helps developers spot energy leaks in their source code.
+Using a statistical spectrum-based analysis and JRAPL\cite{guimaraes2016some,liu_data-oriented_2015}, the tool locates energy-inefficient code fragments.
+According to the authors, it is language and context independent.
 
-% static cost of Vms \cite{kurpicz2016much}
+\subsection{Energy optimization in the execution phase}
+
+The second part of this section will focus on energy optimization in the execution phase, where the goal is to optimize this energy consumption for already made software without making changes to the source code.
 
@@
 
-java collection energy footprint  \cite{pinto2016comprehensive} \cite{fernandes2017assisting}
+java collection energy footprint
 
-java classes energy footprint \cite{hasan2016energy}
+% java classes energy footprint \cite{hasan2016energy}
 

### 2022-09-14 chapters/literature/draft_citations.tex
add : execution environment soa

@@
 
-JRAPL is another example of an energy measurement tool estimating tool that has been utilized in a variety of publications~\cite{guimaraes2016some,liu2015data}.
+JRAPL is another example of an energy measurement tool estimating tool that has been utilized in a variety of publications~\cite{guimaraes2016some,liu_data-oriented_2015}.
 This software enables for the energy usage of Java programs, functions, or even a block of code lines to be profiled and measured.
@@
 
-The second part of this section will focus on energy optimization in the execution phase, where the goal is to optimize this energy consumption for already made software without making changes to the source code.
+The second part of this section will focus on energy optimization in the execution phase, where the goal is to optimize this energy consumption for already developed software without making changes to the source code.The goal is to set up and create an environment in which software can run with the least amount of energy.
+This could involve process scheduling, system tunning and so on.
 
-carbon footprint of training neural network \cite{strubell2019energy}
+We begin with Aequitas\cite{ribic2016aequitas}, a system that allows parallel applications to live on power domains that are co-managed (sharing the same CPU).
+The technology is founded on the premise that coexisting programs can regard power-management hardware as a shared resource and collaborate on power management decisions.
+As a result, it accomplishes its purpose by scheduling these applications with a time slicing technique ( as anexamample round robin).
+The authors claim that their strategy achieves a 12.9\% improvement while incurring only a 2.5\% performance cost.
 
-SPELL, the energy leaks detector tool \cite{pereira2017helping}
+As for virtual machines, \citeauthor{kurpicz2016much}.
+analyzed the total energy consumption of a VM in a data center while emphasizing the static cost versus the dynamic one.In this paper~\cite{kurpicz2016much}, the authros introduced the transparent, reproducible, and predictive cost calculator model EPAVE for VM-based environments.
+The purpose of EPAVE is to provide the static cost of each VM on the server, which includes the air conditioning, power distribution, and the dynamic cost that is related to the VM activities.
 
-impact of energy profiling on software \cite{jagroep2017energy}
+The energy usage of virtual machine allocation and task placement has also been investigated\cite{mishra_energy_2018}.In this article, The authors propose a method for mapping workloads to virtual machines and virtual machines to the physical ones (PMs) in an energy-efficient manner.
+To solve the problem of high heterogeneity of activities and resources, the jobs are categorised based on their resource requirements, and then the relevant VM is found, followed by the appropriate PM where the selected VM can be deployed Using a cloud simulator, the authors claimed The suggested technique saves energy by reducing the number of active PMs while also minimizing the makespan and task rejection rate.
 
-impact of maintainability on software energy evolution \cite{calero2021does}
+% carbon footprint of training neural network \cite{strubell2019energy}
 
-definition \cite{wang1993grpc}
+% SPELL, the energy leaks detector tool \cite{pereira2017helping}
 
-comparison \cite{chamas2017comparing}
+% impact of energy profiling on software \cite{jagroep2017energy}
 
-empirical study \cite{de2021empirical}
+% impact of maintainability on software energy evolution \cite{calero2021does}
 
-impact of website implementation on energy consumption \cite{philippot_characterization_2014} \cite{manotas2013investigating}
+% definition \cite{wang1993grpc}
 
-PHP \cite{benmoussa_new_2019} \cite{das_comparison_2016}
+% comparison \cite{chamas2017comparing}
 
-simulation of web cache \cite{cardenas_performance_2005}
+% empirical study \cite{de2021empirical}
 
-java spring analysis \cite{gajewski_analysis_2019}
+% impact of website implementation on energy consumption \cite{philippot_characterization_2014} \cite{manotas2013investigating}
 
-performance \cite{mishra2021web}
+% PHP \cite{benmoussa_new_2019} \cite{das_comparison_2016}
 
-cross compiler benchmarking \cite{yet2016cross}
+% simulation of web cache \cite{cardenas_performance_2005}
 
-\subsection{Python}
+% java spring analysis \cite{gajewski_analysis_2019}
 
-web frameworks on python \cite{pankiv_concurrent_nodate}
+% performance \cite{mishra2021web}
 
-hope library \cite{akeret_hope_2015}
+% cross compiler benchmarking \cite{yet2016cross}
 
-bytecode effect \cite{ben_asher_effect_2009}
+% \subsection{Python}
 
-numba \cite{crist_dask_2016}
+% web frameworks on python \cite{pankiv_concurrent_nodate}
 
-intel python \cite{li_boosting_2016}
+% hope library \cite{akeret_hope_2015}
 
-python 2 vs 3 \cite{modzelewski_pyston_2020}
+% bytecode effect \cite{ben_asher_effect_2009}
 
-python runtime performances \cite{redondo_comprehensive_2015} \cite{murri_performance_2013}
+% numba \cite{crist_dask_2016}
 
-static vs dynamic languages \cite{pang_what_nodate}
+% intel python \cite{li_boosting_2016}
 
-concurent benchmark framework \cite{pankiv2019concurrent}
+% python 2 vs 3 \cite{modzelewski_pyston_2020}
 
-\subsection{JVM}
+% python runtime performances \cite{redondo_comprehensive_2015} \cite{murri_performance_2013}
 
-java vs python \cite{destefanis_statistical_2016}
+% static vs dynamic languages \cite{pang_what_nodate}
 
-hotspot vs J9 \cite{oi2011power} - same but for big querries \cite{chiba2018towards}
+% concurent benchmark framework \cite{pankiv2019concurrent}
 
-constant overheat of the jvm \cite{lafond2006energy}
+% \subsection{JVM}
 
-infer the energy cost based on the byte code instructions \cite{ma2017biogeography}
+% java vs python \cite{destefanis_statistical_2016}
 
-java collection energy footprint
+% hotspot vs J9 \cite{oi2011power} - same but for big querries \cite{chiba2018towards}
 
-% java classes energy footprint \cite{hasan2016energy}
+% constant overheat of the jvm \cite{lafond2006energy}
 
-framework to reduce java collections \cite{manotas2014seeds}
+% infer the energy cost based on the byte code instructions \cite{ma2017biogeography}
 
-energy consumption of java dev tools \cite{baskar2013experimental}
+% java collection energy footprint
 
-Microbenchmarks on jvm \cite{longo2019reducing} \cite{baskar2013experimental}
+% % java classes energy footprint \cite{hasan2016energy}
 
-android automatic refactoring to reduce energy consumption \cite{banerjee2016automated} \cite{rodriguez2017reducing}
+% framework to reduce java collections \cite{manotas2014seeds}
 
-java vs native c in android \cite{corral2014method}
+% energy consumption of java dev tools \cite{baskar2013experimental}
 
-\subsection{benchmarking}
-%%%%%%%%%%% further references 
-NAS \cite{bailey_nas_nodate}
-statistics \cite{he_statistics-based_2019}
-benchmark crimes \cite{van_der_kouwe_benchmarking_2018}
-microservices \cite{grambow_benchmarking_2020}
-impact of webservers on web applications energy \cite{manotas_investigating_2013}
+% Microbenchmarks on jvm \cite{longo2019reducing} \cite{baskar2013experimental}
+
+% android automatic refactoring to reduce energy consumption \cite{banerjee2016automated} \cite{rodriguez2017reducing}
+
+% java vs native c in android \cite{corral2014method}
+
+% \subsection{benchmarking}
+% %%%%%%%%%%% further references 
+% NAS \cite{bailey_nas_nodate}
+% statistics \cite{he_statistics-based_2019}
+% benchmark crimes \cite{van_der_kouwe_benchmarking_2018}
+% microservices \cite{grambow_benchmarking_2020}
+% impact of webservers on web applications energy \cite{manotas_investigating_2013}
 %%%%%%%%%%%%%%%%%  virtualisation
@@
 
-\cite{marathe2017empirical}
+% \cite{marathe2017empirical}
 

### 2022-09-14 chapters/literature/draft_citations.tex
add :  runtime soa

@@
 
+Besides the impact of the orchestration of virtual machine, some works have been targeting the runtime of specific programming languages.Using the TPC-DS benchmark  , \citeauthor{chiba2018towards} investigated the influence of HOTSPOT \footnote\url{https://openjdk.org/groups/hotspot/} and J9\footnote{\url{https://www.eclipse.org/openj9/}} on the performance of SQL-on-Hadoop systems (SPARK and TEZ) to reveal a three-fold disadvantage that one JVM can have over the other\cite{chiba2018towards}.
+\citeauthor{oi_power_2011} also compared the performance of HOTSPOT and J9.
+They demonstrated in their research~\cite{oi_power_2011} that the relative performance of a JVM is affected by the workload.
+They discovered that HOTSPOT's performance ranged from 44\% to 289\% of J9, while its dynamic power consumption ranged from 2.7W to 7.2W, using the SPECjvm2008~\cite{shiv2009specjvm2008} benchmarks.
+
+As for python, \citeauthor{redondo_comprehensive_2015}compared the performance and memory usage of various Python implementations (CPython, Jython, IronPython, PyPy, and so on) using a 215 set of benchmarks to discover that Python2 performed better with short applications, while Python3 versions covered more tests due to compatibility and the fact that Python2 became obsolete.\cite{redondo_comprehensive_2015}.
+
+% java vs python \cite{destefanis_statistical_2016}
+
+% constant overheat of the jvm \cite{lafond2006energy}
+
+% java collection energy footprint
+
+% % java classes energy footprint \cite{hasan2016energy}
+
+% framework to reduce java collections \cite{manotas2014seeds}
+
+% energy consumption of java dev tools \cite{baskar2013experimental}
+
+% Microbenchmarks on jvm \cite{longo2019reducing} \cite{baskar2013experimental}
+
+% android automatic refactoring to reduce energy consumption \cite{banerjee2016automated} \cite{rodriguez2017reducing}
+
+% java vs native c in android \cite{corral2014method}
+
 % carbon footprint of training neural network \cite{strubell2019energy}
@@
 
-% java vs python \cite{destefanis_statistical_2016}
-
-% hotspot vs J9 \cite{oi2011power} - same but for big querries \cite{chiba2018towards}
-
-% constant overheat of the jvm \cite{lafond2006energy}
-
-% infer the energy cost based on the byte code instructions \cite{ma2017biogeography}
-
-% java collection energy footprint
-
-% % java classes energy footprint \cite{hasan2016energy}
-
-% framework to reduce java collections \cite{manotas2014seeds}
-
-% energy consumption of java dev tools \cite{baskar2013experimental}
-
-% Microbenchmarks on jvm \cite{longo2019reducing} \cite{baskar2013experimental}
-
-% android automatic refactoring to reduce energy consumption \cite{banerjee2016automated} \cite{rodriguez2017reducing}
-
-% java vs native c in android \cite{corral2014method}
-
 % \subsection{benchmarking}
@@
 After that, we will use this system to compare the energy consumption of different programming languages.
-We will extend the work of \citeauthor{pereira2017energy} to a closer distance to production environment by comparing a set of use cases.
+We will extend the work of \citeauthor{pereira_energy_2017} to a closer distance to production environment by comparing a set of use cases.
 Starting by GRPC framework, and a set of Web Frameworks.

### 2022-09-14 chapters/optimization/java-dir/java.tex
update
add : execution environment soa
add : java defition

@@
 
+\subsection{characteristics of JVM}
+\begin{figure}%[ht]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/Jvm_architecture}
+    \caption{JVM architecture}
+    \label{fig:JVM_architecture}
+\end{figure}
+
+Java's portability is a core design goal, which means that Java applications will work exactly the same on any operating system and on any hardware.As we see in figure~\ref{fig:JVM_architecture}, Instead of machine code, this is accomplished by compiling Java language code to an intermediate representation known as Java bytecode.
+Java bytecode instructions are similar to machine code, but they are executed by a JVM rather than directly on the machine.
+As a result, Java programs will run slower and consume more memory than C++ programs.
+However, Just-in-Time (JIT) compilation, which is built into the JVM, improves performance by converting bytecode to machine code at runtime.
+Furthermore, an automatic garbage collector (GC) is used by Java to manage memory throughout the object lifecycle and to recover memory when objects are no longer in use.
+
 \subsection{Goal}
@@
 \begin{compactenum}[\indent\bf RQ\,1:]
-    %  TODO : generalize this format of Rq  to all the thesis 
     \item \emph{What is the impact of existing JVM distributions on the energy consumption of Java-based software services?}
@@
 To answer those research questions, we conduct an empirical study to highlight the impact of this runtime.
-Figure
-
-% \subsection{definition of the JVM}
-% TODO : ADD it to the introduction to 
-% \begin{figure}%[ht]
-%     \centering
-%     \includegraphics[width=\linewidth]{imgs/Jvm_architecture}
-%     \caption{JVM architecture}
-%     \label{fig:hotspot}
-% \end{figure}
 

### 2022-09-14 chapters/benchmarking/conclusion.tex
refactoring and fix missing refrences

@@
-\subsection{Extension}\label{sec:bench_extension}
+%  CONCLUSION au lieu de extension
+
+%% maybe add extension as a part of the conclusion for this chapter 
+\section{Extension}\label{sec:bench_extension}
 We propose to extend the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
@@
 note that some of those metrics are external, such the energy consumption of the system , execution time, execution stack ..etc .
-
+\begin{figure}[!htb]
+    \center{\includegraphics[width=\linewidth]{imgs/benchmarkingprotocol}}
+    \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
+\end{figure}
 Figure~\ref{fig:benchmarkingprotocol} explains in detail the relationship within each component of the protocol.
@@
 Instead of retrieving the data directly from the experiment, we suggest to use an observator to allow the extension of the metrics, this observer in general should be outside the experiment to be able to gather as many informations without impacting the the code.
-\begin{figure}[!htb]
-    \center{\includegraphics[width=\linewidth]{imgs/benchmarkingprotocol}}
-    \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
-\end{figure}
 
-%% TODO : make a generic conclusion , 
-%% maybe add extension as a part of the conclusion for this chapter
-
-\section{Perspectives / representativeness}
 By the end of this study, we have gathered enough guidelines to make the tests more reproducible, accurate and extensible.
@@
 % \end{figure}
+\section{Conclusion}

### 2022-09-14 chapters/benchmarking/draft.tex
refactoring and fix missing refrences

@@
-\section{Motivation}
+\section{Reproducibility within the context of energy}\label{sec:benchmarking_reproducibility}
+\subsection{Introduction}
+
+%  NOTE : add introduction and change this into motivation ?? 
 Empirical measurements are critical to capture the effect of developers' choices on software energy consumption.
@@
 % \end{itemize}
-\section{Reproducibility within the context of energy}\label{sec:benchmarking_reproducibility}
+
 % Now that we have covered the general strucutre of the protocol, this section details the specific aspects of the protocol that are relevant to the energy consumption analysis.
@@
 later we will talk about the energy metrics and how to ensure the accuracy when it comes to the measurement of the energy consumption.
-This section discusses various techniques to encapsulate the systems-under-test in order to assure the reproducibility of an experiment.
+This section addresses various strategies for encapsulating the systems-under-test in order to ensure experiment reproducibility.
 
-\paragraph{Virtual Machines}
+\subsection{Virtual Machines}
 The very first choice should be to use \emph{Virtual Machines} (VM), which allow researchers to select the most appropriate tools, software, and operating system that they are most comfortable with without incurring the cost of changing the actual working environment, giving them more control over dependencies and the execution environment.
@@
 
-\paragraph{Containers}
+\subsection{Containers}
 Another option would be to use something that allows us to benefit from the host OS's isolation while simultaneously simplifying replication as proposed by VM and the direct interface with the hardware provided by the traditional techniques.
@@
 
-Using research from \cite{santos2018does} who examined how adding the Docker layer affected energy consumption.
-\citeauthor{santos2018does} conducted their experiment by running numerous benchmarks both with and without Ddocker.
+Using research from \cite{eddie_antonio_santos_how} who examined how adding the Docker layer affected energy consumption.
+\citeauthor{eddie_antonio_santos_how} conducted their experiment by running numerous benchmarks both with and without Ddocker.
 They contrasted the energy usage and execution time that resulted.
@@
     \center{\includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/idle_energy}}
-    \caption{energy consumption of Idle system with and without docker \cite{santos2018does}}\label{fig:docker_idle}
+    \caption{energy consumption of Idle system with and without docker \cite{eddie_antonio_santos_how}}\label{fig:docker_idle}
 \end{figure}
@@
     \includegraphics[width=.5\linewidth]{imgs/docker_vs_vm_energy_paper/reddis_energy}
-    \caption{Execution time \& energy consumption of Redis with and without Docker~\cite{santos2018does}}\label{fig:docker_reddis}
+    \caption{Execution time \& energy consumption of Redis with and without Docker~\cite{eddie_antonio_santos_how}}\label{fig:docker_reddis}
 \end{figure}

### 2022-09-14 chapters/benchmarking/experiment.tex
refactoring and fix missing refrences

@@
 \section{Improving Accuracy by Taming Energy Variations}\label{sec:taming-the-energy-variation}
-
+\subsection{introduction}
 While the previous section aimed to ensure the reproducibility of our software energy consumption experiments, this section provides a collection of tips and tools to help increase the accuracy of these experiments.
@@
 % We will first start with evaluating the parameters mentioned by \cite{heinrich_predicting}.
-
-In this section, we investigate the following research questions:
+\subsection{Research questions}
+This study will focus on the following research questions:
 \begin{description}
@@
 Then, we compared the job energy consumption to the global RAPL data.
-We calculated the Pearson correlation~\cite{ref1} of the energy consumption and variation between global RAPL and \textsc{PowerAPI}, as illustrated in Figure~\ref{fig:correlation}.
+We calculated the Pearson correlation~\cite{ref1} of the energy consumption and variation between global RAPL and \textsc{PowerAPI}, as illustrated in Figure~\ref{fig:ev_correlation}.
 The job energy consumption and variation are strongly correlated with the global energy consumption and variation with the coefficients 93.6\,\% and 85.3\,\%, respectively.
@@
     \center{\includegraphics[width=.8\linewidth]{imgs/correlation}}
-    \caption{The correlation between the RAPL and the job consumption and variation}\label{fig:correlation}
+    \caption{The correlation between the RAPL and the job consumption and variation}\label{fig:ev_correlation}
 \end{figure}

### 2022-09-14 chapters/literature/draft_citations.tex
refactoring and fix missing refrences

@@
 
-As an example \citeauthor{avgerinou2017trends} evaluated the development of power use effectiveness (PUE) in data centers thkat belongs to various organizations participating in the European code of conduct for energy efficiency program \cite{avgerinou2017trends}.
+As an example \citeauthor{avgerinou_trends_2017} evaluated the development of power use effectiveness (PUE) in data centers thkat belongs to various organizations participating in the European code of conduct for energy efficiency program \cite{avgerinou_trends_2017}.
 The research found a gradual decline in the PUE of data centers, which measures the ratio of the overall energy supplied to the energy used by IT equipment.
@@
 Section~\ref{sec:soa_benchmarking} will discuss the pitfalls and best practices associated with empirical research before applying them to our field of interest.
-After that Section~\ref{sec:energymeasurement} describes software energy measurements.
+After that Section~\ref{sec:soa_energymeasurement} describes software energy measurements.
 It provides examples of hardware and software measuring instruments and describes their differences, benefits, and drawbacks.
 It also examines the sources of energy measurement variations, which represent a significant obstacle to achieving precise readings and a higher accuracy.
-Then, in section \ref{section:soa_optimization}, we will go through some of the previous work on improving the energy consumption of softwares.
+Then, in section \ref{section:soa_energyoptimization}, we will go through some of the previous work on improving the energy consumption of softwares.
 
@@
 
-\section{Energy Measurement}
+\section{Energy Measurement}\label{sec:soa_energymeasurement}
 Now that we have discussed the importance of benchmarking and the different approaches that can be used to evaluate the performance of a system, we will focus on energy measurement, which will be the main metric in this thesis.Therefore,in this section, we will discuss the different approaches that can be used to measure the energy consumption of a system.
@@
 
-\section{Energy optimisation}
+\section{Energy optimisation}\label{section:soa_energyoptimization},
 

### 2022-09-14 chapters/optimization/java-dir/java.tex
refactoring and fix missing refrences

@@
 % In other words, we used the same node to ran all the JVM from the different providers and with numerous configurations for each single benchmark application.
-Moreover, we tried to be very careful, while running our experiments, not to fall in the most common benchmarking "crimes"~\cite{crimes}.
+Moreover, we tried to be very careful, while running our experiments, not to fall in the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
 Every single experiment, therefore, reports on energy metrics obtained from at least $20$ executions of $50$ iterations per benchmark.

### 2022-09-14 chapters/programming_languages/rpc/rpc.tex
reformat text
refactoring and fix missing refrences

@@
 
-% The above figure depicts the overall behavior of each framework based on the size of request (payload).
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/energy_cost_payload}
+    \end{center}
+    \caption{Energy consumption based on the request size}\label{fig:energy_cost_payload}
+\end{figure}
+Figure~\ref{fig:energy_cost_payload} depicts the overall behavior of each framework based on the size of request (payload).
 For each framework, we can distinguish three modes, and they all depend on the payload size:
@@
 [\textsc{RQ}~2:] How do RPC implementations react to the number of clients ?
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/energy_cost_clients}
+    \end{center}
+    \caption{Energy behaviour based on the number of the clients}\label{fig:rpcproenergy_cost_clients}
+\end{figure}
 
 \subsubsection{Power behaviour}
+
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/power_consumption_clients}
+    \end{center}
+    \caption{Average power consumption based on the number of the clients}\label{fig:power_consumption_clients}
+\end{figure}
+
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/power_consumption_payload}
+    \end{center}
+    \caption{Average power consumption based on the request size}\label{fig:power_consumption_payload}
+\end{figure}
+
 Based on the heatmap, we can distinguish two modes:
@@
 \paragraph{RPS}
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/rps_clients}
+    \end{center}
+    \caption{Number of requests per second based on the number of clients}\label{fig:rps_clients}
+\end{figure}
+
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/rps_payload}
+    \end{center}
+    \caption{Number of requests per second based on the request size}\label{fig:rps_payload}
+\end{figure}
 Most of the servers hit their RPS limit after 5 clients and 100 clients for vm based servers, and after this the number keeps constant, which will decrease the average RPS per client.
@@
 The increase in the number of requests per second, does not necessarily mean a lower latency.
-As one can notice in Table~\ref{...}, until the 1000 clients, Go provides the least latency beside .Net.
+As one can notice in Fgirue~\ref{fig:tail99_clients}, until the 1000 clients, Go provides the least latency beside .Net.
 GraalVM provides the highest latency, on average.
 However, Dart tends to become slower when we increase the number of clients, until we pass the 600 simultaneous clients, and there it changes its behaviour, instead of satisfying most the requests it notify the clients directly that the server is saturated, hence a drop in satisfaction ratio, and an amelioration for the average latency.
+
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/tail99_clients}
+    \end{center}
+    \caption{Tail latency (99\%) based on the number of clients}\label{fig:tail99_clients}
+\end{figure}
+
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/tail99_payload}
+    \end{center}
+    \caption{Tail latency (99\%) based on the request size}\label{fig:tail99_payload}
+\end{figure}
 
@@
 
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/energy_cost_clients}
-    \end{center}
-    \caption{Experimental software architecture}\label{fig:rpcprotocol}
-\end{figure}
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/energy_cost_payload}
-    \end{center}
-    \caption{Experimental software architecture}\label{fig:energy_cost_payload}
-\end{figure}
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/power_consumption_clients}
-    \end{center}
-    \caption{Experimental software architecture}\label{fig:power_consumption_clients}
-\end{figure}
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/power_consumption_payload}
-    \end{center}
-    \caption{Experimental software architecture}\label{fig:power_consumption_payload}
-\end{figure}
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rps_clients}
-    \end{center}
-    \caption{Experimental software architecture}\label{fig:rps_clients}
-\end{figure}
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rps_payload}
-    \end{center}
-    \caption{Experimental software architecture}\label{fig:rps_payload}
-\end{figure}
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/tail99_clients}
-    \end{center}
-    \caption{Experimental software architecture}\label{fig:tail99_clients}
-\end{figure}
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/tail99_payload}
-    \end{center}
-    \caption{Experimental software architecture}\label{fig:tail99_payload}
-\end{figure}
-
 \subsection{Threads to Validity}

### 2022-09-14 chapters/benchmarking/conclusion.tex
fix : compilation error

@@
 We propose to extend the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
-Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in section ~\ref{sec:soa_reproducibility}, we propose an abstract model to describe an empirical experiment:
+Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in section ~\ref{sec:benchmarking_reproducibility}, we propose an abstract model to describe an empirical experiment:
 

### 2022-09-14 chapters/optimization/python-dir/python.tex
fix : compilation error

@@
     \includegraphics[width=\linewidth]{imgs/programminglanguangespopularity.png}
-    \caption{PYPL Popularity of Programming Languages~\cite{noauthor_pypl_nodate}.}
+    \caption{PYPL Popularity of Programming Languages~\cite{noauthor_pypl_2018}.}
     \label{fig:pypl}
@@
 
-In particular, Noureddine~\emph{et~al.}~\cite{noureddine_preliminary_2012} in 2012, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
+In particular,\citeauthor{noureddine_preliminary_2012}~\cite{noureddine_preliminary_2012} in 2012, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
 According to their experiments, an interpreted programming language, like Python, can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. \Cref{fig:hannoi}).
@@
 In this chapter, we investigate the energy footprint of Python in its most popular domains of adoption.
-We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016-1} did in~\cite{hasan_energy_2016-1}.
+We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016} did in~\cite{hasan_energy_2016}.
 Then, we measure the energy consumption of several Python implementations to propose a non-intrusive technique to improve energy efficiency.
@@
 \section{Green Web Development}\label{sec:webdev}
+
 Django~\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}}  are the most popular frameworks for web development.
@@
 Therefore,we chose to observe if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
+\subsection{Experiment \& Results}
 To accomplish so, we examined the cost of a single request of the prior website using various ways to extract data from the database.
@@
 
-\paragraph{Conclusion}
+\subsection{Conclusion}
 We conclude this study by stating that the database and ORM selection have a major impact on the website's performance and energy consumption.
@@
 
-\subsection{Python Insights}
+\section{Python Insights}
 This section aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
-Therefore, we start by extending the work of \citeauthor{hasan_energy_2016} and \citeauthor{oliveira_recommending_nodate} to the context of Python environments.
-
+Therefore, we start by extending the work of \citeauthor{hasan_energy_2016} and \citeauthor{oliveira_recommending_nodate} to the context of Python environments \cite{hasan_energy_2016,oliveira_recommending_nodate}.
+\subsection{experiment and results}
 %%%% concurency 

### 2022-09-14 chapters/optimization/python-dir/python_interperters.tex
reformat text
refactoring and fix missing refrences
fix : compilation error

@@
 
-\paragraph{Preliminary studies}
+\subsection{Preliminary studies}
 For the first studies, we used the official version of Python, because the goal was mainly to highlight the impact of the structure of code on energy consumption.
@@
 
-\section{Experimental Protocol}
+\subsection{Experimental Protocol}
 As discussed in the previous chapter, our idea is to design a benchmarking solution that allows practitioners to reproduce and extends our benchmarks.
@@
 
-\subsection{Measurement Context}
+\subsubsection{Measurement Context}
 \paragraph{Hardware settings.}
@@
 For the sake of reproducibility, each experiment runs within a Docker container.
-% for each test we create a docker image.
-\subsection{Key Performance Metrics}
+\subsubsection{Key Performance Metrics}
 Our focus will be mainly on CPU energy consumption as it is ten folds more than the DRAM one, since it is a task-based benchmarking, time is highly correlated within the energy, and it will be only useful to explain some specific energy behaviors, thus we do not put a lot of focus on this metric.
@@
 
-\subsection{Benchmark Preparation}
+\subsubsection{Benchmark Preparation}
 \paragraph{Input workload.}
@@
 
-\section{Results \& Findings}
+\subsection{Results \& Findings}
 This section will be dedicated to the results of the experiments and statistical analysis of these results.
@@
       \centering
-      \includegraphics[width=\linewidth]{imgs/alltomti_performance}
+      \includegraphics[width=1.0\linewidth]{imgs/alltomti_performance}
       \caption{Different interpreter's energy scores.}
@@
 
-\section{Conclusion}
+\subsection{Conclusion}
 One can conclude that the choice of Python interpreter has a significant impact on the programs' energy consumption.

### 2022-09-15 chapters/benchmarking/conclusion.tex
refactoring

@@
 %% maybe add extension as a part of the conclusion for this chapter 
-\section{Extension}\label{sec:bench_extension}
+\section{Benchmarking protocol to extend reproducibility}\label{sec:bench_extension}
+
 We propose to extend the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
@@
 % \end{figure}
-\section{Conclusion}
+\section{Conclusion \note{missing}}

### 2022-09-15 chapters/literature/draft_citations.tex
fix : compilation error
refactoring

@@
-% this sections containes all the references for each section :
-
-% 1. group articles by field
-
 \section{Introduction}
@@
     \includegraphics[width=\linewidth]{imgs/cpu_cost_comparaison}
-    \caption{electrical cost comparasion between two cpus }
+    \caption{Electrical cost comparasion between two cpus }
     \label{fig:soa_comparaisoncpu}
@@
 
-we will place a greater emphasis on the software level in order to decrease the amount of energy that is used.more particularly on the execution phase of the program cycle.
+We will place a greater emphasis on the software level in order to decrease the amount of energy that is used.more particularly on the execution phase of the program cycle.
 We will be proceeding through an empirical analysis of the energy consumption of the software while changing some components of the source code without impacting its behavior.
@@
 
-\subsubsection{Reproducibility}
+\subsubsection{Reproducibility}\label{subsec:soa_reproducibility}
 

### 2022-09-15 chapters/optimization/python-dir/python.tex
refactoring

@@
 This section aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
-Therefore, we start by extending the work of \citeauthor{hasan_energy_2016} and \citeauthor{oliveira_recommending_nodate} to the context of Python environments \cite{hasan_energy_2016,oliveira_recommending_nodate}.
+Therefore, we start by extending the work of \citeauthor{hasan_energy_2016} and
+\citeauthor{oliveira_recommending_nodate} to the context of Python environments
+\cite{hasan_energy_2016,oliveira_recommending_nodate}.
 \subsection{experiment and results}
@@
 
-\paragraph{Conclusion}
+\subsection{Conclusion}
 This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
 
-\subsection{Python \& Multiprocessing}
+\section{Python \& Multiprocessing}
 \Cref{fig:python_multiprocessing} compares the behavior of python programs when we try to introduce parallelism.
@@
 
-\subsection{Python \& Machine Learning}
+\section{Python \& Machine Learning}
 Machine learning is becoming an integral part of our daily lives, growing more potent and energy-hungry each year.
 As machine learning can have a significant impact on climate change, it is vital to investigate mitigation techniques.
-
+\subsection{Experimental Protocol}
+\subsubsection{Measurement context}
 \paragraph{Hardware settings}
@@
 These tests are run on top of a minimal version of Debian-10 to increase the accuracy of the tests by eliminating any unnecessary processes.
-
 \subsubsection{Input Workload}
@@
 
-\section{Results \& Findings}
+\subsection{Results \& Findings}
 As the model's accuracy increases, so does the energy required for the next accuracy increment.
@@
 
-\paragraph{Conclusion}
+\subsection{Conclusion}
 We discovered that when the model's accuracy improves, so does the energy required for a subsequent accuracy gain.
@@
 \input{python_interperters.tex}
-%% TODO : ADD General conclusion for the chapter
+\subsection{threads to validity \note{missing}}
+\section{Conclusion \note{missing}}

### 2022-09-15 chapters/programming_languages/programming_languages.tex
refactoring and fix missing refrences
refactoring
Proofreading

@@
-\chapter{Energy Footprint of Programming Languages}
+\chapter{The Energy Footprint of Programming Frameworks}
 \label{chapter:porgramming_langauges}
-In this chapter, we study how the programming language affects the energy the software consumes.
-We suggest starting with general micro-benchmarking and watching how each programming language performs with the CPU and memory.
-The main goal of this chapter is to advise developers on how to choose a programming language based on their project's needs in order to make their product use the least amount of energy possible.
+\section{Introduction}
+In this chapter, we study how the programming framework affects the energy the software consumes.
+We suggest starting with general micro-benchmarking and watching how each programming framework performs with the CPU and memory.
+The main goal of this chapter is to advise developers on how to choose a programming framework based on their project's needs to make their product use the least amount of energy possible.
 For such a question, no answer is obvious.
-Nonetheless, there are some features we can take from each programming language, such as: 
+Nonetheless, there are some features we can take from each programming framework, such as:
 \begin{itemize}
@@
 \import{\currfiledir/rpc}{rpc}
+\clearpage
 
@@
 \import{\currfiledir/web_frameworks}{web_frameworks}
+
+\section{Conclusion\note{missing}}

### 2022-09-15 chapters/literature/draft_citations.tex
update : python runtimes table
update : some references

@@
 
-In computer science, the objective is essentially the same.Numerous studies have been conducted on energy optimization.
+In computer science, the objective is essentially the same.
+Numerous studies have been conducted on energy optimization.
 Some of these studies concentrate on minimizing energy consumption at the hardware level, while others optimize energy consumption via software.
@@
 % In addition, the utilization expenses of data centers, networks, and terminals are also substantial (19 percent , 16 percent and 20 percent respectively).
-% TODO : ask romain about the graphics for energy distribution
 
@@
 
-We will place a greater emphasis on the software level in order to decrease the amount of energy that is used.more particularly on the execution phase of the program cycle.
+We will place a greater emphasis on the software level in order to decrease the amount of energy that is used, more particularly on the execution phase of the program cycle.
 We will be proceeding through an empirical analysis of the energy consumption of the software while changing some components of the source code without impacting its behavior.
@@
 % While this feature prevents the experimental environment from any undesirable interference from the outside world, this interaction may be required---especially when the experiment is dependent to an external
-%%%% TODO : Elaborate this section and check what should be added 
 % As Stephen M.
@@
 
-% TODO Add : diagram of the possible point where we can measure the energy 
 % depends on the scientific usage there are two apporaches :
@@
 
-On the other hand, GreenMiner \cite{hindle2014greenminer}, %%i dont think that  mobile measurement will be relevent for my case 
+% On the other hand, GreenMiner \cite{hindle2014greenminer}, %%i dont think that  mobile measurement will be relevent for my case 
 powerpack~\cite{ge2009powerpack} in the other hand is an api that synchronizes the data gathered from other monitoring tools such as Watt’s Up Pro, NI and RadioShack pro and the lines of code.
@@
 
-%% TODO : add diagram where the distribution of the energy consumption is more related to the CPU rather than other tools - u can cite the tools dedicated to the GPU since it is is included in the software aggregators ?
-power api, pythoules ...etc
-
-As we see in % two diagrams one for research the other one for power usage 
-The CPU is the part responsible for the most energy consumption in software. Hence, the finer we go to measure this energy consumption, the better it is for our work.
+Accoring to the work of \citeauthor{vasques2019review} and\citeauthor{wang2018modelling}
+The CPU is the part responsible for the most energy consumption in a data center\cite{vasques2019review,wang2018modelling}.
+Hence, the finer we go to measure this energy consumption, the better it is for our work.
 Fortunately, Intel and later AMD proposed a tool that estimates the power consumption of different parts of the CPI based on counter performances.
@@
 Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
-On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the CPU temperature and the energy consumption variation are tightly coupled.
-
-%  TODO : add the corelation value
+On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the CPU temperature and the energy consumption variation are tightly coupled (up to 16\% increase in the variation when the temperature changed from 37.7°C to 74.5°C ).
 

### 2022-09-15 chapters/benchmarking/draft.tex

@@
 \begin{enumerate}
-    \item As previously stated [143] [reference Morabito (2015) and van Kessel et al. (2016)], we require a lightweight orchestrator to limit the overhead on energy usage of our experiments
-          \cite{van2016power}
-          % TODO : ask romain about the title of the paper and the reference to the paper
-          [cite Morabito (2015) and van Kessel et al. (2016)],
-          % cite Power efficiency of hypervisor-based virtuali- zation versus container-based virtualization. University of Amsterdam.
+    \item As previously stated in literature~\cite{van2016power,morabito_power_2015}, we require a lightweight orchestrator to limit the overhead on energy usage of our experiments
     \item We need to communicate with the host OS because we are using hardware sensors to measure the energy consumption.
@@
 
-%  TODO : Rephrase this 
 % A special notice to \href{https://github.com/powerapi-ng/virtualwatts}{virtualwatts}. A framework that enables us to retrieve the energy consumption of a virtual machine.

### 2022-09-16 chapters/literature/draft_citations.tex
correction : soa

@@
 This criterion is extended even to the hardware components of a computer.
-Figure~\ref{fig:soa_comparaisoncpu}\footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}} Presents a comparaison between to Intel CPU, the i9-12900KS and the i9-12900KF.The difference between these two CPUs is that the KS has an unlocked multiplier, allowing it to be overclocked.
+Figure~\ref{fig:soa_comparaisoncpu}\footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}} presents a comparaison between to Intel CPU, the i9-12900KS and the i9-12900KF.
+The difference between these two CPUs is that the KS has an unlocked multiplier, allowing it to be overclocked.
 As a result, the basic consumption is less than the KF.
@@
 In order to do this, we will elaborate a benchmarking process and a set of tools that are intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
-Thus, we will begin by examining the state of the art of empirical analysis and retrieving the best empirical experimentation methodologies in the research field.Then, we will narrow these practices down to computer science so that we can finally adapt them to energy consumption.
+Thus, we will begin by examining the state of the art of empirical analysis and retrieving the best empirical experimentation methodologies in the research field.
+Then, we will narrow these practices down to computer science so that we can finally adapt them to energy consumption.
 
@@
 
-Experiment reproducibility is frequently listed as one of the most difficult challenges faced by researchers.Reproducing an experiment has been one of the major tools science has used to help establish the validity and importance of scientific findings since the Philosophical Transactions of the Royal Society were established in 1665~\cite{hankins1986debate}.
+Experiment reproducibility is frequently listed as one of the most difficult challenges faced by researchers.
+Reproducing an experiment has been one of the major tools science has used to help establish the validity and importance of scientific findings since the Philosophical Transactions of the Royal Society were established in 1665~\cite{hankins1986debate}.
 In point of fact, many of the outcomes are not duplicable,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
@@
 In biology, chimestry, and electronics, they use clean rooms, which are environments where pollutants like dust, airborne microbes, and aerosol particles are filtered out and factors like humidity, air flow, and temperature can be regulated.
-As for empirical analysis, the accuracy can be measured by numeric metrics such as the variance,the standard deviation (STD), and the interval inter quartile (IRQ).
+As for empirical analysis, the accuracy can be measured by numeric metrics such as the variance, the standard deviation (STD), and the interval inter quartile (IRQ).
 Section~\ref{sec:energy_variation} will go over the accuracy in the subject of energy optimisation.
@@
 In Social sciences, this can be achieved by selecting a representative sample size.
-\citeauthor{omair2014sample} presents a guideline on to achieve a such representativeness.
+\citeauthor{omair2014sample}~\cite{omair2014sample} presents a guideline on to achieve a such representativeness.
 As for computer science.
 The field of benchmarking is still in its infancy, and there is no consensus on how to achieve representativeness.
-howerver many attemps have been made to provide a set of benchmarks for specific purpose.
+Howerver many attemps have been made to provide a set of benchmarks for specific purpose.
 For example, the \href{https://www.spec.org/}{Standard Performance Evaluation Corporation} (SPEC) provides a set of benchmarks for CPU performance evaluation.
-this sets covers a wide range of use cases such as  CPU17 for testing the CPU \footnote{\url{https://www.spec.org/cpu2017/}}, SPECviewperf\footnote{\url{https://gwpg.spec.org/benchmarks/benchmark/specviewperf-2020-v3-1/}} for Graphic usage and
+This sets covers a wide range of use cases such as  CPU17 for testing the CPU \footnote{\url{https://www.spec.org/cpu2017/}}, SPECviewperf\footnote{\url{https://gwpg.spec.org/benchmarks/benchmark/specviewperf-2020-v3-1/}} for Graphic usage and
 one can cite stressNg when it comes to benchmark the hardware compoment, the SPEC benchmark when it comes to benchmark the software performance,and SPECPower\footnote{\url{https://www.spec.org/power_ssj2008} }.
@@
 The benchmarks are designed to be small, self-contained, and easy to implement in any language.
-The benchmarks are also ,made in a way to represent of most of the typical real-world workloads in an isolated manner.
+The benchmarks are also made in a way to represent of most of the typical real-world workloads in an isolated manner.
 Dacapo~\cite{blackburn2006dacapo} and renaissance~\cite{renaissance} are another example of a benchmarking set that are created to represent the Java Virtual Machine (JVM) performance.
@@
 %further references 
-First, we start with the work of \citeauthor{mytkowicz2009producing} ,where they made an evaluation  of 133 studies from ASPLOS, PACT, PLDI, and CGO, to find out that none of the experimental findings papers appropriately considered measurement bias.
+First, we start with the work of \citeauthor{mytkowicz2009producing}\cite{mytkowicz2009producing},where they made an evaluation  of 133 studies from ASPLOS, PACT, PLDI, and CGO, to find out that none of the experimental findings papers appropriately considered measurement bias.
 Which can lead derive incorrect results from an experiment if a seemingly insignificant feature of the experimental design is altered.
 They treated this problem by proposing two strategies for detecting measurement bias by using causal analysis and preventing it with setup randomization \cite{mytkowicz2009producing}.
-another study that was published in the book "Measuring computer performance: a practitioner's guide"\cite{lilja2005measuring},\citeauthor{lilja2005measuring} examinated performance indicators and gave in-depth treatment of benchmark program tactics.
-they provided clear explanations of the basic statistical methods required for interpreting measured performance data.
-they also outlined the overall 'design of experiments' method and demonstrated how to collect the most information with the least amount of work.
+another study that was published in the book "Measuring computer performance: a practitioner's guide"\cite{lilja2005measuring}, \citeauthor{lilja2005measuring} examinated performance indicators and gave in-depth treatment of benchmark program tactics.
+They provided clear explanations of the basic statistical methods required for interpreting measured performance data.
+They also outlined the overall "design of experiments" method and demonstrated how to collect the most information with the least amount of work.
 This practical book will appeal to anybody seeking a comprehensive, yet intuitive, grasp of computer system performance analysis.
@@
 
-In their paper\cite{stodden2018empirical}, the authors found that, even though it's a big step in the right direction, journal policies that require authors to give back digital scholarly objects after publication, like the data and code that back up the claims, don't get more than half of these objects back.
+In their paper~\cite{stodden2018empirical}, the authors found that, even though it's a big step in the right direction, journal policies that require authors to give back digital scholarly objects after publication, like the data and code that back up the claims, don't get more than half of these objects back.
 Then, using these artifacts, about a fourth of the published computational claims in the study could be made.
@@
 \section{Energy Measurement}\label{sec:soa_energymeasurement}
-Now that we have discussed the importance of benchmarking and the different approaches that can be used to evaluate the performance of a system, we will focus on energy measurement, which will be the main metric in this thesis.Therefore,in this section, we will discuss the different approaches that can be used to measure the energy consumption of a system.
-Many studies have been conducted to estimate a such energy consumption.
+Now that we have discussed the importance of benchmarking and the different approaches that can be used to evaluate the performance of a system, we will focus on energy measurement, which will be the main metric in this thesis.
+Therefore, in this section, we will discuss the different approaches that can be used to measure the energy consumption of a system.
+Many studies have been conducted to estimate such energy consumption.
 that varies from static analysis of the source code to infer its energy consumption like \citeauthor{pereira_helping_2017} where they provided a tool to highlight the most energy consuming parts of the code \cite{pereira_helping_2017}.
-The key advantage of this approach is that it allows you to estimate a program's energy usage without executing it.
-Unlike program complexity, energy consumption is strongly dependent to the execution environment.
+The key advantage of this approach is that it allows you to estimate a program's energy usage without actually executing it.
+Unlike program complexity, energy consumption is strongly dependent on the execution environment.
 As a result, static analysis may not accurately represent the behavior of the same program when run in a production context.
@@
 dedicated one , they are based on the instrumentation of some parts of the computer , however they have a high cost and they are defficult to scale since they require an instrumentation of the execusion environment.
-according to \citeauthor{hackenberg2014hdeem} there are four main criterions to evaluate an energy measurement tool \cite{hackenberg2014hdeem}
+According to \citeauthor{hackenberg2014hdeem} there are four main criterions to evaluate an energy measurement tool \cite{hackenberg2014hdeem}
 \begin{itemize}
-    \item \em{Spatial granularity}: the more specific the target of monitoring we are able to measure, the more efficient we can do optimization since we will know what causes the pitfalls of the energy consumption.
-    \item \em{Temporal granularity}: same as spatial granularity, temporal granularity helps us to identify the sequence of code that need to be optimized.
-    \item \em{Scalability}: this is mainly related to the cost of the tools and the ease of their integration for our system.
-    \item  \em{Accuracy}: to eliminate extra hazards and get more representative measurement.
+    \item \emph{Spatial granularity}: the more specific the target of monitoring we are able to measure, the more efficient we can do optimization since we will know what causes the pitfalls of the energy consumption.
+    \item \emph{Temporal granularity}: same as spatial granularity, temporal granularity helps us to identify the sequence of code that need to be optimized.
+    \item \emph{Scalability}: this is mainly related to the cost of the tools and the ease of their integration for our system.
+    \item  \emph{Accuracy}: to eliminate extra hazards and get more representative measurement.
 \end{itemize}
 
-We believe that accuracy can be extracted from those criterions since it is a result of the combination of the two first ones. therefore, we will focus more on the first 3 criterions from later on.
-below, we will discuss some of the well known tools that are used in literature.
+We believe that accuracy can be extracted from those criteria since it is a result of the combination of the two first ones.
+Therefore, we will focus more on the first 3 criteria later on.
+Below, we will discuss some of the well known tools that are used in literature.
 \subsection{Hardware Tools}
-Nowadays, most of the high performance computing systems (HPC) implements a tool to report the energy consumption of the nodes for the sake of monitoring and administration.
-Those tools are mainly integrated withing the power supply units (PSU) or the power distribution units (PDU).
+Nowadays, most high-performance computing systems (HPC) implement a tool to report the energy consumption of the nodes for the sake of monitoring and administration.
+Those tools are mainly integrated with the power supply units (PSU) or power distribution units (PDU).
 Then, they provide an interface and a log to follow the history of the energy consumption.
-Despite their scalability and ease of integration.
-Such tools lack both in spatial and temporal granularity since they monitor the whole energy of the nodes, and most of the times they have a very low sampling frequency.
+despite their scalability and ease of integration.
+Such tools lack both spatial and temporal granularity since they monitor the whole energy of the nodes and, most of the time, have a very low sampling frequency.
 Most of those tools are provided directly by the manufacturers.
-Such as \em{IBM EnergyScale technology}~\cite{mccreary2007energyscale,caldeira2014ibm,caldeiraibm} or \em{Dell poweredge}~\cite{lovicott2009thermal}, MEGware Cluststafe~\cite{breitbart2015case}.
+Such as \emph{IBM EnergyScale technology}~\cite{mccreary2007energyscale,caldeira2014ibm,caldeiraibm} or \emph{Dell poweredge}~\cite{lovicott2009thermal}, MEGware Cluststafe~\cite{breitbart2015case}.
 As said earlier, the true purpose of those tools is more monitoring than analyzing the energy consumption.
@@
 
-despite his high temporal granularity, wattsup pro lacks in term of spatial granularity since it monitors the energy consumption of the whole system.
+Despite his high temporal granularity, wattsup pro lacks in term of spatial granularity since it monitors the energy consumption of the whole system.
 To have a finer granularity we need to isolate the energy consumption of each component.
@@
 
-\subsection{Mitigating Energy Variations.}
+\subsection{Mitigating Energy Variations}
 Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
@@
 
-\section{Energy optimisation}\label{section:soa_energyoptimization},
+\section{Energy optimisation}\label{section:soa_energyoptimization}
 
@@
 \subsection{Energy optimization in the conception phase}
+
 The first part of this section will focus on energy optimisation in the conception phase, where the goal is to make the final product use less energy by choosing the best set of programming languages, tools, libraries, etc.
@@
 
-We start with the work of \citeauthor{pereira_energy_2017} where they  did an energy consumption comparison analysis of the most popular programming languages.
-Using the pareto optimum, the paper  provides recommendations on how to combine some of these languages to enhance code quality while taking into account execution time, memory utilization, and energy consumption using the pareto optimum.
+We start with the work of \citeauthor{pereira_energy_2017} \cite{pereira_energy_2017} where they  did an energy consumption comparison analysis of the most popular programming languages.
+Using the Pareto optimum~\cite{hochman1969pareto}, the paper  provides recommendations on how to combine some of these languages to enhance code quality while taking into account execution time, memory utilization, and energy consumption.
 Some of the study's findings demonstrate that interpreted languages, such as Python, have lower energy efficiency than compiled languages, such as C or Rust.
@@
 
-In this paper, we define a ranking of energy efficiency in programming languages.
-We considered a set of computing problems implemented in ten well-known programming languages and monitored the energy consumed when executing each language.
-Our preliminary results show that although the fastest languages tend to be the lowest energy-consuming ones, there are other interesting cases where slower languages are more energy-efficient than faster ones.
-
-\citeauthor{couto2017towards} investigated the influence of programming language choice on the energy consumption of software during execution.In their paper~\cite{couto2017towards}, they examined a set of computing problems written in ten well-known programming languages while observing the energy required when running each language.
+\citeauthor{couto2017towards} ~\cite{couto2017towards}investigated the influence of programming language choice on the energy consumption of software during execution.
+In their paper~\cite{couto2017towards}, they examined a set of computing problems written in ten well-known programming languages while observing the energy required when running each language.
 They also found that there are interesting situations in which slower languages use less energy than faster ones, even though fast languages are usually the ones that use the least energy.
@@
 
-\citeauthor{hasan_energy_2016} examinind the energy consumption of multiple Java data structures, analyzing the bytecode and evaluating the change in energy consumption in various circumstances(research,insertion, deletion, etc.).
-They also simulated best- and worst-case energy usage scenarios in real-world production systems by replacing the LinkedList and ArrayList, and discovered that incorrect collection can result in a 300\% increase in energy consumption\cite{hasan_energy_2016}.
+\citeauthor{hasan_energy_2016}~\cite{hasan_energy_2016} examinind the energy consumption of multiple Java data structures, analyzing the bytecode and evaluating the change in energy consumption in various circumstances(research,insertion, deletion, etc.).
+They also simulated best- and worst-case energy usage scenarios in real-world production systems by replacing the LinkedList and ArrayList, and discovered that incorrect collection can result in a 300\% increase in energy consumption.
 
@@
 
-We begin with Aequitas\cite{ribic2016aequitas}, a system that allows parallel applications to live on power domains that are co-managed (sharing the same CPU).
+We begin with Aequitas~\cite{ribic2016aequitas}, a system that allows parallel applications to live on power domains that are co-managed (sharing the same CPU).
 The technology is founded on the premise that coexisting programs can regard power-management hardware as a shared resource and collaborate on power management decisions.
@@
 
-As for virtual machines, \citeauthor{kurpicz2016much}.
-analyzed the total energy consumption of a VM in a data center while emphasizing the static cost versus the dynamic one.In this paper~\cite{kurpicz2016much}, the authros introduced the transparent, reproducible, and predictive cost calculator model EPAVE for VM-based environments.
+As for virtual machines, \citeauthor{kurpicz2016much}~\cite{kurpicz2016much}.
+analyzed the total energy consumption of a VM in a data center while emphasizing the static cost versus the dynamic one.
+In this paper , the authros introduced the transparent, reproducible, and predictive cost calculator model EPAVE for VM-based environments.
 The purpose of EPAVE is to provide the static cost of each VM on the server, which includes the air conditioning, power distribution, and the dynamic cost that is related to the VM activities.
 
-The energy usage of virtual machine allocation and task placement has also been investigated\cite{mishra_energy_2018}.In this article, The authors propose a method for mapping workloads to virtual machines and virtual machines to the physical ones (PMs) in an energy-efficient manner.
+The energy usage of virtual machine allocation and task placement has also been investigated~\cite{mishra_energy_2018}.
+In this article, The authors propose a method for mapping workloads to virtual machines and virtual machines to the physical ones (PMs) in an energy-efficient manner.
 To solve the problem of high heterogeneity of activities and resources, the jobs are categorised based on their resource requirements, and then the relevant VM is found, followed by the appropriate PM where the selected VM can be deployed Using a cloud simulator, the authors claimed The suggested technique saves energy by reducing the number of active PMs while also minimizing the makespan and task rejection rate.
 
-Besides the impact of the orchestration of virtual machine, some works have been targeting the runtime of specific programming languages.Using the TPC-DS benchmark  , \citeauthor{chiba2018towards} investigated the influence of HOTSPOT \footnote\url{https://openjdk.org/groups/hotspot/} and J9\footnote{\url{https://www.eclipse.org/openj9/}} on the performance of SQL-on-Hadoop systems (SPARK and TEZ) to reveal a three-fold disadvantage that one JVM can have over the other\cite{chiba2018towards}.
-\citeauthor{oi_power_2011} also compared the performance of HOTSPOT and J9.
-They demonstrated in their research~\cite{oi_power_2011} that the relative performance of a JVM is affected by the workload.
+Besides the impact of the orchestration of virtual machine, some works have been targeting the runtime of specific programming languages.Using the TPC-DS benchmark  , \citeauthor{chiba2018towards}~\cite{chiba2018towards} investigated the influence of HOTSPOT \footnote{\url{https://openjdk.org/groups/hotspot}} and J9\footnote{\url{https://www.eclipse.org/openj9}} on the performance of SQL-on-Hadoop systems (SPARK and TEZ) to reveal a three-fold disadvantage that one JVM can have over the other.
+\citeauthor{oi_power_2011}~\cite{oi_power_2011} also compared the performance of HOTSPOT and J9.
+They demonstrated in their research that the relative performance of a JVM is affected by the workload.
 They discovered that HOTSPOT's performance ranged from 44\% to 289\% of J9, while its dynamic power consumption ranged from 2.7W to 7.2W, using the SPECjvm2008~\cite{shiv2009specjvm2008} benchmarks.
 
-As for python, \citeauthor{redondo_comprehensive_2015}compared the performance and memory usage of various Python implementations (CPython, Jython, IronPython, PyPy, and so on) using a 215 set of benchmarks to discover that Python2 performed better with short applications, while Python3 versions covered more tests due to compatibility and the fact that Python2 became obsolete.\cite{redondo_comprehensive_2015}.
+As for Python, \citeauthor{redondo_comprehensive_2015}~\cite{redondo_comprehensive_2015}compared the performance and memory usage of various Python implementations (CPython, Jython, IronPython, PyPy, and so on) using a 215 set of benchmarks to discover that Python2 performed better with short applications, while Python3 versions covered more tests due to compatibility and the fact that Python2 became obsolete.
 

### 2022-09-16 chapters/benchmarking/conclusion.tex
corrections : 2nd chapter

@@
 %% maybe add extension as a part of the conclusion for this chapter 
-\section{Benchmarking protocol to extend reproducibility}\label{sec:bench_extension}
+\section{Conclusion \note{Uncomplete }}
+% TODO : uncomplete 
+By the end of this study, we have gathered enough guidelines to make the tests more reproducible, accurate and extensible.
+We created a set of new tests named \textbf{energy tests} which are more similar to performance tests.
+Thanks to the work of two interns [Mamadou and Adrien] we created a CI/CD platform to measure the energy consumption of Java projects and we could track the evolution of the this energy across different stages of the project.
+In the figure below we see an example of this plugin.
+For more details please visit the gitlab repository\footnote{\url{https://gitlab.inria.fr/mamdiall/J-Joules}}~\footnote{\url{https://gitlab.inria.fr/mamdiall/jjoules-plugin}} .
 
 We propose to extend the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
-Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in section ~\ref{sec:benchmarking_reproducibility}, we propose an abstract model to describe an empirical experiment:
+Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in Section~\ref{sec:benchmarking_reproducibility}, we propose an abstract model to describe an empirical experiment:
 
 \emph{context}: the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor .
-An \emph{orchestrator} which its job to run the experiment.
+An \emph{orchestrator} : which its job to run the experiment.
 The orchestrator provides three interfaces:
 
-\emph{Workload interface}: provides a set of functions that should be implemented by the workload  and called by the each candidate in the experiment.
-Observer interface: provides a set of metrics that are collected by the orchestrator during the expermentation.
-note that some of those metrics are external, such the energy consumption of the system , execution time, execution stack ..etc .
+\emph{Workload interface}: provides a set of functions that should be implemented by the workload and called by each candidate in the experiment.
+The Observer interface provides a set of metrics that are collected by the orchestrator during the experiment.
+Note that some of those metrics are external, such as the energy consumption of the system , execution time, execution stack, etc .
 \begin{figure}[!htb]
@@
 
-By the end of this study, we have gathered enough guidelines to make the tests more reproducible, accurate and extensible.
-We created a set of new tests named \textbf{energy tests} which are more similar to performance tests.
-Thanks to the work of two interns [Mamadou and Adrien] we created a CI/CD platform to measure the energy consumption of Java projects and we could track the evolution of the this energy across different stages of the project.
-In the figure below we see an example of this plugin.
-For more details please visit the gitlab repository\footnote{\url{https://gitlab.inria.fr/mamdiall/J-Joules}}~\footnote{\url{https://gitlab.inria.fr/mamdiall/jjoules-plugin}} .
-
 % \begin{figure}%[!htb]
@@
 % \end{figure}
-\section{Conclusion \note{missing}}

### 2022-09-16 chapters/literature/draft_citations.tex
Fixing typos

--- 2022-09-15 chapters/literature/draft_citations.tex
@@
 As for empirical analysis, the accuracy can be measured by numeric metrics such as the variance,the standard deviation (STD), and the interval inter quartile (IRQ).
-Section~\ref{sec:energy_variation} will go over the accuracy in the subject of energy optimisation.
+Section~\ref{sec:energy_variation} will go over the accuracy in the subject of energy optimization.
 
@@
 
-\section{Energy Variation}\label{sec:energy_variation}
+\section{Energy Variations}\label{sec:energy_variation}
 In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of physical measurements.
@@
 The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
-Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
+Acun~\emph{et~al.}~\cite{acun_variation_2016} found a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
 They also proposed some guidelines to reduce this variation by replacing the old---slower---chips, by load balancing the workload on the CPU cores and leaving one core idle.
 They claimed that the variation between the processor cores is insignificant.
-In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.
+In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating for the uneven effects of power capping.
 
@@
 
-\section{Energy optimisation}\label{section:soa_energyoptimization},
+\section{Energy Optimizations}\label{section:soa_energyoptimization},
 
-We will now turn our focus to the energy optimisation challenge after considering the various ways for measuring energy consumption in computers and understanding the energy variation problem.
-Over the previous decade, there has been a considerable increase of interest in this field, with many papers proposing different approaches to reduce the energy consumption of software applications.
-This section will passthrough the main contributions in this field, and will focus on the two following parts.
+We will now turn our focus to the energy optimization challenge after considering the various ways of measuring energy consumption in computers and understanding the energy variation problem.
+Over the previous decade, there has been a considerable increase in interest in this field, with many papers proposing different approaches to reduce the energy consumption of software applications.
+This section will pass through the main contributions in this field, and will focus on the two following parts.
 
-\subsection{Energy optimization in the conception phase}
-The first part of this section will focus on energy optimisation in the conception phase, where the goal is to make the final product use less energy by choosing the best set of programming languages, tools, libraries, etc.
+\subsection{Energy Optimization in the Design Phase}
+The first part of this section will focus on energy optimizations in the conception phase, where the goal is to make the final product use less energy by choosing the best set of programming languages, tools, libraries, etc.
 It also includes all the work and optimizations that developers can do to the source code to make the software use less energy when it is running.
@@
 We start with the work of \citeauthor{pereira_energy_2017} where they did an energy consumption comparison analysis of the most popular programming languages.
-Using the pareto optimum, the paper  provides recommendations on how to combine some of these languages to enhance code quality while taking into account execution time, memory utilization, and energy consumption using the pareto optimum.
+Using the Pareto optimum, the paper provides recommendations on how to combine some of these languages to enhance code quality while taking into account execution time, memory utilization, and energy consumption using the Pareto optimum.
 Some of the study's findings demonstrate that interpreted languages, such as Python, have lower energy efficiency than compiled languages, such as C or Rust.
@@
 
-\citeauthor{couto2017towards} investigated the influence of programming language choice on the energy consumption of software during execution.In their paper~\cite{couto2017towards}, they examined a set of computing problems written in ten well-known programming languages while observing the energy required when running each language.
+\citeauthor{couto2017towards} investigated the influence of programming language choice on the energy consumption of software during execution.
+In their paper~\cite{couto2017towards}, they examined a set of computing problems written in ten well-known programming languages while observing the energy required when running each language.
 They also found that there are interesting situations in which slower languages use less energy than faster ones, even though fast languages are usually the ones that use the least energy.
@@
 They discovered that there is no statistically significant difference in CPU load between individual measurements( less than 2\%) 2.
-Howerver, Kotlin implementation has never earned the best results in any collection of measurements.
+However, Kotlin implementation has never earned the best results in any collection of measurements.
 
-Other works that have studied the impact of the website technology on the energy conusmptuon,\cite{philippot_characterization_2014,manotas_investigating_2013}.
+Other works that have studied the impact of website technology on energy consumption,\cite{philippot_characterization_2014,manotas_investigating_2013}.
 In their work~\cite{philippot_characterization_2014},\citeauthor{philippot_characterization_2014} measured the computer resources used during the loading of a website in a browser, such as memory utilization and energy consumption, for over 500 websites and proposed some best practices for developers.
@@
 [52] compared the energy consumption of several Java data structures, analyzing the bytecode using the Wala framework8 and assessing the evolution of the energy consumption in different scenarios (insertion at the beginning, iteration, etc.).
-They also used some automated replacement of LinkedList and ArrayList to simulate best-and worst-case energy consumption scenarios in real production applications.
+They also used some automated replacements of LinkedList and ArrayList to simulate best-and worst-case energy consumption scenarios in real production applications.
 Their study showed that using inappropriate collection can cause up to 300\% of energy consumption inefficiency.
@@
 
-\citeauthor{hasan_energy_2016} examinind the energy consumption of multiple Java data structures, analyzing the bytecode and evaluating the change in energy consumption in various circumstances(research,insertion, deletion, etc.).
+\citeauthor{hasan_energy_2016} examined the energy consumption of multiple Java data structures, analyzing the bytecode and evaluating the change in energy consumption in various circumstances(research, insertion, deletion, etc.).
 They also simulated best- and worst-case energy usage scenarios in real-world production systems by replacing the LinkedList and ArrayList, and discovered that incorrect collection can result in a 300\% increase in energy consumption\cite{hasan_energy_2016}.
@@
 Using a statistical spectrum-based analysis and JRAPL\cite{guimaraes2016some,liu_data-oriented_2015}, the tool locates energy-inefficient code fragments.
-According to the authors, it is language and context independent.
+According to the authors, it is language and context-independent.
 
-\subsection{Energy optimization in the execution phase}
-
-The second part of this section will focus on energy optimization in the execution phase, where the goal is to optimize this energy consumption for already developed software without making changes to the source code.The goal is to set up and create an environment in which software can run with the least amount of energy.
+\subsection{Energy Optimizations in the Execution Phase}
+The second part of this section will focus on energy optimization in the execution phase, where the goal is to optimize this energy consumption for already developed software without making changes to the source code.
+The goal is to set up and create an environment in which software can run with the least amount of energy.
 This could involve process scheduling, system tunning and so on.
@@
 The technology is founded on the premise that coexisting programs can regard power-management hardware as a shared resource and collaborate on power management decisions.
-As a result, it accomplishes its purpose by scheduling these applications with a time slicing technique ( as anexamample round robin).
+As a result, it accomplishes its purpose by scheduling these applications with a time slicing technique ( as an example round robin).
 The authors claim that their strategy achieves a 12.9\% improvement while incurring only a 2.5\% performance cost.
@@
 As for virtual machines, \citeauthor{kurpicz2016much}.
-analyzed the total energy consumption of a VM in a data center while emphasizing the static cost versus the dynamic one.In this paper~\cite{kurpicz2016much}, the authros introduced the transparent, reproducible, and predictive cost calculator model EPAVE for VM-based environments.
+analyzed the total energy consumption of a VM in a data center while emphasizing the static cost versus the dynamic one.
+In this paper~\cite{kurpicz2016much}, the authors introduced the transparent, reproducible, and predictive cost calculator model EPAVE for VM-based environments.
 The purpose of EPAVE is to provide the static cost of each VM on the server, which includes the air conditioning, power distribution, and the dynamic cost that is related to the VM activities.
@@
 The energy usage of virtual machine allocation and task placement has also been investigated\cite{mishra_energy_2018}.In this article, The authors propose a method for mapping workloads to virtual machines and virtual machines to the physical ones (PMs) in an energy-efficient manner.
-To solve the problem of high heterogeneity of activities and resources, the jobs are categorised based on their resource requirements, and then the relevant VM is found, followed by the appropriate PM where the selected VM can be deployed Using a cloud simulator, the authors claimed The suggested technique saves energy by reducing the number of active PMs while also minimizing the makespan and task rejection rate.
+To solve the problem of high heterogeneity of activities and resources, the jobs are categorized based on their resource requirements, and then the relevant VM is found, followed by the appropriate PM where the selected VM can be deployed Using a cloud simulator, the authors claimed The suggested technique saves energy by reducing the number of active PMs while also minimizing the makespan and task rejection rate.
 
-Besides the impact of the orchestration of virtual machine, some works have been targeting the runtime of specific programming languages.Using the TPC-DS benchmark  , \citeauthor{chiba2018towards} investigated the influence of HOTSPOT \footnote\url{https://openjdk.org/groups/hotspot/} and J9\footnote{\url{https://www.eclipse.org/openj9/}} on the performance of SQL-on-Hadoop systems (SPARK and TEZ) to reveal a three-fold disadvantage that one JVM can have over the other\cite{chiba2018towards}.
+Besides the impact of the orchestration of virtual machine, some works have been targeting the runtime of specific programming languages.
+Using the TPC-DS benchmark, \citeauthor{chiba2018towards} investigated the influence of HOTSPOT \footnote\url{https://openjdk.org/groups/hotspot/} and J9\footnote{\url{https://www.eclipse.org/openj9/}} on the performance of SQL-on-Hadoop systems (SPARK and TEZ) to reveal a three-fold disadvantage that one JVM can have over the other\cite{chiba2018towards}.
 \citeauthor{oi_power_2011} also compared the performance of HOTSPOT and J9.

### 2022-09-17 chapters/literature/draft_citations.tex
Fix intro

@@
 F) for homes, cars, and electronic products, to provide the consumer an indication of the energy consumption of their devices, which will reflect on their power bill.
+
 This criterion is extended even to the hardware components of a computer.
-Figure~\ref{fig:soa_comparaisoncpu}\footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}} Presents a comparaison between to Intel CPU, the i9-12900KS and the i9-12900KF.The difference between these two CPUs is that the KS has an unlocked multiplier, allowing it to be overclocked.
+Figure~\ref{fig:soa_comparaisoncpu}\footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}} compares Intel CPUs  i9-12900KS and i9-12900KF.
+The difference between these two CPUs is that the KS has an unlocked multiplier, allowing it to be overclocked.
 As a result, the basic consumption is less than the KF.
@@
     \includegraphics[width=\linewidth]{imgs/cpu_cost_comparaison}
-    \caption{Electrical cost comparasion between two cpus }
+    \caption{Electrical cost comparison between two CPUs.}
     \label{fig:soa_comparaisoncpu}

### 2022-09-17 chapters/benchmarking/draft.tex

@@
 Addressing all of them becomes significantly more difficult in studies that compare multiple treatments.
-
-Furthermore, between the exploratory experiment and the published results, there may have been the appearance of new candidates as well as the evolution of others.
-As a result, it is not only necessary to ensure the "reproducibility" of the results, but it is also essential to provide room for new candidates.
-As a result, we'd like to present "Extension", a new approach for leveraging experimentation.
-The ability to give not only the tools needed to replicate the tests but also to integrate more candidates, workloads, or important key performance indicators
+Also, between the exploratory experiment and when the results were made public, new candidates may have come up and others may have changed.
+So, not only is it important to make sure the results can be "reproduced," but it's also important to make room for new candidates.
 This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
@@
 Finally, Section~\ref{sec:bench_extension} presents tools for introducing new candidates, workloads, or other key performance indicators.
-Regarding the "representativeness" of the benchmarks, we will elaborate in the section entitled Perspectives.
+% Regarding the "representativeness" of the benchmarks, we will elaborate in the section entitled Perspectives.
 %% this is might be the part that treats reproducibility for my case
@@
 \subsection{Docker \& Energy}
-Now that we have decided to use container technology to enclose our tests, what effect will this have on the amount of energy our tests use?
+Now that we have decided to use container technology to enclose our tests, what effect will this have on the amount of energy consumed by our tests ?
 %%%%%%%%%
 
-Using research from \cite{eddie_antonio_santos_how} who examined how adding the Docker layer affected energy consumption.
-\citeauthor{eddie_antonio_santos_how} conducted their experiment by running numerous benchmarks both with and without Ddocker.
+Using research from \cite{eddie_antonio_santos_how} who examined how adding the Docker layer affected energy consumption, \citeauthor{eddie_antonio_santos_how} conducted their experiment by running numerous benchmarks both with and without Docker.
 They contrasted the energy usage and execution time that resulted.
@@
 
-Yet, as seen in Figure~\ref{fig:docker_reddis}, Docker increased the execution duration of the benchmark by 50 seconds,which led to a significant rise in the energy usage.
+Yet, as seen in Figure~\ref{fig:docker_reddis}, Docker increased the execution duration of the benchmark by 50 seconds, which led to a significant rise in the energy usage.
 According to the authors, this overhead is primarily caused by the Docker daemon and not by the fact that  the application is in a container.

### 2022-09-17 chapters/literature/draft_citations.tex
Proofreading

@@
 In the majority of fields, the purpose is to minimize the energy consumption of electrical devices and components.
-Modern times even see energy classification (A, B,...
-F) for homes, cars, and electronic products, to provide the consumer an indication of the energy consumption of their devices, which will reflect on their power bill.
+Modern times even see energy classification (A, B\dots F) for homes, cars, and electronic products, to provide the consumer an indication of the energy consumption of their devices, which will reflect on their power bill.
 
 This criterion is extended even to the hardware components of a computer.
-Figure~\ref{fig:soa_comparaisoncpu}\footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}} compares Intel CPUs  i9-12900KS and i9-12900KF.
+Figure~\ref{fig:soa_comparaisoncpu} compares Intel CPUs i9-12900KS and i9-12900KF.\footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}}
 The difference between these two CPUs is that the KS has an unlocked multiplier, allowing it to be overclocked.
@@
 
-As an example \citeauthor{avgerinou_trends_2017} evaluated the development of power use effectiveness (PUE) in data centers thkat belongs to various organizations participating in the European code of conduct for energy efficiency program \cite{avgerinou_trends_2017}.
+As an example, \citeauthor{avgerinou_trends_2017} evaluated the development of power use effectiveness (PUE) in data centers that belongs to various organizations participating in the European code of conduct for energy efficiency program~\cite{avgerinou_trends_2017}.
 The research found a gradual decline in the PUE of data centers, which measures the ratio of the overall energy supplied to the energy used by IT equipment.
@@
 
-We will place a greater emphasis on the software level in order to decrease the amount of energy that is used, more particularly on the execution phase of the program cycle.
+We will place a greater emphasis on the software level to decrease the amount of energy that is used, more particularly in the execution phase of the program cycle.
 We will be proceeding through an empirical analysis of the energy consumption of the software while changing some components of the source code without impacting its behavior.
-In order to do this, we will elaborate a benchmarking process and a set of tools that are intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
-Thus, we will begin by examining the state of the art of empirical analysis and retrieving the best empirical experimentation methodologies in the research field.
+To do this, we will elaborate on a benchmarking process and a set of tools that are intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
+Thus, we will begin by examining the state of the art in empirical analysis and retrieving the best empirical experimentation methodologies in the research field.
 Then, we will narrow these practices down to computer science so that we can finally adapt them to energy consumption.
@@
 It provides examples of hardware and software measuring instruments and describes their differences, benefits, and drawbacks.
-It also examines the sources of energy measurement variations, which represent a significant obstacle to achieving precise readings and a higher accuracy.
-Then, in section \ref{section:soa_energyoptimization}, we will go through some of the previous work on improving the energy consumption of softwares.
+It also examines the sources of energy measurement variations, which represent a significant obstacle to achieving precise readings and higher accuracy.
+Then, in section \ref{section:soa_energyoptimization}, we will go through some of the previous work on improving the energy consumption of software.
 
@@
 A successful benchmark must meet three criteria.
-First, it must be \textbf{reproducible} in order for others to imitate it.
-Second, the findings should be \textbf{accurate},which implies that we should expect the same results each time we run the benchmark.
+First, it must be \textbf{reproducible} for others to imitate it.
+Second, the findings should be \textbf{accurate}, which implies that we should expect the same results each time we run the benchmark.
 Finally, it should \textbf{represent} reality.
@@
 
-Experiment reproducibility is frequently listed as one of the most difficult challenges faced by researchers.Reproducing an experiment has been one of the major tools science has used to help establish the validity and importance of scientific findings since the Philosophical Transactions of the Royal Society were established in 1665~\cite{hankins1986debate}.
-In point of fact, many of the outcomes are not duplicable,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
+Experiment reproducibility is frequently listed as one of the most difficult challenges faced by researchers.
+Reproducing an experiment has been one of the major tools science has used to help establish the validity and importance of scientific findings since the Philosophical Transactions of the Royal Society were established in 1665~\cite{hankins1986debate}.
+Many of the outcomes are not reproducible,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
 As a result of the crisis affecting the majority of empirical studies, most reviews now include reproducibility as a minimum standard for judging scientific merit~\cite{peng2011reproducible}.
@@
 In our case, this means the ability to run the benchmark multiple times with low variation. This can be achieved by controlling the experiment environment, allowing less room for random factors.
-In biology, chimestry, and electronics, they use clean rooms, which are environments where pollutants like dust, airborne microbes, and aerosol particles are filtered out and factors like humidity, air flow, and temperature can be regulated.
-As for empirical analysis, the accuracy can be measured by numeric metrics such as the variance,the standard deviation (STD), and the interval inter quartile (IRQ).
+In biology, chemistry, and electronics, they use clean rooms, which are environments where pollutants like dust, airborne microbes, and aerosol particles are filtered out and factors like humidity, airflow, and temperature can be regulated.
+As for empirical analysis, the accuracy can be measured by numeric metrics such as the variance, the standard deviation (STD), and the interval inter quartile (IRQ).
 Section~\ref{sec:energy_variation} will go over the accuracy in the subject of energy optimization.
@@
 this sets covers a wide range of use cases such as  CPU17 for testing the CPU \footnote{\url{https://www.spec.org/cpu2017/}}, SPECviewperf\footnote{\url{https://gwpg.spec.org/benchmarks/benchmark/specviewperf-2020-v3-1/}} for Graphic usage and
-one can cite stressNg when it comes to benchmark the hardware compoment, the SPEC benchmark when it comes to benchmark the software performance,and SPECPower\footnote{\url{https://www.spec.org/power_ssj2008} }.
-NASA Parallel Benchmarks (NPB~\footnote{\url{https://www.nas.nasa.gov/publications/npb.html}} )
-and HPCchallenge \footnote{\url{https://www.hpcchallenge.org/}} are two other examples of benchmarking sets that are created to represent the high performance computing.
+one can cite StressNg when it comes to benchmark the hardware components, the SPEC benchmark when it comes to benchmark the software performance, and SPECPower\footnote{\url{https://www.spec.org/power_ssj2008} }.
+NASA Parallel Benchmarks (NPB)\footnote{\url{https://www.nas.nasa.gov/publications/npb.html}} and HPCchallenge\footnote{\url{https://www.hpcchallenge.org/}} are two other examples of benchmarking sets that are created to represent the high performance computing.
 As for programming languages we can cite Computer Language Benchmarks Game~\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}, which is a collection of benchmarks for various programming languages.
 The benchmarks are designed to be small, self-contained, and easy to implement in any language.
-The benchmarks are also ,made in a way to represent of most of the typical real-world workloads in an isolated manner.
-Dacapo~\cite{blackburn2006dacapo} and renaissance~\cite{renaissance} are another example of a benchmarking set that are created to represent the Java Virtual Machine (JVM) performance.
+The benchmarks are also made in a way to represent of most of the typical real-world workloads in an isolated manner.
+Dacapo~\cite{blackburn2006dacapo} and renaissance~\cite{renaissance} are other examples of benchmarking sets that are created to represent the Java Virtual Machine (JVM) performance.
 On the other hand, a new sort of test has arisen within the software development life cycle.
 This type is known as stress testing, and it is used to assess the software's robustness and reliability before releasing it to the public.
-gatling~\footnote{\url{https://gatling.io}} and TCPCopy\footnote{\url{https://github.com/session-replay-tools/tcpcopy}} are great  examples of a stress testing tools that are used to test the performance server applications.
+gatling~\footnote{\url{https://gatling.io}} and TCPCopy\footnote{\url{https://github.com/session-replay-tools/tcpcopy}} are great examples of stress testing tools that are used to test the performance of server applications.
 
@@
 %further references 
-First, we start with the work of \citeauthor{mytkowicz2009producing} ,where they made an evaluation  of 133 studies from ASPLOS, PACT, PLDI, and CGO, to find out that none of the experimental findings papers appropriately considered measurement bias.
+First, we start with the work of \citeauthor{mytkowicz2009producing}, where they evaluated 133 studies from ASPLOS, PACT, PLDI, and CGO, to find out that none of the experimental findings papers appropriately considered measurement bias.
 Which can lead derive incorrect results from an experiment if a seemingly insignificant feature of the experimental design is altered.
 They treated this problem by proposing two strategies for detecting measurement bias by using causal analysis and preventing it with setup randomization \cite{mytkowicz2009producing}.
-another study that was published in the book "Measuring computer performance: a practitioner's guide"\cite{lilja2005measuring},\citeauthor{lilja2005measuring} examinated performance indicators and gave in-depth treatment of benchmark program tactics.
+another study that was published in the book "Measuring computer performance: a practitioner's guide"\cite{lilja2005measuring},\citeauthor{lilja2005measuring} examined performance indicators and gave in-depth treatment of benchmark program tactics.
 they provided clear explanations of the basic statistical methods required for interpreting measured performance data.
-they also outlined the overall 'design of experiments' method and demonstrated how to collect the most information with the least amount of work.
+They also outlined the overall design of the experimental method and demonstrated how to collect the most information with the least amount of work.
 This practical book will appeal to anybody seeking a comprehensive, yet intuitive, grasp of computer system performance analysis.
 
-\citeauthor{bukh1992art} wrote a book about computer performance analysis, where they discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems.
-The article \cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistake in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often seen errors.
+\citeauthor{bukh1992art} wrote a book about computer performance analysis, where they discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi-squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems.
+The article \cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistakes in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often seen errors.
 
-Another article~\cite{patil2018training} expands on this line of thought by evaluating the uncertainty caused by replications in new research.
+Another article~\cite{patil2018training} expands on this line of thought by evaluating the uncertainty caused by replications in the new research.
 They offered some strategies to capture uncertainty in inferential investigations, such as cross-study validation and ensemble models.
@@
 
-% as for computer sience. In their paper~\cite{buytaert_statistically_nodate}, the authors  separated the start-up measurement from the steady state when comparing different Java optimizations.
+% as for computer science. In their paper~\cite{buytaert_statistically_nodate}, the authors  separated the start-up measurement from the steady state when comparing different Java optimizations.
 \begin{figure}[!hbt]
@@
 Finally, to make unify the benchmarking methodology across different research works in the field of computer science, we can cite the paper~\cite{stephen_evaluate_2012}.
-In their approach they  divided any empirical experiment into four  components.
+In their approach, they divided any empirical experiment into four components.
 Figure \ref{fig:soa_expermiment_component} presents these components, which are:
@@
     \item \emph{metrics} that specify the attributes to be measured and how to assess them.
-    \item \emph{Data analysis} show how to examine data and evaluate the outcomes of the analysis to offer insight into the assertions that arise from the study.
+    \item \emph{Data analysis} shows how to examine data and evaluate the outcomes of the analysis to offer insight into the assertions that arise from the study.
 \end{enumerate}
 
-The work of this thesis will be based on this approach, since it provides a unified methodology for benchmarking and evaluating the performance of different systems.
+The work of this thesis will be based on this approach since it provides a unified methodology for benchmarking and evaluating the performance of different systems.
 
@@
 \section{Energy Measurement}\label{sec:soa_energymeasurement}
-Now that we have discussed the importance of benchmarking and the different approaches that can be used to evaluate the performance of a system, we will focus on energy measurement, which will be the main metric in this thesis.Therefore,in this section, we will discuss the different approaches that can be used to measure the energy consumption of a system.
-Many studies have been conducted to estimate a such energy consumption.
-that varies from static analysis of the source code to infer its energy consumption like \citeauthor{pereira_helping_2017} where they provided a tool to highlight the most energy consuming parts of the code \cite{pereira_helping_2017}.
+Now that we have discussed the importance of benchmarking and the different approaches that can be used to evaluate the performance of a system, we will focus on energy measurement, which will be the main metric in this thesis.
+Therefore, in this section, we will discuss the different approaches that can be used to measure the energy consumption of a system.
+Many studies have been conducted to estimate such energy consumption that varies from static analysis of the source code to infer its energy consumption like \citeauthor{pereira_helping_2017} where they provided a tool to highlight the most energy consuming parts of the code \cite{pereira_helping_2017}.
 The key advantage of this approach is that it allows you to estimate a program's energy usage without executing it.
-Unlike program complexity, energy consumption is strongly dependent to the execution environment.
+Unlike program complexity, energy consumption is strongly dependent on the execution environment.
 As a result, static analysis may not accurately represent the behavior of the same program when run in a production context.
@@
 dedicated one , they are based on the instrumentation of some parts of the computer , however they have a high cost and they are defficult to scale since they require an instrumentation of the execusion environment.
-according to \citeauthor{hackenberg2014hdeem} there are four main criterions to evaluate an energy measurement tool \cite{hackenberg2014hdeem}
+According to \citeauthor{hackenberg2014hdeem}, there are four main criteria to evaluate an energy measurement tool \cite{hackenberg2014hdeem}:
 \begin{itemize}
-    \item \em{Spatial granularity}: the more specific the target of monitoring we are able to measure, the more efficient we can do optimization since we will know what causes the pitfalls of the energy consumption.
-    \item \em{Temporal granularity}: same as spatial granularity, temporal granularity helps us to identify the sequence of code that need to be optimized.
-    \item \em{Scalability}: this is mainly related to the cost of the tools and the ease of their integration for our system.
-    \item  \em{Accuracy}: to eliminate extra hazards and get more representative measurement.
+    \item \em{Spatial granularity}: the more specific the target of monitoring we can measure, the more efficient we can do optimization since we will know what causes the pitfalls of the energy consumption,
+    \item \em{Temporal granularity}: same as spatial granularity, temporal granularity helps us to identify the sequence of code that need to be optimized,
+    \item \em{Scalability}: this is mainly related to the cost of the tools and the ease of their integration for our system,
+    \item  \em{Accuracy}: to eliminate extra hazards and get a more representative measurement.
 \end{itemize}
 
-We believe that accuracy can be extracted from those criterions since it is a result of the combination of the two first ones. therefore, we will focus more on the first 3 criterions from later on.
-below, we will discuss some of the well known tools that are used in literature.
+We believe that accuracy can be extracted from those criteria since it is a result of the combination of the two first ones. therefore, we will focus more on the first 3 criteria from later on.
+Below, we will discuss some of the well-known tools that are used in literature.
+
 \subsection{Hardware Tools}
-Nowadays, most of the high performance computing systems (HPC) implements a tool to report the energy consumption of the nodes for the sake of monitoring and administration.
-Those tools are mainly integrated withing the power supply units (PSU) or the power distribution units (PDU).
-Then, they provide an interface and a log to follow the history of the energy consumption.
+Nowadays, most high-performance computing systems (HPC) implement a tool to report the energy consumption of the nodes for the sake of monitoring and administration.
+Those tools are mainly integrated within the power supply units (PSU) or the power distribution units (PDU).
+Then, they provide an interface and a log to follow the history of energy consumption.
 Despite their scalability and ease of integration.
-Such tools lack both in spatial and temporal granularity since they monitor the whole energy of the nodes, and most of the times they have a very low sampling frequency.
+Such tools lack both spatial and temporal granularity since they monitor the whole energy of the nodes, and most of the time they have a very low sampling frequency.
 Most of those tools are provided directly by the manufacturers.
 Such as \em{IBM EnergyScale technology}~\cite{mccreary2007energyscale,caldeira2014ibm,caldeiraibm} or \em{Dell poweredge}~\cite{lovicott2009thermal}, MEGware Cluststafe~\cite{breitbart2015case}.
-As said earlier, the true purpose of those tools is more monitoring than analyzing the energy consumption.
+As said earlier, the true purpose of those tools is more monitoring than analyzing energy consumption.
 %% add something to link with wattsup pro 
 WattsUp Pro, is a device that can be installed between the power source of the machine and the system under test.
-It allows a sampling frequency up to 1\,Hz and has a internal memory to store a wide variety of data, such as the maximum voltage, current, that later can be exported via USB port for personal usage or lined to some graph programs like Logger Pro or LabQuest.
+It allows a sampling frequency up to 1\,Hz and has an internal memory to store a wide variety of data, such as the maximum voltage, and current, that later can be exported via USB port for personal usage or lined to some graph programs like Logger Pro or LabQuest.
 The main advantage of this tool is the ability to monitor the energy consumption from a different device which will reduce the risk of interference with the energy consumption of the test % a revoir 
@@
 
-despite his high temporal granularity, wattsup pro lacks in term of spatial granularity since it monitors the energy consumption of the whole system.
+Despite its high temporal granularity, WattsUp Pro lacks in term of spatial granularity since it monitors the energy consumption of the whole system.
 To have a finer granularity we need to isolate the energy consumption of each component.
-For this, we will need more intrusive tools such.
+% For this, we will need more intrusive tools such.
 
@@
 
-PowerMon and its upgraded version powerMon2 \cite{bedard2010powermon} are based on an micro-controller chip that can monitor up to 6 channel (8 for powermon2), simultaneously.
+PowerMon and its upgraded version powerMon2 \cite{bedard2010powermon} are based on a micro-controller chip that can monitor up to 6 channels (8 for powermon2), simultaneously.
 Therefore, we are able to monitor the power consumption of 4 devices at the same time.
 The frequency sample of this tool is up to 50\,Hz, with an accuracy of 1.2\%.
-Powermore2 comes with smaller size that can fit within 3.5 inches rack drive.
+Powermore2 comes in a smaller size that can fit within 3.5 inches rack drive.
 
-PowerInsight~\cite{laros2013powerinsight} is another fine-grained measurement tool that is based on an ARM bagelbone processor~\cite{coley2012beaglebone}, which is able to measure up to 30 channel simultaneously with a frequency of 1KHz per channel.
+PowerInsight~\cite{laros2013powerinsight} is another fine-grained measurement tool that is based on an ARM BeagleBone processor~\cite{coley2012beaglebone}, which can measure up to 30 channels simultaneously with a frequency of 1KHz per channel.
 
 % On the other hand, GreenMiner \cite{hindle2014greenminer}, %%i dont think that  mobile measurement will be relevent for my case 
-powerpack~\cite{ge2009powerpack} in the other hand is an api that synchronizes the data gathered from other monitoring tools such as Watt’s Up Pro, NI and RadioShack pro and the lines of code.
+powerpack~\cite{ge2009powerpack} in the other hand is an API that synchronizes the data gathered from other monitoring tools such as Watt’s Up Pro, NI and RadioShack pro and the lines of code.
 %% maybe remove this one from the list because it is not well known.
 
-Other monitor tools have been realized by the manufactures, such as IBM Power executive~\cite{koomey2011growth}, which allows their customer to monitor the power consumption and thermal behavior of the of BladeCenter systems in the data center.
+Other monitor tools have been realized by the manufacturers, such as IBM Power executive~\cite{koomey2011growth}, which allows their customers to monitor the power consumption and thermal behavior of the of BladeCenter systems in the data center.
 
@@
 
-Figre~\ref{fig:rapl-domaine} shows the different scopes  thhat can be monitored using RAPL.
+Figure~\ref{fig:rapl-domaine} shows the different scopes that can be monitored using RAPL.
 The CPU package can be monitored for both server and desktop processors. However, DRAM is only available on server CPUs, whilst the integrated GPU is only available on desktop processors.
-The advantage of such an approach is that the absence of any intrusive measurement tools.
-Furthermore, they have a high temporal granularity with a sampling frequency up to 1\,KHz~\cite{ilsche_power_2015}.
+The advantage of such an approach is the absence of any intrusive measurement tools.
+Furthermore, they have a high temporal granularity with a sampling frequency of up to 1\,KHz~\cite{ilsche_power_2015}.
 
-With similar approach, we can find NVIDIA reporting tools, such as GPU TESLA \cite{burtscher2014measuring} and the NVML library~\cite{fahad2019comparative}.
+With a similar approach, we can find NVIDIA reporting tools, such as GPU TESLA \cite{burtscher2014measuring} and the NVML library~\cite{fahad2019comparative}.
 
@@
 Software-based measurement tools are based on other hardware tools to monitor energy use.
-Granularity is the core value of these technologies, unlike hardware tools, which only provide the total energy usage of the system/component (computer, server, motherboard, etc.) for most cases.
+Granularity is the core value of these technologies, unlike hardware tools, which only provide the total energy usage of the system/component (computer, server, motherboard, etc.) in most cases.
 Because they are frequently constructed on empirical estimations and data learning methods, they drop in accuracy.
@@
 Only the lightest sensors required for data collection and transmission are put into the monitored devices.
-The SmartWatts formula is then executed on the primary server in order to construct the model that permits assigning the energy usage for each operating control group.
+The SmartWatts formula is then executed on the primary server to construct the model that permits assigning the energy usage for each operating control group.
 PowerAPI is only compatible with Linux on a bare-metal physical computer.
@@
 This program uses power models to predict process energy usage.
-It uses CPU events which are passed from the measured node to a model generator node in order to construct the power model.
-It works by combining a description of the CPU with a list of the hardware events through multiple calibration phases in order to build a robust model.
+It uses CPU events that are passed from the measured node to a model generator node to construct the power model.
+It works by combining a description of the CPU with a list of the hardware events through multiple calibration phases to build a robust model.
 
 Joulemeter~\cite{kothari2009joulemeter,jagroep2017energy} is a Microsoft software that estimates the energy usage of Windows running applications down to the process level by using power models (for CPU, memory, and drives).
-It employs low-overhead power models to infer power consumption from resource utilization during runtime, and it provides power limiting features for virtual machines.
+It employs low-overhead power models to infer power consumption from resource utilization during runtime, and it provides power-limiting features for virtual machines.
 Previous Joulemeter tests~\cite{jagroep2015profiling} demonstrated that the instrument provides a less accurate estimation of energy use that differs greatly from the real one.
 To adjust its models to the hardware on which it operates, Joulemeter must first go through a calibration step.
-It is able to only monitor one process at a time with a frequency of 1\,Hz.
+It can only monitor one process at a time with a frequency of 1\,Hz.
 
 JRAPL is another example of an energy measurement tool estimating tool that has been utilized in a variety of publications~\cite{guimaraes2016some,liu_data-oriented_2015}.
-This software enables for the energy usage of Java programs, functions, or even a block of code lines to be profiled and measured.
+This software enables the energy usage of Java programs, functions, or even a block of code lines to be profiled and measured.
 The measurements are heavily reliant on the data supplied by RAPL.
@@
 Jalen~\cite{noureddine2015monitoring} is another tool that profiles and monitors the energy usage of a Java program.
-Unlike jRAPL, Jalen is able to cover the scope down to the function level.It gathers data using code instrumentation and statistical sampling at a predetermined pace.
+Unlike jRAPL, Jalen can cover the scope down to the function level.
+It gathers data using code instrumentation and statistical sampling at a predetermined pace.
 Because of the overhead that code instrumentation may incur, the authors recommend utilizing the second option.
-Every 10 ms, Jalen records the JVM's stack trace together with the CPU time of threads and computes statistics about method calls.These statistics are then utilized to calculate each method's energy usage.
+Every 10 ms, Jalen records the JVM's stack trace together with the CPU time of threads and computes statistics about method calls.
+These statistics are then utilized to calculate each method's energy usage.
 
@@
 \section{Energy Variations}\label{sec:energy_variation}
-In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should increase the accuracy of physical measurements.
+In theory, using an identical CPU, the same memory configuration, and similar storage and networking capabilities should increase the accuracy of physical measurements.
 Unfortunately, this is not possible when it comes to measuring the energy consumption of a system.
-Applying the benchmarking guidelines and repeating the same experiment with in the same configuration are not sufficient to reproduce the the same energy measurements, not only between identical machines, but even within the same machine.
+Applying the benchmarking guidelines and repeating the same experiment within the same configuration is not sufficient to reproduce the same energy measurements, not only between identical machines, but even within the same machine.
 This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
@@
 Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
-We can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.
+We can observe a large variation of the energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.
 
@@
 This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
-On the one hand, the correlation between the processor temperature and the energy consumption was one of the most explored paths.
+On the one hand, the correlation between processor temperature and energy consumption was one of the most explored paths.
 Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
@@
 In~\cite{ranka_energy_2009}, the authors claimed that energy consumption may vary due to fluctuations caused by the external environment.
-These fluctuations may alter the processor temperature and its energy consumption.
+These fluctuations may alter the processor's temperature and energy consumption.
 However, the temperature inside a data center does not show major variations from one node to another.
@@
 
-As for python, \citeauthor{redondo_comprehensive_2015}compared the performance and memory usage of various Python implementations (CPython, Jython, IronPython, PyPy, and so on) using a 215 set of benchmarks to discover that Python2 performed better with short applications, while Python3 versions covered more tests due to compatibility and the fact that Python2 became obsolete.\cite{redondo_comprehensive_2015}.
+As for python, \citeauthor{redondo_comprehensive_2015} compared the performance and memory usage of various Python implementations (CPython, Jython, IronPython, PyPy, and so on) using a 215 set of benchmarks to discover that Python2 performed better with short applications, while Python3 versions covered more tests due to compatibility and the fact that Python2 became obsolete.\cite{redondo_comprehensive_2015}.
 

### 2022-09-17 chapters/introduction.tex
refactoring
add : missing references

@@
 \newpage
-\chapter{Motivation}
+
+\chapter{Introduction \note{missing}}
+
+\chapter{Motivation \note{missing}}
+
 \label{chapter:introduction}
 
-Nowadays, computers are invading our daily life from work to leisure, from fancy smartphone to an embedded peacemaker that regulates the heartbeat of people.
+Nowadays, computers are invading our daily lives, from work to leisure, from fancy smartphones to an embedded peacemaker that regulates the heartbeat of people.
 As human beings, we are known to use tools to enhance our bodies.
-And thanks to computers we pushed that step even further where now we are using machines to extend our brains.
-Some even argue that one day they will replace us, and therefore we created our own ending.
-Well this is the problem for the future generations.
-For the moment, the main concern of humanity is to keep this planet liveable until we find another alternative.
+And thanks to computers, we pushed that step even further, to the point where now we are using machines to extend our brains.
+Some even argue that one day they will replace us, and therefore we have created our own ending.
+
+Well, this is the problem for the future generations. For the moment, the main concern of humanity is to keep this planet liveable until we find another alternative.
 
@@
 
-\note{add the impact of the ict ..etc}
-
 Researchers are trying to solve this issue through different angles.
-While some scientists are trying to find an alternative green source to generate energy. others focus on reducing it's%whose?
-consumption. In computer science we are concerned with the ladder solution %reference.
+While some scientists are trying to find an alternative green source to generate energy. others focus on reducing this energy consumption.
+In computer science we are concerned with the ladder solution %reference.
 Therefore most of the works are done on tunning the software and the hardware in order to have a more efficient way use this energy. %more efficient way of producing programs from electricity? 

### 2022-09-19 chapters/programming_languages/programming_languages.tex
Proofreading

@@
 
-As we saw in the last chapter, one of the most important things about a test is how well it \textsc{represents} the production environment.
-Therefore, we extend this study to include real-world use cases, with two case studies provided in the parts that follow.
+As we saw in the previous chapter, one of the most important things about a benchmark is how well it \textsc{reflects} the production environment.
+Therefore, we extend our study to cover real-world use cases, including two case studies reported in the following sections.
 
@@
 \import{\currfiledir/rpc}{rpc}
-\clearpage
+% \clearpage
 

### 2022-09-19 chapters/introduction.tex

@@
 \chapter{Introduction \note{missing}}
-
-\chapter{Motivation \note{missing}}
 
@@
 Some even argue that one day they will replace us, and therefore we have created our own ending.
+?
+https://www.iea.org/reports/data-centres-and-data-transmission-networks
 
+The Internet of Things (IoT) is one of the most popular topics in computer science and engineering, and it is expected to be a very important part of our lives in the future
+However, the energy consumption of ICT equipment is also expected to increase, from 4.3 exajoules in 2018 to 5.8 exajoules in 2025, according to [17].
+[17]: https://www.statista.com/statistics/271139/energy-consumption-of-ict-worldwide/
 Well, this is the problem for the future generations. For the moment, the main concern of humanity is to keep this planet liveable until we find another alternative.
+
+The number of data centers is expected to increase from 1.6 million in 2018 to 2.1 million in 2025, according to [16].
+[16]: https://www.statista.com/statistics/271138/number-of-data-centers-worldwide/
+
+The number of people connected to the Internet has increased by 4.4 billion in 2019, reaching 4.54 billion worldwide, or 59.2\% of the world population, according to the Internet World Stats [14].
+[14]: https://www.internetworldstats.com/stats.htm
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{chapters/distribution_of_ict_consumption.png}
+    \caption{Final energy consumption of digital technologies by item in 2017}
+    \label{fig:soa_comparaisoncpu}
+\end{figure}
+
+All these new activities have increased the overall environmental footprint of the Information and Communication Technology (ICT) sector, which is estimated to be responsible for approximately 4\% of the greenhouse gaz (GHG) emissions worldwide in 2020 with a worrying 8\% growth rate, according to the French think tank The Shift Project [137], or 2\% according to [13], a similar number to the aviation sector contribution.
+    [137]: https://www.theshiftproject.org/article/ict-environmental-impact/
+[13]: https://www.theguardian.com/environment/2020/jan/06/tech-industry-emissions-soar-at-double-rate-of-aviation-and-shipping
 

### 2022-09-19 chapters/optimization/python-dir/python.tex
Proofreading

@@
 
-In particular,\citeauthor{noureddine_preliminary_2012}~\cite{noureddine_preliminary_2012} in 2012, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
+In particular, \citeauthor{noureddine_preliminary_2012} in 2012~\cite{noureddine_preliminary_2012}, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
 According to their experiments, an interpreted programming language, like Python, can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. \Cref{fig:hannoi}).
@@
 \subsection{Python Popularity}
-Nowadays, Python seems to attract a large community of developers who are interested in data analysis, web development, system administration, and machine learning.
-According to a survey conducted in 2018 by JetBrains,\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}} one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.
+Nowadays, Python attracts a large community of developers who are interested in data analysis, web development, system administration, and machine learning.
+According to a survey conducted in 2018 by JetBrains,\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}} one can fear that the wide adoption of dynamic programming languages in production, like Python, may critically hamper the power consumption of ICT.
 As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
@@
 As one can notice in \Cref{fig:hannoi}, Python consumes $30$ times more than C or C++.
-The benchmark was done with an implementation of the \fnurl{Tower of Hanoi}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi} of 30 disks.
-
-Python consumes a lot of energy, mainly because it is slow in execution.
-Its flexibility and simplicity caused it to drop off in performance because Python gains its flexibility from being a dynamic language.
-Therefore, it needs an interpreter to execute its programs, which makes them much slower compared to the others that are written in compiled programming languages, such as C and C++ or semi-compiled languages like Java.
+The benchmark was done with an implementation of the \fnurl{Tower of Hanoi}{https://en.wikipedia.org/wiki/Tower_of_Hanoi} of 30 disks.
 
 As shown in \Cref{fig:clbg}, one can observe that, for most of the applications taken for the \emph{Computer Language Benchmark Game} (CLBG), Python takes more time to execute---the only case that he was not the worst one was in the benchmark \textsf{regx-redux} where he beats Go---and in some cases the gap was huge, such as in \textsf{n-body} where Python took around $100$ times more than C++.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}
+
+Python consumes energy, mainly because it is slow in execution.
+Its flexibility and simplicity caused it to drop off in performance because Python gains its flexibility from being a dynamic language.
+Therefore, it requires a faster interpreter to execute its programs to compete against alternatives written in compiled programming languages, such as C and C++ or semi-compiled languages like Java.
 
@@
 To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
-\Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}.
-57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration.
-\paragraph{Remark}
-The options in this survey were not mutually exclusive. As a result, the total of the percentages is greater than $100\%$.
+\Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers.\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}
+57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration.\footnote{The options in this survey were not mutually exclusive.As a result, the total of the percentages is greater than $100\%$.}
 
@@
     \item \emph{What is the energy footprint of Python when used in data science?}
-    \item \emph{Are the Python guidelines energy efficient by construction?}
+    \item \emph{Are the Python guidelines energy-efficient by construction?}
     \item \emph{Can we reduce the energy consumption of Python programs without altering the source code?}
@@
 
-Django~\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}}  are the most popular frameworks for web development.
-According to a Jetbrains poll\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} , Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.
+Django\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}} are the most popular frameworks for web development.
+According to a Jetbrains poll,\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.
 
-In contrast to Flask, which is still a microframework, Django is a high-level web framework that provides a standard method for creating and maintaining complex and scalable database-driven websites quickly and effectively.
-Therefore, this study will focus on The latter one.
+In contrast to Flask, which is a micro-framework, Django is a high-level web framework that provides a standard method for creating and maintaining complex and scalable database-driven websites quickly and effectively.
+Therefore, this study will focus on the latter one.
 
-\subsection{life cycle of a request in Django}
-
-Django is a MVT (Model-View-Template) framework, which means that it follows the MVC (Model-View-Controller) pattern.
-Figure~\ref{fig:django-life-cycle} describes the life cycle of a request in this framework.
-Whenever a request arrives in Django, it is processed by several middlewares, one at a time.
-These middlewares are in charge of authentication, security, and so on.
-Once the request is processed by these middlewares, it is passed to the URL Router, which will simply extract the URL from the request and will try to match it to the defined URLs.
+\subsection{Life-cycle of a Request in Django}
+Django is a MVT (\emph{Model-View-Template}) framework, which means that it follows the MVC (\emph{Model-View-Controller}) pattern.
+Figure~\ref{fig:django-life-cycle} describes the life-cycle of a request in this framework.
+Whenever a request arrives in Django, it is processed by the middleware layers, one at a time.
+These middleware layers are in charge of authentication, security, and so on.
+Once the request is processed by these layers, it is passed to the URL router, which extracts the URL from the request and tries to match it to the defined URLs.
 After getting the matching URL, the corresponding view function is called.
 This function is responsible for treating the request, gathering the data, and then generating the response that will be put inside a template to be returned as a response.
-However, as mentioned before, Django is an MVT model. which offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.
+As Django adopts a MVT model, it also offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.
 
@@
 
-First, we start by investigating the energy consumption of the request-response life cycle in Django, in ordrer to determine which part consumes the most energy.
-To do so, we created a simple Django application that simply returns the response of the request.
-and then tracked its energy consumption using joulehunter~\footnote{\url{https://github.com/powerapi-ng/joulehunter}}.
-
-Joulehunter is an open source library that we developed in order to help practitioners find the energy hotspots in their applications using statistical profilers.
-In the case of Django, it can be added as middleware without any extra setup or changes to the source code.
-The energy consumption of the request-response life cycle in Django is shown in the figure~\ref{fig:django_life_cycle_naive}.As we can notice, 91.4\% of the total energy consumption is spent to resolve the request by retrieving the data while only 5.27\% is spent to render the response.
+First, we start by investigating the energy consumption of the request-response life cycle in Django, to determine which layer consumes the most energy.
+To do so, we created a sample Django application that returns the response of the request.
+We tracked its energy consumption using JouleHunter.\footnote{\url{https://github.com/powerapi-ng/joulehunter}}
+JouleHunter is an open-source library that I developed to help practitioners to identify  energy hotspots in their applications using statistical profilers.
+In the case of Django, it can be added as an extension with no additional setup or change to the source code.
+The energy consumption of the request-response life-cycle in Django is shown in Figure~\ref{fig:django_life_cycle_naive}.
+As we can notice, 91.4\% of the total energy consumption is spent to resolve the request by retrieving the data, while only 5.27\% is spent to render the response.
 
@@
 % We have chosen Django as our web framework.
-Therefore,we chose to observe if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
-\subsection{Experiment \& Results}
+Therefore, we chose to study if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
+% \subsection{Experiment \& Results}
 To accomplish so, we examined the cost of a single request of the prior website using various ways to extract data from the database.
-We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records.
+We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records:
 \begin{enumerate}
@@
 %%%%% have to rephrase this 
-As one can observe in \Cref{fig:django}, the strategy to query stored data has a huge impact on the energy consumption.
+As one can observe in \Cref{fig:django}, the strategy to query stored data has a huge impact on energy consumption.
 As the \textsf{Vanilla} strategy can consume up to $10\times$ more energy than the \textsf{Optimized} one.
-Conversely, the choice of the database does not exhibit a key impact on the total energy despite their different behavior regarding the execution time and the average power.
-This can be useful to support developers in choosing which database engine they can adopt, based on the number of expected requests and the targeted performance.
+Conversely, the choice of the database does not exhibit a key impact on the total energy, despite their different behavior regarding the execution time and the average power.
+This can be useful to support developers in choosing which database engine they can adopt, guided by the number of expected requests and the targeted performance.
 
@@
 Another interesting observation is the impact of the interpreter, as \Cref{fig:django} highlights.
-For example, using the \textsf{Pypy} interpreter reduces the energy consumption, even when we adopting the \textsf{Vanilla} strategy.
+For example, using the \textsf{PyPy} interpreter reduces the energy consumption, even when adopting the \textsf{Vanilla} strategy.
 
-Finally, we run the same experiment with the \textsf{Optimized} strategy
-using joulehunter
-.Figure~\ref{fig:django_profiled_optimized} shows the new graph of the energy consumption.
-As one can notice.
-while the rendering method consumed the same amount of energy as in  the \textsf{Vanilla} strategy (arround 1.3KJ), the resolve part droped by $20\times$.
+Finally, we run the same experiment with the \textsf{Optimized} strategy using JouleHunter.
+Figure~\ref{fig:django_profiled_optimized} depicts the resulting energy consumption.
+As one can notice, while the rendering method consumed the same amount of energy as in the \textsf{Vanilla} strategy (around 1.3\,kJ), the resolve part dropped by $20\times$.
 
-\subsection{Conclusion}
-We conclude this study by stating that the database and ORM selection have a major impact on the website's performance and energy consumption.
-And unlike the rendering portion, there are many options available for the database, the ORM, and the data handling strategy.
+% \subsection{Conclusion}
+We can therefore conclude that the database and ORM selection have a major impact on the website's performance and energy consumption.
+And, unlike the rendering portion, there are many options available for the database, the ORM, and the data handling strategy.
+
 \begin{figure}[hbt]
@@
     \includegraphics[width=\linewidth]{imgs/django_profiled_optimized}
-    \caption{Tree representation of the energy consumption of a single request in django (naive version)}
+    \caption{Tree representation of the energy consumption of a single request in Django (naive version)}
     \label{fig:django_profiled_optimized}
 \end{figure}
-
-\Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
-Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).
-However, due to the increase of cores/threads per CPU, many Python libraries started to take advantage of this hardware feature by allowing concurrent execution of Python processes.
-For example, the multiprocessing library\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}} spawns sub-processes to increase the degree of concurrency of Python applications.
-As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of threads until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.
-
-\begin{figure}[hbt]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-    \caption{Energy consumption of Python multiprocessing depending on the number of exploited threads.}
-    \label{fig:python_multiprocessing}
-\end{figure}
-
-Before reaching the limit of physical cores, we also observe that the scheduler of the operating system tends to favor the execution of processes on the same physical core, by taking advantage of the hyper-thread feature.
-While this strategy aims to save energy by leveraging the ACPI P-states and C-states of unallocated cores, this leads to increase execution times.
 
@@
 \cite{hasan_energy_2016,oliveira_recommending_nodate}.
+
 \subsection{experiment and results}
@@
 
-\subsection{Conclusion}
+\subsection{Synthesis}
 This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
@@
 \section{Python \& Multiprocessing}
+
+\subsubsection{Cost of Parallelism in Python}
+Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
+Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).
+However, due to the increase of cores/threads per CPU, many Python libraries started to take advantage of this hardware feature by allowing concurrent execution of Python processes.
+For example, the multiprocessing library\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}} spawns sub-processes to increase the degree of concurrency of Python applications.
+As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of threads until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
+    \caption{Energy consumption of Python multiprocessing depending on the number of exploited threads.}
+    \label{fig:python_multiprocessing}
+\end{figure}
+
+Before reaching the limit of physical cores, we also observe that the scheduler of the operating system tends to favor the execution of processes on the same physical core, by taking advantage of the hyper-thread feature.
+While this strategy aims to save energy by leveraging the ACPI P-states and C-states of unallocated cores, this leads to increase execution times.
+
 \Cref{fig:python_multiprocessing} compares the behavior of python programs when we try to introduce parallelism.

### 2022-09-21 chapters/introduction.tex
comments

--- 2022-09-17 chapters/introduction.tex
@@
 
+from 4.3 exajoules in 2018 to 5.8 exajoules in 2025\footnote{\url{https://www.statista.com/statistics/271139/energy-consumption-of-ict-worldwide}}.
+
+https://www.iea.org/reports/data-centres-and-data-transmission-networks
+
+The Internet of Things (IoT) is one of the most popular topics in computer science and engineering, and it is expected to be a very important part of our lives in the future
+However, the energy consumption of ICT equipment is also expected to increase,
+
 Well, this is the problem for the future generations. For the moment, the main concern of humanity is to keep this planet liveable until we find another alternative.
+The number of data centers is expected to increase from 1.6 million in 2018 to 2.1 million in 2025, according to [16].
+
+The number of people connected to the Internet has increased by 4.4 billion in 2019, reaching 4.54 billion worldwide, or 59.2\% of the world population, according to the Internet World Stats\footnote{\url{https://www.internetworldstats.com/stats.htm}}.
+
+\begin{figure}
+    \includegraphics[width=\linewidth]{chapters/distribution_of_ict_consumption.png}
+    \caption{Final energy consumption of digital technologies by item in 2017}
+    \label{fig:distribution_of_ict_consumption}
+\end{figure}
+
+All these new activities have increased the overall environmental footprint of the Information and Communication Technology (ICT) sector, which is estimated to be responsible for approximately 4\% of the greenhouse gaz (GHG) emissions worldwide in 2020 with a worrying 8\% growth rate, according to the French think tank The Shift Project [137], or 2\% according to [13], a similar number to the aviation sector contribution.
+
+    [137]: https://www.theshiftproject.org/article/ict-environmental-impact/
+
+[13]: https://www.theguardian.com/environment/2020/jan/06/tech-industry-emissions-soar-at-double-rate-of-aviation-and-shipping
 

### 2022-09-21 chapters/optimization/python-dir/python.tex
comments

@@
 \section{Introduction}
+% TODO : MOVE this to SOA
+
+% new structure : 
+%  introduction  / motivation -> definition of python -> the plan of experiments -> experiments 
 Dynamic programming languages, except Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf. \Cref{fig:pypl}).
@@
 According to their experiments, an interpreted programming language, like Python, can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. \Cref{fig:hannoi}).
+\begin{figure}[htbp]
+    \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
+    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
+    \label{fig:hannoi}
+\end{figure}
+
 In this chapter, we explore the oblivious optimizations that can be applied to Python legacy applications to reduce their energy footprint.
@@
 One type of optimization includes alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
-
-\begin{figure}[htbp]
-    \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
-    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
-    \label{fig:hannoi}
-\end{figure}
 

### 2022-09-22 chapters/benchmarking/benchmarking.tex
refactoring
refactoring and fix missing refrences
update :benchmarking protocol

@@
-\newpage
-\label{chapter:benchmarking}
-\chapter{Benchmarking Protocol for measuring Energy Consumption of stoftwares }
-
+\chapter{A Benchmarking Protocol to Measure Software Energy Consumption}\label{chapter:benchmarking}
+\section{Introduction \note{missing} }
+This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
+First, we go through the three components of a successful benchmark within the energy consumption experiment.
+Section~\ref{sec:benchmarking_reproducibility} concentrates on the "reproducibility" issue and how to deliver reproducible experiments without interfering with the energy measurement,while Section~\ref{sec:taming-the-energy-variation} discusses accuracy in software energy consumption.
 \import{\currfiledir}{draft}
-
-\import{\currfiledir}{reproducibility}
 \import{\currfiledir}{experiment}
-
 \import{\currfiledir}{conclusion}

### 2022-09-22 chapters/benchmarking/conclusion.tex
comments
update :benchmarking protocol

@@
 \section{Conclusion \note{Uncomplete }}
-% TODO : uncomplete 
-By the end of this study, we have gathered enough guidelines to make the tests more reproducible, accurate and extensible.
-We created a set of new tests named \textbf{energy tests} which are more similar to performance tests.
-Thanks to the work of two interns [Mamadou and Adrien] we created a CI/CD platform to measure the energy consumption of Java projects and we could track the evolution of the this energy across different stages of the project.
-In the figure below we see an example of this plugin.
-For more details please visit the gitlab repository\footnote{\url{https://gitlab.inria.fr/mamdiall/J-Joules}}~\footnote{\url{https://gitlab.inria.fr/mamdiall/jjoules-plugin}} .
+In this chapter we :
+- studied two aspects of the three creterias of the emperical research and how they will be applied in the field of energy measurement .
+- we provided a brief comparaison of different methods to increase the reproducibility of the measurement .
+- we studied the factors that can affect the measurement accuracy and how to reduce them .
+- and finally
 
-We propose to extend the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
-Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in Section~\ref{sec:benchmarking_reproducibility}, we propose an abstract model to describe an empirical experiment:
-
-\emph{context}: the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor .
-An \emph{orchestrator} : which its job to run the experiment.
-The orchestrator provides three interfaces:
-
-\emph{Workload interface}: provides a set of functions that should be implemented by the workload and called by each candidate in the experiment.
-The Observer interface provides a set of metrics that are collected by the orchestrator during the experiment.
-Note that some of those metrics are external, such as the energy consumption of the system , execution time, execution stack, etc .
-\begin{figure}[!htb]
-    \center{\includegraphics[width=\linewidth]{imgs/benchmarkingprotocol}}
-    \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
-\end{figure}
-Figure~\ref{fig:benchmarkingprotocol} explains in detail the relationship within each component of the protocol.
-First of all, we should start with the orchestrator which is the kernel of the experiment. Most of the time this kernel should represent the constants parts of the experiment.
-The orchestrator is responsible for the following  for the coordination between the workload and the candidate, therefore it provides an interface that is implemented by each workload.
-Later this interface will be provided to each candidate by the orchestrator.
-Like this we will remove any bias of dependency between the workload and the candidates, which will alow us to extend both parts without any change in the code.
-Instead of retrieving the data directly from the experiment, we suggest to use an observator to allow the extension of the metrics, this observer in general should be outside the experiment to be able to gather as many informations without impacting the the code.
+%  THIS SHOULD BE WITHIN THE perspective chapter 
+% By the end of this study, we have gathered enough guidelines to make the tests more reproducible, accurate and extensible.
+% We created a set of new tests named \textbf{energy tests} which are more similar to performance tests.
+% Thanks to the work of two interns [Mamadou and Adrien] we created a CI/CD platform to measure the energy consumption of Java projects and we could track the evolution of the this energy across different stages of the project.
+% In the figure below we see an example of this plugin.
+% For more details please visit the gitlab repository\footnote{\url{https://gitlab.inria.fr/mamdiall/J-Joules}}~\footnote{\url{https://gitlab.inria.fr/mamdiall/jjoules-plugin}} .
 

### 2022-09-22 chapters/benchmarking/draft.tex
update :benchmarking protocol

@@
 So, not only is it important to make sure the results can be "reproduced," but it's also important to make room for new candidates.
-This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
-First, we go through the three components of a successful benchmark within the energy consumption experiment.
-Section~\ref{sec:benchmarking_reproducibility} concentrates on the "reproducibility" issue and how to deliver reproducible experiments without interfering with the energy measurement.
-Section~\ref{sec:taming-the-energy-variation} discusses accuracy in software energy consumption.
-Finally, Section~\ref{sec:bench_extension} presents tools for introducing new candidates, workloads, or other key performance indicators.
+% TODO :  something missing here  to link between these two paragraphs.
+
 % Regarding the "representativeness" of the benchmarks, we will elaborate in the section entitled Perspectives.
@@
 
+As can be seen, the introduction of the supervisor has increased the impact of Docker on energy that will be applied equitably to all experiments.
+Therefore, when it comes to comparison analysis, it will mitigate its impact automatically.
+Furthermore, because we have access to the host hardware, we do not need to worry about capturing the SUT's energy use.
+As a result, we will use Docker to keep all of our tests distinct, as it will assure reproducibility in a simple and clear manner.
+
+\subsection{Extension}
+Whith a rapid evolution of the software industry, it is important to be able to include new candidates
+We propose to extend the benchmarking framework suggested by the \em{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
+Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in Section~\ref{sec:benchmarking_reproducibility}, we propose an abstract model to describe an empirical experiment:
+
+\em{context}: the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor .
+An \em{orchestrator} : which its job to run the experiment.
+The orchestrator provides three interfaces:
+
+\em{Workload interface}: provides a set of functions that should be implemented by the workload and called by each candidate in the experiment.
+The Observer interface provides a set of metrics that are collected by the orchestrator during the experiment.
+Note that some of those metrics are external, such as the energy consumption of the system , execution time, execution stack, etc .
+\begin{figure}[!htb]
+    \center{\includegraphics[width=\linewidth]{imgs/benchmarkingprotocol}}
+    \caption{Benchmarking protocol}\label{fig:benchmarkingprotocol}
+\end{figure}
+
+Figure~\ref{fig:benchmarkingprotocol} explains in detail the relationship within each component of the protocol.
+First of all, we should start with the orchestrator which is the kernel of the experiment. Most of the time this kernel should represent the constants parts of the experiment.
+The orchestrator is responsible for the following  for the coordination between the workload and the candidate, therefore it provides an interface that is implemented by each workload.
+Later this interface will be provided to each candidate by the orchestrator.
+Like this we will remove any bias of dependency between the workload and the candidates, which will alow us to extend both parts without any change in the code.
+Instead of retrieving the data directly from the experiment, we suggest to use an observator to allow the extension of the metrics, this observer in general should be outside the experiment to be able to gather as many informations without impacting the the code.
+
 \subsection{Conclusion}
-As can be seen, the introduction of the supervisor has increased the impact of Docker on energy that will be applied equitably to all experiments.
-Therefore when it comes to comparison analysis, it will mitigate its impact automatically.
-Furthermore, because we have access to the host Hardware, we do not need to worry about capturing the SUT's energy use.
-As a result, we will use Docker as an encapsulation solution for all of our experiments.
-It will ensure the reproducibility of our experiments in a simple and straightforward manner.
+
+In this section, we have discussed the different options for encapsulating the system-under-test in order to ensure experiment reproducibility.
+We have shown that the use of VMs is not suitable for performance-related tests, while containers are a good alternative.
+and then we discussed the pros and cons of using Docker while measuring the energy consumption of our experiments.
+
 \clearpage

### 2022-09-22 chapters/benchmarking/experiment.tex
update :benchmarking protocol

@@
-\newpage
 % To satisfy those three requirements. we elaborated a benchmarking protocol that aims to help reaserchers and practicioners in their experementations.
@@
 Our main criterion was the CPU configuration.
-The table~\ref{table:g5k}  below, presents a description of the four clusters that have been chosen for our experiments.
+Table~\ref{table:g5k}  below, presents a description of the four clusters that have been chosen for our experiments.
 
@@
 \subsection{Conclusion}
-In this part, we conducted an empirical study of controllable factors that can increase the energy variations on platforms with some of the latest CPU, and for several workloads.
-We provide a set of guidelines that can be implemented and tuned (through the OS GRUB for example), especially with the new data centers isolation trend and the cloud usage, even for scientific and R\&D purposes.
-Our guidelines aim at helping the user in reducing the CPU energy variation during systems benchmarking, and conduct more stable experiments when the variation is critical.
-For example, when comparing the energy consumption of two versions of an algorithm or a software system, where the difference can be tight and need to be measured accurately.
-
-Overall, our results are not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
-The aim of our work is to be able to tame and mitigate this variability along controlled experiments.
-We studied some previously discussed aspects on some recent CPU, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to achieve the variability mitigating purpose.
+In order to increase the accuracy in comparative experiments, we conducted an empirical study of controllable factors that may increase the energy variations on platforms with some of the latest CPUs and for several workloads.
+this study is not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
+Instead, it aims to tame and mitigate this variability along controlled experiments.
+In this study, We investigated some previously discussed aspects on some recent CPU, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to increase the accuracy of energy measurement.
 Some of these factors, like the C-states usage, can reduce the energy variation up to 500\,\% at low workloads, while choosing the wrong cores/PU strategy can cause up to $30\times$ more variability.
-
-We believe that our approach can also be used to study/discover other potential variability factors, and extend our results to alternative CPU generations/brands.
-Most importantly, this should motivate future works on creating a better knowledge on the variability due to CPU manufacturing process and other factors.

### 2022-09-23 chapters/benchmarking/draft.tex
update : introduction of reproducibility

@@
 
-%  NOTE : add introduction and change this into motivation ?? 
 Empirical measurements are critical to capture the effect of developers' choices on software energy consumption.
 To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state of the art~\cite{van_der_kouwe_benchmarking_2018}.
-Furthermore, when it comes to evaluating the energy consumption of software, other challenges occur, such as the impact of hardware, the difference between the machines used for measurements and those that may apply the results, and so on.
-Moreover, considerable advances in computer science have resulted in an increase in the number of obsolete results.
-Addressing all of them becomes significantly more difficult in studies that compare multiple treatments.
-Also, between the exploratory experiment and when the results were made public, new candidates may have come up and others may have changed.
-So, not only is it important to make sure the results can be "reproduced," but it's also important to make room for new candidates.
-% TODO :  something missing here  to link between these two paragraphs.
 
+Second, one should not ignore tremendous progress in computer science, which have led to a rise in the number of obsolete results.
+Furthermore, when it comes to comparative research, the execution environment may have an impact on the study itself.
+Finally, between the exploratory experiment and the publication of the results, new candidates may have emerged and others may have changed.
+
+As a result, it is critical not only to ensure that the results can be "reproduced", so one can test their hypothesis in different environments, but also to provide room for new candidates.
+In this section, we will address these issues and investigate several ways of encapsulating the systems-under-test to ensure experiment reproducibility in the context of energy consumption tests while providing the benefits and drawbacks of each strategy.
+Later in this section, we will also propose a protocol to ensure that the results are not only reproducible but extensible.
+
+% Furthermore, when it comes to evaluating the energy consumption of software, other challenges occur, such as the impact of hardware, the difference between the machines used for measurements and those that may apply the results, and so on.
 % Regarding the "representativeness" of the benchmarks, we will elaborate in the section entitled Perspectives.
@@
 later we will talk about the energy metrics and how to ensure the accuracy when it comes to the measurement of the energy consumption.
-This section addresses various strategies for encapsulating the systems-under-test in order to ensure experiment reproducibility.
 

### 2022-09-23 chapters/benchmarking/draft.tex
conclusion : reproducibiliy

@@
 Furthermore, because we have access to the host hardware, we do not need to worry about capturing the SUT's energy use.
-As a result, we will use Docker to keep all of our tests distinct, as it will assure reproducibility in a simple and clear manner.
+As a result, we will use Docker to keep all of our tests separate, as it will make sure that they can be repeated in a clear and simple way.
 
 \subsection{Extension}
-Whith a rapid evolution of the software industry, it is important to be able to include new candidates
-We propose to extend the benchmarking framework suggested by the \em{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}} to address those limitations.
-Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in Section~\ref{sec:benchmarking_reproducibility}, we propose an abstract model to describe an empirical experiment:
+% TODO : proofread
+\subsubsection{Definition of Extention}
+With the rapid evolution of the software industry, Even ensuring the reproducibility of the same research won't be enough.
+Each day, new versions of the software are released, and new features are added.
+Even more, the goal of the research might evolve.
+As a result, nowadays, especially in comparative studies, it is essential to leave room for expansion.
 
-\em{context}: the hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of the external factor .
-An \em{orchestrator} : which its job to run the experiment.
-The orchestrator provides three interfaces:
+One can extand their experiment through three axes :
+\begin{itemize}
+    \item \textbf{proposed solutions}: Where one will expand their research by including additional solutions and comparing them to earlier ones
+    \item \textbf{evaluation criteria}: This axe's objective is to broaden the evaluation criteria to incorporate additional measures like as performance, cost, and security, among others.
+    \item \textbf{benchmarks}: The purpose of this axe is to enlarge the benchmarks to include other ones in order to broaden the scope of the research.
+\end{itemize}
 
-\em{Workload interface}: provides a set of functions that should be implemented by the workload and called by each candidate in the experiment.
-The Observer interface provides a set of metrics that are collected by the orchestrator during the experiment.
-Note that some of those metrics are external, such as the energy consumption of the system , execution time, execution stack, etc .
-\begin{figure}[!htb]
+\subsubsection{The architecture of the extension}
+\begin{figure}
     \center{\includegraphics[width=\linewidth]{imgs/benchmarkingprotocol}}
@@
 \end{figure}
+To be able to extend the empirical experiments through these axes,
+We propose to enhance the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}}.
+Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in Section~\ref{sec:benchmarking_reproducibility}, we suggest an abstract model to describe an empirical experiment.
+Figure~\ref{fig:benchmarkingprotocol} shows the proposed model while providing a comparaison between the existing solution and the proposed one.
 
-Figure~\ref{fig:benchmarkingprotocol} explains in detail the relationship within each component of the protocol.
-First of all, we should start with the orchestrator which is the kernel of the experiment. Most of the time this kernel should represent the constants parts of the experiment.
-The orchestrator is responsible for the following  for the coordination between the workload and the candidate, therefore it provides an interface that is implemented by each workload.
-Later this interface will be provided to each candidate by the orchestrator.
+The model is composed of different components that are described below,
+\paragraph{Context}
+The hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of  external factors .
+
+\paragraph{orchestrator}
+
+The core design of the experiment which is responsible for running the experiment regrardless of the context.
+In the experiment,it is  the only component that is allowed to interact with the SUT.
+The orchestrator provides three interfaces:
+\begin{itemize}
+    \item \emph{Workload interface} : provides a set of functions that should be implemented by the workload and called by each candidate in the experiment.This interface is responsible for extending the experiment with new benchmarks.
+    \item \emph{The Observer interface}: provides a set of metrics that are collected by the orchestrator during the experiment.
+Implementing the observer interface allows the user to extend the experiment with new metrics.
+    \item \emph{The Candidate interface} : It provides a set of functions that should be implemented by the candidate and called by the orchestrator.
+This interface is responsible for extending the experiment with new solutions.
+\end{itemize}
+
+\paragraph{Data}
+The row data collected collected by the experiment.
+In ourcase it will be provided by the observator, one ore more observators can be used to collect different metrics.
+The purpose of this data is to ensure the \emph{replication} of the experiment which will allow the reader to perform extra analysis without rexucuting the experiment.
+
+\paragraph{Analysis}
+The final part, that should provide the set of methodes and functions used to do the empirical analysis in order to answer the research questions.
+
+% It starts with the orchestrator is the core design of the experiment  for the coordination between the workload and the candidate, therefore it provides an interface that is implemented by each workload.
+% Later this interface will be provided to each candidate by the orchestrator.
 Like this we will remove any bias of dependency between the workload and the candidates, which will alow us to extend both parts without any change in the code.
-Instead of retrieving the data directly from the experiment, we suggest to use an observator to allow the extension of the metrics, this observer in general should be outside the experiment to be able to gather as many informations without impacting the the code.
+% Instead of retrieving the data directly from the experiment, we suggest to use an observator to allow the extension of the metrics, this observer in general should be outside the experiment to be able to gather as many informations without impacting the the code.
 
@@
 
-In this section, we have discussed the different options for encapsulating the system-under-test in order to ensure experiment reproducibility.
-We have shown that the use of VMs is not suitable for performance-related tests, while containers are a good alternative.
-and then we discussed the pros and cons of using Docker while measuring the energy consumption of our experiments.
+In this section, we addressed the first challenge of empirical research, which is the \textbf{reproducibility} of the experiments.
+We started by listing the different options for encapsulating the system-under-test.
+Then, we have shown that the use of VMs is not suitable for performance-related tests nor energy-related ones, while containers are a good alternative.
+Laterm, we discussed the pros and cons of using Docker for experiments related to energy consumption and have shown that it has a constant overhead which is self-mitigated when it comes to comparison analysis.
+Lastly, we have proposed an addition to the benchmarking framework that would allow the experiment to be extended along the three axes already mentioned.
 

### 2022-09-23 chapters/literature/draft_citations.tex
fix: missing references

@@
 It also examines the sources of energy measurement variations, which represent a significant obstacle to achieving precise readings and higher accuracy.
-Then, in section \ref{section:soa_energyoptimization}, we will go through some of the previous work on improving the energy consumption of software.
+Then, in Section~\ref{section:soa_energyoptimization}, we will go through some of the previous work on improving the energy consumption of software.
 
@@
 One of the criteria for supporting reproducibility is the publication of the dataset and the algorithms run on the raw data to derive the results.
-There is even some disagreement about what the terms "reproducibility" or "replicability" by themselves mean \cite{goodman2016does}.
-According to \cite{echtler2018open}, \emph{replicability} extends \emph{reproducibility} with the ability to collect a new raw dataset comparable to the original one by re-executing the experiment under similar conditions, instead of just the ability to get the same results by running the statistical analyses on the original data set.
+There is even some disagreement about what the terms "reproducibility" or "replicability" by themselves mean~\cite{goodman2016does}.
+According to~\cite{echtler2018open}, \emph{replicability} extends \emph{reproducibility} with the ability to collect a new raw dataset comparable to the original one by re-executing the experiment under similar conditions, instead of just the ability to get the same results by running the statistical analyses on the original data set.
 
@@
 Which can lead derive incorrect results from an experiment if a seemingly insignificant feature of the experimental design is altered.
-They treated this problem by proposing two strategies for detecting measurement bias by using causal analysis and preventing it with setup randomization \cite{mytkowicz2009producing}.
-another study that was published in the book "Measuring computer performance: a practitioner's guide"\cite{lilja2005measuring},\citeauthor{lilja2005measuring} examined performance indicators and gave in-depth treatment of benchmark program tactics.
+They treated this problem by proposing two strategies for detecting measurement bias by using causal analysis and preventing it with setup randomization~\cite{mytkowicz2009producing}.
+another study that was published in the book "Measuring computer performance: a practitioner's guide"~\cite{lilja2005measuring},\citeauthor{lilja2005measuring} examined performance indicators and gave in-depth treatment of benchmark program tactics.
 they provided clear explanations of the basic statistical methods required for interpreting measured performance data.
@@
 \citeauthor{bukh1992art} wrote a book about computer performance analysis, where they discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi-squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems.
-The article \cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistakes in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often seen errors.
+The article~\cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistakes in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often seen errors.
 
@@
 Therefore, in this section, we will discuss the different approaches that can be used to measure the energy consumption of a system.
-Many studies have been conducted to estimate such energy consumption that varies from static analysis of the source code to infer its energy consumption like \citeauthor{pereira_helping_2017} where they provided a tool to highlight the most energy consuming parts of the code \cite{pereira_helping_2017}.
+Many studies have been conducted to estimate such energy consumption that varies from static analysis of the source code to infer its energy consumption like \citeauthor{pereira_helping_2017} where they provided a tool to highlight the most energy consuming parts of the code~\cite{pereira_helping_2017}.
 The key advantage of this approach is that it allows you to estimate a program's energy usage without executing it.
@@
 dedicated one , they are based on the instrumentation of some parts of the computer , however they have a high cost and they are defficult to scale since they require an instrumentation of the execusion environment.
-According to \citeauthor{hackenberg2014hdeem}, there are four main criteria to evaluate an energy measurement tool \cite{hackenberg2014hdeem}:
+According to \citeauthor{hackenberg2014hdeem}, there are four main criteria to evaluate an energy measurement tool~\cite{hackenberg2014hdeem}:
 \begin{itemize}
@@
 It allows a sampling frequency up to 1\,Hz and has an internal memory to store a wide variety of data, such as the maximum voltage, and current, that later can be exported via USB port for personal usage or lined to some graph programs like Logger Pro or LabQuest.
-The main advantage of this tool is the ability to monitor the energy consumption from a different device which will reduce the risk of interference with the energy consumption of the test % a revoir 
-\cite{hirst2013watts}
+The main advantage of this tool is the ability to monitor the energy consumption from a different device which will reduce the risk of interference with the energy consumption of the test~\cite{hirst2013watts}
 
@@
 
-PowerMon and its upgraded version powerMon2 \cite{bedard2010powermon} are based on a micro-controller chip that can monitor up to 6 channels (8 for powermon2), simultaneously.
+PowerMon and its upgraded version powerMon2~\cite{bedard2010powermon} are based on a micro-controller chip that can monitor up to 6 channels (8 for powermon2), simultaneously.
 Therefore, we are able to monitor the power consumption of 4 devices at the same time.
@@
 
-With a similar approach, we can find NVIDIA reporting tools, such as GPU TESLA \cite{burtscher2014measuring} and the NVML library~\cite{fahad2019comparative}.
+With a similar approach, we can find NVIDIA reporting tools, such as GPU TESLA~\cite{burtscher2014measuring} and the NVML library~\cite{fahad2019comparative}.
 
@@
 
-Other works that have studied the impact of website technology on energy consumption,\cite{philippot_characterization_2014,manotas_investigating_2013}.
+Other works that have studied the impact of website technology on energy consumption,~\cite{philippot_characterization_2014,manotas_investigating_2013}.
 In their work~\cite{philippot_characterization_2014},\citeauthor{philippot_characterization_2014} measured the computer resources used during the loading of a website in a browser, such as memory utilization and energy consumption, for over 500 websites and proposed some best practices for developers.
@@
 \citeauthor{hasan_energy_2016} examined the energy consumption of multiple Java data structures, analyzing the bytecode and evaluating the change in energy consumption in various circumstances(research, insertion, deletion, etc.).
-They also simulated best- and worst-case energy usage scenarios in real-world production systems by replacing the LinkedList and ArrayList, and discovered that incorrect collection can result in a 300\% increase in energy consumption\cite{hasan_energy_2016}.
+They also simulated best- and worst-case energy usage scenarios in real-world production systems by replacing the LinkedList and ArrayList, and discovered that incorrect collection can result in a 300\% increase in energy consumption~\cite{hasan_energy_2016}.
 
-Other studies\cite{longo_reducing_2019,kumar_energy_2017} looked into the energy use of Java primitive types, string operations, and the use of exceptions, loops, and arrays.
+Other studies~\cite{longo_reducing_2019,kumar_energy_2017} looked into the energy use of Java primitive types, string operations, and the use of exceptions, loops, and arrays.
 \citeauthor{kumar_energy_2017}, for example, examined the energy usage of code snippets and micro-benchmarks and provided several conclusions, such as string concatenation would use less energy than StringBuilder and StringBuffer and static variables tends to consume 60\% more energy than instance variables.
 
-\citeauthor{pereira_helping_2017} \cite{pereira_helping_2017} presented SPELL, a tool that helps developers spot energy leaks in their source code.
-Using a statistical spectrum-based analysis and JRAPL\cite{guimaraes2016some,liu_data-oriented_2015}, the tool locates energy-inefficient code fragments.
+\citeauthor{pereira_helping_2017}~\cite{pereira_helping_2017} presented SPELL, a tool that helps developers spot energy leaks in their source code.
+Using a statistical spectrum-based analysis and JRAPL~\cite{guimaraes2016some,liu_data-oriented_2015}, the tool locates energy-inefficient code fragments.
 According to the authors, it is language and context-independent.
@@
 
-The energy usage of virtual machine allocation and task placement has also been investigated\cite{mishra_energy_2018}.In this article, The authors propose a method for mapping workloads to virtual machines and virtual machines to the physical ones (PMs) in an energy-efficient manner.
+The energy usage of virtual machine allocation and task placement has also been investigated~\cite{mishra_energy_2018}.In this article, The authors propose a method for mapping workloads to virtual machines and virtual machines to the physical ones (PMs) in an energy-efficient manner.
 To solve the problem of high heterogeneity of activities and resources, the jobs are categorized based on their resource requirements, and then the relevant VM is found, followed by the appropriate PM where the selected VM can be deployed Using a cloud simulator, the authors claimed The suggested technique saves energy by reducing the number of active PMs while also minimizing the makespan and task rejection rate.
@@
 Besides the impact of the orchestration of virtual machine, some works have been targeting the runtime of specific programming languages.
-Using the TPC-DS benchmark, \citeauthor{chiba2018towards} investigated the influence of HOTSPOT \footnote\url{https://openjdk.org/groups/hotspot/} and J9\footnote{\url{https://www.eclipse.org/openj9/}} on the performance of SQL-on-Hadoop systems (SPARK and TEZ) to reveal a three-fold disadvantage that one JVM can have over the other\cite{chiba2018towards}.
+Using the TPC-DS benchmark, \citeauthor{chiba2018towards} investigated the influence of HOTSPOT \footnote{\url{https://openjdk.org/groups/hotspot/}} and J9\footnote{\url{https://www.eclipse.org/openj9/}} on the performance of SQL-on-Hadoop systems (SPARK and TEZ) to reveal a three-fold disadvantage that one JVM can have over the other~\cite{chiba2018towards}.
 \citeauthor{oi_power_2011} also compared the performance of HOTSPOT and J9.

### 2022-09-23 chapters/benchmarking/conclusion.tex
add : conclusion to chapter 2

@@
 %% maybe add extension as a part of the conclusion for this chapter 
-\section{Conclusion \note{Uncomplete }}
-In this chapter we :
-- studied two aspects of the three creterias of the emperical research and how they will be applied in the field of energy measurement .
-- we provided a brief comparaison of different methods to increase the reproducibility of the measurement .
-- we studied the factors that can affect the measurement accuracy and how to reduce them .
-- and finally
+\clearpage
+\section{Summary}
+there As seen in Section~\cref{sec:soa_benchmarking} of the state of the art, a successful benchmark faces three challenges: reproducibility, accuracy and representativeness. 
+This chapter has covered two of the three criteria.
+
+The first section covered the reproducibility challenge by studying the existing techniques that ensure reproducibility.
+We have observed that there are two primary methods for encapsulating experimental systems.
+The first is to utilize virtual machines,while the second one is based on containers.
+We established that Docker is more suitable for energy-related studies for three reasons.
+
+\begin{enumerate}
+    \item It is more lightweight than virtual machines.
+    \item It offers the interactivity with the hardware of the host machine which will enable us to gather more metrics.
+    \item It has a constant overhead, a key factor to nullify the encapsulation impact on the energy consumption when performing empirical analyses.
+\end{enumerate}
+
+After settling on the most suitable choice to encapsulate experiments, we highlighted the need to enhance the reproducibility of empirical tests along several axes (benchmarks, metrics, and candidates) for the purpose To keep up with the rapid pace of software development.
+Furthermore, we have provided a model that enables to extend comparative expriment.
+The foundation of this model is the separation of the experiment into multiple independent components: an orchestrator that executes the experiment, one or more observers that collect metrics, the candidates being tested, and the benchmark against which these candidates are being compared.
+
+As for the second section, we addressed the accuracy challenge in the context of energy-based experiments.
+We started by analyzing the impact of the chosen encapsulation method from the previous section on this challenge. to find out that the impact is negligible.
+Then, we conducted an empirical study utilizing some of the well-known benchmarks in literature, \texttt{Stress-ng}~\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} and \texttt{NAS Parallel Benchmark}~\cite{Bailey:1991:NPB:125826.125925}, on a variety of machines with diverse hardware combinations and operating system setups and tunings.
+We've shown that optimizing the operating system can have a big effect on how accurately energy is used.
+This effect can affect the accuracy of the experiments  from an increase of $5\times$ to a penalty of $30\times$.
+
+We have also shown in this section the harm that this taming of the variation can cause to the representativeness of the results,since some aspects that help increase the accuracy of the results in the research environment can't be applied in the production environment, such as turning off the C-states.
+Another impact of increasing this accuracy was the increase in the overall energy consumption of the tests.
+Even if the overall was for all the candidates, this remains an issue to be addressed.
+
+Overall, the goal of this chapter is to establish a reproducible and accurate protocol for conducting energy-related experiments.
+From now on, we shall use this protocol in our studies for the purpose of reducing the energy usage of software.
 

### 2022-09-23 chapters/benchmarking/experiment.tex
fix: missing references
add : conclusion to chapter 2

@@
 
-\begin{table}[!hbt]
+\begin{table}[hbt]
     \centering
@@
 
-\begin{figure}%[!htb]
+\begin{figure}[htb]
     \center{\includegraphics[width=.9\linewidth]{imgs/lstopo}}
@@
 To meet these criteria, we looked at the "state of the art" and found the most common benchmarks used to test the performance of hardware. We then chose the ones that could be scaled up.
-Our candidate is  emph{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925}: one of the most used benchmarks for \emph{HPC}.
+Our candidate is \texttt{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925}: one of the most used benchmarks for \emph{HPC}.
 We used the applications (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
@@
 
-\begin{figure*}%[!htb]
+\begin{figure*}[htb]
     \center{\includegraphics[width=\linewidth]{imgs/all}}
@@
 
-\begin{figure}%[!htb]
+\begin{figure}[htb]
     \center{\includegraphics[width=.9\linewidth]{imgs/stressng}}
@@
 \subsection{Conclusion}
-In order to increase the accuracy in comparative experiments, we conducted an empirical study of controllable factors that may increase the energy variations on platforms with some of the latest CPUs and for several workloads.
-this study is not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
+In order to increase the \textbf{accuracy} in comparative experiments, we conducted an empirical study of controllable factors that may increase the energy variations on platforms with some of the latest CPUs and for several benchmarks.
+This study is not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
 Instead, it aims to tame and mitigate this variability along controlled experiments.
-In this study, We investigated some previously discussed aspects on some recent CPU, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to increase the accuracy of energy measurement.
-Some of these factors, like the C-states usage, can reduce the energy variation up to 500\,\% at low workloads, while choosing the wrong cores/PU strategy can cause up to $30\times$ more variability.
+In this study, We investigated some previously discussed aspects on some recent CPUs, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to increase this accuracy for energy related experiments.
+Some of these factors, like the \emph{C-states} usage, can reduce the energy variation up to 500\,\% at low workloads, while choosing the wrong cores/PU strategy can cause up to $30\times$ more variability.

### 2022-09-23 chapters/optimization/python-dir/python_interperters.tex
corrections : end part
fix: missing references

@@
             A Python library that aims to introduce JIT compiler into the Python code;
-      \item \textbf{Parakeet}~\cite{DBLP:conf/hotpar/RubinsteynHWS12}:
+      \item \textbf{Parakeet}~\cite{rubinsteyn2012parakeet}:
             A runtime accelerator for an array-oriented subset of Python;
@@
 
-We should explain the classification of these runtimes
+% We should explain the classification of these runtimes
 \begin{table}
@@
 In Numba, the developer must target the portion of the code that will be passed to JIT, while the rest of the code will be executed by the regular interpreter (aka CPython).
-
 \paragraph{Remark:}
@@
       \centering
-      \includegraphics[width=1.0\linewidth]{imgs/alltomti_performance}
+      \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{imgs/alltomti_performance}
       \caption{Different interpreter's energy scores.}
@@
 One can conclude that the choice of Python interpreter has a significant impact on the programs' energy consumption.
+While most of the candidates used Python2, we observed that different strategies that were implemented within each candidate have a significant impact on the energy consumption. 
+The major improvement of the energy was related to the use of JIT, which is a technique that compiles the code into machine code at runtime. 
+The effect of the compilation of the python code into machine code was also confirmed with the use of translators such as shedskin.
+
 This investigation is made more intriguing by the absence of a universal solution.

### 2022-09-23 chapters/literature/draft_citations.tex

@@
 
-Other works examined software energy consumption efficiency through source code changes and optimizations.
+% Other works examined software energy consumption efficiency through source code changes and optimizations.
 For example, some papers [42, 122] studied the effect of Java collections on energy consumption, depending on the collection size and/or the most executed tasks on the collection (insertion, removal, search).

### 2022-09-23 chapters/programming_languages/programming_languages.tex
conclusion : Java chapter
Proofreading titles

@@
-\chapter{The Energy Footprint of Programming Frameworks}
+\chapter{Energy Footprint of Distributed Programming Frameworks}
 \label{chapter:porgramming_langauges}
-\section{Introduction}
+% \section{Introduction}
 In this chapter, we study how the programming framework affects the energy the software consumes.
@@
 \import{\currfiledir/rpc}{rpc}
-% \clearpage
 

### 2022-09-23 chapters/optimization/python-dir/python.tex
remove dupliacte figures

@@
 This section aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
-Therefore, we start by extending the work of \citeauthor{hasan_energy_2016} and
-\citeauthor{oliveira_recommending_nodate} to the context of Python environments
-\cite{hasan_energy_2016,oliveira_recommending_nodate}.
+Therefore, we start by extending the work of Hasan~\emph{et~al.} and Oliveira~\emph{et~al.} to the context of Python environments
+\cite{hasan_energy_2016,oliveira_recommending}.
 
@@
 
-\begin{figure}[hbt]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-    \caption{Energy consumption of Python multiprocessing depending on the number of exploited threads.}
-    \label{fig:python_multiprocessing}
-\end{figure}
+% \begin{figure}[hbt]
+%     \centering
+%     \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
+%     \caption{Energy consumption of Python multiprocessing depending on the number of exploited threads.}
+%     \label{fig:python_multiprocessing}
+% \end{figure}
 
@@
 \input{python_interperters.tex}
-\subsection{threads to validity \note{missing}}
-\section{Conclusion \note{missing}}
+% \subsection{threads to validity \note{missing}}
+\section{Summary}:

### 2022-09-23 chapters/benchmarking/benchmarking.tex
Proofreading titles

--- 2022-06-11 chapters/benchmarking/benchmarking.tex
@@
-\newpage
-\label{chapter:benchmarking}
-\chapter{Benchmarking Protocol for measuring Energy Consumption of stoftwares }
+\chapter{Benchmarking Protocol to Measure Software Energy Consumption}\label{chapter:benchmarking}
+% \section{Introduction \note{missing} }
+This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
+First, we go through the three components of a successful benchmark when performing energy-related experiments.
+Section~\ref{sec:benchmarking_reproducibility} focuses on the "reproducibility" challenge---to deliver reproducible experiments without interfering with the energy measurement---while Section~\ref{sec:taming-the-energy-variation} discusses accuracy of software energy consumption.
 
 \import{\currfiledir}{draft}
-
-\import{\currfiledir}{reproducibility}
 \import{\currfiledir}{experiment}
-
 \import{\currfiledir}{conclusion}

### 2022-09-23 chapters/optimization/java-dir/java.tex
refactoring
corrections : end part

@@
 
-\subsection{characteristics of JVM}
+\subsection{Characteristics of JVM}
 \begin{figure}%[ht]
@@
 
-\subsection{Goal}
+\subsection{Research questions}
 In this section, we will investigate the following research questions:
@@
 
-\section{Conclusion}
-The results of our investigations show that many JVMs share energy efficiencies and can be grouped into 3 classes: \textsc{HotSpot}, J9, and \textsc{GraalVM}.
-The 3 selected JVM classes can however report a different energy efficiency for different software and/or workloads, sometimes by a large margin.
-While we did not observe a unique champion when it comes to energy consumption, \textsc{GraalVM} reported the best energy efficiency for a majority of benchmarks.
-Nonetheless, each JVM can achieve the best or the worst depending on the hosted application.
-One cause can be thread management strategies, as observed with J9 when advantageously running \textsf{Avrora}.
-Moreover, some JVM settings can cause energy consumption variations.
-Our experiments showed that the default JIT compiler of the JVM is often near-optimal, in at least ~80\% of our experiments.
-The default GC, however, was outperforming alternative strategies in half of our experiments, with some large gains observed when using some alternative GC depending on the application characteristics.
+\clearpage
+\section{Summary}
 
-Our main conclusions and guidelines can be thus summarized as:
+This chapter describes an empirical investigation of contrast in energy consumption that some of the most well-known and widely supported JVM platforms can exhibit as well as the important factors that can influence this energy consumption.
+During our experiments, we investigated 12 well-known and diverse-purpose Java benchmarks, as well as 52 JVMs, including many versions of 11 different distributions. 
+Our findings revealed that various JVMs share energy efficiency and may be classified into three major classes: \textsc{HotSpot}, \textsc{J9}, and \textsc{GraalVM}.
+However, the three JVM classes can show substantially different energy efficiency depending on the workloads and/or application.
+While there was no clear champion in terms of energy consumption,\textsc{GraalVM} recorded the highest energy efficiency for the most of benchmarks.
+Nevertheless, depending on the workload, each JVM can achieve higher or worse efficiency.
+One explanation could be thread management policy, as seen with \textsc{J9} when running \textsf{Avrora}.
+
+Additionally, numerous JVM settings can cause changes in energy consumption.
+Our experiments revealed that the JVM's default JIT compiler is frequently near-optimal in at least 80\% of the experiments.
+
+On the other hand , the default GC outperformed the other strategies in just half of our experiments, with some significant benefits reported when employing different GCs based on the application characteristics.
+Our primary observations and recommendations can be stated as follows:
 \begin{inparaenum}[\em i)]
@@
 \end{inparaenum}
-To ease the integration of the above guidelines, we propose a tool, named \textsf{J-Referral}, to recommend the most energy-efficient JVM distribution and configuration among more than a hundred considered possibilities.
-It establishes a full report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to help the user to choose the one with the least consumption for Java software.
-% TODO : add or change the graphes of the use case 
-% TODO : add the description and a diagrame for the experimental protocol
+
+To make the above principles easier to implement, we propose \emph{J-Referral}, a tool that recommends the best energy-efficient JVM distribution and configuration from among more than a hundred options.
+It generates a comprehensive report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration in order to assist the user in selecting the one with the lowest consumption for Java applications.

### 2022-09-23 chapters/optimization/python-dir/python_interperters.tex

@@
 
+%  Switch this to conclusion 
 To help practitioners to choose the best implementation for their use case, we have created a chart that describes the behavior of each implementation toward a specific aspect of programming.
@@
 Unfortunately, the \textsc{Tommti} benchmark is geared for general-purpose programming.
-Despite the fact that Nuitka is a compiler, there was no discernible difference in energy consumption.
+Despite the fact that Nuitxka is a compiler, there was no discernible difference in energy consumption.
 It was even more similar to the reference Python implementation than other interpreters.
@@
 The effect of the compilation of the python code into machine code was also confirmed with the use of translators such as shedskin. 
-
 This investigation is made more intriguing by the absence of a universal solution.
 The primary downside is the incompatibility of some of these solutions, which causes us to make concessions when we need a generic answer.
+we did a classification and clusturing of different python implementations based on their energy consumption.

### 2022-09-23 chapters/programming_languages/rpc/rpc.tex
refactoring

@@
 \paragraph{\textsc{RQ}~1: How do RPC implementations consume with regard to the size of the incoming request?}
+
 The purpose of this question is to study the energy consumption of the RPC server when transferring large objects.
@@
 
-\subsection{Threads to Validity}
+% \subsection{Threads to Validity \note{missing}}
 
-\subsection{Conclusion}
+\subsection{Conclusion \note{missing}}

### 2022-09-24 chapters/optimization/python-dir/python.tex
update : python chapter

@@
 \section{Introduction}
-% TODO : MOVE this to SOA
 
-% new structure : 
-%  introduction  / motivation -> definition of python -> the plan of experiments -> experiments 
 Dynamic programming languages, except Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf. \Cref{fig:pypl}).
 However, it remains unclear whether this category of dynamic programming languages can truly compete with compiled ones in terms of power consumption.
-
 \begin{figure}[htbp]
@@
 
-In this chapter, we explore the oblivious optimizations that can be applied to Python legacy applications to reduce their energy footprint.
-As Python is widely adopted by software services deployed in public and private cloud infrastructures, we believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
-More specifically, this chapter focuses on runtime optimizations that can be adopted by developers to leverage the power consumption of Python applications.
-We start by studying the impact of programmer choices, such as the type of data or control structures, on the global energy consumption of the execution code.
-Then, we discuss other factors, such as the levels of concurrency, before we investigate other non-intrusive approaches to optimize the energy consumption of applications.
-One type of optimization includes alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
+In this chapter, we aim to reduce the energy consumption of Python code.
+We first start by presenting the motivation behind our work in Section~\ref{python:sec_motivation}.
+Then we provide some insights on the impact of programmer choices in the second section.
+To do so, we will conduct a set of experiments on different Python aspects, from data structures and iteration mechanisms to the impact of training algorithms in machine learning, passing by the impact of the concurrency strategy and the use of orms in web development.
+Finally, in Section~\ref{python:sec_interpreters},we investigate other non-intrusive approaches to optimizing the energy consumption of applications by comparing multiple Python implementations, which include alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as ahead-of-time (AOT) compilation and just-in-time (JIT) libraries that are maintained by the community.
 
-\section{Motivation}
+\section{Motivation}\label{python:sec_motivation}
 
@@
 
-\subsection{Python Limits}
+\subsection{Python usecases}
 To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
@@
 
-In this chapter, we investigate the energy footprint of Python in its most popular domains of adoption.
+\section{Energy optimization of python on the source code level}\label{python:sec_insights}
+In this section , we investigate the energy footprint of Python in its most popular domains of adoption.
 We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016} did in~\cite{hasan_energy_2016}.
@@
 
-\section{Green Web Development}\label{sec:webdev}
+\subsection{Green Web Development}\label{sec:webdev}
 
@@
 
-\subsection{Life-cycle of a Request in Django}
+\subsubsection{Life-cycle of a Request in Django}
 Django is a MVT (\emph{Model-View-Template}) framework, which means that it follows the MVC (\emph{Model-View-Controller}) pattern.
@@
 
-\section{Python Insights}
+\subsection{Python Insights}
 This section aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
@@
 
-\subsection{experiment and results}
+\subsubsection{experiment and results}
 %%%% concurency 
@@
 
-\subsection{Synthesis}
+\subsubsection{Synthesis}
 This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
 
-\section{Python \& Multiprocessing}
+\subsection{Python \& Multiprocessing}
 
-\subsubsection{Cost of Parallelism in Python}
+% \subsubsection{Cost of Parallelism in Python}
 Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
@@
 
-\section{Python \& Machine Learning}
+\subsection{Python \& Machine Learning}
 Machine learning is becoming an integral part of our daily lives, growing more potent and energy-hungry each year.
 As machine learning can have a significant impact on climate change, it is vital to investigate mitigation techniques.
-\subsection{Experimental Protocol}
-\subsubsection{Measurement context}
-\paragraph{Hardware settings}
+\subsubsection{Experimental Protocol}
+\paragraph{Measurement context}
+\subparagraph{Hardware settings}
 Chifflot 8 from Grid 5000's Lille site was used for all of the trials.
@@
 
-\paragraph{Software settings}
+\subparagraph{Software settings}
 For the sake of reproducibility, each experiment is done within a Docker container using Jupyter lab.
@@
 \subsubsection{Input Workload}
-\paragraph{Models}
+\subparagraph{Models}
 Several models were developed, however, only two were used in the final trials because they achieved 94 percent accuracy in an acceptable length of time.
@@
 
-\paragraph{Datasets}
+\subparagraph{Datasets}
 The CIFAR-10 dataset was the major source of data for the studies.
@@
 
-\subsubsection{candidates}
+\paragraph{candidates}
 The experiments were run with several different CPU and GPU configurations:
@@
 
-\subsubsection{Key Performance Metrics}
+\paragraph{Key Performance Metrics}
 We used Pytorch 1.10.0 to train those models and pyjoules to measure the energy consumption of the GPU and CPU.
@@
 
-\subsection{Results \& Findings}
+\subsubsection{Results \& Findings}
 As the model's accuracy increases, so does the energy required for the next accuracy increment.
@@
 % \subsection{threads to validity \note{missing}}
+\skippage
+
 \section{Summary}: 
+As Python is widely adopted by software services deployed in public and private cloud infrastructures, we believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT but also reduce their cloud invoice for the resources consumed by these services.

### 2022-09-24 chapters/optimization/python-dir/python_interperters.tex
update : python chapter

--- 2022-09-14 chapters/optimization/python-dir/python_interperters.tex
@@
-\section{Python Interpreters}\label{sec:pythoninterpreters}
-%%% note : add this to the section of python interpreters
-Due to the lack of support for most non-conventional Python interpreters, we mainly focus on micro-benchmarks.
-Except for \textsc{PyPy}, most of the Python implementations do not support extra Python libraries, despite those extra implementations being developed to optimize a specific library, such as \textsf{Numba} with \textsf{Numpy}, or \textsf{intelpython} with machine learning algorithms.
-
-\subsection{Preliminary studies}
+\section{Python Interpreters}\label{python:sec_interpreters}
 For the first studies, we used the official version of Python, because the goal was mainly to highlight the impact of the structure of code on energy consumption.
 One main drawback of the previous method is the work to be done to update the existing code base to reduce  energy consumption.
+Despite the availability of numerous tools for code refactoring and automatic code optimization, the optimization process remains a manual effort.
+This is due to the fact that code optimization is a task that requires extensive understanding of both the codebase and the language itself.
+
 To avoid such hustle, we tried to find a non-intrusive approach to make the Python code more eco-friendly without altering its structure.
 Python is an interpreted language, which led many initiatives to implement their own interpreter to improve one or many aspects of the Python code. 
-In the following section, we discuss the impact of those implementations on the energy consumption of python programs, and in which case, one should use a non-conventional interpreter to save the energy consumption of their application.
+Therefore, nn this section, we will examine at how different Python interpreters affect the energy consumption of software,and in which case, one should use a non-conventional interpreter to save the energy consumption of their application.
 
+\subsection{Python runtimes}\label{python:sec_interpreters}
 To do so, we gathered a list of interpreters, transpilers and other optimization libraries that can contribute to reduce the energy consumption of legacy Python applications:
@@
 \subsection{Results \& Findings}
-This section will be dedicated to the results of the experiments and statistical analysis of these results.
-Because we are conducting a comparison study, we first did a Shapiro-Wilk~\cite{shapiro1968comparative} test to see if the data follows a normal distribution.
+This part will be dedicated to the results of the experiments and statistical analysis of these results.
+First and foremost, we want to know whether our samples have a normal distribution. As a result, we ran a Shapiro-Wilk~\cite{shapiro1968comparative} test on our data.
 Despite the fact that some implementations had a $p$-value higher than $a=0.05$ (such as CPython and ActivePython), other implementations like PyPy and Numba presented a $p$-values smaller than 0.01, which leads to reject the hypothesis $H0$ that all the distributions follow a normal distribution.
@@
 The Mann-Whitney U test results for each implementation are shown in Table \ref{table:manwithyu}.
-The first column displays the implementation's average energy usage, while the second provides the p-value of the test when compared to the reference Python implementation.
-If the p-value is less than 0.05, the test result is significant.
-This implies that the given implementation's energy consumption differs significantly from the reference implementation's.As one can notice, most of the implementations differ significantly from the reference implementation.
+The first column shows the average energy usage of the implementation, while the second shows the p-value of the test when compared to the reference Python implementation.
+If the p-value is less than 0.05, the test result is significant, indicating that the energy consumption of the given implementation differs significantly from that of the reference implementation.
+As can be seen, the majority of the implementations deviate significantly from the reference implementation.
 \input{table-envs-based-cpython.tex}
 
-As we want to study the position of these implementations with regards to the references, we proceed with agglomerative \emph{hierarchical cluster analysis} (HCA)~\cite{johnson1967hierarchical}.
+As we want to study the position of these implementations with regards to the reference implementation (aka Cpython2 and Cpython3), we proceed with agglomerative \emph{hierarchical cluster analysis} (HCA)~\cite{johnson1967hierarchical}.
 HCA is a method that groups the different implementations based on their energy consumption.
@@
 In Numba, the developer must target the portion of the code that will be passed to JIT, while the rest of the code will be executed by the regular interpreter (aka CPython).
+\bigskip
 
-\paragraph{Remark:}
-Because GraalPython was still in its early stages, some techniques required an abnormally lengthy time, affecting the clustering algorithm.
-
+%  Switch this to conclusion 
 To help practitioners to choose the best implementation for their use case, we have created a chart that describes the behavior of each implementation toward a specific aspect of programming.
@@
       \centering
-      \includegraphics[width=1.0\linewidth]{imgs/alltomti_performance}
+      \includegraphics[height=\textheight,keepaspectratio]{imgs/alltomti_performance}
       \caption{Different interpreter's energy scores.}
@@
 The lower the energy consumption, the higher the score.
-We adopted a logarithmic scale to help the practitioners to compare numerous implementations due to the large differences between them.
+We adopted a \textsc{logarithmic scale} to help the practitioners to compare numerous implementations due to the large differences between them.
 
@@
 Unlike the previous interpeters, GraalPython exhibits a certain promise when it comes to complex algorithms (nested loops in particular).
+However, it is still in its early stages of development and it is not yet ready for production.
+Since some benchmarks required an abnormally lengthy time, such as array manipulation and sorting.
+
 % Finally, MicroPython is dedicated to embedded systems, so claiming it is a powerful CPU machine will be misleading.
@@
 
-\subsection{Python \& Binary Operations}
-The second part of this study is to look at how different Python implementations work when executing different binary operations.
+% \subsection{Python \& Binary Operations}
+% The second part of this study is to look at how different Python implementations work when executing different binary operations.
 
-\begin{figure}[!hbt]
-      \small
-      \center
-      \begin{tabular}{|lrrrrr||r|}
-            \toprule
-            benchmark    & Addition & Right Rotation & Left Rotation & Or      & XOR     & Average     \\
-            \midrule
-            ActivePython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76      \\
-            CPython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43      \\
-            cpython3     & 595.209  & 685.085        & 563.839       & 657.972 & 655.560 & 531.53      \\
-            \hline
-            Cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26      \\
-            Nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2        \\
-            \hline
-            Numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & \best{9.22} \\
-            Numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & \best{9.59} \\
-            PyPy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & \best{8.13} \\
-            PyPy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & \best{8.21} \\
-            Shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & \best{8.18} \\
-            \bottomrule
-      \end{tabular}
-      \caption{Energy consumption of different implementations using bit operations benchmark (in Joules).}
-      \label{table:bitops}
-\end{figure}
+% \begin{figure}[!hbt]
+%       \small
+%       \center
+%       \begin{tabular}{|lrrrrr||r|}
+%             \toprule
+%             benchmark    & Addition & Right Rotation & Left Rotation & Or      & XOR     & Average     \\
+%             \midrule
+%             ActivePython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76      \\
+%             CPython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43      \\
+%             cpython3     & 595.209  & 685.085        & 563.839       & 657.972 & 655.560 & 531.53      \\
+%             \hline
+%             Cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26      \\
+%             Nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2        \\
+%             \hline
+%             Numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & \best{9.22} \\
+%             Numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & \best{9.59} \\
+%             PyPy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & \best{8.13} \\
+%             PyPy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & \best{8.21} \\
+%             Shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & \best{8.18} \\
+%             \bottomrule
+%       \end{tabular}
+%       \caption{Energy consumption of different implementations using bit operations benchmark (in Joules).}
+%       \label{table:bitops}
+% \end{figure}
 
-Table~\ref{table:bitops} provides a detailed report on the energy consumption of different implementations of the code.
-Besides \emph{Cython}, each implementation treats the binary operation the same.
-As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
-However, when it comes to Cython, the rotation operations tend to consume 5 times more energy than the other operations, and left rotations consume even more than the right ones.
-The reason behind such behavior is the fact that Cython does not have a native implementation for these two operations, so it replaces them with more complex operations, which leads to an overhead of $6s$ for the right rotation and $10s$ for the left, compared to $1.5s$ for the other operations.
-Therefore, it is safe to summarize all the operations into one that we call \emph{binary}.
+% Table~\ref{table:bitops} provides a detailed report on the energy consumption of different implementations of the code.
+% Besides \emph{Cython}, each implementation treats the binary operation the same.
+% As one can see, Table~\ref{table:bitops} shows no difference in the energy consumption of those operations.
+% However, when it comes to Cython, the rotation operations tend to consume 5 times more energy than the other operations, and left rotations consume even more than the right ones.
+% The reason behind such behavior is the fact that Cython does not have a native implementation for these two operations, so it replaces them with more complex operations, which leads to an overhead of $6s$ for the right rotation and $10s$ for the left, compared to $1.5s$ for the other operations.
+% Therefore, it is safe to summarize all the operations into one that we call \emph{binary}.
 
-\begin{figure}
-      \centering
-      \includegraphics[width=\linewidth]{imgs/bitopts_mean}
-      \caption{Summary of the binary operations for different python implementations }
-      \label{fig:bitops}
-\end{figure}
+% \begin{figure}
+%       \centering
+%       \includegraphics[width=\linewidth]{imgs/bitopts_mean}
+%       \caption{Summary of the binary operations for different python implementations }
+%       \label{fig:bitops}
+% \end{figure}
 
-Figure~\ref{fig:bitops} depicts the average energy consumption of different implementations for the binary operations.
-One can notice 3 clusters.
-The first, which is the interpreted one, uses approximately 500 Joules per operation.
-The second is a non-optimized compilation that includes Cython and Nuitka and uses 10 times less energy than the first.
-And the final one, which has been fully optimized, uses 50 percent less energy.
+% Figure~\ref{fig:bitops} depicts the average energy consumption of different implementations for the binary operations.
+% One can notice 3 clusters.
+% The first, which is the interpreted one, uses approximately 500 Joules per operation.
+% The second is a non-optimized compilation that includes Cython and Nuitka and uses 10 times less energy than the first.
+% And the final one, which has been fully optimized, uses 50 percent less energy.
 This category includes JIT libraries, also known as Numba, Python implementations that include JIT, also known as PyPy, and Shedskin, which transpile the source code into C and then use a C/C++ compiler to generate the binary, as opposed to Cython and Nuitka, which directly compile the Python code.
-To conclude this experiment, JIT has a huge impact when it comes to binary operations, and an inadequate runtime selection can lead to a 100x increase in energy consumption.
-Furthermore, we observed that different binary operations exhibit the same energy footprint for all the implementations, except Cython.
+% To conclude this experiment, JIT has a huge impact when it comes to binary operations, and an inadequate runtime selection can lead to a 100x increase in energy consumption.
+% Furthermore, we observed that different binary operations exhibit the same energy footprint for all the implementations, except Cython.
+
+\subsubsection{threads to validity }
+Due to the lack of support for most non-conventional Python interpreters, we mainly focus on micro-benchmarks.
+Except for \textsc{PyPy}, most of the Python implementations do not support extra Python libraries, despite those extra implementations being developed to optimize a specific library, such as \textsf{Numba} with \textsf{Numpy}, or \textsf{intelpython} with machine learning algorithms.
+Furthermore, we excluded from this study runtimes that were unable to complete our benchmarks because we judged that this was the bare minimum for an implementation to be employed in a real-world application.
+Besides this, since the goal of this study was to provide a non-intrusive technique to decrease the energy consumption of software, we opted to omit some of the implementations that required extra work on the sourcecode to run on them, such as \textsf{Cython}.
 
 \subsection{Conclusion}
-One can conclude that the choice of Python interpreter has a significant impact on the programs' energy consumption.
-This investigation is made more intriguing by the absence of a universal solution.
-The primary downside is the incompatibility of some of these solutions, which causes us to make concessions when we need a generic answer.
+
+One can conclude that the Python interpreter used has a considerable impact on the energy consumption of the programs.
+Furthermore, the absence of a universal solution makes this invistigation more intriguing.
+While Python2 was employed by the majority of the candidates, Python3 provided better compatibility coverage.
+Except for Pypy, a standout interpreter, the type of optimization strategy had a substantial impact on energy usage because each class was grouped together regardless of Python version.
+Finally The use of JIT, a technology that converts code into machine code at runtime, was responsible for the significant gain in energy efficiency.
+The usage of translators such as Shedskin confirmed the effect of compiling Python code into machine code.
+Hence, making PyPy, an interpreter that has a built-in JIT, within the same class as the complied ones such as Shedskin

### 2022-09-24 chapters/optimization/python-dir/python.tex
refactor : python experiments order

@@
 However, it remains unclear whether this category of dynamic programming languages can truly compete with compiled ones in terms of power consumption.
-\begin{figure}[htbp]
+\begin{figure}[!htb]
     \includegraphics[width=\linewidth]{imgs/programminglanguangespopularity.png}
@@
 According to their experiments, an interpreted programming language, like Python, can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. \Cref{fig:hannoi}).
-\begin{figure}[htbp]
+\begin{figure}[!htb]
     \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
@@
 
+% TODO proof read this 
 In this chapter, we aim to reduce the energy consumption of Python code.
@@
 Its flexibility and simplicity caused it to drop off in performance because Python gains its flexibility from being a dynamic language.
-Therefore, it requires a faster interpreter to execute its programs to compete against alternatives written in compiled programming languages, such as C and C++ or semi-compiled languages like Java.
+Therefore, it requires a faster interpreter to execute its programs to compete against alternatives written in compiled programming languages, such as C and C++, or semi-compiled languages like Java.
 
@@
 Then, we dive deeper into the energy consumption of Python core structures, before concluding with the impact of the Python interpreter on energy consumption.
-
-\subsection{Green Web Development}\label{sec:webdev}
-
-Django\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}} are the most popular frameworks for web development.
-According to a Jetbrains poll,\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.
-
-In contrast to Flask, which is a micro-framework, Django is a high-level web framework that provides a standard method for creating and maintaining complex and scalable database-driven websites quickly and effectively.
-Therefore, this study will focus on the latter one.
-
-\subsubsection{Life-cycle of a Request in Django}
-Django is a MVT (\emph{Model-View-Template}) framework, which means that it follows the MVC (\emph{Model-View-Controller}) pattern.
-Figure~\ref{fig:django-life-cycle} describes the life-cycle of a request in this framework.
-Whenever a request arrives in Django, it is processed by the middleware layers, one at a time.
-These middleware layers are in charge of authentication, security, and so on.
-Once the request is processed by these layers, it is passed to the URL router, which extracts the URL from the request and tries to match it to the defined URLs.
-After getting the matching URL, the corresponding view function is called.
-This function is responsible for treating the request, gathering the data, and then generating the response that will be put inside a template to be returned as a response.
-As Django adopts a MVT model, it also offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.
-
-\begin{figure}[hbt]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/django_request_lifecycle}
-    \caption{Request-Response life cycle in Django}
-    \label{fig:django-life-cycle}
-\end{figure}
-
-First, we start by investigating the energy consumption of the request-response life cycle in Django, to determine which layer consumes the most energy.
-To do so, we created a sample Django application that returns the response of the request.
-We tracked its energy consumption using JouleHunter.\footnote{\url{https://github.com/powerapi-ng/joulehunter}}
-JouleHunter is an open-source library that I developed to help practitioners to identify  energy hotspots in their applications using statistical profilers.
-In the case of Django, it can be added as an extension with no additional setup or change to the source code.
-The energy consumption of the request-response life-cycle in Django is shown in Figure~\ref{fig:django_life_cycle_naive}.
-As we can notice, 91.4\% of the total energy consumption is spent to resolve the request by retrieving the data, while only 5.27\% is spent to render the response.
-
-\begin{figure}[hbt]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/django_life_cycle_naive}
-    \caption{Tree representation of the energy consumption of a single request in django (naive version)}
-    \label{fig:django_life_cycle_naive}
-\end{figure}
-
-%% REMARQUE the experiment with joulehunter isn't conducted within the same machine that is used for generating the boxplot graphs 
-% We have chosen Django as our web framework.
-Therefore, we chose to study if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
-% \subsection{Experiment \& Results}
-To accomplish so, we examined the cost of a single request of the prior website using various ways to extract data from the database.
-We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records:
-\begin{enumerate}
-    \item \textsf{Vanilla} relies only on the ORM to retrieve the data,
-    \item \textsf{Prefetch} queries the data before being requested,
-    \item \textsf{Optimized} leverages SQL without passing by the ORM.
-\end{enumerate}
-
-%%%%% have to rephrase this 
-As one can observe in \Cref{fig:django}, the strategy to query stored data has a huge impact on energy consumption.
-As the \textsf{Vanilla} strategy can consume up to $10\times$ more energy than the \textsf{Optimized} one.
-Conversely, the choice of the database does not exhibit a key impact on the total energy, despite their different behavior regarding the execution time and the average power.
-This can be useful to support developers in choosing which database engine they can adopt, guided by the number of expected requests and the targeted performance.
-
-\begin{figure}[!hbt]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/django}
-    \caption{Energy behavior resulting from data access strategies.}
-    \label{fig:django}
-\end{figure}
-
-Another interesting observation is the impact of the interpreter, as \Cref{fig:django} highlights.
-For example, using the \textsf{PyPy} interpreter reduces the energy consumption, even when adopting the \textsf{Vanilla} strategy.
-
-Finally, we run the same experiment with the \textsf{Optimized} strategy using JouleHunter.
-Figure~\ref{fig:django_profiled_optimized} depicts the resulting energy consumption.
-As one can notice, while the rendering method consumed the same amount of energy as in the \textsf{Vanilla} strategy (around 1.3\,kJ), the resolve part dropped by $20\times$.
-
-% \subsection{Conclusion}
-We can therefore conclude that the database and ORM selection have a major impact on the website's performance and energy consumption.
-And, unlike the rendering portion, there are many options available for the database, the ORM, and the data handling strategy.
-
-\begin{figure}[hbt]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/django_profiled_optimized}
-    \caption{Tree representation of the energy consumption of a single request in Django (naive version)}
-    \label{fig:django_profiled_optimized}
-\end{figure}
-
-% Another key finding is when we hit the number of physical cores.
-there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
-Basically it favors the hyper threads than the physical core which will lead to some context switches which cause the slow behavior, however in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the energy consumption in this case.
-% in the Chapter %% reference the work of intern from lyon  
-% we discuss a deeper this behavior of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there
-
-\subsection{Python Insights}
-This section aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
-Therefore, we start by extending the work of Hasan~\emph{et~al.} and Oliveira~\emph{et~al.} to the context of Python environments
-\cite{hasan_energy_2016,oliveira_recommending}.
-
-\subsubsection{experiment and results}
-%%%% concurency 
-Another field of investigation is the type of data and control structures that might impact the energy consumption.
-We iterate over a list using 3 methods.
-First, the classical \texttt{for(i in range(len(n)))}.
-However, as we can see here, unlike other programming languages, it requires extra operations, such as determining the length of the collection and then using the iterator range.
-So, we tried the more adapted version \texttt{for(element in collection)}.
-Moreover, in most programming languages.
-the \texttt{for} loop is translated to a \texttt{while} loop ( transformation from D type to B type -- asm -- ), therefore we wanted to compare this with a \texttt{while} version.
-After determining the main ways to iterate over a \texttt{loop}, we run the collection algorithms of different data types to observe if they impact the energy consumption of the code, and we repeated it for the size of the collection.
-
-As one can see in \Cref{fig:pythonloops}, the type of the data has no impact on the energy consumption.
-However, the way one iterates over the collection has a huge factor.
-Interestingly, the \texttt{for in range} loop was by far the optimal one, followed by the regular \texttt{for in collection}, and the \texttt{while} part was the last one with an overhead of 400\% compared to the first option.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/python_iterations}
-    \caption{Comparison of the energy consumption of different Python loops.}
-    \label{fig:pythonloops}
-\end{figure}
-
-%%% putting the idea, later i ll reformulate it 
-The reason behind such a behavior is mainly related to how the Python interpreter is implemented\footnote{\url{https://www.python.org/doc/essays/list2str}},\footnote{\url{https://www.pythonpool.com/for-vs-while-loop-python/}}.
-To reduce the latency of Python applications, most of the built-in functions and operations are written in C, and the same goes for the function \texttt{range}.\footnote{\url{ http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1}}
-Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$ as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object~\footnote{\url{https://wiki.python.org/moin/TimeComplexity}}.
-Therefore, the \texttt{for in range} is creating a new iterator that has the same length as the first one and, for each iteration, requires second access (\texttt{l[i]}) instead of one---explaining the doubled time.
-The \texttt{while} is even slower due to the implicit increment of the variable, which causes an extra operation during the loop.
-To confirm this hypothesis, we tried to construct a new list by editing the elements of the previous one (cf. \Cref{fig:pythontreatement}).
-And, as predicted, the built-in methods are the most energy-saving ones, while the customize \texttt{while} loop is the heaviest.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/python_treatemens}
-    \caption{Comparison of the energy consumption of different methods to convert a list.}
-    \label{fig:pythontreatement}
-\end{figure}
-
-Another interesting finding is the impact of anonymous functions (also known as lambda expressions) on energy consumption.
-The reason is that Python treats these functions as local variables, unlike the predefined ones which are global in our case.
-Therefore, they are faster and consume less energy.  % same reference https://www.python.org/doc/essays/list2str/
-
-\subsubsection{Synthesis}
-This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
-
-\subsection{Python \& Multiprocessing}
-
-% \subsubsection{Cost of Parallelism in Python}
-Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
-Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).
-However, due to the increase of cores/threads per CPU, many Python libraries started to take advantage of this hardware feature by allowing concurrent execution of Python processes.
-For example, the multiprocessing library\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}} spawns sub-processes to increase the degree of concurrency of Python applications.
-As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of threads until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.
-
-% \begin{figure}[hbt]
-%     \centering
-%     \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-%     \caption{Energy consumption of Python multiprocessing depending on the number of exploited threads.}
-%     \label{fig:python_multiprocessing}
-% \end{figure}
-
-Before reaching the limit of physical cores, we also observe that the scheduler of the operating system tends to favor the execution of processes on the same physical core, by taking advantage of the hyper-thread feature.
-While this strategy aims to save energy by leveraging the ACPI P-states and C-states of unallocated cores, this leads to increase execution times.
-
-\Cref{fig:python_multiprocessing} compares the behavior of python programs when we try to introduce parallelism.
-As we know python is a single-threaded program thanks to the GLS (\emph{Global Lock System}) however, due to the increase in the number of cores/threads per CPU, many libraries started to take advantage of this hardware feature by simulating multithreading as multiple instances of the processor.
-% TODO: TO VERIFY
-
-\begin{figure}[hbt]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-    \caption{Energy behavior when using multiprocessing.}
-    \label{fig:python_multiprocessing}
-\end{figure}
-
-As we can see in Figure~\ref{fig:python_multiprocessing}, the energy consumption correlates with the number of threads until reaching the number of physical cores, then losing the benefit from multiprocessing by facing concurency issues imposed by different sub-processes competing for the CPU resources.
-Another finding is observed when hitting the number of physical cores: there is an increase of execution time, but with a reduced energy.
-The reason behind this is the scheduler of the operating system.
-More specifically, it favors hyper-threads instead of physical cores, hence increasing context switches which cause the performance bottleneck.
-Howerver,the other two physical cores are not consuming energy, neither their hyper-threads, which explains the energy savings.
-In the Chapter , %% reference the work of intern from lyon  
-we discuss this behavior of the scheduler more deeply and we assess that it is not due to Python, but is a more generic behavior. %% maybe even this graph should go there
-
-%% todo 
-% add performance 
-% add dram \ a glimps
 
@@
 \end{figure}
+
 \begin{figure}
@@
 
+\subsection{Green Web Development}\label{sec:webdev}
+
+Django\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}} are the most popular frameworks for web development.
+According to a Jetbrains poll,\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.
+
+In contrast to Flask, which is a micro-framework, Django is a high-level web framework that provides a standard method for creating and maintaining complex and scalable database-driven websites quickly and effectively.
+Therefore, this study will focus on the latter one.
+
+\subsubsection{Life-cycle of a Request in Django}
+Django is a MVT (\emph{Model-View-Template}) framework, which means that it follows the MVC (\emph{Model-View-Controller}) pattern.
+Figure~\ref{fig:django-life-cycle} describes the life-cycle of a request in this framework.
+Whenever a request arrives in Django, it is processed by the middleware layers, one at a time.
+These middleware layers are in charge of authentication, security, and so on.
+Once the request is processed by these layers, it is passed to the URL router, which extracts the URL from the request and tries to match it to the defined URLs.
+After getting the matching URL, the corresponding view function is called.
+This function is responsible for treating the request, gathering the data, and then generating the response that will be put inside a template to be returned as a response.
+As Django adopts a MVT model, it also offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django_request_lifecycle}
+    \caption{Request-Response life cycle in Django}
+    \label{fig:django-life-cycle}
+\end{figure}
+
+First, we start by investigating the energy consumption of the request-response life cycle in Django, to determine which layer consumes the most energy.
+To do so, we created a sample Django application that returns the response to the request.
+We tracked its energy consumption using JouleHunter~\footnote{\url{https://github.com/powerapi-ng/joulehunter}}.
+JouleHunter is an open-source library that I developed to help practitioners identify energy hotspots in their applications using statistical profilers.
+In the case of Django, it can be added as an extension with no additional setup or change to the source code.
+The energy consumption of the request-response life-cycle in Django is shown in Figure~\ref{fig:django_life_cycle_naive}.
+As we can notice, 91.4\% of the total energy consumption is spent on resolving the request by retrieving the data, while only 5.27\% is spent on rendering the response.
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django_life_cycle_naive}
+    \caption{Tree representation of the energy consumption of a single request in django (naive version)}
+    \label{fig:django_life_cycle_naive}
+\end{figure}
+
+%% REMARQUE the experiment with joulehunter isn't conducted within the same machine that is used for generating the boxplot graphs 
+% We have chosen Django as our web framework.
+Therefore, we chose to study if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
+% \subsection{Experiment \& Results}
+To accomplish so, we examined the cost of a single request of the prior website using various ways to extract data from the database.
+We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records:
+\begin{enumerate}
+    \item \textsf{Vanilla} relies only on the ORM to retrieve the data,
+    \item \textsf{Prefetch} queries the data before being requested,
+    \item \textsf{Optimized} leverages SQL without passing by the ORM.
+\end{enumerate}
+
+%%%%% have to rephrase this 
+As one can observe in \Cref{fig:django}, the strategy to query stored data has a huge impact on energy consumption.
+As the \textsf{Vanilla} strategy can consume up to $10\times$ more energy than the \textsf{Optimized} one.
+Conversely, the choice of the database does not exhibit a key impact on the total energy, despite their different behavior regarding the execution time and the average power.
+This can be useful to support developers in choosing which database engine they can adopt, guided by the number of expected requests and the targeted performance.
+
+\begin{figure}[!hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django}
+    \caption{Energy behavior resulting from data access strategies.}
+    \label{fig:django}
+\end{figure}
+
+Another interesting observation is the impact of the interpreter, as \Cref{fig:django} highlights.
+For example, using the \textsf{PyPy} interpreter reduces energy consumption, even when adopting the \textsf{Vanilla} strategy.
+We will further discuss the choice of the interpreter in the next section.
+
+Finally, we run the same experiment with the \textsf{Optimized} strategy using JouleHunter.
+Figure~\ref{fig:django_profiled_optimized} depicts the resulting energy consumption.
+As one can notice, while the rendering method consumed the same amount of energy as in the \textsf{Vanilla} strategy (around 1.3\,kJ), the resolve part dropped by $20\times$.
+
+% \subsection{Conclusion}
+We can therefore conclude that the database and ORM selection have a major impact on the website's performance and energy consumption.
+And, unlike the rendering portion, there are many options available for the database, the ORM, and the data handling strategy.
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/django_profiled_optimized}
+    \caption{Tree representation of the energy consumption of a single request in Django (naive version)}
+    \label{fig:django_profiled_optimized}
+\end{figure}
+
+\subsection{Python Insights}
+This part aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
+Therefore, we start by extending the work of Hasan~\emph{et~al.} and Oliveira~\emph{et~al.} to the context of Python environments
+\cite{hasan_energy_2016,oliveira_recommending}.
+
+\subsubsection{experiment and results}
+%%%% concurency 
+Another field of investigation is the type of data and control structures that might impact the energy consumption.
+We iterate over a list using 3 methods.
+First, the classical \texttt{for(i in range(len(n)))}.
+However, as we can see here, unlike other programming languages, it requires extra operations, such as determining the length of the collection and then using the iterator range.
+So, we tried the more adapted version \texttt{for(element in collection)}.
+Moreover, in most programming languages.
+the \texttt{for} loop is translated to a \texttt{while} loop ( transformation from D type to B type -- asm -- ), therefore we wanted to compare this with a \texttt{while} version.
+After determining the main ways to iterate over a \texttt{loop}, we run the collection algorithms of different data types to observe if they impact the energy consumption of the code, and we repeated it for the size of the collection.
+
+As one can see in \Cref{fig:pythonloops}, the type of the data has no impact on the energy consumption.
+However, the way one iterates over the collection has a huge factor.
+Interestingly, the \texttt{for in range} loop was by far the optimal one, followed by the regular \texttt{for in collection}, and the \texttt{while} part was the last one with an overhead of 400\% compared to the first option.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/python_iterations}
+    \caption{Comparison of the energy consumption of different Python loops.}
+    \label{fig:pythonloops}
+\end{figure}
+
+The reason behind such a behavior is mainly related to how the Python interpreter is implemented\footnote{\url{https://www.python.org/doc/essays/list2str}},\footnote{\url{https://www.pythonpool.com/for-vs-while-loop-python/}}.
+To reduce the latency of Python applications, most of the built-in functions and operations are written in C, and the same goes for the function \texttt{range}.\footnote{\url{ http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1}}
+Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$ as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object~\footnote{\url{https://wiki.python.org/moin/TimeComplexity}}.
+Therefore, the \texttt{for in range} is creating a new iterator that has the same length as the first one and, for each iteration, requires second access (\texttt{l[i]}) instead of one---explaining the doubled time.
+The \texttt{while} is even slower due to the implicit increment of the variable, which causes an extra operation during the loop.
+To confirm this hypothesis, we tried to construct a new list by editing the elements of the previous one (cf. \Cref{fig:pythontreatement}).
+And, as predicted, the built-in methods are the most energy-saving ones, while the customize \texttt{while} loop is the heaviest.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/python_treatemens}
+    \caption{Comparison of the energy consumption of different methods to convert a list.}
+    \label{fig:pythontreatement}
+\end{figure}
+
+Another interesting finding is the impact of anonymous functions (also known as lambda expressions) on energy consumption.
+The reason is that Python treats these functions as local variables, unlike the predefined ones, which are global in our case.
+Therefore, they are faster and consume less energy.
+
+\subsubsection{Synthesis}
+This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
+
+\subsection{Python \& Multiprocessing}
+
+% \subsubsection{Cost of Parallelism in Python}
+Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
+Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).
+However, due to the increase of cores/threads per CPU, many Python libraries started to take advantage of this hardware feature by allowing concurrent execution of Python processes.
+For example, the multiprocessing library\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}} spawns sub-processes to increase the degree of concurrency of Python applications.
+As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of threads until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.
+
+% \begin{figure}[hbt]
+%     \centering
+%     \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
+%     \caption{Energy consumption of Python multiprocessing depending on the number of exploited threads.}
+%     \label{fig:python_multiprocessing}
+% \end{figure}
+
+Before reaching the limit of physical cores, we also observe that the scheduler of the operating system tends to favor the execution of processes on the same physical core, by taking advantage of the hyper-thread feature.
+While this strategy aims to save energy by leveraging the ACPI P-states and C-states of unallocated cores, this leads to increase execution times.
+
+\Cref{fig:python_multiprocessing} compares the behavior of python programs when we try to introduce parallelism.
+As we know python is a single-threaded program thanks to the GLS (\emph{Global Lock System}) however, due to the increase in the number of cores/threads per CPU, many libraries started to take advantage of this hardware feature by simulating multithreading as multiple instances of the processor.
+% TODO: TO VERIFY
+
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
+    \caption{Energy behavior when using multiprocessing.}
+    \label{fig:python_multiprocessing}
+\end{figure}
+
+As we can see in Figure~\ref{fig:python_multiprocessing}, the energy consumption correlates with the number of threads until reaching the number of physical cores, then losing the benefit from multiprocessing by facing concurency issues imposed by different sub-processes competing for the CPU resources.
+Another finding is observed when hitting the number of physical cores: there is an increase of execution time, but with a reduced energy.
+The reason behind this is the scheduler of the operating system.
+More specifically, it favors hyper-threads instead of physical cores, hence increasing context switches which cause the performance bottleneck.
+Howerver,the other two physical cores are not consuming energy, neither their hyper-threads, which explains the energy savings.
+In the Chapter , %% reference the work of intern from lyon  
+we discuss this behavior of the scheduler more deeply and we assess that it is not due to Python, but is a more generic behavior. 
+%% maybe even this graph should go there 
+% Another key finding is when we hit the number of physical cores.
+there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
+Basically it favors the hyper threads than the physical core which will lead to some context switches which cause the slow behavior, however in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the energy consumption in this case.
+% in the Chapter %% reference the work of intern from lyon  
+% we discuss a deeper this behavior of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there 
+%% todo 
+% add performance 
+% add dram \ a glimps
+
 \input{python_interperters.tex}
 % \subsection{threads to validity \note{missing}}
-\skippage
-
-\section{Summary}: 
-As Python is widely adopted by software services deployed in public and private cloud infrastructures, we believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT but also reduce their cloud invoice for the resources consumed by these services.
+\clearpage
+\section{Summary} 
+As Python is widely adopted by software services deployed in public and private cloud infrastructures, Imporving the energy efficiency of python based applications will push the carbon emissions of ICT a step further.
+This chapters presents differents approaches to improve the energy efficiency of python based applications.
+First it presents the reasons behind a such choice, then setup a strategy to improve the energy of the most use cases which revealed to be web developement and machine learning.
+For the web developements, we have shown that most of the energy is consumed while treating requests, hence choosing the right methode can decrease the energy consumption by 10 times. 
+After that we, we stdied the energy consumption for the machine learning usecase and we have shown that the energy consumption is not linear with the accuracy of the model.
+therefore a cut in the accuracy can lead to a significant energy saving.
+After that we went deeper to study how data structures, parallelism and iterating mechanism can affect the energy consumption. 
+As for the second section we provided a non intrusive solution to optimize the energy consumption of python based applications.
+without signigicant changes in the code.
+which is the use of alternative python interpreters.
+Section 2 started by a classification and filtering these implemenation, to come up with three major types (compiler,interpreter and extra libraries).Then we run a set of experiments to compare the energy consumption of these interpreters.
+The results shown the absence of a generic solution, and the need to choose the right interpreter for the right use case.
+Some of these solutions can be the best in somes and at the same time the worst in others.
+Finally we have shown that the most efficient way to increase the energy efficiency of python based applications is the use of JIT, regardless of  implementation. 
+We believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT but also reduce their cloud invoice for the resources consumed by these services.

### 2022-09-27 chapters/optimization/python-dir/python.tex
update

@@
 \subsection{Python Gluttony}
-% In the other hand,because of the interpreted nature of Python code, this programming language has paid a high price in terms of performance.  \ref{fig:clbg}, memory and energy consumption.
+% In the other hand,because of the interpreted nature of Python code, this programming language has paid a high price in terms of performance~\ref{fig:clbg}, memory and energy consumption.
 According to~\cite{pinto_energy_2017} and~\cite{noureddine_preliminary_2012}, Python tends to be one more energy hungry programming language.
@@
 \Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers.\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}
-57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration.\footnote{The options in this survey were not mutually exclusive.As a result, the total of the percentages is greater than $100\%$.}
+57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration~\footnote{The options in this survey were not mutually exclusive.As a result, the total of the percentages is greater than $100\%$.}.
 
@@
 To answer these questions, we report on 4 case studies that intend to answer these research questions.
-First, we study the energy behavior of Python in two application contexts: web applications (cf. \Cref{sec:webdev}) and machine learning.
+First, we study the energy behavior of Python in two application contexts: machine learning (cf. \Cref{sec:webdev}) web applications (cf. \Cref{sec:ml}).
 Then, we dive deeper into the energy consumption of Python core structures, before concluding with the impact of the Python interpreter on energy consumption.
 
-\subsection{Python \& Machine Learning}
+\subsection{Python \& Machine Learning}\label{sec:ml}
 Machine learning is becoming an integral part of our daily lives, growing more potent and energy-hungry each year.
@@
 \subsubsection{Results \& Findings}
+% TODO : Missing analysis of the results
 As the model's accuracy increases, so does the energy required for the next accuracy increment.
@@
 
-\begin{figure}[hbt]
+\begin{figure}[!htb]
     \centering
@@
 
-%%%%% have to rephrase this 
 As one can observe in \Cref{fig:django}, the strategy to query stored data has a huge impact on energy consumption.
@@
 
-% \subsection{Conclusion}
 We can therefore conclude that the database and ORM selection have a major impact on the website's performance and energy consumption.
@@
 
-\begin{figure}[hbt]
+\begin{figure}[!hbt]
     \centering
@@
 \subsection{Python Insights}
-This part aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
-Therefore, we start by extending the work of Hasan~\emph{et~al.} and Oliveira~\emph{et~al.} to the context of Python environments
-\cite{hasan_energy_2016,oliveira_recommending}.
+% This part aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
+% Therefore, we start by extending the work of Hasan~\emph{et~al.} and Oliveira~\emph{et~al.} to the context of Python environments
+% \cite{hasan_energy_2016,oliveira_recommending}.
+Another field of investigation is the type of data and control structures that might impact the energy consumption.
 
-\subsubsection{experiment and results}
-%%%% concurency 
-Another field of investigation is the type of data and control structures that might impact the energy consumption.
 We iterate over a list using 3 methods.
@@
 
+\begin{figure}[hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
+    \caption{Energy consumption of Python multiprocessing depending on the number of exploited threads.}
+    \label{fig:python_multiprocessing}
+\end{figure}
+
 % \begin{figure}[hbt]
@@
 
-Before reaching the limit of physical cores, we also observe that the scheduler of the operating system tends to favor the execution of processes on the same physical core, by taking advantage of the hyper-thread feature.
-While this strategy aims to save energy by leveraging the ACPI P-states and C-states of unallocated cores, this leads to increase execution times.
+Before reaching the limit of physical cores, we also observe that the scheduler of the operating system tends to favor the execution of processes on the same physical core by taking advantage of the hyper-thread feature.
+While this strategy aims to save energy by leveraging the ACPI P-states and C-states of unallocated cores, this leads to increased execution times.
 
-\Cref{fig:python_multiprocessing} compares the behavior of python programs when we try to introduce parallelism.
-As we know python is a single-threaded program thanks to the GLS (\emph{Global Lock System}) however, due to the increase in the number of cores/threads per CPU, many libraries started to take advantage of this hardware feature by simulating multithreading as multiple instances of the processor.
-% TODO: TO VERIFY
+\Cref{fig:python_multiprocessing} compares the behavior of Python programs when we try to introduce parallelism.
 
-\begin{figure}[hbt]
-    \centering
-    \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-    \caption{Energy behavior when using multiprocessing.}
-    \label{fig:python_multiprocessing}
-\end{figure}
+As we know, Python is a single-threaded program thanks to the GLS (\emph{Global Lock System}).
+However, due to the increase in the number of cores/threads per CPU, many libraries have started to take advantage of this hardware feature by simulating multi-threading as multiple instances of the processor.
 
@@
 \section{Summary} 
-As Python is widely adopted by software services deployed in public and private cloud infrastructures, Imporving the energy efficiency of python based applications will push the carbon emissions of ICT a step further.
-This chapters presents differents approaches to improve the energy efficiency of python based applications.
-First it presents the reasons behind a such choice, then setup a strategy to improve the energy of the most use cases which revealed to be web developement and machine learning.
-For the web developements, we have shown that most of the energy is consumed while treating requests, hence choosing the right methode can decrease the energy consumption by 10 times. 
-After that we, we stdied the energy consumption for the machine learning usecase and we have shown that the energy consumption is not linear with the accuracy of the model.
-therefore a cut in the accuracy can lead to a significant energy saving.
-After that we went deeper to study how data structures, parallelism and iterating mechanism can affect the energy consumption. 
-As for the second section we provided a non intrusive solution to optimize the energy consumption of python based applications.
-without signigicant changes in the code.
-which is the use of alternative python interpreters.
-Section 2 started by a classification and filtering these implemenation, to come up with three major types (compiler,interpreter and extra libraries).Then we run a set of experiments to compare the energy consumption of these interpreters.
-The results shown the absence of a generic solution, and the need to choose the right interpreter for the right use case.
-Some of these solutions can be the best in somes and at the same time the worst in others.
-Finally we have shown that the most efficient way to increase the energy efficiency of python based applications is the use of JIT, regardless of  implementation. 
-We believe that our contributions will benefit a wide diversity of legacy systems and not only favorably contribute to reducing the carbon emissions of ICT but also reduce their cloud invoice for the resources consumed by these services.
+Because Python is widely used by software services deployed in public and private cloud infrastructures, increasing the energy efficiency of Python-based applications will reduce ICT carbon emissions.
+This chapter discusses various ways for increasing the energy efficiency of Python-based applications.
+It first explains why such a choice was made, and then it develops a method to enhance the energy efficiency of the most commonly employed use cases, which are revealed to include web development and machine learning.
+In web development, we found that the majority of the energy is wasted while processing requests, therefore adopting the proper strategy can reduce energy consumption by tenfold.
+Following that, we looked at the energy usage for the machine learning usecase and discovered that it is not proportional to model accuracy.
+As a result, a reduction in accuracy can result in large energy savings.
+Following that, we dug deeper to investigate how data structures, parallelism, and iterative methods affect energy use.
+In the second section, we presented a simple technique for optimizing the energy usage of Python-based apps. without making substantial changes to the code This is the use of many Python interpreters
+Section 2 started by categorizing and filtering these implementations into three major kinds (compiler, interpreter, and extra libraries).
+Then we conducted a series of experiments to compare the energy consumption of various interpreters.
+The findings indicate the lack of a general solution and the importance of selecting the appropriate interpreter for the appropriate use case.
+In certain circumstances, some of these answers may be the best, while in others, they may be the worst.
+Finally, regardless of implementation, we demonstrated that using JIT is the most efficient technique to boost the energy efficiency of Python-based programs.
+We believe that our contributions will benefit a wide spectrum of legacy systems, not only reducing ICT's carbon footprint, but also lowering their cloud bills for the resources these services need.

### 2022-09-27 chapters/optimization/python-dir/python.tex
add : machine learning analysis

@@
 
-% TODO proof read this 
 In this chapter, we aim to reduce the energy consumption of Python code.
 We first start by presenting the motivation behind our work in Section~\ref{python:sec_motivation}.
-Then we provide some insights on the impact of programmer choices in the second section.
-To do so, we will conduct a set of experiments on different Python aspects, from data structures and iteration mechanisms to the impact of training algorithms in machine learning, passing by the impact of the concurrency strategy and the use of orms in web development.
-Finally, in Section~\ref{python:sec_interpreters},we investigate other non-intrusive approaches to optimizing the energy consumption of applications by comparing multiple Python implementations, which include alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as ahead-of-time (AOT) compilation and just-in-time (JIT) libraries that are maintained by the community.
+In the second section, we share some insights on the impact of programmer choices.
+To do this, we will run a series of experiments to examine the energy behavior of Python code in the most of its use cases.Section\ref{python:sec_insights} will then end by investigating several features of Python structure and their impact on energy consumption.
+After that, in Section~\ref{python:sec_interpreters},we investigate other non-intrusive approaches to optimizing the energy consumption of applications by comparing multiple Python implementations, which include alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as ahead-of-time (AOT) compilation and just-in-time (JIT) libraries that are maintained by the community.
 
@@
 To answer these questions, we report on 4 case studies that intend to answer these research questions.
-First, we study the energy behavior of Python in two application contexts: machine learning (cf. \Cref{sec:webdev}) web applications (cf. \Cref{sec:ml}).
-Then, we dive deeper into the energy consumption of Python core structures, before concluding with the impact of the Python interpreter on energy consumption.
+First, we study the energy behavior of Python in two application contexts: machine learning (cf. \Cref{sec:webdev}) and  web applications (cf. \Cref{sec:ml}).
+Then, we dive deeper into the energy consumption of Python core structures, before concluding with the impact of parallelism on energy consumption.
 
@@
 \subsubsection{Results \& Findings}
-% TODO : Missing analysis of the results
+
 As the model's accuracy increases, so does the energy required for the next accuracy increment.
@@
     \caption{Accuracy along epochs.}
-    \label{fig:p2}
+    \label{fig:eopochvsaccuracy}
 \end{figure}
@@
 
+cref{fig:eopochvsaccuracy}shows the accuracy of the model based on the number of epochs.
+This figure shows that there is a non-significant impact on the choice of the strategy on the accuracy; it is only a matter of the number of epochs.
+Hover, the increase in accuracy is not linear, as can be seen.
+It took only 10 epochs to reach an accuracy of 84\%, but it required more than twice the number of epochs to add axtra 6\% accuracy.
+This highlights the price that one should pay to increase the accuracy of the model.
+This has been confirmed by the energy consumption, as shown in \cref{fig:cum_energy_vs_accuracy}.
+As one can see the training the model up to 90\% accuracy requires three times the energy required for training to 80\% accuracy.
+Moreover, one can notice that this price is mostly payed by the calcul units such as CPU and GPU, while the memory is not that impacted.
+
+Interestingly there were no significant difference between different strategies -arround $1.3\%$- As one can seen in Figures ~\ref{fig:av_energy_setup}.
+Howerver this gap may increase as the number of epochs increases.
+
+On the other hand, as the number of epochs increases, the average power consumption decreases until it reaches a steady state after 10 epochs.
+This could be connected to the caching methods utilized by the CPU and GPU.
+This evolution, however, is still insignificant in comparison to the baseline value.
+This is shown in Figure~\ref{fig:ephoch_power}.
+
+As one can see in figures \ref{fig:ephoch_power}~\ref{fig:epoch_duration}, the average power consumption and the average duration of an epoch don't have a significant variation, as most of them were around 3.7 seconds.
+However, there was an intermediate correlation between these two values if we exclude the strategy based on one core plus its hyper thread.
+The reason why the last strategy exhibits more power even when its duration isn't optimal is the context switched between the core and its hyper thread.
+
 \begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{imgs/power_gpu_baedonepoche}
-    \caption{Evolution of average GPU power along epochs.}
-    \label{fig:p2}
+    \includegraphics[width=\linewidth]{imgs/cumulative_energy_fast10.pdf}
+    \caption{Average energy for training the model based on the strategy.}
+    \label{fig:av_energy_setup}
 \end{figure}
@@
     \centering
-    \includegraphics[width=\linewidth]{imgs/cumulative_energy_fast10}
-    \caption{cumulative energy of fast10 benchmark within different configurations }
-    \label{fig:p2}
+    \includegraphics[width=\linewidth]{imgs/power_gpu_baedonepoche.pdf}
+    \caption{Evolution of average GPU power along epochs.}
+    \label{fig:ephoch_power}
 \end{figure}
 
-\subsection{Conclusion}
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{imgs/epoch_duration.pdf}
+    \caption{Distribtion of the duration of each epoch.}
+    \label{fig:epoch_duration}
+\end{figure}
+\subsubsection{conculsion}
 We discovered that when the model's accuracy improves, so does the energy required for a subsequent accuracy gain.
@@
 Is a 10\% increase in accuracy worth it if we have to spend three times the energy?
-Even while utilizing a GPU consumes more total power, the training time is significantly reduced.
-As a result, the utilization of a GPU is required to lower the energy consumption of the training.
-Some experiments revealed unusual behavior that was not studied during this internship.
-These could be examined in future studies and perhaps used to achieve more energy savings.
-The execution time did not depend on the number of cores employed in several studies.
-This parameter had a significant impact on training time in others.
-As a result, the best core configuration will be determined by the script.
-The next steps in this area could include exploring the reasons for these discrepancies and possibly creating a script that can automatically find the ideal core configuration for a specific script.
+While using the GPU, the CPU has no significant impact on the energy consumption of the training.
+Therefore, one can limit the number of CPU cores used during the execution without sacrificing either the execution time nor the accuracy.
 

### 2022-09-27 chapters/optimization/python-dir/python.tex
update : python loops

@@
 \subsection{Python Insights}
-% This part aims to formulate some actionable insights to optimize the energy consumption of Python applications without altering the source code.
-% Therefore, we start by extending the work of Hasan~\emph{et~al.} and Oliveira~\emph{et~al.} to the context of Python environments
-% \cite{hasan_energy_2016,oliveira_recommending}.
 Another field of investigation is the type of data and control structures that might impact the energy consumption.
+In this study, we will run some basic algortims with different data structures and control structures to see if there is any difference in energy consumption.
+\subsubsection{experimental Protocol}
+All of the tests in this study are done on the grid5000 cluster using the same machine, Python interpreter, and library version.
+Due to PowerAPI's frequency constraint, we measure the energy of 1000 algorithm iterations for each test.
+And, for completeness, we perform each test 20 times, deleting the cache between each run.
+The data is then averaged and presented in graphs.
 
-We iterate over a list using 3 methods.
+\subsubsection{Python loops}
+In the first experiment we want to study the impact of the loop type on the energy consumption. to do so, we run the following code snippet:
+\begin{algorithm}[!htb]
+    \begin{algorithmic}[1]
+        \State $sum \gets 0$
+        \For{$i \gets 0$ to $N$}
+        \State $sum \gets sum+1$
+        \EndFor
+        \State \Return {$sum$}
+    \end{algorithmic}
+\end{algorithm}
+
 First, the classical \texttt{for(i in range(len(n)))}.
@@
 the \texttt{for} loop is translated to a \texttt{while} loop ( transformation from D type to B type -- asm -- ), therefore we wanted to compare this with a \texttt{while} version.
-After determining the main ways to iterate over a \texttt{loop}, we run the collection algorithms of different data types to observe if they impact the energy consumption of the code, and we repeated it for the size of the collection.
+Therefore our candidates will be :
 
-As one can see in \Cref{fig:pythonloops}, the type of the data has no impact on the energy consumption.
-However, the way one iterates over the collection has a huge factor.
-Interestingly, the \texttt{for in range} loop was by far the optimal one, followed by the regular \texttt{for in collection}, and the \texttt{while} part was the last one with an overhead of 400\% compared to the first option.
+\begin{itemize}[]
+    \item \texttt{for(i in range(len(n)))}
+    \item \texttt{for(element in collection)}
+    \item \texttt{while}
+\end{itemize}
 
-\begin{figure}
+For each of these implementations, we will run the same code snippet with different primitive data types : Int, Float, and String.
+\begin{figure}[!htb]
     \centering
@@
 
+Figure~\ref{fig:pythonloops} shows the results of the experiment. As one can notice, the type of the data has negligibale impact on the energy consumption.
+However, the way one iterates over the collection has a huge factor.
+Interestingly, the \texttt{for in range} loop was by far the optimal one, followed by the regular \texttt{for in collection}, and the \texttt{while} part was the last one with an overhead of 400\% compared to the first option.
+
 The reason behind such a behavior is mainly related to how the Python interpreter is implemented\footnote{\url{https://www.python.org/doc/essays/list2str}},\footnote{\url{https://www.pythonpool.com/for-vs-while-loop-python/}}.
@@
 The \texttt{while} is even slower due to the implicit increment of the variable, which causes an extra operation during the loop.
+
 To confirm this hypothesis, we tried to construct a new list by editing the elements of the previous one (cf. \Cref{fig:pythontreatement}).

### 2022-09-27 chapters/optimization/python-dir/python.tex
update : python insights

@@
 
-To confirm this hypothesis, we tried to construct a new list by editing the elements of the previous one (cf. \Cref{fig:pythontreatement}).
-And, as predicted, the built-in methods are the most energy-saving ones, while the customize \texttt{while} loop is the heaviest.
-
 \begin{figure}
@@
 
-Another interesting finding is the impact of anonymous functions (also known as lambda expressions) on energy consumption.
-The reason is that Python treats these functions as local variables, unlike the predefined ones, which are global in our case.
-Therefore, they are faster and consume less energy.
+To confirm this hypothesis, we tried to construct a new list by editing the elements of the previous one (cf. \Cref{fig:pythontreatement}).
+We used four different methods to do so: \texttt{comprehension list}, \texttt{while loop}, \texttt{a map with predefined function}, and \texttt{a map with an anonymous function}.
 
-\subsubsection{Synthesis}
+A comprehension list is a Pythonic way to create a new list by applying a function to each element of the previous one.It is based on the mathematical set-builder notation~\cite{editiondiscrete}.
+
+As for the map function, it is a higher-order function that applies a given function to each element of a collection.
+It is also possible to use an anonymous function, which is a function that is not bound to a name but only to its definition, which in the Python case is called a lambda function/expression.
+
+As one can notice in~\Cref{fig:pythontreatement}, the built-in methods are the most energy-saving ones, while the customize \texttt{while} loop is the heaviest.
+Moreover, when increasing the size of the collection, the map with lambda functions tends to consume less energy than the predetermined one.
+The reason behind such behavior is that Python treats these functions as local variables, unlike the predefined ones, which are global in our case. Therefore, they are faster and consume less energy.
+
+\paragraph*{subsubsection}{Synthesis}
 This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
 
-\subsection{Python \& Multiprocessing}
+\subsubsection{Python \& Multiprocessing}
 

### 2022-09-27 chapters/optimization/python-dir/python.tex
update : multiprocessing p1

@@
 As one can notice in~\Cref{fig:pythontreatement}, the built-in methods are the most energy-saving ones, while the customize \texttt{while} loop is the heaviest.
-Moreover, when increasing the size of the collection, the map with lambda functions tends to consume less energy than the predetermined one.
+Moreover, when increasing the size of the collection, the map with lambda functions tends to consume less energy than the predefined one.
 The reason behind such behavior is that Python treats these functions as local variables, unlike the predefined ones, which are global in our case. Therefore, they are faster and consume less energy.
 
-\paragraph*{subsubsection}{Synthesis}
+\paragraph*{Synthesis}
 This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
@@
 \subsubsection{Python \& Multiprocessing}
+The purpose of this part is to study the impact of concurrency on the energy consumption for Python applications.
+To do so, we run a simple code snippet that computes the sum of the first $N$ numbers using the standard concurrency libraries \texttt{multiprocessing} and \texttt{multihtreading}.
+Using Desktop  machine with Intel(R) Core(TM) i7-4800MQ CPU @ 2.70GHz with four cores and four hyperthreads, we run the following strategies.
+%  N ==5e5
+\begin{itemize}
+    \item Sequentiel: and here we just compute the sum of the first $N$ numbers without any concurency
+    \item Multithreading: and here we use the \texttt{ ThreadPoolExecutor} library to compute the sum of the first $N$ numbers, We devide $N$ by the number of threads and each thread will compute the sum of the numbers in its range.
+For our case we used 4 threads.
+    \item multiprocessing : and in this case, we use the \texttt{ProcessPoolExecutor} library to calculate the sum of the first $N$ numbers.
+We divide $N$ by the number of processes, and each process sums the numbers in its range.
+In this situation, we employed 2,4,8,16 processes, correspondingly.
+\end{itemize}
 
-% \subsubsection{Cost of Parallelism in Python}
+Figure~\ref{fig:python_multiprocessing_strategies} shows the results of the experiment. As one can notice
+\begin{figure}[!hbt]
+    \centering
+    \includegraphics[width=\linewidth]{imgs/python_multiprocessing}
+    \caption{Energy behaviour resulting from diffrent python concurrency strategies (number of workers) }
+    \label{fig:python_multiprocessing_strategies}
+\end{figure}
+
 Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
@@
 
+\clearpage
 \input{python_interperters.tex}

### 2022-09-27 chapters/optimization/python-dir/python.tex
remove : extra figure

@@
 The purpose of this part is to study the impact of concurrency on the energy consumption for Python applications.
-To do so, we run a simple code snippet that computes the sum of the first $N$ numbers using the standard concurrency libraries \texttt{multiprocessing} and \texttt{multihtreading}.
+To do so, we run a simple code snippet that computes the sum of the first $N$ numbers using the standard concurrency libraries \texttt{multiprocessing~\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}}} and \texttt{multihtreading~\footnote{\url{https://docs.python.org/3/library/threading.html}}}.
 Using Desktop  machine with Intel(R) Core(TM) i7-4800MQ CPU @ 2.70GHz with four cores and four hyperthreads, we run the following strategies.
@@
 
-Figure~\ref{fig:python_multiprocessing_strategies} shows the results of the experiment. As one can notice
 \begin{figure}[!hbt]
@@
 
-Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications when introducing parallelism.
+Figure~\ref{fig:python_multiprocessing_strategies} shows the results of the experiment. As one can notice
+
+performance
+- sequentiel took less time then mulit threading
+- 4 process was the optimal one
+-- for cpu power the multi threading reached 13 watts whule the 4 process multiprocessing was 42 watts
+-- for global energy  the sequentiel one took the most, then the multithreading after that the muli processing , while keeping the 4 process always the optimial one
+
+Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications.
 Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).

### 2022-09-28 chapters/optimization/python-dir/python.tex
add : python multi processing

@@
     \caption{Energy behaviour resulting from diffrent python concurrency strategies (number of workers) }
-    \label{fig:python_multiprocessing_strategies}
+    \label{fig:python_multiprocessing}
 \end{figure}
 
-Figure~\ref{fig:python_multiprocessing_strategies} shows the results of the experiment. As one can notice
+Figure~\ref{fig:python_multiprocessing} shows the results of the experiment regarding four metrics: power, execution time , DRAM energy, and CPU energy .
 
-performance
-- sequentiel took less time then mulit threading
-- 4 process was the optimal one
--- for cpu power the multi threading reached 13 watts whule the 4 process multiprocessing was 42 watts
--- for global energy  the sequentiel one took the most, then the multithreading after that the muli processing , while keeping the 4 process always the optimial one
+As one can notice, CPU energy is ten folds higher than the DRAMS one.Therefore we will focus on the CPU for this study.
 
-Additionally, \Cref{fig:python_multiprocessing} reports on the energy consumption of Python applications.
-Python applications are single-threaded systems constrained by the \emph{global lock system} (GLS).
-However, due to the increase of cores/threads per CPU, many Python libraries started to take advantage of this hardware feature by allowing concurrent execution of Python processes.
-For example, the multiprocessing library\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}} spawns sub-processes to increase the degree of concurrency of Python applications.
-As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of threads until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.
+In general, the multi-processing strategy is the most energy efficient one as well as the fastest one.
+However, unexpectedly, the multithreaded option took more time to execute the code snippet than the sequential one.
+Therefore, we will divide our analysis into two parts: the first one will focus on multithreading versus the sequential strategies, while the second one will study the behavior of the multiprocessing strategy.
 
-% \begin{figure}[hbt]
-%     \centering
-%     \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-%     \caption{Energy consumption of Python multiprocessing depending on the number of exploited threads.}
-%     \label{fig:python_multiprocessing}
-% \end{figure}
+\paragraph*{Multithreading vs Sequential}
+The unexpected behavior of the multithreading strategy is due to the fact that the Python interpreter is not thread-safe.
+Because of the \emph{global lock system} (GLS), Python can't run multiple threads at the same time. Therefore, the multithreading strategy is slower than the sequential one.
+On the other hand, a side effect of this is the context switching between threads, which will put the CPU in a lower power state.
+The multi-threading strategy depicts an average power of $13 Watts$ compared to the sequential one, which is around $22 Watts$.
+This difference in power consumption has overcome the lack of performance of the multithreading strategy, which leads to better energy efficiency.
 
-% \begin{figure}[hbt]
-%     \centering
-%     \includegraphics[width=\linewidth]{imgs/multiprocessing_energyvstime}
-%     \caption{Energy consumption of Python multiprocessing depending on the number of exploited threads.}
-%     \label{fig:python_multiprocessing}
-% \end{figure}
+\paragraph*{Multiprocessing}
+Unlike the multi-threading strategy, the multi-processing strategy is based on forking the program into multiple processes, which are independent from each other.
+This allows the Python interpreter to run multiple processes at the same time, which explains the better performance of the multi-processing strategy.
+Howerver, this will increase the Average power consumption of the CPU, which is around $40 Watts$ compared to the sequential one, which is around $22 Watts$.
+nevertheless, the multi-processing strategy is still the most energy efficient one.
+
+Despite, the fact that increasing the number of processes will reduce the execution time because we split the tasks on the number of processes, there is a point where the execution time will increase.
+As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of processes until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.
 
@@
 
-\Cref{fig:python_multiprocessing} compares the behavior of Python programs when we try to introduce parallelism.
-
-As we know, Python is a single-threaded program thanks to the GLS (\emph{Global Lock System}).
-However, due to the increase in the number of cores/threads per CPU, many libraries have started to take advantage of this hardware feature by simulating multi-threading as multiple instances of the processor.
-
-As we can see in Figure~\ref{fig:python_multiprocessing}, the energy consumption correlates with the number of threads until reaching the number of physical cores, then losing the benefit from multiprocessing by facing concurency issues imposed by different sub-processes competing for the CPU resources.
-Another finding is observed when hitting the number of physical cores: there is an increase of execution time, but with a reduced energy.
-The reason behind this is the scheduler of the operating system.
-More specifically, it favors hyper-threads instead of physical cores, hence increasing context switches which cause the performance bottleneck.
-Howerver,the other two physical cores are not consuming energy, neither their hyper-threads, which explains the energy savings.
-In the Chapter , %% reference the work of intern from lyon  
-we discuss this behavior of the scheduler more deeply and we assess that it is not due to Python, but is a more generic behavior.
-%% maybe even this graph should go there 
-% Another key finding is when we hit the number of physical cores.
-there was an increase of execution time but still reduced energy, the reason behind this is the scheduler of the operating system.
-Basically it favors the hyper threads than the physical core which will lead to some context switches which cause the slow behavior, however in the other hand the other two physical cores are not consuming energy, neither their hyper threads which explains the gain of the energy consumption in this case.
-% in the Chapter %% reference the work of intern from lyon  
-% we discuss a deeper this behavior of the scheduler and in that case we confirm that it is not related to python but it is more generic behaviour %% maybe even this graph should go there 
-%% todo 
-% add performance 
-% add dram \ a glimps
+\paragraph*{Synthesis}
+This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to use the multiprocessing strategy, which is the most energy efficient one.
+However, this strategy should be used with caution as it can lead to increased execution times.
+This study demonstrates that the optimal number of processes is equal to the number of physical cores, which is four in our case.
+By using the multithreading strategy, it was also found that sometimes the performance of the application has to be given up in order to save energy.
 

### 2022-09-28 chapters/optimization/python-dir/python.tex
add : conculsion

@@
 In the second section, we share some insights on the impact of programmer choices.
-To do this, we will run a series of experiments to examine the energy behavior of Python code in the most of its use cases.Section\ref{python:sec_insights} will then end by investigating several features of Python structure and their impact on energy consumption.
-After that, in Section~\ref{python:sec_interpreters},we investigate other non-intrusive approaches to optimizing the energy consumption of applications by comparing multiple Python implementations, which include alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as ahead-of-time (AOT) compilation and just-in-time (JIT) libraries that are maintained by the community.
+To do this, we will run a series of experiments to examine the energy behavior of Python code in most of its use cases.Section\ref{python:sec_insights} will then end by investigating several features of Python structure and their impact on energy consumption.
+After that, in Section~\ref{python:sec_interpreters}, we investigate other non-intrusive approaches to optimizing the energy consumption of applications by comparing multiple Python implementations, which include alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as ahead-of-time (AOT) compilation and just-in-time (JIT) libraries that are maintained by the community.
 
@@
 \subsection{Python Gluttony}
-% In the other hand,because of the interpreted nature of Python code, this programming language has paid a high price in terms of performance~\ref{fig:clbg}, memory and energy consumption.
+% On the other hand, because of the interpreted nature of Python code, this programming language has paid a high price in terms of performance~\ref{fig:clbg}, memory and energy consumption.
 According to~\cite{pinto_energy_2017} and~\cite{noureddine_preliminary_2012}, Python tends to be one more energy hungry programming language.
@@
 
-Python consumes energy, mainly because it is slow in execution.
+Python consumes energy mainly because it is slow in execution.
 Its flexibility and simplicity caused it to drop off in performance because Python gains its flexibility from being a dynamic language.
@@
 
-cref{fig:eopochvsaccuracy}shows the accuracy of the model based on the number of epochs.
+cref{fig:eopochvsaccuracy}shows the model's accuracy based on the number of epochs.
 This figure shows that there is a non-significant impact on the choice of the strategy on the accuracy; it is only a matter of the number of epochs.
 Hover, the increase in accuracy is not linear, as can be seen.
-It took only 10 epochs to reach an accuracy of 84\%, but it required more than twice the number of epochs to add axtra 6\% accuracy.
+It took only ten epochs to reach an accuracy of 84\%, but it required more than twice the number of epochs to add axtra 6\% accuracy.
 This highlights the price that one should pay to increase the accuracy of the model.
@@
 \end{figure}
-\subsubsection{conculsion}
+\subsubsection{Synthesis}
 We discovered that when the model's accuracy improves, so does the energy required for a subsequent accuracy gain.
@@
 As one can notice, while the rendering method consumed the same amount of energy as in the \textsf{Vanilla} strategy (around 1.3\,kJ), the resolve part dropped by $20\times$.
+\paragraph{Synthesis}
 
@@
 \paragraph*{Synthesis}
-This study demonstrates that the optimal way to reduce the energy consumption of Python application is to follow the guidelines and to privilege the built-in functions.
+This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to follow the guidelines and to privilege the built-in functions.
 
@@
 \paragraph*{Multiprocessing}
-Unlike the multi-threading strategy, the multi-processing strategy is based on forking the program into multiple processes, which are independent from each other.
-This allows the Python interpreter to run multiple processes at the same time, which explains the better performance of the multi-processing strategy.
-Howerver, this will increase the Average power consumption of the CPU, which is around $40 Watts$ compared to the sequential one, which is around $22 Watts$.
-nevertheless, the multi-processing strategy is still the most energy efficient one.
+Unlike the multi-threading strategy, the multi-processing strategy is based on forking the program into multiple processes, which are independent of each other.
+This allows the Python interpreter to run multiple processes simultaneously, which explains the better performance of the multi-processing strategy.
+However, this will increase the Average power consumption of the CPU, which is around $40 Watts$, compared to the sequential one, which is around $22 Watts$.
+Nevertheless, the multi-processing strategy is still the most energy-efficient one.
 
-Despite, the fact that increasing the number of processes will reduce the execution time because we split the tasks on the number of processes, there is a point where the execution time will increase.
-As one can see in \Cref{fig:python_multiprocessing}, the energy consumption is correlated with the number of processes until reaching the limit of physical cores, when concurrent processes start to compete for the CPU.
+Although increasing the number of processes will reduce the execution time because we split the tasks on the number of processes, there is a point where the execution time will increase.
+As one can see in \Cref{fig:python_multiprocessing}, the energy consumption correlates with the number of processes until the limit of physical cores, when concurrent processes compete for the CPU.
 
@@
 \paragraph*{Synthesis}
-This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to use the multiprocessing strategy, which is the most energy efficient one.
-However, this strategy should be used with caution as it can lead to increased execution times.
+This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to use the multiprocessing strategy, which is the most energy-efficient one.
+However, this strategy should be used cautiously as it can lead to increased execution times.
 This study demonstrates that the optimal number of processes is equal to the number of physical cores, which is four in our case.
@@
 \section{Summary}
-Because Python is widely used by software services deployed in public and private cloud infrastructures, increasing the energy efficiency of Python-based applications will reduce ICT carbon emissions.
-This chapter discusses various ways for increasing the energy efficiency of Python-based applications.
-It first explains why such a choice was made, and then it develops a method to enhance the energy efficiency of the most commonly employed use cases, which are revealed to include web development and machine learning.
-In web development, we found that the majority of the energy is wasted while processing requests, therefore adopting the proper strategy can reduce energy consumption by tenfold.
-Following that, we looked at the energy usage for the machine learning usecase and discovered that it is not proportional to model accuracy.
-As a result, a reduction in accuracy can result in large energy savings.
-Following that, we dug deeper to investigate how data structures, parallelism, and iterative methods affect energy use.
-In the second section, we presented a simple technique for optimizing the energy usage of Python-based apps. without making substantial changes to the code This is the use of many Python interpreters
-Section 2 started by categorizing and filtering these implementations into three major kinds (compiler, interpreter, and extra libraries).
-Then we conducted a series of experiments to compare the energy consumption of various interpreters.
+
+Since many software services use Python in public and private cloud infrastructures, making Python-based apps more energy efficient will lower ICT carbon emissions.
+This chapter discusses various ways of optimizing the energy consumption of Python-based applications.
+It first explains why such a choice is relevant.
+Then it studies the energy behavior of Python within its most commonly employed use cases, which are revealed to be web development and machine learning.
+
+%  TODO : add description about the technology used in the training model 
+First, we studied the energy consumption of a machine learning algorithm using the Cifar-10 benchmark to train a model.
+We discovered that this energy consumption increases exponentially while the model accuracy increases.
+As a result, a reduction in accuracy can result in considerable energy savings.
+We investigated how data structures, parallelism, and iterative methods affect energy use.
+
+As for web development, we looked at a website built with Django, which is one of the most popular web frameworks in the Python community.
+We found that fetching the data from the database consumes most of the energy during the request processing phase.
+Therefore, Django-based web servers can use up to ten times less energy if the developer chooses the write strategy to fetch the data.
+
+After that, we analyzed the impact of several iteration mechanisms in python and their impact on energy usage.
+We found that the built-in functions and the best practices for writing Python code are the most energy-efficient ones.
+This energy efficiency is because most of these built-in functions are written in a lower programming language, C.
+
+Finally, we studied the impact of concurrency on energy consumption for Python applications.
+We found that the multi-processing strategy is the most energy-efficient one.
+However, this strategy should be used cautiously.
+This study demonstrates that the optimal number of processes is equal to the number of physical cores, and exceeding this number will cause the application loses its efficiency in both performance and energy consumption.
+As for the multithreading strategy, We found that the Python interpreter is not thread-safe, which leads to a slower performance than the sequential one.
+However, this also leads to lower power consumption, which sometimes overcomes the lack of performance to make the application more energy-efficient.
+
+In the second section, we presented a simple technique for optimizing the energy usage of Python-based apps. Without making substantial changes to the code, This is the use of many Python interpreters
+Section 2 started by categorizing and filtering these implementations into three significant kinds (compiler, interpreter, and extra libraries).
+Then we conducted a series of experiments to compare the energy consumption of these alternatives.
 The findings indicate the lack of a general solution and the importance of selecting the appropriate interpreter for the appropriate use case.
-In certain circumstances, some of these answers may be the best, while in others, they may be the worst.
+In certain circumstances, some of these answers may be the best, while others may be the worst.
+We found that most interpreters had a similar or worse energy consumption to the official implementation of CPython of their Pythons' version.
+The reason behind such behavior was that some implementations focused on specific usage cases such as machine learning or security.
+In contrast, others focused on the compatibility of the python code with their platforms like Jython and IronPython.
 Finally, regardless of implementation, we demonstrated that using JIT is the most efficient technique to boost the energy efficiency of Python-based programs.
-We believe that our contributions will benefit a wide spectrum of legacy systems, not only reducing ICT's carbon footprint, but also lowering their cloud bills for the resources these services need.
+
+We believe our contributions will benefit a broad spectrum of legacy systems, reducing ICT's carbon footprint and lowering cloud bills for these services' resources.

### 2022-09-28 chapters/optimization/python-dir/python.tex
update : python  chapter

@@
 As one can notice in \Cref{fig:hannoi}, Python consumes $30$ times more than C or C++.
-The benchmark was done with an implementation of the \fnurl{Tower of Hanoi}{https://en.wikipedia.org/wiki/Tower_of_Hanoi} of 30 disks.
+The benchmark was implemented with the \fnurl{Tower of Hanoi}{https://en.wikipedia.org/wiki/Tower_of_Hanoi} of 30 disks.
 
@@
 \Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers.\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}
-57\% of the respondents reported that they use Python for data science, 51\% said they are using it for web development, and around 40\% are using it for system administration~\footnote{The options in this survey were not mutually exclusive.As a result, the total of the percentages is greater than $100\%$.}.
+57\% of the respondents reported that they use Python for data science, and 51\% said they use it for web development.
+Around 40\% are using it for system administration~\footnote{The options in this survey were not mutually exclusive.
+As a result, the total of the percentages is greater than $100\%$.}.
 
@@
 \section{Energy optimization of python on the source code level}\label{python:sec_insights}
-In this section , we investigate the energy footprint of Python in its most popular domains of adoption.
+In this section, we investigate the energy footprint of Python in its most popular domains of adoption.
 We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016} did in~\cite{hasan_energy_2016}.
@@
 First, we study the energy behavior of Python in two application contexts: machine learning (cf. \Cref{sec:webdev}) and web applications (cf. \Cref{sec:ml}).
-Then, we dive deeper into the energy consumption of Python core structures, before concluding with the impact of parallelism on energy consumption.
+Then, we dive deeper into the energy consumption of Python core structures before concluding with the impact of parallelism on energy consumption.
 
@@
 \subparagraph{Software settings}
-For the sake of reproducibility, each experiment is done within a Docker container using Jupyter lab.
-These tests are run on top of a minimal version of Debian-10 to increase the accuracy of the tests by eliminating any unnecessary processes.
+Each experiment is done within a Docker container using Jupyter lab to ensure reproducibility.
+These tests are run on top of a minimal version of Debian-10 to increase the tests' accuracy by eliminating unnecessary processes.
 \subsubsection{Input Workload}
 \subparagraph{Models}
-Several models were developed, however, only two were used in the final trials because they achieved 94 percent accuracy in an acceptable length of time.
+Several models were developed. However, only two were used in the final trials because they achieved 94 percent accuracy in an acceptable length of time.
 David Page's cifar10-fast~\footnote{\url{https://github.com/davidcpage/cifar10-fast}} and Woonhyuk Baek's torch skeleton~\footnote{\url{https://github.com/wbaek/torchskeleton}} are shown here.
@@
 \begin{itemize}
-    \item accuracy : in \%
-    \item execution time : in seconds for both the duration of each epoch and the total duration to a chieve a certain accuracy
-    \item total energy consumption : for the CPU and the  GPU
+    \item accuracy: in \%
+    \item execution time: in seconds for both the duration of each epoch and the total duration to achieve a certain accuracy
+    \item total energy consumption: for the CPU and the  GPU
 \end{itemize}
@@
 
-As the model's accuracy increases, so does the energy required for the next accuracy increment.
+As the model's accuracy increases, so do the energy required for the next accuracy increment.
 
@@
 It took only ten epochs to reach an accuracy of 84\%, but it required more than twice the number of epochs to add axtra 6\% accuracy.
-This highlights the price that one should pay to increase the accuracy of the model.
-This has been confirmed by the energy consumption, as shown in \cref{fig:cum_energy_vs_accuracy}.
-As one can see the training the model up to 90\% accuracy requires three times the energy required for training to 80\% accuracy.
-Moreover, one can notice that this price is mostly payed by the calcul units such as CPU and GPU, while the memory is not that impacted.
+This highlights the price one should pay to increase the model's accuracy.\cref{fig:cum_energy_vs_accuracy} confirm a such claim.
+As one can see, the training of the model up to 90\% accuracy requires three times the energy required for training to 80\% accuracy.
+Moreover, one can notice that this price is paid mainly by the CPU and GPU, while the memory is not that impacted.
 
-Interestingly there were no significant difference between different strategies -arround $1.3\%$- As one can seen in Figures ~\ref{fig:av_energy_setup}.
-Howerver this gap may increase as the number of epochs increases.
+Interestingly there were no significant differences between different strategies -around $1.3\%$- As shown in Figures ~\ref{fig:av_energy_setup}.
+However, this gap may increase as the number of epochs increases.
 
-On the other hand, as the number of epochs increases, the average power consumption decreases until it reaches a steady state after 10 epochs.
+On the other hand, as the number of epochs increases, the average power consumption decreases until it reaches a steady state after ten epochs.
 This could be connected to the caching methods utilized by the CPU and GPU.
-This evolution, however, is still insignificant in comparison to the baseline value.
+However, this evolution is still insignificant compared to the baseline value.
 This is shown in Figure~\ref{fig:ephoch_power}.
@@
 As one can see in figures \ref{fig:ephoch_power}~\ref{fig:epoch_duration}, the average power consumption and the average duration of an epoch don't have a significant variation, as most of them were around 3.7 seconds.
-However, there was an intermediate correlation between these two values if we exclude the strategy based on one core plus its hyper thread.
+However, there was an intermediate correlation between these two values if we excluded the strategy based on one core plus its hyper thread.
 The reason why the last strategy exhibits more power even when its duration isn't optimal is the context switched between the core and its hyper thread.
@@
 \subsubsection{Life-cycle of a Request in Django}
-Django is a MVT (\emph{Model-View-Template}) framework, which means that it follows the MVC (\emph{Model-View-Controller}) pattern.
+Django is an MVT (\emph{Model-View-Template}) framework, which means that it follows the MVC (\emph{Model-View-Controller}) pattern.
 Figure~\ref{fig:django-life-cycle} describes the life-cycle of a request in this framework.
@@
 These middleware layers are in charge of authentication, security, and so on.
-Once the request is processed by these layers, it is passed to the URL router, which extracts the URL from the request and tries to match it to the defined URLs.
+Once these layers process the request, it is passed to the URL router, which extracts the URL from the request and tries to match it to the defined URLs.
 After getting the matching URL, the corresponding view function is called.
 This function is responsible for treating the request, gathering the data, and then generating the response that will be put inside a template to be returned as a response.
-As Django adopts a MVT model, it also offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.
+As Django adopts an MVT model, it also offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.
 
@@
 
-First, we start by investigating the energy consumption of the request-response life cycle in Django, to determine which layer consumes the most energy.
+First, we investigate the energy consumption of the request-response life cycle in Django to determine which layer consumes the most energy.
 To do so, we created a sample Django application that returns the response to the request.
 We tracked its energy consumption using JouleHunter~\footnote{\url{https://github.com/powerapi-ng/joulehunter}}.
-JouleHunter is an open-source library that I developed to help practitioners identify energy hotspots in their applications using statistical profilers.
-In the case of Django, it can be added as an extension with no additional setup or change to the source code.
+JouleHunter is an open-source library that we developed to help practitioners identify energy hotspots in their applications using statistical profilers.
+In the case of Django, Joulehunter will be added as a middleware with no additional setup or change to the source code.
 The energy consumption of the request-response life-cycle in Django is shown in Figure~\ref{fig:django_life_cycle_naive}.
-As we can notice, 91.4\% of the total energy consumption is spent on resolving the request by retrieving the data, while only 5.27\% is spent on rendering the response.
+As we can notice, 91.4\% of the total energy consumption is spent on resolving the request by retrieving the data, while only 5.27\% is consumed on rendering the response.
 
@@
     \includegraphics[width=\linewidth]{imgs/django_life_cycle_naive}
-    \caption{Tree representation of the energy consumption of a single request in django (naive version)}
+    \caption{Tree representation of the energy consumption of a single request in Django (naive version)}
     \label{fig:django_life_cycle_naive}
@@
 % We have chosen Django as our web framework.
-Therefore, we chose to study if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
+Therefore, we chose to study if the choice of the database and the ORM impacts the energy consumption and the website's performance.
 % \subsection{Experiment \& Results}
 To accomplish so, we examined the cost of a single request of the prior website using various ways to extract data from the database.
-We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records:
+We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records and three different ways to fetch the records:
 \begin{enumerate}
@@
 
-As one can observe in \Cref{fig:django}, the strategy to query stored data has a huge impact on energy consumption.
+As one can observe in \Cref{fig:django}, the strategy to query stored data greatly impacts energy consumption.
 As the \textsf{Vanilla} strategy can consume up to $10\times$ more energy than the \textsf{Optimized} one.
@@
 Another interesting observation is the impact of the interpreter, as \Cref{fig:django} highlights.
-For example, using the \textsf{PyPy} interpreter reduces energy consumption, even when adopting the \textsf{Vanilla} strategy.
-We will further discuss the choice of the interpreter in the next section.
+For example, using the \textsf{PyPy} interpreter reduces energy consumption, even when adopting the \textsf{Vanilla} strategy. We will further discuss the choice of an interpreter in the next section.
 
@@
 
-We can therefore conclude that the database and ORM selection have a major impact on the website's performance and energy consumption.
-And, unlike the rendering portion, there are many options available for the database, the ORM, and the data handling strategy.
+We can conclude that the database and ORM selection significantly impact the website's performance and energy consumption.
+Moreover, unlike the rendering portion, many options are available for the database, the ORM, and the data handling strategy.
 
@@
 \subsection{Python Insights}
-Another field of investigation is the type of data and control structures that might impact the energy consumption.
+Another field of investigation is the type of data and control structures that might impact energy consumption.
 In this study, we will run some basic algortims with different data structures and control structures to see if there is any difference in energy consumption.
@@
 Due to PowerAPI's frequency constraint, we measure the energy of 1000 algorithm iterations for each test.
-And, for completeness, we perform each test 20 times, deleting the cache between each run.
+Furthermore, we perform each test 20 times for completeness, deleting the cache between each run.
 The data is then averaged and presented in graphs.
@@
 \subsubsection{Python loops}
-In the first experiment we want to study the impact of the loop type on the energy consumption. to do so, we run the following code snippet:
+In the first experiment, we want to study the impact of the loop type on energy consumption. To do so, we run the following code snippet:
 \begin{algorithm}[!htb]
@@
 First, the classical \texttt{for(i in range(len(n)))}.
-However, as we can see here, unlike other programming languages, it requires extra operations, such as determining the length of the collection and then using the iterator range.
+However, as we can see here, unlike other programming languages, it requires extra operations, such as determining the collection length and then using the iterator range.
 So, we tried the more adapted version \texttt{for(element in collection)}.
@@
 
-For each of these implementations, we will run the same code snippet with different primitive data types : Int, Float, and String.
+We will run the same code snippet for each of these implementations with different primitive data types: Int, Float, and String.
 \begin{figure}[!htb]
@@
 
-Figure~\ref{fig:pythonloops} shows the results of the experiment. As one can notice, the type of the data has negligibale impact on the energy consumption.
+Figure~\ref{fig:pythonloops} shows the results of the experiment. As one can notice, the data type has a negligible impact on energy consumption.
 However, the way one iterates over the collection has a huge factor.
@@
 
-The reason behind such a behavior is mainly related to how the Python interpreter is implemented\footnote{\url{https://www.python.org/doc/essays/list2str}},\footnote{\url{https://www.pythonpool.com/for-vs-while-loop-python/}}.
-To reduce the latency of Python applications, most of the built-in functions and operations are written in C, and the same goes for the function \texttt{range}.\footnote{\url{ http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1}}
+The reason behind such behavior is mainly related to how the Python interpreter is implemented\footnote{\url{https://www.python.org/doc/essays/list2str}},\footnote{\url{https://www.pythonpool.com/for-vs-while-loop-python/}}.
+Most of the built-in functions and operations are written in C to reduce the latency of Python applications, and the same goes for the function \texttt{range}.\footnote{\url{ http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1}}
 Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$ as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object~\footnote{\url{https://wiki.python.org/moin/TimeComplexity}}.
@@
 
-A comprehension list is a Pythonic way to create a new list by applying a function to each element of the previous one.It is based on the mathematical set-builder notation~\cite{editiondiscrete}.
+A comprehension list is a Pythonic way to create a new list by applying a function to each element of the previous one. It is based on the mathematical set-builder notation~\cite{editiondiscrete}.
 
-As for the map function, it is a higher-order function that applies a given function to each element of a collection.
+The map function is a higher-order function that applies a given function to each collection element.
 It is also possible to use an anonymous function, which is a function that is not bound to a name but only to its definition, which in the Python case is called a lambda function/expression.
@@
 As one can notice in~\Cref{fig:pythontreatement}, the built-in methods are the most energy-saving ones, while the customize \texttt{while} loop is the heaviest.
-Moreover, when increasing the size of the collection, the map with lambda functions tends to consume less energy than the predefined one.
+Moreover, when increasing the collection size, the map with lambda functions tends to consume less energy than the predefined one.
 The reason behind such behavior is that Python treats these functions as local variables, unlike the predefined ones, which are global in our case. Therefore, they are faster and consume less energy.
@@
 \paragraph*{Synthesis}
-This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to follow the guidelines and to privilege the built-in functions.
+This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to follow the guidelines and privilege the built-in functions.
 
 \subsubsection{Python \& Multiprocessing}
-The purpose of this part is to study the impact of concurrency on the energy consumption for Python applications.
+The purpose of this part is to study the impact of concurrency on energy consumption for Python applications.
 To do so, we run a simple code snippet that computes the sum of the first $N$ numbers using the standard concurrency libraries \texttt{multiprocessing~\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}}} and \texttt{multihtreading~\footnote{\url{https://docs.python.org/3/library/threading.html}}}.
-Using Desktop  machine with Intel(R) Core(TM) i7-4800MQ CPU @ 2.70GHz with four cores and four hyperthreads, we run the following strategies.
+We run the following strategies using a Desktop machine with Intel(R) Core(TM) i7-4800MQ CPU @ 2.70GHz with four cores and four hyperthreads.
 %  N ==5e5
@@
     \item Sequentiel: and here we just compute the sum of the first $N$ numbers without any concurency
-    \item Multithreading: and here we use the \texttt{ ThreadPoolExecutor} library to compute the sum of the first $N$ numbers, We devide $N$ by the number of threads and each thread will compute the sum of the numbers in its range.
-For our case we used 4 threads.
-    \item multiprocessing : and in this case, we use the \texttt{ProcessPoolExecutor} library to calculate the sum of the first $N$ numbers.
+    \item Multithreading: Here, we use the \texttt{ ThreadPoolExecutor} library to compute the sum of the first $N$ numbers, We devide $N$ by the number of threads, and each thread will compute the sum of the numbers in its range.
+For our case, we used four threads.
+    \item multiprocessing: and in this case, we use the \texttt{ProcessPoolExecutor} library to calculate the sum of the first $N$ numbers.
 We divide $N$ by the number of processes, and each process sums the numbers in its range.
-In this situation, we employed 2,4,8,16 processes, correspondingly.
+In this situation, we employed 2,4,8,16 processes correspondingly.
 \end{itemize}
@@
 
-%  TODO : add description about the technology used in the training model 
-First, we studied the energy consumption of a machine learning algorithm using the Cifar-10 benchmark to train a model.
+First, we studied the energy consumption of a machine learning algorithm using a benchmark of 60000 entries to train the cifar10-fast model.
 We discovered that this energy consumption increases exponentially while the model accuracy increases.

### 2022-09-28 chapters/optimization/python-dir/python_interperters.tex
refactor : python experiments order
update : python  chapter

@@
 \section{Python Interpreters}\label{python:sec_interpreters}
-For the first studies, we used the official version of Python, because the goal was mainly to highlight the impact of the structure of code on energy consumption.
+We used the official version of Python for the first studies because the goal was mainly to highlight the code structure's impact on energy consumption.
 One main drawback of the previous method is the work to be done to update the existing code base to reduce energy consumption.
 Despite the availability of numerous tools for code refactoring and automatic code optimization, the optimization process remains a manual effort.
-This is due to the fact that code optimization is a task that requires extensive understanding of both the codebase and the language itself.
+This is because code optimization is a task that requires extensive understanding of both the codebase and the language itself.
 
 To avoid such hustle, we tried to find a non-intrusive approach to make the Python code more eco-friendly without altering its structure.
-Python is an interpreted language, which led many initiatives to implement their own interpreter to improve one or many aspects of the Python code. 
-Therefore, nn this section, we will examine at how different Python interpreters affect the energy consumption of software,and in which case, one should use a non-conventional interpreter to save the energy consumption of their application.
+Python is an interpreted language, which led to many initiatives to implement their interpreter to improve one or many aspects of the Python code. 
+Therefore, in this section, we will examine at how different Python interpreters affect the energy consumption of software, and in which ca
+se, one should use a non-conventional interpreter to save the energy consumption of their application.
 
 \subsection{Python runtimes}\label{python:sec_interpreters}
-To do so, we gathered a list of interpreters, transpilers and other optimization libraries that can contribute to reduce the energy consumption of legacy Python applications:
+To do so, we gathered a list of interpreters, transpilers, and other optimization libraries that can contribute to reducing the energy consumption of legacy Python applications:
 \begin{enumerate}
@@
             It compiles the most used portions of the Python code into a binary code for better performance.
-            To benefit from these optimizations, the program has to be executed for at least for few seconds so the JIT has enough time to warm up, the JIT optimization are only applied to the code written by the developer and not to external libraries;
+            To benefit from these optimizations, the program has to be executed for at least for few seconds, so the JIT has enough time to warm up; the JIT optimization is only applied to the code written by the developer and not to external libraries;
       \item \textbf{Cython}:\footnote{\url{https://github.com/cython/cython}}
             A static compiler for Python.
-            It translates the Python code into C, and then compiles it using a C compiler.
-            It also supports an extended version of the Python language that allows programmers to call \emph{C functions}, declare \emph{C types} and use static types, which will help the translation of Python objects into native types, such as integers, float.
-            This often means better performances, since native C libraries are almost all the time faster than the Python written once~\cite{pereira_energy_2017};
+            It translates the Python code into C and then compiles it using a C compiler.
+            It also supports an extended version of the Python language that allows programmers to call \emph{C functions}, declare \emph{C types}, and use static types, which will help translate Python objects into native types, such as integers, and float.
+            This often means better performances since native C libraries are almost all the time faster than the Python written once~\cite{pereira_energy_2017};
       \item \textbf{Intel\,Python}:\footnote{\url{https://software.intel.com/en-us/distribution-for-python}}
-            A customized interpreter developed by Intel to enhance performances of Python programs.
+            A customized interpreter developed by Intel to enhance the performance of Python programs.
             It is dedicated to data sciences and high-performance computing.
@@
       \item \textbf{IronPython}:\footnote{\url{https://ironpython.net}}
-            A .Net-based Python interpretation platform written in C\# that is used with the .Net virtual machine or Mono.
+            A .Net-based Python interpretation platform is written in C\# and used with the .Net virtual machine or Mono.
             It benefits from all the optimizations of .Net virtual machines, such as the JIT and garbage collector mechanisms;
@@
             A Python interpreter that is based on GraalVM\footnote{\url{https://www.graalvm.org/docs/why-graal/}} (a universal virtual machine developed by oracle for running applications written in different programming languages).
-            For the time being, it only supports Python~3 and it is still in the experimental stage;
+            For the time being, it only supports Python~3, and it is still in the experimental stage;
       \item \textbf{Jython}:\footnote{\url{https://jython.github.io}}
             An implementation of Python programming language written in Java for the \emph{Java Virtual Machine} (JVM).
-            Similar to IronPython and GraalPython, it leverages the optimization mechanisms provided by the JVM to enhance the Python performances;
+            Similar to IronPython and GraalPython, it leverages the optimization mechanisms provided by the JVM to enhance Python performances;
       \item \textbf{MicroPython}:\footnote{\url{http://micropython.org}}
@@
       \item \textbf{Nuitka}:\footnote{\url{http://nuitka.net/pages/overview.html}}
-            A Python compiler written in Python that generates a binary executable from Python code.
+            A Python compiler is written in Python that generates a binary executable from Python code.
             It translates the Python code into a C program that is then compiled into a binary executable;
       \item \textbf{Numba}:\footnote{\url{https://numba.pydata.org}}
-            A library that includes JIT compiler to enhance the performances of Python functions using the industry-standard LLVM compiler library;
+            A library that includes a JIT compiler to enhance the performances of Python functions using the industry-standard LLVM compiler library;
       \item \textbf{Shedskin}:\footnote{\url{https://github.com/shedskin/shedskin}}
@@
       \item \textbf{Hope}~\cite{akeret_hope_2015}:
-            A Python library that aims to introduce JIT compiler into the Python code;
+            A Python library that aims to introduce a JIT compiler into the Python code;
       \item \textbf{Parakeet}~\cite{rubinsteyn2012parakeet}:
@@
             It also offers an interpreter, called \emph{grumprun}, which can directly execute the Python code.
-            Unfortunately, we cannot use it because the project is already outdated (last commit is in 2017) and it has a lot of limitations in terms of supporting the Python language, such as some built-in functions and standard libraries;
+            Unfortunately, we cannot use it because the project is already outdated (the last commit was in 2017), and it has many limitations in terms of supporting the Python language, such as some built-in functions and standard libraries;
       \item \textbf{Psyco}:\footnote{\url{http://psyco.sourceforge.net}}
@@
 \begin{description}
-      \item[\bf Type] refers to the category of runtime infrastructure that supports the execution of a Python application.
+      \item[\bf Type] refers to the runtime infrastructure category that supports a Python application's execution.
             In particular, we consider 3 types of environments: \emph{Interpreter}, \emph{Compiler} and \emph{Library};
@@
 
-There are other implementations that we did not consider because either the project aborted many years ago or it has very limited support for Python features.
+We did not consider other implementations because either the project aborted many years ago or it has insufficient Python features.
 After the inventory of those implementations, we filtered them.
 To keep only the versions that are still maintained and support most Python features.
-We then classified them into 3 categories depending on their integration with the Python code.
-In \Cref{fig:pythonimplementations}, we describe the implementations that we kept, the version of each implementation and its category.
+We then classified them into three categories depending on their integration with the Python code.
+In \Cref{fig:pythonimplementations}, we describe the implementations we kept, each implementation's version, and its category.
 
@@
 As discussed in the previous chapter, our idea is to design a benchmarking solution that allows practitioners to reproduce and extends our benchmarks.
-This benchmarking solution is also the one we use to answer the research questions addressed in this manuscript.
-%%%% probably ill resue the same things from other chapters
+This benchmarking solution is also used to answer the research questions addressed in this manuscript.
 
@@
 \subsubsection{Key Performance Metrics}
-Our focus will be mainly on CPU energy consumption as it is ten folds more than the DRAM one, since it is a task-based benchmarking, time is highly correlated within the energy, and it will be only useful to explain some specific energy behaviors, thus we do not put a lot of focus on this metric.
+Our focus will be mainly on CPU energy consumption as it is ten folds more than the DRAM one; since it is a task-based benchmarking, time correlates highly with the energy, and it will be only useful to explain some specific energy behaviors.
+Thus we do not put much focus on this metric.
 %% basically i what i want to say is that we we care mostly about energy, because there are studies that were made on the performance, plus the DRAM is not that interested even within the heavy memory  benchmakrs
@@
 \paragraph{Energy measurement.}
-As we know, the energy of a program is integral of its power over time.
+As we know, a program's energy is integral to its power over time.
 For our case, we used Intel \emph{Running Power Average Limit} (RAPL)~\cite{Khan:2018:RAE:3199681.3177754} to collect the power samples of the running tests.
@@
 \paragraph{Input workload.}
-To benchmark our implementations, we employed the \textsc{Tommti} microbenchmarks suite.\footnote{\url{http://www.tommti-systems.de/main-Dateien/reviews/languages/benchmarks.html}}
-\textsc{Tommti} is a set of 13 microbenchmarks that examine common language features, such as arithmetic operations, data structures and input/output manipulations, among others.
-In addition to these microbenchmarks, we implemented some binary operations to investigate the behavior of the previous implementations when it comes to low-level operations that only works with registries.
+We employed the \textsc{Tommti} microbenchmarks suite to benchmark our implementations.\footnote{\url{http://www.tommti-systems.de/main-Dateien/reviews/languages/benchmarks.html}}
+\textsc{Tommti} is a set of 13 microbenchmarks that examine common language features, such as arithmetic operations, data structures, and input/output manipulations, among others.
+In addition to these microbenchmarks, we implemented some binary operations to investigate the behavior of the previous implementations when it comes to low-level operations that only work with registries.
 
-To study the energy behavior of the Python implementations, we have to focus on the effect of the implementations and mitigate any side effect, such as the organization of the code or any extra consumption due to the operating system or third-party libraries.
+To study the energy behavior of the Python implementations, we have to focus on the effect of the implementations and mitigate any side effects, such as the organization of the code or any extra consumption due to the operating system or third-party libraries.
 Therefore, for each benchmark, we took the implementation written in Python2 as a reference and tried to use it in other implementations as it is.
-If it is not supported by Python3, we converted the code using the official library \emph{2to3}.\footnote{\url{https://docs.python.org/3.7/library/2to3.html}}
-In the case of the libraries using \emph{JIT} adding a decorator to the function that we want to optimize was enough, if there are other changes, we assume that they alter the original code which is against our purpose.
+If Python3 does not support it, we converted the code using the official library \emph{2to3}.\footnote{\url{https://docs.python.org/3.7/library/2to3.html}}
+In the case of the libraries using \emph{JIT}, adding a decorator to the function we want to optimize was enough; if there are other changes, we assume that they alter the original code, which is against our purpose.
 Each benchmark is isolated in a Docker container for several reasons:
@@
       \item Deployment: to use the benchmarking machine without extra configurations that may alter the behavior of the operating system toward energy consumption,
-      \item Reproducibility: one of the most frequent benchmark crimes~\cite{DBLP:journals/corr/abs-1801-02381} in research is the lack of reproducibility---by using Docker we ensure that each benchmark has an image that will be publicly accessible.
+      \item Reproducibility: one of the most frequent benchmark crimes~\cite{DBLP:journals/corr/abs-1801-02381} in research is the lack of reproducibility---by using Docker, we ensure that each benchmark has an image that will be publicly accessible.
 \end{itemize}
@@
 In the repository listing the Python implementations under study,\footnote{\url{https://github.com/chakib-belgaid/python-implementations}} we propose a dedicated tool to generate new workloads and new candidates.
-The script \texttt{generator.py} allows practitioners to create new benchmarks by implementing a Python code within different interpreters.
+The script \texttt{generator.py} allows practitioners to create new benchmarks by implementing Python code within different interpreters.
 Then, it generates \texttt{launcher-benchmark.sh} that can be executed to run the associated benchmark.
-Furthermore, all the successful implementations are stored in a separate directory and the ones that failed (\emph{e.g.}, mostly because of compatibility issues) are stored in a recap file called \texttt{benchmarkTest.md}, where \texttt{benchmark} is the name of the new workload.
-To add new candidate runtimes, one should add a base Docker file that contains the new implementation and, if there should be extra manipulation that should be added to the workload files such as adding new decorator or changing some parameters, then they should be added as an extra function in the script \texttt{generator.py}.
+Furthermore, all the successful implementations are stored in a separate directory, and the ones that failed (\emph{e.g.}, mostly because of compatibility issues) are put in a recap file called \texttt{benchmarkTest.md}, where \texttt{benchmark} is the name of the new workload.
+To add new candidate runtimes, one should add a base Docker file that contains the new implementation.
+Suppose extra manipulation should be performed on the workload files, such as adding a new decorator or changing some parameters.
+In that case, they should be added as an extra function in the script \texttt{generator.py}.
 Finally, they should be included in the Python candidates.
@@
 First and foremost, we want to know whether our samples have a normal distribution. As a result, we ran a Shapiro-Wilk~\cite{shapiro1968comparative} test on our data.
-Despite the fact that some implementations had a $p$-value higher than $a=0.05$ (such as CPython and ActivePython), other implementations like PyPy and Numba presented a $p$-values smaller than 0.01, which leads to reject the hypothesis $H0$ that all the distributions follow a normal distribution.
+Even though some implementations had a $p$-value higher than $a=0.05$ (such as CPython and ActivePython), other implementations like PyPy and Numba presented a $p$-values smaller than 0.01, which leads to rejecting the hypothesis $H0$ that all the distributions follow a normal distribution.
 Therefore, to compare the different implementations, we performed the non-parametric Mann-Whitney U test~\cite{zimmerman1987comparative}, with CPython2 and CPython3 as base references.
@@
 HCA is a method that groups the different implementations based on their energy consumption.
-This technique is a bottom-up approach as it starts with each implementation being a single cluster, and then merges the two most similar clusters until all the clusters are merged into a single cluster.
+This technique is a bottom-up approach as it starts with each implementation being a single cluster and then merges the two most similar clusters until all the clusters are merged into a single cluster.
 The similarity between two clusters is measured by the maximum distance between their furthest points.
 The distance between two points is calculated as the Euclidean distance between the two points.
-Figure~\ref{fig:dendogram} displays the Dendogram produced by the HCA.
+Figure~\ref{fig:dendogram} displays the Dendrogram produced by the HCA.
 
-\begin{figure}
+\begin{figure}[!htb]
       \centering
@@
 
-Except GraalPython, one can notice 3 main clusters.
+Except for GraalPython, one can notice three main clusters.
 The first cluster contains the reference implementations (CPython2 and CPython3) and the implementations that are based on interpreters (IronPython, IntelPython, ActivePython, Jython, and Nuitka).
@@
 The second cluster groups the interpreter implementations that are based on a JIT compiler (PyPy2 and PyPy3).
-Closer to them we find Shedskin, which is a translator that converts the Python code into C++ code to be later compiled into binary code.
+Closer to them, we find Shedskin, which is a translator that converts the Python code into C++ code to be later compiled into binary code.
 Table~\ref{table:manwithyu} shows that this cluster is by far the most energy-efficient one.
 While Numba2 and Numba3 are also JIT libraries, they differ from PyPy due to their manual optimization.
-Unlike PyPy, which will decide by itself which part of Python to optimize.
-In Numba, the developer must target the portion of the code that will be passed to JIT, while the rest of the code will be executed by the regular interpreter (aka CPython).
+Unlike PyPy, which decides the code parts to optimize, Numba gives this responsibility to the developer.
 \bigskip
@@
 The lower the energy consumption, the higher the score.
-We adopted a \textsc{logarithmic scale} to help the practitioners to compare numerous implementations due to the large differences between them.
+We adopted a \textsc{logarithmic scale} to help the practitioners compare numerous implementations due to their large differences.
 
@@
 Moreover, IntelPython, and ActivePython all behave similarly to the reference implementation.
-Therefore, one can conclude that the work done on those interpreters is primarily to improve a specific purpose and not the core interpreter.
-ActivePython states that their version is focused on security and prepackaged libraries, which explains why it is slower than other versions due to the addition of this security support.\footnote{\url{https://www.activestate.com/solutions/why-activestate/}}
+Therefore, one can conclude that the work done on those interpreters is primarily to improve a specific purpose, not the core interpreter.
+ActivePython states that its version is focused on security and prepackaged libraries, which explains why it is slower than other versions due to the addition of this security support.\footnote{\url{https://www.activestate.com/solutions/why-activestate/}}
 As for IntelPython, it is designed for machine learning.
 Unfortunately, the \textsc{Tommti} benchmark is geared for general-purpose programming.
-Despite the fact that Nuitxka is a compiler, there was no discernible difference in energy consumption.
+Although Nuitxka is a compiler, there was no discernible difference in energy consumption.
 It was even more similar to the reference Python implementation than other interpreters.
-However, if we look at the Nuitka techniques, we can see that they simply encapsulate the Python code with an interpreter into a single executable.
-Finally, Shedskin reports on the best energy consumption pattern when it comes to arithmetic operations.
-One can conclude it is due to the fact of the native type of the variables, unlike in the JIT, where they are treated as objects in the beginning.
+However, looking at the Nuitka techniques, we can see that they simply encapsulate the Python code with an interpreter into a single executable.
+Finally, Shedskin reports on the best energy consumption pattern in arithmetic operations.
+One can conclude it is due to the native type of the variables, unlike in the JIT, where they are treated as objects initially.
 
-Regarding the other VM-based interpereters, Jython and Ipy lacked in terms of energy optimization, which expected as they were at the beginning of the development stage and the main purpose of such implementation is to link the bytecode generated by Jython and IronPython with their respective virtual machines.
+Regarding the other VM-based interpreters, Jython and Ipy lacked in terms of energy optimization, which was expected as they were at the beginning of the development stage.
+The primary purpose of such implementation is to link the bytecode generated by Jython and IronPython with their respective virtual machines.
 
-Unlike the previous interpeters, GraalPython exhibits a certain promise when it comes to complex algorithms (nested loops in particular).
-However, it is still in its early stages of development and it is not yet ready for production.
-Since some benchmarks required an abnormally lengthy time, such as array manipulation and sorting.
+Unlike the previous interpreters, GraalPython exhibits a specific promise regarding complex algorithms (nested loops in particular).
+However, it is still in its early stages of development.
+It is not yet ready for production since some benchmarks, such as array manipulation and sorting, require an abnormally lengthy time.
 
@@
 Due to the lack of support for most non-conventional Python interpreters, we mainly focus on micro-benchmarks.
-Except for \textsc{PyPy}, most of the Python implementations do not support extra Python libraries, despite those extra implementations being developed to optimize a specific library, such as \textsf{Numba} with \textsf{Numpy}, or \textsf{intelpython} with machine learning algorithms.
-Furthermore, we excluded from this study runtimes that were unable to complete our benchmarks because we judged that this was the bare minimum for an implementation to be employed in a real-world application.
-Besides this, since the goal of this study was to provide a non-intrusive technique to decrease the energy consumption of software, we opted to omit some of the implementations that required extra work on the sourcecode to run on them, such as \textsf{Cython}.
+Except for \textsc{PyPy}, most of the Python implementations do not support extra Python libraries, despite those different implementations being developed to optimize a specific library, such as \textsf{Numba} with \textsf{Numpy}, or \textsf{intelpython} with machine learning algorithms.
+Furthermore, we excluded from this study runtime that could not complete our benchmarks because we judged that this was the bare minimum for an implementation to be employed in a real-world application.
+Besides this, since the goal of this study was to provide a non-intrusive technique to decrease the energy consumption of software, we opted to omit some of the implementations that required extra work on the source code to run on them, such as \textsf{Cython}.
 
@@
 
-One can conclude that the Python interpreter used has a considerable impact on the energy consumption of the programs.
-Furthermore, the absence of a universal solution makes this invistigation more intriguing.
-While Python2 was employed by the majority of the candidates, Python3 provided better compatibility coverage.
-Except for Pypy, a standout interpreter, the type of optimization strategy had a substantial impact on energy usage because each class was grouped together regardless of Python version.
-Finally The use of JIT, a technology that converts code into machine code at runtime, was responsible for the significant gain in energy efficiency.
+One can conclude that the Python interpreter used considerably impacts the programs' energy consumption. Furthermore, the absence of a universal solution makes this investigation more intriguing.
+While most of the candidates employed Python2, Python3 provided better compatibility coverage.
+Except for Pypy, a standout interpreter, the optimization strategy substantially impacted energy usage because each class was grouped regardless of the Python version.
+Finally, JIT, a technology that converts code into machine code at runtime, was responsible for the significant gain in energy efficiency.
 The usage of translators such as Shedskin confirmed the effect of compiling Python code into machine code.
-Hence, making PyPy, an interpreter that has a built-in JIT, within the same class as the complied ones such as Shedskin
+Hence, PyPy, an interpreter with a built-in JIT, is within the same class as the complied ones, such as Shedskin.

### 2022-09-28 chapters/optimization/python-dir/python.tex
update : typo

@@
 According to a survey conducted in 2018 by JetBrains,\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}} one can fear that the wide adoption of dynamic programming languages in production, like Python, may critically hamper the power consumption of ICT.
-As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\,Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mostly for energy considerations.
+As the popularity of such dynamic programming languages partly builds on the wealth and the diversity of their ecosystem (\emph{e.g.}, the NumPY, SciKit\, Learn, and Panda libraries in Python), one cannot reasonably expect that developers will likely move to an alternative programming language mainly for energy considerations.
 Rather, we believe that a better option consists of leveraging the strength of this rich ecosystem to promote energy-efficient solutions to improve the power consumption of legacy software systems.
@@
 \subsection{Python usecases}
-To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
+To reduce the energy consumption of Python, we started by targeting the primary usage of this programming language, which is revealed to be data science and web development.
 \Cref{fig:usecase} illustrates a study published by the JetBrain company on Python developers.\footnote{\url{https://www.jetbrains.com/lp/python-developers-survey-2020}}
@@
 \subparagraph{Datasets}
-The CIFAR-10 dataset was the major source of data for the studies.
+The CIFAR-10 dataset was the primary source of data for the studies.
 It is made up of 60000 32x32 color images grouped into ten categories.
@@
     \includegraphics[width=\linewidth]{imgs/cumulative_energy_vs_accuracy}
-    \caption{Cumulative energy consumption vs accuracy.}
+    \caption{Cumulative energy consumption vs. accuracy.}
     \label{fig:cum_energy_vs_accuracy}
@@
 While using the GPU, the CPU has no significant impact on the energy consumption of the training.
-Therefore, one can limit the number of CPU cores used during the execution without sacrificing either the execution time nor the accuracy.
+Therefore, one can limit the number of CPU cores used during the execution without sacrificing the execution time and accuracy.
 
@@
 Django\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}} are the most popular frameworks for web development.
-According to a Jetbrains poll,\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.
+According to the Jetbrains poll,\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.
 
@@
 \subsection{Python Insights}
-Another field of investigation is the type of data and control structures that might impact energy consumption.
+Another field of investigation is the data and control structures that might impact energy consumption.
 In this study, we will run some basic algortims with different data structures and control structures to see if there is any difference in energy consumption.
@@
     \includegraphics[width=\linewidth]{imgs/python_multiprocessing}
-    \caption{Energy behaviour resulting from diffrent python concurrency strategies (number of workers) }
+    \caption{Energy behavior resulting from different python concurrency strategies (number of workers) }
     \label{fig:python_multiprocessing}
@@
 
-Figure~\ref{fig:python_multiprocessing} shows the results of the experiment regarding four metrics: power, execution time , DRAM energy, and CPU energy .
+Figure~\ref{fig:python_multiprocessing} shows the experiment's results regarding four metrics: power, execution time, DRAM energy, and CPU energy.
 
-As one can notice, CPU energy is ten folds higher than the DRAMS one.Therefore we will focus on the CPU for this study.
+One can notice that CPU energy is tenfold higher than the DRAMS one. Therefore we will focus on the CPU for this study.
 
-In general, the multi-processing strategy is the most energy efficient one as well as the fastest one.
+In general, the multi-processing strategy is the most energy-efficient one as well as the fastest one.
 However, unexpectedly, the multithreaded option took more time to execute the code snippet than the sequential one.
-Therefore, we will divide our analysis into two parts: the first one will focus on multithreading versus the sequential strategies, while the second one will study the behavior of the multiprocessing strategy.
+Therefore, we will divide our analysis into two parts: the first will focus on multithreading versus sequential strategies.
+In contrast, the second one will study the behavior of the multiprocessing strategy.
 
-\paragraph*{Multithreading vs Sequential}
-The unexpected behavior of the multithreading strategy is due to the fact that the Python interpreter is not thread-safe.
-Because of the \emph{global lock system} (GLS), Python can't run multiple threads at the same time. Therefore, the multithreading strategy is slower than the sequential one.
+\paragraph*{Multithreading vs. Sequential}
+The unexpected behavior of the multithreading strategy is that the Python interpreter is not thread-safe.
+Because of the \emph{global lock system} (GLS), Python cannot run multiple threads simultaneously. Therefore, the multithreading strategy is slower than the sequential one.
 On the other hand, a side effect of this is the context switching between threads, which will put the CPU in a lower power state.
 The multi-threading strategy depicts an average power of $13 Watts$ compared to the sequential one, which is around $22 Watts$.
-This difference in power consumption has overcome the lack of performance of the multithreading strategy, which leads to better energy efficiency.
+This difference in power consumption has overcome the multithreading strategy's lack of performance, leading to better energy efficiency.
 
 \paragraph*{Multiprocessing}
-Unlike the multi-threading strategy, the multi-processing strategy is based on forking the program into multiple processes, which are independent of each other.
+Unlike the multi-threading strategy, the multi-processing strategy is based on forking the program into multiple processes independent of each other.
 This allows the Python interpreter to run multiple processes simultaneously, which explains the better performance of the multi-processing strategy.
@@
 
-Before reaching the limit of physical cores, we also observe that the scheduler of the operating system tends to favor the execution of processes on the same physical core by taking advantage of the hyper-thread feature.
+Before reaching the limit of physical cores, we also observe that the operating system's scheduler tends to favor the execution of processes on the same physical core by taking advantage of the hyper-thread feature.
 While this strategy aims to save energy by leveraging the ACPI P-states and C-states of unallocated cores, this leads to increased execution times.
@@
 \paragraph*{Synthesis}
-This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to use the multiprocessing strategy, which is the most energy-efficient one.
-However, this strategy should be used cautiously as it can lead to increased execution times.
+This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to use the multiprocessing strategy, which is the most energy-efficient.
+However, this strategy should be used cautiously, leading to increased execution times.
 This study demonstrates that the optimal number of processes is equal to the number of physical cores, which is four in our case.
-By using the multithreading strategy, it was also found that sometimes the performance of the application has to be given up in order to save energy.
-
 \clearpage

### 2022-09-28 chapters/optimization/python-dir/python.tex
Proofreading python

--- 2022-09-28 chapters/optimization/python-dir/python.tex
@@
 \section{Introduction}
+Dynamic programming languages, except Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf. \Cref{fig:pypl}).
+However, it remains unclear if this category of dynamic programming languages can truly compete with compiled ones in terms of energy consumption.
 
-Dynamic programming languages, except Perl, have surpassed compiled programming languages in terms of popularity among software system developers over the past decade (cf. \Cref{fig:pypl}).
-However, it remains unclear whether this category of dynamic programming languages can truly compete with compiled ones in terms of power consumption.
 \begin{figure}[!htb]
     \includegraphics[width=\linewidth]{imgs/programminglanguangespopularity.png}
-    \caption{PYPL Popularity of Programming Languages~\cite{noauthor_pypl_2018}.}
+    \caption{PYPL popularity of programming languages~\cite{noauthor_pypl_2018}.}
     \label{fig:pypl}
@@
 
-In particular, \citeauthor{noureddine_preliminary_2012} in 2012~\cite{noureddine_preliminary_2012}, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
+In particular, \citeauthor{noureddine_preliminary_2012}~\cite{noureddine_preliminary_2012} in 2012, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic: both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
 According to their experiments, an interpreted programming language, like Python, can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. \Cref{fig:hannoi}).
+
 \begin{figure}[!htb]
     \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
-    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
+    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}.}
     \label{fig:hannoi}
@@
 We first start by presenting the motivation behind our work in Section~\ref{python:sec_motivation}.
-In the second section, we share some insights on the impact of programmer choices.
-To do this, we will run a series of experiments to examine the energy behavior of Python code in most of its use cases.Section\ref{python:sec_insights} will then end by investigating several features of Python structure and their impact on energy consumption.
-After that, in Section~\ref{python:sec_interpreters}, we investigate other non-intrusive approaches to optimizing the energy consumption of applications by comparing multiple Python implementations, which include alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as ahead-of-time (AOT) compilation and just-in-time (JIT) libraries that are maintained by the community.
+Then, we share some insights on the impact of programmer choices.
+To do this, we report on a series of experiments to examine the energy consumption of Python code across various use cases.
+Section~\ref{python:sec_insights} will then end by investigating several features of Python structure and their impact on energy consumption.
+After that, in Section~\ref{python:sec_interpreters}, we investigate other non-intrusive approaches to optimizing the energy consumption of applications by comparing multiple Python runtime implementations, which include alternative interpreters and libraries that are dedicated to optimizing the code without changing its structures, such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
 
@@
 As shown in \Cref{fig:clbg}, one can observe that, for most of the applications taken for the \emph{Computer Language Benchmark Game} (CLBG), Python takes more time to execute---the only case that he was not the worst one was in the benchmark \textsf{regx-redux} where he beats Go---and in some cases the gap was huge, such as in \textsf{n-body} where Python took around $100$ times more than C++.\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}
-
-Python consumes energy mainly because it is slow in execution.
-Its flexibility and simplicity caused it to drop off in performance because Python gains its flexibility from being a dynamic language.
-Therefore, it requires a faster interpreter to execute its programs to compete against alternatives written in compiled programming languages, such as C and C++, or semi-compiled languages like Java.
 
@@
 
-\subsection{Python usecases}
+Python consumes energy mainly because it is slow in execution.
+Its flexibility and simplicity caused it to drop off in performance because Python gains its flexibility from being a dynamic language.
+Therefore, it requires a faster interpreter to execute its programs to compete against alternatives written in compiled programming languages, such as C and C++, or semi-compiled languages, like Java.
+
+\subsection{Python Use Cases}
 To reduce the energy consumption of Python, we started by targeting the main usage of this programming language, which is revealed to be data science and web development.
@@
 57\% of the respondents reported that they use Python for data science, and 51\% said they use it for web development.
-Around 40\% are using it for system administration~\footnote{The options in this survey were not mutually exclusive.
-As a result, the total of the percentages is greater than $100\%$.}.
+Around 40\% are using it for system administration.\footnote{The options in this survey were not mutually exclusive. As a result, the total of the percentages is greater than $100\%$.}
 
@@
 
-\section{Energy optimization of python on the source code level}\label{python:sec_insights}
-In this section, we investigate the energy footprint of Python in its most popular domains of adoption.
+\section{Optimizing the Energy Consumption of Python Applications}\label{python:sec_insights}
+In this section, we study the energy footprint of Python in its most popular domains of adoption.
 We first explore the data and control structures, aiming to reveal some fundamental guidelines, as \citeauthor{hasan_energy_2016} did in~\cite{hasan_energy_2016}.
 Then, we measure the energy consumption of several Python implementations to propose a non-intrusive technique to improve energy efficiency.
-Therefore, this chapter will address the following research questions:
+Overall, this section will address the following research questions:
 \begin{compactenum}[\indent\bf RQ\,1:]
@@
 First, we study the energy behavior of Python in two application contexts: machine learning (cf. \Cref{sec:webdev}) and web applications (cf. \Cref{sec:ml}).
-Then, we dive deeper into the energy consumption of Python core structures before concluding with the impact of parallelism on energy consumption.
+Then, we dive deeper into the energy consumption of Python core structures, before concluding with the impact of parallelism on energy consumption.
 
-\subsection{Python \& Machine Learning}\label{sec:ml}
+\subsection{Python for Machine Learning}\label{sec:ml}
 Machine learning is becoming an integral part of our daily lives, growing more potent and energy-hungry each year.
 As machine learning can have a significant impact on climate change, it is vital to investigate mitigation techniques.
+
 \subsubsection{Experimental Protocol}
-\paragraph{Measurement context}
-\subparagraph{Hardware settings}
-Chifflot 8 from Grid 5000's Lille site was used for all of the trials.
+\paragraph{Measurement Context}
+\subparagraph{Hardware settings:}
+Chifflot\,8 from Grid 5000's Lille site was used for all of the trials.
 The machine is outfitted with two Intel Xeon Gold 6126 CPUs, each having 12 physical cores, 192 GB of RAM, and two 32 GB Tesla V100 GPUs.
 
-\subparagraph{Software settings}
+\subparagraph{Software settings:}
 Each experiment is done within a Docker container using Jupyter lab to ensure reproducibility.
-These tests are run on top of a minimal version of Debian-10 to increase the tests' accuracy by eliminating unnecessary processes.
+These tests are run atop a minimal version of Debian-10 to increase the tests' accuracy by eliminating unnecessary processes.
+
 \subsubsection{Input Workload}
-\subparagraph{Models}
-Several models were developed. However, only two were used in the final trials because they achieved 94 percent accuracy in an acceptable length of time.
-David Page's cifar10-fast~\footnote{\url{https://github.com/davidcpage/cifar10-fast}} and Woonhyuk Baek's torch skeleton~\footnote{\url{https://github.com/wbaek/torchskeleton}} are shown here.
+\subparagraph{Models:}
+Several models were developed.
+However, only two were used in the final trials, because they achieved 94\% accuracy with an acceptable training time.
+David Page's \textsf{cifar10-fast}\footnote{\url{https://github.com/davidcpage/cifar10-fast}} and Woonhyuk Baek's torch skeleton\footnote{\url{https://github.com/wbaek/torchskeleton}} are reported here.
 
-\subparagraph{Datasets}
+\subparagraph{Datasets:}
 The CIFAR-10 dataset was the major source of data for the studies.
-It is made up of 60000 32x32 color images grouped into ten categories.
+It is made up of $60,000$ \texttt{32x32} color images grouped into $10$ categories.
 
 Some experiments were done using the MNIST dataset of handwritten digits to validate the results acquired from the first dataset.
-The model did not need to be updated because the 60000 28x28 grayscale photos were padded.
+The model did not need to be updated because the $60,000$ \texttt{28x28} grayscale photos were padded.
 
-\paragraph{candidates}
+\paragraph{Candidates:}
 The experiments were run with several different CPU and GPU configurations:
@@
 
-\paragraph{Key Performance Metrics}
-We used Pytorch 1.10.0 to train those models and pyjoules to measure the energy consumption of the GPU and CPU.
+\paragraph{Key Performance Metrics:}
+We used Pytorch 1.10.0 to train those models and pyJoules to measure the energy consumption of the GPU and CPU.
+The key performance metrics are, therefore:
 \begin{itemize}
-    \item accuracy: in \%
-    \item execution time: in seconds for both the duration of each epoch and the total duration to achieve a certain accuracy
-    \item total energy consumption: for the CPU and the  GPU
+    \item \emph{accuracy}: in \%
+    \item \emph{execution time}: in seconds for both the duration of each epoch and the total duration to achieve a certain accuracy
+    \item \emph{total energy consumption}: in joules, including the CPU and the GPU.
 \end{itemize}
@@
 \subsubsection{Results \& Findings}
-
-As the model's accuracy increases, so do the energy required for the next accuracy increment.
-
+As the model's accuracy increases, so does the energy required for the next accuracy increment.
 Figure~\ref{fig:cum_energy_vs_accuracy} depicts how the curve steepens as training progresses.
@@
 
+\Cref{fig:eopochvsaccuracy} shows the model's accuracy based on the number of epochs.
+This figure shows that there is a non-significant impact on the choice of the strategy on the accuracy; it is only a matter of the number of epochs.
+However, the increase in accuracy is not linear, as can be observed.
+It took only 10 epochs to reach an accuracy of 84\%, but it required more than twice the number of epochs to add an axtra 6\% accuracy.
+This highlights the price one should pay to increase the model's accuracy.
+
 \begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{imgs/accuracy_basedonepoch}
+    \includegraphics[width=1.1\textwidth]{imgs/accuracy_basedonepoch}
     \caption{Accuracy along epochs.}
@@
 
+\Cref{fig:cum_energy_vs_accuracy} confirm this observation.
+As one can see, the training of the model up to 90\% accuracy requires three times the energy required for training to 80\% accuracy.
+Moreover, one can notice that this price is paid mainly by the CPU and GPU, while the memory is not that impacted.
+
 \begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{imgs/cumulative_energy_vs_accuracy}
+    \includegraphics[width=1.1\textwidth]{imgs/cumulative_energy_vs_accuracy}
     \caption{Cumulative energy consumption vs accuracy.}
@@
 
-cref{fig:eopochvsaccuracy}shows the model's accuracy based on the number of epochs.
-This figure shows that there is a non-significant impact on the choice of the strategy on the accuracy; it is only a matter of the number of epochs.
-Hover, the increase in accuracy is not linear, as can be seen.
-It took only ten epochs to reach an accuracy of 84\%, but it required more than twice the number of epochs to add axtra 6\% accuracy.
-This highlights the price one should pay to increase the model's accuracy.\cref{fig:cum_energy_vs_accuracy} confirm a such claim.
-As one can see, the training of the model up to 90\% accuracy requires three times the energy required for training to 80\% accuracy.
-Moreover, one can notice that this price is paid mainly by the CPU and GPU, while the memory is not that impacted.
-
-Interestingly there were no significant differences between different strategies -around $1.3\%$- As shown in Figures ~\ref{fig:av_energy_setup}.
+Interestingly there were no significant differences between different strategies---around $1.3\%$---as shown in Figure~\ref{fig:av_energy_setup}.
 However, this gap may increase as the number of epochs increases.
-
-On the other hand, as the number of epochs increases, the average power consumption decreases until it reaches a steady state after ten epochs.
-This could be connected to the caching methods utilized by the CPU and GPU.
-However, this evolution is still insignificant compared to the baseline value.
-This is shown in Figure~\ref{fig:ephoch_power}.
-
-As one can see in figures \ref{fig:ephoch_power}~\ref{fig:epoch_duration}, the average power consumption and the average duration of an epoch don't have a significant variation, as most of them were around 3.7 seconds.
-However, there was an intermediate correlation between these two values if we excluded the strategy based on one core plus its hyper thread.
-The reason why the last strategy exhibits more power even when its duration isn't optimal is the context switched between the core and its hyper thread.
 
@@
     \centering
-    \includegraphics[width=\linewidth]{imgs/cumulative_energy_fast10.pdf}
+    \includegraphics[width=1.1\textwidth]{imgs/cumulative_energy_fast10.pdf}
     \caption{Average energy for training the model based on the strategy.}
@@
 
+On the other hand, as the number of epochs increases, the average power consumption decreases until it reaches a steady state after 10 epochs.
+This could be connected to the caching strategies adopted by the CPU and GPU.
+However, this evolution is still insignificant compared to the baseline value.
+This is shown in Figure~\ref{fig:ephoch_power}.
+
 \begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{imgs/power_gpu_baedonepoche.pdf}
+    \includegraphics[width=1.1\textwidth]{imgs/power_gpu_baedonepoche.pdf}
     \caption{Evolution of average GPU power along epochs.}
@@
 
+As one can see in Figures~\ref{fig:ephoch_power} and~\ref{fig:epoch_duration}, the average power consumption and the average duration of an epoch do not have a significant variation, as most of them last around 3.7 seconds.
+However, there is an intermediate correlation between these two values if we excluded the strategy based on one core and its hyper-thread.
+The reason why the last strategy exhibits more power---even when its duration is not optimal---is the context switches between the core and its hyper-thread.
+
 \begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{imgs/epoch_duration.pdf}
-    \caption{Distribtion of the duration of each epoch.}
+    \includegraphics[width=1.1\textwidth]{imgs/epoch_duration.pdf}
+    \caption{Distribution of the durations of each epoch.}
     \label{fig:epoch_duration}
 \end{figure}
+
 \subsubsection{Synthesis}
@@
 
-\subsection{Green Web Development}\label{sec:webdev}
-
-Django\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}} are the most popular frameworks for web development.
+\subsection{Python for Web Development}\label{sec:webdev}
+Django\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}} are the most popular Pyhton frameworks for web development.
 According to a Jetbrains poll,\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.
@@
 
-\subsubsection{Life-cycle of a Request in Django}
+\paragraph{Life-cycle of a request in Django}
 Django is an MVT (\emph{Model-View-Template}) framework, which means that it follows the MVC (\emph{Model-View-Controller}) pattern.
@@
     \centering
-    \includegraphics[width=\linewidth]{imgs/django_request_lifecycle}
+    \includegraphics[width=.9\linewidth]{imgs/django_request_lifecycle}
     \caption{Request-Response life cycle in Django}
@@
 To do so, we created a sample Django application that returns the response to the request.
-We tracked its energy consumption using JouleHunter~\footnote{\url{https://github.com/powerapi-ng/joulehunter}}.
+We tracked its energy consumption using JouleHunter.\footnote{\url{https://github.com/powerapi-ng/joulehunter}}
 JouleHunter is an open-source library that we developed to help practitioners identify energy hotspots in their applications using statistical profilers.
-In the case of Django, Joulehunter will be added as a middleware with no additional setup or change to the source code.
+In the case of Django, JouleHunter is included as a middleware with no additional setup or change to the source code.
 The energy consumption of the request-response life-cycle in Django is shown in Figure~\ref{fig:django_life_cycle_naive}.
@@
     \includegraphics[width=\linewidth]{imgs/django_life_cycle_naive}
-    \caption{Tree representation of the energy consumption of a single request in Django (naive version)}
+    \caption{Tree representation of the energy consumption of a single request in Django (naive version).}
     \label{fig:django_life_cycle_naive}
@@
 % We have chosen Django as our web framework.
-Therefore, we chose to study if the choice of the database and the ORM impacts the energy consumption and the website's performance.
+Therefore, we chose to study if the choice of the database and the ORM impacts the energy consumption and the web application performance.
 % \subsection{Experiment \& Results}
@@
     \includegraphics[width=\linewidth]{imgs/django}
-    \caption{Energy behavior resulting from data access strategies.}
+    \caption{Key performance indicators observed for different data access strategies.}
     \label{fig:django}
@@
 Another interesting observation is the impact of the interpreter, as \Cref{fig:django} highlights.
-For example, using the \textsf{PyPy} interpreter reduces energy consumption, even when adopting the \textsf{Vanilla} strategy. We will further discuss the choice of an interpreter in the next section.
+For example, using the \textsf{PyPy} interpreter reduces energy consumption, even when adopting the \textsf{Vanilla} strategy.
+We will further discuss the choice of Python runtime implementation in the next section.
 
@@
 As one can notice, while the rendering method consumed the same amount of energy as in the \textsf{Vanilla} strategy (around 1.3\,kJ), the resolve part dropped by $20\times$.
-\paragraph{Synthesis}
-
-We can conclude that the database and ORM selection significantly impact the website's performance and energy consumption.
-Moreover, unlike the rendering portion, many options are available for the database, the ORM, and the data handling strategy.
 
@@
     \includegraphics[width=\linewidth]{imgs/django_profiled_optimized}
-    \caption{Tree representation of the energy consumption of a single request in Django (naive version)}
+    \caption{Tree representation of the energy consumption of a request in Django (optimized version)}
     \label{fig:django_profiled_optimized}
@@
 
-\subsection{Python Insights}
+\paragraph{Synthesis}
+We can conclude that the database and ORM selection significantly impact the web application performance and energy consumption.
+Moreover, unlike rendering responses, many alternatives can be considered to store and query the application database.
+
+\subsection{Python for Data Processing}
 Another field of investigation is the type of data and control structures that might impact energy consumption.
-In this study, we will run some basic algortims with different data structures and control structures to see if there is any difference in energy consumption.
-\subsubsection{experimental Protocol}
-All of the tests in this study are done on the grid5000 cluster using the same machine, Python interpreter, and library version.
-Due to PowerAPI's frequency constraint, we measure the energy of 1000 algorithm iterations for each test.
+In this study, we will run some common algorithms with different data structures and control structures to study if there is any difference in energy consumption.
+\subsubsection{Experimental Protocol}
+All of the tests in this study are done on the Grid5000 cluster using the same machine, Python interpreter, and library version.
+Due to PowerAPI's frequency constraint, we measure the energy of $1,000$ algorithm iterations for each test.
 Furthermore, we perform each test 20 times for completeness, deleting the cache between each run.
@@
 
-\subsubsection{Python loops}
-In the first experiment, we want to study the impact of the loop type on energy consumption. To do so, we run the following code snippet:
+\subsubsection{Python Loops}
+In the first experiment, we study the impact of the type of loop on energy consumption.
+To do so, we execute the following code snippet:
 \begin{algorithm}[!htb]
@@
 However, as we can see here, unlike other programming languages, it requires extra operations, such as determining the collection length and then using the iterator range.
-So, we tried the more adapted version \texttt{for(element in collection)}.
-Moreover, in most programming languages.
-the \texttt{for} loop is translated to a \texttt{while} loop ( transformation from D type to B type -- asm -- ), therefore we wanted to compare this with a \texttt{while} version.
-Therefore our candidates will be :
-
+So, we tried an alternative version: \texttt{for(element in collection)}.
+Moreover, in most programming languages, the \texttt{for} loop is translated into a \texttt{while} loop, so we wanted to compare this with a \texttt{while} version.
+Therefore, our candidate loops are defined as:
 \begin{itemize}[]
-    \item \texttt{for(i in range(len(n)))}
-    \item \texttt{for(element in collection)}
-    \item \texttt{while}
+    \item \texttt{for (i in range(len(n)))},
+    \item \texttt{for (element in collection)},
+    \item \texttt{while}.
 \end{itemize}
 
-We will run the same code snippet for each of these implementations with different primitive data types: Int, Float, and String.
+Furthermore, we run the same code snippet for each of these implementations with different primitive data types: \texttt{Int}, \texttt{Float}, and \texttt{String}.
+
 \begin{figure}[!htb]
@@
 
-Figure~\ref{fig:pythonloops} shows the results of the experiment. As one can notice, the data type has a negligible impact on energy consumption.
+Figure~\ref{fig:pythonloops} shows the results of the experiment.
+As one can notice, the data type has a negligible impact on energy consumption.
 However, the way one iterates over the collection has a huge factor.
-Interestingly, the \texttt{for in range} loop was by far the optimal one, followed by the regular \texttt{for in collection}, and the \texttt{while} part was the last one with an overhead of 400\% compared to the first option.
+Interestingly, the \texttt{for in range(...)} loop was by far the optimal one, followed by the regular \texttt{for in collection}, and the \texttt{while} part was the last one with an overhead of 400\% compared to the first option.
 
-The reason behind such behavior is mainly related to how the Python interpreter is implemented\footnote{\url{https://www.python.org/doc/essays/list2str}},\footnote{\url{https://www.pythonpool.com/for-vs-while-loop-python/}}.
-Most of the built-in functions and operations are written in C to reduce the latency of Python applications, and the same goes for the function \texttt{range}.\footnote{\url{ http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1}}
-Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$ as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object~\footnote{\url{https://wiki.python.org/moin/TimeComplexity}}.
-Therefore, the \texttt{for in range} is creating a new iterator that has the same length as the first one and, for each iteration, requires second access (\texttt{l[i]}) instead of one---explaining the doubled time.
+The reason behind such behavior is mainly related to how the Python interpreter is implemented.\footnote{\url{https://www.python.org/doc/essays/list2str}},\footnote{\url{https://www.pythonpool.com/for-vs-while-loop-python/}}
+Most of the built-in functions and operations are written in C, to reduce the latency of Python applications, and the same goes for the function \texttt{range}.\footnote{\url{ http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1}}
+Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$, as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object~\footnote{\url{https://wiki.python.org/moin/TimeComplexity}}.
+Therefore, the \texttt{for in range(...)} is creating a new iterator that has the same length as the first one and, for each iteration, requires second access (\texttt{l[i]}) instead of one---explaining the doubled time.
 The \texttt{while} is even slower due to the implicit increment of the variable, which causes an extra operation during the loop.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{imgs/python_treatemens}
-    \caption{Comparison of the energy consumption of different methods to convert a list.}
-    \label{fig:pythontreatement}
-\end{figure}
 
@@
 We used four different methods to do so: \texttt{comprehension list}, \texttt{while loop}, \texttt{a map with predefined function}, and \texttt{a map with an anonymous function}.
-
-A comprehension list is a Pythonic way to create a new list by applying a function to each element of the previous one. It is based on the mathematical set-builder notation~\cite{editiondiscrete}.
-
+A comprehension list is a Pythonic way to create a new list by applying a function to each element of the previous one.
+It is based on the mathematical set-builder notation~\cite{editiondiscrete}.
 The map function is a higher-order function that applies a given function to each collection element.
@@
 Moreover, when increasing the collection size, the map with lambda functions tends to consume less energy than the predefined one.
-The reason behind such behavior is that Python treats these functions as local variables, unlike the predefined ones, which are global in our case. Therefore, they are faster and consume less energy.
+The reason behind such behavior is that Python treats these functions as local variables, unlike the predefined ones, which are global in our case.
+Therefore, they are faster and consume less energy.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\linewidth]{imgs/python_treatemens}
+    \caption{Comparison of the energy consumption of different methods to convert a list.}
+    \label{fig:pythontreatement}
+\end{figure}
 
@@
 To do so, we run a simple code snippet that computes the sum of the first $N$ numbers using the standard concurrency libraries \texttt{multiprocessing~\footnote{\url{https://docs.python.org/3/library/multiprocessing.html}}} and \texttt{multihtreading~\footnote{\url{https://docs.python.org/3/library/threading.html}}}.
-We run the following strategies using a Desktop machine with Intel(R) Core(TM) i7-4800MQ CPU @ 2.70GHz with four cores and four hyperthreads.
+We run the following strategies using a Desktop machine with Intel(R) Core(TM) i7-4800MQ CPU @ 2.70GHz with four cores and four hyper-threads.
 %  N ==5e5
 \begin{itemize}
-    \item Sequentiel: and here we just compute the sum of the first $N$ numbers without any concurency
-    \item Multithreading: Here, we use the \texttt{ ThreadPoolExecutor} library to compute the sum of the first $N$ numbers, We devide $N$ by the number of threads, and each thread will compute the sum of the numbers in its range.
-For our case, we used four threads.
-    \item multiprocessing: and in this case, we use the \texttt{ProcessPoolExecutor} library to calculate the sum of the first $N$ numbers.
+    \item \textsf{Sequential}: we compute the sum of the first $N$ numbers with no concurency;
+    \item \textsf{Multithreading}: we use the \texttt{ThreadPoolExecutor} library to compute the sum of the first $N$ numbers.
+We divide $N$ by the number of threads, and each thread will compute the sum of the numbers in its range.
+For our case, we used four threads;
+    \item \textsf{Multiprocessing}: we use the \texttt{ProcessPoolExecutor} library to compute the sum of the first $N$ numbers.
 We divide $N$ by the number of processes, and each process sums the numbers in its range.
-In this situation, we employed 2,4,8,16 processes correspondingly.
+In this situation, we considered 2, 4, 8, and 16 processes iteratively.
 \end{itemize}
+
+Figure~\ref{fig:python_multiprocessing} shows the results of the experiment regarding four metrics: power, execution time, DRAM energy, and CPU energy.
+As one can notice, CPU energy is ten folds higher than the DRAM one.
+Therefore, we will focus on the CPU for this study.
 
@@
     \includegraphics[width=\linewidth]{imgs/python_multiprocessing}
-    \caption{Energy behaviour resulting from diffrent python concurrency strategies (number of workers) }
+    \caption{Energy consumption resulting from different Python concurrency strategies (number of workers).}
     \label{fig:python_multiprocessing}
@@
 
-Figure~\ref{fig:python_multiprocessing} shows the results of the experiment regarding four metrics: power, execution time , DRAM energy, and CPU energy .
-
-As one can notice, CPU energy is ten folds higher than the DRAMS one.Therefore we will focus on the CPU for this study.
-
-In general, the multi-processing strategy is the most energy efficient one as well as the fastest one.
-However, unexpectedly, the multithreaded option took more time to execute the code snippet than the sequential one.
-Therefore, we will divide our analysis into two parts: the first one will focus on multithreading versus the sequential strategies, while the second one will study the behavior of the multiprocessing strategy.
+In general, the \textsc{Multiprocessing} strategy is the most energy efficient one, and the fastest one.
+However, unexpectedly, the \textsf{Multithreading} option took more time to execute the code snippet than the \textsf{Sequential} one.
+Therefore, we will divide our analysis into two parts: the first one will focus on \textsf{Multithreading} versus the \textsf{Sequential} strategies, while the second one will study the behavior of the \textsc{Multiprocessing} strategy.
 
 \paragraph*{Multithreading vs Sequential}
-The unexpected behavior of the multithreading strategy is due to the fact that the Python interpreter is not thread-safe.
-Because of the \emph{global lock system} (GLS), Python can't run multiple threads at the same time. Therefore, the multithreading strategy is slower than the sequential one.
+The unexpected behavior of the \textsc{Multiprocessing} strategy is because the Python interpreter is not thread-safe.
+Because of the \emph{global lock system} (GLS), Python cannot run multiple threads at the same time.
+Therefore, the \textsf{Multithreading} strategy is slower than the \textsf{Sequential} one.
 On the other hand, a side effect of this is the context switching between threads, which will put the CPU in a lower power state.
-The multi-threading strategy depicts an average power of $13 Watts$ compared to the sequential one, which is around $22 Watts$.
-This difference in power consumption has overcome the lack of performance of the multithreading strategy, which leads to better energy efficiency.
+The \textsf{Multithreading} strategy depicts an average power of $13 Watts$ compared to the \textsf{Sequential} one, which is around $22 Watts$.
+This difference in power consumption has overcome the lack of performance of the \textsf{Multithreading} strategy, which leads to better energy efficiency.
 
 \paragraph*{Multiprocessing}
-Unlike the multi-threading strategy, the multi-processing strategy is based on forking the program into multiple processes, which are independent of each other.
-This allows the Python interpreter to run multiple processes simultaneously, which explains the better performance of the multi-processing strategy.
-However, this will increase the Average power consumption of the CPU, which is around $40 Watts$, compared to the sequential one, which is around $22 Watts$.
-Nevertheless, the multi-processing strategy is still the most energy-efficient one.
+Unlike the \textsf{Multithreading} strategy, the \textsc{Multiprocessing} strategy is based on forking the program into multiple processes, which are independent of each other. 
+This allows the Python interpreter to run multiple processes simultaneously, hence explaining the better performance of the multi-processing strategy.
+However, this will increase the average power consumption of the CPU, which is around $40 Watts$, compared to the \textsf{Sequential} one, which is around $22 Watts$.
+Nevertheless, the \textsc{Multiprocessing} strategy is still the most energy-efficient one.
 
-Although increasing the number of processes will reduce the execution time because we split the tasks on the number of processes, there is a point where the execution time will increase.
+Although increasing the number of processes will reduce the execution time because we split the tasks on the number of processes, there is a point where the execution time increases.
 As one can see in \Cref{fig:python_multiprocessing}, the energy consumption correlates with the number of processes until the limit of physical cores, when concurrent processes compete for the CPU.
@@
 \paragraph*{Synthesis}
-This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to use the multiprocessing strategy, which is the most energy-efficient one.
-However, this strategy should be used cautiously as it can lead to increased execution times.
+This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to adopt the \textsc{Multiprocessing} strategy, which is the most energy-efficient one.
+However, this strategy should be used carefully as it can lead to increased execution times.
 This study demonstrates that the optimal number of processes is equal to the number of physical cores, which is four in our case.
-By using the multithreading strategy, it was also found that sometimes the performance of the application has to be given up in order to save energy.
+When using the \textsf{Multithreading} strategy, it was also found that sometimes the performance of the application has to be given up to save energy.
 
@@
 \clearpage
-\section{Summary}
-
-Since many software services use Python in public and private cloud infrastructures, making Python-based apps more energy efficient will lower ICT carbon emissions.
+\section{Conclusion}
+As many software services use Python in public and private cloud infrastructures, making Python-based apps more energy efficient will lower ICT carbon emissions.
 This chapter discusses various ways of optimizing the energy consumption of Python-based applications.
 It first explains why such a choice is relevant.
-Then it studies the energy behavior of Python within its most commonly employed use cases, which are revealed to be web development and machine learning.
+Then, it studies the energy behavior of Python within its most commonly employed use cases, which are revealed to be web development and machine learning.
 
-First, we studied the energy consumption of a machine learning algorithm using a benchmark of 60000 entries to train the cifar10-fast model.
+First, we studied the energy consumption of a machine learning algorithm using a benchmark of $60,000$ entries to train the \texttt{cifar10-fast} model.
 We discovered that this energy consumption increases exponentially while the model accuracy increases.
@@
 
-After that, we analyzed the impact of several iteration mechanisms in python and their impact on energy usage.
+After that, we analyzed the impact of several iteration mechanisms in Python and their impact on energy consumption.
 We found that the built-in functions and the best practices for writing Python code are the most energy-efficient ones.
@@
  However, this strategy should be used cautiously.
-This study demonstrates that the optimal number of processes is equal to the number of physical cores, and exceeding this number will cause the application loses its efficiency in both performance and energy consumption.
-As for the multithreading strategy, We found that the Python interpreter is not thread-safe, which leads to a slower performance than the sequential one.
+ This study demonstrates that the optimal number of processes is equal to the number of physical cores, and exceeding this number will cause the application to loose its efficiency in both performance and energy consumption.
+ As for the multithreading strategy, we found that the Python interpreter is not thread-safe, which leads to a slower performance than the sequential one.
  However, this also leads to lower power consumption, which sometimes overcomes the lack of performance to make the application more energy-efficient.
 
-In the second section, we presented a simple technique for optimizing the energy usage of Python-based apps. Without making substantial changes to the code, This is the use of many Python interpreters
-Section 2 started by categorizing and filtering these implementations into three significant kinds (compiler, interpreter, and extra libraries).
-Then we conducted a series of experiments to compare the energy consumption of these alternatives.
-The findings indicate the lack of a general solution and the importance of selecting the appropriate interpreter for the appropriate use case.
-In certain circumstances, some of these answers may be the best, while others may be the worst.
-We found that most interpreters had a similar or worse energy consumption to the official implementation of CPython of their Pythons' version.
-The reason behind such behavior was that some implementations focused on specific usage cases such as machine learning or security.
-In contrast, others focused on the compatibility of the python code with their platforms like Jython and IronPython.
+In the second section, we presented a non-intrusive technique to optimize the energy consumption of Python-based apps without making substantial changes to the code.
+This technique consists in guiding the choice of the Python runtime implementation.
+We started by categorizing and filtering these implementations into three significant classes (compiler, interpreter, and extra libraries).
+Then, we conducted a series of experiments to compare the energy consumption of these alternatives.
+Our findings indicate the lack of a general solution and the importance of tuning the Python runtime depending on applications.
+% In some circumstances, some of these answers may be the best, while others may be the worst.
+We found that most interpreters had a similar or worse energy consumption than the official implementation of CPython.
+The reason behind such behavior was that some implementations focused on specific case studies, such as machine learning or security.
+In contrast, others focused on the compatibility of the Python code with their platforms, like Jython and IronPython.
 Finally, regardless of implementation, we demonstrated that using JIT is the most efficient technique to boost the energy efficiency of Python-based programs.
 
-We believe our contributions will benefit a broad spectrum of legacy systems, reducing ICT's carbon footprint and lowering cloud bills for these services' resources.
+We believe that our contributions will benefit a broad spectrum of legacy systems, reducing ICT's carbon footprint and lowering cloud bills for these services' resources.

### 2022-09-28 chapters/optimization/python-dir/python_interperters.tex
Proofreading python

@@
 \begin{figure}[hbt]
-      \includegraphics[width=.9\linewidth]{imgs/SmartWatts.png}
+      \centering
+      \includegraphics[width=.8\linewidth]{imgs/SmartWatts.png}
       \caption{Benchmarking architecture deployed with \textsc{PowerAPI}.}
@@
       \centering
-      \includegraphics[width=\linewidth]{imgs/dendogram_interpreters}
-      \caption{Dendogram of the different implementations}
+      \includegraphics[width=.8\linewidth]{imgs/dendogram_interpreters}
+      \caption{Dendogram of the different Python runtime implementations.}
       \label{fig:dendogram}
@@
       \includegraphics[height=\textheight,keepaspectratio]{imgs/alltomti_performance}
-      \caption{Different interpreter's energy scores.}
+      \caption{Different Python runtime energy scores.}
       \label{fig:tommi_all}
@@
 Unfortunately, the \textsc{Tommti} benchmark is geared for general-purpose programming.
-Although Nuitxka is a compiler, there was no discernible difference in energy consumption.
+Although Nuitka is a compiler, there was no discernible difference in energy consumption.
 It was even more similar to the reference Python implementation than other interpreters.
@@
 
-\subsubsection{threads to validity }
+\subsubsection{Threats to validity}
 Due to the lack of support for most non-conventional Python interpreters, we mainly focus on micro-benchmarks.
-Except for \textsc{PyPy}, most of the Python implementations do not support extra Python libraries, despite those different implementations being developed to optimize a specific library, such as \textsf{Numba} with \textsf{Numpy}, or \textsf{intelpython} with machine learning algorithms.
+Except for \textsc{PyPy}, most of the Python implementations do not support extra Python libraries, despite those different implementations being developed to optimize a specific library, such as \textsf{Numba} with \textsf{Numpy}, or \textsf{IntelPython} with machine learning algorithms.
 Furthermore, we excluded from this study runtime that could not complete our benchmarks because we judged that this was the bare minimum for an implementation to be employed in a real-world application.
@@
 \subsection{Conclusion}
-
-One can conclude that the Python interpreter used considerably impacts the programs' energy consumption. Furthermore, the absence of a universal solution makes this investigation more intriguing.
-While most of the candidates employed Python2, Python3 provided better compatibility coverage.
-Except for Pypy, a standout interpreter, the optimization strategy substantially impacted energy usage because each class was grouped regardless of the Python version.
+One can conclude that the Python interpreter used considerably impacts the programs' energy consumption.
+Furthermore, the absence of a universal solution makes this investigation more intriguing.
+While most of the candidates employed \textsf{Python2}, \textsf{Python3} provided better compatibility coverage.
+Except for \textsf{Pypy}, a standout interpreter, the optimization strategy substantially impacted energy usage because each class was grouped regardless of the Python version.
 Finally, JIT, a technology that converts code into machine code at runtime, was responsible for the significant gain in energy efficiency.
-The usage of translators such as Shedskin confirmed the effect of compiling Python code into machine code.
-Hence, PyPy, an interpreter with a built-in JIT, is within the same class as the complied ones, such as Shedskin.
+The usage of translators, such as \textsf{Shedskin} confirmed the effect of compiling Python code into machine code.
+Hence, \textsf{PyPy}, an interpreter with a built-in JIT, is within the same class as the complied ones, such as \textsf{Shedskin}.

### 2022-09-28 chapters/benchmarking/conclusion.tex
Proofreading

@@
-%  CONCLUSION au lieu de extension
-
-%% maybe add extension as a part of the conclusion for this chapter 
-\clearpage
-\section{Summary}
+\section{Conclusion}
 there As seen in Section~\cref{sec:soa_benchmarking} of the state of the art, a successful benchmark faces three challenges: reproducibility, accuracy and representativeness. 

### 2022-09-28 chapters/benchmarking/experiment.tex

@@
 As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or \textsf{socket} to designate one of the two sockets and \textsf{PU} for the \emph{processing unit}.
-For our study we consider hyper-threads as distinct \textbf{PUs}.
-
+For our study, we consider hyper-threads as distinct \textbf{PUs}.
 Figure~\ref{fig:topo} shows the detailed topology of a node in the cluster \textsf{Dahu} as an example.
@@
 In most cases, we were seeking to evaluate the \emph{STandard Deviation} (STD), which is the most representative factor of the energy variation.
-We tried to be very careful, while running our experiments, not to fall in the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
+We tried to be very careful, while running our experiments, not to fall into the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
 As we study the STD difference of measurements we observed from empirical experiments, we use the bootstrap method~\cite{efron2000bootstrap} to randomly build multiple subsets of data from the original dataset, and we draw the STD density of those sets, as illustrated in Figure~\ref{fig:docker}.
@@
 We believe this can help to achieve better and more reliable comparisons.
-We mainly consider 3 different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium and high CPU usage when analyzing the studied parameters effect, respectively.
+We mainly consider 3 different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium and high CPU usage when analyzing the effect of the studied parameters, respectively.
 These workloads reflect the ratio of used PU count to the total available PU.
@@
 \subsection{Analysis}\label{subsec:parameters}
-In this part, we aim to establish experimental guidelines to reduce the CPU energy variation.
-We therefore explore many potential factors and parameters that could have a considerable effect on the energy variation.
+In this part, we aim to establish experimental guidelines to reduce CPU energy variation.
+We, therefore, explore many potential factors and parameters that could have a considerable effect on the energy variation.
 
 \subsection{Docker \& Accuracy}
-As the state of the art assesses the impact of Docker on the energy consumption, one can also consider its impact on accuracy.
+As the state of the art assesses the impact of Docker on energy consumption, one can also consider its impact on accuracy.
 In other words:\\
@@
 To answer this question we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
-Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running a same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
-In fact, while Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
+Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running the same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
+While Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
 The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.
@@
 To achieve a robust and reproducible experiment, practitioners often tend to repeat their tests multiple times, in order to analyze the related performance indicators, such as execution time, memory consumption or energy consumption.
-We therefore aim to study the benchmarking protocol to identify how to efficiently iterate the tests to capture a trustable energy consumption evaluation.
+We, therefore, aim to study the benchmarking protocol to identify how to efficiently iterate the tests to capture a trustable energy consumption evaluation.
 % This process of repeating executions should not be random in order to avoid different results because of the CPU behavior, the cache data, the memory size and plenty of other aspects.
@@
 Table~\ref{table:g5k}) for different workloads.
-We note that the results have been executed with different datasets sizes (\textsf{B}, \textsf{C} and \textsf{D} for single process, 50\,\% and 100\,\% respectively) to remedy to the brief execution times at high workloads for small datasets.
-This justifies the scale differences of reported energy consumptions between the 3 modes in Figure~\ref{fig:running-process}.
+We note that the results have been executed with different datasets sizes (\textsf{B}, \textsf{C} and \textsf{D} for a single process, 50\,\% and 100\,\% respectively) to remedy to the brief execution times at high workloads for small datasets.
+This justifies the scale differences in reported energy consumption between the 3 modes in Figure~\ref{fig:running-process}.
 As one can observe, picking one of these strategies does not have a strong impact on the energy variation for most workloads.
@@
 The only exception is the \textsf{reboot} mode at 100\,\% load, where the STD is 150\,\% times worst, due to an important amount of outliers.
-This goes against our expectation, even when setting a warm-up time after reboot to stabilize the OS.
+This goes against our expectations, even when setting a warm-up time after reboot to stabilize the OS.
 
@@
 In Figure~\ref{fig:reboot}, we study the standard deviation of the three modes by constituting $5,000$ random 30-iterations sets from the previous executions set and we compute the STD in each case, considering mainly the 100\,\% workload as the STD was 150\,\% higher for the \textsf{reboot} mode with that load.
-We can observe that the considerable amount of outliers in the \textsf{reboot} mode is not negligible, as the STD density is clearly higher than the two other modes.
-This makes the \textsf{reboot} mode as the less appropriate for the energy variation at high workloads.
+We can observe that the considerable amount of outliers in the \textsf{reboot} mode is not negligible, as the STD density is higher than the two other modes.
+This makes the \textsf{reboot} mode less appropriate for the energy variation at high workloads.
 
@@
 
-    To answer \textsc{RQ~1}, we conclude that the benchmarking protocol \textbf{partially affects} the energy variation, as highlighted by the \textsf{reboot} mode results for high workloads.
+    To answer \textsc{RQ~1}, we conclude that the benchmarking protocol \textbf{partially affects} the energy variation, as highlighted by the \textsf{reboot} mode results in high workloads.
 
@@
 
-We can clearly see the effect that has the \textsf{C-states off} mode when running a single-process application/benchmark.
+We can see the effect that has the \textsf{C-states off} mode when running a single-process application/benchmark.
 The energy consumption varies 5~times less than the default mode.
@@
 The other cores are switched to a low consumption state when C-states are on, the switching operation causes an important energy consumption difference between the cores, and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
-On the other hand, switching off the C-states would keep all the cores---even the unused ones---at a high frequency usage.
+On the other hand, switching off the C-states would keep all the cores---even the unused ones---at a high-frequency usage.
 This highly reduces the variation, but causes up to 50\,\% of extra energy consumption in this test ($Mean_{C-states-off}=11,665 mJ$,$Mean_{C-states-on}=7,641 mJ$).
 
-At a 100\,\% workload, disabling the C-states seems to have no effect on the total energy consumption nor its variation.
-In fact, all the cores are used at 100\,\% and the C-states module would have no effect, as the cores are not idle.
+At a 100\,\% workload, disabling the C-states seems to not affect the total energy consumption nor its variation.
+All the cores are used at 100\,\% and the C-states module would have no effect, as the cores are not idle.
 The same reason would apply for the 50\,\% load, as the hyper-threading is active on all cores, thus causing the usage of most of them.
 For single process workloads, disabling the C-states causes the process to consume 50\,\% more energy as reported in Figure~\ref{fig:c-states}, but reduces the variation by 5~times compared to the \textsf{C-states on} mode.
-This leads to mainly two questions: Can a process pinning method reduce/increase the energy variation? And, how does the energy consumption variation evolve at different PU usage level?
+This leads to mainly two questions: Can a process pinning method reduce/increase the energy variation? And, how does the energy consumption variation evolve at different PU usage levels?
 
@@
 The second strategy (\textsf{S2}) splits the workload on the two sockets so each CPU will handle 50\,\% of the load.
-In this strategy, we only use the core PU and not the hyper-threads PU, so every process would not share his core usage (all the cores are being used).
-The third strategy (\textsf{S3}) consists also on splitting the workload between the two sockets, but considering the usage of the hyper-threads on each core---\emph{i.e.}, half of the cores are being used over the two CPU.
+In this strategy, we only use the core PU and not the hyper-threads PU, so every process would not share its core usage (all the cores are being used).
+The third strategy (\textsf{S3}) consists also of splitting the workload between the two sockets, but considering the usage of the hyper-threads on each core---\emph{i.e.}, half of the cores are being used over the two CPU.
 Figure~\ref{fig:cores-pinning} reports on the energy consumption of the three strategies when running the benchmark \texttt{CG} on the cluster \textsf{Dahu}.
 We can notice the big difference between these three execution modes that we obtained only by changing the PU pinning method (that we acknowledged with more than 100 additional runs over more than 30 machines and with the benchmarks \texttt{LU} and \texttt{EP}).
-For example, \textsf{S2} is the least power consuming strategy.
+For example, \textsf{S2} is the least power-consuming strategy.
 We argue that the reason is related to the isolation of every process on a single physical core, reducing the context switch operations.
-In the first and third strategy, $32$ processes are being scheduled on $16$ physical cores using the hyper-threads PU, which will introduce more context switching, and thus more energy consumption.
+In the first and third strategies, $32$ processes are being scheduled on $16$ physical cores using the hyper-threads PU, which will introduce more context switching, and thus more energy consumption.
 
@@
 The reason for the \textsf{S1} low energy consumption is that one of the two sockets is idle and will likely be in a lower power P-state, even with the disabled C-states.
-The \textsf{S2} case is also low energy consuming because by distributing the threads across all the cores, it completes the task faster than in the other cases.
+The \textsf{S2} case reports on low energy consumption because by distributing the threads across all the cores, it completes the task faster than in the other cases.
 Hence, it consumes less energy.
-The \textsf{S3} is a high consuming strategy because both sockets are being used, but only half the cores are active.
+The \textsf{S3} is a highly consuming strategy because both sockets are being used, but only half the cores are active.
 This means that we pay the energy cost for both sockets being operational and for the experiments taking longer to run because of the recurrent context switching.
@@
 Table~\ref{table:corespinning} reports on additional aggregated results for the STD comparison on four other nodes of the cluster \textsf{Dahu} at 50\,\%, with the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
-In fact, the CPU usage strategy \textsf{S1} is by far the experimentation mode that gave the least variation.
+The CPU usage strategy \textsf{S1} is by far the experimentation mode that gave the least variation.
 The STD is almost 5 times better than the strategy \textsf{S2}, but is up to 10\,\% more energy consuming ($Mean_{S1}=4469 mJ$, $Mean_{S2}=4016 mJ$).
@@
 This variation is 4~times lower when running 4 processes and almost equal to the \textsf{C-states on} mode when running 8 processes.
-In fact, running more processes implies to use more CPU cores, which reduces the idle cores count, so the cores will more likely stay at a higher consumption state even if the C-states mechanism is on.
+In fact, running more processes implies using more CPU cores, which reduces the idle cores count, so the cores will more likely stay at a higher consumption state even if the C-states mechanism is on.
 
-In our case, using 4 PU reduces the variation by 4~times and consumes almost the same energy as keeping the C-states mechanism on ($Mean_{C-states-on}=7,048 mJ$, $Mean_{C-states-off}=7,119 mJ$).
+In our case, using 4 PU reduces the variation by 4 times and consumes almost the same energy as keeping the C-states mechanism on ($Mean_{C-states-on}=7,048 mJ$, $Mean_{C-states-off}=7,119 mJ$).
 This case would be the closest to reality as we do not want to increase the energy consumption while reducing the variation, but using a lower number of PU still results in less variation, even if it increases the overall energy consumption.
@@
 Thus, disabling the C-states is very important to stabilize the measurements in some cases when the variation matters the most.
-Comparing the energy consumptions of two algorithms or two versions of a software systems is an example of use case benefiting from this recommendation.
+Comparing the energy consumption of two algorithms or two versions of a software system is an example of a use case benefiting from this recommendation.
 
@@
 The Turbo Boost---also known as \emph{Dynamic Overclocking}---is a feature that has been incorporated in Intel CPU since the Sandy~Bridge micro-architecture, and is now widely available on all of the Core\,i5, Core\,i7, Core\,i9 and Xeon series.
-It automatically raises some of the CPU cores operating frequency for short periods of time, and thus boost performances under specific constraints.
+It automatically raises some of the CPU cores operating frequency for short periods, and thus boosts performances under specific constraints.
 When demanding tasks are running, the operating system decides on using the highest performance state of the processor.
 
-Disabling or enabling the Turbo~Boost has a direct impact on the CPU frequency behavior, as enabling it allows the CPU to reach higher frequencies in order to execute some tasks for a short period of time.
+Disabling or enabling the Turbo~Boost has a direct impact on the CPU frequency behavior, as enabling it allows the CPU to reach higher frequencies to execute some tasks for a short period.
 However, its usage does not have a trivial impact on the energy variation.
@@
 Table~\ref{table:turboboost} compares the STD when enabling/disabling the Turbo~Boost, where the columns are a combination of workload and benchmark.
-In fact, we only got some minor measurements differences when switching on and off the Turbo~Boost, and where in favor or against the usage of the Turbo~Boost while repeating tests, considering multiple nodes and benchmarks.
+We only got some minor measurement differences when switching on and off the Turbo~Boost, and were in favor or against the usage of the Turbo~Boost while repeating tests, considering multiple nodes and benchmarks.
 This behavior is mainly related to the \emph{thermal design power} (TDP), especially at high workloads executions.
@@
 In this case, the Turbo~Boost cannot offer more power to the CPU because of the CPU thermal restrictions.
-At lower workloads, the tests we conducted proved that the Turbo~Boost is not one of the main reasons of the energy variation.
-In fact, the variation difference is barely noticeable when disabling the Turbo~Boost, which cannot be considered as a result regarding the OS activity and the measurement error margin.
-We cannot affirm that the Turbo~Boost does not have an impact on all the CPU, as we only tested on two recent Xeon CPU (clusters \textsf{Chetemi} and \textsf{Dahu}).
+At lower workloads, the tests we conducted proved that the Turbo~Boost is not one of the main reasons for the energy variation.
+The variation difference is barely noticeable when disabling the Turbo~Boost, which cannot be considered as a result of the OS activity and the measurement error margin.
+We cannot affirm that the Turbo~Boost does not have an impact on all the CPUs, as we only tested on two recent Xeon CPUs (clusters \textsf{Chetemi} and \textsf{Dahu}).
 We confirmed our experiments on these machines 100 times at 5\,\%, 25\,\%, 50\,\% and 100\,\% workloads.
@@
 It has been designed to ease the execution of most tasks with multitasking and resource sharing.
-In some delicate tests and measurements, the OS activity and processes can cause a significant overhead and therefore a potential threat to the validity. % experiments validity
-The purpose behind this experiment is to determine if the sampled consumption can be reliably related to the tested application, especially for low-workload applications where CPU resources are not heavily used by the application.
+In some delicate tests and measurements, the OS activity and processes can cause significant overhead and therefore a potential threat to the validity. % experiments validity
+The purpose of this experiment is to determine if the sampled consumption can be reliably related to the tested application, especially for low-workload applications where CPU resources are not heavily used by the application.
 
-The first way to do is to evaluate the OS idle activity consumption, and to compare it to a low workload running job.
+The first way to do this is to evaluate the OS idle activity consumption and to compare it to a low workload running job.
 Therefore, we ran 100 iterations of a single process benchmark \textsf{EP}, \textsf{LU} and \textsf{CG} on multiple nodes from the cluster \textsf{Dahu}, and compared the energy behavior of the node with its idle state on the same duration.
@@
 In fact, for the three nodes, randomly picked from the cluster \textsf{Dahu}, the idle variation is way more important than when a test was running, even if it is a single process test on a $32$-cores node.
-This result shows that OS idle consumption varies widely, due to the lack of activity and the different CPU frequencies states, but it does not mean that this variation is the main responsible for the overall energy variation.
-The OS behaves differently when a job is running, even if the amount of available cores is more than enough for the OS to keep his idle behavior when running a single process.
+This result shows that OS idle consumption varies widely, due to the lack of activity and the different CPU frequency states, but it does not mean that this variation is the main responsible for the overall energy variation.
+The OS behaves differently when a job is running, even if the amount of available cores is more than enough for the OS to keep its idle behavior when running a single process.
 
@@
 Inspecting the OS idle energy variation is not sufficient to relate the energy variation to the active job.
-In fact, the OS can behave differently regarding the resource usage when running a task.
+The OS can behave differently regarding resource usage when running a task.
 To evaluate the OS and the job energy consumption separately, we used the \textsc{PowerAPI} toolkit.
@@
 Thus, it is possible to isolate the job energy consumption instead of the global energy consumption delivered by RAPL.
-To do so, we ran tests with a single process workload on the cluster \textsf{Dahu}, and used the \textsc{PowerAPI} toolkit to measure the energy consumption.
+To do so, we ran tests with a single process workload on the cluster \textsf{Dahu} and used the \textsc{PowerAPI} toolkit to measure the energy consumption.
 Then, we compared the job energy consumption to the global RAPL data.
@@
 However, this does not completely exclude the OS activity, especially if the jobs have tight interaction with the OS through the signals and system calls.
-This brings a new question on whether applying extra-tuning on a minimal OS would reduce the variation?
+This brings a new question on whether applying extra-tuning on a minimal OS would reduce the variation.
 As well as what is the effect of the Meltdown security patch---that is known to be causing some performance degradation~\cite{Kocher2018spectre,Lipp2018meltdown}---on the energy variation?
@@
 \paragraph{OS Tuning}
-An OS is a pack of running processes and services that might or not be required its execution.
-In fact, even using a minimal version of a Debian Linux, we could list many OS running services and process that could be disabled/stopped without impacting the test execution.
+An OS is a pack of running processes and services that might or not be required for its execution.
+Even using a minimal version of Debian Linux, we could list many OS running services and processes that could be disabled/stopped without impacting the test execution.
 This extra-tuning may not be the same depending on the nature of the test or the OS.
 Thus, we conducted a test with a deeply-tuned OS version.
-We disabled all the services/processes that are not essential to the OS/test running, including the OS networking interfaces and logging modules, and we only kept the strict minimum required to the experiment's execution.
+We disabled all the services/processes that are not essential to the OS/test running, including the OS networking interfaces and logging modules, and we only kept the strict minimum required for the experiment's execution.
 Table~\ref{table:osmin} reports on the aggregated results for running single process measurements with the benchmarks \textsf{CG}, \textsf{LU} and \textsf{EP}, on three servers of the cluster \textsf{Dahu}, before and after tuning the OS.
@@
 We notice that the energy variation varies less than 10\,\% after the extra-tuning.
-We argue that this variation is not substantial, as it is not stable from a node to another.
-Moreover, 10\,\% of variation is not a representative difference, due to many factors that can affect it as the CPU temperature or the measurement errors.
+We argue that this variation is not substantial, as it is not stable from one node to another.
+Moreover, 10\,\% of the variation is not a representative difference, due to many factors that can affect it as the CPU temperature or the measurement errors.
 
@@
 \paragraph{Speculative Executions}
-Meltdown and Spectre are two of the most famous hardware vulnerabilities discovered in 2018, and exploiting them allows a malicious process to access others processes data that is supposed to be private~\cite{Kocher2018spectre,Lipp2018meltdown}.
-They both exploit the speculative execution technique where a process anticipates some upcoming tasks, which are not guaranteed to be executed, when extra resources are available, and revert those changes if not.
+Meltdown and Spectre are two of the most famous hardware vulnerabilities discovered in 2018, and exploiting them allows a malicious process to access others processes' data that is supposed to be private~\cite{Kocher2018spectre,Lipp2018meltdown}.
+They both exploit the speculative execution technique where a process anticipates some upcoming tasks, which are not guaranteed to be executed when extra resources are available, and revert those changes if not.
 Some OS-level patches had been applied to prevent/reduce the criticality of these vulnerabilities.
-On the Linux kernel, the patch has been automatically applied since the version 4.14.12.
+On the Linux kernel, the patch has been automatically applied since version 4.14.12.
 It mitigates the risk by isolating the kernel and the user space and preventing the mapping of most of the kernel memory in the user space.
@@
 Nikolay~\emph{et~al.} have studied in~\cite{DBLP:journals/corr/abs-1801-04329} the impact of patching the OS on the performance.
-The results showed that the overall performance decrease is around 2--3\,\% for most of the benchmarks and real-world applications, only some specific functions can meet a high performance decrease.
+The results showed that the overall performance decrease is around 2--3\,\% for most of the benchmarks and real-world applications, only some specific functions can meet a high-performance decrease.
 In our study, we are interested in the applied patch's impact on the energy variation, as the performance decrease could mean an energy consumption increase.
@@
 These results help us to conclude that the security patch's effect on the energy variation is not substantial and can be absorbed through the error margin for the tested benchmarks.
-In fact, the best case to consider is the benchmark \textsf{LU} where the energy variation is less than 10\,\% when we disable the security patch, but this difference is still moderate.
-The little performance difference discussed in~\cite{Kocher2018spectre,Lipp2018meltdown} may only be responsible of a small variation, which will be absorbed through the measurement tools and external noise error margin in most cases.
+The best case to consider is the benchmark \textsf{LU} where the energy variation is less than 10\,\% when we disable the security patch, but this difference is still moderate.
+The little performance difference discussed in~\cite{Kocher2018spectre,Lipp2018meltdown} may only be responsible for a small variation, which will be absorbed through the measurement tools and external noise error margin in most cases.
 
@@
 Intel microprocessors have noticeably evolved during these last 20 years.
-Most of the new CPU come with new enhancements to the chip density, the maximum Frequency or some optimization features like the C-states or the Turbo~Boost.
-This active evolution caused that different generations of CPU can handle a task differently.
-The aim of this expriment is not to justify the evolution of the variation across CPU versions/generations, but to observe if the user can choose the best node to execute her experiments.
+Most of the new CPUs come with new enhancements to the chip density, the maximum Frequency or some optimization features like the C-states or the Turbo~Boost.
+This active evolution caused different generations of CPUs can handle a task differently.
+The aim of this experiment is not to justify the evolution of the variation across CPU versions/generations, but to observe if the user can choose the best node to execute her experiments.
 Previous papers have discussed the evolution of the energy consumption variation across CPU generations and concluded that the variation is getting higher with the latest CPU generations~\cite{wang_experimental_nodate,marathe_empirical_2017_m}, which makes measurements stability even worse.
-In this experiment, we therefore compare four different generations of CPU with the aim to evaluate the energy variation for each CPU and its correlation with the generation.
-Table~\ref{table:cpus} indicates the characteristics of each of the tested CPU.
+In this experiment, we, therefore, compare four different generations of CPU to evaluate the energy variation for each CPU and its correlation with the generation.
+Table~\ref{table:cpus} indicates the characteristics of each of the tested CPUs.
 
@@
 While the Skylake CPU of the cluster \textsf{Dahu} cause often more energy variation than \textsf{Chetemi} and the Ecotype Broadwell CPU.
-We argue that the hypothesis "\emph{the energy consumption on newer CPU varies more}" could be true or not depending on the compared generations, but most importantly, the chips energy behaviors.
-On the other hand, our experiments showed the lowest energy variation when using the \textsf{Ecotype CPU}, these CPU are not the oldest nor the latest, but are tagged with "\texttt{L}" for their low power/TDP.
+We argue that the hypothesis "\emph{the energy consumption on newer CPU varies more}" could be true or not depending on the compared generations, but most importantly, the chip's energy behaviors.
+On the other hand, our experiments showed the lowest energy variation when using the \textsf{Ecotype CPU}, these CPUs are not the oldest nor the latest, but are tagged with "\texttt{L}" for their low power/TDP.
 This result rises another hypothesis when considering CPU choice, which implies selecting the CPU with a low TDP.
@@
 
-Figure~\ref{fig:cpugen} is an illustration of the aggregated STD density of more than $5,000$-random values sets taken from all the conducted experiments.
-This shows that the cluster \textsf{Paranoia} reports the worst variation in most cases, and that \textsf{Ecotype} is the best cluster to consider to get the least variations, as it has a higher density for small variation values.
+Figure~\ref{fig:cpugen} is an illustration of the aggregated STD density of more than $5,000$-random value sets taken from all the conducted experiments.
+This shows that the cluster \textsf{Paranoia} reports the worst variation in most cases and that \textsf{Ecotype} is the best cluster to consider to get the least variations, as it has a higher density for small variation values.
 
@@
 \subsection{Experimental Guidelines}
-To summarize our experiments, we provide some experimental guidelines in Table~\ref{table:guidelines}, based on the multiple experiments and analysis we did.
+To summarize our experiments, we provide some experimental guidelines in Table~\ref{table:guidelines}, based on the multiple experiments and analyses we did.
 These guidelines constitute a set of minimal requirements or best practices, depending on the workload and the criticality of the energy measurement precision.
-It therefore intends to help practitioners in taming the energy variation on the selected CPU, and conduct the experiments with the least variations.
+It, therefore, intends to help practitioners in taming the energy variation on the selected CPU, and conduct the experiments with the least variations.
 
@@
         \hline
-        Execute all the similar and comparable experiments on a same machine.
+        Execute all the similar and comparable experiments on the same machine.
 Identical machines can exhibit many differences regarding their energy behavior             & Any           & Up to $1.3\times$ \\
@@
 
-Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and its variation reduction at low workloads.
+Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and their variation reduction at low workloads.
 However, it also lists some new factors that we identified along the analysis we conducted in Section, such as the results related to the OS or the reboot mode.
-Some of the guidelines are more useful/efficient for specific workloads, as showed in our experiments.
+Some of the guidelines are more useful/efficient for specific workloads, as shown in our experiments.
 Thus, qualifying the workload before conducting the experiments can help in choosing the proper guidelines to apply.
@@
 
-In order to validate the accuracy of our guidelines among a varied set of benchmarks on one hand, and their effect on the variation between identical machines on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
-Half of these experiments has been performed at a 50\,\% workload and the other half on single process jobs.
+To validate the accuracy of our guidelines among a varied set of benchmarks on one hand, and their effect on the variation between identical machines, on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
+Half of these experiments have been performed at a 50\,\% workload and the other half on single process jobs.
 The choice of these two workloads is related to the optimization guidelines that are mainly effective at low and medium workloads.
-We note that we used the cluster \textsf{Dahu} over \textsf{Ecotype} to highlight the guidelines effect on the nodes where the variation is susceptible to be higher.
+We note that we used the cluster \textsf{Dahu} over \textsf{Ecotype} to highlight the effect of the guidelines on the nodes where the variation is susceptible to be higher.
 
@@
 For medium workloads, the STD, and thus variation, is up to 100\,\% better for the benchmark \textsf{CG}, 20--150\,\% better for the \texttt{pbzip2} application and up to 100\% for \textsc{Stress-NG}.
-We note that the optimized version consumes less energy thanks to an appropriate core pinning method.
+We note that the optimized version consumes fewer energy thanks to an appropriate core pinning method.
 
@@
 However, taming the intra-node variation is a good strategy to identify more relevant mediums and medians, and then perform accurate comparisons between the nodes variation.
-Even though, using the same node is always better, to avoid the extra inter-nodes variation and thus improve the stability of measurements.
+Even though using the same node is always better, it avoids the extra inter-nodes variation and thus improves the stability of measurements.
 
@@
 A number of issues affect the validity of our work.
-For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPU, but still adds an important overhead if we adopt a sampling at high frequency.
+For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPUs, but still adds an important overhead if we adopt a sampling at high frequency.
 The other fine-grained tool we used for measurements is \textsc{PowerAPI}.
-It allows to measure the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
+It allows measuring the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
 The usage of \textsc{PowerAPI} adds an error margin because of the power model built over RAPL.
-The RAPL tool mainly measures the CPU and DRAM energy consumption.
+The RAPL tool mainly measures CPU and DRAM energy consumption.
 However, even running CPU/RAM intensive benchmarks would keep a degree on uncertainty concerning the hard disk and networking energy consumption.
@@
 
-The Intel CPU chip manufacturing process and the materials micro-heterogeneity is one of the biggest issues, as we cannot track or justify some of the energy variation between identical CPU or cores.
+The Intel CPU chip manufacturing process and the material's micro-heterogeneity are one of the biggest issues, as we cannot track or justify some of the energy variations between identical CPUs or cores.
 These CPU/cores might handle frequencies and temperature differently and behave consequently.
@@
 
-\subsection{Conclusion}
-In order to increase the \textbf{accuracy} in comparative experiments, we conducted an empirical study of controllable factors that may increase the energy variations on platforms with some of the latest CPUs and for several benchmarks.
+\subsection{Summary}
+To increase the \textbf{accuracy} in comparative experiments, we conducted an empirical study of controllable factors that may increase the energy variations on platforms with some of the latest CPUs and for several benchmarks.
 This study is not intended to nullify the variability of the CPU, as some of this variability is related to the chip manufacturing process and its thermal behavior.
-Instead, it aims to tame and mitigate this variability along controlled experiments.
-In this study, We investigated some previously discussed aspects on some recent CPUs, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to increase this accuracy for energy related experiments.
+Instead, it aims to tame and mitigate this variability through controlled experiments.
+In this study, We investigated some previously discussed aspects of some recent CPUs, considered new factors that have not been deeply analyzed to the best of our knowledge, and constituted a set of guidelines to increase this accuracy for energy-related experiments.
 Some of these factors, like the \emph{C-states} usage, can reduce the energy variation up to 500\,\% at low workloads, while choosing the wrong cores/PU strategy can cause up to $30\times$ more variability.

### 2022-09-28 chapters/optimization/python-dir/python.tex

@@
 Machine learning is becoming an integral part of our daily lives, growing more potent and energy-hungry each year.
-As machine learning can have a significant impact on climate change, it is vital to investigate mitigation techniques.
+As machine learning can significantly impact climate change, it is vital to investigate mitigation techniques.
 

### 2022-09-28 chapters/programming_languages/rpc/rpc.tex

@@
 \subsection{Introduction}
-With the success of Internet and the emergence of cloud technologies, many communication protocols compete to take the lead.
-In particular, most of the software architectures are now based on multi-services and micro-service technologies.
-And, to cope with the higher versatility of developers, multiple companies choose to open their services to different programming languages.
-Interestingly, this approach intends to take advantage of each programming language to satisfy specific requirements.
+With the success of the Internet and the emergence of cloud technologies, many communication protocols compete to take the lead.
+In particular, most software architectures are now based on multi-services and micro-service technologies.
+And, to cope with the increased versatility of developers, multiple companies choose to open their services to different programming languages.
+This approach is interesting because it wants to use the best parts of each programming language to meet different needs.
 However, the challenge nowadays is to make the bridge between those platforms.
-We have many initiatives, such as OpenAPI that try to create a taxonomy for RESTful APIs, while other approaches implement all the different interfaces of the protocol by themselves, such as RPC.
+We have many initiatives, such as OpenAPI, that try to create a taxonomy for RESTful APIs, while other approaches implement all the different interfaces of the protocol by themselves, such as RPC.
 
@@
 However, we do not observe the same behavior when it comes to efficiency.
-Unlike the other interpreted programming languages, PHP performances could be compared to the compiled ones, such as CPP or GO, and even better to some others, such as Swift.
-JVM-based languages tend to have better performances than the interpreted ones.
-Furthermore, OpenJDK has shown more efficiency than GraalVM [].
+Unlike the other interpreted programming languages, PHP's performance can be compared to the compiled ones, such as CPP or GO, and even better to some others, such as Swift.
+JVM-based languages tend to have better performance than interpreted ones.
+Furthermore, OpenJDK has shown more efficiency than GraalVM.
 Overall, we can have 3 groups when it comes the cost of each request:
@@
 \begin{enumerate}
-    \item Drop in performances without an increased power, such as .Net core, Java micronaut, Crystal, and Dart.
+    \item drop in performances without an increased power source, such as Net core, Java Micronaut, Crystal, and Dart.
     In this case, the server keeps using the same resources, and sometimes less, because it takes more time to handle the fewer requests.
     This class of languages tends to be the most energy-consuming when it comes to the cost per request;
-    \item Increase in power without affecting the performances, such as Go or .Net.
-          The energy consumption of a single request, is affected slightly but still increases;
-    \item Increase in power and drop in performances.
-          Despite the increase of power consumption, the server becomes slightly slower, which increases the energy cost per request.
+    \item increase in power without affecting the performance, such as Go or.Net.
+    The energy consumption of a single request is affected slightly but still increases.
+    \item Increase in power and a drop in performance.
+    Despite the increase in power consumption, the server becomes slightly slower, which increases the energy cost per request.
     This cost is still better than the first case, which concludes that the servers in the first category are on the verge of breaking.
@@
 
-We can mention the case of Elixir that keeps scaling despite the lack of performances compared to other compiled languages (Go, CPP).
+We can mention the case of Elixir that keeps scaling despite the lack of performance compared to other compiled languages (Go, CPP)..
 
 \subsubsection{Broken state mode }
-Only four of the 25 configurations could parse the 10 MB files, and only 1 from those could achieve a 76\% acceptance rate which is Elixir, the other 3 had less than 3\% success rate (Rust, Swift and Dart).
+Only four of the 25 configurations could parse the 10 MB files, and only one of those could achieve a 76\% acceptance rate, which is Elixir.
+The other 3 had less than a 3\% success rate (Rust, Swift, and Dart).
 The rest could be divided into two categories:
@@
 
-In each programming category, we observe both energy-efficient and energy-greedy behaviours.
-Therefore, we conclude that it depends more on the implementation of the library itself, rather than the category of the programming language.
-Scala and Kotlin are an excellent example to support this hypothesis, as both of them run on the same virtual machine as Java (OpenJDK 16.1).
+In each programming category, we observe both energy-efficient and energy-greedy behaviors.
+Therefore, we conclude that it depends more on the implementation of the library itself than on the category of the programming language.
+Scala and Kotlin are excellent examples to support this hypothesis, as both of them run on the same virtual machine as Java (OpenJDK 16.1).
 Yet, their average power is 130\% higher than the Java implementation.
@@
 \paragraph{Stressed mode}
-Although the same classes remained the same, not all the languages had the same evolution and here we can clearly observe a correlation with the category of the programming language rather than the implementation itself.
-We can clearly highlight that VM-based languages have a significant increase (double) in the average power consumption after they receive more than 100 concurrent clients.
+Although the same classes remained the same, not all the languages had the same evolution, and here we can clearly observe a correlation with the category of the programming language rather than the implementation itself.
+We can easily point out that the average amount of power used by VM-based languages doubles when they have more than 100 clients at the same time.
 Except PHP, all the interpreted and compiled languages preserved their energetic behaviour.
 Our hypothesis points to the JIT, since it compiles the code and makes it run faster, hence stressing the CPU.
-A interesting behaviour has been noticed for the GraalVM: the decrease of energy consumption when increasing the number of the clients.
-This is related to the drop of the performances, which was probably due to the bottleneck situation where the GraalVM could not handle more than 100 clients simultaneously.
+An interesting behavior has been noticed for the GraalVM: the decrease in energy consumption when increasing the number of clients.
+This is related to the performance drop, which was probably due to the bottleneck situation where the GraalVM could not handle more than 100 clients simultaneously.
 
 \subsubsection{Performance Behaviour}
-In this section, we study only the number of requests per seconds processed by the server without looking at its energy.
+In this section, we study only the number of requests per second processed by the server without looking at its energy.
 We consider three observable variables:
@@
 \paragraph{Satisfaction ratio}
-Most of the considered frameworks satisfy all the requests, by either reducing the number of requests per second or by increasing the processing time.
+Most of the considered frameworks satisfy all the requests by either reducing the number of requests per second or increasing the processing time.
 However, there are some frameworks that have chosen a different approach, such as Dart or Scala, where the choice was to keep a certain limit of latency even if not all the requests are answered.
-Furthermore, we tend to observe this behaviour among other frameworks, such as Python or Asynchronous NodeJs, when the number of the client exceeds 800.
+Furthermore, we tend to observe this behavior among other frameworks, such as Python or Asynchronous NodeJS, when the number of clients exceeds 800.
 
@@
 \end{figure}
-Most of the servers hit their RPS limit after 5 clients and 100 clients for vm based servers, and after this the number keeps constant, which will decrease the average RPS per client.
-.Net server is the most performant, followed by Java and Go, while Python and Ruby are the least performant.
+Most of the servers hit their RPS limit after 5 clients and 100 clients for VM-based servers, and after this the number keeps constant, which will decrease the average RPS per client.
+. The Net server is the most performant, followed by Java and Go, while Python and Ruby are the least performant.
 
 \paragraph{Tail Latency}
-The increase in the number of requests per second, does not necessarily mean a lower latency.
-As one can notice in Fgirue~\ref{fig:tail99_clients}, until the 1000 clients, Go provides the least latency beside .Net.
-GraalVM provides the highest latency, on average.
-However, Dart tends to become slower when we increase the number of clients, until we pass the 600 simultaneous clients, and there it changes its behaviour, instead of satisfying most the requests it notify the clients directly that the server is saturated, hence a drop in satisfaction ratio, and an amelioration for the average latency.
+Even if the number of requests per second goes up, that doesn't mean the latency will go down.
+As one can notice in figure~\ref{fig:tail99_clients}, until the 1000 clients, Go provides the least latency besides.Net.
+GraalVM provides the highest average latency.
+However, Dart tends to become slower when we increase the number of clients, until we pass the 600 simultaneous clients, and then it changes its behavior.
+Instead of satisfying most of the requests, it notifies the clients directly that the server is saturated, resulting in a drop in satisfaction ratio and an amelioration in average latency.
 
@@
 Now, after we made a separation between the energy and the performances, we have seen that most of the performance servers tend to be energy-greedy, so we propose to investigate this trade-off between energy and performances.
-To do so, we report on an average cost of a single request, in Joules.
+To do so, we report the average cost of a single request in joules.
 Except GraalVM when the cost of the a single request increases with when we add more clients, all the frameworks report on a constant cost, Java, .net and go are the most energy efficient, while Python and Ruby may cost up to 10x more.
-Therefore, we conclude that the number of clients does not impact the energy significantly.
-Then, we study how the payload size of the requests impact the energy consumption of the framework.
+Therefore, we conclude that the number of clients does not significantly impact the energy consumption.
+Then, we study how the payload size of the requests impacts the energy consumption of the framework.
 

### 2022-09-28 chapters/programming_languages/rpc/rpc.tex
update : rpc 1st part

@@
 In particular, most software architectures are now based on multi-services and micro-service technologies.
-And, to cope with the increased versatility of developers, multiple companies choose to open their services to different programming languages.
+Moreover, to cope with the increased versatility of developers, multiple companies choose to open their services to different programming languages.
 This approach is interesting because it wants to use the best parts of each programming language to meet different needs.
-However, the challenge nowadays is to make the bridge between those platforms.
-We have many initiatives, such as OpenAPI, that try to create a taxonomy for RESTful APIs, while other approaches implement all the different interfaces of the protocol by themselves, such as RPC.
+However, the challenge nowadays is to make the bridge between these platforms.
+We have many initiatives, such as OpenAPI, that try to create a taxonomy for RESTful APIs, while other approaches implement all the different interfaces of the protocol by themselves, such as The Remote Procedure Call (aka RPC) protocol.
+In this Section
 
-\subsection{Research Questions}
+\subsubsection{Research Questions}
 In this section, we first explore the ease of implementation of this protocol, and then we will try to answer the following research questions:
 \begin{description}
-    \item[\textsc{RQ}~1:] How do RPC implementations consume with regard to the size of the incoming request?
-    \item[\textsc{RQ}~2:] How do RPC implementations consume with regard to the number of concurrent clients?
+    \item[\textsc{RQ}~1:] How do RPC implementations consume energy concerning the size of the incoming request?
+    \item[\textsc{RQ}~2:] How do RPC implementations consume energy regarding the number of concurrent clients?
 \end{description}
@@
 \subsection{Experimental Protocol}
+To answer these questions, we will define an interface using the Proto2~\footnote{\url{https://developers.google.com/protocol-buffers}}language, and then we will implement it in multiple servers, each one using a different programming language.
+We will then compare the energy consumption of each implementation using a custom version of \texttt{GHZ}~\footnote{\url{https://github.com/chakib-belgaid/energy_ghz}}, a tool that allows us to stress test our servers.
+This fork of \texttt{GHZ} allows us to measure the server's energy consumption during the test's execution and synchronize it with other performance statistics such as Tail latency and the number of requests per second the server can reach.
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=.8\linewidth]{imgs/rpcprotocol}
+    \end{center}
+    \caption{Experimental software architecture.}\label{fig:rpcprotocol}
+\end{figure}
+Figure~\ref{fig:rpcprotocol} show the experimental protocol. We will use different messages inserted in the file \texttt{helloworld.proto}. 
+The rest of this part will be dedicated to explaining each aspect of the experiment.
 \subsubsection{Measurement Context}
@@
 All the experiments are run on the cluster \textsf{paravance} of the G5K platform.
-This cluster is composed of 72 identical machines, each one is equipped with 2 Intel Xeon E5-2630 V3, with 128\,GB of RAM.
+This cluster comprises 72 identical machines; each is equipped with 2 Intel Xeon E5-2630 V3, with 128\,GB of RAM.
 For more accuracy, our SUT (\emph{System Under Test}) runs a minimal version of Debian\,9 (4.9.0 kernel version), which enforces the core processes required for our experiment.
 Furthermore, we used Docker containers technology for reproducibility of the experiments and the isolation of the servers.
-
 \paragraph{Client \& server environments.}
@@
 \paragraph{Energy measurements.}
-To report on the energy consumption, we used HwPC sensor [TOCITE], which is based on Intel RAPL technology, one of the most accurate tools to measure the energy consumption of the CPU and DRAM [TOCITE].
-For better accuracy, we ran the HwPC sensor with a frequency of 1~Hz, and we used the same machine for all the experiments to reduce the variability [TOCITE].
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=.8\linewidth]{imgs/rpcprotocol}
-    \end{center}
-    \caption{Experimental software architecture.}\label{fig:rpcprotocol}
-\end{figure}
+To report on energy consumption, we used the \texttt{HWPC} sensor\footnote{\url{https://github.com/powerapi-ng/hwpc-sensor}}, based on Intel \texttt{RAPL} technology, one of the most accurate tools to measure the CPU's and DRAM's energy consumption~\cite{hackenberg2013power,hackenberg2015energy}.
+For better accuracy, we ran the HwPC sensor with a frequency of 1~Hz, and we used the same machine for all the experiments to reduce the variability~\cite{ilsche_power_2015}.
 
 \paragraph{Performances.}
-For better accuracy and more details, We use an updated version of the open source RPC benchmarking tool, named GHZ (\url{https://ghz.sh/}).
+For better accuracy and more details, We updated the open source RPC benchmarking tool GHZ (\url{https://ghz.sh/}).
 The modified version allows us to monitor the average power for each request from both the server and the client sides.
@@
 The purpose of the experiment is to analyze the behavior of different GRPC implementations.
-Therefore, we have two kind of worlkoads: \emph{number of clients} is the number of concurrent clients that we want to benchmark (handled by the GHZ client) and the \emph{payload} reflects the size of each request (varies from 50 Bytes up to 10 MB).
+Therefore, we have two kinds of workloads: \emph{number of clients} is the number of concurrent clients that we want to benchmark (handled by the GHZ client), and the \emph{payload} reflects the size of each request (varies from 50 Bytes up to 10 MB).
 The client consumes the protocol description found in the file \texttt{helloworld.proto} to generate an implementation for the message and then forks multiple instances that send the same request to the server.
@@
 Each server uses 16 cores and is limited to 512\,MB of RAM.
-Each implementation is packaged as a Docker image, which will allow us to add new implementations easily.
+Each implementation is packaged as a Docker image, allowing us to add new implementations easily.
 
@@
 Adding a new RPC \textbf{candidate} can be achieved by creating a new Docker image and putting it in a new folder named after the language.
-As for the \textbf{workload}, it can be extended easily by adding new files in the folder \textbf{payload}.
+As for the \textbf{workload}, it can be extended easily by adding new files in the folder \textbf{payload} and by changing the number of clients in the benchmarking tool.
 
 \subsection{Results \& Findings}
-\paragraph{\textsc{RQ}~1: How do RPC implementations consume with regard to the size of the incoming request?}
+\paragraph{\textsc{RQ}~1: How do RPC implementations consume energy concerning the size of the incoming request?}
 

### 2022-09-28 chapters/optimization/python-dir/python.tex
update : rpc typos

@@
 Several models were developed.
-However, only two were used in the final trials, because they achieved 94\% accuracy with an acceptable training time.
+However, only two were used in the final trials because they achieved 94\% accuracy with an acceptable training time.
 David Page's \textsf{cifar10-fast}\footnote{\url{https://github.com/davidcpage/cifar10-fast}} and Woonhyuk Baek's torch skeleton\footnote{\url{https://github.com/wbaek/torchskeleton}} are reported here.
@@
 \subsubsection{Results \& Findings}
-As the model's accuracy increases, so does the energy required for the next accuracy increment.
+As the model's accuracy increases, so do the energy required for the next accuracy increment.
 Figure~\ref{fig:cum_energy_vs_accuracy} depicts how the curve steepens as training progresses.
@@
 However, the increase in accuracy is not linear, as can be observed.
-It took only 10 epochs to reach an accuracy of 84\%, but it required more than twice the number of epochs to add an axtra 6\% accuracy.
+It took only 10 epochs to reach an accuracy of 84\%, but it required more than twice the number of epochs to add an extra 6\% accuracy.
 This highlights the price one should pay to increase the model's accuracy.
@@
     \includegraphics[width=1.1\textwidth]{imgs/cumulative_energy_vs_accuracy}
-    \caption{Cumulative energy consumption vs accuracy.}
+    \caption{Cumulative energy consumption vs. accuracy.}
     \label{fig:cum_energy_vs_accuracy}
@@
 As one can see in Figures~\ref{fig:ephoch_power} and~\ref{fig:epoch_duration}, the average power consumption and the average duration of an epoch do not have a significant variation, as most of them last around 3.7 seconds.
-However, there is an intermediate correlation between these two values if we excluded the strategy based on one core and its hyper-thread.
-The reason why the last strategy exhibits more power---even when its duration is not optimal---is the context switches between the core and its hyper-thread.
+However, these two values have an intermediate correlation if we exclude the strategy based on one core and its hyper-thread.
+The last strategy exhibits more power- even when its duration is not optimal- because the context switches between the core and its hyper-thread.
 
@@
 % We have chosen Django as our web framework.
-Therefore, we chose to study if the choice of the database and the ORM impacts the energy consumption and the web application performance.
+Therefore, we chose to study if the choice of the database and the ORM impact energy consumption and web application performance.
 % \subsection{Experiment \& Results}

### 2022-09-28 chapters/programming_languages/programming_languages.tex
update : python chapter
update : rpc typos

@@
 \label{chapter:porgramming_langauges}
-% \section{Introduction}
-In this chapter, we study how the programming framework affects the energy the software consumes.
+
+% TODO : repphrase this
+
+after looking at the studies of django we wanted to expand our study to other programming languages and frameworks.
+the idea behind this study is to expand the work of perrira et al. from the microbenchmarking level to get closer to real world applications.
+first we will start with studying the RPC protocol since it offers a multi langauges interfaces. therfore we will reduce the developer expertise bias 
+after that we will enlarge this study to cover most of the popular webframeworks.
+in this study we will focus on the energy footprint of the webframeworks and the RPC protocol in the context of an unlimited time, therefore the factor time will have less impact unlike the microbenchmarking study.
+
+This chapter studies how the programming framework affects the energy the software consumes.
 We suggest starting with general micro-benchmarking and watching how each programming framework performs with the CPU and memory.
@@
 
-\section{Conclusion\note{missing}}
+\section{Conclusion \note{missing}}

### 2022-09-28 chapters/programming_languages/rpc/rpc.tex
update : rpc typos

@@
 
+% TODO : change php to the interpreted part 
 \subsection{Results \& Findings}
@@
 Therefore, our boundary condition is only the number of requests received by the server.
-For this experiment, we investigate 4 observable variables:
+For this experiment, we investigate four observable variables:
 \begin{enumerate}
     \item the \textsf{average power consumption} during the scenario indicates the overall behavior of the server when working for long durations,
-    \item the \textsf{tail latency} for the 99th percentile, which indicates how efficient is the server,
+    \item the \textsf{tail latency} for the 99th percentile, which indicates how efficient the server is,
     \item the \textsf{average number of requests per second}, which indicates the average number of clients that the server can handle,
-    \item the \textsf{average energy cost} of a single request: unlike the first indicator, this one shows how green is the implementation considering performance.
+    \item the \textsf{average energy cost} of a single request: unlike the first indicator, this one shows how green the implementation is considering performance.
 \end{enumerate}
@@
 \end{figure}
-Figure~\ref{fig:energy_cost_payload} depicts the overall behavior of each framework based on the size of request (payload).
+Figure~\ref{fig:energy_cost_payload} depicts the overall behavior of each framework based on the request size (payload).
 For each framework, we can distinguish three modes, and they all depend on the payload size:
 \begin{enumerate}
-    \item \textsf{Stress free} mode when the server has enough resources to satisfy the requests, as they require a memory less than a certain threshold (depends on the language and the platform),
+    \item \textsf{Stress-free} mode when the server has enough resources to satisfy the requests, as they require a memory less than a certain threshold (depends on the language and the platform),
     \item \textsf{Escalation} mode when the requests tend to be bigger, but the server can still manage to handle them, at the price of a change in the energy and performance behaviors,
@@
 
-\subsubsection{Stress Free Mode}
-In this mode, the compiled languages tend to consume fewer resources (average power).
+\subsubsection{Stress-Free Mode}
+The compiled languages consume fewer resources (average power) in this mode.
 JVM-based languages tend to consume more energy, especially Scala.
-However, we do not observe the same behavior when it comes to efficiency.
-Unlike the other interpreted programming languages, PHP's performance can be compared to the compiled ones, such as CPP or GO, and even better to some others, such as Swift.
+However, we do not observe the same behavior regarding efficiency.
+Unlike the other interpreted programming languages, PHP's performance can be compared to the compiled ones, such as CPP or GO, and even better to others, such as Swift.
 JVM-based languages tend to have better performance than interpreted ones.
 Furthermore, OpenJDK has shown more efficiency than GraalVM.
-Overall, we can have 3 groups when it comes the cost of each request:
+Overall, we can have three groups when it comes to the cost of each request:
 \begin{itemize}
@@
     \item drop in performances without an increased power source, such as Net core, Java Micronaut, Crystal, and Dart.
-    In this case, the server keeps using the same resources, and sometimes less, because it takes more time to handle the fewer requests.
+    In this case, the server uses the same resources, sometimes less, because it takes more time to handle fewer requests.
     This class of languages tends to be the most energy-consuming when it comes to the cost per request;
@@
 \begin{itemize}
-    \item \textsf{Timeout} where requests took too much time that the client canceled them, in this category we find most of dynamic codes, such as OpenJDK and Kotlin,
-    \item \textsf{Size of request} exceeded the maximum size when the implementation could not handle requests with large size, as observed with .Net, Go, .Net core, CPP, PHP, Scala, Nodejs, Ruby, Python.
+    \item \textsf{Timeout} where requests took too much time that the client canceled them, in this category, we find most of the dynamic codes, such as OpenJDK and Kotlin,
+    \item \textsf{Size of request} exceeded the maximum size when the implementation could not handle requests with a large size, as observed with .Net, Go, .Net core, CPP, PHP, Scala, Nodejs, Ruby, and Python.
 \end{itemize}
 
-[\textsc{RQ}~2:] How do RPC implementations react to the number of clients ?
+[\textsc{RQ}~2:] How do RPC implementations react to the number of clients?
 \begin{figure}[!hbt]
@@
 
-\subsubsection{Power behaviour}
+\subsubsection{Power behavior}
 
@@
     \item \textsf{Low number of clients} when the number of concurrent clients is below 100,
-    \item \textsf{Moderate to high number of clients} when the number of clients exceeds 100.
+    \item \textsf{Moderate to a high number of clients} when the number of clients exceeds 100.
 \end{itemize}
@@
 In each programming category, we observe both energy-efficient and energy-greedy behaviors.
-Therefore, we conclude that it depends more on the implementation of the library itself than on the category of the programming language.
-Scala and Kotlin are excellent examples to support this hypothesis, as both of them run on the same virtual machine as Java (OpenJDK 16.1).
-Yet, their average power is 130\% higher than the Java implementation.
+Therefore, we conclude that it depends more on the library's implementation than the programming language category.
+Scala and Kotlin are excellent examples to support this hypothesis, as both run on the same virtual machine as Java (OpenJDK 16.1).
+However, their average power is 130\% higher than the Java implementation.
 
 \paragraph{Stressed mode}
-Although the same classes remained the same, not all the languages had the same evolution, and here we can clearly observe a correlation with the category of the programming language rather than the implementation itself.
-We can easily point out that the average amount of power used by VM-based languages doubles when they have more than 100 clients at the same time.
-Except PHP, all the interpreted and compiled languages preserved their energetic behaviour.
-Our hypothesis points to the JIT, since it compiles the code and makes it run faster, hence stressing the CPU.
-An interesting behavior has been noticed for the GraalVM: the decrease in energy consumption when increasing the number of clients.
-This is related to the performance drop, which was probably due to the bottleneck situation where the GraalVM could not handle more than 100 clients simultaneously.
+Although the same classes remained the same, not all languages had the same evolution.
+Here we can observe a correlation with the category of the programming language rather than the implementation itself.
+We can quickly point out that the average amount of power used by VM-based languages doubles when they have more than 100 clients simultaneously.
+Except for PHP, all the interpreted and compiled languages preserved their energetic behavior.
+Our hypothesis points to the JIT since it compiles the code and makes it run faster, hence stressing the CPU.
+An interesting behavior has been noticed for the GraalVM: decreasing energy consumption when increasing the number of clients.
+This is related to the performance drop, probably due to the bottleneck situation where the GraalVM could not handle more than 100 clients simultaneously.
 
@@
 \paragraph{Satisfaction ratio}
-Most of the considered frameworks satisfy all the requests by either reducing the number of requests per second or increasing the processing time.
-However, there are some frameworks that have chosen a different approach, such as Dart or Scala, where the choice was to keep a certain limit of latency even if not all the requests are answered.
+Most of the considered frameworks satisfy all the requests by reducing the number of requests per second or increasing the processing time.
+However, some frameworks have chosen a different approach, such as Dart or Scala, where the choice was to keep a specific latency limit even if not all the requests are answered.
 Furthermore, we tend to observe this behavior among other frameworks, such as Python or Asynchronous NodeJS, when the number of clients exceeds 800.
@@
 \end{figure}
-Most of the servers hit their RPS limit after 5 clients and 100 clients for VM-based servers, and after this the number keeps constant, which will decrease the average RPS per client.
+Most of the servers hit their RPS limit after five clients and 100 clients for VM-based servers, and after this, the number keeps constant, which will decrease the average RPS per client.
 . The Net server is the most performant, followed by Java and Go, while Python and Ruby are the least performant.
@@
 \paragraph{Tail Latency}
-Even if the number of requests per second goes up, that doesn't mean the latency will go down.
+Even if the number of requests per second increases, that does not mean the latency will go down.
 As one can notice in figure~\ref{fig:tail99_clients}, until the 1000 clients, Go provides the least latency besides.Net.
 GraalVM provides the highest average latency.
-However, Dart tends to become slower when we increase the number of clients, until we pass the 600 simultaneous clients, and then it changes its behavior.
+However, Dart tends to become slower when we increase the number of clients until we pass the 600 simultaneous clients, and then it changes its behavior.
 Instead of satisfying most of the requests, it notifies the clients directly that the server is saturated, resulting in a drop in satisfaction ratio and an amelioration in average latency.
@@
 \subsubsection{Energy Per Request}
-Now, after we made a separation between the energy and the performances, we have seen that most of the performance servers tend to be energy-greedy, so we propose to investigate this trade-off between energy and performances.
+After separating the energy and the performances, we have seen that most performance servers tend to be energy-greedy, so we propose investigating this trade-off between energy and performance.
 To do so, we report the average cost of a single request in joules.
-Except GraalVM when the cost of the a single request increases with when we add more clients, all the frameworks report on a constant cost, Java, .net and go are the most energy efficient, while Python and Ruby may cost up to 10x more.
-Therefore, we conclude that the number of clients does not significantly impact the energy consumption.
+Except for GraalVM, when the cost of the single request increases when we add more clients, all the frameworks report on an ongoing cost; Java, .Net, and go are the most energy efficient, while Python and Ruby may cost up to 10x more.
+Therefore, we conclude that the number of clients does not significantly impact energy consumption.
 Then, we study how the payload size of the requests impacts the energy consumption of the framework.

### 2022-09-29 chapters/programming_languages/rpc/rpc.tex
update : rpc

@@
 \subsection{Experimental Protocol}
+
 To answer these questions, we will define an interface using the Proto2~\footnote{\url{https://developers.google.com/protocol-buffers}}language, and then we will implement it in multiple servers, each one using a different programming language.
@@
 All the experiments are run on the cluster \textsf{paravance} of the G5K platform.
-This cluster comprises 72 identical machines; each is equipped with 2 Intel Xeon E5-2630 V3, with 128\,GB of RAM.
+This cluster comprises 72 identical machines equipped with 2 Intel Xeon E5-2630 V3, with 128\,GB of RAM.
 For more accuracy, our SUT (\emph{System Under Test}) runs a minimal version of Debian\,9 (4.9.0 kernel version), which enforces the core processes required for our experiment.
@@
 \paragraph{Client \& server environments.}
-To limit the impact of the network on the experiments, we run both the client and the server on the same machine.
+To limit networks' impact on the experiments, we run both the client and the server on the same machine.
 However, we isolate each part on a separate CPU socket to reduce the noise that the client might have on the server and \emph{vice-versa}.
@@
 
-% TODO : change php to the interpreted part 
 \subsection{Results \& Findings}
+\paragraph{Preparing the Data}
+Before we start our Analysis, we will first clean the data from the outliers using the interquartile range (IQR) method~\cite{leys2013detecting}.
+For each implementation, we will calculate the Q3 and Q1 of the population; then, we will exclude all the values outside the range
+$ [Q1 - 1.5 \times IQR, Q3 + 1.5 \times IQR ]$.
+
 \paragraph{\textsc{RQ}~1: How do RPC implementations consume energy concerning the size of the incoming request?}
@@
 To eliminate extra factors, we let the server handle the rate at which it can answer each request.
-However, we put a 20 seconds timeout deadline for each request.
+However, we put a 20 seconds timeout deadline for each category of requests.
 Therefore, our boundary condition is only the number of requests received by the server.
@@
     \item the \textsf{average number of requests per second}, which indicates the average number of clients that the server can handle,
-    \item the \textsf{average energy cost} of a single request: unlike the first indicator, this one shows how green the implementation is considering performance.
+    \item the \textsf{average energy cost} of a single request: unlike the first indicator, this one shows how green the implementation is, considering performance.
 \end{enumerate}
@@
 \end{figure}
+
 Figure~\ref{fig:energy_cost_payload} depicts the overall behavior of each framework based on the request size (payload).
@@
 \end{itemize}
-
+%  TODO : prove that there is no correlation between energy consumption and the number of clients 
 \subsubsection{Escalation Mode}
@@
           The energy consumption of a single request is affected slightly but still increases.
-    \item Increase in power and a drop in performance.
+    \item Increase in power while dropping in performance.
           Despite the increase in power consumption, the server becomes slightly slower, which increases the energy cost per request.
@@
 \begin{itemize}
-    \item \textsf{Timeout} where requests took too much time that the client canceled them, in this category, we find most of the dynamic codes, such as OpenJDK and Kotlin,
+    \item \textsf{Timeout} where requests took too much time that the client canceled them in this category, we find most of the dynamic codes, such as OpenJDK and Kotlin,
     \item \textsf{Size of request} exceeded the maximum size when the implementation could not handle requests with a large size, as observed with .Net, Go, .Net core, CPP, PHP, Scala, Nodejs, Ruby, and Python.
@@
 \end{figure}
-Most of the servers hit their RPS limit after five clients and 100 clients for VM-based servers, and after this, the number keeps constant, which will decrease the average RPS per client.
+Most of the servers hit their RPS limit after five clients and 100 clients for VM-based servers, and after this, the number keeps constant, decreasing the average RPS per client.
 . The Net server is the most performant, followed by Java and Go, while Python and Ruby are the least performant.

### 2022-09-30 chapters/programming_languages/rpc/rpc.tex
update : rpc RQ1

@@
 \begin{description}
+    \item[\textsc{RQ}~2:] How do RPC implementations consume energy regarding the number of concurrent clients?
     \item[\textsc{RQ}~1:] How do RPC implementations consume energy concerning the size of the incoming request?
-    \item[\textsc{RQ}~2:] How do RPC implementations consume energy regarding the number of concurrent clients?
 \end{description}
@@
 
-\paragraph{\textsc{RQ}~1: How do RPC implementations consume energy concerning the size of the incoming request?}
+\subsubsection{[\textsc{RQ}~1:] How do RPC implementations react to the number of clients?}
+
+In this part, we will analyze the behavior of the different implementations regarding the number of clients.
+
+\paragraph{Power behavior}
+\begin{figure}[!htb]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/power_consumption_clients}
+    \end{center}
+    \caption{Average power consumption based on the number of the clients}\label{fig:power_consumption_clients}
+\end{figure}
+First, we start with the overall power consumption for each server implementation.
+Figure~\ref{fig:power_consumption_clients} shows a heatmap of the average power consumption of each implementation.
+As one can notice, there are two main scenarios.
+The first one is when the number of clients is lower than 100, which we will call \textsf{lite scenario}, and the second one, the \textsf{stressed scenario}, is when the number of clients is more exceeds 100.
+
+\subparagraph{Lite scenario}
+The benchmarked implementations can be grouped into two classes:
+\begin{enumerate}
+    \item \textsf{Energy-efficient frameworks} where most of the framework's power consumption is around 33 Watts.
+    \item \textsf{Energy-greedy frameworks} where the average power consumption is higher than 45 Watts.
+\end{enumerate}
+In each programming category, we observe both energy-efficient and energy-greedy behaviors.
+Therefore, we conclude that it depends more on the library's implementation than the programming language category.
+Scala and Kotlin are excellent examples to support this hypothesis, as both run on the same virtual machine as Java (OpenJDK 16.1).
+However, their average power is 130\% higher than the Java implementation.
+
+\subparagraph{Stressed scenario}
+Although the same classes remained the same, not all languages had the same evolution.
+Here we can observe a correlation with the category of the programming language rather than the implementation itself.
+We can quickly point out that the average amount of power used by VM-based languages doubles when they have more than 100 clients simultaneously.
+Except for PHP, all the interpreted and compiled languages preserved their energetic behavior.
+Our hypothesis points to the JIT since it compiles the code and makes it run faster, hence stressing the CPU.
+An interesting behavior has been noticed for the GraalVM: decreasing energy consumption when increasing the number of clients.
+This is related to the performance drop, probably due to the bottleneck situation where the GraalVM could not handle more than 100 clients simultaneously.
+
+\paragraph{Performance Behaviour}
+this paragraph will cover the number of requests per second processed per server without looking at its energy.
+Here, we consider three observable variables:
+\begin{itemize}
+    \item \textsf{Satisfaction ratio}: how many requests have been satisfied among the total requests,
+    \item \textsf{Request Per Seconds}: The number of the requests that have been answered from the server,
+    \item \textsf{Tail Latency at 99\%}: one of the best metrics to evaluate the performances of a server.
+\end{itemize}
+
+\subparagraph{Satisfaction ratio}
+Most of the considered frameworks satisfy all the requests by reducing the number of requests per second or increasing the processing time.
+However, some frameworks, such as Dart or Scala, have taken a different approach, maintaining a specific latency limit even if not all requests are answered.
+Furthermore, we tend to observe this behavior among other frameworks, such as Python or Asynchronous NodeJS, when the number of clients exceeds 800.
+To dive more into this behavior, the reader can refer to the following  GitHub repository \url{https://github.com/chakib-belgaid/grpc_analysis}
+
+\subparagraph{RPS}
+
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/rps_clients}
+    \end{center}
+    \caption{Number of requests per second based on the number of clients}\label{fig:rps_clients}
+\end{figure}
+%  TODO : REname the frameworks based on their dockerfile 
+Figure~\ref{fig:rps_clients} presents the number of requests per second for each implementation. The heatmap is colored in a logarithmic scale to visualize the results better.
+As one can observe, most of the servers hit their RPS limit after five clients and 100 clients for VM-based servers, and after this, the number keeps constant, decreasing the average RPS per client.
+.Net server is the most performant, followed by Java and Go, while Python and Ruby are the least performant.
+
+\subparagraph{Tail Latency}
+Even if the number of requests per second increases, that does not mean the latency will go down.
+As one can notice in figure~\ref{fig:tail99_clients}, until the 1000 clients, Go provides the least latency besides.Net.
+GraalVM provides the highest average latency.
+However, Dart tends to become slower when we increase the number of clients until we pass the 600 simultaneous clients, and then it changes its behavior.
+Instead of satisfying most of the requests, it notifies the clients directly that the server is saturated, resulting in a drop in satisfaction ratio and an amelioration in average latency.
+
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/tail99_clients}
+    \end{center}
+    \caption{Tail latency (99\%) based on the number of clients}\label{fig:tail99_clients}
+\end{figure}
+
+\paragraph{Energy Per Request}
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/energy_cost_clients}
+    \end{center}
+    \caption{The cost of a single request based on the number of clients (mj)}\label{fig:energy_cost_clients}
+\end{figure}
+After separating the energy and the performances, we have seen that most performance servers tend to be energy-greedy, so we propose investigating this trade-off between energy and performance.
+To do so, we report the average cost of a single request in millijoules in Figure~\ref{fig:energy_cost_clients}.
+
+Except for GraalVM (aka Java AOT), the cost of the single request decreases when we add more clients.Java, .Net, and go are the most energy efficient, while Python and Ruby may cost up to 10x more. 
+Despite its low power, textsfCrystal exhibits greedy behavior when dealing with requests from fewer clients.
+Such behavior is caused by the low number of requests per second, which also results in high latency.
+Therefore, we conclude that the number of clients does not significantly impact energy consumption.
+Then, we study how the payload size of the requests impacts the energy consumption of the framework.
+
+\paragraph{Synthesis}: 
+Depending on the number of clients, most frameworks exhibit two distinct behaviors.
+When dealing with a small number of clients, frameworks, regardless of programming category, tend to be energy-efficient and performant.
+However, as the number of clients grows, the frameworks become more energy-hungry and less performant.
+Interpreted languages suffer the most, particularly in terms of latency and the cost of a single request.
+\textsf{Crystal}, interestingly, acted. It behaves oppositely. The cost of a single request increases as the number of clients decreases.
+
+\subsubsection{\textsc{RQ}~2: How do RPC implementations consume energy concerning the size of the incoming request?}
 
@@
 
-[\textsc{RQ}~2:] How do RPC implementations react to the number of clients?
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/energy_cost_clients}
-    \end{center}
-    \caption{Energy behaviour based on the number of the clients}\label{fig:rpcproenergy_cost_clients}
-\end{figure}
+% \subsection{Threads to Validity \note{missing}}
 
-\subsubsection{Power behavior}
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/power_consumption_clients}
-    \end{center}
-    \caption{Average power consumption based on the number of the clients}\label{fig:power_consumption_clients}
-\end{figure}
+\subsection{Conclusion \note{missing}}
 
@@
 
-Based on the heatmap, we can distinguish two modes:
-\begin{itemize}
-    \item \textsf{Low number of clients} when the number of concurrent clients is below 100,
-    \item \textsf{Moderate to a high number of clients} when the number of clients exceeds 100.
-\end{itemize}
-
-\paragraph{Lite mode}
-The benchmarked implementations can be grouped into two classes:
-\begin{enumerate}
-    \item \textsf{Energy-efficient frameworks} where most of the framework's power consumption is around 33 Watts.
-    \item \textsf{Energy-greedy frameworks} where the average power consumption is higher than 37 Watts.
-\end{enumerate}
-
-In each programming category, we observe both energy-efficient and energy-greedy behaviors.
-Therefore, we conclude that it depends more on the library's implementation than the programming language category.
-Scala and Kotlin are excellent examples to support this hypothesis, as both run on the same virtual machine as Java (OpenJDK 16.1).
-However, their average power is 130\% higher than the Java implementation.
-
-\paragraph{Stressed mode}
-Although the same classes remained the same, not all languages had the same evolution.
-Here we can observe a correlation with the category of the programming language rather than the implementation itself.
-We can quickly point out that the average amount of power used by VM-based languages doubles when they have more than 100 clients simultaneously.
-Except for PHP, all the interpreted and compiled languages preserved their energetic behavior.
-Our hypothesis points to the JIT since it compiles the code and makes it run faster, hence stressing the CPU.
-An interesting behavior has been noticed for the GraalVM: decreasing energy consumption when increasing the number of clients.
-This is related to the performance drop, probably due to the bottleneck situation where the GraalVM could not handle more than 100 clients simultaneously.
-
-\subsubsection{Performance Behaviour}
-In this section, we study only the number of requests per second processed by the server without looking at its energy.
-We consider three observable variables:
-\begin{itemize}
-    \item \textsf{Satisfaction ratio}: how many requests have been satisfied among the total requests,
-    \item \textsf{Request Per Seconds}: The number of the requests that have been answered from the server,
-    \item \textsf{Tail Latency at 99\%}: one of the best metrics to evaluate the performances of a server.
-\end{itemize}
-
-\paragraph{Satisfaction ratio}
-Most of the considered frameworks satisfy all the requests by reducing the number of requests per second or increasing the processing time.
-However, some frameworks have chosen a different approach, such as Dart or Scala, where the choice was to keep a specific latency limit even if not all the requests are answered.
-Furthermore, we tend to observe this behavior among other frameworks, such as Python or Asynchronous NodeJS, when the number of clients exceeds 800.
-
-\paragraph{RPS}
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rps_clients}
-    \end{center}
-    \caption{Number of requests per second based on the number of clients}\label{fig:rps_clients}
-\end{figure}
-
 \begin{figure}[!hbt]
@@
 \end{figure}
-Most of the servers hit their RPS limit after five clients and 100 clients for VM-based servers, and after this, the number keeps constant, decreasing the average RPS per client.
-. The Net server is the most performant, followed by Java and Go, while Python and Ruby are the least performant.
-
-\paragraph{Tail Latency}
-Even if the number of requests per second increases, that does not mean the latency will go down.
-As one can notice in figure~\ref{fig:tail99_clients}, until the 1000 clients, Go provides the least latency besides.Net.
-GraalVM provides the highest average latency.
-However, Dart tends to become slower when we increase the number of clients until we pass the 600 simultaneous clients, and then it changes its behavior.
-Instead of satisfying most of the requests, it notifies the clients directly that the server is saturated, resulting in a drop in satisfaction ratio and an amelioration in average latency.
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/tail99_clients}
-    \end{center}
-    \caption{Tail latency (99\%) based on the number of clients}\label{fig:tail99_clients}
-\end{figure}
-
 \begin{figure}[!hbt]
@@
 \end{figure}
-
-\subsubsection{Energy Per Request}
-After separating the energy and the performances, we have seen that most performance servers tend to be energy-greedy, so we propose investigating this trade-off between energy and performance.
-To do so, we report the average cost of a single request in joules.
-Except for GraalVM, when the cost of the single request increases when we add more clients, all the frameworks report on an ongoing cost; Java, .Net, and go are the most energy efficient, while Python and Ruby may cost up to 10x more.
-Therefore, we conclude that the number of clients does not significantly impact energy consumption.
-Then, we study how the payload size of the requests impacts the energy consumption of the framework.
-
-% \subsection{Threads to Validity \note{missing}}
-
-\subsection{Conclusion \note{missing}}

### 2022-09-30 chapters/programming_languages/rpc/rpc.tex
update : figuges

@@
 \begin{enumerate}
-    \item the \textsf{average power consumption} during the scenario indicates the overall behavior of the server when working for long durations,
-    \item the \textsf{tail latency} for the 99th percentile, which indicates how efficient the server is,
-    \item the \textsf{average number of requests per second}, which indicates the average number of clients that the server can handle,
-    \item the \textsf{average energy cost} of a single request: unlike the first indicator, this one shows how green the implementation is, considering performance.
+    \item the \textsf{average power consumption} during the scenario indicates the overall behavior of the server when working for long durations, Figure~\ref{fig:power_consumption_payload},
+    \item the \textsf{tail latency} for the 99th percentile, which indicates how efficient the server is, Figure~\ref{fig:tail99_payload}
+    \item the \textsf{average number of requests per second}, which indicates the average number of clients that the server can handle, Figure~\ref{fig:rps_payload}
+    \item the \textsf{average energy cost} of a single request: unlike the first indicator, this one shows how green the implementation is, considering performance, Figure~\ref{fig:energy_cost_payload}
 \end{enumerate}
+
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/power_consumption_payload}
+    \end{center}
+    \caption{Average power consumption based on the request size}\label{fig:power_consumption_payload}
+\end{figure}
+
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/tail99_payload}
+    \end{center}
+    \caption{Tail latency (99\%) based on the request size}\label{fig:tail99_payload}
+\end{figure}
+
+\begin{figure}[!hbt]
+    \begin{center}
+        \includegraphics[width=1.2\linewidth]{imgs/rps_payload}
+    \end{center}
+    \caption{Number of requests per second based on the request size}\label{fig:rps_payload}
+\end{figure}
 
@@
 
-Figure~\ref{fig:energy_cost_payload} depicts the overall behavior of each framework based on the request size (payload).
+\input{correlation_table.tex}
+
 For each framework, we can distinguish three modes, and they all depend on the payload size:
@@
     \item \textsf{Escalation} mode when the requests tend to be bigger, but the server can still manage to handle them, at the price of a change in the energy and performance behaviors,
-    \item \textsf{Broken state} mode when the requests are much heavier and the server break—like 10 MB.%ROMAIN unclear to me what this means
+    \item \textsf{Broken state} mode when the size of requests exceeds 1MB, most of the servers cannot handle them, and they tend to crash.
+After 10MB, except Elixir, no server could handle those requests.
 \end{enumerate}
@@
 \end{itemize}
-%  TODO : prove that there is no correlation between energy consumption and the number of clients 
+
 \subsubsection{Escalation Mode}
@@
 
-We can mention the case of Elixir that keeps scaling despite the lack of performance compared to other compiled languages (Go, CPP)..
+We can mention the case of \textsf{Elixir} that keeps scaling despite the lack of performance compared to other compiled languages (Go, CPP).
 
@@
 
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/power_consumption_payload}
-    \end{center}
-    \caption{Average power consumption based on the request size}\label{fig:power_consumption_payload}
-\end{figure}
-
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/rps_payload}
-    \end{center}
-    \caption{Number of requests per second based on the request size}\label{fig:rps_payload}
-\end{figure}
-\begin{figure}[!hbt]
-    \begin{center}
-        \includegraphics[width=1.2\linewidth]{imgs/tail99_payload}
-    \end{center}
-    \caption{Tail latency (99\%) based on the request size}\label{fig:tail99_payload}
-\end{figure}
+\clearpage

### 2022-09-30 chapters/programming_languages/programming_languages.tex
update : figuges
update : typos in thechempower

--- 2022-09-23 chapters/programming_languages/programming_languages.tex
@@
 \label{chapter:porgramming_langauges}
-% \section{Introduction}
-In this chapter, we study how the programming framework affects the energy the software consumes.
+
+% TODO : repphrase this
+
+After looking at Django's studies, we wanted to expand our study to other programming languages and frameworks.
+The idea behind this study is to expand the work of Pereira et al. from the micro benchmarking level to get closer to real-world applications.
+First, we will start with studying the RPC protocol since it offers a multi-language interface. Therefore we will reduce the developer expertise bias
+after that; we will enlarge this study to cover the most popular web frameworks.
+In this study, we will focus on the energy footprint of the web frameworks and the RPC protocol in the context of an unlimited time.
+Therefore the factor time will have less impact, unlike the micro benchmarking study.
+
+This chapter studies how the programming framework affects the energy the software consumes.
 We suggest starting with general micro-benchmarking and watching how each programming framework performs with the CPU and memory.

### 2022-10-01 chapters/benchmarking/conclusion.tex
correct : typos

--- 2022-09-22 chapters/benchmarking/conclusion.tex
@@
-%  CONCLUSION au lieu de extension
+\section{Conclusion}
+As seen in Section~\ref{sec:soa_benchmarking} of state of the art, a successful benchmark faces three challenges: reproducibility, accuracy, and representativeness. 
+This chapter has covered two of the three criteria.
 
-%% maybe add extension as a part of the conclusion for this chapter 
-\section{Conclusion \note{Uncomplete }}
-In this chapter we :
-- studied two aspects of the three creterias of the emperical research and how they will be applied in the field of energy measurement .
-- we provided a brief comparaison of different methods to increase the reproducibility of the measurement .
-- we studied the factors that can affect the measurement accuracy and how to reduce them .
-- and finally
+The first section covered the reproducibility challenge by studying the existing reproducibility techniques.
+We have observed that there are two primary methods for encapsulating experimental systems, the first is to utilize virtual machines, and the second is based on containers.
+We established that Docker is more suitable for energy-related studies for three reasons.
+
+\begin{enumerate}
+    \item It is more lightweight than virtual machines.
+    \item It offers interactivity with the hardware of the host machine, which will enable us to gather more metrics.
+    \item It has a constant overhead, a key factor to nullify the encapsulation impact on the energy consumption when performing empirical analyses.
+\end{enumerate}
+
+After settling on the most suitable choice to encapsulate experiments, we highlighted the need to enhance the reproducibility of empirical tests along several axes (benchmarks, metrics, and candidates) to keep up with the rapid pace of software development.
+Furthermore, we have provided a model that enables comparative experiment \emph{extension}.
+The foundation of this model is separating the experiment into multiple independent components: an orchestrator that executes the experiment, one or more observers that collect metrics, the candidates being tested, and the benchmark against which these candidates are being compared.
+
+As for the second section, we addressed the accuracy challenge in energy-based experiments.
+We started by analyzing the impact of the chosen encapsulation method from the previous section on this challenge to determine that the impact is negligible.
+Then, we conducted an empirical study utilizing some of the well-known benchmarks in literature, \texttt{Stress-ng}~\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} and \texttt{NAS Parallel Benchmark}~\cite{Bailey:1991:NPB:125826.125925}, on a variety of machines with diverse hardware combinations and operating system setups and tunings.
+We have shown that optimizing the operating system can significantly affect how accurately energy is used.
+This effect can affect the accuracy of the experiments from an increase of $5\times$ to a penalty of $30\times$.
+
+We have also shown in this section the harm that this taming of the variation can cause to the representativeness of the results since some aspects that help increase the accuracy of the results in the research environment cannot be applied in the production environment, such as turning off the C-states.
+Another impact of increasing this accuracy was the increase in the overall energy consumption of the tests.
+Even if the overall was for all the candidates, this remains an issue to be addressed.
+
+This chapter aims to establish a reproducible and accurate protocol for conducting energy-related experiments. From now on, we shall use this protocol in our studies to reduce software energy usage.
 

### 2022-10-01 chapters/literature/draft_citations.tex
correct : typos

@@
 Efficiency in energy usage is a well-known topic.
-In the majority of fields, the purpose is to minimize the energy consumption of electrical devices and components.
-Modern times even see energy classification (A, B\dots F) for homes, cars, and electronic products, to provide the consumer an indication of the energy consumption of their devices, which will reflect on their power bill.
+In most fields, the purpose is to minimize the energy consumption of electrical devices and components.
+Modern times even see energy classification (A, B\dots F) for homes, cars, and electronic products to provide the consumer an indication of the energy consumption of their devices, which will reflect on their power bill.
 
@@
 We will be proceeding through an empirical analysis of the energy consumption of the software while changing some components of the source code without impacting its behavior.
-To do this, we will elaborate on a benchmarking process and a set of tools that are intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
-Thus, we will begin by examining the state of the art in empirical analysis and retrieving the best empirical experimentation methodologies in the research field.
-Then, we will narrow these practices down to computer science so that we can finally adapt them to energy consumption.
+To do this, we will elaborate on a benchmarking process and a set of tools intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
+Thus, we will begin by examining state of the art in empirical analysis and retrieving the best empirical experimentation methodologies in the research field.
+Then, we will narrow these practices down to computer science so we can finally adapt them to energy consumption.
 
@@
 It provides examples of hardware and software measuring instruments and describes their differences, benefits, and drawbacks.
-It also examines the sources of energy measurement variations, which represent a significant obstacle to achieving precise readings and higher accuracy.
+It also examines the sources of energy measurement variations, representing a significant obstacle to achieving precise readings and higher accuracy.
 Then, in Section~\ref{section:soa_energyoptimization}, we will go through some of the previous work on improving the energy consumption of software.
@@
 Finally, it should \textbf{represent} reality.
-In other words, the experiment's findings should be applicable outside of the research lab as well.
+In other words, the experiment's findings should also be applicable outside the research lab.
 The aim of \textbf{representativeness} in this thesis is the manufacturing environment.
@@
 
-Experiment reproducibility is frequently listed as one of the most difficult challenges faced by researchers.
-Reproducing an experiment has been one of the major tools science has used to help establish the validity and importance of scientific findings since the Philosophical Transactions of the Royal Society were established in 1665~\cite{hankins1986debate}.
+Experiment reproducibility is frequently listed as one of the most difficult challenges researchers face.
+Reproducing an experiment has been one of the significant tools science has used to help establish the validity and importance of scientific findings since the Philosophical Transactions of the Royal Society were established in 1665~\cite{hankins1986debate}.
 Many of the outcomes are not reproducible,\footnote{Trouble at the lab, The Economist, 19 October 2013;  www.economist.com/news/briefing/ 21588057-scientists-think-science-self-correcting- alarming-degree-it-not-trouble.} which led to a \emph{replication crisis}.
@@
 In Social sciences, this can be achieved by selecting a representative sample size.
-\citeauthor{omair2014sample} presents a guideline on to achieve a such representativeness.
-As for computer science.
-The field of benchmarking is still in its infancy, and there is no consensus on how to achieve representativeness.
-howerver many attemps have been made to provide a set of benchmarks for specific purpose.
+\citeauthor{omair2014sample} presents a guideline on achieving such representativeness.
+As for computer science, the field of benchmarking is still in its infancy, and there is no consensus on how to achieve representativeness.
+However, many attempts have been made to provide a set of benchmarks for a specific purpose.
 For example, the \href{https://www.spec.org/}{Standard Performance Evaluation Corporation} (SPEC) provides a set of benchmarks for CPU performance evaluation.
-this sets covers a wide range of use cases such as  CPU17 for testing the CPU \footnote{\url{https://www.spec.org/cpu2017/}}, SPECviewperf\footnote{\url{https://gwpg.spec.org/benchmarks/benchmark/specviewperf-2020-v3-1/}} for Graphic usage and
+This sets covers a wide range of use cases such as  CPU17 for testing the CPU \footnote{\url{https://www.spec.org/cpu2017/}}, SPECviewperf\footnote{\url{https://gwpg.spec.org/benchmarks/benchmark/specviewperf-2020-v3-1/}} for Graphic usage and
 one can cite StressNg when it comes to benchmark the hardware components, the SPEC benchmark when it comes to benchmark the software performance, and SPECPower\footnote{\url{https://www.spec.org/power_ssj2008} }.
@@
 As for programming languages we can cite Computer Language Benchmarks Game~\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}, which is a collection of benchmarks for various programming languages.
-The benchmarks are designed to be small, self-contained, and easy to implement in any language.
-The benchmarks are also made in a way to represent of most of the typical real-world workloads in an isolated manner.
+The benchmarks are designed to be minor, self-contained, and easy to implement in any language.
+The benchmarks are also made to represent most of the typical real-world workloads in an isolated manner.
 Dacapo~\cite{blackburn2006dacapo} and renaissance~\cite{renaissance} are other examples of benchmarking sets that are created to represent the Java Virtual Machine (JVM) performance.
@@
 to find out that Tier-1 papers commit an average of five benchmarking crimes.
-To analyze the magnitude of the phenomenon, they have identified a set of 22 "benchmarking crimes" that threaten the validity of the system.
+To analyze the magnitude of the phenomenon, they have identified a set of 22 "benchmarking crimes" that threaten the system's validity.
 
@@
 They treated this problem by proposing two strategies for detecting measurement bias by using causal analysis and preventing it with setup randomization~\cite{mytkowicz2009producing}.
-another study that was published in the book "Measuring computer performance: a practitioner's guide"~\cite{lilja2005measuring},\citeauthor{lilja2005measuring} examined performance indicators and gave in-depth treatment of benchmark program tactics.
-they provided clear explanations of the basic statistical methods required for interpreting measured performance data.
+Another study that was published in the book "Measuring computer performance: a practitioner's guide"~\cite{lilja2005measuring},\citeauthor{lilja2005measuring} examined performance indicators and gave an in-depth treatment of benchmark program tactics.
+They clearly explained the basic statistical methods required for interpreting measured performance data.
 They also outlined the overall design of the experimental method and demonstrated how to collect the most information with the least amount of work.
-This practical book will appeal to anybody seeking a comprehensive, yet intuitive, grasp of computer system performance analysis.
+This practical book will appeal to anybody seeking a comprehensive yet intuitive grasp of computer system performance analysis.
 
@@
 
-In their paper~\cite{stodden2018empirical}, the authors found that, even though it's a big step in the right direction, journal policies that require authors to give back digital scholarly objects after publication, like the data and code that back up the claims, don't get more than half of these objects back.
+In their paper~\cite{stodden2018empirical}, the authors found that, even though it's a big step in the right direction, journal policies that require authors to give back digital scholarly objects after publication, like the data and code that back up the claims, do not get more than half of these objects back.
 Then, using these artifacts, about a fourth of the published computational claims in the study could be made.
-They suggested putting out the claims in the literature and the digital scholarly objects that back them up at the same time.
+They suggested putting out the claims in the literature and the digital scholarly objects that back them up simultaneously.
 
@@
 \end{figure}
-Finally, to make unify the benchmarking methodology across different research works in the field of computer science, we can cite the paper~\cite{stephen_evaluate_2012}.
+Finally, to unify the benchmarking methodology across different research works in the field of computer science, we can cite the paper~\cite{stephen_evaluate_2012}.
 In their approach, they divided any empirical experiment into four components.
@@
 \begin{enumerate}
-    \item \emph{measurement contexts} that indicate the software and hardware components that will alter or remain constant during the experiment.
-    \item \emph{workloads} which identify the benchmarks to use in the experiment, as well as their inputs;
-    \item \emph{metrics} that specify the attributes to be measured and how to assess them.
+    \item \emph{measurement contexts} indicates the software and hardware components that will alter or remain constant during the experiment.
+    \item \emph{workloads}  identify the benchmarks to use in the experiment, as well as their inputs;
+    \item \emph{metrics} specify the attributes to be measured and how to assess them.
     \item \emph{Data analysis} shows how to examine data and evaluate the outcomes of the analysis to offer insight into the assertions that arise from the study.
@@
 Many studies have been conducted to estimate such energy consumption that varies from static analysis of the source code to infer its energy consumption like \citeauthor{pereira_helping_2017} where they provided a tool to highlight the most energy consuming parts of the code~\cite{pereira_helping_2017}.
-The key advantage of this approach is that it allows you to estimate a program's energy usage without executing it.
+The essential advantage of this approach is that it allows practitioners to estimate a program's energy usage without executing it.
 Unlike program complexity, energy consumption is strongly dependent on the execution environment.
@@
 To address the issue of representativeness, many researchers measure the energy consumption of programs as they run. As a result, we will get more accurate results.
-There are various tools for measuring energy, and they cover a wide range of applications depending on how accurate and precise the results must be on the one hand, and the price that practitioners are prepared to pay for such accuracy and precision on the other.
+There are various tools for measuring energy, and they cover a wide range of applications depending on how accurate and precise the results must be on the one hand and the price that practitioners are prepared to pay for such accuracy and precision on the other.
 
@@
     \item \em{Temporal granularity}: same as spatial granularity, temporal granularity helps us to identify the sequence of code that need to be optimized,
-    \item \em{Scalability}: this is mainly related to the cost of the tools and the ease of their integration for our system,
+    \item \em{Scalability}: this is mainly related to the cost of the tools and the ease of their integration into our system,
     \item  \em{Accuracy}: to eliminate extra hazards and get a more representative measurement.
@@
 
-We believe that accuracy can be extracted from those criteria since it is a result of the combination of the two first ones. therefore, we will focus more on the first 3 criteria from later on.
-Below, we will discuss some of the well-known tools that are used in literature.
+We believe that accuracy can be extracted from these criteria by combining the first two results. Therefore, we will focus more on the first three criteria later on.
+Below, we will discuss some of the well-known tools used in literature.
 
 \subsection{Hardware Tools}
-Nowadays, most high-performance computing systems (HPC) implement a tool to report the energy consumption of the nodes for the sake of monitoring and administration.
+Nowadays, most high-performance computing systems (HPC) implement a tool to report the nodes' energy consumption for monitoring and administration.
 Those tools are mainly integrated within the power supply units (PSU) or the power distribution units (PDU).
-Then, they provide an interface and a log to follow the history of energy consumption.
-Despite their scalability and ease of integration.
-Such tools lack both spatial and temporal granularity since they monitor the whole energy of the nodes, and most of the time they have a very low sampling frequency.
+Then, they provide an interface and a log to follow the energy consumption history.
+Despite their scalability and ease of integration, such tools lack both spatial and temporal granularity since they monitor the whole energy of the nodes, and most of the time, they have a shallow sampling frequency.
 Most of those tools are provided directly by the manufacturers.
@@
 %% add something to link with wattsup pro 
-WattsUp Pro, is a device that can be installed between the power source of the machine and the system under test.
-It allows a sampling frequency up to 1\,Hz and has an internal memory to store a wide variety of data, such as the maximum voltage, and current, that later can be exported via USB port for personal usage or lined to some graph programs like Logger Pro or LabQuest.
+WattsUp Pro is a device that can be installed between the power source of the machine and the system under test.
+It allows a sampling frequency of up to 1\,Hz.
+It has an internal memory to store a wide variety of data, such as the maximum voltage and current, that can later be exported via USB port for personal usage or lined to graph programs like Logger Pro or LabQuest.
 The main advantage of this tool is the ability to monitor the energy consumption from a different device which will reduce the risk of interference with the energy consumption of the test~\cite{hirst2013watts}
-
-Despite its high temporal granularity, WattsUp Pro lacks in term of spatial granularity since it monitors the energy consumption of the whole system.
-To have a finer granularity we need to isolate the energy consumption of each component.
+Despite its high temporal sampling, WattsUp Pro lacks spatial granularity since it monitors the energy consumption of the whole system.
+To have finer granularity, we need to isolate the energy consumption of each component.
 % For this, we will need more intrusive tools such.
@@
 
-PowerMon and its upgraded version powerMon2~\cite{bedard2010powermon} are based on a micro-controller chip that can monitor up to 6 channels (8 for powermon2), simultaneously.
-Therefore, we are able to monitor the power consumption of 4 devices at the same time.
+PowerMon and its upgraded version powerMon2~\cite{bedard2010powermon} are based on a micro-controller chip that can simultaneously monitor up to 6 channels (8 for powermon2).
+Therefore, we can monitor the power consumption of 4 devices at the same time.
 The frequency sample of this tool is up to 50\,Hz, with an accuracy of 1.2\%.
-Powermore2 comes in a smaller size that can fit within 3.5 inches rack drive.
+Powermore2 is smaller and can fit within 3.5 inches rack drive.
 
@@
 % On the other hand, GreenMiner \cite{hindle2014greenminer}, %%i dont think that  mobile measurement will be relevent for my case 
-powerpack~\cite{ge2009powerpack} in the other hand is an API that synchronizes the data gathered from other monitoring tools such as Watt’s Up Pro, NI and RadioShack pro and the lines of code.
+powerpack~\cite{ge2009powerpack} in the other hand, is an API that synchronizes the data gathered from other monitoring tools such as Watt’s Up Pro, NI and RadioShack pro and the lines of code.
 %% maybe remove this one from the list because it is not well known.
@@
 Fortunately, Intel and later AMD proposed a tool that estimates the power consumption of different parts of the CPI based on counter performances.
-RAPL (\emph{RUnning Average Power Limit}) \cite{hackenberg2013power,hackenberg2015energy} is a set of registers that was introduced by Intel in their CPU since Sandy bridge generation, and later it was followed by AMD, since Family 17h Zen.
+RAPL (\emph{RUnning Average Power Limit}) \cite{hackenberg2013power,hackenberg2015energy} is a set of registers that Intel introduced in their CPU since Sandy bridge generation, and later it was followed by AMD since Family 17h Zen.
 
@@
 Figure~\ref{fig:rapl-domaine} shows the different scopes that can be monitored using RAPL.
-The CPU package can be monitored for both server and desktop processors. However, DRAM is only available on server CPUs, whilst the integrated GPU is only available on desktop processors.
-The advantage of such an approach is the absence of any intrusive measurement tools.
+The CPU package can be monitored for both server and desktop processors. However, DRAM is only available on server CPUs, while the integrated GPU is only available on desktop processors.
+The advantage of such an approach is the absence of intrusive measurement tools.
 Furthermore, they have a high temporal granularity with a sampling frequency of up to 1\,KHz~\cite{ilsche_power_2015}.
@@
 
-Many software measuring tools understand the behavior of a power model and provide estimates of energy usage.
-This model is then used to allocate the observed energy consumption among different execution entities, such as processes, control groups, threads, or even code lines.
+Many software measuring tools understand a power model's behavior and estimate energy usage.
+This model is then used to allocate the observed energy consumption among different execution entities, such as processes, control groups, threads, or code lines.
 
 The first examples of software measurement tools are PowerAPI~\cite{colmant2018next}, SmartWatts formula~\cite{fieni2020smartwatts} and SelfWatts~\cite{fieni2021selfwatts}.
-These tools collect global energy consumption measurements from RAPL and use other system events such as cache misses/hits and CPU frequency evolution (DVFS) via a sensor to construct a power model of the control groups (system control groups, docker containers, kubernetes pods, etc.) using a Ridge regression.
-The model continuously learns and improves its precision of real-time energy usage data with a maximum frequency of 100 Hz.
+These tools collect global energy consumption measurements from RAPL and use other system events such as cache misses/hits and CPU frequency evolution (DVFS) via a sensor to construct a power model of the control groups (system control groups, docker containers, Kubernetes pods, etc.) using a Ridge regression.
+The model continuously learns and improves its real-time energy usage data precision with a maximum frequency of 100 Hz.
 The instrument has a decentralized, lightweight design.
 Only the lightest sensors required for data collection and transmission are put into the monitored devices.
-The SmartWatts formula is then executed on the primary server to construct the model that permits assigning the energy usage for each operating control group.
+The SmartWatts formula is then executed on the primary server to construct the model that permits assigning the energy usage for each functional control group.
 PowerAPI is only compatible with Linux on a bare-metal physical computer.
@@
 This program uses power models to predict process energy usage.
-It uses CPU events that are passed from the measured node to a model generator node to construct the power model.
+It uses CPU events passed from the measured node to a model generator node to construct the power model.
 It works by combining a description of the CPU with a list of the hardware events through multiple calibration phases to build a robust model.
@@
 Joulemeter~\cite{kothari2009joulemeter,jagroep2017energy} is a Microsoft software that estimates the energy usage of Windows running applications down to the process level by using power models (for CPU, memory, and drives).
-It employs low-overhead power models to infer power consumption from resource utilization during runtime, and it provides power-limiting features for virtual machines.
+It employs low-overhead power models to infer power consumption from resource utilization during runtime and provides power-limiting features for virtual machines.
 Previous Joulemeter tests~\cite{jagroep2015profiling} demonstrated that the instrument provides a less accurate estimation of energy use that differs greatly from the real one.
@@
 Another process-level energy usage measuring tool is Jolinar~\cite{islam2016measuring,noureddine2016jolinar}.
-The tool does not need a calibration phase and relies on pre-established power models based on hardware metrics (TDP, disk I/O rate, and so on).
+The tool does not need a calibration phase and relies on pre-established power models based on hardware metrics like TDP, disk I/O rate, and CPU frequency.
 These settings must be identified and supplied by the user for his machine.
-Jolinar is only capable of measuring the energy consumption of one application at a time.
-At the end of the execution, the tool provides the CPU, DRAM, and disk energy usage of the main process.
+Jolinar can only measure one application's energy consumption at a time.
+At the end of the execution, the tool provides the process's CPU, DRAM, and disk energy usage.
 Jalen~\cite{noureddine2015monitoring} is another tool that profiles and monitors the energy usage of a Java program.
@@
 Unfortunately, this is not possible when it comes to measuring the energy consumption of a system.
-Applying the benchmarking guidelines and repeating the same experiment within the same configuration is not sufficient to reproduce the same energy measurements, not only between identical machines, but even within the same machine.
-This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.
+Applying the benchmarking guidelines and repeating the same experiment within the same configuration is insufficient to reproduce the exact energy measurements between identical machines and even within the same machine.
+This difference---also called \emph{energy variation} (EV)---has seriously threatened the accuracy of experimental evaluations.
 
 Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
-We can observe a large variation of the energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.
+We can observe a consedirable variation of the energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.
 
@@
 Some researchers started investigating the hardware impact of the energy variation of power consumption.
-As an example, one can cite~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the main cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
+As an example, one can cite~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the leading cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
 \cite{heinrich_predicting} described this variation as a set of parameters, such as CPU Frequency and the thermal effect.
 \subsection{Studying Hardware Factors}
-This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}, but has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
+This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}.
+However, it has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
 On the one hand, the correlation between processor temperature and energy consumption was one of the most explored paths.
-Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors can exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
+Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors could exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
 On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the CPU temperature and the energy consumption variation are tightly coupled (up to 16\% increase in the variation when the temperature changed from 37.7°C to 74.5°C ).
 
-This exposes the processor temperature as a delicate factor to consider while comparing energy consumption variations across a set of homogeneous processors.%, but also, if a same workload can cause the CPUs to reach different temperatures.
+This exposes the processor temperature as a delicate factor when comparing energy consumption variations across homogeneous processors.
 
-The ambient temperature was also discussed in many papers as a candidate factor for the energy variation of a processor.
-In~\cite{ranka_energy_2009}, the authors claimed that energy consumption may vary due to fluctuations caused by the external environment.
+The ambient temperature was also discussed in many papers as an essential candidate factor for the energy variation of a processor.
+In~\cite{ranka_energy_2009}, the authors claimed that energy consumption might vary due to fluctuations caused by the external environment.
 These fluctuations may alter the processor's temperature and energy consumption.
-However, the temperature inside a data center does not show major variations from one node to another.
+However, the temperature inside a data center does not show significant variations from one node to another.
 In~\cite{el_mehdi_diouri_your_2013}, El~Mehdi~Dirouri~\emph{et~al.} showed that switching the spot of two servers does not affect their energy consumption.
-Moreover, changing hardware components, such as the hard drive, the memory or even the power supply, does not affect the energy variation of a node, making it mainly related to the processor.
+Moreover, changing hardware components, such as the hard drive, the memory, or even the power supply, does not affect the energy variation of a node, making it mainly related to the processor.
 This result was recently assessed by~\cite{wang_potential_2018}, where the rack placement and power supply introduced a maximum of $2.8\,\%$ variation in the observed energy consumption.
@@
 \citeauthor{inadomi_analyzing_2015}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
-All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
+All three tools recorded the same $10\,\%$ of energy variation that was supposedly related to the manufacturing process.
 
 \subsection{Mitigating Energy Variations}
-Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
+Acknowledging the energy variation problem on processors, many papers proposed contributions to reduce and mitigate this variation.
 In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
 The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
-Acun~\emph{et~al.}~\cite{acun_variation_2016} found a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
-They also proposed some guidelines to reduce this variation by replacing the old---slower---chips, by load balancing the workload on the CPU cores and leaving one core idle.
+Acun~\emph{et~al.}~\cite{acun_variation_2016} found a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
+They also proposed some guidelines to reduce this variation by replacing the old---slower---chips, load balancing the workload on the CPU cores, and leaving one core idle.
 They claimed that the variation between the processor cores is insignificant.
-In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating for the uneven effects of power capping.
+In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system could deal with the energy variation by compensating for the uneven effects of power capping.
 
@@
 
-We will now turn our focus to the energy optimization challenge after considering the various ways of measuring energy consumption in computers and understanding the energy variation problem.
-Over the previous decade, there has been a considerable increase in interest in this field, with many papers proposing different approaches to reduce the energy consumption of software applications.
-This section will pass through the main contributions in this field, and will focus on the two following parts.
+We will now focus on the energy optimization challenge after considering the various ways of measuring energy consumption in computers and understanding the energy variation problem.
+Over the previous decade, there has been considerable interest in this field, with many papers proposing different approaches to reduce the energy consumption of software applications.
+This section will pass through the main contributions in this field and focus on the following two parts.
 
 \subsection{Energy Optimization in the Design Phase}
-The first part of this section will focus on energy optimizations in the conception phase, where the goal is to make the final product use less energy by choosing the best set of programming languages, tools, libraries, etc.
-It also includes all the work and optimizations that developers can do to the source code to make the software use less energy when it is running.
+The first part of this section will focus on energy optimizations in the conception phase.
+The goal is to make the final product use less energy by choosing the best programming languages, tools, libraries, etc.
+It also includes all the work and optimizations that developers can do to the source code to make the software use less energy when running.
 
 We start with the work of \citeauthor{pereira_energy_2017} where they did an energy consumption comparison analysis of the most popular programming languages.
-Using the Pareto optimum, the paper provides recommendations on how to combine some of these languages to enhance code quality while taking into account execution time, memory utilization, and energy consumption using the Pareto optimum.
+The paper recommends combining some of these languages to enhance code quality while considering execution time, memory utilization, and energy consumption using the Pareto optimum~\cite{hochman1969pareto}.
 Some of the study's findings demonstrate that interpreted languages, such as Python, have lower energy efficiency than compiled languages, such as C or Rust.
-The research also offers language combinations that developers might use together to improve energy economy, execution speed, and memory utilization.
-This paper's findings were based on the game benchmark, which is the most famous set of benchmarks that compares several programming languages.
+The research also offers language combinations that developers might use to improve energy economy, execution speed, and memory utilization.
+This paper's findings were based on the game benchmark, the most famous benchmark comparing several programming languages.
 
-In this paper, we define a ranking of energy efficiency in programming languages.
+This paper defines a ranking of energy efficiency in programming languages.
 We considered a set of computing problems implemented in ten well-known programming languages and monitored the energy consumed when executing each language.
@@
 In their paper~\cite{couto2017towards}, they examined a set of computing problems written in ten well-known programming languages while observing the energy required when running each language.
-They also found that there are interesting situations in which slower languages use less energy than faster ones, even though fast languages are usually the ones that use the least energy.
+They also found exciting situations where slower languages use less energy than faster ones, even though fast languages usually use the least energy.
 Finally, they produced an energy efficiency rating of programming languages.
-The paper~\cite{bujnowski2020java} compared the energy, performance and database response time of web applications written in Java versus those written in Kotlin.
-They discovered that there is no statistically significant difference in CPU load between individual measurements( less than 2\%) 2.
+The paper~\cite{bujnowski2020java} compared the energy, performance, and database response time of web applications written in Java versus those written in Kotlin.
+They discovered no statistically significant difference in CPU load between individual measurements( less than 2\%) 2.
 However, Kotlin implementation has never earned the best results in any collection of measurements.
 
-Other works that have studied the impact of website technology on energy consumption,~\cite{philippot_characterization_2014,manotas_investigating_2013}.
+Other works have studied the impact of website technology on energy consumption,~\cite{philippot_characterization_2014,manotas_investigating_2013}.
 In their work~\cite{philippot_characterization_2014},\citeauthor{philippot_characterization_2014} measured the computer resources used during the loading of a website in a browser, such as memory utilization and energy consumption, for over 500 websites and proposed some best practices for developers.
@@
 \citeauthor{hasan_energy_2016} examined the energy consumption of multiple Java data structures, analyzing the bytecode and evaluating the change in energy consumption in various circumstances(research, insertion, deletion, etc.).
-They also simulated best- and worst-case energy usage scenarios in real-world production systems by replacing the LinkedList and ArrayList, and discovered that incorrect collection can result in a 300\% increase in energy consumption~\cite{hasan_energy_2016}.
+They also simulated best- and worst-case energy usage scenarios in real-world production systems by replacing the LinkedList and ArrayList and discovered that incorrect collection could result in a 300\% increase in energy consumption~\cite{hasan_energy_2016}.
 
 Other studies~\cite{longo_reducing_2019,kumar_energy_2017} looked into the energy use of Java primitive types, string operations, and the use of exceptions, loops, and arrays.
-\citeauthor{kumar_energy_2017}, for example, examined the energy usage of code snippets and micro-benchmarks and provided several conclusions, such as string concatenation would use less energy than StringBuilder and StringBuffer and static variables tends to consume 60\% more energy than instance variables.
+\citeauthor{kumar_energy_2017}, examined the energy usage of code snippets and micro-benchmarks and provided several conclusions, such as string concatenation would use less energy than StringBuilder and StringBuffer and static variables tends to consume 60\% more energy than instance variables.
 
@@
 \subsection{Energy Optimizations in the Execution Phase}
-The second part of this section will focus on energy optimization in the execution phase, where the goal is to optimize this energy consumption for already developed software without making changes to the source code.
-The goal is to set up and create an environment in which software can run with the least amount of energy.
-This could involve process scheduling, system tunning and so on.
+The second part of this section will focus on energy optimization in the execution phase.
+The goal is to optimize this energy consumption for already developed software without changing the source code.
+The goal is to set up and create an environment where software can run with the least amount of energy.
+This could involve process scheduling, system tunning, and so on.
 
-We begin with Aequitas~\cite{ribic2016aequitas}, a system that allows parallel applications to live on power domains that are co-managed (sharing the same CPU).
+We begin with Aequitas~\cite{ribic2016aequitas}, a system that allows parallel applications to live on co-managed power domains (sharing the same CPU).
 The technology is founded on the premise that coexisting programs can regard power-management hardware as a shared resource and collaborate on power management decisions.
-As a result, it accomplishes its purpose by scheduling these applications with a time slicing technique ( as an example round robin).
+As a result, it accomplishes its purpose by scheduling these applications with a time-slicing technique (for example, round robin).
 The authors claim that their strategy achieves a 12.9\% improvement while incurring only a 2.5\% performance cost.
 
-As for virtual machines, \citeauthor{kurpicz2016much}.
-analyzed the total energy consumption of a VM in a data center while emphasizing the static cost versus the dynamic one.
+As for virtual machines, \citeauthor{kurpicz2016much} analyzed the total energy consumption of a VM in a data center while emphasizing the fixed cost versus the dynamic one.
 In this paper~\cite{kurpicz2016much}, the authors introduced the transparent, reproducible, and predictive cost calculator model EPAVE for VM-based environments.
-The purpose of EPAVE is to provide the static cost of each VM on the server, which includes the air conditioning, power distribution, and the dynamic cost that is related to the VM activities.
+The purpose of EPAVE is to provide the static cost of each VM on the server, including the air conditioning, power distribution, and the dynamic cost related to the VM activities.
 
 The energy usage of virtual machine allocation and task placement has also been investigated~\cite{mishra_energy_2018}.In this article, The authors propose a method for mapping workloads to virtual machines and virtual machines to the physical ones (PMs) in an energy-efficient manner.
-To solve the problem of high heterogeneity of activities and resources, the jobs are categorized based on their resource requirements, and then the relevant VM is found, followed by the appropriate PM where the selected VM can be deployed Using a cloud simulator, the authors claimed The suggested technique saves energy by reducing the number of active PMs while also minimizing the makespan and task rejection rate.
+In order to solve the problem of high heterogeneity of activities and resources, the jobs are categorized based on their resource requirements.
+Then the relevant VM is found, followed by the appropriate PM where the selected VM can be deployed.
+Using a cloud simulator, the authors claimed that The suggested technique saves energy by reducing the number of active PMs while minimizing the makespan and task rejection rate.
 
-Besides the impact of the orchestration of virtual machine, some works have been targeting the runtime of specific programming languages.
+Besides the impact of virtual machine orchestration, some works have been targeting the runtime of specific programming languages.
 Using the TPC-DS benchmark, \citeauthor{chiba2018towards} investigated the influence of HOTSPOT \footnote{\url{https://openjdk.org/groups/hotspot/}} and J9\footnote{\url{https://www.eclipse.org/openj9/}} on the performance of SQL-on-Hadoop systems (SPARK and TEZ) to reveal a three-fold disadvantage that one JVM can have over the other~\cite{chiba2018towards}.
 \citeauthor{oi_power_2011} also compared the performance of HOTSPOT and J9.
-They demonstrated in their research~\cite{oi_power_2011} that the relative performance of a JVM is affected by the workload.
+They demonstrated in their research~\cite{oi_power_2011} that the workload affects the relative performance of a JVM.
 They discovered that HOTSPOT's performance ranged from 44\% to 289\% of J9, while its dynamic power consumption ranged from 2.7W to 7.2W, using the SPECjvm2008~\cite{shiv2009specjvm2008} benchmarks.
 
-As for python, \citeauthor{redondo_comprehensive_2015} compared the performance and memory usage of various Python implementations (CPython, Jython, IronPython, PyPy, and so on) using a 215 set of benchmarks to discover that Python2 performed better with short applications, while Python3 versions covered more tests due to compatibility and the fact that Python2 became obsolete.\cite{redondo_comprehensive_2015}.
+As for python, \citeauthor{redondo_comprehensive_2015} compared the performance and memory usage of various Python implementations (CPython, Jython, IronPython, PyPy, and so on) using a 215 set of benchmarks to discover that Python2 performed better with short applications.
+At the same time, Python3 versions covered more tests due to compatibility and the fact that Python2 became obsolete.\cite{redondo_comprehensive_2015}.
 
@@
 \section{Conclusion}
-As we have seen in the state of the art, many methods have been proposed to reduce the energy footprint of the the ICT, which they can be applied in different parts of the lifecycle of the program, from consumption to the execution.
-Furthermore, the execution phase took attention of many researchers because it is the part where the most energy is consumed.
+As we have seen in state of the art, many methods have been proposed to reduce the ICT's energy footprint, which can be applied in different parts of the program's lifecycle, from consumption to execution.
+Furthermore, the execution phase took the attention of many researchers because it is the part where the most energy is consumed.
 This thesis will focus on that aspect as well.
-However, unlike the most work that have been done on the hardware aspect, we will target the software impact on this energy, starting from the choice of the programming language up to how to tune some features of a frameworks in order to make the software consumes less energy.
-To do so we use the empirical approach due to its consistency for the moment.
-Unlike the performance, which is essentially related to the complexity of the algorithm, the energy consumption is more impacted by the hardware.
-Therefore, to optimize the energy consumption we choose a spiral method based on 3 phases:
+However, unlike most work done on the hardware aspect, we will target the software impact on this energy, starting from the choice of the programming language to how to tune some features of a framework to make the software consumes less energy.
+To do so, we use the empirical approach due to its consistency for the moment.
+Unlike the performance, which is essentially related to the algorithm's complexity, the energy consumption is more impacted by the hardware.
+Therefore, to optimize the energy consumption, we choose a spiral method based on 3 phases:
 \begin{enumerate}
-    \item execute the code,
+    \item executes the code,
     \item measure the program,
-    \item infer the guidelines.
+    \item infers the guidelines.
 \end{enumerate}
@@
 
-The aim of this work is to present a set of guidelines to create a benchmarking system to to measure the energy consumption of different programs.
+%  TODO : rephrase this to fit the order of the chapters 
+This work aims to present a set of guidelines to create a benchmarking system to measure different programs' energy consumption.
 After that, we will use this system to compare the energy consumption of different programming languages.
-We will extend the work of \citeauthor{pereira_energy_2017} to a closer distance to production environment by comparing a set of use cases.
-Starting by GRPC framework, and a set of Web Frameworks.
-Finally, we will discuss the impact of the execution environment on the energy consumption of two of the most famous programming languages Java and Python and present how can tunning the Virtual Machine can reduce the energy consumption.
+We will extend the work of \citeauthor{pereira_energy_2017} to a closer distance to the production environment by comparing a set of use cases.
+
+% we will  starting with the GRPC framework and a set of Web Frameworks.
+% Finally, we will discuss the impact of the execution environment on the energy consumption of two of the most famous programming languages Java and Python, and present how tunning the Virtual Machine can reduce energy consumption.

### 2022-10-01 chapters/benchmarking/draft.tex
proofread : experiment.tex

@@
 
-Second, one should not ignore tremendous progress in computer science, which have led to a rise in the number of obsolete results.
-Furthermore, when it comes to comparative research, the execution environment may have an impact on the study itself.
-Finally, between the exploratory experiment and the publication of the results, new candidates may have emerged and others may have changed.
+Second, one should not ignore tremendous progress in computer science, which has led to a rise in the number of obsolete results.
+Furthermore, when it comes to comparative research, the execution environment may impact the study itself.
+Finally, between the exploratory experiment and the publication of the results, new candidates may have emerged, and others may have changed.
 
-As a result, it is critical not only to ensure that the results can be "reproduced", so one can test their hypothesis in different environments, but also to provide room for new candidates.
+As a result, it is critical to ensure that the results can be \emph{reproduced}, so one can test their hypothesis in different environments and provide room for new candidates.
 In this section, we will address these issues and investigate several ways of encapsulating the systems-under-test to ensure experiment reproducibility in the context of energy consumption tests while providing the benefits and drawbacks of each strategy.
-Later in this section, we will also propose a protocol to ensure that the results are not only reproducible but extensible.
+Later in this section, we will propose a protocol to ensure that the results are reproducible and extensible.
 
@@
 \subsection{Virtual Machines}
-The very first choice should be to use \emph{Virtual Machines} (VM), which allow researchers to select the most appropriate tools, software, and operating system that they are most comfortable with without incurring the cost of changing the actual working environment, giving them more control over dependencies and the execution environment.
-Furthermore, adopting a VM addresses the \emph{replication crisis} since virtual images allow even the most sophisticated architecture to be simply replicated by instantiating a copy of the image.
+The first choice should be to use \emph{Virtual Machines} (VM).
+This technology enables researchers to choose the most appropriate tools, software, and operating system for their needs without incurring the cost of changing the working environment, giving them more control over dependencies and the execution environment.
+Furthermore, adopting a VM addresses the \emph{replication crisis} since virtual images allow even the most sophisticated architecture to be replicated by instantiating a copy of the image.
 
 This option, however, comes at a cost.
-Because of the hypervisor, software will be built on two kernels: one for the virtual machine (guest) and one for the host machine, resulting in a visible overhead and a negative influence on the performance of the system-under-test.
+Because of the hypervisor, the software will be built on two kernels: one for the virtual machine (guest) and one for the host machine, resulting in a visible overhead and a negative influence on the performance of the system-under-test.
 As a result, we cannot use VM for performance-related tests.
-Isolation is another drawback of VM: while this feature protects the experimental setting from unwanted interference from the outside world, it is possible that this interaction is required—-especially if the experiment is dependent on an external source such as energy monitors.
+Isolation is another drawback of VM: while this feature protects the experimental setting from unwanted interference from the outside world, this interaction may be required—-especially if the experiment is dependent on an external source such as energy monitors.
 
@@
     \item \textsf{Type\,1}: runs on the hardware directly.
-          It is primarily utilized by cloud providers where there is no host OS and only VMs that run on the open-source Xen or VMware ESX hypervisors.
+          It is primarily utilized by cloud providers with no host OS and only VMs that run on the open-source Xen or VMware ESX hypervisors.
 
     \item \textsf{Type\,2}: runs on top of the host operating system and is most commonly found on personal computers.
-VMware server and virtualBox are notable examples of this category, and most researchers' experiments use them.
-          However, because of the two operating systems, the applications are typically slower.
+VMware servers and VirtualBox are notable examples of this category, and most researchers' experiments use them.
+          However, the applications are typically slower because of the two operating systems.
 
-    \item \textsf{Containers}: run their operating systems on the host kernel rather than their own , which makes them smaller, faster, and more efficient in terms of hardware utilization.
+    \item \textsf{Containers}: run their operating systems on the host kernel rather than their own, which makes them smaller, faster, and more efficient in terms of hardware utilization.
 One can cite \emph{Docker}, \emph{Linux LXC}, or \emph{LXD}~\cite{abuabdo_virtualization_2019}.
@@
 \begin{enumerate}
-    \item As previously stated in literature~\cite{van2016power,morabito_power_2015}, we require a lightweight orchestrator to limit the overhead on energy usage of our experiments
-    \item We need to communicate with the host OS because we are using hardware sensors to measure the energy consumption.
+    \item As previously stated in literature~\cite{van2016power,morabito_power_2015}, we require a lightweight orchestrator to limit the overhead on the energy usage of our experiments
+    \item We need to communicate with the host OS because we are using hardware sensors to measure energy consumption.
 \end{enumerate}
@@
 \subsection{Docker \& Energy}
-Now that we have decided to use container technology to enclose our tests, what effect will this have on the amount of energy consumed by our tests ?
+Now that we have decided to use container technology to enclose our tests, what effect will this have on the amount of energy consumed by our tests?
 %%%%%%%%%
@@
 They contrasted the energy usage and execution time that resulted.
-The first step was to observe the effects of the orchestrator and the Docker deamon while there was no work to be done.
-Then, they use three benchmarks in their experiments: Wordpress, Reddis and PostgreSQL.
+The first step was to observe the effects of the orchestrator and the Docker daemon while there was no work to be done.
+Then, they used three benchmarks in their experiments: WordPress, Redis, and PostgreSQL.
 The values below show the system under test's energy consumption while it is idle.
-Docker has an overhead of roughly $1,000$  joules, as seen in Figure~\ref{fig:docker_idle} .
+Docker has an overhead of roughly $1,000$  joules, as seen in Figure~\ref{fig:docker_idle}.
 
@@
 
-Yet, as seen in Figure~\ref{fig:docker_reddis}, Docker increased the execution duration of the benchmark by 50 seconds, which led to a significant rise in the energy usage.
-According to the authors, this overhead is primarily caused by the Docker daemon and not by the fact that  the application is in a container.
+However, as seen in Figure~\ref{fig:docker_reddis}, Docker increased the execution duration of the benchmark by 50 seconds, which led to a significant rise in energy usage.
+According to the authors, the Docker daemon is primarily responsible for this overhead, not the fact that the application is running in a container.
 
-Furthermore, they calculated the cost of this extra energy, which was less than 0.15\$ in the worst-case scenario, which is insignificant in comparison to the benefits of Docker for isolation and reproducibility.
+Furthermore, they calculated the cost of this extra energy, which was less than 0.15\$ in the worst-case scenario, which is insignificant compared to the benefits of Docker for isolation and reproducibility.
 
-To summarize, Docker-based software tends to consume more energy since it takes longer to execute.
+To summarize, Docker-based software consumes more energy since it takes longer to execute.
 The execution of the Docker daemon causes an increase in average power consumption of only \textbf{2\,Watts}.
@@
 
-As can be seen, the introduction of the supervisor has increased the impact of Docker on energy that will be applied equitably to all experiments.
+As can be seen, the addition of the supervisor has increased Docker's impact on energy, which will be distributed equitably across all experiments.
 Therefore, when it comes to comparison analysis, it will mitigate its impact automatically.
 Furthermore, because we have access to the host hardware, we do not need to worry about capturing the SUT's energy use.
-As a result, we will use Docker to keep all of our tests separate, as it will make sure that they can be repeated in a clear and simple way.
+As a result, we will use Docker to keep all of our tests separate, ensuring that they can be repeated clearly and simply.
 
@@
 \subsubsection{Definition of Extention}
-With the rapid evolution of the software industry, Even ensuring the reproducibility of the same research won't be enough.
-Each day, new versions of the software are released, and new features are added.
+With the rapid evolution of the software industry, Even ensuring the reproducibility of the same research will not be enough.
+Each day, new software versions are released, and new features are added.
 Even more, the goal of the research might evolve.
@@
 
-One can extand their experiment through three axes :
+One can expand their experiment through three axes :
 \begin{itemize}
     \item \textbf{proposed solutions}: Where one will expand their research by including additional solutions and comparing them to earlier ones
-    \item \textbf{evaluation criteria}: This axe's objective is to broaden the evaluation criteria to incorporate additional measures like as performance, cost, and security, among others.
-    \item \textbf{benchmarks}: The purpose of this axe is to enlarge the benchmarks to include other ones in order to broaden the scope of the research.
+    \item \textbf{evaluation criteria}: This axe's objective is to broaden the evaluation criteria to incorporate additional measures like performance, cost, and security, among others.
+    \item \textbf{benchmarks}: This axe aims to enlarge the benchmarks to include others to broaden the research scope.
 \end{itemize}
@@
 We propose to enhance the benchmarking framework suggested by the \emph{Collaboratory on Experimental Evaluation of Software and Systems in Computer Science}\footnote{\url{http://evaluate.inf.usi.ch/}}.
-Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in Section~\ref{sec:benchmarking_reproducibility}, we suggest an abstract model to describe an empirical experiment.
-Figure~\ref{fig:benchmarkingprotocol} shows the proposed model while providing a comparaison between the existing solution and the proposed one.
+Instead of only presenting the four primary aspects of their guidelines that were mentioned previously in Section~\ref{sec:benchmarking_reproducibility}, we suggest an abstract model describing an empirical experiment.
+Figure~\ref{fig:benchmarkingprotocol} shows the proposed model while comparing the existing solution and the proposed one.
 
@@
 \paragraph{Context}
-The hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers have a better judgement on the results and to be able reproduce the experiment while diminishing the impact of  external factors .
+The hardware and the software configuration for the actual experiment, the purpose of this part is to provide extra information to help readers better judge the results and reproduce the experiment while diminishing the impact of external factors.
 
@@
 
-The core design of the experiment which is responsible for running the experiment regrardless of the context.
-In the experiment,it is  the only component that is allowed to interact with the SUT.
+The core design of the experiment is responsible for running the experiment regardless of the context.
+In the experiment, it is the only component that is allowed to interact with the SUT.
 The orchestrator provides three interfaces:
 \begin{itemize}
-    \item \emph{Workload interface} : provides a set of functions that should be implemented by the workload and called by each candidate in the experiment.This interface is responsible for extending the experiment with new benchmarks.
-    \item \emph{The Observer interface}: provides a set of metrics that are collected by the orchestrator during the experiment.
+    \item \emph{Workload interface}: provides a set of functions that the workload should implement to be called by each candidate in the experiment.
+This interface is responsible for extending the experiment with new benchmarks.
+    \item \emph{Observer interface}: provides a set of metrics that the orchestrator collects during the experiment.
 Implementing the observer interface allows the user to extend the experiment with new metrics.
-    \item \emph{The Candidate interface} : It provides a set of functions that should be implemented by the candidate and called by the orchestrator.
+    \item \emph{The Candidate interface}: It provides a set of functions that the candidate should implement to called  by the orchestrator.
 This interface is responsible for extending the experiment with new solutions.
@@
 \paragraph{Data}
-The row data collected collected by the experiment.
-In ourcase it will be provided by the observator, one ore more observators can be used to collect different metrics.
-The purpose of this data is to ensure the \emph{replication} of the experiment which will allow the reader to perform extra analysis without rexucuting the experiment.
+The row data collected by the experiment.
+In our case, it will be provided by the \emph{observator}.
+One or more observators can be used to collect different metrics.
+This data aims to ensure the \emph{replication}, allowing the reader to perform extra analysis without executing the experiment a second time.
 
 \paragraph{Analysis}
-The final part, that should provide the set of methodes and functions used to do the empirical analysis in order to answer the research questions.
+The final part should provide the set of methods and functions used to do the empirical analysis to answer the research questions.
 
@@
 In this section, we addressed the first challenge of empirical research, which is the \textbf{reproducibility} of the experiments.
-We started by listing the different options for encapsulating the system-under-test.
-Then, we have shown that the use of VMs is not suitable for performance-related tests nor energy-related ones, while containers are a good alternative.
-Laterm, we discussed the pros and cons of using Docker for experiments related to energy consumption and have shown that it has a constant overhead which is self-mitigated when it comes to comparison analysis.
+We started by listing the different options for encapsulating the system under test.
+Then, we have shown that using VMs is unsuitable for performance or energy-related tests, while containers are a good alternative.
+Later, we discussed the benefits and drawbacks of using Docker for energy consumption experiments.
+We have demonstrated that it has a constant overhead that is self-mitigated when compared.
 Lastly, we have proposed an addition to the benchmarking framework that would allow the experiment to be extended along the three axes already mentioned.

### 2022-10-01 chapters/benchmarking/experiment.tex
proofread : experiment.tex

@@
 
-We used the platform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708}, which is a large, flexible testbed for experiment-driven research that is spread out across all of France, to meet these needs.
-Grid5000 has several clusters that are made up of 4 to 124 identical machines with different configurations for each cluster.
-For our experiment, we looked at four groups.
-Our main criterion was the CPU configuration.
-Table~\ref{table:g5k}  below, presents a description of the four clusters that have been chosen for our experiments.
+We used the platform Grid5000 (G5K)~\cite{grid5000,margery:hal-00965708}, a large, flexible testbed for experiment-driven research spread across France, to meet these needs.
+Grid5000 has several clusters comprising 4 to 124 identical machines with different configurations for each cluster.
+We looked at four groups for our experiment, and our main criterion was the CPU configuration.
+Table~\ref{table:g5k} below presents a description of the four clusters that have been chosen for our experiments.
 
@@
 As most of the nodes are equipped with two sockets (physical processors), we use the acronym \textsf{CPU} or \textsf{socket} to designate one of the two sockets and \textsf{PU} for the \emph{processing unit}.
-For our study, we consider hyper-threads as distinct \textbf{PUs}.
+Our study considers hyper-threads as distinct \textbf{PUs}.
 Figure~\ref{fig:topo} shows the detailed topology of a node in the cluster \textsf{Dahu} as an example.
@@
 
-To meet these criteria, we looked at the "state of the art" and found the most common benchmarks used to test the performance of hardware. We then chose the ones that could be scaled up.
+To meet these criteria, we looked at the "state of the art" and found the most common benchmarks used to test hardware performance. We then chose the ones that could be scaled up.
 Our candidate is \texttt{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925}: one of the most used benchmarks for \emph{HPC}.
@@
 Our metric for the accuracy of the test is the Standard deviation aka \textbf{STD} of the energy consumption. Therefore whether the tests consume more or less energy is out of our scope.
-To study this variation we need first a tool to measure the energy consumption.
+We first need a tool to measure energy consumption to study this variation.
 For this we used \textsc{PowerAPI}~\cite{DBLP:journals/jss/ColmantRKSFS18}, which is a power monitoring toolkit that is based on \emph{Intel Running Average Power Limit} (RAPL) \cite{Khan:2018:RAE:3199681.3177754}.
@@
 The energy sensor collects RAPL reports and stores them in a remote \textsc{MongoDB} instance, allowing us to perform \emph{post-mortem} analysis in a dedicated environment.
-Using Docker makes the deployment process easier on the one hand, and provides us with a built-in control group encapsulation of the conducted tests on the other hand.
-This allows \textsc{PowerAPI} to measure all the running containers, even the RAPL sensor consumption, as it is isolated in a container.
+Using Docker makes the deployment process easier on the one hand and provides us with a built-in control group encapsulation of the conducted tests on the other hand.
+Docker allows \textsc{PowerAPI} to measure all the running containers, even the RAPL sensor consumption, as it is isolated in a container.
 
 Every experiment is conducted on $100$ iterations, on multiple nodes and using the 3 NPB benchmarks we mentioned, with a warmup phase of 10 iterations for each experiment.
-In most cases, we were seeking to evaluate the \emph{STandard Deviation} (STD), which is the most representative factor of the energy variation.
-We tried to be very careful, while running our experiments, not to fall into the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
+In most cases, we sought to evaluate the \emph{Standard Deviation} (STD), which is the most representative factor of the energy variation.
+While running our experiments, we tried not to fall into the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
 As we study the STD difference of measurements we observed from empirical experiments, we use the bootstrap method~\cite{efron2000bootstrap} to randomly build multiple subsets of data from the original dataset, and we draw the STD density of those sets, as illustrated in Figure~\ref{fig:docker}.
-Given the space constraints, we report on aggregated results for nodes, benchmarks and workloads.
+Given the space constraints, we report on aggregated results for nodes, benchmarks, and workloads.
 However, the raw data we collected is available through the public repository we published.\footnote{\url{https://github.com/anonymous-data/Energy-Variation}}
 We believe this can help to achieve better and more reliable comparisons.
-We mainly consider 3 different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium and high CPU usage when analyzing the effect of the studied parameters, respectively.
+We mainly consider three different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium, and high CPU usage when analyzing the effect of the studied parameters, respectively.
 These workloads reflect the ratio of used PU count to the total available PU.
@@
 
-To answer this question we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on 3 nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
+To answer this question, we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on three nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
 Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running the same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
@@
 \subsubsection{\textsc{RQ}~1: Benchmarking Protocol}
-To achieve a robust and reproducible experiment, practitioners often tend to repeat their tests multiple times, in order to analyze the related performance indicators, such as execution time, memory consumption or energy consumption.
+To achieve a robust and reproducible experiment, practitioners often repeat their tests multiple times to analyze the related performance indicators, such as execution time, memory consumption, or energy consumption.
 We, therefore, aim to study the benchmarking protocol to identify how to efficiently iterate the tests to capture a trustable energy consumption evaluation.
@@
 In this first experiment, we investigate if changing the testing protocol affects the energy variation.
-To achieve this, we considered 3 execution modes:
+To achieve this, we considered three execution modes:
 In the \textsf{"normal"} mode, we iteratively run the benchmark 100 times without any extra command, while the \textsf{"sleep"} mode suspends the execution script for 60\,seconds between iterations.
-Finally, the \textsf{"reboot"} mode automatically reboots the machine after each iteration.
-The difference between the \textsf{normal} and \textsf{sleep} modes intends to highlight that the CPU needs some rest before starting another iteration, especially for an intense workload.
-Putting the CPU into sleep for several seconds could give it some time to reach a lower frequency state or/and reduce its temperature, which could have an impact on the energy variation.
-The \textsf{reboot} mode, on the other hand, is the most straightforward way to reset the machine state after every iteration.
-It could also be beneficial to reset the CPU frequency and temperature, the stored data, the cache or the CPU registries.
-However, the reboot task takes a considerable amount of time, so rebooting the node after every single operation is not the fastest nor the most eco-friendly solution, but it deserves to be checked to investigate if it effectively enhances the overall energy variation or not.
+Finally, after each iteration, the \textsf{"reboot"} mode automatically reboots the machine.
+The difference between the \textsf{normal} and \textsf{sleep} modes highlights that the CPU needs rest before starting another iteration, especially for an intense workload.
+Putting the CPU into sleep for several seconds could give it time to reach a lower frequency state or/and reduce its temperature, which could impact the energy variation.
+On the other hand, the \textsf{reboot} mode is the most straightforward way to reset the machine state after every iteration.
+It could also be beneficial to reset the CPU frequency and temperature, the stored data, the cache, or the CPU registries.
+However, the reboot task takes a considerable amount of time, so rebooting the node after every single operation is not the fastest nor the most eco-friendly solution.
+However, it deserves to be checked to investigate if it effectively enhances the overall energy variation or not.
 
@@
 Table~\ref{table:g5k}) for different workloads.
-We note that the results have been executed with different datasets sizes (\textsf{B}, \textsf{C} and \textsf{D} for a single process, 50\,\% and 100\,\% respectively) to remedy to the brief execution times at high workloads for small datasets.
-This justifies the scale differences in reported energy consumption between the 3 modes in Figure~\ref{fig:running-process}.
-As one can observe, picking one of these strategies does not have a strong impact on the energy variation for most workloads.
-In fact, all the strategies seem to exhibit the same variation with all the workloads we considered---\emph{i.e.}, the STD is tightly close between the three modes.
-The only exception is the \textsf{reboot} mode at 100\,\% load, where the STD is 150\,\% times worst, due to an important amount of outliers.
+We note that the results have been executed with different datasets sizes (\textsf{B}, \textsf{C} and \textsf{D} for a single process, 50\,\% and 100\,\% respectively) to remedy the brief execution times at high workloads for small datasets.
+This justifies the scale differences in reported energy consumption between the three modes in Figure~\ref{fig:running-process}.
+One can observe that picking one of these strategies does not strongly impact the energy variation for most workloads.
+All the strategies exhibit the same variation with all the workloads we considered---\emph{i.e.}, the STD is tightly close between the three modes.
+The only exception is the \textsf{reboot} mode at 100\,\% load, where the STD is 150\,\% times worst due to a substantial amount of outliers.
 This goes against our expectations, even when setting a warm-up time after reboot to stabilize the OS.
@@
 
-In Figure~\ref{fig:reboot}, we study the standard deviation of the three modes by constituting $5,000$ random 30-iterations sets from the previous executions set and we compute the STD in each case, considering mainly the 100\,\% workload as the STD was 150\,\% higher for the \textsf{reboot} mode with that load.
+In Figure~\ref{fig:reboot}, we study the standard deviation of the three modes by constituting $5,000$ random 30-iterations sets from the previous executions set, and we compute the STD in each case, considering mainly the 100\,\% workload as the STD was 150\,\% higher for the \textsf{reboot} mode with that load.
 We can observe that the considerable amount of outliers in the \textsf{reboot} mode is not negligible, as the STD density is higher than the two other modes.
@@
 
-    To answer \textsc{RQ~1}, we conclude that the benchmarking protocol \textbf{partially affects} the energy variation, as highlighted by the \textsf{reboot} mode results in high workloads.
+    To answer \textsc{RQ~1}, we conclude that the benchmarking protocol \textbf{partially affects} the energy variation, as highlighted by the \textsf{reboot} mode, results in high workloads.
 
@@
 \subsubsection{\textsc{RQ}~2: Processor Features}
-The C-states provide the ability to switch the CPU between more or less consuming states upon activities.
-Turning the C-states on or off have been subject of many discussions~\cite{5463056}, because of its dynamic frequency mechanism but, to the best of our knowledge, there have been no fully conducted C-states behavior analysis on CPU energy variation.
+The C-states allow switching the CPU between more or less consuming states upon activities.
+Turning the C-states on or off have been the subject of many discussions~\cite{5463056} because of its dynamic frequency mechanism, but, to the best of our knowledge, there has been no entirely conducted C-states behavior analysis on CPU energy variation.
 
@@
 Figure~\ref{fig:c-states} depicts the results of the experiments we executed on three nodes of the cluster \textsf{Dahu}.
-On each node, we ran the same set of benchmarks with two modes: \textsf{C-states on}, which is the default mode, and \textsf{C-states off}.
+On each node, we ran the same benchmarks with two modes: \textsf{C-states on}, which is the default mode, and \textsf{C-states off}.
 Each iteration includes 100 executions of the same benchmark at a given workload, with three workload levels.
@@
 In this case, only one CPU core is used among $2\times16$ physical cores.
-The other cores are switched to a low consumption state when C-states are on, the switching operation causes an important energy consumption difference between the cores, and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
+The other cores are switched to a low consumption state when C-states are on.
+The switching operation causes a vital energy consumption difference between the cores and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
 On the other hand, switching off the C-states would keep all the cores---even the unused ones---at a high-frequency usage.
@@
 
-At a 100\,\% workload, disabling the C-states seems to not affect the total energy consumption nor its variation.
-All the cores are used at 100\,\% and the C-states module would have no effect, as the cores are not idle.
+At a 100\,\% workload, disabling the C-states seems not to affect the total energy consumption nor its variation.
+All the cores are used at 100\,\%, and the C-states module would have no effect, as the cores are not idle.
 The same reason would apply for the 50\,\% load, as the hyper-threading is active on all cores, thus causing the usage of most of them.
 For single process workloads, disabling the C-states causes the process to consume 50\,\% more energy as reported in Figure~\ref{fig:c-states}, but reduces the variation by 5~times compared to the \textsf{C-states on} mode.
-This leads to mainly two questions: Can a process pinning method reduce/increase the energy variation? And, how does the energy consumption variation evolve at different PU usage levels?
+This leads to two questions: Can a process pinning method reduce/increase the energy variation? Moreover, how does the energy consumption variation evolve at different PU usage levels?
 
@@
 In this strategy, we only use the core PU and not the hyper-threads PU, so every process would not share its core usage (all the cores are being used).
-The third strategy (\textsf{S3}) consists also of splitting the workload between the two sockets, but considering the usage of the hyper-threads on each core---\emph{i.e.}, half of the cores are being used over the two CPU.
+The third strategy (\textsf{S3}) also involves splitting the workload between the two sockets, but with the hyper-threads on each core used, i.e., half of the cores are used across the two CPUs.
+
 Figure~\ref{fig:cores-pinning} reports on the energy consumption of the three strategies when running the benchmark \texttt{CG} on the cluster \textsf{Dahu}.
@@
 We argue that the reason is related to the isolation of every process on a single physical core, reducing the context switch operations.
-In the first and third strategies, $32$ processes are being scheduled on $16$ physical cores using the hyper-threads PU, which will introduce more context switching, and thus more energy consumption.
+In the first and third strategies, $32$ processes are being scheduled on $16$ physical cores using the hyper-threads PU, which will introduce more context switching, thus, more energy consumption.
 
@@
 
-We note that even if the first and third strategies are very similar (both use hyper-threads, but only on one CPU for the first and on two CPU for the third), the gap between them is considerable variation-wise, as the variation is 30 times lower in the first strategy ($STD_{S1}=116 mJ$,$STD_{S3}=3,452 mJ$).
-This shows that the usage of the hyper-threads technology is not the main reason behind the variation, the first strategy has even less variation than the second one and still uses the hyper-threading.
+Even though the first and third strategies are very similar (both use hyper-threads, but only on one CPU for the first and two CPUs for the third), the variation gap between them is significant, as the variation in the first strategy is 30 times lower ($STD_{S1}=116 mJ$,$STD_{S3}=3,452 mJ$).
+This shows that the hyper-threads technology is not the main reason behind the variation; the first strategy has even less variation than the second one and still uses the hyper-threading.
 
 The reason for the \textsf{S1} low energy consumption is that one of the two sockets is idle and will likely be in a lower power P-state, even with the disabled C-states.
-The \textsf{S2} case reports on low energy consumption because by distributing the threads across all the cores, it completes the task faster than in the other cases.
+The \textsf{S2} case reports low energy consumption because distributing the threads across all the cores completes the task faster than in the other cases.
 Hence, it consumes less energy.
-The \textsf{S3} is a highly consuming strategy because both sockets are being used, but only half the cores are active.
+The \textsf{S3} is a highly consuming strategy because both sockets are used, but only half the cores are active.
 This means that we pay the energy cost for both sockets being operational and for the experiments taking longer to run because of the recurrent context switching.
 
-Our hypothesis regarding the worst results that we observed when using the third strategy is the recurrent context switching, added to the OS scheduling that could reschedule processes from a socket to another, which invalids the cache usage as a process can not take profit of the socket local L3 cache when it moves from a CPU to another (cf.
+Our hypothesis for the worst results observed when using the third strategy is that recurrent context switching, which is added to the OS scheduling and can reschedule processes from one socket to another, invalidates cache usage because a process cannot benefit from the socket local L3 cache when it moves from one CPU to another.
+(cf.
 Figure~\ref{fig:topo}).
@@
 On the other hand, the strategy \textsf{S3} is the worst, where the energy consumption can be up to 5 times higher than the strategy \textsf{S2} ($Mean_{S2}=4016 mJ$, $Mean_{S3}=21645 mJ$) and the variation is much worst (30 times compared to the first strategy).
-These results allow us to have a better understanding of the different processes-to-PU pinning strategies, where isolating the workload on a single CPU is the best strategy.
-Using the hyper-threads PU on multiple sockets seems to be a bad recommendation, while keeping the hyper-threading enabled on the machine is not problematic, as long as the processes are correctly pinned on the PU.
-Our experiments show that running one hyper-thread per core is not always the best to do, at the opposite of the claims of~\cite{marathe_empirical_2017_m}.
+These results allow us to understand better the different processes-to-PU pinning strategies, where isolating the workload on a single CPU is the best strategy.
+Using the hyper-threads PU on multiple sockets seems to be a bad recommendation.
+Keeping the hyper-threading enabled on the machine is not problematic as long as the processes are correctly pinned on the PU.
+Our experiments show that running one hyper-thread per core is not always the best to do, opposite the claims of~\cite{marathe_empirical_2017_m}.
 
@@
 Figure~\ref{fig:process-tresh} compares the aggregated energy variation when the C-states are on and off using 2, 4 and 8 processes for the benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP}.
-This figure confirms that disabling the CPU C-states does not decrease the variation for all the workloads, as we can clearly observe, the variation is increasing along with the number of processes.
+This figure confirms that disabling the CPU C-states does not decrease the variation for all the workloads; as we can clearly observe, the variation is increasing along with the number of processes.
 When running only 2 processes, turning off the C-states reduces the STD up to 6~times, but consumes 20\,\% more energy ($Mean_{C-states-on}=10,334 mJ$, $Mean_{C-states-off}=12,594 mJ$).
-This variation is 4~times lower when running 4 processes and almost equal to the \textsf{C-states on} mode when running 8 processes.
+This variation is 4~times lower when running four processes and almost equal to the \textsf{C-states on} mode when running eight processes.
 In fact, running more processes implies using more CPU cores, which reduces the idle cores count, so the cores will more likely stay at a higher consumption state even if the C-states mechanism is on.
 
-In our case, using 4 PU reduces the variation by 4 times and consumes almost the same energy as keeping the C-states mechanism on ($Mean_{C-states-on}=7,048 mJ$, $Mean_{C-states-off}=7,119 mJ$).
-This case would be the closest to reality as we do not want to increase the energy consumption while reducing the variation, but using a lower number of PU still results in less variation, even if it increases the overall energy consumption.
+In our case, using 4 PU reduces the variation by four times and consumes almost the same energy as keeping the C-states mechanism on ($Mean_{C-states-on}=7,048 mJ$, $Mean_{C-states-off}=7,119 mJ$).
+This case would be the closest to reality as we do not want to increase energy consumption while reducing the variation.
+However, using a lower number of PU still results in less variation, even if it increases the overall energy consumption.
 
@@
 We note that disabling the C-states is not recommended in production environments, as it introduces extra energy consumption for low workloads (around 50\,\% in our case for a single process job).
-However, our goal is not to optimize the energy consumption, but to minimize the energy variation.
+However, our goal is not to optimize energy consumption but to minimize the energy variation.
 Thus, disabling the C-states is very important to stabilize the measurements in some cases when the variation matters the most.
-Comparing the energy consumption of two algorithms or two versions of a software system is an example of a use case benefiting from this recommendation.
+Comparing the energy consumption of two algorithms or two software system versions is an example of a use case benefiting from this recommendation.
 
 \paragraph{Turbo Boost}
-The Turbo Boost---also known as \emph{Dynamic Overclocking}---is a feature that has been incorporated in Intel CPU since the Sandy~Bridge micro-architecture, and is now widely available on all of the Core\,i5, Core\,i7, Core\,i9 and Xeon series.
-It automatically raises some of the CPU cores operating frequency for short periods, and thus boosts performances under specific constraints.
-When demanding tasks are running, the operating system decides on using the highest performance state of the processor.
+The Turbo Boost---also known as \emph{Dynamic Overclocking}---is a feature that has been incorporated in Intel CPU since the Sandy~Bridge micro-architecture, and is now widely available on all of the Core\,i5, Core\,i7, Core\,i9, and Xeon series.
+It automatically raises some of the CPU cores operating frequency for short periods, thus boosting performance under specific constraints.
+When demanding tasks run, the operating system decides to use the processor's highest performance state.
 
-Disabling or enabling the Turbo~Boost has a direct impact on the CPU frequency behavior, as enabling it allows the CPU to reach higher frequencies to execute some tasks for a short period.
-However, its usage does not have a trivial impact on the energy variation.
+Disabling or enabling the Turbo~Boost  impacts the CPU frequency behavior directly, as enabling it allows the CPU to reach higher frequencies to execute some tasks for a short period.
+However, its usage does not significantly impact the energy variation.
 Acun~\emph{et~al.}~\cite{acun_variation_2016} tried to track the Turbo~Boost impact on the Ivy~Bridge and the Sandy~Bridge architectures.
 They concluded that it is one of the main responsible for the energy variation, as it increases the variation from 1\,\% to 16\,\%.
-In our study, we included a Turbo~Boost experiment in our testbed, to check this property on the recent Xeon Gold processors, covering various workloads.
+In our study, we included a Turbo~Boost experiment in our testbed to check this property on the recent Xeon Gold processors, covering various workloads.
 
-The experiment we conducted showed that disabling the Turbo Boost does not exhibit any considerable positive or negative effect on the energy variation.
-Table~\ref{table:turboboost} compares the STD when enabling/disabling the Turbo~Boost, where the columns are a combination of workload and benchmark.
-We only got some minor measurement differences when switching on and off the Turbo~Boost, and were in favor or against the usage of the Turbo~Boost while repeating tests, considering multiple nodes and benchmarks.
-This behavior is mainly related to the \emph{thermal design power} (TDP), especially at high workloads executions.
-When a CPU is used at its maximum capacity, the cores would be heating up very fast and would hit the maximum TDP limit.
+The experiment showed that disabling the Turbo Boost does not exhibit any considerable positive or negative effect on the energy variation.
+Table~\ref{table:turboboost} compares the STD when enabling/disabling the Turbo~Boost, where the columns combine workload and benchmark.
+We only had minor measurement differences when switching on and off the Turbo~Boost. We favored or against using the Turbo~Boost while repeating tests, considering multiple nodes and benchmarks.
+This behavior is mainly related to the \emph{thermal design power} (TDP), especially at high workloads.
+When a CPU is used at its maximum capacity, the cores would heat up very fast and hit the maximum TDP limit.
 In this case, the Turbo~Boost cannot offer more power to the CPU because of the CPU thermal restrictions.
 At lower workloads, the tests we conducted proved that the Turbo~Boost is not one of the main reasons for the energy variation.
-The variation difference is barely noticeable when disabling the Turbo~Boost, which cannot be considered as a result of the OS activity and the measurement error margin.
+The variation difference is barely noticeable when disabling the Turbo~Boost, which cannot be considered due to the OS activity and the measurement error margin.
 We cannot affirm that the Turbo~Boost does not have an impact on all the CPUs, as we only tested on two recent Xeon CPUs (clusters \textsf{Chetemi} and \textsf{Dahu}).
@@
 The aggregated results, illustrated in Figure~\ref{fig:os-idle}, depict that the idle energy variation is up to 140\,\% worst than when running a job, even if it consumes 120\,\% less energy ($Mean_{Job}=8,746 mJ$, $Mean_{Idle}=3,927 mJ$).
-In fact, for the three nodes, randomly picked from the cluster \textsf{Dahu}, the idle variation is way more important than when a test was running, even if it is a single process test on a $32$-cores node.
-This result shows that OS idle consumption varies widely, due to the lack of activity and the different CPU frequency states, but it does not mean that this variation is the main responsible for the overall energy variation.
+In fact, for the three nodes randomly picked from the cluster \textsf{Dahu}, the idle variation is way more important than when a test was running, even if it is a single process test on a $32$-cores node.
+This result shows that OS idle consumption varies widely due to the lack of activity and the different CPU frequency states.
+However, it does not mean that this variation is responsible for the overall energy variation.
 The OS behaves differently when a job is running, even if the amount of available cores is more than enough for the OS to keep its idle behavior when running a single process.
@@
 
-Inspecting the OS idle energy variation is not sufficient to relate the energy variation to the active job.
+Inspecting the OS idle energy variation is insufficient to relate the energy variation to the active job.
 The OS can behave differently regarding resource usage when running a task.
@@
 We calculated the Pearson correlation~\cite{ref1} of the energy consumption and variation between global RAPL and \textsc{PowerAPI}, as illustrated in Figure~\ref{fig:ev_correlation}.
-The job energy consumption and variation are strongly correlated with the global energy consumption and variation with the coefficients 93.6\,\% and 85.3\,\%, respectively.
-However, this does not completely exclude the OS activity, especially if the jobs have tight interaction with the OS through the signals and system calls.
+The job energy consumption and variation are strongly correlated with the global energy consumption and variation with the coefficients 93.6\,\%, and 85.3\,\%, respectively.
+However, this does not entirely exclude the OS activity, especially if the jobs have tight interaction with the OS through the signals and system calls.
 This brings a new question on whether applying extra-tuning on a minimal OS would reduce the variation.
-As well as what is the effect of the Meltdown security patch---that is known to be causing some performance degradation~\cite{Kocher2018spectre,Lipp2018meltdown}---on the energy variation?
+As well as the Meltdown security patch --- known to be causing some performance degradation~\cite{Kocher2018spectre,Lipp2018meltdown}--- affects the energy variation?
 
@@
 Even using a minimal version of Debian Linux, we could list many OS running services and processes that could be disabled/stopped without impacting the test execution.
-This extra-tuning may not be the same depending on the nature of the test or the OS.
+This extra-tuning may not be the same, depending on the nature of the test or the OS.
 Thus, we conducted a test with a deeply-tuned OS version.
-We disabled all the services/processes that are not essential to the OS/test running, including the OS networking interfaces and logging modules, and we only kept the strict minimum required for the experiment's execution.
+We disabled all the services/processes that are not essential to the OS/test running, including the OS networking interfaces and logging modules.
+We only kept the strict minimum required for the experiment's execution.
 Table~\ref{table:osmin} reports on the aggregated results for running single process measurements with the benchmarks \textsf{CG}, \textsf{LU} and \textsf{EP}, on three servers of the cluster \textsf{Dahu}, before and after tuning the OS.
@@
 We argue that this variation is not substantial, as it is not stable from one node to another.
-Moreover, 10\,\% of the variation is not a representative difference, due to many factors that can affect it as the CPU temperature or the measurement errors.
+Moreover, 10\,\% of the variation is not a representative difference due to many factors that can affect it as the CPU temperature or the measurement errors.
 
@@
 \paragraph{Speculative Executions}
-Meltdown and Spectre are two of the most famous hardware vulnerabilities discovered in 2018, and exploiting them allows a malicious process to access others processes' data that is supposed to be private~\cite{Kocher2018spectre,Lipp2018meltdown}.
+Meltdown and Spectre are two of the most famous hardware vulnerabilities discovered in 2018, and exploiting them allows a malicious process to access other processes' data that is supposed to be private~\cite{Kocher2018spectre,Lipp2018meltdown}.
 They both exploit the speculative execution technique where a process anticipates some upcoming tasks, which are not guaranteed to be executed when extra resources are available, and revert those changes if not.
-Some OS-level patches had been applied to prevent/reduce the criticality of these vulnerabilities.
+Some OS-level patches have been applied to prevent/reduce the criticality of these vulnerabilities.
 On the Linux kernel, the patch has been automatically applied since version 4.14.12.
@@
 Nikolay~\emph{et~al.} have studied in~\cite{DBLP:journals/corr/abs-1801-04329} the impact of patching the OS on the performance.
-The results showed that the overall performance decrease is around 2--3\,\% for most of the benchmarks and real-world applications, only some specific functions can meet a high-performance decrease.
+The results showed that the overall performance decrease is around 2--3\,\% for most of the benchmarks and real-world applications; only some specific functions can meet a high-performance decrease.
 In our study, we are interested in the applied patch's impact on the energy variation, as the performance decrease could mean an energy consumption increase.
@@
 The best case to consider is the benchmark \textsf{LU} where the energy variation is less than 10\,\% when we disable the security patch, but this difference is still moderate.
-The little performance difference discussed in~\cite{Kocher2018spectre,Lipp2018meltdown} may only be responsible for a small variation, which will be absorbed through the measurement tools and external noise error margin in most cases.
+The little performance difference discussed in~\cite{Kocher2018spectre,Lipp2018meltdown} might only be responsible for a slight variation, which will be absorbed through the measurement tools and external noise error margin in most cases.
 
@@
 Intel microprocessors have noticeably evolved during these last 20 years.
-Most of the new CPUs come with new enhancements to the chip density, the maximum Frequency or some optimization features like the C-states or the Turbo~Boost.
-This active evolution caused different generations of CPUs can handle a task differently.
-The aim of this experiment is not to justify the evolution of the variation across CPU versions/generations, but to observe if the user can choose the best node to execute her experiments.
+Most new CPUs have enhancements to the chip density, the maximum Frequency, or some optimization features like the C-states or the Turbo~Boost.
+This dynamic evolution caused different generations of CPUs can handle a task differently.
+This experiment aims not to justify the evolution of the variation across CPU versions/generations but to observe if the user can choose the best node to execute her experiments.
 Previous papers have discussed the evolution of the energy consumption variation across CPU generations and concluded that the variation is getting higher with the latest CPU generations~\cite{wang_experimental_nodate,marathe_empirical_2017_m}, which makes measurements stability even worse.
 In this experiment, we, therefore, compare four different generations of CPU to evaluate the energy variation for each CPU and its correlation with the generation.
-Table~\ref{table:cpus} indicates the characteristics of each of the tested CPUs.
+Table~\ref{table:cpus} indicates the characteristics of each tested CPU.
 
@@
 \begin{mdframed}[skipabove=\topsep,skipbelow=\topsep]
-    We conclude on \textbf{affirming \textsc{RQ}~4}, as selecting the right CPU can help to get less variations.
+    We conclude on \textbf{affirming \textsc{RQ}~4}, as selecting the right CPU can help get fewer variations.
 \end{mdframed}
@@
 \subsection{Experimental Guidelines}
-To summarize our experiments, we provide some experimental guidelines in Table~\ref{table:guidelines}, based on the multiple experiments and analyses we did.
+To summarize our experiments, we provide some practical guidelines in Table~\ref{table:guidelines}, based on the multiple experiments and analyses we did.
 These guidelines constitute a set of minimal requirements or best practices, depending on the workload and the criticality of the energy measurement precision.
-It, therefore, intends to help practitioners in taming the energy variation on the selected CPU, and conduct the experiments with the least variations.
+Therefore, it intends to help practitioners tame the energy variation on the selected CPU and conduct experiments with the tiniest variations.
 
@@
 Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and their variation reduction at low workloads.
-However, it also lists some new factors that we identified along the analysis we conducted in Section, such as the results related to the OS or the reboot mode.
-Some of the guidelines are more useful/efficient for specific workloads, as shown in our experiments.
-Thus, qualifying the workload before conducting the experiments can help in choosing the proper guidelines to apply.
-Other studied factors are not been mentioned in the guidelines, like the Turbo~Boost or the Speculative execution, due to the small effect that has been observed in our study.
+However, it also lists some new factors we identified along the analysis we conducted in Section, such as the results related to the OS or the reboot mode.
+Some guidelines are more useful/efficient for specific workloads, as shown in our experiments.
+Thus, qualifying the workload before conducting the experiments can help choose the proper guidelines.
+Other studied factors, such as Turbo~Boost or Speculative execution, are not mentioned in the guidelines due to the small effect observed in our study.
 
-To validate the accuracy of our guidelines among a varied set of benchmarks on one hand, and their effect on the variation between identical machines, on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
-Half of these experiments have been performed at a 50\,\% workload and the other half on single process jobs.
+To validate the accuracy of our guidelines among a varied set of benchmarks on the one hand and their effect on the variation between identical machines, on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
+Half of these experiments have been performed at a 50\,\% workload, and the other half on single process jobs.
 The choice of these two workloads is related to the optimization guidelines that are mainly effective at low and medium workloads.
@@
 This variation can be up to 30\,\% in modern CPU~\cite{wang_experimental_nodate}.
-However, taming the intra-node variation is a good strategy to identify more relevant mediums and medians, and then perform accurate comparisons between the nodes variation.
+However, taming the intra-node variation is a good strategy for identifying more relevant mediums and medians and then accurately comparing the nodes' variation.
 Even though using the same node is always better, it avoids the extra inter-nodes variation and thus improves the stability of measurements.
@@
 \subsection{Threats to Validity}
-A number of issues affect the validity of our work.
-For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPUs, but still adds an important overhead if we adopt a sampling at high frequency.
-The other fine-grained tool we used for measurements is \textsc{PowerAPI}.
+Several issues affect the validity of our work.
+For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPUs but still adds a significant overhead if we adopt a sampling at high frequency.
+The other fine-grained measurement tool we used is \textsc{PowerAPI}.
 It allows measuring the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
@@
 
-The Intel CPU chip manufacturing process and the material's micro-heterogeneity are one of the biggest issues, as we cannot track or justify some of the energy variations between identical CPUs or cores.
+The Intel CPU chip manufacturing process and the material's micro-heterogeneity are among the most significant issues, as we cannot track or justify some energy variations between identical CPUs or cores.
 These CPU/cores might handle frequencies and temperature differently and behave consequently.

### 2022-10-01 chapters/programming_languages/rpc/rpc.tex
add : correlation table
update : typos in thechempower
add : rpc comclusion

@@
 %  TODO : REname the frameworks based on their dockerfile
+
 Figure~\ref{fig:rps_clients} presents the number of requests per second for each implementation. The heatmap is colored in a logarithmic scale to visualize the results better.
@@
 
-\paragraph{Synthesis}: 
-Depending on the number of clients, most frameworks exhibit two distinct behaviors.
+% \paragraph{Synthesis}:
+% Depending on the number of clients, most frameworks exhibit two distinct behaviors.
 When dealing with a small number of clients, frameworks, regardless of programming category, tend to be energy-efficient and performant.
-However, as the number of clients grows, the frameworks become more energy-hungry and less performant.
+% However, as the number of clients grows, the frameworks become more energy-hungry and less performant.
 Interpreted languages suffer the most, particularly in terms of latency and the cost of a single request.
-\textsf{Crystal}, interestingly, acted. It behaves oppositely. The cost of a single request increases as the number of clients decreases.
+% \textsf{Crystal}, interestingly, acted. It behaves oppositely. The cost of a single request increases as the number of clients decreases.
 
@@
 \end{enumerate}
+
+% NOTE : discuss this with Romain later (90% chances that it is not needed)
+% \paragraph{overall analysis}
+% \input{correlation_table.tex}
+% in order to reduce the space of research, we processed to do a correlation test between the different variables. We used the Pearson Spearman rank correlation~\cite{zar2005spearman}.
+
+% Table~\ref{table:correlation} presents the correlation between the different observable variables.
+As one can see,  There is a high correlation between the number of requests, the 99\% tail latency, and the average energy cost.
+This means that the more requests are answered, the less energy is consumed, and the more latency is observed.
+However, the correlation between the average power consumption and the other variables is low, which means that the power consumption is not directly related to the other variables.
+% As one can see, except for rust-based implementation, the correlation between the energy consumption per request and the tail latency was extremely high. 
+% Therefore we will focus on two main variables 
+% \begin{itemize}
+%     \item average power consumption - that will represent the energetic behavior of the server 
+%     \item average request per second: that will be our representative variable to estimate the performance 
+% \end{itemize}
+
+% We chose these two variables because they are both expressed in seconds, which will narrow our analysis to the present behavior of the server at any given time and not along the whole execution process.
+
+% As one can notice in the first column of table~\ref{table:correlation}
+% There are three main categories of servers. 
+% \begin{enumerate}
+%     \item Positive correlation between the energy and performance :  
+% \end{enumerate}
+
+% \paragraph{metric based analysis}
 
@@
 \end{figure}
-
-\input{correlation_table.tex}
 
@@
 
-\subsection{Conclusion \note{missing}}
+\subsection{Conclusion}
+
+This section presents a study on the impact of programming languages on the energy for request handling. To do that, we used an implementation of 25 frameworks based on 17 programming languages. 
+These servers use the official RPC protocol, reducing programmers' bias toward specific languages.
+
+The results show that the number of clients had a binary effect on the servers. 
+With a small number of clients, there was no typical behavior between different categories.
+On the other hand, when dealing with a more significant number of clients,  interpreted languages and compiled ones dropped in terms of performance while keeping the same average power, unlike the Vm~based ones that increased their average power while keeping the same performances.
+Overall the second strategy had a better impact on the single request cost than the first one.
+
+As for the evolution of the query size, we have seen three modes of behavior for each framework, depending on the size of the requests: the stress-free mode, the escalation mode, and the broken state mode.
+We start with the stress-free mode. The leader class was the compiled languages, while the VM-based ones tended to consume more energy. 
+Finally, with the escalation mode.
+one could observe three strategies to deal with the increase in request size.
+The first strategy was to drop in performance while keeping the same power; the second was to increase the average power consumption without impacting the performance; the third was a combination of both previous strategies with an increase in power and a drop in performance.
+Interestingly the third strategy had a lower cost per request than the first one.
+Finally, most implementations could not handle requests heavier than 10 MB for the broken mode. Elixir was an exception to this limit despite its low performance compared to other languages
+
+This study shows the absence of a universal language when dealing with RPC requests.
+Moreover, each implementation had a different way of dealing with scalability.
+While some chose to increase the average power consumption, others dropped in performance.
+Both of these strategies led to a higher cost of a single request.
 

### 2022-10-02 chapters/programming_languages/programming_languages.tex
add : introduction to programming langauges

--- 2022-09-23 chapters/programming_languages/programming_languages.tex
@@
 \label{chapter:porgramming_langauges}
-% \section{Introduction}
-In this chapter, we study how the programming framework affects the energy the software consumes.
-We suggest starting with general micro-benchmarking and watching how each programming framework performs with the CPU and memory.
-The main goal of this chapter is to advise developers on how to choose a programming framework based on their project's needs to make their product use the least amount of energy possible.
-For such a question, no answer is obvious.
-Nonetheless, there are some features we can take from each programming framework, such as:
-\begin{itemize}
-    \item performance,
-    \item community support,
-    \item scalability,
-    \item energy consumption,
-    \item memory usage.
-\end{itemize}
 
-As we saw in the previous chapter, one of the most important things about a benchmark is how well it \textsc{reflects} the production environment.
-Therefore, we extend our study to cover real-world use cases, including two case studies reported in the following sections.
+% After looking at Django's studies, we wanted to expand our study to other programming languages and frameworks.
+% The idea behind this study is to expand the work of Pereira et al. from the micro benchmarking level to get closer to real-world applications.
+% First, we will start with studying the RPC protocol since it offers a multi-language interface. Therefore we will reduce the developer expertise bias
+% after that; we will enlarge this study to cover the most popular web frameworks.
+In this study, we will focus on the energy footprint of the web frameworks and the RPC protocol in the context of an unlimited time.
+Therefore the factor time will have less impact, unlike the micro benchmarking study.
+
+% This chapter studies how the programming framework affects the energy the software consumes.
+% We suggest starting with general micro-benchmarking and watching how each programming framework performs with the CPU and memory.
+% The main goal of this chapter is to advise developers on choosing a programming framework based on their project's needs to make their product use the least amount of energy possible.
+% For such a question, no answer is obvious.
+% Nonetheless, there are some features we can take from each programming framework, such as:
+% \begin{itemize}
+%     \item performance,
+%     \item community support,
+%     \item scalability,
+%     \item energy consumption,
+%     \item memory usage.
+% \end{itemize}
+
+\section*{Introduction}
+Nowadays, web applications dominate online systems.
+From Google to Facebook, web applications are widely deployed across organizations and continuously accessed by end-users for personal and professional daily tasks.
+In practice, the development of these web applications heavily relies on a vast ecosystem of \emph{web frameworks}, intended to ease and foster the development process.
+
+However, once deployed, the applications developed with such web frameworks do not exhibit the same performance as reported by the \emph{Web Framework Benchmarks} peri odically published by the \textsc{TechEmpower} company~\footnote{\url{https://www.techempower.com/benchmarks}}.
+
+Thanks to such benchmarks, developers can make informed decisions on the most efficient technology to adopt to implement their web applications.
+
+Unfortunately, when selecting a web framework, developers and benchmark providers focus primarily on popularity and performance criteria, with little regard for the resource consumption implications of their choice.
+This is especially unfortunate, given that developers increasingly turn to cloud providers to host their web applications.
+In such a context, the energy consumption of web applications is a critical factor, as it directly impacts the cost of hosting the applications.
+While cloud providers provide a convenient elastic provision of resources to scale based on application requirements, this convenience may come at a high cost to their business.
+Indeed, the energy consumption of cloud data centers is a primary concern for cloud providers, as it represents a significant portion of their operational costs.
+Beyond the economic cost of web applications, one can also question the global impact of web applications on worldwide carbon emissions.
+
+Given the tremendous success of web applications, their deployment has severely increased over the last few years, thus causing a rebound effect on the power consumption of server infrastructure hosted or supported by cloud providers.
+
+While one can challenge the relevance of features that developers continuously deploy to keep engaging end-users, reconciling economic and environmental concerns remains an open challenge to address.
+
+Given this context, this chapter intends to address this challenge by investigating the energy footprint of web frameworks.
+In particular, we aim to support the developers of web applications with relevant guidelines that can help them to choose the web framework that is not only the most popular or provides the best performance but also exhibits a low energy footprint.
+
+By minimizing the energy consumed to process user requests with no service quality penalty, developers can reduce the operational cost of their web applications and contribute to reducing worldwide carbon emissions from ICT.
+
+To achieve this objective, we consider two study cases.
+First, we study the impact of programming languages while handling requests from the Remote Procedure Call protocol (RPC).
+RPC is a protocol that allows a computer to call a procedure stored in another address space (usually on another computer on a shared network) without the user having to explicitly code the details for this remote interaction.
+We compare multiple implementations of \textsc{gRPC} library, a new RPC library developed by Google and based on \textsc{Protocol Buffers}, a popular serialization protocol.
+As for section~\ref{sec:webframework}, we study the impact of web framework stacks on energy consumption.
+To do so, we implement a simple web application using the most popular web frameworks and compare their performance, latency, and energy consumption.
+We leverage the \textsc{TechEmpower} \emph{Web Framework Benchmarks} to incorporate server-side energy measurements obtained from a software-defined power meter, named \textsc{PowerAPI}~\cite{fieni2020smartwatts}.
+These measurements are then analyzed in depth to understand the critical criteria that can impact the power consumption of web frameworks and derive guidelines for supporting developers in picking the most energy-efficient web frameworks according to their requirements.
 
@@
 
-\section{Investigating Web Application Frameworks}
+\section{Investigating Web Application Frameworks}\label{sec:webframework}
 \import{\currfiledir/web_frameworks}{web_frameworks}
 
-\section{Conclusion\note{missing}}
+\section{Conclusion \note{missing}}

### 2022-10-02 chapters/programming_languages/rpc/rpc.tex
add : introduction to programming langauges

@@
-\subsection{Introduction}
 With the success of the Internet and the emergence of cloud technologies, many communication protocols compete to take the lead.
@@
 We have many initiatives, such as OpenAPI, that try to create a taxonomy for RESTful APIs, while other approaches implement all the different interfaces of the protocol by themselves, such as The Remote Procedure Call (aka RPC) protocol.
-In this Section
+In order to provide a universal implementation of this protocol, Google developed GRPC, a contemporary open-source RPC framework that may run in any environment.
+Pluggable load balancing, tracing, health monitoring, and authentication support may efficiently link services within and across data centers.
+It is also valuable for the final mile of distributed computing, connecting devices, mobile apps, and browsers to backend services.
 
@@
 
-To answer these questions, we will define an interface using the Proto2~\footnote{\url{https://developers.google.com/protocol-buffers}}language, and then we will implement it in multiple servers, each one using a different programming language.
+To answer these questions, we will define an interface using the Proto2\footnote{\url{https://developers.google.com/protocol-buffers}} language, and then we will implement it in multiple servers, each one using a different programming language.
 We will then compare the energy consumption of each implementation using a custom version of \texttt{GHZ}~\footnote{\url{https://github.com/chakib-belgaid/energy_ghz}}, a tool that allows us to stress test our servers.
@@
     \end{center}
-    \caption{The cost of a single request based on the number of clients (mj)}\label{fig:energy_cost_clients}
+    \caption{The cost of a single request based on the number of clients (mJ)}\label{fig:energy_cost_clients}
 \end{figure}
@@
 % As one can see, except for rust-based implementation, the correlation between the energy consumption per request and the tail latency was extremely high. 
-% Therefore we will focus on two main variables 
+% Therefore, we will focus on two main variables 
 % \begin{itemize}

### 2022-10-03 chapters/programming_languages/rpc/rpc.tex
add : conclusion to webframeworks section

@@
 \begin{description}
-    \item[\textsc{RQ}~2:] How do RPC implementations consume energy regarding the number of concurrent clients?
-    \item[\textsc{RQ}~1:] How do RPC implementations consume energy concerning the size of the incoming request?
+    \item[\textsc{RQ}~1:] How do RPC implementations consume energy regarding the number of concurrent clients?
+    \item[\textsc{RQ}~2:] How do RPC implementations consume energy concerning the size of the incoming request?
 \end{description}

### 2022-10-03 chapters/optimization/java-dir/java.tex
add : example

@@
 \section{Tools and contributions}
-\textsf{J-Referral} is an open-source tool designed to assist developers and practitioners in selecting the most energy-efficient JVM configuration for their software..
+\textsf{J-Referral} is an open-source tool designed to assist developers and practitioners in selecting the most energy-efficient JVM configuration for their software.
 \Cref{table:j-referral} illustrates an example of the final report returned by \textsf{J-Referral}.

### 2022-10-04 chapters/literature/draft_citations.tex
add : summary for programmming langaugess

@@
 To do this, we will elaborate on a benchmarking process and a set of tools intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
-Thus, we will begin by examining state of the art in empirical analysis and retrieving the best empirical experimentation methodologies in the research field.
+Thus, we will begin by examining the state of the art in empirical analysis and retrieving the best empirical experimentation methodologies in the research field.
 Then, we will narrow these practices down to computer science so we can finally adapt them to energy consumption.
@@
 It allows a sampling frequency of up to 1\,Hz.
-It has an internal memory to store a wide variety of data, such as the maximum voltage and current, that can later be exported via USB port for personal usage or lined to graph programs like Logger Pro or LabQuest.
+It has an internal memory to store a wide variety of data, such as the maximum voltage and current, that can later be exported via a USB port for personal usage or lined to graph programs like Logger Pro or LabQuest.
 The main advantage of this tool is the ability to monitor the energy consumption from a different device which will reduce the risk of interference with the energy consumption of the test~\cite{hirst2013watts}
@@
 Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
-We can observe a consedirable variation of the energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.
+We can observe a considerable variation of the energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.
 
@@
 The research also offers language combinations that developers might use to improve energy economy, execution speed, and memory utilization.
-This paper's findings were based on the game benchmark, the most famous benchmark comparing several programming languages.
-
-This paper defines a ranking of energy efficiency in programming languages.
-We considered a set of computing problems implemented in ten well-known programming languages and monitored the energy consumed when executing each language.
-Our preliminary results show that although the fastest languages tend to be the lowest energy-consuming ones, there are other interesting cases where slower languages are more energy-efficient than faster ones.
+This paper's findings were based on the game benchmark\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}, the most famous benchmark comparing several programming languages.
 
@@
 \section{Conclusion}
-As we have seen in state of the art, many methods have been proposed to reduce the ICT's energy footprint, which can be applied in different parts of the program's lifecycle, from consumption to execution.
+As we have seen in the state of the art, many methods have been proposed to reduce the ICT's energy footprint, which can be applied in different parts of the program's lifecycle, from consumption to execution.
 Furthermore, the execution phase took the attention of many researchers because it is the part where the most energy is consumed.

### 2022-10-04 chapters/programming_languages/programming_languages.tex
add : example
add : summary for programmming langaugess

@@
 \section{Investigating Web Application Frameworks}\label{sec:webframework}
+
 \import{\currfiledir/web_frameworks}{web_frameworks}
 
-\section{Conclusion \note{missing}}
+\section{Summary}
+
+this chapter aims to extend the work of \citeauthor{pereira_energy_2017}~\cite{pereira_energy_2017} by studying the energy footprint on other aspects than micro-benchmarking.
+The main focus of this chapter is the behavior of programming languages within the context of web servers. To do so, we have chosen to study cases.
+The first experiment is the impact of programming languages when handling the RPC protocol.
+We used the official implementation of \textsc{gRPC} library to build 25 frameworks using 17 programming languages.
+In the second experiment, we look at how web frameworks affect energy consumption by creating a basic web application and comparing its performance, latency, and energy usage using the most common web frameworks.
+Overall we had $261$ implementation using $35$ and three different databases.
+For each implementation, we considered seven typical scenarios.
+Each scenario is tested with several stress levels to study the scalability of the web frameworks.
+Therefore the final results are an $8,750$ use case.
+
+Our studies revealed the absence of a clear winner in terms of energy efficiency.
+However, when we take into consideration the cost of a single request.
+C++ and Rust were the top-ranked.
+These results confrim the work done by \citeauthor{pereira_energy_2017}.
+On the other hand, Unlike the results shown in the paper \cite{pereira_energy_2017}, \textsf{PHP} exhibits a higher performance while keeping an average power consumption
+This result is due to the nature of \textsf{PHP} language, which is dedicated to the web and not general-purpose programming. 
+While most of the frameworks consumed around $100 watts$, \textsf{JAVA} based frameworks consumed around $220 watts$.
+which made it one of the most energy-consuming in the long run, especially when considering the times when there is low traffic.
+
+Finally, this study revealed that each framework had different behaviors toward scalability.
+Some of them increased the average power consumption while trying to keep up with the performance.
+Others, on the other hand, could keep the average power consumption while decreasing the performance.
+Furthermore, there was a case where there was a mix between both strategies.
+
+As there is no final answer to the question of which framework is the most energy-efficient, we provide developers with a tool to help them compare between frameworks based on their needs.
+The criteria we used to compare frameworks are the scenario type, the stress level, and the database type.
+The tool is available at \url{https://github.com/chakib-belgaid/greenboard}

### 2022-10-04 chapters/optimization/python-dir/python.tex
Proofreading

@@
 
-As for web development, we looked at a website built with Django, which is one of the most popular web frameworks in the Python community.
+As for web development, we study a website built with Django, which is one of the most popular web frameworks in the Python community.
 We found that fetching the data from the database consumes most of the energy during the request processing phase.
@@
 However, this strategy should be used cautiously.
- This study demonstrates that the optimal number of processes is equal to the number of physical cores, and exceeding this number will cause the application to loose its efficiency in both performance and energy consumption.
+This study demonstrates that the optimal number of processes is equal to the number of physical cores, and exceeding this number will cause the application to lose its efficiency in both performance and energy consumption.
 As for the multithreading strategy, we found that the Python interpreter is not thread-safe, which leads to a slower performance than the sequential one.

### 2022-10-04 chapters/optimization/java-dir/java.tex
Proofreading conclusions

@@
 \clearpage
-\section{Summary}
-
+\section{conclusion}
 This chapter describes an empirical investigation of contrast in energy consumption that some of the most well-known and widely supported JVM platforms can exhibit as well as the important factors that can influence this energy consumption.
@@
 
-On the other hand , the default GC outperformed the other strategies in just half of our experiments, with some significant benefits reported when employing different GCs based on the application characteristics.
+On the other hand, the default GC outperformed the other strategies in just half of our experiments, with some significant benefits reported when employing different GCs based on the application characteristics.
 Our primary observations and recommendations can be stated as follows:
@@
 To make the above principles easier to implement, we propose \emph{J-Referral}, a tool that recommends the best energy-efficient JVM distribution and configuration from among more than a hundred options.
-It generates a comprehensive report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration in order to assist the user in selecting the one with the lowest consumption for Java applications.
+It generates a comprehensive report on the energy consumption of both CPU and DRAM components for each JVM distribution and/or configuration to assist the user in selecting the one with the lowest consumption for Java applications.

### 2022-10-04 chapters/programming_languages/programming_languages.tex
remove : link
Proofreading conclusions

--- 2022-10-02 chapters/programming_languages/programming_languages.tex
@@
 \section{Investigating Web Application Frameworks}\label{sec:webframework}
+
 \import{\currfiledir/web_frameworks}{web_frameworks}
 
-\section{Conclusion \note{missing}}
+\section{Conclusion}
+This chapter aims to extend the work of \citeauthor{pereira_energy_2017}~\cite{pereira_energy_2017} by studying the energy footprint on other aspects than micro-benchmarking.
+The main focus of this chapter is the behavior of programming languages within the context of web applications.
+To do so, we have covered two case studies.
+The first experiment is the impact of programming languages when developing RPC services.
+We used the official implementation of \textsc{gRPC} library to build $25$ frameworks using $17$ programming languages.
+In the second experiment, we look at how web frameworks affect energy consumption by creating a basic web application and comparing its performance, latency, and energy usage using the most common web frameworks.
+Overall we had $261$ implementation using $35$ and three different databases.
+For each implementation, we considered $7$ representative scenarios.
+Each scenario is tested with several stress levels to study the scalability of the web frameworks.
+Therefore, the final results cover $8,750$ use cases.
+
+Our studies revealed the absence of a clear winner in terms of energy efficiency.
+However, when we take into consideration the cost of a single request.
+C++ and Rust were the top-ranked, hence assessing the work done by \citeauthor{pereira_energy_2017}.
+On the other hand, Unlike the results shown in the paper \cite{pereira_energy_2017}, \textsf{PHP} exhibits a higher performance while keeping a high power consumption on average.
+This result is due to the nature of \textsf{PHP} language, which is dedicated to the web and not general-purpose programming. 
+While most of the frameworks consumed around $100 watts$, \textsf{Java}-based frameworks consumed around $220 watts$.
+This made \textsf{Java} one of the most energy-consuming in the context of online services, especially when considering the periods with low traffic.
+
+Finally, this study revealed that each framework had different behaviors towards scalability.
+Some of them increased the average power consumption while trying to keep up with the performance.
+Others, on the other hand, could keep the average power consumption while decreasing the performance.
+Furthermore, there was a case where there was a mix between both strategies.
+
+As there is no final answer to the question of which framework is the most energy-efficient, we provide developers with a tool to help them compare between frameworks based on their needs.
+The criteria we used to compare frameworks are the scenario type, the stress level, and the database type.
+
+% The tool is available at \url{https://github.com/chakib-belgaid/greenboard}

### 2022-10-04 chapters/programming_languages/rpc/rpc.tex
add : summary for programmming langaugess
Proofreading conclusions

@@
 
-\subsection{Conclusion}
+\subsection{Summary}
 
-This section presents a study on the impact of programming languages on the energy for request handling. To do that, we used an implementation of 25 frameworks based on 17 programming languages.
+This section reports on a study on the impact of programming languages on the energy for request handling.
+To do that, we used an implementation of $25$ frameworks based on $17$ programming languages.
 These servers use the official RPC protocol, reducing programmers' bias toward specific languages.
@@
 With a small number of clients, there was no typical behavior between different categories.
-On the other hand, when dealing with a more significant number of clients,  interpreted languages and compiled ones dropped in terms of performance while keeping the same average power, unlike the Vm~based ones that increased their average power while keeping the same performances.
-Overall the second strategy had a better impact on the single request cost than the first one.
+On the other hand, when dealing with a more significant number of clients, interpreted languages and compiled ones dropped in terms of performance while keeping the same average power, unlike the VM-based ones that increased their average power while keeping the same performances.
+Overall, the second strategy had a better impact on the single request cost than the first one.
 
 As for the evolution of the query size, we have seen three modes of behavior for each framework, depending on the size of the requests: the stress-free mode, the escalation mode, and the broken state mode.
-We start with the stress-free mode. The leader class was the compiled languages, while the VM-based ones tended to consume more energy.
-Finally, with the escalation mode.
-one could observe three strategies to deal with the increase in request size.
-The first strategy was to drop in performance while keeping the same power; the second was to increase the average power consumption without impacting the performance; the third was a combination of both previous strategies with an increase in power and a drop in performance.
-Interestingly the third strategy had a lower cost per request than the first one.
-Finally, most implementations could not handle requests heavier than 10 MB for the broken mode. Elixir was an exception to this limit despite its low performance compared to other languages
+We start with the stress-free mode.
+The leader class was the compiled languages, while the VM-based ones tended to consume more energy.
+Finally, with the escalation mode, one could observe three strategies to deal with the increase in request size.
+The first strategy tends to drop performances while consuming the same power; the second was to increase the average power consumption without impacting the performances; the third was a combination of both previous strategies with an increase in power and a drop in performance.
+Interestingly, the third strategy had a lower cost per request than the first one.
+Finally, most implementations could not handle requests heavier than 10 MB for the broken mode.
+Elixir was an exception to this limit despite its low performance compared to other languages.
 

### 2022-10-06 chapters/introduction.tex
update : introduction

--- 2022-09-17 chapters/introduction.tex
@@
-\newpage
-
-\chapter{Introduction \note{missing}}
-
-\chapter{Motivation \note{missing}}
-
+\chapter{Introduction }
 \label{chapter:introduction}
 
-Nowadays, computers are invading our daily lives, from work to leisure, from fancy smartphones to an embedded peacemaker that regulates the heartbeat of people.
+Nowadays, computers are invading our daily lives, from work to leisure, from fancy smartphones to embedded peacemakers that regulate the heartbeat of people.
 As human beings, we are known to use tools to enhance our bodies.
-And thanks to computers, we pushed that step even further, to the point where now we are using machines to extend our brains.
-Some even argue that one day they will replace us, and therefore we have created our own ending.
+Moreover, thanks to computers, we pushed that step even further, to the point where now we are using machines to extend our brains, from equation solvers to tools to recommend to us where we should invest our money, what we should eat, and even who fits best as our partner.
+One major aspect of computers that became omnipresent in our lives is the Internet, which is a network of networks that connects millions of computers all over the world.
+According to Internet World Stats\footnote{\url{https://www.internetworldstats.com/stats.htm}}.
+The number of people connected to the Internet has increased by 4.4 billion in 2019, reaching 4.54 billion worldwide, or 59.2\% of the world population.
 
-Well, this is the problem for the future generations. For the moment, the main concern of humanity is to keep this planet liveable until we find another alternative.
+The internet has evolved from a place where government researchers share information in the 60s to a means of Communication at the beginning of the century, and now it is a place where we can find almost anything we want, from information to entertainment, from social media to e-commerce.
 
-Unfortunately those machines doesn't help that much. as according to .... ....
+At present, a large chunk of the global economy and most governments have shifted their operations to the Internet, at least partially and sometimes wholly; this includes online shops, banking, advertising, video and music consumption, and even public functions.
 
-Researchers are trying to solve this issue through different angles.
-While some scientists are trying to find an alternative green source to generate energy. others focus on reducing this energy consumption.
-In computer science we are concerned with the ladder solution %reference.
-Therefore most of the works are done on tunning the software and the hardware in order to have a more efficient way use this energy. %more efficient way of producing programs from electricity? 
-%say about hardware (i.e. superconductive materials with zero resistance)
-In this thesis we are trying to find a way to make the energy consumption of the computer to be as low as possible. %on the software level
-%More precisely, the purpose of this thesis is to highlight (attract attention to?) the impact of programers' choices on energy comsumption of software while running the code and to advise them with the strategies to write the greenest possible code.
-Our approach is to reduce the energy consumption of the software services by changing certain parameters, such as platform, programming language, and/or the design pattern.
+Moreover, Due to The pandemic caused by COVID-19 disease, the world has been forced to adapt to a new way of living, which has been accelerated by the Internet.
+The Internet has become a necessity for people to work, study~\cite{naresh2020education}, and even health consultations \cite{liaw2021primary}.
 
-The best way to do so is to formulate a theory behind the energy consumption of algorithm, such as the complexity and the o notation.
-Unfortunately this is not possible in the current state of the art.
+On the other hand, as humanity, we face a major challenge, which is climate change.
+The Intergovernmental Panel on Climate Change (IPCC) has warned that the world has only 12 years to limit the global temperature rise to 1.5\degree C and that the world has to reach net-zero emissions by 2050 to avoid the worst effects of climate change.
+The IPCC has also warned that the world has to reduce its emissions by 45\% by 2030 to reach the 1.5\degree C target~\cite{portner2022climate}.
+
+To survive, we have come up with three solutions,
+The first one includes finding a new planet that we can populate and live on\footnote{\url{https://en.wikipedia.org/wiki/Interstellar_(film)}}, which is called the Planetary Migration~\cite{mapstone2022cyanobacteria}.
+Meanwhile, the second solution is to provide new sources of energy, such as nuclear energy, wind energy, and even fusion energy~\cite{gross1984fusion}.
+The third solution is to reduce our emissions, which is the main focus of this thesis.
+
+While there are many fields where one can optimize energy consumption.
+Our focus is on the Information and Communication Technology (ICT) sector, which is expected to account for around 4\% of global greenhouse gas (GHG) emissions in 2020, with an alarming 8\% growth rate, according to the French think tank The Shift Project\footnote{\url{ https://www.theshiftproject.org/article/ict-environmental-impact/}}.
+According to \emph{Statista}, the Energy consumption of ICT increased from 4.3 exajoules in 2018 to 5.8 exajoules in 2025\footnote{\url{https://www.statista.com/statistics/271139/energy-consumption-of-ict-worldwide}}.
+\begin{figure}[!h]
+    \includegraphics[width=\linewidth]{chapters/distribution_of_ict_consumption.png}
+    \caption{Final energy consumption of digital technologies by item in 2017}
+    \label{fig:distribution_of_ict_consumption}
+\end{figure}
+
+Figure~\ref{fig:distribution_of_ict_consumption} shows the distribution of ICT consumption in 2018, where the largest chunk of the energy consumption is due to data usage, aka 55\% of the energy consumed where 35\% of this energy is used by data centers.
+Reducing energy consumption means reducing the impact 19\% of the ICT energy consumption has on the environment.
+
+In 2020, the market for data center services was worth 48.9 billion\$. It is thought that this number will go up to 105.6 billion\$ by 2026~\cite{inshakova2022data}. This growth is caused mainly by:
+\begin{itemize}
+    \item Shift to remote lifestyle: work, education, and entertainment.
+    \item Increase in the number of connected devices (IoT)
+    \item Development of data-hungry technologies such as Machine learning, AI, Big data, and so on.
+    \item Edge computing and 5G.
+\end{itemize}
+
+With this increase in the number of data centers comes an increase in energy consumption, which is a major problem for the environment. 
+in 2018 datacenters consumed around 205 terawatt-hours (TWh)\cite{schneider2021world} which is equivalent to the energy consumption of 1\% of the total world's electricity.
+This ratio increased up to 1.5 \% in 2020 according to the Journal of Science~\cite{mytton2021data}.Figure~\ref{fig:data_centers_power_distribution}, as one can see, While 40\% of the energy consumed by data centers is used for cooling, another 40\% is used by the servers themselves.
+Therefore optimizing these two aspects can have a major impact on the energy consumption of data centers.
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.6\linewidth]{chapters/data_centers_power_distribution}
+    \caption{Distribtuion of power consumption in a data-center\cite{rong2016optimizing}}
+    \label{fig:data_centers_power_distribution}
+\end{figure}
+
+Researchers are trying to reduce the energy consumption of data centers through different angles.
+Some of the works are focused on the hardware side, such as using new hardware architectures that are more energy-friendly, such as the use of GPUs instead o ARM processors instead of CPUs~\cite{aroca2012towards}.
+Others are trying to optimize the cooling system, this can be achieved by using more efficient cooling systems, putting data centers in cold locations or under water~\cite{simon2018project}, or even using the waste heat for other purposes such as heating buildings\cite{bouzel2021distributed,cao2021carbon}.
+
+A third approach is to optimize the software, by making software more energy-efficient.
+In this thesis, we focus on this approach, and we try to optimize the software by reducing the number of computations that are done by the software.
+
+The best way to do so is to formulate a theory behind the energy consumption of algorithms, such as the complexity and the o notation.
+Unfortunately, this is not possible in the current state of the art.
 due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.
+Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on the execution environment.
+Therefore, for the moment we will start by formulating some hypotheses and explore them using empirical analysis.
 
-Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on execution environment.
-Therefore, for the moment we will start by formulating some hypothesis and explore them using empirical analysis.
+\subsection*{Objectives}
 
-Our work will be presented through the following chapters:
-\begin{enumerate}
-    \item \ref{chapter:literature_review}: Where we discuss the work done on the energy consumption and optimization in software engineering
-    \item \ref{chapter:benchmarking}: It will present a set of guidelines and tools to help practitioners measure the energy consumption of their algorithms.
-    \item : we will present the impact of programming languages on the energy consumption of the algorithms.
-    \item : it will discuss the behaviour of python and the possibile ways to tune it in ordrer to reduce the energy consumpion
-    \item : will present a study on java programming language and the impact of the JVM choice on the energy consumpion
-    \item : as a prespective we introduce the impact of parallelisme on the enrgy consumption on time agnostic cases
-\end{enumerate}
+% \url{https://www.statista.com/statistics/871513/worldwide-data-created/}
 
-....
+% https://www.iea.org/reports/data-centres-and-data-transmission-networks
+
+\begin{figure}[!h]
+    \caption{Our position in the IT sustainability research}
+    \label{fig:thesis_position}
+    \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_position.pdf}
+\end{figure}
+
+% Our work will be presented in the following chapters:
+% \begin{enumerate}
+%     \item \ref{chapter:literature_review}: Where we discuss the work done on energy consumption and optimization in software engineering
+%     \item \ref{chapter:benchmarking}: It will present a set of guidelines and tools to help practitioners measure the energy consumption of their algorithms.
+%     \item : it will discuss the behavior of python and the possible ways to tune it in order to reduce the energy consumption
+%     \item : will present a study on java programming language and the impact of the JVM choice on the energy consumption
+%     \item : we will present the impact of programming languages on the energy consumption of the algorithms especially when it comes to web services.
+%     \item : as a perspective, we introduce the impact of parallelism on the energy consumption on time agnostic cases
+% \end{enumerate}
 
 \section{Motivation}
-....
 

### 2022-10-06 chapters/introduction.tex
update : introduction

@@
 Unfortunately, this is not possible in the current state of the art.
-due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.
+Due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.
 Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on the execution environment.
 Therefore, for the moment we will start by formulating some hypotheses and explore them using empirical analysis.
+Figure~\ref{fig:thesis_position} highlights this thesis's position on the sustainability of ICT; while this thesis only addresses a small portion of ICT's energy usage, we feel it is a step in the right direction for additional solutions to mature in order to preserve humanity.
 
 \subsection*{Objectives}
-
+As seen in  our thesis is positioned in the brick of software optimization in the field of energy consumption. 
+Our main objective is to reduce the energy consumption using 
 % \url{https://www.statista.com/statistics/871513/worldwide-data-created/}
-
 % https://www.iea.org/reports/data-centres-and-data-transmission-networks

### 2022-10-06 chapters/introduction.tex
add : draft : of objective

--- 2022-10-06 chapters/introduction.tex
@@
-\chapter{Introduction }
+\chapter{Introduction}
 \label{chapter:introduction}
@@
 \begin{figure}[!h]
-    \includegraphics[width=\linewidth]{chapters/distribution_of_ict_consumption.png}
+    \includegraphics[width=0.8\linewidth]{chapters/distribution_of_ict_consumption.png}
     \caption{Final energy consumption of digital technologies by item in 2017}
@@
 Unfortunately, this is not possible in the current state of the art.
-due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.
+Due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.
 Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on the execution environment.
 Therefore, for the moment we will start by formulating some hypotheses and explore them using empirical analysis.
-
-\subsection*{Objectives}
-
-% \url{https://www.statista.com/statistics/871513/worldwide-data-created/}
-
-% https://www.iea.org/reports/data-centres-and-data-transmission-networks
-
+Figure~\ref{fig:thesis_position} highlights this thesis's position on the sustainability of ICT\@; while this thesis only addresses a small portion of ICT's energy usage, we feel it is a step in the right direction for additional solutions to mature in order to preserve humanity.
 \begin{figure}[!h]
@@
 \end{figure}
+\subsection*{Objectives}
+
+The purpose of this thesis is to help developers build more energy-efficient software.
+Unfortunately, when it comes to the energy consumption of programs, there is a lack of awareness and knowledge among software developers~\cite{ournani2020reducing,pang2015programmers,pinto2014mining}.
+This is mainly due to the lack of tools that can help developers understand the energy consumption of their programs.
+Therefore we aim, to provide clear, understandable, and easy-to-use tools and guidelines that can help developers reduce the energy consumption of their programs.
+
+First, we create a benchmarking protocol that serves researchers and practitioners, experimenting with their approaches and solutions to reduce the energy consumption of programs. 
+Then.
+using this protocol, we analyze and try to optimize the energy consumption of one of the most popular programming languages, Python.
+We first study python within its most use cases, machine learning, and web Development.
+Then we try to optimize the energy consumption on the outer side, this can be achieved by targeting the interpreter itself.
+
+After that, we continue our non-intrusive approach, by targeting the impact of the virtual machine on the JAVA code.
+This not only helps practitioners improve their software energy without changing their code, but is also beneficial for reducing the energy consumption of the legacy code with almost zero cost.
+
+Finally, we will shift our focus to other programming languages and compare the energy consumption of different programming languages. 
+We consider two main use cases. The RPC protocol, and web servers.
+
+% \url{https://www.statista.com/statistics/871513/worldwide-data-created/}
+% https://www.iea.org/reports/data-centres-and-data-transmission-networks
 
@@
 % \end{enumerate}
-
-\section{Motivation}
 

### 2022-10-07 chapters/introduction.tex
update : centring graphics
add : diagram of contributions
add : contributions

@@
 \begin{figure}[!h]
+    \centering
     \includegraphics[width=0.8\linewidth]{chapters/distribution_of_ict_consumption.png}
@@
     \label{fig:thesis_position}
+    \centering
     \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_position.pdf}
@@
 
-\section{Research Contributions}
+% \subsection*{Contributions}
 
+\clearpage
+\section{Contributions}
+
+\subsection*{Conferences}
+\nobibliography*
 \begin{enumerate}
-
-    \item Introduces
-    \item Shows how ....
-    \item Proposes ...
-
+    \item \bibentry{ournani2020taming}
+    \item \bibentry{ournani2021evaluating}
 \end{enumerate}
 
-\section{Publications}
+\subsection*{Tools}
+\begin{itemize}
+    \item \bibentry{Belgaid_JouleHunter_an_2021}
+    \item \bibentry{Belgaid_JRefferal_Which_JVM_2021}
+    \item \bibentry{Belgaid_Jouleit_a_2020}
+    \item \bibentry{Belgaid_Ijoules_Python_library_2020}
+    \item \bibentry{Belgaid_Pyjoules_Python_library_2019}
+\end{itemize}
 
-\begin{enumerate}
-    \item ...
+\subsection*{Posters}
+\begin{itemize}
+    \item \emph{PyRAPL\@: Make your python green again}
+\end{itemize}
+\subsection*{Future Work}
+\begin{itemize}
+    \item Reducing the energy consumption of Python using non-intrusive techniques
+    \item Empirical analysis on the energy consumption of different web frameworks
+    \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol)
+    \item THow do ORMs affect how much energy is used? A case study of Django and Flask
+\end{itemize}
 
-    \item ...
-
-    \item ...
-
-\end{enumerate}
 \cleardoublepage

### 2022-10-07 chapters/introduction.tex
update : reformulating some messages

@@
 Moreover, Due to The pandemic caused by COVID-19 disease, the world has been forced to adapt to a new way of living, which has been accelerated by the Internet.
-The Internet has become a necessity for people to work, study~\cite{naresh2020education}, and even health consultations \cite{liaw2021primary}.
+The Internet has become a necessity for people to work, study~\cite{naresh2020education}, and even health consultations~\cite{liaw2021primary}.
 
@@
 in 2018 datacenters consumed around 205 terawatt-hours (TWh)\cite{schneider2021world} which is equivalent to the energy consumption of 1\% of the total world's electricity.
-This ratio increased up to 1.5 \% in 2020 according to the Journal of Science~\cite{mytton2021data}.Figure~\ref{fig:data_centers_power_distribution}, as one can see, While 40\% of the energy consumed by data centers is used for cooling, another 40\% is used by the servers themselves.
+This ratio increased up to 1.5 \% in 2020 according to the Journal of Science~\cite{mytton2021data}.
+Figure~\ref{fig:data_centers_power_distribution}, as one can see, While 40\% of the energy consumed by data centers is used for cooling, another 40\% is used by the servers themselves.
 Therefore optimizing these two aspects can have a major impact on the energy consumption of data centers.
@@
 Figure~\ref{fig:thesis_position} highlights this thesis's position on the sustainability of ICT\@; while this thesis only addresses a small portion of ICT's energy usage, we feel it is a step in the right direction for additional solutions to mature in order to preserve humanity.
+
 \begin{figure}[!h]
@@
 \subsection*{Objectives}
-
 The purpose of this thesis is to help developers build more energy-efficient software.
@@
 
+d
+In order to reduce the energy consumption of software, we can use three strategies.
+The first one consists of \emph{reducing energy consumption by reducing the execution time of the program}.
+The most intuitive way for developers, as the optimization of software's performance, is a key metric for most programs.
+
 First, we create a benchmarking protocol that serves researchers and practitioners, experimenting with their approaches and solutions to reduce the energy consumption of programs.
-Then.
-using this protocol, we analyze and try to optimize the energy consumption of one of the most popular programming languages, Python.
+Then, using this protocol, we analyze and try to optimize the energy consumption of one of the most popular programming languages, Python.
 We first study python within its most use cases, machine learning, and web Development.
 Then we try to optimize the energy consumption on the outer side, this can be achieved by targeting the interpreter itself.
-
 After that, we continue our non-intrusive approach, by targeting the impact of the virtual machine on the JAVA code.
 This not only helps practitioners improve their software energy without changing their code, but is also beneficial for reducing the energy consumption of the legacy code with almost zero cost.
-
 Finally, we will shift our focus to other programming languages and compare the energy consumption of different programming languages.
 We consider two main use cases. The RPC protocol, and web servers.
+
+% we have three scenarios, to reduce energy consumption :
+% 1. win-win, reducing energy consumption by enhancing the performance: the first one is done by default by most of the developers and performance engineers, which is to optimize the code itself
+% 3.
+win neutral, reducing energy consumption while keeping the same performance: the second one is finding reducing the energy consumption while trying to keep the same performance, which the work of this theses
+% 2.
+win loose, reducing energy consumption by reducing the performance: optimizing  the energy consumption by sacrificing the performance of the program itself which is what we call  green faas  we will talk about this in the perspective section
 
@@
 %     \item \ref{chapter:literature_review}: Where we discuss the work done on energy consumption and optimization in software engineering
-%     \item \ref{chapter:benchmarking}: It will present a set of guidelines and tools to help practitioners measure the energy consumption of their algorithms.
-%     \item : it will discuss the behavior of python and the possible ways to tune it in order to reduce the energy consumption
-%     \item : will present a study on java programming language and the impact of the JVM choice on the energy consumption
-%     \item : we will present the impact of programming languages on the energy consumption of the algorithms especially when it comes to web services.
-%     \item : as a perspective, we introduce the impact of parallelism on the energy consumption on time agnostic cases
+%     \item \ref{chapter: benchmarking}: It will present a set of guidelines and tools to help practitioners measure the energy consumption of their algorithms.
+%     \item: it will discuss the behavior of python and the possible ways to tune it in order to reduce the energy consumption
+%     \item: will present a study on java programming language and the impact of the JVM choice on the energy consumption
+%     \item: we will present the impact of programming languages on the energy consumption of the algorithms especially when it comes to web services.
+%     \item: as a perspective, we introduce the impact of parallelism on energy consumption in time-agnostic cases
 % \end{enumerate}
+% \subsection*{Contributions}
 
-% \subsection*{Contributions}
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_contributions.pdf}
+    \caption{Contributions done during this thesis }
+    \label{fig:thesis_contributions}
+\end{figure}
 
@@
     \item \bibentry{Belgaid_Jouleit_a_2020}
-    \item \bibentry{Belgaid_Ijoules_Python_library_2020}
+          % \item \bibentry{Belgaid_Ijoules_Python_library_2020}
     \item \bibentry{Belgaid_Pyjoules_Python_library_2019}
@@
     \item \emph{PyRAPL\@: Make your python green again}
+          % TODO : ask romain about the date and the name of the event 
 \end{itemize}

### 2022-10-07 chapters/introduction.tex
add : introduction

@@
 
-d
 In order to reduce the energy consumption of software, we can use three strategies.
@@
 The most intuitive way for developers, as the optimization of software's performance, is a key metric for most programs.
+There fore in this case, reducing the energy consumption of software is just a happy side effect of optimizing the performance of the software.
 
-First, we create a benchmarking protocol that serves researchers and practitioners, experimenting with their approaches and solutions to reduce the energy consumption of programs.
-Then, using this protocol, we analyze and try to optimize the energy consumption of one of the most popular programming languages, Python.
+The second strategy is \emph{optimizing the energy while trying to keep performance}.
+In this strategy, we aim to reduce the energy consumption of software without impacting its performance of the software.
+This is a more challenging task, as it requires a deeper understanding of the software's energy consumption.
+The main focus of this thesis is to provide tools and guidelines that can help developers in this task.
+
+The final strategy consists of \emph{deleberately reducing the performance of the software in order to reduce its energy consumption}.
+The goal of this method is to reduce energy usage by sacrificing execution time.
+This strategy is useful while dealing with services since the task has an unlimited execution time.
+The second part is to use some.
+opportunistic schedulers, which will stop the software if it is estimated to spend more energy.
+This strategy will be discussed first in Chapter~\ref{chapter:porgramming_langauges}, then we will continue with this strategy in the perspectives section.
+
+\subsection*{Organization}
+Figure\ref{fig:thesis_contributions} summarizes the work done in this thesis.
+As mentioned in the previous section, this thesis aims to assist developers to reduce the energy consumption of their programs in the execution phase.
+To do so, we will use an empirical approach.
+Therefore we will need some tools to help us in this task.
+We will start by providing a tool that can help developers measure the energy consumption of their programs.
+The remainder of the document is summarized below.
+
+\begin{itemize}
+    \item \cref{chapter:literature_review} provides a literature review of the energy consumption of software.
+This chapter will help us understand the current state of the art and the challenges that we will face in this thesis.
+\end{itemize}
+
+% First, we create a benchmarking protocol that serves researchers and practitioners, experimenting with their approaches and solutions to reduce the energy consumption of programs.
+% Then, using this protocol, we analyze and try to optimize the energy consumption of one of the most popular programming languages, Python.
 We first study python within its most use cases, machine learning, and web Development.
 Then we try to optimize the energy consumption on the outer side, this can be achieved by targeting the interpreter itself.
-After that, we continue our non-intrusive approach, by targeting the impact of the virtual machine on the JAVA code.
+% After that, we continue our non-intrusive approach, by targeting the impact of the virtual machine on the JAVA code.
 This not only helps practitioners improve their software energy without changing their code, but is also beneficial for reducing the energy consumption of the legacy code with almost zero cost.
-Finally, we will shift our focus to other programming languages and compare the energy consumption of different programming languages.
-We consider two main use cases. The RPC protocol, and web servers.
+% Finally, we will shift our focus to other programming languages and compare the energy consumption of different programming languages.
+% We consider two main use cases. The RPC protocol, and web servers.
 

### 2022-10-09 chapters/introduction.tex
update : introduction

@@
 \subsection*{Organization}
-Figure\ref{fig:thesis_contributions} summarizes the work done in this thesis.
-As mentioned in the previous section, this thesis aims to assist developers to reduce the energy consumption of their programs in the execution phase.
-To do so, we will use an empirical approach.
-Therefore we will need some tools to help us in this task.
-We will start by providing a tool that can help developers measure the energy consumption of their programs.
-The remainder of the document is summarized below.
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_contributions.pdf}
+    \caption{Contributions done during this thesis }
+    \label{fig:thesis_contributions}
+\end{figure}
+
+Figure~\ref{fig:thesis_contributions} summarizes the work completed for this thesis As mentioned in the previous section, our goal is to help developers reduce the energy consumption of their programs during execution.
+We will use an empirical approach to accomplish this.
+As a result, we will require some tools to assist us in this task.
+To begin, we will provide a tool to assist developers in measuring the energy consumption of their programs.
+The rest of the document is outlined below.
 
@@
     \item \cref{chapter:literature_review} provides a literature review of the energy consumption of software.
-This chapter will help us understand the current state of the art and the challenges that we will face in this thesis.
+First, it introduces the characteristics of empirical studies, their challenges, and how to overcome these challenges in the field of computer sciences, then its presents different approaches to measure the energy consumption of software, after that it discusses the work done to mitigate the energy variation for more a better accuracy, and finally it presents the work done to reduce this energy consumption.
+    \item \cref{chapter:benchmarking} This chapter will discuss two facets of empirical tests and how we adapt them to energy consumption measurements.
+First, we will discuss the reproducibility challenge and how to overcome it with the usage of containers, then we will push this approach further by proposing and new protocol to make tests not only reproducible but able to be extended with other features to overcome the rapid pace of software evolution.
+The second part of this chapter will discuss the aspect of accuracy when it comes to energy consumption measurements.
+This section demonstrates how energy measurements might differ and produce conflicting results for the same work when run on identical machines or even the same machine.
+Then it will provides some solutions to reduce this energy variation and improve the accuracy of the measurements.
+    \item After setting the ground for energy benchmarks in the previous chapter, we start to discuss energy consumption one of the most popular programming languages in the world Python, in \cref{chapter:python}.
+This chapter starts by presenting how much python costs in terms of energy consumption compared to other languages.
+then it provides some statistics about its popularity and typical use cases.
+Then in section \ref{python:sec_insights} we will present the impact of python on energy consumption in three typical use cases, machine learning, web servers, and data manipulation.
+For each use case, we compare the energy consumption of several approaches and provide guidelines on how to optimize the energy consumption of python programs in these use cases.
+Finally, we provide a nonintrusive way to optimize energy consumption without altering the code of the program.
+We will do this by using a different implementation of the python interpreter.
+This not only allows developers to spend less time optimizing their code but also allows them to use it on the legacy code that they cannot modify.
+    \item Motivated by the outcomes of the nonintrusive optimization, we follow this strategy on one of the most popular legacy code base applications programming languages, JAVA.
+In \cref{chapter:java}, we try to optimize the java code, using by changing the default JVM implementation.
+We will compare the energy consumption of $12$ benchmarks using $52$ implementation, each benchmark is dedicated to a typical use case.
+After that, we study two of the JVM features, JIT, AND GC, and show their impact on the energy consumption of JAVA code.
+    \item On the other hand \cref{chapter:porgramming_langauges} will present a different aspect of the impact of energy consumption.
+In this chapter, we will analyze the energy consumption of services, where the execution time is not bounded.
+To do so, we will present two study cases.
+    multiple programming languges , microservices architecture , web servers , the impact of energy consumption within a timless context  i \note{haven't this part yet }
+    \item Finally, we conclude our work in \cref{chapter:conclusion} by summarizing the work done in this thesis and discussing the future work that can be done to further reduce the energy consumption of software.
 \end{itemize}
@@
 
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_contributions.pdf}
-    \caption{Contributions done during this thesis }
-    \label{fig:thesis_contributions}
-\end{figure}
-
-\clearpage
 \section{Contributions}
-
+The contributions of this thesis are summarized as follows:
 \subsection*{Conferences}

### 2022-10-10 chapters/introduction.tex
update : soa in introduction

@@
     \item \cref{chapter:literature_review} provides a literature review of the energy consumption of software.
-First, it introduces the characteristics of empirical studies, their challenges, and how to overcome these challenges in the field of computer sciences, then its presents different approaches to measure the energy consumption of software, after that it discusses the work done to mitigate the energy variation for more a better accuracy, and finally it presents the work done to reduce this energy consumption.
+First, it introduces the challenges that one can encounter when running an empirical study, \emph{reproducibiliy}, \emph{accuracy}, and \emph{representativeness}, the ways in the state of the art overcome these challenges in the field of computer sciences.
+Then, it narrows down to the energy consumption of software and the challenges that one can encounter when measuring the energy consumption of software.
+In this section, we provide the literature approaches used to measure energy consumption and classify them, compare them and discuss their pros and cons when used for our research.
+After that, we will tackle the challenge of accuracy within the field of energy consumption, and discuss how the literature tried to overcome these challenges.
+Finally, we will present some of the most recent works in the field of software energy consumption optimization.
+    % We conclude this chapter by providing a simple method that we will use for the rest of the thesis to optimize the energy consumption of software.
+\textbf{BMO} : benchmark , measure and optimize.
     \item \cref{chapter:benchmarking} This chapter will discuss two facets of empirical tests and how we adapt them to energy consumption measurements.
@@
 This section demonstrates how energy measurements might differ and produce conflicting results for the same work when run on identical machines or even the same machine.
-Then it will provides some solutions to reduce this energy variation and improve the accuracy of the measurements.
+Then it will provide some solutions to reduce this energy variation and improve the accuracy of the measurements.
     \item After setting the ground for energy benchmarks in the previous chapter, we start to discuss energy consumption one of the most popular programming languages in the world Python, in \cref{chapter:python}.
@@
 We will compare the energy consumption of $12$ benchmarks using $52$ implementation, each benchmark is dedicated to a typical use case.
-After that, we study two of the JVM features, JIT, AND GC, and show their impact on the energy consumption of JAVA code.
+After that, we study two of the JVM features, JIT, AND GC, and show their impact on the energy consumption of the JAVA code.
     \item On the other hand \cref{chapter:porgramming_langauges} will present a different aspect of the impact of energy consumption.
@@
 To do so, we will present two study cases.
-    multiple programming languges , microservices architecture , web servers , the impact of energy consumption within a timless context  i \note{haven't this part yet }
+    multiple programming languages, microservices architecture, web servers, the impact of energy consumption within a timeless context  \note{I haven't this part yet }
     \item Finally, we conclude our work in \cref{chapter:conclusion} by summarizing the work done in this thesis and discussing the future work that can be done to further reduce the energy consumption of software.

### 2022-10-11 chapters/abstract.tex
add : abstract

@@
-\thispagestyle{empty} % this line to make the hide the page number of the first page of the abstract. it is a UCCS requirement
+% ************************** Thesis Abstract *****************************
+% Use `abstract' as an option in the document class to print only the titlepage and the abstract.
+\begin{abstract}
+    Wildfires, droughts, heatwaves, floods, melting ice, and other natural calamities have grown alarmingly in recent years, indicating strong evidence of global warming.
+Climate change is already impacting the entire planet and will only intensify in the future.
+According to NASA, global temperatures rose by around 1.1 degrees Celsius between 1901 and 2020, mainly due to increased greenhouse gases, which entrap the energy emitted by the sun's rays.
+While some argue that this happened numerous times in the past due to changes in the Earth's orbit, resulting primarily in five ice ages, this is not the case.
+Carbon dioxide levels have risen 250 times faster in the last century than in the previous 650,000 years.
+Furthermore, greenhouse gas concentrations have reached their highest point in 2 million years.
+    If current trends continue, we will reach +3.2 degrees Celsius by the end of the century, with disastrous repercussions for the Earth.
+However, according to UN reports, we can still rescue ourselves provided we limit the temperature rise below 1.5 degrees Celsius.
+As a result, immediate action is required.
+Many agreements, like the UN Framework Convention on Climate Change and the Paris Agreement, aiming to reach net zero emissions by 2050.
+They propose three approaches: reducing emissions, adapting to climate impacts, and financing necessary adaptations.
+   This thesis will follow the first approach, focusing on reducing emissions.
+We will target the ICT sector since it accounts for 2\% of global emissions.
+While this is a small percentage, it is still a significant amount.
+In fact, the ICT sector is the fastest-growing emitter of greenhouse gases, mainly due to the increasing number of data centers.
+In 2019, the number of data centers worldwide was 1.5 million, which is expected to reach 2.5 million by 2025.
+Therefore, reducing data center energy consumption is crucial to lowering the ICT sector's emissions.
+There are three main ways to do this: improving the hardware's efficiency, reducing the cooling systems' energy consumption, or reducing the energy consumption of the servers themselves.
+We will focus on the last approach, which we believe is the cheapest since it does not require any physical changes to data centers.
+   Our goal is to help developers produce more eco-responsible software by providing tools and guidelines to lower the energy consumption of software running on servers.
+To do so, we decided to follow an empirical approach that consists of three steps, test, measure, then optimize; the reason behind such a decision is to keep up with the rapid pace of the software industry.
+Indeed, the software industry is evolving quickly, and keeping up with the latest technologies is challenging.
+Therefore, we decided to provide tools and protocols to help practitioners test their hypotheses instead of just giving them guidelines since we believe that some of the approaches made during this thesis might already be obsolete when published.
+   Due to the urgency of the matter, we decided to harness the most popular yet energy-hungry programming language, Python.
+In this thesis, we will analyze the energy behavior of Python code during its most commonly used cases.
+Then we will provide a nonintrusive method to reduce its energy consumption.
+Then we will extend this strategy to another programming language famous for its legacy code base, Java, to show that we can still reduce the energy consumption of already running applications without paying a considerable price.
+   Finally, we will alter the way we see the problem.
+Instead of trying to optimize one single application, what if we reduced the energy consumption of the data center as a whole entity?
+Thanks to the micro-services architecture, one application can be produced using many services, each independent of the other.
+This releases us from the bond of using a single programming language as the monolithic application does.
+And with this, one can use multiple programming languages and take advantage of the strengths of each one for a specific scenario.
+   The last chapter will analyze the energy behavior of several programming languages regarding web services while expanding the principle of timeless applications further in the perspective section.
 
-this is the abstract of my thesis
+\end{abstract}
+% Recent years had a worrisome increase in the number of wildfires, droughts, heatwaves, floods, melting ice and other natural disatesrs giving us clear sign of global warming.
+The effects of climate change are already being felt on the whole planet and will continue to worsen in the future.
+According to NASA, the global temperatures increased by around 1.1 °C from 1901 to 2020.
+some argues that this cycle has been already happend before, resulting five major ice ages before, and their main cause was the change of distance between earth and sun resulting in an overall alteration in the earth temperature, and the rise of the carbon dioxide level in the atmosphere, this is the first time that we exceed the level of 300 parts / per million was in 1950 within a span of 800000 years, even worse, this number has risen up 420.
+This level of carbon dioxide is the first within the last 800000.
+% Global warming is the gradual increase of the Earth's average surface temperature.
+It is caused by the increased levels of greenhouse gases in the atmosphere, which trap energy from the sun and cause the Earth to warm which traps the energy of solar rays in earth leading to an increase on the overall temperature.
+% Carbon dioxide from human activities is increasing about 250 times faster than it did from natural sources after the last Ice Age
+% Greenhouse gas concentrations are at their highest levels in 2 million years, and the rate of increase is unprecedented in at least the last 800,000 years.
+% it is not too late, according to scientists if we keep the rise of temperature below 1.5 degrees Celsius, we can still avoid the worst effects of climate change.
+howerver it is projected that the temperature will increase by 3.2 by 2100.
+% According to the Sustainable Development Goals, the UN Framework Convention on Climate Change and the Paris Agreement.
+Three broad categories of action are: cutting emissions, adapting to climate impacts and financing required adjustments.
+% While a growing coalition of countries is committing to net zero emissions by 2050.

### 2022-10-12 chapters/abstract.tex
Proofreading abstract

--- 2021-08-30 chapters/abstract.tex
@@
-\thispagestyle{empty} % this line to make the hide the page number of the first page of the abstract. it is a UCCS requirement
+% ************************** Thesis Abstract *****************************
+% Use `abstract' as an option in the document class to print only the titlepage and the abstract.
+\begin{abstract}
+    % Wildfires, droughts, heatwaves, floods, melting ice, and other natural calamities have grown alarmingly in recent years, indicating strong evidence of global warming.
+Climate change is already impacting the entire planet and will only intensify in the future.
+According to NASA, global temperatures rose by around 1.1 degrees Celsius between 1901 and 2020, mainly due to increased greenhouse gases, which entrap the energy emitted by the sun's rays.
+While some argue that this happened numerous times in the past due to changes in the Earth's orbit, resulting primarily in five ice ages, this is not the case.
+Carbon dioxide levels have risen 250 times faster in the last century than in the previous 650,000 years.
+Furthermore, greenhouse gas concentrations have reached their highest point in 2 million years.
+    % If current trends continue, we will reach +3.2 degrees Celsius by the end of the century, with disastrous repercussions for the Earth.
+However, according to UN reports, we can still rescue ourselves provided we limit the temperature rise below 1.5 degrees Celsius.
+As a result, immediate action is required.
+Many agreements, like the UN Framework Convention on Climate Change and the Paris Agreement, aiming to reach net zero emissions by 2050.
+They propose three approaches: reducing emissions, adapting to climate impacts, and financing necessary adaptations.
+    The ICT sector is claimed to account for 2\% of global emissions.
+    While this may be perceived as a small percentage, the success of ICT technologies inevitably contributes to increase the emissions of greenhouse gases.
+    % the ICT sector is the fastest-growing emitter of greenhouse gases, mainly due to the increasing number of data centers. 
+    In 2019, the number of data centers worldwide was 1.5 million, which is expected to reach 2.5 million by 2025.
+    In this context, reducing data center energy consumption is crucial to lowering the ICT sector's emissions.
+    There are three main ways to achieve this goal: improving the hardware's efficiency, reducing the cooling systems' energy consumption, or reducing the energy consumption of the servers themselves.
+    This thesis focuses on the last approach, which I believe is the most affordable one, as it does not require any physical changes to data centers.
+    My objective is to support developers in producing more sustainable software services through the delivery of tools and guidelines to lower the energy consumption of software running on servers.
+    To do so, I decided to follow an empirical approach that consists of three steps: test, measure, then optimize.
+    The reason behind such a decision is to keep up with the rapid pace of the software industry.
+    Indeed, the software industry is evolving quickly, and keeping up with the latest technologies is challenging.
+    Therefore, I decided to provide tools and protocols to help practitioners to test their hypotheses, instead of only reporting them guidelines as I believe that some of the insights shared as part of this thesis might already be obsolete when published.
+    Due to the urgency of the matter, I decided to harness the most popular, yet energy-hungry programming language, Python.
 
-this is the abstract of my thesis
+    In this thesis, I will analyze the energy behavior of Python code during its most commonly used cases.
+    Then, I will provide a non-intrusive method to reduce its energy consumption.
+    Then, I will extend this strategy to another programming language famous for its legacy code base, Java, to show that we can still reduce the energy consumption of already running applications without paying a considerable price.
+    Finally, I will adopt a more systemic approach.
+    Instead of trying to optimize one single application, can one reduce the energy consumption of the data center as a whole entity?
+    Thanks to the micro-services architecture, one application can be produced using many services, each independent of the other.
+    This releases us from the bond of adopting a single programming language as the monolithic application does.
+    And with this, one can use multiple programming languages and take advantage of the strengths of each one for a specific scenario. 
+    The last chapter will analyze the energy behavior of several programming languages regarding web services while expanding the principle of timeless applications further in the perspective section.
+\end{abstract}
+% Recent years had a worrisome increase in the number of wildfires, droughts, heatwaves, floods, melting ice and other natural disatesrs giving us clear sign of global warming.
+The effects of climate change are already being felt on the whole planet and will continue to worsen in the future.
+According to NASA, the global temperatures increased by around 1.1 °C from 1901 to 2020.
+some argues that this cycle has been already happend before, resulting five major ice ages before, and their main cause was the change of distance between earth and sun resulting in an overall alteration in the earth temperature, and the rise of the carbon dioxide level in the atmosphere, this is the first time that we exceed the level of 300 parts / per million was in 1950 within a span of 800000 years, even worse, this number has risen up 420.
+This level of carbon dioxide is the first within the last 800000.
+% Global warming is the gradual increase of the Earth's average surface temperature.
+It is caused by the increased levels of greenhouse gases in the atmosphere, which trap energy from the sun and cause the Earth to warm which traps the energy of solar rays in earth leading to an increase on the overall temperature.
+% Carbon dioxide from human activities is increasing about 250 times faster than it did from natural sources after the last Ice Age
+% Greenhouse gas concentrations are at their highest levels in 2 million years, and the rate of increase is unprecedented in at least the last 800,000 years.
+% it is not too late, according to scientists if we keep the rise of temperature below 1.5 degrees Celsius, we can still avoid the worst effects of climate change.
+howerver it is projected that the temperature will increase by 3.2 by 2100.
+% According to the Sustainable Development Goals, the UN Framework Convention on Climate Change and the Paris Agreement.
+Three broad categories of action are: cutting emissions, adapting to climate impacts and financing required adjustments.
+% While a growing coalition of countries is committing to net zero emissions by 2050.

### 2022-10-12 chapters/abstract.tex
update : abstract ( en+fr)

--- 2021-08-30 chapters/abstract.tex
@@
-\thispagestyle{empty} % this line to make the hide the page number of the first page of the abstract. it is a UCCS requirement
+% ************************** Thesis Abstract *****************************
+% Use `abstract' as an option in the document class to print only the titlepage and the abstract.
 
-this is the abstract of my thesis
+\begin{abstract}
+The ICT sector is claimed to account for 2\% of global emissions. Even though this might seem like a small number, the success of ICT technologies will always lead to more greenhouse gas emissions. 
+In 2019, the number of data centers worldwide was 1.5 million, which is expected to reach 2.5 million by 2025. 
+In this situation, lowering the emissions from the ICT sector depends on reducing the amount of energy used by data centers. 
+There are three main methods to achieve this goal: improving the hardware's efficiency, lowering the cooling systems' energy consumption, or decreasing the energy consumption of the servers themselves.
+This thesis focuses on the last approach, which I believe is the most affordable one, as it does not require any physical changes to data centers. 
+My goal is to assist developers in making more eco-friendly software services by providing them with tools and guidelines to create software that runs on servers while consuming less energy. 
+To do so, I decided to pursue an empirical approach consisting of three steps: test, measure, and optimize.
+
+The reason for such a decision is to follow the rapid pace of the software industry. 
+In fact, the software industry has one of the fastest growth rates, which makes it challenging to keep up with the newest technologies. 
+So, instead of just reporting my insights, I gave practitioners the means and protocols to allow them to test their hypotheses. 
+I believe that some of the insights shared as part of this thesis might already have become obsolete when published.
+
+Due to the urgency of the climate change issue, I decided first to harness the most popular yet energy-hungry programming language, Python. 
+Therefore, I started by analyzing Python's code's energy behavior during its most commonly used cases. 
+Then, I provided a non-intrusive technique to reduce its energy consumption. 
+After that, I extended this strategy to another programming language famous for its legacy code base, Java, to show that we can still reduce the energy consumption of already running applications without paying a considerable price.
+
+Finally, I adopted a more systemic approach. 
+Instead of optimizing one single application, can one reduce the energy consumption of the data center as a whole entity?
+Thanks to the micro-services architecture, one application can be constructed using many services, each independent of the other.
+This type of architecture releases us from the bond of adopting a single programming language as the monolithic application does.
+And with this, one can use multiple programming languages and take advantage of the strengths of each one for a specific scenario.
+The last chapter analyzed the energy behavior of several programming languages regarding web services while opening a new path toward sustainability within timeless applications.
+\end{abstract}
+
+% Recent years had a worrisome increase in the number of wildfires, droughts, heatwaves, floods, melting ice and other natural disatesrs giving us clear sign of global warming.
+The effects of climate change are already being felt on the whole planet and will continue to worsen in the future.
+According to NASA, the global temperatures increased by around 1.1 °C from 1901 to 2020.
+some argues that this cycle has been already happend before, resulting five major ice ages before, and their main cause was the change of distance between earth and sun resulting in an overall alteration in the earth temperature, and the rise of the carbon dioxide level in the atmosphere, this is the first time that we exceed the level of 300 parts / per million was in 1950 within a span of 800000 years, even worse, this number has risen up 420.
+This level of carbon dioxide is the first within the last 800000.
+% Global warming is the gradual increase of the Earth's average surface temperature.
+It is caused by the increased levels of greenhouse gases in the atmosphere, which trap energy from the sun and cause the Earth to warm which traps the energy of solar rays in the earth leading to an increase on the overall temperature.
+% Carbon dioxide from human activities is increasing about 250 times faster than it did from natural sources after the last Ice Age
+% Greenhouse gas concentrations are at their highest levels in 2 million years, and the rate of increase is unprecedented in at least the last 800,000 years.
+% it is not too late, according to scientists if we keep the rise of temperature below 1.5 degrees Celsius, we can still avoid the worst effects of climate change.
+howerver it is projected that the temperature will increase by 3.2 by 2100.
+% According to the Sustainable Development Goals, the UN Framework Convention on Climate Change and the Paris Agreement.
+Three broad categories of action are: cutting emissions, adapting to climate impacts and financing required adjustments.
+% While a growing coalition of countries is committing to net zero emissions by 2050.  
+\clearpage
+\selectlanguage{french}
+% \begin{otherlanguage}{french} 
+\begin{abstract_fr}
+Le secteur des TIC serait responsable de 2\% des émissions mondiales.
+Même si ce chiffre peut paraître faible, le succès des technologies TIC entraînera toujours une augmentation des émissions de gaz à effet de serre.
+En 2019, le nombre de centres de données dans le monde était de 1,5 million, et devrait atteindre 2,5 millions d'ici 2025.
+Dans cette situation, la diminution des émissions du secteur des TIC dépend de la réduction de la consommation énergétique des centres de données.
+Il existe trois méthodes principales pour atteindre cet objectif : améliorer l'efficacité du matériel, faire baisser la consommation énergétique des systèmes de refroidissement, ou diminuer la consommation énergétique des serveurs eux-mêmes.
+Cette thèse se focalise sur la dernière approche, qui me semble être la plus abordable, car elle ne nécessite aucun changement physique dans les centres de données.
+
+Mon objectif est d'aider les développeurs à créer des logiciels plus écologiques en leur fournissant des outils et des directives pour créer des applications qui tournent sur des serveurs tout en consommant moins d'énergie.
+Pour ce faire, j'ai décidé d'adopter une approche empirique en trois étapes : tester, mesurer et optimiser.
+
+La raison d'une telle décision est de suivre le rythme rapide de l'industrie du logiciel.
+En fait, le secteur du logiciel connaît l'un des taux de croissance les plus rapides, ce qui rend difficile de suivre les nouvelles technologies.
+Ainsi, au lieu de me contenter de rapporter juste mes réflexions, j'ai fourni aux praticiens les moyens et les protocoles nécessaires pour leur permettre de tester leurs hypothèses.
+Je suis convaincu que certaines des conclusions partagées dans le cadre de cette thèse pourraient déjà être obsolètes au moment de leur publication.
+
+Vue l'urgence de la question du changement climatique, j'ai décidé d'abord d'exploiter le langage de programmation le plus populaire et pourtant le plus gourmand en énergie, Python.
+ J'ai donc commencé par analyser le comportement énergétique du code Python dans ses utilisations les plus courantes.
+ J'ai ensuite proposé une technique non intrusive pour réduire sa consommation énergétique.
+ Ensuite, j'ai étendu cette stratégie à un autre langage de programmation célèbre pour sa base de code ancienne, Java, pour montrer que nous pouvons encore réduire la consommation d'énergie d'applications déjà en cours d'exécution sans payer un prix considérable.
+
+Enfin, j'ai adopté une approche plus systémique.
+Au lieu d'optimiser une seule application, peut-on réduire la consommation d'énergie du centre de données dans son ensemble ?
+Grâce à l'architecture micro-services, une application peut être construite à l'aide de nombreux services, tous indépendants les uns des autres.
+Ce type d'architecture nous libère de l'obligation d'adopter un seul langage de programmation comme le fait l'application monolithique. 
+Et avec cela, on peu employer plusieurs langages de programmation et profiter des forces de chacun d'eux pour un scénario spécifique. 
+Le dernier chapitre a analysé le comportement énergétique de plusieurs langages de programmation concernant les services Web tout en ouvrant une nouvelle voie vers la durabilité au sein des applications intemporelles.
+\end{abstract_fr}
+% \end{otherlanguage}
+\selectlanguage{english}

### 2022-10-12 chapters/introduction.tex
update : introduction

@@
 After that, we study two of the JVM features, JIT, AND GC, and show their impact on the energy consumption of the JAVA code.
-    \item On the other hand \cref{chapter:porgramming_langauges} will present a different aspect of the impact of energy consumption.
-In this chapter, we will analyze the energy consumption of services, where the execution time is not bounded.
-To do so, we will present two study cases.
-    multiple programming languages, microservices architecture, web servers, the impact of energy consumption within a timeless context  \note{I haven't this part yet }
+    \item In contrast, \cref{chapter:porgramming_langauges} will use the flexibility provided by the micro-services architecture~\cite{dmitry2014micro} to analyze each programming language's energy behavior in light of various web scenarios to optimize the energy use of web services.
+We will first examine the effects of the various programming languages when dealing with the Remote Procedure Call (RPC) Protocol.
+In this instance, we will use two scaling factors: the number of concurrent clients and the size of the requests.
+Then we will compare $261$ web frameworks, each implementing the same website using seven use cases.
+This analysis aims to look at how each technology uses energy in different kinds of web situations.
     \item Finally, we conclude our work in \cref{chapter:conclusion} by summarizing the work done in this thesis and discussing the future work that can be done to further reduce the energy consumption of software.
@@
 
-\subsection*{Posters}
-\begin{itemize}
-    \item \emph{PyRAPL\@: Make your python green again}
-          % TODO : ask romain about the date and the name of the event 
-\end{itemize}
+% \subsection*{Posters}
+% \begin{itemize}
+%     \item \emph{PyRAPL\@: Make your python green again}
+%           % TODO : ask romain about the date and the name of the event 
+% \end{itemize}
 \subsection*{Future Work}

### 2022-10-12 chapters/programming_languages/programming_languages.tex
update : introduction

@@
 
-However, once deployed, the applications developed with such web frameworks do not exhibit the same performance as reported by the \emph{Web Framework Benchmarks} peri odically published by the \textsc{TechEmpower} company~\footnote{\url{https://www.techempower.com/benchmarks}}.
+However, once deployed, the applications developed with such web frameworks do not exhibit the same performance as reported by the \emph{Web Framework Benchmarks} periodically published by the \textsc{TechEmpower} company~\footnote{\url{https://www.techempower.com/benchmarks}}.
 
@@
 In the second experiment, we look at how web frameworks affect energy consumption by creating a basic web application and comparing its performance, latency, and energy usage using the most common web frameworks.
-Overall we had $261$ implementation using $35$ and three different databases.
+Overall we had $261$ implementation using $35$ distinct programming languages and three different databases.
 For each implementation, we considered $7$ representative scenarios.

### 2022-10-12 chapters/conclusion.tex
add : contributions
update : reformulating some messages
update : introduction
begin  : conclusion

@@
-\newpage
+\clearpage
 \chapter{Discussion and Conclusion}
-\label{chapter:Discussion_Future_Work}
+\label{chapter:conclusion}
 
-\section{Conclusion}
-\label{section:Conclusion}
+This manuscript made numerous contributions to measuring and reducing software's energy consumption. We used a three-step strategy to lower software energy:  benchmarking, measuring, and optimizing. 
+We started with the benchmarking phase.
+Chapter 2 discussed the challenges of a successful benchmarking strategy: reproducibility, accuracy, and representativeness.
+We reasoned that "docker" would be the best fit to ensure that energy studies could be reproduced.
+We then extended this reproducibility to an evolving protocol that helps researchers keep up with the rapid pace of software development.
+Then we targeted the accuracy by studying the hardware and software factors that can impact the energy variation and how practitioners can tune them to harness this energy variation.
+
+After establishing a robust benchmarking protocol to create energy-based experiments, we shifted our focus to the optimization side.
+We opted to start with Python, the most popular yet energy-hungry programming language.
+As a result, we began by examining the energy behavior of Python code in its most common usage.
+Then we presented a non-intrusive method to lessen its energy use.
+Following that, we applied the same strategy to another programming language known for its legacy code base, Java, to prove that we can still cut the energy usage of existing running applications without incurring high costs.
+
+Lastly, we used the flexibility of the micro-services architecture to look at how each programming language uses energy in different web scenarios.
+We first examined the effects of the various programming languages when dealing with the Remote Procedure Call (RPC) Protocol.
+Then we extended this study to a more practical application by comparing $261$ web frameworks, each implementing the same website using seven use cases.
+Then, we provided practitioners with a dashboard to determine which stack is best for a given situation.
 
@@
 
-This section will describe the contributions of this thesis.  These can be summarized as follows:
+The contributions made by this thesis are covered in this section. The following is a summary of them:
 
+\subsection*{Articles}
 \begin{enumerate}
-    \item \textbf{First Idea:}
-          We proposed ...
-    \item \textbf{Second Idea:}
-          We investigated ...
-    \item \textbf{Third Idea:}
-          We addressed ...
+      \item \textbf{Taming energy consumption variations in systems benchmarking}: We investigate the phenomenon of variation when measuring the energy consumption of experiments in this study.
+In this paper, we discuss various hardware and software factors that can amplify variations in recorded energy measures, with a focus on the following research questions:
+      \begin{description}
+            \item[\textsc{RQ}~1:] Does the benchmarking protocol affect the energy variation?
+            \item[\textsc{RQ}~2:] How important is the impact of the processor features on the energy variation?
+            \item[\textsc{RQ}~3:] What effect does the operating system have on energy variation?
+            \item[\textsc{RQ}~4:] Does the choice of processor make a difference in reducing the energy variation?
+      \end{description}
 
+      This contribution shows how processor features can significantly affect the variation in energy use between the benchmarking protocol and the operating system.
+Finally, this study presents several guidelines for controllable parameters that practitioners could easily change to increase the accuracy of their experiments.
+
+      \bibentry{ournani2020taming}
+
+      \item \textbf{Evaluating the impact of Java virtual machines on energy consumption}: In this article, we thoroughly investigate how JVMs affect software energy usage.
+In order to address the following research concerns, we reveal through this study several trials on hundreds of JVMs versions provided by various providers:
+      \begin{description}
+            \item[\textsc{RQ}~1:] What is the impact of existing JVM distributions on the energy consumption of Java-based software services?
+            \item [\textsc{RQ}~1:] What are the relevant JVM settings that can reduce the energy consumption of a given software service?
+        \end{description}
+        The findings demonstrate that choosing the right JVM platform can significantly reduce energy usage depending on the software and use case.
+This optimization can also be achieved by properly configuring the JIT and GC parameters.
+
+      \bibentry{ournani2021evaluating}
 \end{enumerate}
 
-\section{Limitations and Challenges}
-\label{section:limitations}
-
-\section{Future Work}
+\section{Future Work\note{missing}}
 .... Some potential areas for future efforts could include the following:

### 2022-10-12 chapters/conclusion.tex
add : software contributions

@@
 \subsection*{Articles}
+this part summarizes the contributions that have been already accepted in conferences and journals.
+
 \begin{enumerate}
@@
 
+\subsection{Softwares and Tools}
+While the benchmarking and optimization parts were presented as chapters in this thesis, the measurement part was primarily based on developing tools that measure the energy consumption for a given use case.
+The following is a summary of the tools that were developed during this thesis:
+\begin{itemize}
+      \item \textbf{Jouleit} \url{github.com/powerapi-ng/jouleit}: a tool that can be used to monitor energy consumption for any Linux program, this tool was used to compare the energy consumption of different JVMs.
+      \item \textbf{JRefferal} \url{github.com/chakib-belgaid/jreferral}: a tool that allows the user to explore the JVM settings and their impact on the energy consumption of a given Java program.
+This tool was the result of the second article of this thesis.
+      \item \textbf{PyJoules} \url{pypi.org/project/pyJoules}: is a software toolkit to measure the energy footprint of a host machine along the execution of a piece of Python code.
+It can measure the energy consumption on the level of script, function, and bloc of code.
+      \item \textbf{JouleHunter} \url{pypi.org/project/joulehunter}: an energy profiler for python applications.
+It can be used to highlight the functions that consume the most energy in a given Python program.
+Its main usage is to help developers do an exploratory analysis of their application to scope the functions that should be optimized to be then targeted by PyJoules.
+      \item \textbf{GreenBoard} \url{github.com/chakib-belgaid/greenboard}: A dashboard designed to help developers choose the best stack for their web application.
+It is based on the results of the third article of the last chapter.
+
+\end{itemize}
+
+\subsection{Future Submissions}
+\begin{itemize}
+      \item Reducing the energy consumption of Python using non-intrusive techniques
+      \item Empirical analysis on the energy consumption of different web frameworks
+      \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol)
+      \item THow do ORMs affect how much energy is used? A case study of Django and Flask
+\end{itemize}
+
 \section{Future Work\note{missing}}

### 2022-10-13 chapters/introduction.tex
begin  : conclusion
Proofreading introduction

@@
 Nowadays, computers are invading our daily lives, from work to leisure, from fancy smartphones to embedded peacemakers that regulate the heartbeat of people.
-As human beings, we are known to use tools to enhance our bodies.
+As human beings, we are known to use tools to empower our bodies.
 Moreover, thanks to computers, we pushed that step even further, to the point where now we are using machines to extend our brains, from equation solvers to tools to recommend to us where we should invest our money, what we should eat, and even who fits best as our partner.
 One major aspect of computers that became omnipresent in our lives is the Internet, which is a network of networks that connects millions of computers all over the world.
-According to Internet World Stats\footnote{\url{https://www.internetworldstats.com/stats.htm}}.
-The number of people connected to the Internet has increased by 4.4 billion in 2019, reaching 4.54 billion worldwide, or 59.2\% of the world population.
+According to Internet World Stats,\footnote{\url{https://www.internetworldstats.com/stats.htm}} the number of people connected to the Internet has increased by 4.4 billion in 2019, reaching 4.54 billion worldwide, or 59.2\% of the world population.
 
-The internet has evolved from a place where government researchers share information in the 60s to a means of Communication at the beginning of the century, and now it is a place where we can find almost anything we want, from information to entertainment, from social media to e-commerce.
+The Internet has evolved from a place where government researchers share information in the 60s to a means of Communication at the beginning of the century, and now it is a place where we can find almost anything we want, from information to entertainment, from social media to e-commerce.
 
@@
 
-Moreover, Due to The pandemic caused by COVID-19 disease, the world has been forced to adapt to a new way of living, which has been accelerated by the Internet.
+Moreover, due to the pandemic caused by COVID-19 disease, the world has been forced to adapt to a new way of living, which has been accelerated by the Internet.
 The Internet has become a necessity for people to work, study~\cite{naresh2020education}, and even health consultations~\cite{liaw2021primary}.
@@
 On the other hand, as humanity, we face a major challenge, which is climate change.
-The Intergovernmental Panel on Climate Change (IPCC) has warned that the world has only 12 years to limit the global temperature rise to 1.5\degree C and that the world has to reach net-zero emissions by 2050 to avoid the worst effects of climate change.
+The \emph{Intergovernmental Panel on Climate Change} (IPCC) has warned that the world has only 12 years to limit the global temperature rise to 1.5\degree C and that the world has to reach net-zero emissions by 2050 to avoid the worst effects of climate change.
 The IPCC has also warned that the world has to reduce its emissions by 45\% by 2030 to reach the 1.5\degree C target~\cite{portner2022climate}.
 
-To survive, we have come up with three solutions,
-The first one includes finding a new planet that we can populate and live on\footnote{\url{https://en.wikipedia.org/wiki/Interstellar_(film)}}, which is called the Planetary Migration~\cite{mapstone2022cyanobacteria}.
+To survive, we have come up with three solutions.
+The first one includes finding a new planet that we can populate and live on,\footnote{\url{https://en.wikipedia.org/wiki/Interstellar_(film)}} which is known as the Planetary Migration~\cite{mapstone2022cyanobacteria}.
 Meanwhile, the second solution is to provide new sources of energy, such as nuclear energy, wind energy, and even fusion energy~\cite{gross1984fusion}.
@@
 While there are many fields where one can optimize energy consumption.
-Our focus is on the Information and Communication Technology (ICT) sector, which is expected to account for around 4\% of global greenhouse gas (GHG) emissions in 2020, with an alarming 8\% growth rate, according to the French think tank The Shift Project\footnote{\url{ https://www.theshiftproject.org/article/ict-environmental-impact/}}.
-According to \emph{Statista}, the Energy consumption of ICT increased from 4.3 exajoules in 2018 to 5.8 exajoules in 2025\footnote{\url{https://www.statista.com/statistics/271139/energy-consumption-of-ict-worldwide}}.
+Our focus is on the \emph{Information and Communication Technology} (ICT) sector, which is expected to account for around 4\% of \global \emph{GreenHouse Gas} (GHG) emissions in 2020, with an alarming 8\% growth rate, according to the French think tank The Shift Project.\footnote{\url{ https://www.theshiftproject.org/article/ict-environmental-impact/}}
+According to \emph{Statista}, the energy consumption of ICT increased from 4.3 exajoules in 2018 to 5.8 exajoules in 2025.\footnote{\url{https://www.statista.com/statistics/271139/energy-consumption-of-ict-worldwide}}
+
 \begin{figure}[!h]
     \centering
-    \includegraphics[width=0.8\linewidth]{chapters/distribution_of_ict_consumption.png}
+    \includegraphics[width=.7\linewidth]{chapters/distribution_of_ict_consumption.png}
     \caption{Final energy consumption of digital technologies by item in 2017}
@@
 
-In 2020, the market for data center services was worth 48.9 billion\$. It is thought that this number will go up to 105.6 billion\$ by 2026~\cite{inshakova2022data}. This growth is caused mainly by:
+In 2020, the market for data center services was worth 48.9 billion\$.
+It is thought that this number will go up to 105.6 billion\$ by 2026~\cite{inshakova2022data}.
+This growth is caused mainly by:
 \begin{itemize}
-    \item Shift to remote lifestyle: work, education, and entertainment.
-    \item Increase in the number of connected devices (IoT)
-    \item Development of data-hungry technologies such as Machine learning, AI, Big data, and so on.
-    \item Edge computing and 5G.
+    \item shift to remote lifestyle: work, education, and entertainment,
+    \item increase in the number of connected devices (IoT),
+    \item development of data-hungry technologies such as Machine learning, AI, Big data, and so on,
+    \item edge computing and 5G.
 \end{itemize}
@@
 With this increase in the number of data centers comes an increase in energy consumption, which is a major problem for the environment.
-in 2018 datacenters consumed around 205 terawatt-hours (TWh)\cite{schneider2021world} which is equivalent to the energy consumption of 1\% of the total world's electricity.
+in 2018, data centers consumed around 205 terawatt-hours (TWh)~\cite{schneider2021world}, which is equivalent to the energy consumption of 1\% of the total world's electricity.
 This ratio increased up to 1.5 \% in 2020 according to the Journal of Science~\cite{mytton2021data}.
-Figure~\ref{fig:data_centers_power_distribution}, as one can see, While 40\% of the energy consumed by data centers is used for cooling, another 40\% is used by the servers themselves.
-Therefore optimizing these two aspects can have a major impact on the energy consumption of data centers.
+In Figure~\ref{fig:data_centers_power_distribution}, one can see that 40\% of the energy consumed by data centers is used for cooling, while another 40\% is used by the servers themselves.
+Therefore, optimizing these two aspects can have a major impact on the energy consumption of data centers.
+
 \begin{figure}[!h]
@@
     \includegraphics[width=0.6\linewidth]{chapters/data_centers_power_distribution}
-    \caption{Distribtuion of power consumption in a data-center\cite{rong2016optimizing}}
+    \caption{Distribution of power consumption in a data center~\cite{rong2016optimizing}}
     \label{fig:data_centers_power_distribution}
@@
 Some of the works are focused on the hardware side, such as using new hardware architectures that are more energy-friendly, such as the use of GPUs instead o ARM processors instead of CPUs~\cite{aroca2012towards}.
-Others are trying to optimize the cooling system, this can be achieved by using more efficient cooling systems, putting data centers in cold locations or under water~\cite{simon2018project}, or even using the waste heat for other purposes such as heating buildings\cite{bouzel2021distributed,cao2021carbon}.
+Others are trying to optimize the cooling system, this can be achieved by using more efficient cooling systems, putting data centers in cold locations or under water~\cite{simon2018project}, or even using the waste heat for other purposes, such as heating buildings~\cite{bouzel2021distributed,cao2021carbon}.
 
@@
 Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on the execution environment.
-Therefore, for the moment we will start by formulating some hypotheses and explore them using empirical analysis.
-Figure~\ref{fig:thesis_position} highlights this thesis's position on the sustainability of ICT\@; while this thesis only addresses a small portion of ICT's energy usage, we feel it is a step in the right direction for additional solutions to mature in order to preserve humanity.
+Therefore, for the moment, we will start by formulating some hypotheses and exploring them using empirical analysis.
+Figure~\ref{fig:thesis_position} highlights this thesis's position on the sustainability of ICT\@; while this thesis only addresses a small portion of ICT's energy usage, we feel it is a step in the right direction for additional solutions to mature to preserve humanity.
 
 \begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.7\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_position.pdf}
     \caption{Our position in the IT sustainability research}
     \label{fig:thesis_position}
-    \centering
-    \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_position.pdf}
 \end{figure}
-\subsection*{Objectives}
+\subsection{Objectives}
 The purpose of this thesis is to help developers build more energy-efficient software.
@@
 This is mainly due to the lack of tools that can help developers understand the energy consumption of their programs.
-Therefore we aim, to provide clear, understandable, and easy-to-use tools and guidelines that can help developers reduce the energy consumption of their programs.
+Therefore, we aim to provide clear, understandable, and easy-to-use tools and guidelines that can help developers reduce the energy consumption of their programs.
 
-In order to reduce the energy consumption of software, we can use three strategies.
+To reduce the energy consumption of software, we can use three strategies.
 The first one consists of \emph{reducing energy consumption by reducing the execution time of the program}.
 The most intuitive way for developers, as the optimization of software's performance, is a key metric for most programs.
-There fore in this case, reducing the energy consumption of software is just a happy side effect of optimizing the performance of the software.
+Therefore, in this case, reducing the energy consumption of software becomes a side effect of optimizing the performance of the software.
 
@@
 
-The final strategy consists of \emph{deleberately reducing the performance of the software in order to reduce its energy consumption}.
+The final strategy consists of \emph{deliberately reducing the performance of the software to reduce its energy consumption}. 
 The goal of this method is to reduce energy usage by sacrificing execution time.
 This strategy is useful while dealing with services since the task has an unlimited execution time.
-The second part is to use some.
-opportunistic schedulers, which will stop the software if it is estimated to spend more energy.
+The second part is to use some. opportunistic schedulers, which will stop the software if it is estimated to spend more energy. 
 This strategy will be discussed first in Chapter~\ref{chapter:porgramming_langauges}, then we will continue with this strategy in the perspectives section.
 
-\subsection*{Organization}
+\subsection{Organization}
 \begin{figure}[!h]
     \centering
-    \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_contributions.pdf}
-    \caption{Contributions done during this thesis }
+    \includegraphics[width=.7\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_contributions.pdf}
+    \caption{Contributions reported in this thesis}
     \label{fig:thesis_contributions}
@@
 To begin, we will provide a tool to assist developers in measuring the energy consumption of their programs.
-The rest of the document is outlined below.
+The rest of the manuscript is outlined below.
 
@@
     \item \cref{chapter:literature_review} provides a literature review of the energy consumption of software.
-First, it introduces the challenges that one can encounter when running an empirical study, \emph{reproducibiliy}, \emph{accuracy}, and \emph{representativeness}, the ways in the state of the art overcome these challenges in the field of computer sciences.
+    First, it introduces the challenges that one can encounter when running an empirical study, \emph{reproducibility}, \emph{accuracy}, and \emph{representativeness}, the ways in the state of the art overcome these challenges in the field of computer sciences.
     Then, it narrows down to the energy consumption of software and the challenges that one can encounter when measuring the energy consumption of software.
@@
 After that, we will tackle the challenge of accuracy within the field of energy consumption, and discuss how the literature tried to overcome these challenges.
-Finally, we will present some of the most recent works in the field of software energy consumption optimization.
+Finally, we will present some of the most recent works in the field of software energy consumption optimization;
           % We conclude this chapter by providing a simple method that we will use for the rest of the thesis to optimize the energy consumption of software.
 \textbf{BMO} : benchmark , measure and optimize.
-    \item \cref{chapter:benchmarking} This chapter will discuss two facets of empirical tests and how we adapt them to energy consumption measurements.
+    \item \cref{chapter:benchmarking} will discuss two facets of empirical tests and how we adapt them to energy consumption measurements.
     First, we will discuss the reproducibility challenge and how to overcome it with the usage of containers, then we will push this approach further by proposing and new protocol to make tests not only reproducible but able to be extended with other features to overcome the rapid pace of software evolution.
@@
     This section demonstrates how energy measurements might differ and produce conflicting results for the same work when run on identical machines or even the same machine. 
-Then it will provide some solutions to reduce this energy variation and improve the accuracy of the measurements.
-    \item After setting the ground for energy benchmarks in the previous chapter, we start to discuss energy consumption one of the most popular programming languages in the world Python, in \cref{chapter:python}.
-This chapter starts by presenting how much python costs in terms of energy consumption compared to other languages.
-then it provides some statistics about its popularity and typical use cases.
-Then in section \ref{python:sec_insights} we will present the impact of python on energy consumption in three typical use cases, machine learning, web servers, and data manipulation.
+    Then, it will provide some solutions to reduce this energy variation and improve the accuracy of the measurements;
+    \item After setting the ground for energy benchmarks in the previous chapter, we discuss energy consumption one of the most popular programming languages in the world, Python, in \cref{chapter:python}.
+    This chapter reports on how much python costs in terms of energy consumption compared to other languages.
+    Then, it provides some statistics about its popularity and typical use cases and studies the impact of Python on energy consumption in three typical use cases, machine learning, web servers, and data manipulation.
     For each use case, we compare the energy consumption of several approaches and provide guidelines on how to optimize the energy consumption of python programs in these use cases.
-Finally, we provide a nonintrusive way to optimize energy consumption without altering the code of the program.
-We will do this by using a different implementation of the python interpreter.
-This not only allows developers to spend less time optimizing their code but also allows them to use it on the legacy code that they cannot modify.
-    \item Motivated by the outcomes of the nonintrusive optimization, we follow this strategy on one of the most popular legacy code base applications programming languages, JAVA.
+    Finally, we provide a non-intrusive way to optimize energy consumption without altering the code of the program.
+    We will do this by using a different implementation of the Python interpreter.
+    This not only allows developers to spend less time optimizing their code but also allows them to use it on the legacy code that they cannot modify;
+    \item Motivated by the outcomes of the non-intrusive optimization, we follow this strategy on one of the most popular legacy code base applications programming languages, Java.
     In \cref{chapter:java}, we try to optimize the java code, using by changing the default JVM implementation.
     We will compare the energy consumption of $12$ benchmarks using $52$ implementation, each benchmark is dedicated to a typical use case.
-After that, we study two of the JVM features, JIT, AND GC, and show their impact on the energy consumption of the JAVA code.
+    After that, we study two of the JVM features, JIT and GC, and show their impact on the energy consumption of the Java code;
     \item In contrast, \cref{chapter:porgramming_langauges} will use the flexibility provided by the micro-services architecture~\cite{dmitry2014micro} to analyze each programming language's energy behavior in light of various web scenarios to optimize the energy use of web services.
-We will first examine the effects of the various programming languages when dealing with the Remote Procedure Call (RPC) Protocol.
+    We will first examine the effects of the various programming languages when dealing with the \emph{Remote Procedure Call} (RPC) Protocol.
     In this instance, we will use two scaling factors: the number of concurrent clients and the size of the requests.
-Then we will compare $261$ web frameworks, each implementing the same website using seven use cases.
+Then, we will compare $261$ web frameworks, each implementing the same website using seven use cases.
     This analysis aims to look at how each technology uses energy in different kinds of web situations.
@@
 \begin{enumerate}
-    \item \bibentry{ournani2020taming}
-    \item \bibentry{ournani2021evaluating}
+    \item \bibentry{ournani2020taming},
+    \item \bibentry{ournani2021evaluating}.
 \end{enumerate}
@@
 \begin{itemize}
-    \item \bibentry{Belgaid_JouleHunter_an_2021}
-    \item \bibentry{Belgaid_JRefferal_Which_JVM_2021}
-    \item \bibentry{Belgaid_Jouleit_a_2020}
+    \item \bibentry{Belgaid_JouleHunter_an_2021},
+    \item \bibentry{Belgaid_JRefferal_Which_JVM_2021},
+    \item \bibentry{Belgaid_Jouleit_a_2020},
           % \item \bibentry{Belgaid_Ijoules_Python_library_2020}
-    \item \bibentry{Belgaid_Pyjoules_Python_library_2019}
+    \item \bibentry{Belgaid_Pyjoules_Python_library_2019}.
 \end{itemize}
@@
 \begin{itemize}
-    \item Reducing the energy consumption of Python using non-intrusive techniques
-    \item Empirical analysis on the energy consumption of different web frameworks
-    \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol)
-    \item THow do ORMs affect how much energy is used? A case study of Django and Flask
+    \item Reducing the energy consumption of Python using non-intrusive techniques,
+    \item Empirical analysis on the energy consumption of different web frameworks,
+    \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol),
+    \item THow do ORMs affect how much energy is used? A case study of Django and Flask.
 \end{itemize}

### 2022-10-13 chapters/conclusion.tex
typo
Proofreading

--- 2022-10-12 chapters/conclusion.tex
@@
 
-This manuscript made numerous contributions to measuring and reducing software's energy consumption. We used a three-step strategy to lower software energy:  benchmarking, measuring, and optimizing. 
+This manuscript reports on several contributions to measuring and reducing software energy consumption.
+We used a three-step strategy to lower software energy: benchmarking, measuring, and optimizing.
 We started with the benchmarking phase.
-Chapter 2 discussed the challenges of a successful benchmarking strategy: reproducibility, accuracy, and representativeness.
-We reasoned that "docker" would be the best fit to ensure that energy studies could be reproduced.
+Chapter~\ref{chapter:literature_review} discussed the challenges of a successful benchmarking strategy: reproducibility, accuracy, and representativeness.
+We concluded that software containers, like "Docker", would be the best fit to ensure that energy studies could be reproduced.
 We then extended this reproducibility to an evolving protocol that helps researchers keep up with the rapid pace of software development.
-Then we targeted the accuracy by studying the hardware and software factors that can impact the energy variation and how practitioners can tune them to harness this energy variation.
+Then, we targeted the accuracy by studying the hardware and software factors that can impact the energy variation and how practitioners can tune them to harness this energy variation.
 
@@
 As a result, we began by examining the energy behavior of Python code in its most common usage. 
-Then we presented a non-intrusive method to lessen its energy use.
+Then, we presented a non-intrusive method to lessen its energy use.
 Following that, we applied the same strategy to another programming language known for its legacy code base, Java, to prove that we can still cut the energy usage of existing running applications without incurring high costs.
@@
 Lastly, we used the flexibility of the micro-services architecture to look at how each programming language uses energy in different web scenarios.
-We first examined the effects of the various programming languages when dealing with the Remote Procedure Call (RPC) Protocol.
-Then we extended this study to a more practical application by comparing $261$ web frameworks, each implementing the same website using seven use cases.
+We first examined the effects of the various programming languages when dealing with the \emph{Remote Procedure Call} (RPC) protocol.
+Then, we extended this study to a more practical application by comparing $261$ web frameworks, each implementing the same website using seven use cases.
 Then, we provided practitioners with a dashboard to determine which stack is best for a given situation.
@@
 \label{section:SummaryofContributions}
+The contributions reported in this thesis are covered in this section.
+The following is a summary of them:
 
-The contributions made by this thesis are covered in this section. The following is a summary of them:
+\subsection{Published Articles}
+This part summarizes the contributions that have been already accepted in conferences and journals.
 
-\subsection*{Articles}
 \begin{enumerate}
-      \item \textbf{Taming energy consumption variations in systems benchmarking}: We investigate the phenomenon of variation when measuring the energy consumption of experiments in this study.
+      \item \textbf{Taming energy consumption variations in systems benchmarking}:
+      We investigate the phenomenon of variation when measuring the energy consumption of experiments in this study.
       In this paper, we discuss various hardware and software factors that can amplify variations in recorded energy measures, with a focus on the following research questions:
@@
 
-      \item \textbf{Evaluating the impact of Java virtual machines on energy consumption}: In this article, we thoroughly investigate how JVMs affect software energy usage.
-In order to address the following research concerns, we reveal through this study several trials on hundreds of JVMs versions provided by various providers:
+      \item \textbf{Evaluating the impact of Java virtual machines on energy consumption}:
+      In this paper, we thoroughly investigate how JVMs affect software energy usage.
+      To address the following research concerns, we reveal through this study several trials on hundreds of JVMs versions provided by various providers:
       \begin{description}
@@
 \end{enumerate}
+
+\subsection{Softwares and Tools}
+While the benchmarking and optimization parts were presented as chapters in this thesis, the measurement part was primarily based on developing tools that measure the energy consumption for a given use case.
+The following is a summary of the tools that were developed during this thesis:
+\begin{itemize}
+      \item \textbf{Jouleit} (\url{github.com/powerapi-ng/jouleit}): a tool that can be used to monitor energy consumption for any Linux program, this tool was used to compare the energy consumption of different JVMs;
+      \item \textbf{JRefferal} (\url{github.com/chakib-belgaid/jreferral}): a tool that allows the user to explore the JVM settings and their impact on the energy consumption of a given Java program.
+This tool was the result of the second article of this thesis;
+      \item \textbf{PyJoules} (\url{pypi.org/project/pyJoules}) is a software toolkit to measure the energy footprint of a host machine along the execution of a piece of Python code.
+It can measure the energy consumption on the level of script, function, and bloc of code;
+      \item \textbf{JouleHunter} (\url{pypi.org/project/joulehunter}): an energy profiler for python applications.
+It can be used to highlight the functions that consume the most energy in a given Python program.
+Its main usage is to help developers do an exploratory analysis of their application to scope the functions that should be optimized to be then targeted by PyJoules;
+      \item \textbf{GreenBoard} (\url{github.com/chakib-belgaid/greenboard}): a dashboard designed to help developers choose the best stack for their web application.
+It is based on the results of the third article of the last chapter.
+\end{itemize}
+
+\subsection{Future Submissions}
+\begin{itemize}
+      \item Reducing the energy consumption of Python using non-intrusive techniques,
+      \item Empirical analysis on the energy consumption of different web frameworks,
+      \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol),
+      \item THow do ORMs affect how much energy is used? A case study of Django and Flask.
+\end{itemize}
 

### 2022-10-13 chapters/introduction.tex
fiux : compilation error

@@
 While there are many fields where one can optimize energy consumption.
-Our focus is on the \emph{Information and Communication Technology} (ICT) sector, which is expected to account for around 4\% of \global \emph{GreenHouse Gas} (GHG) emissions in 2020, with an alarming 8\% growth rate, according to the French think tank The Shift Project.\footnote{\url{ https://www.theshiftproject.org/article/ict-environmental-impact/}}
+Our focus is on the \emph{Information and Communication Technology} (ICT) sector, which is expected to account for around 4\% of global \emph{GreenHouse Gas} (GHG) emissions in 2020, with an alarming 8\% growth rate, according to the French think tank The Shift Project.\footnote{\url{ https://www.theshiftproject.org/article/ict-environmental-impact/}}
 According to \emph{Statista}, the energy consumption of ICT increased from 4.3 exajoules in 2018 to 5.8 exajoules in 2025.\footnote{\url{https://www.statista.com/statistics/271139/energy-consumption-of-ict-worldwide}}

### 2022-10-13 chapters/conclusion.tex
add : notes to be discussed in the future works

@@
 \begin{enumerate}
-      \item ...
-      \item ...
-      \item ...
+      \item  vulgarisation of the energy consumption -> abstract numbers aren't enough 
+      \item  Evolution of the energy consumption according to python version 
+      \item Django vs Dlask vs Fast API, how do the ORMs affect the energy consumption while interacting with different web servers
+      \item Deeper analysis for some web frameworks ( what is the reason behind an increase in power consumption while scaling ) 
+      \item automation of the greenboard dashboard - the ability to run extra tests and add them to the dashboard 
+      \item implementing multiple physical clients to techempower benchmarks platform - synchronisation problem - 
+      \item a deeper analysis between the JVM optimizations and the JVM based programming languages 
+      \item using github copilot ( codelabs) to compare the energy consumption of generated code -> extend the work of programming langauges 
+      \item higher level progrmaming languges and the optimized target -> haxe -> java , check with other higher level lanagues 
+      \item impact of LLVM on the energy consumption of the target language  compilation flags ..etc
 
@@
 
+\subsection*{Representativeness} : 
+What is the purpose of doing optimizations if they cannot be applied to real-world applications? 
+\begin{enumerate}
+      \item CI/CD energy tests  - joulediff - 
+      \item Green commits 
+      \item lazy code, sacrificing the performance in order to get a better energy consumption --> green faas 
+      \item more representative benchmarks, using several CPU saturation levels 
+      \item measuring the energy impact instead of the row energy consumption.
+( the highway analogy when a single application consumes less energy, however the total energy consumption of the system increases )
+\end{enumerate}
 \vfill \strut  % to fill the rest of the page with blank lines

### 2022-10-14 chapters/introduction.tex
add : notes to be discussed in the future works
update: introduction

@@
 According to \emph{Statista}, the energy consumption of ICT increased from 4.3 exajoules in 2018 to 5.8 exajoules in 2025.\footnote{\url{https://www.statista.com/statistics/271139/energy-consumption-of-ict-worldwide}}
-
 \begin{figure}[!h]
@@
     \includegraphics[width=.7\linewidth]{chapters/distribution_of_ict_consumption.png}
-    \caption{Final energy consumption of digital technologies by item in 2017}
+    \caption{Final energy consumption of digital technologies by item in 2019(The Shift Project – Forecast Model 2021\cite{shift_2021})}
     \label{fig:distribution_of_ict_consumption}
@@
 
-Figure~\ref{fig:distribution_of_ict_consumption} shows the distribution of ICT consumption in 2018, where the largest chunk of the energy consumption is due to data usage, aka 55\% of the energy consumed where 35\% of this energy is used by data centers.
-Reducing energy consumption means reducing the impact 19\% of the ICT energy consumption has on the environment.
+Figure~\ref{fig:distribution_of_ict_consumption} shows the distribution of ICT consumption in 2018, where the largest chunk of the energy consumption is due to data usage, aka 63\% of the energy consumed where 22\% of this energy is used by data centers.
+Reducing energy consumption means reducing the impact 14\% of the ICT energy consumption has on the environment.
 
@@
 Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on the execution environment.
-Therefore, for the moment, we will start by formulating some hypotheses and exploring them using empirical analysis.
+Therefore, for the moment, we start by formulating some hypotheses and exploring them using empirical analysis.
 Figure~\ref{fig:thesis_position} highlights this thesis's position on the sustainability of ICT\@; while this thesis only addresses a small portion of ICT's energy usage, we feel it is a step in the right direction for additional solutions to mature to preserve humanity.
@@
 \end{figure}
-\subsection{Objectives}
+\section{Objectives}
 The purpose of this thesis is to help developers build more energy-efficient software.
@@
 The second part is to use some. opportunistic schedulers, which will stop the software if it is estimated to spend more energy.
-This strategy will be discussed first in Chapter~\ref{chapter:porgramming_langauges}, then we will continue with this strategy in the perspectives section.
+This strategy is discussed first in Chapter~\ref{chapter:porgramming_langauges}, then we continue with this strategy in the perspectives section.
 
-\subsection{Organization}
+\section{Organization}
 \begin{figure}[!h]
@@
 Figure~\ref{fig:thesis_contributions} summarizes the work completed for this thesis As mentioned in the previous section, our goal is to help developers reduce the energy consumption of their programs during execution.
-We will use an empirical approach to accomplish this.
-As a result, we will require some tools to assist us in this task.
-To begin, we will provide a tool to assist developers in measuring the energy consumption of their programs.
+We use an empirical approach to accomplish this.
+As a result, we require some tools to assist us in this task.
+To begin, we provide a tool to assist developers in measuring the energy consumption of their programs.
 The rest of the manuscript is outlined below.
@@
 \begin{itemize}
-    \item \cref{chapter:literature_review} provides a literature review of the energy consumption of software.
+    \item Chapter~\ref{chapter:literature_review} provides a literature review of the energy consumption of software.
           First, it introduces the challenges that one can encounter when running an empirical study, \emph{reproducibility}, \emph{accuracy}, and \emph{representativeness}, the ways in the state of the art overcome these challenges in the field of computer sciences.
@@
           In this section, we provide the literature approaches used to measure energy consumption and classify them, compare them and discuss their pros and cons when used for our research.
-After that, we will tackle the challenge of accuracy within the field of energy consumption, and discuss how the literature tried to overcome these challenges.
-Finally, we will present some of the most recent works in the field of software energy consumption optimization;
+After that, we tackle the challenge of accuracy within the field of energy consumption, and discuss how the literature tried to overcome these challenges.
+Finally, we  present some of the most recent works in the field of software energy consumption optimization;
           % We conclude this chapter by providing a simple method that we will use for the rest of the thesis to optimize the energy consumption of software.
 \textbf{BMO} : benchmark , measure and optimize.
-    \item \cref{chapter:benchmarking} will discuss two facets of empirical tests and how we adapt them to energy consumption measurements.
-          First, we will discuss the reproducibility challenge and how to overcome it with the usage of containers, then we will push this approach further by proposing and new protocol to make tests not only reproducible but able to be extended with other features to overcome the rapid pace of software evolution.
-          The second part of this chapter will discuss the aspect of accuracy when it comes to energy consumption measurements.
+    \item Chapter~\ref{chapter:benchmarking} discusses two facets of empirical tests and how we adapt them to energy consumption measurements.
+          First, we discuss the reproducibility challenge and how to overcome it with the usage of containers, then we push this approach further by proposing and new protocol to make tests not only reproducible but able to be extended with other features to overcome the rapid pace of software evolution.
+          The second part of this chapter discusses the aspect of accuracy when it comes to energy consumption measurements.
           This section demonstrates how energy measurements might differ and produce conflicting results for the same work when run on identical machines or even the same machine.
-          Then, it will provide some solutions to reduce this energy variation and improve the accuracy of the measurements;
-    \item After setting the ground for energy benchmarks in the previous chapter, we discuss energy consumption one of the most popular programming languages in the world, Python, in \cref{chapter:python}.
-          This chapter reports on how much python costs in terms of energy consumption compared to other languages.
+          Then, it provides some solutions to reduce this energy variation and improve the accuracy of the measurements;
+    \item After setting the ground for energy benchmarks in the previous chapter, we discuss the energy consumption of Python, one of the world's most popular programming languages.
+          Chapter~\ref{chapter:python} reports on how much python costs in terms of energy consumption compared to other languages.
           Then, it provides some statistics about its popularity and typical use cases and studies the impact of Python on energy consumption in three typical use cases, machine learning, web servers, and data manipulation.
@@
           Finally, we provide a non-intrusive way to optimize energy consumption without altering the code of the program.
-          We will do this by using a different implementation of the Python interpreter.
+          We achieve this by using a different implementation of the Python interpreter.
           This not only allows developers to spend less time optimizing their code but also allows them to use it on the legacy code that they cannot modify;
     \item Motivated by the outcomes of the non-intrusive optimization, we follow this strategy on one of the most popular legacy code base applications programming languages, Java.
-          In \cref{chapter:java}, we try to optimize the java code, using by changing the default JVM implementation.
-          We will compare the energy consumption of $12$ benchmarks using $52$ implementation, each benchmark is dedicated to a typical use case.
+          In Chapter~\ref{chapter:java}, we try to optimize the Java code, using by changing the default JVM implementation.
+          We compare the energy consumption of $12$ benchmarks using $52$ implementation, each benchmark is dedicated to a typical use case.
           After that, we study two of the JVM features, JIT and GC, and show their impact on the energy consumption of the Java code;
-    \item In contrast, \cref{chapter:porgramming_langauges} will use the flexibility provided by the micro-services architecture~\cite{dmitry2014micro} to analyze each programming language's energy behavior in light of various web scenarios to optimize the energy use of web services.
-          We will first examine the effects of the various programming languages when dealing with the \emph{Remote Procedure Call} (RPC) Protocol.
-          In this instance, we will use two scaling factors: the number of concurrent clients and the size of the requests.
-Then, we will compare $261$ web frameworks, each implementing the same website using seven use cases.
+    \item In contrast, Chapter~\ref{chapter:porgramming_langauges}  uses the flexibility provided by the micro-services architecture~\cite{dmitry2014micro} to analyze each programming language's energy behavior in light of various web scenarios to optimize the energy use of web services.
+          We first examine the effects of the various programming languages when dealing with the \emph{Remote Procedure Call} (RPC) Protocol.
+          In this instance, we use two scaling factors: the number of concurrent clients and the size of the requests.
+Then, we compare $261$ web frameworks, each implementing the same website using seven use cases.
           This analysis aims to look at how each technology uses energy in different kinds of web situations.
-    \item Finally, we conclude our work in \cref{chapter:conclusion} by summarizing the work done in this thesis and discussing the future work that can be done to further reduce the energy consumption of software.
+    \item Finally, we conclude our work in Chapter~\ref{chapter:conclusion} by summarizing the work done in this thesis and discussing the future work that can be done to further reduce the energy consumption of software.
 \end{itemize}
@@
 \subsection*{Tools}
+% \begin{itemize}
+%     \item \textbf{Jouleit} (\url{github.com/powerapi-ng/jouleit}):  \bibentry{Belgaid_JouleHunter_an_2021},
+%     \item \bibentry{Belgaid_JRefferal_Which_JVM_2021},
+%     \item \bibentry{Belgaid_Jouleit_a_2020},
+%           % \item \bibentry{Belgaid_Ijoules_Python_library_2020}
+%     \item \bibentry{Belgaid_Pyjoules_Python_library_2019}.
+% \end{itemize}
 \begin{itemize}
-    \item \bibentry{Belgaid_JouleHunter_an_2021},
-    \item \bibentry{Belgaid_JRefferal_Which_JVM_2021},
-    \item \bibentry{Belgaid_Jouleit_a_2020},
-          % \item \bibentry{Belgaid_Ijoules_Python_library_2020}
-    \item \bibentry{Belgaid_Pyjoules_Python_library_2019}.
+    \item \textbf{Jouleit} (\url{github.com/powerapi-ng/jouleit}): a tool that can be used to monitor energy consumption for any Linux program, this tool was used to compare the energy consumption of different JVMs;
+          \\
+          \bibentry{Belgaid_Jouleit_a_2020},
+    \item \textbf{JRefferal} (\url{github.com/chakib-belgaid/jreferral}): a tool that allows the user to explore the JVM settings and their impact on the energy consumption of a given Java program.
+This tool was the result of the second article of this thesis;
+          \\
+          \bibentry{Belgaid_JRefferal_Which_JVM_2021},
+    \item \textbf{PyJoules} (\url{pypi.org/project/pyJoules}) is a software toolkit to measure the energy footprint of a host machine along the execution of a piece of Python code.
+It can measure the energy consumption on the level of script, function, and bloc of code;
+          \\
+          \bibentry{Belgaid_Pyjoules_Python_library_2019},
+    \item \textbf{JouleHunter} (\url{pypi.org/project/joulehunter}): an energy profiler for python applications.
+It can be used to highlight the functions that consume the most energy in a given Python program.
+Its main usage is to help developers do an exploratory analysis of their application to scope the functions that should be optimized to be then targeted by PyJoules;
+          \\
+          \bibentry{Belgaid_JouleHunter_an_2021},
+    \item \textbf{GreenBoard} (\url{github.com/chakib-belgaid/greenboard}): a dashboard designed to help developers choose the best stack for their web application.
+It is based on the results of the third article of the last chapter.
 \end{itemize}
 
-% \subsection*{Posters}
+% \subsection*{Future Work}
 % \begin{itemize}
-%     \item \emph{PyRAPL\@: Make your python green again}
-%           % TODO : ask romain about the date and the name of the event 
+%     \item Reducing the energy consumption of Python using non-intrusive techniques,
+%     \item Empirical analysis on the energy consumption of different web frameworks,
+%     \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol),
+%     \item How do ORMs affect how much energy is used? A case study of Django and Flask.
 % \end{itemize}
-\subsection*{Future Work}
-\begin{itemize}
-    \item Reducing the energy consumption of Python using non-intrusive techniques,
-    \item Empirical analysis on the energy consumption of different web frameworks,
-    \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol),
-    \item THow do ORMs affect how much energy is used? A case study of Django and Flask.
-\end{itemize}
 
-\cleardoublepage
+% \cleardoublepage

### 2022-10-14 chapters/literature/draft_citations.tex
update : centring graphics
add : introduction
update: introduction

@@
 
-\newpage
+\clearpage
 \section{Benchmarking}\label{sec:soa_benchmarking}
@@
 \begin{figure}[!hbt]
-    \centering{\includegraphics{imgs/power-planes-rapl}}
+    \centering
+    \includegraphics[width=0.8\linewidth]{imgs/power-planes-rapl}
     \caption{INTEL RAPL scopes
@@
 The first examples of software measurement tools are PowerAPI~\cite{colmant2018next}, SmartWatts formula~\cite{fieni2020smartwatts} and SelfWatts~\cite{fieni2021selfwatts}.
-These tools collect global energy consumption measurements from RAPL and use other system events such as cache misses/hits and CPU frequency evolution (DVFS) via a sensor to construct a power model of the control groups (system control groups, docker containers, Kubernetes pods, etc.) using a Ridge regression.
+These tools collect global energy consumption measurements from RAPL and use other system events such as cache misses/hits and CPU frequency scaling (DVFS) via a sensor to construct a power model of the control groups (system control groups, docker containers, Kubernetes pods, etc.) using a Ridge regression.
 The model continuously learns and improves its real-time energy usage data precision with a maximum frequency of 100 Hz.
@@
 Unlike the performance, which is essentially related to the algorithm's complexity, the energy consumption is more impacted by the hardware.
-Therefore, to optimize the energy consumption, we choose a spiral method based on 3 phases:
+Therefore, to optimize energy consumption, we choose a spiral method based on 3 phases:
 \begin{enumerate}
-    \item executes the code,
-    \item measure the program,
-    \item infers the guidelines.
+    \item execute the code,
+    \item measure program's energy consumption,
+    \item infer the guidelines.
 \end{enumerate}

### 2022-10-14 chapters/conclusion.tex
update : some ideas about perspectives
add : some perspectives

@@
 
-\subsection{Published Articles}
+\subsection{Published Papers}
 This part summarizes the contributions that have been already accepted in conferences and journals.
@@
       \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol),
-      \item THow do ORMs affect how much energy is used? A case study of Django and Flask.
+      \item How do ORMs affect how much energy is used? A case study of Django .
 \end{itemize}
 
-\section{Future Work\note{missing}}
-.... Some potential areas for future efforts could include the following:
+\section{Future Work\note{ In Draft }}
+While our contributions are a good start to the energy-aware software engineering field, there are still many challenges to overcome.
+This is just the tip of the iceberg.
+The following are some of the challenges that we would like to address in the future:
+
+\subsection{Short Term Challenges}
+Before starting a journey looking for a new mine, one should first look at the resources available to him.
+In the same way, we need to first look at the resources available to us before we start looking for new challenges.
+The following are some of the challenges that we can address in the short term:
+
+\paragraph{The evolution of python interpreters}
+In Chapter \ref{chapter:python} we compared several python interpeters.
+However, most of these alternative solutions were based on python2 which is now deprecated.
+On the other hand, the default python interpreter (aka .
+CPython) has included many features and optimization since our last study, such as the introduction of Python introduced the \texttt{dataclasses} in version 3.7 (PEP 557\footnote{\url{https://peps.python.org/pep-0557/}}) that can be used to reduce the memory footprint of python objects,
+the new parser in python 3.9~\footnote{\url{https://docs.python.org/3/whatsnew/3.9.html}}, the user type alias in python3.10~\footnote{\url{https://peps.python.org/pep-0613/}}.
+And the most interesting changes for us occurred in CPython 3.11 which is claimed to be 25\% faster than python 3.10 \footnote{\url{https://github.com/faster-cpython/ideas}}.
+We need to re-evaluate the impact of these changes on the energy consumption of python programs.
+
+\paragraph{The impact of ORMS}
+We continue with our work with Python, and this time we will delve deeper into the impact of the ORMS, As shown in Section\ref{sec:webdev}, The ORMs are the most energy-consuming part of a web application.
+We intend to widen this analysis to other python ORMS such as SQLAlchemy\link{https://www.sqlalchemy.org/} and Peewee\link{http://docs.peewee-orm.com/en/latest/}.
+We will also explore the real relationship between the ORM, the database, and the web framework.
+
+\paragraph{Python and machine learning}
+In chapter \ref{chapter:python}, we showed that training accuracy has a huge cost on the energy using a single model.
+However, in the real world, we often use multiple models to solve a problem.
+We will explore the impact of using multiple models on the energy consumption of a machine learning application As well as compare the energy consumption of different machine learning frameworks such as PyTorch\link{https://pytorch.org/}, scikit-learn\link{https://scikit-learn.org/stable/} and TensorFlow \link{https://www.tensorflow.org/}.
+
+\paragraph{ Which is the greenest couple? a comparative study of JVMs and JVM based programming languages}
+As we have seen in Chapter \ref{chapter:java} the choice of the JVM can greatly impact the energy consumption of the the program.
+On the other hand, we have seen in Chapter \ref{chapter:porgramming_langauges} that other JVM-based languages depicted half power consumption of the JAVA code, was it only because of the web framework ?
+or there is a real difference between the JVMs and the JVM based languages?
+We will explore this question by comparing the energy consumption of different JVMs and JVM-based languages such as Scala, Kotlin, Groovy, and Clojure.
 
@@
       \item  vulgarisation of the energy consumption -> abstract numbers aren't enough
-      \item  Evolution of the energy consumption according to python version 
-      \item Django vs Dlask vs Fast API, how do the ORMs affect the energy consumption while interacting with different web servers
       \item Deeper analysis for some web frameworks ( what is the reason behind an increase in power consumption while scaling )
       \item automation of the greenboard dashboard - the ability to run extra tests and add them to the dashboard
-      \item implementing multiple physical clients to techempower benchmarks platform - synchronisation problem - 
+      \item implementing multiple physical clients to techempower benchmarks platform - synchronization problem -
       \item a deeper analysis between the JVM optimizations and the JVM based programming languages
-      \item using github copilot ( codelabs) to compare the energy consumption of generated code -> extend the work of programming langauges 
-      \item higher level progrmaming languges and the optimized target -> haxe -> java , check with other higher level lanagues 
-      \item impact of LLVM on the energy consumption of the target language  compilation flags ..etc
+      \item using github copilot (codelabs) to compare the energy consumption of generated code -> extend the work of programming langauges
+      \item higher level programming languages and the optimized target -> haxe -> java, check with other higher level languages
 
@@
 
-\subsection*{Representativeness} : 
+\subsection*{Representativeness}:
 What is the purpose of doing optimizations if they cannot be applied to real-world applications?
@@
 ( the highway analogy when a single application consumes less energy, however the total energy consumption of the system increases )
+      \item https://www.oneapi.io/open-source/
 \end{enumerate}

### 2022-10-15 chapters/conclusion.tex
add : extra perspectives

@@
 
-\paragraph{ Which is the greenest couple? a comparative study of JVMs and JVM based programming languages}
-As we have seen in Chapter \ref{chapter:java} the choice of the JVM can greatly impact the energy consumption of the the program.
+\paragraph{ Which is the greenest couple? a comparative study of JVMs and JVM-based programming languages}
+As we have seen in Chapter \ref{chapter:java} the choice of the JVM can greatly impact the energy consumption of the program.
 On the other hand, we have seen in Chapter \ref{chapter:porgramming_langauges} that other JVM-based languages depicted half power consumption of the JAVA code, was it only because of the web framework ?
@@
 
+\paragraph{the impact of programming languages on energy consumption, do translators helps in reducing the energy cost of a program?}
+One of the hardest challenges when comparing the energy consumption of multiple programming languages was the bias of the expertise of the programmer.
+One solution was to create a reference benchmark that allows one to compare several programming lanagues \link{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}, this was used for many researchers such as the work of \citeauthor{couto2017towards} \cite{couto2017towards}, others were to use some basic algorithms like \cite{noureddine_preliminary_2012} where they compare the energy consumption using the Hanoi tower problem \link{https://en.wikipedia.org/wiki/Tower_of_Hanoi} using different programming languages using the same algorithm.or other simple benchmarks using the Rosettacode base \link{https://rosettacode.org/wiki/Rosetta_Code}
+however this scope is limited to single algorithms.
+and does not help cover the production mode, this is why we shifted to the web frameworks.
+Now that are semantinc analysers issued from the openai project\link{https://openai.com/} like the automatic test generator ponicode\link{https://www.ponicode.com/} and the ai code geneator \link{https://github.com/features/copilot}
+an interesting new feature provided by the GitHub copilot is github copilot labs\link{https://githubnext.com/projects/copilot-labs/} is the ability to automatically translate the code from one language to another.
+We will explore the impact of this feature on the energy consumption of a program.
+
 \begin{enumerate}
@@
       \item Deeper analysis for some web frameworks ( what is the reason behind an increase in power consumption while scaling )
-      \item automation of the greenboard dashboard - the ability to run extra tests and add them to the dashboard
-      \item implementing multiple physical clients to techempower benchmarks platform - synchronization problem -
-      \item a deeper analysis between the JVM optimizations and the JVM based programming languages
-      \item using github copilot (codelabs) to compare the energy consumption of generated code -> extend the work of programming langauges
-      \item higher level programming languages and the optimized target -> haxe -> java, check with other higher level languages
 

### 2022-10-15 chapters/conclusion.tex
add : some drafts for perspectives

@@
 
+\paragraph{raising awareness of the energy consumption of the Softwares}
+While this thesis's main focus was to optimize energy consumption using comparative studies, it was easier to say that approach X is greener than approach Y no matter which metric we were using.
+however, it won't be the case for developers while measuring the energy consumption of their programs, some tried to give labels such as (A,B,C ..etc) while others transtaled this raw metrics in an equivalent of a fuel consumption (e.g.
+1 liter of fuel per 1000 lines of code).
+most of the approaches where to use the carbon emmision, such as\cite{patterson2021carbon} or ecograder~\link{https://ecograder.com/}
+we will explore these approaches and see how they can be used to rise the awareness of the energy consumption of their code
+
 \begin{enumerate}
-      \item  vulgarisation of the energy consumption -> abstract numbers aren't enough
+
       \item Deeper analysis for some web frameworks ( what is the reason behind an increase in power consumption while scaling )

### 2022-10-16 chapters/conclusion.tex
add : ci cd perspective draft

--- 2022-10-15 chapters/conclusion.tex
@@
 One of the hardest challenges when comparing the energy consumption of multiple programming languages was the bias of the expertise of the programmer.
-One solution was to create a reference benchmark that allows one to compare several programming lanagues \link{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}, this was used for many researchers such as the work of \citeauthor{couto2017towards} \cite{couto2017towards}, others were to use some basic algorithms like \cite{noureddine_preliminary_2012} where they compare the energy consumption using the Hanoi tower problem \link{https://en.wikipedia.org/wiki/Tower_of_Hanoi} using different programming languages using the same algorithm.or other simple benchmarks using the Rosettacode base \link{https://rosettacode.org/wiki/Rosetta_Code}
-however this scope is limited to single algorithms.
+One solution was to create a reference benchmark that allows one to compare several programming languages \link{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}, this was used for many researchers such as the work of \citeauthor{couto2017towards} \cite{couto2017towards}, others were to use some basic algorithms like \cite{noureddine_preliminary_2012} where they compare the energy consumption using the Hanoi tower problem \link{https://en.wikipedia.org/wiki/Tower_of_Hanoi} using different programming languages using the same algorithm.or other simple benchmarks using the Rosettacode base \link{https://rosettacode.org/wiki/Rosetta_Code}
+however, this scope is limited to single algorithms.
 and does not help cover the production mode, this is why we shifted to the web frameworks.
-Now that are semantinc analysers issued from the openai project\link{https://openai.com/} like the automatic test generator ponicode\link{https://www.ponicode.com/} and the ai code geneator \link{https://github.com/features/copilot}
-an interesting new feature provided by the GitHub copilot is github copilot labs\link{https://githubnext.com/projects/copilot-labs/} is the ability to automatically translate the code from one language to another.
+Now that our semanticc analyzers issued from the OpenAI project\link{https://openai.com/} like the automatic test generator ponicode\link{https://www.ponicode.com/} and the ai code generator \link{https://github.com/features/copilot}
+an interesting new feature provided by the GitHub copilot is GitHub copilot labs\link{https://githubnext.com/projects/copilot-labs/} is the ability to automatically translate the code from one language to another.
 We will explore the impact of this feature on the energy consumption of a program.
 
-\begin{enumerate}
-      \item  vulgarisation of the energy consumption -> abstract numbers aren't enough
-      \item Deeper analysis for some web frameworks ( what is the reason behind an increase in power consumption while scaling )
-
-\end{enumerate}
+\paragraph{raising awareness of the energy consumption of the Softwares}
+While this thesis's main focus was to optimize energy consumption using comparative studies, it was easier to say that approach X is greener than approach Y no matter which metric we were using.
+however, it won't be the case for developers while measuring the energy consumption of their programs, some tried to give labels such as (A,B,C ..etc) while others translated these raw metrics in an equivalent of fuel consumption (e.g.
+1 liter of fuel per 1000 lines of code).
+most of the approaches were to use the carbon emission, such as\cite{patterson2021carbon} or ecograder~\link{https://ecograder.com/}
+we will explore these approaches and see how they can be used to rise awareness of the energy consumption of their code
 
 \subsection*{Representativeness}:
+
+As we have seen in the state of the art, a successful benchmark meets three criteria, reproducibility, accuracy, and representativeness, During this thesis we discussed two of these criteria, reproducibility, and accuracy, however, we did not discuss the representativeness of our benchmarks.
+We relied on state-of-the-art benchmarks to represent real-world applications.
+However, the gap between these benchmarks and the industry is getting bigger and bigger due to the extreme pace of the software development.
+In this Section, we will tackle the issue of representativeness and discuss how we can improve our protocol to fit real-world applications.
 What is the purpose of doing optimizations if they cannot be applied to real-world applications?
+
+First, we will consider taking advantage of the popularity of CI/CD among developers to provide some insights into the energy consumption of their code.
+We did provide a sonar tool with our intern that can provide the energy consumption of JAVA-based code repository, figure~\ref{fig:JunitSonarplugin} shows the work of our intern that used a sonar plugin to highlight the energy evolution of java application across the time.
+This prototype got even more sophisticated to become Joule-dff\link{https://github.com/davidson-consulting/diff-jjoules}  is a tool to be added in the project continuous integration in order to highlight the energy evolution of the java programs.
+We intend to push this even further to detect the  commits responsible for an increase or the optimization within the source code
+
+\begin{figure}[!h]
+      \centering
+      \includegraphics[width=0.8\linewidth]{chapters/JunitSonarplugin}
+      \caption{Sonar plugin for Junit}
+      \label{fig:JunitSonarplugin}
+\end{figure}
+
 \begin{enumerate}
-      \item CI/CD energy tests  - joulediff -
-      \item Green commits
+
       \item lazy code, sacrificing the performance in order to get a better energy consumption --> green faas
@@
 ( the highway analogy when a single application consumes less energy, however the total energy consumption of the system increases )
-      \item https://www.oneapi.io/open-source/
 \end{enumerate}

### 2022-10-17 chapters/conclusion.tex
add : drafts

@@
 
+we urge energy-efficiency benchmark developers to report measurements at nonpeak activity levels for a more complete characterization of a system's energy behavior\cite{barroso2007case}.
+
+\subsubsection*{greenFaas}
+servers operate most of the time at between 10 and 50 percent of their maximum utilization levels \cite{barroso2007case}
+
 \begin{enumerate}

### 2022-10-17 chapters/conclusion.tex
Proofreading conclusion

--- 2022-10-15 chapters/conclusion.tex
@@
       \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol),
-      \item How do ORMs affect how much energy is used? A case study of Django .
+      \item How do ORMs affect how much energy is used? A case study of Django.
 \end{itemize}
@@
 This is just the tip of the iceberg.
-The following are some of the challenges that we would like to address in the future:
+The following perspectives are some of the challenges that we would like to address in the future.
 
-\subsection{Short Term Challenges}
-Before starting a journey looking for a new mine, one should first look at the resources available to him.
+\subsection{Short-Term Challenges}
+Before starting a journey looking for a new mine, one should first look at the available resources.
 In the same way, we need to first look at the resources available to us before we start looking for new challenges.
-The following are some of the challenges that we can address in the short term:
+The following are some of the challenges that we can address in the short term.
 
-\paragraph{The evolution of python interpreters}
-In Chapter \ref{chapter:python} we compared several python interpeters.
-However, most of these alternative solutions were based on python2 which is now deprecated.
-On the other hand, the default python interpreter (aka .
-CPython) has included many features and optimization since our last study, such as the introduction of Python introduced the \texttt{dataclasses} in version 3.7 (PEP 557\footnote{\url{https://peps.python.org/pep-0557/}}) that can be used to reduce the memory footprint of python objects,
-the new parser in python 3.9~\footnote{\url{https://docs.python.org/3/whatsnew/3.9.html}}, the user type alias in python3.10~\footnote{\url{https://peps.python.org/pep-0613/}}.
-And the most interesting changes for us occurred in CPython 3.11 which is claimed to be 25\% faster than python 3.10 \footnote{\url{https://github.com/faster-cpython/ideas}}.
-We need to re-evaluate the impact of these changes on the energy consumption of python programs.
+\paragraph{Studying the evolution of Python interpreters.}
+In Chapter~\ref{chapter:python}, we compared several python interpreters.
+However, most of these alternative solutions were based on Python\,2, which is now deprecated. 
+On the other hand, the default python interpreter (aka CPython) has included many features and optimization since our last study, such as the introduction of Python introduced the \texttt{dataclasses} in version 3.7 (PEP~557)\footnote{\url{https://peps.python.org/pep-0557/}} that can be used to reduce the memory footprint of Python objects, the new parser in Python\,3.9,\footnote{\url{https://docs.python.org/3/whatsnew/3.9.html}} the user type alias in Python\,3.10.\footnote{\url{https://peps.python.org/pep-0613/}}
+And, the most interesting changes for us occurred in CPython 3.11, which is claimed to be 25\% faster than Python\,3.10.\footnote{\url{https://github.com/faster-cpython/ideas}}.
+We need to re-evaluate the impact of these changes on the energy consumption of Python programs.
 
-\paragraph{The impact of ORMS}
-We continue with our work with Python, and this time we will delve deeper into the impact of the ORMS, As shown in Section\ref{sec:webdev}, The ORMs are the most energy-consuming part of a web application.
-We intend to widen this analysis to other python ORMS such as SQLAlchemy\link{https://www.sqlalchemy.org/} and Peewee\link{http://docs.peewee-orm.com/en/latest/}.
+\paragraph{Studying the impact of ORMs.}
+We continue with our work with Python, and this time we will dive deeper into the impact of the ORMs.
+As shown in Section~\ref{sec:webdev}, the ORMs are the most energy-consuming part of a web application.
+We intend to widen this analysis to other Python ORMs, such as \textsc{SQLAlchemy}\link{https://www.sqlalchemy.org/} and \textsc{Peewee}.\link{http://docs.peewee-orm.com/en/latest/}
 We will also explore the real relationship between the ORM, the database, and the web framework.
 
-\paragraph{Python and machine learning}
-In chapter \ref{chapter:python}, we showed that training accuracy has a huge cost on the energy using a single model.
-However, in the real world, we often use multiple models to solve a problem.
-We will explore the impact of using multiple models on the energy consumption of a machine learning application As well as compare the energy consumption of different machine learning frameworks such as PyTorch\link{https://pytorch.org/}, scikit-learn\link{https://scikit-learn.org/stable/} and TensorFlow \link{https://www.tensorflow.org/}.
+\paragraph{Studying the impact of Machine learning.}
+In Chapter~\ref{chapter:python}, we showed that training accuracy has a huge cost on the energy using a single model.
+However, in practice, we often use multiple models to address a given problem.
+We will explore the impact of using multiple models on the energy consumption of a machine learning application, as well as compare the energy consumption of different machine learning frameworks, such as \textsc{PyTorch},\link{https://pytorch.org/} \textsc{SciKit-Learn},\link{https://scikit-learn.org/stable/} and \textsc{TensorFlow}.\link{https://www.tensorflow.org/}
 
-\paragraph{ Which is the greenest couple? a comparative study of JVMs and JVM-based programming languages}
-As we have seen in Chapter \ref{chapter:java} the choice of the JVM can greatly impact the energy consumption of the program.
-On the other hand, we have seen in Chapter \ref{chapter:porgramming_langauges} that other JVM-based languages depicted half power consumption of the JAVA code, was it only because of the web framework ?
-or there is a real difference between the JVMs and the JVM based languages?
-We will explore this question by comparing the energy consumption of different JVMs and JVM-based languages such as Scala, Kotlin, Groovy, and Clojure.
+\paragraph{Studying the impact of JVM bytecode.}
+As we have seen in Chapter~\ref{chapter:java}, the choice of the JVM can greatly impact the energy consumption of hosted applications.
+On the other hand, we have seen in Chapter~\ref{chapter:porgramming_langauges} that other JVM-based languages depicted half power consumption of the Java code.
+Was it only because of the adopted web framework? 
+Or is there a real difference between the JVMs and the bytecode handled by JVM-based languages? 
+We will explore this question by comparing the energy consumption of different JVMs and JVM-based languages, such as Scala, Kotlin, Groovy, and Clojure.
 
-\paragraph{the impact of programming languages on energy consumption, do translators helps in reducing the energy cost of a program?}
+\paragraph{Revisiting the impact of programming languages.}
 One of the hardest challenges when comparing the energy consumption of multiple programming languages was the bias of the expertise of the programmer.
-One solution was to create a reference benchmark that allows one to compare several programming lanagues \link{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}, this was used for many researchers such as the work of \citeauthor{couto2017towards} \cite{couto2017towards}, others were to use some basic algorithms like \cite{noureddine_preliminary_2012} where they compare the energy consumption using the Hanoi tower problem \link{https://en.wikipedia.org/wiki/Tower_of_Hanoi} using different programming languages using the same algorithm.or other simple benchmarks using the Rosettacode base \link{https://rosettacode.org/wiki/Rosetta_Code}
-however this scope is limited to single algorithms.
-and does not help cover the production mode, this is why we shifted to the web frameworks.
-Now that are semantinc analysers issued from the openai project\link{https://openai.com/} like the automatic test generator ponicode\link{https://www.ponicode.com/} and the ai code geneator \link{https://github.com/features/copilot}
-an interesting new feature provided by the GitHub copilot is github copilot labs\link{https://githubnext.com/projects/copilot-labs/} is the ability to automatically translate the code from one language to another.
+One solution was to create a reference benchmark that allows one to compare several programming languages.\link{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}
+This benchmark has been adopted by many researchers, such as the work of \citeauthor{couto2017towards}~\cite{couto2017towards}, while others were to use some basic algorithms, like \cite{noureddine_preliminary_2012} who compare the energy consumption using the Hanoi tower problem\link{https://en.wikipedia.org/wiki/Tower_of_Hanoi} using different programming languages, or even simpler benchmarks, like the Rosettacode base.\link{https://rosettacode.org/wiki/Rosetta_Code}
+However, this scope is limited to single algorithms and does not help to cover production-like conditions, this is why we shifted to the web frameworks.
+Nonetheless, semantic analyzers issued from the OpenAI project\link{https://openai.com/} like the automatic test generator \textsc{Ponicode}\link{https://www.ponicode.com/} and AI code generator,\link{https://github.com/features/copilot} an interesting new feature provided by the GitHub copilot is GitHub copilot labs\link{https://githubnext.com/projects/copilot-labs/} is the ability to automatically translate the code from one language to another.
 We will explore the impact of this feature on the energy consumption of a program.
@@
 \paragraph{raising awareness of the energy consumption of the Softwares}
-While this thesis's main focus was to optimize energy consumption using comparative studies, it was easier to say that approach X is greener than approach Y no matter which metric we were using.
-however, it won't be the case for developers while measuring the energy consumption of their programs, some tried to give labels such as (A,B,C ..etc) while others transtaled this raw metrics in an equivalent of a fuel consumption (e.g.
-1 liter of fuel per 1000 lines of code).
-most of the approaches where to use the carbon emmision, such as\cite{patterson2021carbon} or ecograder~\link{https://ecograder.com/}
-we will explore these approaches and see how they can be used to rise the awareness of the energy consumption of their code
+While this thesis's main focus was to optimize energy consumption using comparative studies, it was easier to say that approach \texttt{X} is greener than approach \texttt{Y}, no matter which metric we were using.
+However, it will not be the case for developers when measuring the energy consumption of their programs.
+Some tried to give labels, such as (A,B,C\dots), while others translated these raw metrics in an equivalent of fuel consumption (e.g., 1 liter of fuel per 1000 lines of code).
+Most of the approaches were to use the carbon emission, such as~\cite{patterson2021carbon} or ecograder~\link{https://ecograder.com/} as a pivot metrics to quantify the environmental cost of osftware.
+We intend to explore these approaches and study how they can be used to increase awareness of the energy consumption of their code.
+
+\subsection*{Representativeness}:
+As reported in the state of the art, a successful benchmark meets three criteria: reproducibility, accuracy, and representativeness.
+This PhD thesis mostly discussed two of these criteria: reproducibility, and accuracy.
+However, we did not discuss the representativeness of our benchmarks.
+We relied on state-of-the-art benchmarks to represent real-world applications.
+Nevertheless, the gap between these benchmarks and the industry is getting bigger and bigger, due to the extreme pace of software development.
+In the future, we would like to tackle the issue of representativeness and discuss how we can improve our protocol to fit real-world applications.
+What is the purpose of doing optimizations if they cannot be applied to real-world applications?
+
+First, we will consider taking advantage of the popularity of CI/CD among developers to provide some insights into the energy consumption of their code.
+We did prototype a sonar tool that can report on the energy consumption of Java-based code repository.
+Figure~\ref{fig:JunitSonarplugin} shows the work of our intern who used a Sonar plugin to highlight the energy evolution of Java applications along time.
+This prototype got even more sophisticated to become Joule-dff\link{https://github.com/davidson-consulting/diff-jjoules}, which is a tool to be added in a continuous integration to highlight the energy evolution of the Java programs.
+We intend to push this even further to detect the commits responsible for an increase or the optimization within the source code.
+
+\begin{figure}[!h]
+      \centering
+      \includegraphics[width=0.8\linewidth]{chapters/JunitSonarplugin}
+      \caption{Sonar plugin for Junit}
+      \label{fig:JunitSonarplugin}
+\end{figure}
 
@@
 
-      \item Deeper analysis for some web frameworks ( what is the reason behind an increase in power consumption while scaling )
-
-\end{enumerate}
-
-\subsection*{Representativeness}:
-What is the purpose of doing optimizations if they cannot be applied to real-world applications?
-\begin{enumerate}
-      \item CI/CD energy tests  - joulediff -
-      \item Green commits
       \item lazy code, sacrificing the performance in order to get a better energy consumption --> green faas

### 2022-10-18 chapters/conclusion.tex
add : extra figures for greenfaas

@@
 Figure~\ref{fig:JunitSonarplugin} shows the work of our intern who used a Sonar plugin to highlight the energy evolution of Java applications along time.
-This prototype got even more sophisticated to become Joule-dff\link{https://github.com/davidson-consulting/diff-jjoules}, which is a tool to be added in a continuous integration to highlight the energy evolution of the Java programs.
+This prototype got even more sophisticated to become Joule-dff\link{https://github.com/davidson-consulting/diff-jjoules}, which is a tool to be added in continuous integration to highlight the energy evolution of the Java programs.
 We intend to push this even further to detect the commits responsible for an increase or the optimization within the source code.
@@
 
+\begin{figure}[!h]
+      \centering
+      \caption{Evolution of CPU power}
+      \label{fig:power_evolution_greenfaas}
+      \includegraphics[width=\linewidth]{chapters/power_evolution_greenfaas}
+
+\end{figure}
+
+\begin{figure}[!h]
+      \centering
+      \caption{Evolution of CPU behavior during multiple contexts }
+      \label{fig:green_faas}
+      \includegraphics[width=\linewidth]{chapters/green_faas_duration}
+      \includegraphics[width=\linewidth]{chapters/green_faas_energy}
+      \includegraphics[width=\linewidth]{chapters/green_faas_power}
+      \includegraphics[width=\linewidth]{chapters/green_faas_impact}
+      \includegraphics[width=\linewidth]{chapters/soa_energy_efficiency}
+
+\end{figure}
+
+exp graph coloration benchmark
+\url{https://github.com/chakib-belgaid/lazy-green-code-test}
+
+setup
+Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz
+8 cores 16 GB RAM
 \begin{enumerate}

### 2022-10-18 chapters/conclusion.tex
Proofreading

@@
 As reported in the state of the art, a successful benchmark meets three criteria: reproducibility, accuracy, and representativeness.
-This PhD thesis mostly discussed two of these criteria: reproducibility, and accuracy.
+This PhD thesis mostly discussed two of these criteria: reproducibility and accuracy.
 However, we did not discuss the representativeness of our benchmarks.
@@
 First, we will consider taking advantage of the popularity of CI/CD among developers to provide some insights into the energy consumption of their code.
-We did prototype a sonar tool that can report on the energy consumption of Java-based code repository.
+We did prototype a Sonar plugin that can report on the energy consumption of Java-based code repository.
 Figure~\ref{fig:JunitSonarplugin} shows the work of our intern who used a Sonar plugin to highlight the energy evolution of Java applications along time.
-This prototype got even more sophisticated to become Joule-dff\link{https://github.com/davidson-consulting/diff-jjoules}, which is a tool to be added in continuous integration to highlight the energy evolution of the Java programs.
+This prototype got even more sophisticated to become Joule-diff,\link{https://github.com/davidson-consulting/diff-jjoules} which is a tool to be included in a CI/CD to highlight the energy evolution of the Java programs.
 We intend to push this even further to detect the commits responsible for an increase or the optimization within the source code.
@@
       \includegraphics[width=0.8\linewidth]{chapters/JunitSonarplugin}
-      \caption{Sonar plugin for Junit}
+      \caption{Sonar Energy plugin for JUnit}
       \label{fig:JunitSonarplugin}
@@
 
-we urge energy-efficiency benchmark developers to report measurements at nonpeak activity levels for a more complete characterization of a system's energy behavior\cite{barroso2007case}.
+We urge energy-efficiency benchmark developers to report measurements at non-peak activity levels for a more complete characterization of a system's energy behavior~\cite{barroso2007case}.
 
-\subsubsection*{greenFaas}
-servers operate most of the time at between 10 and 50 percent of their maximum utilization levels \cite{barroso2007case}
+\subsubsection*{GreenFaaS}
+servers operate most of the time at between 10 and 50 percent of their maximum utilization levels~\cite{barroso2007case}.
 
@@
       \includegraphics[width=\linewidth]{chapters/power_evolution_greenfaas}
-
 \end{figure}
@@
       \includegraphics[width=\linewidth]{chapters/soa_energy_efficiency}
-
 \end{figure}
@@
 \begin{enumerate}
-
-      \item lazy code, sacrificing the performance in order to get a better energy consumption --> green faas
+      \item lazy code, sacrificing the performance to get a better energy consumption --> green faas
       \item more representative benchmarks, using several CPU saturation levels

### 2022-10-18 chapters/conclusion.tex
add : first part of greenfaas

--- 2022-10-18 chapters/conclusion.tex
@@
 As reported in the state of the art, a successful benchmark meets three criteria: reproducibility, accuracy, and representativeness.
-This PhD thesis mostly discussed two of these criteria: reproducibility, and accuracy.
+This Ph.D. thesis mostly discussed two of these criteria: reproducibility, and accuracy.
 However, we did not discuss the representativeness of our benchmarks.
@@
 In the future, we would like to tackle the issue of representativeness and discuss how we can improve our protocol to fit real-world applications.
-What is the purpose of doing optimizations if they cannot be applied to real-world applications?
+After all, what is the purpose of doing optimizations if they cannot be applied to real-world applications?
 
@@
 We did prototype a sonar tool that can report on the energy consumption of Java-based code repository.
-Figure~\ref{fig:JunitSonarplugin} shows the work of our intern who used a Sonar plugin to highlight the energy evolution of Java applications along time.
+Figure~\ref{fig:JunitSonarplugin} shows the work of our intern who used a Sonar plugin to highlight the energy evolution of Java applications over time.
 This prototype got even more sophisticated to become Joule-dff\link{https://github.com/davidson-consulting/diff-jjoules}, which is a tool to be added in continuous integration to highlight the energy evolution of the Java programs.
 We intend to push this even further to detect the commits responsible for an increase or the optimization within the source code.
-
 \begin{figure}[!h]
@@
 
-we urge energy-efficiency benchmark developers to report measurements at nonpeak activity levels for a more complete characterization of a system's energy behavior\cite{barroso2007case}.
+\subsubsection{Energy impact vs Energy consumption}
+To make the software more eco-friendly, both state-of-the-art and this thesis focused on reducing software energy consumption.
+To accomplish this, we sought to isolate the energy consumption of the software under test as much as possible, primarily by running the tests on an isolated machine that runs only a few services.
+Others tried to infer the energy consumption of a single process using additional metrics that vary from the CPU usage, such as \cite{noureddine-issta-2016} to more sophisticated models like~\cite{fieni2020smartwatts,fieni2021selfwatts}.
 
-\subsubsection*{greenFaas}
-servers operate most of the time at between 10 and 50 percent of their maximum utilization levels \cite{barroso2007case}
+However, by the end of this thesis, I realized that isolating the energy consumption by running the tests on an isolated machine does not represent real-world conditions; many processes are running in parallel.
+On the other hand, our benchmark can compete with other processes for resources such as CPU, memory, and network.
+Therefore, this program can increase the energy consumption of other processes running on the same machine, although its energy consumption is low.
+
+This is why I've started to work on a new project called \textsc{Energy Impact} that aims to measure the energy impact of software.
+According to the work of \citeauthor{barroso2007case}~\cite{barroso2007case}, servers operate most of the time at between 10 and 50 percent of their maximum utilization levels.
+Moreover, as shown in Figure~\ref{fig:soa_energy_efficiency} with 50\% of the servers' capacity, the server operates only with  75\% of its maximum power.
+In other words, it will cost us only 25\% of the energy to operate the extra 50\% of the servers' capacity.
+
+\begin{figure}[!h]
+      \centering
+      \includegraphics[width=0.9\linewidth,keepaspectratio]{chapters/soa_energy_efficiency}
+      \caption{Power efficiency vs power consumption\cite{barroso2007case}}
+      \label{fig:soa_energy_efficiency}
+\end{figure}
 
@@
       \includegraphics[width=\linewidth]{chapters/power_evolution_greenfaas}
+\end{figure}
 
+In order to confirm whether this hypothesis still holds. We run the following experiment.
+First, we took a machine with minimal services, then created a process that stresses one core by doing some mathematical operations, and each second we introduced a copy of this process.
+after a while, we removed these copies one by one.
+Figure~\ref{fig:power_evolution_greenfaas} displays the machine's power consumption during this experiment.
+As one can notice, the introduction of the first copies had a greater cost than the other ones, despite not sharing the same core.
+And the more we approach the peak of power, the less significance the introduction of a new process has on power consumption to be almost zero after the 9th copy.
+One should note that this experiment was conducted on a machine with the following configuration:
+Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz, with 4 physical cores and 4 hyperthreads, 16GB RAM, and 687GB HDD.
+Therefore with the introduction of the eighth copy, we have already saturated the machine's capacity, and the introduction of the ninth copy had no impact on the power consumption.
+
+\begin{figure}[!h]
+      \centering
+      \caption{Evolution of CPU behavior during multiple contexts }
+      \label{fig:green_faas}
+      \includegraphics[width=\linewidth]{chapters/green_faas_duration}
 \end{figure}
@@
       \label{fig:green_faas}
-      \includegraphics[width=\linewidth]{chapters/green_faas_duration}
       \includegraphics[width=\linewidth]{chapters/green_faas_energy}
+\end{figure}
+
+\begin{figure}[!h]
+      \centering
+      \caption{Evolution of CPU behavior during multiple contexts }
+      \label{fig:green_faas}
       \includegraphics[width=\linewidth]{chapters/green_faas_power}
+\end{figure}
+
+\begin{figure}[!h]
+      \centering
+      \caption{Evolution of CPU behavior during multiple contexts }
+      \label{fig:green_faas}
       \includegraphics[width=\linewidth]{chapters/green_faas_impact}
-      \includegraphics[width=\linewidth]{chapters/soa_energy_efficiency}
-
 \end{figure}
@@
 
-setup
-Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz
-8 cores 16 GB RAM
 \begin{enumerate}

### 2022-10-18 chapters/conclusion.tex
add : power analysis

@@
 And the more we approach the peak of power, the less significance the introduction of a new process has on power consumption to be almost zero after the 9th copy.
+
 One should note that this experiment was conducted on a machine with the following configuration:
-Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz, with 4 physical cores and 4 hyperthreads, 16GB RAM, and 687GB HDD.
+Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz, with 4 physical cores and 4 hyper threads, 16GB RAM, and 687GB HDD.
 Therefore with the introduction of the eighth copy, we have already saturated the machine's capacity, and the introduction of the ninth copy had no impact on the power consumption.
+\begin{figure}[!h]
+      \centering
+      \includegraphics[width=\linewidth]{chapters/green_faas_duration}
+      \caption{Evolution of CPU behavior during multiple contexts }
+      \label{fig:green_faas_duration}
+\end{figure}
+
+To push this experiment further, we created a stressing tool that allows us to fix the number of cores and \% usage of each core to simulate a production context.
+we then run a benchmark( graph coloration algrothim\link{https://github.com/chakib-belgaid/lazy-green-code-test/tree/master/exp01}) using multiple configurations.
+Figure~\ref{fig:green_faas_duration} displays the overall performance of the benchmark. As one can see, the execution time remained the same when the context was using 3 cores at any saturation level,
+this is because we have 4 physical CPU cores, therefore each process was running a different core,(3 cores were used by the context and the other one was used by the benchmark).
+However, behavior changes when the context consumes more than 3 cores.
+CPU saturation tends to have a great impact on the performance of the benchmark.
+when the benchmark was using a hyper thread of the same core as one of the context workers it took almost double the time to finish the same job.
+As when we saturated only 50\% of the CPU capacity, this increase didn't happen until the context used 7 cores, and in the worst-case scenario ( 10 extra processes) it took only 75 seconds to finish the job.
+
+\begin{figure}[!h]
+      \centering
+      \includegraphics[width=\linewidth]{chapters/green_faas_power}
+      \caption{Evolution of CPU behavior during multiple contexts }
+      \label{fig:green_faas_power}
+\end{figure}
+
+On the other hand, as we can see in Figure~\ref{fig:green_faas_power}, the average power behavior of the system had different behavior.
+With the 100\% saturation scenario, the power consumption hit its peak as soon as we hit the 4 core usage 3 from the context and 1 from the benchmark), and it kept fluctuating until the 8th core saturation, after that it was in a constant decrease.
+This was because the processes were competing with each other for cache usage, which led to a significant number of cache defaults that slowed down the CPU frequency which led to a decrease in power consumption.
+
+However, when we used 50\% of the CPU capacity, the power kept increasing by adding the number of extra processes.
+As for the scenarios where we used more than 50\% of the CPU capacity, the power consumption kept increasing until it hit its peak then it start to decrease; this number of processes needed to reach the peak varies according to the saturation level.
 
@@
       \caption{Evolution of CPU behavior during multiple contexts }
-      \label{fig:green_faas}
-      \includegraphics[width=\linewidth]{chapters/green_faas_duration}
+      \label{fig:green_faas_energy}
+      \includegraphics[width=\linewidth]{chapters/green_faas_energy}
 \end{figure}
+
+On the other hand, as shown in Figure~\ref{fig:green_faas_energy} the energy started increasing from the beginning, this is due to an increase in power that we will discuss
 
@@
       \caption{Evolution of CPU behavior during multiple contexts }
-      \label{fig:green_faas}
-      \includegraphics[width=\linewidth]{chapters/green_faas_energy}
-\end{figure}
-
-\begin{figure}[!h]
-      \centering
-      \caption{Evolution of CPU behavior during multiple contexts }
-      \label{fig:green_faas}
-      \includegraphics[width=\linewidth]{chapters/green_faas_power}
-\end{figure}
-
-\begin{figure}[!h]
-      \centering
-      \caption{Evolution of CPU behavior during multiple contexts }
-      \label{fig:green_faas}
+      \label{fig:green_faas_impact}
       \includegraphics[width=\linewidth]{chapters/green_faas_impact}
@@
 
-exp graph coloration benchmark
 \url{https://github.com/chakib-belgaid/lazy-green-code-test}

### 2022-10-18 chapters/conclusion.tex
add : energy impact

@@
 
-In order to confirm whether this hypothesis still holds. We run the following experiment.
+In order to confirm whether this hypothesis still holds. We run the following experiment\link.
 First, we took a machine with minimal services, then created a process that stresses one core by doing some mathematical operations, and each second we introduced a copy of this process.
 after a while, we removed these copies one by one.
+The source code of experiment can be found in the following GitHub repository\link{https://github.com/chakib-belgaid/lazy-green-code-test}
 Figure~\ref{fig:power_evolution_greenfaas} displays the machine's power consumption during this experiment.
@@
 One should note that this experiment was conducted on a machine with the following configuration:
-Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz, with 4 physical cores and 4 hyper threads, 16GB RAM, and 687GB HDD.
+Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz, with 4 physical cores and 4 hyper threads, 16GB RAM, and 687GB HDD. As for the operating system, we used Ubuntu 20.04.2 LTS.
+
 Therefore with the introduction of the eighth copy, we have already saturated the machine's capacity, and the introduction of the ninth copy had no impact on the power consumption.
@@
 As for the scenarios where we used more than 50\% of the CPU capacity, the power consumption kept increasing until it hit its peak then it start to decrease; this number of processes needed to reach the peak varies according to the saturation level.
-
 \begin{figure}[!h]
@@
 
-On the other hand, as shown in Figure~\ref{fig:green_faas_energy} the energy started increasing from the beginning, this is due to an increase in power that we will discuss
+When we try to measure the energy consumption of this experiment, we find that the scenario where we used 100\% of the energy consumption was the worst case by faar, and it kept increasing whenever we increase the number of core usage.
+Figure~\ref{fig:green_faas_energy} present the energy consumption of the system during the experiment.
 
-\begin{figure}[!h]
+However, if we calculate the energy impact of the benchmark which we define as: 
+\begin{equation}
+      Ipact_{Benchmark} =E_{Benchmark} - E_{System_before}
+\end{equation}
+
+Wich can be approximated by: 
+\begin{equation}
+      Ipact_{Benchmark} =E_{Benchmark} - P_{System_before} * T_{Benchmark}
+\end{equation}
+One can notice that the energy impact of our benchmark did not have the same behavior as the system's energy consumption.
+Sometimes introducing the benchmark had a \emph{negative} impact on the system's energy consumption, which made it even greener.
+This is because it reduces power consumption.This is shown in Figure~\ref{fig:green_faas_impact}
+
+We are aware that during this analysis, we have not considered the execution time of the context, and it will be affected by the benchmark - mainly, it will increase when we use more cores and/or higher CPU saturation -.
+However, this action was motivated by the fact that most of the could Software are web services, as we have seen in chapter\ref{chapter:porgramming_langauges} where the execution time is not bounded, and each service can run for an unlimited amount of time.
+
+\section{Conclusion}
+This experiment shows that the context can have a significant impact on the energy consumption of the system, and sometimes measuring only the energy consumption of an application does not represent the real impact of this application on the energy consumption of the whole system, especially when we are dealing with timeless Softwares aka services.
+In the future, we tend to extend this experiment with more data, such as the energy consumption of the application itself instead of the whole system using \cite{fieni2020smartwatts} and comparing the strategies that were mentioned in this thesis using multiple contexts instead of just a minimal version of the operating system.
+
+\begin{figure}[!t]
       \centering
       \caption{Evolution of CPU behavior during multiple contexts }
+      \includegraphics[width=\linewidth]{chapters/green_faas_impact}\
       \label{fig:green_faas_impact}
-      \includegraphics[width=\linewidth]{chapters/green_faas_impact}
 \end{figure}
 
-\url{https://github.com/chakib-belgaid/lazy-green-code-test}
-
-\begin{enumerate}
-      \item lazy code, sacrificing the performance to get a better energy consumption --> green faas
-      \item more representative benchmarks, using several CPU saturation levels
-      \item measuring the energy impact instead of the row energy consumption.
+% \begin{enumerate}
+%       \item lazy code, sacrificing the performance to get a better energy consumption --> green faas
+%       \item more representative benchmarks, using several CPU saturation levels
+%       \item measuring the energy impact instead of the row energy consumption.
 ( the highway analogy when a single application consumes less energy, however the total energy consumption of the system increases )
-\end{enumerate}
+% \end{enumerate}
 \vfill \strut  % to fill the rest of the page with blank lines

### 2022-10-18 chapters/benchmarking/conclusion.tex
update : conculsion

@@
 \section{Conclusion}
-As seen in Section~\ref{sec:soa_benchmarking} of state of the art, a successful benchmark faces three challenges: reproducibility, accuracy, and representativeness. 
+As seen in Section~\ref{sec:soa_benchmarking} of state of art, a successful benchmark faces three challenges: reproducibility, accuracy, and representativeness. 
 This chapter has covered two of the three criteria.

### 2022-10-18 chapters/conclusion.tex
update : conculsion

@@
 Chapter~\ref{chapter:literature_review} discussed the challenges of a successful benchmarking strategy: reproducibility, accuracy, and representativeness.
-We concluded that software containers, like "Docker", would be the best fit to ensure that energy studies could be reproduced.
+We concluded that software containerization technologies, like "Docker", are the best fit to ensure that energy studies can be reproduced.
 We then extended this reproducibility to an evolving protocol that helps researchers keep up with the rapid pace of software development.
@@
 
-\subsection{Published Papers}
-This part summarizes the contributions that have been already accepted in conferences and journals.
+\paragraph*{Benchmarking Protocol to Measure Software Energy Consumption}
+During this chapter, we tackled two of the main challenges faced by researchers when using empirical analysis to measure software energy consumption: \emph{reproducibility}, \emph{accuracy}.
+First, we discussed the challenge of reproducibility by looking at the existing techniques.
+We have seen two main methods to encapsulate experimental systems.
+The first is to use virtual machines, and the second is to use containers.
+We established that Docker is more suitable for energy-related studies.
+Then we extended this reproducibility to an evolving protocol that helps researchers keep up with the rapid pace of software development.
 
-\begin{enumerate}
-      \item \textbf{Taming energy consumption variations in systems benchmarking}:
-            We investigate the phenomenon of variation when measuring the energy consumption of experiments in this study.
-            In this paper, we discuss various hardware and software factors that can amplify variations in recorded energy measures, with a focus on the following research questions:
+Finally, We examined the phenomenon of variation when measuring the energy consumption of experiments in order to increase their accuracy. To do so, we investigated the following research questions.:
 \begin{description}
@@
 
-            This contribution shows how processor features can significantly affect the variation in energy use between the benchmarking protocol and the operating system.
-            Finally, this study presents several guidelines for controllable parameters that practitioners could easily change to increase the accuracy of their experiments.
+We discovered that processor features and the operating system can have a significant impact on the variation in energy consumption between the benchmarking protocol.
 
+Finally, we presented several guidelines for controllable parameters that practitioners could easily modify to improve the precision of their experiments.
+
+The results of this work have been published in the following paper:
+\\
 \bibentry{ournani2020taming}
 
-      \item \textbf{Evaluating the impact of Java virtual machines on energy consumption}:
-            In this paper, we thoroughly investigate how JVMs affect software energy usage.
-            To address the following research concerns, we reveal through this study several trials on hundreds of JVMs versions provided by various providers:
-            \begin{description}
-                  \item[\textsc{RQ}~1:] What is the impact of existing JVM distributions on the energy consumption of Java-based software services?
-                  \item [\textsc{RQ}~1:] What are the relevant JVM settings that can reduce the energy consumption of a given software service?
-            \end{description}
-            The findings demonstrate that choosing the right JVM platform can significantly reduce energy usage depending on the software and use case.
-            This optimization can also be achieved by properly configuring the JIT and GC parameters.
+\paragraph*{Impact of Energy-saving Strategies in the Python Ecosystem}
+During this chapter, we tackled the challenge of reducing the energy consumption of Python code.
+To do so, we investigated the energy behavior of the python during its most common usage:
 
-            \bibentry{ournani2021evaluating}
-\end{enumerate}
+\begin{itemize}
+      \item \emph{machine learning}:  we studied the energy consumption of a machine learning algorithm using a benchmark of $60,000$ entries to train the \texttt{cifar10-fast} model.
+To confirm that the energy consumption of a model is highly impacted by its accuracy, increasing the accuracy from 85\% to 95\% can lead to double the energy consumption.
+      \item \emph{web development}: We examined the Django-based application and discovered that fetching data from the database consumes most of the energy.
+As a result, we compared three methods for retrieving data from two different databases.
+We discovered that using an optimized strategy can reduce energy consumption by $20\times$.
+      \item \emph{Data Processing}: In this section, we studied the energy consumption of data structures, iteration algorithms, and concurrency.
+following the guidelines and prioritizing built-in functions is the best way to reduce the energy consumption of Python applications.
+      As for the concurrency, we found out that the \textsf{Multithreading} strategy can lead to a decrease in energy consumption despite the increase in execution time; the \textsf{Multiprocessing} was the optimal option, however, it is greatly affected by the number of processes.
+\end{itemize}
+After that, We presented a non-intrusive technique for optimizing the energy consumption of Python-based apps without requiring significant code changes.
+This technique involves using a different Python runtime implementation.
+We began by categorizing and filtering these implementations into three major classes (compiler, interpreter, and extra libraries).
+Then we ran a series of tests to compare the energy consumption of these options.
+Our findings point to the lack of a general solution and the importance of customizing the Python runtime for each application.
+We discovered that while most interpreters consumed as much or more energy than the official CPython implementation, the JIT-based ones tend to consume significantly less energy.
 
-\subsection{Softwares and Tools}
-While the benchmarking and optimization parts were presented as chapters in this thesis, the measurement part was primarily based on developing tools that measure the energy consumption for a given use case.
-The following is a summary of the tools that were developed during this thesis:
+This work resulted in the development of two tools to measure the energy consumption of Python code: 
 \begin{itemize}
-      \item \textbf{Jouleit} (\url{github.com/powerapi-ng/jouleit}): a tool that can be used to monitor energy consumption for any Linux program, this tool was used to compare the energy consumption of different JVMs;
-      \item \textbf{JRefferal} (\url{github.com/chakib-belgaid/jreferral}): a tool that allows the user to explore the JVM settings and their impact on the energy consumption of a given Java program.
-This tool was the result of the second article of this thesis;
       \item \textbf{PyJoules} (\url{pypi.org/project/pyJoules}) is a software toolkit to measure the energy footprint of a host machine along the execution of a piece of Python code.
@@
 Its main usage is to help developers do an exploratory analysis of their application to scope the functions that should be optimized to be then targeted by PyJoules;
-      \item \textbf{GreenBoard} (\url{github.com/chakib-belgaid/greenboard}): a dashboard designed to help developers choose the best stack for their web application.
-It is based on the results of the third article of the last chapter.
      \end{itemize}
 
-\subsection{Future Submissions}
-\begin{itemize}
-      \item Reducing the energy consumption of Python using non-intrusive techniques,
-      \item Empirical analysis on the energy consumption of different web frameworks,
-      \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol),
-      \item How do ORMs affect how much energy is used? A case study of Django.
-\end{itemize}
+\paragraph*{Impact of Java Virtual Machine Configurations on Energy Consumption}
 
-\section{Future Work\note{ In Draft }}
+In this chapter, we thoroughly explored how JVMs affect software energy consumption by investing the following research questions:
+\begin{description}
+      \item[\textsc{RQ}~1:] What is the impact of existing JVM distributions on the energy consumption of Java-based software services?
+      \item [\textsc{RQ}~1:] What are the relevant JVM settings that can reduce the energy consumption of a given software service?
+\end{description}
+To do so, we compared the energy consumption of $12$ benchmarks using $52$ implementation, each benchmark is dedicated to a typical use case.
+After that, we studied two of the JVM features, JIT and GC, and show their impact on the energy consumption of the Java code;
+The findings demonstrate that choosing the right JVM platform can significantly reduce energy usage depending on the software and use case, and this optimization can also be achieved by properly configuring the JIT and GC parameters.
+
+This work resulted in the development of software called \textbf{JRefferal} (\url{github.com/chakib-belgaid/jreferral}) that allows the user to explore the JVM settings and their impact on the energy consumption of a given Java program and a publication of the following paper :
+\\
+\bibentry{ournani2021evaluating}
+
+\paragraph*{Energy Footprint of Distributed Programming Frameworks}
+Finally, in Chapter~\ref{chapter:porgramming_langauges}, we extended the work of \citeauthor{pereira_energy_2017}~\cite{pereira_energy_2017} by studying the energy footprint on other aspects than micro-benchmarking by investigating the impact of the programming language on the energy consumption of web services.
+In the first study, we used the official implementation of \textsc{gRPC} library to build $25$ frameworks using $17$ programming languages. 
+As for the second study, we looked at how web frameworks affect energy consumption by creating a basic web application and comparing its performance, latency, and energy usage using the most common web frameworks.
+We employed $261$ implementation using $35$ distinct programming languages and three different databases.
+For each implementation, we considered $7$ representative scenarios.
+Each scenario is tested with several stress levels to study the scalability of the web frameworks.
+Therefore, the final results cover $8,750$ use cases.
+
+Both studies revealed that each framework had different behaviors toward scalability.
+While the absence of a clear winner, we found out that most compiled languages are more energy-efficient because of their performance, Java tends to depict more power, and the interpreted languages lacked in terms of performance while keeping a low power consumption.
+PHP was an exception as its performance was compared to the compiled languages.
+The results of this work resulted in creating a dashboard to help developers choose the right programming language and framework for their application based on the context.
+The dashboard can be found at (\url{github.com/chakib-belgaid/greenboard}):
+
+% \subsection{Future Submissions}
+% \begin{itemize}
+%       \item Reducing the energy consumption of Python using non-intrusive techniques,
+%       \item Empirical analysis on the energy consumption of different web frameworks,
+%       \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol),
+%       \item How do ORMs affect how much energy is used? A case study of Django.
+% \end{itemize}
+
+\section{Future Work}
 While our contributions are a good start to the energy-aware software engineering field, there are still many challenges to overcome.
@@
 
-\paragraph{raising awareness of the energy consumption of the Softwares}
+\paragraph{raising awareness of the energy consumption of the Software}
 While this thesis's main focus was to optimize energy consumption using comparative studies, it was easier to say that approach \texttt{X} is greener than approach \texttt{Y}, no matter which metric we were using.
@@
 after a while, we removed these copies one by one.
-The source code of experiment can be found in the following GitHub repository\link{https://github.com/chakib-belgaid/lazy-green-code-test}
+The source code of the experiment can be found in the following GitHub repository\link{https://github.com/chakib-belgaid/lazy-green-code-test}
 Figure~\ref{fig:power_evolution_greenfaas} displays the machine's power consumption during this experiment.

### 2022-10-18 chapters/conclusion.tex
Proofreading

--- 2022-10-18 chapters/conclusion.tex
@@
 
-\paragraph{raising awareness of the energy consumption of the Softwares}
-While this thesis's main focus was to optimize energy consumption using comparative studies, it was easier to say that approach \texttt{X} is greener than approach \texttt{Y}, no matter which metric we were using.
-However, it will not be the case for developers when measuring the energy consumption of their programs.
-Some tried to give labels, such as (A,B,C\dots), while others translated these raw metrics in an equivalent of fuel consumption (e.g., 1 liter of fuel per 1000 lines of code).
-Most of the approaches were to use the carbon emission, such as~\cite{patterson2021carbon} or ecograder~\link{https://ecograder.com/} as a pivot metrics to quantify the environmental cost of osftware.
+\paragraph{Raising awareness about the energy consumption of software}
+As the core focus of this PhD thesis was to optimize energy consumption using comparative studies, it was straightforward to claim that approach \texttt{X} is greener than \texttt{Y}, no matter the metrics we were using.
+However, this cannot be the case for developers when measuring the energy consumption of their programs.
+Some tried to give labels, such as (A,B,C\dots), while others translated these raw metrics in an equivalent of fuel consumption (e.g., 1 liter of fuel per $1,000$ lines of code).
+Most of the approaches were to use the carbon emission, such as~\cite{patterson2021carbon} or ecograder,\link{https://ecograder.com/} as a pivot metrics to quantify the environmental cost of software.
 We intend to explore these approaches and study how they can be used to increase awareness of the energy consumption of their code.
 
-\subsection*{Representativeness}:
+\subsection*{Representativeness}
 As reported in the state of the art, a successful benchmark meets three criteria: reproducibility, accuracy, and representativeness.

### 2022-10-18 chapters/conclusion.tex
update : conclusion

--- 2022-10-18 chapters/conclusion.tex
@@
 Both studies revealed that each framework had different behaviors toward scalability.
-While the absence of a clear winner, we found out that most compiled languages are more energy-efficient because of their performance, Java tends to depict more power, and the interpreted languages lacked in terms of performance while keeping a low power consumption.
-PHP was an exception as its performance was compared to the compiled languages.
-The results of this work resulted in creating a dashboard to help developers choose the right programming language and framework for their application based on the context.
+While the absence of a clear winner, we found out that most compiled languages are more energy-efficient because of their performance.
+Java tends to depict more power, and the interpreted languages lacked performance while maintaining low power consumption.
+PHP was an exception, as its performance was compared to the compiled languages.
+This work resulted in creating a dashboard to help developers choose the correct programming language and framework for their application based on the context.
 The dashboard can be found at (\url{github.com/chakib-belgaid/greenboard}):
@@
 \begin{equation}
-      Ipact_{Benchmark} =E_{Benchmark} - E_{System_before}
+      Impact_{Benchmark} =E_{Benchmark} - E_{Before}
+\end{equation}
+Wich can be approximated by: 
+\begin{equation}
+      Impact_{Benchmark} =E_{Benchmark} - P_{Before} * T_{Benchmark}
 \end{equation}
 
-Wich can be approximated by: 
-\begin{equation}
-      Ipact_{Benchmark} =E_{Benchmark} - P_{System_before} * T_{Benchmark}
-\end{equation}
+Where $Impact_{Benchmark}$ is the energetical impact of the benchmark, $E_{Benchmark}$ is the energy of the system when the benchmark is running, $E_{Before}$ is the energy of the system before the benchmark is running, $P_{Before}$ is the power of the system before the benchmark is running, and $T_{Benchmark}$ is the time of the benchmark execution.
+
 One can notice that the energy impact of our benchmark did not have the same behavior as the system's energy consumption.
@@
 However, this action was motivated by the fact that most of the could Software are web services, as we have seen in chapter\ref{chapter:porgramming_langauges} where the execution time is not bounded, and each service can run for an unlimited amount of time.
-
-\section{Conclusion}
-This experiment shows that the context can have a significant impact on the energy consumption of the system, and sometimes measuring only the energy consumption of an application does not represent the real impact of this application on the energy consumption of the whole system, especially when we are dealing with timeless Softwares aka services.
-In the future, we tend to extend this experiment with more data, such as the energy consumption of the application itself instead of the whole system using \cite{fieni2020smartwatts} and comparing the strategies that were mentioned in this thesis using multiple contexts instead of just a minimal version of the operating system.
-
 \begin{figure}[!t]
@@
 \end{figure}
+\paragraph*{Synthesis}
+This experiment shows that the context can have a significant impact on the energy consumption of the system, and sometimes measuring only the energy consumption of an application does not represent the real impact of this application on the energy consumption of the whole system, especially when we are dealing with timeless Softwares aka services.
+In the future, we tend to extend this experiment with more data, such as the energy consumption of the application itself instead of the whole system using \cite{fieni2020smartwatts} and comparing the strategies that were mentioned in this thesis using multiple contexts instead of just a minimal version of the operating system.
 

### 2022-10-18 chapters/conclusion.tex

--- 2022-10-18 chapters/conclusion.tex
@@
 
-\paragraph{raising awareness of the energy consumption of the Software}
-While this thesis's main focus was to optimize energy consumption using comparative studies, it was easier to say that approach \texttt{X} is greener than approach \texttt{Y}, no matter which metric we were using.
-However, it will not be the case for developers when measuring the energy consumption of their programs.
-Some tried to give labels, such as (A,B,C\dots), while others translated these raw metrics in an equivalent of fuel consumption (e.g., 1 liter of fuel per 1000 lines of code).
-Most of the approaches were to use the carbon emission, such as~\cite{patterson2021carbon} or ecograder~\link{https://ecograder.com/} as a pivot metrics to quantify the environmental cost of osftware.
+\paragraph{Raising awareness about the energy consumption of software}
+As the core focus of this PhD thesis was to optimize energy consumption using comparative studies, it was straightforward to claim that approach \texttt{X} is greener than \texttt{Y}, no matter the metrics we were using.
+However, this cannot be the case for developers when measuring the energy consumption of their programs.
+Some tried to give labels, such as (A,B,C\dots), while others translated these raw metrics in an equivalent of fuel consumption (e.g., 1 liter of fuel per $1,000$ lines of code).
+Most of the approaches were to use the carbon emission, such as~\cite{patterson2021carbon} or ecograder,\link{https://ecograder.com/} as a pivot metrics to quantify the environmental cost of software.
 We intend to explore these approaches and study how they can be used to increase awareness of the energy consumption of their code.
 
-\subsection*{Representativeness}:
+\subsection*{Representativeness}
 As reported in the state of the art, a successful benchmark meets three criteria: reproducibility, accuracy, and representativeness.

### 2022-10-18 chapters/optimization/python-dir/python.tex

@@
 Figure~\ref{fig:python_multiprocessing} shows the results of the experiment regarding four metrics: power, execution time, DRAM energy, and CPU energy.
-As one can notice, CPU energy is ten folds higher than the DRAM one.
+As one can notice, CPU energy is ten folds higher than DRAM one.
 Therefore, we will focus on the CPU for this study.
@@
 This study demonstrates that the optimal way to reduce the energy consumption of Python applications is to adopt the \textsc{Multiprocessing} strategy, which is the most energy-efficient one.
-However, this strategy should be used carefully as it can lead to increased execution times.
+However, this strategy should be used carefully as it can lead to increase power consumption.
 This study demonstrates that the optimal number of processes is equal to the number of physical cores, which is four in our case.
@@
 Then, we conducted a series of experiments to compare the energy consumption of these alternatives.
-Our findings indicate the lack of a general solution and the importance of tuning the Python runtime depending on applications.
+Our findings indicate the lack of a general solution and the importance of tuning the Python runtime depending on the application.
 % In some circumstances, some of these answers may be the best, while others may be the worst.

### 2022-10-18 chapters/conclusion.tex
Proofrading

@@
 As reported in the state of the art, a successful benchmark meets three criteria: reproducibility, accuracy, and representativeness.
-This Ph.D. thesis mostly discussed two of these criteria: reproducibility, and accuracy.
+This PhD thesis mostly discussed two of these criteria: reproducibility, and accuracy.
 However, we did not discuss the representativeness of our benchmarks.
@@
 Figure~\ref{fig:JunitSonarplugin} shows the work of our intern who used a Sonar plugin to highlight the energy evolution of Java applications over time.
-This prototype got even more sophisticated to become Joule-dff\link{https://github.com/davidson-consulting/diff-jjoules}, which is a tool to be added in continuous integration to highlight the energy evolution of the Java programs.
+This prototype got even more sophisticated to become Joule-diff,\link{https://github.com/davidson-consulting/diff-jjoules} which is a tool to be included in a CI/CD workflow to highlight the energy evolution of the Java programs.
 We intend to push this even further to detect the commits responsible for an increase or the optimization within the source code.
+
 \begin{figure}[!h]
       \centering
-      \includegraphics[width=0.8\linewidth]{chapters/JunitSonarplugin}
+      \includegraphics[width=.8\linewidth]{chapters/JunitSonarplugin}
       \caption{Sonar Energy plugin for JUnit}
@@
 
-\subsubsection{Energy impact vs Energy consumption}
-To make the software more eco-friendly, both state-of-the-art and this thesis focused on reducing software energy consumption.
+\subsubsection{Energy impact Vs. Energy consumption}
+To develop more eco-friendly software, both state-of-the-art and this thesis focused on reducing software energy consumption.
 To accomplish this, we sought to isolate the energy consumption of the software under test as much as possible, primarily by running the tests on an isolated machine that runs only a few services.
-Others tried to infer the energy consumption of a single process using additional metrics that vary from the CPU usage, such as \cite{noureddine-issta-2016} to more sophisticated models like~\cite{fieni2020smartwatts,fieni2021selfwatts}.
+Others tried to infer the energy consumption of a single process using additional metrics that vary from the CPU usage, such as~\cite{noureddine-issta-2016} to more sophisticated models like~\cite{fieni2020smartwatts,fieni2021selfwatts}.
 
@@
 
-This is why I've started to work on a new project called \textsc{Energy Impact} that aims to measure the energy impact of software.
+This is why I have started to work on a new project called \textsc{Energy Impact} that aims to measure the energy impact of software.
 According to the work of \citeauthor{barroso2007case}~\cite{barroso2007case}, servers operate most of the time at between 10 and 50 percent of their maximum utilization levels.
@@
       \centering
-      \includegraphics[width=0.9\linewidth,keepaspectratio]{chapters/soa_energy_efficiency}
-      \caption{Power efficiency vs power consumption\cite{barroso2007case}}
+      \includegraphics[width=.7\linewidth,keepaspectratio]{chapters/soa_energy_efficiency}
+      \caption{Power efficiency Vs. power consumption~\cite{barroso2007case}}
       \label{fig:soa_energy_efficiency}
@@
 
+To confirm if this hypothesis still holds, we run the following experiment.
+First, we provision a machine with minimal services, then created a process that stresses one core by doing some mathematical operations, and each second we introduced a copy of this process.
+After a while, we removed these copies one by one.
+The source code of the experiment can be found in the following GitHub repository.\link{https://github.com/chakib-belgaid/lazy-green-code-test}
+Figure~\ref{fig:power_evolution_greenfaas} reports on the machine power consumption during this experiment.
+As one can notice, the introduction of the first copies had a greater cost than the other ones, despite not sharing the same core.
+And the more we approach the peak of power, the less significance the introduction of a new process has on power consumption to be almost zero after the $9^{th}$ copy.
+
 \begin{figure}[!h]
       \centering
+      \includegraphics[width=\linewidth]{chapters/power_evolution_greenfaas}
       \caption{Evolution of CPU power}
       \label{fig:power_evolution_greenfaas}
-      \includegraphics[width=\linewidth]{chapters/power_evolution_greenfaas}
 \end{figure}
 
-In order to confirm whether this hypothesis still holds. We run the following experiment\link.
-First, we took a machine with minimal services, then created a process that stresses one core by doing some mathematical operations, and each second we introduced a copy of this process.
-after a while, we removed these copies one by one.
-The source code of the experiment can be found in the following GitHub repository\link{https://github.com/chakib-belgaid/lazy-green-code-test}
-Figure~\ref{fig:power_evolution_greenfaas} displays the machine's power consumption during this experiment.
-As one can notice, the introduction of the first copies had a greater cost than the other ones, despite not sharing the same core.
-And the more we approach the peak of power, the less significance the introduction of a new process has on power consumption to be almost zero after the 9th copy.
+One should note that this experiment was conducted on a machine with the following configuration: Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz, with 4 physical cores and 4 hyper threads, 16GB RAM, and 687GB HDD.
+As for the operating system, we used Ubuntu 20.04.2 LTS.
 
-One should note that this experiment was conducted on a machine with the following configuration:
-Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz, with 4 physical cores and 4 hyper threads, 16GB RAM, and 687GB HDD. As for the operating system, we used Ubuntu 20.04.2 LTS.
+Therefore, with the introduction of the $8^{th}$ copy, we have already saturated the machine capacity, and the introduction of the $9^{th}$ copy had no impact on the power consumption.
 
-Therefore with the introduction of the eighth copy, we have already saturated the machine's capacity, and the introduction of the ninth copy had no impact on the power consumption.
+To push this experiment further, we created a stress tool to fix the number of cores and \% usage of each core to simulate a production context. 
+We then run a benchmark (graph coloration algorithm)\link{https://github.com/chakib-belgaid/lazy-green-code-test/tree/master/exp01} using multiple configurations.
+Figure~\ref{fig:green_faas_duration} displays the overall performance of the benchmark.
+As one can see, the execution time remained the same when the context was using 3 cores at any saturation level, because we have 4 physical CPU cores, therefore each process was running a different core (3 cores were used by the context and the other one was used by the benchmark).
+However, behavior changes when the context consumes more than 3 cores.
+CPU saturation tends to have a greater impact on the performance of the benchmark.
+When the benchmark was using a hyper-thread of the same core as one of the context workers, it took almost double the time to finish the same job.
+As when we saturated only 50\% of the CPU capacity, this increase did not happen until the context used 7 cores and, in the worst-case scenario (10 extra processes), it took only 75 seconds to finish the job.
+
 \begin{figure}[!h]
@@
 
-To push this experiment further, we created a stressing tool that allows us to fix the number of cores and \% usage of each core to simulate a production context.
-we then run a benchmark( graph coloration algrothim\link{https://github.com/chakib-belgaid/lazy-green-code-test/tree/master/exp01}) using multiple configurations.
-Figure~\ref{fig:green_faas_duration} displays the overall performance of the benchmark. As one can see, the execution time remained the same when the context was using 3 cores at any saturation level,
-this is because we have 4 physical CPU cores, therefore each process was running a different core,(3 cores were used by the context and the other one was used by the benchmark).
-However, behavior changes when the context consumes more than 3 cores.
-CPU saturation tends to have a great impact on the performance of the benchmark.
-when the benchmark was using a hyper thread of the same core as one of the context workers it took almost double the time to finish the same job.
-As when we saturated only 50\% of the CPU capacity, this increase didn't happen until the context used 7 cores, and in the worst-case scenario ( 10 extra processes) it took only 75 seconds to finish the job.
+On the other hand, as we can see in Figure~\ref{fig:green_faas_power}, the average power behavior of the system had different behavior.
+With the 100\% saturation scenario, the power consumption hit its peak as soon as we hit the 4 core usage 3 from the context and 1 from the benchmark), and it kept fluctuating until the $8^{th}$ core saturation, after that, it was in a constant decrease.
+This was because the processes were competing with each others for cache usage, which led to a significant number of cache misses that slowed down the CPU frequency and led to a decrease in power consumption.
 
@@
 
-On the other hand, as we can see in Figure~\ref{fig:green_faas_power}, the average power behavior of the system had different behavior.
-With the 100\% saturation scenario, the power consumption hit its peak as soon as we hit the 4 core usage 3 from the context and 1 from the benchmark), and it kept fluctuating until the 8th core saturation, after that it was in a constant decrease.
-This was because the processes were competing with each other for cache usage, which led to a significant number of cache defaults that slowed down the CPU frequency which led to a decrease in power consumption.
-
 However, when we used 50\% of the CPU capacity, the power kept increasing by adding the number of extra processes.
 As for the scenarios where we used more than 50\% of the CPU capacity, the power consumption kept increasing until it hit its peak then it start to decrease; this number of processes needed to reach the peak varies according to the saturation level.
+
+When we try to measure the energy consumption of this experiment, we find that the scenario where we used 100\% of the energy consumption was the worst case by far, and it kept increasing whenever we increase the number of core usage.
+Figure~\ref{fig:green_faas_energy} present the energy consumption of the system during the experiment.
+
 \begin{figure}[!h]
@@
 
-When we try to measure the energy consumption of this experiment, we find that the scenario where we used 100\% of the energy consumption was the worst case by faar, and it kept increasing whenever we increase the number of core usage.
-Figure~\ref{fig:green_faas_energy} present the energy consumption of the system during the experiment.
-
 However, if we calculate the energy impact of the benchmark which we define as: 
@@
 \end{equation}
-Wich can be approximated by: 
+and can be approximated by:
 \begin{equation}
@@
 \end{equation}
-
-Where $Impact_{Benchmark}$ is the energetical impact of the benchmark, $E_{Benchmark}$ is the energy of the system when the benchmark is running, $E_{Before}$ is the energy of the system before the benchmark is running, $P_{Before}$ is the power of the system before the benchmark is running, and $T_{Benchmark}$ is the time of the benchmark execution.
+where $Impact_{Benchmark}$ is the energy impact of the benchmark, $E_{Benchmark}$ is the energy of the system when the benchmark is running, $E_{Before}$ is the energy of the system before the benchmark is running, $P_{Before}$ is the power of the system before the benchmark is running, and $T_{Benchmark}$ is the time of the benchmark execution.
 
 One can notice that the energy impact of our benchmark did not have the same behavior as the system's energy consumption.
-Sometimes introducing the benchmark had a \emph{negative} impact on the system's energy consumption, which made it even greener.
-This is because it reduces power consumption.This is shown in Figure~\ref{fig:green_faas_impact}
+Sometimes, introducing the benchmark had a \emph{negative} impact on the system's energy consumption, which made it even greener.
+This is because it reduces power consumption.
+This is shown in Figure~\ref{fig:green_faas_impact}.
 
-We are aware that during this analysis, we have not considered the execution time of the context, and it will be affected by the benchmark - mainly, it will increase when we use more cores and/or higher CPU saturation -.
-However, this action was motivated by the fact that most of the could Software are web services, as we have seen in chapter\ref{chapter:porgramming_langauges} where the execution time is not bounded, and each service can run for an unlimited amount of time.
 \begin{figure}[!t]
@@
 \end{figure}
+
+We are aware that during this analysis, we have not considered the execution time of the context, and it will be affected by the benchmark---mainly, it will increase when we use more cores and/or higher CPU saturation.
+However, this action was motivated by the fact that most of the cloud software are web services, as we have seen in Chapter~\ref{chapter:porgramming_langauges} where the execution time is not bounded, and each service can run for an unlimited amount of time.
+
 \paragraph*{Synthesis}

### 2022-10-18 chapters/conclusion.tex
Proofreading

@@
 \clearpage
-\chapter{Discussion and Conclusion}
+\chapter{Conclusion \& Perspectives}
 \label{chapter:conclusion}
@@
 
-Finally, We examined the phenomenon of variation when measuring the energy consumption of experiments in order to increase their accuracy. To do so, we investigated the following research questions.:
+Finally, We examined the phenomenon of variation when measuring the energy consumption of experiments to increase their accuracy.
+To do so, we investigated the following research questions:
 \begin{description}
@@
 To do so, we investigated the energy behavior of the python during its most common usage:
-
 \begin{itemize}
@@
 We discovered that using an optimized strategy can reduce energy consumption by $20\times$.
-      \item \emph{Data Processing}: In this section, we studied the energy consumption of data structures, iteration algorithms, and concurrency.
+      \item \emph{data processing}: In this section, we studied the energy consumption of data structures, iteration algorithms, and concurrency.
 following the guidelines and prioritizing built-in functions is the best way to reduce the energy consumption of Python applications.
@@
 It can be used to highlight the functions that consume the most energy in a given Python program.
-Its main usage is to help developers do an exploratory analysis of their application to scope the functions that should be optimized to be then targeted by PyJoules;
+Its main usage is to help developers do an exploratory analysis of their application to scope the functions that should be optimized to be then targeted by PyJoules.
      \end{itemize}
@@
 
-In this chapter, we thoroughly explored how JVMs affect software energy consumption by investing the following research questions:
+In this chapter, we thoroughly explored how JVMs affect software energy consumption by investing in the following research questions:
 \begin{description}
@@
 
-This work resulted in the development of software called \textbf{JRefferal} (\url{github.com/chakib-belgaid/jreferral}) that allows the user to explore the JVM settings and their impact on the energy consumption of a given Java program and a publication of the following paper :
+This work resulted in the development of software called \textbf{JReferral} (\url{github.com/chakib-belgaid/jreferral}) that allows the user to explore the JVM settings and their impact on the energy consumption of a given Java program and a publication of the following paper:
 \\
@@
 While the absence of a clear winner, we found out that most compiled languages are more energy-efficient because of their performance.
-Java tends to depict more power, and the interpreted languages lacked performance while maintaining low power consumption.
+Java tends to depict more power, and the interpreted languages lacked performance but maintained low power consumption.
 PHP was an exception, as its performance was compared to the compiled languages. 
 This work resulted in creating a dashboard to help developers choose the correct programming language and framework for their application based on the context.
-The dashboard can be found at (\url{github.com/chakib-belgaid/greenboard}):
+The dashboard can be found at \url{github.com/chakib-belgaid/greenboard}.
 

### 2022-10-18 chapters/conclusion.tex
update

@@
 As we have seen in Chapter~\ref{chapter:java}, the choice of the JVM can greatly impact the energy consumption of hosted applications.
-On the other hand, we have seen in Chapter~\ref{chapter:porgramming_langauges} that other JVM-based languages depicted half power consumption of the Java code.
+On the other hand, we have seen in Chapter~\ref{chapter:porgramming_langauges} that other JVM-based languages depicted half the power consumption of the Java code.
 Was it only because of the adopted web framework?
@@
 \paragraph{Raising awareness about the energy consumption of software}
-As the core focus of this PhD thesis was to optimize energy consumption using comparative studies, it was straightforward to claim that approach \texttt{X} is greener than \texttt{Y}, no matter the metrics we were using.
+As the core focus of this Ph.D.
+thesis was to optimize energy consumption using comparative studies, it was straightforward to claim that approach \texttt{X} is greener than \texttt{Y}, no matter the metrics we were using.
 However, this cannot be the case for developers when measuring the energy consumption of their programs.
 Some tried to give labels, such as (A,B,C\dots), while others translated these raw metrics in an equivalent of fuel consumption (e.g., 1 liter of fuel per $1,000$ lines of code).
-Most of the approaches were to use the carbon emission, such as~\cite{patterson2021carbon} or ecograder,\link{https://ecograder.com/} as a pivot metrics to quantify the environmental cost of software.
+Most of the approaches were to use the carbon emission, such as~\cite{patterson2021carbon} or ecograder,\link{https://ecograder.com/} as a pivot metric to quantify the environmental cost of the software.
 We intend to explore these approaches and study how they can be used to increase awareness of the energy consumption of their code.
@@
 First, we will consider taking advantage of the popularity of CI/CD among developers to provide some insights into the energy consumption of their code.
-We did prototype a sonar tool that can report on the energy consumption of Java-based code repository.
+We did prototype a sonar tool that can report on the energy consumption of a Java-based code repository.
 Figure~\ref{fig:JunitSonarplugin} shows the work of our intern who used a Sonar plugin to highlight the energy evolution of Java applications over time.
@@
 With the 100\% saturation scenario, the power consumption hit its peak as soon as we hit the 4 core usage 3 from the context and 1 from the benchmark), and it kept fluctuating until the $8^{th}$ core saturation, after that, it was in a constant decrease.
-This was because the processes were competing with each others for cache usage, which led to a significant number of cache misses that slowed down the CPU frequency and led to a decrease in power consumption.
+This was because the processes were competing with each other for cache usage, which led to a significant number of cache misses that slowed down the CPU frequency and led to a decrease in power consumption.
 

### 2022-10-18 chapters/introduction.tex
proof read and fix some typos

@@
           In this section, we provide the literature approaches used to measure energy consumption and classify them, compare them and discuss their pros and cons when used for our research.
-After that, we tackle the challenge of accuracy within the field of energy consumption, and discuss how the literature tried to overcome these challenges.
+After that, we tackle the challenge of accuracy within the field of energy consumption and discuss how the literature tried to overcome these challenges.
 Finally, we  present some of the most recent works in the field of software energy consumption optimization;

### 2022-10-18 chapters/optimization/java-dir/java.tex
proof read and fix some typos

@@
 \section{Introduction}
-As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.In fact, in 2022, Java will be second only to Python, according to PYPY~\footnote{\url{https://pypl.github.io/PYPL.html}}.
-Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language.According to the TIOBE index, Java was the most frequently used language from 2002 until 2017, and it remained in the top 5 after that~\footnote{\url{https://www.tiobe.com/tiobe-index/}}.
+As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.
+In fact, in 2022, Java will be second only to Python, according to PYPY~\footnote{\url{https://pypl.github.io/PYPL.html}}.
+Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language.
+According to the TIOBE index, Java was the most frequently used language from 2002 until 2017, and it remained in the top 5 after that~\footnote{\url{https://www.tiobe.com/tiobe-index/}}.
 In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performance, Java applications can be at the same time one of the most energy-efficient or hungry solutions.
@@
 
-Java's portability is a core design goal, which means that Java applications will work exactly the same on any operating system and on any hardware.As we see in figure~\ref{fig:JVM_architecture}, Instead of machine code, this is accomplished by compiling Java language code to an intermediate representation known as Java bytecode.
+Java's portability is a core design goal, which means that Java applications will work the same on any operating system and any hardware.
+As we see in figure~\ref{fig:JVM_architecture}, Instead of machine code, this is accomplished by compiling Java language code to an intermediate representation known as Java bytecode.
 Java bytecode instructions are similar to machine code, but they are executed by a JVM rather than directly on the machine.
 As a result, Java programs will run slower and consume more memory than C++ programs.
-However, Just-in-Time (JIT) compilation, which is built into the JVM, improves performance by converting bytecode to machine code at runtime.
+However, the Just-in-Time (JIT) compilation, which is built into the JVM, improves performance by converting bytecode to machine code at runtime.
 Furthermore, an automatic garbage collector (GC) is used by Java to manage memory throughout the object lifecycle and to recover memory when objects are no longer in use.
@@
 \subsection{Metrics and Measurement}
-Since the goal of this study is the green aspect of JVM, our key metric will be the energy consumed by job completed for each JVM configuration.
+Since the goal of this study is the green aspect of JVM, our key metric will be the energy consumed by the job completed for each JVM configuration.
 In addition to the energy consumption, we collected additional metrics to explain the reasons behind the behavior of each experiment.
@@
 The package is available in the GitHub repository.\footnote{\url{https://github.com/chakib-belgaid/jvm-comparaison}}
-To add extra \textbf{jvm candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the jvm to be used.
+To add extra \textbf{jvm candidates} for the benchmark applications, we added a new configuration file \texttt{jvm.sh} in the root directory of the repository, where we put the name and the version of the JVM to be used.
 For the \textbf{input workload},the benchmarks should be provided in the \texttt{benchmarks} directory.
@@
 
-For faster experiments, we propose \textsc{Jreferral}\footnote{\url{https://github.com/chakib-belgaid/jreferral}}, an open source tool that automatically compares the JVM energy consumption and recommends the most energy-efficient configuration for a given Java application.
+For faster experiments, we propose \textsc{Jreferral}\footnote{\url{https://github.com/chakib-belgaid/jreferral}}, an open-source tool that automatically compares the JVM energy consumption and recommends the most energy-efficient configuration for a given Java application.
 We further discuss this tool in Section~\ref{section:SummaryofContributions}.
@@
 After about $150$ seconds, the power differences between the 3 JVMs is barely noticeable.
-One can, however, notice the effect of the JIT, as \textsc{HotSpot} takes the advantage over \textsc{GraalVM} and becomes more energy efficient.
+One can, however, notice the effect of the JIT, as \textsc{HotSpot} takes advantage over \textsc{GraalVM} and becomes more energy efficient.
 In total, \textsc{HotSpot} completes $597$ requests against $510$ for \textsc{GraalVM} and $381$ for \textsc{J9}, as shown in \Cref{table:power/request}.
@@
 \subsubsection{Multithreading}
-The purpose of this phase is to investigate the impact JVM thread management strategies on energy consumption.
+The purpose of this phase is to investigate the impact of JVM thread management strategies on energy consumption.
 This encompasses exploring if the management strategies of application-level parallelism (so-called \emph{threads}) result in different energy efficiencies, depending on JVM distributions.
@@
 Investigating such a hypothesis requires a selection of highly parallel and CPU-intensive benchmarks, which is one of the main criteria for our benchmark selection.
-As no tool can accurately monitor the energy consumption at a thread level, we monitor the global power consumption and CPU utilization during the execution using RAPL for the energy, and several Linux tools for the CPU-utilization (\texttt{htop}, \texttt{cpufreq}).
+As no tool can accurately monitor the energy consumption at a thread level, we monitor the global power consumption and CPU utilization during the execution using RAPL for the energy, and several Linux tools for the CPU usage (\texttt{htop}, \texttt{cpufreq}).
 %The experiments showed differences in energy consumption between the 3 JVMs and even between 2 version on the HotSpot VM.
@@
 
-\Cref{fig:threads} delivers a closer look to the thread allocation strategies adopted by JVM.
+\Cref{fig:threads} delivers a closer look at the thread allocation strategies adopted by JVM.
 First, \Cref{fig:avrora_threads} illustrates the active threads count evolution over time (excluding the JVM-related threads, usually 1 or 2 extra threads depending on the execution phase) for \textsf{Avrora}.
@@
 
-In conclusion, JVMs thread management can sometimes constitute a key factor that impacts software energy consumption.
-However, we suggest checking and comparing JVMs before deploying a software, especially if the target application is parallel and multi-threaded.
+In conclusion, JVM thread management can sometimes constitute a key factor that impacts software energy consumption.
+However, we suggest checking and comparing JVMs before deploying software, especially if the target application is parallel and multi-threaded.
 
@@
 We also considered both the community and economy configurations (no enterprise).
-A JIT+AOT (\emph{Ahead Of Time}) disabling experiment has also been considered for all of the 3 JVM platforms.
+A JIT+AOT (\emph{Ahead Of Time}) disabling experiment has also been considered for all the 3 JVM platforms.
 \Cref{table:JIT} reports on the energy consumption of the experiments we conducted for most of the benchmarks and JIT configurations under study.
@@
 The \textsf{warm} configuration delivers the closest results to the best results observed with the default configuration.
-Moreover, choosing a low minimum count of method calls seems to have a negative effect on the execution time and energy consumption.
+Moreover, choosing a low minimum count of method calls seems to have a negative effect on execution time and energy consumption.
 The only parameter that can give better performance than the default configuration in some cases is the number of parallel JIT threads---using 3 and 7 parallel threads---but is not statistically significant.
@@
 However, some considerations should be taken into account when choosing the GC, such as the garbage collection time, the throughput, etc.
-Other settings are less trivial to determine, such as tenure age, memory size, and GC threads count.
+Other settings are less trivial to determine, such as tenure age, memory size, and GC thread count.
 Experiments should thus be conducted on the software to tune the most convenient GC configuration to achieve better energy efficiency in production.
@@
 \textsf{Zip4J} runs a large file compression, while \textsf{K-nucleotide} extracts a DNA sequence, and updates a hashtable of k-nucleotide keys to count specific values.
-The short report presented in \Cref{table:j-referral} shows the ratio of potential energy saving between the most and least energy consuming tested JVM (40\% and 70\% energy savings for \textsf{Zip4J} and \textsf{K-nucleotide}, respectively).
+The short report presented in \Cref{table:j-referral} shows the ratio of potential energy saving between the most and least energy-consuming tested JVM (40\% and 70\% energy savings for \textsf{Zip4J} and \textsf{K-nucleotide}, respectively).
 Options are available for \textsf{J-Referral} to obtain much more detailed reports including execution time, DRAM usage, split DRAM vs. CPU consumption, etc.

### 2022-10-18 chapters/optimization/python-dir/python.tex
proof read and fix some typos

@@
 \begin{itemize}
-    \item \textsf{Sequential}: we compute the sum of the first $N$ numbers with no concurency;
+    \item \textsf{Sequential}: we compute the sum of the first $N$ numbers with no concurrency;
     \item \textsf{Multithreading}: we use the \texttt{ThreadPoolExecutor} library to compute the sum of the first $N$ numbers.
@@
 We divide $N$ by the number of processes, and each process sums the numbers in its range.
-In this situation, we considered 2, 4, 8, and 16 processes iteratively.
+In this situation, we considered the 2, 4, 8, and 16 processes iteratively.
 \end{itemize}

### 2022-10-18 chapters/programming_languages/rpc/rpc.tex
proof read and fix some typos

@@
 For more accuracy, our SUT (\emph{System Under Test}) runs a minimal version of Debian\,9 (4.9.0 kernel version), which enforces the core processes required for our experiment.
-Furthermore, we used Docker containers technology for reproducibility of the experiments and the isolation of the servers.
+Furthermore, we used Docker containers technology for the reproducibility of the experiments and the isolation of the servers.
 \paragraph{Client \& server environments.}
@@
 \paragraph{Performances.}
-For better accuracy and more details, We updated the open source RPC benchmarking tool GHZ (\url{https://ghz.sh/}).
+For better accuracy and more details, We updated the open-source RPC benchmarking tool GHZ (\url{https://ghz.sh/}).
 The modified version allows us to monitor the average power for each request from both the server and the client sides.
@@
     \item \textsf{Satisfaction ratio}: how many requests have been satisfied among the total requests,
-    \item \textsf{Request Per Seconds}: The number of the requests that have been answered from the server,
+    \item \textsf{Request Per Seconds}: The number of requests that have been answered from the server,
     \item \textsf{Tail Latency at 99\%}: one of the best metrics to evaluate the performances of a server.
@@
 
-Figure~\ref{fig:rps_clients} presents the number of requests per second for each implementation. The heatmap is colored in a logarithmic scale to visualize the results better.
+Figure~\ref{fig:rps_clients} presents the number of requests per second for each implementation. The heatmap is colored with a logarithmic scale to visualize the results better.
 As one can observe, most of the servers hit their RPS limit after five clients and 100 clients for VM-based servers, and after this, the number keeps constant, decreasing the average RPS per client.
@@
     \item \textsf{Broken state} mode when the size of requests exceeds 1MB, most of the servers cannot handle them, and they tend to crash.
-After 10MB, except Elixir, no server could handle those requests.
+After 10MB, except for Elixir, no server could handle those requests.
 \end{enumerate}
@@
 With a small number of clients, there was no typical behavior between different categories.
-On the other hand, when dealing with a more significant number of clients, interpreted languages and compiled ones dropped in terms of performance while keeping the same average power, unlike the VM-based ones that increased their average power while keeping the same performances.
+On the other hand, when dealing with a more significant number of clients, interpreted languages and compiled ones dropped in terms of performance while keeping the same average power, unlike the VM-based ones that increased their average power while keeping the same performance.
 Overall, the second strategy had a better impact on the single request cost than the first one.

### 2022-10-18 chapters/benchmarking/benchmarking.tex

@@
 First, we go through the three components of a successful benchmark when performing energy-related experiments.
-Section~\ref{sec:benchmarking_reproducibility} focuses on the "reproducibility" challenge---to deliver reproducible experiments without interfering with the energy measurement---while Section~\ref{sec:taming-the-energy-variation} discusses accuracy of software energy consumption.
+Section~\ref{sec:benchmarking_reproducibility} focuses on the "reproducibility" challenge---to deliver reproducible experiments without interfering with the energy measurement---while Section~\ref{sec:taming-the-energy-variation} discusses the \emph{accuracy} of software energy consumption.
 

### 2022-10-18 chapters/benchmarking/draft.tex
update : abstract ( en+fr)

@@
 Empirical measurements are critical to capture the effect of developers' choices on software energy consumption.
-To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state of the art~\cite{van_der_kouwe_benchmarking_2018}.
+To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state-of-the-art~\cite {van_der_kouwe_benchmarking_2018}.
 
-Second, one should not ignore tremendous progress in computer science, which has led to a rise in the number of obsolete results.
+Second, one should not ignore the tremendous progress in computer science, which has led to a rise in the number of obsolete results.
 Furthermore, when it comes to comparative research, the execution environment may impact the study itself.
@@
 \subsection{Docker Vs. Virtual Machine}
-Despite the fact that Type 1 is more performant than Type 2, the latter is the most used in research, as most researchers tend to conduct their experiments on their own machines.
+Even though Type 1 is more performant than Type 2, the latter is the most used in research, as most researchers tend to conduct their experiments on their machines.
 Docker, on the other hand, is the most well-known container technology.
@@
 
-\subsection{Extension}
-% TODO : proofread
 \subsubsection{Definition of Extention}
@@
 Implementing the observer interface allows the user to extend the experiment with new metrics.
-    \item \emph{The Candidate interface}: It provides a set of functions that the candidate should implement to called  by the orchestrator.
+    \item \emph{The Candidate interface}: It provides a set of functions that the candidate should implement to be called by the orchestrator.
 This interface is responsible for extending the experiment with new solutions.

### 2022-10-18 chapters/benchmarking/experiment.tex

@@
 
-Every experiment is conducted on $100$ iterations, on multiple nodes and using the 3 NPB benchmarks we mentioned, with a warmup phase of 10 iterations for each experiment.
+Every experiment is conducted on $100$ iterations, on multiple nodes, and using the 3 NPB benchmarks we mentioned, with a warmup phase of 10 iterations for each experiment.
 In most cases, we sought to evaluate the \emph{Standard Deviation} (STD), which is the most representative factor of the energy variation.
@@
 \subsection{Docker \& Accuracy}
-As the state of the art assesses the impact of Docker on energy consumption, one can also consider its impact on accuracy.
+As the state-of-the-art assesses the impact of Docker on energy consumption, one can also consider its impact on accuracy.
 In other words:\\
@@
 
-To answer this question, we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on three nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
-Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running the same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
+To answer this question, we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG}, and \textsf{EP} in a Docker container and a flat binary format on three nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
+Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running the same compiled version of the benchmarks at 5\,\%, 50\,\%, and 100\,\% workloads.
 While Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
@@
 In this case, only one CPU core is used among $2\times16$ physical cores.
-The other cores are switched to a low consumption state when C-states are on.
+The other cores are switched to a low-consumption state when C-states are on.
 The switching operation causes a vital energy consumption difference between the cores and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
@@
 Even though the first and third strategies are very similar (both use hyper-threads, but only on one CPU for the first and two CPUs for the third), the variation gap between them is significant, as the variation in the first strategy is 30 times lower ($STD_{S1}=116 mJ$,$STD_{S3}=3,452 mJ$).
-This shows that the hyper-threads technology is not the main reason behind the variation; the first strategy has even less variation than the second one and still uses the hyper-threading.
+This shows that the hyper-threads technology is not the main reason behind the variation; the first strategy has even less variation than the second one and still uses hyper-threading.
 
@@
 
-The first way to do this is to evaluate the OS idle activity consumption and to compare it to a low workload running job.
-Therefore, we ran 100 iterations of a single process benchmark \textsf{EP}, \textsf{LU} and \textsf{CG} on multiple nodes from the cluster \textsf{Dahu}, and compared the energy behavior of the node with its idle state on the same duration.
+The first way to do this is to evaluate the OS idle activity consumption and compare it to a low-workload running job.
+Therefore, we ran 100 iterations of a single process benchmark \textsf{EP}, \textsf{LU}, and \textsf{CG} on multiple nodes from the cluster \textsf{Dahu}, and compared the energy behavior of the node with its idle state on the same duration.
 The aggregated results, illustrated in Figure~\ref{fig:os-idle}, depict that the idle energy variation is up to 140\,\% worst than when running a job, even if it consumes 120\,\% less energy ($Mean_{Job}=8,746 mJ$, $Mean_{Idle}=3,927 mJ$).
@@
 We argue that this variation is not substantial, as it is not stable from one node to another.
-Moreover, 10\,\% of the variation is not a representative difference due to many factors that can affect it as the CPU temperature or the measurement errors.
+Moreover, 10\,\% of the variation is not a representative difference due to many factors that can affect it as CPU temperature or measurement errors.
 
@@
 Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and their variation reduction at low workloads.
-However, it also lists some new factors we identified along the analysis we conducted in Section, such as the results related to the OS or the reboot mode.
+However, it also lists some new factors we identified along the analysis we conducted in the previous Section, such as the results related to the OS or the reboot mode.
 Some guidelines are more useful/efficient for specific workloads, as shown in our experiments.
@@
 
-To validate the accuracy of our guidelines among a varied set of benchmarks on the one hand and their effect on the variation between identical machines, on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
+To validate the accuracy of our guidelines among a varied set of benchmarks on the one hand and their effect on the variation between identical machines, on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
 Half of these experiments have been performed at a 50\,\% workload, and the other half on single process jobs.
@@
 The RAPL tool mainly measures CPU and DRAM energy consumption.
-However, even running CPU/RAM intensive benchmarks would keep a degree on uncertainty concerning the hard disk and networking energy consumption.
+However, even running CPU/RAM intensive benchmarks would keep a degree of uncertainty concerning the hard disk and networking energy consumption.
 In addition, the operating system adds a layer of confusion and uncertainty.

### 2022-10-18 chapters/literature/draft_citations.tex

@@
 \citeauthor{bukh1992art} wrote a book about computer performance analysis, where they discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi-squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems.
-The article~\cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistakes in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often seen errors.
+The article~\cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistakes in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often-seen errors.
 
@@
 
-Other monitor tools have been realized by the manufacturers, such as IBM Power executive~\cite{koomey2011growth}, which allows their customers to monitor the power consumption and thermal behavior of the of BladeCenter systems in the data center.
+Other monitor tools have been realized by the manufacturers, such as IBM Power executive~\cite{koomey2011growth}, which allows their customers to monitor the power consumption and thermal behavior of the BladeCenter systems in the data center.
 
@@
 Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
-We can observe a considerable variation of the energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.
+We can observe a considerable variation of energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.
 
@@
 This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}.
-However, it has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
+However, it has also been the subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
 On the one hand, the correlation between processor temperature and energy consumption was one of the most explored paths.
@@
 
-As for the impact of the source code on the energy consumption, we can cite~\cite{pinto_comprehensive_2016,fernandes_assisting_2017} where the authors investigated the impact of Java collections on energy usage based on collection size and its usage (insertion, removal, search) and They provide some insights into the energy efficiency of specific collections under various scenarios.
+As for the impact of the source code on energy consumption, we can cite~\cite{pinto_comprehensive_2016,fernandes_assisting_2017} where the authors investigated the impact of Java collections on energy usage based on collection size and its usage (insertion, removal, search) and They provide some insights into the energy efficiency of specific collections under various scenarios.
 
@@
 We begin with Aequitas~\cite{ribic2016aequitas}, a system that allows parallel applications to live on co-managed power domains (sharing the same CPU).
-The technology is founded on the premise that coexisting programs can regard power-management hardware as a shared resource and collaborate on power management decisions.
+The technology is founded on the premise that coexisting programs can regard power-management hardware as a shared resource and collaborate on power-management decisions.
 As a result, it accomplishes its purpose by scheduling these applications with a time-slicing technique (for example, round robin).
@@
 
-%  TODO : rephrase this to fit the order of the chapters 
 This work aims to present a set of guidelines to create a benchmarking system to measure different programs' energy consumption.
-After that, we will use this system to compare the energy consumption of different programming languages.
-We will extend the work of \citeauthor{pereira_energy_2017} to a closer distance to the production environment by comparing a set of use cases.
+Following that, we will investigate the energy consumption of two well-known programming languages: Java and Python.
+Finally, we extend this analysis to cover various programming languages
+In this study, We will extend the work of \citeauthor{pereira_energy_2017} to a closer distance to the production environment.
 

### 2022-10-18 chapters/optimization/python-dir/python_interperters.tex

@@
 Despite the availability of numerous tools for code refactoring and automatic code optimization, the optimization process remains a manual effort.
-This is because code optimization is a task that requires extensive understanding of both the codebase and the language itself.
+This is because code optimization is a task that requires an extensive understanding of both the codebase and the language itself.
 
-To avoid such hustle, we tried to find a non-intrusive approach to make the Python code more eco-friendly without altering its structure.
+To avoid a such hustle, we tried to find a non-intrusive approach to make the Python code more eco-friendly without altering its structure.
 Python is an interpreted language, which led to many initiatives to implement their interpreter to improve one or many aspects of the Python code. 
-Therefore, in this section, we will examine at how different Python interpreters affect the energy consumption of software, and in which ca
-se, one should use a non-conventional interpreter to save the energy consumption of their application.
+Therefore, in this section, we will examine how different Python interpreters affect the energy consumption of software, and in which case, one should use a non-conventional interpreter to save the energy consumption of their application.
 
-\subsection{Python runtimes}\label{python:sec_interpreters}
+\subsection{Python Runtimes}\label{python:sec_interpreters}
 To do so, we gathered a list of interpreters, transpilers, and other optimization libraries that can contribute to reducing the energy consumption of legacy Python applications:
@@
       \item \textbf{Pyjion}:\footnote{\url{https://github.com/microsoft/pyjion}}
-            A JIT API for CPython, same purpose as Parakeet and Hope;
+            A JIT API for CPython, the same purpose as Parakeet and Hope;
       \item \textbf{Pyston}:\footnote{\url{https://blog.pyston.org}}
@@
       \item \textbf{Grumpy}:\footnote{\url{https://github.com/google/grumpy}}
-            A source-to-source transpiler that translates the Python code into Go before being compiled to a binary executable.
+            A source-to-source transpiler that translates the Python code into Go before being compiled into a binary executable.
             It also offers an interpreter, called \emph{grumprun}, which can directly execute the Python code.
@@
 
-We did not consider other implementations because either the project aborted many years ago or it has insufficient Python features.
+We did not consider other implementations because either the project was aborted many years ago or it has insufficient Python features.
 After the inventory of those implementations, we filtered them.
@@
 First and foremost, we want to know whether our samples have a normal distribution. As a result, we ran a Shapiro-Wilk~\cite{shapiro1968comparative} test on our data.
-Even though some implementations had a $p$-value higher than $a=0.05$ (such as CPython and ActivePython), other implementations like PyPy and Numba presented a $p$-values smaller than 0.01, which leads to rejecting the hypothesis $H0$ that all the distributions follow a normal distribution.
+Even though some implementations had a $p$-value higher than $a=0.05$ (such as CPython and ActivePython), other implementations like PyPy and Numba presented $p$-values smaller than 0.01, which leads to rejecting the hypothesis $H0$ that all the distributions follow a normal distribution.
 Therefore, to compare the different implementations, we performed the non-parametric Mann-Whitney U test~\cite{zimmerman1987comparative}, with CPython2 and CPython3 as base references.
@@
 The second cluster groups the interpreter implementations that are based on a JIT compiler (PyPy2 and PyPy3).
-Closer to them, we find Shedskin, which is a translator that converts the Python code into C++ code to be later compiled into binary code.
+Closer to them, we find Shedskin, which is a translator that converts Python code into C++ code to be later compiled into binary code.
 Table~\ref{table:manwithyu} shows that this cluster is by far the most energy-efficient one.
@@
 
-\subsection{Conclusion}
+\subsection{Summary}
 One can conclude that the Python interpreter used considerably impacts the programs' energy consumption.

### 2022-10-18 chapters/programming_languages/rpc/tail_latency_clients.tex

\begin{tabular}{lllrrrrrrrrrrrr}
{} & {} & {payload} & {1005} & {905} & {805} & {705} & {605} & {505} & {405} & {305} & {205} & {105} & {5} & {1} \\
{languageGroup} & {languageType} & {name} & {} & {} & {} & {} & {} & {} & {} & {} & {} & {} & {} & {} \\
\multirow[c]{5}{*}{1} & \multirow[c]{5}{*}{compiled} & cpp grpc st & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 15.0%, transparent 15.0%) 1.752 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 13.4%, transparent 13.4%) 1.724 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 11.2%, transparent 11.2%) 1.900 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 10.8%, transparent 10.8%) 1.842 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 11.3%, transparent 11.3%) 1.898 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 10.8%, transparent 10.8%) 1.843 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 12.6%, transparent 12.6%) 2.038 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 10.6%, transparent 10.6%) 1.811 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 11.0%, transparent 11.0%) 1.844 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 9.0%, transparent 9.0%) 1.881 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 1.804 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.0%, transparent 0.0%) 5.359 \\
 &  & crystal grpc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 27.5%, transparent 27.5%) 2.998 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 28.4%, transparent 28.4%) 3.261 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 23.9%, transparent 23.9%) 3.233 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 26.2%, transparent 26.2%) 3.415 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 24.4%, transparent 24.4%) 3.244 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 26.7%, transparent 26.7%) 3.457 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 31.8%, transparent 31.8%) 4.009 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 42.8%, transparent 42.8%) 5.055 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 67.0%, transparent 67.0%) 7.433 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 13.167 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 202.268 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 754.372 \\
 &  & go grpc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 5.0%, transparent 5.0%) 0.758 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.1%, transparent 4.1%) 0.766 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.761 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.763 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.768 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.768 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.770 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.772 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.775 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.808 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 2.056 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 7.307 \\
 &  & rust tonic st & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 20.9%, transparent 20.9%) 2.337 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 19.2%, transparent 19.2%) 2.313 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 16.0%, transparent 16.0%) 2.407 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 14.4%, transparent 14.4%) 2.208 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 13.1%, transparent 13.1%) 2.087 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 12.9%, transparent 12.9%) 2.051 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 11.6%, transparent 11.6%) 1.939 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 12.5%, transparent 12.5%) 2.003 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 12.4%, transparent 12.4%) 1.985 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 9.1%, transparent 9.1%) 1.896 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 2.037 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 5.788 \\
 &  & swift grpc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 38.4%, transparent 38.4%) 4.086 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 41.3%, transparent 41.3%) 4.590 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 33.3%, transparent 33.3%) 4.208 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 44.2%, transparent 44.2%) 5.257 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 47.3%, transparent 47.3%) 5.605 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 46.1%, transparent 46.1%) 5.421 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 45.7%, transparent 45.7%) 5.439 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 48.4%, transparent 48.4%) 5.617 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 48.7%, transparent 48.7%) 5.610 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 38.3%, transparent 38.3%) 5.515 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.9%, transparent 1.9%) 5.445 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.4%, transparent 0.4%) 8.337 \\
\multirow[c]{2}{*}{2} & \multirow[c]{2}{*}{vm .net} & csharp grpc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 7.2%, transparent 7.2%) 0.981 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 6.2%, transparent 6.2%) 0.978 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 2.1%, transparent 2.1%) 0.953 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 2.1%, transparent 2.1%) 0.948 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.9%, transparent 1.9%) 0.937 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.8%, transparent 1.8%) 0.927 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.8%, transparent 1.8%) 0.935 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 2.0%, transparent 2.0%) 0.938 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 2.2%, transparent 2.2%) 0.959 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 2.0%, transparent 2.0%) 1.019 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.6%, transparent 0.6%) 2.962 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.7%, transparent 0.7%) 10.767 \\
 &  & dotnet grpc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 5.2%, transparent 5.2%) 0.775 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.3%, transparent 4.3%) 0.786 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.773 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.768 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.766 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.765 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 0.759 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.759 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 0.756 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 0.774 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.4%, transparent 0.4%) 2.594 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.4%, transparent 0.4%) 8.450 \\
\multirow[c]{9}{*}{3} & \multirow[c]{9}{*}{vm java} & java aot & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 43.5%, transparent 43.5%) 4.595 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 41.8%, transparent 41.8%) 4.638 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 40.7%, transparent 40.7%) 4.978 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 40.0%, transparent 40.0%) 4.825 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 38.7%, transparent 38.7%) 4.720 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 36.3%, transparent 36.3%) 4.431 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 33.9%, transparent 33.9%) 4.228 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 34.1%, transparent 34.1%) 4.175 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 28.3%, transparent 28.3%) 3.564 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 15.3%, transparent 15.3%) 2.661 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.9%, transparent 0.9%) 3.458 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.7%, transparent 0.7%) 10.831 \\
 &  & java grpc g1gc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 5.1%, transparent 5.1%) 0.763 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.1%, transparent 4.1%) 0.767 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.765 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.769 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.771 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.774 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.768 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.763 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.766 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 0.783 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.0%, transparent 0.0%) 1.760 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.0%, transparent 0.0%) 5.420 \\
 &  & java grpc pgc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.8%, transparent 4.8%) 0.742 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.0%, transparent 4.0%) 0.752 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.755 & \colorgreen \width10em \height80% 0.737 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 0.751 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 0.750 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.0%, transparent 0.0%) 0.750 & \colorgreen \width10em \height80% 0.740 & \colorgreen \width10em \height80% 0.741 & \colorgreen \width10em \height80% 0.767 & \colorgreen \width10em \height80% 1.699 & \colorgreen \width10em \height80% 5.268 \\
 &  & java grpc sgc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.7%, transparent 4.7%) 0.726 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 3.9%, transparent 3.9%) 0.748 & \colorgreen \width10em \height80% 0.737 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 0.747 & \colorgreen \width10em \height80% 0.741 & \colorgreen \width10em \height80% 0.742 & \colorgreen \width10em \height80% 0.747 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 0.754 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.774 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.4%, transparent 0.4%) 0.815 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.0%, transparent 0.0%) 1.782 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.0%, transparent 0.0%) 5.363 \\
 &  & java grpc she & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 5.3%, transparent 5.3%) 0.788 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.4%, transparent 4.4%) 0.797 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.5%, transparent 0.5%) 0.791 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.5%, transparent 0.5%) 0.785 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.5%, transparent 0.5%) 0.796 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.5%, transparent 0.5%) 0.791 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 0.782 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.4%, transparent 0.4%) 0.776 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.4%, transparent 0.4%) 0.783 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.793 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.0%, transparent 0.0%) 1.779 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.1%, transparent 0.1%) 5.944 \\
 &  & java grpc zgc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 5.3%, transparent 5.3%) 0.791 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.5%, transparent 4.5%) 0.804 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.6%, transparent 0.6%) 0.797 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.6%, transparent 0.6%) 0.795 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.5%, transparent 0.5%) 0.793 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.5%, transparent 0.5%) 0.793 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.4%, transparent 0.4%) 0.787 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.4%, transparent 0.4%) 0.776 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.6%, transparent 0.6%) 0.796 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 0.796 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.0%, transparent 0.0%) 1.733 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.0%, transparent 0.0%) 5.562 \\
 &  & java micronaut & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 5.6%, transparent 5.6%) 0.815 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.8%, transparent 4.8%) 0.833 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.9%, transparent 0.9%) 0.835 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.9%, transparent 0.9%) 0.824 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.1%, transparent 1.1%) 0.857 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.8%, transparent 0.8%) 0.827 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.8%, transparent 0.8%) 0.834 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.8%, transparent 0.8%) 0.821 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.9%, transparent 0.9%) 0.835 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.8%, transparent 0.8%) 0.863 & nan & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.2%, transparent 0.2%) 7.134 \\
 &  & kotlin grpc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 9.5%, transparent 9.5%) 1.205 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 8.2%, transparent 8.2%) 1.184 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.1%, transparent 4.1%) 1.164 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.0%, transparent 4.0%) 1.151 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 3.9%, transparent 3.9%) 1.143 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 3.6%, transparent 3.6%) 1.107 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 3.7%, transparent 3.7%) 1.128 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 3.7%, transparent 3.7%) 1.115 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 3.8%, transparent 3.8%) 1.122 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 3.1%, transparent 3.1%) 1.155 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.5%, transparent 0.5%) 2.669 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.3%, transparent 0.3%) 7.594 \\
 &  & scala akka & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 24.2%, transparent 24.2%) 2.670 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 21.2%, transparent 21.2%) 2.523 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 24.5%, transparent 24.5%) 3.294 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 27.8%, transparent 27.8%) 3.582 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 28.3%, transparent 28.3%) 3.654 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 27.8%, transparent 27.8%) 3.568 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 25.5%, transparent 25.5%) 3.360 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 28.0%, transparent 28.0%) 3.562 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 27.3%, transparent 27.3%) 3.465 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 19.3%, transparent 19.3%) 3.159 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 2.1%, transparent 2.1%) 5.976 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 2.0%, transparent 2.0%) 20.122 \\
4 & vm beam & elixir grpc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 15.8%, transparent 15.8%) 1.837 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 15.0%, transparent 15.0%) 1.889 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 11.0%, transparent 11.0%) 1.879 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 10.8%, transparent 10.8%) 1.836 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 10.6%, transparent 10.6%) 1.833 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 10.4%, transparent 10.4%) 1.800 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 9.8%, transparent 9.8%) 1.756 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 10.0%, transparent 10.0%) 1.751 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 9.6%, transparent 9.6%) 1.704 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 7.2%, transparent 7.2%) 1.662 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.3%, transparent 1.3%) 4.385 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.8%, transparent 0.8%) 11.448 \\
\multirow[c]{6}{*}{5} & \multirow[c]{6}{*}{interpreted} & dart grpc & \colorgreen \width10em \height80% 0.259 & \colorgreen \width10em \height80% 0.343 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.0%, transparent 0.0%) 0.739 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 8.2%, transparent 8.2%) 1.576 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 49.7%, transparent 49.7%) 5.846 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 59.6%, transparent 59.6%) 6.796 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 59.3%, transparent 59.3%) 6.837 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 60.5%, transparent 60.5%) 6.837 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 59.1%, transparent 59.1%) 6.645 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 46.4%, transparent 46.4%) 6.520 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 2.5%, transparent 2.5%) 6.649 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.7%, transparent 0.7%) 10.768 \\
 &  & node grpc st & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 51.5%, transparent 51.5%) 5.394 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 49.5%, transparent 49.5%) 5.427 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 36.2%, transparent 36.2%) 4.517 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 46.8%, transparent 46.8%) 5.523 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 46.2%, transparent 46.2%) 5.491 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 47.5%, transparent 47.5%) 5.564 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 44.7%, transparent 44.7%) 5.330 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 44.3%, transparent 44.3%) 5.200 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 43.3%, transparent 43.3%) 5.067 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 34.1%, transparent 34.1%) 5.000 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.9%, transparent 1.9%) 5.567 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.5%, transparent 0.5%) 8.719 \\
 &  & node grpcjs st & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 52.9%, transparent 52.9%) 5.533 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 61.0%, transparent 61.0%) 6.611 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 58.5%, transparent 58.5%) 6.842 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 56.9%, transparent 56.9%) 6.553 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 73.4%, transparent 73.4%) 8.279 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 75.2%, transparent 75.2%) 8.379 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 71.7%, transparent 71.7%) 8.104 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 72.6%, transparent 72.6%) 8.053 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 72.6%, transparent 72.6%) 7.992 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 54.9%, transparent 54.9%) 7.578 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 2.6%, transparent 2.6%) 6.900 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 0.7%, transparent 0.7%) 10.644 \\
 &  & php grpc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 31.8%, transparent 31.8%) 3.431 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 30.7%, transparent 30.7%) 3.503 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 26.2%, transparent 26.2%) 3.467 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 26.4%, transparent 26.4%) 3.437 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 26.2%, transparent 26.2%) 3.436 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 26.3%, transparent 26.3%) 3.413 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 26.1%, transparent 26.1%) 3.425 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 26.5%, transparent 26.5%) 3.414 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 26.7%, transparent 26.7%) 3.412 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 21.4%, transparent 21.4%) 3.422 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.5%, transparent 1.5%) 4.721 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.8%, transparent 1.8%) 18.743 \\
 &  & python grpc & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 81.9%, transparent 81.9%) 8.416 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 76.7%, transparent 76.7%) 8.226 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 11.166 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 10.959 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 11.016 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 10.894 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 11.010 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 10.812 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 98.8%, transparent 98.8%) 10.608 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 76.3%, transparent 76.3%) 10.226 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.3%, transparent 4.3%) 10.263 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.3%, transparent 1.3%) 14.892 \\
 &  & ruby grpc & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 10.224 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 10.623 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 73.9%, transparent 73.9%) 8.441 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 97.5%, transparent 97.5%) 10.701 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 95.9%, transparent 95.9%) 10.599 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 97.0%, transparent 97.0%) 10.592 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 96.1%, transparent 96.1%) 10.612 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 99.0%, transparent 99.0%) 10.711 & \colorred \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 100.0%, transparent 100.0%) 10.731 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 80.8%, transparent 80.8%) 10.782 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 4.0%, transparent 4.0%) 9.623 & \width10em \height80% \backgroundlinear-gradient(90deg, #aed6f1 1.2%, transparent 1.2%) 14.042 \\
\end{tabular}

### 2022-10-19 chapters/conclusion.tex
remove : green faas part

@@
 This is why I have started to work on a new project called \textsc{Energy Impact} that aims to measure the energy impact of software.
-According to the work of \citeauthor{barroso2007case}~\cite{barroso2007case}, servers operate most of the time at between 10 and 50 percent of their maximum utilization levels.
-Moreover, as shown in Figure~\ref{fig:soa_energy_efficiency} with 50\% of the servers' capacity, the server operates only with 75\% of its maximum power.
-In other words, it will cost us only 25\% of the energy to operate the extra 50\% of the servers' capacity.
-
-\begin{figure}[!h]
-      \centering
-      \includegraphics[width=.7\linewidth,keepaspectratio]{chapters/soa_energy_efficiency}
-      \caption{Power efficiency Vs. power consumption~\cite{barroso2007case}}
-      \label{fig:soa_energy_efficiency}
-\end{figure}
-
-To confirm if this hypothesis still holds, we run the following experiment.
-First, we provision a machine with minimal services, then created a process that stresses one core by doing some mathematical operations, and each second we introduced a copy of this process.
-After a while, we removed these copies one by one.
-The source code of the experiment can be found in the following GitHub repository.\link{https://github.com/chakib-belgaid/lazy-green-code-test}
-Figure~\ref{fig:power_evolution_greenfaas} reports on the machine power consumption during this experiment.
-As one can notice, the introduction of the first copies had a greater cost than the other ones, despite not sharing the same core.
-And the more we approach the peak of power, the less significance the introduction of a new process has on power consumption to be almost zero after the $9^{th}$ copy.
-
-\begin{figure}[!h]
-      \centering
-      \includegraphics[width=\linewidth]{chapters/power_evolution_greenfaas}
-      \caption{Evolution of CPU power}
-      \label{fig:power_evolution_greenfaas}
-\end{figure}
-
-One should note that this experiment was conducted on a machine with the following configuration: Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz, with 4 physical cores and 4 hyper threads, 16GB RAM, and 687GB HDD.
-As for the operating system, we used Ubuntu 20.04.2 LTS.
-
-Therefore, with the introduction of the $8^{th}$ copy, we have already saturated the machine capacity, and the introduction of the $9^{th}$ copy had no impact on the power consumption.
-
-To push this experiment further, we created a stress tool to fix the number of cores and \% usage of each core to simulate a production context. 
-We then run a benchmark (graph coloration algorithm)\link{https://github.com/chakib-belgaid/lazy-green-code-test/tree/master/exp01} using multiple configurations.
-Figure~\ref{fig:green_faas_duration} displays the overall performance of the benchmark.
-As one can see, the execution time remained the same when the context was using 3 cores at any saturation level, because we have 4 physical CPU cores, therefore each process was running a different core (3 cores were used by the context and the other one was used by the benchmark).
-However, behavior changes when the context consumes more than 3 cores.
-CPU saturation tends to have a greater impact on the performance of the benchmark.
-When the benchmark was using a hyper-thread of the same core as one of the context workers, it took almost double the time to finish the same job.
-As when we saturated only 50\% of the CPU capacity, this increase did not happen until the context used 7 cores and, in the worst-case scenario (10 extra processes), it took only 75 seconds to finish the job.
-
-\begin{figure}[!h]
-      \centering
-      \includegraphics[width=\linewidth]{chapters/green_faas_duration}
-      \caption{Evolution of CPU behavior during multiple contexts }
-      \label{fig:green_faas_duration}
-\end{figure}
-
-On the other hand, as we can see in Figure~\ref{fig:green_faas_power}, the average power behavior of the system had different behavior.
-With the 100\% saturation scenario, the power consumption hit its peak as soon as we hit the 4 core usage 3 from the context and 1 from the benchmark), and it kept fluctuating until the $8^{th}$ core saturation, after that, it was in a constant decrease.
-This was because the processes were competing with each other for cache usage, which led to a significant number of cache misses that slowed down the CPU frequency and led to a decrease in power consumption.
-
-\begin{figure}[!h]
-      \centering
-      \includegraphics[width=\linewidth]{chapters/green_faas_power}
-      \caption{Evolution of CPU behavior during multiple contexts }
-      \label{fig:green_faas_power}
-\end{figure}
-
-However, when we used 50\% of the CPU capacity, the power kept increasing by adding the number of extra processes.
-As for the scenarios where we used more than 50\% of the CPU capacity, the power consumption kept increasing until it hit its peak then it start to decrease; this number of processes needed to reach the peak varies according to the saturation level.
-
-When we try to measure the energy consumption of this experiment, we find that the scenario where we used 100\% of the energy consumption was the worst case by far, and it kept increasing whenever we increase the number of core usage.
-Figure~\ref{fig:green_faas_energy} present the energy consumption of the system during the experiment.
-
-\begin{figure}[!h]
-      \centering
-      \caption{Evolution of CPU behavior during multiple contexts }
-      \label{fig:green_faas_energy}
-      \includegraphics[width=\linewidth]{chapters/green_faas_energy}
-\end{figure}
-
-However, if we calculate the energy impact of the benchmark which we define as: 
-\begin{equation}
-      Impact_{Benchmark} =E_{Benchmark} - E_{Before}
-\end{equation}
-and can be approximated by:
-\begin{equation}
-      Impact_{Benchmark} =E_{Benchmark} - P_{Before} * T_{Benchmark}
-\end{equation}
-where $Impact_{Benchmark}$ is the energy impact of the benchmark, $E_{Benchmark}$ is the energy of the system when the benchmark is running, $E_{Before}$ is the energy of the system before the benchmark is running, $P_{Before}$ is the power of the system before the benchmark is running, and $T_{Benchmark}$ is the time of the benchmark execution.
-
-One can notice that the energy impact of our benchmark did not have the same behavior as the system's energy consumption.
-Sometimes, introducing the benchmark had a \emph{negative} impact on the system's energy consumption, which made it even greener.
-This is because it reduces power consumption.
-This is shown in Figure~\ref{fig:green_faas_impact}.
-
-\begin{figure}[!t]
-      \centering
-      \caption{Evolution of CPU behavior during multiple contexts }
-      \includegraphics[width=\linewidth]{chapters/green_faas_impact}\
-      \label{fig:green_faas_impact}
-\end{figure}
-
-We are aware that during this analysis, we have not considered the execution time of the context, and it will be affected by the benchmark---mainly, it will increase when we use more cores and/or higher CPU saturation.
-However, this action was motivated by the fact that most of the cloud software are web services, as we have seen in Chapter~\ref{chapter:porgramming_langauges} where the execution time is not bounded, and each service can run for an unlimited amount of time.
 
 \paragraph*{Synthesis}
-This experiment shows that the context can have a significant impact on the energy consumption of the system, and sometimes measuring only the energy consumption of an application does not represent the real impact of this application on the energy consumption of the whole system, especially when we are dealing with timeless Softwares aka services.
-In the future, we tend to extend this experiment with more data, such as the energy consumption of the application itself instead of the whole system using \cite{fieni2020smartwatts} and comparing the strategies that were mentioned in this thesis using multiple contexts instead of just a minimal version of the operating system.
+I believe that the context can have a significant impact on the energy consumption of the system, and sometimes measuring only the energy consumption of an application does not represent its real impact on the energy consumption of the whole system, especially when we are dealing with timeless Software aka services.
+In the future, we tend to explore this behavior from different angles, such as the energy consumption of the application itself instead of the whole system using \cite{fieni2020smartwatts}, comparing the strategies that were mentioned in this thesis using multiple contexts instead of just a minimal version of the operating system, and studying the impact of deliberately putting the application on hold when we have a lower CPU frequency to reduce its impact using some opportunistic schedulers.
 

### 2022-10-19 chapters/green_faas.tex
remove : green faas part

According to the work of \citeauthor{barroso2007case}~\cite{barroso2007case}, servers operate most of the time at between 10 and 50 percent of their maximum utilization levels.
Moreover, as shown in Figure~\ref{fig:soa_energy_efficiency} with 50\% of the servers' capacity, the server operates only with 75\% of its maximum power.
In other words, it will cost us only 25\% of the energy to operate the extra 50\% of the servers' capacity.

\begin{figure}[!h]
    \centering
    \includegraphics[width=.7\linewidth,keepaspectratio]{chapters/soa_energy_efficiency}
    \caption{Power efficiency Vs. power consumption~\cite{barroso2007case}}
    \label{fig:soa_energy_efficiency}
\end{figure}

To confirm if this hypothesis still holds, we run the following experiment.
First, we provision a machine with minimal services, then created a process that stresses one core by doing some mathematical operations, and each second we introduced a copy of this process.
After a while, we removed these copies one by one.
The source code of the experiment can be found in the following GitHub repository.\link{https://github.com/chakib-belgaid/lazy-green-code-test}
Figure~\ref{fig:power_evolution_greenfaas} reports on the machine power consumption during this experiment.
As one can notice, the introduction of the first copies had a greater cost than the other ones, despite not sharing the same core.
And the more we approach the peak of power, the less significance the introduction of a new process has on power consumption to be almost zero after the $9^{th}$ copy.

\begin{figure}[!h]
    \centering
    \includegraphics[width=\linewidth]{chapters/power_evolution_greenfaas}
    \caption{Evolution of CPU power}
    \label{fig:power_evolution_greenfaas}
\end{figure}

One should note that this experiment was conducted on a machine with the following configuration: Intel(R) Core(TM) i7-3740QM CPU @ 2.70GHz, with 4 physical cores and 4 hyper threads, 16GB RAM, and 687GB HDD.
As for the operating system, we used Ubuntu 20.04.2 LTS.

Therefore, with the introduction of the $8^{th}$ copy, we have already saturated the machine capacity, and the introduction of the $9^{th}$ copy had no impact on the power consumption.

To push this experiment further, we created a stress tool to fix the number of cores and \% usage of each core to simulate a production context.
We then run a benchmark (graph coloration algorithm)\link{https://github.com/chakib-belgaid/lazy-green-code-test/tree/master/exp01} using multiple configurations.
Figure~\ref{fig:green_faas_duration} displays the overall performance of the benchmark.
As one can see, the execution time remained the same when the context was using 3 cores at any saturation level, because we have 4 physical CPU cores, therefore each process was running a different core (3 cores were used by the context and the other one was used by the benchmark).
However, behavior changes when the context consumes more than 3 cores.
CPU saturation tends to have a greater impact on the performance of the benchmark.
When the benchmark was using a hyper-thread of the same core as one of the context workers, it took almost double the time to finish the same job.
As when we saturated only 50\% of the CPU capacity, this increase did not happen until the context used 7 cores and, in the worst-case scenario (10 extra processes), it took only 75 seconds to finish the job.

\begin{figure}[!h]
    \centering
    \includegraphics[width=\linewidth]{chapters/green_faas_duration}
    \caption{Evolution of CPU behavior during multiple contexts }
    \label{fig:green_faas_duration}
\end{figure}

On the other hand, as we can see in Figure~\ref{fig:green_faas_power}, the average power behavior of the system had different behavior.
With the 100\% saturation scenario, the power consumption hit its peak as soon as we hit the 4 core usage 3 from the context and 1 from the benchmark), and it kept fluctuating until the $8^{th}$ core saturation, after that, it was in a constant decrease.
This was because the processes were competing with each other for cache usage, which led to a significant number of cache misses that slowed down the CPU frequency and led to a decrease in power consumption.

\begin{figure}[!h]
    \centering
    \includegraphics[width=\linewidth]{chapters/green_faas_power}
    \caption{Evolution of CPU behavior during multiple contexts }
    \label{fig:green_faas_power}
\end{figure}

However, when we used 50\% of the CPU capacity, the power kept increasing by adding the number of extra processes.
As for the scenarios where we used more than 50\% of the CPU capacity, the power consumption kept increasing until it hit its peak then it start to decrease; this number of processes needed to reach the peak varies according to the saturation level.

When we try to measure the energy consumption of this experiment, we find that the scenario where we used 100\% of the energy consumption was the worst case by far, and it kept increasing whenever we increase the number of core usage.
Figure~\ref{fig:green_faas_energy} present the energy consumption of the system during the experiment.

\begin{figure}[!h]
    \centering
    \caption{Evolution of CPU behavior during multiple contexts }
    \label{fig:green_faas_energy}
    \includegraphics[width=\linewidth]{chapters/green_faas_energy}
\end{figure}

However, if we calculate the energy impact of the benchmark which we define as:
\begin{equation}
    Impact_{Benchmark} =E_{Benchmark} - E_{Before}
\end{equation}
and can be approximated by:
\begin{equation}
    Impact_{Benchmark} =E_{Benchmark} - P_{Before} * T_{Benchmark}
\end{equation}
where $Impact_{Benchmark}$ is the energy impact of the benchmark, $E_{Benchmark}$ is the energy of the system when the benchmark is running, $E_{Before}$ is the energy of the system before the benchmark is running, $P_{Before}$ is the power of the system before the benchmark is running, and $T_{Benchmark}$ is the time of the benchmark execution.

One can notice that the energy impact of our benchmark did not have the same behavior as the system's energy consumption.
Sometimes, introducing the benchmark had a \emph{negative} impact on the system's energy consumption, which made it even greener.
This is because it reduces power consumption.
This is shown in Figure~\ref{fig:green_faas_impact}.

\begin{figure}[!t]
    \centering
    \caption{Evolution of CPU behavior during multiple contexts }
    \includegraphics[width=\linewidth]{chapters/green_faas_impact}\
    \label{fig:green_faas_impact}
\end{figure}

We are aware that during this analysis, we have not considered the execution time of the context, and it will be affected by the benchmark---mainly, it will increase when we use more cores and/or higher CPU saturation.
However, this action was motivated by the fact that most of the cloud software is web services, as we have seen in Chapter~\ref{chapter:porgramming_langauges} where the execution time is not bounded, and each service can run for an unlimited amount of time.

\paragraph*{Synthesis}
This experiment shows that the context can have a significant impact on the energy consumption of the system, and sometimes measuring only the energy consumption of an application does not represent the real impact of this application on the energy consumption of the whole system, especially when we are dealing with timeless Softwares aka services.
In the future, we tend to extend this experiment with more data, such as the energy consumption of the application itself instead of the whole system using \cite{fieni2020smartwatts} and comparing the strategies that were mentioned in this thesis using multiple contexts instead of just a minimal version of the operating system.

### 2022-11-04 chapters/introduction.tex
update : typos

@@
 Moreover, thanks to computers, we pushed that step even further, to the point where now we are using machines to extend our brains, from equation solvers to tools to recommend to us where we should invest our money, what we should eat, and even who fits best as our partner.
-One major aspect of computers that became omnipresent in our lives is the Internet, which is a network of networks that connects millions of computers all over the world.
+One significant aspect pushing computers to become omnipresent in our lives is the Internet, a network connecting millions of computers worldwide.
 According to Internet World Stats,\footnote{\url{https://www.internetworldstats.com/stats.htm}} the number of people connected to the Internet has increased by 4.4 billion in 2019, reaching 4.54 billion worldwide, or 59.2\% of the world population.
 
-The Internet has evolved from a place where government researchers share information in the 60s to a means of Communication at the beginning of the century, and now it is a place where we can find almost anything we want, from information to entertainment, from social media to e-commerce.
+The Internet has evolved from a place where government researchers share information in the 60s to Communication support at the beginning of the century.
+Now it is where we can find almost anything we want, from information to entertainment, social media to e-commerce.
 
@@
 
-Moreover, due to the pandemic caused by COVID-19 disease, the world has been forced to adapt to a new way of living, which has been accelerated by the Internet.
-The Internet has become a necessity for people to work, study~\cite{naresh2020education}, and even health consultations~\cite{liaw2021primary}.
+Moreover, due to the pandemic caused by COVID-19 disease, the world has been forced to adopt a new living style, in which the Internet has become a necessity for people to work, study~\cite{naresh2020education}, and even health consultations~\cite{liaw2021primary}.
 
-On the other hand, as humanity, we face a major challenge, which is climate change.
+On the other hand, as humanity, we face a significant challenge, which is climate change.
 The \emph{Intergovernmental Panel on Climate Change} (IPCC) has warned that the world has only 12 years to limit the global temperature rise to 1.5\degree C and that the world has to reach net-zero emissions by 2050 to avoid the worst effects of climate change.
@@
 The first one includes finding a new planet that we can populate and live on,\footnote{\url{https://en.wikipedia.org/wiki/Interstellar_(film)}} which is known as the Planetary Migration~\cite{mapstone2022cyanobacteria}.
-Meanwhile, the second solution is to provide new sources of energy, such as nuclear energy, wind energy, and even fusion energy~\cite{gross1984fusion}.
+Meanwhile, the second solution is to provide new energy sources, such as nuclear energy, wind energy, and even fusion energy~\cite{gross1984fusion}.
 The third solution is to reduce our emissions, which is the main focus of this thesis.
@@
 
-With this increase in the number of data centers comes an increase in energy consumption, which is a major problem for the environment.
-in 2018, data centers consumed around 205 terawatt-hours (TWh)~\cite{schneider2021world}, which is equivalent to the energy consumption of 1\% of the total world's electricity.
-This ratio increased up to 1.5 \% in 2020 according to the Journal of Science~\cite{mytton2021data}.
+With this increase in the number of data centers comes an increase in energy consumption, which is a significant environmental problem.
+In 2018, data centers consumed around 205 terawatt-hours (TWh)~\cite{schneider2021world}, equivalent to the energy consumption of 1\% of the entire world's electricity.
+This ratio increased up to 1.5\% in 2020 according to the Journal of Science~\cite{mytton2021data}.
 In Figure~\ref{fig:data_centers_power_distribution}, one can see that 40\% of the energy consumed by data centers is used for cooling, while another 40\% is used by the servers themselves.
-Therefore, optimizing these two aspects can have a major impact on the energy consumption of data centers.
+Therefore, optimizing these two aspects can significantly impact the energy consumption of data centers.
 
@@
 Researchers are trying to reduce the energy consumption of data centers through different angles.
-Some of the works are focused on the hardware side, such as using new hardware architectures that are more energy-friendly, such as the use of GPUs instead o ARM processors instead of CPUs~\cite{aroca2012towards}.
-Others are trying to optimize the cooling system, this can be achieved by using more efficient cooling systems, putting data centers in cold locations or under water~\cite{simon2018project}, or even using the waste heat for other purposes, such as heating buildings~\cite{bouzel2021distributed,cao2021carbon}.
+Some of the works are focused on the hardware side, such as using new hardware architectures that are more energy-friendly, such as using GPUs instead o ARM processors instead of CPUs~\cite{aroca2012towards}.
+Others are trying to optimize the cooling system; this can be achieved by using more efficient cooling systems, putting data centers in cold locations or under water~\cite{simon2018project}, or even using the waste heat for other purposes, such as heating buildings~\cite{bouzel2021distributed,cao2021carbon}.
 
-A third approach is to optimize the software, by making software more energy-efficient.
-In this thesis, we focus on this approach, and we try to optimize the software by reducing the number of computations that are done by the software.
+A third approach is to optimize the software by making software more energy-efficient.
+In this thesis, we focus on this approach and try to optimize the software by reducing the number of computations done by the software.
 
@@
 Unfortunately, this is not possible in the current state of the art.
-Due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.
+Due to the lack of knowledge about the energy consumption of the algorithms and the strong correlation between this consumption and the hardware configuration.
 Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on the execution environment.
@@
 The goal of this method is to reduce energy usage by sacrificing execution time.
-This strategy is useful while dealing with services since the task has an unlimited execution time.
-The second part is to use some. opportunistic schedulers, which will stop the software if it is estimated to spend more energy.
-This strategy is discussed first in Chapter~\ref{chapter:porgramming_langauges}, then we continue with this strategy in the perspectives section.
+This strategy is helpful while dealing with services since the task has an unlimited execution time.
+The second part is to use some opportunistic schedulers, which will stop the software if it is estimated to spend more energy.
+This strategy is discussed first in Chapter~\ref{chapter:porgramming_langauges}, then we continue this strategy in the perspectives section.
 
@@
       \item Chapter~\ref{chapter:benchmarking} discusses two facets of empirical tests and how we adapt them to energy consumption measurements.
-          First, we discuss the reproducibility challenge and how to overcome it with the usage of containers, then we push this approach further by proposing and new protocol to make tests not only reproducible but able to be extended with other features to overcome the rapid pace of software evolution.
-          The second part of this chapter discusses the aspect of accuracy when it comes to energy consumption measurements.
+            First, we discuss the reproducibility challenge and how to overcome it with the usage of containers; then we push this approach further by proposing and new protocol to make tests not only reproducible but able to be extended with other features to overcome the rapid pace of software evolution.
+            The second part of this chapter discusses the aspect of accuracy regarding energy consumption measurements.
             This section demonstrates how energy measurements might differ and produce conflicting results for the same work when run on identical machines or even the same machine.
@@
             Then, it provides some statistics about its popularity and typical use cases and studies the impact of Python on energy consumption in three typical use cases, machine learning, web servers, and data manipulation.
-          For each use case, we compare the energy consumption of several approaches and provide guidelines on how to optimize the energy consumption of python programs in these use cases.
-          Finally, we provide a non-intrusive way to optimize energy consumption without altering the code of the program.
+            For each use case, we compare the energy consumption of several approaches and provide guidelines on optimizing the energy consumption of python programs in these use cases.
+            Finally, we provide a non-intrusive way to optimize energy consumption without altering the program's code.
             We achieve this by using a different implementation of the Python interpreter.
@@
             In Chapter~\ref{chapter:java}, we try to optimize the Java code, using by changing the default JVM implementation.
-          We compare the energy consumption of $12$ benchmarks using $52$ implementation, each benchmark is dedicated to a typical use case.
+            We compare the energy consumption of $12$ benchmarks using $52$ implementation, and each benchmark is dedicated to a typical use case.
             After that, we study two of the JVM features, JIT and GC, and show their impact on the energy consumption of the Java code;
@@
 Then, we compare $261$ web frameworks, each implementing the same website using seven use cases.
-          This analysis aims to look at how each technology uses energy in different kinds of web situations.
-    \item Finally, we conclude our work in Chapter~\ref{chapter:conclusion} by summarizing the work done in this thesis and discussing the future work that can be done to further reduce the energy consumption of software.
+            This analysis aims to look at how each technology uses energy in different web situations.
+      \item Finally, we conclude our work in Chapter~\ref{chapter:conclusion} by summarizing the work done in this thesis and discussing the future work that can be done to reduce the energy consumption of software further.
 \end{itemize}

### 2022-11-21 chapters/optimization/python-dir/python.tex
add : citing the interns work

@@
 As machine learning can significantly impact climate change, it is vital to investigate mitigation techniques.
+This part highlights the work of \citeauthor{kaminetzky_energy_2021}~\cite{kaminetzky_energy_2021} during his internship at inria
 
@@
     \includegraphics[width=1.1\textwidth]{imgs/accuracy_basedonepoch}
-    \caption{Accuracy along epochs.}
+    \caption{Accuracy along epochs~\cite{kaminetzky_energy_2021}.}
     \label{fig:eopochvsaccuracy}
@@
     \includegraphics[width=1.1\textwidth]{imgs/cumulative_energy_vs_accuracy}
-    \caption{Cumulative energy consumption vs. accuracy.}
+    \caption{Cumulative energy consumption vs. accuracy~\cite{kaminetzky_energy_2021}.}
     \label{fig:cum_energy_vs_accuracy}
@@
     \includegraphics[width=1.1\textwidth]{imgs/cumulative_energy_fast10.pdf}
-    \caption{Average energy for training the model based on the strategy.}
+    \caption{Average energy for training the model based on the strategy~\cite{kaminetzky_energy_2021}.}
     \label{fig:av_energy_setup}
@@
     \includegraphics[width=1.1\textwidth]{imgs/power_gpu_baedonepoche.pdf}
-    \caption{Evolution of average GPU power along epochs.}
+    \caption{Evolution of average GPU power along epochs~\cite{kaminetzky_energy_2021}.}
     \label{fig:ephoch_power}
@@
     \includegraphics[width=1.1\textwidth]{imgs/epoch_duration.pdf}
-    \caption{Distribution of the durations of each epoch.}
+    \caption{Distribution of the durations of each epoch~\cite{kaminetzky_energy_2021}.}
     \label{fig:epoch_duration}
@@
 % \subsection{Experiment \& Results}
-To accomplish so, we examined the cost of a single request of the prior website using various ways to extract data from the database.
+To accomplish so, we examined the cost of a single request from the prior website using various ways to extract data from the database.
 We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records and three different ways to fetch the records:
@@
 In this study, we will run some common algorithms with different data structures and control structures to study if there is any difference in energy consumption.
+This part displays the work of \citeauthor{randrianaina_optimisation_2019}\cite{randrianaina_optimisation_2019} during his internship at Inria.
+
 \subsubsection{Experimental Protocol}
@@
     \includegraphics[width=\linewidth]{imgs/python_iterations}
-    \caption{Comparison of the energy consumption of different Python loops.}
+    \caption{Comparison of the energy consumption of different Python loops~\cite{randrianaina_optimisation_2019}.}
     \label{fig:pythonloops}
@@
     \includegraphics[width=.9\linewidth]{imgs/python_treatemens}
-    \caption{Comparison of the energy consumption of different methods to convert a list.}
+    \caption{Comparison of the energy consumption of different methods to convert a list~\cite{randrianaina_optimisation_2019}.}
     \label{fig:pythontreatement}

### 2022-12-06 chapters/literature/draft_citations.tex
update : university logo

@@
     \item \em{Scalability}: this is mainly related to the cost of the tools and the ease of their integration into our system,
-    \item  \em{Accuracy}: to eliminate extra hazards and get a more representative measurement.
+    \item  \em{Precision}: to eliminate extra hazards and get a more representative measurement.
 \end{itemize}
 
-We believe that accuracy can be extracted from these criteria by combining the first two results. Therefore, we will focus more on the first three criteria later on.
+We believe that precision can be extracted from these criteria by combining the first two results. Therefore, we will focus more on the first three criteria later on.
 Below, we will discuss some of the well-known tools used in literature.

### 2022-12-16 chapters/acknowledgement.tex
add : aknowlegement

% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}

    I dedicate this thesis to my mother, who has never ceased supporting me since my birth and will continue to do so in the future, and to my father, who has always believed in me and encouraged me to challenge myself constantly.
    I also want to convey my deepest gratitude to my advisors, Romain Rouvoy and Lionel Seinturier, for allowing me to complete my Ph.D. study. Thank you for all the advice and discussions that helped me complete this project. Thank you for your unwavering support in helping me overcome the obstacles; without your help, I would never have overcome these challenges. I genuinely appreciate everything you've done to make this journey joyful for me.
    I also thank Davidson Consulting for providing me with the opportunity to work on this project by providing financial and technical support for this thesis. Thanks to David Olivier for his constant interest in my work.
    Also, I'd like to thank Pierre Boulet, Chantal Taconet, Olivier Barais, and Thomas Degeul, who are on my thesis committee, for their time and insightful feedback, which will help me keep this research going.

    I would also like to thank the Spirals team and its many talents who welcomed, guided me, and inspired me with their professional and personal achievements; special thanks to Zakaria Ournani for his constant emotional and technical support during this whole journey.

    Lastly, I would like to warmly thank my friends who helped, guided, and encouraged me during this adventure. You believed in me even when I doubted myself, and for this, I will be eternally grateful to you. A special mention goes to my dear friend Oudjedi Damerdji Soufiane, who inspired me with his endeavors during his fight, May your memory remain with us. And my dear girlfriend, who has always been there to support me.

\end{acknowledgements}

### 2022-12-16 title_page_fr.tex
Update : title pages

\thispagestyle{empty}

% \begin{figure}[ht]
%     \minipage{0.76\textwidth}
%     \includegraphics[width=.3\linewidth, keepaspectratio]{Figs/logo-davidson.jpeg}
%     \label{EscudoUABC}
%     \endminipage
%     \minipage{0.32\textwidth}
%     \includegraphics[width=.3\linewidth, keepaspectratio]{Figs/logo-davidson.jpeg}
%     \label{EscudoFC}
%     \endminipage
% \end{figure}

\begin{figure}[!ht]
    \includegraphics[width=.24\linewidth, keepaspectratio]{Figs/logo-davidson.jpeg}\,
    \includegraphics[width=.24\linewidth]{Figs/logo-inria}\,
    \includegraphics[width=.24\linewidth,trim={0 3em 0 0 }]{Figs/logo-univ-lille}\,
    \includegraphics[width=.24\linewidth]{Figs/logo-cristal}\,

\end{figure}
\begin{center}
    \LARGE
    \textbf{Éco-développement} : une approche empirique pour réduire la consommation énergétique des logiciels

    \vspace{.8cm}
    \Large
    \textbf{Mohammed Chakib \textsc{BELGAID}}
    \vspace{0.8cm}
    \large

    Une dissertation pour L'obtension d'un diplome de \emph{Doctorat} en \textsc{Informatique}  \\

    \vspace{0.5cm}
    \large
    \textbf{Directeur}\\
    \minipage{0.4\textwidth}
    \large
    \vspace{.5cm}
    Pr. Romain \textsc{ROUVOY}
    \endminipage
    \vspace{0.5cm}
    \large
    \\
    \textbf{Co-Directeur}\\
    \minipage{0.4\textwidth}
    \large
    \vspace{.5cm}
    Pr. Lionel \textsc{SEINTURIER}
    \endminipage
    \vspace{0.5cm}
    %  reporters  
    \\
    \large
    \textbf{Examinateurs}\\
    \minipage{0.4\textwidth}
    \large
    \vspace{.5cm}
    Pr. Pierre \textsc{BOULET} \newline
    Dr. Thomas \textsc{DEGUEULE}
    \endminipage
    \vspace{.5cm}
    \\
    \large
    \textbf{Rapporteurs}\\
    \minipage{0.4\textwidth}
    \large
    \vspace{.5cm}
    Pr. Olivier \textsc{BARAIS} \newline
    Pr. Chantal \textsc{TACONET}
    \endminipage
    \vspace{0.5cm}
    \\
    \large
    \textbf{Invité}\\
    \minipage{0.4\textwidth}
    \large
    \vspace{.5cm}
    Mr. David \textsc{OLIVIER}\\
    \endminipage
    \vspace{0.5cm}
    \\
    \Large
    \emph{Université de Lille}
    \minipage{0.75\textwidth}
    \'Ecole Doctorale MADIS
    \endminipage
    \minipage{0.25\textwidth}
    \emph{December} 2022
    \endminipage

\end{center}

### 2022-12-20 title_page_fr.tex
update to final version
fix bold text in the titile

@@
     \LARGE
-    \textbf{Éco-développement} : une approche empirique pour réduire la consommation énergétique des logiciels
+    Eco-développement : une approche empirique pour réduire la consommation~énergétique des logiciels
 
@@
 
-    \vspace{0.5cm}
     \large
-    \textbf{Directeur}\\
-    \minipage{0.4\textwidth}
-    \large
+    \textbf{Directeurs}\\
+    \minipage{0.55\textwidth}
+    \RaggedRight
     \vspace{.5cm}
-    Pr. Romain \textsc{ROUVOY}
+    \normalsize
+    Pr. \emph{Romain} \textsc{ROUVOY}, Université de Lille,\newline
+    Pr. \emph{Lionel} \textsc{SEINTURIER}, Université de Lille
     \endminipage
-    \vspace{0.5cm}
-    \large
-    \\
-    \textbf{Co-Directeur}\\
-    \minipage{0.4\textwidth}
-    \large
-    \vspace{.5cm}
-    Pr. Lionel \textsc{SEINTURIER}
-    \endminipage
-    \vspace{0.5cm}
+    \vspace{0.6cm}
+    % \large
+    % \\
+    % \textbf{Co-Director}\\
+    % \minipage{0.4\textwidth}
+    % \large
+    % \vspace{.5cm}
+    % Pr. Lionel \textsc{SEINTURIER}
+    % \endminipage
+    % \vspace{0.5cm}
     %  reporters
-    \\
+
     \large
     \textbf{Examinateurs}\\
-    \minipage{0.4\textwidth}
-    \large
+    \minipage{0.55\textwidth}
+    \normalsize
+
     \vspace{.5cm}
-    Pr. Pierre \textsc{BOULET} \newline
-    Dr. Thomas \textsc{DEGUEULE}
+
+    Pr. \emph{Pierre} \textsc{BOULET}, Université de Lille,\newline
+    Dr. \emph{Thomas} \textsc{DEGUEULE}, Université de Bordeaux
     \endminipage
@@
     \textbf{Rapporteurs}\\
-    \minipage{0.4\textwidth}
-    \large
+    \minipage{0.55\textwidth}
+
     \vspace{.5cm}
-    Pr. Olivier \textsc{BARAIS} \newline
-    Pr. Chantal \textsc{TACONET}
+    \normalsize
+    Pr. \emph{Olivier} \textsc{BARAIS}, Université de Rennes~1,\newline
+    Pr. \emph{Chantal} \textsc{TACONET}, Télécom SudParis
+
     \endminipage
-    \vspace{0.5cm}
+    \vspace{0.6cm}
     \\
@@
     \textbf{Invité}\\
-    \minipage{0.4\textwidth}
+    \minipage{0.55\textwidth}
+    \normalsize
+    \vspace{.5cm}
+    M. \emph{David} OLIVIER, Davidson consulting
+    \endminipage
+    \vspace{0.8cm}
+    \\
     \large
-    \vspace{.5cm}
-    Mr. David \textsc{OLIVIER}\\
-    \endminipage
-    \vspace{0.5cm}
-    \\
-    \Large
     \emph{Université de Lille}
